### 2021-06-24 ch1intro/ch1intro.tex
gearing up for intense work.

%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Sediment transport and landscape evolution}
\label{ch:Introduction}

Landscapes evolve across geologial time due to energy and mass contributed from above and below the Earth's surface.

\section{Stochastic properties of sediment movement}

\section{Historical survey: statistical physics in sediment transport}

\section{Philosophy of the statistical physics approach}

\section{What's left?}

\endinput

Any text after an \endinput is ignored.
You could put scraps here or things in progress.

### 2021-06-24 ch2flipFlop/jgr/rev-one.tex
gearing up for intense work.

\documentclass[draft]{agujournal2018}
\journalname{JGR: Earth Surface}

\usepackage{apacite}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{comment}
\usepackage{wrapfig}
\usepackage{lipsum}
\usepackage{booktabs} % for wrapping text around tabulars in 
\usepackage{ragged2e}
\usepackage{lineno}
\usepackage[utf8]{inputenc}

% reduce white space after tables
%\usepackage[belowskip=-15pt,aboveskip=0pt]{caption}
\setlength{\intextsep}{10pt plus 2pt minus 2pt}
\setlength{\textfloatsep}{30pt}

\raggedbottom

\justifying
\linenumbers

% there is a bug with align environments in conjunction with lineno
% the patch is 
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
	\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
	\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
	\renewenvironment{#1}%
	{\linenomath\csname old#1\endcsname}%
	{\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
	\patchAmsMathEnvironmentForLineno{#1}%
	\patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
	\patchBothAmsMathEnvironmentsForLineno{equation}%
	\patchBothAmsMathEnvironmentsForLineno{align}%
	\patchBothAmsMathEnvironmentsForLineno{flalign}%
	\patchBothAmsMathEnvironmentsForLineno{alignat}%
	\patchBothAmsMathEnvironmentsForLineno{gather}%
	\patchBothAmsMathEnvironmentsForLineno{multline}%
}
% from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation?noredirect=1&lq=1

\begin{document}

\title{Joint stochastic bedload transport and bed elevation model: variance regulation and power-law rests}
\authors{J. Kevin Pierce\affil{1}\thanks{Vancouver, British Columbia, Canada}, and Marwan A. Hassan\affil{1}}
\affiliation{1}{Department of Geography, University of British Columbia}
\correspondingauthor{James K. Pierce}{kpierce@alumni.ubc.ca}

\begin{keypoints}
\item Two species stochastic population model describes fluvial bedload particle activity and local bed elevations
\item Model predicts elevation dependent transport statistics and power-law distributed resting times for sediment undergoing burial
\item Results imply elevation changes shift particle activity fluctuations by up to 90\% and sediment rests can generate anomalous bedload diffusion
\end{keypoints}

\begin{abstract}
We describe the joint dynamics of bedload transport and bed elevation changes with a stochastic population model, and we analyze (1) the dependence of bedload flux statistics on local bed elevations and (2) resting time distributions for sediment undergoing burial in the fluctuating sedimentary bed.
The model exhibits suppression of the statistical moments of the bedload flux the bed is in a locally aggraded state and enhancement of the moments when the bed is in a degraded state.
This variance regulation effect is contingent on collective entrainment, whereby moving grains destabilize stationary grains in a positive feedback. 
When collective entrainment is turned off, the variance regulation disappears.
Return times from above in the bed elevation time-series provide heavy tailed power-law distributions of resting times with tail behavior characterized by the mean erosion rate and the active layer depth.
These results imply bedload statistics measurements on relatively short timescales can be strongly biased by bed elevation changes, and they support the growing consensus that sediment burial generates heavy tailed sediment resting times that ultimately generate anomalous bedload diffusion.
\end{abstract}

\section{Introduction}
The transport characteristics of coarse grains moving under a turbulent flow ultimately control a wide set processes within rivers, including the export of contaminants \citep{Malmon2005,Macklin2006}, the success of ecological restoration efforts \citep{Gaeuman2017}, and the response of channel morphology to disturbances \citep{Hassan2017}.
Although the displacements of individual grains are certainly a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}. In response, investigators have developed a stochastic concept of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.

Essentially two types of bedload transport model have been developed from this concept. 
The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2018,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Ancey2008,Furbish2012a}.
In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}. 
Differences between the random-walk motions of one grain and the next imply a spreading apart of tracer grains as they transport downstream: bedload tracers undergo diffusion.

Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
Resting times have puzzled researchers because early experiments show exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments show heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016, Bradley2017, Liu2019}.
A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}; this hypothesis permits surface grains to retain exponential resting times. 
Conceptually, when grains rest on the surface, material transported from upstream can deposit on top of them, preventing entrainment until its removal, driving up resting times and imparting a heavy tail to the distribution.
To our knowledge, \citet{Martin2014} have provided the only direct support for this hypothesis by tracking grains through complete cycles of burial and exhumation using a narrow flume with glass walls.
They observed heavy-tailed resting times of buried grains and described their results with a mathematical model similar to an earlier effort by \citet{Voepel2013}.
Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation time-series \citep{Redner2007}. Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, and neither model explicitly includes bedload transport.
Models of sedimentary bed evolution incorporating sediment transport processes might enhance understanding of sediment resting times.

The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015a,Ma2014}.
A recent surge in research activity has generated rapid progress and spawned many new inquiries in this subject. 
For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014,Ma2014, Lee2018}.
While they deemed this feedback necessary to model realistic activity fluctuations, the implications of this collective entrainment term on bed topography and particle activity changes has not been fully explored.

In this work, we present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
Our model extends the \citet{Ancey2008} model to describe the interplay between bedload flux and bed elevation fluctuations in a control volume.
This development permits a systematic study of the repercussions of collective entrainment, and it frames bed elevation changes as a direct consequence of the sediment transport process.
Our model has two key assumptions: (1) bedload erosion and deposition can be characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and (2) these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
We study statistical characteristics of bedload transport, bed elevation, and resting times of sediment undergoing burial using a mixture of numerical simulations and analytical approximations. We introduce the stochastic model in section \ref{sec:model}, and we solve it in section \ref{sec:solution} with a mixture of numerical and analytical techniques. We discern several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates and the local bed elevation. We present these features in section \ref{sec:results}. We conclude with the implications of our results and speculate on topics for future research in sections \ref{sec:discussion} and \ref{sec:conclusion}.

\section{Stochastic model of bedload transport and bed elevations}
\label{sec:model}

We prescribe a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:definition}. We define $m$ relative to the mean number of grains within the control volume, so that it can be either positive or negative. $n$ is always a positive integer including $0$.
For simplicity, we consider all particles as approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
Following \citet{Ancey2008}, we prescribe four events that can occur at any instant to modify the populations $n$ and $m$, and we characterize these events using probabilities per unit time (rates).
These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
The four events are depicted as arrows in figure \ref{fig:definition}.
As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
\begin{figure}[!ht]
	\includegraphics[width=\linewidth,keepaspectratio]{definition.pdf}
	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains. Migration, entrainment, and deposition are represented by arrows, and the instantaneous bed elevation is depicted by dotted lines. The bed is displayed in a degraded state, where $m<0$. The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using (\ref{eq:ele}).}
	\label{fig:definition}
\end{figure}
The populations $n$ and $m$ provide the bulk bedload flux $q_s$ and the local bed elevation $z$.
The mean bedload transport rate is given by $q_s = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
\sum_{n,m}nP(n,m) $ is the mean number of grains in motion \citep{Charru2004, Ancey2008, Furbish2012a}.
The bed elevation is related to $m$ through the packing geometry of the bed.
To quantify this, we introduce a packing fraction $\phi$ of grains in the bed \citep{Bennett1972}, and for simplicity we consider the bed as two-dimensional \citep{Einstein1950, Paintal1971}. The deviation from the mean bed elevation is then
\begin{equation} z(m) = \frac{\pi a^2}{\phi L}m = z_1 m. \label{eq:ele}\end{equation}
The constant $z_1 = \pi a^2/(\phi L)$ is an important scale of the problem. 
$z_1$ is the magnitude of bed elevation change in an average sense across the control volume associated with the addition or removal of a single grain.

Bed elevation changes modify the likelihood of entrainment and deposition in a negative feedback \citep{Sawai1987, Wong2007}; that is, aggradation increases the likelihood of entrainment, while degradation increases the likelihood of deposition.
\citet{Wong2007} concluded that bed elevation changes induce an exponential variation in entrainment and deposition probabilities, while \citet{Sawai1987} concluded that the variation is linear.
For simplicity, we incorporate the scaling of \citet{Sawai1987} and note its equivalence to the \citet{Wong2007} scaling when bed elevation changes are small.
Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, we expect the erosion and deposition feedbacks to have the same strength.
That is, as bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.
Merging these ideas with those of \citet{Ancey2008}, we write the four possible transitions with local bed elevation-dependent entrainment and deposition rates as
\begin{align}
&R_{MI}(n+1|n) = \nu && &\text{migration in}, \label{eq:rate1}\\
&R_E(n+1,m-1|n,m)=(\lambda + \mu n)[1 + \kappa m], && &\text{entrainment}, \label{eq:rate2}\\
&R_D(n-1,m+1|n,m)=\sigma n [1- \kappa m ], && &\text{deposition}. \label{eq:rate3} \\
&R_{MO}(n-1|n) =\gamma n && &\text{migration out} \label{eq:rate4}.
\end{align}
In equations (\ref{eq:rate2}) and (\ref{eq:rate3}), $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
$\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
Away from this elevation, entrainment and deposition are alternatively suppressed and enhanced depending on the sign of $m$, constituting a feedback between bed elevation changes and erosion and deposition.
We refer to $\kappa$ as a coupling constant since it controls the strength of this feedback. We later demonstrate the relationship
\begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
\end{equation}
 where $l$ is a characteristic length scale of bed elevation change that we interpret as the active layer depth \citep{Wong2007,Church2017}.
All four rates are independent of the past history of the populations and depend only on the current populations $(n,m)$. 
As a result, the model is Markovian \citep{Cox1965, VanKampen1992}, meaning time intervals between any two subsequent transitions are exponentially distributed \citep{Gillespie2007}.

We write the master equation for the probability flow using the forward Kolmogorov equation $\partial P(n,m;t)/\partial t = 
\sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1992, Ancey2008} as 
\begin{multline}
 \frac{\partial P}{\partial t}(n,m;t) =  
\nu P(n-1,m;t) + [\lambda(m+1) + \mu(n-1)][1+\kappa(m+1)]P(n-1,m+1;t)\\  
+ \sigma(n+1)[1-\kappa(m-1)]P(n+1,m-1;t) + \gamma(n+1) P(n+1,m;t) \\
- 
\{ \nu + \lambda+ \mu n (1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}P(n,m;t).
 \label{eq:master}
\end{multline}
The joint probability distribution $P(n,m;t)$ solving this equation fully characterizes the statistics of $n$ and $m$ -- proxies for the bedload flux and local bed elevation.
The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
We anticipate that solutions of (\ref{eq:master}) will adjust from the initial conditions to a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium conditions. Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
This Master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
In our context, the two populations are the moving and stationary grains in the volume.

\section{Model solutions}
\label{sec:solution}

Unfortunately, equation (\ref{eq:master}) does not appear to admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
In response to this difficulty, we resort to numerical methods and analytical approximations, simulating equation (\ref{eq:master}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1992, Gillespie2007} and solving it approximately with mean field and Fokker-Planck approaches \citep{Haken1983,Gardiner1983}. The simulation algorithm is described in \ref{sec:numerical}, and analytical approximations are described in \ref{sec:analytical}.

\subsection{Numerical simulations}
\label{sec:numerical}

The Gillespie algorithm leverages the defining property of a Markov process: when transition rates are independent of history, time intervals between transitions are \begin{wraptable}{l}{0.5\textwidth}
	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ from \citet{Ancey2008}. Units are $s^{-1}$ (probability/time). In our model, bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
	\begin{tabular}{cccccc} \\ 
		\toprule  
		flow & $\nu$ & $\lambda$ & $\mu$ & $\sigma$ & $\gamma$ \\
		\midrule
		(a) & 5.45  & 6.59  & 3.74 & 4.67 & 0.77 \\
		\midrule
		(g) & 7.74  & 8.42  & 4.34 & 4.95 & 0.56 \\
		\midrule
		(i) & 15.56 & 22.07 & 3.56 & 4.52 & 0.68 \\
		\midrule
		(l) & 15.52 & 14.64 & 4.32 & 4.77 & 0.48 \\
		\bottomrule
	\end{tabular}
\end{wraptable}exponentially distributed \citep{Cox1965}.
As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}). The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
We provide additional background on the stochastic simulation method in the supplementary material and refer the reader to \citet{Gillespie2007} for more detail.

Using this method, we simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, we used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
In all simulations, we take the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and we set $L = 22.5$cm and $a = 0.3$cm in accord with the \citet{Ancey2008} experiments.
Each simulation was run for $250$ hours of virtual time, a period selected to ensure neat convergence of particle activity and bed elevation statistics.

\subsection{Approximate solutions}
\label{sec:analytical}

We approximately decouple the $n$ and $m$ dynamics in equation (\ref{eq:master}) using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$. These inequalites mean many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
We concentrate on steady state conditions $\partial P/\partial t = 0$ and introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to equation (\ref{eq:master}), with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$.
This provides the steady state equation
\begin{multline}
 0 =  
\nu A(n-1|m)M(m) + [\lambda + \mu(n-1)][1+\kappa(m+1)]A(n-1|m+1)M(m+1)\\  
+ \sigma(n+1)[1-\kappa(m-1)]A(n+1,m-1)M(m-1) + \gamma(n+1)A(n+1|m)M(m) \\
- 
\{ \nu + [\lambda+ \mu n ](1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}A(n,m)M(m).
\label{eq:decomp}
\end{multline}
Summing this equation over $n$ provides a still exact description of the distribution of bed elevations $M(m)$ in terms of the conditional mean particle activity $\langle n | m \rangle = \sum_{n}nA(n|m)$:
\begin{multline}
 0 =  [\lambda + \mu\langle n | m+1\rangle][1+\kappa(m+1)]M(m+1)
\\+ \sigma\langle n | m-1\rangle[1-\kappa(m-1)]M(m-1) \\
- 
\{  [\lambda+ \mu \langle n | m\rangle ](1+\kappa m) +  \sigma  \langle n | m\rangle( 1- \kappa m) \}M(m).
\label{eq:approxele}
\end{multline}
Unforunately, these two equations are no easier to solve than the original master equation, since the coupling between $n$ and $m$ is not reduced in equation (\ref{eq:decomp}).

The simplest approximation to these equations holds that $\kappa$ is so small that the dynamics of $n$ are totally independent of $m$: $A(n|m) = A(n)$. Taking this limit in equation (\ref{eq:decomp}), summing over $m$, and using $\langle m \rangle = 0$ reproduces the \citet{Ancey2008} particle activity model.
As shown by \citet{Ancey2008}, this has solution
\begin{equation} A(n) = \frac{\Gamma(r+n)}{\Gamma(r)n!}p^r(1-p)^n.\label{eq:ancey}\end{equation}
which is a negative binomial distribution for the particle activity with parameters $r=(\nu+\lambda)/\mu$ and $p=1-\mu/(\sigma+\gamma).$
This result implies $\langle n | m \rangle = \langle n \rangle$, so with the definitions of $E$ and $D$ and the equilibrium condition $E=D$, equation (\ref{eq:approxele}) provides
\begin{equation}0 \approx [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1)-2M(m). \label{eq:ou} \end{equation}
This mean field equation matches the discrete Ornstein-Uhlenbeck model of bed elevation changes developed by \citet{Martin2014}.
We summarize that the independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model we present in a mean field approximation when $\kappa$ is insignificant.

In the supplementary information we show the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within equation (\ref{eq:ou}) provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
As we will demonstrate in section \ref{sec:results}, and as we have already suggested with equation (\ref{eq:active}), this is a poor approximation to the bed elevation variance. Nevertheless, this approximation does capture the Gaussian shape of the bed elevation distribution.
The essential issue with this mean field approach is that the conditional mean particle activity $\langle n | m \rangle$ varies significantly with $m$ in actuality, especially when collective entrainment contributes to the mean entrainment rate $E$. We will discuss these points subsequently when developing more refined approximations and presenting numerical results.

A more careful approximate solution to equation (\ref{eq:approxele}) can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into equation (\ref{eq:approxele}) in order to close the equation for $m$ without solving equation (\ref{eq:decomp}).
From numerical simulations we determine that
\begin{equation}
\langle n | m  \rangle \approx \langle n \rangle \Big( 1 - \frac{2\kappa m}{1-\mu/\sigma}\Big) \label{eq:closure}
\end{equation}
 captures the general features of the conditional mean particle activity.
As we show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
\begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in equation (\ref{eq:active}).
$M_0$ is a normalization constant. 
As we will demonstrate, this closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:master}), at least for the flow parameters in table \ref{tab:anceyparams}.

\section{Results}
\label{sec:results}

From the initial conditions, all simulations show a rapid attainment of steady-state stochastic dynamics of $n$ and $m$ which support a time-independent joint distribution $P(n,m)$. We show an elevation time-series in the bottom panel of figure \ref{fig:definition}. In order to describe the implications of coupling bedload transport to bed elevation changes, we present the numerical and analytical results for the probability distributions of bedload transport and bed elevations in section \ref{sec:pdf} and the statistical moments of these quantities in section \ref{sec:mom}. We isolate the effects of collective entrainment on bed elevation changes in section \ref{sec:colent}, and we present the resting times of sediment undergoing burial in section \ref{sec:rtcdf}.

\subsection{Probability distributions of bedload transport and bed elevations}
\label{sec:pdf}

We compute this joint distribution by counting occurrences of the states $(n,m)$ in the simulated time series.
From this joint distribution we compute marginal distributions $P(n)$ and $P(m)$ by summing over $m$ and $n$ respectively.
A representative subset of these marginal distributions is displayed in figure \ref{fig:pdfs} alongside the approximate results of equations (\ref{eq:ancey}) and (\ref{eq:ou2}).
\begin{figure}[t]
	\includegraphics[width=\linewidth,keepaspectratio]{distributions.pdf}
	\caption{Panel (a) presents the probability distribution of particle activity $n$ and panel (b) presents the probability distribution of the relative number of particles $m$ for a representative subset of simulations. These distributions represent different flows from table \ref{tab:anceyparams}, distinguished by color, and different values of the active layer depth $l$ (equivalently the coupling constant $\kappa$), distinguished by the marker style. The mean field theories (mft) of equations (\ref{eq:ancey}) and (\ref{eq:ou2}) are displayed as solid black lines.}
	\label{fig:pdfs}
\end{figure}
The mean field equation (\ref{eq:ancey}) for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match equation (\ref{eq:ou2}), having Gaussian profiles consistent with our assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.

\subsection{Statistical moments}
\label{sec:mom}

We calculate the moments of $n$ and $m$ by summing over $P(n,m)$. 
The $j$th order unconditional moment of the particle activity $n$ derives from
\begin{equation} \langle n^j \rangle = \sum_{n}n^jP(n),\end{equation}
while the $j$th order moment of $n$ held conditional on $m$ is
\begin{equation} \langle n^j|m \rangle = \sum_{n}n^j P(n,m) .\end{equation}
\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[width=0.5\textwidth,keepaspectratio]{variance.pdf}
	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as posited in equation (\ref{eq:active}): $\sigma_m^2 \approx (l/z_1)^2$. }
	\label{fig:var}
\end{wrapfigure}
We observe no dependence of the moments of $m$ on the value of $n$. 
The mean elevation is always $\langle m \rangle = 0 $ due to our initial assumption of symmetry in the entrainment and deposition rate scaling with $m$. 
Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in equation (\ref{eq:ou2}); this result supports our earlier assertion that $l$ is a characteristic length scale of bed elevation fluctuations.
The close correspondence between the mean field approximation and the numerical simulations in figure (\ref{fig:pdfs}a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result. We find them to be identical within numerical uncertainty.

The coupling between bed elevation changes and the erosion and deposition rates develop a strong dependence of the particle activity on $m$. Figure (\ref{fig:condmoms}) displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
\begin{figure}[!ht]
	\includegraphics[width=\linewidth,keepaspectratio]{momentsuppression.pdf}
	\caption{The shifts between particle activity moments conditioned on instantaneous elevations and their over-all mean values. Panel (a) indicates the mean particle activity shift versus the bed elevation measured in units of $\sigma_z=l$. This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$. The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations. These results indicate that bedload statistics measurements on short timesclaes could be severely biased by departures from the mean bed elevation.}
	\label{fig:condmoms}
\end{figure}
Figure (\ref{fig:condmoms}a) demonstrates that the \citet{Ancey2008} flow conditions support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry in this quantity.
Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
For the parameters of the \citet{Ancey2008} experiments, figure (\ref{fig:condmoms}b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
We summarize that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.

\subsection{Collective entrainment and bedload activity fluctuations}
\label{sec:colent}
Noting that bed elevations regulate the particle activity moments, we now study the influence of collective entrainment on this effect by modifying the relative proportion of the individual to collective contributions in the mean entrainment rate $E=\lambda + \sigma \langle n \rangle $.
Using the equilibrium condition $E=D$, we determine the fraction of entrainment due to the collective process is $f = \mu\langle n \rangle/E = \mu/\sigma$. Using this fraction, we can hold $E$ constant and modify the prevalence of the collective entrainment process by setting $\lambda = E(1-f)$ and $\mu= \sigma f$. As we interpolate $f$ between zero and one, the particle activity component of the master equation \ref{eq:master} interpolates from a purely Poissonian model \citep{Ancey2006} to a negative binomial model \citep{Ancey2008}, isolating the imprint of collective entrainment on particle activity statistics over a dynamic sedimentary bed.
\begin{figure}[!ht]
	\includegraphics[width=\linewidth,keepaspectratio]{colent-suppression.pdf}
	\caption{The shift of the mean particle activity in panel (a) and its fluctuations in panel (b) with departures of the bed elevation from its mean. All simulations are at flow condition (g) from table \ref{tab:anceyparams} except $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment. Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (\ref{eq:ancey}) at large departures from the mean bed elevation. Panel (b) shows particle activity fluctuations suppressed by 90\% when $z\approx -3l$ and collective entrainment is the dominant process. When collective entrainment is absent, meaning $\mu/\sigma=0$, this moment regulation effect vanishes: it is a consequence of collective entrainment.}
	\label{fig:colent}
\end{figure}
Figure \ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed. When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect we introduced in section \ref{sec:mom}.

\subsection{Resting times of sediment undergoing burial}
\label{sec:rtcdf}

\begin{figure}[!ht]
	\includegraphics[width=\linewidth,keepaspectratio]{rtcdf.pdf}
	\caption{Resting time statistics scale differently with transport conditions and the bed elevation variance. Panel (a) shows differing flow conditions at a fixed $l$ value, while panel (c) shows fixed flow conditions at differing $l$. When scaled by $T_0$ (\ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d). In panels (b) and (d), the black dotted lines indicate a power law decay of the collapsed tails having parameter $\alpha\approx1.18$ .}
	\label{fig:cdfs}
\end{figure}

Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the time-series of $m$ \citep[e.g.,][]{Redner2007}.
Following \citet{Voepel2013} and \citet{Martin2014}, we concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
These are the return times from above of the sedimentary bed conditional to the elevation $m'$.
Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
Using the marginal probability distribution of bed elevations $P(m)$ (figure \ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}: 
\begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
A representative subset of these results are displayed in figure \ref{fig:cdfs}.
Comparing panels \ref{fig:cdfs}(a) and \ref{fig:cdfs}(c) shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic timescale $T_0$ is found to collapse away both variations.
We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
This is the mean number of grains leaving the bed per unit time is $E$, and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}). 
Therefore, bed elevations change with a characteristic speed $v = z_1 E$.
Since the range of elevation deviations is $l$ (figure \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
\begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
When scaling the resting time by this $T_0$, we obtain the collapse shown in figure \ref{fig:cdfs}.
Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
These distributions are sufficiently heavy tailed to violate the central limit theorem and drive anomalous super-diffusion of bedload, a result which supports the earlier conclusions of \citet{Voepel2013} and \citet{Martin2014}.

\section{Discussion}
\label{sec:discussion}

\subsection{Context}
Einstein developed the first stochastic models of bedload tracer diffusion \citep{Einstein1937} and the bedload flux \citep{Einstein1950}, and his ideas can be viewed as the nexus of an entire paradigm of research that extends into the present day \citep[e.g.,][]{Hubbell1964, Nakagawa1976,Hassan1991,Ancey2008, Wu2019}.
These models aim to predict bedload transport characteristics from stochastic concepts of individual particle motions.
With some exceptions \citep{Yang1971,Nakagawa1980,Pelosi2016,Wu2019,Wu2019a}, existing descriptions are spatially one-dimensional, concentrating on the motion of grains in the downstream direction without including the vertical dimension wherein local bed elevation changes imply sediment burial \citep{Voepel2013,Martin2014} and change the mobility of surface grains \citep{Yang1971,Nakagawa1980}.

\subsection{Contributions}

In this paper, we have built on earlier works \citep{Ancey2008,Martin2014} to include the vertical dimension of bed elevation dynamics, study the interplay between bedload transport and bed elevation fluctuations, and investigate resting time distributions of sediment undergoing burial.
To our knowledge, this model is the first description of bedload transport and bed elevations as a coupled stochastic population model based on individual grains.
Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015a} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.

Our inclusion of coupling between the bed elevation and entrainment and deposition rates revealed a novel dependence of particle activity on bed elevation changes,  highlighting a new consequence of the collective entrainment process \citep{Ancey2008,Lee2018}. This coupling develops a significant variation of the particle activity moments with deviations of the bed from its mean elevation. We isolated the role of collective entrainment in this bedload activity regulation, and pointed out that particle activity variations with bed elevations dissapear in the absence of collective entrainment.
Finally, we obtained resting times for sediment undergoing burial within the sedimentary bed by analyzing return times from above in the bed elevation time-series. We found heavy-tailed power law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
The distribution tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.

As our model builds on earlier works describing particle activity and bed elevation changes independently, it also reduces to these works in simplified limits when the coupling between the particle activity and bed elevation vanishes.
With the mean field approach in section \ref{sec:analytical}, we derived the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity as simplified limits of our coupled model. While the mean field description of bed elevations over-predicts the bed elevation variance by approximately a factor of two, it does capture the Gaussian shape of the bed elevation distribution, and its conclusions on the tail characteristics of resting time distributions for sediment undergoing burial are identical to ours within the numerical uncertainty: \citet{Martin2014} described a power-law distribution with tail parameter $\alpha \approx 1$ which falls neatly within our estimation $\alpha = 1.18 \pm 0.32$. In addition to our original contributions, we have corroborated the models of \citet{Ancey2008} and \citet{Martin2014} from an alternate perspective, showing their results to be mostly robust when accounting for bed elevation changes.

\subsection{Next steps}

The model we have presented computes statistical characteristics of the bedload particle activity and bed elevation within a control volume by assuming all particles on the bed surface have similar mobility characteristics while sediment transport and bed topography are in equilibrium. In actuality, particles span a range of sizes, and spatial organization occurs both in the forces imparted to particles by the flow \citep{Shih2017,Amir2014} and in the mobility characteristics of particles on the bed surface \citep{Charru2004, Hassan2008, Nelson2014}.
Together, these factors may generate spatial correlations in particle activities that models concentrating on a single control volume will be unable to capture. 
Models chaining multiple control volumes together have shown spatial correlations in the particle activity as a result of collective entrainment \citep{Heyman2014, Ancey2015}, and similar approaches have also been applied to study correlations in turbulent flows \citep{Gardiner1983}. In light of this work, we consider the model we have presented as a preliminary step toward a multiple-cell model of particle activities and bed elevation changes with potential to express spatial correlations between longitudinal profile and particle activity statistics.

Like \citet{Martin2014}, we obtained heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016}. 
Our resting time distributions show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2000} which is within range of our results.
Divergent mean resting time distributions present a paradox, since they imply all particles should eventually be immobile, violating the equilibrium transport assumption.
\citet{Voepel2013} demonstrated that a bounded random walk for bed elevations provides a power-law distribution that eventually transitions to a faster thin-tailed decay, allowing for power-law scaling like our result and \citet{Martin2014} without this divergent mean paradox. 
One resolution to this issue could come from a spatially distributed model with multiple cells. Neighboring locations might bound excessive local elevation changes through granular relaxations from gradients above the angle of repose. In this interpretation, divergent mean power law resting time distributions may be relics of single cell models for bed elevation changes. We should always expect a maximum depth to which the bed can degrade relative to neighboring locations; this could temper the power law tail without required the reflecting boundaries used by \citet{Voepel2013}.

Finally, we studied probability distribution functions and first and second moments of the particle activity and bed elevation, making novel conclusions about coordination between the statistical characteristics of these quantites which deserve experimental testing.
In the last decade, particle tracking experiments have emerged \citep{Lajeunesse2010,Roseberry2012,Martin2014, Fathel2015,Heyman2016,Liu2019}, that allow joint resolution of bed elevations and bedload transport. 
A suitably designed experiment could test our prediction that bed elevations regulate particle activity statistics, as essentially represented in figures \ref{fig:condmoms} and \ref{fig:colent}. 
However, we have left many other statistical characteristics of bedload transport for future studies. 
For example, the dependence of bedload transport \citep{Saletti2015a,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic. Statistical quantities can either be monoscaling or multiscaling across the observation scale \citep{Sornette2000}, and we currently lack physical understanding and general conclusions about the scale dependence of particle activity and bed elevation signals.
The model we have presented shows statistical monoscaling for both quantites \citep[e.g.][]{Saletti2015a}, whereas other experiments indicate statistical multiscaling \citep{Aberle2006,Singh2009,Singh2012}. We consider this topic to go beyond the scope of the present work, and we have focused on statistical characteristics at the highest temporal resolutions, with no averaging over the observation scale.

\section{Conclusion}
\label{sec:conclusion}

We developed a stochastic model for particle activity and local bed elevations including feedbacks between elevation changes and the erosion and deposition rates.
This model includes collective entrainment, whereby moving particles tend to destabilize stationary ones.
We analyzed this model using a mixture of numerical and analytical methods and provided two key results:
\begin{enumerate}
\item Resting times for sediment undergoing burial lie on a heavy-tailed power law distributions with tail parameter $\alpha \approx 1.2$;
\item Collective entrainment generates a statistical regulation effect, whereby bed elevation changes modify the mean and variance of the particle activity by as much as 90\%: this effect vanishes when collective entrainment is absent.
\end{enumerate}
These results imply measurements of bedload transport statistics could be severely biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
Next steps are to generalize our model to a multi-cell framework and to study spatial correlations in bed elevation and particle activity statistics.

\acknowledgments
The simulation code is available at
\sloppy
\url{https://drive.google.com/file/d/1bWEX_ZHlcGRKCxtRIadnSbhh4EpQlsoh/view?usp=sharing}.
This research was supported by an NSERC Discovery Grant to M. Hassan.
We would like to thank Shawn Chartrand and Conor McDowell for helpful discussions.

\bibliography{biblio}
\end{document}

### 2021-06-24 clean.sh

@@ -1,24 +1,28 @@
 #!/bin/bash
 # this cleans out all of the auxilary and log files after compilation
-rm -f ch5expt/ch5expt.aux
+rm -f .abs/abs.aux
+rm -f .ack/ack.aux
+rm -f .ch1intro/ch1intro.aux
+rm -f .ch2flipFlop/ch2flipFlop.aux
+rm -f .ch2flipFlop/ch2flipFlop.log
+rm -f .ch3bedEle/ch3bedEle.aux
+rm -4 .ch3bedEle/ch3bedEle.log
+rm -f .ch4downDiff/ch4downDiff.aux
+rm -f .ch4downDiff/ch4downDiff.log
+rm -f .ch5conc/ch5conc.aux
+rm -f .ch5conc/ch5conc.log
+rm -f .laySum/laySum.aux
+rm -f .laySum/laySum.log
+rm -f .pref/pref.aux
+rm -f .pref/pref.log
 rm -f appendix.aux
-rm -f diss.toc
-rm -f laySum/laySum.aux
-rm -f ch3downDiff/ch3downDiff.aux
-rm -f ch2bedEle/ch2bedEle.aux
-rm -f diss.brf
-rm -f ack/ack.aux
-rm -f diss.log
-rm -f diss.lof
-rm -f ch6conc/ch6conc.aux
 rm -f diss.aux
-rm -f pref/pref.aux
-rm -f abs/abs.aux
-rm -f diss.lot
-rm -f ch4simul/ch4simul.aux
-rm -f diss.blg
 rm -f diss.bbl
-rm -f diss.synctex.gz
-rm -f ch1intro/ch1intro.aux
+rm -f diss.blg
+rm -f diss.brf
+rm -f diss.lof
+rm -f diss.log
+rm -f diss.lot
 rm -f diss.out
-rm -f glos/glos.aux
+rm -f diss.synctex.gz
+rm -f diss.toc

### 2021-06-24 formerPapers/grlRrevision2/pointbypoint/point-by-point.tex
got one chapter in. some issues.

\documentclass[11pt]{article}
% General document formatting
\usepackage[margin=0.75in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{titling}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{booktabs} % for wrapping tabulars in accord with
\bibliographystyle{agu}

%\usepackage[math]{kurier}
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}
\newcommand\El{\mathcal{L}}
\newcommand\tg{\tilde{g}}
\newcommand\tG{\tilde{G}}
\begin{document}

	\title{Response to reviewers: ``Back to Einstein: burial-induced three-range diffusion in fluvial sediment transport''}
	\author{J. Kevin Pierce \& Marwan A. Hassan}
	\maketitle

\section*{Associate Editor}
I agree with the reviewers that the revisions have improved the manuscript but there are new points raised by the reviewers, in particular the comments of reviewer 1 on the number of regimes identified and the need for a consistent story, rather than a timeline style of presentation of the results, and edits from reviewer 2.

\paragraph{Reply:}
We modified the manuscript in response to every comment of both reviewers. 
We clarified the number of regimes identified (four variance scaling ranges, three of which are diffusive) at L20, 202, 226, and 273. We removed the timeline style presentation of the results, and we incorated all edits from reviewer \#2.

\section*{Reviewer \#1}

\begin{enumerate}
\item 'However, this context changes the main story and we now have a manuscript with “three range diffusion” in the title, but “four range” in the key points and the key figure- figure 2. Also, lines 221-226 describe three ranges, then the 4 th range is mentioned after the list but not counted.'
\paragraph{Reply:} 
We reconciled the inconsistency you pointed out between "three range" and "four range" within the revised manuscript. There are four variance scaling ranges but only three ``diffusion'' ranges. The fourth ``geomorphic'' variance scaling range does not involve the movement of any particles (at least with permanent burial), so it isn't a diffusion range. We clarified this distinction in the revised manuscript at L20, 202, 226, and 273 so we could write ``three-range diffusion'' without ambiguity.

\item I suggest that the authors remove their timeline of discovery (eg lines 240-250 “in the preliminary studies for this paper we found that a two-state....” “these realizations and the need for a physical model...” )
\paragraph{Reply:}
We removed the timeline of discovery and rephrased the underlying point that the Nikora et al. suggestion of a two state model with heavy-tailed rests does not generate three diffusion ranges.

\item  All previous descriptions of the global regime fit within this model depending on the timescale/tail relationship between rest and burial. Similarly, existence of a geomorphic range (or at least the slope on figure 2) is a function of permanent burial or not. The multi state model generalizes all random walk findings and highlights the physical process timescales that are significant. If there are other physical processes that are relevant, they can be
added. All of this should be reflected in Figure 2-
\paragraph{Reply:} 
We comprehensively revised figure 2 to include all components of your thoughtful summary. We highlighted the variation of global range diffusion with $k_1/\kappa$ (ratio of burial time to resting time), we illustrated the anticipated result of impermanent burial in the geomorphic range, and we added an additional panel with a schematic of bedload trajectories through motion, rest, and burial states to indicate the physical processes that generate each diffusion range.
\end{enumerate}

We appreciate your attention toward improving our manuscript and hope we have addressed your concerns sufficiently.

\section*{Reviewer \#2}

\subsection*{General Comments}

The revised manuscript appropriately addresses the first set of reviews by placing a larger emphasis on the geophysical aspects of the problem. The revised manuscript is suitable for publication following minor revisions or a few changes that could potentially happen within the copy-editing phase.

\subsection*{Specific Comments:}
\begin{enumerate}
\item Abstract. The abstract would benefit from some general editing. 7/12 lines are introducing the problem rather than serving as a synopsis of the paper. Some of this would be better suited to the introduction section.
\paragraph{Reply:} We removed these lines and replaced them with a synopsis of the paper. The first several lines of the abstract now introduce the problem, the middle lines state what we did to solve the problem, and the final lines summarize our key findings.

\item Ln. 64. This is a bit of a miss citaiton here. Both Bradley (2017) and Phillips et al. (2013) only consider time above a transport threshold thus specifically excluding the time spent resting during low flow between floods.
\paragraph{Reply:} We modified this sentence from the terminology ``between subsequent floods'' to represent our point that sampling timescales of tracer positions in these studies are at least as long as individual floods.

\item Ln. 278-282. In Weeks and Swinney (1998) long time subdiffusion is uncommon for asymmetric random walks requiring very heavy-tailed waiting times. Given re-entrainment from burial is likely in all but aggrading systems is long term subdiffusion really a viable solution for natural rivers?
\paragraph{Reply:} We are unsure whether long-time sub-diffusion is viable, so we modified this section to better communicate the existing support for both possibilities. The field study of  \citet{Olinde2015} shows sufficiently heavy-tailed resting times for long-time sub-diffusion, but their methods do not discern whether these rests are due to burial or some other process. Several models including our recent JGR paper show that burial times under equilibrium transport conditions are not sufficiently heavy-tailed for sub-diffusion \citep{Pierce2020,Martin2014,Voepel2013}. Similarly, \citet{Bradley2017} presents resting times that were insufficient for long-time sub-diffusion. The evidence mostly points toward long-time super-diffusion, but the limited data leave us not confident enough to refute the Nikora et al. hypothesis.
\end{enumerate}

Thank you sincerely for your constructive comments and attention toward our work.

\bibliography{biblio}
\end{document}

### 2021-06-24 abs.tex
got one chapter in. some issues.

--- 2021-06-24 ch1intro/ch1intro.tex
@@
 
-\chapter{Sediment transport and landscape evolution}
-\label{ch:Introduction}
+\chapter{Abstract}
 
-Landscapes evolve across geologial time due to energy and mass contributed from above and below the Earth's surface.
-
-\section{Stochastic properties of sediment movement}
-
-\section{Historical survey: statistical physics in sediment transport}
-
-\section{Philosophy of the statistical physics approach}
-
-\section{What's left?}
-
-\endinput
-
-Any text after an \endinput is ignored.
-You could put scraps here or things in progress.
+Write an abstract.

### 2021-06-24 ack.tex
got one chapter in. some issues.

%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Acknowledgments}

Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way. Tyler cannon. . .

Shawn Chartrand, Conor McDowell, Matteo Saletti, Will Booker, David Adams, Leo King, Tobias Mueller, Nisreen Ghazi,

Adam Payne, Adam Collins, Megat Denney, Andrew Rice, Evan Graber, Scott Ferris, Craig Tenney, and the rest of the WVU physics lounge family.

Andy Osadetz, Josh Plankeel,

My partner Mary for putting up with the late nights.

### 2021-06-24 ch2flipFlop.tex
got one chapter in. some issues.

%%!TEX root = diss.tex

\chapter{Calculation of the sediment flux}

\section{Introduction}
\section{Theoretical development}

### 2021-06-24 ch4downDiff.tex
got one chapter in. some issues.

%%!TEX root = diss.tex

\chapter{Inclusion of sediment burial}
\label{ch:downDiff}

Many environmental problems including channel morphology \citep{Hassan2017}, contaminant transport \citep{Macklin2006}, and aquatic habitat restoration \citep{Gaeuman2017} rely on our ability to predict the diffusion characteristics of coarse sediment tracers in river channels.
Diffusion is quantified by the time dependence of the positional variance $\sigma_x^2$ of a group of tracers.
With the scaling $\sigma_x^2 \propto t$, the diffusion is said to be normal, since this is found in the classic problems \citep{Taylor1920}.
However, with the scaling $\sigma_x^2 \propto t^\gamma$ with $\gamma \neq 1$, the diffusion is said to be anomalous \citep{Sokolov2012}, with $\gamma>1$ defining super-diffusion and $\gamma<1$ defining sub-diffusion \citep{Metzler2000}.
\citet{Einstein1937} developed one of the earliest models of bedload diffusion to describe a series of flume experiments \citep{Ettema2014}.
Interpreting individual bedload trajectories as a sequence of random steps and rests, Einstein originally concluded that a group of bedload tracers undergoes normal diffusion.

More recently, Nikora et al. realized coarse sediment tracers can show either normal or anomalous diffusion depending on the length of time they have been tracked \citep{Nikora2001a,Nikora2002}.
From numerical simulations and experimental data, Nikora et al. discerned ``at least three'' scaling ranges $\sigma_x^2 \propto t^\gamma$ as the observation time increased.
They associated the first range with ``local'' timescales less than the interval between subsequent collisions of moving grains with the bed, the second with ``intermediate'' timescales less than the interval between successive resting periods of grains, and the third with ``global'' timescales composed of many intermediate timescales.
Nikora et al. proposed super-diffusion in the local range, anomalous or normal diffusion in the intermediate range, and sub-diffusion in the global range.
They attributed these ranges to ``differences in the physical processes which govern the local, intermediate, and global trajectories'' of grains \citep{Nikora2001a}, and they called for a physically based model to explain the diffusion characteristics \citep{Nikora2002}.

Experiments support the Nikora et al. conclusion of multiple scaling ranges \citep{Martin2012,Fathel2016}, but they do not provide consensus on the expected number of ranges or their scaling properties.
This lack of consensus probably stems from resolution issues.
For example, experiments have tracked only moving grains, resolving the local range \citep{Furbish2012,Furbish2017,Fathel2016}; grains resting on the bed surface between movements, resolving the intermediate range \citep{Einstein1937,Yano1969,Nakagawa1976}; grains either moving or resting on the bed surface, likely resolving local and intermediate ranges \citep{Martin2012}; or grains resting on the surface after floods, likely resolving the global range \citep{Phillips2013,Bradley2017}. 
At long timescales, a significant fraction of tracers bury under the bed surface \citep{Hassan1991,Hassan2013,Ferguson2002a,Haschenburger2013,Papangelakis2016}, meaning burial dominates long term diffusion characteristics \citep{Bradley2017,Martin2014,Voepel2013}, possibly at global or even longer ``geomorphic'' timescales \citep{Hassan2017} than Nikora et al. originally considered.
As a result, three diffusion ranges can be identified by patching together multiple datasets \citep{Zhang2012,Nikora2002}, but they are not resolved by any one dataset.

Newtonian bedload trajectory models also show multiple diffusion ranges, although they also do not provide consensus on the expected number of ranges or their scaling properties. 
The majority of these models predict two ranges of diffusion (local and intermediate) without predicting a global range.
Among these, \citet{Nikora2001a} used synthetic turbulence \citep{Kraichnan1970} with a discrete element method for the granular phase \citep{Cundall1979}; \citet{Bialik2012} used synthetic turbulence with a random collision model \citep{Sekine1992}; and \citet{Fan2016} used a Langevin equation with probabilistic rests.
To our knowledge, only \citet{Bialik2015} have claimed to capture all three ranges from a Newtonian approach.
They incorporated a second resting mechanism into their earlier model \citep{Bialik2012}, implicitly suggesting that three diffusion ranges could result from two distinct timescales of sediment rest.
However, Newtonian approaches have not evaluated the effect of sediment burial on tracer diffusion, probably due to the long simulation timescales required.

Random walk bedload diffusion models constructed in the spirit of \citet{Einstein1937} provide an alternative to the Newtonian approach and can include a second timescale of rest by incorporating sediment burial.
Einstein originally modeled bedload trajectories as instantaneous steps interrupted by durations of rest lying on statistical distributions \citep{Hassan1991}, but this generates only one range of normal diffusion \citep{Einstein1937,Hubbell1964,Nakagawa1976}.
Recently, researchers have generalized Einstein's model in a few different ways to describe multiple diffusion ranges.
\citet{Lisle1998} and \citet{Lajeunesse2018} promoted Einstein's instantaneous steps to motion intervals with random durations and a constant velocity, providing two diffusion ranges -- local and intermediate.
\citet{Wu2019} retained Einstein's instantaneous steps but included the possibility that grains can become permanently buried as they rest on the bed, also providing two diffusion ranges -- intermediate and global. 
These earlier works suggest the minimal required components to model three bedload diffusion ranges: (1) exchange between motion and rest intervals and (2) the sediment burial process.

In this study, we incorporate these two components into Einstein's original approach to describe three diffusion ranges with a physically based model, as called for by \citet{Nikora2002}.
Einstein was a giant in river geophysics and fostered an entire paradigm of research leveraging and generalizing his stochastic methods \citep{Hubbell1964, Yano1969, Yang1971, Gordon1972, Nakagawa1976,Paintal1971}.
Einstein's model can be viewed as a pioneering application of the continuous time random walk (CTRW) developed by \citet{Montroll1965} in condensed matter physics to describe the diffusion of charge carriers in solids.
To incorporate motion intervals and sediment burial, we utilize the multi-state CTRW developed by \citet{Weiss1976, Weiss1994} that extends the CTRW of \citet{Montroll1965}.
Below, we develop and solve the model in section \ref{sec:model}. Then, we discuss the predictions of our model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in sections \ref{sec:discussion} and \ref{sec:conclusion}.

\section{Bedload trajectories as a multi-state random walk}
\label{sec:model}
\subsection{Model assumptions}
\label{sec:assumptions}
We construct a three-state random walk where the states are motion, surface rest, and burial, and we label these states as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
Our target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if we know it started with the initial distribution $p(x,0)=\delta(x)$.
We characterize times spent moving or resting on the surface by exponential distributions $\psi_2(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}. We expect our conclusions will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
We consider grains in motion to have characteristic velocity $v$ \citep{Lisle1998,Lajeunesse2018}, and we model burial as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried, meaning $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
We specify the initial conditions with probabilities $\theta_1$ and $\theta_2$ to be in rest and motion at $t=0$, and we require $\theta_1+\theta_2=1$ for normalization.

\subsection{Governing equations}
Using these assumptions, we derive the governing equations for the set of probabilities $\om_{ij}(x,t)$ that a transition occurs from state $i$ to state $j$ at position $x$ and time $t$ using the statistical physics approach to multi-state random walks \citep{Weiss1994,Schmidt2007,Weeks1998}.
Denoting by $g_{ij}(x,t)$ the probability for a particle to displace by $x$ in a time $t$ within the state $i$ before it transitions to the state $j$, the transition probabilities $\om_{ij}(x,t)$ sum over all possible paths to the state $i$ from previous locations and times:
\be \om_{ij}(x,t) = \theta_i g_{ij}(x,t) + \sum_{k=0}^2 \int_0^x dx' \int_0^t dt' \om_{ki}(x',t')g_{ij}(x-x',t-t').\label{eq:g1}\ee
Defining another set of probabilities $G_i(x,t)$ that a particle displaces by a distance $x$ in a time $t$ within the state $i$ and possibly remains within the state, we perform a similar sums over paths for the probabilities to be in the state $i$ at $x$, $t$: 
\be p_i(x,t) = \theta_i G_i(x,t) + \sum_{k=0}^2 \int_0^x dx' \int_0^t dt' \om_{ki}(x',t')G_i(x-x',t-t').\label{eq:g2}\ee
Finally, the overall probability to be at position $x$ at time $t$ is
\be p(x,t) = \sum_{k=0}^2 p_k(x,t) \ee
This joint density is completely determined from the solutions of equations (\ref{eq:g1}-\ref{eq:g2}). We only need to specify the distributions $g_{ij}$ and $G_i$.

\subsection{Joint probability distribution}
\label{sec:solution}

We construct these distributions from the assumptions described in section \ref{sec:assumptions}.
Since particles resting on the bed surface bury in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, we obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$. Since motions have velocity $v$ for times distributed as $\psi_2(t)$, we have $g_{21}(x,t) = \delta(x-vt)k_2e^{-k_2 t}$.
Since burial is quasi-permanent, all other $g_{ij} = 0$.
The $G_i$ are constructed in the same way except using the cumulative probabilities $\int_t^\infty dt'\psi_i(t) = e^{-k_i t}$, since these characterize motions and rests that are ongoing \citep{Weiss1994}.
We obtain $G_1(x,t) = \delta(x)e^{-k_1t}$ and $G_2(x,t) = \delta(x-vt)e^{-k_2 t}$.

To solve equations (\ref{eq:g1}-\ref{eq:g2}) with these $g_{ij}$ and $G_i$, we take Laplace transforms in space and time $(x,t \rightarrow 
\eta,s) $ using a method similar to \citet{Weeks1998} to unravel the convolution structure of these equations, eventually obtaining
\be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}, \label{eq:nicedist}\ee
where $k' = k_1 + k_2$. Inverting this result using known Laplace transforms \citep{Prudnikov1986,Arfken1985} obtains
\begin{align}
	\begin{split}
		p(x,t) = \theta_1&\Big[1-\frac{k_1}{\kappa+k_1}\Big(1-e^{-(\kappa+k_1)t}\Big)\Big]\delta(x) \\ &+ \frac{1}{v}e^{-\Omega \tau - \xi}\Big(\theta_1\Big[k_1\mathcal{I}_0\big(2\sqrt{\xi\tau}\big) + k_2\sqrt{\frac{\tau}{\xi}}\mathcal{I}_1\big(2\sqrt{\xi\tau}\big)\Big] \\ 
		&+ \theta_2\Big[k_1\delta(\tau) + k_2 \mathcal{I}_0\big(2\sqrt{\xi\tau}\big)+k_1 \sqrt{\frac{\xi}{\tau}}\mathcal{I}_1\big(2\sqrt{\xi\tau}\big)\Big]\Big) \\
		&+ \frac{1}{v}\frac{\kappa k_2}{\kappa + k_1}e^{-\kappa \xi/(\kappa + k_1)}\Big[(\theta_1/\Omega)\mathcal{Q}_2(\xi/\Omega,\Omega\tau) + \theta_2 \mathcal{Q}_1(\xi/\Omega,\Omega\tau)\Big]
		\label{eq:pdf}
	\end{split}
\end{align}
for the joint distribution that a tracer is found at position $x$ at time $t$.
This result generalizes the earlier results of \citet{Lisle1998} and \citet{Einstein1937} to include sediment burial.
In this equation, we used the shorthand notations $\xi = k_2 x/v$, $\tau = k_1(t-x/v)$, and $\Omega = (\kappa+k_1)/k_1$ \citep{Lisle1998}. The $\mathcal{I}_\nu$ are modified Bessel functions of the first kind and the $\mathcal{Q}_\mu$ are generalized Marcum Q-functions defined by $\mathcal{Q}_\mu(x,y) = \int_0^y e^{-z-x}(z/x)^{(\mu-1)/2}\mathcal{I}_{\mu-1}(2\sqrt{xz})dz $ and originally devised for radar detection theory \citep{Marcum1960,Temme1996}. 
The Marcum Q-functions derive from the burial process.
Since we assumed resting grains could bury with an exponential probability while the resting probability follows a modified Bessel distribution \citep{Einstein1937,Lisle1998}, burial develops the Q-function convolution structure.

\begin{figure}[t]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/pdf-plot.png}
	\caption{Joint distributions for a grain to be at position $x$ at time $t$ are displayed for the choice $k_1=0.1$, $k_2=1.0$, $v=2.0$. Grains are considered initially at rest ($\theta_1=1$, $\theta_2=0$). The solid lines are the analytical distribution in equation (\ref{eq:pdf}), while the points are numerically simulated, showing the correctness of our derivations. Colors pertain to different times. Units are unspecified, since we aim to demonstrate the general characteristics of $p(x,t)$. Panel (a) shows the case $\kappa=0$ -- no burial. In this case, the joint distribution tends toward Gaussian at large times \citep{Einstein1937,Lisle1998}. Panel (b) shows the case when grains have rate $\kappa = 0.01$ to become buried while resting. Because of burial, the joint distribution tends toward a more uniform distribution than Gaussian.
		\label{fig:pdfs}}
\end{figure}

Figure \ref{fig:pdfs} depicts the distribution (\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
When grains do not become buried, as in panel (a) of figure \ref{fig:pdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion and satisfying the central limit theorem.
When grains become buried, as in panel (b) of figure \ref{fig:pdfs},  the Q-function terms prevent the distribution from approaching a Gaussian at large timescales, exemplifying anomalous diffusion \citep{Weeks1998} and violating the central limit theorem \citep{Metzler2000,Schumer2009}.

\subsection{Positional variance}

To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, we derive the first two moments of position by taking derivatives with respect to $\eta$ of the Laplace space distribution (\ref{eq:nicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}, and we use these to calculate the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$. 
The first two moments are
\begin{align}
	\bra x(t) \ket &= A_1 e^{(b-a)t}+B_1e^{-(a+b)t}+C_1, \label{eq:mean}\\
	\bra x^2(t) \ket &= A_2(t)e^{(b-a)t}+B_2(t)e^{-(a+b)t}+C_2, \label{eq:second}
\end{align}
so the variance is 
\be \sigma_x^2(t) = A(t)e^{(b-a)t} + B(t)e^{-(a+b)t} + C(t). \label{eq:var}\ee
In these equations, $a = (\kappa + k_1+k_2)/2$ and $b = \sqrt{a^2-\kappa k_2}$ are effective rates having dimensions of inverse time, while the $A$, $B$, and $C$ factors are provided in table \ref{table:params}.

%\\
%		&\hspace{1cm}
\begin{table}[!h]
	\centering
	%\begin{wraptable}{r}{width=5.5cm}
	\caption{Abbreviations used in the expressions of the mean (\ref{eq:mean}), second moment (\ref{eq:second}) and variance (\ref{eq:var}) of bedload tracers.}
	\label{table:params}
	\small
	\begin{tabular}{c}
		\toprule
		$\begin{aligned}[t]
			&A_1 = \frac{v}{2b}\big[\theta_2+\frac{k_1+\theta_2\kappa}{b-a}\big] \\
			&B_1 = -\frac{v}{2b}\big[\theta_2-\frac{k_1+\theta_2 \kappa}{a+b}\big] \\
			&C_1 =  -\frac{v}{2b}\big[\frac{k_1+\theta_2 \kappa}{b-a}+\frac{k_1+\theta_2 \kappa}{a+b}\big]\\
			&A_2(t) = \frac{v^2}{2b^3}\Big[(bt-1)[k_1+\theta_2(2\kappa + k_1 + b-a)]+\theta_2b 
			+ \frac{(\kappa+k_1)(\theta_2\kappa+k_1)}{(b-a)^2}[(bt-1)(b-a)-b]\Big]\\
			&B_2(t) = \frac{v^2}{2b^3}\Big[(bt+1)[k_1 + \theta_2(2\kappa+k_1-a-b)]+\theta_2b
			-\frac{(\kappa+k_1)(\theta_2\kappa+k_1)}{(a+b)^2}[(bt+1)(a+b)+b]\Big]\\
			&C_2 = \frac{v^2}{2b^3}(\kappa+k_1)(\theta_2 \kappa + k_1)\Big[\frac{2b-a}{(b-a)^2}+\frac{a+2b}{(a+b)^2}\Big]\\
			&A(t) = A_2(t)-2A_1C_1 - A_1^2\exp[(b-a)t]\\
			&B(t) = B_2(t)-2B_1C_1 - B_1^2\exp[-(a+b)t]\\
			&C(t) = C_2-C_1^2-2A_1B_1\exp[-2at]\\			
		\end{aligned}$\\
		\bottomrule
	\end{tabular}
	%\end{wraptable}
	\vspace{-0.5cm}
\end{table}
The positional variance (\ref{eq:var}) is plotted in figure \ref{fig:var} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
We interpret ``$\gg$'' to mean ``of at least an order of magnitude greater''.
These conditions are most relevant to tracers in gravel-bed rivers, since they represent that grains are initially at rest \citep{Hassan1991,Wu2019}, motions are typically much shorter than rests \citep{Einstein1937,Hubbell1964}, and burial requires a much longer time than typical rests  \citep{Ferguson2002,Hassan1994,Haschenburger2013}.
Figure \ref{fig:var} demonstrates that under these conditions the variance (\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that we identify as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers. 
We suggest to call the fourth range geomorphic, since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.

\begin{figure}[h!]	
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/diffusion.png}
	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (\ref{eq:var}) with mean motion time $1.5$ s, resting time $30.0$ s, and movement velocity $0.1$ m/s -- values comparable to laboratory experiments transporting small ($5$ mm) gravels \citep{Lajeunesse2010,Martin2012}. The burial timescale is $7200.0$s (two hours), and grains start from rest ($\theta_1=1$). The solid line is equation (\ref{eq:var}), and the points are numerically simulated. Panel (b) demonstrates four distinct scaling ranges of $\sigma_x^2$: local, intermediate, global, and geomorphic. The first three are diffusive. Three crossover times $\tau_L$, $\tau_I$, and $\tau_G$ divide the ranges. Within each range, a slope key demonstrates the scaling $\sigma_x^2 \propto t^\gamma$. Panel (a) demonstrates that different mixtures of motion, rest, and burial states generate the ranges. At local timescales, grains usually either rest or move; at intermediate timescales, they transition between rest and motion; at global timescales, they transition between rest, motion, and burial; and at geomorphic timescales, all grains bury. Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as we discuss in section \ref{sec:discussion}.
	}
	\label{fig:var}
\end{figure}

\subsection{Diffusion exponents}

Two limiting cases of equation (\ref{eq:var}) provide the scaling exponents $\gamma$ of the diffusion $\sigma_x^2 \propto t^\gamma$ in each range. Limit (1) represents times so short a negligible amount of sediment burial has occurred, $t\ll 1/\kappa$, while limit (2) represents times so long motion intervals appear as instantaneous steps of mean length $l=v/k_2$, $1/k_2 \rightarrow 0$ while $v/k_2 = \text{constant}$.
Limit (1) provides local exponent $2 \leq \gamma \leq 3$ depending on the initial conditions $\theta_i$, and intermediate exponent $\gamma=1$.
If grains start in motion or rest exclusively, meaning one $\theta_i = 0$, the local exponent is $\gamma=3$, while if grains start in a mixture of motion and rest states, meaning neither $\theta_i$ is zero, the local exponent is $\gamma=2$.
Limit (2) provides global exponent $1 \leq \gamma \leq 3$ depending on the relative importance of $\kappa$ and $k_1$.
In the extreme $k_1/\kappa \ll 1$, we find $\gamma=1$ in the global range, while in the opposite extreme $k_1/\kappa \rightarrow \infty$ we find $\gamma=3$.
We summarize when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, equation (\ref{eq:var}) implies:
\begin{enumerate}
	\item local range super-diffusion with $2<\gamma<3$ depending on whether grains start from purely motion or rest ($\gamma=3$) or from a mixture of both states ($\gamma=2$),
	\item intermediate range normal diffusion $\gamma=1$ independent of model parameters, and
	\item global range super-diffusion $1<\gamma<3$ depending on whether burial happens relatively slowly ($\gamma \rightarrow 1$) or quickly ($\gamma \rightarrow 3$) compared to surface resting times.
\end{enumerate}
Finally, the burial of all tracers generates a geomorphic range of no diffusion.

\section{Discussion}
\label{sec:discussion}

\subsection{Local and intermediate ranges with comparison to earlier work}

We extended \citet{Einstein1937} by including motion and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010}, before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
\citet{Nikora2002} highlighted the need for such a physical description, although they suggested to use a two-state random walk between motion and rest states with heavy-tailed resting times, and they did not discuss sediment burial.
However, other works have demonstrated that a two-state walk with heavy-tailed rests provides two diffusion ranges -- not three \citep{Weeks1996,Fan2016}, and although heavy-tailed resting times have been documented for surface particles \citep{Liu2019,Fraccarollo2019}, they are more often associated with buried particles \citep{Martin2012,Martin2014,Voepel2013,Olinde2015,Pelosi2016, Pierce2020}, while surface particles retain light-tailed resting times \citep{Einstein1937,Yano1969,Ancey2006,Nakagawa1976}.
Accordingly, we developed a random walk model of bedload trajectories with light-tailed surface resting times that incorporates sediment burial.

The local and intermediate range diffusion characteristics resulting from our model correspond closely to the original Nikora et al. concepts, while our global range has a different origin than Nikora et al. envisioned.
\citet{Nikora2001a} explained that local diffusion results from the non-fractal (smooth) characteristics of bedload trajectories between subsequent interactions with the bed,  while intermediate diffusion results from the fractal (rough) characteristics of bedload trajectories after many interactions with the bed.
Our model represents these conclusions: non-fractal (and super-diffusive) bedload trajectories exist on timescales short enough that each grain is either resting or moving, while fractal (and normal-diffusive) bedload trajectories exist on timescales when grains are actively switching between motion and rest states.
We conclude that local and intermediate ranges stem from the interplay between motion and rest timescales, as demonstrated by earlier two-state random walk models \citep{Lisle1998,Lajeunesse2017} and by all Newtonian models that develop sequences of motions and rests \citep{Nikora2001a, Bialik2012}, even those including heavy-tailed rests \citep{Fan2016}.

\subsection{Global and geomorphic ranges with next steps for research}

Nikora et al. explained that divergent resting times generate a sub-diffusive global range.
However, studies have demonstrated that divergent resting times can generate super-diffusion in asymmetric random walks \citep{Weeks1996,Weeks1998}, and both experiments \citep{Bradley2017,Bradley2010} and models \citep{Pelosi2016,Wu2019,Wu2019a} of bedload tracers undergoing burial have demonstrated global range super-diffusion.
While our results also show global range super-diffusion, they do not necessarily refute the Nikora et al. conclusion of sub-diffusion at long timescales.
We assumed sediment burial was a permanent condition which developed a non-diffusive geomorphic range.
In actuality, burial is a temporary condition, because bed scour can exhume buried sediment back into transport \citep{Wu2019a}, probably after heavy-tailed intervals \citep{Voepel2013,Martin2014,Pierce2020}.
We anticipate that a generalization of our model to include heavy-tailed timescales between burial and exhumation would develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in figure \ref{fig:var}.
If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested by equilibrium transport models \citep{Voepel2013, Martin2014, Pierce2020} and laboratory experiments \citep{Martin2014,Martin2012}, we expect a super-diffusive geomorphic range \citep{Weeks1998}.
However, if they decay slower than $T^{-1/2}$, as implicitly suggested by the data of \citet{Olinde2015}, we expect a genuinely sub-diffusive geomorphic range \citep{Weeks1998}, leaving Nikora et al. with the final word on long-time sub-diffusion.

The analytical solution of bedload diffusion in equation (\ref{eq:var}) reduces exactly to the analytical solutions of the \citet{Lisle1998} and \citet{Lajeunesse2017} models in the limit without burial ($\kappa \rightarrow 0$), the \citet{Wu2019} model in the limit of instantaneous steps ($k_2 \rightarrow \infty$ and $l = v/k_2$), and the original \citet{Einstein1937} model in the limit of instantaneous steps without burial.
These reductions demonstrate that the majority of recent bedload diffusion models, whether developed from Exner-type equations \citep{Wu2019,Pelosi2014,Pelosi2016} or advection-diffusion equations \citep{Lisle1998,Lajeunesse2017}, can be viewed equivalently as continuous-time random walks applied to individual bedload trajectories.
Within random walk theory, sophisticated descriptions of transport with variable velocities \citep{Zaburdaev2008,Masoliver1994}, correlated motions \citep{Escaff2018,Vicsek2012a}, and anomalous diffusion \citep{Masoliver2016,Fa2014,Metzler2014} have been developed.
Meanwhile, in bedload transport research, variable velocities \citep{Lajeunesse2010,Furbish2012,Heyman2016}, correlated motions \citep{Heyman2014,Lee2018,Saletti2020}, and anomalous diffusion \citep{Fathel2016,Bradley2017,Schumer2009} constitute open research issues.
We believe further developing the linkage between existing bedload models and random walk concepts could rapidly progress our understanding.

\section{Conclusion}
\label{sec:conclusion}
We developed a random walk model to describe sediment tracers transporting through a river channel as they gradually become buried, providing a physical description of the local, intermediate, and global diffusion ranges identified by \citet{Nikora2002}.
Pushing their ideas somewhat further, we proposed a geomorphic range to describe diffusion characteristics at timescales larger than the global range when burial and exhumation both moderate downstream transport.
At base level, our model demonstrates that (1) durations of sediment motions, (2) durations of sediment rest, and (3) the sediment burial process are sufficient to develop three diffusion ranges that terminate when all tracers become buried.
A next step is to incorporate exhumation to better understand the geomorphic range.
Ultimately, we emphasize that the multi-state random walk formalism used in this paper implicitly underlies most existing bedload diffusion models and provides a powerful tool for researchers targeting landscape-scale understanding from statistical concepts of the underlying grain-scale dynamics.

### 2021-06-24 formerPapers/ens-rev/introduction.tex
got one chapter in. some issues.

\section{Introduction: birth-death models for the probability distribution of bedload flux}

In river science, a fundamental problem is the determination of the bedload flux, or the rate of downstream movement of bedload grains \citep{Ballio2014}.
Since bedload transport has strong feedback with stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of environmental considerations. 
These considerations span a wide range, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}. 
Unfortunately, existing approaches to compute the bedload flux are inadequate. 
Predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.

Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
The bedload flux is in a cyclical feedback with its controls: these controls are the details of turbulence and bed organization \citep{Jerolmack2005}.

As a result, bedload fluxes exhibit wide fluctuations, even as average characterizations of flow and bed organization remain steady \citep{Ancey2014}.
In the most controlled laboratory experiments, with steady flows driving uniform glass beads under conditions which do not favor bedform development, instantaneous fluxes are often as much as 200\% mean values \citep{Bohm2004, Ancey2008, Heyman2014, Heyman2016}. 
In natural streams, with temporally or spatially varying grain size distributions \citep{Lisle1992, Chen2008}, unsteady flows \citep{Mao2012, FerrerBoix2015}, a variety of dynamic bed surface structures \citep{Hassan2008, Venditti2017}, variable sediment supply \citep{Madej2009, Elgueta2018}, lateral adjustment \citep{Pitlick2013, Redolfi2018}, and migrating bedforms \citep{Gomez1989, Dhont2018} all imparting additional sources of variabilility through their feedbacks with the flux, instantaneous fluxes can be even larger. 
% Iseya1987, Lisle1989, Kuhnle1998

Apparently, bedload fluxes have statistical characteristics.
Einstein was the first to recognize the statistical character of bedload transport \citep{Einstein1937}. 
He understood transport as a random switching between states of motion and rest \citep{Einstein1942, Einstein1950}. 
He called the transition from rest to motion entrainment, and he characterized it with a probability which was linked to extreme events in fluid turbulence \citep{Einstein1949, Einstein1950}. 
He called the transition from motion to rest deposition, and he considered it an implicit function of the bedload flux \citep{Einstein1942, Einstein1950}. 
By convolving over sequences of entrainment and deposition, employing some semi-emprical arguments, Einstein developed a formula for the average bedload flux \citep{Einstein1950}.

Many investigators have criticized and developed Einstein's ideas \citep{Paintal1971, Yalin1972, Shen1980, Lisle1998, Papanicolaou2002, Sun2000, Ancey2006, Armanini2015}. 
These revisions focus on the more ad hoc elements of Einstein's derivations, and they lend a more mechanistic basis, a firmer mathematical foundation, and more generality to the approach. 
Although all of these authors accepted the statistical character of bedload motion, none developed theories for a probability distribution for the bedload flux. 
This is desirable since its higher moments give unambiguous measurements of the magnitude of bedload fluctuations.

The extension of the Einstein approach to obtain a probability distribution of the bedload flux, with fluctuations stemming from the random character of entrainment and deposition, was first explored by \citet{Lisle1998, Sun2000}, and it was further developed by \citet{Ancey2006}, and the more recent work I will discuss extensively in this review.   
All of these authors revisited Einstein's ideas from a foundation in the stochastic mathematics which became formalized somewhat after Einstein's work \citep[e.g.][]{Cox1965}, treating the transitions between motion and rest states as random events characterized by probabilities. 
This enabled them to apply the formalism of continuous time Markov processes to derive a probability distribution of the bedload flux, with a mean value which was an improved version of the \citet{Einstein1950} formula, very similar to the revised formula of \citet{Yalin1972}, and a variance which provided an unambiguous prediction of the magnitude of bedload fluctuations.

Deterministic processes trace the evolution of some set of variables $\{x_1(t),x_2(t),\dots\}$ through time. 
Stochastic processes, in contrast, trace the evolution of a probability distribution for random variables $P_{X_1,X_2,\dots}(x_1,x_2,\dots;t)$ through time. 
The Markov moniker refers to the amount of memory in the process.
If the probability distribution of the variables of interest can be predicted in the future using only distribution functions from the present, the process is Markovian. 
Otherwise, if predicting the future requires knowledge of the entire history, the process is non-Markovian \citep{Cox1965, VanKampen1992}. 
Markov birth-death processes consider the probabilistic creation and annhilation of members of a population \citep{Cox1965, VanKampen1992}.

To model the bedload flux with such a process, the population considered is the number of moving particles within a control volume over the bed. 
This population is subject to creation by entrainment and annhilation by deposition \citep{Ancey2008, Turowski2009, Heyman2013, Ma2014,  Ancey2014a, Ancey2015}. 
This approach is exciting because Markov birth-death processes are highly studied in physics, chemistry, and population ecology, so there are many extensions readily available \citep{Bailey1968, Cox1965, Pielou1977, VandenBroek2012, VanKampen1992, Gillespie1992, Field2010, Mendez2015}. 
At the same time, birth-death modeling of the bedload flux remains relatively undeveloped: birth-death models are all one-dimensional and designed for a single grain size.

In this article, I review the literature on Markov birth-death models of the bedload flux.
In the review I assume some familiarity with discrete state continuous time Markov processes, which could be gleaned from reading the appropriate chapter in \citet{Cox1965}. 
These theories are exciting because they describe the mean bedload flux and its fluctuations within a volume \citep{Ancey2006, Ancey2008, Turowski2009}, statistical properties of the flux at a point in space \citep{Heyman2013, Ma2014}, and the spatial and temporal correlations in the bedload flux within a reach of channel \citep{Ancey2014, Ancey2015}. 
All of these considerations are new, and they are unique within the river science literature in that they do not ignore the statistical character of bedload transport. 
They initiate a theoretical framework from which the feedbacks between the arrangement of bed particles, turbulence, and channel morphology which lead to bedload fluctuations can be more carefully studied and understood. 
In reverse, the capacity of bedload models to describe fluctuations provides an additional benchmark against which to judge models, a task which is notably difficult \citep{Iverson2013}.

One of the first birth-death models of the bedload flux I will review is \citet{Ancey2006}: these authors generalized \citet{Einstein1950} using a birth-death framework to obtain a statistical distribution of the bedload flux within a control volume. 
They found the fluctuations predicted by the Einstein-like model were not large enough to describe experimental data.  
In order to match the wide fluctuations of experimental data, \citet{Ancey2008} reworked their earlier control volume model to include a collective entrainment effect.
Collective entrainment is a mathematical term charaterizing the correlations between moving grains due to turbulence, collisions, or granular avalanche effects.

The \citet{Ancey2008} model has been generalized and applied in a handful of followup works. 
\citet{Turowski2009} generalized it to include limited sediment availability. 
\citet{Heyman2013} and \citet{Ma2014} studied how the \citet{Ancey2008} model could be interpreted to describe the statistics of the bedload flux at a fixed point in space, rather than just a control volume like that considered by \citet{Einstein1950}. 
Control volumes are less desirable because transport is practically easier to measure at a point in space than it is within a control volume.

\citet{Ancey2014, Ancey2015} generalized the \citet{Ancey2008} model in order to understand spatial characteristics of bedload transport. 
They considered a reach of channel as an array of adjacent control volumes (cells), and they coupled the \citet{Ancey2008} model between adjacent cells. 
They mapped this generalized birth-death model for the bedload flux distributed over an array of control volumes onto an advection diffusion equation for the bedload flux. 
Some reprecussions of this model were further explored in context of experimental data by \citet{Heyman2014}.

All of these works contain many simplifying assumptions, and many of these were carefully highlighted by their authors. 
These studies were selected for this review because, taken together, they define a research trajectory which can be extrapolated to suggest future research topics. 
To be sure, there is a long road ahead before Markov models can accomodate all of the complexities of natural river channels, some of which I've mentioned. 
The goal of this review is to indicate this trajectory in order to motivate future research: this is pursued in sections \ref{sec:background} and \ref{sec:review}. 
I discuss some of the future research topics I believe the trajectory points out in section \ref{sec:scope}, and I develop a few of these ideas myself in \ref{sec:extension}. 
Along the way, I have somewhat changed the mathematical treatment and the notation from some of the original papers in order to highlight the historical progression between each work, and to illustrate that these works all share the same theme.

My own extensions concern: (1) birth-death modeling of the fractional transport rates of a sediment mixture -- earlier works concentrate on a single grain size \citep{Ancey2006,Ancey2008,Turowski2009}; (2) the bedload flux of a single grain size when bed elevations vary, modifying entrainment and deposition characteristics -- earlier works concentrate on a flat bed with no bedforms \citep{Ancey2006,Ancey2008,Turowski2009}; and (3) a model for bedload flux including two-dimensional diffusion -- earlier works concetrate on one dimensional diffusion \citep{Ancey2014,Ancey2015,Heyman2014}. 
Once these extensions are presented, in section \ref{sec:conclusion} I recap the literature review, my identifications of underlying assumptions and future research directions, and my own extended Markov birth-death models of the bedload flux.

### 2021-06-24 formerPapers/ens-rev/reviewsections/ancey2006.tex
got one chapter in. some issues.

\subsection{Formalizing Einstein: a foundation in Markov process theory} 
\label{sec:ancey2006}

A significant theoretical development was made in stochastic modeling of the bedload flux by \citet{Ancey2006}. 
They reinterpreted the concepts of \citet{Einstein1950} to develop a new model of the bedload flux expressed as a Markov birth-death processes. 
The motivation for the \citet{Ancey2006} work was a series of experiments: \citet{Bohm2004} measured the bedload flux of glass beads at the outlet of a narrow flume using a percussion sensor, while \citet{Ancey2006} gathered a higher resolution dataset by filming particle motion with a video camera through the plexiglass sidewall of a narrow flume. 
They obtained bedload flux and motion characteristics using image processing. 
Conditions were highly controlled: uniform 6mm diameter glass beads were used as sediment. The flume was 6.5mm wide, so transport was quasi 1D. 
It only occurred in the downstream direction, and experiments were conducted under steady state conditions when the rate of sediment input was matched by the yield at the outlet. 
Despite this idealized level of control, measured fluxes had large fluctuations. 
Instantaneous values were as much as four times mean values.

The prevalence of large bedload fluctuations even under these idealized laboratory conditions motivated \citet{Ancey2006} to revisit Einstein's assumptions to develop a model of the bedload flux as a random variable. 
The assumptions underlying their model were: 
\begin{enumerate}
\item The sediment system can be characterized using a control volume with a size much larger than the scale of a single grain, similar to \citet{Einstein1950}. A series of volumes as depicted in \ref{fig:yalin} is not necessary.
\item The motion of each grain is independent of every other, and can be characterized as a random succession of motion and rest intervals. 
\item Transitions between motion and rest states are characterized by probabilities per unit time, or rates. 
\item Choosing a particle at random from the system, the possible transitions of that particle are sufficiently determined by its present state, i.e., motion or rest, and knowledge of its complete history of motion/rest transitions does not improve knowledge of its future state. This is the Markov property. 
\end{enumerate}
These assumptions are enough to develop a birth-death model for the bedload flux. 
Within this model, bedload fluctuations are due to variations in the number of moving particles within the control volume. 
As we will see, \citet{Ancey2006} obtained a probability distribution for the bedload flux. 
The mean of this probability distribution is the Einstein-like formula of Yalin \ref{eq:yalinflux}.  
The variance of this distribution is an unambiguous determination of the expected magnitude of bedload fluctuations.

\subsubsection{Theoretical development: the number of moving particles within a control volume}

Following \citet{Ancey2006} we consider a control volume as in figure \ref{fig:anceygeometry}, and we assume that a particle's dynamics are independent of its history and of the dynamics of all other particles.  
The probability of entrainment or deposition of a particle at each instant only depends on whether a particle is in motion or at rest at that instant. 
Hence, the stochastic dynamics of each particle satisfy the Markov property: it is independent of the past dynamics and only contingent on the current state of the particle. 
Since there are no interactions between particles to take account of, their dynamics, which are random successions between motion and rest states, can be considered separately. 
At the end, bedload transport within the control volume can be understood as the net contributions from all individual particles within it.

Therefore, we consider an individual particle within the control volume. 
We denote the probability that this particle is in the rest state at some time $t$ by $\pi_0(t)$ and the probability that it is in the motion state by $\pi_1(t)$.
We consider that within a small time interval $\delta t$, if the particle is at rest, the probability for the particle to transition into motion (entrainment) is $\sigma^{-1}\delta t$. Similarly, if the particle is in motion, the probability that it transitions into rest (deposition) is $\tau^{-1}\delta t$. 
Hence $\sigma^{-1}$ is the rate of entrainment of an individual grain, and $\tau^{-1}$ is the rate of deposition. 
Since these rates have dimensions of inverse time, $\sigma$ can be interpreted as a timescale of rest, and $\tau$ can be interpreted as a timescale of motion.

\begin{wrapfigure}{l}{0.5\textwidth}
  \includegraphics[width=.98\linewidth]{./figures/anceygeometry.png}
  \caption{A control volume $V$ of downstream length $L$ is defined over a region of mobile bed. The number $n$ of particles in motion is considered a random variable due to entrainment and deposition processes. Particles in motion move with velocity $u_p$. \label{fig:anceygeometry}}
\end{wrapfigure}

Using these transition rates and the standard method of setting up a master equation for a birth-death Markov process \citep[e.g.]{Cox1965, Pielou1977, Gillespie1992} develops the following equations describing the flow of probability through time: 
\begin{align} 
\dot{\pi}_0(t) &= \tau^{-1} \pi_1(t) - \sigma^{-1} \pi_0(t) \label{eq:anceymaster2006}\\
\dot{\pi}_1(t) &= \sigma^{-1} \pi_0(t) - \tau^{-1} \pi_1(t).
\end{align}
This two state stochastic process, illustrated in figure \ref{fig:telegraph} is known as the random telegraph process \citep{Gillespie1992, Gardiner1983}, and it is often invoked as a pedagogical example in textbooks on stochastic processes.

\begin{wrapfigure}{r}{0.25\textwidth}
  \includegraphics[width=.98\linewidth]{./figures/telegraph.png}
  \caption{Each particle in the control volume switches between motion and rest states with transition rates $\sigma^{-1}$ and $\tau^{-1}$.\label{fig:telegraph}}
\end{wrapfigure}

These equations can be solved for the probability that a particle is at rest -- $\pi_0(t)$ -- or in motion -- $\pi_1(t)$ -- by elementary techniques. 
If the particle is initially at rest, so that $\pi_0(0)=1$ and $\pi_1(0)=0$, the solution is 
\begin{align} \label{eq:anc2006}
\pi_0(t) &=  e^{-t/t_c} + \frac{\tau}{\sigma + \tau}(1-e^{-t/t_c}),\\
\pi_1(t) &= \frac{\sigma}{\sigma + \tau}(1-e^{-t/t_c}).
\end{align}
Here $t_c = \sigma \tau/(\sigma + \tau)$ is a correlation timescale which is the geometric mean of the motion and rest timescales. 
We can see that, after a time $t\gg t_c$, when $e^{-t/t_c} \approx 0$, the trajectory of the particle is independent of (forgets) the initial condition and settles into a stationary state with no time dependence. 
In this steady state, the probability that the particle is in motion is 
\be \pi_1 = \frac{\tau}{\sigma+\tau} = \xi,\ee
while the probability that the particle is at rest is 
\be \pi_0 = \frac{\sigma}{\sigma+\tau} = 1-\xi.\ee

We can interpret $\xi$ as the fraction of time spent in the motion state by the particle on average. 
To see this, we consider the distribution of residence times in the motion state. 
Let $R(T_m)$ be the probability the particle has been in motion for time $T_m$.
Then $R(T_m+\delta t)$ is the probability the particle remains in motion for another period $\delta t$.
This is just the probability the particle was in motion at $T_m$, times the probability that the particle did not deposit in $\delta t$. 
In symbols, $ R(T_m + \delta t) = R(T_m)(1-\tau^{-1}\delta t). $ 
As $\delta t \rightarrow 0$, this develops the differential equation 
\be \frac{d}{dT_m}R(T_m) = -\frac{1}{\tau}R(T_m),\ee
and the solution is the exponential distribution: 
\be R(T_m) = \tau e^{-T_m/\tau}.\ee
Hence the mean time spent in motion is $\tau$. 
By a similar argument, the distribution of resting times is exponential with parameter $\sigma$, so that $\sigma$ is the mean time spent at rest. 
Accordingly $\xi = \tau/(\sigma+\tau)$ represents a fraction of time spent in motion by the particle. 
This is an important link: the parameters $\tau$ and $\sigma$ are observable quantities.
The rate paramters entering the \citet{Ancey2006} model reviewed in this section are summarized in table \ref{tab:symbols2006}. 
\begin{comment}
\begin{wraptable}{r}{0.5\linewidth}
\vspace{-10pt}
\caption{The parameters used in section \ref{sec:ancey2006}}\label{tab:symbols2006}
\begin{tabular}{p{3.3cm}p{3.3cm}}\\
\toprule  
Symbol & Meaning \\
\midrule
$\sigma$ & average time spent at rest \\  
$\mu$ & average time spent in motion\\  
$\xi $ & $\mu/(\sigma+\mu)$ \\ 
\bottomrule
\end{tabular}
\end{wraptable} 
\end{comment}

We have followed \citet{Ancey2006} to a description of the random start-stop motions of an individual particle. 
The probability that the particle is found in motion is $\xi$ and the probability that it is found at rest is $1-\xi$. 
Hence an observation of whether or not the particle is in motion is a Bernoulli trial with known probabilities, like flipping a coin.
These two probabilities are ratios of observable quantities.

Now in order to get at the bedload flux, consider that on the bed surface within the control volume there are $N$ particles at any time, each switching randomly from motion to rest as an equilibrium telegraph process. 
We ask, of the $N$ particles, what is the probability that $n$ of them are in motion, $P(n)$?
The sum of $N$ Bernoulli trials is a binomial distribution \citep{Feller1968}, so the probability is
\be P(n) = {N \choose n} \xi^{N-n} (1-\xi)^n \sim \text{Bi}(\xi N, \xi(1-\xi)N). \label{eq:anceybinomial}\ee
This is just the probability that $N-n$ particles are at rest with probability $1-\xi$ while $n$ are in motion with probability $\xi$. 
The factor ${N \choose n}$ represents the number of ways to select the $n$ moving particles from the $N$ total particles.
$\text{Bi}(a,b)$ is a shorthand for a binomial distribution of mean $a$ and variance $b$. 
Hence \citet{Ancey2006} developed a probability distribution for the number of particles in motion in a control volume using the conceptual picture of \citet{Einstein1950}.

Now we outline two important limits of the binomial particle activity distribution as the number $N$ of particles available for motion in the control volume becomes very large.
The first limit is to consider that as $N$ becomes large, the probability $\xi$ that any of these $N$ particles is observed in motion remains fixed. 
In this case, the binomial distribution, equation \ref{eq:anceybinomial}, tends to a Gaussian distribution: 
\be P(n) \sim \mathcal{N}[N\xi, N\xi(1-\xi)]. \label{eq:normaldistflux} \ee
This follows from applying Stirling's formula $\ln(N!) \approx N\ln(N)-N$. 
The second way is to consider that as $N$ becomes large, the probability $\xi$ that any one of these $N$ particles is observed in motion decreases accordingly to leave the product $N\xi \coloneqq \lambda$ fixed. 
Under this limit, the binomial distribution tends to a Poisson distribution: 
\be P(n) \sim \frac{\lambda^n}{n!}e^{-\lambda}. \label{eq:ancey2006poissonlimit}\ee
This follows from taking only the highest order of $N$ from the binomial coefficients, and using one of the definitions of the exponential function: $\lim_{N\rightarrow \infty}(1-\frac{\lambda}{N})^{N} = e^{-\lambda}.$
We will use these limits to evaluate the bedload flux from the particle activity distribution derived in the \citet{Ancey2006} model, and to make contact between this early birth-death model of the bedload flux, and the more sophisticated models which follow it.

\subsubsection{From the number of moving particles to the bedload flux}
\label{sec:fluxdef}
Usually, the bedload flux is defined as the number of moving particles crossing a surface $S$ perpendicular to the flow direction: $q_s = \int_S \textbf{u}_p \cdot d\textbf{S}$, where $d\textbf{S}$ is an area increment with direction perpendicular to surface $S$ \citep{Ballio2014}.  
The surface $S$ used within the definition of the flux does not have an obvious relationship to the control volume in which we considered the random number of particles in motion. 
We need to connect the properties of bedload motion on a surface to the number of active particles within a volume.

The surface integral definition of flux just stated is often used in continuum field theories such as classical electrodynamics and kinetic theory, but it is not well suited to discrete particles. 
\citet{Ancey2006} and \citet{Furbish2012} have advocated an alternative formulation based on ensemble averaging. 
This alternative definition hinges on the probability $P[\textbf{u}_p | \textbf{x},t]$ that a particle contacts the control surface $S$ at position $\textbf{x}$ and time $t$ with velocity $\textbf{u}_p$. 
This conditional probability is considered to result from a very large collection of identical systems selected at random moments in their evolution: that is, the conditional probability is an ensemble quantity \citep{Kittel1958}.

In terms of this ensemble probability, the flux (per unit width) is \citep{Ancey2006}: 
\be q_s = \frac{1}{w}\int_{\text{all } \textbf{u}_p} \int_S P[\textbf{u}_p | \textbf{x},t] \textbf{u}_p \cdot d \textbf{S} d \textbf{u}_p . \label{eq:anceyflux} \ee
I have slightly modified this equation with the factor of $w^{-1}$. 
In steady conditions, when the probability is independent of time -- $\partial P / \partial t = 0$, the ensemble definition of the flux \ref{eq:anceyflux} can be related to the number of moving particles in the control volume we derived in the previous section. 
This link is only approximate, which I believe an under-emphasized point in the literature.

The approximation is developed by swapping the ensemble interpretation of $P$ by a frequentist interpretation.
This is done by integrating along the control volume.  
If the control volume is sufficiently long, $L \rightarrow \infty$, it can be seen as a stack of very many independent cross sectional surfaces. 
These comprise a stack of replicas of $S$, and at an instant, each surface constitutes one realization (particles intersecting $S$ with some set of positions and velocities) contributing to $P$. 
We can define $P$ by counting occurences along this stack of replica surfaces. 
This becomes an integral across the control volume.

With this concept of replica surfaces in mind, we can formalize the link with symbols. 
The $n$ particles of radius $a$ will be distributed throughout the volume at some set of positions $\textbf{x}_i$ with some set of velocities $\textbf{u}_i$, where $i=0,1,\dots, n$.
Thus the cloud of particles within the control volume at some instant can be represented by its density in a position-velocity phase space: 
\be \rho(\textbf{x},\textbf{u}_p) = \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i). \ee  
Here $M_a(\textbf{z})$ is a marker function with definition 
\be 
M_a(\textbf{z}) = 
\begin{cases}
1 & \text{if } |\textbf{z}]< a \\
0 & \text{otherwise}
\end{cases}. 
\ee 
It is $1$ within a sphere of radius $a$ around its argument, and zero otherwise. 
Its volume integral is $\int M_a(\textbf{x}) dV = 4\pi a^3/3 = \nu_p$, the particle volume. 
There are other applications of such functions in \citet{Furbish2012} and \citet{Ballio2014}.

Using this density, for sufficiently large $L$, we can write the conditional probability $P[\textbf{u}_p | \textbf{x}]$ approximately as an integral over the stack of replica surfaces.   
This is an exchange of the ensemble interpretation of $P$ for a frequentist interpretation: 
\be P[\textbf{u}_p|\textbf{x}] \approx \frac{1}{L} \int_0^L dx \rho(\textbf{x},\textbf{u}_p). \ee
The integral runs along the downstream coordinate.
This relationship is only exact as $L \rightarrow \infty$.
Presumably, for fixed $L$, the quality of this approximation would scale with spatial correlations among bedload positions and velocities.
The more correlated bedload positions and velocities are, the worse the approximation will become.

Using this swap, the bedload flux becomes 
\be q_s \approx \frac{1}{w}\int_{\text{all } \textbf{u}_p} \int_S \frac{1}{L} \int_0^L dx \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i) \textbf{u}_p \cdot \textbf{k} dS d \textbf{u}_p  = \frac{\nu_p}{Lw} \sum_{i=1}^n \textbf{u}_i\cdot \textbf{k} \ee
It is expressed in terms of the number of particles $n$ within the control volume. $n$ is the random variable we considered earlier as the result of $N$ telegraph processes.
If we write the downstream component of each particle's velocity as $\textbf{u}_i \cdot \textbf{k} = u_{i}$, the bedload flux takes the simple looking form 
\be q_s \approx \frac{\nu_p}{Lw}\sum_{i=1}^n u_i \label{eq:fluxsum} \ee
in terms of the number of particles $n$ within the (necessarily large) control volume and their downstream velocities $u_i$, where $i=1,2,\dots,n$. 
This equation becomes exact as $L\rightarrow \infty$.

There are options for further analysis of the flux in this way:  
\begin{enumerate}
\item  We can consider the downstream velocities of particles as random variables governed by a probability distribution \citep{Roseberry2012, Furbish2016}, in which case the bedload flux is a sum of a random number $n$ of random variables $u_i$. 
In stochastic processes parlance, a sum of a random number of random variables is called a compound stochastic process \citep{Feller1968}. 
Treating $q_s$ as a compound process has not been fully explored, to my knowledge.

\item We can consider that bedload particles, once set in motion, will all move with approximately the same average velocity, so that the sum \ref{eq:fluxsum} simplifies.
This idea has some theoretical support \citep[e.g.][]{Ancey2003}, although it conflicts with experimental findings \citep{Heyman2016}, so it should be considered critically.  
This is the approach taken by \citet{Ancey2006}.  
\end{enumerate}

For now, we will follow \citet{Ancey2006} to consider that all particles move at approximately the same downstream particle velocity $u_s$, so that the bedload flux becomes 
\be q_s(n) \approx \frac{\nu_p u_s}{Lw} n. \label{eq:fluxy}\ee
The flux is directly proportional to the number $n$ of moving particles within the control volume. 
The bedload flux is a random quantity which inherits its statistical properties from $n$.

\subsubsection{The bedload flux as a random variable: Einstein revisited, Yalin an intermediary}

Now we develop the probability distribution of the bedload flux by combining the results we have obtained so far: (1) the probability distribution of the number of moving particles within a control volume equation \ref{eq:anceybinomial}, and (2) the relationship between the number of moving particles and the bedload flux equation \ref{eq:fluxy}. 
Using the law for transformation of discrete random variables \citep[e.g.][]{Feller1968},
\be P(q_s) \approx \sum_{n=0}^N P(n) \delta_{n,Lw q_s/(\nu_p u_s)} ,\label{eq:above}\ee
but this is not valid. 
Since the bedload flux is only approximately given by \ref{eq:fluxy} there is no guarantee that $q_s Lw /(\nu_p u_s)$ will be an integer, and this transform cannot be carried through.

Instead, we can take advantage of the large $L$ assumption embedded into the equation \ref{eq:fluxy}. 
Because $L$ is necessarily large and $N \propto L$, it follows that $N$ is also large when \ref{eq:fluxy} is valid. 
Therefore, we can also take a limit of the binomial distribution $P(n)$ as the number $N$ of particles in the control volume becomes very large, while the probability that any individual particle is observed in motion remains fixed. 
We can use the resulting distribution, equation \ref{eq:normaldistflux}, with \ref{eq:fluxy}, to derive the bedload flux distribution without the mathematical ambiguity of \ref{eq:above}. 
The probability distribution for the bedload flux becomes, using the law for transformations of continuous random variables \citep[e.g.][]{Feller1968} with \ref{eq:fluxy} and \ref{eq:normaldistflux}:
\be P(q_s) \approx \int_0^\infty dn \delta(n - L w q_s/(\nu_p u_s)) P(n) \approx \mathcal{N}\big(N\xi [\nu_p u_s/(Lw)], N\xi(1-\xi)[\nu_p u_s/(Lw)]^2\big) \label{eq:anceygauss}\ee
\citet{Ancey2006} obtain a normal distribution for the bedload flux.

Now we discuss the connection of the \citet{Ancey2006} formalism to the earlier work of \citet{Einstein1950}. 
From equation \ref{eq:anceygauss}, the mean flux is 
\be \bra q_s \ket = \nu_p u_s \xi \frac{N}{Lw}.\ee
According to \citet{Einstein1950}, the ratio $N/(Lw)$ is related to the particle size via $N/(Lw) = 1/ a^2$. Since the particle volume is $\nu_p = 4\pi a^3/3$, the mean flux becomes 
\be \bra q_s \ket = \frac{4}{3}\pi a u_s \xi. \ee
Einstein famously considered that the average jump distance of a particle within a single motion interval was proportional to the particle size \citep{Einstein1950, Yalin1972}. 
If we denote the jump distance as $l$ to distinguish it from the control volume size $L$, this means $a \propto l$.
Similarly, Einstein held that the mean transport rate scaled with an inverse characteristic timescale.  
Originally, it was expressed in terms of the particle size and the settling velocity \citep{Einstein1950}. 
Later on, \citet{Yalin1972} concluded it was a timescale of turbulent fluctuations, and there have been similar interpretations by other investigators \citep{Paintal1971, Cheng2004, Armanini2015}. 
Apparently, in this case, to make the link from \citet{Ancey2006} to \citet{Einstein1950}, the timescale should be 
$\tau = k a/u_s$, where $k$ is a constant of proportionality.  
With these substitutions, the bedload flux becomes 
\be \bra q_s \ket = A \frac{a l}{\tau} \xi ,\ee
where $A$ is a constant of proportionality. 
This is similar to Yalin's reinterpretation of Einstein, equation \ref{eq:yalinflux}. 
In the \citet{Ancey2006} model, $\xi$, the fraction of time spent in motion by a single particle, plays the role of $p$, which was the probability of (only) one entrainment in a unit of time.  
In contrast to the $p$ and $\tau$ of \citet{Einstein1950}, the parameters of this model, $\xi$ and $\tau$, are measurable quantities with a clear physical basis. $\xi$ is the fraction of time spent in motion by an individual particle. $\tau$ is the ratio of particle size and the average downstream velocity of sediment in motion.  
We can conclude that \citet{Ancey2006} has reframed and extended Einstein's theory using stochastic mathematics, treating the bedload flux as a statistical quantity. 
The parameters of the refined model have a sound physical interpretation.

Now we will examine the magnitude of bedload fluctuations expected from the model. 
We can examine the variance of the bedload flux, and compare this to the mean of the flux, to measure the strength of fluctuations. 
However, a clearer conclusion emerges from comparing the mean and variance of the particle activity, derived from the distributions \ref{eq:anceybinomial} or either of its large $N$ limits. 
The ratio of the variance to the mean is 
\be \frac{\bra \delta n^2 \ket}{\bra n \ket} = 1 - \xi. \label{eq:2006flucts} \ee
We conclude that according to the \citet{Ancey2006} model, because $\xi>0$, the variance in the number of active particles within the observation window should never exceed the mean.

All of the experiments of \citet{Ancey2006} violated this condition, undermining the model, and leading us to examine its assumptions.  
\citet{Ancey2006} hypothesized that the inability of the birth-death model they developed to describe experimental fluctuations was due to the lack of interaction between particles as they transition between motion and rest states. 
They noted that particles preferentially moved in groups, and that the entrainment of one particle led to subsequent entrainment of neighboring particles which were contingent on it for stability. 
Therefore they extended their 2006 model to include a collective motion effect in \citet{Ancey2008}. 
As it turns out, this inclusion leads to a negative binomial distribution of the bedload flux.
This distribution has a sufficiently heavy tail to describe experimental bedload fluctuations. 
We focus on the \citet{Ancey2008} model and the collective entrainment process it introduces in the next section.

### 2021-06-24 formerPapers/ens-rev/reviewsections/ancey2014.tex
got one chapter in. some issues.

\subsection{The diffusion of bedload from a microstructural basis}

\citet{Ancey2008} developed a generalized version of \citet{Einstein1950} which took account of a new process to describe the large fluctuations seen in experimental bedload flux signals: collective entrainment. 
\citet{Turowski2009} generalized this model to finite sediment availability, while \citet{Heyman2013} and \citet{Ma2014} used the \citet{Ancey2008} model in its original form to understand the distribution of waiting times between emigration events and the bedload flux probability distribution at a point in space, respectively. 
These two works developed new understanding of spatial and temporal aspects of bedload flux: \citet{Heyman2013} demonstrated that the waiting time between successive emigration events expressed contributions from slow and fast timescales related to individual and collective entrainment, respectively. 
\citet{Ma2014} highlighted the role of observation timescale on the magnitude of flutuations, and discriminated three distinct scaling regimes for bedload fluctuations which are contingent on a dimensionless number $Ra(\delta t)$, where $\delta t$ is the timescale of observation.

However, up to this point in early 2014, stochastic birth-death models had focused on a single region of space -- a control volume \citep{Einstein1950, Ancey2006, Ancey2008}, or a single point in space -- the downstream end of the control volume, or the plane across which emigration events happen \citep{Heyman2013, Ma2014b}.
This approach is obviously oversimplified. Variability of sediment transport rates is almost a defining feature of them \citep{Hassan2008, Venditti2016, Nelson2014}. 
There is a need to take account of spatial differences in entrainment and deposition characteristics in calculating bedload fluxes. 
Additionally, there are a wide set of issues centered around bedload diffusion-- or the spreading of bedload particles as they are tracked through time and subjected to random entrainment and deposition events \citep{Hassan2016}. Indeed, this study of bedload diffusion was the original research directive of \citet{Einstein1937}.

Often, bedload diffusion has been understood through deterministic partial differential equation models. These advection diffusion equations have been derived by considering conservation of mass \citep{}, but they were not very well connected to the underlying stochastic dynamics of bedload transport. 
Generating this connection between a microstructural stochastic model like \citet{Ancey2008} and the continuum diffusion equation treatment of bedload diffusion \citep[e.g.][]{Parker2002} was the directive of the next papers on birth-death modeling \citep{Ancey2014,Ancey2015}. 
These papers took a conceptually straightforward extension of the \citet{Ancey2008} model which leads to difficult mathematics.

Rather than considering a single control volume, as in \citet{Ancey2008}, \citet{Ancey2014,Ancey2015} considered an infinite array of adjacent control volumes (cells), indexed by $i=1,2,\dots,M$. 
Bedload particles can entrain and deposit within each cell, and they can also migrate between cells. 
Each of these transitions are characterized by probabilities per unit time in generalization of \citet{Ancey2008} across spatial extent. 
The $i$th cell has a random number of particles $N_i$ in motion within it. 
Therefore the state of the system at an instant of time is fully characterized by the set of these numbers: $\textbf{n} = (n_1(t), n_2(t), \dots n_M(t))$.

One target is the grand probability distribution of the number of active particles within each cell at an arbitrary time: $P(\textbf{n};t)$. 
Another target is the continuum limit: letting the length $\Delta x$ of each cell shrink to zero develops an advection-diffusion equation for bedload diffusion. 
Ancey et al termed this a "Stochastic interpretation of bedload diffusion": and the latter target has much more fundamental scope.
It explores the link between microscale stochastic models of bedload transport and mesoscale deterministic models based upon advection-diffusion equations.

Now I'll sketch the \citet{Ancey2014, Ancey2015} derivation of a birth-death-migration model across an array of cells. 
There are essentially three transitions to take account of: 
\begin{enumerate}
\item Entrainment can occur within each cell ($i$) at probability per unit time $(\lambda_i + \mu_i n_i ) \delta t$
\item Deposition can occur within each cell ($i$) at probability per unit time $ \sigma_i n_i \delta t$
\item Migration can occur from cell $i$ to cell $i+1$ (downstream) at probability per unit time $\nu_i$
\end{enumerate} 
Now some notation is introduced to write the effect of these transitions on $\textbf{n}$ in shorthand. 
If the change of state due to one of these transitions is written as $\Delta \textbf{n}$, then the effect of these transitions can be characterized using vectors $\textbf{r}_i^j$ and $\textbf{r}_i^\pm$. The vectors $ \textbf{r}_i^j$ are of the same dimension as $\textbf{n}$ and all but two of their entries are zero: $r_i = 1$, $r_j = -1$, $r_k=0$ for $k\neq i,j$. The vectors $\textbf{r}_i^\pm$ are the same dimension as $\textbf{n}$ and all but their $i$th entry is zero: $r_i = \pm 1$, $r_j = 0$, $j\neq i$.

Thus the transition probabilities of each process can be written: 
\begin{enumerate}
\item $ p_i^3 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^+; t+ \delta t) = (\lambda_i + \mu_i N_i) \delta t $ -- entrainment
\item $ p_i^2 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^-; t+\delta t) = \sigma_i N_i \delta t $ -- deposition
\item $ p_i^1 = \text{Prob}(\textbf{N} = \textbf{n} \textbf{n} + \textbf{r}_i^{j-1}; t+ \delta t) = \nu_{i-1}N_{i-1} \delta t $ -- migration
\end{enumerate}

These transition probabilities develop a master equation analogous to the one from \citet{Ancey2008}, but generalized to a collection of $M$ cells: 
\begin{multline} \frac{\partial}{\partial t} P(\textbf{n};t) = \sum_{i=1}^M (n_i+1)[P(\textbf{n}+\textbf{r}_{i+1}^j,t) \nu_i + P(\textbf{n}+\textbf{r}_i^+,t)\sigma_i ]\\
+ P(\textbf{n}+\textbf{r}_i^-,t)(\lambda_i + \mu_i(n_i-1))\\
+ P(\textbf{n}+\textbf{r}_i^{j-1},t)\nu_{i-1}n_{i-1}\\
- P(\textbf{n},t)(\nu_{i-1} n_{i-1}+\lambda_i + \mu_i n_{i+1} + \nu_i n_i + \sigma_i n_i ) \end{multline} 
Information about solving this equation is sparse in the literature. 
The closest approach has been in ecology: these birth/death/migration-type stochastic models have been considered in context of ecological population dynamics since at least the sixties \citep{Bailey1968}. 
One exact solution of a very similar stochastic model has been attained using path integral approaches borrowed from quantum field theory \citep{Field2010}.
By any metric, the analytical barrier this multidimensional master equation presents is not trivial, although definitely numerical solutions are possible \citep[e.g.][]{Gillespie1992}.

In order to link this equation to spatial diffusion, \citet{Ancey2014, Ancey2015} resorted to a Poisson space representation \citep{Gardiner1983}. 
In effect, the Poisson transformation converts the discrete random variable $\textbf{N}$ to a continuous random variable $\textbf{a}$ (the Poisson rate). 
The Poisson transform is 
\be P(\textbf{n},t) = \prod_i\int \frac{e^{-a_i}a_i^n}{n!}f(\textbf{a},t)d\textbf{a}, \ee
where $\textbf{a} = (a_i) \geq 0$ for $i=1,2,\dots$ and $f(\textbf{a},t)$ is the multivariable probability density of the continuous vector $\textbf{a}$.

\citet{Ancey2014, Ancey2015} used this transformation to obtain a Langevin equation representation of the dynamics of the random variable $\textbf{a}$.
That is, they obtained a stochastic differential equation representation of $\textbf{a}$: 
\be da_i(t) = (\lambda_i - a_i(\sigma_i-\mu_i) + \nu_{i-1} a_{i-1} - \nu_i a_i)dt + \sqrt{2 \mu_i a_i}dW_i(t),\ee
and the $W_i(t)$ is a white noise term (Wiener Process) on cell $i$. 
The mean steady state solution of this Langevin equation is 
\be \bra a \ket_ss = \frac{\lambda}{\sigma-\mu} \ee 
for all $i$.

Now \citet{Ancey2014,Ancey2015} sought out to find an advection diffusion equation as the limit of cell size $\Delta x \rightarrow 0$, so they went on to impose this limit on the Langevin equation, thereby obtained an advection diffusion equation. 
Unfortunately, the limit is more subtle than expected. 
This did not really work for reasons I need to understand better, and it did not work in ways I need to learn to explain better. 
More study is needed on spatially varying birth-death models, from me, and from others especially.

### 2021-06-24 formerPapers/ens-rev/reviewsections/heyman2013.tex
got one chapter in. some issues.

\subsection{The waiting time between emigration events: Bedload flux from waiting time statistics}

These studies are concerned with linking the birth-death formulation, which in the original interpretation deals with transport characteristics within an observation window \citep{Ancey2008}, to the transport characteristics at a point -- the number of particles crossing a plane perpendicular to the flow direction.

\citet{Heyman2013} was concerned with the statistics of the time interval between successive emigration events, while \citet{Ma2014b} focused on actually counting the number of emigration events in a unit time: that is, they computed the bedload flux over the downstream boundary of the control volume \citet{Ancey2008}.

The waiting time between successive transport events is significant in a broad range of transport studies. 
First, there are probabilistic transport rate formulations which use the waiting time distribution as their input \citep{Turowski2010}. 
Second, the waiting time between successive transport events, as we shall show, reflects back on the underlying entrainment and deposition characteristics, so it provides fundamental information into the somewhat mysterious physics of bedload transport.

The original \citet{Einstein1937,Einstein1950} approach reveals that the waiting time between successive transport events should be an exponentially distributed random variable. 
However, Heyman et al (2013) demonstrates that the inclusion of collective entrainment due to \citet{Ancey2008} modifies the waiting time distribution in a non-trivial way. 
The waiting time between successive entrainment events has contributions at two different scales: first, there is a shorter timescale of individual entrainment events, controlled by fluid turbulence and essentially recognized within the \citet{Einstein1937} formalism, and second, there is a longer timescale due to collective entrainment.
These authors showed that within the \citet{Ancey2008} birth-death model, at low transport rates, when entrainment and motion are highly intermittent, the short timescale and the long timescale are well separated. 
While at higher transport rates, these two timescales blend together as individual and collective entrainment contributions become indistinguishable.

I'll now sketch the \citet{Heyman2013} derivation of the statistics of the waiting time between successive emigration events, and show how the two different timescales due to individual and collective entrainment arise. 
For \citet{Heyman2013} the variable of interest is $S_k = \sum_{i=0}^k T-i$: $S_k$ is the time when the $k$th emigration event occurs. $T_1, T_2, \dots $ are the waiting times: the time from $t=0$ to the first emigration event is $T_1$. The time from $T_1$ to the second emigration event is $T_2$, and so on. 
For a Markov process such as the \citet{Ancey2008} model, the waiting times are independant and identically distributed. 
Within the Einstein model we have already seen that the waiting time probability distribution is expoential. 
However, in the \citet{Ancey2008} formalism there is no reason to expect an exponential distribution.

Heyman et al define $F_n(t)$ as the probability that there are $n$ particles in motion and no emigration event occurred in time t: 
\be F_n(t) = Pr(T>t,N(t)=n).\ee
By extension, the probability that there are any number of particles and no emigration event occurred in time $t$ is 
\be F(t) = Pr(T>t) = \sum_{n=0}^\infty F_n(t) .\ee

Now $F(t+\Delta t)$ is equivalent to the probability that any other event but emigration occurs in $\Delta t$. 
Therefore we can write a master equation for $F(t)$: 
\be F_n(t+\Delta t) = F_n(t)[1-(\lambda + n(\sigma+\mu+\gamma))\Delta t] + F_{n+1}(t) \sigma(n+1)\Delta t + F_{n-1}(t)\lambda + \mu(n-1)\Delta t + o(\Delta t).\ee
This equation holds for $n\geq 1$. At $n=0$ deposition and collective entrainment processes are not possible, therefore 
\be F_0(t+\Delta t) = F_0(t) [ 1-\lambda \Delta t] + F_1(t) \sigma \Delta t + o(\Delta t) \ee 
Again dividing by $\Delta t$ and taking $\Delta t \rightarrow 0$ gives the differential difference equations
\begin{align} 
F_0'(t) &= \lambda F_0(t) + \sigma F_1(t) \\
F_n'(t) &= -(\lambda + n(\sigma + \mu + \gamma))F_n(t) + \sigma(n+1)F_{n+1}(t) + [\lambda + \mu(n-1)]F_{n-1}(t) \text{ if } n\geq 1 \end{align} 
for the waiting time distribution.

Summing all of the terms gives the simple equation 
\be \sum_{n=0}^\infty F_n'(t) = -\gamma \sum_{n=0}^\infty nF_n(t), \ee
which, denoting the pdf of waiting times $T$ as $f_T(t)$ gives the simple relationship: 
\be f_T(t) = -F'(t) = \gamma\bra F_n(t)\ket. \ee
Therefore the probability distribution function of waiting times between emigration events, $f_T(t)$, is an average of the probability that there are $n$ particles and no emigration event occurred in $t$ over all $n$.

The general solution of $F_n(t)$ can be obtained from the differential equations with a generating function: 
\be G(z,t) = \sum_{n=0}^\infty F_n(t) z^n. \ee
Multipling the differential equation by $z^n$ and summing over all $n$ gives
\be \frac{\partial G}{\partial t} = (\sigma + \mu z^2 - z(\alpha + \mu) ) \frac{\partial G}{\partial t} + (z-1)\lambda G, \ee
where the shorthand $\alpha = \gamma + \sigma $ has been introduced.

This equation can be solved with the method of characteristics, giving a closed form solution for $G(z,t)$. A useful property of generating functions (and the source of their name) is $\bra F_n \ket = \partial G(z,t) /\partial t |_{z=1}$, therefore the probability distribution of waiting times between emigration events in the \citet{Ancey2008} birth-death model of bedload flux is 
\be f_T(t) = \gamma(z_1-z_2)^{\lambda/\mu} \big( \frac{\alpha-\mu}{A(t)-B(t)} \big)^{\lambda/\mu + 1} e^{-\lambda (1-z_2)t}\big( \frac{\lambda/\mu+1)B(t)}{A(t)-B(t)} [(1-z_2)e^{-(z_1-z_2)t}+z_1-1]+e^{-\mu(z_1-z_2)t}-1 \big)  \ee 
When collective entrainment is turned off ($\mu=0$), the probability distribution becomes $f_T(t) = \frac{\alpha}{\lambda \gamma} e^{-\alpha t/(\lambda \gamma)}$: the exponential waiting time between emigration events is recovered.

This probability distribution for waiting times between emigration events in the \citet{Ancey2008} model has, in general, contributions from two timescales. The relatively common individual entrainment process imparts a 'fast' timescale to the probability distribution of emigration waiting time, while the relatively rare collective entrainment process imparts a 'slower' timescale. 
When collective entrainment is turned off, only the fast timescale is present in the pdf. When the entrainment rate is relatively low, the fast timescale is much shorter than the slow timescale, so that the timescales are well separated and the waiting time distribution is not well approximated by any common distribution in the literature (Gamma, exponential): the behavior of the \citet{Ancey2008} model at low transport rates has defied prediction by any other framework.

The most significant point of the \citet{Heyman2013} paper is that the probability distribution function of waiting times between emigration events can be used to calibrate the \citet{Ancey2008} birth death model to flume experiments. 
This is because short emigration times control $\alpha-\mu$ and long emigration times control $\lambda$. 
The evaluation of waiting times between emigration events supports a calibration of the stochastic birth-death model of \citet{Ancey2008} on flume studies where the transport rate is measured at the outlet of the flume, but the erosion, deposition, or emigration rates are not necessarily measured. 
In contrast to \citet{Ma2014b} the mathematics are not much beyond the original formulation of the \citet{Ancey2008} model, so this work may be more readily accepted by the research community.

### 2021-06-24 formerPapers/ens-rev/reviewsections/heyman2014.tex
got one chapter in. some issues.

\subsection{Spatial correlations in bedload transport: }

\citet{Heyman2014} used the \citet{Ancey2014} model to examine a set of experiments resolving spatial correlations of bedload transport. 
They found a stochastic differential equation for the mean of the poisson rate $\eta$: 
They related this to a point process framework which allowed them to analyze spatial correlations in the bedload rate and relate this to their birth-death model. 
Somehow this whole thing is contingent on their two-point spatial correlation process. 
They state that the whole problem is linear so that it'd be straightforward to extend it to 2d. 
slowing down ...

### 2021-06-24 formerPapers/ens-rev/reviewsections/ma2014.tex
got one chapter in. some issues.

\subsection{The bedload flux at a point: the counting statistics of emigration events}

Up to this point, the Markovian approaches to understand a fluctuating bedload rate were based upon analyzing the stochastic number of active particles within a control volume. 
The bedload flux definition considered for these studies was somewhat unconventional. 
Typically it is defined as the rate of bedload crossing a vertical plane perpendicular to the flow direction \citep{Ballio2014}.

The studies of \citet{Heyman2013} and \citet{Ma2014b} demonstrated that the birth-death model of \citet{Ancey2008} can also be used to define a flux at a position in space. 
Although the birth-death model of \citet{Ancey2008} counts the number of moving particles at each instant of time, it does not count the number of emigration events which contribute to this number. 
Counting the number of emigration events in an interval of time is a much more difficult problem.
Ma et al (2014) approached this counting problem using a path integral approach: this approach is essentially outlined in \citet{Ohkubo2009}. 
The path integral approach draws heavily from the stochastic simulation algorithm of \citet{Gillespie1977}.

The conventional definition of the bedload flux is an integral of particles through a surface perpendicular to the flow: 
\be Q_s(t) = \int \int_A dA \textbf{u}_s \cdot \textbf{n}, \ee
where $\textbf{n}$ is a unit vector perpendicular to the average flow direction and $\textbf{u}_s$ is the velocity of bed load. 
This instantaneous flux is very difficult to measure, so usually it is averaged over some time interval $\delta t$: 
\be Q_s(t;\delta t) = \frac{1}{\delta t} \int \int_A \int_t^{t+\delta t} \textbf{u}_s \cdot \textbf{n} d\tau dA = \frac{\delta V_s}{\delta t} = v_s \frac{\delta S(t;\delta t)}{\delta t}. \ee
$Q_s(t;\delta t)$ is the volume flux averaged over $\delta t$; $\delta V_s$ is the volume of bed load particles passing through $A$ within $\delta t$; $S(t)$ is the cumulative number of particles passing through the cross section since t=0, and $\delta S(t;dt)$ is the number of particles passing through the surface between $t$ and $t+\delta t$, i.e. $\delta S(t; \delta t) = S(t+\delta t) - S(t)$. 
This equation means the bedload flux at $t$, when viewed in a time-averaged sense, is the number of particles crossing the surface between $t$ and $t+\delta t$. 
Because volume units can always be multiplied in later, the bedload flux is written 
\be q_s(t) = \frac{\delta S(t; \delta t}{\delta t}. \ee

This bedload flux is a random variable.
A fluctuation is defined strictly as $q_s' = q_s - \bra q_s \ket$, and the magnitude of fluctuations is defined without ambiguity by the variance $\bra q_s'^2 \ket$. 
This $\delta S(t;dt)/\delta t$ is the number of emigration events between $t$ and $t+\delta t$, so if this number could be counted, the bedload flux would be known at the downstream boundary of the control volume: the bedload flux would be known at a point in space.

The path integral formulation performs this counting problem. 
For the sake of computation, the time interval t is discretized as $i \Delta t = t$, $i=0,1,2,\dots$.
In the end, the limit $\Delta t \rightarrow 0$ will be taken. 
If the increments $\Delta t$ are sufficiently small, Ma et al argued that emigration events during one of these intervals will be approximately Poissonian so that
\be P[\Delta S_i = n | N_i(t) = N_i] = \exp[-f_3(N_i)\Delta t] \frac{(f_3(N_i) \Delta t)^N}{n!}. \ee
This equation expresses the probability that $n$ particles emigrate between $t$ and $t+\delta t$ is a Poisson distribution with rate parameter $f_3(N_i)\Delta t$, where $f_3(N_i) = \gamma N_i$ is the emigration rate.

Now Ma et al argued that because the entrainment rate $f_1 = \lambda + \mu N \~ O(L)$, the deposition rate $f_2 = \sigma N \~ O(L)$, and the emigration rate $f_3 = \gamma N \~ O(1)$, if the observation window is sufficiently large, entrainment and deposition processes dominate, so that the number of moving particles $N$ is perturbed only weakly by emigration. 
In this limit of sufficiently large observation window size, it then holds that
\be P(\sum \Delta S_i = n | N_1,N_2, \dots) = \exp[-\sum f_3(N_i)\Delta t]\frac{(f_3(N_i)\Delta t)^n}{n!}, \ee
which in the continuum limit becomes a path integral counting emigration events: 
\be P[\delta S = n |N(t): t_0<t<t_0+\delta t] = \exp[-\int f_3(N(\tau))d\tau]\frac{[\int f_3[N(\tau)]d\tau ]^n}{n!}.\ee
Denoting the integral over the process $f_3(N(\tau))$ as $\alpha = \int_{t_0}^{t_0+\delta t} d\tau f_3[N(\tau)]$, the probability for $n$ emigration events in $\delta t$ is written 
\be P(\delta S = n) = \int d\alpha \frac{e^{-\alpha}\alpha^n}{n!}P(\alpha,t). \ee
From this difficult derivation a simple formula emerges: the probability of $n$ emigration events in $\delta t$ is an integral over all possible realizations of the process $N(t)$, where each process is given a weight $e^{-\alpha}\alpha^n/n! P(\alpha,t) d\alpha$. This is called a Cox process. It's a generalized Poisson process where the parameter $\alpha$ is also a random variable. It represents a superposition of a series of Poisson distributions taking every possible path of $N(t)$ into account. 
This is a stochastic description of the bedload transport flux based on the counting statistics of emigration events.

Notably, because of the time integral of $N(t)$ involved in this relationship, it has a memory. It is non-Markovian. 
This character reveals a very interesting fact: the fluctuations of the transport rate depend on the time scale of observation. 
In fact, Ma et al derived the expression for the variance: 
\be \delta t \text{var} [q_s(\delta t)] = \gamma^2 \frac{\lambda (\sigma + \gamma)}{(\sigma + \gamma -\mu)^2} 2 t_c [\delta t - t_c (1-e^{-\delta t/t_c}] + \frac{\gamma\lambda}{\sigma + \gamma -\mu}\delta t \label{eq:3regimes} \ee

where $t_c = 1/(\sigma + \gamma - \mu)$ is the autocorrelation time of $N(t)$, representing memory in the system. 
Apparently, the variance of the bedload flux at a point depends on the observation or averaging time-scale $\delta t$.

In fact, Ma et al discriminated three regimes of behavior from \ref{eq:3regimes}:   
\be
    \delta t \text{var}[q_s(\delta t)] \approx 
\begin{cases}
    \frac{\gamma \lambda}{\sigma + \gamma - \mu}, & 0<\delta t \ll t_l \\
    \delta t \frac{\gamma^2 \lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}   & t_l<\delta t< t_c \\
    2 \gamma^2 \frac{\lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}t_c + \frac{\gamma \lambda}{\sigma + \gamma - \mu}   & t_c \ll \delta t < \infty
\end{cases}
\ee
When this three-regime equation is divided by the mean bedload rate $\text{mean}[q_s(\delta t)]$, the behavior is discriminated by a dimensionless number Ma et al called $Ra$: 
\be Ra(\delta t) \approx \begin{cases} 1 & 0<delta t \ll t_l \\
\frac{delta t}{t_l} & t_l < \delta t < t_c \\ 
\frac{2 t_c}{t_l} + 1 & t_c \ll \delta t < \infty \end{cases} \ee

The essential point is that Ma et al derived three regimes of fluctuations from the \citet{Ancey2008} model. 
Depending on the sampling interval $\delta t$, three different scaling relations are possible for the magnitude of fluctuations in the bedload flux. 
These intervals are discriminated by two timescales $t_c$ and $t_l$. 
There are probably deep consequences of this research which remain to be fully understood. 
At this stage, it's sufficient to note that fluctuations differ with observation timescale, and that this effect can be discriminated into stages with a dimensionless number $Ra$.

The origin of the two timescales still needs to be discussed.

There is another way to get information about the bedload flux at the downstream boundary of the control volume defined by \citet{Ancey2008}, and this was examined by \citet{Heyman2013, Heyman2016}. 
Rather than actually counting the number of emigration events with the complicated path integral approach, Heyman et al. took a somewhat simpler angle: they examined the waiting time between successive emigration events.

### 2021-06-24 formerPapers/ens-rev/scraps/classical.tex
got one chapter in. some issues.

\section{Foundational Markovian transport theories}

\subsection*{Einstein's 1937 Thesis: the dispersion of bed load grains}

Before 1937 Einstein added colored stones to an artificial stream channel and investigated their movement characteristics downstream \citep{Einstein1937}. 
He subsequently developed a probabilistic model describing the statistics of downstream particle travel distance with time. 
In this way he derived that particle travel distance with time should be a Poisson distribution. 
Einstein's introduction justifies his probabilistic approach: 
\begin{quote}
The (experimental) results demonstrated clearly that even under the same experimental conditions stones having essentially identical characteristics were transported widely varying distances ... (yet) similar curves were obtained when one plotted the frequency of particles as a function of the path. 
Consequently, it seemed reasonable to approach the subject of particle movement as a probability problem.  
\end{quote}

Einstein's model, developed for low transport rates, has three key assumptions: (1) sediment motion is an alternating series of rest and motion intervals; (2) rest and motion can be characterized by random variables governed by probability distributions; and (3) because the time spent in motion is small relative to the time spent at rest, motion can be considered instantaneous. 
Within this conceptual picture, particle motions trace out stair step trajectories through the space time plane. 
Some example trajectories are indicated in \ref{fig:einsteinTrajectories}.
Although each trajectory is different, they define a probability distribution when many are collected.

\begin{wrapfigure}{L}{0.5\textwidth}
  \includegraphics[width=.98\linewidth]{./figures/einsteintrajectories.png}
  \caption{The travel distance - time plane with three particle trajectories indicated as they were conceptualized by Einstein in terms of alternate start-stop motions. Each corner on the stair step trajectories is a transition between motion and rest states. The time spent at rest (resting time) and the distance traveled in motion (hop distance) are random variables drawn from probability distributions. These random motions coordinate so that motions are statistically described by a Poisson distribution.}
  \label{fig:einsteinTrajectories}
\end{wrapfigure}

Einstein considered that the random hop distances and travel times were described by exponential distributions: 
\begin{align}
P_L(l) &= l_0 e^{l/l_0} && \text{(hop distance)} \label{eq:hop}\\
P_T(t) &= T_0 e^{t/T_0} && \text{(travel time)} \label{eq:rest}
\end{align}
Here $l_0$ and $\tau_0$ are the mean hop distance and travel time, respectively.  
Einstein's key insight was that the distribution of downstream particle positions $P(x,t)$ at each space-time point $(x,t)$ can be computed as a path integral over all possible stair-step trajectories through the $(x,t)$ plane. 
Particle trajectories from $(0,0)$ to $(x,t)$ can differ in the number of steps, the height of each step (hop distance), and the width of each step (resting time). 
Einstein's argument to derive the resulting particle travel distance distribution is now briefly summarized.

Consider that a large number of particles are released into the water stream at the space-time point $(x=0,t=0)$, so they are initially in a state of motion. 
Subsequently, these particles will transition to the resting state so the fraction of particles at each point $(x_0,t=0)$ is 
$$P(x = x_0,t=0) = P_L(x_0).$$

When particles transition once more into a state of motion and back into a state of rest, a series of transitions Einstein referred to as a double phase, the fraction of particles at each space time point $(x_0 + x_1,t_1)$ will be 
$$P(x = x_0 + x_1, t = t_1) = P_L(x_1)P_T(t1).$$
When particles undergo two double phases, the fraction of particles at each space time point $(x = x_0 + x_1 + x_2, t_1 + t_2)$ is
\begin{align}
 P(x &= x_0 + x_1 + x_2 , t = t_0 + t_1 + t_2) = \\   &P_L(x_2)P_T(t_2)P_L(x_1)P_T(t_1)P_L(x_0)P_T(t_0). 
\end{align} 
Now generalizing, the fraction of particles at $(x,t)$ after $n$ double phases (rest-motion-rest transitions) is 
$$P(x = x_0 + \cdots + x_n, t = t_0+\cdots + t_n) = \prod_{i=0}^n P_L(x_i)P_T(t_i). \label{eq:pathIntegral}$$

Therefore, considering that the fraction of particles at time $(x,t)$ may have arrived there in any number $n$ of double phases, with any set of hop distances $x_i$ satisfying $x = x_0 + \cdots + x_n$ and any set of resting times $t_i$ satisfying $t = t_0 + \cdots + t_n$, the fraction of the introduced particles arriving at $(x,t)$ will result from a path integral 
\be P(x,t) = \sum_{n=0}^\infty \int dt_n \int dx_n \cdots \int dt_0 \int dx_0 \prod_{i=0}^n P_L(x_i)P_T(t_i). \ee

This integral over all stair step trajectories from $(x=0,t=0)$ to $(x,t)$ is performed in an appendix. The result is that the distribution of distances $x$ traveled by sediment grains in time $t$ is a Poisson distribution: 
\be P(x,t) = e^{-x/l_0 - T/t_0}\sum_{n=0}^\infty \frac{(x/l_0)^n}{n!}\frac{(t/t_0)^n}{n!}. \ee

This is a stochastic model of the downstream movement of tracer particles with time. 
This model is Markovian in the sense that the movement of a particle from a point $(x_0,t_0)$ to some subsequent point $(x_1,t_1)$ can be computed probabilistically with no prior information about the particle's movement at any $x<x_0$ or $t<t_0$. 
The formulation is not contingent on memory.

Several authors have generalized this formulation or clarified aspects of it. One generalization has been made concerning the velocity of particles in motion. 
As noted, the travel time is considered to be instantaneous, meaning there are vertical jumps in the $x,t$ plane (\ref{fig:einsteinTrajectories}).
That is, the velocity of particles in motion is considered to be infinite under Einstein's assumptions, which is obviously a limitation of the model under certain conditions.  
\citet{Lisle1998} approached the Einstein model from a slightly different perspective in order to explain the travel distance of sediment under sheet flow due to excessive rain fall in arid regions, where the time spent by particles in motion is not small relative to resting times. 
This formulation will be discussed in a subsequent section.

Other authors have been concerned with the distributions for hop distance and resting time which Einstein presented. \citet{Hubbell1964} took the straightforward generalization from \ref{eq:pathIntegral} for the case where the hop distance was gamma distributed instead of exponential. \citet{Ganti2010} considered other distributions.

These investigations have been followed up with fervor, as they relate to the spreading of clouds of tracer particles throughout river systems.
This category of investigations into so-called tracer diffusion was obviously initiated by Einstein with the theoretical development just sketched. 
blah blah cite a bunch of diffusion stuff
discuss obvious scope for generalization of the travel distance

\subsection*{Einstein's 1950 technical report: the stochastic transport rate of bed material}

After Einstein completed his 1937 probabilistic travel distance formulation, he did not really carry it forward, except that similar ideas are evident when he considers particle travel distance in his 1950 work.
Einstein became concerned primarily with the modeling of the sediment transport rate from a probabilistic basis.

One aspect of this work was to investiate the hydraulic components of particle motion \citep{Einstein1949}. Another, which is probably the most significant component of Einstein's scientific legacy, is his stochastic model of the sediment transport rate $q_s$, which he defined to be the number of particles passing a surface perpindicular to the channel flow in a unit time, although other definitions are available and will be considered \citep{Ballio2014}.

Einstein's 1950 model generated a transport rate of bedload $q_s$ from phenomenological arguments involving randomness as a fundamental feature.
Essentially his derivation of $q_s$ was ad hoc: it was constructed from intuition and somewhat loose reasoning in order to describe the experimental data which was available at the time. 
In light of standard methodologies in the theory of turbulence, a modern perspective on the derivation is to see it as a scaling and dimensional analysis argument \citep{}.

One can ask, "what are the relevant length, mass, and time scales of the problem, and how should they relate to form a dimensionally consistent $q_s$ with the correct behavior to describe experiments?" 
This appears to be essentially what Einstein did. 
The scales Einstein identified include a time scale $\tau$ which characterizes the typical exchange time of particles on the bed. 
This timescale is considered to be the interval between the erosion of a particle at a location and its subsequent replacement by a different deposited particle.  
Einstein also identified three relevant length scales. 
These are $a$, the particle radius; $l$, the particle hop distance; and $L$, the length of an Eulerian control cell within which particle entrainment and deposition are considered. These are summarized in table \ref{tab:einsteinScales}.

\begin{wraptable}{r}{0.5\textwidth}
\caption{The relevant scales in the sediment transport rate equation as considered by Einstein 1950. $[L]$ denotes a length scale and $[t]$ denotes a time scale.}
\label{tab:einsteinScales}
\begin{center}
\begin{tabular}{l|c|r}
\textbf{scale}: & \textbf{unit}: & \textbf{description}: \\ 
\hline
$a$ & $[l]$ & radius of sediment grains \\
$L$ & $[l]$ & size of control volume \\
$\tau$ & $[t]$ & sediment exchange timescale \\
$q_s$ & $1/([l][t])$ & sediment transport rate

\end{tabular} 
\end{center} 
\end{wraptable}

Einstein incorporated randomness as a fundamental feature of his transport model by considering that each particle on the bed surface had probability $p$ to entrain within a unit $\tau$ of time. 
Further, he reasoned there would be  
$$ n_s = \frac{L}{2a} \propto \frac{L}{a} $$ 
particles on the bed surface, an assumption which figure \ref{fig:surfaceParticles} outlines in cartoon form.

\begin{wrapfigure}{l}{0.5\textwidth}
  \includegraphics[width=.98\linewidth]{./figures/einsteingeometry.png}
  \caption{Einstein's assumption of N = L/(2a) surface particles available for entrainment is equivalent to a 2D conceptual picture of perfect spheres stacked in a perfect cubic structure.}
  \label{fig:surfaceParticles}
\end{wrapfigure}

He considered that the rate of erosion within the control volume should be proportional to the number of particles available for entrainment and the probability $p/\tau$ that a particle entrains in a unit of time: 
$$ E \propto \frac{L}{a}\frac{p}{\tau}. $$

Of course, mass conservation requires sediment transport is governed by the difference between entrainment and deposition: 
$$ \frac{\partial q_s}{\partial x} = E - D. $$
When transport is in equilibrium, one steady state solution to this differential equation is 
$$ E = D.$$

Einstein considered that the deposition rate of particles $D$ should depend on the sediment transport rate and the particle volume, since particle volume will control the forces on particles, and the sediment transport rate will control the number of particles available for deposition. One dimensionally consistent combination is 
$$ D \propto \frac{q_s}{a^3}.$$
Therefore, using the relations for $E$ and $D$ just assembled, 
$$ q_s \propto a^3 E \propto \frac{L}{2}\frac{a^2}{\tau} p. $$

Now the entrainment rate per unit area is expressed in terms of the relevant scales and the entrainment probability $p$. 
One scale which is ambiguous in meaning is the size of the Eulerian control cell $L$. Einstein stated that the most reasonable choice for the control cell size $L$ should be the mean travel distance.

As in his 1937 work, \textit{travel distance} is intended to mean the summation of individual hop distances over the possible switching combinations between motion and rest, and this is the step in which is 1937 work is at least partially carried forward. 
If the particle travels distance $l$ in each hop as assumed, and the entrainment probability into a hop is $p$, so that the deposition probability from a hop is $1-p$, the mean travel distance will be 
$$ \overline{l} = \sum_{n=0}^{\infty} (1-p)p^n(1+n)l = \frac{1}{1-p} l.$$ 
The reasoning here is that a fraction $1-p$ of particles in motion deposit after traveling distance $l$, while a fraction $(1-p)^2p$ of particles deposit after traveling distance $2l$.
Generalizing, a fraction $(1-p)p^n$ deposit after traveling a distance $(n+1)l$, so that the mean travel distance is the weighted sum above.

Therefore, Einstein's sediment transport rate is 
$$ q_s = A \frac{p}{1-p} \frac{a^2}{\tau}. $$ 
The constant of proportionality and exchange time scale are its calibration parameters.

A huge body of work within the stochastic sediment transport literature has been focused on deriving the entrainment probably per unit time $p/\tau$.

Eistein himself considered that the probability $p$ should be proportional to the interval of time in which the instantaneous lift force surpassed the weight of a bed particle. 
Many investigators followed up this topic and it has been beaten to death \citep{Grass1970, Paintal1969, Paintal1971} etc

He also prescribed that $\tau$ should be proportional to the particle sedimentation velocity. This is a reasonable assumption of a characteristic velocity scale of particle exchange, but it has been viewed as total conjecture with no empirical evidence available to support or refute this prescription.

Einstein assumed the exchange time-scale would be related to the particle radius and its sedimentation velocity. Einstein and subsequent researchers did not offer much justification for this assumption. \citep{Cheng2004} and \citet{Paintal1971} employed the bed shear velocity to develop this timescale.

There have been generalizations.

How is this markov? As presented here this theory is obviously phenomenological and heuristic. 
Einstein's physical and mechanical reasoning is clearly evident reading the original work, but the arguments leading to $q_s$ certainly lack rigor or much clear foundation in earlier work.

One especially surprising characteristic of the $q_s$ equation is that it does not explicitly contain the velocity of the fluid. 
\citet{Ancey2008} identified this discrepancy and ammended Einstein's theory using the Markov birth-death process. 
They obtain a very similar $q_s$ except that it is explicitly dependent on fluid velocity. 
Also Einstein did not explicitly derive a distribution of sediment transport rates $q_s$. Ancey does this using birth-death processes. 
This work will be reviewed at some point in this thing.

### 2021-06-24 formerPapers/grlOlderVersion/addressing_comments.tex
got one chapter in. some issues.

\documentclass[11pt]{article}
% General document formatting
\usepackage[margin=0.75in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{titling}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{booktabs} % for wrapping tabulars in accord with
\bibliographystyle{agu}

%\usepackage[math]{kurier}
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}
\newcommand\El{\mathcal{L}}
\newcommand\tg{\tilde{g}}
\newcommand\tG{\tilde{G}}
\begin{document}

	\title{Replies to reviewers: ``Back to Einstein: How to include sediment burial in bedload diffusion models?''}
	\author{James K. Pierce \& Marwan A. Hassan}
	\maketitle

\section*{Associate Editor}

Both reviewers appreciate the modelling approach but comment on the need to focus more on the physics than the model itself. Some references are missing and in general the authors should take some time to rethink the way the paper is written so that it is clear what this model does with respect to the existing literature and how it fits in it and how does this model advance our understanding of sediment transport. Although I don't expect the main conclusions of the paper to change, it would be good for the authors to have the time to change the writing approach and make sure the overall message of the paper justifies urgent publication.

\section*{Reviewer \#1}
\subsection*{Summary:}
In this work, a CTRW model that explicitly incorporates motion, rest, and burial states is used to explain changing diffusive regimes of gravel tracers in rivers. In particular, different diffusive states are related to the mean duration of particle motion, mean duration of particle rest, and mean duration of particle rest prior to burial. This is a nice example of theoretical stochastic models pointing toward physical explanations of particle transport observations.
This work is appropriate for GRL after some revisions:

\paragraph{Reply:}
Thank you reviewer \#1 for your careful reading and commentary. Your summary shows perfect understanding and we found all of your comments apt and constructive. Therefore, we responded to every comment with changes in the manuscript. We identify three major themes in your comments: (1) you indicated that the manuscript's literature review needed to more carefully outline why we expect three diffusion ranges and what earlier investigations found; (2) you pointed out several oversights --  we used the term ``scale-dependence'' without really defining it, we did not cite the papers of Furbish and coworkers, and we did not specify that we were discussing the transport of a group of tracer grains; and (3) you raised the point that many of the mathematical details included in the manuscript are not essential to the physical story. We responded to (1) and (2) by totally reworking the manuscript's literature review, carefully defining ``scale dependence'' and minimizing its usage, citing all relevant papers by Furbish and coworkers, and emphasizing that our topic is the diffusion of tracers. We have responded to (3) by moving the majority of the mathematical details into a supplement, including all three appendices and a large chunk of section 2. Thank you again for your careful comments that have improved the manuscript.

\subsection*{Specific Comments:}
\begin{enumerate}
\item L. 37 diffusion is induced by transport characteristics of grains, but also the conditions they encounter

\paragraph{Reply:} We have moved this sentence to be the first of the introduction and included the suggested revision.

\item L. 43 Did Einstein really conclude that bedload transport shows normal diffusion? More specifically, no mathematical tools existed that would suggest that it was possible for rest times to not be described by a mean.

\paragraph{Reply:} Yes, Einstein describes normal diffusion of bedload in equation 19 of his 1937 PhD thesis \citep{Einstein1937}.
He also presents flume measurements of the sediment resting time and step length distributions, which he found to be exponential. In his introduction, he carefully describes his realization that the sediment resting time cannot be characterized by its mean value. He used the exponential resting time distribution extensively in his analysis, which is a pioneering application of random walk theory that precedes the classic work of \citet{Montroll1965} by nearly three decades.

\item L . 54 and elsewhere- I think that the use of "scale dependence" in the entire manuscript needs to be clarified. Scale dependence as a problem most commonly refers to model parameters changing with time or space scale - usually this is because the model is incorrectly applied- for example using the diffusion model for a superdiffusive process results in the D coefficient being scale dependent. Here, we have a process whose scaling regime itself changes after crossing characteristic timescales and we need a model that can accommodate this- as demonstrated in this manuscript. This needs to be clarified
\paragraph{Reply:} Thank you for raising this point. We have clarified our use of ``scale dependence'' at L. 58, so we could use it without ambiguity in the rest of the manuscript.

\item L. 54 Rather than the lit review just saying that others have shown different types of scale dependence, please specify what they found. There is comparison to Nikora's 3 scales- but these have different origins than those described in this manuscript. For example the smallest scale includes correlation of velocity in Nikora's work.
\paragraph{Reply:}

We totally reworked this section to specify what various studies have found. We split the literature review into one paragraph about experimental studies and another about modelling studies. As we discuss after L. 54, at base level Nkora attributed the smallest scale to the smoothness of bedload trajectories in between subsequent resting periods. This timescale is related to the correlation of (virtual) velocity, so there is no essential difference.

\item Section 3 and Table 1. The mathematical solution and description of functions used to solve PDEs is not relevant to the story. I suggest moving much of this section to a supplement and dedicating more of the paper to the sediment transport/geomorphic story. For example, Nikora's global timescale is subdiffusive, while yours is superdiffusive. Why? How your model relates to all the others you mention is more important to the reader and how they should model tracer transport.
\paragraph{Reply:} 
We moved the majority of this information to a supplement as you suggest, and we have added some text to the discussion to clarify these issues. However, we note that the Nikora data was assembled by patching together several datasets, meaning we don't have total faith in these diffusion ranges to begin with. For example, the local range was resolved by numerical simulations, the intermediate range was resolved by the Duck creek data, and the global range was resolved by the \citet{Drake1988} data.

\item Reference to the detailed bedload lab studies and detailed statistical descriptions in a variety of manuscripts by Mark Schmeeckle and David Furbish are completely missing from this manuscript and make statements about transport and resting times not having been studied incorrect. Incorporating their large body of work into this manuscript will strengthen it.
\paragraph{Reply:} We added citations to \citet{Fathel2015}, \cite{Furbish2016}, \citet{Roseberry2012}, and \citet{Furbish2012b} as they directly support our use of exponential models for resting and motion intervals.

\item Mention that what is being modeled is the spreading of a cloud of tracers, rather than bulk transport generally. This will help the reader understand that line 217- all tracers become buried. "All particles" can not be buried by definition- some have to be at the top.
\paragraph{Reply:} We added the requested clarification in many places in the manuscript, perhaps most usefully in L7 (first key point) and L 27 (first sentence of introduction).

\item Key points- the key points should tell us takeaways... instead of we apply the model... perhaps talk about the characteristic timescales and what they represent. Have more of a physical focus than a model focus.
\paragraph{Reply:}

We totally revised the key points to address this shortcoming. We sincerely appreciate your careful attention and constructive comments, reviewer \# 1.
\end{enumerate}

\section*{Reviewer \#2}
\subsection*{Summary}
This paper presents a derivation of a continuous time random walk model for bed load particle diffusion that proposes four scaling regimes. As far as I can tell the manuscript is technically correct, though it is not clear who the target audience is or how it advances our knowledge of sediment transport over other bed load diffusion treatments.

\paragraph{Reply:}
We thank you for the careful reading, reviewer \# 2. We have added a clear statement of who the target audience is at L. 24 and also modified the first paragraph to clarify this point. We have modified the work to place its key contribution to bedload diffusion understanding front and center: it is the first model of three diffusion ranges describing the propositions of Nikora. We synthesize the literature to argue that three diffusion ranges is correct in L. 61-93, then we point out that have developed the first analytical model of three diffusion ranges at L. 9, L. 269, and L. 395.

We believe the key contribution of our paper is that it's the first model deriving three stages of bedload diffusion following the expectation for three set out by \citet{Nikora2001a} and existing experimental data. We'd like to highlight that \citet{Wu2019}, \citet{Wu2019a}, and \citet{Lajeunesse2018} have all highlighted the need for this extension.

\subsection*{General Comments}
This is a challenging review for me. On one hand the manuscript appears, to my knowledge, technically correct, however I am not sure that I really understand how this advances the field over other recent attempts at this problem. This is another attempt at approximating bed load transport as a diffusion problem and similar to past attempts at the problem it reveals several diffusion regimes for a conceptual set of particles traveling with like sized particles. It is not clear what the correct number of diffusion regimes actually is for bed load (two, three, four)? From my understanding of previous papers (most of which are cited within the manuscript), the number of scaling regimes and the numerical values of the scaling exponents depends on the types of distributions used for the rest and travel durations and the particular sets of equations used within the model. Here they are taken as exponential without justification, I suspect because it makes the derivation simpler? Furbish and colleagues provide numerous examples of what these distributions may be from a mathematical perspective and support them with experimental data. Perhaps it would be useful to consider their advances as the starting point for the model application here. Given the data supported theoretical distributions, would you still derive three (four counting the new longer time constant one) regimes? It is not clear to me how challenging it would be to incorporate their results, but it would provide a stronger rational for the results presented here.

My recommendations for making this manuscript stand apart could take several different routes. Option 1 - edit the manuscript to place the new capabilities of this model front and center. What does this model now allow us to do that we could not derive from past models of this problem (i.e. what does the analytical solution allow us to do moving forward that the many previous numerical solutions can't do). As an example, as a non mathematician I can not tell if the analytical derivation is particulary novel or a standard application of mathematical concepts. Option 2 would be to utilize some form of field of laboratory data to demonstrate that the four scaling regimes are indeed representative of bed load transport rather then being the result of a set of mathematical equations that may represent a simplified conceptual form of a process like bed load transport. These are only suggestions and I am sure there are other routes that could be taken to clarify the impact of the contribution.

It is hard for me to rate the article for above a science category 3 in its current form, however given that it appears correct it is up to the authors to convince the editor that this paper is indeed an advance worthy of rapid publication. The presentation of the manuscript is fine and the writing is easy to follow, the figures are fine as they are but they don't necessarily add much to the paper as is.

\paragraph{Reply:}
You've presented us a unique perspective on the weak points in the work, and we've worked to concentrate on the key points you've raised and act on them.
We hope you don't mind our summarizing of key points from your general comments and replying to each in turn. 
\begin{enumerate} 
\item The target audience is unclear.
\paragraph{Reply:}
We have added text at L. 24 and between L. 29 and L. 32 to illustrate the target audience and some practical problems this fundamental research lends perspective to.

\item The impact of the work is not obvious.
\paragraph{Reply:}
We have rewritten the paper to emphasize its essential contributions to bedload diffusion understanding. This is the first analytical description of bedload diffusion across the local, intermediate, and global timescales introduced by \citet{Nikora2001a}. The model thus demonstrates the required components for three bedload diffusion ranges: these are (1) the duration of sediment motions, (2) mobile-immobile switching, and (3) a sediment trapping process.
In addition, the work re-frames earlier studies \citep[e.g.,][]{Wu2019,Lisle1998,Lajeunesse2018,Einstein1937} within a comprehensive mathematical formalism built out of a fusion of condensed matter \citep[e.g.,][]{Weiss1994} and chemical physics \citep[e.g.][]{Schmidt2007} ideas.

\item The expected number of diffusion ranges is unclear.
\paragraph{Reply:}
This is an uncertainty shared by the research community that we set out to address. Your comment indicated we failed to achieve this in the initial manuscript, so we worked to make the points clear in the revised manuscript. Please see L. 61-98. In short, cross comparing experimental studies demonstrates three diffusion ranges, while scale limitations prevent any one experiment from demonstrating three diffusion ranges. Meanwhile, models show all one or two diffusion ranges, but only one Newtonian simulation approach has shown all three \citep{Bialik2012}.

\item You believe the type of distributions might determine the number of regimes rather than the number of states (i.e., motion, rest, burial).

\paragraph{Reply:}
Random walk models show asymptotic diffusion characteristics that are independent of the functional form of the input distributions, provided light tailed distributions are not swapped for a heavy tailed ones \citep{Weiss1994,Weeks1998}.
In the preliminary stages of this work, we attempted to model Nikora's three diffusion ranges using heavy tailed sojourn times in a two state model. This is actually suggested in the conclusion of \citet{Nikora2001a}, but we determined from numerical simulations that it did not work, and in any case such a model would not be analytically solvable, so we moved on to try incorporating the sediment burial process. In the end, we came to the conclusion that three ranges stem from three states: motion, rest, and burial. We do not believe heavy tailed distributions are sufficient.
We note \citet{Lisle1998}, \citet{Lajeunesse2018}, and \citet{Wu2019} are all two state random walk models showing two ranges of diffusion, whereas our three state model shows three ranges of diffusion.

\item You take issue with the simple assumption of exponential sojourn times.

\paragraph{Reply:}
We have added text to the manuscript and a handful of citations to Dr. Furbish and coworkers to justify our choice of these distributions between L. 179 and L. 185.
In short, the exponential assumption does make the derivation easier because it makes the stochastic model Markovian \citep[e.g.,][]{Cox1965}. However, we emphasize that prevalent models in the literature have simplified even further than we have by neglecting all statistical moments except the first two \citep[e.g.][]{Wu2019}. This leads to highly simplified Gaussian probability distributions of the tracer cloud, in contrast to the less idealized Bessel distributions in the manuscript.

\item You believe we have neglected the works of Furbish and coworkers.
\paragraph{Reply:}
\citet{Furbish2012a}, \citet{Roseberry2012}, \citet{Fathel2015}, and \citet{Furbish2016} all provide direct support for our use of exponential travel and resting time distributions and we have incorporated these citations into the text. Our work is totally aligned with these papers, as they mostly confirm and build up the much earlier findings of \citet{Einstein1937}.

\item You are unsure if changing the distributions from exponential will affect the essential conclusion of three diffusion ranges.
\paragraph{Reply:}
We still expect $3$ $(+1)$ regimes regardless of the distributions chosen, provided they are light tailed and there is a sufficient separation of motion, rest, and burial timescales.

\item You wonder how challenging it would be to incorporate different distributions.
\paragraph{Reply:}
Choosing other light tailed distributions would be quite easy in numerical simulations, but difficult or impossible using analytical methods.

\item You recommend placing the capabilities of the model front and center to address the new features of the model.
\paragraph{Reply:}
We have added reference to the key contribution of this paper (a first analytical description of Nikora's three diffusion ranges) at L. 9, L. 101, L. 271, and L. 397.

\item You are unsure of the novelty of the analytical derivation and its relative merits to earlier numerical approaches.
\paragraph{Reply:}
We are unaware of any other model of more than two diffusion ranges in the broader physics literature, while the interdisciplinary impact of anomalous diffusion research has been carefully highlighted \citep[e.g.,][]{Sokolov2012,Metzler2014}. We are fusing ideas from chemical physics \citep[e.g.,][]{Schmidt2007} and condensed matter physics \citep[e.g.,][]{Weiss1994} to describe an open problem in river science, and we have not seen an analysis quite like this in the bedload transport literature. For this reason, we chose to describe our mathematical analysis with sufficient depth to be easily understood by our audience.
Nevertheless, we have acted on your comment and moved a great deal of the mathematical details to a supplement.

\item You suggest field or laboratory data might strengthen the paper.
\paragraph{Reply:}
We have added discussion between L. 61 and L. 73 to indicate why experimental resolution of three diffusion ranges has been so elusive. We do not believe existing technology allows such a resolution. This is part of the reason our model explaining three diffusion ranges from a set of simple assumptions is a useful contribution to the research community.

\end{enumerate}
In summary, we appreciate your careful reading and constructive comments, and we sincerely thank you for your attention toward improving our work. Now we will turn to your specific comments.

\subsection*{Specific Comments}
\begin{enumerate}

\item 
Throughout paper - Single authors seem to have initials cited in text. See N. D. Bradley ln. 32. May be a result of his initials being reported two ways in your references. Likely an error in the reference manager software.

\paragraph{Reply:}
Thanks! We fixed this. It was a Mendeley problem as suggested.

\item 
Ln. 55-63. Is their a physical justification why we should expect three regimes? Why do these studies, taken collectively, produce conflicting results? If this paper could synthesize this literature and demonstrate (a) why three regimes is correct and (b) why the other studies do not produce a consensus picture of the problem that would be a very useful contribution.

\paragraph{Reply:}
The present work attempts to convince readers the necessary ingredients of three stage diffusion are (1) a motion interval, (2) a mobile-immobile switching, and (3) a trapping process. We believe the introduction now outlines why we expect three ranges in line with Nikora, despite the incomplete physical motivation provided in the early Nikora papers and highlighted after L. 54.

\item 
Ln. 151. What is the justification for the exponential rest distribution functions? Likewise please provide a justification for the exponential durations of a motion in line 157-158.

\paragraph{Reply:}

Similarly \citet{Furbish2016}, \citet{Furbish2017}, \citet{Ancey2006}, \citet{Ancey2008}, \citet{Wu2019}, \citet{Lisle1998}, \citet{Lajeunesse2018} have all used, measured, or derived exponential motion durations.
We have discussed this carefully between L. 179 and L. 186.

\item 
Figure 2. These timescales seem a bit short for bed load transport. The geomorphic timescale is on the order of 8 hrs to ~10 days (less than one snowmelt season in the Rockies). I understand that this is for parameters derived within a small scale flume, but even a recirculating flume might run well past these timescales.

\paragraph{Reply:}
The geomorphic timescale has $T_G \propto \sqrt{T_b}$, where $T_b = 1/\kappa$ is the mean time required for sediment resting on the surface to become buried. In figure 2 we arbitrarily chose $T_b = 2$hrs, implying the geomorphic timescale of order $8$hrs to $10$days you mention. Were we to increase $T_b$ to $200$hrs, the geomorphic timescale would increase to $80$hrs to $100$ days. A burial timescale of $200$hrs still seems reasonable enough compared to field studies of vertical exchange of sediment such as \citet{Ferguson2002} and \citet{Haschenburger2011}. We are unsure of a realistic value for this timescale, as no measurements of it exist, so we highlight this uncertainty at L. 386 as a topic for future research.

\item 
Ln. 245. Is there a physical explanation for the super ballistic scaling within the local region? This was also observed within the recent work of Lajeunesse et al. (2018 - Advection and dispersion of bed load tracers, Earth Surface Dynamics), though they suggest that only two regimes are necessary.

\paragraph{Reply:}
We added the physical explanation for super ballistic local range diffusion at L. 54. Interestingly, \citet{Lajeunesse2018} is also mathematically equivalent to \citet{Lisle1998}. They suggest two only two regimes are necessary, but they also state ``Here, we neglect this mechanism [sediment burial] and restrict our analysis to steady and uniform sediment transport.'' We have included sediment burial and derived the third stage proposed by Nikora, in addition to the two derived by Lisle et al. and Lajeunesse et al.

\item 
Ln. 305. It isn't clear how the model confines the valid range for these types of studies. Please provide an example or additional explanation.

\paragraph{Reply:}
We added an additional sentence to clarify this point. 
\end{enumerate}
We sincerely appreciate your time reviewer \#2. Your comments greatly improved the manuscript.

\bibliography{biblio}
\end{document}

### 2021-06-24 formerPapers/grlRrevision2/rwdiffu.tex
got one chapter in. some issues.

--- 2021-06-24 ch4downDiff.tex
@@
-%%!TEX root = diss.tex
+\documentclass[grl]{agujournal2018}
+%\documentclass[]{agujournal2018}
+\usepackage{apacite}
+\usepackage{natbib}
 
-\chapter{Inclusion of sediment burial}
-\label{ch:downDiff}
+\usepackage{wrapfig}
+\usepackage{url}
+\usepackage{lineno}
+\usepackage{trackchanges}
+\soulregister\citet7
+\soulregister\citep7
+\soulregister\ref7
+\addeditor{JKP}
+
+\draftfalse
+\journalname{Geophysical Research Letters}
+
+%custom packages
+\usepackage{amsmath,amssymb,amsfonts,amsthm}
+\usepackage{comment}
+\usepackage{booktabs}
+% custom commands
+\newcommand\be{\begin{equation}}
+\newcommand\ee{\end{equation}} 
+\newcommand\bra{\langle}
+\newcommand\ket{\rangle}
+\newcommand\om{\omega}
+\newcommand\tom{\tilde{\omega}}
+\newcommand\tg{\tilde{g}}
+\newcommand\tp{\tilde{p}}
+\newcommand\tG{\tilde{G}}
+\newcommand\El{\mathcal{L}}
+%\usepackage{layouts}
+%\printinunitsof{in}\prntlen{\textwidth} % check scales in the document
+\linenumbers
+
+% THIS IS THE LINENO PATCH TO WORK WITH AMSMATH
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+	\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+	\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+	\renewenvironment{#1}%
+	{\linenomath\csname old#1\endcsname}%
+	{\csname oldend#1\endcsname\endlinenomath}}% 
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+	\patchAmsMathEnvironmentForLineno{#1}%
+	\patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+	\patchBothAmsMathEnvironmentsForLineno{equation}%
+	\patchBothAmsMathEnvironmentsForLineno{align}%
+	\patchBothAmsMathEnvironmentsForLineno{flalign}%
+	\patchBothAmsMathEnvironmentsForLineno{alignat}%
+	\patchBothAmsMathEnvironmentsForLineno{gather}%
+	\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
+% the patch comes from http://phaseportrait.blogspot.com/2007/08/lineno-and-amsmath-compatibility.html
+
+\begin{document}
+
+\title{Back to Einstein: burial-induced three-range diffusion in fluvial sediment transport}
+
+\authors{J. Kevin Pierce \affil{1}and Marwan A. Hassan\affil{1}}
+\affiliation{1}{Department of Geography \\University of British Columbia}
+\correspondingauthor{Kevin Pierce}{kpierce@alumni.ubc.ca}
+
+\begin{keypoints}
+\item Random walk model describes coarse gravel tracers spreading out within a river as they gradually become buried
+\item Interchange of grains between motion, surface rest, and burial states generates three diffusion ranges as the observation time increases
+\item Sediment burial dominates long-time properties and exhumation may develop a fourth ``geomorphic'' range of tracer diffusion
+\end{keypoints}
+
+\begin{abstract}
+Individual grains move through gravel bed rivers in cycles of motion and rest, so tracer grains diffuse as they transport downstream.
+Tracer experiments demonstrate at least three diffusion ranges as observation times increase, with different spreading rates in each range.
+Up till now, the generating processes of these ranges remains unclear.
+In this work, we develop a random walk model of individual bedload trajectories including motion, rest, and burial processes.
+The model describes three bedload diffusion ranges that terminate in a fourth non-diffusive range when all tracers become buried.
+Using the model, we attribute three-range tracer diffusion to the interplay between motion, rest, and burial processes, and we relate the multi-range diffusion characteristics to the timescales of these processes.
+\end{abstract}
+
+\section{Introduction}
 
@@
 Our target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if we know it started with the initial distribution $p(x,0)=\delta(x)$.
-We characterize times spent moving or resting on the surface by exponential distributions $\psi_2(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
+We characterize times spent moving or resting on the surface by exponential distributions $\psi_2(t)&=k_2e^{-k_2 t}$ and $\psi_1(t) &= k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
 We expect our conclusions will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
@@
 \begin{figure}[t]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/pdf-plot.png}
+	\includegraphics[width=\linewidth,keepaspectratio]{pdf-plot.png}
 	\caption{Joint distributions for a grain to be at position $x$ at time $t$ are displayed for the choice $k_1=0.1$, $k_2=1.0$, $v=2.0$.
@@
 \begin{figure}[h!]	
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/diffusion.png}
+	\includegraphics[width=\linewidth,keepaspectratio]{diffusion.png}
 	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (\ref{eq:var}) with mean motion time $1.5$ s, resting time $30.0$ s, and movement velocity $0.1$ m/s -- values comparable to laboratory experiments transporting small ($5$ mm) gravels \citep{Lajeunesse2010,Martin2012}.
@@
 Our model represents these conclusions: non-fractal (and super-diffusive) bedload trajectories exist on timescales short enough that each grain is either resting or moving, while fractal (and normal-diffusive) bedload trajectories exist on timescales when grains are actively switching between motion and rest states.
-We conclude that local and intermediate ranges stem from the interplay between motion and rest timescales, as demonstrated by earlier two-state random walk models \citep{Lisle1998,Lajeunesse2017} and by all Newtonian models that develop sequences of motions and rests \citep{Nikora2001a, Bialik2012}, even those including heavy-tailed rests \citep{Fan2016}.
+We conclude that local and intermediate ranges stem from the interplay between motion and rest timescales, as demonstrated by earlier two-state random walk models \citep{Lisle1998,Lajeunesse2018} and by all Newtonian models that develop sequences of motions and rests \citep{Nikora2001a, Bialik2012}, even those including heavy-tailed rests \citep{Fan2016}.
 
@@
 
-The analytical solution of bedload diffusion in equation (\ref{eq:var}) reduces exactly to the analytical solutions of the \citet{Lisle1998} and \citet{Lajeunesse2017} models in the limit without burial ($\kappa \rightarrow 0$), the \citet{Wu2019} model in the limit of instantaneous steps ($k_2 \rightarrow \infty$ and $l = v/k_2$), and the original \citet{Einstein1937} model in the limit of instantaneous steps without burial.
-These reductions demonstrate that the majority of recent bedload diffusion models, whether developed from Exner-type equations \citep{Wu2019,Pelosi2014,Pelosi2016} or advection-diffusion equations \citep{Lisle1998,Lajeunesse2017}, can be viewed equivalently as continuous-time random walks applied to individual bedload trajectories.
+The analytical solution of bedload diffusion in equation (\ref{eq:var}) reduces exactly to the analytical solutions of the \citet{Lisle1998} and \citet{Lajeunesse2018} models in the limit without burial ($\kappa \rightarrow 0$), the \citet{Wu2019} model in the limit of instantaneous steps ($k_2 \rightarrow \infty$ and $l = v/k_2$), and the original \citet{Einstein1937} model in the limit of instantaneous steps without burial.
+These reductions demonstrate that the majority of recent bedload diffusion models, whether developed from Exner-type equations \citep{Wu2019,Pelosi2014,Pelosi2016} or advection-diffusion equations \citep{Lisle1998,Lajeunesse2018}, can be viewed equivalently as continuous-time random walks applied to individual bedload trajectories.
 Within random walk theory, sophisticated descriptions of transport with variable velocities \citep{Zaburdaev2008,Masoliver1994}, correlated motions \citep{Escaff2018,Vicsek2012a}, and anomalous diffusion \citep{Masoliver2016,Fa2014,Metzler2014} have been developed.
@@
 Ultimately, we emphasize that the multi-state random walk formalism used in this paper implicitly underlies most existing bedload diffusion models and provides a powerful tool for researchers targeting landscape-scale understanding from statistical concepts of the underlying grain-scale dynamics.
+
+\appendix
+
+\acknowledgments
+K.
+Pierce acknowledges thoughtful exchanges with Eduardo Daly and Eric Weeks during the early stages of this work.
+Both authors would like to thank Matteo Saletti, Conor McDowell, Shawn Chartrand, and Will Booker for helpful discussions.
+M.
+Hassan is supported by an NSERC Discovery grant.
+The Python simulation code is freely available at \sloppy
+\url{https://zenodo.org/record/3659291#.Xj3p7XVKjIU}.
+
+\bibliography{biblio.bib}
+\end{document}

### 2021-06-24 formerPapers/jgr/supp.tex
got one chapter in. some issues.

\documentclass[11pt]{article}
% General document formatting
\usepackage[margin=0.75in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{titling}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{algpseudocode}
\usepackage{mathtools} % for colon equals

\usepackage{booktabs} % for wrapping tabulars in accord with
\bibliographystyle{agu}
\setlength{\droptitle}{-5em}   % This is your set screw

%\usepackage[math]{kurier}
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}
\newcommand\El{\mathcal{L}}
\newcommand\tg{\tilde{g}}
\newcommand\tG{\tilde{G}}
\date{} % make the date field blank

% for pseudo code -- example from https://tex.stackexchange.com/questions/163768/write-pseudo-code-in-latex
\makeatletter
\def\BState{\State\hskip-\ALG@thistlm}
\makeatother

\begin{document}
\title{Supplementary Material for ``Joint stochastic bedload transport and bed elevation model: variance regulation and power-law rests" by J. Kevin Pierce and Marwan A. Hassan}
\maketitle

\section{Numerical simulation algorithm}

The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1992} and \citet{Gillespie2007}.
The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
In the following sections, we begin by demonstrating the time intervals $\tau$ between subsequent transitions (i.e., the intervals between transition times) are exponentially distributed within the model we develop in the main text. Then we describe the SSA as a consequence of this property.

\subsection{Times between transitions of any kind}
Our joint stochastic description of bedload and bed elevation changes is characterized by a set of states $(n,m)$ where $n$ and $m$ are positive integers. 
Our description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in equations (2-5) in the main text.
From the state $(n,m)$, the rate (probability per unit time) for any transition to occur is the sum over all possibilities:
\begin{multline} A(n,m) = R_{MI}(n+1,m|n,m) + R_E(n+1,m-1|n,m) \\+ R_D(n-1,m+1|n,m) + R_{MO}(n-1,m|n,m).\end{multline}
Using this, the probability that no transition occurs from the state $(n,m)$ in a small time interval $\delta \tau$ is $1-A(n,m)\delta \tau$. If we denote by $Q(\tau|n,m)$ the probability density that a transition of any kind occurs from the state $(n,m)$ after a time $\tau$, we can express the probability density that a transition happens after a slightly larger time $\tau + \delta \tau$ as 
\be Q(\tau+\delta \tau|n,m) = \big[1-A(n,m)\delta \tau\big]Q(\tau|n,m).\ee
This equation invokes the Markov property, since it does not involve the past history of states $(n,m)$. Taking $\delta\tau \rightarrow 0 $ we find the master equation $\frac{d}{d\tau}Q(\tau|n,m) = -A(n,m)Q(\tau|n,m)$, from which we conclude the time $\tau$ between subsequent transitions is distributed as 
\be Q(\tau|n,m) = A(n,m)e^{-A(n,m)\tau}. \label{eq:exp}\ee
Therefore we have shown the time $\tau$ to the next transition from a state $(n,m)$ is exponentially distributed with mean value $\bar{\tau} = 1/A(n,m).$ In deriving this result, we used the normalization condition $1 = \int_0^\infty Q(\tau|n,m)d\tau.$

This result implies if the stochastic process transitioned to the state $(n,m)$ at a time $t$, the next transition will occur at a time $t+\tau$ with $\tau$ a random variable drawn from the exponential distribution \ref{eq:exp}.
Therefore, we can determine the times between subsequent transitions by drawing exponentially distributed random numbers.

\subsection{Selection of transitions that occur}

So far, we have determined how to step the time from one transition to the next, but we have not specified how to step the state variables $n$ and $m$ at each transition time. 
That is, we have not specified the type of transitions that occur at each time step.
Intuitively, this will depend on the relative magnitudes of the rates from equations (2-5) in the main text: the transition with the highest rate is most likely to occur. This is formalized by generating the ratios
\be S = \Bigg\{\frac{R_{MI}(n+1,m|n,m)}{ A(n,m)},\frac{R_E(n+1,m-1|n,m)}{ A(n,m)},\frac{R_D(n-1,m+1|n,m)}{ A(n,m)},\frac{R_{MO}(n-1,m|n,m)}{ A(n,m)}\Bigg\}. \label{eq:rel}\ee
By construction, $\text{sum}\{S\}=1$.
By forming cumulative sums of the four ratios, we partition the unit interval $[0,1]$ into four chunks, each associated with a transition -- either migration in, entrainment, deposition, or migration out. The transitions with the highest rates have the largest associated chunks. To randomly select the transition that occurs at a transition time, we draw a random number on $[0,1]$ and find which chunk it falls in.

In summary, to step the process through a single transition, we draw the time interval to the next transition from the distribution (\ref{eq:exp}), then draw a uniform random from $[0,1]$ and use it to select the transition that occurs from the cumulative sum of the ratios in (\ref{eq:rel}). The SSA simply iterates this random number generation/selection process to generate exact realizations of the stochastic process: these are series of $n$ and $m$ through time from which we can calculate any statistics of interest.

\subsection{Pseudo code for the Gillespie SSA}

To initialize a simulation, we specify the initial conditions $n_0$ and $m_0$, the model parameters for use in equations (2-7) in the main paper, and the desired simulation duration $t_\text{max}$. The SSA uses these inputs to generate time series of $n$ and $m$ as follows:

\rule{\linewidth}{1pt}

\begin{algorithmic}
	\State $t = 0$ 
	\State $n = n_0$\Comment{Set the initial state $(n_0,m_0)$}
	\State $m =  m_0$

	\While{$t<t_{\text{max}}$;}{	\Comment{Simulation will go until $t$ surpasses $t_\text{max}$}
	\State record $(n,m,t)$ \Comment{Build time series of $n$ and $m$}		
	\State draw $\tau$ from eq. (\ref{eq:exp}) \Comment{Select time to next transition}
	\State $t \coloneqq t+\tau$
	\State draw a random number $r$ in $[0,1]$ \Comment{Select type of transition that occurs}
	\State compute the ratios $r_1,r_2,r_3,r_4$ in eq. (\ref{eq:rel})
	\State form the cumulative sums $r_i = \sum_{1\leq j\leq i}r_j$ 
	\Comment{Now enact the transition}
	\If {$0\leq r<r_1$}	\Comment{Migration in}
	\State $n \coloneqq n+1$
	\ElsIf {$r_1\leq r < r_2$} 	\Comment{Entrainment}
	\State $n \coloneqq n + 1$
	\State $m \coloneqq m-1 $
	\ElsIf {$r_2\leq r < r_3$} 	\Comment{Deposition}
	\State $ n \coloneqq n - 1$
	\State $ m \coloneqq m + 1$
	\ElsIf {$r_3 \leq r \leq 1$} \Comment{ Migration out}
	\State $ n \coloneqq n-1$
	\EndIf
}
	\EndWhile
\end{algorithmic}

\section{Approximate solutions of the Master equation}

\subsection{Mean field solution of particle activity}

By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation (8) of the manuscript over all values of $m$, we obtain the mean field equation for the particle activity: 
\be 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1)+\gamma(n+1)A(n+1)-(\nu  + \lambda + \mu n + \sigma n + \gamma n)A(n).\ee
This can be solved by introducing the generating function \citep{Cox1965} $G(x) = \sum_n x^n A(n)$, providing
\be 0 = (\nu+\lambda)(x-1)G + [\mu x^2 + \sigma + \gamma - (\mu + \sigma + \gamma)x]\frac{\partial G}{\partial x},\ee
which is separable and integrates for 
\be G(x) = \Bigg( \frac{\gamma + \sigma -\mu}{\gamma + \sigma - \mu x}\Bigg)^{\frac{\nu + \lambda}{\mu}}\ee
after applying the normalization condition $G(1)=1$.
From the definition of $G$ we can determine $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in equation (10) of the manuscript.

\subsection{Mean field solution of bed elevations}

From the negative binomial distribution we have $\langle n |m \rangle = \langle n \rangle$, so equation (9) provides 
\begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m). \end{multline}
Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives equation (11) of the manuscript
\be 0 = [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1) - 2M(m).\ee
The \citet{Martin2014} Ornstein-Uhlenbeck model.

We solve this using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$ This works since $z_1$ is small. Introducing $z$ and writing $\overline{\kappa}=\kappa/z_1$ gives
\be 0 = [1+\overline{\kappa}(z+z_1)]M(z+z_1) + [1-\overline{\kappa}(z-z_1)]M(z-z_1) - 2M(z).\ee
$\overline{\kappa}$ should not depend on $z_1$ since the magnitude of the feedbacks between bed elevations and entrainment and deposition rates depends on elevation changes, and not on the size of grains or the length of the control volume.
Expanding the entire first and second terms to second order $z_1$ provides the Fokker-Planck equation
\be 0 = -2\overline{\kappa}z_1[zM(z)]' + z_1^2 M''(z). \ee
Taking into account that this distribution should be normalizable, $\lim_{z\rightarrow \pm \infty}M(z) = 0$, we find
\be M = M_0e^{-\kappa (z/z_1)^2}\ee
as stated in the manuscript.

\subsection{Closure equation approach for bed elevations}
We set out to determine an approximate relationship to close $\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$.
\be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
for the mean particle activity conditional to the elevation $m$ into into manuscript equation (9), noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
\be 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
\ee
At this point, we consider $c$ an undetermined positive constant that may depend on the entrainment, migration, and deposition rates.
Taking the Fokker-Planck expansion and requiring the distribution $M(m)$ to vanish at infinity for normalization yields
\be 0 = 2\kappa m \Big( 2 + \frac{(\sigma-\mu)c}{E}\Big)M(m) + \Big\{
\Big[2-\kappa m \frac{(\sigma+\mu)c}{E} \Big]M(m)\Big\}'.\ee
Finally, integrating, expanding to first order in $\kappa$, and exponentiating to solve for $M$, we find
\be M(m) = M_0 \exp\Big\{-\kappa m^2 \Big[1 + \frac{(\sigma-\mu)c}{2E}\Big]\Big\}.\ee
Since the numerical solutions show $\sigma_m^2 = \frac{1}{4\kappa}$, we determine the closure relation
\be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \ee
corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation (12) of the manuscript that is plotted with the numerical simulations in manuscript figure (4a).

\bibliography{biblio}
\end{document}

### 2021-06-24 pref.tex
got one chapter in. some issues.

--- 2021-06-24 ack.tex
@@
 
-\chapter{Acknowledgments}
+\chapter{Preface}
 
-Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way.
-Tyler cannon.
-.
-.
-
-Shawn Chartrand, Conor McDowell, Matteo Saletti, Will Booker, David Adams, Leo King, Tobias Mueller, Nisreen Ghazi,
-
-Adam Payne, Adam Collins, Megat Denney, Andrew Rice, Evan Graber, Scott Ferris, Craig Tenney, and the rest of the WVU physics lounge family.
-
-Andy Osadetz, Josh Plankeel,
-
-My partner Mary for putting up with the late nights.
+A preface should be written indicating what is to come and what the point of this whole thing is.

### 2021-06-24 clean.sh

@@ -1,20 +1,20 @@
 #!/bin/bash
 # this cleans out all of the auxilary and log files after compilation
-rm -f .abs/abs.aux
-rm -f .ack/ack.aux
-rm -f .ch1intro/ch1intro.aux
-rm -f .ch2flipFlop/ch2flipFlop.aux
-rm -f .ch2flipFlop/ch2flipFlop.log
-rm -f .ch3bedEle/ch3bedEle.aux
-rm -4 .ch3bedEle/ch3bedEle.log
-rm -f .ch4downDiff/ch4downDiff.aux
-rm -f .ch4downDiff/ch4downDiff.log
-rm -f .ch5conc/ch5conc.aux
-rm -f .ch5conc/ch5conc.log
-rm -f .laySum/laySum.aux
-rm -f .laySum/laySum.log
-rm -f .pref/pref.aux
-rm -f .pref/pref.log
+rm -f abs.aux
+rm -f ack.aux
+rm -f ch1intro.aux
+rm -f ch2flipFlop.aux
+rm -f ch2flipFlop.log
+rm -f ch3bedEle.aux
+rm -f ch3bedEle.log
+rm -f ch4downDiff.aux
+rm -f ch4downDiff.log
+rm -f ch5conc.aux
+rm -f ch5conc.log
+rm -f laySum.aux
+rm -f laySum.log
+rm -f pref.aux
+rm -f pref.log
 rm -f appendix.aux
 rm -f diss.aux
 rm -f diss.bbl

### 2021-06-24 ch2flipFlop.tex
got another component in there.. ch 2 . Now to add the langevin paper and get all figures working.

@@
 \section{Introduction}
-\section{Theoretical development}
+
+A relativity weak flow shearing a bed of sediment entrains individual particles into a state of motion controlled by turbulent forcing and intermittent collisions with other grains at rest on the bed, generating wide fluctuations in the sediment velocity \citep{Heyman2016,Fathel2015}.
+Bed load particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered divots on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
+Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2019,Valyrakis2014}, particles are once again entrained, and the cycle repeats.
+Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and rest.
+This process has proven itself extremely challenging to describe mathematically, given the technicality of the stochastic physics required \citep{Pierce2020,Ancey2020}.
+
+To date, descriptions of bed load transport have therefore simplified the problem in various ways to enable progress.
+The foundational work is due to Einstein, who considering bed load motions as instantaneous, so he could describe bed load transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
+Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E l$, where $k_E$ is the rate at which an individual bed particle entrains into motion, and $l$ is the mean length of each downstream step.
+Later, he applied these ideas to calculate the mean downstream flux of many particles \citep{Einstein1950}.
+Einstein reasoned that if the density of resting particles on the bed is $\rho_b$, the overall areal entrainment rate of particles can be written $E = \rho_b k_E$, so the mean downstream sediment flux can be expressed as $\bra q_s \ket = \rho_b \bra u \ket  = El $.
+
+Many researchers have since refined Einstein's approach to provide more realistic descriptions of individual particle motions than Einstein's instantaneous step model.
+One set of efforts has concentrated on particle motions only, calculating the downstream velocity distributions of moving particles using Langevin-type equations to describe turbulent flow and collision forces \citep{Ancey2014,Fan2014,Pierce2021}.
+These models do not yet include transitions between motion and rest.
+Another set of efforts has concentrated on including both motion and rest phases while promoting Einstein's instantaneous steps into finite periods of motion \citep{Lisle1998,Lajeunesse2017,Pierce2020b}, but due to mathematical challenges, these models characterize particle motions by a constant velocity, rather than the fluctuating velocities that real sediment particles exhibit.
+
+Researchers have also refined stochastic formulations of the sediment flux beyond the description of the mean flux provided by Einstein \citep{Furbish2012}.
+Experiments demonstrate that the sediment flux exhibits wide fluctuations due to (1) variations in the number of moving particles and (2) variations in the velocities of moving particles \citep{Ancey2008, Ancey2014}.
+As a result of these sediment tranport fluctuations, measurements of the mean sediment flux depend on the timescale over which they are collected \citep{Saletti2014,Dhont2019,Singh2009}, giving a scale-dependent character to the mean sediment flux.
+To date, very few models have calculated the probability distribution of the bed load sediment flux, and among these, even fewer have described any observation-scale dependence of the sediment flux \citep{Ancey2019}.
+
+This survey reveals two major issues in need of research attention.
+First, we do not yet have the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux probability distribution and the dependence of the moments of this distribution on the observation time.
+Here, we develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a Langevin-type equation of motion. 
+This stochastic equation includes alternation between motion and rest at random intervals, and the motion state includes stochastic forcing that ascribes fluctuating velocities to moving particles.
+Using the probability distribution of particle position generated by this model, we construct a formalism to derive analytically the probability distribution of the sediment flux, and this distribution includes an explicit observation-scale dependence.
+Below, we develop the new formalism in sec.
+\ref{sec:mod}, solve it in sec.
+\ref{sec:res}, and we discuss the implications of our results and future research ideas in secs.
+\ref{sec:disc} and \ref{sec:conc}.
+
+\section{Model development \label{sec:mod}}
+We consider an infinite one-dimensional domain populated with sediment particles on the surface of a sedimentary bed.
+We consider that the flow is weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common.
+The flow is in constrast strong enough so that particles are in motion.
+We label the downstream coordinate as $x$, so that the downstream velocity of a moving particle is $\dot{x}$, and we describe all sediment particles as independent from one another, but governed by the same underlying dynamical equations, meaning we neglect any influence of sediment size or shape or spatial variations in the overlying fluid flow.
+
+\subsection{Dynamical equation for bed load sediment transport}
+From these assumptions, our first target is to write an equation of motion for the individual sediment particle encompassing two features.
+First, particles should alternate between motion and rest.
+The transition rate from rest to motion is called entrainment and occurs with probability per unit time (or rate) $k_E$, while the transition from motion to rest is called deposition and occurs with rate $k_D$.
+Second, particles in motion should move with mean velocity $V$ and some fluctuations around this velocity.
+The simplest equation of motion including these features is
+\be \dot{x}(t) = [V + \sqrt{2D}\xi(t)]\eta(t).  \label{eq:langevin} \ee
+Here $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles, and $\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion, and $\eta=0$, representing rest.
+Here, $V$ is the mean particle velocity, and $D$ is a diffusivity [units $L^2/T$] of moving particles.
+The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$.
+We write $k=k_E+k_D$ as a shorthand.
+
+%We will show later that $D=0$ reproduces the model of Lisle et al \citep{Lisle1998} for the sediment particle alternating between motion and rest with a constant velocity, while the further limit $k_D \rightarrow \infty$ as $V/k_D \rightarrow l$ reproduces the original theory of Einstein \citep{Einstein1937}.
+\subsection{Derivation of the master equation for $P(x,t)$}
+The solution of equation \ref{eq:langevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives the trajectory of a single particle.
+Averaging over the ensemble of all such trajectories from different realizations of the noises will obtain the probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$.
+This distribution, by construction, will generalize earlier models which did not include velocity fluctuations among moving particles \citep{Lisle1998,Lajeunesse2017}.
+
+We form the desired probability distribution of position as $ P(y,t) = \big\bra \delta(y-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
+\ref{eq:langevin} and the average is over both noises, but this symbolic equation is not yet useful as taking these averages directly is a challenging mathematical problem \citep{Hanggi1978}.
+A simpler approach is to conduct the necessary averages in Fourier space. Integrating eq. 1, using its solution in the probability distribution, then Fourier transforming gives
+\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
+Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen1978} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Loginov1978}, in a method similar to \citep{Balakrishnan1993}, provides the Fourier-space master equation
+\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
+and inverse Fourier transforming provides the master equation
+\be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0. \label{eq:master}\ee
+This is a diffusion-like equation governing the probability distribution of position for individual particles as they transport downstream through a sequence of motions and rests, with the movement velocity being a random variable.
+One can see in particular that taking an the entrainment rate $k_E$ very large, meaning that all particles are generally moving, implies an advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations.
+Otherwise, with $k_E$ of similar order as $k_D$, there is a finite probability that the particle is at rest, and the advection-diffusion process is often interrupted by deposition, giving rise to the additional terms in eq.
+\ref{eq:master}.
+\section{Formalism for the downstream sediment flux}
+Now we express the probability distribution of the sediment flux using the probability distribution of particle position $P(x,t)$ provided as the solution of equation \ref{eq:master}.
+We apply a modified version of the approach recently developed by Banerjee and coworkers for \citep{Banerjee2020}.
+The basic idea, as depicted in Figure 1, is that we distribute particles in all states of motion along a domain at random locations at $x<0$, then we calculate using $P(x,t)$ the rate of particle arrival to $x>0$ within the sampling time $T$.
+\begin{figure}
+	\centerline{\includegraphics{./figures/ch2/figure1.png}}
+	\caption{The left panel indicates the configuration for the flux.
+The particle trajectories within are calculated from equation \ref{eq:langevin}, demonstrating alternation between rest and motion with fluctuating velocity.
+Particles begin their transport with positions $-L\leq x \leq 0$ at $t=0$, and as depicted in the right panel, the flux is calculated as the number of particles $N_>(t)$ which lie to the right of $x=0$ at the observation time $t$, divided by $t$: $Q(t) = t^{-1}N_>(t)$.
+We calculate the probability distribution of $Q$ over all realizations of the trajectories and initial positions as $L\rightarrow \infty$}
+	\label{fig:fig1}
+\end{figure}
+
+The rate of particles crossing the surface in an observation time $T$ is
+\be Q(T) = \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T). \ee
+In this equation, $\mathcal{I}_i(T)$ is an indicator function which is $1$ whenever the $i$th particle has passed our control surface and $0$ otherwise.
+All particles which have not crossed the control surface (or which have crossed and then crossed back) contribute nothing to the flux.
+The probability distribution of the flux is then
+\be P(Q|T) = \Big \bra \delta\Big( Q - \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T) \Big) \Big\ket. \ee
+The average is over the initial conditions of each particle and the ensemble of trajectories for each particle.
+Taking the Laplace transform over $Q$ (i.e. forming the characteristic function) obtains
+\begin{align} \tilde{P}(s|T) &= \Bigg\bra \int_0^\infty dQ e^{-s Q}\delta\Big( Q - \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T) \Big)\Bigg\ket \\
+	&=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N \mathcal{I}_i(T)\Big)\Big\ket \\
+	&=  \prod_{i=1}^N \Big\bra \exp\Big(-\frac{s}{T}\mathcal{I}_i(T)\Big)\Big\ket \\
+	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra \mathcal{I}_i(T) \big\ket \Big] \end{align}
+This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $ \mathcal{I}_i(T)$ is either $1$ or $0$ ($e^{ax} = 1-(1-e^a)x$ if $x=0,1$).
+The average over initial conditions and possible trajectories for the $i$th particle can be written
+\be \bra \mathcal{I}_i(t) \ket = \frac{1}{L}\int_L^0 dx' \int_0^\infty dx \mathcal{P}(x,t|x') =  \frac{1}{L}\int_0^L dx' \int_0^\infty dx \mathcal{P}(x,t|-x'), \ee
+where $\mathcal{P}(x,t|x')$ is the probability density that the particle is found at position $x$ at time $t$ given it was initially at $x'$ at time $0$.
+This is the part of the flux that depends on the particle dynamics (ie instantenous velocities and entrainment/deposition characteristics).
+
+Inserting (7) into (6) and taking the limit as $L\rightarrow \infty$ and $N \rightarrow \infty$ while the density of particles $\rho = N/L$ remains constant provides
+\be \tilde{P}(s|T) = \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\mu(T) \Big)^N = \exp \Big[ -\big(1-e^{-s/T}\big)\mu(T) \Big].\ee
+where $\mu(T) = \rho \int_0^\infty dx \int_0^\infty dx' \mathcal{P}(x,T|x')$ is a rate constant which encodes the particle dynamics.
+This expression is the characteristic function of a Poisson distribution.
+Expanding in $e^{-s/T}$ and inverting the Laplace transform provides the distribution of the flux
+\be P(Q|T) = \sum_{k=0}^\infty \frac{\mu(T)^k}{k!}e^{-\mu(T)}\delta(Q-\frac{k}{T}).\ee
+This equation implies that the mean flux is $\bra Q \ket(T) = \int_0^\infty dQ Q P(Q|T) = \mu(T)/T$, and similarly the variance is $\sigma_Q^2(T) = \mu(T)/T^2$.
+The conclusion is that if the flux is considered as a time averaged number of particles crossing a control surface, the mean flux is always Poissonian no matter how particles move, provided they do not interact with one another.
+
+\section{Results \label{sec:res}}
+\subsection{Derivation of the position probability distribution and its moments}
+\begin{figure}
+	\centerline{\includegraphics{figures/ch2/figure2slopeKey.pdf}}
+	\caption{The left panel shows the probability distribution of position evolving through time.
+From the initial state, which is a mixture of moving and resting particles, the distribution splits at short times into contributions from Delta function-like stationary particles and Normal-like moving particles.
+The right panel demonstrates the resulting spreading characteristics of particles.
+This short-time splitting noted in the left panel gives rise to ballistic diffusion at short timescales, followed by normal diffusion, as exemplified by equation \ref{eq:var}.}
+	\label{fig:fig1}
+\end{figure}
+\be P(x,0) = \delta(x) \ee
+\be \pt P(x,0) = -\frac{Vk_E}{k}\delta'(x) \ee
+These initial conditions come from the initial state
+\be P(x,0) = \lim_{t\rightarrow 0 } \frac{k_E}{k} \delta(x-Vt) + \frac{k_D}{k}\delta(x).\ee
+
+\be \bar{\tilde{P}}(g,s) = \frac{s + k + Dg^2 - i g V k_D/k}{s(s+k)+(Dg^2-igV)(s+k_E)}. \ee
+\be \tilde{P}(x,s) = \frac{-D\px^2 + Vk_D/k\px + s + k}{VR(s+k_E)}\exp\Big[\frac{Vx}{2D} - \frac{V|x|}{2D}R \Big]. \label{eq:laplace}\ee
+\be R = \sqrt{1 + \frac{4D}{V^2}\frac{s(s+k)}{s+k_E}}\ee
+\begin{multline} P(x,t) = \big[-D\px^2 + Vk_D/k \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du \end{multline}
+The mean is $ \bra x \ket = k_E V t/k$.
+The variance is 
+\be \sigma_x^2 = 2\Big[ \frac{k_EV^2 k_D}{k^3} + \frac{k_ED}{k}\Big]\Big( \frac{1}{k}e^{-kt} - \frac{1}{k} + t \Big) \label{eq:var}\ee
+\subsection{Calculation of the flux}
+\be \mu(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t).\ee
+
+Taking the Laplace transform,
+\be \tilde{\mu}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s).\ee
+
+\begin{multline} 
+	\mu(t) = \rho \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
+	\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{1}{2}\Big)e^{-V^2 u/4D} + \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du.
+\end{multline}
+\subsection{Connection to earlier work}
+\section{Discussion \label{sec:disc}}
+\subsection{The role of stochasticity in landscape evolution}
+\subsection{Methods to calculate the sediment flux}
+\subsection{Outlook and future research}
+\section{Conclusion \label{sec:conc}}

### 2021-06-24 ch3bedEle.tex
got another component in there.. ch 2 . Now to add the langevin paper and get all figures working.

--- 2021-06-24 ch2flipFlop/jgr/rev-one.tex
@@
-\documentclass[draft]{agujournal2018}
-\journalname{JGR: Earth Surface}
+%%!TEX root = diss.tex
 
-\usepackage{apacite}
-\usepackage{amsmath,amssymb,amsfonts,amsthm}
-\usepackage{comment}
-\usepackage{wrapfig}
-\usepackage{lipsum}
-\usepackage{booktabs} % for wrapping text around tabulars in 
-\usepackage{ragged2e}
-\usepackage{lineno}
-\usepackage[utf8]{inputenc}
+\chapter{Analysis of the bed elevation}
+\label{ch:downDiff}
 
-% reduce white space after tables
-%\usepackage[belowskip=-15pt,aboveskip=0pt]{caption}
-\setlength{\intextsep}{10pt plus 2pt minus 2pt}
-\setlength{\textfloatsep}{30pt}
-
-\raggedbottom
-
-\justifying
-\linenumbers
-
-% there is a bug with align environments in conjunction with lineno
-% the patch is 
-\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
-	\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
-	\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
-	\renewenvironment{#1}%
-	{\linenomath\csname old#1\endcsname}%
-	{\csname oldend#1\endcsname\endlinenomath}}% 
-\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
-	\patchAmsMathEnvironmentForLineno{#1}%
-	\patchAmsMathEnvironmentForLineno{#1*}}%
-\AtBeginDocument{%
-	\patchBothAmsMathEnvironmentsForLineno{equation}%
-	\patchBothAmsMathEnvironmentsForLineno{align}%
-	\patchBothAmsMathEnvironmentsForLineno{flalign}%
-	\patchBothAmsMathEnvironmentsForLineno{alignat}%
-	\patchBothAmsMathEnvironmentsForLineno{gather}%
-	\patchBothAmsMathEnvironmentsForLineno{multline}%
-}
-% from https://tex.stackexchange.com/questions/43648/why-doesnt-lineno-number-a-paragraph-when-it-is-followed-by-an-align-equation?noredirect=1&lq=1
-
-\begin{document}
-
-\title{Joint stochastic bedload transport and bed elevation model: variance regulation and power-law rests}
-\authors{J. Kevin Pierce\affil{1}\thanks{Vancouver, British Columbia, Canada}, and Marwan A. Hassan\affil{1}}
-\affiliation{1}{Department of Geography, University of British Columbia}
-\correspondingauthor{James K. Pierce}{kpierce@alumni.ubc.ca}
-
-\begin{keypoints}
-\item Two species stochastic population model describes fluvial bedload particle activity and local bed elevations
-\item Model predicts elevation dependent transport statistics and power-law distributed resting times for sediment undergoing burial
-\item Results imply elevation changes shift particle activity fluctuations by up to 90\% and sediment rests can generate anomalous bedload diffusion
-\end{keypoints}
-
-\begin{abstract}
-We describe the joint dynamics of bedload transport and bed elevation changes with a stochastic population model, and we analyze (1) the dependence of bedload flux statistics on local bed elevations and (2) resting time distributions for sediment undergoing burial in the fluctuating sedimentary bed.
-The model exhibits suppression of the statistical moments of the bedload flux the bed is in a locally aggraded state and enhancement of the moments when the bed is in a degraded state.
-This variance regulation effect is contingent on collective entrainment, whereby moving grains destabilize stationary grains in a positive feedback. 
-When collective entrainment is turned off, the variance regulation disappears.
-Return times from above in the bed elevation time-series provide heavy tailed power-law distributions of resting times with tail behavior characterized by the mean erosion rate and the active layer depth.
-These results imply bedload statistics measurements on relatively short timescales can be strongly biased by bed elevation changes, and they support the growing consensus that sediment burial generates heavy tailed sediment resting times that ultimately generate anomalous bedload diffusion.
-\end{abstract}
-
-\section{Introduction}
 The transport characteristics of coarse grains moving under a turbulent flow ultimately control a wide set processes within rivers, including the export of contaminants \citep{Malmon2005,Macklin2006}, the success of ecological restoration efforts \citep{Gaeuman2017}, and the response of channel morphology to disturbances \citep{Hassan2017}.
@@
 Essentially two types of bedload transport model have been developed from this concept. 
-The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2018,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Ancey2008,Furbish2012a}.
+The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2017,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Ancey2008,Furbish2012a}.
 In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}.
@@
 As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
-\begin{figure}[!ht]
-	\includegraphics[width=\linewidth,keepaspectratio]{definition.pdf}
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition.pdf}
 	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains.
@@
 A representative subset of these marginal distributions is displayed in figure \ref{fig:pdfs} alongside the approximate results of equations (\ref{eq:ancey}) and (\ref{eq:ou2}).
-\begin{figure}[t]
-	\includegraphics[width=\linewidth,keepaspectratio]{distributions.pdf}
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/distributions.pdf}
 	\caption{Panel (a) presents the probability distribution of particle activity $n$ and panel (b) presents the probability distribution of the relative number of particles $m$ for a representative subset of simulations.
@@
 	\centering
-	\includegraphics[width=0.5\textwidth,keepaspectratio]{variance.pdf}
+	\includegraphics[width=0.5\textwidth,keepaspectratio]{./figures/ch3/variance.pdf}
 	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as posited in equation (\ref{eq:active}): $\sigma_m^2 \approx (l/z_1)^2$. }
@@
 Figure (\ref{fig:condmoms}) displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
-\begin{figure}[!ht]
-	\includegraphics[width=\linewidth,keepaspectratio]{momentsuppression.pdf}
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/momentsuppression.pdf}
 	\caption{The shifts between particle activity moments conditioned on instantaneous elevations and their over-all mean values.
@@
 As we interpolate $f$ between zero and one, the particle activity component of the master equation \ref{eq:master} interpolates from a purely Poissonian model \citep{Ancey2006} to a negative binomial model \citep{Ancey2008}, isolating the imprint of collective entrainment on particle activity statistics over a dynamic sedimentary bed.
-\begin{figure}[!ht]
-	\includegraphics[width=\linewidth,keepaspectratio]{colent-suppression.pdf}
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/colent-suppression.pdf}
 	\caption{The shift of the mean particle activity in panel (a) and its fluctuations in panel (b) with departures of the bed elevation from its mean.
@@
 
-\begin{figure}[!ht]
-	\includegraphics[width=\linewidth,keepaspectratio]{rtcdf.pdf}
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/rtcdf.pdf}
 	\caption{Resting time statistics scale differently with transport conditions and the bed elevation variance.
@@
 Next steps are to generalize our model to a multi-cell framework and to study spatial correlations in bed elevation and particle activity statistics.
-
-\acknowledgments
-The simulation code is available at
-\sloppy
-\url{https://drive.google.com/file/d/1bWEX_ZHlcGRKCxtRIadnSbhh4EpQlsoh/view?usp=sharing}.
-This research was supported by an NSERC Discovery Grant to M. Hassan.
-We would like to thank Shawn Chartrand and Conor McDowell for helpful discussions.
-
-\bibliography{biblio}
-\end{document}

### 2021-06-25 abs.tex
eh

@@
 
-Write an abstract.
+A central task within Earth science is to understand and predict the evolution of landscapes due to flowing water.
+As rainfall channelizes and flows downhill, it carves out basins and etches in networks of intersecting channels.
+In the highlands of these networks, water arranges boulders and gravels into an intricate array of patterns -- steps, pools, bars, and riffles, which cooperate with vegetation both live and dead to set the stage for much of Earth's biota.
+In the lowlands, channels laden with old mountain sediments, weathered and abraded to fine particles, splay across floodplains and drift between configurations through millennia.
+
+Sediment transport is a main driver of all fluvial dynamics.
+Channel evolution ultimately occurs because individual sediment grains move from one location to another.
+Yet in a majority of modelling studies, landscapes are represented as continua, where the locations of individual grains are averaged away, and sediment transport is represented as a steady stream of mass, rather than the intermittent movements of individual grains
+Useful as this continuum approach may be, many fluvial phenomena are not well-suited for it.
+Sediment transport rates are known to fluctuate widely through space and time, and these fluctuations are understood to initiate bedform development and control channel widths.
+The largest grains in small mountain channels are known to confer the most stability, while violating any assumption of being small compared to the scales of interest.
+To understand fluvial dynamics, we need the capability to model discrete grains, not just continua.
+
+In this thesis, I present four or five years of my theoretical research into the movements of individual sediment grains in river channels.
+This sub-field of Earth science, with a focus on grain-scale process, has existed at least since Hans Albert Einstein in 1937, and within it, modellers have traditionally compromised on severe approximations to balance realism against mathematical difficulty.
+In this enterprise, perfectly flat beds which do not change shape, spherical grains, infinite movement velocities, and turbulence free flows have been typical assumptions to make progress, even though they have little basis in reality.
+The research presented here takes on more mathematical difficulty than before to introduce more realism, adding some bricks to the fortress, and introducing new methods to the field which should work well for other people to make more bricks later.
+I hope you find it useful!
+Happy reading.

### 2021-06-25 ack.tex
eh

@@
 \chapter{Acknowledgments}
-
 Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way.
@@
 .
+
+The people whose benevolence spans generations: both those I know: Joe, Joyce, Ama, Ivory Nichols, and Parry Steele, and those I was unlucky to miss.
+Maragret  Squire Avis and Oddessa Mathis.
+Joe, Joyce, Ama, and Parry Steele.
+My
 
@@
 My partner Mary for putting up with the late nights.
+
+Mike church and olav slaymaker for the inspiring seminars.
+
+marwan hassan for being one of the most emotionally intelligent people I've ever met.

### 2021-06-25 appendix.tex
eh

--- 2021-06-24 ch3bedEle.tex
@@
 
-\chapter{Analysis of the bed elevation}
-\label{ch:downDiff}
-
-The transport characteristics of coarse grains moving under a turbulent flow ultimately control a wide set processes within rivers, including the export of contaminants \citep{Malmon2005,Macklin2006}, the success of ecological restoration efforts \citep{Gaeuman2017}, and the response of channel morphology to disturbances \citep{Hassan2017}.
-Although the displacements of individual grains are certainly a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}.
-In response, investigators have developed a stochastic concept of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
-
-Essentially two types of bedload transport model have been developed from this concept. 
-The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2017,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Ancey2008,Furbish2012a}.
-In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}.
-Differences between the random-walk motions of one grain and the next imply a spreading apart of tracer grains as they transport downstream: bedload tracers undergo diffusion.
-
-Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
-Resting times have puzzled researchers because early experiments show exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments show heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016, Bradley2017, Liu2019}.
-A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}; this hypothesis permits surface grains to retain exponential resting times.
-Conceptually, when grains rest on the surface, material transported from upstream can deposit on top of them, preventing entrainment until its removal, driving up resting times and imparting a heavy tail to the distribution.
-To our knowledge, \citet{Martin2014} have provided the only direct support for this hypothesis by tracking grains through complete cycles of burial and exhumation using a narrow flume with glass walls.
-They observed heavy-tailed resting times of buried grains and described their results with a mathematical model similar to an earlier effort by \citet{Voepel2013}.
-Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation time-series \citep{Redner2007}.
-Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, and neither model explicitly includes bedload transport.
-Models of sedimentary bed evolution incorporating sediment transport processes might enhance understanding of sediment resting times.
-
-The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
-These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015a,Ma2014}.
-A recent surge in research activity has generated rapid progress and spawned many new inquiries in this subject. 
-For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014,Ma2014, Lee2018}.
-While they deemed this feedback necessary to model realistic activity fluctuations, the implications of this collective entrainment term on bed topography and particle activity changes has not been fully explored.
-
-In this work, we present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
-Our model extends the \citet{Ancey2008} model to describe the interplay between bedload flux and bed elevation fluctuations in a control volume.
-This development permits a systematic study of the repercussions of collective entrainment, and it frames bed elevation changes as a direct consequence of the sediment transport process.
-Our model has two key assumptions: (1) bedload erosion and deposition can be characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and (2) these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
-We study statistical characteristics of bedload transport, bed elevation, and resting times of sediment undergoing burial using a mixture of numerical simulations and analytical approximations.
-We introduce the stochastic model in section \ref{sec:model}, and we solve it in section \ref{sec:solution} with a mixture of numerical and analytical techniques.
-We discern several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates and the local bed elevation.
-We present these features in section \ref{sec:results}.
-We conclude with the implications of our results and speculate on topics for future research in sections \ref{sec:discussion} and \ref{sec:conclusion}.
-
-\section{Stochastic model of bedload transport and bed elevations}
-\label{sec:model}
-
-We prescribe a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:definition}.
-We define $m$ relative to the mean number of grains within the control volume, so that it can be either positive or negative.
-$n$ is always a positive integer including $0$.
-For simplicity, we consider all particles as approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
-Following \citet{Ancey2008}, we prescribe four events that can occur at any instant to modify the populations $n$ and $m$, and we characterize these events using probabilities per unit time (rates).
-These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
-The four events are depicted as arrows in figure \ref{fig:definition}.
-As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
-\begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition.pdf}
-	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains.
-Migration, entrainment, and deposition are represented by arrows, and the instantaneous bed elevation is depicted by dotted lines.
-The bed is displayed in a degraded state, where $m<0$.
-The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using (\ref{eq:ele}).}
-	\label{fig:definition}
-\end{figure}
-The populations $n$ and $m$ provide the bulk bedload flux $q_s$ and the local bed elevation $z$.
-The mean bedload transport rate is given by $q_s = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
-\sum_{n,m}nP(n,m) $ is the mean number of grains in motion \citep{Charru2004, Ancey2008, Furbish2012a}.
-The bed elevation is related to $m$ through the packing geometry of the bed.
-To quantify this, we introduce a packing fraction $\phi$ of grains in the bed \citep{Bennett1972}, and for simplicity we consider the bed as two-dimensional \citep{Einstein1950, Paintal1971}.
-The deviation from the mean bed elevation is then
-\begin{equation} z(m) = \frac{\pi a^2}{\phi L}m = z_1 m. \label{eq:ele}\end{equation}
-The constant $z_1 = \pi a^2/(\phi L)$ is an important scale of the problem. 
-$z_1$ is the magnitude of bed elevation change in an average sense across the control volume associated with the addition or removal of a single grain.
-
-Bed elevation changes modify the likelihood of entrainment and deposition in a negative feedback \citep{Sawai1987, Wong2007}; that is, aggradation increases the likelihood of entrainment, while degradation increases the likelihood of deposition.
-\citet{Wong2007} concluded that bed elevation changes induce an exponential variation in entrainment and deposition probabilities, while \citet{Sawai1987} concluded that the variation is linear.
-For simplicity, we incorporate the scaling of \citet{Sawai1987} and note its equivalence to the \citet{Wong2007} scaling when bed elevation changes are small.
-Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, we expect the erosion and deposition feedbacks to have the same strength.
-That is, as bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.
-Merging these ideas with those of \citet{Ancey2008}, we write the four possible transitions with local bed elevation-dependent entrainment and deposition rates as
-\begin{align}
-	&R_{MI}(n+1|n) = \nu && &\text{migration in}, \label{eq:rate1}\\
-	&R_E(n+1,m-1|n,m)=(\lambda + \mu n)[1 + \kappa m], && &\text{entrainment}, \label{eq:rate2}\\
-	&R_D(n-1,m+1|n,m)=\sigma n [1- \kappa m ], && &\text{deposition}. \label{eq:rate3} \\
-	&R_{MO}(n-1|n) =\gamma n && &\text{migration out} \label{eq:rate4}.
-\end{align}
-In equations (\ref{eq:rate2}) and (\ref{eq:rate3}), $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
-$\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
-At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
-Away from this elevation, entrainment and deposition are alternatively suppressed and enhanced depending on the sign of $m$, constituting a feedback between bed elevation changes and erosion and deposition.
-We refer to $\kappa$ as a coupling constant since it controls the strength of this feedback. We later demonstrate the relationship
-\begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
-\end{equation}
-where $l$ is a characteristic length scale of bed elevation change that we interpret as the active layer depth \citep{Wong2007,Church2017}.
-All four rates are independent of the past history of the populations and depend only on the current populations $(n,m)$. 
-As a result, the model is Markovian \citep{Cox1965, VanKampen1992}, meaning time intervals between any two subsequent transitions are exponentially distributed \citep{Gillespie2007}.
-
-We write the master equation for the probability flow using the forward Kolmogorov equation $\partial P(n,m;t)/\partial t = 
-\sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1992, Ancey2008} as 
-\begin{multline}
-	\frac{\partial P}{\partial t}(n,m;t) =  
-	\nu P(n-1,m;t) + [\lambda(m+1) + \mu(n-1)][1+\kappa(m+1)]P(n-1,m+1;t)\\  
-	+ \sigma(n+1)[1-\kappa(m-1)]P(n+1,m-1;t) + \gamma(n+1) P(n+1,m;t) \\
-	- 
-	\{ \nu + \lambda+ \mu n (1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}P(n,m;t).
-	\label{eq:master}
-\end{multline}
-The joint probability distribution $P(n,m;t)$ solving this equation fully characterizes the statistics of $n$ and $m$ -- proxies for the bedload flux and local bed elevation.
-The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
-We anticipate that solutions of (\ref{eq:master}) will adjust from the initial conditions to a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium conditions.
-Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
-This Master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
-In our context, the two populations are the moving and stationary grains in the volume.
-
-\section{Model solutions}
-\label{sec:solution}
-
-Unfortunately, equation (\ref{eq:master}) does not appear to admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
-The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
-In response to this difficulty, we resort to numerical methods and analytical approximations, simulating equation (\ref{eq:master}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1992, Gillespie2007} and solving it approximately with mean field and Fokker-Planck approaches \citep{Haken1983,Gardiner1983}.
-The simulation algorithm is described in \ref{sec:numerical}, and analytical approximations are described in \ref{sec:analytical}.
-
-\subsection{Numerical simulations}
-\label{sec:numerical}
-
-The Gillespie algorithm leverages the defining property of a Markov process: when transition rates are independent of history, time intervals between transitions are \begin{wraptable}{l}{0.5\textwidth}
-	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ from \citet{Ancey2008}.
-Units are $s^{-1}$ (probability/time).
-In our model, bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
-	\begin{tabular}{cccccc} \\ 
-		\toprule  
-		flow & $\nu$ & $\lambda$ & $\mu$ & $\sigma$ & $\gamma$ \\
-		\midrule
-		(a) & 5.45  & 6.59  & 3.74 & 4.67 & 0.77 \\
-		\midrule
-		(g) & 7.74  & 8.42  & 4.34 & 4.95 & 0.56 \\
-		\midrule
-		(i) & 15.56 & 22.07 & 3.56 & 4.52 & 0.68 \\
-		\midrule
-		(l) & 15.52 & 14.64 & 4.32 & 4.77 & 0.48 \\
-		\bottomrule
-	\end{tabular}
-\end{wraptable}exponentially distributed \citep{Cox1965}.
-As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
-then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}).
-The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
-This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
-We provide additional background on the stochastic simulation method in the supplementary material and refer the reader to \citet{Gillespie2007} for more detail.
-
-Using this method, we simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
-These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
-For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, we used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
-Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
-In all simulations, we take the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and we set $L = 22.5$cm and $a = 0.3$cm in accord with the \citet{Ancey2008} experiments.
-Each simulation was run for $250$ hours of virtual time, a period selected to ensure neat convergence of particle activity and bed elevation statistics.
-
-\subsection{Approximate solutions}
-\label{sec:analytical}
-
-We approximately decouple the $n$ and $m$ dynamics in equation (\ref{eq:master}) using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
-These inequalites mean many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
-We concentrate on steady state conditions $\partial P/\partial t = 0$ and introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to equation (\ref{eq:master}), with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$.
-This provides the steady state equation
-\begin{multline}
-	0 =  
-	\nu A(n-1|m)M(m) + [\lambda + \mu(n-1)][1+\kappa(m+1)]A(n-1|m+1)M(m+1)\\  
-	+ \sigma(n+1)[1-\kappa(m-1)]A(n+1,m-1)M(m-1) + \gamma(n+1)A(n+1|m)M(m) \\
-	- 
-	\{ \nu + [\lambda+ \mu n ](1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}A(n,m)M(m).
-	\label{eq:decomp}
-\end{multline}
-Summing this equation over $n$ provides a still exact description of the distribution of bed elevations $M(m)$ in terms of the conditional mean particle activity $\langle n | m \rangle = \sum_{n}nA(n|m)$:
-\begin{multline}
-	0 =  [\lambda + \mu\langle n | m+1\rangle][1+\kappa(m+1)]M(m+1)
-	\\+ \sigma\langle n | m-1\rangle[1-\kappa(m-1)]M(m-1) \\
-	- 
-	\{  [\lambda+ \mu \langle n | m\rangle ](1+\kappa m) +  \sigma  \langle n | m\rangle( 1- \kappa m) \}M(m).
-	\label{eq:approxele}
-\end{multline}
-Unforunately, these two equations are no easier to solve than the original master equation, since the coupling between $n$ and $m$ is not reduced in equation (\ref{eq:decomp}).
-
-The simplest approximation to these equations holds that $\kappa$ is so small that the dynamics of $n$ are totally independent of $m$: $A(n|m) = A(n)$.
-Taking this limit in equation (\ref{eq:decomp}), summing over $m$, and using $\langle m \rangle = 0$ reproduces the \citet{Ancey2008} particle activity model.
-As shown by \citet{Ancey2008}, this has solution
-\begin{equation} A(n) = \frac{\Gamma(r+n)}{\Gamma(r)n!}p^r(1-p)^n.\label{eq:ancey}\end{equation}
-which is a negative binomial distribution for the particle activity with parameters $r=(\nu+\lambda)/\mu$ and $p=1-\mu/(\sigma+\gamma).$
-This result implies $\langle n | m \rangle = \langle n \rangle$, so with the definitions of $E$ and $D$ and the equilibrium condition $E=D$, equation (\ref{eq:approxele}) provides
-\begin{equation}0 \approx [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1)-2M(m). \label{eq:ou} \end{equation}
-This mean field equation matches the discrete Ornstein-Uhlenbeck model of bed elevation changes developed by \citet{Martin2014}.
-We summarize that the independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model we present in a mean field approximation when $\kappa$ is insignificant.
-
-In the supplementary information we show the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within equation (\ref{eq:ou}) provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
-As we will demonstrate in section \ref{sec:results}, and as we have already suggested with equation (\ref{eq:active}), this is a poor approximation to the bed elevation variance.
-Nevertheless, this approximation does capture the Gaussian shape of the bed elevation distribution.
-The essential issue with this mean field approach is that the conditional mean particle activity $\langle n | m \rangle$ varies significantly with $m$ in actuality, especially when collective entrainment contributes to the mean entrainment rate $E$.
-We will discuss these points subsequently when developing more refined approximations and presenting numerical results.
-
-A more careful approximate solution to equation (\ref{eq:approxele}) can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into equation (\ref{eq:approxele}) in order to close the equation for $m$ without solving equation (\ref{eq:decomp}).
-From numerical simulations we determine that
-\begin{equation}
-	\langle n | m  \rangle \approx \langle n \rangle \Big( 1 - \frac{2\kappa m}{1-\mu/\sigma}\Big) \label{eq:closure}
-\end{equation}
-captures the general features of the conditional mean particle activity.
-As we show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
-\begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
-representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in equation (\ref{eq:active}).
-$M_0$ is a normalization constant. 
-As we will demonstrate, this closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:master}), at least for the flow parameters in table \ref{tab:anceyparams}.
-
-\section{Results}
-\label{sec:results}
-
-From the initial conditions, all simulations show a rapid attainment of steady-state stochastic dynamics of $n$ and $m$ which support a time-independent joint distribution $P(n,m)$.
-We show an elevation time-series in the bottom panel of figure \ref{fig:definition}.
-In order to describe the implications of coupling bedload transport to bed elevation changes, we present the numerical and analytical results for the probability distributions of bedload transport and bed elevations in section \ref{sec:pdf} and the statistical moments of these quantities in section \ref{sec:mom}.
-We isolate the effects of collective entrainment on bed elevation changes in section \ref{sec:colent}, and we present the resting times of sediment undergoing burial in section \ref{sec:rtcdf}.
-
-\subsection{Probability distributions of bedload transport and bed elevations}
-\label{sec:pdf}
-
-We compute this joint distribution by counting occurrences of the states $(n,m)$ in the simulated time series.
-From this joint distribution we compute marginal distributions $P(n)$ and $P(m)$ by summing over $m$ and $n$ respectively.
-A representative subset of these marginal distributions is displayed in figure \ref{fig:pdfs} alongside the approximate results of equations (\ref{eq:ancey}) and (\ref{eq:ou2}).
-\begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/distributions.pdf}
-	\caption{Panel (a) presents the probability distribution of particle activity $n$ and panel (b) presents the probability distribution of the relative number of particles $m$ for a representative subset of simulations.
-These distributions represent different flows from table \ref{tab:anceyparams}, distinguished by color, and different values of the active layer depth $l$ (equivalently the coupling constant $\kappa$), distinguished by the marker style.
-The mean field theories (mft) of equations (\ref{eq:ancey}) and (\ref{eq:ou2}) are displayed as solid black lines.}
-	\label{fig:pdfs}
-\end{figure}
-The mean field equation (\ref{eq:ancey}) for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match equation (\ref{eq:ou2}), having Gaussian profiles consistent with our assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.
-
-\subsection{Statistical moments}
-\label{sec:mom}
-
-We calculate the moments of $n$ and $m$ by summing over $P(n,m)$. 
-The $j$th order unconditional moment of the particle activity $n$ derives from
-\begin{equation} \langle n^j \rangle = \sum_{n}n^jP(n),\end{equation}
-while the $j$th order moment of $n$ held conditional on $m$ is
-\begin{equation} \langle n^j|m \rangle = \sum_{n}n^j P(n,m) .\end{equation}
-\begin{wrapfigure}{r}{0.5\textwidth}
-	\centering
-	\includegraphics[width=0.5\textwidth,keepaspectratio]{./figures/ch3/variance.pdf}
-	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as posited in equation (\ref{eq:active}): $\sigma_m^2 \approx (l/z_1)^2$. }
-	\label{fig:var}
-\end{wrapfigure}
-We observe no dependence of the moments of $m$ on the value of $n$. 
-The mean elevation is always $\langle m \rangle = 0 $ due to our initial assumption of symmetry in the entrainment and deposition rate scaling with $m$. 
-Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in equation (\ref{eq:ou2}); this result supports our earlier assertion that $l$ is a characteristic length scale of bed elevation fluctuations.
-The close correspondence between the mean field approximation and the numerical simulations in figure (\ref{fig:pdfs}a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
-We find them to be identical within numerical uncertainty.
-
-The coupling between bed elevation changes and the erosion and deposition rates develop a strong dependence of the particle activity on $m$.
-Figure (\ref{fig:condmoms}) displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
-\begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/momentsuppression.pdf}
-	\caption{The shifts between particle activity moments conditioned on instantaneous elevations and their over-all mean values.
-Panel (a) indicates the mean particle activity shift versus the bed elevation measured in units of $\sigma_z=l$.
-This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$.
-The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
-These results indicate that bedload statistics measurements on short timesclaes could be severely biased by departures from the mean bed elevation.}
-	\label{fig:condmoms}
-\end{figure}
-Figure (\ref{fig:condmoms}a) demonstrates that the \citet{Ancey2008} flow conditions support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
-The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry in this quantity.
-Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
-For the parameters of the \citet{Ancey2008} experiments, figure (\ref{fig:condmoms}b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
-We summarize that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
-
-\subsection{Collective entrainment and bedload activity fluctuations}
-\label{sec:colent}
-Noting that bed elevations regulate the particle activity moments, we now study the influence of collective entrainment on this effect by modifying the relative proportion of the individual to collective contributions in the mean entrainment rate $E=\lambda + \sigma \langle n \rangle $.
-Using the equilibrium condition $E=D$, we determine the fraction of entrainment due to the collective process is $f = \mu\langle n \rangle/E = \mu/\sigma$.
-Using this fraction, we can hold $E$ constant and modify the prevalence of the collective entrainment process by setting $\lambda = E(1-f)$ and $\mu= \sigma f$.
-As we interpolate $f$ between zero and one, the particle activity component of the master equation \ref{eq:master} interpolates from a purely Poissonian model \citep{Ancey2006} to a negative binomial model \citep{Ancey2008}, isolating the imprint of collective entrainment on particle activity statistics over a dynamic sedimentary bed.
-\begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/colent-suppression.pdf}
-	\caption{The shift of the mean particle activity in panel (a) and its fluctuations in panel (b) with departures of the bed elevation from its mean.
-All simulations are at flow condition (g) from table \ref{tab:anceyparams} except $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment.
-Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (\ref{eq:ancey}) at large departures from the mean bed elevation.
-Panel (b) shows particle activity fluctuations suppressed by 90\% when $z\approx -3l$ and collective entrainment is the dominant process.
-When collective entrainment is absent, meaning $\mu/\sigma=0$, this moment regulation effect vanishes: it is a consequence of collective entrainment.}
-	\label{fig:colent}
-\end{figure}
-Figure \ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed.
-When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect we introduced in section \ref{sec:mom}.
-
-\subsection{Resting times of sediment undergoing burial}
-\label{sec:rtcdf}
-
-\begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/rtcdf.pdf}
-	\caption{Resting time statistics scale differently with transport conditions and the bed elevation variance.
-Panel (a) shows differing flow conditions at a fixed $l$ value, while panel (c) shows fixed flow conditions at differing $l$.
-When scaled by $T_0$ (\ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d).
-In panels (b) and (d), the black dotted lines indicate a power law decay of the collapsed tails having parameter $\alpha\approx1.18$ .}
-	\label{fig:cdfs}
-\end{figure}
-
-Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the time-series of $m$ \citep[e.g.,][]{Redner2007}.
-Following \citet{Voepel2013} and \citet{Martin2014}, we concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
-These are the return times from above of the sedimentary bed conditional to the elevation $m'$.
-Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
-Using the marginal probability distribution of bed elevations $P(m)$ (figure \ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}:
-\begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
-A representative subset of these results are displayed in figure \ref{fig:cdfs}.
-Comparing panels \ref{fig:cdfs}(a) and \ref{fig:cdfs}(c) shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
-However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic timescale $T_0$ is found to collapse away both variations.
-We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
-This is the mean number of grains leaving the bed per unit time is $E$, and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}). 
-Therefore, bed elevations change with a characteristic speed $v = z_1 E$.
-Since the range of elevation deviations is $l$ (figure \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
-\begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
-When scaling the resting time by this $T_0$, we obtain the collapse shown in figure \ref{fig:cdfs}.
-Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
-These distributions are sufficiently heavy tailed to violate the central limit theorem and drive anomalous super-diffusion of bedload, a result which supports the earlier conclusions of \citet{Voepel2013} and \citet{Martin2014}.
-
-\section{Discussion}
-\label{sec:discussion}
-
-\subsection{Context}
-Einstein developed the first stochastic models of bedload tracer diffusion \citep{Einstein1937} and the bedload flux \citep{Einstein1950}, and his ideas can be viewed as the nexus of an entire paradigm of research that extends into the present day \citep[e.g.,][]{Hubbell1964, Nakagawa1976,Hassan1991,Ancey2008, Wu2019}.
-These models aim to predict bedload transport characteristics from stochastic concepts of individual particle motions.
-With some exceptions \citep{Yang1971,Nakagawa1980,Pelosi2016,Wu2019,Wu2019a}, existing descriptions are spatially one-dimensional, concentrating on the motion of grains in the downstream direction without including the vertical dimension wherein local bed elevation changes imply sediment burial \citep{Voepel2013,Martin2014} and change the mobility of surface grains \citep{Yang1971,Nakagawa1980}.
-
-\subsection{Contributions}
-
-In this paper, we have built on earlier works \citep{Ancey2008,Martin2014} to include the vertical dimension of bed elevation dynamics, study the interplay between bedload transport and bed elevation fluctuations, and investigate resting time distributions of sediment undergoing burial.
-To our knowledge, this model is the first description of bedload transport and bed elevations as a coupled stochastic population model based on individual grains.
-Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
-Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015a} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.
-
-Our inclusion of coupling between the bed elevation and entrainment and deposition rates revealed a novel dependence of particle activity on bed elevation changes,  highlighting a new consequence of the collective entrainment process \citep{Ancey2008,Lee2018}.
-This coupling develops a significant variation of the particle activity moments with deviations of the bed from its mean elevation.
-We isolated the role of collective entrainment in this bedload activity regulation, and pointed out that particle activity variations with bed elevations dissapear in the absence of collective entrainment.
-Finally, we obtained resting times for sediment undergoing burial within the sedimentary bed by analyzing return times from above in the bed elevation time-series.
-We found heavy-tailed power law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
-The distribution tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.
-
-As our model builds on earlier works describing particle activity and bed elevation changes independently, it also reduces to these works in simplified limits when the coupling between the particle activity and bed elevation vanishes.
-With the mean field approach in section \ref{sec:analytical}, we derived the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity as simplified limits of our coupled model.
-While the mean field description of bed elevations over-predicts the bed elevation variance by approximately a factor of two, it does capture the Gaussian shape of the bed elevation distribution, and its conclusions on the tail characteristics of resting time distributions for sediment undergoing burial are identical to ours within the numerical uncertainty: \citet{Martin2014} described a power-law distribution with tail parameter $\alpha \approx 1$ which falls neatly within our estimation $\alpha = 1.18 \pm 0.32$.
-In addition to our original contributions, we have corroborated the models of \citet{Ancey2008} and \citet{Martin2014} from an alternate perspective, showing their results to be mostly robust when accounting for bed elevation changes.
-
-\subsection{Next steps}
-
-The model we have presented computes statistical characteristics of the bedload particle activity and bed elevation within a control volume by assuming all particles on the bed surface have similar mobility characteristics while sediment transport and bed topography are in equilibrium.
-In actuality, particles span a range of sizes, and spatial organization occurs both in the forces imparted to particles by the flow \citep{Shih2017,Amir2014} and in the mobility characteristics of particles on the bed surface \citep{Charru2004, Hassan2008, Nelson2014}.
-Together, these factors may generate spatial correlations in particle activities that models concentrating on a single control volume will be unable to capture. 
-Models chaining multiple control volumes together have shown spatial correlations in the particle activity as a result of collective entrainment \citep{Heyman2014, Ancey2015}, and similar approaches have also been applied to study correlations in turbulent flows \citep{Gardiner1983}.
-In light of this work, we consider the model we have presented as a preliminary step toward a multiple-cell model of particle activities and bed elevation changes with potential to express spatial correlations between longitudinal profile and particle activity statistics.
-
-Like \citet{Martin2014}, we obtained heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
-This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016}. 
-Our resting time distributions show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2000} which is within range of our results.
-Divergent mean resting time distributions present a paradox, since they imply all particles should eventually be immobile, violating the equilibrium transport assumption.
-\citet{Voepel2013} demonstrated that a bounded random walk for bed elevations provides a power-law distribution that eventually transitions to a faster thin-tailed decay, allowing for power-law scaling like our result and \citet{Martin2014} without this divergent mean paradox.
-One resolution to this issue could come from a spatially distributed model with multiple cells.
-Neighboring locations might bound excessive local elevation changes through granular relaxations from gradients above the angle of repose.
-In this interpretation, divergent mean power law resting time distributions may be relics of single cell models for bed elevation changes.
-We should always expect a maximum depth to which the bed can degrade relative to neighboring locations; this could temper the power law tail without required the reflecting boundaries used by \citet{Voepel2013}.
-
-Finally, we studied probability distribution functions and first and second moments of the particle activity and bed elevation, making novel conclusions about coordination between the statistical characteristics of these quantites which deserve experimental testing.
-In the last decade, particle tracking experiments have emerged \citep{Lajeunesse2010,Roseberry2012,Martin2014, Fathel2015,Heyman2016,Liu2019}, that allow joint resolution of bed elevations and bedload transport.
-A suitably designed experiment could test our prediction that bed elevations regulate particle activity statistics, as essentially represented in figures \ref{fig:condmoms} and \ref{fig:colent}. 
-However, we have left many other statistical characteristics of bedload transport for future studies. 
-For example, the dependence of bedload transport \citep{Saletti2015a,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic.
-Statistical quantities can either be monoscaling or multiscaling across the observation scale \citep{Sornette2000}, and we currently lack physical understanding and general conclusions about the scale dependence of particle activity and bed elevation signals.
-The model we have presented shows statistical monoscaling for both quantites \citep[e.g.][]{Saletti2015a}, whereas other experiments indicate statistical multiscaling \citep{Aberle2006,Singh2009,Singh2012}.
-We consider this topic to go beyond the scope of the present work, and we have focused on statistical characteristics at the highest temporal resolutions, with no averaging over the observation scale.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-We developed a stochastic model for particle activity and local bed elevations including feedbacks between elevation changes and the erosion and deposition rates.
-This model includes collective entrainment, whereby moving particles tend to destabilize stationary ones.
-We analyzed this model using a mixture of numerical and analytical methods and provided two key results:
-\begin{enumerate}
-	\item Resting times for sediment undergoing burial lie on a heavy-tailed power law distributions with tail parameter $\alpha \approx 1.2$;
-	\item Collective entrainment generates a statistical regulation effect, whereby bed elevation changes modify the mean and variance of the particle activity by as much as 90\%: this effect vanishes when collective entrainment is absent.
-\end{enumerate}
-These results imply measurements of bedload transport statistics could be severely biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
-Next steps are to generalize our model to a multi-cell framework and to study spatial correlations in bed elevation and particle activity statistics.
+\chapter{Mathematical Compendia}

### 2021-06-25 ch5langevin.tex
eh

%%!TEX root = diss.tex

\chapter{Collisional model of sediment velocity distributions}
\label{ch:langevin}
\section{Introduction}

Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
Due to these fluctuations, measured transport rates often show extremely slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude.
These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions.
In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations, quantifying the dependence of measurements on the observation scale, and linking bulk transport characteristics to the ``microscopic" dynamics of individual grains.
Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Boltzman, Lajeunesse2020} and bedform initiation \citep{Ancey2014,Bohorquez2016} provide additional motivation for these stochastic formulations.
Many stochastic approaches express downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume.
In these approaches, the instantaneous velocity distribution of sediment particles therefore becomes an object of crucial importance.
Unfortunately, current understanding of this distribution remains limited, and no mechanistic models have yet been presented that describe the full range of experimental observations. Here, we take some steps toward rectifying these shortcomings.

High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears.
One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}. 
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$. Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Sezilles2014} were viscous.
A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads having higher Stokes numbers ($St \sim 10-500$). In all Gaussian cases, flows are turbulent ($Re>5000$) and super-critical ($Fr>1$).
Two other experiments display velocity distributions that are intermediate between exponential and Gaussian and appear more like a Gamma distribution \citep{Houssais2012, Liu2019}. The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions. They resolved the velocity distributions for larger grains only. The \citet{Liu2019} experiments used uniformly-graded sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, we can summarize that the shape of the velocity distribution does not consistentely relate to whether a flow is super or sub-critical ($Fr$), whether sediment grains are natural (sand, gravel) or synthetic (beads), or whether the flow is laminar or turbulent ($Re$). However, the typical Stokes numbers of particles do seem to increase monotonically from exponential velocity experiments (where $St \sim 1$), to intermediate (Gamma-like) experiments (where $St \sim 10$), and eventually to Gaussian experiments (where $St \sim 10^2$.) Apparently, the shape of streamwise bed load velocity distributions depends on the particle size.

Existing models of streamwise bed load velocities can be divided into computational and statistical physics categories.
Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modelling particles as spheres interacting through repulsive forces, and the flow using direct simulation of the Navier-Stokes equations or some related approximation (such as large eddy simulation or the St-Venant equations). When streamwise particle velocities have been analyzed in such simulations, they show exponential tails \citep{Gonzalez2017,Furbish2013} that agree with only a subset of the experimental data.
Statistical physics models have incorporated stochastic driving and resisting terms into the Newtonian dynamics of individual grains to develop a Langevin-like description of bed load particle motions.
For the downstream velocity $u(t)$, \citet{Fan2014} wrote $\dot{u} = F -\gamma \text{sgn}{u} + \xi(t)$ where $F$ represents the steady component of the fluid forcing, the term involving $\gamma$ is a quasi-static (Coulomb-type) friction term representing momentum dissipation by particle-bed collisions, and $\xi(t)$ is a Gaussian white noise representing variability in these forces. This model provides exponential velocity distributions which agree with one subset of experiments. \citet{Ancey2014} took a similar approach that includes different forces, solving $\dot{u} = \gamma(\overline{u}-u) + \xi(t)$. Here the term involving $\gamma$ is similar to a Stokes drag, except it involves the mean sediment velocity $\overline{u}$, not the fluid velocity as for a ``real" Stokes drag. $\xi(t)$ is again a Gaussian white noise representing fluctuations, and the model provides Gaussian velocity distributions that agree with another subset of experiments. 
While these models build insight into grain-scale sediment transport mechanics and provide powerful techniques with which to approach the problem, they have not yet provided a comprehensive explanation for the range of streamwise velocity distributions resolved in experiments.

Here, motivated by the realization that experimental particle velocity distributions vary systematically with the grain size, as summarized above, we hypothesize that the shape of the streamwise velocity distribution is controlled by the momentum dissipation characteristics of particle-bed collisions.
It is well-known that the elasticity of granular collisions within viscous fluids depends on the particle size.
In addition, the velocity distributions of granular gases are known to develop increasingly heavier tails than the Gaussian (Boltzmann) form of elastic gases as the inelastiscity of particle-particle collisions is increased.
Taking inspiration from this established knowledge, we develop below a model for sediment grains in transport as they undergo inelastic particle-bed collisions. Our intentions are to test the hypothesis that particle-bed collision characteristics explain the range of experimentally-observed streamwise bed load particle velocity distributions, and to introduce more realistic forces into earlier statistical physics descriptions of individual bed load particle dynamics. 
We develop the model and explain our assumptions in section \ref{sec:model}, then we present the analytical solution and major results in \ref{sec:results}. We finally discuss the implications of these results, summarize our findings, and suggest ideas for further research in sections \ref{sec:discussion} and \ref{sec:conclusion}.

\section{Model}
\label{sec:model}

Figure \ref{fig:1} indicates the configuration we have in mind. Nearly spherical and cohesionless particles of diameter $d$ and mass $m$ moving as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow. The flow is just strong enough to drive grains into rarefied transport of a kind typical in gravel-bed rivers: particles saltate along the bed in sequences of collisions between events of erosion and deposition; moving particles collide often with stationary particles, but rarely with other moving particles.
Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces. In contrast to the computational physics approach, we do not aim to characterize the exact timeseries of the forces on an individual particle. Instead, we model the ensemble of possible force timeseries that particles could conceivably experience. Each possibility implies a different velocity timeseries $u(t)$ in the downstream direction.
Our objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
We include the most realistic article-bed collision and fluid forces we can while still allowing for analytical solutions.

Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Williams2020}.
The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}. Here, we introduce a restitution-like coefficient $\varepsilon$ as indicated in figure \ref{fig:1}. This ranges from $\varepsilon=0$ for completely inelastic collisions to $\varepsilon=1$ for completely elastic collisions.
If the streamwise velocity just prior to a collision is $u$, just after the collision it becomes $\varepsilon u$. 
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig1Concept.png}}
	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces. During saltation, pre-collisional streamwise velocities $u$ are transformed to postcollisional velocities $\ve u < u$.}
	\label{fig:fig1}
\end{figure}
Since this quantity combines effects of particle shape and collision geometry and should vary from one collision to the next, we consider that the fraction of streamwise momentum dissipated per collision $\varepsilon$ lies on a statistical distribution $\rho(\varepsilon)$.
Similar ideas are available in the granular physics literature \citep{Serrero2015}.
Further assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, we write the collision force in the downstream direction as
\be F_C(u,t) = - m u \sum_{k=1}^{N_\nu(t)}(1-\varepsilon_k)\delta(t-\tau_k). \label{eq:col} \ee
Here, $N_\nu(t)$ is the number of collisions in time $t$, the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are elasticity coefficients characterizing the amount by which each collision slows the particle down.
This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momentum. This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions. These conditions are always satisfied for the idealized saltation-type motion depicted in figure \ref{eq:1}.

Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Rey_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Rey_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
At realistic $\Rey_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Rey_p$ formulas \citep{Schmeeckle2007,Mei2006}.
A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Rey_p$ can be written $F_D = \frac{\pi}{8}
\rho_f d^2 C_D(\Rey_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
In the present model we set $C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1995,Armenio2001,Dalche2015}.

Drag forces have been argued to fluctuate rapidly compared to the inertial response times of coarse sediment grains \citep{Fan2014}.
The magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
Using these ideas, we make two key simplifications of the drag force above.
First, we split the drag $F_D$ into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, we represent drag fluctuations as a Gaussian white noise characterized by a particle diffusivity $D$ \citep{Fan2014,Ancey2014}. 
Defining $\bar{V}$ as a representative slip velocity which we specify more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, we express the fluid forces as
\be F_D(t) = \frac{\pi}{8}
\rho_f d^2 \bar{C}_D \bar{V}^2 + \sqrt{2 D } \eta(t). \ee
In this drag force \ref{eq:drag} and the collision force \ref{eq:col}, the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions. This set of possiblities defines a statistical ensemble.

\subsection{Langevin equation}

With the above forces, we express the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics as
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier stocahstic bed load models with an episodic term which provides a more realistic representation of particle-bed collisions during saltation.
It represents a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulent generates "diffusion".
The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,vandenBroek1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1984,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweiss2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice since the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional.
Given this integration rule, the remaining issues are to obtain the integro-differential equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

\subsection{Chapman-Komogorov equation and particle-bed collision integral}

We derive the equation governing the streamwise velocity distribution $P(u,t)$ from a simple limiting argument in appendix \ref{sec:appendixA}, finding
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + \mathcal{I}_c(u,t). \label{eq:master} \ee
In this equation, we introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$. The term
\be \mathcal{I}_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.
Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Such equations combining are known as Chapman-Komogorov equations \citep{Gardiner1983}. 
Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral of the Boltzmann equation in kinetic theory and granular gases \citep{Duderstadt1971, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) goes to infinity. Mathematically, it represents the probability distribution of the product between $\varepsilon$ and $u$ \citep[c.f.][]{Feller1968}.

Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so we make one further approximation.
We assume the distribution of dissipation coefficients $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$. This allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
\be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
In what follows, we drop all but the first two terms to obtain the leading order contribution of particle-bed collisions to the velocity distribution. Higher orders could always be included later by perturbation theory. We solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$. Scaling the time in Equation \ref{eq:master}, we can see this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions.

\section{Results}
\label{sec:results}

\subsection{Derivation of the velocity distribution}
\label{sec:solution}
Hereafter we drop the prime on the most common streamwise restitution coefficient $\ve'$. With the truncation to two terms, equation \ref{eq:master} gives 
\be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
which is now a non-local ordinary differential equation. Such equations have seen some attention in the mathematics literature, where they are called pantograph equations \citep{Hall1970,Zaidi2015,Bhalekar2017} for their relationship to a current collection device used on electric trains \citep{Ockendon1971}.
In the appendix we solve equation \ref{eq:governer} using Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(-d \lambda_-^2 \ve^{-2m} + \gamma \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in the appendix; they are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors are $K_\pm$ are 
\be K_\pm = d(\lambda_+-\lambda_-)\prod_{l=1}^\infty (-d\lambda_\pm^2 \ve^{2l} + \gamma \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution appears quite complicated, one can verify that this is a normalized probability distribution which has very simple limiting behaviors as the most common dissipation coefficient $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.

It is rather simple to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
The first moment is
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve},\ee
which is scales weakly with the mean fluid drag and sharply with the rate and typical elasticity of collisions.
The second moment is
\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to them. In contrast, velocity fluctuations depend sharply on the parameters representing particle-bed collisions.

Figure \ref{fig:fig2} depicts velocity characteristics for different realizations of the fluid and collisional forces.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig2pdfs.png}}
	\caption{Left panels show velocity realizations as gray traces. Velocities are calculated from Monte Carlo simulations. Individual realizations are singled out as black traces. Particle-bed collisions imply sudden downward-velocity jumps. Flow forces generate fluctuating positive accelerations between collisions. Right panels show simulated histograms of particle velocities and exact solutions from equation \ref{eq:steadystate}.}
	\label{fig:fig2}
\end{figure}
We can see an apparent transition from exponential-like to Gaussian-like velocity distributions as typical collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$). In between, the full distribution \ref{eq:steadystate} resembles a Gamma distribution, although it is not a Gamma distribution.

\subsection{Exponential and Gaussian regimes}
\label{sec:modelcomparison}

In fact, the apparent transition in figure \ref{fig:fig2} can be made rigorous: despite its complex appearance, simple Gaussian and exponential distributions appear as rigorous mathematical limits of equation \ref{eq:steadystate}. When particle-bed collisions are completely inelastic, \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic, \ref{eq:steadystate} becomes Gaussian. Figure \ref{fig:fig3} demonstrates more closely the approach of the distribution toward these limits.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig3asymptotic.png}}
	\caption{The particle velocity distribution approaches an exponential distribution in (a) as particle-bed collisions become extremely elastic ($\ve \rightarrow 1$), and it approaches a Gaussian in (b) as they become extremely inelastic ($\ve \rightarrow 0$). On the abscissa, the mean sediment velocity is standardized by its mean $\bar{u}$ and standard deviation $\sigma_u$. }
	\label{fig:fig3}
\end{figure}

The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is rather easy to see. Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for that with $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to notational differences:
\be P(u) = \frac{d}{\sqrt{\gamma^2 + 4d}}e^{\frac{\gamma u }{2 d} - \frac{\sqrt{\gamma^2 + 4 d}|u|}{2d}}. \ee
Thus, for bed load transport conditions with typically very inelastic particle-bed collisions, we can expect exponential-like velocities and large deviations from a Gaussian behavior.

The Gaussian limit as $\ve \rightarrow 1$ of \ref{eq:steadystate} is more difficult to evaluate. The challenge is that the statistical moments \ref{eq:mean} and \ref{eq:2ndmom} diverge at the same time as the denominator factors of the distribution \ref{eq:steadystate}. In appendix \ref{sec:appendixB} we return instead to the original equation \ref{eq:governer} to evaluate this elastic limit, obtaining
\be P(u) = \frac{1}{\sqrt{2\pi\sigma_u^2}}e^{-\frac{(u-\bar{u})^2}{2\sigma_u^2}}. \label{eq:gaussian}\ee
This result is identical to the velocity distribution derived by \citet{Ancey2014}, up to notation.

\subsection{Comparison with experimental data}
\label{sec:experimentcomparison}

Now we compare the analytical distribution \ref{eq:steadystate} with the available experimental data.
Upfront, we point out that the distribution above has free parameters and this is only a proof of concept that the velocity distribution is capable of fitting the available experimental data; it is not a proof that this is the underlying mechanism for these data blahblahblah that was a good writing day.

\begin{table}
	\begin{center}
		\def~{\hphantom{0}}
		\begin{tabular}{lccc}
			$a/d$  & $M=4$   &   $M=8$ & Callan  \\[3pt]
			0.1   & 1.56905 & ~~1.56~ & 1.56904\\
			0.3   & 1.50484 & ~~1.504 & 1.50484\\
			0.55  & 1.39128 & ~~1.391 & 1.39131\\
			0.7   & 1.32281 & ~10.322 & 1.32288\\
			0.913 & 1.34479 & 100.351 & 1.35185\\
		\end{tabular}
		\caption{Values of $kd$ at which trapped modes occur when $\rho(\theta)=a$.}
		\label{tab:kd}
	\end{center}
\end{table}

\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig4expComparison.png}}
	\caption{The features of the four possible modes corresponding to
		(\textit{a}) periodic\protect\\ and (\textit{b}) half-periodic solutions.}
	\label{fig:fig4}
\end{figure}

\section{Discussion}
\label{sec:discussion}

We developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions,
generalizes earlier approaches available in the literature which did not treat episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, we have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise. Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Coleman2019,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are relatively long compared to the timescales of turbulent fluctuations \citep{}.
We modelled particle-bed collision forces as a sequence of instantaneous impulses where the intervals between successive collisions were characterized as exponential random variables. The effect of each collision on the streamwise particle velocities was parameterized by a restitution-like coefficient.
Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the setting here is somewhat different than grains in air. Because particles within a viscous flow interact at a distance, we should expect the collision model will become poor when the time between subsequent collisions becomes small. Therefore, although the model seems appropriate for saltation, it could become questionable for the ``reptation" transport mode when the times between subsequent particle-bed collisions are short. Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport. Unfortunately, numerical methods would likely be required to interpret the resulting equations, in contrast to the analytical approaches applied here.

Sediment transport experiments reveal correlations between particle size and the shape of the bed load velocity distribution.
Experiments with smaller particles tend to give exponential distributions, and those with larger particles give Gaussian distributions.
In fluid dynamics, the dissipation characteristics of particle-particle collisions in viscous flows are known to depend on the particle size and approach velocity through the Stokes number. 
In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic particles generate non-Gaussian distributions.
Taken together, these ideas suggest that we might relate the shape of the particle velocity distribution to particle size.
We can estimate typical Stokes numbers of colliding bed load particles in experiments as $...$, using with the flow shear velocity and mean streamwise sediment velocity to calculate $V$. Estimating in this way, transport experiments with exponential velocities have $St \sim 1-10$, those with neither exponential nor Gaussian velocities have $St \sim 10-100$, and those with Gaussian velocities have $St > 100$.
In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2001,Joseph2001,Yang2006}.
Although collision geometry and grain shape certainly complicate the narrative, these values of $St$ are consistent with our model conclusion that the shape of the particle velocity depends on the elasticity of collisions.

In real channels, grain sizes often span a wide range. A major implication of the dependence of the shape of the particle velocity distribution on grain size is that different grain sizes in a mixture will impart distinct fluctuation signatures to the overall bulk transport rate. Even in the absence of sorting effects and differential mobility, smaller grains can be expected to carry more control over the largest fluctuations in the overall transport rate, since their velocity distributions have wider tails.

For example, in a mixture of small and large grains, neglecting any sorting effects whereby the mobility of small grains is contingent on the mobility of the large grains, small grains will have exponential velocities with relatively wide fluctuations, while large grains will have Gaussian velocities with relatively narrow fluctuations. Considering the over-all flux then as the number of moving particles times their velocities, transport fluctuations

aeolean transport extension maybe
particle size distributions - interesting implications
connection to computational physics approach
connection to the stochastic description of the flux

\section{Conclusion}
We have demonstrated that particle-bed collisions control the shape of the particle velocity distribution.  
\label{sec:conclusion}

\appendix

\section{Derivation of Master Equation}
To derive the master equation from \ref{eq:langevin}, we temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 d} h$ with $h$ distributed as $f(h)$. We will later take a GWN limit on this noise. With this assumption, integrating \ref{eq:langevin} over a small time interval $\delta t$ considering the Ito interpretation for the collision term provides
\be     
u(t+\delta t) =
\begin{cases}
	u(t) + \gamma \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
	u(t) + \sqrt{2 d} h &  \text{with probability } r \delta t \\
	\ve u(t) &  \text{with probability } \nu \delta t
\end{cases}
\ee

Considering the probability $P(u,t+\delta)$ as sum over possible paths from $P(u,t)$ develops 
\begin{align} P(u,t+\delta t) =
	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\gamma \delta t) P(w,t)\\ 
	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 d}h) P(w,t) \\ 
	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
\end{align}
Evaluating all integrals over $\delta$-functions provides 
\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanKampen1983}. This process finally obtains the master equation (\ref{eq:master}).
\section{Derivation of Steady-state solution}

Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
where
\be q(z) = d z^2 - i \gamma z + 1 . \label{eq:q} \ee
Recursing $N+1$ times provides
\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
\be \lambda_\pm = \frac{\gamma}{2d}\Big[1 \pm \sqrt{ 1 + 4d/\gamma^2} \Big].\ee
Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
where the coefficients $R_l^\pm$ are the residues of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
The Fourier transform (\ref{eq:recursion}) has a beautiful feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
Taking this limit and evaluating the residues provides 
\begin{multline} \tilde{P}(s) = \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
	\\+ \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
Finally, inverting the Fourier transforms term by term with contour integration and incorporating (\ref{eq:q}) provides the steady-state solution (\ref{eq:steadystate}).

\section{Calculation of the moments}
Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
\be 0 = Dk (k-1)\langle u^{k-2} \rangle + \Gamma k \langle u^{k-1} \rangle + \nu (\ve^k-1)\langle u^k \rangle. \ee
$k=1$ provides the mean
\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}\ee
while $k=2$ provides the second moment
\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
leading to the velocity variance
\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
\section{Weak and strong collision limits}
Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities. The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
Following \cite{Hall1970}, this suggests normalizing the distribution $P(u)$ using 
\be z = \frac{u - \bar{u}}{\sigma_u} \ee
and 
\be Q(z) = \sigma_u P(u)\ee
to seek a differential equation for $Q(z)$ with manageable behavior as $\ve \rightarrow 1$.
Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" Master equation
\be (1-\ve^2) \frac{d}{2d + \gamma} Q''(z) - \frac{\gamma\sqrt{1-\ve^2}}{\sqrt{2d+\gamma^2}}Q'(z) - Q(z) + \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\gamma\sqrt{1-\ve^2}}{\ve\sqrt{2d+\gamma^2}} \Big]\Big) = 0 \ee
This equation remains exact and is only a change of variables from (\ref{eq:master}).
Now we approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$. This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-06-25 ch6conc.tex
eh

%%!TEX root = ../diss.tex

\chapter{Summary and future work}
\label{ch:conc}

\section{Key contributions}

\subsection{Probability distribution of the sediment flux}

\subsection{Inclusion of velocity fluctuations into Einstein's model of individual particle trajectories}

\subsection{Quantification of the control of bed elevation fluctuations on sediment transport fluctuations}

\subsection{Understanding of how sediment burial affects the downstream spreading of sediment tracer particles}

\section{Models and the real world}

\section{Closure: Complexity vs Realism}

### 2021-06-25 clean.sh

@@ -9,8 +9,6 @@ rm -f ch3bedEle.aux
 rm -f ch3bedEle.log
 rm -f ch4downDiff.aux
 rm -f ch4downDiff.log
-rm -f ch5conc.aux
-rm -f ch5conc.log
 rm -f laySum.aux
 rm -f laySum.log
 rm -f pref.aux
@@ -26,3 +24,8 @@ rm -f diss.lot
 rm -f diss.out
 rm -f diss.synctex.gz
 rm -f diss.toc
+rm -f ch6conc.aux
+rm -f ch6conc.log
+rm -f ch5langevin.aux
+rm -r ch5langevin.log
+

### 2021-06-25 ch1intro.tex
bout to move this bish

%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Sediment transport and landscape evolution}
\label{ch:Introduction}

\section{Geomorphology traditions}
Geomorphology can be defined as the study of Earth's evolution by tectonics, weathering, and erosion \citep{Chorley1984}. 
This science has long been characterized by observation and description \citep{Gilbert1877, Tight1903, Hack1960}. Only relatively recently did the science turn to quantitative methods \citep{Bagnold1941, Strahler1952, Scheidegger1961, Leopold1962}, with researchers working to frame observations in terms of underlying processes and to describe them with mathematical models modified from physics \citep{Church2005,Roades1996,Chorley1965}.

This quantitative shift or incorporation of physics methods into geomorphology has been criticized as overly reductionist \citep{Slaymaker2020}.
Many of the problems addressed within physics involve closed systems with perfect order \citep{GoldSteinClassicalMechanics,SakuraiQM,JacksonElectrodynamics}. 
These systems stand in contrast to the systems in geomorphology. Earth's surface is open and highly disordered. The surface receives fluxes of mass and energy across its boundaries by tectonics, climate, transpiration, and sunlight, and its component parts vary widely in their characteristics, activities, and compositions. With this complexity \citep{Chorley1971}, we can conclude that a complete reduction of geomorphology to physics is likely impossible.

There are nevertheless lessons physics can exchange with geomorphology and sediment transport theory. This is the central theme of this thesis. Before presenting original research, the first chapter summarizes key developments in quantitative geomorphology models which have resulted from this exchange. The section is organized along two classifications. First, I distinguish earlier works as ``stochastic" or ``deterministic". Stochastic models introduce randomness or noise as a means to crudely represent the complexity of natural systems (open boundaries, diversity in component parts) within idealized mathematical models of nature. Deterministic models involve Newtonian equations without noise to represent nature. Second, I distinguish earlier models as ``discrete" or ``continuous". These terms reflect whether the degrees of freedom of individual particles are retained in mathematical models, or whether they are averaged away. When grains are tracked, a model is discrete. When they are discussed as a flow, it is continuous. We can therefore discuss a ``discrete and stochastic" model, or a ``continuous and deterministic" model, and so on.
With these distinctions in mind, we launch into a survey of the foundation upon which this thesis builds: physics-inspired, process-based models of sediment transport and landscape evolution.

\section{Bedload transport foundations}

Sediment moves in rivers when the tractive force of the water flow overcomes any resisting forces by gravity and friction.

\subsection{Bulk bedload flux}

\subsection{Bagnold's energy balance}

One of the most influential approaches from physical arguments is due to Bagnold \citep{Bagnold1956,Bagnold1966}.
Bagnold described sediment transport as a process which ultimately converts flow energy to heat.
He assumed that the flow power $P_f$ available to move sediment scaled as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move. 
He reasoned that if the average downstream flux of particles is $q$, and particles move on average with velocity proportional to the fluid velocity near the bed, then the power $P_g$ required to sustain particle motion scales as $P_g \propto q/\tau^{1/2}. $ Balancing fluid propulsion against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
\be q = k(\tau-\tau_c)\tau^{1/2}, \label{eq:bagnold}\ee
which has shown good correspondence with laboratory data after careful calibration of the constant factors $k$ and $\tau_c$.
The large shear stress limit $q \sim \tau^{3/2}$ is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972}

\subsection{Discrete particle models}

\subsection{The probabilistic approach of Einstein}

\section{From transport to landscape evolution}

\subsection{Exner's approach}

Exner was probably the first to describe landscape evolution as a mathematical result of sediment transport \citep{Exner1925}, when he related the topographic elevation to the sediment flux via
\be (1-\lambda)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \ee
This equation links the temporal evolution of the land elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux. The parameter $\lambda$ is the bed porosity.

Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuous fields. This representation can be interpreted as an average over the detailed locations of individual grains, and it is expected to be an excellent approximation whenever the scales of interest are large compared to the size of the averaging window, which should be taken much larger than the size of the largest grains being modelled \citep{Coleman2009}.

Yet there are many contexts on Earth's surface where the scale of interest is not much larger than the largest grains involved in the system. We can wonder, for example, in mountain channels, how large boulders contribute to the formation of steps \citep{Zimmerman2008,Saletti2020}, or other structures with sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}. In these cases, individual grains are comparable to the scales of interest, and the continuum approximation breaks down.

\subsection{Tusjimoto and beyond}
Explain what tsujimoto essentially did to couple Einstein into Exner, giving the entrainment form of the exner equation.
Cite Parker and Furbish who provided refinements of it.
Describe the inherently nonlocal characteristics of sediment flux.
Indicate the major strengths and weaknesses -  key weakness being that, if the sediment flux is a fluctuating quantity, as formulated by Einstein, then

\section{Realism versus complexity}

\section{The point of the thesis}

\subsection{Open problems in the stochastic paradigm}

\subsection{Quantities of interest}
Explain what the pdf of position is
explain what the variance represents
explain that the moments more generally are of interest
explain that the sediment flux is key to landscape evolution. not just in mean value, but also in its fluctuations.

\subsection{Tools of the trade}

\subsubsection{The Langevin equation}
When particles are subjected to fluctuating forces, their Newtonian equation of motion can be written

In this equation, $F$ is the deterministic or steady component of the forcing, while $\xi$ is the fluctuating component. 
The classic example is due to Einstein, and 
\subsubsection{Master equations and probability flow}

\subsubsection{Idealized noises}

\subsection{Organization of the thesis}

Four main problems. Ch 2 is about how to calculate the sediment flux and to understnad particle motion with fluctuating velocities.
Ch3 is a sketch of how we can incorporate bed elevation changes into sediment transport modelling based on stochastic theories.
Ch4 is the

\section{Introduction: birth-death models for the probability distribution of bedload flux}

In river science, a fundamental problem is the determination of the bedload flux, or the rate of downstream movement of bedload grains \citep{Ballio2014}.
Since bedload transport has strong feedback with stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of environmental considerations. 
These considerations span a wide range, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}. 
Unfortunately, existing approaches to compute the bedload flux are inadequate. 
Predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.

Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
The bedload flux is in a cyclical feedback with its controls: these controls are the details of turbulence and bed organization \citep{Jerolmack2005}.

As a result, bedload fluxes exhibit wide fluctuations, even as average characterizations of flow and bed organization remain steady \citep{Ancey2014}.
In the most controlled laboratory experiments, with steady flows driving uniform glass beads under conditions which do not favor bedform development, instantaneous fluxes are often as much as 200\% mean values \citep{Bohm2004, Ancey2008, Heyman2014, Heyman2016}. 
In natural streams, with temporally or spatially varying grain size distributions \citep{Lisle1992, Chen2008}, unsteady flows \citep{Mao2012, FerrerBoix2015}, a variety of dynamic bed surface structures \citep{Hassan2008, Venditti2017}, variable sediment supply \citep{Madej2009, Elgueta2018}, lateral adjustment \citep{Pitlick2013, Redolfi2018}, and migrating bedforms \citep{Gomez1989, Dhont2018} all imparting additional sources of variabilility through their feedbacks with the flux, instantaneous fluxes can be even larger. 
% Iseya1987, Lisle1989, Kuhnle1998

Apparently, bedload fluxes have statistical characteristics.
Einstein was the first to recognize the statistical character of bedload transport \citep{Einstein1937}. 
He understood transport as a random switching between states of motion and rest \citep{Einstein1942, Einstein1950}. 
He called the transition from rest to motion entrainment, and he characterized it with a probability which was linked to extreme events in fluid turbulence \citep{Einstein1949, Einstein1950}. 
He called the transition from motion to rest deposition, and he considered it an implicit function of the bedload flux \citep{Einstein1942, Einstein1950}. 
By convolving over sequences of entrainment and deposition, employing some semi-emprical arguments, Einstein developed a formula for the average bedload flux \citep{Einstein1950}.

Many investigators have criticized and developed Einstein's ideas \citep{Paintal1971, Yalin1972, Shen1980, Lisle1998, Papanicolaou2002, Sun2000, Ancey2006, Armanini2015}. 
These revisions focus on the more ad hoc elements of Einstein's derivations, and they lend a more mechanistic basis, a firmer mathematical foundation, and more generality to the approach. 
Although all of these authors accepted the statistical character of bedload motion, none developed theories for a probability distribution for the bedload flux. 
This is desirable since its higher moments give unambiguous measurements of the magnitude of bedload fluctuations.

The extension of the Einstein approach to obtain a probability distribution of the bedload flux, with fluctuations stemming from the random character of entrainment and deposition, was first explored by \citet{Lisle1998, Sun2000}, and it was further developed by \citet{Ancey2006}, and the more recent work I will discuss extensively in this review.   
All of these authors revisited Einstein's ideas from a foundation in the stochastic mathematics which became formalized somewhat after Einstein's work \citep[e.g.][]{Cox1965}, treating the transitions between motion and rest states as random events characterized by probabilities. 
This enabled them to apply the formalism of continuous time Markov processes to derive a probability distribution of the bedload flux, with a mean value which was an improved version of the \citet{Einstein1950} formula, very similar to the revised formula of \citet{Yalin1972}, and a variance which provided an unambiguous prediction of the magnitude of bedload fluctuations.

Deterministic processes trace the evolution of some set of variables $\{x_1(t),x_2(t),\dots\}$ through time. 
Stochastic processes, in contrast, trace the evolution of a probability distribution for random variables $P_{X_1,X_2,\dots}(x_1,x_2,\dots;t)$ through time. 
The Markov moniker refers to the amount of memory in the process.
If the probability distribution of the variables of interest can be predicted in the future using only distribution functions from the present, the process is Markovian. 
Otherwise, if predicting the future requires knowledge of the entire history, the process is non-Markovian \citep{Cox1965, VanKampen1992}. 
Markov birth-death processes consider the probabilistic creation and annhilation of members of a population \citep{Cox1965, VanKampen1992}.

To model the bedload flux with such a process, the population considered is the number of moving particles within a control volume over the bed. 
This population is subject to creation by entrainment and annhilation by deposition \citep{Ancey2008, Turowski2009, Heyman2013, Ma2014,  Ancey2014a, Ancey2015}. 
This approach is exciting because Markov birth-death processes are highly studied in physics, chemistry, and population ecology, so there are many extensions readily available \citep{Bailey1968, Cox1965, Pielou1977, VandenBroek2012, VanKampen1992, Gillespie1992, Field2010, Mendez2015}. 
At the same time, birth-death modeling of the bedload flux remains relatively undeveloped: birth-death models are all one-dimensional and designed for a single grain size.

In this article, I review the literature on Markov birth-death models of the bedload flux.
In the review I assume some familiarity with discrete state continuous time Markov processes, which could be gleaned from reading the appropriate chapter in \citet{Cox1965}. 
These theories are exciting because they describe the mean bedload flux and its fluctuations within a volume \citep{Ancey2006, Ancey2008, Turowski2009}, statistical properties of the flux at a point in space \citep{Heyman2013, Ma2014}, and the spatial and temporal correlations in the bedload flux within a reach of channel \citep{Ancey2014, Ancey2015}. 
All of these considerations are new, and they are unique within the river science literature in that they do not ignore the statistical character of bedload transport. 
They initiate a theoretical framework from which the feedbacks between the arrangement of bed particles, turbulence, and channel morphology which lead to bedload fluctuations can be more carefully studied and understood. 
In reverse, the capacity of bedload models to describe fluctuations provides an additional benchmark against which to judge models, a task which is notably difficult \citep{Iverson2013}.

One of the first birth-death models of the bedload flux I will review is \citet{Ancey2006}: these authors generalized \citet{Einstein1950} using a birth-death framework to obtain a statistical distribution of the bedload flux within a control volume. 
They found the fluctuations predicted by the Einstein-like model were not large enough to describe experimental data.  
In order to match the wide fluctuations of experimental data, \citet{Ancey2008} reworked their earlier control volume model to include a collective entrainment effect.
Collective entrainment is a mathematical term charaterizing the correlations between moving grains due to turbulence, collisions, or granular avalanche effects.

The \citet{Ancey2008} model has been generalized and applied in a handful of followup works. 
\citet{Turowski2009} generalized it to include limited sediment availability. 
\citet{Heyman2013} and \citet{Ma2014} studied how the \citet{Ancey2008} model could be interpreted to describe the statistics of the bedload flux at a fixed point in space, rather than just a control volume like that considered by \citet{Einstein1950}. 
Control volumes are less desirable because transport is practically easier to measure at a point in space than it is within a control volume.

\citet{Ancey2014, Ancey2015} generalized the \citet{Ancey2008} model in order to understand spatial characteristics of bedload transport. 
They considered a reach of channel as an array of adjacent control volumes (cells), and they coupled the \citet{Ancey2008} model between adjacent cells. 
They mapped this generalized birth-death model for the bedload flux distributed over an array of control volumes onto an advection diffusion equation for the bedload flux. 
Some reprecussions of this model were further explored in context of experimental data by \citet{Heyman2014}.

All of these works contain many simplifying assumptions, and many of these were carefully highlighted by their authors. 
These studies were selected for this review because, taken together, they define a research trajectory which can be extrapolated to suggest future research topics. 
To be sure, there is a long road ahead before Markov models can accomodate all of the complexities of natural river channels, some of which I've mentioned. 
The goal of this review is to indicate this trajectory in order to motivate future research: this is pursued in sections \ref{sec:background} and \ref{sec:review}. 
I discuss some of the future research topics I believe the trajectory points out in section \ref{sec:scope}, and I develop a few of these ideas myself in \ref{sec:extension}. 
Along the way, I have somewhat changed the mathematical treatment and the notation from some of the original papers in order to highlight the historical progression between each work, and to illustrate that these works all share the same theme.

\section{Background: the foundational stochastic model of bedload transport} 
\label{sec:background}

\citet{Einstein1950} can be credited with the first attempt to understand the bedload flux as random quantity.  
Einstein understood the movement of individual bedload grains as a random succession of motion and rest intervals.
He termed the transition from rest to motion entrainment, and characterized it by a probability related to fluid turbulence. 
When particles are set in motion, Einstein considered they moved in discrete jumps of the same average distance. 
We define the bedload flux as the volume of grains passing a stream cross section per unit time.
It is a volume of grains per time per unit width of stream, so its units are $\text{area}/\text{time}$.  
Therefore, Einstein's mean flux formula is derived from summing the alternate start-stop motions of all bed particles which cross a surface in a time interval.

The conceptual picture Einstein considered is depicted in figure \ref{fig:yalin}. 
All bed particles are identical and are considered to have the same average geometry when at rest on the bed, meaning their entrainment characteristics are identical. 
For continuity with the rest of the review, we consider sediment particles are spheres of radius $a$, although \citet{Einstein1950} was more general. 
We can follow Einstein to compute the mean number of these particles crossing a flow cross-sectional surface $S$ in an interval of time. 
This will construct the mean bedload flux $\bra q_s \ket$.
In subsequent sections, we consider $q_s$ a random variable, and we will review approaches to derive its full probability distribution.
From these contemporary models, an Einstein-like mean flux formula emerges from taking the mean of the probability distribution for $q_s$.  
Hence we use averaging brackets $\bra \cdot \ket$ for the sake of continuity.

\begin{figure}
	\centering
	\includegraphics[width=.98\linewidth]{./figures/ch1/yalindrawing.png}
	\caption{Einstein's conceptual picture (adapted from \citet{Yalin1972}). Particles move in discrete jumps of length $L$ from left to right through an array of adjacent control volumes. The bedload flux is the volume of bedload particles crossing the surface $S$ per unit width and time. \label{fig:yalin}}
\end{figure}

Einstein's derivation goes as follows: we partition the channel into a sequence of identical control volumes $V$. 
Each volume has dimensions $L$, $w$, and $h$ ($V=lwh$). 
$L$ is the downstream length of each control volume, and it is also the average jump distance of particles. 
$w$ and $h$ are the width and height of each control volume.
Let $P_n$ be the probability that any individual grain on the bed is entrained \textit{at least} $n$ times during the time interval $T$. 
That is, let $P_n$ be the probability that an individual grain undergoes \textit{at least} $n$ jumps of length $L$ in an interval $T$, meaning it travels a distance $nL$ or more in $T$. 
Considering that each control volume contains $N_V$ grains at rest within it, it follows that on average
\be N_V P_n. \ee
grains are displaced \textit{at least} a distance $nL$ from a control volume within the time interval $T$.

Now we derive the mean flux. 
Note that the grains passing through a cross-sectional surface $S$ (of area $lw$) in the time $T$ could have come from any upstream location. 
Therefore, the number of grains crossing $S$ in $T$ is 
\be \sum_{n=1}^\infty N_V P_n. \ee
This is the net arrival of grains in the interval $T$ from all upstream control volumes. 
Multiplying by the particle volume $\nu_p$ and dividing by the timescale $T$ and channel width $w$ gives the mean bedload flux (volume per width per time):
\be \bra q_s \ket = \frac{\nu_p N_V}{w T}\sum_{n=1}^\infty P_n = A \frac{a^2}{T} \sum_{n=1}^\infty P_n. \label{eq:einstein1}\ee 
In the second equality, we have introduced $\nu_p = 4\pi a^3/3$, and two assumptions of Einstein. First, is the reasonable assumption that the number $N_V$ of particles on the surface within $V$ is proportional to the ratio of bed surface and particle areas: $N_V \propto Lw/a^2$, and second is the assumption that the travel distance $L$ is proportional to particle size: $L \propto a$. The latter assumption is more difficult to justify \citep[e.g.][]{Yalin1972}. 
$A$ is a constant of proportionality.

Einstein implies the probability that an individual grain travels a distance $nL$ or more in $T$ is equivalent to the probability that an individual grain travels at least a distance $L$ successively $n$ times in $T$. 
Hence $P_n = P_1^n$ by the law of multiplication of probabilities, so that $\sum_{n=1}^\infty P_n = \sum_{n=1}^\infty P_1^n = P_1/(1-P_1)$, using the geometric series (noting $P_1 < 1$ because it is a probability). 
Owing to this substitution, Einstein's mean bedload formula takes the form 
\be  \bra q_s \ket = A \frac{a^2}{T}\frac{P_1}{1-P_1}, \label{eq:einstein2} \ee
where $A$ is a constant of proprotionality.  
The $P_1/(1-P_1)$ structure of Einstein's formula is a distinctive characteristic. 
In order to use the apply the Einstein formula to predict bedload transport in natural streams or experiments, the timescale $T$ and the probability $P_1$ that at least one grain entrains in $T$ must be determined. 
Then the formula can be calibrated to a given setting by adjusting the constant $A$.

In the original work, Einstein considered the timescale $T$ proportional to the "time of the grain" formed by the grain's size $a$ and its settling velocity $w$ in still fluid: $T \propto a/w$. 
We will revisit this assumption shortly. 
Regarding the probability $P_1$, one of Einstein's most influential and enduring ideas within hydraulic engineering is his conception of the probability $P_1$ that at least one particle entrains in $T$. 
He considered that entrainment is driven by the fluctuating lift force imparted by the turbulent fluid, and it is resisted by the submerged weight of grains. 
Therefore, he formulated $P_1$ as an exceedance probability of the lift force over weight, so he wrote $P_1$ as an integral over the probability distribution of fluid shear stress. 
Einstein developed an alternative perspective on the incipient motion of particles that embraced turbulence and did not rely on the critical shear stress concept \citep[e.g.][]{Shields1936}.

This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly random) resisting quantities has been generalized and extended a great deal. 
\citet{Paintal1971} made a significant extension by including random granular geometry of bed particles into the force balance. 
More recently, \citet{Tregnaghi2012} ammended the theory to include the impulse concept: the experimentally verified idea that force magnitude and duration are of equal importance for entrainment \citep[e.g.][]{Diplas2008, Celik2014}. 
Refined theories of the entrainment probability, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2008, Dey2018}, and they are a subject of intensive ongoing research.

As noted by \citet{Yalin1972}, who wrote a careful review and revision of \citet{Einstein1950}, the Einstein theory for $\bra q_s \ket$ is theoretically sound, at least to equation \ref{eq:einstein2}, and provided Einstein's probabilities $P_n$ are interpreted in a careful way. 
I have mimiced Yalin to say $P_n$ is the probability of \textit{at least} $n$ jumps in $T$. 
Had I said $P_n$ was the probability of (only) $n$ jumps in $T$, the relationship $P_n = P_1^n$ would not follow, so the Einstein flux would not develop the notable $P_1/(1-P_1)$ form as in \ref{eq:einstein2}. 
Einstein was not as clear as he could have been on this distinction. 
Notably, the more general birth-death theories which reproduce Einstein as a limiting case \citep[e.g.][]{Ancey2006} use probability concepts relating to (only) one entrainment in a time interval. 
Therefore, to link back to Einstein from these contemporary theories, we need to follow \citet{Yalin1972} to express the mean flux $\bra q_s \ket$ in terms of the probabilities of (only) $n$ entrainments in a time interval, which we can denote by a lower case letter $p_n$. 
These two probabilties, $P_n$ (at least) and $p_n$ (only), have probably been conflated often within the literature.

Yalin appealed to scaling arguments in order to refine Einstein's theory. 
He highlighted that Einstein's prescription of $T \propto a/w$, where $w$ is the settling velocity, was not supported by any experimental evidence or theoretical argument. 
Yalin noted that within Einstein's theory, "$T$ appears in connection to the study of detachment of individual grains due to turbulent fluctuations", from which he concluded that $T$ should be actually be measured in terms of a period $\tau$ of turbulent fluctuations: 
\be T = N_\tau \tau, \label{eq:yalintime}\ee
where $N_\tau$ is the dimensionless number of turbulent fluctuations in $T$ (which is considered unknown but very large).

Yalin introduces a probability $p$ of (only) one entrainment during an individual turbulent fluctuation. 
Since the number $N_\tau$ of turbulent fluctuations in $T$ is very large, he considers this probability is very small, so that the probability of (only) $n$ detachments in $T$ can be computed with the Poisson distribution: 
\be p_n = \frac{(N_\tau p)^n}{n!}e^{-N_\tau p}. \label{eq:yalinpoisson} \ee
The probability $P_n$ of \textit{at least} $n$ detachments in $T$ can be expressed in terms of the probability $p_n$ of (only) $n$ detachments in $T$ as 
\be P_n = \sum_{i=n}^\infty p_i. \label{eq:yalinsum}\ee
Owing to equations \ref{eq:yalintime}, \ref{eq:yalinpoisson}, and \ref{eq:yalinsum}, the mean bedload flux \ref{eq:einstein1} becomes
\be  \bra q_s \ket = A \frac{a^2}{N_\tau \tau} \sum_{n=1}^\infty \sum_{i=n}^\infty p_i = A \frac{a^2}{N_\tau \tau} \sum_{n=1}^\infty n p_n = A \frac{a L}{\tau} p. \label{eq:yalinflux} \ee
The unknown number of turbulent fluctuations $N_\tau$ cancels. 
Hence, Yalin's ammended Einstein formula is proportional to $p$, rather than $P_1/(1-P_1)$ as in \citet{Einstein1950}. 
In Yalin's mind, the probability $p$ is the probability of entrainment of only one grain in a small interval $\tau$ related to the period of turbulent fluctuations.

When Einstein's conceptual picture of bedload transport is revisited from a Markov process framework, the probability distribution $P(q_s)$ of the bedload rate can be derived from his assumptions \citep{Ancey2006}. 
This distribution implies a mean bedload formula $\bra q_s \ket = \sum q_s P(q_s)$ formally similar to Yalin's Einstein-like formula \ref{eq:yalinflux}, and it provides additional information about the magnitude of bedload fluctuations, which are characterized without ambiguity by the variance $\bra \delta q_s^2 \ket$, where $\delta q_s = q_s - \bra q_s \ket$ is the deviation of the bedload flux from the mean.
Jumping ahead a little, the fluctuations $\bra (\delta q_s)^2 \ket$ derived in this way are not large enough to match experimental data \citep{Ancey2006}, and this is because \citet{Einstein1950}, \citet{Yalin1972}, and \citet{Ancey2006} each assumed the transitions between rest and motion were independent between all particles.

\citet{Ancey2008} fixed this problem by describing bedload transport within a more general birth-death process framework \citep{Cox1965}. 
They included a collective interaction into the entrainment rates of bed particles in their model, and with this inclusion they derived bedload fluctuations of realistic magnitude. 
Their collective interaction biases entrainment, making it more likely when particles are already moving, a point we will discuss.  
It leads to clouds of active particles, and a wide tail on the bedload flux probability distribution: both of these features are in accord with experimental observations \citep{Drake1988, Ancey2006, Ancey2008}. 
I'll start the review with \citet{Ancey2006}, which is essentially \citet{Einstein1950} revisited from the birth-death framework, lending it more generality by treating the bedload flux as a probability distribution.   
The mean of this probability distribution reproduces equation \ref{eq:yalinflux}, which is a satisfying historical coherence. 
Followups and refinements of the \citet{Ancey2006} work constitute the bulk of the review in section \ref{sec:review}.

\section{Scope: identifying future research directions}
\label{sec:scope}

\subsection{recap} 
Einstein developed first stochastic theories describing the diffusion and flux of bedload. 
Ancey 2006, building on earlier work from Lisle 1998 and the many investigators who criticized and revised Einstein, formalized Einstein's assumptions over a foundation in Markov process theory, and extended his work by treating the bedload flux within a control volume as the cooperation of many independent two-state Markov processes: each bed particle within the control volume makes random transitions between motion and rest. 
Thus he derived a statistical distribution of the bedload rate which reproduces Einstein's result when the mean is taken. 
However, the magnitude of fluctuations predicted by this model are too small.

Ancey 2008 extended the Einstein theory by linking the transition rates between particles within the observation window with a collective entrainment term: they prescribed that the transition rate from rest to motion depends on the number of active particles. 
This is a simple way to include the effects of coherent turbulence and impact-based entrainment: it is an experimental observation that moving particles tend to come in waves \citep{Drake1988}. 
The Ancey 2008 model was demonstrated capable of describing fluctuations, and all of its parameters have physical meanings, but notably there is no clear suggestion as to how collective and individual entrainment processes can be separated within experiments. 
Clearly, approaches to derive entrainment probabilities are focused on the individual entrainment rate \citep{Dey2018}, and there have been no developments with regard to computing the collective entrainment rate: there is no clear physical model of collective entrainment yet.

Turowski 2009 extended the Ancey work to take account of limited supply, in order to describe semi-alluvial channels where alluvial deposits lie on top of bedrock. 
He continued to work within the Einstein paradigm by describing the transport rate within a control volume, rather than by counting the number of particles leaving the control volume: although the Ancey 2008 work essentially outlines this possibiilty. The transport rate at a point should be considered as the number of emigration events in a unit time.

Heyman et al 2013 and Ma et al 2014 went on to consider the statistics of emigration: they tried to discern what could be learned about the transport rate at a fixed point from the control volume formalism of Ancey et al and Einstein before that. 
Heyman et al was concerned with the statistics of the waiting time between successive emigration events. 
The original Einstein (1937, 1950) assumptions generate an exponential waiting time ditsribution with a timescale related to the entrainment rate.  
However, Heyman et al (2013) showed that when collective entrainment effects were included there are two timescales: one fast timescale related to the individual entrainment rate, and one slower timescale related to collective entrainment effects. 
Thus, the distribution of waiting times between successive emigration events, which is a useful concept for alternative stochastic models of the bedload flux distribution \citep[e.g.][]{Turowski2010}, is a more subtle object if collective entrainment occurs. 
These effects are accentuated at low transport rates, where the fast and slow timescales are more disparate. 
At high rates, the timescales become comparable and the waiting time between successive emigration events blends into an Einstein-like exponential.

All of these approaches considered bedload transport within a control volume or at a point, but indeed, a very large set of contemporary river science studies are concerned with the diffusion or spreading of bed material through a downstream reach of river \citep{Hassan2016}. 
Bedload diffusion is not captured by a model at a single location: often, it has been described using an advection diffusion equation. 
These advection diffusion equations follow from considerations of mass balance within river channels, but they were not derived from any underlying microstructural model until \citep{Ancey2014}. 
\citet{Ancey2014} extended the previous control volume model \citep{Ancey2008} to an array of control volumes, where emigration from the $i$th volume is immigration to the $i+1$th volume.

In an approach widely used in chemical physics \citep{Gardiner1983} and ecology \citep{Bailey1968}, \citet{Ancey2015, Ancey2015} derived an advection diffusion equation from their microstructural model in a limit as the size of each control volume goes to zero while the number becomes infinite. 
Thus, in their interpretation, bedload diffusion emerges in the continuum limit of coupled birth-death immigration-emigration models. 
They rederived the famous Exner equation of bedload diffusion.

Taken together, this research exhibits a coherent progression from the original work of Einstein to a more mathematically and physically based set of more general approaches. 
Within these approaches the bedload flux is understood as a random quantity. 
The mean bedload flux and the magnitude of its fluctuations are modelled in an unambiguous way \citep{Ancey2006, Ancey2008} considering no limitations in sediment availability, and it was shown that collective effects are a necessary inclusion to properly describe fluctuations in bedload transport \citep{Ancey2008}. 
An extension to a more realistic situation of finite supply was developed \citep{Turowski2009}, and the relationships between these control volume based models and the definition of bedload flux as particles crossing a plane perpendicular to the flow were developed \citep{Heyman2013, Ma2014b, Ballio2014}.

These approaches did not address spatial correlations or variations in bedload fluxes, although the problem of bedload diffusing through a reach of channel is of contemporary significance \citep{Hassan2016}. 
This extension was the most recent development in birth-death modelling of bedload \citep{Ancey2014, Ancey2015}. 
By extending the \citet{Ancey2008} model to an array of adjacent control volumes, an advection-diffusion equation describing the concentration of bedload in motion was developed, providing the first connection between a microscale stochastic model of bedload transport with a macroscopic advection-diffusion formulation. 
\citet{Heyman2014} examined the spatial correlations between local bedload fluxes expressed by the \citet{Ancey2014} model. 
This is the state of the art of bedload flux models.

\subsection{Assumptions, calibraiton problems, and unclear aspects: A criticism of the birth-death approach}

First, the various glaring assumptions of the birth-death approaches are highlighted. 
\begin{enumerate}
	\item Models only consider one particle size, while natural streams exhibit wide distributions of sizes and bedload transport expresses a wide set of effects related to particle size segregation \citep{Wilcock2003, Parker1982, Chen2008}. 
	\item There is no clear means to discriminate collective and individual entrainment processes within experiments. 
	\item There is only a one-way feedback in these models: bedload is considered subordinate to the fluid flow, and the effect of bedload transport back onto the fluid flow is considered negligible. In fact, there are measurable influences of bedload transport on the fluid phase \citep{}; it's not yet clear whether a one-way coupled scheme such as this can actually describe natural streams across a realistic range of conditions. 
	\item Collective entrainment is somewhat of a catch-all term with the physical mechanisms which may contribute to it unresolved. These may include collective entrianment, the formation and disintegration of particle clusters, interactions between the motion of grains of idfferent size, turbulent structure, local avalanche behavior, and bed form migration. 
	\item there have been no theories developed to compute the collective entrainment rate from any simplified mechanical model, although there is a very large set of work concerned with calculating the individual entrainment rate from considerations of fluid turbulence and random granular arrangement \citep{Einstein1949, Einstein1950, Grass1970, Paintal1971, Cheng1998, Wu2004, Dey2008, Tregnaghi2012, Dey2018}. Calculating the collective entrainment rate from underlying principles appears on the surface very difficult. If it is considered to stem from collisions of moving grains with stationary grains, the collective entrainment probability will be related to the probability of collisions with the granular bed. If it is considered to stem from granular avalanches, where coherent turbulent structures initiate collections or clusters of grains into motion simultaneously, the collective entrainment probability will depend on the collective dynamics of the granular assembly -- leading immediately into the murky physics of force balance within granular assemblies.
	\item Einstein-like assume a clean division between rest and motion states, which is no doubt an idealization. Bedload transport makes a continuous transition from the idealized start-stop motions of Einstein-like models to a less idealized granular flow or creep, where all particles move together in a coorelated fluid-like flow \citep{}, meaning bedload models should only hold at relatively low mobility stages. 
	\item At the same time, the divison of motion into only two categories may be flawed. A wide collection of studies have highlighted different modes of motion. \citet{Einstein1950} considered that bedload was particles moving "in a rolling, sliding, or saltating mode". What if rolling, sliding, and saltating modes of motion were treated independently? In an \citet{Ancey2006} model of independent particles switching between states, this would require a four-state model. There are too many transition probabilities within such a model probably to calibrate the model from experiments. Again, in order to consider such a scenario we would need more knowledge of underlying mechanics.

\end{enumerate}

Incorporating multiple grain sizes in birth-death models is possible, in principle. 
As a first approximation, it is easy to introduce multiple grain sizes without including interactions between grain sizes in the entrainment and deposition probabilities for each size fraction. 
However, spatial and temporal heterogeniety in the bed surface characteristics are a hallmark of bedload transport of gravel mixtures \citep{Hassan2008}. 
This leaves models such as \citet{Ancey2008} somewhat without basis when multiple grain sizes are considered. 
The concept of \citet{Turowski} is essential to include when multiple grain sizes are considered: the bed surface state determines which grains can entrain \citep[e.g.][]{Wilcock2003, Parker1982}, and presumably which grains can deposit, as well. 
The entrainment and deposition rates of each size fraction will need to depend on the bed surface state, and fractional transport will set up spatial hetereogenieties in the bed surface state, so that an \citet{Ancey2014} type model, incorporating the possiblity of spatial hetereogeniety, should be mixed with a \citet{Turowski2009} type model to incorporate the effect of the bed surface state on the entrainment and deposition rates of each size fraction. 
This extension will not be easy, and it will introduce many undetermined parameters: to pin down this transport model of multiple interacting size fractions will take a lot of work.

These stochastic models are admittedly difficult to calibrate. 
Their calibration requires a large dataset which is prohibitive to measure within natural streams. 
Therefore, their applicability in real streams remains limited until the inputs of stochastic models can be computed from physical theories based upon practically measurable quantities. 
This quest for relationships to compute the inputs of stochastic models from measurable quantities has been called "stochastic closure" for the analogous problem in turbulence \citep{Heyman2016}. 
The difficult issues precluding the application of stochastic models to natural streams should not be downplayed: much more work is needed.

\endinput

Any text after an \endinput is ignored.
You could put scraps here or things in progress.

### 2021-06-26 ch1intro.tex
lots of progress on the introduction .

@@
 With these distinctions in mind, we launch into a survey of the foundation upon which this thesis builds: physics-inspired, process-based models of sediment transport and landscape evolution.
+
+\section{theories of individual particle movements}
+
+A fundamental problem in sediment transport modelling is to predict the downstream movement of an individual particle.
+At first glance, this seems a basic problem in general physics, but digging in exposes severe challenges.
+Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not known \citep{Furbish1997}.
+Downstream movement is resisted by frictional collisions between moving particles and the bed.
+Since the bed is a granular surface, its surface geometry is very difficult to characterize \citep{Gordon1972}, and the outcome of collisions varies from one to the next \citep{Sekine1992}
+Entrainment and deposition provide an additional layer of complexity.
+Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966}, so they can deposit \citep{Charru2004}.
+These pockets are not permanent shelter.
+Rearrangement of the surrounding bed can reexpose sheltered particles to the flow, and sufficiently strong turbulent fluctuations can overcome the shelter, even if the surrounding particles do not move \citep{Valyrakis2012,Celik2014}, so particles can entrain back into motion.
+Ultimately, particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}.
+These buried particles cannot move again until those burying them have \citep{Nakagawa1981}.
+
+Predicting the movements of individual particles therefore involves many contributing processes, and our understanding of each of these processes remains incomplete.
+\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize this complexity, although it has been slightly revised more recently \citep{Campagnol2013, Hassan2017, Pierce2020}.
+Nikora et al divided the downstream trajectory of an individual particle into three timescales, or ``ranges", termed local, intermediate, and global.
+The local range refers to the period of motion between subsequent interactions with the bed, when the particle accelerates downstream within the flow, the intermediate range reflects particle motions through sequences of collisions, and the global range refers to particle motion between subsequent entrainment and disentrainment events.
+\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried. 
+This set of timescales -- local, intermediate, global, and geomorphic -- provides a structure by which to organize wide modelling literature on bedload trajectories and to understand the approximations which have been made on the fundamental problem of predicting the downstream movements of individual grains.
+
+\subsection{The tracer problem}
+The original motivation to understand individual particle motions was probably to understand the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research today \citep{Hassan2017,Pretzlav2021}.
+A common measurement technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
+In principle, if tracers are a proxy for the population of grains in a stream, one can estimate the velocities of tracers, then multiply by an estimate of the number of grains available for motion in the stream to calculate the overall flux \citep{Yano1969,Nakagawa1976,Hassan2017}.
+Unfortunately, this does not really work, due to the distinct behavior of tracers over the timescales identified above.
+Apparently, tracer particles slow down through time within streams \citep{Ferguson2002,Haschenberger2012}.
+This means it is not clear how measured tracer velocities can be mapped back to bulk sediment fluxes, and it exposes a need for further understanding of how exactly particle motion characteristics differ across the local, intermediate, global, and geomorphic timescales.
+
+This need to understand tracer movements is not however the sole point of research into individual particle motions.
+Modern science trains us to reduce systems to their component parts in order to glean better understanding of the system as a whole.
+It is conceptually clear that sediment transport and geomorphology are intimately linked to the movements of individual grains.
+If we understand them better, research progress will probably follow.
+
+\subsection{ Einstein 1937}
+Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}, motivated (at least as far as his research supervisor Meyer-Peter was concerned) by the need to understand the efficiency of Helley-Smith samplers for the management of sedimentation \citep{Ettema2004}.
+Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to a probabilistic approach to characterize their motion.
+Einstein's basic insight was to represent particle motions as an alternating sequence of motions and rests having random characteristics.
+As his interest was on the global range of particle motion, and the duration of particle motions is usually short compared to rests, Einstein approximated individual motions (between entrainment and deposition) as instantaneous.
+With this setup, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
+Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
+Einstein's experiments indicated that both step lengths and resting times were drawn from exponential distributions.
+His target was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
+He formulated the problem mathematically with an infinite series of convolution integrals, in an elegant display of mathematical physics.
+
+For connection to later work in the thesis, I will summarize Einstein's work slightly differently.
+In effect, if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
+\be \dot{x}(t) = \eta(t), \label{eq:einlangevin}\ee
+where $\dot{x} = dx/dt$, and $\eta(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $l$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
+A particular realization of this noise can be written
+\be \eta(t) = \sum{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
+where $N(t)$ is the number of particle entrainments in time $t$, $the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are distributed according to $P(s) = l^{-1}\exp(-s/l).$
+Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
+Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1956}, but with a particular random driving term (equation \ref{eq:einrando}) chosen to basically represent the phenomenon under consideration.
+
+The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1984,Moss1989}.
+Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
+\be  \big(l \px \pt + k_E l \px + pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
+This equation can be solved by standard methods (series solutions or transform calculus) to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
+\be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/l}\theta(x) \sqrt{\frac{k_E t}{lx}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{l}}\Big) \ee
+Here, $\mathcal{I}$ is a modified Bessel function. 
+The probability distribution eq. \ref{eq:eindist} fully characterizes the dynamics of an individual particle alternating through steps and rests.
+
+This distribution displays both advective and diffusive characteristics.
+One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^l$ and integrating over space. This gives the mean position of the particle
+\be \langle x \rangle (t) = k_E l t, \ee
+so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E l$, given by the entrainment rate times the mean step length.
+The rate at which one particle spreads out from another due to differences in their motion characteristics can be represented by the variance of position, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$.
+This gives
+\be \sigma_x^2(t) = 2 k_E l^2 t, \ee
+so particles in Einstein's model spread apart as a normal diffusion process \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E l^2.$
+
+\subsection{First extensions of the Einstein model}
+
+Einstein's model of particle movement seems to have been overlooked for several decades, or possibly overshadowed by his later work, which uses its essential ideas for more obviously practical purposes.
+Eventually, curiosity into individual particle motions was re-sparked in the context of cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}, so Einstein's model received a great deal of new attention.
+
+The first set of modifications to Einstein's approach in this period concerned the form of the distributions used for step lengths and resting times.
+Einsteins flume experiments demonstrated that particle step lengths followed exponential distributions \citep{1937}, and this was confirmed by subsequent studies \citep{Yano1969, Nakagawa1976}.
+Yet these studies were all conducted in steady flows in artificial flumes, not real streams.
+When sediment transport is observed before and after a flood, the step length within a flood is better described by a Gamma distribution, this being the sum across the many individual steps that occurred during the flood \citep{Hassan1991}.
+When sediment transport occurs over dunes, similar modifications were made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965}.
+
+\subsection{ Inclusion of the movement duration}
+\label{sec:lisle}
+
+Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescale of individual particle movements. 
+Yet the advent of high speed camera experiments of bed load transport produced data on the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which Einstein never intended to model.
+In the local range, particles move with a velocity that fluctuates due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
+Einstein's infinite movement velocity assumption sets aside these details.
+
+Studies by \citet{Gordon1972}, \citet{Lisle1998}, and \citet{Lajeunesse2017} incorporated the duration of sediment motion.
+They approximated particle velocities as constant, and assumed that the motion time is also an exponentially distributed random variable, this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion after an initial entrainment.
+The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ is 
+\be \dot{x} = V\eta(t), \ee
+where the noise is now a so-called ``dichotomous Markov noise" \citep{Bena2006}, which is essentially a random switch or telegraph-type signal that alternates between ``on" ($\eta(t) = 1$) and ``off" ($\eta(t) = 0$) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996} as displayed in figure \ref{fig:lislefig} panel (a).
+Some particle trajectories solving equation \ref{eq:lislelangevin} are displayed in \ref{fig:lislefig} panel (b).
+
+This time, the governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt' \rangle$ becomes \citep{Balakrishnan1993}
+\be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\ee
+where $k_E$ and $k_D$ are the entrainment and deposition rates, $V$ is the particle velocity during the motion phase, and $k = k_E+k_D$.
+This partial differential equation is called an asymmetric telegrapher's equation in mathematical physics \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is rarely encountered in the literature.
+
+For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
+\be \ee
+Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase.
+The variance, however, is different.
+Computing $\sigma_x^2$ provides
+\be \sigma_x^2(t) = ,\ee
+which is a non-trivial result.
+At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2014}.
+At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
+In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not.
+
+\subsection{The Newtonian approach}
+
+Some authors have attempted to model local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
+Early efforts impart particles with time-averaged fluid forces, typically linked to a logarithmic flow velocity profile \citep{VanRijn1984}, and some include simplified collision forces that modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992}.
+Later on, researchers began to include granular interactions among particles to model collisions using the discrete element method \citep{Cundall1979, Haff1993}.
+The early works utilizing this approach used a two dimensional domain with a highly simplified flow model \citep{Jiang1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow.
+The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, using particles as the boundary condition for the fluid \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowincklel2014,Elghannay2017,Yousefi2020}.
+A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2011, Azema2012, Wachs2019}.
+These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytically tractable models remain necessary for holistic understanding of sediment transport.
+
+\subsection{Langevin models for the sediment velocity distribution}
+
+The final set of models relevant to this thesis are the couple of analytical models that have been developed to describe particle velocities in the local and intermediate ranges when velocities fluctuate due to turbulence and particle-bed collisions.
+These models are indended to apply only between entrainment and deposition.
+In contrast to the constant velocities introduced in section \ref{sec:lisle}, the velocities of bed load particles fluctuate through time and are best represented by statistical distributions. 
+Among experimental studies on bedload velocities, two dominant conclusions have emerged.
+One subset of observations indicates that bedload velocities lie on exponential distributions \citep{Lajeunesse2010,Furbish2012,Fathel2015}, and another subset indicates Gaussian distributions \citep{Martin2012,Ancey2014,Heyman2016}.
+
+\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities with a mechanistic model including a noisy driving term to represent fluid turbulence.
+They wrote, for the streamwise particle velocity $u$, a Langevin equation
+\be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
+This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of particle velocity fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
+The fluid drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
+The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1984,VanKampen2007}
+\be \pt P(u,t) = \Delta\partial_u\Big[ \text{sgn}(u) P \Big] + D \partial_u^2 P,\ee
+implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by equation \ref{eq:fanlangevin} is
+\be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
+This is the (two-sided) exponential distribution observed in one subset of the aforementioned experiments.
+
+In a similar approach, \citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments.
+They wrote for the streamwise velocity
+\be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
+where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance.
+The timescale $t_r$ is a relaxation time over which velocity fluctuations decay.
+This is the well-known Ornstein-Uhlenbeck process of stochastic physics \citep{Gardiner1983}, and its Fokker-Planck equation is
+\be \pt P(u,t) = -\partial_u\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2} \partial_u^2 P.\ee
+This time, the steady state solution becomes
+\be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
+which is the Gaussian velocity distribution from the other subset of the experiments.
+
+These models produce successful descriptions of disparate experimental results, although their components are not clearly interpretable in terms of physical expecatations.
+For example, granular interactions were included in the Fan et al model as Coulomb friction, and while such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo subsequent collisions with the rough granular bed: although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
+In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{CLift1978}.
+We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
 

### 2021-06-27 ch1intro.tex
finished a few figures and made lots of progress on the introduction.

@@
 
-\section{theories of individual particle movements}
+\section{Theories of individual particle movement}
 
@@
 
-\subsection{The tracer problem}
-The original motivation to understand individual particle motions was probably to understand the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research today \citep{Hassan2017,Pretzlav2021}.
+\subsection{Motivation: tracers and basic understanding}
+The original motivation to understand individual particle motions was probably to understand the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
 A common measurement technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
-In principle, if tracers are a proxy for the population of grains in a stream, one can estimate the velocities of tracers, then multiply by an estimate of the number of grains available for motion in the stream to calculate the overall flux \citep{Yano1969,Nakagawa1976,Hassan2017}.
-Unfortunately, this does not really work, due to the distinct behavior of tracers over the timescales identified above.
-Apparently, tracer particles slow down through time within streams \citep{Ferguson2002,Haschenberger2012}.
-This means it is not clear how measured tracer velocities can be mapped back to bulk sediment fluxes, and it exposes a need for further understanding of how exactly particle motion characteristics differ across the local, intermediate, global, and geomorphic timescales.
+In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion in the stream to calculate the overall sediment flux \citep{Yano1969,Nakagawa1976,Hassan2017}.
+In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales.
+Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenberger2012}.
+This means it is not clear how measured tracer velocities can be mapped back to bulk sediment fluxes, exposing a need for further research into how exactly particle motion characteristics evolve through time.
 
 This need to understand tracer movements is not however the sole point of research into individual particle motions.
-Modern science trains us to reduce systems to their component parts in order to glean better understanding of the system as a whole.
-It is conceptually clear that sediment transport and geomorphology are intimately linked to the movements of individual grains.
-If we understand them better, research progress will probably follow.
+It is conceptually clear that bulk sediment fluxes ultimately result from the aggregated movements of individual grains, and the prediction of the sediment flux has long been acknowledged as a challenging problem with no clear solution, which impedes geomorphology and engineering understanding \citep{Ancey2020,Ancey2020a}.
+Better understanding of individual particle motions will support this research.
 
-\subsection{ Einstein 1937}
+\subsection{Einstein 1937}
 Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}, motivated (at least as far as his research supervisor Meyer-Peter was concerned) by the need to understand the efficiency of Helley-Smith samplers for the management of sedimentation \citep{Ettema2004}.
-Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to a probabilistic approach to characterize their motion.
-Einstein's basic insight was to represent particle motions as an alternating sequence of motions and rests having random characteristics.
-As his interest was on the global range of particle motion, and the duration of particle motions is usually short compared to rests, Einstein approximated individual motions (between entrainment and deposition) as instantaneous.
-With this setup, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
+Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to probabilistic methods to characterize their transport.
+Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
+As his interest was on the global range of particle motion, and the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
+With this configuration, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
 Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
-Einstein's experiments indicated that both step lengths and resting times were drawn from exponential distributions.
-His target was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
+Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
 He formulated the problem mathematically with an infinite series of convolution integrals, in an elegant display of mathematical physics.
+The mathematics Einstein took on were a pioneering application of the continuous time random walk, which was not formalized until much later \citep{Montroll1965}.
 
-For connection to later work in the thesis, I will summarize Einstein's work slightly differently.
+For connection to later work in the thesis, and to provide additional perspective on Einstein's work which is not available in the literature, I will summarize Einstein's work slightly differently than he originally formulated it, using a method based on a stochastic dynamical equation.
 In effect, if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
-\be \dot{x}(t) = \eta(t), \label{eq:einlangevin}\ee
-where $\dot{x} = dx/dt$, and $\eta(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $l$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
-A particular realization of this noise can be written
-\be \eta(t) = \sum{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
-where $N(t)$ is the number of particle entrainments in time $t$, $the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are distributed according to $P(s) = l^{-1}\exp(-s/l).$
+\be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
+where $\dot{x} = dx/dt$, and $\mu(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
+A particular realization of this pulsed noise can be written
+\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
+where $N(t)$ is the number of particle entrainments in time $t$, distributed as a Poisson distribution $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
 Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
@@
 
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
+	\caption{Panel (a) indicates the representation of Einstein's model as an idealized ``white shot noise", as indicated in eq.
+\ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$).
+}
+	\label{fig:einsteinfig}
+\end{figure}
+
 The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1984,Moss1989}.
 Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
-\be  \big(l \px \pt + k_E l \px + pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
-This equation can be solved by standard methods (series solutions or transform calculus) to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
-\be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/l}\theta(x) \sqrt{\frac{k_E t}{lx}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{l}}\Big) \ee
+\be  \big(\ell \px \pt + k_E \ell \px + pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
+This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1986a} reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
+\be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/\ell}\theta(x) \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big) \ee
 Here, $\mathcal{I}$ is a modified Bessel function. 
@@
 This distribution displays both advective and diffusive characteristics.
-One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^l$ and integrating over space. This gives the mean position of the particle
-\be \langle x \rangle (t) = k_E l t, \ee
-so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E l$, given by the entrainment rate times the mean step length.
+One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
+\be \langle x \rangle (t) = k_E \ell t, \ee
+so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$, given by the entrainment rate times the mean step length.
 The rate at which one particle spreads out from another due to differences in their motion characteristics can be represented by the variance of position, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$.
 This gives
-\be \sigma_x^2(t) = 2 k_E l^2 t, \ee
-so particles in Einstein's model spread apart as a normal diffusion process \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E l^2.$
+\be \sigma_x^2(t) = 2 k_E \ell^2 t, \ee
+so particles in Einstein's model spread apart as a normal diffusion process \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$
 
@@
 
-Einstein's model of particle movement seems to have been overlooked for several decades, or possibly overshadowed by his later work, which uses its essential ideas for more obviously practical purposes.
-Eventually, curiosity into individual particle motions was re-sparked in the context of cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}, so Einstein's model received a great deal of new attention.
+Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest, but it is clearly oversimplified from the reality of particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
+
+Einstein's model of particle movement seems to have been overlooked for the first several decades after its development, or possibly overshadowed by Einstein's later work, which uses its essential ideas for more obviously practical purposes.
+Eventually, curiosity into individual particle motions was re-sparked by cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}.
+In relation to these issues, Einstein's model received a great deal of new attention which generated some generalizations of the model.
 
@@
 Einsteins flume experiments demonstrated that particle step lengths followed exponential distributions \citep{1937}, and this was confirmed by subsequent studies \citep{Yano1969, Nakagawa1976}.
-Yet these studies were all conducted in steady flows in artificial flumes, not real streams.
-When sediment transport is observed before and after a flood, the step length within a flood is better described by a Gamma distribution, this being the sum across the many individual steps that occurred during the flood \citep{Hassan1991}.
-When sediment transport occurs over dunes, similar modifications were made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965}.
+Yet these studies were all conducted in steady flows in artificial flumes, not real streams, where sediment transport conditions can be different.
+When sediment transport is observed before and after a flood, rather than during steady flow, the step lengths are better described by a Gamma distribution, this being the sum across the numerous individual steps that occurred during the flood \citep{Hassan1991}.
+When sediment transport occurs in the presence of dunes or other bedforms, similar modifications have been made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965} and the embedding of particles within bedforms (which increases the periods of rest) \citep{Yang1971,Nakagawa1980}.
 
-\subsection{ Inclusion of the movement duration}
+\subsection{Inclusion of the movement duration}
 \label{sec:lisle}
 
-Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescale of individual particle movements. 
-Yet the advent of high speed camera experiments of bed load transport produced data on the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which Einstein never intended to model.
-In the local range, particles move with a velocity that fluctuates due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
-Einstein's infinite movement velocity assumption sets aside these details.
+Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles embed in the subsurface or other sedimentary deposits (geomorphic range).
+Starting in the 1970s, the advent of high speed camera experiments of bed load transporthave produced data on the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which \citet{Einstein1937} never intended to describe.
+In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
+In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
+Einstein's infinite movement velocity assumption excludes the timescales over which these processes occur.
 
-Studies by \citet{Gordon1972}, \citet{Lisle1998}, and \citet{Lajeunesse2017} incorporated the duration of sediment motion.
-They approximated particle velocities as constant, and assumed that the motion time is also an exponentially distributed random variable, this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion after an initial entrainment.
-The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ is 
+Studies by \citet{Gordon1972}, \citet{Lisle1998}, and \citet{Lajeunesse2017} have generalized the Einstein theory to include the duration of sediment motion.
+They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are also (along with rests) exponentially distributed random variables, this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion (between entrainment and deposition).
+The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written as
 \be \dot{x} = V\eta(t), \ee
@@
 This partial differential equation is called an asymmetric telegrapher's equation in mathematical physics \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is rarely encountered in the literature.
+
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/lisleConcept.pdf}
+	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise eq.
+\ref{eq:lislerando}, while panel (b) shows the slanted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (which have mean duration $1/k_E$).
+}
+	\label{fig:lislefig}
+\end{figure}
 
@@
 
-\subsection{Langevin models for the sediment velocity distribution}
+\subsection{Mechanistic-stochastic models for the sediment velocity distribution}
 
@@
 These models produce successful descriptions of disparate experimental results, although their components are not clearly interpretable in terms of physical expecatations.
-For example, granular interactions were included in the Fan et al model as Coulomb friction, and while such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo subsequent collisions with the rough granular bed: although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
-In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{CLift1978}.
+For example, granular interactions were included in the Fan et al model as a Coulomb friction term.
+While such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo intermittent collisions with the granular bed.
+Although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
+In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{Clift1978}.
 We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
+
+\subsection{Next steps for individual particle motions}
+
+\section{The bulk bedload flux}
+
+\subsection{The definition of the flux}
+
+\subsection{The scaling arguments of Bagnold}
+
+One of the most influential formulations of the bedload flux is due to Bagnold \citep{Bagnold1956,Bagnold1966}, who derived a formula for the mean sediment flux using energy balance arguments.
+Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
+He assumed that the flow power $P_f$ available to move sediment scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
+Considering that the average downstream flux of particles is $q$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ required to sustain particle motion scales as $P_g \propto q/\tau^{1/2}.$
+ Balancing flow energy against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
+\be q = k(\tau-\tau_c)\tau^{1/2}, \label{eq:bagnold}\ee
+which has shown good correspondence with laboratory data at large transport rates, given careful calibration of the constant factors $k$ and $\tau_c$.
+
+The large shear stress limit $q \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1998}.
+The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of the details have turned out to be incorrect.
+For example, Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and in reality, the flow power dissipated by sediment transport shows only a weak correlation the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
+These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$, which is related in Bagnold's formalism to the angle of repose of sediment, and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
+Finally, Bagnold's formulation faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}, which is a topic for another thesis.
+The difficulties with Bagnold's approach led many followers to revise his theory, merging his approach, which is based on Bagnold's characteristically sharp physical insight, with new experimental conclusions on the energetics of sediment transport \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
+
+\subsection{Einstein's probabilistic approach}
+
+Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
+Einstein's formulated his original model of individual grains in transport \citep{Einstein1937} in terms of particle entrainment and deposition, supplemented with the idea that grains move downstream through a sequence of instantaneous steps.
+Later, \citet{Einstein1942,Einstein1950} formulated the bulk sediment flux using these same ideas, providing an alternative to Bagnold based on probabilistic arguments.
+The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
+ \begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
+	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
+left to right through an array of adjacent control volumes. The bedload flux is the volume (or number) of bedload particles crossing
+the surface $\mathcal{S}$ per unit width and time.}
+	\label{fig:lislefig}
+\end{figure}
+
+To derive the mean bedload flux, Einstein partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
+Each control volume has downstream length $\ell$ which is also the average particle step length.
+Denoting by $P_n$ the probability that an individual grain undergoes at least $n$ jumps of length $\ell$ in a time interval $T$, meaning it travels at least a distance $n \ell$, and that there is a density $\rho$ of particles at rest on the bed, it follows that on average $\rho \ell P_n$ particles will displace a distance $n \ell$ or more from each control volume within the time interval $T$.
+As a result, since grains crossing $\mathcal{S}$ in a time $T$ could have come from any upstream location, the number of grains crossing $\mathcal{S}$ in $T$ is a sum over all control volumes: $\sum_{n=1}^\infty \rho \ell P_n$.
+Dividing by the time $T$ to get the average rate of grains crossing $\mathcal{S}$ gives the mean flux
+\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty P_n. \label{eq:einflux} \ee
+The reamining quantity to evaluate is $P_n$, the probability a particle entrains \textit{at least} $n$ times in a time $T$.
+
+Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, but this approach has been critiqued by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
+Instead, one should calculate $P_n$ as an exceedance probability.
+If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution:
+\be p_n = \frac{(k_E T)^n}{n!}e^{-k_E T}.\ee
+This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^n}{n!}e^{-k_E T}, $
+so Einstein's mean sediment flux (eq. \ref(eq:einflux}) becomes
+\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
+Noticing that $\rho k_E$ is the entrainment rate of a single grain multiplied by the density of grains available for entrainment on the bed, we can write 
+\be q = E \ell \ee
+as the central result of Einstein's theory.
+Here, the quantity $E = \rho k_E$ is the areal entrainment rate, representing the number of grains entrained per unit bed area \citep{Wilcock1997,Furbish2012}.
+
+Alongside the central result $q=E\ell$ for the mean flux, one of Einstein's most influential and enduring ideas was his formulation of the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
+The original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
+The entrainment rate is calculated from the exceedance probability of the turbulent lift over the weight, providing an alternative to the critical shear stress concept which explicitly incorporates turbulent fluctuations in the fluid flow.
+This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
+random) resisting quantities \citep{Grass1970} is a key part of Einstein's legacy. \citet{Paintal1971} made a significant extension
+by including the random supporting geometry of bed particles into the force balance. More recently, \citet{Tregnaghi2012}
+ammended the theory to include both force magnitude and duration \citep{Diplas2008, Valyrakis2012, Celik2014}.
+Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and remain under active development.
+
+\subsection{Fusing Einstein and Bagnold}
+
+Although Einstein worked to relate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character which is not clearly linked to the underlying fluid-granular physics.
+The erosion-deposition model originally developed by \citet{Charru2004,Charru2006,Charru2006a} relates the central parameters of the Einstein model to properties of the flow using scaling relations \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
+This can be characterized as a mixture of the Einstein and Bagnold strategies.
+
+Derived by evaluating mass balance within a control volume, the erosion-deposition model is
+\be \pt \gamma +  \px V \gamma = E - D. \label{eq:charru}\ee
+In this equation, $\gamma$ is the ``particle activity", which is the number of moving particles per unit area \citep{Furbish2012}, $V$ is the ensemble average movement velocity of sediment grains (which in unsteady conditions may depend on space and time), $E$ is the entrainment rate (the number of particles transitioning to motion per unit area and time), and $D$ is the deposition rate (the number of particles coming to rest on the bed per unit area and time).
+
+Scaling arguments provide relations for $E$, $D$, and $V$ in terms of the fluid shear stress $\tau$, particle size $d$, particle settling velocity $V_s$, and the critical shear stress $\tau_c$:
+\be E = a \frac{\tau-\tau_c}{d^3 V_s}, \label{eq:charru1}\ee
+\be D  = b\frac{\gamma V_s}{d}, \ee
+\be V = c + d(\sqrt{\tau}-\sqrt{\tau_c}). \label{eq:charru2}\ee
+The constant coefficients $a$, $b$, $c$, and $d$ are determined experimentally.
+
+Equation \ref{eq:charru} indicates that the mean flux in steady transport conditions is the implicit solution to the equation $E = D.$
+Using the scaling relations \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
+\be \gamma \propto \frac{\tau-\tau_c}{d^2 V_s^2}.\ee
+Expressing the mean flux as $q = \gamma V$, in the control volume interpretation, the relationship between flux and bed shear stress becomes
+\be q = \frac{A}{d^2 V_s^2}\big(\tau-\tau_c\big)\big[c + d(\sqrt{\tau}-\sqrt{\tau_c})\big]. \ee
+This recovers the Bagnold scaling $q \propto \tau^{3/2}$ at large bed shear stresses.
+
+\subsection{The nonlocal formulation}
+
+Einstein's model of the bulk bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009}.
+\citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
+\be q(x,t) = \int_0^\infty dx' F(x-x',t)E(x',t). \ee
+In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a just-entrained particle steps a distance $l$ before deposition.
+
+\citet{Furbish2012} generalized the Parker model to include a finite duration of motion.
+They wrote
+\be q(x,t) = \int_0^\infty dx' \int_0^\infty dt' F(x',t') E(x-x',t-t'), \label{eq:furbo}\ee
+where, $F(x,t)$ represents the probability density that particles move \textit{at least} a distance x in time $t$ (right after entrainment).
+For a simple example of the Furbish et al formalism, consider particles that move with a constant velocity $V$ and have a deposition rate $k_D$. 
+Then the probability density that a particle moves \textit{exactly} a distance $x$ in time $t$ is $f(x,t) = \delta(x-Vt)k_D\exp(-k_D t), $ so the probability that it moves 
+\textit{at least} a distance $x$ in $t$ is
+\be F(x,t) = \int_0^t \delta(x-Vt)k_D\exp(-k_D t) = \theta(Vt-x)k_D\exp(-k_D t).\ee
+Considering uniform conditions with a density $\rho$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the bulk entrainment rate can be expressed as $E=\rho k_E$, and equation \ref{eq:furbo} provides a mean flux
+\be q = \rho k_E \int_0^\infty dx' \int_0^\infty dt' \theta(Vt-x)k_D\exp(-k_D t) = \rho k_E V/k_D. \label{eq:furbflux}\ee
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$. Equation \ref{eq:furbflux} provides another perspective on Einstein's result $q= E\ell$.
+
+Furbish et al also provided a local approximation of the nonlocal flux expressed equation \ref{eq:furbo}.
+This result is useful for unsteady conditions when spatial gradients occur in the particle activity.
+
+\subsection{The connection to landscape evolution}
+
+Exner was probably the first to write the mathematical relationship between sediment transport and topographic change \citep{Exner1925}. He wrote
+\be (1-\lambda)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \ee
+This equation links the temporal evolution of the land elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux $q$. The parameter $\lambda$ is the bed porosity.
+
+Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuous fields.
+This representation can be interpreted as an average over the detailed locations of individual grains, and it is expected valid whenever the scales of interest are large compared to the size of the averaging window, which should be taken much larger than the size of the largest grains being modelled \citep{Coleman2009}.
+Yet there are many contexts on Earth's surface where the scale of interest is not much larger than the largest grains involved in the system.
+We can wonder, for example, in mountain channels, how large boulders control to the formation of steps \citep{Church2007,Zimmerman2008,Saletti2020}, or other structures with sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
+In these cases, individual grains are comparable to the scales of interest, and the continuum description provided by the Exner equation seems questionable.
+
+\subsection{Fluctuations and scale dependence}
+
+\subsection{Birth death models}
+
+\subsection{Queueing theories}
+
+\subsection{Next steps for the bedload flux}
+
+\endinput
 
@@
 
-Sediment moves in rivers when the tractive force of the water flow overcomes any resisting forces by gravity and friction.
+In river science, a fundamental problem is the determination of the bedload flux, or the rate of downstream movement of coarse sediment grains \citep{Ballio2014}.
+Since bedload transport has strong feedback with stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of environmental considerations. 
+These considerations span a wide range, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}. 
+Unfortunately, no satisfying approach currently exists to compute the bedload flux \citep{Ancey2020,Ancey2020a}, and predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
+
+Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
+Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
+The same turbulence and bed sediment organization details which correlate with the bedload flux are also modified by it.
+Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, while moving sediment affects turublence characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload transport modifies the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
+The bedload flux is in a cyclical feedback with its control variables, the details of turbulence and bed organization \citep{Jerolmack2005}.
 
@@
 
+Perhaps a surprising outcome of over a century of bedload transport research is that no one definition of the sediment flux has yet been agreed upon.
+Today, there are two main competing (or complementary) approaches to define the bedload flux. 
+The first definition is reminiscent of continuum mechanics or electrodynamics \citep{Landau, Jackson19xx}, and formulates the flux as a kind of current of sediment across (normal to) a control surface $\matchal{S}$ \citep{Furbish2012,Heyman2016}:
+\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\textbf{\mathcal{S}}. \ee
+This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field over the scales of interest \citep{Heyman2016}.
+Lately, this concentration has been interpreted as an ensemble average \citep{Furbish2012,Ballio2014}.
+The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $\mathcal{V}$:
+\be q = \frac{1}{L} \sum_{i\in\mathcal{V}} u_i.\ee
+Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
+
 \subsection{Bagnold's energy balance}
-
-One of the most influential approaches from physical arguments is due to Bagnold \citep{Bagnold1956,Bagnold1966}.
-Bagnold described sediment transport as a process which ultimately converts flow energy to heat.
-He assumed that the flow power $P_f$ available to move sediment scaled as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
-He reasoned that if the average downstream flux of particles is $q$, and particles move on average with velocity proportional to the fluid velocity near the bed, then the power $P_g$ required to sustain particle motion scales as $P_g \propto q/\tau^{1/2}.
-$ Balancing fluid propulsion against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
-\be q = k(\tau-\tau_c)\tau^{1/2}, \label{eq:bagnold}\ee
-which has shown good correspondence with laboratory data after careful calibration of the constant factors $k$ and $\tau_c$.
-The large shear stress limit $q \sim \tau^{3/2}$ is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972}
 
@@
 
-Exner was probably the first to describe landscape evolution as a mathematical result of sediment transport \citep{Exner1925}, when he related the topographic elevation to the sediment flux via
-\be (1-\lambda)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \ee
-This equation links the temporal evolution of the land elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux. The parameter $\lambda$ is the bed porosity.
-
-Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuous fields.
-This representation can be interpreted as an average over the detailed locations of individual grains, and it is expected to be an excellent approximation whenever the scales of interest are large compared to the size of the averaging window, which should be taken much larger than the size of the largest grains being modelled \citep{Coleman2009}.
-
-Yet there are many contexts on Earth's surface where the scale of interest is not much larger than the largest grains involved in the system.
-We can wonder, for example, in mountain channels, how large boulders contribute to the formation of steps \citep{Zimmerman2008,Saletti2020}, or other structures with sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
-In these cases, individual grains are comparable to the scales of interest, and the continuum approximation breaks down.
-
 \subsection{Tusjimoto and beyond}
@@
 
-\section{Realism versus complexity}
-
-\section{The point of the thesis}
-
-\subsection{Open problems in the stochastic paradigm}
-
-\subsection{Quantities of interest}
-Explain what the pdf of position is
-explain what the variance represents
-explain that the moments more generally are of interest
-explain that the sediment flux is key to landscape evolution. not just in mean value, but also in its fluctuations.
-
-\subsection{Tools of the trade}
-
-\subsubsection{The Langevin equation}
-When particles are subjected to fluctuating forces, their Newtonian equation of motion can be written
-
-In this equation, $F$ is the deterministic or steady component of the forcing, while $\xi$ is the fluctuating component. 
-The classic example is due to Einstein, and 
-\subsubsection{Master equations and probability flow}
-
-\subsubsection{Idealized noises}
+\section{Topics and organization of the thesis}
 
@@
 The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
-Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
+Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turublence characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
 The bedload flux is in a cyclical feedback with its controls: these controls are the details of turbulence and bed organization \citep{Jerolmack2005}.
@@
 Many investigators have criticized and developed Einstein's ideas \citep{Paintal1971, Yalin1972, Shen1980, Lisle1998, Papanicolaou2002, Sun2000, Ancey2006, Armanini2015}. 
-These revisions focus on the more ad hoc elements of Einstein's derivations, and they lend a more mechanistic basis, a firmer mathematical foundation, and more generality to the approach. 
+These revisions focus on the more simplified aspects of Einstein's derivations, and they lend a more mechanistic basis, a firmer mathematical foundation, and more generality to the approach. 
 Although all of these authors accepted the statistical character of bedload motion, none developed theories for a probability distribution for the bedload flux. 
@@
 
-\endinput
-
 Any text after an \endinput is ignored.

### 2021-06-27 ch1intro.tex
got a lot of progress on the intro.

@@
 
-\section{Geomorphology traditions}
-Geomorphology can be defined as the study of Earth's evolution by tectonics, weathering, and erosion \citep{Chorley1984}. 
-This science has long been characterized by observation and description \citep{Gilbert1877, Tight1903, Hack1960}.
-Only relatively recently did the science turn to quantitative methods \citep{Bagnold1941, Strahler1952, Scheidegger1961, Leopold1962}, with researchers working to frame observations in terms of underlying processes and to describe them with mathematical models modified from physics \citep{Church2005,Roades1996,Chorley1965}.
-
-This quantitative shift or incorporation of physics methods into geomorphology has been criticized as overly reductionist \citep{Slaymaker2020}.
-Many of the problems addressed within physics involve closed systems with perfect order \citep{GoldSteinClassicalMechanics,SakuraiQM,JacksonElectrodynamics}. 
-These systems stand in contrast to the systems in geomorphology.
-Earth's surface is open and highly disordered.
-The surface receives fluxes of mass and energy across its boundaries by tectonics, climate, transpiration, and sunlight, and its component parts vary widely in their characteristics, activities, and compositions.
-With this complexity \citep{Chorley1971}, we can conclude that a complete reduction of geomorphology to physics is likely impossible.
-
-There are nevertheless lessons physics can exchange with geomorphology and sediment transport theory.
-This is the central theme of this thesis.
-Before presenting original research, the first chapter summarizes key developments in quantitative geomorphology models which have resulted from this exchange.
-The section is organized along two classifications.
-First, I distinguish earlier works as ``stochastic" or ``deterministic".
-Stochastic models introduce randomness or noise as a means to crudely represent the complexity of natural systems (open boundaries, diversity in component parts) within idealized mathematical models of nature.
-Deterministic models involve Newtonian equations without noise to represent nature.
-Second, I distinguish earlier models as ``discrete" or ``continuous".
-These terms reflect whether the degrees of freedom of individual particles are retained in mathematical models, or whether they are averaged away.
-When grains are tracked, a model is discrete.
-When they are discussed as a flow, it is continuous.
-We can therefore discuss a ``discrete and stochastic" model, or a ``continuous and deterministic" model, and so on.
-With these distinctions in mind, we launch into a survey of the foundation upon which this thesis builds: physics-inspired, process-based models of sediment transport and landscape evolution.
+Landscales change when sediment moves.
 
@@
 
-A fundamental problem in sediment transport modelling is to predict the downstream movement of an individual particle.
-At first glance, this seems a basic problem in general physics, but digging in exposes severe challenges.
+A basic (but not simple) problem in sediment transport modelling is to predict the downstream movement of an individual particle.
+At first glance, maybe at the idealistic beginnings of one's PhD, this seems an elementary problem in general physics, but then some challenges appear.
 Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not known \citep{Furbish1997}.
@@
 Predicting the movements of individual particles therefore involves many contributing processes, and our understanding of each of these processes remains incomplete.
-\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize this complexity, although it has been slightly revised more recently \citep{Campagnol2013, Hassan2017, Pierce2020}.
+\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize the challenges and research progress, although it has been slightly revised in more recent studies \citep{Campagnol2013, Hassan2017, Pierce2020}.
 Nikora et al divided the downstream trajectory of an individual particle into three timescales, or ``ranges", termed local, intermediate, and global.
@@
 \subsection{Mechanistic-stochastic models for the sediment velocity distribution}
+\label{sec:langevin}
 
@@
 
-\subsection{Next steps for individual particle motions}
+\section{The bedload transport problem}
 
-\section{The bulk bedload flux}
+In river science, a fundamental problem is the determination of the bulk bedload flux, or the rate of downstream movement of coarse sediment grains \citep{Ballio2014}.
+Since bedload transport controls stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of problems, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}.
+Unfortunately, no satisfying approach currently exists to compute the bedload flux \citep{Ancey2020,Ancey2020a}, and predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
+
+Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
+Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
+The same turbulence and bed sediment organization details which correlate with the bedload flux are also modified by it.
+Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turublence characteristics \citep{Singh2010, Santos2014, Liu2016}.
+Bedload transport is controlled by the configuration \citep{Miller1966, Paintal1971,} and consolidation \citep{Johnson2016,Allen2018,Masteller2019} of grains on the bed surface, but bedload transport also modifies the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
+The bedload flux controls its control variables. It is in a cyclical feedback with the details of turbulence and bed organization \citep{Jerolmack2005}, so predicting it is challenging.
 
@@
 
+Perhaps a surprising observation of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
+Today, there are two main competing (or complementary) approaches to define the bedload flux. 
+The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\matchal{S}$ \citep{Furbish2012,Heyman2016}: 
+\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\textbf{\mathcal{S}}. \ee
+This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field over the scales of interest \citep{Heyman2016}.
+Lately, this concentration has been interpreted as an ensemble average \citep{Furbish2012,Ballio2014}.
+The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $\mathcal{V}$:
+\be q = \frac{1}{L} \sum_{i\in\mathcal{V}} u_i. \label{eq:controlflux}\ee
+Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
+
+An alternative statistical formulation of the sediment flux provides an approximate correspondence between these two definitions \citep{Ancey2006,Furbish2012}.
+This alternative definition hinges on the probability $P[\textbf{u}_p | \textbf{x},t]$ that a particle contacts a control surface $S$ at position $\textbf{x}$ and time $t$ with velocity $\textbf{u}_p$.
+This conditional probability is considered to result from a very large collection of identical systems selected at random moments in their evolution: that is, the conditional probability is an ensemble quantity \citep{Kittel1958}.
+
+In terms of this ensemble probability, the flux is \citep{Ancey2006}: 
+\be q_s = \int_{\text{all } \textbf{u}_p} \int_S P[\textbf{u}_p | \textbf{x},t] \textbf{u}_p \cdot d \textbf{S} d \textbf{u}_p . \label{eq:anceyflux} \ee
+In steady conditions, when the probability is independent of time -- $\partial P / \partial t = 0$, the ensemble definition of the flux \ref{eq:anceyflux} can be approximated by the number of moving particles in the control volume.
+
+An approximate connection to a control volume flux can be derived by swapping the ensemble average for a spatial average, by integrating along a control volume.  
+If the control volume is sufficiently long, $L \rightarrow \infty$, it can be seen as a stack of very many independent cross sectional surfaces. 
+These comprise a stack of replicas of $\mathcal{S}$, and at an instant, if spatial correlations in bedload transport are weak, each surface constitutes one configuration of particles intersecting $S$ with some set of positions and velocities contributing to $P$.
+One can then define $P$ by counting occurrences along this stack of replica surfaces with an integral across the control volume.
+
+With this concept of replica surfaces in mind, we can formalize the link with symbols. 
+Consider $n$ particles distributed throughout a volume at some set of positions $\textbf{x}_i$ with some set of velocities $\textbf{u}_i$, where $i=0,1,\dots, n$,
+The cloud of particles can be represented by its (discrete) density in a position-velocity phase space: 
+\be \rho(\textbf{x},\textbf{u}_p) = \sum_{i=1}^n M(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i). \ee  
+Here $M(\textbf{z})$ is a marker function, which is $1$ if $\textbf{z}$ is inside the particle, and $0$ otherwise.
+The marker's volume integral is $\int M(\textbf{x}) dV = \nu_p$, the particle volume.
+
+Using this density at sufficiently large $L$, the conditional probability $P[\textbf{u}_p | \textbf{x}]$ can be written approximately as an integral over the stack of replica surfaces:
+\be P[\textbf{u}_p|\textbf{x}] \approx \frac{1}{L} \int_0^L dx \rho(\textbf{x},\textbf{u}_p). \ee
+The integral runs along the downstream coordinate, and the equality becomes exact as $L \rightarrow \infty$.
+With this swap of ensemble averaging for spatial averaging, the bedload flux becomes 
+\be q_s \approx \int_{\text{all} \textbf{u}_p} \int_S \frac{1}{L} \int_0^L dx \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i) \textbf{u}_p \cdot \textbf{k} dS d \textbf{u}_p = \frac{\nu_p}{L}\sum_{i=1}^n u_i , \ee
+so there is some correspondence between the surface and volume definitions of the flux.
+Yet this is never exact, except in the unphysical case of sediment transport with no spatial correlations (vanishing particle size, discontinuous trajectories).
+Apparently, the control volume and control surface formulations of the sediment flux are not equivalent, so we see in the literature two parallel streams of research.
+
 \subsection{The scaling arguments of Bagnold}
 
-One of the most influential formulations of the bedload flux is due to Bagnold \citep{Bagnold1956,Bagnold1966}, who derived a formula for the mean sediment flux using energy balance arguments.
+One of the most influential formulations of the bedload flux is due to Bagnold \citep{Bagnold1956,Bagnold1966}, who derived a formula for the mean sediment flux using an energy balance approach.
 Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
@@
 The large shear stress limit $q \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1998}.
-The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of the details have turned out to be incorrect.
-For example, Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and in reality, the flow power dissipated by sediment transport shows only a weak correlation the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
-These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$, which is related in Bagnold's formalism to the angle of repose of sediment, and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
-Finally, Bagnold's formulation faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}, which is a topic for another thesis.
-The difficulties with Bagnold's approach led many followers to revise his theory, merging his approach, which is based on Bagnold's characteristically sharp physical insight, with new experimental conclusions on the energetics of sediment transport \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
+The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of the details have since turned out to be incorrect.
+Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and the flow power dissipated by sediment transport shows only a weak correlation the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
+These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$ and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
+More generally, Bagnold's formulation faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}, which is a topic for another thesis.
+The difficulties with Bagnold's approach led to many revisions of his theory which kept Bagnold's sharp physical insights, but included new experimental conclusions on the energetics of sediment transport as they became available \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
 
@@
 Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
-Einstein's formulated his original model of individual grains in transport \citep{Einstein1937} in terms of particle entrainment and deposition, supplemented with the idea that grains move downstream through a sequence of instantaneous steps.
-Later, \citet{Einstein1942,Einstein1950} formulated the bulk sediment flux using these same ideas, providing an alternative to Bagnold based on probabilistic arguments.
+Einstein formulated his original model of individual grains in transport \citep{Einstein1937} in terms of particle entrainment and deposition using the conceptual picture of grains moving downstream through a sequence of instantaneous steps.
+Later, \citet{Einstein1942,Einstein1950} calculated the bulk sediment flux with these same probabilistic ideas, providing an alternative to the Bagnold scaling approach.
 The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
@@
 	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
-left to right through an array of adjacent control volumes. The bedload flux is the volume (or number) of bedload particles crossing
-the surface $\mathcal{S}$ per unit width and time.}
+left to right through an array of adjacent control volumes. The bedload flux is the rate of bedload particles crossing
+the surface $\mathcal{S}$ per unit width and time, collected from all upstream control volumes.}
 	\label{fig:lislefig}
@@
 
-To derive the mean bedload flux, Einstein partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
+He partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
 Each control volume has downstream length $\ell$ which is also the average particle step length.
-Denoting by $P_n$ the probability that an individual grain undergoes at least $n$ jumps of length $\ell$ in a time interval $T$, meaning it travels at least a distance $n \ell$, and that there is a density $\rho$ of particles at rest on the bed, it follows that on average $\rho \ell P_n$ particles will displace a distance $n \ell$ or more from each control volume within the time interval $T$.
+Denoting by $P_n$ the probability that an individual grain undergoes at least $n$ jumps of length $\ell$ in a time interval $T$, meaning it travels at least a distance $n \ell$, and considering that there is a density $\rho$ of particles at rest on the bed, it follows that on average $\rho \ell P_n$ particles will displace a distance $n \ell$ or more from within each control volume during the time interval $T$.
 As a result, since grains crossing $\mathcal{S}$ in a time $T$ could have come from any upstream location, the number of grains crossing $\mathcal{S}$ in $T$ is a sum over all control volumes: $\sum_{n=1}^\infty \rho \ell P_n$.
-Dividing by the time $T$ to get the average rate of grains crossing $\mathcal{S}$ gives the mean flux
+Dividing by the time $T$ to get the average rate of grains crossing $\mathcal{S}$ provides the mean flux:
 \be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty P_n. \label{eq:einflux} \ee
-The reamining quantity to evaluate is $P_n$, the probability a particle entrains \textit{at least} $n$ times in a time $T$.
+The final quantity to evaluate is $P_n$, the probability a particle entrains \textit{at least} $n$ times in a time $T$.
 
-Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, but this approach has been critiqued by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
-Instead, one should calculate $P_n$ as an exceedance probability.
-If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution:
+Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$, but this approach has been revised by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
+Some of these authors have argued that instead, one should calculate $P_n$ as an exceedance probability.
+If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution \citep{Cox1965}:
 \be p_n = \frac{(k_E T)^n}{n!}e^{-k_E T}.\ee
@@
 \be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
-Noticing that $\rho k_E$ is the entrainment rate of a single grain multiplied by the density of grains available for entrainment on the bed, we can write 
-\be q = E \ell \ee
-as the central result of Einstein's theory.
-Here, the quantity $E = \rho k_E$ is the areal entrainment rate, representing the number of grains entrained per unit bed area \citep{Wilcock1997,Furbish2012}.
+Noticing that $\rho k_E$ is the entrainment rate of a single grain multiplied by the density of grains available for entrainment on the bed, we can summarize the Einstein theory as 
+\be q = E \ell, \ee
+where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit bed area \citep{Wilcock1997,Furbish2012}.
 
-Alongside the central result $q=E\ell$ for the mean flux, one of Einstein's most influential and enduring ideas was his formulation of the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
+Alongside this central result $q=E\ell$ for the mean flux, one of Einstein's most influential and enduring ideas was his formulation of the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
 The original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
@@
 ammended the theory to include both force magnitude and duration \citep{Diplas2008, Valyrakis2012, Celik2014}.
-Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and remain under active development.
+Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.
 
-\subsection{Fusing Einstein and Bagnold}
+\subsection{Fusing Einstein and Bagnold: The erosion-deposition model}
 
-Although Einstein worked to relate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character which is not clearly linked to the underlying fluid-granular physics.
-The erosion-deposition model originally developed by \citet{Charru2004,Charru2006,Charru2006a} relates the central parameters of the Einstein model to properties of the flow using scaling relations \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
+Although Einstein worked to relate the single-particle entrainment rate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character which is not clearly linked to the underlying fluid-granular physics.
+The erosion-deposition model originally developed by \citet{Charru2004,Charru2006,Charru2006a} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
 This can be characterized as a mixture of the Einstein and Bagnold strategies.
 
-Derived by evaluating mass balance within a control volume, the erosion-deposition model is
+Derived by evaluating a mass balance within a control volume, the erosion-deposition model is
 \be \pt \gamma +  \px V \gamma = E - D. \label{eq:charru}\ee
-In this equation, $\gamma$ is the ``particle activity", which is the number of moving particles per unit area \citep{Furbish2012}, $V$ is the ensemble average movement velocity of sediment grains (which in unsteady conditions may depend on space and time), $E$ is the entrainment rate (the number of particles transitioning to motion per unit area and time), and $D$ is the deposition rate (the number of particles coming to rest on the bed per unit area and time).
+In this equation, $\gamma$ is the ``particle activity", which is the number of moving particles per unit area \citep{Furbish2012}, $V$ is the ensemble averaged movement velocity of sediment grains (which in unsteady conditions may depend on space and time), $E$ is the areal entrainment rate (the number of particles transitioning to motion per unit area and time), and $D$ is the areal deposition rate (the number of particles coming to rest on the bed per unit area and time).
 
-Scaling arguments provide relations for $E$, $D$, and $V$ in terms of the fluid shear stress $\tau$, particle size $d$, particle settling velocity $V_s$, and the critical shear stress $\tau_c$:
+Scaling arguments provide relations for $E$, $D$, and $V$ in terms of the fluid shear stress $\tau$, particle size $d$, particle settling velocity $V_s$, and critical shear stress $\tau_c$:
 \be E = a \frac{\tau-\tau_c}{d^3 V_s}, \label{eq:charru1}\ee
@@
 In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a just-entrained particle steps a distance $l$ before deposition.
+This approach builds on Einstein because it can handle non-uniform conditions by space and time dependence of $E$ and $F$.
 
@@
 where, $F(x,t)$ represents the probability density that particles move \textit{at least} a distance x in time $t$ (right after entrainment).
-For a simple example of the Furbish et al formalism, consider particles that move with a constant velocity $V$ and have a deposition rate $k_D$. 
+For a simple example of the Furbish et al formalism, consider that particles move with a constant velocity $V$ and have a deposition rate $k_D$. 
 Then the probability density that a particle moves \textit{exactly} a distance $x$ in time $t$ is $f(x,t) = \delta(x-Vt)k_D\exp(-k_D t), $ so the probability that it moves 
@@
 
-Furbish et al also provided a local approximation of the nonlocal flux expressed equation \ref{eq:furbo}.
-This result is useful for unsteady conditions when spatial gradients occur in the particle activity.
-
-\subsection{The connection to landscape evolution}
+\subsection{Landscape evolution}
 
 Exner was probably the first to write the mathematical relationship between sediment transport and topographic change \citep{Exner1925}. He wrote
-\be (1-\lambda)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \ee
-This equation links the temporal evolution of the land elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux $q$. The parameter $\lambda$ is the bed porosity.
+\be (1-\phi)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \label{eq:exner}\ee
+This equation links the temporal evolution of the land elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux $q$. The parameter $\phi$ is the bed porosity.
 
 Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuous fields.
-This representation can be interpreted as an average over the detailed locations of individual grains, and it is expected valid whenever the scales of interest are large compared to the size of the averaging window, which should be taken much larger than the size of the largest grains being modelled \citep{Coleman2009}.
-Yet there are many contexts on Earth's surface where the scale of interest is not much larger than the largest grains involved in the system.
-We can wonder, for example, in mountain channels, how large boulders control to the formation of steps \citep{Church2007,Zimmerman2008,Saletti2020}, or other structures with sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
-In these cases, individual grains are comparable to the scales of interest, and the continuum description provided by the Exner equation seems questionable.
+This representation can be interpreted as a spatial average over the detailed positions of individual grains, and it is expected to be valid whenever the scales of interest are large compared to the size of the averaging window, which is required to be be taken much larger than the size of the largest grains in the modelling domain \citep{Coleman2009}.
+Yet there are many Earth surface processes where the scale of interest is not much larger than the largest grains involved in the system.
+We can wonder, for example, how large boulders in mountain channels control the formation of steps \citep{Church2007,Zimmerman2008,Saletti2020}, or other bed structures having sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
+In these cases, individual grains are comparable to the scales of interest, and the continuum description provided by the Exner equation is not applicable.
+
+\citet{Nakawaga1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation based upon Einstein's concepts of entrainment and deposition rates.
+According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition:
+\be \partial_x q(x,t) = E - D. \ee 
+Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described in an Einstein-like framework:
+\be (1-\phi) px z(x,t) = D - E.\ee
+In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the result of entrainment at all upstream locations, giving
+\be (1-\phi) px z(x,t) = \int_0^\infty \dx' int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
+This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
 
@@
 
-\subsection{Birth death models}
+Bedload flux signals exhibit characteristically large fluctuations \cite{Gomez1991,Bunte2005,Recking2012}, but all models reviewed to this point describe only mean sediment fluxes, and not the strength of bedload transport fluctuations, correlations, or other statistical characteristics of the flux.
+A full statistical characterization of sediment transport is desirable because transport fluctuations are responsible for the initiation of bedforms \citep{Jerolmack2005,Bohorquez2016} and are beginning to show links to basic geomorphology considerations like the maintenance of stable channel widths \citep{Abramian2019,Abramian2020}.
+This is not to mention that because sediment transport fluctuations are ubiquitous, models which describe only mean bedload fluxes signals are conceptually incomplete.
 
-\subsection{Queueing theories}
+An additional reason to characterize fluctuations in bed load fluxes is because in reality, sediment flux measurements always involve a sampling interval.
+This introduces questions of the convergence of transport rate measurements \citep{Dhont2018,Ancey2020}.
+Fluctuations originate from processes which span a vast range of characteristic timescales.
+These can occur over a matter of seconds, such as differences in the velocities of grains due to a series of collisions \citep{Benavides2021}, to many hundreds of hours, such as the migrations of dunes and bedload sheets \citep{Hammamori1962,Hoey1992,Guala2014} or the cycling between aggradation and degradation in pools \citep{Dhont2018}.
+Such processes introduce temporal correlations into sediment transport signals which control the sampling interval over which measurements of the mean sediment flux converge \citep{Saletti2015,Singh2009,Singh2012}.
+To date, very few modelling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
 
-\subsection{Next steps for the bedload flux}
+\subsection{Birth death models for bedload flucuations}
+
+The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random variable.
+They derived the probability distribution of this variable by counting the number of moving particles in a control volume. 
+This number changes through time as a result of entrainment and deposition.
+To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
+
+Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where entrainment is ``birth" and deposition is ``death".
+They formulated the probability mass function $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
+\be \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) - [n+1]\sigma P(n+1,t) - \big[\lambda + n (\sigma + \mu)\big],\ee
+whose terms describe entrainment at rate $\lambda$, collective entrainment at rate $\mu$, and deposition at rate $\sigma$. 
+The final term encodes the possibility that $n$ does not change at a given instant.
+These coupled equations (one for each $n=0,1,\dots$) can be solved by generating functions \citep{Cox1965,Ancey2008}, providing the steady-state distribution
+\be P(n) = \frac{\Gamma(r+n)}{\Gamma(r) n!}p^r(1-p)^n, \label{eq:negbin}\ee
+where $r = \lambda/\mu$ and $p = 1-\mu/\sigma$.
+This is a negative binomial distribution, which is a wide-tailed generalization of the Poisson distribution.
+
+From eq. \ref{eq:negbin}, the mean number of moving particles is $\langle n \rangle = \lambda/(\sigma-\mu)$, and the variance is
+\be \sigma_n^2 = \frac{\lambda \sigma}{(\sigma-\mu)^2}. \ee
+Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\sigma/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to 1, as the distribution \ref{eq:negbin} limits to a Poisson distribution \citep{Ancey2006}.
+
+The probability distribution of the bedload flux can be computed by the control volume form, eq.
+\ref{eq:controlflux} given, in addition, the probability distribution of particle velocites, which can be determined experimentally \citep{Lajeunesse2012,Fathel2015,Heyman2016} or calculated using the Langevin models described in section \ref{sec:langevin}.
+Assuming that the particle activity in the control volume and the velocities of moving particles are completely independent, which may be a good assumption for a large enough control volume, the probability distribution $Q(q)$ of the flux is \citep{Ancey2008,Ancey2014,Ancey2020}
+\be Q(q) = L \sum_{k=1}^\infty P(k)G(U) ,\ee
+where $P(n)$ comes from eq. \ref{eq:negbin} and $G(U)$ is the probability that the sum of $k$ particle velocities at some instant is equal to $U$.
+
+\subsection{Queueing theories for scale dependence}
+
+The final approach to summarize is that of \citet{Ancey2020}, which calculates the sediment flux probability distribution from an ``arrival time distribution".
+This is the distribution of times between subsequent arrivals of particles to a control surface.
+This approach is reminiscient of queueing theory, which evalues the wait times of customers in a line given a distribution of service times \citep{Cox1961}.
+
+\citet{Ancey2020} considered two different arrival time distributions to calculate the time-averaged sediment flux probability distribution.
+The first is a simple exponential distribution, $P(t) = \Lambda \exp(-\Lambda t).$ Here the mean time between particle arrivals is $1/\Lambda$.
+In this case, the flux can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of section \ref{sec:einstein}:
+\be q(t) = \frac{1}{t}\int_0^t dt' \mu(t'). \ee
+Here, the noise is 
+\be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
+where $N(t)$ is Poisson distributed with rate $\Lambda$. 
+This flux is a random variable.
+Its probability distribution can be derived by evaluating $P(q,t) = \bra \delta(q-\int_0^t \mu(t')/t) \ket$, which involves an average over all possible realizations of the noise in eq.
+\ref{eq:poissfluxnoise}, providing \citep{VanKampen2007}
+\be  \ee
+
+\section{Open problems}
 
 \endinput
-
-\section{Bedload transport foundations}
-
-In river science, a fundamental problem is the determination of the bedload flux, or the rate of downstream movement of coarse sediment grains \citep{Ballio2014}.
-Since bedload transport has strong feedback with stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of environmental considerations. 
-These considerations span a wide range, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}. 
-Unfortunately, no satisfying approach currently exists to compute the bedload flux \citep{Ancey2020,Ancey2020a}, and predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
-
-Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
-Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
-The same turbulence and bed sediment organization details which correlate with the bedload flux are also modified by it.
-Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, while moving sediment affects turublence characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload transport modifies the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
-The bedload flux is in a cyclical feedback with its control variables, the details of turbulence and bed organization \citep{Jerolmack2005}.
-
-\subsection{Bulk bedload flux}
-
-Perhaps a surprising outcome of over a century of bedload transport research is that no one definition of the sediment flux has yet been agreed upon.
-Today, there are two main competing (or complementary) approaches to define the bedload flux. 
-The first definition is reminiscent of continuum mechanics or electrodynamics \citep{Landau, Jackson19xx}, and formulates the flux as a kind of current of sediment across (normal to) a control surface $\matchal{S}$ \citep{Furbish2012,Heyman2016}:
-\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\textbf{\mathcal{S}}. \ee
-This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field over the scales of interest \citep{Heyman2016}.
-Lately, this concentration has been interpreted as an ensemble average \citep{Furbish2012,Ballio2014}.
-The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $\mathcal{V}$:
-\be q = \frac{1}{L} \sum_{i\in\mathcal{V}} u_i.\ee
-Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
-
-\subsection{Bagnold's energy balance}
-
-\subsection{Discrete particle models}
-
-\subsection{The probabilistic approach of Einstein}
-
-\section{From transport to landscape evolution}
-
-\subsection{Exner's approach}
-
-\subsection{Tusjimoto and beyond}
-Explain what tsujimoto essentially did to couple Einstein into Exner, giving the entrainment form of the exner equation.
-Cite Parker and Furbish who provided refinements of it.
-Describe the inherently nonlocal characteristics of sediment flux.
-Indicate the major strengths and weaknesses -  key weakness being that, if the sediment flux is a fluctuating quantity, as formulated by Einstein, then
-
-\section{Topics and organization of the thesis}
-
-\subsection{Organization of the thesis}
-
-Four main problems. Ch 2 is about how to calculate the sediment flux and to understnad particle motion with fluctuating velocities.
-Ch3 is a sketch of how we can incorporate bed elevation changes into sediment transport modelling based on stochastic theories.
-Ch4 is the
-
-\section{Introduction: birth-death models for the probability distribution of bedload flux}
-
-In river science, a fundamental problem is the determination of the bedload flux, or the rate of downstream movement of bedload grains \citep{Ballio2014}.
-Since bedload transport has strong feedback with stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of environmental considerations. 
-These considerations span a wide range, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}. 
-Unfortunately, existing approaches to compute the bedload flux are inadequate. 
-Predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
-
-Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
-Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
-The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
-Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turublence characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
-The bedload flux is in a cyclical feedback with its controls: these controls are the details of turbulence and bed organization \citep{Jerolmack2005}.
-
-As a result, bedload fluxes exhibit wide fluctuations, even as average characterizations of flow and bed organization remain steady \citep{Ancey2014}.
-In the most controlled laboratory experiments, with steady flows driving uniform glass beads under conditions which do not favor bedform development, instantaneous fluxes are often as much as 200\% mean values \citep{Bohm2004, Ancey2008, Heyman2014, Heyman2016}.
-In natural streams, with temporally or spatially varying grain size distributions \citep{Lisle1992, Chen2008}, unsteady flows \citep{Mao2012, FerrerBoix2015}, a variety of dynamic bed surface structures \citep{Hassan2008, Venditti2017}, variable sediment supply \citep{Madej2009, Elgueta2018}, lateral adjustment \citep{Pitlick2013, Redolfi2018}, and migrating bedforms \citep{Gomez1989, Dhont2018} all imparting additional sources of variabilility through their feedbacks with the flux, instantaneous fluxes can be even larger.
-% Iseya1987, Lisle1989, Kuhnle1998
-
-Apparently, bedload fluxes have statistical characteristics.
-Einstein was the first to recognize the statistical character of bedload transport \citep{Einstein1937}. 
-He understood transport as a random switching between states of motion and rest \citep{Einstein1942, Einstein1950}. 
-He called the transition from rest to motion entrainment, and he characterized it with a probability which was linked to extreme events in fluid turbulence \citep{Einstein1949, Einstein1950}. 
-He called the transition from motion to rest deposition, and he considered it an implicit function of the bedload flux \citep{Einstein1942, Einstein1950}. 
-By convolving over sequences of entrainment and deposition, employing some semi-emprical arguments, Einstein developed a formula for the average bedload flux \citep{Einstein1950}.
-
-Many investigators have criticized and developed Einstein's ideas \citep{Paintal1971, Yalin1972, Shen1980, Lisle1998, Papanicolaou2002, Sun2000, Ancey2006, Armanini2015}. 
-These revisions focus on the more simplified aspects of Einstein's derivations, and they lend a more mechanistic basis, a firmer mathematical foundation, and more generality to the approach. 
-Although all of these authors accepted the statistical character of bedload motion, none developed theories for a probability distribution for the bedload flux. 
-This is desirable since its higher moments give unambiguous measurements of the magnitude of bedload fluctuations.
-
-The extension of the Einstein approach to obtain a probability distribution of the bedload flux, with fluctuations stemming from the random character of entrainment and deposition, was first explored by \citet{Lisle1998, Sun2000}, and it was further developed by \citet{Ancey2006}, and the more recent work I will discuss extensively in this review.
-All of these authors revisited Einstein's ideas from a foundation in the stochastic mathematics which became formalized somewhat after Einstein's work \citep[e.g.][]{Cox1965}, treating the transitions between motion and rest states as random events characterized by probabilities.
-This enabled them to apply the formalism of continuous time Markov processes to derive a probability distribution of the bedload flux, with a mean value which was an improved version of the \citet{Einstein1950} formula, very similar to the revised formula of \citet{Yalin1972}, and a variance which provided an unambiguous prediction of the magnitude of bedload fluctuations.
-
-Deterministic processes trace the evolution of some set of variables $\{x_1(t),x_2(t),\dots\}$ through time. 
-Stochastic processes, in contrast, trace the evolution of a probability distribution for random variables $P_{X_1,X_2,\dots}(x_1,x_2,\dots;t)$ through time. 
-The Markov moniker refers to the amount of memory in the process.
-If the probability distribution of the variables of interest can be predicted in the future using only distribution functions from the present, the process is Markovian. 
-Otherwise, if predicting the future requires knowledge of the entire history, the process is non-Markovian \citep{Cox1965, VanKampen1992}. 
-Markov birth-death processes consider the probabilistic creation and annhilation of members of a population \citep{Cox1965, VanKampen1992}.
-
-To model the bedload flux with such a process, the population considered is the number of moving particles within a control volume over the bed. 
-This population is subject to creation by entrainment and annhilation by deposition \citep{Ancey2008, Turowski2009, Heyman2013, Ma2014,  Ancey2014a, Ancey2015}. 
-This approach is exciting because Markov birth-death processes are highly studied in physics, chemistry, and population ecology, so there are many extensions readily available \citep{Bailey1968, Cox1965, Pielou1977, VandenBroek2012, VanKampen1992, Gillespie1992, Field2010, Mendez2015}.
-At the same time, birth-death modeling of the bedload flux remains relatively undeveloped: birth-death models are all one-dimensional and designed for a single grain size.
-
-In this article, I review the literature on Markov birth-death models of the bedload flux.
-In the review I assume some familiarity with discrete state continuous time Markov processes, which could be gleaned from reading the appropriate chapter in \citet{Cox1965}. 
-These theories are exciting because they describe the mean bedload flux and its fluctuations within a volume \citep{Ancey2006, Ancey2008, Turowski2009}, statistical properties of the flux at a point in space \citep{Heyman2013, Ma2014}, and the spatial and temporal correlations in the bedload flux within a reach of channel \citep{Ancey2014, Ancey2015}.
-All of these considerations are new, and they are unique within the river science literature in that they do not ignore the statistical character of bedload transport. 
-They initiate a theoretical framework from which the feedbacks between the arrangement of bed particles, turbulence, and channel morphology which lead to bedload fluctuations can be more carefully studied and understood.
-In reverse, the capacity of bedload models to describe fluctuations provides an additional benchmark against which to judge models, a task which is notably difficult \citep{Iverson2013}.
-
-One of the first birth-death models of the bedload flux I will review is \citet{Ancey2006}: these authors generalized \citet{Einstein1950} using a birth-death framework to obtain a statistical distribution of the bedload flux within a control volume.
-They found the fluctuations predicted by the Einstein-like model were not large enough to describe experimental data.  
-In order to match the wide fluctuations of experimental data, \citet{Ancey2008} reworked their earlier control volume model to include a collective entrainment effect.
-Collective entrainment is a mathematical term charaterizing the correlations between moving grains due to turbulence, collisions, or granular avalanche effects.
-
-The \citet{Ancey2008} model has been generalized and applied in a handful of followup works. 
-\citet{Turowski2009} generalized it to include limited sediment availability. 
-\citet{Heyman2013} and \citet{Ma2014} studied how the \citet{Ancey2008} model could be interpreted to describe the statistics of the bedload flux at a fixed point in space, rather than just a control volume like that considered by \citet{Einstein1950}.
-Control volumes are less desirable because transport is practically easier to measure at a point in space than it is within a control volume.
-
-\citet{Ancey2014, Ancey2015} generalized the \citet{Ancey2008} model in order to understand spatial characteristics of bedload transport. 
-They considered a reach of channel as an array of adjacent control volumes (cells), and they coupled the \citet{Ancey2008} model between adjacent cells. 
-They mapped this generalized birth-death model for the bedload flux distributed over an array of control volumes onto an advection diffusion equation for the bedload flux. 
-Some reprecussions of this model were further explored in context of experimental data by \citet{Heyman2014}.
-
-All of these works contain many simplifying assumptions, and many of these were carefully highlighted by their authors. 
-These studies were selected for this review because, taken together, they define a research trajectory which can be extrapolated to suggest future research topics. 
-To be sure, there is a long road ahead before Markov models can accomodate all of the complexities of natural river channels, some of which I've mentioned. 
-The goal of this review is to indicate this trajectory in order to motivate future research: this is pursued in sections \ref{sec:background} and \ref{sec:review}. 
-I discuss some of the future research topics I believe the trajectory points out in section \ref{sec:scope}, and I develop a few of these ideas myself in \ref{sec:extension}. 
-Along the way, I have somewhat changed the mathematical treatment and the notation from some of the original papers in order to highlight the historical progression between each work, and to illustrate that these works all share the same theme.
-
-\section{Background: the foundational stochastic model of bedload transport} 
-\label{sec:background}
-
-\citet{Einstein1950} can be credited with the first attempt to understand the bedload flux as random quantity.  
-Einstein understood the movement of individual bedload grains as a random succession of motion and rest intervals.
-He termed the transition from rest to motion entrainment, and characterized it by a probability related to fluid turbulence. 
-When particles are set in motion, Einstein considered they moved in discrete jumps of the same average distance. 
-We define the bedload flux as the volume of grains passing a stream cross section per unit time.
-It is a volume of grains per time per unit width of stream, so its units are $\text{area}/\text{time}$.  
-Therefore, Einstein's mean flux formula is derived from summing the alternate start-stop motions of all bed particles which cross a surface in a time interval.
-
-The conceptual picture Einstein considered is depicted in figure \ref{fig:yalin}. 
-All bed particles are identical and are considered to have the same average geometry when at rest on the bed, meaning their entrainment characteristics are identical. 
-For continuity with the rest of the review, we consider sediment particles are spheres of radius $a$, although \citet{Einstein1950} was more general. 
-We can follow Einstein to compute the mean number of these particles crossing a flow cross-sectional surface $S$ in an interval of time. 
-This will construct the mean bedload flux $\bra q_s \ket$.
-In subsequent sections, we consider $q_s$ a random variable, and we will review approaches to derive its full probability distribution.
-From these contemporary models, an Einstein-like mean flux formula emerges from taking the mean of the probability distribution for $q_s$.  
-Hence we use averaging brackets $\bra \cdot \ket$ for the sake of continuity.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=.98\linewidth]{./figures/ch1/yalindrawing.png}
-	\caption{Einstein's conceptual picture (adapted from \citet{Yalin1972}).
-Particles move in discrete jumps of length $L$ from left to right through an array of adjacent control volumes.
-The bedload flux is the volume of bedload particles crossing the surface $S$ per unit width and time.
-\label{fig:yalin}}
-\end{figure}
-
-Einstein's derivation goes as follows: we partition the channel into a sequence of identical control volumes $V$. 
-Each volume has dimensions $L$, $w$, and $h$ ($V=lwh$). 
-$L$ is the downstream length of each control volume, and it is also the average jump distance of particles. 
-$w$ and $h$ are the width and height of each control volume.
-Let $P_n$ be the probability that any individual grain on the bed is entrained \textit{at least} $n$ times during the time interval $T$. 
-That is, let $P_n$ be the probability that an individual grain undergoes \textit{at least} $n$ jumps of length $L$ in an interval $T$, meaning it travels a distance $nL$ or more in $T$. 
-Considering that each control volume contains $N_V$ grains at rest within it, it follows that on average
-\be N_V P_n. \ee
-grains are displaced \textit{at least} a distance $nL$ from a control volume within the time interval $T$.
-
-Now we derive the mean flux. 
-Note that the grains passing through a cross-sectional surface $S$ (of area $lw$) in the time $T$ could have come from any upstream location. 
-Therefore, the number of grains crossing $S$ in $T$ is 
-\be \sum_{n=1}^\infty N_V P_n. \ee
-This is the net arrival of grains in the interval $T$ from all upstream control volumes. 
-Multiplying by the particle volume $\nu_p$ and dividing by the timescale $T$ and channel width $w$ gives the mean bedload flux (volume per width per time):
-\be \bra q_s \ket = \frac{\nu_p N_V}{w T}\sum_{n=1}^\infty P_n = A \frac{a^2}{T} \sum_{n=1}^\infty P_n. \label{eq:einstein1}\ee 
-In the second equality, we have introduced $\nu_p = 4\pi a^3/3$, and two assumptions of Einstein.
-First, is the reasonable assumption that the number $N_V$ of particles on the surface within $V$ is proportional to the ratio of bed surface and particle areas: $N_V \propto Lw/a^2$, and second is the assumption that the travel distance $L$ is proportional to particle size: $L \propto a$.
-The latter assumption is more difficult to justify \citep[e.g.][]{Yalin1972}.
-$A$ is a constant of proportionality.
-
-Einstein implies the probability that an individual grain travels a distance $nL$ or more in $T$ is equivalent to the probability that an individual grain travels at least a distance $L$ successively $n$ times in $T$.
-Hence $P_n = P_1^n$ by the law of multiplication of probabilities, so that $\sum_{n=1}^\infty P_n = \sum_{n=1}^\infty P_1^n = P_1/(1-P_1)$, using the geometric series (noting $P_1 < 1$ because it is a probability).
-Owing to this substitution, Einstein's mean bedload formula takes the form 
-\be  \bra q_s \ket = A \frac{a^2}{T}\frac{P_1}{1-P_1}, \label{eq:einstein2} \ee
-where $A$ is a constant of proprotionality.  
-The $P_1/(1-P_1)$ structure of Einstein's formula is a distinctive characteristic. 
-In order to use the apply the Einstein formula to predict bedload transport in natural streams or experiments, the timescale $T$ and the probability $P_1$ that at least one grain entrains in $T$ must be determined.
-Then the formula can be calibrated to a given setting by adjusting the constant $A$.
-
-In the original work, Einstein considered the timescale $T$ proportional to the "time of the grain" formed by the grain's size $a$ and its settling velocity $w$ in still fluid: $T \propto a/w$. 
-We will revisit this assumption shortly. 
-Regarding the probability $P_1$, one of Einstein's most influential and enduring ideas within hydraulic engineering is his conception of the probability $P_1$ that at least one particle entrains in $T$.
-He considered that entrainment is driven by the fluctuating lift force imparted by the turbulent fluid, and it is resisted by the submerged weight of grains. 
-Therefore, he formulated $P_1$ as an exceedance probability of the lift force over weight, so he wrote $P_1$ as an integral over the probability distribution of fluid shear stress. 
-Einstein developed an alternative perspective on the incipient motion of particles that embraced turbulence and did not rely on the critical shear stress concept \citep[e.g.][]{Shields1936}.
-
-This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly random) resisting quantities has been generalized and extended a great deal. 
-\citet{Paintal1971} made a significant extension by including random granular geometry of bed particles into the force balance. 
-More recently, \citet{Tregnaghi2012} ammended the theory to include the impulse concept: the experimentally verified idea that force magnitude and duration are of equal importance for entrainment \citep[e.g.][]{Diplas2008, Celik2014}.
-Refined theories of the entrainment probability, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2008, Dey2018}, and they are a subject of intensive ongoing research.
-
-As noted by \citet{Yalin1972}, who wrote a careful review and revision of \citet{Einstein1950}, the Einstein theory for $\bra q_s \ket$ is theoretically sound, at least to equation \ref{eq:einstein2}, and provided Einstein's probabilities $P_n$ are interpreted in a careful way.
-I have mimiced Yalin to say $P_n$ is the probability of \textit{at least} $n$ jumps in $T$. 
-Had I said $P_n$ was the probability of (only) $n$ jumps in $T$, the relationship $P_n = P_1^n$ would not follow, so the Einstein flux would not develop the notable $P_1/(1-P_1)$ form as in \ref{eq:einstein2}.
-Einstein was not as clear as he could have been on this distinction. 
-Notably, the more general birth-death theories which reproduce Einstein as a limiting case \citep[e.g.][]{Ancey2006} use probability concepts relating to (only) one entrainment in a time interval. 
-Therefore, to link back to Einstein from these contemporary theories, we need to follow \citet{Yalin1972} to express the mean flux $\bra q_s \ket$ in terms of the probabilities of (only) $n$ entrainments in a time interval, which we can denote by a lower case letter $p_n$.
-These two probabilties, $P_n$ (at least) and $p_n$ (only), have probably been conflated often within the literature.
-
-Yalin appealed to scaling arguments in order to refine Einstein's theory. 
-He highlighted that Einstein's prescription of $T \propto a/w$, where $w$ is the settling velocity, was not supported by any experimental evidence or theoretical argument. 
-Yalin noted that within Einstein's theory, "$T$ appears in connection to the study of detachment of individual grains due to turbulent fluctuations", from which he concluded that $T$ should be actually be measured in terms of a period $\tau$ of turbulent fluctuations:
-\be T = N_\tau \tau, \label{eq:yalintime}\ee
-where $N_\tau$ is the dimensionless number of turbulent fluctuations in $T$ (which is considered unknown but very large).
-
-Yalin introduces a probability $p$ of (only) one entrainment during an individual turbulent fluctuation. 
-Since the number $N_\tau$ of turbulent fluctuations in $T$ is very large, he considers this probability is very small, so that the probability of (only) $n$ detachments in $T$ can be computed with the Poisson distribution:
-\be p_n = \frac{(N_\tau p)^n}{n!}e^{-N_\tau p}. \label{eq:yalinpoisson} \ee
-The probability $P_n$ of \textit{at least} $n$ detachments in $T$ can be expressed in terms of the probability $p_n$ of (only) $n$ detachments in $T$ as 
-\be P_n = \sum_{i=n}^\infty p_i. \label{eq:yalinsum}\ee
-Owing to equations \ref{eq:yalintime}, \ref{eq:yalinpoisson}, and \ref{eq:yalinsum}, the mean bedload flux \ref{eq:einstein1} becomes
-\be  \bra q_s \ket = A \frac{a^2}{N_\tau \tau} \sum_{n=1}^\infty \sum_{i=n}^\infty p_i = A \frac{a^2}{N_\tau \tau} \sum_{n=1}^\infty n p_n = A \frac{a L}{\tau} p. \label{eq:yalinflux} \ee
-The unknown number of turbulent fluctuations $N_\tau$ cancels. 
-Hence, Yalin's ammended Einstein formula is proportional to $p$, rather than $P_1/(1-P_1)$ as in \citet{Einstein1950}. 
-In Yalin's mind, the probability $p$ is the probability of entrainment of only one grain in a small interval $\tau$ related to the period of turbulent fluctuations.
-
-When Einstein's conceptual picture of bedload transport is revisited from a Markov process framework, the probability distribution $P(q_s)$ of the bedload rate can be derived from his assumptions \citep{Ancey2006}.
-This distribution implies a mean bedload formula $\bra q_s \ket = \sum q_s P(q_s)$ formally similar to Yalin's Einstein-like formula \ref{eq:yalinflux}, and it provides additional information about the magnitude of bedload fluctuations, which are characterized without ambiguity by the variance $\bra \delta q_s^2 \ket$, where $\delta q_s = q_s - \bra q_s \ket$ is the deviation of the bedload flux from the mean.
-Jumping ahead a little, the fluctuations $\bra (\delta q_s)^2 \ket$ derived in this way are not large enough to match experimental data \citep{Ancey2006}, and this is because \citet{Einstein1950}, \citet{Yalin1972}, and \citet{Ancey2006} each assumed the transitions between rest and motion were independent between all particles.
-
-\citet{Ancey2008} fixed this problem by describing bedload transport within a more general birth-death process framework \citep{Cox1965}. 
-They included a collective interaction into the entrainment rates of bed particles in their model, and with this inclusion they derived bedload fluctuations of realistic magnitude. 
-Their collective interaction biases entrainment, making it more likely when particles are already moving, a point we will discuss.  
-It leads to clouds of active particles, and a wide tail on the bedload flux probability distribution: both of these features are in accord with experimental observations \citep{Drake1988, Ancey2006, Ancey2008}.
-I'll start the review with \citet{Ancey2006}, which is essentially \citet{Einstein1950} revisited from the birth-death framework, lending it more generality by treating the bedload flux as a probability distribution.
-The mean of this probability distribution reproduces equation \ref{eq:yalinflux}, which is a satisfying historical coherence. 
-Followups and refinements of the \citet{Ancey2006} work constitute the bulk of the review in section \ref{sec:review}.
-
-\section{Scope: identifying future research directions}
-\label{sec:scope}
-
-\subsection{recap} 
-Einstein developed first stochastic theories describing the diffusion and flux of bedload. 
-Ancey 2006, building on earlier work from Lisle 1998 and the many investigators who criticized and revised Einstein, formalized Einstein's assumptions over a foundation in Markov process theory, and extended his work by treating the bedload flux within a control volume as the cooperation of many independent two-state Markov processes: each bed particle within the control volume makes random transitions between motion and rest.
-Thus he derived a statistical distribution of the bedload rate which reproduces Einstein's result when the mean is taken. 
-However, the magnitude of fluctuations predicted by this model are too small.
-
-Ancey 2008 extended the Einstein theory by linking the transition rates between particles within the observation window with a collective entrainment term: they prescribed that the transition rate from rest to motion depends on the number of active particles.
-This is a simple way to include the effects of coherent turbulence and impact-based entrainment: it is an experimental observation that moving particles tend to come in waves \citep{Drake1988}. 
-The Ancey 2008 model was demonstrated capable of describing fluctuations, and all of its parameters have physical meanings, but notably there is no clear suggestion as to how collective and individual entrainment processes can be separated within experiments.
-Clearly, approaches to derive entrainment probabilities are focused on the individual entrainment rate \citep{Dey2018}, and there have been no developments with regard to computing the collective entrainment rate: there is no clear physical model of collective entrainment yet.
-
-Turowski 2009 extended the Ancey work to take account of limited supply, in order to describe semi-alluvial channels where alluvial deposits lie on top of bedrock. 
-He continued to work within the Einstein paradigm by describing the transport rate within a control volume, rather than by counting the number of particles leaving the control volume: although the Ancey 2008 work essentially outlines this possibiilty.
-The transport rate at a point should be considered as the number of emigration events in a unit time.
-
-Heyman et al 2013 and Ma et al 2014 went on to consider the statistics of emigration: they tried to discern what could be learned about the transport rate at a fixed point from the control volume formalism of Ancey et al and Einstein before that.
-Heyman et al was concerned with the statistics of the waiting time between successive emigration events. 
-The original Einstein (1937, 1950) assumptions generate an exponential waiting time ditsribution with a timescale related to the entrainment rate.  
-However, Heyman et al (2013) showed that when collective entrainment effects were included there are two timescales: one fast timescale related to the individual entrainment rate, and one slower timescale related to collective entrainment effects.
-Thus, the distribution of waiting times between successive emigration events, which is a useful concept for alternative stochastic models of the bedload flux distribution \citep[e.g.][]{Turowski2010}, is a more subtle object if collective entrainment occurs.
-These effects are accentuated at low transport rates, where the fast and slow timescales are more disparate. 
-At high rates, the timescales become comparable and the waiting time between successive emigration events blends into an Einstein-like exponential.
-
-All of these approaches considered bedload transport within a control volume or at a point, but indeed, a very large set of contemporary river science studies are concerned with the diffusion or spreading of bed material through a downstream reach of river \citep{Hassan2016}.
-Bedload diffusion is not captured by a model at a single location: often, it has been described using an advection diffusion equation. 
-These advection diffusion equations follow from considerations of mass balance within river channels, but they were not derived from any underlying microstructural model until \citep{Ancey2014}. 
-\citet{Ancey2014} extended the previous control volume model \citep{Ancey2008} to an array of control volumes, where emigration from the $i$th volume is immigration to the $i+1$th volume.
-
-In an approach widely used in chemical physics \citep{Gardiner1983} and ecology \citep{Bailey1968}, \citet{Ancey2015, Ancey2015} derived an advection diffusion equation from their microstructural model in a limit as the size of each control volume goes to zero while the number becomes infinite.
-Thus, in their interpretation, bedload diffusion emerges in the continuum limit of coupled birth-death immigration-emigration models. 
-They rederived the famous Exner equation of bedload diffusion.
-
-Taken together, this research exhibits a coherent progression from the original work of Einstein to a more mathematically and physically based set of more general approaches. 
-Within these approaches the bedload flux is understood as a random quantity. 
-The mean bedload flux and the magnitude of its fluctuations are modelled in an unambiguous way \citep{Ancey2006, Ancey2008} considering no limitations in sediment availability, and it was shown that collective effects are a necessary inclusion to properly describe fluctuations in bedload transport \citep{Ancey2008}.
-An extension to a more realistic situation of finite supply was developed \citep{Turowski2009}, and the relationships between these control volume based models and the definition of bedload flux as particles crossing a plane perpendicular to the flow were developed \citep{Heyman2013, Ma2014b, Ballio2014}.
-
-These approaches did not address spatial correlations or variations in bedload fluxes, although the problem of bedload diffusing through a reach of channel is of contemporary significance \citep{Hassan2016}.
-This extension was the most recent development in birth-death modelling of bedload \citep{Ancey2014, Ancey2015}. 
-By extending the \citet{Ancey2008} model to an array of adjacent control volumes, an advection-diffusion equation describing the concentration of bedload in motion was developed, providing the first connection between a microscale stochastic model of bedload transport with a macroscopic advection-diffusion formulation.
-\citet{Heyman2014} examined the spatial correlations between local bedload fluxes expressed by the \citet{Ancey2014} model. 
-This is the state of the art of bedload flux models.
-
-\subsection{Assumptions, calibraiton problems, and unclear aspects: A criticism of the birth-death approach}
-
-First, the various glaring assumptions of the birth-death approaches are highlighted. 
-\begin{enumerate}
-	\item Models only consider one particle size, while natural streams exhibit wide distributions of sizes and bedload transport expresses a wide set of effects related to particle size segregation \citep{Wilcock2003, Parker1982, Chen2008}.
-	\item There is no clear means to discriminate collective and individual entrainment processes within experiments. 
-	\item There is only a one-way feedback in these models: bedload is considered subordinate to the fluid flow, and the effect of bedload transport back onto the fluid flow is considered negligible.
-In fact, there are measurable influences of bedload transport on the fluid phase \citep{}; it's not yet clear whether a one-way coupled scheme such as this can actually describe natural streams across a realistic range of conditions.
-	\item Collective entrainment is somewhat of a catch-all term with the physical mechanisms which may contribute to it unresolved.
-These may include collective entrianment, the formation and disintegration of particle clusters, interactions between the motion of grains of idfferent size, turbulent structure, local avalanche behavior, and bed form migration.
-	\item there have been no theories developed to compute the collective entrainment rate from any simplified mechanical model, although there is a very large set of work concerned with calculating the individual entrainment rate from considerations of fluid turbulence and random granular arrangement \citep{Einstein1949, Einstein1950, Grass1970, Paintal1971, Cheng1998, Wu2004, Dey2008, Tregnaghi2012, Dey2018}.
-Calculating the collective entrainment rate from underlying principles appears on the surface very difficult.
-If it is considered to stem from collisions of moving grains with stationary grains, the collective entrainment probability will be related to the probability of collisions with the granular bed.
-If it is considered to stem from granular avalanches, where coherent turbulent structures initiate collections or clusters of grains into motion simultaneously, the collective entrainment probability will depend on the collective dynamics of the granular assembly -- leading immediately into the murky physics of force balance within granular assemblies.
-	\item Einstein-like assume a clean division between rest and motion states, which is no doubt an idealization.
-Bedload transport makes a continuous transition from the idealized start-stop motions of Einstein-like models to a less idealized granular flow or creep, where all particles move together in a coorelated fluid-like flow \citep{}, meaning bedload models should only hold at relatively low mobility stages.
-	\item At the same time, the divison of motion into only two categories may be flawed.
-A wide collection of studies have highlighted different modes of motion.
-\citet{Einstein1950} considered that bedload was particles moving "in a rolling, sliding, or saltating mode".
-What if rolling, sliding, and saltating modes of motion were treated independently?
-In an \citet{Ancey2006} model of independent particles switching between states, this would require a four-state model.
-There are too many transition probabilities within such a model probably to calibrate the model from experiments.
-Again, in order to consider such a scenario we would need more knowledge of underlying mechanics.
-
-\end{enumerate}
-
-Incorporating multiple grain sizes in birth-death models is possible, in principle. 
-As a first approximation, it is easy to introduce multiple grain sizes without including interactions between grain sizes in the entrainment and deposition probabilities for each size fraction. 
-However, spatial and temporal heterogeniety in the bed surface characteristics are a hallmark of bedload transport of gravel mixtures \citep{Hassan2008}. 
-This leaves models such as \citet{Ancey2008} somewhat without basis when multiple grain sizes are considered. 
-The concept of \citet{Turowski} is essential to include when multiple grain sizes are considered: the bed surface state determines which grains can entrain \citep[e.g.][]{Wilcock2003, Parker1982}, and presumably which grains can deposit, as well.
-The entrainment and deposition rates of each size fraction will need to depend on the bed surface state, and fractional transport will set up spatial hetereogenieties in the bed surface state, so that an \citet{Ancey2014} type model, incorporating the possiblity of spatial hetereogeniety, should be mixed with a \citet{Turowski2009} type model to incorporate the effect of the bed surface state on the entrainment and deposition rates of each size fraction.
-This extension will not be easy, and it will introduce many undetermined parameters: to pin down this transport model of multiple interacting size fractions will take a lot of work.
-
-These stochastic models are admittedly difficult to calibrate. 
-Their calibration requires a large dataset which is prohibitive to measure within natural streams. 
-Therefore, their applicability in real streams remains limited until the inputs of stochastic models can be computed from physical theories based upon practically measurable quantities. 
-This quest for relationships to compute the inputs of stochastic models from measurable quantities has been called "stochastic closure" for the analogous problem in turbulence \citep{Heyman2016}. 
-The difficult issues precluding the application of stochastic models to natural streams should not be downplayed: much more work is needed.
-
-Any text after an \endinput is ignored.
-You could put scraps here or things in progress.

### 2021-06-27 pref.tex
eh
got a lot of progress on the intro.

@@
 
-\chapter{Preface}
+\chapter{Preface: Geomorphology traditions}
+Geomorphology can be defined as the study of Earth's evolution by tectonics, weathering, and erosion. 
+This science has long been characterized by observation and description.
+Only relatively recently did the science turn to quantitative methods, with researchers working to frame observations in terms of underlying processes and to describe them with mathematical models modified from physics \citep{Church2005}.
 
-A preface should be written indicating what is to come and what the point of this whole thing is.
+The overreliance on physics methods in geomorphology has been rightly criticized as overly reductionist \citep{Slaymaker2020}.
+In physics, many of the greatest achievements describe closed systems with perfect order, but these idealized systems stand in contrast to those encountered in geomorphology.
+Earth's surface is open, and it is highly disordered.
+The surface receives fluxes of mass and energy across its boundaries by tectonics, climate, transpiration, and sunlight, and its component parts vary widely in their characteristics, activities, and compositions.
+Earth's problems concern biota, society, and the deepest issues in the human sphere.
+With such complexity, a pure reduction of geomorphology to physics is not a goal to work toward.
+
+But we can borrow ways of thinking.
+An extremely successful approach in physics has been to construct idealized models without any intention of realism, study them deeply to obtain complete understanding, and then, decades or centuries later, to embellish these models with more sophisticated features to describe real-world phenomena.
+The example which comes to mind is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow ends up fundamental in the deepest inquiries of theoretical physics from superconductivity to cosmological inflation \citep{Fetter2003,Liddle2000}.
+This thesis applies this start dumb approach, inspired by physics, to some problems in Earth science.
+The target is to build archetypes, not describe the Earth's surface in all its grandeur. 
+That's for later (and not for me!)

### 2021-06-27 ch1intro.tex
drafted the hardest part of the intro . . .

@@
 
-Landscales change when sediment moves.
+Landscapes evolve when water, wind, and ice, driven by gravity, grade the uplifted relics of forces from below Earth's surface.
+Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands as they grind them to sand through a machine of denundation.
+Earth's biota colonize these networks which form conduits for migration, transmitting water and organic material while biota convert sediments to soils, staging the joint evolution of life and landscapes which has persisted over geological time.
+Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis.
+Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and river engineering strategies more than ever before.
+
+In this context river geomorphology as a scientific practice is shifting toward more quantitative methods to enable concrete predictions about the natural world.
+Sediment transport in river channels is especially amenable to this quantitative approach since it is basically the result of fluid and granular processes.
+Modelling sediment transport is extremely challenging because fluid and granular physics are notoriously difficult subfields of classical physics.
+Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains.
+When particles are coarse as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload" by bouncing, rolling, and sliding along the bed surface.
+In these conditions, fluid turbulence and the irregular bed surface become important controls over sediment dynamics.
+
+Bedload transport exerts considerable influence over stream morphology and stability \citep{Church2006,Hassan2008,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Zimmerman2007,Comiti2012, McKenzie2018, Eaton2020}.
+As a result, a longstanding problem in river science is to determine the bedload flux, or downtream rate of movement of bedload sediment grains \citep{duBoys1879, }.
+Unfortunately, existing approaches to compute the bedload flux are inadequate as predictions regularly deviate by orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012, Dhont2018}, and this is despite well over a century of concerted effort \citep{Gomez1991}.
+This state of affairs indicates that new research approaches are needed to accelerate research progress \citep{Ancey2020a,Ancey2020}.
+
+Predicting bedload fluxes is challenging because transport is not always well correlated to average characteristics of the flow and bed material. 
+Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
+The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
+Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Strom2008}, which encourages further transport in a positive feedback \citep{Ancey2008,Heyman2016,Lee2018}.
+The bedload flux is thus in a cyclical feedback with its controls in the details of turbulence and bed organization \citep{Jerolmack2005}.
+
+The approach taken in this thesis is to consider the bedload flux as an aggregate result of many individual transported grains.
+This departs from the traditional strategy of correlating transport rates to the mean flow and sedimentary characteristics \citep{Gilbert1917,MeyerPeter1948}.
+Since the trajectories of individual grains are governed by Newtonian mechanics, this perspective provides a wide foothold on the problem.
+The Newtonian approach is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time, in part due to the turbulent flow and the erratic interactions of moving particles with the bed.
+To address this complication, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
+These efforts build on an an earlier literature to provide more realistic descriptions of the trajectories of individual grains and the bedload fluxes they generate.
+
+The thesis begins with a review of the most relevant pieces of this earlier literature on which my research builds.
+The rest of this chapter summarizes earlier works, rephrasing them when necessary to indicate common themes and to show continuity with my own research which follows in the subsequent chapters.
+In particular, the common theme to watch for is that many earlier models rely implicitly on the idealized noises of nonequilibrium statistical physics, Gaussian white noise, white Poisson noise, white dichotomous noise, and so on \citep{Cox1965,VanKampen1992,Gardiner1983,Kubo,Risken1984}.
+The review in this chapter focuses on two main themes: first, models of the movements of individual particles, and second, models of the over-all bedload flux which results from these individual movements.
+An outline of my own research which follows comes at the end of these two sections.
 
@@
 A basic (but not simple) problem in sediment transport modelling is to predict the downstream movement of an individual particle.
-At first glance, maybe at the idealistic beginnings of one's PhD, this seems an elementary problem in general physics, but then some challenges appear.
+At first glance, this seems an elementary problem in general physics, but challenges appear.
 Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not known \citep{Furbish1997}.
@@
 Entrainment and deposition provide an additional layer of complexity.
-Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966}, so they can deposit \citep{Charru2004}.
-These pockets are not permanent shelter.
-Rearrangement of the surrounding bed can reexpose sheltered particles to the flow, and sufficiently strong turbulent fluctuations can overcome the shelter, even if the surrounding particles do not move \citep{Valyrakis2012,Celik2014}, so particles can entrain back into motion.
+Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and they deposit \citep{Charru2004}.
+These pockets are not permanent shelter, because rearrangement of the surrounding bed can re-expose particles to the flow, and sufficiently strong turbulent fluctuations can overcome shelter, even if it is not disturbed \citep{Valyrakis2012,Celik2014}, and particles entrain.
 Ultimately, particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}.
-These buried particles cannot move again until those burying them have \citep{Nakagawa1981}.
+These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}.
 
-Predicting the movements of individual particles therefore involves many contributing processes, and our understanding of each of these processes remains incomplete.
-\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize the challenges and research progress, although it has been slightly revised in more recent studies \citep{Campagnol2013, Hassan2017, Pierce2020}.
+Individual trajectories of particles therefore involve a number of contributing processes, and we have limitations in our understanding of every one of these.
+\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize earlier research by the subset of these processes which it attempts to incorporate, although this conceptual model been slightly revised in some recent studies \citep{Campagnol2013, Hassan2017, Pierce2020}.
+
 Nikora et al divided the downstream trajectory of an individual particle into three timescales, or ``ranges", termed local, intermediate, and global.
@@
 We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
-
-\section{The bedload transport problem}
-
-In river science, a fundamental problem is the determination of the bulk bedload flux, or the rate of downstream movement of coarse sediment grains \citep{Ballio2014}.
-Since bedload transport controls stream morphology \citep{Church2006, Recking2016}, its prediction is useful to a wide range of problems, from aquatic habitat restoration to energy production \citep{Kondolf2014, Wohl2015a}.
-Unfortunately, no satisfying approach currently exists to compute the bedload flux \citep{Ancey2020,Ancey2020a}, and predictions regularly deviate by two orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
-
-Predicting bedload fluxes is challenging because transport is not always well correlated to average characterizations of flow and bed material. 
-Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
-The same turbulence and bed sediment organization details which correlate with the bedload flux are also modified by it.
-Turbulent impulses induce sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turublence characteristics \citep{Singh2010, Santos2014, Liu2016}.
-Bedload transport is controlled by the configuration \citep{Miller1966, Paintal1971,} and consolidation \citep{Johnson2016,Allen2018,Masteller2019} of grains on the bed surface, but bedload transport also modifies the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Hassan2008}.
-The bedload flux controls its control variables. It is in a cyclical feedback with the details of turbulence and bed organization \citep{Jerolmack2005}, so predicting it is challenging.
 
@@
 
-\subsection{Queueing theories for scale dependence}
+\subsection{Renewal theories for scale dependence}
 
-The final approach to summarize is that of \citet{Ancey2020}, which calculates the sediment flux probability distribution from an ``arrival time distribution".
-This is the distribution of times between subsequent arrivals of particles to a control surface.
-This approach is reminiscient of queueing theory, which evalues the wait times of customers in a line given a distribution of service times \citep{Cox1961}.
+The final approach to summarize is that of \citet{Ancey2020}, which calculates the sediment flux probability distribution from an ``arrival time distribution" representing the distribution of times between subsequent arrivals of particles to a control surface.
+This distribution changes depending on the observation time $T$ used to calculate the averaged flux.
+This is the phenomenon of scale dependence.
 
-\citet{Ancey2020} considered two different arrival time distributions to calculate the time-averaged sediment flux probability distribution.
-The first is a simple exponential distribution, $P(t) = \Lambda \exp(-\Lambda t).$ Here the mean time between particle arrivals is $1/\Lambda$.
-In this case, the flux can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of section \ref{sec:einstein}:
-\be q(t) = \frac{1}{t}\int_0^t dt' \mu(t'). \ee
+ \begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/anceyRenewal.pdf}
+	\caption{The scale-dependent sediment flux is described as a renewal process.
+Panel (a) indicates the arrivals of particles at rate $\Lambda$, while panel (b) shows an ensemble of realizations of the sediment flux versus the observation time $T$.
+Note that different realizations of the flux converge toward the same value at large observation times, whereas at small observation times, uncertainty in the flux is large.
+This is observation-scale dependence.
+The nature of this convergence depends on the particle dynamics in an as yet unspecified way.
+}
+	\label{fig:ancey}
+\end{figure}
+
+\citet{Ancey2020} considered two different arrival time distributions to calculate the probability distribution of the time-averaged sediment flux.
+Only the first of these is reviewed here.
+This is an exponential distribution: $P(t) = \Lambda \exp(-\Lambda t),$ so the mean time between particle arrivals is $1/\Lambda$.
+In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of section \ref{sec:einstein}:
+\be q(T) = \frac{1}{T}\int_0^T dt' \mu(t'). \label{eq:flucflux}\ee
 Here, the noise is 
 \be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
-where $N(t)$ is Poisson distributed with rate $\Lambda$. 
-This flux is a random variable.
-Its probability distribution can be derived by evaluating $P(q,t) = \bra \delta(q-\int_0^t \mu(t')/t) \ket$, which involves an average over all possible realizations of the noise in eq.
-\ref{eq:poissfluxnoise}, providing \citep{VanKampen2007}
+as indicated in figure \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda t$. 
+The flux in eq.
+\ref{eq:flucflux} is a random variable as indicated in \ref{fig:ancey} panel (b).
+Its probability distribution, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
+This equation entails an average over all possible realizations of the noise in eq.
+\ref{eq:poissfluxnoise}, producing \citep{VanKampen2007}
 \be  \ee
 
-\section{Open problems}
+\section{Open problems and thesis outline}
 

### 2021-06-27 pref.tex
drafted the hardest part of the intro . . .

--- 2021-06-24 pref.tex
@@
 
-\chapter{Preface}
+\chapter{Preface: Complementary traditions}
 
-A preface should be written indicating what is to come and what the point of this whole thing is.
+This thesis deals with the transport of coarse sediment in flowing water from the perspective that sediment transport results from the movements of individual grains.
+The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.
+
+Geomorphology has long been characterized by observation, description, and the building of conceptual models, not mathematical ones.
+Only relatively recently has the science turned toward quantitative methods, with researchers working to frame observations in terms of underlying processes and to describe them with methods adapted from physics.
+
+An over-reliance on physics methods in geomorphology can be criticized as simplistic and reductionist.
+After all, in physics, many of the greatest achievements describe closed systems with perfect order, and these stand in contrast to those encountered in Earth's landscapes.
+Earth's surface is open, and it is highly disordered.
+It receives fluxes of mass and energy across its boundaries by tectonics, climate, transpiration, and the sun, and its component parts vary widely in their characteristics, activities, and compositions.
+
+With such complexity, a complete reduction of geomorphology to physics is unlikely, but we can borrow ways of thinking.
+An extremely successful approach in physics has been to construct idealized models with little intention of direct realism, study them deeply to obtain complete understanding, and then, decades or centuries later, to embellish these models with more sophisticated features to describe real-world phenomena.
+An example is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow ends up incorporated in the deepest inquiries of theoretical physics, from quantum matter to cosmological inflation.
+This thesis applies this start simple approach to some problems in Earth science.
+The objective is to build up simple archetypical models which can be embellished later, not to solve \textbf{the model} of fluvial geomorphology. 
+That's for later (and not for me!) CHEERS.

### 2021-06-29 abs.tex
got some stuff figured out .

@@
 \chapter{Abstract}
+
+Bedload transport is the movement of coarse sediment grains in a river channel by rolling, sliding, and bouncing.
+Motivated by the need to predict river morphodynamics, a central problem in river science is to calculate the downstream flux of grains moving as bedload.
+This problem is usually formulated in terms of continuum mechanics, but this approach is questionable considering that bedload grains are coarse, they rarely reach densities approximating a continuum.
+This thesis contains four projects that abandon the continuum hypothesis to interpret sediment transport in terms of individual grains.
+First, I develop a new theory to calculate the sediment flux from the trajectories of individual grains as they alternate between motion and rest.
+This leads to a sediment flux probability distribution which predicts mean sediment fluxes that depend on the timescale over which they are observed, explaining why measurements of the bedload flux can give such poor results.
+Second, I evaluate the interplay between bed elevation changes and sediment transport rates, finding that because aggradation and degradation oppose changes in the sediment flux, bed elevation changes buffer sediment transport fluctuations.
+Third, I incorporate the process of sediment burial into a model of downstream sediment trajectories, demonstrating that burial eventually halts downstream transport.
+This has implications for the transport of contaminants in river channels.
+Finally, using concepts borrowed from the theory of granular gases, I develop a stochastic model of individual particles moving downstream in the turbulent flow through a sequence of collisions with the bed surface.
+This indicates that the velocity characteristics of sediment particles are tuned by the amount of dissipation particles experience when colliding with the sedimentary bed.
+This research has implications across river science, where there is increasing acknowledgement of the shortcomings of the continuum hypothesis for predicting dynamics in mountain streams.
+This work extends a wide body of work on individual particle motions which originates from the 1930s, and unifies it with more recent approaches to calculate the bulk bedload flux which have not, until now, been related to the movement characteristics of individual grains.
+
+\endinput
 

### 2021-06-29 ch1intro.tex
got some stuff figured out .

@@
 
-Landscapes evolve when water, wind, and ice, driven by gravity, grade the uplifted relics of forces from below Earth's surface.
+Landscapes evolve when water, wind, and ice, driven by gravity, erode the uplifted relics of forces from below Earth's surface.
 Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands as they grind them to sand through a machine of denundation.
-Earth's biota colonize these networks which form conduits for migration, transmitting water and organic material while biota convert sediments to soils, staging the joint evolution of life and landscapes which has persisted over geological time.
-Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis.
-Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and river engineering strategies more than ever before.
+Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
+Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
+Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.
 
-In this context river geomorphology as a scientific practice is shifting toward more quantitative methods to enable concrete predictions about the natural world.
-Sediment transport in river channels is especially amenable to this quantitative approach since it is basically the result of fluid and granular processes.
-Modelling sediment transport is extremely challenging because fluid and granular physics are notoriously difficult subfields of classical physics.
+In this context river geomorphology as a scientific discipline is shifting toward more quantitative methods that enable concrete predictions about the natural world.
+Sediment transport in river channels is especially amenable to this quantitative approach since it is basically the result of fluid and granular physics \citep{Frey2011}.
 Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains.
-When particles are coarse as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload" by bouncing, rolling, and sliding along the bed surface.
-In these conditions, fluid turbulence and the irregular bed surface become important controls over sediment dynamics.
+When particles are coarse as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface.
+In bedload transport conditions, fluid turbulence and the irregular bed surface inherit special control over sediment dynamics.
 
-Bedload transport exerts considerable influence over stream morphology and stability \citep{Church2006,Hassan2008,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Zimmerman2007,Comiti2012, McKenzie2018, Eaton2020}.
-As a result, a longstanding problem in river science is to determine the bedload flux, or downtream rate of movement of bedload sediment grains \citep{duBoys1879, }.
-Unfortunately, existing approaches to compute the bedload flux are inadequate as predictions regularly deviate by orders of magnitude from measured values \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012, Dhont2018}, and this is despite well over a century of concerted effort \citep{Gomez1991}.
-This state of affairs indicates that new research approaches are needed to accelerate research progress \citep{Ancey2020a,Ancey2020}.
+Bedload transport exerts particular influence over channel morphology and stability \citep{Church2006,Hassan2008,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Zimmerman2007,Comiti2012, McKenzie2018, Eaton2020}.
+A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
+Unfortunately, existing approaches to compute the bedload flux are inadequate, with predictions regularly deviating from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012, Dhont2018}.
+Considering that this is despite well over a century of research into bedload transport \citep{Gomez1991}, we can conclude that new research approaches are sorely needed \citep{Ancey2020a,Ancey2020}.
 
@@
 Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Strom2008}, which encourages further transport in a positive feedback \citep{Ancey2008,Heyman2016,Lee2018}.
-The bedload flux is thus in a cyclical feedback with its controls in the details of turbulence and bed organization \citep{Jerolmack2005}.
+Bedload fluxes are thus in a cyclical feedback with their controls in the details of turbulence and bed organization \citep{Jerolmack2005}.
 
 The approach taken in this thesis is to consider the bedload flux as an aggregate result of many individual transported grains.
-This departs from the traditional strategy of correlating transport rates to the mean flow and sedimentary characteristics \citep{Gilbert1917,MeyerPeter1948}.
-Since the trajectories of individual grains are governed by Newtonian mechanics, this perspective provides a wide foothold on the problem.
-The Newtonian approach is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time, in part due to the turbulent flow and the erratic interactions of moving particles with the bed.
-To address this complication, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
+This departs from the traditional strategy of correlating transport rates to the mean flow and sedimentary characteristics, which considers the granular phase as a continuum \citep{Gilbert1917,MeyerPeter1948}.
+Since the trajectories of individual grains are governed by Newtonian mechanics, this alternative perspective provides traction on the problem.
+This Newtonian approach is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time, in part due to the turbulent flow and the erratic interactions of moving particles with the bed.
+To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
 These efforts build on an an earlier literature to provide more realistic descriptions of the trajectories of individual grains and the bedload fluxes they generate.
@@
 The thesis begins with a review of the most relevant pieces of this earlier literature on which my research builds.
-The rest of this chapter summarizes earlier works, rephrasing them when necessary to indicate common themes and to show continuity with my own research which follows in the subsequent chapters.
-In particular, the common theme to watch for is that many earlier models rely implicitly on the idealized noises of nonequilibrium statistical physics, Gaussian white noise, white Poisson noise, white dichotomous noise, and so on \citep{Cox1965,VanKampen1992,Gardiner1983,Kubo,Risken1984}.
-The review in this chapter focuses on two main themes: first, models of the movements of individual particles, and second, models of the over-all bedload flux which results from these individual movements.
-An outline of my own research which follows comes at the end of these two sections.
+This review rephrases works when necessary to indicate common themes and to show continuity with my own research which follows in the subsequent chapters.
 
@@
 Downstream movement is resisted by frictional collisions between moving particles and the bed.
-Since the bed is a granular surface, its surface geometry is very difficult to characterize \citep{Gordon1972}, and the outcome of collisions varies from one to the next \citep{Sekine1992}
+Since the bed is a granular surface, its geometry is very difficult to characterize \citep{Gordon1972}, and the outcome of collisions varies from one to the next \citep{Sekine1992}
 Entrainment and deposition provide an additional layer of complexity.
-Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and they deposit \citep{Charru2004}.
-These pockets are not permanent shelter, because rearrangement of the surrounding bed can re-expose particles to the flow, and sufficiently strong turbulent fluctuations can overcome shelter, even if it is not disturbed \citep{Valyrakis2012,Celik2014}, and particles entrain.
+Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and enable their deposition \citep{Charru2004}.
+These pockets are not permanent shelter, because rearrangement of the surrounding bed by conventional entrainment or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if the pocket geometry is not disturbed \citep{Valyrakis2012,Celik2014}.
+As a result, particles generally alternate through sequences of entrainment and deposition.
 Ultimately, particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}.
@@
 
-Individual trajectories of particles therefore involve a number of contributing processes, and we have limitations in our understanding of every one of these.
-\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize earlier research by the subset of these processes which it attempts to incorporate, although this conceptual model been slightly revised in some recent studies \citep{Campagnol2013, Hassan2017, Pierce2020}.
-
+Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales.
+\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize this complexity.
 Nikora et al divided the downstream trajectory of an individual particle into three timescales, or ``ranges", termed local, intermediate, and global.
@@
 \citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried. 
-This set of timescales -- local, intermediate, global, and geomorphic -- provides a structure by which to organize wide modelling literature on bedload trajectories and to understand the approximations which have been made on the fundamental problem of predicting the downstream movements of individual grains.
+This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize transport models of individual grains.
 
 \subsection{Motivation: tracers and basic understanding}
-The original motivation to understand individual particle motions was probably to understand the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
+The original motivation to understand individual particle motions was probably to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
 A common measurement technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
@@
 
-This need to understand tracer movements is not however the sole point of research into individual particle motions.
-It is conceptually clear that bulk sediment fluxes ultimately result from the aggregated movements of individual grains, and the prediction of the sediment flux has long been acknowledged as a challenging problem with no clear solution, which impedes geomorphology and engineering understanding \citep{Ancey2020,Ancey2020a}.
-Better understanding of individual particle motions will support this research.
-
 \subsection{Einstein 1937}
-Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}, motivated (at least as far as his research supervisor Meyer-Peter was concerned) by the need to understand the efficiency of Helley-Smith samplers for the management of sedimentation \citep{Ettema2004}.
+Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}.
 Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to probabilistic methods to characterize their transport.
@@
 Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
-He formulated the problem mathematically with an infinite series of convolution integrals, in an elegant display of mathematical physics.
-The mathematics Einstein took on were a pioneering application of the continuous time random walk, which was not formalized until much later \citep{Montroll1965}.
+In the process, Einstein formulated an example of what is now called the continuous time random walk, but this was not was not formalized until much later \citep{Montroll1965}.
 
-For connection to later work in the thesis, and to provide additional perspective on Einstein's work which is not available in the literature, I will summarize Einstein's work slightly differently than he originally formulated it, using a method based on a stochastic dynamical equation.
-In effect, if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
+Einstein's assumptions mean that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
 \be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
-where $N(t)$ is the number of particle entrainments in time $t$, distributed as a Poisson distribution $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
+where $N(t)$ is the number of particle entrainments in time $t$, distributed as a Poisson distribution $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
 Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
-Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1956}, but with a particular random driving term (equation \ref{eq:einrando}) chosen to basically represent the phenomenon under consideration.
-
+Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1956}, but with random driving term (equation \ref{eq:einrando}).
 \begin{figure}[!htbp]
@@
 \be  \big(\ell \px \pt + k_E \ell \px + pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
-This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1986a} reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
+This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1986a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
 \be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/\ell}\theta(x) \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big) \ee
@@
 so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$, given by the entrainment rate times the mean step length.
-The rate at which one particle spreads out from another due to differences in their motion characteristics can be represented by the variance of position, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$.
+The rate at which one particle spreads out from another due to differences in their trajectories can be represented by the variance, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$.
 This gives
 \be \sigma_x^2(t) = 2 k_E \ell^2 t, \ee
-so particles in Einstein's model spread apart as a normal diffusion process \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$
-
-\subsection{First extensions of the Einstein model}
-
-Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest, but it is clearly oversimplified from the reality of particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
-
-Einstein's model of particle movement seems to have been overlooked for the first several decades after its development, or possibly overshadowed by Einstein's later work, which uses its essential ideas for more obviously practical purposes.
-Eventually, curiosity into individual particle motions was re-sparked by cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}.
-In relation to these issues, Einstein's model received a great deal of new attention which generated some generalizations of the model.
-
-The first set of modifications to Einstein's approach in this period concerned the form of the distributions used for step lengths and resting times.
-Einsteins flume experiments demonstrated that particle step lengths followed exponential distributions \citep{1937}, and this was confirmed by subsequent studies \citep{Yano1969, Nakagawa1976}.
-Yet these studies were all conducted in steady flows in artificial flumes, not real streams, where sediment transport conditions can be different.
-When sediment transport is observed before and after a flood, rather than during steady flow, the step lengths are better described by a Gamma distribution, this being the sum across the numerous individual steps that occurred during the flood \citep{Hassan1991}.
-When sediment transport occurs in the presence of dunes or other bedforms, similar modifications have been made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965} and the embedding of particles within bedforms (which increases the periods of rest) \citep{Yang1971,Nakagawa1980}.
+so particles in Einstein's model spread apart as a normal diffusion process $\sigma_x^2 = 2 D_\text{eff} t$ \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$
 
@@
 
-Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles embed in the subsurface or other sedimentary deposits (geomorphic range).
-Starting in the 1970s, the advent of high speed camera experiments of bed load transporthave produced data on the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which \citet{Einstein1937} never intended to describe.
+Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles become embedded in sedimentary deposits (geomorphic range).
+The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
 In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
@@
 
-Studies by \citet{Gordon1972}, \citet{Lisle1998}, and \citet{Lajeunesse2017} have generalized the Einstein theory to include the duration of sediment motion.
+Studies by \citet{Gordon1972}, \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include these timescales.
 They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are also (along with rests) exponentially distributed random variables, this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion (between entrainment and deposition).
-The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written as
+The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
 \be \dot{x} = V\eta(t), \ee
-where the noise is now a so-called ``dichotomous Markov noise" \citep{Bena2006}, which is essentially a random switch or telegraph-type signal that alternates between ``on" ($\eta(t) = 1$) and ``off" ($\eta(t) = 0$) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996} as displayed in figure \ref{fig:lislefig} panel (a).
+where the noise is now a ``dichotomous Markov noise" \citep{Bena2006}, which is essentially a random switch or telegraph-type signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996} as displayed in figure \ref{fig:lislefig} panel (a).
 Some particle trajectories solving equation \ref{eq:lislelangevin} are displayed in \ref{fig:lislefig} panel (b).
 
-This time, the governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt' \rangle$ becomes \citep{Balakrishnan1993}
+The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt' \rangle$ becomes \citep{Balakrishnan1993}
 \be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\ee
 where $k_E$ and $k_D$ are the entrainment and deposition rates, $V$ is the particle velocity during the motion phase, and $k = k_E+k_D$.
-This partial differential equation is called an asymmetric telegrapher's equation in mathematical physics \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is rarely encountered in the literature.
-
+This partial differential equation is called an asymmetric telegrapher's equation \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is not often encountered.
 \begin{figure}[!htbp]
@@
 	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise eq.
-\ref{eq:lislerando}, while panel (b) shows the slanted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (which have mean duration $1/k_E$).
+\ref{eq:lislerando}, while panel (b) shows the tilted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (with mean duration $1/k_E$).
 }
@@
 At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
-In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not.
+In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, since fluctuations of the velocity between entrainment and deposition have been neglected.
 
@@
 A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2011, Azema2012, Wachs2019}.
-These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytically tractable models remain necessary for holistic understanding of sediment transport.
+These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are desired for further insight into the problem.
 
@@
 
-The final set of models relevant to this thesis are the couple of analytical models that have been developed to describe particle velocities in the local and intermediate ranges when velocities fluctuate due to turbulence and particle-bed collisions.
-These models are indended to apply only between entrainment and deposition.
-In contrast to the constant velocities introduced in section \ref{sec:lisle}, the velocities of bed load particles fluctuate through time and are best represented by statistical distributions. 
-Among experimental studies on bedload velocities, two dominant conclusions have emerged.
-One subset of observations indicates that bedload velocities lie on exponential distributions \citep{Lajeunesse2010,Furbish2012,Fathel2015}, and another subset indicates Gaussian distributions \citep{Martin2012,Ancey2014,Heyman2016}.
+Some models have been developed to describe particle velocities in the local and intermediate ranges when velocities fluctuate due to turbulence and particle-bed collisions.
+Experimental studies on bedload velocities have provided two dominant conclusions.
+One subset of observations indicates that bedload velocities follow exponential distributions \citep{Lajeunesse2010,Furbish2012,Fathel2015}, and another subset indicates Gaussian distributions \citep{Martin2012,Ancey2014,Heyman2016}.
 
@@
 \be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
-This is the (two-sided) exponential distribution observed in one subset of the aforementioned experiments.
+This is the (two-sided) exponential distribution observed in one subset of experiments.
 
-In a similar approach, \citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments.
-They wrote for the streamwise velocity
+\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
 \be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
@@
 The timescale $t_r$ is a relaxation time over which velocity fluctuations decay.
-This is the well-known Ornstein-Uhlenbeck process of stochastic physics \citep{Gardiner1983}, and its Fokker-Planck equation is
-\be \pt P(u,t) = -\partial_u\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2} \partial_u^2 P.\ee
-This time, the steady state solution becomes
+This time, the Fokker-Planck equation is
+\be \pt P(u,t) = -\partial_u\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2} \partial_u^2 P,\ee 
+and the steady state solution becomes
 \be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
@@
 
-These models produce successful descriptions of disparate experimental results, although their components are not clearly interpretable in terms of physical expecatations.
-For example, granular interactions were included in the Fan et al model as a Coulomb friction term.
-While such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo intermittent collisions with the granular bed.
-Although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
-In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{Clift1978}.
-We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
-
 \subsection{The definition of the flux}
 
-Perhaps a surprising observation of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
+Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
 Today, there are two main competing (or complementary) approaches to define the bedload flux. 
-The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\matchal{S}$ \citep{Furbish2012,Heyman2016}: 
-\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\textbf{\mathcal{S}}. \ee
-This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field over the scales of interest \citep{Heyman2016}.
-Lately, this concentration has been interpreted as an ensemble average \citep{Furbish2012,Ballio2014}.
+The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\mathcal{S}$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
+\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\mathcal{\textbf{S}}. \ee
+This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.
 The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $\mathcal{V}$:
@@
 Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
-
-An alternative statistical formulation of the sediment flux provides an approximate correspondence between these two definitions \citep{Ancey2006,Furbish2012}.
-This alternative definition hinges on the probability $P[\textbf{u}_p | \textbf{x},t]$ that a particle contacts a control surface $S$ at position $\textbf{x}$ and time $t$ with velocity $\textbf{u}_p$.
-This conditional probability is considered to result from a very large collection of identical systems selected at random moments in their evolution: that is, the conditional probability is an ensemble quantity \citep{Kittel1958}.
-
-In terms of this ensemble probability, the flux is \citep{Ancey2006}: 
-\be q_s = \int_{\text{all } \textbf{u}_p} \int_S P[\textbf{u}_p | \textbf{x},t] \textbf{u}_p \cdot d \textbf{S} d \textbf{u}_p . \label{eq:anceyflux} \ee
-In steady conditions, when the probability is independent of time -- $\partial P / \partial t = 0$, the ensemble definition of the flux \ref{eq:anceyflux} can be approximated by the number of moving particles in the control volume.
-
-An approximate connection to a control volume flux can be derived by swapping the ensemble average for a spatial average, by integrating along a control volume.  
-If the control volume is sufficiently long, $L \rightarrow \infty$, it can be seen as a stack of very many independent cross sectional surfaces. 
-These comprise a stack of replicas of $\mathcal{S}$, and at an instant, if spatial correlations in bedload transport are weak, each surface constitutes one configuration of particles intersecting $S$ with some set of positions and velocities contributing to $P$.
-One can then define $P$ by counting occurrences along this stack of replica surfaces with an integral across the control volume.
-
-With this concept of replica surfaces in mind, we can formalize the link with symbols. 
-Consider $n$ particles distributed throughout a volume at some set of positions $\textbf{x}_i$ with some set of velocities $\textbf{u}_i$, where $i=0,1,\dots, n$,
-The cloud of particles can be represented by its (discrete) density in a position-velocity phase space: 
-\be \rho(\textbf{x},\textbf{u}_p) = \sum_{i=1}^n M(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i). \ee  
-Here $M(\textbf{z})$ is a marker function, which is $1$ if $\textbf{z}$ is inside the particle, and $0$ otherwise.
-The marker's volume integral is $\int M(\textbf{x}) dV = \nu_p$, the particle volume.
-
-Using this density at sufficiently large $L$, the conditional probability $P[\textbf{u}_p | \textbf{x}]$ can be written approximately as an integral over the stack of replica surfaces:
-\be P[\textbf{u}_p|\textbf{x}] \approx \frac{1}{L} \int_0^L dx \rho(\textbf{x},\textbf{u}_p). \ee
-The integral runs along the downstream coordinate, and the equality becomes exact as $L \rightarrow \infty$.
-With this swap of ensemble averaging for spatial averaging, the bedload flux becomes 
-\be q_s \approx \int_{\text{all} \textbf{u}_p} \int_S \frac{1}{L} \int_0^L dx \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i) \textbf{u}_p \cdot \textbf{k} dS d \textbf{u}_p = \frac{\nu_p}{L}\sum_{i=1}^n u_i , \ee
-so there is some correspondence between the surface and volume definitions of the flux.
-Yet this is never exact, except in the unphysical case of sediment transport with no spatial correlations (vanishing particle size, discontinuous trajectories).
-Apparently, the control volume and control surface formulations of the sediment flux are not equivalent, so we see in the literature two parallel streams of research.
 
@@
  \begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.png}
 	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
@@
 This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^n}{n!}e^{-k_E T}, $
-so Einstein's mean sediment flux (eq. \ref(eq:einflux}) becomes
+so Einstein's mean sediment flux (eq. \ref{eq:einflux}) becomes
 \be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
@@
 
-Although Einstein worked to relate the single-particle entrainment rate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character which is not clearly linked to the underlying fluid-granular physics.
+Although Einstein worked to relate the single-particle entrainment rate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character and are not clearly linked to any underlying fluid-granular physics.
 The erosion-deposition model originally developed by \citet{Charru2004,Charru2006,Charru2006a} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
@@
 
-Derived by evaluating a mass balance within a control volume, the erosion-deposition model is
+The erosion-deposition model is
 \be \pt \gamma +  \px V \gamma = E - D. \label{eq:charru}\ee
+This is derived by evaluating a mass balance within a control volume.
 In this equation, $\gamma$ is the ``particle activity", which is the number of moving particles per unit area \citep{Furbish2012}, $V$ is the ensemble averaged movement velocity of sediment grains (which in unsteady conditions may depend on space and time), $E$ is the areal entrainment rate (the number of particles transitioning to motion per unit area and time), and $D$ is the areal deposition rate (the number of particles coming to rest on the bed per unit area and time).
@@
 \be \gamma \propto \frac{\tau-\tau_c}{d^2 V_s^2}.\ee
-Expressing the mean flux as $q = \gamma V$, in the control volume interpretation, the relationship between flux and bed shear stress becomes
+Expressing the mean flux as $q = \gamma V$, the relationship between flux and bed shear stress becomes
 \be q = \frac{A}{d^2 V_s^2}\big(\tau-\tau_c\big)\big[c + d(\sqrt{\tau}-\sqrt{\tau_c})\big]. \ee
@@
 
-Einstein's model of the bulk bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009}.
-\citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
+Einstein's model of the bulk bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009,Tucker2014}.
+\citet{Nakagawa1976} and later \citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
 \be q(x,t) = \int_0^\infty dx' F(x-x',t)E(x',t). \ee
 In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a just-entrained particle steps a distance $l$ before deposition.
-This approach builds on Einstein because it can handle non-uniform conditions by space and time dependence of $E$ and $F$.
+In general, this approach can handle non-uniform conditions by including space and time dependence in $E$ and $F$.
 
@@
 \be q(x,t) = \int_0^\infty dx' \int_0^\infty dt' F(x',t') E(x-x',t-t'), \label{eq:furbo}\ee
-where, $F(x,t)$ represents the probability density that particles move \textit{at least} a distance x in time $t$ (right after entrainment).
+where $F(x,t)$ represents the probability density that particles move \textit{at least} a distance x in time $t$ (right after entrainment).
 For a simple example of the Furbish et al formalism, consider that particles move with a constant velocity $V$ and have a deposition rate $k_D$. 
@@
 \be q = \rho k_E \int_0^\infty dx' \int_0^\infty dt' \theta(Vt-x)k_D\exp(-k_D t) = \rho k_E V/k_D. \label{eq:furbflux}\ee
-Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$. Equation \ref{eq:furbflux} provides another perspective on Einstein's result $q= E\ell$.
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$. Equation \ref{eq:furbflux} therefore provides another perspective on Einstein's result $q= E\ell$.
 
@@
 
-Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuous fields.
-This representation can be interpreted as a spatial average over the detailed positions of individual grains, and it is expected to be valid whenever the scales of interest are large compared to the size of the averaging window, which is required to be be taken much larger than the size of the largest grains in the modelling domain \citep{Coleman2009}.
+Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuum fields.
+This representation can be interpreted as a spatial average over the detailed positions of individual grains, and it is expected to be valid whenever the scales of interest are large compared to the size of the averaging window, which is must be taken much larger than the largest grains in the modelling domain \citep{Coleman2009}.
 Yet there are many Earth surface processes where the scale of interest is not much larger than the largest grains involved in the system.
@@
 In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the result of entrainment at all upstream locations, giving
-\be (1-\phi) px z(x,t) = \int_0^\infty \dx' int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
+\be (1-\phi) px z(x,t) = \int_0^\infty dx' int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
 This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
@@
 
-Bedload flux signals exhibit characteristically large fluctuations \cite{Gomez1991,Bunte2005,Recking2012}, but all models reviewed to this point describe only mean sediment fluxes, and not the strength of bedload transport fluctuations, correlations, or other statistical characteristics of the flux.
-A full statistical characterization of sediment transport is desirable because transport fluctuations are responsible for the initiation of bedforms \citep{Jerolmack2005,Bohorquez2016} and are beginning to show links to basic geomorphology considerations like the maintenance of stable channel widths \citep{Abramian2019,Abramian2020}.
-This is not to mention that because sediment transport fluctuations are ubiquitous, models which describe only mean bedload fluxes signals are conceptually incomplete.
-
-An additional reason to characterize fluctuations in bed load fluxes is because in reality, sediment flux measurements always involve a sampling interval.
-This introduces questions of the convergence of transport rate measurements \citep{Dhont2018,Ancey2020}.
-Fluctuations originate from processes which span a vast range of characteristic timescales.
-These can occur over a matter of seconds, such as differences in the velocities of grains due to a series of collisions \citep{Benavides2021}, to many hundreds of hours, such as the migrations of dunes and bedload sheets \citep{Hammamori1962,Hoey1992,Guala2014} or the cycling between aggradation and degradation in pools \citep{Dhont2018}.
-Such processes introduce temporal correlations into sediment transport signals which control the sampling interval over which measurements of the mean sediment flux converge \citep{Saletti2015,Singh2009,Singh2012}.
-To date, very few modelling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
+The sediment flux . . . .. has fluctuations and scale dependence. 
+An early approach to calculate the sediment flux distribution is due to \citet{Hamamori1962}, who evaluated the sediment flux as the result of migrating bedforms.
+This approach has been extended to include bedforms with a range of sizes \citep{Guala2014,Nikora1984}.
 
@@
 \subsection{Renewal theories for scale dependence}
-
-The final approach to summarize is that of \citet{Ancey2020}, which calculates the sediment flux probability distribution from an ``arrival time distribution" representing the distribution of times between subsequent arrivals of particles to a control surface.
+\label{sec:renewal}
+The final approach to summarize that of calculating the sediment flux probability distribution from an ``arrival time distribution" representing the distribution of times between subsequent arrivals of particles to a control surface \citep{Turowski2010,Ancey2020}.
 This distribution changes depending on the observation time $T$ used to calculate the averaged flux.
 This is the phenomenon of scale dependence.
+The statistics machinery of counting arrivals is called renewal theory \citep{Cox1967}.
 
@@
 \end{figure}
-
-\citet{Ancey2020} considered two different arrival time distributions to calculate the probability distribution of the time-averaged sediment flux.
-Only the first of these is reviewed here.
+Earlier studies have considered two different arrival time distributions to calculate the probability distribution of the time-averaged sediment flux, but only one of these is reviewed here.
 This is an exponential distribution: $P(t) = \Lambda \exp(-\Lambda t),$ so the mean time between particle arrivals is $1/\Lambda$.
@@
 \ref{eq:flucflux} is a random variable as indicated in \ref{fig:ancey} panel (b).
-Its probability distribution, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
+Its probability distribution which is contingent on the observation time $T$ can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
 This equation entails an average over all possible realizations of the noise in eq.
 \ref{eq:poissfluxnoise}, producing \citep{VanKampen2007}
-\be  \ee
+\be P(q|T) = \sum_{l=0}^\infty \frac{\big(\Lambda T\big)^l}{l!}e^{-\Lambda T}\delta(q-\frac{l}{T}),\ee
+which is a Poisson distribution.
 
-\section{Open problems and thesis outline}
+The mean flux derived from this scheme ($\langle q \rangle(T) = \int_0^\infty dq P(q|T)$ is
+\be \langle q \rangle = \Lambda, \ee
+which is independent of observation scale, while the magnitude of bedload transport fluctuations scales with $1/T$: 
+\be \sigma_q^2 = \frac{\Lambda^2}{T}.\ee
+Even this simple model gives a non-trivial conclusion that the relative uncertainty in a measurement of bedload transport depends on observation time: $\sigma_Q/\langle q \rangle \propto T^{-1/2}.$
+The limitation of these approaches is that they do not obviously relate to the dynamics of individual particles. In renewal models, the distribution of arrival times is heuristic.
+
+\section{Summary}
+
+This literature review has indicated that successful bedload transport descriptions have been developed to describe individual particles and bulk bedload fluxes.
+All of the works reviewed so far are either mean field models that do not include fluctuations, or stochastic models that calculate probability distributions for the quantities of interest.
+The common thread of all of these models is that they take advantage of various devices to side-step the complex interplay of turbulence and granular physics which control bedload transport.
+In the case of mean field models, the devices are heuristic scaling arguments and semi-empirical formulas, while in the case of stochastic models, they are simplified dynamical equations driven by idealized noises, like sequences of pulses, alternating switches, and erratic fluctuations.
+
+\section{Outline of the thesis}
+\subsection{Problem 1: The scale-dependent flux from individual particle dynamics}
+
+The original model by \citet{Einstein1937} describing particle trajectories when the movement velocity is infinite was subsequently improved to include a finite yet constant velocity \citep{Gordon1972,Lisle1998,Lajeunesse2018}.
+Particle velocities fluctuate during movements between entrainment and deposition, so this should also be accounted for in bedload trajectory models.
+In addition, the scale-dependence of the sediment flux has been calculated from renewal models \citep{Turowski2010,Ancey2020}, but these overlook the details of individual particle motions, so it remains unclear how the sediment flux distribution and its scale dependence originate from individual particle trajectories.
+
+Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories which includes velocity fluctuations in the motion phase, then demonstrates how these trajectories can be used to calculate scale-dependent probability distribution of the bedload flux, building on all of these earlier studies.
+
+\subsection{Problem 2: Feedbacks between sediment transport fluctuations and bed elevation change}
+
+Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Ancey2008,Heyman2013,Ancey2014}, which violates conservation of mass and the Exner equation \citep{Exner1925,Tsujimoto1976}, and precludes the possibility that sediment can become buried in the subsurface of the bed.
+Although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2012,Hassan2017}, our understanding of how long sediment can remain buried is limited.
+
+Chapter \ref{ch:burial} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movements characteristics of grains.
+
+\subsection{Problem 3: The effect of sediment burial on downstream transport of sediment tracers}
+
+The model of sediment transport as an alterating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface, and it provides a useful description of bedload transport on the intermediate and global timescales of \citet{Nikora2001,Nikora2002}.
+Tracers become buried, and this affects their motion characteristics at global and geomorphic timescales \citep{Hassan2017}.
+These phenomena are not yet included in Einstein-type models that treat individual sediment trajectories as a random walk \citep[e.g.][]{Lisle1998}.
+
+Chapter \ref{ch:downdiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2018} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.
+
+\subsection{Problem 4: The control of particle-bed collisions over bedload particle velocity distributions}
+
+Finally, the Langevin models of \citet{Fan2014} and \citet{Ancey2014} describe the velocity distributions of sediment moving downstream in the local and intermediate ranges in between entrainment and deposition.
+These models describe two different end-member distributions from among the experimental record on bedload velocities. One develops exponential velocities, and the other develops Gaussian velocities.
+Since the forces included in these models are phenomenological, we can assume the issue is that particle-bed collisions are not adequately represented.
+
+Chapter \ref{ch:langevin} presents a stochastic Langevin model including episodic particle-bed collisions formulated by analogy with the theory of granular gases \citep{Brilliantov2004}.
+This produces a satisfying description of all different bedload velocity distributions which have been reported in experimental studies.
 
 \endinput
+
+An alternative statistical formulation of the sediment flux provides an approximate correspondence between these two definitions \citep{Ancey2006,Furbish2012}.
+This hinges on the probability $P[\textbf{u}_p | \textbf{x},t]$ that a particle contacts a control surface $S$ at position $\textbf{x}$ and time $t$ with velocity $\textbf{u}_p$. 
+This conditional probability is considered to result from a very large collection of identical systems selected at random moments in their evolution: that is, the conditional probability is an ensemble quantity \citep{Kittel1958}.
+
+In terms of this ensemble probability, the flux is \citep{Ancey2006}: 
+\be q_s = \int_{\text{all } \textbf{u}_p} \int_S P[\textbf{u}_p | \textbf{x},t] \textbf{u}_p \cdot d \textbf{S} d \textbf{u}_p . \label{eq:anceyflux} \ee
+In steady conditions, when the probability is independent of time -- $\partial P / \partial t = 0$.
+
+An approximate connection to a control volume flux can be derived by swapping the ensemble average for a spatial average by integrating along a control volume.  
+If the control volume is sufficiently long, $L \rightarrow \infty$, it can be seen as a stack of very many independent cross sectional surfaces. 
+These comprise a stack of replicas of $\mathcal{S}$, and at an instant, if spatial correlations in bedload transport are weak, each surface constitutes one configuration of particles intersecting $S$ with some set of positions and velocities contributing to $P$.
+One can then define $P$ by counting occurrences along this stack of replica surfaces with an integral across the control volume.
+
+With this concept of replica surfaces in mind, we can formalize the link with symbols. 
+Consider $n$ particles distributed throughout a volume at some set of positions $\textbf{x}_i$ with some set of velocities $\textbf{u}_i$, where $i=0,1,\dots, n$,
+The cloud of particles can be represented by its (discrete) density in a position-velocity phase space: 
+\be \rho(\textbf{x},\textbf{u}_p) = \sum_{i=1}^n M(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i). \ee  
+Here $M(\textbf{z})$ is a marker function, which is $1$ if $\textbf{z}$ is inside the particle, and $0$ otherwise.
+The marker's volume integral is $\int M(\textbf{x}) dV = \nu_p$, the particle volume.
+
+Using this density at sufficiently large $L$, the conditional probability $P[\textbf{u}_p | \textbf{x}]$ can be written approximately as an integral over the stack of replica surfaces:
+\be P[\textbf{u}_p|\textbf{x}] \approx \frac{1}{L} \int_0^L dx \rho(\textbf{x},\textbf{u}_p). \ee
+The integral runs along the downstream coordinate, and the equality becomes exact as $L \rightarrow \infty$.
+With this swap of ensemble averaging for spatial averaging, the bedload flux becomes 
+\be q_s \approx \int_{\text{all} \textbf{u}_p} \int_S \frac{1}{L} \int_0^L dx \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i) \textbf{u}_p \cdot \textbf{k} dS d \textbf{u}_p = \frac{\nu_p}{L}\sum_{i=1}^n u_i , \ee
+so there is some correspondence between the surface and volume definitions of the flux.
+Yet this is never exact, except in the unphysical case of sediment transport with no spatial correlations (vanishing particle size, discontinuous trajectories).
+Apparently, the control volume and control surface formulations of the sediment flux are not equivalent, so we see in the literature two parallel streams of research.
+
+\subsection{First extensions of the Einstein model}
+
+Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest, but it is clearly oversimplified from the reality of particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
+
+Einstein's model of particle movement seems to have been overlooked for the first several decades after its development, or possibly overshadowed by Einstein's later work, which uses its essential ideas for more obviously practical purposes.
+Eventually, curiosity into individual particle motions was re-sparked by cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}.
+In relation to these issues, Einstein's model received a great deal of new attention which generated some generalizations of the model.
+
+The first set of modifications to Einstein's approach in this period concerned the form of the distributions used for step lengths and resting times.
+Einsteins flume experiments demonstrated that particle step lengths followed exponential distributions \citep{1937}, and this was confirmed by subsequent studies \citep{Yano1969, Nakagawa1976}.
+Yet these studies were all conducted in steady flows in artificial flumes, not real streams, where sediment transport conditions can be different.
+When sediment transport is observed before and after a flood, rather than during steady flow, the step lengths are better described by a Gamma distribution, this being the sum across the numerous individual steps that occurred during the flood \citep{Hassan1991}.
+When sediment transport occurs in the presence of dunes or other bedforms, similar modifications have been made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965} and the embedding of particles within bedforms (which increases the periods of rest) \citep{Yang1971,Nakagawa1980}.
+
+A full statistical characterization of sediment transport is desirable because transport fluctuations are responsible for the initiation of bedforms \citep{Jerolmack2005,Bohorquez2016} and are beginning to show links to basic geomorphology considerations like the maintenance of stable channel widths \citep{Abramian2019,Abramian2020}.
+This is not to mention that because sediment transport fluctuations are ubiquitous, models which describe only mean bedload fluxes signals are conceptually incomplete.
+
+An additional reason to characterize fluctuations in bed load fluxes is because in reality, sediment flux measurements always involve a sampling interval.
+This introduces questions of the convergence of transport rate measurements \citep{Dhont2018,Ancey2020}.
+Fluctuations originate from processes which span a vast range of characteristic timescales.
+These can occur over a matter of seconds, such as differences in the velocities of grains due to a series of collisions \citep{Benavides2021}, to many hundreds of hours, such as the migrations of dunes and bedload sheets \citep{Hammamori1962,Hoey1992,Guala2014} or the cycling between aggradation and degradation in pools \citep{Dhont2018}.
+Such processes introduce temporal correlations into sediment transport signals which control the sampling interval over which measurements of the mean sediment flux converge \citep{Saletti2015,Singh2009,Singh2012}.
+To date, very few modelling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
+
+These models produce successful descriptions of disparate experimental results, although their components are not clearly interpretable in terms of physical expecatations.
+For example, granular interactions were included in the Fan et al model as a Coulomb friction term.
+While such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo intermittent collisions with the granular bed.
+Although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
+In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{Clift1978}.
+We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
+
+One overarching methodology is the use of idealized noises in combination, where each represents a separate mechanical process affecting the movements of particles.
+This takes on more mathematical complexity than earlier studies, but it also introduces more realism.
+
+Regarding the movements of individual particles, the division into local, intermediate, global, and geomorphic ranges by \citet{Nikora2001,Nikora2002} and \citet{Hassan2017} provides a conceptual scheme by which to organize the current state of research.
+Einstein's original description \citep{Einstein1937} and those of his early followers \citep{} described only the global range associated with timescales over which particles cycle through motion and rest.
+This limitation was overcome in later works which included a finite duration of sediment motion in order to describe both intermediate and global scales, and Newtonian models have also attained descriptions of additional scales, such as local and intermediate \citep{}.
+Meanwhile, Langevin models have derived the velocity characteristics of particles within the local range \citep{}, although they do not manage to account for the full range of experimentally-observed distributions.
+The major shortcoming exposed by this framing in terms of the timescales of particle motion is that no models have yet resolved particles across the full set of timescales, so all of these descriptions are essentially incomplete.
+
+Models of the sediment flux face a similar set of shortcomings.
+Einstein and Bagnold founded complementary perspectives of the sediment flux.
+Both were mean field theories, but one counted particles crossing a surface using probabilistic arguments, while the other considered an average energy balance in a control volume.
+Research has come a long way since these studies, but the essential divide between surface and volume-based flux calculations still exists. These definitions are not equivalent.
+The erosion-deposition model takes steps toward joining these two mean-field theories into a common framework, though it involves many parameters which must be experimentally calibrated.
+A paradigm shift in this research stream came from the recognitition that the sediment flux is a fluctuating quantity best characterized by a probability distribution, and from the acknowledgement that measured values of the sediment flux depend on observation scale, meaning it is a scale-dependent quantity.
+Although simple models of the sediment flux probability distribution have been developed, they stand somewhat apart from the detailed dynamics of individual particles.
+Rather than describing particles moving downstream through cycles of entrainment and deposition as in \citet{Einstein1937}, these methods characterize particle motions by herustic entrainment and deposition rates, or interarrival time distributions, which are somewhat abstract objects, not concretely related.
+Despite the considerable progress these approaches have brought, the lack of connection to the underlying particle mechanics is a real shortcoming which would ideally be rectified.

### 2021-06-29 ch2flipFlop.tex
eh
got some stuff figured out .

--- 2021-06-24 ch2flipFlop.tex
@@
 \chapter{Calculation of the sediment flux}
+\label{ch:flux}
+\section{Introduction}
 
-\section{Introduction}
-\section{Theoretical development}
+A relativity weak flow shearing a bed of sediment entrains individual particles into a state of motion controlled by turbulent forcing and intermittent collisions with other grains at rest on the bed, generating wide fluctuations in the sediment velocity \citep{Heyman2016,Fathel2015}.
+Bed load particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered pockets on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
+Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2019,Valyrakis2014}, particles are once again entrained, and the cycle repeats.
+Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and rest.
+This process is difficult to describe mathematically given the technicality of the stochastic physics involved \citep{Furbish2017,Ancey2020}.
+
+To date, descriptions of bed load transport have therefore simplified the problem in various ways to enable progress.
+The foundational work is due to Einstein, who considering bed load motions as instantaneous so he could describe bed load transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
+Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E l$, where $k_E$ is the rate at which an individual bed particle entrains into motion, and $l$ is the mean length of each downstream step.
+Later, Einstein applied these ideas to calculate the mean downstream flux due to the movements of many such particles \citep{Einstein1950}.
+Einstein reasoned that if the density of resting particles on the bed is $\rho_b$, the overall areal entrainment rate of particles can be written $E = \rho_b k_E$, giving a mean sediment flux $\bra q_s \ket = \rho_b \bra u \ket  = El $.
+
+Many researchers have since refined Einstein's approach to provide more realistic descriptions of individual particle motions than Einstein's instantaneous step model.
+One set of efforts has neglected particle deposition process to calculate the downstream velocity distributions of moving particles after entrainment using mechanistic equations with terms representing turbulent fluid forces and particle-bed collisions \citep{Ancey2014,Fan2014}.
+Another set of efforts has described particles alternating between motion and rest, considering that motions occur with a constant velocity \citep{Lisle1998,Lajeunesse2017,Pierce2020b}. 
+No models have yet been formulated which describe particles alternating between motion and rest, while motions have a realistic fluctuating velocity.
+
+A set of studies has developed stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Furbish2012}. 
+Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Ancey2008,Ancey2014}, the migration of bedforms and sediment waves \citep{Hamamori1962,Guala2014,Hoey1992,Recking2012}, and a host of other processes \citep{Dhont2018}.
+Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale-dependence \citep{Saletti2014,Dhont2019,Singh2009,Turowski2010,Ancey2020}.
+To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020,Turowski2010}.
+Among those that do, none have yet formulated the sediment flux in terms of the downstream trajectories of individual grains.
+
+This survey provides context for the two problems addressed in this chapter.
+First, we do not yet have the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution, and the dependence of its statistical moments on the observation time.
+Here, we develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a Langevin-type dynamical equation which combines earlier approaches \citep{Fan2014,Ancey2014,Lisle1998,Lajeunesse2017}
+This stochastic equation includes alternation between motion and rest at random intervals while particles in motion have velocity that fluctuate around a mean value.
+Using the probability distribution of particle position generated by this model, we construct a formalism to derive analytically the probability distribution of the sediment flux.
+This distribution exhibits observation-scale dependence as a result of velocity fluctuations among moving sediment.
+Below, we develop the new formalism in sec.
+\ref{sec:mod}, solve it in sec.
+\ref{sec:res}, and we discuss the implications of our results and future research ideas in secs.
+\ref{sec:disc} and \ref{sec:conc}.
+
+\section{Model development \label{sec:mod}}
+We consider an infinite one-dimensional domain populated with sediment particles on the surface of a sedimentary bed.
+We consider that the flow is weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common.
+The flow is in constrast strong enough so that particles move.
+We label the downstream coordinate as $x$, so that the downstream velocity of a moving particle is $\dot{x}$, and we describe all sediment particles as independent from one another but governed by the same underlying dynamical equations, meaning we neglect any influence of sediment size or shape or spatial variations in the overlying fluid flow.
+
+\subsection{Dynamical equation for bed load sediment transport}
+From these assumptions, our first target is to write an equation of motion for the individual sediment particle encompassing two features.
+First, particles should alternate between motion and rest.
+The transition rate from rest to motion is called entrainment and occurs with probability per unit time (or rate) $k_E$, while the transition from motion to rest is called deposition and occurs with rate $k_D$.
+Second, particles in motion should move with mean velocity $V$ and some fluctuations around this velocity.
+The simplest equation of motion including these features is
+\be \dot{x}(t) = [V + \sqrt{2D}\xi(t)]\eta(t).  \label{eq:langevin} \ee
+Here $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles, and $\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion (with mean duration $1/k_D$), and $\eta=0$, representing rest (with mean duration $1/k_E$).
+Here, $V$ is the mean particle velocity, and $D$ is a diffusivity [units $L^2/T$] describing velocity fluctuations among moving particles.
+The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$.
+We write $k=k_E+k_D$ as a shorthand.
+
+\subsection{Derivation of the master equation for $P(x,t)$}
+The solution of equation \ref{eq:langevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives the trajectory of one particle.
+Averaging over the ensemble of all such trajectories from different realizations of the noises will obtain the probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$.
+This distribution, by construction, will generalize earlier models of bedload alternation between motion and rest which did not include velocity fluctuations among moving particles \citep{Lisle1998,Lajeunesse2017}.
+
+We form the desired probability distribution of position as $ P(y,t) = \big\bra \delta(y-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
+\ref{eq:langevin} and the average is over both noises, but this symbolic equation is not yet useful as taking averages over both noises directly is a challenging mathematical problem \citep{Hanggi1978}.
+A simpler approach is to conduct the averages in Fourier space. Integrating eq. 1, using its solution in the probability distribution, then Fourier transforming gives
+\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
+Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen1978} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Loginov1978} in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
+\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
+and inverse Fourier transforming provides the desired real-space master equation
+\be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0. \label{eq:master}\ee
+This is a diffusion-like equation governing the probability distribution of position for individual particles as they transport downstream through a sequence of motions and rests, with the movement velocity being a fluctuating quantity.
+One can see in particular that taking an the entrainment rate $k_E$ very large, meaning that all particles are generally moving, implies an advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
+Otherwise, with $k_E$ of similar order as $k_D$, there is a finite probability that the particle is at rest, and the advection-diffusion process is often interrupted by deposition, giving rise to the additional terms in eq.
+\ref{eq:master} reminiscient of earlier motion-rest models with constant velocity \citep{Lisle1998,Lajeunesse2017}.
+
+\section{Formalism for the downstream sediment flux}
+The probability distribution of the sediment flux is calculated using the probability distribution of particle position $P(x,t)$ provided as the solution of equation \ref{eq:master}.
+This method is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
+The basic idea, as depicted in Figure 1, is that particles are initially distributed in all states of motion along a domain at random locations to the left of $x=0$.
+The flux is calculated using the number of particles that end up to the right of $x=0$ during the sampling time $T$.
+\begin{figure}
+	\centerline{\includegraphics{./figures/ch2/figure1.pdf}}
+	\caption{The left panel indicates the configuration for the flux.
+The particle trajectories within are calculated from equation \ref{eq:langevin}, demonstrating alternation between rest and motion with fluctuating velocity.
+Particles begin their transport with positions $-L\leq x \leq 0$ at $t=0$, and as depicted in the right panel, the flux is calculated as the number of particles $N_>(t)$ which lie to the right of $x=0$ at the observation time $t$, divided by $t$: $Q(t) = t^{-1}N_>(t)$.
+We calculate the probability distribution of $Q$ over all realizations of the trajectories and initial positions as $L\rightarrow \infty$}
+	\label{fig:fig1}
+\end{figure}
+
+The rate of particles crossing the surface in an observation time $T$ is
+\be Q(T) = \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T). \ee
+In this equation, $\mathcal{I}_i(T)$ is an indicator function which is $1$ whenever the $i$th particle has passed our control surface and $0$ otherwise.
+All particles which have not crossed the control surface (or which have crossed and then crossed back) contribute nothing to the flux.
+The probability distribution of the flux is then
+\be P(Q|T) = \Big \bra \delta\Big( Q - \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T) \Big) \Big\ket. \ee
+The average is over the initial conditions of each particle and the ensemble of trajectories for each particle.
+Taking the Laplace transform over $Q$ (i.e. forming the characteristic function) obtains
+\begin{align} \tilde{P}(s|T) &= \Bigg\bra \int_0^\infty dQ e^{-s Q}\delta\Big( Q - \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T) \Big)\Bigg\ket \\
+	&=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N \mathcal{I}_i(T)\Big)\Big\ket \\
+	&=  \prod_{i=1}^N \Big\bra \exp\Big(-\frac{s}{T}\mathcal{I}_i(T)\Big)\Big\ket \\
+	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra \mathcal{I}_i(T) \big\ket \Big] \end{align}
+This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $ \mathcal{I}_i(T)$ is either $1$ or $0$ ($e^{ax} = 1-(1-e^a)x$ if $x=0,1$).
+The average over initial conditions and possible trajectories for the $i$th particle can be written
+\be \bra \mathcal{I}_i(t) \ket = \frac{1}{L}\int_L^0 dx' \int_0^\infty dx \mathcal{P}(x,t|x') =  \frac{1}{L}\int_0^L dx' \int_0^\infty dx \mathcal{P}(x,t|-x'), \ee
+where $\mathcal{P}(x,t|x')$ is the probability density that the particle is found at position $x$ at time $t$ given it was initially at $x'$ at time $0$.
+This is the part of the flux that depends on the particle dynamics (ie instantenous velocities and entrainment/deposition characteristics).
+
+Inserting (7) into (6) and taking the limit as $L\rightarrow \infty$ and $N \rightarrow \infty$ while the density of particles $\rho = N/L$ remains constant provides
+\be \tilde{P}(s|T) = \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\mu(T) \Big)^N = \exp \Big[ -\big(1-e^{-s/T}\big)\mu(T) \Big].\ee
+where $\mu(T) = \rho \int_0^\infty dx \int_0^\infty dx' \mathcal{P}(x,T|x')$ is a rate constant which encodes the particle dynamics.
+This expression is the characteristic function of a Poisson distribution \citep{Cox1965}.
+Expanding in $e^{-s/T}$ and inverting the Laplace transform provides the probability distribution of the flux, contingent on the sampling time $T$:
+\be P(Q|T) = \sum_{k=0}^\infty \frac{\mu(T)^k}{k!}e^{-\mu(T)}\delta(Q-\frac{k}{T}).\ee
+This equation implies that the mean flux is $\bra Q \ket(T) = \int_0^\infty dQ Q P(Q|T) = \mu(T)/T$, and similarly the variance is $\sigma_Q^2(T) = \mu(T)/T^2$. 
+The conclusion is that if the flux is considered as a time averaged number of particles crossing a control surface, the mean flux is always Poissonian, no matter how particles move, provided the particles are independent and do not interact.
+
+\section{Results \label{sec:res}}
+\subsection{The position probability distribution and its moments}
+\begin{figure}
+	\centerline{\includegraphics{figures/ch2/figure2_slopeKey.pdf}}
+	\caption{Panel (a) indicates the probability distribution of particle position as it evolves through time.
+From the initial mixture of motion and rest states, particles advect downstream as they diffuse due to differences in their fluctuating velocities and exchange between motion and rest.
+Panel (b) shows the resulting particle diffusion.
+At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard diffusion process (as advection is irrelevant on these timescales).
+For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
+Finally longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, exemplified by exchange between motion and rest.
+All results are scaled by the mean hop length $l=V/k_D$ and the autocorrelation time $\tau=1/k$ of the motion/rest alternation.}
+	\label{fig:fig1}
+\end{figure}
+The master equation \ref{eq:master} describes the evolution of the probability distribution of position through time.
+Loosely speaking the solution of this equation is expected to be some combination of the \citet{Einstein1937} theory for particle transport with the Gaussian propagator of the advection-diffusion equation \citep{Gillespie1992}.
+
+Because the master equation is second order in time, it requires initial conditions for $P$ and $\pt P$.
+Considering that particles start at rest in a mixture of motion and rest states, with a fraction $k_E/k$ starting in motion and a fraction $k_D/k$ starting in rest, these conditions derive from the state
+\be P(x,0) = \lim_{t\rightarrow 0 } \frac{k_E}{k} \sqrt{\frac{1}{4\pi D t}} \exp\Big[-\frac{(x-Vt)^2}{4Dt}\Big]+ \frac{k_D}{k}\delta(x)\ee
+which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep{Weiss2002}.
+
+The master equation \ref{eq:master} is solved by transform calculus in the appendix, providing
+\begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du.
+\label{eq:monster}
+ \end{multline}
+This integral term encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator.
+The integral term convolves the probability that the particle has been in motion for a period $u$ out of a time $t$ with the probability density that a particle has travelled a distance $x$ in time $u$.
+The prefactor involving partial derivatives can be understood as adapting this distribution to the initial conditions.
+
+The moments of position from this distribution could be calculated by integrating equation \ref{eq:monster}, but this is challenging.
+Instead, we calculate the governing equation of the moments directly from the master equation \ref{eq:master}.
+Multipling this equation by $x^l$ and integrating over all $x$ provides
+\be \pt^2 m_l -V l \pt m_{l-1} -k_E V l m_{l-1} + k \pt m_l - D l (l-1) \pt m_{l-2} - k_E D l(l-1) m_{l-2} = 0,\ee
+where $\bra x^l \ket = m_l$. 
+For $l = 1$, this equation generates the mean position $ \bra x \ket(t) = k_E V t/k$, which is unaffected by diffusion (since Gaussian velocity fluctuations are symmetric).
+The case $l=2$ provides the second moment, which provides the variance of position
+\be \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) + 2\frac{k_E D}{k}t.\ee
+At short times $t\ll k^{-1}$, this becomes
+\be \sigma_x^2 \sim \frac{k_E k_D V^2}{k^2} t^2 + \frac{2 k_E D }{k}t,\ee
+so it scales as $\sigma_x^2 \sim t^2$ when $t \gg \frac{2Dk}{V^2 k_D}$ and $\sigma_x^2 \sim t$ when $t \ll \frac{2Dk}{V^2 k_D}$.
+Similarly at long times $t\gg k^{-1}$, 
+\be \sigma_x^2 \sim \frac{2k_Ek_D V^2}{k^3}t,\ee
+so we have a non-trivial multi-scale diffusion phenomenon. 
+Provided that $2D/V^2 \ll k_D/k^2$, we therefore have
+\be \sigma_x^2 \sim 
+\begin{cases}
+	 t, & t \ll \frac{2Dk}{V^2 k_D}, \\ 
+	 t^2, &  \frac{2Dk}{V^2 k_D} \ll t \ll \frac{1}{k}, \\
+	 t, & t\gg \frac{1}{k}.
+\end{cases}\ee
+Note in the physical condition when $k\approx k_D$, the condition for the existence of three ranges becomes $ \Pe \ll 1,$ where
+$\Pe = 2 D k_D/V^2.$
+The small Peclet limit, when three diffusion ranges are obtained, is characteristic of bedload sediment transport where velocity fluctuations are typically small compared to mean downstream movement velocites.
+
+\subsection{Calculation of the flux}
+
+From the formalism in section xxxx, the central parameter of the sediment flux distribution is 
+\be \mu(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t).\ee
+This represents the rate of particles crossing $x=0$ at the observation time $T$ given they started somewhere to left of $x=0$ at $T=0$.
+
+It is convenient to calculate this quantity in Laplace space.   
+Taking the Laplace transform,
+\be \tilde{\mu}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s),\ee
+then performing both integrals provides
+\be  \tilde{\mu}(s) = -\frac{\phi D}{V R(s+k_E)} + \frac{2D\phi}{VR(1-R)(s+k_E)} + \frac{4D^2(s+k)}{V^3R(1-R)^2(s+k_E)}. \label{eq:laplacefluxrate}\ee
+$\mu(t)$ derives as the inverse Laplace transform of this equation. 
+This calculation is performed in the appendix, providing
+\begin{multline} 
+\mu(t) = \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
+\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{k_D}{2 k}\Big)e^{-V^2 u/4D} + \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du.
+\label{eq:fluxy}
+\end{multline}
+In this equation, the notation $\cev{\partial}$ means that the derivative acts from the left of all terms in which it is involved, as in $f(t)\cev{partial}_t g(t) = \partial_t f(t)g(t).$
+
+This is a complicated result for the rate constant in the sediment flux \ref{eq:fluxdist}, but this complexity is not surprising given that equation \ref{eq:langevin} involves two interacting diffusion processes.
+As displayed in \ref{fig:fig3}, as a result of eq.
+\ref{eq:fluxy} the mean flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $Q_0 = E l$ as the observation time becomes much larger than $1/(k_D \Pe)$.
+\begin{figure}
+	\centerline{\includegraphics{figures/ch2/figure3_slopeKey.pdf}}
+	\caption{}
+	\label{fig:fig3}
+\end{figure}
+The convergence time of the flux thus scales with the inverse Peclet number.
+When particle velocity fluctuations during motions are larger compared to advection, the flux becomes slow to converge, in general.
+
+\subsection{Connection to earlier work}
+
+The mathematical form of our results is complicated, although the ideas are conceptually clear.
+Particles alternate through motion and rest, and particle motions have a fluctuating velocity, not a deterministic velocity.
+Nevertheless, turning off velocity fluctuations by setting $D=0$ re-derives the results of \citet{Lisle1998}, who formulated sediment transport as a random walk between motion and rest states.
+Further taking the velocity infinite as the time spent in motion vanishes, while their product $V/k_D = \ell$, the mean step distance, rederives the central result of \citet{Einstein1937}, given originally in \ref{eq:einsteinpdf}.
+
+In either of these limit cases, the mean sediment flux becomes $\bra q \ket = E \ell$, with no dependence on observation scale.
+This indicates that scale dependence in the sediment flux is a consequence of velocity fluctuations among moving particles, at least when observation timescales are relatively short.
+
+\section{Discussion \label{sec:disc}}
+
+This chapter has generalized the random walk description of individual sediment trajectories b \citet{Lisle1998} and \citet{Lajeunesse2018} to include velocity fluctuations in the motion state.
+Using the probability distribution of particle position derived from this generalized model, I demonstrated how to calculate the sediment flux probability distribution. 
+This is a Poisson distribution, reflecting the conclusion that when sediment particles do not interact, and when they all undergo the same dynamics, a Poisson distributed-flux derives as an inevitable consequence.
+
+This is an interesting conclusion because in reality, sediment flux distributions are rarely Poissonian \citep{Ancey2008,Turowski2010,Dhont2010,Saletti2015}.
+Generally they have wide tails representing the possibility of extreme fluctuations which have been associated with burstiness or intermittency in sediment flux timeseries \citep{Singh2009, Benavides2021}.
+We conclude that non-interacting models of the sediment flux are probably not capable of describing realistic sediment transport fluctuations, which actually harks back to the realization of \citet{Ancey2006} which led them to introduce the collective entrainment process to generate realistic fluctuations \citep{Ancey2008}.
+Apparently, real sediment grains have correlated motions.
+Whether this is because of entrainment by particle-bed collisions \citep{Lee2018,Heyman2016}, or waves of entrainment caused by episodic losses of stability in the bed \citep{Heyman2013,Dhont2018,Benavides2021} remains an open question deserving of further research.
+
+\subsection{The role of stochasticity in landscape evolution}
+
+As reflected in this work and a vast number now of experimental studies, the sediment flux is a stochastic quantity. 
+The conventional view of landscape evolution applies the Exner equation, which evaluates the mass exchange between moving particles with the bed.
+The Exner approach assumes that the bed and sediment flux are continuous, deterministic quantities, so that topographic evolution is itself determinsitic as well.
+
+This continuous view of landscapes neglects the granularity which can moderate
+
+If sediment transport is stochastic, a deterministic Exner equation would not always be expected to correspond to reality.
+Here I explore the implications of a fluctuating sediment flux within the Exner formulation.
+In one dimension, the Exner equation was given in section \ref{sec:exner} as 
+\be \pt z(x,t) = -\px q(x,t). \ee
+Considering the sediment flux as a quantity which fluctuates both in space and time, this gives
+
+\subsection{Methods to calculate the sediment flux}
+
+An early approach to calculate the sediment flux distribution is due to \citet{Hamamori1962}, who evaluated the sediment flux as the result of migrating bedforms.
+This approach has been extended to include bedforms with a range of sizes \citep{Guala2014,Nikora1984}.
+
+Earlier approaches to calculate the probability distribution of the sediment flux have counted the fluctuating number of particles in a control volume with population models \citep{Ancey2008,Ancey2014}, or they have evaluated the average rate of particles crossing a control surface using the interarrival time distribution of particles from upstream \citep{Turowski2010,Ancey2020,Heyman2013}, in what is basically an application of renewal theory \citep{Cox1962}.
+These approaches can be described as ``kinematic", in that they do not explicitly account for the trajectories of individual particles as they move downstream.
+
+Here, I have demonstrated how to describe the sediment flux probability distribution from a dynamical perspective, rather than a kinematic one, using stochastic dynamical equations to describe the downstream trajectories of individual grains (eq.
+\ref{eq:langevin}.
+In effect, the interarrival time distribution is replaced by the probability distributions of position for all individual particles available for motion, with their initial positions and velocities considered as random variables.
+Particles moving downstream are counted as they cross the control surface, giving rise to a parameter $\Lambda(T)$, dependent on observation scale, which is analogous to the heuristic rate constant in the renewal theories that were reviewed in section \ref{sec:renewal} \citep{Turowski2010,Ancey2020}
+This new description of the sediment flux from the grain scale particle dynamics, rather than characteristic kinematics derived from them, represents considerable progress toward a mechanistic understanding of sediment transport from statistical physics.
+
+\subsection{Outlook and future research}
+
+The phrasing of sediment fluxes the topographies they moderate as statistical quantities, having mean configurations and fluctuations, provides some context on the severe challenges encountered in predicting landscape change with deterministic equations \citep{}.
+What we observe, looking at Earth's landscapes, may be a single realization, drawn from a spectrum of possibilities, whose precise form evades description. 
+The stochastic perspective allows us to
+
+The sediment flux probability distribution in equation \ref{eq:fluxdist} represents a Poisson distribution with an observation-scale dependent rate.
+Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
+This contrasts with many bedload transport experiments with wide-tailed fluctuations \citep{Ancey2008,Heyman2016,Fathel2015}, especially for weak flow conditions near initiation of particle motion \citep{Benavides2021}, exposing a need to generalize a mechanistic theory of the sediment flux developed here to provide wider fluctuations.
+
+There are essentially two mechanisms available by which wide fluctuations can arise \citep{Goh2008}.
+Both would be ultimately reflected in non-exponential interarrival times of particles \citep[e.g.][]{Turowski2010,Heyman2013}, although their origins would be unambigiously linked to the underlying particle transport mechanics.
+The first possibility is to introduce correlations between particle movements \citep{Ancey2008,Heyman2013,Ancey2014}, such as might arise when turbulent flow fluctuations entrain many particles at once \citep{Cameron2020} or drive waves of high velocity particles downstream simultaneously \citep{Nino1996}.
+In a theory such as the one I presented here, this first mechanism for wide fluctuations could be described by temporally unsteady velocities, entrainment or deposition rates, or diffusion coefficients in the Langevin model \ref{eq:langevin}.
+The second possibilty is to introduce interactions between moving particles, such as might result from collisions \citep{Lee2018} or wake vortices \citep{Schmeeckle2014} inducing particle entrainment, leading to collective waves of motion \citep{Ancey2014}.
+Including such effects would require a many-body generalization of the Langevin and master equations \ref{eq:langevin} and \ref{eq:master}, where the probability distribution of position is promoted to a joint form $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$, which is puzzling, since the number of particles $N(t)$ in the distribution fluctuates.
+The model of \citet{Ancey2014} hints at this possibility, although its starting point is not a dynamical equation as in \ref{eq:master}.
+Some hints as to how to generalize our work to include interactions between particles to describe wide-tailed sediment fluxes can be gleaned from literature on reaction-diffusion processes \citep{Hernandez2004,Cardy2006} and interacting gases \citep{Chapman1970,Brilliantov2004}.
+The further inclusion of interacting particle models into bedload transport theory would be an exciting development.
+
+\section{Conclusion \label{sec:conc}}
+
+This chapter has introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motion and rest, including fluctuating movement velocities in the motion state.
+This description of particle trajectories was connected to the probability distribution of the bedload sediment flux, linking this probability distribution to the individual particle dynamics for the first time.
+Notably, the bedload flux distribution calculated in this way is scale dependent, so long as particle velocities are fluctuating, indicating that bedload velocity fluctuations slow the convergence of measurements of averaged bedload fluxes, and relating the timescales over which mean bedload fluxes to the mean particle transport velocity and the strength of its fluctuations.
+
+\citet{Einstein1937} introduced the concept of describing individual particle motions, but this description was not clearly linked to the mechanics of bedload transport that involve turbulent flow and particle-particle contacts and collisions.
+In the thesis up to now I have emphasized that Einstein's model and the works which followed it can actually be viewed as dynamical equations, reminiscent of Newton's third law, where the time derivatives of particle position equate to stochastic driving terms, force-like quantities, which provide phenomenological representations of particle movement.
+These stochastic driving terms are always highly simplified and idealized forces -- sequences of pulses, alternating switches, and up to now, they have always been used in isolation. 
+In this work I combined these idealized driving forces to provide a more realistic description of particle trajectories moving downstream through motion and rest.
+On the one hand, particle velocities alternate through entrainment and deposition, and on the other, they fluctuate during the period of motion.
+This strategy of combining idealised noises to approximate the complex reality of fluid turbulence, grain-grain collisions, and the entrainment and deposition processes these generate has potential to be extremely useful in future research.

### 2021-06-29 ch5langevin.tex
got some stuff figured out .

@@
 In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations, quantifying the dependence of measurements on the observation scale, and linking bulk transport characteristics to the ``microscopic" dynamics of individual grains.
-Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Boltzman, Lajeunesse2020} and bedform initiation \citep{Ancey2014,Bohorquez2016} provide additional motivation for these stochastic formulations.
-Many stochastic approaches express downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume.
-In these approaches, the instantaneous velocity distribution of sediment particles therefore becomes an object of crucial importance.
-Unfortunately, current understanding of this distribution remains limited, and no mechanistic models have yet been presented that describe the full range of experimental observations.
-Here, we take some steps toward rectifying these shortcomings.
+Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Boltzman, Lajeunesse2020} and bedform initiation \citep{Ancey2014,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
+One subset of these methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume.
+These approaches highlight the instantaneous velocity distribution of sediment particles as an object of crucial importance.
+Unfortunately, current understanding of bedload velocity distributions remains limited.
+We have as of yet no consensus on the shape of the bedload velocity distribution (is it Exponential, Gaussian, Gamma, or some other shape?), and no mechanistic models have yet been developed to describe the full range of experimental observations.
+Here, we develop a stochastic model of particle velocities to addresses these issues.
 
-High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears.
-One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}. 
+Particle trajectories are a compromise between turbulent drag and particle-bed collision \citep{Wiberg1985}.
+We therefore anticipate that the Stokes number, which is known to characterize colliding bodies within a viscous fluid \citep{}, is an important dimensionless parameter of the bedload velocity distribution.
+The Stokes number is defined as ... where ....
+
+High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears in a given set of hydraulic and sedimentary conditions.One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
 These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$.
@@
 
-We developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
+Here, I developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
 The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions,
@@
 Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, we have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise.
-Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Coleman2019,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are relatively long compared to the timescales of turbulent fluctuations \citep{}.
+Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Coleman2019,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
 We modelled particle-bed collision forces as a sequence of instantaneous impulses where the intervals between successive collisions were characterized as exponential random variables.
@@
 Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the setting here is somewhat different than grains in air.
-Because particles within a viscous flow interact at a distance, we should expect the collision model will become poor when the time between subsequent collisions becomes small.
-Therefore, although the model seems appropriate for saltation, it could become questionable for the ``reptation" transport mode when the times between subsequent particle-bed collisions are short.
-Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport.
-Unfortunately, numerical methods would likely be required to interpret the resulting equations, in contrast to the analytical approaches applied here.
+Because particles within a viscous flow in general interact at a distance, we should expect the collision model will become poor when the time between subsequent collisions becomes small.
+Therefore, although the model seems appropriate for saltation, it should be critically examined for ``reptation", when the times between subsequent particle-bed collisions are short.
+Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport, although these might require numerical methods for their solution.
+The theory of granular gases provides some indication of the types of more nuanced collision models which are possible \citep{Brilliantov2004,other..}.
 
@@
 In fluid dynamics, the dissipation characteristics of particle-particle collisions in viscous flows are known to depend on the particle size and approach velocity through the Stokes number. 
-In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic particles generate non-Gaussian distributions.
+In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic grains generate non-Gaussian distributions \citep{Chapman1970,Brilliantov2004}.
 Taken together, these ideas suggest that we might relate the shape of the particle velocity distribution to particle size.
@@
 Even in the absence of sorting effects and differential mobility, smaller grains can be expected to carry more control over the largest fluctuations in the overall transport rate, since their velocity distributions have wider tails.
+
+\section{Implications for landscape evolution}
+
+Many phenomena of landscape evolution depend on the velocities of individual grains.
+The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Skylar2004,Tingian,Turowski2020}, on which they have a sensitive nonlinear dependence.
+It is not much of a stretch to imagine that bedrock erosion could be dominated by the largest impact velocities that originate from the extreme tail of the bedload velocity distribution.
+Here, we have shown that the weight of this tail depends sensitively on the dissipation characteristics of particle-bed collisions, suggesting that another feedback, that between dissipation and ...
+may be at play in the evolution of bedrock canyons.
+
+\section{Rivers, hillslopes, and dunes}
+
+It has long been recognized, probably since \citet{Bagnold1941} and certainly today \citep{}, that there are deep analogies between tranpsort phenomena in air and water.
+In an ideal perspective, we might imagine writing the governing equations for transport in water as a function of the fluid viscosity, then tuning the viscosity between air and water to obtain a description applying to both spheres.
 

### 2021-06-29 pref.tex
got some stuff figured out .

--- 2021-06-24 pref.tex
@@
 
-\chapter{Preface}
+\chapter{Preface: complementary traditions}
 
-A preface should be written indicating what is to come and what the point of this whole thing is.
+This thesis deals with the transport of coarse sediment in flowing water from the perspective that sediment transport results from the movements of individual grains.
+The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.
+
+Geomorphology has long been characterized by observation, description, and the building of conceptual models, not mathematical ones.
+Only relatively recently has the science turned toward quantitative methods, with researchers working to frame observations in terms of underlying processes and to describe them with methods adapted from physics.
+
+Given the complexity of geomorphology problems, a complete reduction of geomorphology to physics would be narrow-minded, but we can nonetheless borrow ways of thinking.
+An extremely successful approach in physics has been to construct idealized models with little intention of direct realism, study them deeply to obtain complete understanding, and then, decades or centuries later, to embellish these models with more sophisticated features to describe real-world phenomena.
+An example is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow shows up in the deepest inquiries of theoretical physics, from quantum matter to cosmological inflation.
+This thesis applies the simple harmonic oscillator approach to some problems in Earth science.
+The objective is to build up simple archetypes which can be embellished later, not to build \textbf{the model} of fluvial geomorphology. 
+That's for later. (And it's not for me!) CHEERS.

### 2021-07-02 ack.tex
got another intro figure constructed.

--- 2021-06-24 ack.tex
@@
 
-Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way.
-Tyler cannon.
-.
-.
+Attending university long enough to earn a PhD is an unbelieveable priviledge, and I only got here with an incredible amount of guidance and support.
 
-Shawn Chartrand, Conor McDowell, Matteo Saletti, Will Booker, David Adams, Leo King, Tobias Mueller, Nisreen Ghazi,
+Most of all, to mom, dad, Kelsey, Kim, and Gugs, you've formed all of the best parts of my personality, and it's impossible to explain how thankful I am for that.
 
-Adam Payne, Adam Collins, Megat Denney, Andrew Rice, Evan Graber, Scott Ferris, Craig Tenney, and the rest of the WVU physics lounge family.
+Mom, during my childhood, you worked full time, taught piano lessons, earned Master and Doctoral degrees, played organ every Sunday, consoled us through every difficulty, and engaged us in as many extracurricular activities as we were willing to take on.
+I cannot imagine a better example of work ethic, open-mindedness, and devotion. I hope I picked a few things up.
 
-Andy Osadetz, Josh Plankeel,
+Dad, I've watched you solve an insane range of problems without hesitation, from worn out brake pads to broken arms.
+By age 10 I must have helped with copper plumbing, woodwork, car repair, home wiring, and tree felling.
+I had personal tours of West Virginia's backroads and civil engineering works, and I learned more about science paddling the South Branch with you than I ever did in school.
+To you, water currents, fish habits, electrons, hellgrammites, engines, heat, and radio waves are all deeply important, and any man-made problems are easy enough to solve with time and the right tools.
+Thanks for sharing the perspective of a problem solver and knowledge seeker! And of course so much more.
 
-My partner Mary for putting up with the late nights.
+Kelsey and Kim, my earliest memories are sitting in your bedroom floor struggling through your books, asking you for help every couple of paragraphs.
+You made me good at reading and taught me to enjoy it, which is the foundation of all my learning.
+Plus, we had a lot of hilarious times.
+I know it goes without saying, but thanks.
+
+Gugs, your approach to life is clean and selfless, and I cannot imagine a single way you could have been more reliable, stable, or invested in my development.
+I'll never be able to thank you enough!
+One day I'll take up your 4am wakeups and adopt your completely measured and calm responses to life's challenges.
+Until then, I'll keep thinking about I'd ideally do in terms of what you always do, and if we're lucky I might even occasionally pull it off.
+
+Another family member who deserves special mention is my great aunt Alice.
+Thanks so much for paying for my undergraduate education.
+We hardly met, but we're family, and that was enough.
+It really means a lot.
+
+Thanks to my early educators who put in the extra care to keep me on track, Jonathan Ramey, Linda Afzhalirhad, Brandon Willard, Linda Mendez, Marty Ojeda, Phyllis Adkins, and Red Hamlin among many others.
+Thanks also to the later ones at Southern WV Community College (SWVCC) and West Virginia University (WVU) whose lessons fostered my way of thinking and provided me with many of the tools I used for the research in this thesis.
+At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
+Mindy in particular transformed my view of the world into one based on mechanics, and provided me with a deep understanding of general physics and calculus.
+Everyone there always encouraged me to continue my education.
+You all made SWVCC a personal transformation and an unbelievable privilege.
+
+At WVU, I'd like to thank Alan Bristow, Tudor Stanescu, and Leo Golubovic.
+Alan and Tudor, thanks for providing my first opportunities in research.
+Leo, you are the key person in my intellectual development.
+You inspired me to learn mechanics, statistical physics, and mathematical physics at a high level, and I can't thank you enough.
+Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr.
+Golubovic would like this one." I wish I had taken more advantage of the opportunity to speak to you outside of classes.
+I'll feel deeply successful if I can one day become as inspiring an educator as you are.
+
+To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Andrew Rice, Andrew Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times.
+I'll never forget Scott's quaternion obsession, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
+I have yet to experience teamwork and shared exploration like we had in the undergrad lounge.
+I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, and that they have some Andrew Rice imitator to flip some cars for em (mike jones).
+Because of you all, WVU was magic for me.
+
+Now we come to Marwan Hassan.
+About five years ago now I wandered into Marwan's office saying I liked rivers and wanted to describe sediment movement with statistical physics.
+I realize now when I said ``statistical physics" your eyes lit up with thoughts of Einstein, Nakagawa, Yano, Furbish, Hubbell, and all of the names I now know well.
+I was lucky to use the right trigger words.
+Marwan, thanks for taking the risk and taking me on from outside the field, for all of the guidance, for loaning me so many papers and books, giving me freedom to explore ideas, and most of all for showing me that it's ok to be human in science.
+At this point I also want to thank Yusuf for trying your patience to its limits over a long enough period so that things didn't seem so bad once I came along for this short time together.
+
+My friends in geography have also been great.
+Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, David, Jiamei, Xingyu, Niannian, and Emma, plus the outsiders, Will, Dave, Rose, and Anya.
+In particular, thanks to Shawn for all of the chats and shared speculations, and Matteo and Conor -- friendships forged through suffering!
+We'll have to keep in touch.
+
+Thank you Rui Ferreira and Brett Eaton!
+Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics.
+Your kindness and wide-ranging interests beyond hydraulics are inspiring.
+Thanks for the thoughtful comments on my PhD.
+I'm not forgetting what I owe you!
+Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand modelling in light of the real world and its complexity.
+Although I'm not sure if I understood your lessons as well as you would have liked, I have a list of your favorite books on my computer, and they're a first task for as soon as I defend.
+Maybe they'll bring me closer to your way of thinking.
+
+I'd be amiss not to mention David Furbish.
+Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation.
+At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes".
+Now I have a thesis on it.
+Later on, David recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social repercussions of disinformation, and a whole range of questions including the meaning of probability in landscape evolution, probability as being just as fundamental as mass, momentum, and energy, and fringe ideas in cosmology.
+These are my people!
+David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me hop on the spacecraft.
+
+Among UBC Geographers, thanks especially to Nina Hewitt!
+It was always a pleasure and maybe a sigh of relief to TA your courses, which are so uniquely organized and student-focused.
+UBC Geography is a much greater department with you in it.
+I'm so thankful to have been associated with your influence on the undergrads.
+I'll be happy if I can teach like you one day!
+
+UBC Geography is successful because it has excellent staff, Vincent Kujala, Sandy Lapsky, Bret Peterson, Suzanne Lawrence, Connie Cheung, Jose Aparicio, thank you all so much! 
+Vincent showed me the magic of bash, Sandy handled conference expenses and got me back in an office during the pandemic when my productivity was floundering, Bret put out the tech fires, Suzanne organized the paychecks and appointments, and Jose handled every online teaching challenge.
+You all saved me a ton of time, and I really appreciate it.
+
+The kind, open-minded, and collaborative atmosphere I experienced here is clearly a legacy from the former generation of UBC's physical geographers, some whom I had the pleasure of meeting and learning from, like Mike Church and Olav Slaymaker, and others whom I didn't.
+Thanks Mike and Olav for your inspiring seminars and the academic environment you cultivated.
+
+Finally, Mary!
+My PhD has not been great in terms of work-life balance.
+Thanks for sticking it out, putting up with my late bedtimes, my early wakeups, and all of those social events I didn't come to.
+It's time!
+I'm done.
+The pandemic also seems to be winding down.
+We can finally get to Aus!
+I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella.
+After that, we're a step closer to settling somewhere and determining once and for all whether our future cat will be white or orange.
+
+Completing a PhD looks like a personal achievement, but it's really not.
+It's an expression of 30 years of influence and guidance from other people at every stage.
+I'm kinda tired because it's the end.
+Maybe because of it, I find it easy to drift into daydreams on the parts of my life I've really loved, Watoga and Williams River with my family, reading encyclopedias in Gugs' living room, lining up just left of center with my dad and bracing on the paddle for the hit, nighttime fireflies under my great grandpa's tree, and sliding down rocks at Kumbrabow with my sisters.
+The daydream shifts to years later, as I flip on the light in the physics lounge and surprise Megat who was asleep on the couch, I finish a problem set with friends, discuss Tsujimoto with Marwan (``those guys did everything!"), or just walk around Kitsilano with Mary at the edge of dark.
+It's all one thing!
+Thank you all for every part of it.

### 2021-07-02 ch1intro.tex
got another intro figure constructed.

@@
 
-Landscapes evolve when water, wind, and ice, driven by gravity, erode the uplifted relics of forces from below Earth's surface.
-Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands as they grind them to sand through a machine of denundation.
-Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
+Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes uplifted by tectonics.
+Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands, grinding boulders to sand in a machine of abrasion.
+Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
 Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
@@
 
-In this context river geomorphology as a scientific discipline is shifting toward more quantitative methods that enable concrete predictions about the natural world.
-Sediment transport in river channels is especially amenable to this quantitative approach since it is basically the result of fluid and granular physics \citep{Frey2011}.
-Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains.
-When particles are coarse as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface.
-In bedload transport conditions, fluid turbulence and the irregular bed surface inherit special control over sediment dynamics.
+In this context river geomorphology as a scientific discipline has shifted toward more quantitative methods that enable concrete predictions about the natural world \citep{Church2005}.
+Sediment transport in river channels is especially amenable to this quantitative approach since it is mechanistically the result of fluid and granular physics.
+Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
+When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}.
+In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.
 
-Bedload transport exerts particular influence over channel morphology and stability \citep{Church2006,Hassan2008,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Zimmerman2007,Comiti2012, McKenzie2018, Eaton2020}.
+Bedload transport exerts unique influence over channel morphology and stability \citep{Church2006,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Hassan2008, Eaton2020}.
 A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
-Unfortunately, existing approaches to compute the bedload flux are inadequate, with predictions regularly deviating from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012, Dhont2018}.
-Considering that this is despite well over a century of research into bedload transport \citep{Gomez1991}, we can conclude that new research approaches are sorely needed \citep{Ancey2020a,Ancey2020}.
+Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
+Given that this problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.
 
@@
 Local fluxes can range through orders of magnitude as details of turbulent fluctuations and bed organization vary, while average characterizations of flow and sediment remain constant \citep{Sumer2003, Charru2004, Hassan2008, Venditti2017}.
-The same turbulence and sediment organization details which correlate with the bedload flux also interact with it. 
-Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}, and bedload fluxes modify the stability and arrangement of bed surface grains \citep{Kirchener1990, Charru2004, Strom2008}, which encourages further transport in a positive feedback \citep{Ancey2008,Heyman2016,Lee2018}.
-Bedload fluxes are thus in a cyclical feedback with their controls in the details of turbulence and bed organization \citep{Jerolmack2005}.
+The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
+Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}.
+The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018}.
+Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values.
 
-The approach taken in this thesis is to consider the bedload flux as an aggregate result of many individual transported grains.
-This departs from the traditional strategy of correlating transport rates to the mean flow and sedimentary characteristics, which considers the granular phase as a continuum \citep{Gilbert1917,MeyerPeter1948}.
-Since the trajectories of individual grains are governed by Newtonian mechanics, this alternative perspective provides traction on the problem.
-This Newtonian approach is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time, in part due to the turbulent flow and the erratic interactions of moving particles with the bed.
-To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
-These efforts build on an an earlier literature to provide more realistic descriptions of the trajectories of individual grains and the bedload fluxes they generate.
+The approach taken in this thesis is to consider bedload transport as an aggregate result of many individual transported grains whose movements have probabilistic characteristics.
+This departs from the traditional strategy of correlating transport rates to the average flow and sedimentary characteristics \citep{MeyerPeter1948, Parker1990, Wilcock2001}.
+Because the trajectories of individual grains are governed by Newtonian mechanics, this approach brings reliable tools to the problem, but it is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time due to fluid turbulence and the erratic interactions of moving particles with the bed.
 
-The thesis begins with a review of the most relevant pieces of this earlier literature on which my research builds.
-This review rephrases works when necessary to indicate common themes and to show continuity with my own research which follows in the subsequent chapters.
+To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and the body of earlier literature which inspired my approach.
+To provide context and motivation, the thesis begins by reviewing this literature, rephrasing earlier work as necessary to indicate common themes and show continuity with my own developments which follow in subsequent chapters.
 
@@
 
-A basic (but not simple) problem in sediment transport modelling is to predict the downstream movement of an individual particle.
-At first glance, this seems an elementary problem in general physics, but challenges appear.
-Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not known \citep{Furbish1997}.
-Downstream movement is resisted by frictional collisions between moving particles and the bed.
-Since the bed is a granular surface, its geometry is very difficult to characterize \citep{Gordon1972}, and the outcome of collisions varies from one to the next \citep{Sekine1992}
+A basic problem in sediment transport theory is to predict the downstream movement of an individual particle.
+At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
+Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Furbish1997}.
+Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the outcomes of these encounters are variable \citep{Sekine1992}.
+
 Entrainment and deposition provide an additional layer of complexity.
-Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and enable their deposition \citep{Charru2004}.
-These pockets are not permanent shelter, because rearrangement of the surrounding bed by conventional entrainment or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if the pocket geometry is not disturbed \citep{Valyrakis2012,Celik2014}.
-As a result, particles generally alternate through sequences of entrainment and deposition.
-Ultimately, particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}.
-These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}.
+Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and cause their deposition \citep{Charru2004}.
+These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2012,Celik2014}.
+As a result, particles generally alternate through sequences of entrainment and deposition, but determining precisely when these events occur is impractical.
 
-Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales.
-\citet{Nikora2001,Nikora2002} provided a conceptual model of individual particle motions which helps to organize this complexity.
-Nikora et al divided the downstream trajectory of an individual particle into three timescales, or ``ranges", termed local, intermediate, and global.
-The local range refers to the period of motion between subsequent interactions with the bed, when the particle accelerates downstream within the flow, the intermediate range reflects particle motions through sequences of collisions, and the global range refers to particle motion between subsequent entrainment and disentrainment events.
-\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried. 
-This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize transport models of individual grains.
+Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}, 
+These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Church1994,Ferguson2002}.
+
+Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales, from seconds to years, and each of these processes eludes precise description.
+
+\citet{Nikora2001,Nikora2002} provided a conceptual framework which helps to organize this complexity.
+Nikora et al divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
+The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
+The intermediate range reflects particle motions through sequences of bed encounters, when particles alternately accelerate and decellerate. 
+The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
+\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017,Pretzslav2021}.
+This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize the literature describing trajectories of individual grains.
 
 \subsection{Motivation: tracers and basic understanding}
-The original motivation to understand individual particle motions was probably to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
-A common measurement technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
-In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion in the stream to calculate the overall sediment flux \citep{Yano1969,Nakagawa1976,Hassan2017}.
+An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
+A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
+In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997,Ferguson2002}.
 In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales.
 Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenberger2012}.
-This means it is not clear how measured tracer velocities can be mapped back to bulk sediment fluxes, exposing a need for further research into how exactly particle motion characteristics evolve through time.
+This obscures the mapping between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.
 
@@
 
-Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest, but it is clearly oversimplified from the reality of particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
+Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest.
+Despite its success, it oversimplifies particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
 

### 2021-07-02 ch2flipFlop.tex
got another intro figure constructed.

@@
 	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra \mathcal{I}_i(T) \big\ket \Big] \end{align}
-This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $ \mathcal{I}_i(T)$ is either $1$ or $0$ ($e^{ax} = 1-(1-e^a)x$ if $x=0,1$).
+This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $ \mathcal{I}_i(T)$ is either $1$ or $0$, so that $e^{ax} = 1-(1-e^a)x$.
 The average over initial conditions and possible trajectories for the $i$th particle can be written
@@
 
-This chapter has generalized the random walk description of individual sediment trajectories b \citet{Lisle1998} and \citet{Lajeunesse2018} to include velocity fluctuations in the motion state.
-Using the probability distribution of particle position derived from this generalized model, I demonstrated how to calculate the sediment flux probability distribution. 
-This is a Poisson distribution, reflecting the conclusion that when sediment particles do not interact, and when they all undergo the same dynamics, a Poisson distributed-flux derives as an inevitable consequence.
+This chapter has generalized the earlier descriptions of individual sediment trajectories \citep[e.g.][]{Lisle1998,Lajeunesse2018} to include velocity fluctuations in the motion state.
+Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
+Calculated in this way, the sediment flux distribution is observation-scale dependent, describing how sediment fluxes depend on the period over which they are measured.
 
-This is an interesting conclusion because in reality, sediment flux distributions are rarely Poissonian \citep{Ancey2008,Turowski2010,Dhont2010,Saletti2015}.
-Generally they have wide tails representing the possibility of extreme fluctuations which have been associated with burstiness or intermittency in sediment flux timeseries \citep{Singh2009, Benavides2021}.
-We conclude that non-interacting models of the sediment flux are probably not capable of describing realistic sediment transport fluctuations, which actually harks back to the realization of \citet{Ancey2006} which led them to introduce the collective entrainment process to generate realistic fluctuations \citep{Ancey2008}.
-Apparently, real sediment grains have correlated motions.
-Whether this is because of entrainment by particle-bed collisions \citep{Lee2018,Heyman2016}, or waves of entrainment caused by episodic losses of stability in the bed \citep{Heyman2013,Dhont2018,Benavides2021} remains an open question deserving of further research.
+\subsection{Fluctuations and collective motions}
 
-\subsection{The role of stochasticity in landscape evolution}
+The sediment flux probability distribution in equation \ref{eq:fluxdist} represents a Poisson distribution with an observation-scale dependent rate.
+Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
+In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2010,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
+This exposes a need to generalize a mechanistic theory of the sediment flux I developed here to produce wider fluctuations.
 
-As reflected in this work and a vast number now of experimental studies, the sediment flux is a stochastic quantity. 
-The conventional view of landscape evolution applies the Exner equation, which evaluates the mass exchange between moving particles with the bed.
-The Exner approach assumes that the bed and sediment flux are continuous, deterministic quantities, so that topographic evolution is itself determinsitic as well.
+Descriptions of sediment transport based on population dynamics in a control volume have realistically-wide fluctuations by incorporating a positive feedback between the number of moving particles and the particle entrainment rate called collective entrainment \citep{Ancey2008,Ancey2014}.
+This feedback generates spatially-correlated waves of moving particles \citep{Ancey2014, Heyman2015} and produces non-exponential interarrival time distributions \citep{Heyman2013} which imply wide-tailed flux distributions when incorporated in renewal theory \citep{Turowski2010,Ancey2020}.
+Collective entrainment has been attributed to particle-particle interactions, such as small granular avalanches and collision-induced entrainment, and to fluid-particle interactions, such as large eddies entraining particles as they sweep downstream \citep{Heyman2014b,Ancey2014,Lee2018}.
 
-This continuous view of landscapes neglects the granularity which can moderate
+We can imagine generalizing the description in equation \ref{eq:Langevin} to include such fluid-particle interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
+Particle-particle interactions would be more challenging to include.
+The dynamical equation \ref{eq:langevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1956}.
+The resulting joint distribution of particle positions and velocities ($P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$) might be formulated by analogy to the theory of reaction diffusion systems \citep{Cardy2006}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
+By analogy to collective entrainment, interacting models which generalize the approach established in this chapter should be capable of producing wide sediment transport fluctuations.
 
-If sediment transport is stochastic, a deterministic Exner equation would not always be expected to correspond to reality.
-Here I explore the implications of a fluctuating sediment flux within the Exner formulation.
-In one dimension, the Exner equation was given in section \ref{sec:exner} as 
-\be \pt z(x,t) = -\px q(x,t). \ee
-Considering the sediment flux as a quantity which fluctuates both in space and time, this gives
+\subsection{Measurement protocol for the sediment flux}
 
-\subsection{Methods to calculate the sediment flux}
+\subsection{Stochasticity in landscape evolution}
 
-An early approach to calculate the sediment flux distribution is due to \citet{Hamamori1962}, who evaluated the sediment flux as the result of migrating bedforms.
-This approach has been extended to include bedforms with a range of sizes \citep{Guala2014,Nikora1984}.
+\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{eq:exner}), writing
+\be (1-\phi) \pt h(x,t) = -\px q(x,t). \ref{eq:stocEx}\ee
+In this one-dimensional equation, $h(x,t)$ is a continuous field representing the longitudinal channel profile.
+Traditionally, the Exner equation is evaluated considering the flux as a deterministic quantity \citep{Parker2007,Viparelli2011,An2017}, which provides a solitary trajectory of the profile through time.
+
+Considering the flux in the Exner equation as a stochastic quantity requires us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2014}.
+In contrast to the traditional analysis, a stochastic Exner equation with fluctuating $q(x,t)$ generates an ensemble of possibilities for the channel profile $z(x)$ at each instant.
+This set of possibilities is characterized by a probability distribution $P(\{z(x)\},t)$ for each possible profile.
+Mathematically, since the channel profile is a field, evaluating such a probability distribution of channel profiles would require statistical field theory \citep{Kardar2007}.
+A starting point on this analysis might be $P(\{z(x)\},t) = \bra \delta[z(x,t)-h(x,t)]\ket,$ where $h(x,t)$ involves the stochastic flux $q(x,t)$ from equation \ref{eq:stocEx}.
+In this equation, $\delta[f(x)]$ is a Dirac delta functional, which generalizes the delta function $\delta(x)$ from a single-valued parameter $x$ to a continuous field $f(x)$.
+Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2004,Nelson2004}, but they have not been introduced in geomorphology.
+
+Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
+The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1960}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
+Progress in stochastic approaches to river science \citep[e.g.][]{Furbish2021c,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask again with these classic works to what extent fluctuations shape Earth's surface.
+
+\section{Summary and Conclusion \label{sec:conc}}
+
+This chapter introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motions and rests.
+The motion phase was enriched to include velocity fluctuations, forming the first physical model of this type.
+The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying trajectories of individual particles.
+The measurement timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations in terms of a Peclet number.
+
+These results generalize the bedload trajectory models of \citet{Einstein1937}, \citet{Lisle1998}, and \citet{Lajeunesse2018} to include fluctuating movement velocities, and they reframe earlier renewal models of the sediment flux probability distribution \citep{Lajeunesse1010,Ancey2020} to explicitly include the transport dynamics of individual sediment grains.
+They further quantify the dependence of the sediment flux on the observation scale, and provide guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in simplified configurations when bedforms are absent.
+Next steps are to include interactions between particles to produce wider sediment transport fluctuations, and to evaluate the effects of sediment transport fluctuations on channel morphodynamics.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\endinput
+Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
+The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1960}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
+Progress in stochastic approaches to river science \citep[e.g.][]{Furbish2021c,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask again with these classic works to what extent fluctuations shape Earth's surface.
+
+This chapter upscaled from individual particle motions to calculate the sediment flux.
+We can imagine going further, to upscale from the sediment flux to understand channel morphodynamics.
+The sediment flux is stochastic as a result of variability in the individual particle trajectories forming it.
+As a result, the channel morphodynamics should also adopt this stochasticity, although very few studies have evaluated this possibility \citep{Bohorquez2016, Jerolmack2005}.
 
@@
 The further inclusion of interacting particle models into bedload transport theory would be an exciting development.
-
-\section{Conclusion \label{sec:conc}}
-
-This chapter has introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motion and rest, including fluctuating movement velocities in the motion state.
-This description of particle trajectories was connected to the probability distribution of the bedload sediment flux, linking this probability distribution to the individual particle dynamics for the first time.
-Notably, the bedload flux distribution calculated in this way is scale dependent, so long as particle velocities are fluctuating, indicating that bedload velocity fluctuations slow the convergence of measurements of averaged bedload fluxes, and relating the timescales over which mean bedload fluxes to the mean particle transport velocity and the strength of its fluctuations.
-
-\citet{Einstein1937} introduced the concept of describing individual particle motions, but this description was not clearly linked to the mechanics of bedload transport that involve turbulent flow and particle-particle contacts and collisions.
-In the thesis up to now I have emphasized that Einstein's model and the works which followed it can actually be viewed as dynamical equations, reminiscent of Newton's third law, where the time derivatives of particle position equate to stochastic driving terms, force-like quantities, which provide phenomenological representations of particle movement.
-These stochastic driving terms are always highly simplified and idealized forces -- sequences of pulses, alternating switches, and up to now, they have always been used in isolation. 
-In this work I combined these idealized driving forces to provide a more realistic description of particle trajectories moving downstream through motion and rest.
-On the one hand, particle velocities alternate through entrainment and deposition, and on the other, they fluctuate during the period of motion.
-This strategy of combining idealised noises to approximate the complex reality of fluid turbulence, grain-grain collisions, and the entrainment and deposition processes these generate has potential to be extremely useful in future research.

### 2021-07-02 ch5langevin.tex
got another intro figure constructed.

@@
 These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Rey_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
-At realistic $\Rey_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Rey_p$ formulas \citep{Schmeeckle2007,Mei2006}.
-A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Rey_p$ can be written $F_D = \frac{\pi}{8}
+At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Rey_p$ formulas \citep{Schmeeckle2007,Mei2006}.
+A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
 \rho_f d^2 C_D(\Rey_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.

### 2021-07-02 pref.tex
got another intro figure constructed.

@@
 
-This thesis deals with the transport of coarse sediment in flowing water from the perspective that sediment transport results from the movements of individual grains.
+This thesis deals with the transport of coarse sediment in flowing water from the perspective that sediment transport results from the movements of individual grains, and it is entirely the original research of Kevin Pierce.
+The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
+The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \citet{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor.
+Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to the submission of the thesis.
+
 The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.
-
 Geomorphology has long been characterized by observation, description, and the building of conceptual models, not mathematical ones.
@@
 An example is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow shows up in the deepest inquiries of theoretical physics, from quantum matter to cosmological inflation.
-This thesis applies the simple harmonic oscillator approach to some problems in Earth science.
+This thesis applies this ``simple harmonic oscillator approach" to some problems in Earth science.
 The objective is to build up simple archetypes which can be embellished later, not to build \textbf{the model} of fluvial geomorphology. 
-That's for later. (And it's not for me!) CHEERS.
+That's for later! (And not for me.) CHEERS.

### 2021-07-02 ack.tex
getting the appendices structured. made lots of progress on figures.

--- 2021-06-24 ack.tex
@@
 
-Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way.
-Tyler cannon.
-.
-.
+Attending university long enough to earn a PhD is an unbelievable privilege, and I only got here with an incredible amount of guidance and support.
 
-Shawn Chartrand, Conor McDowell, Matteo Saletti, Will Booker, David Adams, Leo King, Tobias Mueller, Nisreen Ghazi,
+Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele (Gugs), you've formed all of the best parts of my personality and continually guided me toward this point, and it's impossible to explain how thankful I am for that.
 
-Adam Payne, Adam Collins, Megat Denney, Andrew Rice, Evan Graber, Scott Ferris, Craig Tenney, and the rest of the WVU physics lounge family.
+Mom, during my childhood, you worked full time, taught piano lessons, earned Master and Doctoral degrees, played organ every Sunday, consoled us through every difficulty, and engaged us in as many extracurricular activities as we were willing to take on.
+I cannot imagine a better role model of work ethic, open-mindedness, and devotion. I hope I picked a few things up.
 
-Andy Osadetz, Josh Plankeel,
+Dad, I've watched you solve an insane diversity of problems without hesitation, from easy ones, like worn out brake pads or blown engines, to hard ones, like burnt eyes and broken arms.
+By age 10 I must have helped with copper plumbing, woodwork, car repair, home wiring, and tree felling.
+I had unintentional lessons in water currents, fish habits, electrons, engines, wings, hellgrammites, heat transfer, and radio waves.
+Canoeing the South Branch together taught me more science than school ever did.
+The overarching message is that phenomena can be understood, and most any problems can be solved, at least with the right tools.
+It's incredibly useful in a PhD!
+Thanks for this and so much more.
 
-My partner Mary for putting up with the late nights.
+Kelsey and Kim, my earliest memories are sitting in your bedroom floor struggling through your books, asking for help every couple of paragraphs.
+You made me good at reading and taught me to enjoy it, and this is the basis of everything.
+Kim, you were my first role model, and I think of you all the time.
+Kelsey, I know it goes without saying, but thanks for a lifetime of guidance.
+You're now, as always, five steps ahead and far more successful.
+On the one hand, I'm deeply proud at your progress through life, with a stable career and now two beautiful kids, but on the other, I'm flatly shocked by it.
+I'll be playing catch up forever!
+
+Gugs, your approach to life is clean and selfless, and you've been so exceptionally reliable and invested in my development, more like a second mother than a typical grandmother.
+I'll never be able to thank you enough!
+It's always my goal to take up your 4am wakeups and completely measured responses to all of life's challenges. I'm making a bit of progress. Let's hope I can get there!
+
+Finally with regard to family, I'd like to thank my great aunt Alice.
+Thanks so much for paying my undergrad education!
+We hardly met before you passed, but we're family, and that was enough.
+It really means a lot.
+
+Next comes education.
+Thanks to my early teachers and mentors who put in the extra care to keep me on track and shield me from harm, Jonathan Ramey, Linda Afzhalirhad, Brandon Willard, Linda Mendez, Marty Ojeda, Phyllis Adkins, and Red Hamlin among many others.
+Thanks also to the later ones at Southern WV Community College (SWVCC) and West Virginia University (WVU) whose lessons fostered my way of thinking and provided me with many of the tools I used for the research in this thesis.
+At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
+Mindy in particular transformed my view of the world into one based on mathematics with her incredibly effective teaching, Anne dismantled it into chemistry, and Tex reconstructed it into artistic interpretation, which will always transcend science.
+Everyone encouraged me to continue my education.
+You all made SWVCC a transformation and an unbelievable privilege.
+
+At WVU, I'd like to thank Alan Bristow, Tudor Stanescu, and Leo Golubovic.
+Alan and Tudor, thanks you for providing my first opportunities in research.
+(Sorry I was bad at it!) Leo, you are the key person in my development as a researcher.
+You inspired me to learn mechanics, statistical physics, and mathematical physics at a high level, and I can't thank you enough.
+Thanks to you, I read Mathews, Landau, and Goldstein, and I saw that mechanics was really what I wanted to do.
+Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr.
+Golubovic would like this one." I wish I had taken more advantage of the opportunity to speak to you outside of classes.
+I'll feel deeply successful if I can one day become as inspiring an educator as you are.
+I hope we can chat in Morgantown sometime!
+
+To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times.
+I'll never forget Scott's quaternions and borderline quantum consciousness beliefs, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
+I have yet to experience teamwork and shared exploration like we had in the undergrad lounge.
+I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, and that they have some Andrew Rice imitator to flip some cars for em (mike jones).
+May that live on forever.
+Because of you all, WVU was magic for me.
+
+Now we come to Marwan Hassan.
+About five years ago now I wandered into Marwan's office saying I liked rivers and wanted to describe sediment transport with statistical physics.
+I realize now when I said ``statistical physics" Marwan's eyes lit up with thoughts of Einstein, Nakagawa, Yano, Hubbell, Crickmore, and so many other names I now know well.
+I was lucky to use the right trigger words.
+Marwan, thanks for taking the risk and taking me on from outside the field, for all of the guidance, for loaning me so many papers and books, giving me freedom to explore ideas, and most of all for showing me that it's ok to be human in science.
+At this point I also want to thank Yusuf for trying your patience to its limits over a long enough period so that things didn't seem so bad once I came along for this short time together.
+
+My friends in geography have also been great.
+Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, David, Jiamei, Xingyu, Niannian, and Emma, plus the outsiders, Will, Dave, Rose, and Anya.
+In particular, thanks to Shawn for all of the shared speculations, and Matteo and Conor -- friendships forged through suffering!
+We'll have to keep in touch and transform our shared venting sessions into well-rounded friendships and productive research collaborations.
+
+Thank you Rui Ferreira and Brett Eaton!
+Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics.
+Your kindness, deep fluid and granular physics knowledge, and wide-ranging interests beyond physical science are inspiring.
+Thanks for the thoughtful comments and advice on my PhD.
+I'm not forgetting what I owe you!
+But I am impressed at your concise way of speaking and unbounded kindness.
+Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand modelling in light of the real world and its complexity.
+Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I can read something besides articles!
+Maybe they'll bring me closer to your way of thinking.
+
+I'd be amiss not to mention David Furbish.
+Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation.
+At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes".
+Now I have a thesis on it.
+Later on, David recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social repercussions of disinformation, and a whole range of questions in natural philosophy, pondering implications of stochastic landscape evolution, comparing probability to mass, momentum, and energy, and evaluating fringe ideas in cosmology.
+These are my people!
+David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me on board the spacecraft.
+
+Among UBC Geographers, thanks especially to Nina Hewitt!
+It was always a pleasure and sometimes a sigh of relief to TA your courses, which are so uniquely organized and student-focused.
+UBC Geography is a much greater department with you in it.
+I'm so thankful to have been a part of your process to transform undergrads into capable scientists.
+I'll be happy if I can teach like you one day!
+
+Finally, Mary!
+My PhD has not been great in terms of work-life balance.
+Thanks for sticking it out, putting up with my late bedtimes, my early wakeups, and all of those social events I skipped.
+It's time!
+I'm done.
+The pandemic also seems to be winding down.
+We can finally get to Aus!
+I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella.
+After that, we're a step closer to settling once and for all whether our cat will be white or orange.
+I'm leaning toward white again.
+
+Completing a PhD looks like a personal achievement, but it's really not.
+It's an expression of 30 years of influence and guidance from other people at every stage, my parents, sisters, grandmother, and a huge network of teachers and mentors.
+I'm kinda tired because it's the end, so it's natural to drift into daydreams of Watoga and Williams River with my family, reading encyclopedias in Gugs' living room, lining up just left of center with my dad and bracing on the paddle for the hit, calm discussions with Johnathan Ramey about my education, the glowing fireflies under my great grandpa's tree, ``cross-up, cross-up, squeeze" in Mindy's class, or laughing about a broken watermelon at Kumbrabow with my sisters.
+It's easy too to shift years later, flipping on the light in the physics lounge at 3am and surprising Megat who was asleep on the couch, or finishing a problem set with Payne and snapping the book shut, discussing Tsujimoto with Marwan, or roaming through Kitsilano with Mary at the edge of dark.
+The PhD is all of it!
+Thank you all for every part of it.
+University is finished.

### 2021-07-02 appendix.tex
getting the appendices structured. made lots of progress on figures.

@@
 
-\chapter{Mathematical Compendia}
+\chapter{Appendices}
+\section{Calculations of chapter 1:}
+\section{Calculations of chapter 2: From particle dynamics to the sediment flux}
+\section{Calculations of chapter 3:}
+
+\section{Calculations of Chapter 4: }
+
+\section{Strategies for Monte Carlo simulation}
+
+\subsection{Derivation of Master Equation}
+To derive the master equation from \ref{eq:langevin}, we temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 d} h$ with $h$ distributed as $f(h)$.
+We will later take a GWN limit on this noise.
+With this assumption, integrating \ref{eq:langevin} over a small time interval $\delta t$ considering the Ito interpretation for the collision term provides
+\be     
+u(t+\delta t) =
+\begin{cases}
+	u(t) + \gamma \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
+	u(t) + \sqrt{2 d} h &  \text{with probability } r \delta t \\
+	\ve u(t) &  \text{with probability } \nu \delta t
+\end{cases}
+\ee
+
+Considering the probability $P(u,t+\delta)$ as sum over possible paths from $P(u,t)$ develops 
+\begin{align} P(u,t+\delta t) =
+	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\gamma \delta t) P(w,t)\\ 
+	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 d}h) P(w,t) \\ 
+	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
+\end{align}
+Evaluating all integrals over $\delta$-functions provides 
+\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
+Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanKampen1983}.
+This process finally obtains the master equation (\ref{eq:master}).
+\subsection{Derivation of Steady-state solution}
+
+Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
+\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
+where
+\be q(z) = d z^2 - i \gamma z + 1 . \label{eq:q} \ee
+Recursing $N+1$ times provides
+\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
+The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
+\be \lambda_\pm = \frac{\gamma}{2d}\Big[1 \pm \sqrt{ 1 + 4d/\gamma^2} \Big].\ee
+Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
+\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
+where the coefficients $R_l^\pm$ are the residues of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
+\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
+The Fourier transform (\ref{eq:recursion}) has a beautiful feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
+Taking this limit and evaluating the residues provides 
+\begin{multline} \tilde{P}(s) = \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
+	\\+ \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
+Finally, inverting the Fourier transforms term by term with contour integration and incorporating (\ref{eq:q}) provides the steady-state solution (\ref{eq:steadystate}).
+
+\subsection{Calculation of the moments}
+Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
+\be 0 = Dk (k-1)\langle u^{k-2} \rangle + \Gamma k \langle u^{k-1} \rangle + \nu (\ve^k-1)\langle u^k \rangle. \ee
+$k=1$ provides the mean
+\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}\ee
+while $k=2$ provides the second moment
+\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
+leading to the velocity variance
+\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
+\subsection{Weak and strong collision limits}
+Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
+The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
+Following \cite{Hall1970}, this suggests normalizing the distribution $P(u)$ using 
+\be z = \frac{u - \bar{u}}{\sigma_u} \ee
+and 
+\be Q(z) = \sigma_u P(u)\ee
+to seek a differential equation for $Q(z)$ with manageable behavior as $\ve \rightarrow 1$.
+Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" Master equation
+\be (1-\ve^2) \frac{d}{2d + \gamma} Q''(z) - \frac{\gamma\sqrt{1-\ve^2}}{\sqrt{2d+\gamma^2}}Q'(z) - Q(z) + \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\gamma\sqrt{1-\ve^2}}{\ve\sqrt{2d+\gamma^2}} \Big]\Big) = 0 \ee
+This equation remains exact and is only a change of variables from (\ref{eq:master}).
+Now we approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
+\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
+which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$.
+This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-07-02 ch1intro.tex
getting the appendices structured. made lots of progress on figures.

@@
 Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
-\be  \big(\ell \px \pt + k_E \ell \px + pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
+\be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
 This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1986a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
@@
 
-The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt' \rangle$ becomes \citep{Balakrishnan1993}
+The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt') \rangle$ becomes \citep{Balakrishnan1993}
 \be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\ee
@@
 The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1984,VanKampen2007}
-\be \pt P(u,t) = \Delta\partial_u\Big[ \text{sgn}(u) P \Big] + D \partial_u^2 P,\ee
+\be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
 implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by equation \ref{eq:fanlangevin} is
@@
 This time, the Fokker-Planck equation is
-\be \pt P(u,t) = -\partial_u\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2} \partial_u^2 P,\ee 
+\be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
 and the steady state solution becomes
@@
 which is the Gaussian velocity distribution from the other subset of the experiments.
+
+These two models of bedload velocity distributions are summarized in figure \ref{fig:fanAncey}.
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
+	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag.
+Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
+	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d).
+The latter has relatively narrow and symmetric fluctuations, with the mode well above $u=0$.}
+	\label{fig:fanAncey}
+\end{figure}
 
@@
  \begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.png}
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
 	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from

### 2021-07-02 ch2flipFlop.tex
getting the appendices structured. made lots of progress on figures.

@@
 
-\chapter{Calculation of the sediment flux}
+\chapter{From particle dynamics to the sediment flux}
 \label{ch:flux}
@@
 
-\section{Model development \label{sec:mod}}
+\section{Description of motion-rest alternation with velocity fluctuations \label{sec:mod}}
 We consider an infinite one-dimensional domain populated with sediment particles on the surface of a sedimentary bed.
@@
 
-\subsection{Dynamical equation for bed load sediment transport}
+\subsection{Dynamical equation for grain-scale sediment transport}
 From these assumptions, our first target is to write an equation of motion for the individual sediment particle encompassing two features.
@@
 
-\subsection{Derivation of the master equation for $P(x,t)$}
+\subsection{Derivation of the master equation for the sediment position distribution $P(x,t)$}
 The solution of equation \ref{eq:langevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives the trajectory of one particle.
@@
 
-\subsection{Calculation of the flux}
+\subsection{Calculation of the sediment flux}
 
@@
 \ref{eq:fluxy} the mean flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $Q_0 = E l$ as the observation time becomes much larger than $1/(k_D \Pe)$.
-\begin{figure}
-	\centerline{\includegraphics{figures/ch2/figure3_slopeKey.pdf}}
+
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{figures/ch2/figure3_slopeKey.pdf}
 	\caption{}
-	\label{fig:fig3}
+	\label{fig:fluxconvergence}
 \end{figure}
@@
 
-\subsection{Connection to earlier work}
+\subsection{Reduction to earlier work}
 

### 2021-07-02 ch3bedEle.tex
got another intro figure constructed.
getting the appendices structured. made lots of progress on figures.

@@
 
-\chapter{Analysis of the bed elevation}
-\label{ch:downDiff}
+\chapter{Analysis of bed elevation change and sediment transport fluctuations}
+\label{ch:bedEle}
 
@@
 \begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition.pdf}
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition-combo.pdf}
 	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains.
@@
 
-\subsection{Numerical simulations}
+\subsection{Numerical study of the joint model}
 \label{sec:numerical}
@@
 
-\subsection{Approximate solutions}
+\subsection{Approximate solutions of the joint model}
 \label{sec:analytical}
@@
 
-\subsection{Statistical moments}
+\subsection{Statistical moments of bed elevation and the particle activity}
 \label{sec:mom}
@@
 
-\subsection{Context}
+\subsection{Context of the research}
 Einstein developed the first stochastic models of bedload tracer diffusion \citep{Einstein1937} and the bedload flux \citep{Einstein1950}, and his ideas can be viewed as the nexus of an entire paradigm of research that extends into the present day \citep[e.g.,][]{Hubbell1964, Nakagawa1976,Hassan1991,Ancey2008, Wu2019}.
@@
 
-\subsection{Contributions}
+\subsection{New contributions}
 
@@
 
-\subsection{Next steps}
+\subsection{Next steps for research}
 
@@
 
-\section{Conclusion}
+\section{Summary and conclusion}
 \label{sec:conclusion}

### 2021-07-02 ch5langevin.tex
getting the appendices structured. made lots of progress on figures.

@@
 
-\chapter{Collisional model of sediment velocity distributions}
+\chapter{Collisional Langevin model of the bedload sediment velocity distributions}
 \label{ch:langevin}
@@
 Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
-Due to these fluctuations, measured transport rates often show extremely slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude.
-These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions.
-In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations, quantifying the dependence of measurements on the observation scale, and linking bulk transport characteristics to the ``microscopic" dynamics of individual grains.
-Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Boltzman, Lajeunesse2020} and bedform initiation \citep{Ancey2014,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
-One subset of these methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume.
-These approaches highlight the instantaneous velocity distribution of sediment particles as an object of crucial importance.
-Unfortunately, current understanding of bedload velocity distributions remains limited.
-We have as of yet no consensus on the shape of the bedload velocity distribution (is it Exponential, Gaussian, Gamma, or some other shape?), and no mechanistic models have yet been developed to describe the full range of experimental observations.
-Here, we develop a stochastic model of particle velocities to addresses these issues.
+Due to these fluctuations, measured transport rates often show slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude \citep{Ancey2020}.
+These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{}.
+In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations \citep{Ancey2008}, quantifying the dependence of measurements on the observation scale \citep{Turowski2010}, and linking bulk transport characteristics to the ``microstructural" dynamics of individual grains \citep{Ancey2014}.
+Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
+One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume \citep{Ancey2008,Furbish2012}.
+These approaches rely on the instantaneous velocity distribution of sediment particles \citep{Lajeunesse2010}, but our understanding of these distributions unfortunately remains limited.
+We have as of yet no consensus on the shape of the bedload velocity distribution (is it Exponential, Gaussian, Gamma, or some other shape?), and although models have described some extreme endmember distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
+In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.
 
@@
 
-\section{Model}
+\section{Mechanistic description of particle velocities}
 \label{sec:model}
@@
 
-Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Rey_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
-These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Rey_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
-At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Rey_p$ formulas \citep{Schmeeckle2007,Mei2006}.
+Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Re_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
+These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
+At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Re_p$ formulas \citep{Schmeeckle2007,Mei2006}.
 A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
-\rho_f d^2 C_D(\Rey_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
-In the present model we set $C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1995,Armenio2001,Dalche2015}.
+\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
+In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1995,Armenio2001,Dalche2015}.
 
@@
 
-\subsection{Langevin equation}
+\subsection{Langevin equation for collisional transport}
 
@@
 
-\subsection{Derivation of the velocity distribution}
+\subsection{Derivation of the bedload velocity distribution}
 \label{sec:solution}
@@
 
-\subsection{Exponential and Gaussian regimes}
+\subsection{Exponential and Gaussian regimes: limits to earlier work}
 \label{sec:modelcomparison}
@@
 
-\section{Rivers, hillslopes, and dunes}
+\section{Analogies to hillslope and wind-driven transport}
 
@@
 \label{sec:conclusion}
-
-\appendix
-
-\section{Derivation of Master Equation}
-To derive the master equation from \ref{eq:langevin}, we temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 d} h$ with $h$ distributed as $f(h)$.
-We will later take a GWN limit on this noise.
-With this assumption, integrating \ref{eq:langevin} over a small time interval $\delta t$ considering the Ito interpretation for the collision term provides
-\be     
-u(t+\delta t) =
-\begin{cases}
-	u(t) + \gamma \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
-	u(t) + \sqrt{2 d} h &  \text{with probability } r \delta t \\
-	\ve u(t) &  \text{with probability } \nu \delta t
-\end{cases}
-\ee
-
-Considering the probability $P(u,t+\delta)$ as sum over possible paths from $P(u,t)$ develops 
-\begin{align} P(u,t+\delta t) =
-	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\gamma \delta t) P(w,t)\\ 
-	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 d}h) P(w,t) \\ 
-	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
-\end{align}
-Evaluating all integrals over $\delta$-functions provides 
-\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
-Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanKampen1983}.
-This process finally obtains the master equation (\ref{eq:master}).
-\section{Derivation of Steady-state solution}
-
-Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
-\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
-where
-\be q(z) = d z^2 - i \gamma z + 1 . \label{eq:q} \ee
-Recursing $N+1$ times provides
-\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
-The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
-\be \lambda_\pm = \frac{\gamma}{2d}\Big[1 \pm \sqrt{ 1 + 4d/\gamma^2} \Big].\ee
-Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
-\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
-where the coefficients $R_l^\pm$ are the residues of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
-\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
-The Fourier transform (\ref{eq:recursion}) has a beautiful feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
-Taking this limit and evaluating the residues provides 
-\begin{multline} \tilde{P}(s) = \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
-	\\+ \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
-Finally, inverting the Fourier transforms term by term with contour integration and incorporating (\ref{eq:q}) provides the steady-state solution (\ref{eq:steadystate}).
-
-\section{Calculation of the moments}
-Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
-\be 0 = Dk (k-1)\langle u^{k-2} \rangle + \Gamma k \langle u^{k-1} \rangle + \nu (\ve^k-1)\langle u^k \rangle. \ee
-$k=1$ provides the mean
-\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}\ee
-while $k=2$ provides the second moment
-\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
-leading to the velocity variance
-\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
-\section{Weak and strong collision limits}
-Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
-The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
-Following \cite{Hall1970}, this suggests normalizing the distribution $P(u)$ using 
-\be z = \frac{u - \bar{u}}{\sigma_u} \ee
-and 
-\be Q(z) = \sigma_u P(u)\ee
-to seek a differential equation for $Q(z)$ with manageable behavior as $\ve \rightarrow 1$.
-Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" Master equation
-\be (1-\ve^2) \frac{d}{2d + \gamma} Q''(z) - \frac{\gamma\sqrt{1-\ve^2}}{\sqrt{2d+\gamma^2}}Q'(z) - Q(z) + \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\gamma\sqrt{1-\ve^2}}{\ve\sqrt{2d+\gamma^2}} \Big]\Big) = 0 \ee
-This equation remains exact and is only a change of variables from (\ref{eq:master}).
-Now we approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
-\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
-which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$.
-This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-07-02 app_ch2.tex
got symbols working in the toc.

%%!TEX root = diss.tex

\chapter{Ch. 2 calculations}

\section{Derivation of the master equation}
\section{Solution for the position probability distribution}
\section{Moments of position}
\section{Calculation of the flux rate constant}

### 2021-07-02 app_ch3.tex
got symbols working in the toc.

--- 2021-06-24 formerPapers/jgr/supp.tex
@@
-\documentclass[11pt]{article}
-% General document formatting
-\usepackage[margin=0.75in]{geometry}
-\usepackage[parfill]{parskip}
-\usepackage[utf8]{inputenc}
-\usepackage{subfig}         % side-by-side figures 
-% Related to math
-\usepackage{amsmath,amssymb,amsfonts,amsthm}
-\usepackage{graphicx}
-\usepackage{natbib}
-\usepackage{titling}
-\usepackage{hyperref}
-\usepackage{wrapfig}
-\usepackage{algpseudocode}
-\usepackage{mathtools} % for colon equals
+%%!TEX root = diss.tex
 
-\usepackage{booktabs} % for wrapping tabulars in accord with
-\bibliographystyle{agu}
-\setlength{\droptitle}{-5em}   % This is your set screw
-
-%\usepackage[math]{kurier}
-\newcommand\be{\begin{equation}} % shortcut to start eq envs 
-\newcommand\ee{\end{equation}}   % shortcut to end eq envs
-\newcommand\ol{\overline}        % shortcut to draw overline 
-\newcommand\bra{\langle}
-\newcommand\ket{\rangle}
-\newcommand\El{\mathcal{L}}
-\newcommand\tg{\tilde{g}}
-\newcommand\tG{\tilde{G}}
-\date{} % make the date field blank
-
-% for pseudo code -- example from https://tex.stackexchange.com/questions/163768/write-pseudo-code-in-latex
-\makeatletter
-\def\BState{\State\hskip-\ALG@thistlm}
-\makeatother
-
-\begin{document}
-\title{Supplementary Material for ``Joint stochastic bedload transport and bed elevation model: variance regulation and power-law rests" by J. Kevin Pierce and Marwan A. Hassan}
-\maketitle
+\chapter{Ch. 3 calculations}
 
@@
 corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation (12) of the manuscript that is plotted with the numerical simulations in manuscript figure (4a).
-
-\bibliography{biblio}
-\end{document}

### 2021-07-02 app_ch4.tex
got symbols working in the toc.

%%!TEX root = diss.tex

\chapter{Ch. 4 calculations}

\section{Calculation of the distribution function}
\label{sec:appendixA}

Owing to the convolution structure  of manuscript equations (1-3), their solution is a formidable problem.
Luckily, we have the device of Laplace transforms.
These project integro-differential equations into an alternate space in which convolutions are unraveled \citep[e.g.,][]{Arfken1985}.
The double Laplace transform of a joint probability distribution $p(x,t)$ is defined by 
\be \tilde{p}(\eta,s) = \int_0^\infty dx e^{-\eta x}\int_0^\infty dt e^{-st} p(x,t). \label{eq:doubletransform}\ee
The Laplace-transformed moments of $x$ are linked to derivatives of the double transformed distribution (\ref{eq:doubletransform}) \citep[cf.,][]{Berezhkovskii2002}.
Equation (\ref{eq:doubletransform}) implies
\be \bra \tilde{x}(s)^k \ket = (-)^k\partial_\eta^k \tilde{p}(\eta,s)\Big|_{\eta=0}.\label{eq:momenttrick}\ee
The operator $\bra \circ \ket$ denotes the ensemble average \citep[e.g.,][]{Kittel1958}.
This means we can compute the variance of position as $\sigma_x^2(t) = \bra x^2 \ket - \bra x \ket^2 = \El^{-1} \{\bra\tilde{x}^2 \ket;t\} - \El^{-1} \{\bra\tilde{x} \ket;t\}^2$, where $\El^{-1}$ denotes the inverse Laplace transform \citep[e.g.,][]{Arfken1985}. This is a powerful tool, since we can use it to derive the positional variance without integrating the distribution in equation (7) of the manuscript.

Double transforming manuscript equations (1-3) using the definition (\ref{eq:doubletransform}) gives

\begin{alignat}{2}
	&\tom_{1T}(\eta,s) &&= \theta_1 \tg_1(\eta,s) + \tom_2(\eta,s)\tg_1(\eta,s)-\tom_{1F}(\eta,s),\\
	&\tom_{1F}(\eta,s) &&= \theta_1\tg_1(\eta,s+\kappa) + \tom_2(\eta,s)\tg_1(\eta,s+\kappa),\\
	&\tom_2(\eta,s) &&= \theta_2 \tg_2(\eta,s) + \tom_{1F}(\eta,s)\tg_2(\eta,s).
\end{alignat}
This algebraic system solves for 
\begin{alignat}{2}
	&\tom_{1T}(\eta,s) &&= \frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\big\{\tg_1(\eta,s)-\tg_1(\eta,s+\kappa) \big\}, \label{eq:A} \\
	&\tom_{1F}(\eta,s) &&= \frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\tg_1(\eta,s+\kappa),\\
	&\tom_{2}(\eta,s) &&= \frac{\theta_2 + \theta_1 \tg_1(\eta,s+\kappa)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\tg_2(\eta,s). 
\end{alignat}
Double transforming manuscript equations (4-6) gives
\begin{align}
	\tp_0(\eta,s) &= \frac{1}{s}\tom_{1T}(\eta,s),\\
	\tp_1(\eta,s) &= \theta_1 \tG_1(\eta,s) + \tom_2(\eta,s) \tG_1(\eta,s),\\
	\tp_2(\eta,s) &= \theta_2 \tG_2(\eta,s) + \tom_{1F}(\eta,s)\tG_2(\eta,s).\label{eq:Z}
\end{align}
The total probability is $p(x,t) = p_0(x,t) + p_1(x,t) + p_2(x,t)$. Using equations (\ref{eq:A}-\ref{eq:Z}) this becomes, in the double Laplace representation, 
\begin{multline}
	\tp(\eta,s) = \frac{1}{s}\frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\big\{\tg_1(\eta,s)-\tg_1(\eta,s+\kappa) \big\} \\
	+\frac{\theta_1\big[\tG_1(\eta,s) + \tg_1(\eta,s+\kappa)\tG_2(\eta,s)\big]+ \theta_2\big[\tG_2(\eta,s) + \tg_2(\eta,s)\tG_1(\eta,s)\big]}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}. \\
	\label{eq:lap}
\end{multline}
Plugging the propagators outlined in manuscript equations (8-9) into equation (\ref{eq:lap}) gives 
\be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}.\label{eq:nicedist}\ee
In this equation, $k'=k_1+k_2$, and we have used the normalization requirement of the initial probabilities: $\theta_1 + \theta_2 = 1.$
The double inverse transform of this equation provides the distribution $p(x,t)$.
We invert the transform over $\eta$ first.
Using the results 15.103 (transform of exponential), 15.123 (transform of derivative), and 15.141 (transform of Dirac delta function) from \citet{Arfken1985} provides 
\begin{multline} \tp(x,s) = \theta_1 \frac{s+\kappa}{s(s+\kappa + k_1)}\delta(x) + \frac{1}{v} \Big(\frac{(s+\kappa+k')s+\kappa k_2}{s(s+\kappa+k_1)} \\- \frac{\theta_1(s+\kappa)[s(s+\kappa+k_1)+\kappa k_2]}{s(s+\kappa+k_1)^2}\Big)
	\exp\Big[-\frac{(s+\kappa+k')s+\kappa k_2}{s+\kappa+k_1}\frac{x}{v}\Big].\end{multline}
Inverting the remaining transform over $s$, applying results 15.152 (substitution), 15.164 (translation), and 15.175 (transform of $te^{kt}$) from \citet{Arfken1985}, and defining the shorthand notations $\tau = k_1(t-x/v)$, $\xi = k_2 x/v$, and $\Omega = (\kappa + k_1)/k_1$, gives the simpler form 
\begin{multline}
	p(x,t) = \theta_1\Big[1-\frac{k_1}{\kappa + k_1}\big(1-e^{-(\kappa + k_1)t}\big)\Big]\delta(x) + \frac{1}{v}\exp[\Omega \tau - \xi]\\
	\times \El^{-1}\Big\{\Big( \theta_2 + \frac{\theta_1k_1+\theta_2 k_2}{s}+\frac{\theta_1k_1k_2}{s^2} + \frac{\theta_2\kappa k_2}{s(s-\kappa-k_1)} + \frac{\theta_1\kappa k_1 k_2}{s^2(s-\kappa-k_1)}\Big)\\
	\times\exp\big[\frac{k_1 \xi}{s}\big];\tau/k_1\Big\}.
\end{multline}
Using entries 2.2.2.1, 2.2.2.8, and 1.1.1.13 from \citet{Prudnikov1992a} in conjunction with the definition of the Marcum Q-function $ \mathcal{P}_\mu(x,t)$ \citep[e.g.,][]{Temme1996}, and inserting the Heaviside functions to account for the fact that grains can neither travel backwards nor at speeds exceeding $v$, we finally arrive at manuscript equation (10) for the joint distribution $p(x,t)$.

\section{Calculation of the moments}
\label{sec:appendixB}

We compute the first two moments of position $x$ and ultimately its variance using equation (\ref{eq:momenttrick}). The first two derivatives of the double Laplace transformed distribution (\ref{eq:nicedist}) are
\be \partial_\eta \tp(\eta,s) = -v \frac{1}{s}\frac{[(s+\kappa + k')s + \kappa k_2][\theta_2(s+\kappa) + k_1]}{[\eta v(s+\kappa +k_1) + (s+ \kappa + k')s+\kappa k_2]^2},\ee
\be \partial_\eta^2 \tp(\eta,s) = 2v^2 \frac{1}{s} \frac{(s+\kappa+k_1)[(s+\kappa + k')s+\kappa k_2][\theta_2(s+\kappa) + k_1]}{[\eta v(s+\kappa + k_1) + (s+\kappa + k')s+ \kappa k_2]^3}.\ee
Evaluating these at $\eta=0$ and applying equation (\ref{eq:momenttrick}) provides the Laplace transformed moments
\be  \frac{\bra\tilde{x}(s)\ket} {v} = \frac{1}{s}\frac{\theta_2(s+\kappa)+k_1}{(s+\kappa+k')s+\kappa k_2} = \frac{1}{s} \frac{\theta_2(s+\kappa)+k_1}{(s+a+b)(s+a-b)}\label{eq:lapmean},\ee
\be \frac{\bra \tilde{x}^2(s) \ket}{2v^2} = \frac{1}{s} \frac{(s+\kappa+k_1)(\theta_2(s+\kappa)+k_1)}{[(s+\kappa+k')s+\kappa k_2]^2}=  \frac{1}{s}\frac{(s+\kappa+k_1)(\theta_2(s+\kappa)+k_1)}{(s+a+b)^2(s+a-b)^2}.\label{eq:lapsecondmom}\ee
The parameters $a= (\kappa+k')/2$ and $b^2 = a^2 -\kappa k_2$ were introduced to factorize the denominators.
These equations can be inverted using the properties 15.164 (translation), 15.11.1 (integration), and 15.123 (differentiation) from  \citet{Arfken1985} after expansion in partial fractions.
For the mean, the calculation is
\begin{align}
	\frac{2b}{v}\bra x \ket &= \big[\theta_2 + (k_1+\theta_2 \kappa)\int_0^t dt\big]\El^{-1}\Big\{ \frac{1}{s+a-b}-\frac{1}{s+a+b};t\Big\}\\
	&= \Big[\theta_2 + \frac{k_1+\theta_2\kappa}{b-a}\Big]e^{(b-a)t} - \Big[\theta_2 - \frac{k_1+\theta_2\kappa}{a+b}\Big]e^{-(a+b)t} - \Big[\frac{k_1+\theta_2\kappa}{b-a} + \frac{k_1+\theta_2\kappa}{a+b}\Big].
\end{align}
This equation rearranges to manuscript equation (11).
The second moment (\ref{eq:lapsecondmom}) is 
\begin{multline}
	\frac{2b^2}{v^2}\bra x^2 \ket = \Big[\theta_2(\delta(t) + \partial_t) + (\theta_2(2\kappa + k_1)+k_1) + (\kappa+k_1)(\theta_2\kappa+k_1)\int_0^t dt \Big] \\
	\times \El^{-1}\Big\{ \frac{1}{(s+a-b)^2} + \frac{1}{(s+a+b)^2}-\frac{1}{b(s+a-b)}+\frac{1}{b(s+a+b)};t\Big\}.
\end{multline}
This becomes 
\begin{multline}
	\frac{2b^3}{v^2}\bra x^2 \ket = \Big[\theta_2\partial_t + [\theta_2(2\kappa+k_1)+k_1] + (\kappa+k_1)(\theta_2\kappa+k_1)\int_0^tdt\Big]\\
	\times \Big((bt-1)e^{(b-a)t}+(bt+1)e^{-(a+b)t}\Big)
\end{multline}
which evaluates to manuscript equation (12).
Finally, $\sigma_x^2 = \bra x^2 \ket - \bra x \ket^2$ derives the variance in manuscript equation (13).

\section{Limiting behavior of the moments}
\label{sec:appendixC}

We determine the diffusion exponents $\gamma$ in the local, intermediate, and global ranges using the two limiting cases described in the discussion of the manuscript.
Limit (1) is $\kappa \rightarrow 0$. We take this limit in equations (\ref{eq:lapmean}) and (\ref{eq:lapsecondmom}) with initial condition $\theta_1=1$ to obtain
\begin{align}
	\bra \tilde{x} \ket &= vk_1 \frac{1 }{s^2(s+k')}, \label{eq:li1}\\
	\bra \tilde{x}^2 \ket &= 2v^2k_1 \frac{s+k_1}{s^3(s+k')^2}. \label{eq:li2}
\end{align}
Inverting these equations provides the variance
\be \sigma_x^2 = 2v^2\frac{k_1}{k'^4}\Big(k_1\big[\frac{1}{2} - k'te^{-k't} - \frac{1}{2} e^{-2k't}\big] + k_2\big[-2+k't + (2+k't)e^{-k't}\big]\Big).\label{eq:li}\ee
This result encodes two ranges of diffusion and can also be derived from the governing equations of the \citet{Lisle1998} and \citet{Lajeunesse2018} models.
Expanding for small $t$ provides $\sigma_x^2(t) = v^2k_1t^3/3$ -- local range super-diffusion.
Expanding for large $t$ provides $\sigma_x^2(t) = 2v^2k_1k_2t/k'^3$ -- intermediate range normal diffusion.

We further investigate limit (1) for arbitrary initial conditions.
By applying Tauberian theorems, we assert the $ t \rightarrow 0$ variance is determined by the $s\rightarrow \infty$ limits of (\ref{eq:lapmean}) and (\ref{eq:lapsecondmom}) \citep[e.g.,][]{Weiss1994, Weeks1998}.  
Expanding these equations in powers of $1/s$ and inverting the resulting transforms gives
\begin{align} \bra x \ket &= v \theta_2 t + \frac{1}{2}v(\theta_1k_1-\theta_2k_2)t^2 + O(t^3),\\
	\bra x^2 \ket &= v^2\theta_2 t^2 + \frac{1}{3}v^2(\theta_1k_1-2\theta_2k_2)t^3+ O(t^4).
\end{align}
This equation highlights the effect of initial conditions on the diffusion characteristics of the local range:
\be \sigma_x^2(t) \sim v^2\theta_1\theta_2t^2 + \frac{1}{3}v^2(\theta_1k_1+\theta_2k_2)t^3.\label{eq:init}\ee
We have taken only leading order terms for any option of $\theta_1$ and $\theta_2$.
Equation (\ref{eq:init}) shows local range exponent $\gamma=2$ when initial conditions are mixed (both are non-zero) and $\gamma=3$ when initial conditions are pure (one is zero).

Limit (2) is $1/k_2 \rightarrow 0$ and $v\rightarrow \infty$ while $v/k_2 = l$. Under this limit, equations (\ref{eq:lapmean}) and (\ref{eq:lapsecondmom}) provide
\begin{align}
	\bra \tilde{x} \ket &= k_1l\frac{1}{s(s+\kappa)},\\
	\bra \tilde{x}^2 \ket &= 2l^2k_1 \frac{s+\kappa+k_1}{s(s+\kappa)^2}.
\end{align}
Inverting these equations and introducing the variables $c=lk_1$ (an effective velocity) and $D_d = l^2k_1$ (a diffusivity) provides positional variance
\be \sigma_x^2(t) = \frac{2D_d(1-e^{-\kappa t})}{\kappa} + \frac{(1-e^{-2\kappa t}-2e^{-\kappa t}\kappa t)c^2}{\kappa^2}. \label{eq:wuvar}\ee
This is mathematically identical to the key result of \citet{Wu2019}.
Expanding for small $t$ provides $\sigma_x^2(t) = 2D_d t$ -- intermediate range normal diffusion, while sending $t\rightarrow \infty$ provides $\sigma_x^2 = (2D_d\kappa + c^2)/\kappa^2$ -- a constant variance in the geomorphic range.
The global range is characterized by competition between terms in equation (\ref{eq:wuvar}), and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.,][]{Wu2019}.
Finally, both equations (\ref{eq:li}) and (\ref{eq:wuvar}) reduce to the Einstein result $\sigma_x^2(t) = 2D_d t$ in further simplified limits.

### 2021-07-02 app_ch5.tex
got symbols working in the toc.

--- 2021-07-02 appendix.tex
@@
 
-\chapter{Appendices}
-\section{Calculations of chapter 1:}
-\section{Calculations of chapter 2: From particle dynamics to the sediment flux}
-\section{Calculations of chapter 3:}
+\chapter{Ch. 5 calculations}
 
-\section{Calculations of Chapter 4: }
-
-\section{Strategies for Monte Carlo simulation}
-
-\subsection{Derivation of Master Equation}
+\section{Derivation of Master Equation}
 To derive the master equation from \ref{eq:langevin}, we temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 d} h$ with $h$ distributed as $f(h)$.
@@
 This process finally obtains the master equation (\ref{eq:master}).
-\subsection{Derivation of Steady-state solution}
+\section{Derivation of Steady-state solution}
 
@@
 
-\subsection{Calculation of the moments}
+\section{Calculation of the moments}
 Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
@@
 \be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
-\subsection{Weak and strong collision limits}
+\section{Weak and strong collision limits}
 Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.

### 2021-07-02 ch6conc.tex
got symbols working in the toc.

@@
 \label{ch:conc}
+\section{The overall strategy}
+\subsection{Langevin and master equations}
+\subsection{Idealized noises and their combinations}
 
@@
 
-\subsection{Probability distribution of the sediment flux}
+\subsection{Probability distribution of the sediment flux from micromechanics of particle trajectories}
 
@@
 
-\subsection{Quantification of the control of bed elevation fluctuations on sediment transport fluctuations}
+\subsection{Quantification of the control of bed elevation fluctuations over sediment transport fluctuations}
 
-\subsection{Understanding of how sediment burial affects the downstream spreading of sediment tracer particles}
+\subsection{Predicting how sediment burial affects the downstream spreading of sediment tracer particles}
 
@@
 
-\section{Closure: Complexity vs Realism}
+\section{Conclusion and next steps}

### 2021-07-02 pref.tex
got symbols working in the toc.

--- 2021-06-29 pref.tex
@@
 
-\chapter{Preface: complementary traditions}
+\chapter{Preface}
 
-This thesis deals with the transport of coarse sediment in flowing water from the perspective that sediment transport results from the movements of individual grains.
+This thesis is entirely the original research of Kevin Pierce, including all figures, writing, and calculations.
+The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
+The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \citet{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
+Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to the thesis submission.
+
+This thesis deals with the transport of coarse sediment in flowing water from the perspective it fundamentally results from the movements of individual grains.
 The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.

### 2021-07-03 app_ch4.tex
made some progress on the nightmare citations.

@@
 \be \sigma_x^2 = 2v^2\frac{k_1}{k'^4}\Big(k_1\big[\frac{1}{2} - k'te^{-k't} - \frac{1}{2} e^{-2k't}\big] + k_2\big[-2+k't + (2+k't)e^{-k't}\big]\Big).\label{eq:li}\ee
-This result encodes two ranges of diffusion and can also be derived from the governing equations of the \citet{Lisle1998} and \citet{Lajeunesse2018} models.
+This result encodes two ranges of diffusion and can also be derived from the governing equations of the \citet{Lisle1998} and \citet{Lajeunesse2017} models.
 Expanding for small $t$ provides $\sigma_x^2(t) = v^2k_1t^3/3$ -- local range super-diffusion.

### 2021-07-03 app_ch5.tex
made some progress on the nightmare citations.

@@
 \begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
-Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanKampen1983}.
+Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanDenBroeck1983}.
 This process finally obtains the master equation (\ref{eq:master}).

### 2021-07-03 ch1intro.tex
made some progress on the nightmare citations.

@@
 
-Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes uplifted by tectonics.
+Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes generated by tectonics.
 Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands, grinding boulders to sand in a machine of abrasion.
@@
 
-In this context river geomorphology as a scientific discipline has shifted toward more quantitative methods that enable concrete predictions about the natural world \citep{Church2005}.
+In this context river geomorphology as a scientific discipline has shifted toward more quantitative methods that enable concrete predictions about the natural world \citep{Church2005,Church2010}.
 Sediment transport in river channels is especially amenable to this quantitative approach since it is mechanistically the result of fluid and granular physics.
 Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
-When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}.
+When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947,Turowski2020}.
 In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.
@@
 Bedload transport exerts unique influence over channel morphology and stability \citep{Church2006,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Hassan2008, Eaton2020}.
-A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
+A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement, since this relates to morphological change.
 Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
@@
 Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}.
-The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018}.
-Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values.
+The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018, Masteller2019, Pretzlav2020}.
+Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values \citep{Ancey2020b}.
 
@@
 This departs from the traditional strategy of correlating transport rates to the average flow and sedimentary characteristics \citep{MeyerPeter1948, Parker1990, Wilcock2001}.
-Because the trajectories of individual grains are governed by Newtonian mechanics, this approach brings reliable tools to the problem, but it is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time due to fluid turbulence and the erratic interactions of moving particles with the bed.
+Because the trajectories of individual grains are governed by Newtonian mechanics, this approach brings powerful tools to the problem, but it is nonetheless complicated because the forces driving and resisting sediment motion vary through space and time due to fluid turbulence and the erratic interactions of moving particles with the bed.
 
-To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and the body of earlier literature which inspired my approach.
-To provide context and motivation, the thesis begins by reviewing this literature, rephrasing earlier work as necessary to indicate common themes and show continuity with my own developments which follow in subsequent chapters.
+To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and drawing upon a wide interdisciplinary literature.
+To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes and show continuity with my own developments which follow in subsequent chapters.
 
@@
 At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
-Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Furbish1997}.
-Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the outcomes of these encounters are variable \citep{Sekine1992}.
+Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Coleman1967,Schmeeckle2007}.
+Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Martin2013}.
 
 Entrainment and deposition provide an additional layer of complexity.
-Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and cause their deposition \citep{Charru2004}.
-These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2012,Celik2014}.
-As a result, particles generally alternate through sequences of entrainment and deposition, but determining precisely when these events occur is impractical.
+Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and induce their deposition \citep{Charru2004}.
+These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2010,Celik2014}.
+As a result, particles generally alternate through sequences of entrainment and deposition, but the exact times at which these alternations occur is impossible to characterize.
 
-Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1972}, 
-These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Church1994,Ferguson2002}.
+Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}, 
+These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Hassan1994,Ferguson2002}.
 
@@
 An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
-A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1966, Ashmore2020}.
-In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997,Ferguson2002}.
+A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
+In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
 In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales.
-Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenberger2012}.
-This obscures the mapping between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.
+Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013}.
+This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.
 
@@
 Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
-Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1956}, but with random driving term (equation \ref{eq:einrando}).
+Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein2001}, but with random driving term (equation \ref{eq:einrando}).
 \begin{figure}[!htbp]
@@
 
-The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1984,Moss1989}.
+The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1989,Moss1989}.
 Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
 \be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
-This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1986a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
+This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1992a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
 \be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/\ell}\theta(x) \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big) \ee
@@
 \be \sigma_x^2(t) = 2 k_E \ell^2 t, \ee
-so particles in Einstein's model spread apart as a normal diffusion process $\sigma_x^2 = 2 D_\text{eff} t$ \citep{Sokolev2002}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$
+so particles in Einstein's model spread apart as a normal diffusion process $\sigma_x^2 = 2 D_\text{eff} t$ \citep{Sokolov2012}, with an effective diffusivity $D_\text{eff} = k_E \ell^2.$
 
@@
 Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles become embedded in sedimentary deposits (geomorphic range).
-The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1970,Francis1972,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
+The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
 In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
@@
 which is a non-trivial result.
-At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2014}.
+At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}.
 At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
@@
 Early efforts impart particles with time-averaged fluid forces, typically linked to a logarithmic flow velocity profile \citep{VanRijn1984}, and some include simplified collision forces that modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992}.
-Later on, researchers began to include granular interactions among particles to model collisions using the discrete element method \citep{Cundall1979, Haff1993}.
-The early works utilizing this approach used a two dimensional domain with a highly simplified flow model \citep{Jiang1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+Later on, researchers began to include granular interactions among particles to model collisions using the discrete element method \citep{Cundall1979a, Haff1993}.
+The early works utilizing this approach used a two dimensional domain with a highly simplified flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
 The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow.
-The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, using particles as the boundary condition for the fluid \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowincklel2014,Elghannay2017,Yousefi2020}.
-A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2011, Azema2012, Wachs2019}.
+The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, using particles as the boundary condition for the fluid \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
+A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2012, Azema2012, Wachs2019}.
 These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are desired for further insight into the problem.
@@
 The fluid drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
-The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1984,VanKampen2007}
+The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007}
 \be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
@@
 \be q = E \ell, \ee
-where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit bed area \citep{Wilcock1997,Furbish2012}.
+where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit bed area \citep{Wilcock1997a,Furbish2012}.
 
@@
 by including the random supporting geometry of bed particles into the force balance. More recently, \citet{Tregnaghi2012}
-ammended the theory to include both force magnitude and duration \citep{Diplas2008, Valyrakis2012, Celik2014}.
+ammended the theory to include both force magnitude and duration \citep{Diplas2008, Valyrakis2013, Celik2014}.
 Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.
@@
 Although Einstein worked to relate the single-particle entrainment rate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character and are not clearly linked to any underlying fluid-granular physics.
-The erosion-deposition model originally developed by \citet{Charru2004,Charru2006,Charru2006a} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
+The erosion-deposition model originally developed by \citet{Charru2004,Charru2006} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2017}.
 This can be characterized as a mixture of the Einstein and Bagnold strategies.
@@
 
-Einstein's model of the bulk bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009,Tucker2014}.
+Einstein's model of the bulk bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009,Tucker2010a,Martin2012}.
 \citet{Nakagawa1976} and later \citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
@@
 
-\citet{Furbish2012} generalized the Parker model to include a finite duration of motion.
+\citet{Furbish2012,Furbish2017} generalized the Parker model to include a finite duration of motion.
 They wrote
@@
 
-Exner was probably the first to write the mathematical relationship between sediment transport and topographic change \citep{Exner1925}. He wrote
+\citet{Exner1925} was probably the first to write the mathematical relationship between sediment transport and topographic change. He wrote
 \be (1-\phi)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \label{eq:exner}\ee
@@
 Yet there are many Earth surface processes where the scale of interest is not much larger than the largest grains involved in the system.
-We can wonder, for example, how large boulders in mountain channels control the formation of steps \citep{Church2007,Zimmerman2008,Saletti2020}, or other bed structures having sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
+We can wonder, for example, how large boulders in mountain channels control the formation of steps \citep{Chin1999,Zimmermann2010, Comiti2012, Saletti2020}, or other bed structures having sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
 In these cases, individual grains are comparable to the scales of interest, and the continuum description provided by the Exner equation is not applicable.
 
-\citet{Nakawaga1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation based upon Einstein's concepts of entrainment and deposition rates.
+\citet{Nakagawa1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation based upon Einstein's concepts of entrainment and deposition rates.
 According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition:
@@
 The probability distribution of the bedload flux can be computed by the control volume form, eq.
-\ref{eq:controlflux} given, in addition, the probability distribution of particle velocites, which can be determined experimentally \citep{Lajeunesse2012,Fathel2015,Heyman2016} or calculated using the Langevin models described in section \ref{sec:langevin}.
+\ref{eq:controlflux} given, in addition, the probability distribution of particle velocites, which can be determined experimentally \citep{Lajeunesse2010,Fathel2015,Heyman2016} or calculated using the Langevin models described in section \ref{sec:langevin}.
 Assuming that the particle activity in the control volume and the velocities of moving particles are completely independent, which may be a good assumption for a large enough control volume, the probability distribution $Q(q)$ of the flux is \citep{Ancey2008,Ancey2014,Ancey2020}
@@
 
-The original model by \citet{Einstein1937} describing particle trajectories when the movement velocity is infinite was subsequently improved to include a finite yet constant velocity \citep{Gordon1972,Lisle1998,Lajeunesse2018}.
+The original model by \citet{Einstein1937} describing particle trajectories when the movement velocity is infinite was subsequently improved to include a finite yet constant velocity \citep{Gordon1972,Lisle1998,Lajeunesse2017}.
 Particle velocities fluctuate during movements between entrainment and deposition, so this should also be accounted for in bedload trajectory models.
@@
 
-Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Ancey2008,Heyman2013,Ancey2014}, which violates conservation of mass and the Exner equation \citep{Exner1925,Tsujimoto1976}, and precludes the possibility that sediment can become buried in the subsurface of the bed.
-Although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2012,Hassan2017}, our understanding of how long sediment can remain buried is limited.
+Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Ancey2008,Heyman2013,Ancey2014}, which violates conservation of mass and the Exner equation \citep{Exner1925,Tsujimoto1978}, and precludes the possibility that sediment can become buried in the subsurface of the bed.
+Although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2002,Hassan2017}, our understanding of how long sediment can remain buried is limited.
 
@@
 
-Chapter \ref{ch:downdiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2018} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.
+Chapter \ref{ch:downdiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.
 

### 2021-07-03 ch2flipFlop.tex
made some progress on the nightmare citations.

@@
 Bed load particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered pockets on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
-Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2019,Valyrakis2014}, particles are once again entrained, and the cycle repeats.
+Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, particles are once again entrained, and the cycle repeats.
 Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and rest.
@@
 Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Ancey2008,Ancey2014}, the migration of bedforms and sediment waves \citep{Hamamori1962,Guala2014,Hoey1992,Recking2012}, and a host of other processes \citep{Dhont2018}.
-Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale-dependence \citep{Saletti2014,Dhont2019,Singh2009,Turowski2010,Ancey2020}.
+Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale-dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
 To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020,Turowski2010}.
@@
 The master equation \ref{eq:master} describes the evolution of the probability distribution of position through time.
-Loosely speaking the solution of this equation is expected to be some combination of the \citet{Einstein1937} theory for particle transport with the Gaussian propagator of the advection-diffusion equation \citep{Gillespie1992}.
+Loosely speaking the solution of this equation is expected to be some combination of the \citet{Einstein1937} theory for particle transport with the Gaussian propagator of the advection-diffusion equation \citep{Morse1953a}.
 
@@
 
-This chapter has generalized the earlier descriptions of individual sediment trajectories \citep[e.g.][]{Lisle1998,Lajeunesse2018} to include velocity fluctuations in the motion state.
+This chapter has generalized the earlier descriptions of individual sediment trajectories \citep[e.g.][]{Lisle1998,Lajeunesse2017} to include velocity fluctuations in the motion state.
 Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
@@
 Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
-In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2010,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
+In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
 This exposes a need to generalize a mechanistic theory of the sediment flux I developed here to produce wider fluctuations.
@@
 This feedback generates spatially-correlated waves of moving particles \citep{Ancey2014, Heyman2015} and produces non-exponential interarrival time distributions \citep{Heyman2013} which imply wide-tailed flux distributions when incorporated in renewal theory \citep{Turowski2010,Ancey2020}.
-Collective entrainment has been attributed to particle-particle interactions, such as small granular avalanches and collision-induced entrainment, and to fluid-particle interactions, such as large eddies entraining particles as they sweep downstream \citep{Heyman2014b,Ancey2014,Lee2018}.
+Collective entrainment has been attributed to particle-particle interactions, such as small granular avalanches and collision-induced entrainment, and to fluid-particle interactions, such as large eddies entraining particles as they sweep downstream \citep{Heyman2014a,Ancey2014,Lee2018}.
 
@@
 Particle-particle interactions would be more challenging to include.
-The dynamical equation \ref{eq:langevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1956}.
-The resulting joint distribution of particle positions and velocities ($P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$) might be formulated by analogy to the theory of reaction diffusion systems \citep{Cardy2006}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
+The dynamical equation \ref{eq:langevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+The resulting joint distribution of particle positions and velocities ($P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$) might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
 By analogy to collective entrainment, interacting models which generalize the approach established in this chapter should be capable of producing wide sediment transport fluctuations.
@@
 
-Considering the flux in the Exner equation as a stochastic quantity requires us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2014}.
+Considering the flux in the Exner equation as a stochastic quantity requires us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2016}.
 In contrast to the traditional analysis, a stochastic Exner equation with fluctuating $q(x,t)$ generates an ensemble of possibilities for the channel profile $z(x)$ at each instant.
@@
 In this equation, $\delta[f(x)]$ is a Dirac delta functional, which generalizes the delta function $\delta(x)$ from a single-valued parameter $x$ to a continuous field $f(x)$.
-Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2004,Nelson2004}, but they have not been introduced in geomorphology.
+Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2001,Nelson2004}, but they have not been introduced in geomorphology.
 
 Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
-The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1960}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
+The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1959,Wolman1978}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
 Progress in stochastic approaches to river science \citep[e.g.][]{Furbish2021c,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask again with these classic works to what extent fluctuations shape Earth's surface.
@@
 
-These results generalize the bedload trajectory models of \citet{Einstein1937}, \citet{Lisle1998}, and \citet{Lajeunesse2018} to include fluctuating movement velocities, and they reframe earlier renewal models of the sediment flux probability distribution \citep{Lajeunesse1010,Ancey2020} to explicitly include the transport dynamics of individual sediment grains.
+These results generalize the bedload trajectory models of \citet{Einstein1937}, \citet{Lisle1998}, and \citet{Lajeunesse2017} to include fluctuating movement velocities, and they reframe earlier renewal models of the sediment flux probability distribution \citep{Lajeunesse2010,Ancey2020} to explicitly include the transport dynamics of individual sediment grains.
 They further quantify the dependence of the sediment flux on the observation scale, and provide guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in simplified configurations when bedforms are absent.

### 2021-07-03 ch3bedEle.tex
made some progress on the nightmare citations.

@@
 Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
-Resting times have puzzled researchers because early experiments show exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments show heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016, Bradley2017, Liu2019}.
+Resting times have puzzled researchers because early experiments show exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments show heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
 A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}; this hypothesis permits surface grains to retain exponential resting times.
@@
 The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
-These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015a,Ma2014}.
+These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015,Ma2014}.
 A recent surge in research activity has generated rapid progress and spawned many new inquiries in this subject. 
@@
 All four rates are independent of the past history of the populations and depend only on the current populations $(n,m)$. 
-As a result, the model is Markovian \citep{Cox1965, VanKampen1992}, meaning time intervals between any two subsequent transitions are exponentially distributed \citep{Gillespie2007}.
+As a result, the model is Markovian \citep{Cox1965, VanKampen2007}, meaning time intervals between any two subsequent transitions are exponentially distributed \citep{Gillespie2007}.
 

### 2021-07-03 ch4downDiff.tex
got diffusion.png working in ch 4 .
got another component in there.. ch 2 . Now to add the langevin paper and get all figures working.
getting the appendices structured. made lots of progress on figures.
made some progress on the nightmare citations.

@@
 
-\chapter{Inclusion of sediment burial}
+\chapter{Burial-induced three-range diffusion in sediment transport}
 \label{ch:downDiff}
@@
 However, with the scaling $\sigma_x^2 \propto t^\gamma$ with $\gamma \neq 1$, the diffusion is said to be anomalous \citep{Sokolov2012}, with $\gamma>1$ defining super-diffusion and $\gamma<1$ defining sub-diffusion \citep{Metzler2000}.
-\citet{Einstein1937} developed one of the earliest models of bedload diffusion to describe a series of flume experiments \citep{Ettema2014}.
+\citet{Einstein1937} developed one of the earliest models of bedload diffusion to describe a series of flume experiments \citep{Ettema2004}.
 Interpreting individual bedload trajectories as a sequence of random steps and rests, Einstein originally concluded that a group of bedload tracers undergoes normal diffusion.
@@
 The majority of these models predict two ranges of diffusion (local and intermediate) without predicting a global range.
-Among these, \citet{Nikora2001a} used synthetic turbulence \citep{Kraichnan1970} with a discrete element method for the granular phase \citep{Cundall1979}; \citet{Bialik2012} used synthetic turbulence with a random collision model \citep{Sekine1992}; and \citet{Fan2016} used a Langevin equation with probabilistic rests.
+Among these, \citet{Nikora2001a} used synthetic turbulence \citep{Kraichnan1970} with a discrete element method for the granular phase \citep{Cundall1979a}; \citet{Bialik2012} used synthetic turbulence with a random collision model \citep{Sekine1992}; and \citet{Fan2016} used a Langevin equation with probabilistic rests.
 To our knowledge, only \citet{Bialik2015} have claimed to capture all three ranges from a Newtonian approach.
@@
 Recently, researchers have generalized Einstein's model in a few different ways to describe multiple diffusion ranges.
-\citet{Lisle1998} and \citet{Lajeunesse2018} promoted Einstein's instantaneous steps to motion intervals with random durations and a constant velocity, providing two diffusion ranges -- local and intermediate.
+\citet{Lisle1998} and \citet{Lajeunesse2017} promoted Einstein's instantaneous steps to motion intervals with random durations and a constant velocity, providing two diffusion ranges -- local and intermediate.
 \citet{Wu2019} retained Einstein's instantaneous steps but included the possibility that grains can become permanently buried as they rest on the bed, also providing two diffusion ranges -- intermediate and global.
@@
 \label{sec:model}
-\subsection{Model assumptions}
+\subsection{Assumptions of the burial model}
 \label{sec:assumptions}
@@
 We expect our conclusions will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
-We consider grains in motion to have characteristic velocity $v$ \citep{Lisle1998,Lajeunesse2018}, and we model burial as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried, meaning $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
+We consider grains in motion to have characteristic velocity $v$ \citep{Lisle1998,Lajeunesse2017}, and we model burial as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried, meaning $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
 We specify the initial conditions with probabilities $\theta_1$ and $\theta_2$ to be in rest and motion at $t=0$, and we require $\theta_1+\theta_2=1$ for normalization.
@@
 
-\subsection{Joint probability distribution}
+\subsection{Joint probability distribution of particle position with burial}
 \label{sec:solution}
@@
 
-\begin{figure}[t]
+\begin{figure}[!htbp]
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/pdf-plot.png}
@@
 
-\subsection{Positional variance}
+\subsection{Downstream diffusion}
 
@@
 
-\begin{figure}[h!]	
+\begin{figure}[!htbp]	
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/diffusion.png}
@@
 
-\subsection{Diffusion exponents}
+\subsection{Diffusion exponents and three range scaling}
 
@@
 \citet{Nikora2002} highlighted the need for such a physical description, although they suggested to use a two-state random walk between motion and rest states with heavy-tailed resting times, and they did not discuss sediment burial.
-However, other works have demonstrated that a two-state walk with heavy-tailed rests provides two diffusion ranges -- not three \citep{Weeks1996,Fan2016}, and although heavy-tailed resting times have been documented for surface particles \citep{Liu2019,Fraccarollo2019}, they are more often associated with buried particles \citep{Martin2012,Martin2014,Voepel2013,Olinde2015,Pelosi2016, Pierce2020}, while surface particles retain light-tailed resting times \citep{Einstein1937,Yano1969,Ancey2006,Nakagawa1976}.
+However, other works have demonstrated that a two-state walk with heavy-tailed rests provides two diffusion ranges -- not three \citep{Weeks1996,Fan2016}, and although heavy-tailed resting times have been documented for surface particles \citep{Liu2019,Fraccarollo2019}, they are more often associated with buried particles \citep{Martin2012,Martin2014,Voepel2013,Olinde2015,Pelosi2016, Pierce2020a}, while surface particles retain light-tailed resting times \citep{Einstein1937,Yano1969,Ancey2006,Nakagawa1976}.
 Accordingly, we developed a random walk model of bedload trajectories with light-tailed surface resting times that incorporates sediment burial.
@@
 We anticipate that a generalization of our model to include heavy-tailed timescales between burial and exhumation would develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in figure \ref{fig:var}.
-If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested by equilibrium transport models \citep{Voepel2013, Martin2014, Pierce2020} and laboratory experiments \citep{Martin2014,Martin2012}, we expect a super-diffusive geomorphic range \citep{Weeks1998}.
+If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested by equilibrium transport models \citep{Voepel2013, Martin2014, Pierce2020a} and laboratory experiments \citep{Martin2014,Martin2012}, we expect a super-diffusive geomorphic range \citep{Weeks1998}.
 However, if they decay slower than $T^{-1/2}$, as implicitly suggested by the data of \citet{Olinde2015}, we expect a genuinely sub-diffusive geomorphic range \citep{Weeks1998}, leaving Nikora et al.

### 2021-07-03 ch5langevin.tex
got symbols working in the toc.
made some progress on the nightmare citations.

@@
 These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$.
-Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Sezilles2014} were viscous.
+Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Seizilles2014} were viscous.
 A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
@@
 
-Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Williams2020}.
+Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Williams2021}.
 The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}.
@@
 Since this quantity combines effects of particle shape and collision geometry and should vary from one collision to the next, we consider that the fraction of streamwise momentum dissipated per collision $\varepsilon$ lies on a statistical distribution $\rho(\varepsilon)$.
-Similar ideas are available in the granular physics literature \citep{Serrero2015}.
+Similar ideas are available in the granular physics literature \citep{Serero2015}.
 Further assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, we write the collision force in the downstream direction as
@@
 \rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
-In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1995,Armenio2001,Dalche2015}.
+In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1997,Armenio2001,Dalche2015}.
 
@@
 The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
-Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,vandenBroek1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2019}.
-One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1984,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweiss2011}.
+Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2019}.
+One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
 Here, the Ito interpretation (lower endpoint integration rule) is the physical choice since the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional.
@@
 Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
-This term is analogous to the collision integral of the Boltzmann equation in kinetic theory and granular gases \citep{Duderstadt1971, Brilliantov2004}.
+This term is analogous to the collision integral of the Boltzmann equation in kinetic theory and granular gases \citep{Duderstadt1979, Brilliantov2004}.
 Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) goes to infinity.
@@
 In what follows, we drop all but the first two terms to obtain the leading order contribution of particle-bed collisions to the velocity distribution.
-Higher orders could always be included later by perturbation theory.
+Higher orders could always be included later by perturbation theory \citep{Morse1953}.
 We solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$.
-Scaling the time in Equation \ref{eq:master}, we can see this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions.
+Scaling the time in Equation \ref{eq:master} by $1/\nu$, we can see this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions.
 
@@
 Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, we have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise.
-Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Coleman2019,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
 We modelled particle-bed collision forces as a sequence of instantaneous impulses where the intervals between successive collisions were characterized as exponential random variables.
@@
 Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport, although these might require numerical methods for their solution.
-The theory of granular gases provides some indication of the types of more nuanced collision models which are possible \citep{Brilliantov2004,other..}.
+The theory of granular gases provides some indication of the types of more nuanced collision models which are possible \citep{Brilliantov2004}.
 
@@
 Estimating in this way, transport experiments with exponential velocities have $St \sim 1-10$, those with neither exponential nor Gaussian velocities have $St \sim 10-100$, and those with Gaussian velocities have $St > 100$.
-In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2001,Joseph2001,Yang2006}.
+In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2011,Joseph2001,Yang2006}.
 Although collision geometry and grain shape certainly complicate the narrative, these values of $St$ are consistent with our model conclusion that the shape of the particle velocity depends on the elasticity of collisions.
@@
 
-\section{Implications for landscape evolution}
+\subsection{Implications for landscape evolution}
 
 Many phenomena of landscape evolution depend on the velocities of individual grains.
-The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Skylar2004,Tingian,Turowski2020}, on which they have a sensitive nonlinear dependence.
+The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Sklar2004,Tingian}, on which they have a sensitive nonlinear dependence.
 It is not much of a stretch to imagine that bedrock erosion could be dominated by the largest impact velocities that originate from the extreme tail of the bedload velocity distribution.
@@
 
-\section{Analogies to hillslope and wind-driven transport}
+\subsection{Analogies to hillslope and wind-driven transport}
 

### 2021-07-03 ch1intro.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
-Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2020}.
+Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2021}.
 Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.
@@
 Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
-When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947,Turowski2020}.
+When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947,Gomez1991}.
 In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.
@@
 Bedload transport exerts unique influence over channel morphology and stability \citep{Church2006,Recking2016}, in part because the coarsest grains in a river provide a partially-immobile skeleton upon which sedimentary deposits can develop \citep{Hassan2008, Eaton2020}.
-A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement, since this relates to morphological change.
+A longstanding problem in river science is therefore to determine the bedload flux, or the rate of coarse sediment movement.
 Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
-Given that this problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.
+Given that the problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.
 
@@
 The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
-\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017,Pretzslav2021}.
+\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017,Pretzlav2021}.
 This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize the literature describing trajectories of individual grains.
@@
 Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
-Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein2001}, but with random driving term (equation \ref{eq:einrando}).
+Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1997}, but with random driving term (equation \ref{eq:einrando}).
 \begin{figure}[!htbp]
@@
 The governing equation of the distribution $P(x,t)$ to find the particle at $x$ can be calculated as an ensemble average of $\delta(x-x(t))$ over all possible realizations of the noise \citep{Risken1989,Moss1989}.
-Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
+Different methods exist to compute such averages \citep{Hanggi1978, Hanggi1984a, Balakrishnan1993, VanDenBroeck1983}, but whatever the approach, the governing equation for the distribution comes out as
 \be  \big(\ell \px \pt + k_E \ell \px + \pt \big)P(x,t) = 0. \label{eq:einmaster}\ee
@@
 
-The large shear stress limit $q \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1998}.
+The large shear stress limit $q \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1990}.
 The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of the details have since turned out to be incorrect.
@@
 An early approach to calculate the sediment flux distribution is due to \citet{Hamamori1962}, who evaluated the sediment flux as the result of migrating bedforms.
-This approach has been extended to include bedforms with a range of sizes \citep{Guala2014,Nikora1984}.
+This approach has been extended to include bedforms with a range of sizes \citep{Guala2014,Nikora1997}.
 
@@
 The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random variable.
-They derived the probability distribution of this variable by counting the number of moving particles in a control volume. 
+They derived the probability distribution of the flux by counting the number of moving particles in a control volume. 
 This number changes through time as a result of entrainment and deposition.
@@
 whose terms describe entrainment at rate $\lambda$, collective entrainment at rate $\mu$, and deposition at rate $\sigma$. 
-The final term encodes the possibility that $n$ does not change at a given instant.
+The final term includes the possibility that $n$ stays constant.
 These coupled equations (one for each $n=0,1,\dots$) can be solved by generating functions \citep{Cox1965,Ancey2008}, providing the steady-state distribution
@@
 The probability distribution of the bedload flux can be computed by the control volume form, eq.
-\ref{eq:controlflux} given, in addition, the probability distribution of particle velocites, which can be determined experimentally \citep{Lajeunesse2010,Fathel2015,Heyman2016} or calculated using the Langevin models described in section \ref{sec:langevin}.
+\ref{eq:controlflux} given the probability distribution of particle velocites, which can be determined experimentally \citep{Lajeunesse2010,Fathel2015,Heyman2016} or calculated using the Langevin models described in section \ref{sec:langevin}.
 Assuming that the particle activity in the control volume and the velocities of moving particles are completely independent, which may be a good assumption for a large enough control volume, the probability distribution $Q(q)$ of the flux is \citep{Ancey2008,Ancey2014,Ancey2020}
@@
 \label{sec:renewal}
-The final approach to summarize that of calculating the sediment flux probability distribution from an ``arrival time distribution" representing the distribution of times between subsequent arrivals of particles to a control surface \citep{Turowski2010,Ancey2020}.
+The final approach to summarize that of calculating the sediment flux probability distribution from an ``arrival time distribution" representing the statistics of times between subsequent arrivals of particles to a control surface \citep{Turowski2010,Ancey2020}.
 This distribution changes depending on the observation time $T$ used to calculate the averaged flux.
 This is the phenomenon of scale dependence.
-The statistics machinery of counting arrivals is called renewal theory \citep{Cox1967}.
+The statistics machinery of counting arrivals is called renewal theory \citep{Cox1962}.
 

### 2021-07-03 ch2flipFlop.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 \be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
-Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen1978} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Loginov1978} in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
+Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen2007} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Shapiro1978} in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
 \be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
@@
 Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
-The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1959,Wolman1978}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
-Progress in stochastic approaches to river science \citep[e.g.][]{Furbish2021c,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask again with these classic works to what extent fluctuations shape Earth's surface.
+Early process-based geomorphologists acknowledged variability \citep{Strahler1952,Langbein1964}, although their main efforts were to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent conditions to summarize climatic variability \citep{Wolman1959,Wolman1978}, characteristic grain sizes to overcome sorting processes \citep{Parker1982,Andrews1983}, and averaged flow conditions to avoid flow turbulence \citep{MeyerPeter1948,Bagnold1954}.
+Recent progress in stochastic approaches \citep[e.g.][]{Furbish2021a,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask with more precision to what extent fluctuations shape Earth's surface.
 

### 2021-07-03 ch3bedEle.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
-These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015,Ma2014}.
+These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015}.
 A recent surge in research activity has generated rapid progress and spawned many new inquiries in this subject. 
@@
 We write the master equation for the probability flow using the forward Kolmogorov equation $\partial P(n,m;t)/\partial t = 
-\sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1992, Ancey2008} as 
+\sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1991, Ancey2008} as 
 \begin{multline}
@@
 The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
-In response to this difficulty, we resort to numerical methods and analytical approximations, simulating equation (\ref{eq:master}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1992, Gillespie2007} and solving it approximately with mean field and Fokker-Planck approaches \citep{Haken1983,Gardiner1983}.
+In response to this difficulty, we resort to numerical methods and analytical approximations, simulating equation (\ref{eq:master}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007} and solving it approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
 The simulation algorithm is described in \ref{sec:numerical}, and analytical approximations are described in \ref{sec:analytical}.
@@
 Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
-Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015a} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.
+Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.
 
@@
 Like \citet{Martin2014}, we obtained heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
-This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016}. 
-Our resting time distributions show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2000} which is within range of our results.
+This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016a}. 
+Our resting time distributions show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of our results.
 Divergent mean resting time distributions present a paradox, since they imply all particles should eventually be immobile, violating the equilibrium transport assumption.
@@
 However, we have left many other statistical characteristics of bedload transport for future studies. 
-For example, the dependence of bedload transport \citep{Saletti2015a,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic.
-Statistical quantities can either be monoscaling or multiscaling across the observation scale \citep{Sornette2000}, and we currently lack physical understanding and general conclusions about the scale dependence of particle activity and bed elevation signals.
-The model we have presented shows statistical monoscaling for both quantites \citep[e.g.][]{Saletti2015a}, whereas other experiments indicate statistical multiscaling \citep{Aberle2006,Singh2009,Singh2012}.
+For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic.
+Statistical quantities can either be monoscaling or multiscaling across the observation scale \citep{Sornette2006}, and we currently lack physical understanding and general conclusions about the scale dependence of particle activity and bed elevation signals.
+The model we have presented shows statistical monoscaling for both quantites \citep[e.g.][]{Saletti2015}, whereas other experiments indicate statistical multiscaling \citep{Aberle2006,Singh2009,Singh2012}.
 We consider this topic to go beyond the scope of the present work, and we have focused on statistical characteristics at the highest temporal resolutions, with no averaging over the observation scale.

### 2021-07-03 ch4downDiff.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 We assumed sediment burial was a permanent condition which developed a non-diffusive geomorphic range.
-In actuality, burial is a temporary condition, because bed scour can exhume buried sediment back into transport \citep{Wu2019a}, probably after heavy-tailed intervals \citep{Voepel2013,Martin2014,Pierce2020}.
+In actuality, burial is a temporary condition, because bed scour can exhume buried sediment back into transport \citep{Wu2019a}, probably after heavy-tailed intervals \citep{Voepel2013,Martin2014}.
 We anticipate that a generalization of our model to include heavy-tailed timescales between burial and exhumation would develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in figure \ref{fig:var}.

### 2021-07-03 ch5langevin.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
-At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Re_p$ formulas \citep{Schmeeckle2007,Mei2006}.
+At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
 A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
 \rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
-In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1997,Armenio2001,Dalche2015}.
+In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1997,Armenio2001}.
 
@@
 \be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
-which is now a non-local ordinary differential equation.
-Such equations have seen some attention in the mathematics literature, where they are called pantograph equations \citep{Hall1970,Zaidi2015,Bhalekar2017} for their relationship to a current collection device used on electric trains \citep{Ockendon1971}.
+which is now a non-local ordinary differential equation. have seen some attention in the mathematics literature, where they are called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
 In the appendix we solve equation \ref{eq:governer} using Laplace transforms, providing
@@
 \be K_\pm = d(\lambda_+-\lambda_-)\prod_{l=1}^\infty (-d\lambda_\pm^2 \ve^{2l} + \gamma \lambda_\pm \ve^{l} + 1). \ee
-Although this velocity distribution appears quite complicated, one can verify that this is a normalized probability distribution which has very simple limiting behaviors as the most common dissipation coefficient $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
+Although this velocity distribution appears quite complicated, one can verify that this is a normalized probability distribution which has very simple limiting behaviors as the mode dissipation constant $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
 
@@
 Many phenomena of landscape evolution depend on the velocities of individual grains.
-The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Sklar2004,Tingian}, on which they have a sensitive nonlinear dependence.
+The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Sklar2004,Li2020,Turowski2021}, on which they have a sensitive dependence.
 It is not much of a stretch to imagine that bedrock erosion could be dominated by the largest impact velocities that originate from the extreme tail of the bedload velocity distribution.

### 2021-07-03 ch6conc.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 \section{The overall strategy}
+
+This thesis has presented a review and four chapters which generalize and extend earlier work to describe sediment transport as a stochastic process.
+In the research, I characterized the movements of grains along a wider range of timescales than before, and I developed new a new formalism using these movements to predict the sediment flux.
+
 \subsection{Langevin and master equations}
+
+Probability
+
 \subsection{Idealized noises and their combinations}

### 2021-07-03 pref.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.

@@
 The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
-The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \citet{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
+The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
 Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to the thesis submission.

### 2021-07-04 abs.tex
some progress..

--- 2021-06-25 abs.tex
@@
 \chapter{Abstract}
+
+Because coarse grains control river stability, a central problem in river science is to predict bedload transport, when coarse sediment moves downstream by bouncing, rolling, and sliding.
+This problem is usually approached with continuum mechanics, but this is questionable considering that coarse sediment grains rarely move in densities approximating a continuum.
+An alternative approach based on describing the trajectories of individual grains with statistical physics has become increasingly popular in recent decades, but many fundamental issues remain in this formulation.
+In particular, the connection between individual particle movements and transport rates remains unclear, descriptions of particle trajectories remain highly simplified, feedbacks between topography and sediment transport remain challenging to analyze, processes such as sediment burial known to control the long-term movements of individual grains have not been evaluated, and even fundamental characteristics such as downstream velocities of grains remain incompletely understood.
+
+This thesis presents four projects completed in the course of my PhD which address these problems.
+First, I demonstrate how to calculate the sediment flux from the dynamics of individual grains and describe the trajectories of grains alternating through rests and motions having fluctuating velocities, linking the sediment flux to the particle dynamics and describing these with more detail.
+Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting the residence times of particles buried within the bed.
+Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
+Finally I describe particles propelled downstream through episodic collisions with the bed surface with a granular gas-like approach, fitting all experimental data on downstream sediment velocities.
+I conclude by summarizing these developments, discussing their implications for the statistical mechanics sediment transport framework, and suggesting how we can use this progress to better understand landscapes.
+
+\endinput
+
+This research provides solutions to longstanding problems within the statistical approach to sediment transport.
+
+and the mathematical tools applied in this thesis should be of interest to researchers of many other problems. Most
+This research has broad implications across river science where there is increasing acknowledgement of the shortcomings of the continuum hypothesis for predicting dynamics in mountain streams.
+This work extends a wide body of work on individual particle motions which originates from the 1930s, and unifies it with more recent approaches to calculate the bulk bedload flux which have not, until now, been related to the movement characteristics of individual grains.
 

### 2021-07-04 ack.tex
some progress..

--- 2021-06-25 ack.tex
@@
 \chapter{Acknowledgments}
-Thanks Leo Golubovic, Mindy Saunders, Calisa and Jim Pierce, Kim and Kelsey Pierce, Johnathan Ramey, Linda Mendez, Tex Wood, Charles Wood, and all of the Appalachian intelligensia who paved my way.
-Tyler cannon.
+
+Attending university long enough to earn a PhD is an unbelievable privilege, and I only had access to it thanks to an incredible amount of guidance and support.
+
+Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma (Gugs) Steele, you've formed all of the best parts of my personality and continually guided me toward this point, and it's impossible to explain how thankful I am for that.
+
+Mom, during my childhood, you worked full time, taught piano lessons, earned Master and Doctoral degrees, played organ every Sunday, consoled us through every difficulty, and engaged us in as many extracurricular activities as we were willing to take on.
+I cannot imagine a better role model of work ethic, open-mindedness, and devotion. I hope I picked a few things up!
+
+Dad, in my childhood I watched you solve an insane diversity of problems without hesitation, from easy ones, like worn out brake pads or stuck chainsaws, to hard ones, like burnt eyes and broken arms.
+On canoeing trips I got lessons in water currents, fish habits, tree species, electrons, engines, wings, hellgrammites, heat transfer, world history, radio waves, and probably everything else I could think of.
+Years of free science lessons!
+You taught me more about science and problem solving than school ever did.
+"Any problem can be fixed with the right tools" is basically the Jim Pierce philosophy.
+It it's incredibly useful in a PhD!
+Thanks for this and so much more.
+
+Kelsey and Kim, my earliest memories are sitting in your bedroom floor struggling through your books, asking for help every couple of paragraphs.
+You made me good at reading and taught me to enjoy it, and this is the basis of everything.
+Kim, you were my first role model, and I think of you all the time.
+Kelsey, I know it goes without saying, but thanks for a lifetime of guidance.
+You're now, as always (all state chorus, math field day, spelling bees, piano skills, grades, .
 .
 .
+), five steps ahead and far more successful.
+On the one hand, I'm deeply proud at your progress through life, with a stable career and now two beautiful kids, but on the other, I'm flatly shocked by it.
+I'll be playing catch up forever!
 
-The people whose benevolence spans generations: both those I know: Joe, Joyce, Ama, Ivory Nichols, and Parry Steele, and those I was unlucky to miss.
-Maragret  Squire Avis and Oddessa Mathis.
-Joe, Joyce, Ama, and Parry Steele.
-My
+Gugs, you're easily the most level-headed, good-natured, and reliable person I've ever had contact with, and to think I was lucky enough to be about half raised by you!
+You've been my second mother, my counsellor, at times my daily chauffeur, and my best role model of grace, consistency, and even-handedness.
+I'll never be able to thank you enough!
+One day I'll finally take on your 5am wakeups and figure out how to copycat your completely measured responses to all of life's challenges.
 
-Shawn Chartrand, Conor McDowell, Matteo Saletti, Will Booker, David Adams, Leo King, Tobias Mueller, Nisreen Ghazi,
+Finally with regard to family, I'd like to thank my great aunt Alice.
+Thanks so much for paying my undergrad education!
+We hardly met before you passed, but we're family, and that was enough.
+It really means a lot.
 
-Adam Payne, Adam Collins, Megat Denney, Andrew Rice, Evan Graber, Scott Ferris, Craig Tenney, and the rest of the WVU physics lounge family.
+Next comes the mentors in education.
+Thanks to my early teachers who put in the extra care to keep me on track, Jonathan Ramey, Damon Spurlock, Linda Mendez, Linda Afzhalirhad, Phyllis Adkins, Brandon Willard, Marty Ojeda, Larry Wolford, and Red Hamlin among many others.
 
-Andy Osadetz, Josh Plankeel,
+Thanks also to my professors at Southern WV Community College (SWVCC) and West Virginia University (WVU) who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
+At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
+Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science.
+Everyone encouraged me to continue my education.
+SWVCC was a personal transformation and a great privilege.
 
-My partner Mary for putting up with the late nights.
+At WVU, I'd like to thank Alan Bristow, Tudor Stanescu, and Leo Golubovic in particular.
+Alan and Tudor, thanks you for providing my first opportunities in research.
+(Sorry I was bad at it!) Leo, you are the key person in my development as a researcher.
+Your ability as an educator and a motivator is incredible.
+Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr.
+Golubovic would like this one!" I wish I had taken more advantage of the opportunity to speak with you outside of classes.
+I'll be successful if I can one day be as inspiring to someone as you were to me.
 
-Mike church and olav slaymaker for the inspiring seminars.
+To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times.
+I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
+I have yet to experience teamwork and mutual exploration like we had in the undergrad lounge.
+Math and physics are fun!
+I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator to flip some cars for em (mike jones), and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level.
+May the lounge live on forever, in Dirac's name, amen.
 
-marwan hassan for being one of the most emotionally intelligent people I've ever met.
+Now we come to Marwan Hassan.
+About five years ago now I wandered into Marwan's office saying I liked rivers and wanted to describe sediment transport with statistical physics.
+I realize now when I said ``statistical physics" Marwan's eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well.
+I was lucky to use the right trigger words.
+Marwan, thanks for taking the risk on me, for all of the guidance, for lending me hundreds of books and papers, for giving me freedom to explore ideas, and most of all for showing me that it's ok (and secretly expected) to be human in science.
+
+My friends in geography have also been great.
+Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, Jiamei, Xingyu, Niannian, and Emma, plus the outsider Eaton-ites, Will, Dave, Rose, and Anya.
+In particular thanks to Shawn for all of the shared speculations, and Matteo and Conor -- friendships forged through suffering!
+We'll have to keep in touch and form productive research collaborations once Matteo comes back from retirement.
+
+Thank you Rui Ferreira and Brett Eaton!
+Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics.
+Your kindness, physics knowledge, and wide-ranging interests beyond physical science are inspiring.
+Thanks for the thoughtful comments and advice on my PhD.
+I'm not forgetting what I owe you!
+I will always be impressed at your concise way of speaking and unbounded kindness.
+Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modelling in light of real world complexity.
+Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done.
+Maybe they'll bring me closer to your way of thinking!
+
+I'd be amiss not to mention David Furbish.
+Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation.
+At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes".
+Now I have a thesis on it.
+David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy from stochastic landscape evolution, to probability as a physical quantity like energy and momentum and fringe ideas in cosmology.
+These are my people!
+David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.
+
+Among UBC Geographers I'd like to thank Nina Hewitt!
+It was always a pleasure and sometimes a sigh of relief to TA your courses, which are so uniquely organized and student-focused.
+UBC Geography is a much greater department with you in it.
+I'm so thankful to have been a part of your process transforming undergrads into scientists.
+I'll be happy if I can teach like you one day!
+
+Finally, Mary!
+My PhD has not been great in terms of work-life balance.
+Thanks for sticking it out, putting up with my late bedtimes, my early wakeups, the rambly science jargon, and all of those social events I skipped.
+It's time!
+I'm done.
+The pandemic also seems to be winding down.
+We can finally get to Aus!
+I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella.
+I hope to take up still lives with Marianne's coaching.
+After that, we're a step closer to settling once and for all what colour cat we're gonna get.
+I'm personally now leaning toward white.
+
+Completing a PhD looks like a personal achievement, but it's really not.
+It's an expression of 30 years of influence and guidance from other people at every stage, my parents, my sisters, my grandmother, and a huge network of teachers and mentors.
+I'm kinda tired because it's the end, so it's natural to drift on to recollections of my favorite times in life, Watoga and Williams River with my family, encyclopedias in Gugs' living room, watching eagles over the South Branch with dad, discussions with Johnathan Ramey about my education, the glowing fireflies under my great grandpa's tree, ``cross-up, cross-up, squeeze" in Mindy's calc class, or laughter about stolen watermelons with my sisters.
+I can skip years later and flip on the light in the physics lounge to find Megat asleep on the couch, visit Gugs on a trip back home, discuss Tsujimoto with Marwan, or walk through Kitsilano with Mary at the edge of dark.
+The PhD is all of it!
+Thank you all for every part of it.

### 2021-07-04 ch1intro.tex
some progress..

@@
 
-Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes generated by tectonics.
-Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands, grinding boulders to sand in a machine of abrasion.
+Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes produced by tectonics.
+Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands.
 Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
 Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2021}.
-Geologic records display unprecedented recent shifts in ancient climatic, denundational, and biotic patterns, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.
+Research demonstrates unprecedented recent shifts in ancient climatic \citep{Sivan2004,Slater2021}, denundational \citep{Hooke2000,Szabo2010}, and biotic patterns \citep{Walther2002,Willis2009}, requiring effective aquatic habitat restoration, contaminant management, and engineering strategies more than ever before.
 
@@
 Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
-When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947,Gomez1991}.
-In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Frey2009,Frey2011}.
+When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}.
+In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Ferreira2015}.
 
@@
 For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
-\be \ee
+\be  \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) \ee
 Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase.
-The variance, however, is different.
+The variance picks up a two range scaling:
 Computing $\sigma_x^2$ provides
@@
 the surface $\mathcal{S}$ per unit width and time, collected from all upstream control volumes.}
-	\label{fig:lislefig}
+	\label{fig:einsteinFluxConcept}
 \end{figure}
@@
 Although Einstein worked to relate the single-particle entrainment rate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character and are not clearly linked to any underlying fluid-granular physics.
-The erosion-deposition model originally developed by \citet{Charru2004,Charru2006} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2017}.
+The erosion-deposition model originally developed by \citet{Charru2004,Charru2006} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2018}.
 This can be characterized as a mixture of the Einstein and Bagnold strategies.
@@
 \subsection{Landscape evolution}
+\label{sec:landscape}
 

### 2021-07-04 ch2flipFlop.tex
some progress..

@@
 \be P(x,0) = \lim_{t\rightarrow 0 } \frac{k_E}{k} \sqrt{\frac{1}{4\pi D t}} \exp\Big[-\frac{(x-Vt)^2}{4Dt}\Big]+ \frac{k_D}{k}\delta(x)\ee
-which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep{Weiss2002}.
+which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep{Weiss2002a}.
 
@@
 
-We can imagine generalizing the description in equation \ref{eq:Langevin} to include such fluid-particle interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
+We can imagine generalizing the description in equation \ref{eq:flipfloplangevin} to include such fluid-particle interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
 Particle-particle interactions would be more challenging to include.
-The dynamical equation \ref{eq:langevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+The dynamical equation \ref{eq:flipfloplangevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
 The resulting joint distribution of particle positions and velocities ($P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$) might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
@@
 
-\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{eq:exner}), writing
+\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{sec:landscape}), writing
 \be (1-\phi) \pt h(x,t) = -\px q(x,t). \ref{eq:stocEx}\ee
@@
 Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
-Early process-based geomorphologists acknowledged variability \citep{Strahler1952,Langbein1964}, although their main efforts were to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent conditions to summarize climatic variability \citep{Wolman1959,Wolman1978}, characteristic grain sizes to overcome sorting processes \citep{Parker1982,Andrews1983}, and averaged flow conditions to avoid flow turbulence \citep{MeyerPeter1948,Bagnold1954}.
-Recent progress in stochastic approaches \citep[e.g.][]{Furbish2021a,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask with more precision to what extent fluctuations shape Earth's surface.
 

### 2021-07-04 ch6conc.tex
some progress..

--- 2021-07-02 ch6conc.tex
@@
 \label{ch:conc}
-\section{The overall strategy}
+\section{Overall Methodology}
+
+This thesis has presented a review and four chapters which generalize and extend earlier work to describe sediment transport as a stochastic process.
+In the research, I have characterized the movements of grains along a wider range of timescales than before, connected a wide literature of earlier research by framing it as different applications of Langevin and Master equations, and extended these earlier works to link
+
 \subsection{Langevin and master equations}
+
+Probability
+
 \subsection{Idealized noises and their combinations}
@@
 
-\section{Models and the real world}
+\section{Conclusion}
 
-\section{Conclusion and next steps}
+This thesis embraced variability in grain scale transport to form some new descriptions of sediment transport processes in river channels.
+The remaining question is how variability in sediment transport impacts landscapes.
+
+The founders of process-based geomorphology generally accepted the role of variability in geomorphology \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes despite it, including competent conditions to summarize climatic fluctuations \citep{Wolman1959,Wolman1978}, characteristic grain sizes to overcome sorting processes \citep{Parker1982,Andrews1983}, and averaged flow conditions to avoid turbulence \citep{MeyerPeter1948,Bagnold1954}.
+
+The stochastic models of sediment transport developed in this thesis suggest a methodology we can use to step beyond averaged descriptions of landscape evolution, propagate noises through the governing equations, and revisit this old question of how variability shapes Earth's surface.

### 2021-07-04 laySum.tex
some progress..

%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

%% https://www.grad.ubc.ca/current-students/dissertation-thesis-preparation/preliminary-pages
%% 
%% LAY SUMMARY Effective May 2017, all theses and dissertations must
%% include a lay summary.  The lay or public summary explains the key
%% goals and contributions of the research/scholarly work in terms that
%% can be understood by the general public. It must not exceed 150
%% words in length.

\chapter{Lay Summary}

Predicting the movement of gravel through river channels is important to manage environmental issues involving rivers, since moving gravel can destabilize river channels.
A majority of models assume gravel flows as if it were a heavy fluid, but this description is crude considering how gravel actually moves.
In this thesis, I instead describe gravel movement as the result of individual grains bouncing and rolling along the streambed.
Using this concept, I calculate the downstream flow rate of gravel, study how this flow rate depends on the shape of the bed, and predict how quickly individual grains move downstream.
This research produces a more realistic description of how gravel moves in river channels which helps us understand how rivers change through time.

\endinput
The lay or public summary explains the key goals and contributions of
the research\slash{}scholarly work in terms that can be understood by the
general public. It must not exceed 150 words in length.

### 2021-07-04 pref.tex
some progress..

--- 2021-07-02 pref.tex
@@
 The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
-The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \citet{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
-Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to the thesis submission.
+The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
+Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to submission of the thesis.
+
+\endinput
+The
 
@@
 This thesis applies this ``simple harmonic oscillator approach" to some problems in Earth science.
+As a result, there are big assumptions: sediment grains all have the same size, and water flows are constant through time. 
 The objective is to build up simple archetypes which can be embellished later, not to build \textbf{the model} of fluvial geomorphology. 

### 2021-07-04 abs.tex
rolling on the conclusion

@@
 Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
-Finally I describe particles propelled downstream through episodic collisions with the bed surface with a granular gas-like approach, fitting all experimental data on downstream sediment velocities.
+Finally I describe particles propelled downstream through episodic collisions with the bed surface submissionusing a granular gas-like theory, fitting all experimental data on downstream sediment velocities.
 I conclude by summarizing these developments, discussing their implications for the statistical mechanics sediment transport framework, and suggesting how we can use this progress to better understand landscapes.

### 2021-07-04 app_ch3.tex
made some progress on the nightmare citations.
rolling on the conclusion

@@
 
-\chapter{Ch. 3 calculations}
+\chapter{Calculations involved in the bed elevation model}
 
@@
 The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
-It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1992} and \citet{Gillespie2007}.
+It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1991} and \citet{Gillespie2007}.
 The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.

### 2021-07-04 ch1intro.tex
rolling on the conclusion

@@
 In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales.
-Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013}.
-This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale.
+Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
+This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
 
-\subsection{Einstein 1937}
+\subsection{Einstein's random walk model}
 Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}.
@@
 For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
-\be  \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) \ee
+\be  P(x,t) =  \ee
 Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase.
@@
 Computing $\sigma_x^2$ provides
-\be \sigma_x^2(t) = ,\ee
+\be \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) ,\ee
 which is a non-trivial result.
@@
 which is the Gaussian velocity distribution from the other subset of the experiments.
-
 These two models of bedload velocity distributions are summarized in figure \ref{fig:fanAncey}.
@@
 \subsection{The definition of the flux}
-
+\begin{figure}[!htbp]
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.pdf}
+	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$.
+Here, particles are shown moving from time $t-dt$ (faint white particles) to time $t$ (grey particles).
+Each particle's velocity component $u_i$ in the downstream direction is indicated.
+At the instant $t$, the surface definition of the flux \ref{eq:surfflux} gives $q = \frac{1}{L}(u_7+u_8).$, while the volume definition \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_6)$,
+	showing the non-equivalence of these two definitions.}
+	\label{fig:fluxdefs}
+\end{figure}
 Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
@@
 The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\mathcal{S}$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
-\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\mathcal{\textbf{S}}. \ee
+\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\mathcal{\textbf{S}}. \label{eq:surfflux} \ee
 This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.
-The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $\mathcal{V}$:
-\be q = \frac{1}{L} \sum_{i\in\mathcal{V}} u_i. \label{eq:controlflux}\ee
+The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $V$:
+\be \Phi = \frac{1}{L} \sum_{i\in V} u_i. \label{eq:controlflux}\ee
 Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
@@
 left to right through an array of adjacent control volumes. The bedload flux is the rate of bedload particles crossing
-the surface $\mathcal{S}$ per unit width and time, collected from all upstream control volumes.}
+the surface $S$ per unit width and time, collected from all upstream control volumes.}
 	\label{fig:einsteinFluxConcept}
@@
 Denoting by $P_n$ the probability that an individual grain undergoes at least $n$ jumps of length $\ell$ in a time interval $T$, meaning it travels at least a distance $n \ell$, and considering that there is a density $\rho$ of particles at rest on the bed, it follows that on average $\rho \ell P_n$ particles will displace a distance $n \ell$ or more from within each control volume during the time interval $T$.
-As a result, since grains crossing $\mathcal{S}$ in a time $T$ could have come from any upstream location, the number of grains crossing $\mathcal{S}$ in $T$ is a sum over all control volumes: $\sum_{n=1}^\infty \rho \ell P_n$.
-Dividing by the time $T$ to get the average rate of grains crossing $\mathcal{S}$ provides the mean flux:
+As a result, since grains crossing $S$ in a time $T$ could have come from any upstream location, the number of grains crossing $S$ in $T$ is a sum over all control volumes: $\sum_{n=1}^\infty \rho \ell P_n$.
+Dividing by the time $T$ to get the average rate of grains crossing $S$ provides the mean flux:
 \be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty P_n. \label{eq:einflux} \ee
@@
 \be \partial_x q(x,t) = E - D. \ee 
-Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described in an Einstein-like framework:
-\be (1-\phi) px z(x,t) = D - E.\ee
+Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described in Einstein's variables:
+\be (1-\phi) \px z(x,t) = D - E.\ee
 In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the result of entrainment at all upstream locations, giving
-\be (1-\phi) px z(x,t) = \int_0^\infty dx' int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
+\be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
 This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
@@
 
-Chapter \ref{ch:burial} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movements characteristics of grains.
+Chapter \ref{ch:ch3} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movements characteristics of grains.
 
@@
 
-Chapter \ref{ch:downdiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.
+Chapter \ref{ch:downDiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.
 

### 2021-07-04 ch5langevin.tex
rolling on the conclusion

@@
 
-\chapter{Collisional Langevin model of the bedload sediment velocity distributions}
+\chapter{Collisional Langevin model of bedload sediment velocity distributions}
 \label{ch:langevin}
@@
 Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
-Due to these fluctuations, measured transport rates often show slow convergence through time, and predicted rates often deviate from measured values by several orders of magnitude \citep{Ancey2020}.
-These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{}.
-In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting fluctuations \citep{Ancey2008}, quantifying the dependence of measurements on the observation scale \citep{Turowski2010}, and linking bulk transport characteristics to the ``microstructural" dynamics of individual grains \citep{Ancey2014}.
+Due to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
+These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
+In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations \citep{Ancey2008} and quantifying the dependence of measurements on the observation scale \citep{Turowski2010}.
 Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
 One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume \citep{Ancey2008,Furbish2012}.
-These approaches rely on the instantaneous velocity distribution of sediment particles \citep{Lajeunesse2010}, but our understanding of these distributions unfortunately remains limited.
-We have as of yet no consensus on the shape of the bedload velocity distribution (is it Exponential, Gaussian, Gamma, or some other shape?), and although models have described some extreme endmember distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
+This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although our understanding of these distributions unfortunately remains limited.
+We have as of yet no consensus on the shape of the bedload velocity distribution, and although models have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
 In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.
@@
 The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
-Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2019}.
+Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2016,Dubkov2019}.
 One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.

### 2021-07-04 ch6conc.tex
rolling on the conclusion

--- 2021-07-02 ch6conc.tex
@@
 \label{ch:conc}
-\section{The overall strategy}
+\section{Overall methodology of the thesis}
+
+The research in this thesis generalizes and extends earlier work to describe sediment transport as a stochastic process.
+In the research, I have characterized the movements of grains along a wider range of timescales than before, framed existing research in terms of Langevin and Master equations, and produced more detailed descriptions of the bedload transport phenomenon, addressing several long-recognized shortcomings of earlier stochastic models.
+In particular I have filled in the connection between sediment fluxes and individual praticle motions, included particle velocity fluctuations in a descriptions of individual particle motions through rest and motion, evaluated the role of bed elevation changes on sediment transport rate fluctuations, and included the process of sediment burial in descriptions of downstream particle movement.
+
 \subsection{Langevin and master equations}
+
+The over-arching strategy in all of these extentions has been to identify control and response variables, represent variable controls by idealized noises (entrainment and deposition events, particle burial events, turbulent forces, or particle-bed collisions), and then to write dynamical equations relating the respones variables to these stochastic control variables.
+This approach phrases sedimentary dynamics in terms of Langevin-like equations, stochastic analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for a the response variable of interest in the problem and the force $F$ is a stochastic idealization of the control variables.
+
+Once such a stochastic dynamical equation was written for a given problem, its solutions for all realizations of the noise which provide a given outcome were counted () to form a probability distribution of the response variable.
+These distributions are generally given as solutions to integro-differential equations (), and the distributions were determined as solutions to these equations with transform calculus ().
+
+In every case, the result is a probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (), particle velocity (), bed elevation (), and sediment transport rate ().
+
 \subsection{Idealized noises and their combinations}
+
+A major challenge in this research is that only a handful noises in statistical physics are comprehensively understood \citep{}, so the available options in stochastic modelling are constrained.
+The noises used in this thesis include white Gaussian, Poisson, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively. 
+Turbulence was described using Gaussian noise (); instantaneous steps (), particle arrivals to a control surface (), and particle-bed collisions () were described with Poisson noise; and alternation between motion and rest was represented with dichotomous noise ().
+Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise () as required.
+This inclusion of multiple noise sources has not yet to my knowledge been pursued in earlier stochastic models of sediment transport.
+
+River science offers no guarantee that the idealized white noises we understand best are sufficient to describe its phenomena.
+White noise spectra are an idealization which is probably never represented in nature \citep{Gardiner1983}.
+In contrast, colored spectra in which fluctuations distribute power unevenly across frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of turbulence \citep{Kolmogorov1941,Nikora2000} and avalanches \citep{Bak1987,Jensen1998}.
+In river science, many phenomena exhibit colored spectra, like the size distribution of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment fluxes \citep{Dhont2018,Chartrand2021}.
+Dynamical equations driven by colored noise are mathematically extremely challenging to solve and require case-by-case analysis \citep{Hanggi1978,Luczka2005,Hanggi2007}.
+This means even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.
 
@@
 
-\subsection{Probability distribution of the sediment flux from micromechanics of particle trajectories}
+\subsection{Probability distribution of the sediment flux from micromechanics of particle transport}
 
@@
 
-\subsection{Predicting how sediment burial affects the downstream spreading of sediment tracer particles}
+\subsection{Predicting how sediment burial affects the downstream transport of sediment particles}
 
-\section{Models and the real world}
+\section{Limitations and future research directions}
 
-\section{Conclusion and next steps}
+\subsection{Morphology-transport interplay}
+
+\subsection{Grain size distributions}
+
+\subsection{Transport stage}
+
+\subsection{Continuum descriptions of sedimentary beds}
+
+\section{Conclusion}
+
+This thesis accepted variability in grain scale bedload transport to form some new descriptions of sediment transport processes in river channels.
+
+The founders of process-based geomorphology always acknowledged variability in geomorphology \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes despite it, such as competent conditions to summarize climatic fluctuations \citep{Wolman1959,Wolman1978}, global characteristics of rivers and basins like channel orders and network topologies to avoid variable local characteristics, and averaged flow conditions to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}.
+
+The stochastic models of sediment transport developed in this thesis demonstrate a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through equations governing landscape evolution, and revisit the old question of how variability shapes Earth's surface.

### 2021-07-04 pref.tex
rolling on the conclusion

@@
 The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
-The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
-Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to submission of the thesis.
+The thesis includes two published works, Chapters \ref{ch:ch3} and \ref{ch:downDiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
+Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to thesis submission.
 

### 2021-07-05 ack.tex
lots of progress. Still lots of work to do though.

@@
 
-Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma (Gugs) Steele, you've formed all of the best parts of my personality and continually guided me toward this point, and it's impossible to explain how thankful I am for that.
+Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele, you've formed all of the best parts of my personality and continually guided me toward this point, and it's impossible to explain how thankful I am for that.
+Regarding family I will also always be thankful for my great aunt Alice who graciously paid for my undergraduate education.
+
+Thanks to my early teachers who put in the extra care to keep me on track, Jonathan Ramey, Damon Spurlock, Linda Mendez, Linda Afzhalirhad, Phyllis Adkins, Brandon Willard, Marty Ojeda, and Red Hamlin among many others.
+
+Thanks also to my professors at Southern WV Community College (SWVCC) and West Virginia University (WVU) who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
+At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
+Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science.
+Everyone encouraged me to continue my education.
+SWVCC was a personal transformation and a great privilege.
+
+At WVU, I'd like to thank Alan Bristow, Tudor Stanescu, and Leo Golubovic in particular.
+Alan and Tudor, thank you for providing my first opportunities in research.
+Leo, you are the key person in my development as a researcher.
+Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr.
+Golubovic would like this one!" I'll be successful if I can one day be as influential an educator as you've been.
+
+To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times.
+I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
+I have yet to experience teamwork and mutual exploration like we had in the undergrad lounge.
+Math and physics are fun!
+I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator for comic relief, and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level.
+May the lounge live on forever!
+
+Now we come to Marwan Hassan.
+About five years ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics.
+I realize now when I said ``statistical physics" your eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well.
+I was lucky to use the right trigger words.
+Marwan, thanks for taking the risk on me, for all of the guidance, for lending me hundreds of books and papers, for giving me freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.
+
+My friends in geography have also been great.
+Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, Jiamei, Xingyu, Niannian, and Emma, plus the outsider Eaton-ites, Will, Dave, Rose, and Anya.
+In particular thanks to Shawn for all of the shared speculations, and Matteo and Conor -- friendships forged through suffering!
+We'll have to keep in touch and form productive research collaborations once Matteo returns from retirement.
+
+Thank you Rui Ferreira and Brett Eaton!
+Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics.
+Your kindness, physics knowledge, and wide-ranging interests beyond physical science are inspiring.
+Thanks for the thoughtful comments and advice on my PhD.
+I'm not forgetting what I owe you!
+I will always be impressed at your concise way of speaking and unbounded kindness.
+Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modelling in light of real world complexity.
+Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done.
+Maybe they'll bring me closer to your way of thinking!
+
+I'd be amiss not to mention David Furbish.
+Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation.
+At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes".
+Now I have a thesis on it.
+David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy from stochastic landscape evolution, to probability as a physical quantity like energy and momentum and fringe ideas in cosmology.
+These are my people!
+David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.
+
+Among UBC Geographers I'd like to thank Nina Hewitt in particular.
+It was always a pleasure and sometimes a sigh of relief to TA your courses, which are so uniquely organized and student-focused.
+UBC Geography is a much greater department with you in it.
+
+Finally, Mary!
+My PhD has not been great in terms of work-life balance.
+Thanks for sticking it out, putting up with my late bedtimes, early wakeups, the rambling science jargon, and all of the social events I skipped.
+It's time!
+I'm done.
+I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella.
+I hope to take up still-lives with Marianne's coaching.
+After that, we're a step closer to settling what colour cat we're gonna get.
+I'm personally now leaning toward white.
+
+Completing a PhD looks like a personal achievement, but it's really not.
+It's an expression of 30 years of influence and guidance from other people at every stage, my parents, my sisters, my grandmother, and a huge network of teachers and mentors.
+Thank you all for every part of it.
+
+5
+\endinput
 
@@
 Next comes the mentors in education.
-Thanks to my early teachers who put in the extra care to keep me on track, Jonathan Ramey, Damon Spurlock, Linda Mendez, Linda Afzhalirhad, Phyllis Adkins, Brandon Willard, Marty Ojeda, Larry Wolford, and Red Hamlin among many others.
 
-Thanks also to my professors at Southern WV Community College (SWVCC) and West Virginia University (WVU) who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
-At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
-Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science.
-Everyone encouraged me to continue my education.
-SWVCC was a personal transformation and a great privilege.
-
-At WVU, I'd like to thank Alan Bristow, Tudor Stanescu, and Leo Golubovic in particular.
-Alan and Tudor, thanks you for providing my first opportunities in research.
-(Sorry I was bad at it!) Leo, you are the key person in my development as a researcher.
-Your ability as an educator and a motivator is incredible.
-Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr.
-Golubovic would like this one!" I wish I had taken more advantage of the opportunity to speak with you outside of classes.
-I'll be successful if I can one day be as inspiring to someone as you were to me.
-
-To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times.
-I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
-I have yet to experience teamwork and mutual exploration like we had in the undergrad lounge.
-Math and physics are fun!
-I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator to flip some cars for em (mike jones), and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level.
-May the lounge live on forever, in Dirac's name, amen.
-
-Now we come to Marwan Hassan.
-About five years ago now I wandered into Marwan's office saying I liked rivers and wanted to describe sediment transport with statistical physics.
-I realize now when I said ``statistical physics" Marwan's eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well.
-I was lucky to use the right trigger words.
-Marwan, thanks for taking the risk on me, for all of the guidance, for lending me hundreds of books and papers, for giving me freedom to explore ideas, and most of all for showing me that it's ok (and secretly expected) to be human in science.
-
-My friends in geography have also been great.
-Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, Jiamei, Xingyu, Niannian, and Emma, plus the outsider Eaton-ites, Will, Dave, Rose, and Anya.
-In particular thanks to Shawn for all of the shared speculations, and Matteo and Conor -- friendships forged through suffering!
-We'll have to keep in touch and form productive research collaborations once Matteo comes back from retirement.
-
-Thank you Rui Ferreira and Brett Eaton!
-Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics.
-Your kindness, physics knowledge, and wide-ranging interests beyond physical science are inspiring.
-Thanks for the thoughtful comments and advice on my PhD.
-I'm not forgetting what I owe you!
-I will always be impressed at your concise way of speaking and unbounded kindness.
-Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modelling in light of real world complexity.
-Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done.
-Maybe they'll bring me closer to your way of thinking!
-
-I'd be amiss not to mention David Furbish.
-Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation.
-At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes".
-Now I have a thesis on it.
-David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy from stochastic landscape evolution, to probability as a physical quantity like energy and momentum and fringe ideas in cosmology.
-These are my people!
-David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.
-
-Among UBC Geographers I'd like to thank Nina Hewitt!
-It was always a pleasure and sometimes a sigh of relief to TA your courses, which are so uniquely organized and student-focused.
-UBC Geography is a much greater department with you in it.
-I'm so thankful to have been a part of your process transforming undergrads into scientists.
-I'll be happy if I can teach like you one day!
-
-Finally, Mary!
-My PhD has not been great in terms of work-life balance.
-Thanks for sticking it out, putting up with my late bedtimes, my early wakeups, the rambly science jargon, and all of those social events I skipped.
-It's time!
-I'm done.
-The pandemic also seems to be winding down.
-We can finally get to Aus!
-I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella.
-I hope to take up still lives with Marianne's coaching.
-After that, we're a step closer to settling once and for all what colour cat we're gonna get.
-I'm personally now leaning toward white.
-
-Completing a PhD looks like a personal achievement, but it's really not.
-It's an expression of 30 years of influence and guidance from other people at every stage, my parents, my sisters, my grandmother, and a huge network of teachers and mentors.
 I'm kinda tired because it's the end, so it's natural to drift on to recollections of my favorite times in life, Watoga and Williams River with my family, encyclopedias in Gugs' living room, watching eagles over the South Branch with dad, discussions with Johnathan Ramey about my education, the glowing fireflies under my great grandpa's tree, ``cross-up, cross-up, squeeze" in Mindy's calc class, or laughter about stolen watermelons with my sisters.

### 2021-07-05 ch1intro.tex
lots of progress. Still lots of work to do though.

@@
 Earth's biota colonize these networks which form conduits for migration, transmitting water, genetic information, and organic material alike, while biota and chemical decomposition convert sediments to soils, staging the joint evolution of life and landscapes which has occurred across geological time.
+
 Human impacts on these old patterns have become severe, in what has been characterized as an environmental and social crisis \citep{Slaymaker2021}.
@@
 \be \dot{x} = V\eta(t), \ee
-where the noise is now a ``dichotomous Markov noise" \citep{Bena2006}, which is essentially a random switch or telegraph-type signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996} as displayed in figure \ref{fig:lislefig} panel (a).
+where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph-type signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
+This noise is displayed in figure \ref{fig:lislefig} panel (a).
 Some particle trajectories solving equation \ref{eq:lislelangevin} are displayed in \ref{fig:lislefig} panel (b).
@@
 Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
+These two definitions, $q$ defined with a surface and $\Phi$ defined with a volume, are generally not equivalent \citep{Ancey2020}.
 
@@
 Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
-He assumed that the flow power $P_f$ available to move sediment scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
-Considering that the average downstream flux of particles is $q$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ required to sustain particle motion scales as $P_g \propto q/\tau^{1/2}.$
+He assumed that the flow power $P_f$ available to move sediment in some volume scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
+Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ in a volume required to sustain particle motion scales as $P_g \propto q/\tau^{1/2}.$
  Balancing flow energy against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
-\be q = k(\tau-\tau_c)\tau^{1/2}, \label{eq:bagnold}\ee
+\be \Phi = k(\tau-\tau_c)\tau^{1/2}, \label{eq:bagnold}\ee
 which has shown good correspondence with laboratory data at large transport rates, given careful calibration of the constant factors $k$ and $\tau_c$.
 
-The large shear stress limit $q \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1990}.
-The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of the details have since turned out to be incorrect.
+The large shear stress limit $\Phi \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1990}.
+The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of Bagnold's assumptions have since turned out to be incorrect.
 Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and the flow power dissipated by sediment transport shows only a weak correlation the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
 These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$ and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
-More generally, Bagnold's formulation faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}, which is a topic for another thesis.
-The difficulties with Bagnold's approach led to many revisions of his theory which kept Bagnold's sharp physical insights, but included new experimental conclusions on the energetics of sediment transport as they became available \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
+More generally, Bagnold's formulation faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}.
+The difficulties with Bagnold's approach led to many revisions of his theory which kept Bagnold's physical reasoning, but includes more correct details of energetics of sediment transport \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
 
@@
 
-Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$, but this approach has been revised by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
-Some of these authors have argued that instead, one should calculate $P_n$ as an exceedance probability.
+Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$ from a geometric series, but this approach has been criticized by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
+Some of these authors have argued convincingly that instead, one should calculate $P_n$ as an exceedance probability.
 If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution \citep{Cox1965}:
@@
 This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^n}{n!}e^{-k_E T}, $
-so Einstein's mean sediment flux (eq. \ref{eq:einflux}) becomes
+so Einstein's mean sediment flux across the control surface (eq. \ref{eq:einflux}) becomes
 \be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
@@
 \be q = E \ell, \ee
-where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit bed area \citep{Wilcock1997a,Furbish2012}.
+where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit streambed area \citep{Wilcock1997a,Furbish2012}.
 
-Alongside this central result $q=E\ell$ for the mean flux, one of Einstein's most influential and enduring ideas was his formulation of the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
+Alongside this central result $q=E\ell$ for the mean flux, Einstein formulated the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
 The original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
@@
 This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
-random) resisting quantities \citep{Grass1970} is a key part of Einstein's legacy. \citet{Paintal1971} made a significant extension
-by including the random supporting geometry of bed particles into the force balance. More recently, \citet{Tregnaghi2012}
-ammended the theory to include both force magnitude and duration \citep{Diplas2008, Valyrakis2013, Celik2014}.
+random) resisting quantities \citep{Grass1970} another of Einstein's key contributions.
+\citet{Paintal1971} made a significant extension
+by including the random supporting geometry of bed particles into the force balance.
+More recently, \citet{Tregnaghi2012}
+amended the theory to include both force magnitude and duration, incorporating the recognition that both are relevant for particle entrainment \citep{Diplas2008, Valyrakis2013, Celik2014}.
 Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.
@@
 \be \gamma \propto \frac{\tau-\tau_c}{d^2 V_s^2}.\ee
-Expressing the mean flux as $q = \gamma V$, the relationship between flux and bed shear stress becomes
-\be q = \frac{A}{d^2 V_s^2}\big(\tau-\tau_c\big)\big[c + d(\sqrt{\tau}-\sqrt{\tau_c})\big]. \ee
-This recovers the Bagnold scaling $q \propto \tau^{3/2}$ at large bed shear stresses.
+Expressing the mean flux as $\Phi = \gamma V$, the relationship between flux and bed shear stress becomes
+\be \Phi = \frac{A}{d^2 V_s^2}\big(\tau-\tau_c\big)\big[c + d(\sqrt{\tau}-\sqrt{\tau_c})\big]. \ee
+This recovers the Bagnold scaling $\Phi \propto \tau^{3/2}$ at large bed shear stresses.
 
@@
 \be q(x,t) = \int_0^\infty dx' F(x-x',t)E(x',t). \ee
-In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a just-entrained particle steps a distance $l$ before deposition.
+In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a just-entrained particle steps a distance $x$ before deposition.
 In general, this approach can handle non-uniform conditions by including space and time dependence in $E$ and $F$.
@@
 where $F(x,t)$ represents the probability density that particles move \textit{at least} a distance x in time $t$ (right after entrainment).
+
 For a simple example of the Furbish et al formalism, consider that particles move with a constant velocity $V$ and have a deposition rate $k_D$. 
@@
 \be q = \rho k_E \int_0^\infty dx' \int_0^\infty dt' \theta(Vt-x)k_D\exp(-k_D t) = \rho k_E V/k_D. \label{eq:furbflux}\ee
-Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$. Equation \ref{eq:furbflux} therefore provides another perspective on Einstein's result $q= E\ell$.
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$.
+ Equation \ref{eq:furbflux} therefore provides another perspective on Einstein's central result $q= E\ell$.
 
@@
 
-\citet{Exner1925} was probably the first to write the mathematical relationship between sediment transport and topographic change. He wrote
+\citet{Exner1925} was probably the first to write the mathematical formula linking sediment transport to topographic change.
+He wrote
 \be (1-\phi)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \label{eq:exner}\ee
@@
 
-Within the Exner equation, the elevation $z$ and sediment flux $q$ are represented as continuum fields.
-This representation can be interpreted as a spatial average over the detailed positions of individual grains, and it is expected to be valid whenever the scales of interest are large compared to the size of the averaging window, which is must be taken much larger than the largest grains in the modelling domain \citep{Coleman2009}.
-Yet there are many Earth surface processes where the scale of interest is not much larger than the largest grains involved in the system.
-We can wonder, for example, how large boulders in mountain channels control the formation of steps \citep{Chin1999,Zimmermann2010, Comiti2012, Saletti2020}, or other bed structures having sizes comparable to channel widths, like ribs or stone cells \citep{Hassan2008,Venditti2017}.
-In these cases, individual grains are comparable to the scales of interest, and the continuum description provided by the Exner equation is not applicable.
-
-\citet{Nakagawa1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation based upon Einstein's concepts of entrainment and deposition rates.
-According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition:
+\citet{Nakagawa1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation using Einstein's entrainment and deposition rates.
+According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition rates:
 \be \partial_x q(x,t) = E - D. \ee 
-Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described in Einstein's variables:
+Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described with Einstein's rates:
 \be (1-\phi) \px z(x,t) = D - E.\ee
-In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the result of entrainment at all upstream locations, giving
+In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the eventual result of entrainment from all upstream locations, giving
 \be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
@@
 
-The sediment flux . . . .. has fluctuations and scale dependence. 
-An early approach to calculate the sediment flux distribution is due to \citet{Hamamori1962}, who evaluated the sediment flux as the result of migrating bedforms.
-This approach has been extended to include bedforms with a range of sizes \citep{Guala2014,Nikora1997}.
+Sediment transport rates always fluctuate \citep{Kuhnle1988,Hoey1992,Recking2012}, even under the most controlled laboratory conditions available \citep{Ancey2006,Roseberry2012}.
+At short timescales, fluctuations arise from the intermittent arrivals of individual grains \citep{Bohm2004,Ballio2018}.
+At moderate timescales they emerge from waves of moving grains \citep{Heyman2014} or the episodic entrainment of clusters of grains \citep{Strom2004,Papanicolaou2018}.
+At the longest timescales, they originate from migrating bedforms \citep{Guala2014} or cycles of aggradation and degradation in pools \citep{Dhont2018}.
+These mechanisms occur in conjunction with other sources of bedload transport fluctuations like grain size sorting \citep{Iseya1987,Cudden2003}, flow discharge variations \citep{Wong2006,Mao2012,Redolfi2018}, or sediment supply perturbations \citep{Lisle1993,Madej2009,Elgueta-Astaburuaga2019}.
+
+Despite widespread acceptance that sediment transport rates always fluctuate, we still have relatively little understanding of how to describe these fluctuations.
+\citet{Hamamori1962} was probably the first to calculate a probability distribution for the sediment flux, but this subject fell into relative obscurity for a long period, until it was eventually revisited \citep{Nikora1997,Ancey2006}.
+Now there are several approaches to calculate sediment flux distributions which are described below.
+
+Recognition that sediment transport rates fluctuate requires careful interpretation of sediment transport measurements. 
+In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Chartrand2018}, impact sensors \citep{Rickenmann2007}, computer vision \citep{Roseberry2012}, or sediment traps \citep{Papangelakis2016}.
+
+This averaging process in context of fluctuations introduces scale-dependence to sediment transport measurements, whereby averaged transport characteristics depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
+A fundamental question raised by scale-dependence is how large averaging windows must be for sediment flux measurements to converge.
+To date, very few mathematical models have addressed this issue. Several that have are reviewed below.
 
@@
 In the case of mean field models, the devices are heuristic scaling arguments and semi-empirical formulas, while in the case of stochastic models, they are simplified dynamical equations driven by idealized noises, like sequences of pulses, alternating switches, and erratic fluctuations.
+This latter methodology is applied throughout the thesis.
 
@@
 
-The original model by \citet{Einstein1937} describing particle trajectories when the movement velocity is infinite was subsequently improved to include a finite yet constant velocity \citep{Gordon1972,Lisle1998,Lajeunesse2017}.
+The original model by \citet{Einstein1937} describing particle trajectories with infinite movement velocity was subsequently improved to include a constant velocity \citep{Gordon1972,Lisle1998,Lajeunesse2017}.
 Particle velocities fluctuate during movements between entrainment and deposition, so this should also be accounted for in bedload trajectory models.
@@
 
-Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories which includes velocity fluctuations in the motion phase, then demonstrates how these trajectories can be used to calculate scale-dependent probability distribution of the bedload flux, building on all of these earlier studies.
+Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories which includes velocity fluctuations in the motion phase, then demonstrates how these trajectories can be used to calculate scale-dependent probability distribution of the bedload flux, building up and linking together all of these earlier studies.
 
@@
 
-Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Ancey2008,Heyman2013,Ancey2014}, which violates conservation of mass and the Exner equation \citep{Exner1925,Tsujimoto1978}, and precludes the possibility that sediment can become buried in the subsurface of the bed.
+Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Heyman2013,Ancey2014}, which violates conservation of mass and the Exner equation \citep{Exner1925,Tsujimoto1978}, and precludes the possibility that sediment can become buried within the bed.
 Although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2002,Hassan2017}, our understanding of how long sediment can remain buried is limited.
@@
 Rather than describing particles moving downstream through cycles of entrainment and deposition as in \citet{Einstein1937}, these methods characterize particle motions by herustic entrainment and deposition rates, or interarrival time distributions, which are somewhat abstract objects, not concretely related.
-Despite the considerable progress these approaches have brought, the lack of connection to the underlying particle mechanics is a real shortcoming which would ideally be rectified.
+Despite the considerable progress these approaches have brought, the lack of connection to the underlying particle mechanics requires attention.

### 2021-07-05 ch2flipFlop.tex
rolling on the conclusion
lots of progress. Still lots of work to do though.

@@
 A set of studies has developed stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Furbish2012}. 
-Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Ancey2008,Ancey2014}, the migration of bedforms and sediment waves \citep{Hamamori1962,Guala2014,Hoey1992,Recking2012}, and a host of other processes \citep{Dhont2018}.
+Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
 Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale-dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
@@
 First, we do not yet have the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution, and the dependence of its statistical moments on the observation time.
-Here, we develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a Langevin-type dynamical equation which combines earlier approaches \citep{Fan2014,Ancey2014,Lisle1998,Lajeunesse2017}
-This stochastic equation includes alternation between motion and rest at random intervals while particles in motion have velocity that fluctuate around a mean value.
-Using the probability distribution of particle position generated by this model, we construct a formalism to derive analytically the probability distribution of the sediment flux.
+
+Here, I develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a Langevin-type dynamical equation in a combination of earlier approaches \citep{Fan2014,Ancey2014,Lisle1998,Lajeunesse2017}.
+This dynamical equation describes alternation between motion and rest at random intervals while particles in motion have velocity that fluctuates around a mean value.
+Using the probability distribution of particle position generated by this model, I construct a formalism to derive analytically the probability distribution of the sediment flux.
 This distribution exhibits observation-scale dependence as a result of velocity fluctuations among moving sediment.
-Below, we develop the new formalism in sec.
+Below, I develop the new formalism in sec.
 \ref{sec:mod}, solve it in sec.
-\ref{sec:res}, and we discuss the implications of our results and future research ideas in secs.
+\ref{sec:res}, and I discuss the implications of my results and future research ideas in secs.
 \ref{sec:disc} and \ref{sec:conc}.
@@
 \section{Description of motion-rest alternation with velocity fluctuations \label{sec:mod}}
-We consider an infinite one-dimensional domain populated with sediment particles on the surface of a sedimentary bed.
-We consider that the flow is weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common.
-The flow is in constrast strong enough so that particles move.
-We label the downstream coordinate as $x$, so that the downstream velocity of a moving particle is $\dot{x}$, and we describe all sediment particles as independent from one another but governed by the same underlying dynamical equations, meaning we neglect any influence of sediment size or shape or spatial variations in the overlying fluid flow.
+Consider an infinite one-dimensional domain populated with sediment particles on the surface of a sedimentary bed.
+Consider that the flow is weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common.
+The flow is in contrast strong enough so that particles move.
+Label the downstream coordinate as $x$, so that the downstream velocity of a moving particle is $\dot{x}$, and we describe all sediment particles as independent from one another but governed by the same underlying dynamical equations, meaning we neglect any influence of sediment size or shape or spatial variations in the overlying fluid flow.
 
@@
 We write $k=k_E+k_D$ as a shorthand.
+This problem is similar to the ``randomly flashing diffusion" developed by Luczka and coworkers \citep{uczka1992,Luczka1993,Luczka1995}, although it includes a more complex asymmetric noise ($k_E\neq k_D$) and involves particle rest.
 
@@
 \citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{sec:landscape}), writing
-\be (1-\phi) \pt h(x,t) = -\px q(x,t). \ref{eq:stocEx}\ee
+\be (1-\phi) \pt h(x,t) = -\px q(x,t). \label{eq:stocEx}\ee
 In this one-dimensional equation, $h(x,t)$ is a continuous field representing the longitudinal channel profile.

### 2021-07-05 ch5langevin.tex
lots of progress. Still lots of work to do though.

@@
 
-With the above forces, we express the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics as
+With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics becomes
 \be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
-This equation replaces the steady friction terms of earlier stocahstic bed load models with an episodic term which provides a more realistic representation of particle-bed collisions during saltation.
+This equation replaces the steady friction terms of earlier models () with an episodic term that provides a more realistic representation of particle-bed collisions during saltation.
 It represents a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
-Collisions introduce ``jumps" in velocity while turbulent generates "diffusion".
-The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
-Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Luczka1995,Daly2010,Mau2014,Dubkov2016,Dubkov2019}.
+Collisions introduce ``jumps" in velocity while turbulent generates "diffusion", and the collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
+Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
 One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
@@
 
-We derive the equation governing the streamwise velocity distribution $P(u,t)$ from a simple limiting argument in appendix \ref{sec:appendixA}, finding
-\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + \mathcal{I}_c(u,t). \label{eq:master} \ee
+The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix \ref{sec:appendixA}, providing
+\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
 In this equation, we introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$. The term
@@
 Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
-This term is analogous to the collision integral of the Boltzmann equation in kinetic theory and granular gases \citep{Duderstadt1979, Brilliantov2004}.
+This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}.
 Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) goes to infinity.
-Mathematically, it represents the probability distribution of the product between $\varepsilon$ and $u$ \citep[c.f.][]{Feller1968}.
+Mathematically, the collision integral term represents the probability distribution of the product between $\varepsilon$ and $u$ \citep[c.f.][]{Feller1968}.
 
-Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so we make one further approximation.
-We assume the distribution of dissipation coefficients $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$.
-This allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
+Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so I introduce one further approximation.
+Assuming the distribution of dissipation coefficients $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$ allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
 Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
@@
 where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
-In what follows, we drop all but the first two terms to obtain the leading order contribution of particle-bed collisions to the velocity distribution.
+In what follows, I drop all but the first two terms in this expansion to obtain the leading order contribution of particle-bed collisions to the velocity distribution.
 Higher orders could always be included later by perturbation theory \citep{Morse1953}.
-We solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$.
-Scaling the time in Equation \ref{eq:master} by $1/\nu$, we can see this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions.
+I solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$.
+Equation \ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg1/\nu$).
 
@@
 \label{sec:solution}
-Hereafter we drop the prime on the most common streamwise restitution coefficient $\ve'$. With the truncation to two terms, equation \ref{eq:master} gives 
+Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
+With the truncation to two terms, equation \ref{eq:master} gives 
 \be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
-which is now a non-local ordinary differential equation. have seen some attention in the mathematics literature, where they are called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
-In the appendix we solve equation \ref{eq:governer} using Laplace transforms, providing
+which is now a functional differential equation that is nonlocal in velocity due to the last term on the right hand side.
+Such equations seen some attention in the mathematics literature, where they are called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
+In the appendix I solve equation \ref{eq:governer} using Laplace transforms, providing
 \begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
@@
 \be K_\pm = d(\lambda_+-\lambda_-)\prod_{l=1}^\infty (-d\lambda_\pm^2 \ve^{2l} + \gamma \lambda_\pm \ve^{l} + 1). \ee
-Although this velocity distribution appears quite complicated, one can verify that this is a normalized probability distribution which has very simple limiting behaviors as the mode dissipation constant $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
+Although this velocity distribution appears quite complicated, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode dissipation constant $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
 
-It is rather simple to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
-The first moment is
+It is possible to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
+From this method the mean velocity is
 \be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve},\ee
@@
 \be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
-This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to them.
-In contrast, velocity fluctuations depend sharply on the parameters representing particle-bed collisions.
+This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to fluid properties.
+In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are the leading control on particle velocity fluctuations.
 
-Figure \ref{fig:fig2} depicts velocity characteristics for different realizations of the fluid and collisional forces.
+Figure \ref{fig:fig2} depicts the velocity characteristics of particles for different realizations of the fluid and collisional forces.
 \begin{figure}
@@
 We can see an apparent transition from exponential-like to Gaussian-like velocity distributions as typical collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
-In between, the full distribution \ref{eq:steadystate} resembles a Gamma distribution, although it is not a Gamma distribution.
+In between, the full distribution \ref{eq:steadystate} resembles a Gamma distribution, although it is not exactly a Gamma distribution.
 
@@
 
-In fact, the apparent transition in figure \ref{fig:fig2} can be made rigorous: despite its complex appearance, simple Gaussian and exponential distributions appear as rigorous mathematical limits of equation \ref{eq:steadystate}.
+In fact, the apparent transition from exponential to Gaussian in figure \ref{fig:fig2} can be formalized: despite its complex appearance, simple Gaussian and exponential distributions appear as rigorous mathematical limits of equation \ref{eq:steadystate}.
 When particle-bed collisions are completely inelastic, \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic, \ref{eq:steadystate} becomes Gaussian.
@@
 The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is rather easy to see.
-Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for that with $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to notational differences:
+Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for that with $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
 \be P(u) = \frac{d}{\sqrt{\gamma^2 + 4d}}e^{\frac{\gamma u }{2 d} - \frac{\sqrt{\gamma^2 + 4 d}|u|}{2d}}. \ee
@@
 The challenge is that the statistical moments \ref{eq:mean} and \ref{eq:2ndmom} diverge at the same time as the denominator factors of the distribution \ref{eq:steadystate}.
-In appendix \ref{sec:appendixB} we return instead to the original equation \ref{eq:governer} to evaluate this elastic limit, obtaining
+In appendix \ref{sec:appendixB} I return instead to the original equation \ref{eq:governer} to evaluate this elastic limit, obtaining
 \be P(u) = \frac{1}{\sqrt{2\pi\sigma_u^2}}e^{-\frac{(u-\bar{u})^2}{2\sigma_u^2}}. \label{eq:gaussian}\ee
@@
 
-Now we compare the analytical distribution \ref{eq:steadystate} with the available experimental data.
-Upfront, we point out that the distribution above has free parameters and this is only a proof of concept that the velocity distribution is capable of fitting the available experimental data; it is not a proof that this is the underlying mechanism for these data blahblahblah that was a good writing day.
+A variety of velocity distributions have been observed in bedload transport experiments, ranging from exponential to Gaussian in shape. 
+This section compares these with the theoretical result eq. \ref{eq:steadystate}.
+Comparing eq.
+\ref{eq:steadystate} with experimental data requires values for the steady component of the drag force $\Gamma$, the particle mass $m$, the magnitude of turbulent drag fluctuations $D$, the rate of particle-bed collisions $\nu$, and the most common fraction of momentum $\varepsilon$ dissipated in each collision.
 
+Below, I compare this formula with the results of six different experiments.
+In each case, the particle mass is computed from experimental parameters assuming spherical sediment as $m = \pi \rho_s d^3/6$, where $\rho_s$ is the sediment density and $d$ is the particle diameter.
+The steady component of the drag force is estimated as
+\be \Gamma =  \frac{\pi}{8}\rho C_D(\Rey_p) d^2 u_\ast^2,\ee
+where $\rho$ is the fluid density, $C_D(\Rey_p)$ is the drag coefficient dependent on particle Reynolds number, and $u_\ast$ is the shear velocity of the flow, considered as a proxy for the average slip velocity between moving particles and the flow.
+The magnitude $D$ of turbulent fluctuations is eliminated with the mean and variance of the particle velocity.
+The remaining coefficients in the model, the dissipation per collision $\ve$ and the collision rate $\nu$,, are treated as calibration parameters and are tuned to provide best fit between the distribution \ref{eq:steadystate} and the experimental data.
+
+The parameters for each experiment and the best-fit calibration parameters are collated in table \ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in figure \ref{fig:fig4ch5}.
 \begin{table}
@@
 		\def~{\hphantom{0}}
-		\begin{tabular}{lccc}
-			$a/d$  & $M=4$   &   $M=8$ & Callan  \\[3pt]
-			0.1   & 1.56905 & ~~1.56~ & 1.56904\\
-			0.3   & 1.50484 & ~~1.504 & 1.50484\\
-			0.55  & 1.39128 & ~~1.391 & 1.39131\\
-			0.7   & 1.32281 & ~10.322 & 1.32288\\
-			0.913 & 1.34479 & 100.351 & 1.35185\\
+		\begin{tabular}{l|ccccccc}
+			Experiment & $d$ [$mm$] & $u_\ast$ [$cm/s$]  & $\Rey_p$ [-] & $\St$ [-] & $\Fro$ [-] & $\ve$ [-] & $\nu$ [$s^{-1}$] \\
+			\toprule 
+			\textit{(a) Fathel et al} & 0.5 & 2.0 & 9.97 & 2.9 & 0.3 & 0.08 & 24. \\ 
+			\textit{(b) Lajeu. et al} & 2.2 & 4.4 & 99.  & 29.  	& 1.5 & 0.21 & 2.6 \\ 
+			\textit{(c) Liu et al}    & 1.1 & 8.6 & 94.  & 29.  & 0.3 & 0.28 & 34.  \\ 
+			\textit{(d) Heyman et al} & 6.4 & 9.7 & 620. & 180. & 1.3 & 0.96 & 13. \\ 
+			\textit{(e) Ancey et al}  & 8.0 & 7.4 & 590. & 170. & 2.1 & 0.92 & 4.8 \\ 
+			\textit{(f) Martin et al} & 7.1 & 5.8 & 410. & 120. & 3.7 & 0.89 & 2.1  \\ 
 		\end{tabular}
-		\caption{Values of $kd$ at which trapped modes occur when $\rho(\theta)=a$.}
-		\label{tab:kd}
+		\caption{Parameters used to fit the distributions in figure \ref{fig:fig4ch5}.
+The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St$, and Froude number $\Fro$.
+The final two columns involve values that were tuned to fit the theoretical distribution \ref{eq:steadystate} to the experimental data, the dissipation $\ve$ and the collision rate $\nu$.
+Units are indicated in square brackets.
+One can see a generally increasing relationship between particle size and elasticity, and likewise a decreasing relationship between particle size and collision frequency.
+}
+		\label{tab:calib}
 	\end{center}
 \end{table}
-
+In every case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model \ref{eq:langevin} captures most of the underlying physics.
 \begin{figure}
 	\centerline{\includegraphics{./figures/ch5/Fig4expComparison.png}}
-	\caption{The features of the four possible modes corresponding to
-		(\textit{a}) periodic\protect\\ and (\textit{b}) half-periodic solutions.}
-	\label{fig:fig4}
+	\caption{This figure compares the available experimental data on velocity distributions and the theoretical velocity distribution \ref{eq:steadystate} with parameters $\nu$, $\varepsilon$, and $D$ treated as calibration parameters.
+In all cases there is good agreement, indicating that the model is capable of the range of distributions exhibited by the experimental data.} \label{fig:fig4ch5}
 \end{figure}
@@
 
-\subsection{Analogies to hillslope and wind-driven transport}
-
-It has long been recognized, probably since \citet{Bagnold1941} and certainly today \citep{}, that there are deep analogies between tranpsort phenomena in air and water.
-In an ideal perspective, we might imagine writing the governing equations for transport in water as a function of the fluid viscosity, then tuning the viscosity between air and water to obtain a description applying to both spheres.
-
-For example, in a mixture of small and large grains, neglecting any sorting effects whereby the mobility of small grains is contingent on the mobility of the large grains, small grains will have exponential velocities with relatively wide fluctuations, while large grains will have Gaussian velocities with relatively narrow fluctuations.
-Considering the over-all flux then as the number of moving particles times their velocities, transport fluctuations
-
-aeolean transport extension maybe
-particle size distributions - interesting implications
-connection to computational physics approach
-connection to the stochastic description of the flux
-
 \section{Conclusion}

### 2021-07-05 ch6conc.tex
lots of progress. Still lots of work to do though.

--- 2021-07-03 ch6conc.tex
@@
 \label{ch:conc}
-\section{The overall strategy}
 
-This thesis has presented a review and four chapters which generalize and extend earlier work to describe sediment transport as a stochastic process.
-In the research, I characterized the movements of grains along a wider range of timescales than before, and I developed new a new formalism using these movements to predict the sediment flux.
+The research in this thesis generalizes and extends earlier work to describe sediment transport as a stochastic process.
+In the research, I have characterized the movements of grains along a wider range of timescales than before, framed existing research in terms of Langevin and Master equations, and produced more detailed descriptions of the bedload transport phenomenon, addressing several long-recognized shortcomings of earlier stochastic models.
+In particular I have filled in the connection between sediment fluxes and individual praticle motions, included particle velocity fluctuations in a descriptions of individual particle motions through rest and motion, evaluated the role of bed elevation changes on sediment transport rate fluctuations, and included the process of sediment burial in descriptions of downstream particle movement.
+
+\section{Overall methodology of the thesis}
 
@@
 
-Probability
+The overarching strategy in all of these developments has been to identify control and response variables, represent variable controls by idealized noises (entrainment and deposition events, particle burial events, turbulent forces, or particle-bed collisions), and then to write dynamical equations relating the respones variables to these stochastic control variables.
+This phrases sedimentary dynamics in terms of Langevin-like equations, stochastic analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for the response variable of interest in the problem and the force $F$ is a stochastic idealization of the control variables.
+
+Once the stochastic dynamical equation was written for a given problem, its solutions were counted () across realizations of the control variables to form a probability distribution of the response variable.
+These distributions were generally governed by integro-differential equations (), and the distributions solving these equations were obtained with transform calculus ().
+
+In every case, the result is a probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (), particle velocity (), bed elevation (), and sediment transport rate ().
+The same basic strategy should be applicable to a wide range of problems in geomorphology where variability matters \citep{Furbish2021a}.
 
@@
 
+A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modelling are constrained.
+The noises used in this thesis include white Gaussian, Poisson, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
+Turbulence was described using Gaussian noise (); while instantaneous steps (), particle arrivals to a control surface (), and particle-bed collisions () were described with Poisson noise.
+Alternation between motion and rest was represented with dichotomous noise ().
+Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise as required ().
+This inclusion of multiple noise sources has not yet to my knowledge been pursued in any earlier stochastic models of sediment transport.
+
+River science offers no guarantee that these idealized white noises which we happen to understand best are sufficient to describe its phenomena.
+White noises are an idealization which is probably never represented in nature \citep{Gardiner1983,Kubo1972}.
+In contrast, colored spectra in which fluctuations distribute power unevenly across frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
+In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
+It will be problematic if these phenomena require colored noise for their description, because dynamical equations driven by colored noise can be extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
+Even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.
+
 \section{Key contributions}
 
-\subsection{Probability distribution of the sediment flux from micromechanics of particle trajectories}
+\subsection{Calculation of the probability distribution of the sediment flux from micromechanics of particle transport}
+
+The first major contribution of this thesis is in chapter ().
+Here, I formulated the probability distribution of the sediment flux from the trajectories of individual particles moving downstream.
+This work unifies the sediment trajectory models originating from Einstein () with the renewal theory approach to calculate the sediment flux ().
+The striking feature of this formulation is that, as a result of the particle dynamics, the mean bedload flux becomes scale-dependent, whereby the expected magnitude of the flux depends on the time-period over which it is observed ().
+
+An expectation that scale-dependence derives from individual particle trajectories was described theoretically by \citet{Ballio2018}, but this had not been described in a mathematical model until now.
+Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (), and an emerging body of research was producing the full probability distribution of the flux, but without referencing individual movement characteristic ().
+This work unifies these two research themes and indicates a path toward a statistical mechanics formulation of bed load sediment transport based on the grain scale physics.
 
@@
 
+Second, in chapter () I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
+Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (), and this was later improved to include the duration of motion ().
+Until this thesis, movement velocities in analytical models were considered constant () which contrasts with our experience (). 
+Although some numerical models have described motion/rest cycles with velocity fluctuations \citep{Fan2016,Bialik2012,Schmeeckle2014}, they had not resolved the novel three-range diffusion characteristics these dynamics imply.
+
+This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state (); it introduces a dimensionless Peclet number as an important characteristic of bedload sediment transport (); and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains ().
+These developments constitute a new understanding of bedload movement across its timescales ().
+
 \subsection{Quantification of the control of bed elevation fluctuations over sediment transport fluctuations}
 
-\subsection{Predicting how sediment burial affects the downstream spreading of sediment tracer particles}
+Third, chapter () modified the birth-death description of bedload transport () to include feedbacks between the local bed elevation and the entrainment and deposition rates.
+This allowed for a mathematical investigation of (1) how bed elevation changes affect sediment transport rates and (2) how bed elevation changes control the residence times of particles buried in the bed.
+Sediment transport fluctuations have come under increasing scrutiny with the resurgence of stochastic models in sediment transport (), while the burial times of particles are crucial for using sediment tracers to predict sediment transport ().
+Earlier birth-death models had generally considered that entrainment and deposition rates of particles remain constant even though these processes imply bed degradation and aggradation respectively, which are known to modify the entrainment and deposition rates in a negative feedback.
 
-\section{Models and the real world}
+This work demonstrates that this negative feedback buffers bedload transport fluctuations whenever collective entrainment occurs, meaning the magnitude of bedload transport fluctuations depends on the rate of bed elevation change ().
+The residence times of buried particles are random variables that lie on heavy-tailed power-law distributions.
+These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to predict the downstream sediment flux in applications by tracking sediment tracer particles ().
 
-\section{Conclusion and next steps}
+\subsection{Characterization of how sediment burial affects the downstream transport of sediment particles}
+
+Fourth, chapter () presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzslav2021}.
+
+Until this work, the mechanisms which produce the different spreading rates of particles across the scaling ranges identified by Nikora and coworkers had been uncertain for several decades (), and earlier models works had included what were believed to be the required features without describing three or more scaling ranges ().
+The work further demonstrates that many approaches to describe individual particle motions are just clever reproductions of the continuous time random walk formalism from physics, indicating underlying unity within a diverse body of research and bringing powerful tools from physics to the sediment transport problem.
+
+\subsection{Description of how particle-bed collisions control movement velocities of grains}
+
+Finally, chapter () displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
+This work provides the first comprehensive description of all bedload velocity distributions observed in experiments (), while earlier works had described only particular end-member cases ().
+
+This work includes ideas from granular gas theory into sediment transport modelling and provides a concrete example of the control grain-scale processes exert over large scale landscape evolution.
+In particular, since the work implies the shape of the bedload particle velocity distribution may depend on particle size, this has implications for mechanistic understanding of how abrasion by collisions of saltating bedload occurs (), which relates ultimately to rate at which Earth's bedrock canyons incise.
+
+\section{Limitations and future research directions}
+
+The objective of this thesis was to produce new understanding of how to describe bedload transport using statistical physics.
+
+In some cases these limitations specific to the models produced in this thesis, but in most, they are shared in common with a majority of other models in the same research paradigm \citep{Ancey2020,Furbish2021a}.
+
+\subsection{Landscape dynamics and channel morphology}
+
+The first limitation concerns the assumption, implicit in every chapter of this thesis, that sediment transport characteristics are steady through time and space.
+This assumption contrasts with conditions in real gravel-bed river channels, where riffles, bars, and steps control the movements of individual grains \citep{}, sediment is supplied in episodic bursts from mass movements \citep{}, woody debris forces sediment deposition \citep{}, and flow conditions vary through time with weather \citep{}, climate change \citep{}, and seasonality \citep{}.
+
+Here I have considered extremely idealized conditions where sediment transport and landscape dynamics are not linked, except for chapter 2 where bed elevations were coupled to sediment entrainment and deposition.
+
+\subsection{Grain size distributions}
+
+Sediment in real rivers spans a range of sizes, while I have modelled sediment transport of one grain size only, characterized by
+
+\subsection{Rivers, hillslopes, and dunes}
+
+It has long been recognized, probably since \citet{Bagnold1941} and certainly today \citep{}, that there are deep analogies between tranpsort phenomena in air and water.
+In an ideal perspective, we might imagine writing the governing equations for transport in water as a function of the fluid viscosity, then tuning the viscosity between air and water to obtain a description applying to both spheres.
+
+For example, in a mixture of small and large grains, neglecting any sorting effects whereby the mobility of small grains is contingent on the mobility of the large grains, small grains will have exponential velocities with relatively wide fluctuations, while large grains will have Gaussian velocities with relatively narrow fluctuations.
+Considering the over-all flux then as the number of moving particles times their velocities, transport fluctuations
+
+aeolean transport extension maybe
+particle size distributions - interesting implications
+connection to computational physics approach
+connection to the stochastic description of the flux
+
+\section{Conclusion}
+
+This thesis applied statistical mechanics methods to describe the movements of individual grains along streambeds.
+The research has produced new understanding of the relationships between the overall sediment transport rates which control channel evolution and the movements of individual grains.
+The work embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as their fluctuations.
+
+The founders of process geomorphology always acknowledged variability in landscapes \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes despite it, like averaged flow conditions to avoid including fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to summarize climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid wide grain size distributions \citep{Parker1982,Andrews1983}.
+
+The stochastic models of sediment transport developed in this thesis hint toward a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit this old question in geomorphology: How does variability shape Earth's surface?

### 2021-07-05 pref.tex
lots of progress. Still lots of work to do though.

--- 2021-07-04 pref.tex
@@
 The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
-The thesis includes two published works, Chapters \ref{ch:bedEle} and \ref{ch:downdiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
+The thesis includes two published works, Chapters \ref{ch:ch3} and \ref{ch:downDiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
 Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to submission of the thesis.

### 2021-07-06 app_ch2.tex
rolling on the conclusion
chapter 1 is flawless. gradually changing the notation and rewriting chapter 2. really getting bogged down on it.

@@
 
-\chapter{Ch. 2 calculations}
+\chapter{Calculations involved in dynamical sediment flux model}
+\label{sec:appendixA}
+\section{Derivation of the master equation for particle position}
+\label{sec:appAmaster}
+The probability distribution of position in section \ref{sec: satisfies $P(x,t) = \bra\bra \delta(x-x(t))\ket_\xi \ket_\eta$, where the averages are over all realizations of the two independent noises.
+Integrating the Langevin equation, substituting it, and taking the Fourier transform in space gives
+\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi\ee
+This can be rearranged as
+\be \tilde{P}(g,t) = \Big\bra \exp\Big[i g V \int_0^t du \eta(u)\Big] \Big\bra \exp\Big[ i g \sqrt{2D}\int_0^t du \xi(u) \eta(u)]\Big]\Big\ket_\xi \Big\ket_\eta \ee
+Using the classic identity for an average over white noise of an exponential  from Gardiner 1983 (which derives easily from the reduction of gaussian cumulants to 2 point correlation functions) gives, after realizing that $\eta^2 = \eta$,
+\be \tilde{P}(g,t) = \Big\bra \exp\Big[(igV-g^2D) \int_0^t du \eta(u) \Big]\Big\ket_\eta \ee
+Calling $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ as shorthands and taking a time derivative gives
+\be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket.\ee
+Taking a second time derivative opens up the possibility of using the Furutsu-Novikov formula: 
+\be G^{-1}\pt \tilde{P}(g,t) = \pt \bra \eta F\ket = G\bra \eta F \ket + k\big[ \bra \eta \ket \tilde{P}(g,t) - G^{-1} \pt \tilde{P}(g,t) \big] \ee
+for which the best reference is gardiner. This finally gives
+\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
+and inverse Fourier transforming provides the master equation
+\be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0 \ee
+\section{Solution for the position probability distribution}
+\label{sec:fluccymastersol}
+The initial conditions for which to solve this equation are 
+\be P(x,0) = \delta(x) \ee
+\be \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]\ee
+These initial conditions come from the initial state
+\be P(x,0) = \lim_{t\rightarrow 0 } \frac{k_E}{k} \sqrt{\frac{1}{4\pi D t}} \exp\Big[-\frac{(x-Vt)^2}{4Dt}\Big]+ \frac{k_D}{k}\delta(x)\ee
+characterizing the potential for a particle to start in motion, where it undergoes advection and diffusion, and rest, where it stays put.
+In the derivation of the initial condition, it is convenient to use the Fokker-Planck equation satisfied by the Gaussian Green's function to convert the time derivative to a space derivative.
+Applying the rules above provides
+\be \bar{\tilde{P}}(g,s) = \frac{s + k +\phi  Dg^2 - i g V \phi}{s(s+k)+(Dg^2-igV)(s+k_E)}, \ee
+where $\varphi = k_D/k$.
+This appears correct as it reduces to Einstein and Brownian problems in the appropriate limits.
+The numerator terms encode the initial conditions. The denominator terms are where the real structure of the solution is contained.
 
-\section{Derivation of the master equation}
-\section{Solution for the position probability distribution}
+The numerator terms involving $g$ are easily expressed as first and second derivatives with respect to $x$.
+The problem, then is to calculate the inverse Fourier integral of the denominator.
+This can be conducted by finding the zeros of the denominator, expanding in partial fractions, then applying the contour integral
+\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g + i c} dg = \theta(x)e^{-c x}\ee
+that I've used to solve earlier problems. Rearranging the governing equation gives
+\be \tilde{\hat{P}}(g,s) = \frac{\phi D g^2 - ig V\phi  + s + k}{D(s+k_E)}\frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}}\ee
+The roots of the denominator are at
+\be g_\pm = \frac{iV}{2D}\Big[ 1  \pm R \Big],\ee
+where
+\be R = \sqrt{1 + \frac{4D}{V^2}\frac{s(s+k)}{s+k_E}}\ee
+The partial fractions expansion is therefore
+\be \frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}} = \frac{D}{i V R}\Big[ \frac{1}{g-g_+} - \frac{1}{g-g_-}\Big]\ee
+Giving 
+\be \tilde{P}(x,s) = \frac{-\phi D\px^2 + V\phi\px + s + k}{VR(s+k_E)}\Bigg[ \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_+} - \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_-}\Bigg]\ee
+or eventually
+\be \tilde{P}(x,s) = \frac{-\phi D\px^2 + V\phi \px + s + k}{VR(s+k_E)}\Bigg[ \theta(x) \exp\Big(\frac{Vx}{2D}(1-R)\Big) + \theta(-x)\exp\Big( \frac{Vx}{2D}(1+R)\Big) \Bigg].\ee
+which simplifies a bit
+\be \tilde{P}(x,s) = \frac{-\phi D\px^2 + V\phi \px + s + k}{VR(s+k_E)}\exp\Big[\frac{Vx}{2D} - \frac{V|x|}{2D}R \Big], \label{eq:laplace}\ee
+where as a reminder, $\phi = k_D/k$.
+
+An incredible propery from the Bateman manuscript by Erdelyi vol. 1 pg. 133 is
+\be \mathcal{L}^{-1} \Big\{ \frac{1}{s} \tilde{g}(s-a/s)\Big\} = \int_0^t \mathcal{I}_0\Big(2\sqrt{au(t-u)}\Big) g(u) du
+\ee
+which can be derived by taking a Laplace transform and using u-substitution.
+
+The problem is
+\be P(x,t) = \mathcal{L}^{-1}\Bigg\{ \frac{1}{V} \big[-\phi D\px^2 + V\phi \px + k + s \big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(-\frac{V|x|R}{2D} \Big)}{(s+k_E)R} \Bigg\}(t) \ee
+Using the shift property, this becomes
+\be P(x,t) = \frac{1}{V}e^{-k_E t} \mathcal{L}^{-1}\Bigg\{ \big[-\phi D\px^2 + V \phi \px + k_D + s\big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(\frac{-V|x|R_\ast}{2D}\Big)}{s R_\star} \Bigg\},\ee
+where 
+\be R_\ast = \sqrt{1 + \frac{4D(k_D-k_E)}{V^2}+ \frac{4D}{V^2}\Big(s-\frac{k_Ek_D}{s}\Big)}. \ee
+Using the property that $\mathcal{L}^{-1}\big\{s \tilde{f}\big\} = (\delta(t) + \pt)f(t)$ gives
+\be P(x,t) = \frac{1}{V} e^{-k_E t} \big[ -\phi D\px^2 + V\phi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big] \mathcal{L}^{-1}\Bigg\{ \frac{-\exp\Big[\frac{-V|x|R_\ast}{2D}\Big]}{sR_\ast} \Bigg\}(t) \ee
+Using the Erdelyi shifting property from Bateman manuscript gives 
+\begin{multline}  P(x,t) = \frac{1}{V} e^{-k_E t}  \big[ -\phi D\px^2 + V\phi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big] \int_0^t du \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) \\ \times \mathcal{L}^{-1}\Bigg\{ \frac{\exp\Big[ \frac{-V|x|}{2D}\sqrt{a+b s}\Big]}{\sqrt{a + b s}}\Bigg\}(u) \end{multline}
+where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$.
+Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator, the remaining transform can be evaluated from tables or mathematica, eventually giving (after integration over and setting to 1 of the fake parameter),
+\begin{multline} P(x,t) = \big[-\phi D\px^2 + V\phi \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du \end{multline}
+We see here a convolution between a modified Bessel function describing the probability that a particle is at rest, and a Gaussian propagator, damped by deposition, describing the drift and diffusion of moving particles.
+Probably the $\mathcal{I}_0$ term represents the probability that the particle has been in motion for a time $u$ after a time $t$ has elapsed.
+The expansion of this formula by evaluating the derivatives and so on gives nothing nice.
+it becomes a sum of mainly three terms.
+Two are convolution integrals over bessels $I_1$ and $I_0$ with elementary functions also factors inside the integrals.
+One is a delta function decaying with $k_D$, and another uis a Gaussian propagator decaying with $k_E$.
+
 \section{Moments of position}
+
+Multiplying the master equation by $x^l$ and integrating over all $x$ gives
+\be \pt^2 m_l -V l \pt m_{l-1} -k_E V l m_{l-1} + k \pt m_l - D l (l-1) \pt m_{l-2} - k_E D l(l-1) m_{l-2} = 0,\ee
+where $\bra x^l \ket = m_l$. 
+For $l=1$ this generates
+\be (\pt^2+ k \pt )\bra x \ket = k_E V \ee
+which has solution $ \bra x \ket = k_E V t/k$ given the initial conditions of the problem.
+The general solution is $\alpha_2 + k_E V t /k + \alpha_1 e^{-k t}/k$.
+The first moment is totally unaffected by diffusion.
+For $l=2$, the governing equation is
+\be (\pt^2 + k \pt)\bra x^2 \ket = 2V(\pt + k_E)\bra x \ket + 2 k_E D.\ee
+Plugging in the result for the first moment develops
+\be (\pt^2 + k\pt ) \bra x^2 \ket = \frac{2V^2 k_E}{k}[1+k_E t] + 2k_E D. \ee
+This time the general solution is
+\be \bra x^2 \ket = \frac{2k_E}{k}\Big[\frac{V^2k_D}{k^2} + D\Big]t + \Big(\frac{Vk_E t}{k}\Big)^2  + C_1 e^{-kt}/k + C_2.\ee
+This is subject to the conditions $\bra x^2 \ket(0) = 0$ and $\pt \bra x^2 \ket (0) = 2k_E D/k$, derived by the initial conditions used for the pdf, giving a variance
+\be \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) + 2\frac{k_E D}{k}t.\ee
+At short times $t\ll k^{-1}$, this becomes
+\be \sigma_x^2 \sim \frac{k_E k_D V^2}{k^2} t^2 + \frac{2 k_E D }{k}t,\ee
+so it scales as $\sigma_x^2 \sim t^2$ when $t \gg \frac{2Dk}{V^2 k_D}$ and $\sigma_x^2 \sim t$ when $t \ll \frac{2Dk}{V^2 k_D}$.
+Similarly at long times $t\gg k^{-1}$, 
+\be \sigma_x^2 \sim \frac{2k_Ek_D V^2}{k^3}t,\ee
+so we have another non-trivial multi-scale diffusion phenomenon. Provided that $2D/V^2 \ll k_D/k^2$, we therefore have
+\be \sigma_x^2 \sim 
+\begin{cases}
+	t, & t \ll \frac{2Dk}{V^2 k_D}, \\ 
+	t^2, &  \frac{2Dk}{V^2 k_D} \ll t \ll \frac{1}{k}, \\
+	t, & t\gg \frac{1}{k}.
+\end{cases}\ee
+Note in the physical condition when $k\approx k_D$, the condition for the existence of three ranges becomes $ \Pe \ll 1,$ where
+$\Pe = 2 D k_D/V^2.$
+
 \section{Calculation of the flux rate constant}
+
+The central object required for the flux is 
+\be \mu(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t).\ee
+This represents, up to a proportionality factor, the probability that a particle starts at $x_i$, somewhere to the left of $x=0$, then manages to cross $x=0$ by time $t$.
+This can be demonstrated as the rate of the Poisson flux distribution (although it's divided by the density).
+Taking the Laplace transform,
+\be \tilde{\mu}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s).\ee
+This means the starting point is integrating equation \ref{eq:laplace}. Noting that $x+x_i$ is always positive,
+The first integration gives ($\phi = k_D/k$)
+\be \tilde{\mu}(s) = \int_0^\infty dx_i \exp\Big[\frac{V(1-R)x_i}{2D}\Big]\Big(\phi \frac{1-R}{2R(s+k_E)} - \frac{\phi}{R(s+k_E)} - \frac{2D(s+k)}{V^2R(1-R)(s+k_E)}\Big), \ee
+and the second integration gives
+\be  \tilde{\mu}(s) = -\frac{\phi D}{V R(s+k_E)} + \frac{2D\phi}{VR(1-R)(s+k_E)} + \frac{4D^2(s+k)}{V^3R(1-R)^2(s+k_E)}. \label{eq:laplacefluxrate}\ee
+The problem is to take the inverse transform of this.
+
+Starting from \ref{eq:laplacefluxrate}, taking the inverse transform, converting the $s$ factor in the numerator of the last term to $\pt + \delta(t)$, and using the shift property gives
+\be \mu(t) = \Bigg\{ - \frac{\phi D}{VR_\star s} + \frac{2D\phi}{VR_\ast(1-R_\star)s} + \frac{4D^2\big( \cev{\pt} + k\big)}{V^3 R_\star(1-R_\star)^2s}\Bigg\}.\ee
+Here the notation $\cev{\pt}$ means the derivative acts from the left on all terms multiplying it.
+Now using the Erdelyi formula again provides
+\be \mu(t) = e^{-k_E t}\int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_D u(t-u)}\Big)K(u)du,\ee
+where 
+\be K(u) = \Bigg\{  - \frac{\phi D}{V\sqrt{a+bs}}+ \frac{2D\phi}{V\sqrt{a+bs}(1-\sqrt{a+bs})} + \frac{4D^2\big( \cev{\pt} + k\big)}{V^3 \sqrt{a+bs}(1-\sqrt{a+bs})^2} \Bigg\}(u)
+\ee
+Using the shift property again provides
+\be K(u) = e^{-au/b} \Bigg\{  - \frac{\phi D}{V\sqrt{bs}}+ \frac{2D\phi}{V\sqrt{bs}(1-\sqrt{bs})} + \frac{4D^2\big( \cev{\pt} + k\big)}{V^3\sqrt{bs}(1-\sqrt{bs})^2} \Bigg\}(u) \ee
+The three transforms within are (from mathematica):
+\be \Bigg\{ - \frac{\phi D}{V\sqrt{bs}} \Bigg\}(u) = -\frac{\phi}{2}\sqrt{\frac{D}{\pi u}} \ee
+\be \Bigg\{\frac{2D\phi}{V\sqrt{bs}(1-\sqrt{bs})} \Bigg\}(u) = -\frac{k_D V}{2k}e^{V^2u/4D}\erfc\Bigg(-\sqrt{\frac{V^2u}{4D}}\Bigg) \ee
+\be \Bigg\{  \frac{4D^2\big( \cev{\pt} + k\big)}{V^3\sqrt{bs}(1-\sqrt{bs})^2} \Bigg\}(u) = \big(\cev{\partial}_t+k\big)\Bigg[\sqrt{\frac{Du}{\pi}} + \frac{Vu}{2}e^{V^2 u/4D}\erfc\Bigg(-\sqrt{\frac{V^2u}{4D}}\Bigg)\Bigg] \ee
+so the net result is
+\begin{multline} 
+	\mu(t) = \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
+	\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{k_D}{2 k}\Big)e^{-V^2 u/4D} + \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du.
+\end{multline}
+Evaluating the derivatives and doing algebra gives finally
+\begin{multline}	\mu(t) = 
+	e^{-k_D t} \Bigg\{ \sqrt{\frac{Dt}{\pi}} e^{-V^2t/4D} + \frac{Vt}{2} \erfc\Bigg(-\sqrt{\frac{V^2 t}{4D}}\Bigg)\Bigg\} \\
+	+ e^{-k_E t} \int_0^t \mathcal{I}_0 \Big( 2\sqrt{k_E k_D u(t-u) } \Big) e^{-(k_D-k_E)u} \Bigg(\Big(k_D u-\frac{k_D}{2k}\Big) \sqrt{\frac{D}{\pi u}}e^{-V^2u/4D} + \frac{Vk_D}{2}(u-k^{-1})\erfc\Bigg[-\sqrt{\frac{V^2 u}{4D}}\Bigg] \Bigg)du
+	\\ 
+	+ e^{-k_E t} \int_0^t e^{-(k_D-k_E)u} \sqrt{\frac{k_Ek_Du}{t-u}}\mathcal{I}_1\Big( 2\sqrt{k_E k_D u(t-u) } \Big) \Bigg(\sqrt{\frac{Du}{\pi}}e^{-V^2u/4D} + \frac{Vu}{2}\erfc\Bigg[-\sqrt{\frac{V^2 u}{4D}}\Bigg] \Bigg)du
+\end{multline}
+This one was hard work... that $k_D/k$ term multiplying $\mci_0$ was a pain and depends sensitively on getting the correct initial conditions at the starting point.

### 2021-07-06 ch1intro.tex
chapter 1 is flawless. gradually changing the notation and rewriting chapter 2. really getting bogged down on it.

@@
 
-To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics and drawing upon a wide interdisciplinary literature.
+To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
 To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes and show continuity with my own developments which follow in subsequent chapters.
@@
 \section{Theories of individual particle movement}
+\label{sec:trajmodels}
 
-A basic problem in sediment transport theory is to predict the downstream movement of an individual particle.
+A basic problem in sediment transport theory is to predict the downstream movement of individual particles.
 At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
-Particles moving as bedload are driven downstream by a turbulent fluid flow \citep{}, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Coleman1967,Schmeeckle2007}.
-Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Martin2013}.
+Particles moving as bedload are driven downstream by a turbulent fluid flow, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Maxey1983,Schmeeckle2007}.
+Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.
 
 Entrainment and deposition provide an additional layer of complexity.
-Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and induce their deposition \citep{Charru2004}.
+Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and cause deposition \citep{Charru2004}.
 These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2010,Celik2014}.
-As a result, particles generally alternate through sequences of entrainment and deposition, but the exact times at which these alternations occur is impossible to characterize.
+As a result, particles generally alternate through sequences of entrainment and deposition, but the exact times at which these alternations occur is difficult to characterize \citep{Einstein1937}.
 
@@
 
-Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales, from seconds to years, and each of these processes eludes precise description.
+Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales, from seconds to years \citep{Pretzlav2021}, and each of these processes defies exact description.
 
@@
 The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
-\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017,Pretzlav2021}.
+\citet{Hassan2017} added an additional range, referred to as ``geomorphic", to reflect the even longer period over which particles become buried within the subsurface or embedded among sedimentary deposits \citep{Bradley2017}.
 This set of timescales -- local, intermediate, global, and geomorphic -- are used below to organize the literature describing trajectories of individual grains.
@@
 In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales.
-Apparently, tracer particle velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
+Apparently, tracer velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
 This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
@@
 \subsection{Einstein's random walk model}
-Einstein was probably the first to focus on the movements of individual particles through streams \citep{Einstein1937}.
+\label{sec:einwalk}
+Einstein was probably the first to model the movements of individual particles through streams \citep{Einstein1937}.
 Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to probabilistic methods to characterize their transport.
 Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
-As his interest was on the global range of particle motion, and the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
-With this configuration, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
+As his interest was on the global range of particle motion when many motion-rest alternations have occurred, and the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
+With this approximation, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
 Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
@@
 
-Einstein's assumptions mean that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
+Einstein's assumptions imply that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
 \be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
-where $N(t)$ is the number of particle entrainments in time $t$, distributed as a Poisson distribution $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
+where $N(t)$ is the number of particle entrainments in time $t$, a Poisson random variable distributed as $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
 Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
@@
 This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1992a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
-\be P(x,t) = \delta(x) e^{-k_E t} + e^{-k_E t - x/\ell}\theta(x) \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big) \ee
-Here, $\mathcal{I}$ is a modified Bessel function. 
-The probability distribution eq. \ref{eq:eindist} fully characterizes the dynamics of an individual particle alternating through steps and rests.
+\be P(x,t) = \Big[\delta(x) e^{-k_E t} + e^{-k_E t - x/\ell} \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big)\Big]\theta(x)\theta(t) \label{eq:eindist} \ee
+Here, $\mathcal{I}_\nu$ is a modified Bessel function of order $\nu$. 
+The probability distribution eq. \ref{eq:eindist} fully characterizes the downstream movement of an individual particle alternating randomly through steps and rests.
 
-This distribution displays both advective and diffusive characteristics.
+This distribution displays both advection and diffusion.
+Advection occurs because particles move downstream.
+Diffusion occurs because step lengths and instants when entrainment events occur vary from one particle to the next.
 One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
 \be \langle x \rangle (t) = k_E \ell t, \ee
-so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$, given by the entrainment rate times the mean step length.
-The rate at which one particle spreads out from another due to differences in their trajectories can be represented by the variance, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$.
-This gives
+so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$ given by the product of the entrainment rate and mean step length.
+The rate at which particles spread apart due to differences in their trajectories can be represented by the variance of position, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$. This gives
 \be \sigma_x^2(t) = 2 k_E \ell^2 t, \ee
@@
 Einstein's model provides an adequate description of global range particle transport when the period of interest is much larger than the timescales of individual particle movements (local and intermediate ranges) and much smaller than the timescales over which particles become embedded in sedimentary deposits (geomorphic range).
+
 The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
@@
 In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
-Einstein's infinite movement velocity assumption excludes the timescales over which these processes occur.
+Einstein's instantaneous movement assumption obscures the timescales over which these processes occur.
 
-Studies by \citet{Gordon1972}, \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include these timescales.
-They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are also (along with rests) exponentially distributed random variables, this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion (between entrainment and deposition).
+Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales.
+They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
 The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
-\be \dot{x} = V\eta(t), \ee
-where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph-type signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
+\be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
+where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
 This noise is displayed in figure \ref{fig:lislefig} panel (a).
-Some particle trajectories solving equation \ref{eq:lislelangevin} are displayed in \ref{fig:lislefig} panel (b).
+Some particle trajectories solving equation \ref{eq:lislelangevin} are displayed in figure \ref{fig:lislefig} panel (b).
 
 The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt') \rangle$ becomes \citep{Balakrishnan1993}
-\be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\ee
-where $k_E$ and $k_D$ are the entrainment and deposition rates, $V$ is the particle velocity during the motion phase, and $k = k_E+k_D$.
+\be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\label{eq:lislemaster}\ee
+where $k_E$ and $k_D$ are the entrainment and deposition rates, $V$ is the particle velocity during the motion state, and $k = k_E+k_D$.
 This partial differential equation is called an asymmetric telegrapher's equation \citep{Rossetto2018}, and although the symmetric analogue of this equation is well-studied \citep{Weiss2002a, Masoliver2017}, the asymmetric problem is not often encountered.
@@
 
-For the intitial condition that particles have a probability $k_E/k$ to start in motion, the solution of equation \ref{eq:telegraph} is \citep{Lisle1998}
-\be  P(x,t) =  \ee
-Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x \rangle (t) = k_E V t/k$, representing movement at effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion multiplied by the velocity during the motion phase.
-The variance picks up a two range scaling:
+For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of equation \ref{eq:lislemaster} is \citep{Lisle1998}
+\begin{multline}  P(x,t) =  e^{-\chi-\tau} \Big[ \frac{k_E}{V}\delta(\tau) + \frac{k_E}{V} \sqrt{\frac{\chi}{\tau}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_D}{V}\mathcal{I}_0(2\sqrt{\chi\tau}) \\
+	+ \frac{k_Ek_D}{kV}\sqrt{\frac{\tau}{\chi}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_Ek_D}{kV} \delta(\chi)
+	\Big]\theta(\chi)\theta(\tau), \label{eq:lisledist}
+\end{multline}
+where $\chi = k_D x/V$ and $\tau = k_E(t-x/V)$ are shorthand notations.
+Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with an effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state.
+In contrast, the variance develops a two range scaling.
 Computing $\sigma_x^2$ provides
-\be \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) ,\ee
+\be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) , \label{eq:lislevar}\ee
 which is a non-trivial result.
@@
 At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
-In this model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, since fluctuations of the velocity between entrainment and deposition have been neglected.
+The rapid spreading of particles at short timescales occurs because some are completely stationary while others move with velocity $V$.
+The eventual transition to normal diffusion occurs because particles eventually become well-mixed among motion and rest states.
+An analogous diffusion phenomenon is described in \citet{Taylor1920}.
+
+In this trajectory model, both intermediate and global ranges are adequately represented, but local and geomorphic are not, the former since fluctuations of the velocity between entrainment and deposition have been neglected, and the latter since particle burial was not incorporated.
 
@@
 
-Some authors have attempted to model local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
-Early efforts impart particles with time-averaged fluid forces, typically linked to a logarithmic flow velocity profile \citep{VanRijn1984}, and some include simplified collision forces that modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992}.
-Later on, researchers began to include granular interactions among particles to model collisions using the discrete element method \citep{Cundall1979a, Haff1993}.
-The early works utilizing this approach used a two dimensional domain with a highly simplified flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+Some authors have attempted modelled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
+Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included heuristic collision forces to modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992,Nino1998}.
+
+Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep{Cundall1979a, Haff1993}.
+The early works utilizing this approach used a two dimensional domain with a highly simplified ``slab" flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+
 The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow.
-The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, using particles as the boundary condition for the fluid \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
-A next step in this inquiry is to include non-spherical particles, and the foundation for this inquiry is now built \citep{Wachs2012, Azema2012, Wachs2019}.
-These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are desired for further insight into the problem.
+The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, while particles provide the boundary condition for the flow \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
+These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
 
@@
 
-Some models have been developed to describe particle velocities in the local and intermediate ranges when velocities fluctuate due to turbulence and particle-bed collisions.
-Experimental studies on bedload velocities have provided two dominant conclusions.
+Some approaches have described particle velocities in the local and intermediate ranges when velocities fluctuate due to turbulence and particle-bed collisions using Langevin equations, whereby particles are driven by Gaussian white noise \citep{Kubo1978}.
+Experimental studies on bedload velocities have provided two major conclusions for the shape of the particle velocity distribution.
 One subset of observations indicates that bedload velocities follow exponential distributions \citep{Lajeunesse2010,Furbish2012,Fathel2015}, and another subset indicates Gaussian distributions \citep{Martin2012,Ancey2014,Heyman2016}.
@@
 	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d).
-The latter has relatively narrow and symmetric fluctuations, with the mode well above $u=0$.}
+The latter has relatively narrow and symmetric fluctuations, with a positive mode well above $u=0$.}
 	\label{fig:fanAncey}
 \end{figure}
+Although these descriptions successfully describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying reason why bedload transport has shown multiple velocity distributions remains an open question.
+A comprehensive model of the full range of bedload transport observations is a desirable target.
 
@@
 Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
-These two definitions, $q$ defined with a surface and $\Phi$ defined with a volume, are generally not equivalent \citep{Ancey2020}.
+These two definitions, $q$ defined with a surface and $\Phi$ defined with a volume, are generally not equivalent \citep{Ancey2020a}.
 
@@
 
-One of the most influential formulations of the bedload flux is due to Bagnold \citep{Bagnold1956,Bagnold1966}, who derived a formula for the mean sediment flux using an energy balance approach.
+One of the most influential formulations of the bedload flux is due to \citet{Bagnold1956,Bagnold1966}, who derived a formula for the mean sediment flux using an energy balance approach.
 Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
-He assumed that the flow power $P_f$ available to move sediment in some volume scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
-Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ in a volume required to sustain particle motion scales as $P_g \propto q/\tau^{1/2}.$
+He assumed that the flow power per unit volume $P_f$ available to move sediment scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
+Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ per unit volume required to sustain particle motion scales as $P_g \propto \Phi/\tau^{1/2}.$
 Balancing flow energy against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
@@
 The large shear stress limit $\Phi \sim \tau^{3/2}$ of Bagnold's formula is shared in common with many other empirical formulas describing the mean downstream flux of bedload \citep[e.g.][]{MeyerPeter1948, Yalin1972, Wilcock2003, Parker1990}.
-The distinguishing feature of Bagnold's formula is its derivation from mechanical principles, although many of Bagnold's assumptions have since turned out to be incorrect.
+A distinguishing feature of Bagnold's formula is its derivation from mechanistic reasoning, although many of Bagnold's assumptions have since turned out to be incorrect.
 Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and the flow power dissipated by sediment transport shows only a weak correlation the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
 These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$ and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
-More generally, Bagnold's formulation faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}.
-The difficulties with Bagnold's approach led to many revisions of his theory which kept Bagnold's physical reasoning, but includes more correct details of energetics of sediment transport \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
+Bagnold's formulation also faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}.
+The difficulties with Bagnold's approach led to many revisions of his theory, often based on more carefully incorporating the properties of individual particle motions into the sediment transport energy budget \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
 
@@
 Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
-Einstein formulated his original model of individual grains in transport \citep{Einstein1937} in terms of particle entrainment and deposition using the conceptual picture of grains moving downstream through a sequence of instantaneous steps.
-Later, \citet{Einstein1942,Einstein1950} calculated the bulk sediment flux with these same probabilistic ideas, providing an alternative to the Bagnold scaling approach.
+\citet{Einstein1937} formulated his original model of individual particle trajectories in terms of entrainment and deposition rates using the idealization of grains moving downstream through a sequence of instantaneous steps.
+Later, \citet{Einstein1942,Einstein1950} calculated the sediment flux with these same probabilistic ideas, providing an alternative to Bagnold's scaling approach.
 The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
@@
 	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
-left to right through an array of adjacent control volumes. The bedload flux is the rate of bedload particles crossing
+left to right through an array of adjacent control volumes. The bedload flux is the rate of particles crossing
 the surface $S$ per unit width and time, collected from all upstream control volumes.}
@@
 
-Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$ from a geometric series, but this approach has been criticized by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
-Some of these authors have argued convincingly that instead, one should calculate $P_n$ as an exceedance probability.
-If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$!) is a Poisson distribution \citep{Cox1965}:
+Einstein originally constructed this probability by assuming that each particle had $n$ independent entrainment opportunities in the period $T$, each with probability $p$, so that $P_n = p^n$, giving $q \propto p/(1-p)$ by a geometric series, but this approach has been criticized by \citet{Yalin1972} and others \citep{Paintal1971,Cheng2004,Armanini2015,Armanini2017}.
+Some authors have argued convincingly that instead, one should calculate $P_n$ as an exceedance probability.
+The central critique is that there is a distinct different between \textit{exactly} and \textit{at least} $n$ entrainment events in $T$.
+If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$) is a Poisson distribution \citep{Cox1965}:
 \be p_n = \frac{(k_E T)^n}{n!}e^{-k_E T}.\ee
@@
 
-Alongside this central result $q=E\ell$ for the mean flux, Einstein formulated the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle.
-The original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
-The entrainment rate is calculated from the exceedance probability of the turbulent lift over the weight, providing an alternative to the critical shear stress concept which explicitly incorporates turbulent fluctuations in the fluid flow.
-This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
-random) resisting quantities \citep{Grass1970} another of Einstein's key contributions.
-\citet{Paintal1971} made a significant extension
-by including the random supporting geometry of bed particles into the force balance.
-More recently, \citet{Tregnaghi2012}
-amended the theory to include both force magnitude and duration, incorporating the recognition that both are relevant for particle entrainment \citep{Diplas2008, Valyrakis2013, Celik2014}.
-Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.
+Einstein's formulation requires expressions of the single particle entrainment rate $k_E$ and step length $\ell$ in terms of the flow and sediment characteristics.
+Although Einstein and many followers attempted to formulate these connections \citep[e.g.][]{Einstein1950,Grass1970,Paintal1971,Tregnaghi2012,Dey2018}, there are still no comprehensive solutions.
 
@@
 
-Although Einstein worked to relate the single-particle entrainment rate $k_E$ to the flow, other parameters of Einstein's model ($\rho$, $\ell$) retain a heuristic character and are not clearly linked to any underlying fluid-granular physics.
-The erosion-deposition model originally developed by \citet{Charru2004,Charru2006} modifies the Einstein model and relates its parameters to properties of the flow using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2018}.
-This can be characterized as a mixture of the Einstein and Bagnold strategies.
+One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}.
+Their ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's erosion rate $E$ and the complementary deposition rate $D$.
+The model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2018}, in what can be characterized as a mixture of the Einstein and Bagnold strategies.
 
@@
 \be \pt \gamma +  \px V \gamma = E - D. \label{eq:charru}\ee
-This is derived by evaluating a mass balance within a control volume.
-In this equation, $\gamma$ is the ``particle activity", which is the number of moving particles per unit area \citep{Furbish2012}, $V$ is the ensemble averaged movement velocity of sediment grains (which in unsteady conditions may depend on space and time), $E$ is the areal entrainment rate (the number of particles transitioning to motion per unit area and time), and $D$ is the areal deposition rate (the number of particles coming to rest on the bed per unit area and time).
+In this equation, $\gamma$ is the ``particle activity", which is the number of moving particles per unit area \citep{Furbish2012}, $V$ is the ensemble averaged movement velocity of sediment grains (which in unsteady conditions may depend on space and time), $E$ is the areal entrainment rate (the number of particles transitioning into motion per unit area and time), and $D$ is the areal deposition rate (the number of particles coming to rest on the bed per unit area and time).
 
@@
 \be V = c + d(\sqrt{\tau}-\sqrt{\tau_c}). \label{eq:charru2}\ee
-The constant coefficients $a$, $b$, $c$, and $d$ are determined experimentally.
+The constant coefficients $a$, $b$, $c$, and $d$ are calibrated in experiments.
 
@@
 \subsection{The nonlocal formulation}
+\label{sec:nonlocal}
 
-Einstein's model of the bulk bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009,Tucker2010a,Martin2012}.
+Einstein's model of the bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009,Tucker2010a,Martin2012}.
 \citet{Nakagawa1976} and later \citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
 \be q(x,t) = \int_0^\infty dx' F(x-x',t)E(x',t). \ee
-In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a just-entrained particle steps a distance $x$ before deposition.
+In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a particle entrained at $t=0$ steps a distance $x$ before deposition at time $t$.
 In general, this approach can handle non-uniform conditions by including space and time dependence in $E$ and $F$.
@@
 \textit{at least} a distance $x$ in $t$ is
-\be F(x,t) = \int_0^t \delta(x-Vt)k_D\exp(-k_D t) = \theta(Vt-x)k_D\exp(-k_D t).\ee
-Considering uniform conditions with a density $\rho$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the bulk entrainment rate can be expressed as $E=\rho k_E$, and equation \ref{eq:furbo} provides a mean flux
-\be q = \rho k_E \int_0^\infty dx' \int_0^\infty dt' \theta(Vt-x)k_D\exp(-k_D t) = \rho k_E V/k_D. \label{eq:furbflux}\ee
+\be F(x,t) = \int_x^\infty \delta(x-Vt)k_D\exp(-k_D t) dx = \theta(Vt-x)k_D\exp(-k_D t).\ee
+Considering uniform conditions with a density $\rho_s$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the areal entrainment rate can be expressed as $E=\rho_s k_E$, and equation \ref{eq:furbo} provides a mean flux
+\be q = \rho_s k_E \int_0^\infty dx' \int_0^\infty dt \theta(Vt-x)k_D\exp(-k_D t) = \rho_s k_E V/k_D. \label{eq:furbflux}\ee
 Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$.
- Equation \ref{eq:furbflux} therefore provides another perspective on Einstein's central result $q= E\ell$.
+Equation \ref{eq:furbflux} therefore provides another perspective on Einstein's central result $q= E\ell$, but with the added possibility of describing unsteady conditions \citep{Furbish2012}.
 
@@
 This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
+The major limitation of these approaches are that they assume the landscape is a continuous field.
+This can be interpreted as an average over the detailed configurations of individual grains \citep{Coleman2009}.
+This averaging process is expected to obscure the physics whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Schobe2021}.
 
@@
 
-The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random variable.
-They derived the probability distribution of the flux by counting the number of moving particles in a control volume. 
-This number changes through time as a result of entrainment and deposition.
+The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random process.
+They derived the probability distribution of the flux by counting the number of moving particles in a control volume, considering that this number changes through time as a result of particles migrating into the volume from upstream, entraining, depositing, and migrating out of the volume to downstream.
 To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
@@
 Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where entrainment is ``birth" and deposition is ``death".
-They formulated the probability mass function $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
-\be \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) - [n+1]\sigma P(n+1,t) - \big[\lambda + n (\sigma + \mu)\big],\ee
-whose terms describe entrainment at rate $\lambda$, collective entrainment at rate $\mu$, and deposition at rate $\sigma$. 
-The final term includes the possibility that $n$ stays constant.
-These coupled equations (one for each $n=0,1,\dots$) can be solved by generating functions \citep{Cox1965,Ancey2008}, providing the steady-state distribution
+They formulated the probability distribution $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
+\be \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) - [n+1]\alpha P(n+1,t) - \big[\lambda + n (\alpha + \mu)\big],\ee
+whose terms describe particle birth (entrainment and migration in) at rate $\lambda$, collective entrainment at rate $\mu$, and particle death (deposition and migration out) at rate $\alpha$.
+The final term encodes the possibility that $n$ stays constant.
+These coupled equations (one for each $n=0,1,\dots$) can be solved by generating functions \citep{Cox1965}, providing the steady-state distribution
 \be P(n) = \frac{\Gamma(r+n)}{\Gamma(r) n!}p^r(1-p)^n, \label{eq:negbin}\ee
-where $r = \lambda/\mu$ and $p = 1-\mu/\sigma$.
+where $r = \lambda/\mu$ and $p = 1-\mu/\alpha$.
 This is a negative binomial distribution, which is a wide-tailed generalization of the Poisson distribution.
 
-From eq. \ref{eq:negbin}, the mean number of moving particles is $\langle n \rangle = \lambda/(\sigma-\mu)$, and the variance is
-\be \sigma_n^2 = \frac{\lambda \sigma}{(\sigma-\mu)^2}. \ee
-Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\sigma/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to 1, as the distribution \ref{eq:negbin} limits to a Poisson distribution \citep{Ancey2006}.
+From eq. \ref{eq:negbin}, the mean number of moving particles is $\langle n \rangle = \lambda/(\alpha-\mu)$, and the variance is
+\be \sigma_n^2 = \frac{\lambda \alpha}{(\alpha-\mu)^2}. \ee
+Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\alpha/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, as the distribution \ref{eq:negbin} limits to a Poisson distribution where the mean and variance are equal \citep{Ancey2006}.
 
-The probability distribution of the bedload flux can be computed by the control volume form, eq.
-\ref{eq:controlflux} given the probability distribution of particle velocites, which can be determined experimentally \citep{Lajeunesse2010,Fathel2015,Heyman2016} or calculated using the Langevin models described in section \ref{sec:langevin}.
-Assuming that the particle activity in the control volume and the velocities of moving particles are completely independent, which may be a good assumption for a large enough control volume, the probability distribution $Q(q)$ of the flux is \citep{Ancey2008,Ancey2014,Ancey2020}
-\be Q(q) = L \sum_{k=1}^\infty P(k)G(U) ,\ee
-where $P(n)$ comes from eq. \ref{eq:negbin} and $G(U)$ is the probability that the sum of $k$ particle velocities at some instant is equal to $U$.
+The probability distribution of the bedload flux can be computed by the control volume form eq.
+\ref{eq:controlflux} given the probability distribution of particle velocites $P(u)$, if one assumes velocity and activity are independent.
+Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep[e.g.][]{Ancey2014}
+\be P(\Phi) = L \sum_{k=0}^\infty P_k(L\Phi) P(n=k). \ee
+One limitation of this approach is that in reality, entrainment and deposition modify the bed elevation, which in turn modifies the rates of entrainment and deposition \citep{Sawai1987,Wong2007}. 
+This highlights a need to evaluate bed elevation change in a birth-death sediment transport model.
 
@@
 \label{sec:renewal}
-The final approach to summarize that of calculating the sediment flux probability distribution from an ``arrival time distribution" representing the statistics of times between subsequent arrivals of particles to a control surface \citep{Turowski2010,Ancey2020}.
-This distribution changes depending on the observation time $T$ used to calculate the averaged flux.
-This is the phenomenon of scale dependence.
-The statistics machinery of counting arrivals is called renewal theory \citep{Cox1962}.
+The final bedload transport model to summarize calculates the sediment flux probability distribution from an ``arrival time distribution" representing the statistics of times between subsequent arrivals of particles to a control surface \citep{Turowski2010, Heyman2013,Ancey2020}.
+This distribution changes depending on the observation time $T$ used to calculate the averaged flux, characteristic of scale dependence.
+In statistics, the problem of counting arrival events is called renewal theory \citep{Cox1962}.
 
@@
 \end{figure}
-Earlier studies have considered two different arrival time distributions to calculate the probability distribution of the time-averaged sediment flux, but only one of these is reviewed here.
-This is an exponential distribution: $P(t) = \Lambda \exp(-\Lambda t),$ so the mean time between particle arrivals is $1/\Lambda$.
-In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of section \ref{sec:einstein}:
+Earlier studies have considered different arrival time distributions to calculate the probability distribution of the time-averaged sediment flux, but only the simplest exponential case is reviewed here.
+The exponential inter-arrival time distribution is $P(t) = \Lambda \exp(-\Lambda t),$ so the rate of particle arrivals is $\Lambda$.
+In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of section \ref{sec:einwalk}:
 \be q(T) = \frac{1}{T}\int_0^T dt' \mu(t'). \label{eq:flucflux}\ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
-as indicated in figure \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda t$. 
+as indicated in figure \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$. 
 The flux in eq.
-\ref{eq:flucflux} is a random variable as indicated in \ref{fig:ancey} panel (b).
-Its probability distribution which is contingent on the observation time $T$ can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
-This equation entails an average over all possible realizations of the noise in eq.
-\ref{eq:poissfluxnoise}, producing \citep{VanKampen2007}
+\ref{eq:flucflux} is a random variable as indicated in figure \ref{fig:ancey} panel (b).
+Its probability distribution $P(q|T)$, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
+This equation entails an average over all possible realizations of the noise in eq. \ref{eq:poissfluxnoise}, producing \citep{VanKampen2007}
 \be P(q|T) = \sum_{l=0}^\infty \frac{\big(\Lambda T\big)^l}{l!}e^{-\Lambda T}\delta(q-\frac{l}{T}),\ee
-which is a Poisson distribution.
+which is a scale-dependent Poisson distribution for the sediment flux across the control surface.
 
-The mean flux derived from this scheme ($\langle q \rangle(T) = \int_0^\infty dq P(q|T)$ is
-\be \langle q \rangle = \Lambda, \ee
-which is independent of observation scale, while the magnitude of bedload transport fluctuations scales with $1/T$: 
-\be \sigma_q^2 = \frac{\Lambda^2}{T}.\ee
-Even this simple model gives a non-trivial conclusion that the relative uncertainty in a measurement of bedload transport depends on observation time: $\sigma_Q/\langle q \rangle \propto T^{-1/2}.$
-The limitation of these approaches is that they do not obviously relate to the dynamics of individual particles. In renewal models, the distribution of arrival times is heuristic.
+The mean flux derived from this renewal scheme -- $\langle q (T)\rangle = \int_0^\infty dq P(q|T)$ -- is
+\be \langle q (T) \rangle = \Lambda, \ee
+which is independent of the observation scale, while the magnitude of bedload transport fluctuations scales with $1/T$: 
+\be \sigma_q(T)^2 = \frac{\Lambda^2}{T}.\ee
+Even this simple model gives a non-trivial conclusion that the relative uncertainty in a measurement of bedload transport depends on the observation time: $\sigma_q(T)/\langle q(T) \rangle \propto T^{-1/2}.$ The uncertainty diverges as the sampling time becomes short.
+
+The limitation of renewal approaches to calculate the flux is that they do not obviously relate to the dynamics of individual particles.
+In renewal models as they are phrased now, the inter-arrival time distribution is not mechanistic.
 
@@
 
-This literature review has indicated that successful bedload transport descriptions have been developed to describe individual particles and bulk bedload fluxes.
+This literature review has indicated that successful bedload transport descriptions have been developed to describe individual particles and bedload fluxes, although many limitations are left in need of further research attention.
+
 All of the works reviewed so far are either mean field models that do not include fluctuations, or stochastic models that calculate probability distributions for the quantities of interest.
-The common thread of all of these models is that they take advantage of various devices to side-step the complex interplay of turbulence and granular physics which control bedload transport.
+The common thread of all of these models is that they take advantage of various devices to side-step the complex interplay of turbulence and granular physics which ultimately control bedload transport.
+
 In the case of mean field models, the devices are heuristic scaling arguments and semi-empirical formulas, while in the case of stochastic models, they are simplified dynamical equations driven by idealized noises, like sequences of pulses, alternating switches, and erratic fluctuations.
-This latter methodology is applied throughout the thesis.
+
+This latter methodology is applied throughout the thesis to many of the limitations highlighted during the above review. A detailed outline of the problems to be addressed follows below.
 
@@
 
-The original model by \citet{Einstein1937} describing particle trajectories with infinite movement velocity was subsequently improved to include a constant velocity \citep{Gordon1972,Lisle1998,Lajeunesse2017}.
-Particle velocities fluctuate during movements between entrainment and deposition, so this should also be accounted for in bedload trajectory models.
-In addition, the scale-dependence of the sediment flux has been calculated from renewal models \citep{Turowski2010,Ancey2020}, but these overlook the details of individual particle motions, so it remains unclear how the sediment flux distribution and its scale dependence originate from individual particle trajectories.
+The original model by \citet{Einstein1937} describing particle trajectories with instantaneous movements was subsequently improved to include a finite motion time with a constant velocity \citep{Lisle1998,Lajeunesse2017}.
+In reality, particle velocities fluctuate during movements between entrainment and deposition, so there is a need to go one step further and account also for velocity fluctuations of moving particles in Einstein-type models.
 
-Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories which includes velocity fluctuations in the motion phase, then demonstrates how these trajectories can be used to calculate scale-dependent probability distribution of the bedload flux, building up and linking together all of these earlier studies.
+In addition, although the scale-dependent sediment flux has been calculated from renewal models \citep{Turowski2010,Ancey2020}, these approaches are not based on the dynamics of individual particles in transport, so it remains unclear how the sediment flux distribution and its scale dependence originate from the transport characteristics of individual grains.
+
+Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories which includes velocity fluctuations in the motion state.
+The chapter then demonstrates how these trajectories can be used to calculate the scale-dependent probability distribution of the bedload flux, linking together studies from \citet{Einstein1937} to \citet{Ancey2020b} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
 
@@
 
-Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Heyman2013,Ancey2014}, which violates conservation of mass and the Exner equation \citep{Exner1925,Tsujimoto1978}, and precludes the possibility that sediment can become buried within the bed.
-Although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2002,Hassan2017}, our understanding of how long sediment can remain buried is limited.
+Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Heyman2013,Ancey2014}, which violates conservation of mass.
 
-Chapter \ref{ch:ch3} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movements characteristics of grains.
+Bed elevation changes produce sediment burial, and although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2002,Hassan2017}, our understanding of how long sediment can remain buried is limited.
+
+Chapter \ref{ch:ch3} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movement characteristics of individual grains.
 
@@
 
-The model of sediment transport as an alterating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface, and it provides a useful description of bedload transport on the intermediate and global timescales of \citet{Nikora2001,Nikora2002}.
-Tracers become buried, and this affects their motion characteristics at global and geomorphic timescales \citep{Hassan2017}.
-These phenomena are not yet included in Einstein-type models that treat individual sediment trajectories as a random walk \citep[e.g.][]{Lisle1998}.
+Modelling sediment transport as an alterating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface and provides a useful description of bedload transport across intermediate and global timescales, but it does not resolve local or geomorphic timescales.
 
-Chapter \ref{ch:downDiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial to develop a description of sediment trajectories over intermediate, global, and geomorphic timescales.
+At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017}, but burial has not yet been incorporated into Einstein-type models of particle trajectories.
+
+Chapter \ref{ch:downDiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial.
+This provides the first description of sediment trajectories over local, intermediate, global, and geomorphic timescales.
 
@@
 
-Finally, the Langevin models of \citet{Fan2014} and \citet{Ancey2014} describe the velocity distributions of sediment moving downstream in the local and intermediate ranges in between entrainment and deposition.
-These models describe two different end-member distributions from among the experimental record on bedload velocities. One develops exponential velocities, and the other develops Gaussian velocities.
-Since the forces included in these models are phenomenological, we can assume the issue is that particle-bed collisions are not adequately represented.
+Finally, the Langevin models of \citet{Fan2014} and \citet{Ancey2014} describe the velocity distributions of sediment moving downstream in the local and intermediate ranges between entrainment and deposition.
+
+These models describe two different end-member sediment velocity distributions, but our understanding of why these different distributions arise is incomplete, and no mathematical models have been developed that describe the full range of distributions observed in experiments.
 
 Chapter \ref{ch:langevin} presents a stochastic Langevin model including episodic particle-bed collisions formulated by analogy with the theory of granular gases \citep{Brilliantov2004}.
-This produces a satisfying description of all different bedload velocity distributions which have been reported in experimental studies.
+The episodic representation improves on the static Coulomb-like friction representing collisions in earlier models.
+The improved model is demonstrated to describe all of the different bedload velocity distributions which have been reported in experiments.
 
@@
 Despite the considerable progress these approaches have brought, the lack of connection to the underlying particle mechanics requires attention.
+
+Alongside this central result $q=E\ell$ for the mean flux, Einstein formulated the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle, building insight into the grain-scale mechanics of entrainment.
+His original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
+The entrainment rate is calculated from the exceedance probability of the turbulent lift over the weight, providing an alternative to the critical shear stress concept which explicitly incorporates turbulent fluctuations in the fluid flow.
+This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
+random) resisting quantities \citep{Grass1970} another of Einstein's key contributions.
+\citet{Paintal1971} made a significant extension
+by including the random supporting geometry of bed particles into the force balance.
+More recently, \citet{Tregnaghi2012}
+amended the theory to include both force magnitude and duration, incorporating the recognition that both are relevant for particle entrainment \citep{Diplas2008, Valyrakis2013, Celik2014}.
+Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.

### 2021-07-06 ch2flipFlop.tex
chapter 1 is flawless. gradually changing the notation and rewriting chapter 2. really getting bogged down on it.

@@
 Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, particles are once again entrained, and the cycle repeats.
-Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and rest.
-This process is difficult to describe mathematically given the technicality of the stochastic physics involved \citep{Furbish2017,Ancey2020}.
+Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and periods of rest.
+This process is difficult to describe mathematically given the technical stochastic physics required \citep{Furbish2017,Ancey2020}.
 
@@
 The foundational work is due to Einstein, who considering bed load motions as instantaneous so he could describe bed load transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
-Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E l$, where $k_E$ is the rate at which an individual bed particle entrains into motion, and $l$ is the mean length of each downstream step.
+Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E \ell$, where $k_E$ is the rate at which an individual bed particle entrains into motion, and $\ell$ is the mean length of each downstream step.
 Later, Einstein applied these ideas to calculate the mean downstream flux due to the movements of many such particles \citep{Einstein1950}.
-Einstein reasoned that if the density of resting particles on the bed is $\rho_b$, the overall areal entrainment rate of particles can be written $E = \rho_b k_E$, giving a mean sediment flux $\bra q_s \ket = \rho_b \bra u \ket  = El $.
+Einstein reasoned that if the density of resting particles on the bed is $\rho_b$, the overall areal entrainment rate of particles can be written $E = \rho_b k_E$, giving a mean sediment flux $\bra q \ket = \rho_b \bra u \ket  = E \ell $.
 
 Many researchers have since refined Einstein's approach to provide more realistic descriptions of individual particle motions than Einstein's instantaneous step model.
-One set of efforts has neglected particle deposition process to calculate the downstream velocity distributions of moving particles after entrainment using mechanistic equations with terms representing turbulent fluid forces and particle-bed collisions \citep{Ancey2014,Fan2014}.
-Another set of efforts has described particles alternating between motion and rest, considering that motions occur with a constant velocity \citep{Lisle1998,Lajeunesse2017,Pierce2020b}. 
-No models have yet been formulated which describe particles alternating between motion and rest, while motions have a realistic fluctuating velocity.
+One set of efforts has neglected the particle deposition process to calculate the downstream velocity distributions of moving particles after entrainment using mechanistic equations with terms representing turbulent fluid forces and particle-bed collisions \citep{Ancey2014,Fan2014}.
+Another set of works have described particles alternating between motion and rest, considering that motions occur with a constant velocity \citep{Lisle1998,Lajeunesse2017,Pierce2020b}. 
+No models have yet been formulated which describe particles alternating between motion and rest with motions having realistic fluctuating velocities.
 
-A set of studies has developed stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Furbish2012}. 
-Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
+Parallel to these efforts to better describe individual particle trajectories, a complementary set of studies have developed stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Turowski2010,Furbish2012a,Ancey2020}.
+Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006,Furbish2012a}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
 Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale-dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
@@
 This survey provides context for the two problems addressed in this chapter.
-First, we do not yet have the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution, and the dependence of its statistical moments on the observation time.
+First, we do not yet have the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution and the dependence of its statistical moments on the observation time.
 
-Here, I develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a Langevin-type dynamical equation in a combination of earlier approaches \citep{Fan2014,Ancey2014,Lisle1998,Lajeunesse2017}.
+Here, I develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a stochastic Langevin equation.
 This dynamical equation describes alternation between motion and rest at random intervals while particles in motion have velocity that fluctuates around a mean value.
 Using the probability distribution of particle position generated by this model, I construct a formalism to derive analytically the probability distribution of the sediment flux.
-This distribution exhibits observation-scale dependence as a result of velocity fluctuations among moving sediment.
+This distribution exhibits observation-scale dependence as a result of velocity fluctuations among moving particles.
 Below, I develop the new formalism in sec.
@@
 \section{Description of motion-rest alternation with velocity fluctuations \label{sec:mod}}
-Consider an infinite one-dimensional domain populated with sediment particles on the surface of a sedimentary bed.
-Consider that the flow is weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common.
-The flow is in contrast strong enough so that particles move.
-Label the downstream coordinate as $x$, so that the downstream velocity of a moving particle is $\dot{x}$, and we describe all sediment particles as independent from one another but governed by the same underlying dynamical equations, meaning we neglect any influence of sediment size or shape or spatial variations in the overlying fluid flow.
+The starting point for this analysis is an idealized one-dimensional domain, infinite in extent, populated with sediment particles on the surface of a sedimentary bed.
+Particles are set in motion by the turbulent flow and move downstream until they deposit, and the cycle repeats.
+The downstream coordinate is $x$, so that $\dot{x}$ describes a velocity in the downstream direction.
+
+The flow is considered weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common, characteristic of rarefied transport conditions \citep[e.g.][]{Kumaran2006,Furbish2017}.
+Any variations like particle size, shape, or spatial variations in the overlying fluid flow are neglected for mathematical simplicity.
+These conditions allow for all particles to be described as independent from one another but governed by the same underlying dynamical equations.
 
 \subsection{Dynamical equation for grain-scale sediment transport}
-From these assumptions, our first target is to write an equation of motion for the individual sediment particle encompassing two features.
+From these assumptions, the first target is to write an equation of motion for the individual sediment particle encompassing two features.
 First, particles should alternate between motion and rest.
 The transition rate from rest to motion is called entrainment and occurs with probability per unit time (or rate) $k_E$, while the transition from motion to rest is called deposition and occurs with rate $k_D$.
-Second, particles in motion should move with mean velocity $V$ and some fluctuations around this velocity.
+Second, particles in motion should move with mean velocity $V$ and their instantaneous velocities should jitter around this mean through time due to turbulent drag and particle-bed collisions.
+
 The simplest equation of motion including these features is
-\be \dot{x}(t) = [V + \sqrt{2D}\xi(t)]\eta(t).  \label{eq:langevin} \ee
-Here $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles, and $\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion (with mean duration $1/k_D$), and $\eta=0$, representing rest (with mean duration $1/k_E$).
-Here, $V$ is the mean particle velocity, and $D$ is a diffusivity [units $L^2/T$] describing velocity fluctuations among moving particles.
+\be \dot{x}(t) = [V + \sqrt{2D}\xi(t)]\eta(t).  \label{eq:flippylangevin} \ee
+Here $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles.
+$\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion (with mean duration $1/k_D$), and $\eta=0$, representing rest (with mean duration $1/k_E$).
+
+The mean particle velocity is $V$, and $D$ is a diffusivity [units $L^2/T$] describing velocity fluctuations among moving particles.
 The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$.
-We write $k=k_E+k_D$ as a shorthand.
-This problem is similar to the ``randomly flashing diffusion" developed by Luczka and coworkers \citep{uczka1992,Luczka1993,Luczka1995}, although it includes a more complex asymmetric noise ($k_E\neq k_D$) and involves particle rest.
+The notation $k=k_E+k_D$ is a shorthand used throughout the thesis.
+
+The Langevin equation \ref{eq:flippylangevin} is a generalization of the model by \citet{Lisle1998} and \citet{Lajeunesse2017}, presented in section \ref{sec:lisle}, to include velocity fluctuations in the motion state.
+It is similar to the ``randomly flashing diffusion" developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}, although it includes a more complex asymmetric noise ($k_E\neq k_D$).
+This equation describes the downstream movement of particles alternating through motion and rest with velocity fluctuations in the motion state.
+This is a more realistic description of bed load trajectories than earlier works in the Einstein paradigm reviewed in section \ref{sec:trajmodels}.
 
 \subsection{Derivation of the master equation for the sediment position distribution $P(x,t)$}
-The solution of equation \ref{eq:langevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives the trajectory of one particle.
-Averaging over the ensemble of all such trajectories from different realizations of the noises will obtain the probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$.
+The solution of equation \ref{eq:flippylangevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives one possible trajectory of a particle.
+Averaging over the ensemble of all possible trajectories from different realizations of the noises will obtain the probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$.
 This distribution, by construction, will generalize earlier models of bedload alternation between motion and rest which did not include velocity fluctuations among moving particles \citep{Lisle1998,Lajeunesse2017}.
 
-We form the desired probability distribution of position as $ P(y,t) = \big\bra \delta(y-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
-\ref{eq:langevin} and the average is over both noises, but this symbolic equation is not yet useful as taking averages over both noises directly is a challenging mathematical problem \citep{Hanggi1978}.
-A simpler approach is to conduct the averages in Fourier space. Integrating eq. 1, using its solution in the probability distribution, then Fourier transforming gives
-\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
-Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen2007} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Shapiro1978} in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
-\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
-and inverse Fourier transforming provides the desired real-space master equation
-\be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0. \label{eq:master}\ee
-This is a diffusion-like equation governing the probability distribution of position for individual particles as they transport downstream through a sequence of motions and rests, with the movement velocity being a fluctuating quantity.
+The desired probability distribution of position is formed as $ P(x',t) = \big\bra \delta(x'-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
+\ref{eq:flippylangevin} and the average is over both noises.
+This symbolic equation is not directly useful as taking averages over both noises is a challenging mathematical problem \citep{Hanggi1978}.
+This calculation is completed in the appendix section \ref{sec:appAmaster}, providing the master equation
+\be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0 \label{eq:flippymaster}\ee
+for the probability distribution of position.
+
+The master equation \ref{eq:flippymaster} is a diffusion-like equation governing the probability distribution of position for individual particles as they transport downstream through a sequence of motions and rests, with the movement velocity being a fluctuating quantity.
+
 One can see in particular that taking an the entrainment rate $k_E$ very large, meaning that all particles are generally moving, implies an advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
 Otherwise, with $k_E$ of similar order as $k_D$, there is a finite probability that the particle is at rest, and the advection-diffusion process is often interrupted by deposition, giving rise to the additional terms in eq.
-\ref{eq:master} reminiscient of earlier motion-rest models with constant velocity \citep{Lisle1998,Lajeunesse2017}.
+\ref{eq:flippymaster} reminiscent of equation \ref{eq:lislemaster} from earlier simpler motion-rest models.
+When diffusion is turned off ($D=0$), equation \ref{eq:flippymaster} becomes identical to equation \ref{eq:lislemaster}, as expected.
 
 \section{Formalism for the downstream sediment flux}
-The probability distribution of the sediment flux is calculated using the probability distribution of particle position $P(x,t)$ provided as the solution of equation \ref{eq:master}.
+
+The probability distribution of the sediment flux can be calculated using the probability distribution of particle position $P(x,t)$ provided as the solution of equation \ref{eq:flippymaster}.
 This method is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
-The basic idea, as depicted in Figure 1, is that particles are initially distributed in all states of motion along a domain at random locations to the left of $x=0$.
-The flux is calculated using the number of particles that end up to the right of $x=0$ during the sampling time $T$.
+It generalizes the nonlocal formulation (\ref{sec:nonlocal}) to provide the probability distribution of the flux, and it generalizes the renewal approach (sec.
+\ref{sec:renewal}) to include the mechanics of individual particles in downstream transport.
+
+The basic idea, as depicted in Figure \ref{fig:flipflopfig1}, is to initially distribute $N$ particles in all states of motion along a domain of length $L$ at some random initial locations $x_i$ to the left of $x=0$.
+Later, the number of particles $N$ and the size $L$ of the domain will be limited to infinity such that their ratio $\rho=N/L$ remains constant. 
+This limit will provide a configuration similar to the one constructed in the nonlocal formation (sec. \ref{sec:nonlocal}).
+
+From this initial configuration, particles move downstream through time according to equation \ref{eq:flippylangevin}.
+The flux is calculated as the average rate of particles crossing to the right of $x=0$ after a sampling time $T$.
 \begin{figure}
@@
 The particle trajectories within are calculated from equation \ref{eq:langevin}, demonstrating alternation between rest and motion with fluctuating velocity.
-Particles begin their transport with positions $-L\leq x \leq 0$ at $t=0$, and as depicted in the right panel, the flux is calculated as the number of particles $N_>(t)$ which lie to the right of $x=0$ at the observation time $t$, divided by $t$: $Q(t) = t^{-1}N_>(t)$.
-We calculate the probability distribution of $Q$ over all realizations of the trajectories and initial positions as $L\rightarrow \infty$}
-	\label{fig:fig1}
+Particles begin their transport with positions $-L\leq x \leq 0$ at $t=0$, and as depicted in the right panel, the flux is calculated with the number of particles $N_>(T)$ which lie to the right of $x=0$ at the observation time $t=T$, divided by $T$: $q(T) = N_>(T)/T$.
+The probability distribution of $q(T)$ is determined from all possible realizations of the trajectories and initial positions as $N,L$ tend to infinity while the density of particles $\rho=N/L$ to the left of the control surface remains fixed.}
+	\label{fig:flipflopfig1}
 \end{figure}
+This time-averaged flux is
+\be q(T) = \frac{1}{T}\sum_{i=1}^N I_i(T). \label{eq:flippyflux} \ee
+In this equation, $I_i(T)$ is an indicator function which is $1$ if the $i$th particle has passed the control surface ($x=0$) at the observation time $T$ and $0$ otherwise.
+All particles which have not crossed the control surface (or which have crossed and then crossed back) contribute nothing to the flux.
 
-The rate of particles crossing the surface in an observation time $T$ is
-\be Q(T) = \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T). \ee
-In this equation, $\mathcal{I}_i(T)$ is an indicator function which is $1$ whenever the $i$th particle has passed our control surface and $0$ otherwise.
-All particles which have not crossed the control surface (or which have crossed and then crossed back) contribute nothing to the flux.
-The probability distribution of the flux is then
-\be P(Q|T) = \Big \bra \delta\Big( Q - \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T) \Big) \Big\ket. \ee
-The average is over the initial conditions of each particle and the ensemble of trajectories for each particle.
-Taking the Laplace transform over $Q$ (i.e. forming the characteristic function) obtains
-\begin{align} \tilde{P}(s|T) &= \Bigg\bra \int_0^\infty dQ e^{-s Q}\delta\Big( Q - \frac{1}{T}\sum_{i=1}^N \mathcal{I}_i(T) \Big)\Bigg\ket \\
-	&=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N \mathcal{I}_i(T)\Big)\Big\ket \\
-	&=  \prod_{i=1}^N \Big\bra \exp\Big(-\frac{s}{T}\mathcal{I}_i(T)\Big)\Big\ket \\
-	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra \mathcal{I}_i(T) \big\ket \Big] \end{align}
-This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $ \mathcal{I}_i(T)$ is either $1$ or $0$, so that $e^{ax} = 1-(1-e^a)x$.
-The average over initial conditions and possible trajectories for the $i$th particle can be written
-\be \bra \mathcal{I}_i(t) \ket = \frac{1}{L}\int_L^0 dx' \int_0^\infty dx \mathcal{P}(x,t|x') =  \frac{1}{L}\int_0^L dx' \int_0^\infty dx \mathcal{P}(x,t|-x'), \ee
-where $\mathcal{P}(x,t|x')$ is the probability density that the particle is found at position $x$ at time $t$ given it was initially at $x'$ at time $0$.
-This is the part of the flux that depends on the particle dynamics (ie instantenous velocities and entrainment/deposition characteristics).
+The probability distribution of the flux is then the average of equation \ref{eq:flippyflux} across all possible initial configurations of particles and their trajectories
+\be P(q|T) = \Big \bra \delta\Big( q - \frac{1}{T}\sum_{i=1}^N I_i(T) \Big) \Big\ket. \ee
+To evaluate these averages, taking the Laplace transform over $q$ (i.e. forming the characteristic function of $P(q|T)$) obtains
+\begin{align} \tilde{P}(s|T) &=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N I_i(T)\Big)\Big\ket \\
+	&=  \prod_{i=1}^N \Big\bra \exp\Big(-\frac{s}{T}I_i(T)\Big)\Big\ket \\
+	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra I_i(T) \big\ket \Big] \label{eq:flippycharacteristic}\end{align}
+This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $e^{a I} = 1-(1-e^a)I$ because $I$ is either $0$ or $1$.
 
-Inserting (7) into (6) and taking the limit as $L\rightarrow \infty$ and $N \rightarrow \infty$ while the density of particles $\rho = N/L$ remains constant provides
-\be \tilde{P}(s|T) = \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\mu(T) \Big)^N = \exp \Big[ -\big(1-e^{-s/T}\big)\mu(T) \Big].\ee
-where $\mu(T) = \rho \int_0^\infty dx \int_0^\infty dx' \mathcal{P}(x,T|x')$ is a rate constant which encodes the particle dynamics.
-This expression is the characteristic function of a Poisson distribution \citep{Cox1965}.
-Expanding in $e^{-s/T}$ and inverting the Laplace transform provides the probability distribution of the flux, contingent on the sampling time $T$:
-\be P(Q|T) = \sum_{k=0}^\infty \frac{\mu(T)^k}{k!}e^{-\mu(T)}\delta(Q-\frac{k}{T}).\ee
-This equation implies that the mean flux is $\bra Q \ket(T) = \int_0^\infty dQ Q P(Q|T) = \mu(T)/T$, and similarly the variance is $\sigma_Q^2(T) = \mu(T)/T^2$. 
-The conclusion is that if the flux is considered as a time averaged number of particles crossing a control surface, the mean flux is always Poissonian, no matter how particles move, provided the particles are independent and do not interact.
+The average over initial conditions and possible trajectories for the $i$th particle involved in this characteristic function can be written
+\be \bra I_i(t) \ket = \frac{1}{L}\int_L^0 dx' \int_0^\infty dx P(x-x',t) \label{eq:indicator} \ee
+where $P(x,t)$ solves equation \ref{eq:flippymaster}.
+Equation \ref{eq:indicator} describes the probability that the $ith$ particle is found to the right of control surface by time $T$ provided it started at a random location somewhere to the left.
+These $\bra I_i(t) \ket$ are the components of the flux that depend on the particle dynamics.
+
+Inserting equation \ref{eq:indicator} into equation \ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, obtaining the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
+\begin{align} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T)T \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T)T \Big].
+\label{eq:flippychar} \end{align}
+where $\Lambda(T) = \frac{\rho}{T} \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate constant, analogous to the rate constant in the renewal process models of section \ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
+
+The equation \ref{eq:flippychar} is the characteristic function of a Poisson distribution \citep{Cox1965}.
+Expanding in $e^{-s/T}$ and inverting the Laplace transform provides a key equation of this chapter, the probability distribution of the flux, held contingent on the sampling time $T$:
+\be P(q|T) = \sum_{l=0}^\infty \frac{\big[\Lambda(T)T\big]^l}{l!}e^{-\Lambda(T)T}\delta(q-\frac{l}{T}). \label{sec:flippydist}\ee
+This equation implies that the mean flux is $\bra q(T) \ket = \int_0^\infty q P(q|T) dq = \Lambda(T)$, and similarly the variance is $\sigma_q^2(T) = \Lambda(T)/T$.
+For the case when $\Lambda(T)$ does not depend on time, this equation is identical to the renewal theory approach reviewed in section \ref{sec:renewal}.
+
+This section has demonstrated that the scale-dependent sediment flux probability distribution can be calculated using the probabilistic dynamics of individual particles as a starting point.
+This constructs a mechanistic foundation beneath the renewal theory approach reviewed in section \ref{sec:renewal}.
 
@@
 	\caption{Panel (a) indicates the probability distribution of particle position as it evolves through time.
-From the initial mixture of motion and rest states, particles advect downstream as they diffuse due to differences in their fluctuating velocities and exchange between motion and rest.
+From the initial mixture of motion and rest states, particles advect downstream as they diffuse apart from one another due to differences in their velocities and exchange times between motion and rest.
 Panel (b) shows the resulting particle diffusion.
-At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard diffusion process (as advection is irrelevant on these timescales).
+At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process (as advection is irrelevant on these timescales).
 For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
-Finally longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, exemplified by exchange between motion and rest.
+Finally longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
 All results are scaled by the mean hop length $l=V/k_D$ and the autocorrelation time $\tau=1/k$ of the motion/rest alternation.}
@@
 
-Because the master equation is second order in time, it requires initial conditions for $P$ and $\pt P$.
+Because the master equation is second order in time, it requires initial conditions for both $P$ and $\pt P$.
 Considering that particles start at rest in a mixture of motion and rest states, with a fraction $k_E/k$ starting in motion and a fraction $k_D/k$ starting in rest, these conditions derive from the state
@@
 
-The master equation \ref{eq:master} is solved by transform calculus in the appendix, providing
-\begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du.
-\label{eq:monster}
+The master equation \ref{eq:master} is solved by transform calculus in the appendix section \ref{sec:fluccymastersol}, providing
+\begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t) +  \pt \big]\\
+	\times \int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)-k_D u} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[- \frac{(x-Vu)^2}{4Du}\Big] du, \label{eq:monster}
  \end{multline}
-This integral term encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator.
-The integral term convolves the probability that the particle has been in motion for a period $u$ out of a time $t$ with the probability density that a particle has travelled a distance $x$ in time $u$.
-The prefactor involving partial derivatives can be understood as adapting this distribution to the initial conditions.
+where $\varphi=k_D/k$ is the probability the particle starts at rest.
+This equation generalizes the earlier result \citep[e.g.][]{Lisle1998} for alternation between motions and rests, eq.
+\ref{eq:lisledist}, to include velocity fluctuations in the motion state.
 
+The integral in this equation encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the modified Bessel probability density that the particle has been in motion for a period $u$ out of a time $t$ with the advection-diffusion propagator that a particle has travelled a distance $x$ in time $u$.
+The prefactor of this integral term can be understood as adapting this distribution to the initial conditions.
+
+\subsection{The moments of particle position through motion-rest alternation}
 The moments of position from this distribution could be calculated by integrating equation \ref{eq:monster}, but this is challenging.
-Instead, we calculate the governing equation of the moments directly from the master equation \ref{eq:master}.
+Instead, we can calculate the governing equation of the moments directly from the master equation \ref{eq:master}.
 Multipling this equation by $x^l$ and integrating over all $x$ provides
@@
 Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2001,Nelson2004}, but they have not been introduced in geomorphology.
-
-Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
+Such a description would be an exciting extension of the methodology applied in this thesis so far to better understand landscape evolution.
 
@@
 This chapter introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motions and rests.
-The motion phase was enriched to include velocity fluctuations, forming the first physical model of this type.
+The motion phase included velocity fluctuations, forming the first analytical model of this type.
 The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying trajectories of individual particles.
@@
 The further inclusion of interacting particle models into bedload transport theory would be an exciting development.
+
+\endinput
+
+A simpler approach is to conduct the averages in Fourier space. Integrating eq. 1, using its solution in the probability distribution, then Fourier transforming provides
+\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
+Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen2007} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Shapiro1978}, in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
+\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
+and inverse Fourier transforming provides the desired real-space master equation

### 2021-07-06 ch6conc.tex
chapter 1 is flawless. gradually changing the notation and rewriting chapter 2. really getting bogged down on it.

@@
 This work includes ideas from granular gas theory into sediment transport modelling and provides a concrete example of the control grain-scale processes exert over large scale landscape evolution.
-In particular, since the work implies the shape of the bedload particle velocity distribution may depend on particle size, this has implications for mechanistic understanding of how abrasion by collisions of saltating bedload occurs (), which relates ultimately to rate at which Earth's bedrock canyons incise.
+In particular, since the work implies the shape of the bedload particle velocity distribution may depend on particle size, this has implications for mechanistic understanding of how abrasion by collisions of saltating bedload occurs (), which relates ultimately to the rates at which Earth's bedrock canyons incise.
 
@@
 
-The objective of this thesis was to produce new understanding of how to describe bedload transport using statistical physics.
-
-In some cases these limitations specific to the models produced in this thesis, but in most, they are shared in common with a majority of other models in the same research paradigm \citep{Ancey2020,Furbish2021a}.
+This thesis has produced new understanding of how to describe bedload transport using statistical physics, but its approach has many limitations deserving of future research attention.
+In some cases these are specific to the models produced in the thesis, but in others, they are shared in common with a majority of models in the stochastic sediment transport research paradigm \citep{Ancey2020,Furbish2021a}.
 
@@
 
-The first limitation concerns the assumption, implicit in every chapter of this thesis, that sediment transport characteristics are steady through time and space.
-This assumption contrasts with conditions in real gravel-bed river channels, where riffles, bars, and steps control the movements of individual grains \citep{}, sediment is supplied in episodic bursts from mass movements \citep{}, woody debris forces sediment deposition \citep{}, and flow conditions vary through time with weather \citep{}, climate change \citep{}, and seasonality \citep{}.
+The first limitation concerns the assumption, implicit in every chapter of this thesis, that sediment transport characteristics are steady and uniform in time and space.
+This assumption contrasts with conditions in real gravel-bed rivers, where riffles, bars, and steps coordinate the movements of individual grains \citep{Ashmore1998,McDowell2020}, sediment is supplied in episodic bursts from mass movements \citep{Benda1990, Muller2018}, woody debris coordinates sediment movement \citep{Eaton2012,Reid2019}, and variable flow conditions modify bed texture and sediment availability \citep{Mao2012,Phillips2018}.
 
-Here I have considered extremely idealized conditions where sediment transport and landscape dynamics are not linked, except for chapter 2 where bed elevations were coupled to sediment entrainment and deposition.
+To date, very little work has concentrated on stochastic sediment transport models in unsteady conditions \citep[e.g.][]{Bohorquez2016}.
+The most obvious scheme to address unsteadiness is to introduce time and space dependence to movement velocities, diffusivities, or entrainment and deposition rates, but how exactly one should express these rates in terms of time and space is a matter for speculation, and would be contingent on a given context given our limited understanding of how these parameters relate to the flow hydraulics \citep[e.g.][]{Heyman2016}.
+
+Future studies might investigate in more depth the linkage between flow hydraulics, the geometric arrangement of grains on the bed, and the bed topography with the entrainment and deposition rates of individual particles to develop the foundational knowledge to incorporate temporally or spatially variable flow and sedimentary conditions into stochastic models of sediment transport.
 
@@
 
-Sediment in real rivers spans a range of sizes, while I have modelled sediment transport of one grain size only, characterized by
+The second major limitation of this work concerns grain size.
+In actuality, sediments in rivers spans a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.
 
-\subsection{Rivers, hillslopes, and dunes}
+There have been a few works on stochastic modelling of sediment transport with multiple grain sizes \citep{Sun2000,Parker2000}, but these approaches are not spatially distributed, so there is as yet no stochastic method to understand sorting processes.
 
-It has long been recognized, probably since \citet{Bagnold1941} and certainly today \citep{}, that there are deep analogies between tranpsort phenomena in air and water.
-In an ideal perspective, we might imagine writing the governing equations for transport in water as a function of the fluid viscosity, then tuning the viscosity between air and water to obtain a description applying to both spheres.
+Future studies should revisit the stochastic framework applied to multiple grain sizes.
+Simplified experimental geometries based on bimodal sediment beds \citep[e.g.][]{Houssais2012} would be a great context to revisit these issues.
+Extending the motion-rest model of chapter () to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
 
-For example, in a mixture of small and large grains, neglecting any sorting effects whereby the mobility of small grains is contingent on the mobility of the large grains, small grains will have exponential velocities with relatively wide fluctuations, while large grains will have Gaussian velocities with relatively narrow fluctuations.
-Considering the over-all flux then as the number of moving particles times their velocities, transport fluctuations
+\subsection{The full range of geophysical flows}
 
-aeolean transport extension maybe
-particle size distributions - interesting implications
-connection to computational physics approach
-connection to the stochastic description of the flux
+The final limitation I will mention is that all of the efforts in this thesis were concentrated on weak bedload transport, where densities of moving particles are low enough that they may interact with the static bed () but never with each other.
+This concentration is justified because weak bedload transport conditions are typical of gravel bed rivers \citep{Ashworth1989,Warburton1992}, but it is nonetheless a limitation given the diversity of processes which move sediment over Earth's surface.
+
+Earth's landforms are shaped by numerous transport phenomena from booming debris flows \citep{Iverson1997} to barely perceptible hillslope creep \citep{Deshpande2021}.
+Different phenomena are basically distinguished by the relative importance of fluid, granular, and gravitational forces in sustaining them \citep{Jerolmack2019}.
+
+Weak bedload transport is characterized by fluid forces small in comparison to gravity, and collision forces comparable to gravity.
+Viewed in this way, the work in this thesis represents a minute region of the vast parameter space spanned by Earth's geophysical flows, and geomorphology requires characterization of them all.
+Future studies should continue the effort \citep{Furbish2021a} to describe these processes as different expressions of the same basic building blocks. 
+Statistical mechanics has some versatile blocks.
 
@@
 
-This thesis applied statistical mechanics methods to describe the movements of individual grains along streambeds.
-The research has produced new understanding of the relationships between the overall sediment transport rates which control channel evolution and the movements of individual grains.
-The work embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as their fluctuations.
+This thesis described the movements of individual grains along streambeds using probabilistic methods.
+The research has related the overall sediment transport rates responsible for channel evolution to the movements of individual grains.
+At base level, the thesis embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as the magnitude of their fluctuations.
 
-The founders of process geomorphology always acknowledged variability in landscapes \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes despite it, like averaged flow conditions to avoid including fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to summarize climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid wide grain size distributions \citep{Parker1982,Andrews1983}.
+The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes without including it, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climate fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving grain size distributions \citep{Parker1982,Andrews1983}.
 
-The stochastic models of sediment transport developed in this thesis hint toward a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit this old question in geomorphology: How does variability shape Earth's surface?
+The stochastic descriptions of sediment transport developed in this thesis hint toward a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?
+
+\endinput
+
+Its closest neighbours may be rarefied hillslope transport, where solitary grains tumble down hillslopes \citep{Williams2021}, and intense bedload transport, where particles creep downstream in a dense granular flow, supported by collisions with other moving grains \citep{Frey2014}.
+These phenomena differ only by viscosity or density, not mechanically.
+
+\citet{Ashworth1989} and \citet{Warburton1992} identified three phases of bedload transport in gravel-bed streams that occur as the flow increases.
+Phase 1 is characterized by movement of sand only, while gravel remains stationary; phase 2 involves partial mobility of the smaller gravel sizes; and phase 3 is the full mobility of all grain sizes represented on the bed.
+Phase 2 is most common in gravel-bed rivers, but phase 3 has

### 2021-07-07 ack.tex
this is a nightmare

@@
 
-Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele, you've formed all of the best parts of my personality and continually guided me toward this point, and it's impossible to explain how thankful I am for that.
+Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele, you've formed all of the best parts of my personality and guided me toward this point, and it's impossible to explain how thankful I am for that.
 Regarding family I will also always be thankful for my great aunt Alice who graciously paid for my undergraduate education.
@@
 
-Thanks also to my professors at Southern WV Community College (SWVCC) and West Virginia University (WVU) who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
+Thanks also to my professors at Southern WV Community College who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
 At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
 Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science.
-Everyone encouraged me to continue my education.
 SWVCC was a personal transformation and a great privilege.
@@
 I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
-I have yet to experience teamwork and mutual exploration like we had in the undergrad lounge.
-Math and physics are fun!
 I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator for comic relief, and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level.
@@
 Now we come to Marwan Hassan.
-About five years ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics.
+Some time ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics.
 I realize now when I said ``statistical physics" your eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well.
@@
 Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done.
-Maybe they'll bring me closer to your way of thinking!
+Maybe they'll bring me closer to your way of thinking.
 
@@
 Now I have a thesis on it.
-David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy from stochastic landscape evolution, to probability as a physical quantity like energy and momentum and fringe ideas in cosmology.
-These are my people!
+David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy.
 David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.
@@
 
-5
 \endinput

### 2021-07-07 app_ch2.tex
this is a nightmare

--- 2021-07-02 app_ch2.tex
@@
 
-\chapter{Ch. 2 calculations}
+\chapter{Calculations involved in dynamical sediment flux model}
+\label{sec:appendixA}
+\section{Derivation of the master equation for particle position}
+\label{sec:appAmaster}
+Here I derive the master equation \ref{eq:flippymaster} for the probability distribution of particle position from the Langevin equation \ref{eq:flippylangevin}.
+This is closely based off of the approach in \citet{Balakrishnan1993}.
 
-\section{Derivation of the master equation}
+The master equation satisfies $P(x,t) = \bra\bra \delta(x-x(t))\ket_\xi \ket_\eta$, where the averages are over all realizations of the two independent noises.
+It is most convenient to take these averages after taking Fourier transforms and manipulating time derivatives of the resulting equations.
+
+Integrating the Langevin equation \ref{eq:flippylangevin}, substituting the solution in the above definition, and taking the Fourier transform in space gives
+\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi\ee
+This can be rearranged as
+\be \tilde{P}(g,t) = \Big\bra \exp\Big[i g V \int_0^t du \eta(u)\Big] \Big\bra \exp\Big[ i g \sqrt{2D}\int_0^t du \xi(u) \eta(u)]\Big]\Big\ket_\xi \Big\ket_\eta .\ee
+
+Using the classic identity for an average over white noise of an exponential \citep{Balakrishnan1993,VanKampen2007} gives, after recognizing that the dichotomous noise satisfies $\eta^2 = \eta$ since $\eta$ is either $0$ or $1$,
+\be \tilde{P}(g,t) = \Big\bra \exp\Big[(igV-g^2D) \int_0^t du \eta(u) \Big]\Big\ket_\eta. \label{eq:logi}\ee
+
+Now I will take time derivatives in order to apply the so-called Furutsu-Novikov formula to evaluate the ensemble average over dichotomous noise \citep{Shapiro1978}.
+Applied to the dichotomous noise in figure \ref{fig:lislefig}, the Furutsu-Novikov formula is, for an arbitrary functional $F$ of the noise,
+\be \pt \bra \eta(t)F[\eta(t)] \ket = \bra \eta \pt F \ket + k \big[\bra \eta \ket \bra F \ket -\bra \eta F \ket \big]  \label{eq:furuti}.\ee
+
+Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking as time derivative of eq. \ref{eq:logi} gives
+\be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket. \label{eq:furut}\ee
+Taking a second time derivative opens up the possibility of using the Furutsu-Novikov formula \ref{eq:furuti}: 
+\be G^{-1}\pt \tilde{P}(g,t) = \pt \bra \eta F\ket = G\bra \eta F \ket + k\big[ \bra \eta \ket \tilde{P}(g,t) - G^{-1} \pt \tilde{P}(g,t) \big]. \ee
+Applying eq. \ref{eq:furut} finally gives
+\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
+and inverse Fourier transforming provides the master equation \ref{eq:flippymaster}.
+
 \section{Solution for the position probability distribution}
-\section{Moments of position}
-\section{Calculation of the flux rate constant}
+\label{sec:fluccymastersol}
+The position probability distribution can be obtained from eq.
+\ref{flippymaster} for the initial conditions $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ using Fourier transforms in space and Laplace transforms in time.
+Taking these transforms provides
+\be \bar{\tilde{P}}(g,s) = \frac{s + k +\varphi  Dg^2 - i g V \varphi}{s(s+k)+(Dg^2-igV)(s+k_E)}, \ee
+where $\varphi = k_D/k$ is the probability a particle starts at rest.
+
+The numerator terms encode the initial conditions. The denominator terms are where the real structure of the solution is contained.
+The numerator terms involving $g$ are easily expressed as first and second derivatives with respect to $x$. The problem, then is to calculate the inverse Fourier integral of the denominator.
+This can be conducted by finding the zeros of the denominator, expanding in partial fractions, then applying the contour integral \citep[e.g.][]{Arfken1985}
+\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g + i c} dg = \theta(x)e^{-c x}.\ee
+
+Rearranging the governing equation gives
+\be \tilde{\hat{P}}(g,s) = \frac{\varphi D g^2 - ig V\varphi  + s + k}{D(s+k_E)}\frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}}\ee
+The roots of the denominator are at
+\be g_\pm = \frac{iV}{2D}\Big[ 1  \pm R \Big],\ee
+where
+\be R = \sqrt{1 + \frac{4D}{V^2}\frac{s(s+k)}{s+k_E}}\ee
+The partial fractions expansion is therefore
+\be \frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}} = \frac{D}{i V R}\Big[ \frac{1}{g-g_+} - \frac{1}{g-g_-}\Big]\ee
+Giving 
+\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi\px + s + k}{VR(s+k_E)}\Bigg[ \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_+} - \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_-}\Bigg]\ee
+or eventually
+\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\Bigg[ \theta(x) \exp\Big(\frac{Vx}{2D}(1-R)\Big) + \theta(-x)\exp\Big( \frac{Vx}{2D}(1+R)\Big) \Bigg].\ee
+This simplifies slightly to
+\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\exp\Big[\frac{Vx}{2D} - \frac{V|x|}{2D}R \Big], \label{eq:laplaceapp}\ee
+where as a reminder, $\varphi = k_D/k$.
+
+Taking the inverse Laplace transform symbolically provides the desired probability density function
+\be P(x,t) = \mathcal{L}^{-1}\Bigg\{ \frac{1}{V} \big[-\varphi D\px^2 + V\varphi \px + k + s \big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(-\frac{V|x|R}{2D} \Big)}{(s+k_E)R} \Bigg\}(t). \ee
+Using the shift property of Laplace transforms \citep[e.g.][]{Arfken1985}, this becomes
+\be P(x,t) = \frac{1}{V}e^{-k_E t} \mathcal{L}^{-1}\Bigg\{ \big[-\varphi D\px^2 + V \varphi \px + k_D + s\big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(\frac{-V|x|R_\ast}{2D}\Big)}{s R_\star} \Bigg\},\ee
+where 
+\be R_\ast = \sqrt{1 + \frac{4D(k_D-k_E)}{V^2}+ \frac{4D}{V^2}\Big(s-\frac{k_Ek_D}{s}\Big)}. \ee
+Using the property that $\mathcal{L}^{-1}\big\{s \tilde{f}\big\} = (\delta(t) + \pt)f(t)$ \citep{Arfken1985} gives
+\be P(x,t) = \frac{1}{V} e^{-k_E t} \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big] \mathcal{L}^{-1}\Bigg\{ \frac{-\exp\Big[\frac{-V|x|R_\ast}{2D}\Big]}{sR_\ast} \Bigg\}(t) \ee
+Finally, using the identity
+\be \mathcal{L}^{-1} \Big\{ \frac{1}{s} \tilde{g}(s-a/s)\Big\} = \int_0^t \mathcal{I}_0\Big(2\sqrt{au(t-u)}\Big) g(u) du
+\ee
+from \citet{Bateman1953}, pg. 133 (which can be derived by either u-substitution or series expansion in powers of $a$) provides:
+\begin{multline}  P(x,t) = \frac{1}{V} e^{-k_E t}  \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big]\\ \times \int_0^t du \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) \\ \times \mathcal{L}^{-1}\Bigg\{ \frac{\exp\Big[ \frac{-V|x|}{2D}\sqrt{a+b s}\Big]}{\sqrt{a + b s}}\Bigg\}(u)  \label{eq:magic}\end{multline}
+where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$ are shorthands.
+
+Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator, the remaining transform can be evaluated from standard tables \citep{Prudnikov1992a}, eventually giving (after integration over and setting to 1 of the fake parameter),
+\begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du, \end{multline}
+as reported in equation \ref{eq:monster}.
+
+\section{Calculation of the scale-dependent rate function $\Lambda(T)$}
+\label{sec:fluxconstant}
+The central object required to calculate the sediment flux probability distribution in equation \ref{eq:flippyflux} is 
+\be \Lambda(T) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,T).\ee
+This represents the rate at which particles starts at $x_i$, anywhere to the left of $x=0$, and manage to cross $x=0$ by time $T$.
+
+Again it is easiest to conduct the necessary calculus after Laplace transforming. This gives 
+\be \tilde{\Lambda}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s).\ee
+The Laplace transform of the rate function therefore follows after integrating equation \ref{eq:laplaceapp} twice.
+
+Noting that $x+x_i$ is always positive,
+The first integration gives
+\begin{multline} \tilde{\Lambda}(s) = \rho \int_0^\infty dx_i \exp\Big[\frac{V(1-R)x_i}{2D}\Big]\\ 
+	\times\Big(\varphi \frac{1-R}{2R(s+k_E)} - \frac{\varphi}{R(s+k_E)} - \frac{2D(s+k)}{V^2R(1-R)(s+k_E)}\Big), \end{multline}
+and the second integration provides
+\begin{multline}  \tilde{\Lambda}(s) = -\frac{\rho \varphi D}{V R(s+k_E)}  + \frac{2\rho D\varphi}{VR(1-R)(s+k_E)} \\ + \frac{4\rho D^2(s+k)}{V^3R(1-R)^2(s+k_E)}. 
+	\label{eq:laplacefluxrate}\end{multline}
+
+Taking the inverse transform, converting the $s$ factor in the numerator of the last term to $\pt + \delta(t)$, and using the shift property gives \citep[e.g.][]{Arfken1985}
+\be \Lambda(t) = \rho \mathcal{L}^{-1}\Bigg\{ - \frac{\varphi D}{VR_\star s} + \frac{2D\varphi}{VR_\ast(1-R_\star)s} + \frac{4D^2\big( \cev{\pt} + k\big)}{V^3 R_\star(1-R_\star)^2s}\Bigg\}.\ee
+Here the notation $\cev{\pt}$ means the derivative acts from the left on all terms multiplying it (as in $f(t)\cev{\pt} g(t) = \pt [f(t)g(t)]$).
+
+After some work, using eq.
+\ref{eq:magic} and tabulated Laplace transform pairs \citep[e.g.][]{Arfken1985,Prudnikov1992a} to perform the inverse Laplace transforms, the net result for the rate constant is
+\begin{multline} 
+	\Lambda(t) = \rho \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
+	\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{k_D}{2 k}\Big)e^{-V^2 u/4D}  \\ + \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du,
+\end{multline}
+as reported in equation \ref{eq:fluxy}.
+
+\label{sec:fluxlimits}
+Equation \ref{eq:fluxy} is difficult to work with.
+To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep[e.g.][]{Weiss1994} to invert the Laplace-transformed rate function of equation \ref{eq:laplacefluxrate} at the opposite extreme of $s$.
+
+For example, at short times, expanding equation \ref{eq:laplacefluxrate} as $s\rightarrow \infty$ gives
+\be \tilde{\Lambda}(s) = \frac{\rho k_E V}{2 k s^2} + \frac{\rho k_E}{k}\sqrt{\frac{D}{4 s^3}} \ee
+which inverts to
+\be \Lambda(t) \sim \frac{\rho k_E V T}{2 k} + \frac{\rho k_E}{k}\sqrt{\frac{D T}{\pi}},\ee
+giving the small $T$ behavior.
+This has two scaling limits within it. Provided that $T \ll 4D/V^2/\pi < 2 D/V^2$, the scaling goes as $\Lambda(T) \sim T^{-1/2}$. But if $T\gg 4D/V^2/\pi$, it goes as $\Lambda(t) \sim T$.
+
+For large times, taking $s\rightarrow 0 $ gives
+\be \tilde{\Lambda}(s) = \frac{\rho k_E V }{ks^2}, \ee
+and this inverts to $\Lambda(T) = \rho k_E V T/k$ as the long time solution, providing a mean flux equivalent to the Einstein model (sec. \ref{sec:lisle}).
+These limits are summarized in equation \ref{eq:fluccolimitcases}.

### 2021-07-07 ch1intro.tex
this is a nightmare

@@
 \subsection{Birth death models for bedload flucuations}
-
+\label{sec:birthdeath}
 The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random process.

### 2021-07-07 ch2flipFlop.tex
this is a nightmare

@@
 From these assumptions, the first target is to write an equation of motion for the individual sediment particle encompassing two features.
+\begin{figure}
+	\centerline{\includegraphics{./figures/ch2/fig0mod.pdf}}
+	\caption{Panel (a) sketches a realization of the noise in equation \ref{eq:flippylangevin}, while panel (b) shows the trajectory derived from it (in blue) alongside other possible trajectories.
+Keys in panel (a) demonstrate the average movement time $1/k_D$ and rest time $1/k_E$, while the key in panel (b) shows the average movement velocity $V$.
+Velocity fluctuations lead to tilted stair-step trajectories with unsteady slopes in the $x-t$ plane.
+This can be compared with figure \ref{fig:lislefig} which does not include fluctuations.}
+	\label{fig:fluxxy0}
+\end{figure}
 First, particles should alternate between motion and rest.
@@
 $\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion (with mean duration $1/k_D$), and $\eta=0$, representing rest (with mean duration $1/k_E$).
+Some trajectories produced by this equation are sketched in figure \ref{fig:fluxxy0}.
 
@@
 \subsection{Derivation of the master equation for the sediment position distribution $P(x,t)$}
+\label{sec:floppymastereq}
 The solution of equation \ref{eq:flippylangevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives one possible trajectory of a particle.
@@
 \section{Formalism for the downstream sediment flux}
+\label{sec:flippyflux}
 
@@
 Inserting equation \ref{eq:indicator} into equation \ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, obtaining the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
-\begin{align} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T)T \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T)T \Big].
+\begin{align} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T) \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T) \Big].
 \label{eq:flippychar} \end{align}
-where $\Lambda(T) = \frac{\rho}{T} \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate constant, analogous to the rate constant in the renewal process models of section \ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
+where $\Lambda(T) = \rho \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate constant, similar to the rate constant in the renewal process models of section \ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
 
@@
 Expanding in $e^{-s/T}$ and inverting the Laplace transform provides a key equation of this chapter, the probability distribution of the flux, held contingent on the sampling time $T$:
-\be P(q|T) = \sum_{l=0}^\infty \frac{\big[\Lambda(T)T\big]^l}{l!}e^{-\Lambda(T)T}\delta(q-\frac{l}{T}). \label{sec:flippydist}\ee
-This equation implies that the mean flux is $\bra q(T) \ket = \int_0^\infty q P(q|T) dq = \Lambda(T)$, and similarly the variance is $\sigma_q^2(T) = \Lambda(T)/T$.
-For the case when $\Lambda(T)$ does not depend on time, this equation is identical to the renewal theory approach reviewed in section \ref{sec:renewal}.
-
-This section has demonstrated that the scale-dependent sediment flux probability distribution can be calculated using the probabilistic dynamics of individual particles as a starting point.
-This constructs a mechanistic foundation beneath the renewal theory approach reviewed in section \ref{sec:renewal}.
+\be P(q|T) = \sum_{l=0}^\infty \frac{\Lambda(T)^l}{l!}e^{-\Lambda(T)}\delta(q-\frac{l}{T}). \label{eq:flippydist}\ee
+This equation implies that the mean flux is $\bra q(T) \ket = \int_0^\infty q P(q|T) dq = \Lambda(T)/T$, and similarly the variance is $\sigma_q^2(T) = \Lambda(T)/T^2$.
+For the case when $\Lambda(T)$ is proportional to the observation time ($\Lambda \propto T$), this equation is identical to the renewal theory approach reviewed in section \ref{sec:renewal}.
+This demonstrates that the renewal approach can be formulated with more generality by considering the dynamics of individual particles as a starting point.
 
 \section{Results \label{sec:res}}
-\subsection{The position probability distribution and its moments}
+\subsection{Position probability distribution of sediment particles}
 \begin{figure}
 	\centerline{\includegraphics{figures/ch2/figure2_slopeKey.pdf}}
-	\caption{Panel (a) indicates the probability distribution of particle position as it evolves through time.
+	\caption{Panel (a) indicates the probability distribution of particle position (eq.
+\ref{eq:monster}) as it evolves through time.
 From the initial mixture of motion and rest states, particles advect downstream as they diffuse apart from one another due to differences in their velocities and exchange times between motion and rest.
-Panel (b) shows the resulting particle diffusion.
+Panel (b) shows the resulting particle diffusion (eq.
+\ref{eq:diffo}).
 At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process (as advection is irrelevant on these timescales).
 For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
-Finally longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
-All results are scaled by the mean hop length $l=V/k_D$ and the autocorrelation time $\tau=1/k$ of the motion/rest alternation.}
-	\label{fig:fig1}
+Finally at times longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
+All results are scaled by the mean hop length $l=V/k_D$ and the autocorrelation time $\tau=1/k$ of the motion/rest alternation.
+In both plots, the lines are analytical results while the points are the results of Monte Carlo simulations based on evaluating cumulative transition probabilities on a small timestep \citep[e.g.][]{Barik2006}.}
+	\label{fig:flippyfig1}
 \end{figure}
-The master equation \ref{eq:master} describes the evolution of the probability distribution of position through time.
-Loosely speaking the solution of this equation is expected to be some combination of the \citet{Einstein1937} theory for particle transport with the Gaussian propagator of the advection-diffusion equation \citep{Morse1953a}.
+The master equation \ref{eq:flippymaster} describes the evolution of the probability distribution of position through time.
+Loosely speaking the solution of this equation is expected to be some combination of the \citet{Einstein1937} theory for particle transport (section \ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep{Morse1953a}.
 
 Because the master equation is second order in time, it requires initial conditions for both $P$ and $\pt P$.
-Considering that particles start at rest in a mixture of motion and rest states, with a fraction $k_E/k$ starting in motion and a fraction $k_D/k$ starting in rest, these conditions derive from the state
+Considering that particles start at rest in a mixture of motion and rest states, with a fraction $k_E/k$ starting in motion and a fraction $k_D/k$ starting in rest, these conditions derive from the initial state
 \be P(x,0) = \lim_{t\rightarrow 0 } \frac{k_E}{k} \sqrt{\frac{1}{4\pi D t}} \exp\Big[-\frac{(x-Vt)^2}{4Dt}\Big]+ \frac{k_D}{k}\delta(x)\ee
-which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep{Weiss2002a}.
+which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep[c.f.][]{Weiss2002a}.
 
-The master equation \ref{eq:master} is solved by transform calculus in the appendix section \ref{sec:fluccymastersol}, providing
+The master equation \ref{eq:flippymaster} is solved by transform calculus in section \ref{sec:fluccymastersol} of the appendices, providing
 \begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t) +  \pt \big]\\
@@
 This equation generalizes the earlier result \citep[e.g.][]{Lisle1998} for alternation between motions and rests, eq.
-\ref{eq:lisledist}, to include velocity fluctuations in the motion state.
+\ref{eq:lisledist}, to include velocity fluctuations within the motion state.
+This is another key result of this chapter.
 
-The integral in this equation encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the modified Bessel probability density that the particle has been in motion for a period $u$ out of a time $t$ with the advection-diffusion propagator that a particle has travelled a distance $x$ in time $u$.
-The prefactor of this integral term can be understood as adapting this distribution to the initial conditions.
+The distribution \ref{eq:monster} is shown evolving through time in figure \ref{fig:flippyfig1} panel (a), where the advection and diffusion characteristics of equation \ref{eq:flippymaster} are both clearly visible.
+The integral in equation \ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the modified Bessel probability density that the particle has been in motion for a period $u$ out of a time $t$ with the advection-diffusion propagator that a particle has travelled a distance $x$ in time $u$.
+The prefactors of this integral term can be understood as adapting this distribution to the initial conditions.
 
 \subsection{The moments of particle position through motion-rest alternation}
-The moments of position from this distribution could be calculated by integrating equation \ref{eq:monster}, but this is challenging.
-Instead, we can calculate the governing equation of the moments directly from the master equation \ref{eq:master}.
-Multipling this equation by $x^l$ and integrating over all $x$ provides
-\be \pt^2 m_l -V l \pt m_{l-1} -k_E V l m_{l-1} + k \pt m_l - D l (l-1) \pt m_{l-2} - k_E D l(l-1) m_{l-2} = 0,\ee
-where $\bra x^l \ket = m_l$. 
+The moments of position produced by equation \ref{eq:monster} could be derived by integrating it directly, but this is difficult.
+Instead, the moments can be calculated directly from the master equation \ref{eq:flippymaster} \citep{Cox1965}.
+
+Multipling equation \ref{eq:flippymaster} by $x^l$ and integrating over all $x$ provides
+\begin{multline} \pt^2 \bra x^l \ket -V l \pt \bra x^{l-1} \ket -k_E V l \bra x^{l-1} \ket + k \pt \bra x^{l} \ket\\- D l (l-1) \pt \bra x^{l-2} \ket - k_E D l(l-1) \bra x^{l-2} \ket = 0.\end{multline}
 For $l = 1$, this equation generates the mean position $ \bra x \ket(t) = k_E V t/k$, which is unaffected by diffusion (since Gaussian velocity fluctuations are symmetric).
-The case $l=2$ provides the second moment, which provides the variance of position
-\be \sigma_x^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) + 2\frac{k_E D}{k}t.\ee
-At short times $t\ll k^{-1}$, this becomes
-\be \sigma_x^2 \sim \frac{k_E k_D V^2}{k^2} t^2 + \frac{2 k_E D }{k}t,\ee
-so it scales as $\sigma_x^2 \sim t^2$ when $t \gg \frac{2Dk}{V^2 k_D}$ and $\sigma_x^2 \sim t$ when $t \ll \frac{2Dk}{V^2 k_D}$.
-Similarly at long times $t\gg k^{-1}$, 
-\be \sigma_x^2 \sim \frac{2k_Ek_D V^2}{k^3}t,\ee
-so we have a non-trivial multi-scale diffusion phenomenon. 
-Provided that $2D/V^2 \ll k_D/k^2$, we therefore have
+The case $l=2$ provides the second moment, implying the variance of position
+\be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) + 2\frac{k_E D}{k}t. \label{eq:minor} \ee
+This equation describes a non-trivial multi-scale diffusion phenomenon, whereby the rate at which particles spread apart depends on how long their dynamics have been ongoing.
+
+Provided that $2D/V^2 \ll k_D/k^2$, one can determine that equation \ref{eq:minor} produces three different scaling behaviors:
 \be \sigma_x^2 \sim 
@@
 	 t^2, &  \frac{2Dk}{V^2 k_D} \ll t \ll \frac{1}{k}, \\
-	 t, & t\gg \frac{1}{k}.
+	 t, & t\gg \frac{1}{k}. \label{eq:diffo}
 \end{cases}\ee
-Note in the physical condition when $k\approx k_D$, the condition for the existence of three ranges becomes $ \Pe \ll 1,$ where
-$\Pe = 2 D k_D/V^2.$
-The small Peclet limit, when three diffusion ranges are obtained, is characteristic of bedload sediment transport where velocity fluctuations are typically small compared to mean downstream movement velocites.
+Note in the physical condition when $k\approx k_D$, which is satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the condition for existence of three ranges becomes $ \Pe \gg 1,$ where
+$\Pe =V^2/(2k_D D)$ is a Peclet number \citep[e.g.][]{Heyman2014} which measures the relative importance of advection and diffusion in the movements of particles.
+
+The large Peclet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles very rarely move upstream \citep[e.g.][]{Fathel2015}.
+Incidentally, this limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed.
+Figure \ref{fig:flippyfig1} panel (b) sketches this three-range diffusion.
+
+Below, this anomalous diffusion behavior \citep[c.f.][]{Sokolov2012} is shown to affect the rate constant $\Lambda(t)$ in the sediment flux.
+As a result of these multi-scale particle dynamics, the sediment flux becomes scale-dependent in a richer way than the renewal model of section \ref{sec:renewal}.
 
@@
 
-From the formalism in section xxxx, the central parameter of the sediment flux distribution is 
-\be \mu(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t).\ee
-This represents the rate of particles crossing $x=0$ at the observation time $T$ given they started somewhere to left of $x=0$ at $T=0$.
+From the formalism in section \ref{sec:flippyflux}, the central parameter of the sediment flux distribution is the rate function
+\be \Lambda(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t). \label{eq:flatform}\ee
+This represents the number of particles crossing $x=0$ at the observation time $T$ given they started somewhere to left of $x=0$ at $T=0$.
 
-It is convenient to calculate this quantity in Laplace space.   
-Taking the Laplace transform,
-\be \tilde{\mu}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s),\ee
-then performing both integrals provides
-\be  \tilde{\mu}(s) = -\frac{\phi D}{V R(s+k_E)} + \frac{2D\phi}{VR(1-R)(s+k_E)} + \frac{4D^2(s+k)}{V^3R(1-R)^2(s+k_E)}. \label{eq:laplacefluxrate}\ee
-$\mu(t)$ derives as the inverse Laplace transform of this equation. 
-This calculation is performed in the appendix, providing
+Using the probability distribution of position \ref{eq:monster}, the integrals in equation \ref{eq:flatform} are performed in section \ref{sec:fluxconstant} of the appendix, providing the rate constant
 \begin{multline} 
-\mu(t) = \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
-\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{k_D}{2 k}\Big)e^{-V^2 u/4D} + \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du.
+\Lambda(t) = \rho \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
+\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{k_D}{2 k}\Big)e^{-V^2 u/4D} \\+ \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du.
 \label{eq:fluxy}
 \end{multline}
-In this equation, the notation $\cev{\partial}$ means that the derivative acts from the left of all terms in which it is involved, as in $f(t)\cev{partial}_t g(t) = \partial_t f(t)g(t).$
+In this equation, the notation $\cev{\partial_t}$ means that the partial time derivative acts from the left of all terms in which it is involved, as in $f(t)\cev{\partial}_t g(t) = \partial_t [f(t)g(t)].$
 
-This is a complicated result for the rate constant in the sediment flux \ref{eq:fluxdist}, but this complexity is not surprising given that equation \ref{eq:langevin} involves two interacting diffusion processes.
-As displayed in \ref{fig:fig3}, as a result of eq.
-\ref{eq:fluxy} the mean flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $Q_0 = E l$ as the observation time becomes much larger than $1/(k_D \Pe)$.
-
+This is an intricate result for the rate constant in the sediment flux \ref{eq:flippydist}.
+This mathematical complexity may not be surprising given that equation \ref{eq:flippylangevin} involves two interacting diffusion processes.
+As displayed in figure \ref{fig:fluxconvergence}, as a result of eq.
+\ref{eq:fluxy} the mean sediment flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $q_0 = \rho k_E V/k = E \ell$ as the observation time becomes much larger than $1/(k_D \Pe)$.
 \begin{figure}[!htbp]
 	\includegraphics[width=\linewidth,keepaspectratio]{figures/ch2/figure3_slopeKey.pdf}
-	\caption{}
+	\caption{The mean sediment flux is plotted for different values of the Peclet number $\Pe = V^2/(2k_D D)$, characterizing the relative strength of particle velocity fluctuations during motion.
+The flux is normalized by the prediction $q_0 = E \ell$ of the Einstein theory (sec.
+\ref{sec:einflux}).
+For nonzero velocity fluctuations in the motion state (finite $\Pe$), the Einstein limit is approached as the observation time $T$ grows.
+Stronger velocity fluctuations (smaller $\Pe$) slow the convergence to this limit.
+		In all cases, satisfactory convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by equation \ref{eq:fluccolimitcases}.
+Plotted lines are the analytical result eq.
+\ref{eq:fluxy}, while the points are the results of Monte Carlo simulations.}
 	\label{fig:fluxconvergence}
 \end{figure}
-The convergence time of the flux thus scales with the inverse Peclet number.
-When particle velocity fluctuations during motions are larger compared to advection, the flux becomes slow to converge, in general.
 
-\subsection{Reduction to earlier work}
+Further insight into the rate constant of the sediment flux distribution can be gained by investigating extreme cases of the observation time.
+As shown in appendix section \ref{sec:fluxlimits}, equation \ref{eq:fluxy} takes on relatively simple forms at extreme values of $T$: 
+\be \Lambda(T) =
+\begin{cases}
+ \frac{\rho k_E}{k}\sqrt{\frac{D T}{\pi}}	& T \ll (k_D \Pe)^{-1} \\
+ \frac{\rho k_E V T}{k} & T \gg (k_D \Pe)^{-1}. 
+\end{cases}
+\label{eq:fluccolimitcases}
+\ee
 
-The mathematical form of our results is complicated, although the ideas are conceptually clear.
-Particles alternate through motion and rest, and particle motions have a fluctuating velocity, not a deterministic velocity.
-Nevertheless, turning off velocity fluctuations by setting $D=0$ re-derives the results of \citet{Lisle1998}, who formulated sediment transport as a random walk between motion and rest states.
-Further taking the velocity infinite as the time spent in motion vanishes, while their product $V/k_D = \ell$, the mean step distance, rederives the central result of \citet{Einstein1937}, given originally in \ref{eq:einsteinpdf}.
-
-In either of these limit cases, the mean sediment flux becomes $\bra q \ket = E \ell$, with no dependence on observation scale.
-This indicates that scale dependence in the sediment flux is a consequence of velocity fluctuations among moving particles, at least when observation timescales are relatively short.
+Since the mean flux is $\bra q(T) \ket = \Lambda(T)/T$, the condition on the observation time so that the flux converges to the Einstein value $q_0 = \rho k_E V/k = E \ell$ can be expressed as $T\gg (k_D \Pe)^{-1}$.
+It is related to the Peclet number and is proportional to the time particles spend in motion ($1/k_D$).
+As a result, when particle velocity fluctuations during motions become large comparable to the mean particle velocity, equation \ref{eq:fluccolimitcases} indicates that the flux becomes slow to converge to the Einstein result.
 
@@
 This chapter has generalized the earlier descriptions of individual sediment trajectories \citep[e.g.][]{Lisle1998,Lajeunesse2017} to include velocity fluctuations in the motion state.
-Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
-Calculated in this way, the sediment flux distribution is observation-scale dependent, describing how sediment fluxes depend on the period over which they are measured.
+
+Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches (sec.
+\ref{sec:renewal}) more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
+
+This method to calculate the sediment flux distribution can also be interpreted as a generalization of the nonlocal formulation (sec.
+\ref{sec:nonlocal}) due to \citet{Parker2000,Furbish2012a} to describe sediment flux distributions, rather than just mean values.
 
@@
 
-The sediment flux probability distribution in equation \ref{eq:fluxdist} represents a Poisson distribution with an observation-scale dependent rate.
+The sediment flux probability distribution in equation \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
 Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
 In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
-This exposes a need to generalize a mechanistic theory of the sediment flux I developed here to produce wider fluctuations.
+This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider fluctuations.
 
-Descriptions of sediment transport based on population dynamics in a control volume have realistically-wide fluctuations by incorporating a positive feedback between the number of moving particles and the particle entrainment rate called collective entrainment \citep{Ancey2008,Ancey2014}.
-This feedback generates spatially-correlated waves of moving particles \citep{Ancey2014, Heyman2015} and produces non-exponential interarrival time distributions \citep{Heyman2013} which imply wide-tailed flux distributions when incorporated in renewal theory \citep{Turowski2010,Ancey2020}.
-Collective entrainment has been attributed to particle-particle interactions, such as small granular avalanches and collision-induced entrainment, and to fluid-particle interactions, such as large eddies entraining particles as they sweep downstream \citep{Heyman2014a,Ancey2014,Lee2018}.
+Descriptions of sediment transport based on population dynamics in a control volume have produced realistically-wide fluctuations by incorporating a positive feedback between the number of moving particles and the particle entrainment rate called collective entrainment \citep{Ancey2008,Ancey2014}.
+This feedback generates waves of moving particles \citep{Ancey2014, Heyman2015} and produces non-exponential inter-arrival time distributions \citep{Heyman2013} which imply wide-tailed flux distributions when incorporated in the renewal theory \citep[e.g.][]{Turowski2010,Ancey2020}.
 
-We can imagine generalizing the description in equation \ref{eq:flipfloplangevin} to include such fluid-particle interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
-Particle-particle interactions would be more challenging to include.
+Collective entrainment has been attributed to particle-particle interactions, such as small granular avalanches and collision-induced entrainment, and to fluid-particle interactions, such as large eddies \citep[e.g.][]{Cameron2020} entraining particles en masse as they sweep downstream \citep{Ancey2014,Lee2018}, leading to spatial correlations in sediment transport \citep{Heyman2014}.
+
+Such interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here. 
 The dynamical equation \ref{eq:flipfloplangevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+
 The resulting joint distribution of particle positions and velocities ($P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$) might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
-By analogy to collective entrainment, interacting models which generalize the approach established in this chapter should be capable of producing wide sediment transport fluctuations.
+
+Models which generalize the approach established in this chapter to include collective motions should be capable of producing wide sediment transport fluctuations, but their development will be a challenging next step for research.
+The present chapter only lays the basic groundwork.
 
@@
 
-\subsection{Stochasticity in landscape evolution}
+Figure \ref{fig:fluxconvergence} indicates that the sediment flux described by equation \ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q = E\ell$ predicted by the Einstein theory.
+Because particles move for durations $1/k_D$ on average, the variance of velocity among moving particles can be written $\sigma_V^2 = 2D/k_D$.
+In terms of this quantity, the convergence condition is  $T\gg \sigma_V^2/(k_D V^2),$ now phrased in terms of easily measurable quantities -- the velocity of particles, the average time spent in motion, and the magnitude of velocity fluctuations.
 
-\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{sec:landscape}), writing
-\be (1-\phi) \pt h(x,t) = -\px q(x,t). \label{eq:stocEx}\ee
-In this one-dimensional equation, $h(x,t)$ is a continuous field representing the longitudinal channel profile.
-Traditionally, the Exner equation is evaluated considering the flux as a deterministic quantity \citep{Parker2007,Viparelli2011,An2017}, which provides a solitary trajectory of the profile through time.
+This simple result provides the minimal time required to resolve mean sediment transport rates in terms of measurable properties of particle motions.
+For example, in the experiments of \citet{Martin2012}, this condition implies convergence times of approximately $T \gg 15 s$.
+With wide sediment velocity fluctuations and long particle movement periods, this convergence condition could become plausibly long.
+In this case, the velocity fluctuations of particles will obscure running mean transport rate data from instruments like light tables and acoustic sensors, which have often relied on short averaging intervals \citep{Zimmermann2008,Tsakiris2014,Mendes2016}.
 
-Considering the flux in the Exner equation as a stochastic quantity requires us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2016}.
-In contrast to the traditional analysis, a stochastic Exner equation with fluctuating $q(x,t)$ generates an ensemble of possibilities for the channel profile $z(x)$ at each instant.
-This set of possibilities is characterized by a probability distribution $P(\{z(x)\},t)$ for each possible profile.
-Mathematically, since the channel profile is a field, evaluating such a probability distribution of channel profiles would require statistical field theory \citep{Kardar2007}.
-A starting point on this analysis might be $P(\{z(x)\},t) = \bra \delta[z(x,t)-h(x,t)]\ket,$ where $h(x,t)$ involves the stochastic flux $q(x,t)$ from equation \ref{eq:stocEx}.
-In this equation, $\delta[f(x)]$ is a Dirac delta functional, which generalizes the delta function $\delta(x)$ from a single-valued parameter $x$ to a continuous field $f(x)$.
-Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2001,Nelson2004}, but they have not been introduced in geomorphology.
-Such a description would be an exciting extension of the methodology applied in this thesis so far to better understand landscape evolution.
+\subsection{Limitations and next steps}
+
+It has long been recognized that many other processes, like bedform migration \citep{Hamamori1962,Guala2014}, particle sorting processes \citep{}, or entrainment of clusters \citep{Strom2004,Papanicolaou2018} also lend variability to the sediment flux.
+Understanding these processes has been a longstanding problem, in part because different sources of fluctuations mix within measured transport signals \citep[e.g.][]{Hoey1992,Singh2009,Saletti2015,Dhont2018}.
+
+Although the present model has analyzed on individual particles and does not include these processes, we can wonder if better understanding fluctuations due to individual particles will aid understanding of more complex sources.
+An attempt to subtract the signature of individual particle dynamics from spectra of bedload transport may be a useful step to understand more complex morphological sources of sediment transport variability in channels.
+A comprehensive understanding of the linkage between individual particle dynamics and sediment fluxes, as I have worked toward here, would be the foundation of this approach to the problem.
 
 \section{Summary and Conclusion \label{sec:conc}} 
+This chapter introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motions and rests.
+The motion phase included velocity fluctuations, providing the first analytical model of this type.
 
-This chapter introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motions and rests.
-The motion phase included velocity fluctuations, forming the first analytical model of this type.
-The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying trajectories of individual particles.
-The measurement timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations in terms of a Peclet number.
+The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying dynamics of individual particles.
+The observation timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations in terms of a Peclet number.
 
 These results generalize the bedload trajectory models of \citet{Einstein1937}, \citet{Lisle1998}, and \citet{Lajeunesse2017} to include fluctuating movement velocities, and they reframe earlier renewal models of the sediment flux probability distribution \citep{Lajeunesse2010,Ancey2020} to explicitly include the transport dynamics of individual sediment grains.
-They further quantify the dependence of the sediment flux on the observation scale, and provide guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in simplified configurations when bedforms are absent.
-Next steps are to include interactions between particles to produce wider sediment transport fluctuations, and to evaluate the effects of sediment transport fluctuations on channel morphodynamics.
+
+The work further quantifies the dependence of the sediment flux on the observation scale, and provide guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in simplified conditions such as this model assumes.
+The next step is to include collective motions to produce wider sediment transport fluctuations.
 
@@
 and inverse Fourier transforming provides the desired real-space master equation
+
+At short times $t\ll k^{-1}$, this becomes
+\be \sigma_x(t)^2 \sim \frac{k_E k_D V^2}{k^2} t^2 + \frac{2 k_E D }{k}t, \label{eq:major}\ee
+so provided that $t\ll k^{-1}$, eq. \ref{eq:major} scales as $\sigma_x^2 \sim t^2$ when $t \gg \frac{2Dk}{V^2 k_D}$ and $\sigma_x^2 \sim t$ when $t \ll \frac{2Dk}{V^2 k_D}$.
+Similarly at long times $t\gg k^{-1}$, eq \ref{eq:minor} becomes
+\be \sigma_x^2 \sim \frac{2k_Ek_D V^2}{k^3}t. \ee
+
+Calculated in this way, the sediment flux distribution is observation-scale dependent, describing how sediment fluxes depend on the period over which they are measured.
+I demonstrated that scale-dependence results from velocity fluctuations in the motion state during the movements of individual particles, and I introduced a Peclet number ($\Pe$) to characterize the relative strength of these fluctuations.
+
+We can imagine generalizing the description in equation \ref{eq:flipfloplangevin} to include time-dependent phenomena interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
+
+\subsection{Stochasticity in landscape evolution}
+
+\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{sec:landscape}), writing
+\be (1-\phi) \pt h(x,t) = -\px q(x,t). \label{eq:stocEx}\ee
+In this one-dimensional equation, $h(x,t)$ is a continuous field representing the longitudinal channel profile.
+Traditionally, the Exner equation is evaluated considering the flux as a deterministic quantity \citep{Parker2007,Viparelli2011,An2017}, which provides a solitary trajectory of the bed profile $h(x)$ through time.
+
+Considering the flux in the Exner equation as a stochastic quantity invites us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2016}.
+In contrast to the traditional analysis, a stochastic Exner equation with fluctuating $q(x,t)$ generates an ensemble of possibilities for the channel profile $z(x)$ at each instant.
+This set of possibilities is characterized by a probability distribution $P(\{z(x)\},t)$ for each possible profile.
+
+Mathematically, since the profile is a field, evaluating such a probability distribution of channel profiles requires statistical field theory \citep{Kardar2007}.
+A starting point on this analysis might be $P(\{z(x)\},t) = \bra \delta[z(x,t)-h(x,t)]\ket,$ where $h(x,t)$ involves the stochastic flux $q(x,t)$ from equation \ref{eq:stocEx}.
+In this equation, $\delta[f(x)]$ is a Dirac delta functional, which generalizes the delta function $\delta(x)$ from a single-valued parameter $x$ to a continuous field $f(x)$.
+Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2001,Nelson2004}, but they have not been introduced in geomorphology.
+Such a probabilistic description of bed profiles would be an interesting extension of the methodology applied in this thesis so far to understand landscape evolution in an alternative way.

### 2021-07-07 ch3bedEle.tex
rolling on the conclusion
this is a nightmare

@@
 \chapter{Analysis of bed elevation change and sediment transport fluctuations}
-\label{ch:bedEle}
+\label{ch:ch3}
 
 The transport characteristics of coarse grains moving under a turbulent flow ultimately control a wide set processes within rivers, including the export of contaminants \citep{Malmon2005,Macklin2006}, the success of ecological restoration efforts \citep{Gaeuman2017}, and the response of channel morphology to disturbances \citep{Hassan2017}.
-Although the displacements of individual grains are certainly a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}.
-In response, investigators have developed a stochastic concept of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
+Although the displacements of individual grains are a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}.
+In response, researchers have developed a stochastic concepts of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modelled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
 
 Essentially two types of bedload transport model have been developed from this concept.
-The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2017,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Ancey2008,Furbish2012a}.
+The first type provides the probabilistic dynamics of a small population of tracer grains as they transport downstream \citep{Einstein1937,Hubbell1964, Nakagawa1976,Martin2012,Lajeunesse2017,Wu2019}, while the second provides the statistics of the number of moving grains (``the particle activity'') within a control volume \citep{Einstein1950,Ancey2006,Furbish2012a}.
 In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}.
@@
 Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
-Resting times have puzzled researchers because early experiments show exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments show heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
-A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}; this hypothesis permits surface grains to retain exponential resting times.
+Resting times have puzzled researchers because early experiments showed exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments showed heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
+A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}. This permits surface grains to retain exponential resting times. 
 Conceptually, when grains rest on the surface, material transported from upstream can deposit on top of them, preventing entrainment until its removal, driving up resting times and imparting a heavy tail to the distribution.
-To our knowledge, \citet{Martin2014} have provided the only direct support for this hypothesis by tracking grains through complete cycles of burial and exhumation using a narrow flume with glass walls.
+
+\citet{Martin2014} have provided the only direct support for this hypothesis by tracking grains through complete cycles of burial and exhumation in a narrow flume.
 They observed heavy-tailed resting times of buried grains and described their results with a mathematical model similar to an earlier effort by \citet{Voepel2013}.
 Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation time-series \citep{Redner2007}.
-Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, and neither model explicitly includes bedload transport.
+Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, while neither model explicitly describes bedload transport.
 Models of sedimentary bed evolution incorporating sediment transport processes might enhance understanding of sediment resting times.
@@
 The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
-These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Ancey2008, Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are measured or calculated \citep{Singh2009,Singh2012,Saletti2015}.
-A recent surge in research activity has generated rapid progress and spawned many new inquiries in this subject. 
-For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014,Ma2014, Lee2018}.
-While they deemed this feedback necessary to model realistic activity fluctuations, the implications of this collective entrainment term on bed topography and particle activity changes has not been fully explored.
+These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are observed \citep{Singh2009,Singh2012,Saletti2015}.
+A recent surge in research activity has generated rapid progress in this subject. 
+For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014, Lee2018}.
+While this feedback was deemed necessary to model realistic activity fluctuations, the implications of collective entrainment for the bed topography have not been fully explored.
 
-In this work, we present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
-Our model extends the \citet{Ancey2008} model to describe the interplay between bedload flux and bed elevation fluctuations in a control volume.
-This development permits a systematic study of the repercussions of collective entrainment, and it frames bed elevation changes as a direct consequence of the sediment transport process.
-Our model has two key assumptions: (1) bedload erosion and deposition can be characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and (2) these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
-We study statistical characteristics of bedload transport, bed elevation, and resting times of sediment undergoing burial using a mixture of numerical simulations and analytical approximations.
-We introduce the stochastic model in section \ref{sec:model}, and we solve it in section \ref{sec:solution} with a mixture of numerical and analytical techniques.
-We discern several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates and the local bed elevation.
-We present these features in section \ref{sec:results}.
-We conclude with the implications of our results and speculate on topics for future research in sections \ref{sec:discussion} and \ref{sec:conclusion}.
+In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
+This model extends the birth-death model approach reviewed in section \ref{sec:elebirthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
+This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried in the bed subsurface.
+The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
+
+Below, section \ref{sec:elemodel} introduces the stochastic model and section \ref{sec:elesolution} solves it with a mixture of numerical and analytical techniques.
+Model solutions illustrate several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates with bed elevation change.
+Section \ref{sec:eleresults} present these features, while sections \ref{sec:elediscussion} and \ref{sec:eleconclusion} conclude with the implications of these features and the suggestions they raise for future research.
 
 \section{Stochastic model of bedload transport and bed elevations}
-\label{sec:model}
+\label{sec:elemodel}
 
-We prescribe a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:definition}.
-We define $m$ relative to the mean number of grains within the control volume, so that it can be either positive or negative.
+The model in this chapter is formulated by considering a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:eledefinition}.
+The value $m$ is defined relative to the mean number of grains within the control volume, so that it can be either positive or negative.
 $n$ is always a positive integer including $0$.
-For simplicity, we consider all particles as approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
-Following \citet{Ancey2008}, we prescribe four events that can occur at any instant to modify the populations $n$ and $m$, and we characterize these events using probabilities per unit time (rates).
+
+All particles are considered approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
+Following \citet{Ancey2008}, basically four events can occur at any instant to modify the populations $n$ and $m$. These events are characterized using probabilities per unit time (rates).
 These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
-The four events are depicted as arrows in figure \ref{fig:definition}.
+The four events are depicted as arrows in figure \ref{fig:eledefinition}.
 As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
@@
 The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using (\ref{eq:ele}).}
-	\label{fig:definition}
+	\label{fig:eledefinition}
 \end{figure}
-The populations $n$ and $m$ provide the bulk bedload flux $q_s$ and the local bed elevation $z$.
-The mean bedload transport rate is given by $q_s = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
+
+The populations $n$ and $m$ provide the volumetric bedload flux $\Phi$ and the local bed elevation $z$.
+The mean transport rate is given by $\Phi = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
 \sum_{n,m}nP(n,m) $ is the mean number of grains in motion \citep{Charru2004, Ancey2008, Furbish2012a}.
 The bed elevation is related to $m$ through the packing geometry of the bed.
-To quantify this, we introduce a packing fraction $\phi$ of grains in the bed \citep{Bennett1972}, and for simplicity we consider the bed as two-dimensional \citep{Einstein1950, Paintal1971}.
-The deviation from the mean bed elevation is then
+This relationship depends on the packing fraction $\phi$ of grains in the bed \citep{Bennett1972}.
+Considering the bed as two-dimensional \citep{Einstein1950, Paintal1971}, the deviation from the mean bed elevation can be expressed as
 \begin{equation} z(m) = \frac{\pi a^2}{\phi L}m = z_1 m. \label{eq:ele}\end{equation}
-The constant $z_1 = \pi a^2/(\phi L)$ is an important scale of the problem. 
+The constant $z_1 = \pi a^2/(\phi L)$ is an important scale in the problem. 
 $z_1$ is the magnitude of bed elevation change in an average sense across the control volume associated with the addition or removal of a single grain.
 
-Bed elevation changes modify the likelihood of entrainment and deposition in a negative feedback \citep{Sawai1987, Wong2007}; that is, aggradation increases the likelihood of entrainment, while degradation increases the likelihood of deposition.
+Bed elevation changes modify the likelihood of entrainment and deposition in a negative feedback \citep{Sawai1987, Wong2007}.
+Aggradation increases the likelihood of entrainment, while degradation increases the likelihood of deposition.
 \citet{Wong2007} concluded that bed elevation changes induce an exponential variation in entrainment and deposition probabilities, while \citet{Sawai1987} concluded that the variation is linear.
-For simplicity, we incorporate the scaling of \citet{Sawai1987} and note its equivalence to the \citet{Wong2007} scaling when bed elevation changes are small.
-Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, we expect the erosion and deposition feedbacks to have the same strength.
-That is, as bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.
-Merging these ideas with those of \citet{Ancey2008}, we write the four possible transitions with local bed elevation-dependent entrainment and deposition rates as
+For simplicity, this chapter utilizes the scaling of \citet{Sawai1987}. This scaling is equivalent to the  \citet{Wong2007} scaling when bed elevation changes are small.
+Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, the erosion and deposition feedbacks are considered to have the same strength.
+As bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.
+
+Merging these ideas with those of \citet{Ancey2008} produces expressions for the four possible transitions with local bed elevation-dependent entrainment and deposition rates:
 \begin{align}
@@
 At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
-Away from this elevation, entrainment and deposition are alternatively suppressed and enhanced depending on the sign of $m$, constituting a feedback between bed elevation changes and erosion and deposition.
-We refer to $\kappa$ as a coupling constant since it controls the strength of this feedback. We later demonstrate the relationship
-\begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
-\end{equation}
-where $l$ is a characteristic length scale of bed elevation change that we interpret as the active layer depth \citep{Wong2007,Church2017}.
+Away from this elevation, entrainment and deposition are alternatively suppressed or enhanced depending on the sign of $m$, constituting a feedback between bed elevations and erosion / deposition.
+
 All four rates are independent of the past history of the populations and depend only on the current populations $(n,m)$. 
@@
 
-We write the master equation for the probability flow using the forward Kolmogorov equation $\partial P(n,m;t)/\partial t = 
+The master equation for the probability flow can be written using the Komogorov equation
+ $\partial P(n,m;t)/\partial t = 
 \sum_{n',m'} [R(n,m|n',m')P(n',m';t)-R(n',m'|n,m)P(n,m;t)]$ \citep{Cox1965, Gillespie1991, Ancey2008} as 
@@
 	\{ \nu + \lambda+ \mu n (1+\kappa m) +  \sigma n ( 1- \kappa m) + \gamma n \}P(n,m;t).
-	\label{eq:master}
+	\label{eq:elemaster}
 \end{multline}
-The joint probability distribution $P(n,m;t)$ solving this equation fully characterizes the statistics of $n$ and $m$ -- proxies for the bedload flux and local bed elevation.
+The joint distribution $P(n,m;t)$ that solves this equation fully characterizes the statistics of $n$ and $m$ -- proxies for the bedload flux and local bed elevation.
+
 The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
-We anticipate that solutions of (\ref{eq:master}) will adjust from the initial conditions to a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium conditions.
+The solutions of (\ref{eq:elemaster}) should be expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
 Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
 This Master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
-In our context, the two populations are the moving and stationary grains in the volume.
+In context of this chapter, the two populations are the moving and stationary grains in the control volume.
 
 \section{Model solutions}
-\label{sec:solution}
+\label{sec:elesolution}
 
-Unfortunately, equation (\ref{eq:master}) does not appear to admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
+Unfortunately, equation (\ref{eq:elemaster}) does not admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
 The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
-In response to this difficulty, we resort to numerical methods and analytical approximations, simulating equation (\ref{eq:master}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007} and solving it approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
-The simulation algorithm is described in \ref{sec:numerical}, and analytical approximations are described in \ref{sec:analytical}.
+
+Because the model is Markovian, it is nevertheless possible to numerically simulate equation (\ref{eq:elemaster}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007}. 
+In conditions when the coupling constant $\kappa$ between the entrainment/deposition rates and the bed elevation is weak, it is also possible to solve the master equation approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
+The Gillespie simulation algorithm is described below in \ref{sec:numerical}, while these analytical approximations are described in \ref{sec:analytical}.
 
@@
 
-The Gillespie algorithm leverages the defining property of a Markov process: when transition rates are independent of history, time intervals between transitions are \begin{wraptable}{l}{0.5\textwidth}
+The Gillespie algorithm takes advantage of the defining property of a Markov process: when transition rates are independent of history, time intervals between transitions are
+exponentially distributed \citep{Cox1965}.
+
+As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
+then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}).
+The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
+This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
+The appendix chapter \ref{ch:gill} provides additional background on the stochastic simulation method.
+
+Using this method, I simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
+These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
+
+\begin{wraptable}{r}{0.6\textwidth}
 	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ from \citet{Ancey2008}.
 Units are $s^{-1}$ (probability/time).
-In our model, bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
+Bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
 	\begin{tabular}{cccccc} \\ 
@@
 	\end{tabular}
-\end{wraptable}exponentially distributed \citep{Cox1965}.
-As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
-then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}).
-The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
-This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
-We provide additional background on the stochastic simulation method in the supplementary material and refer the reader to \citet{Gillespie2007} for more detail.
-
-Using this method, we simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
-These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
-For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, we used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
+\end{wraptable}
+For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, I used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
 Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
-In all simulations, we take the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and we set $L = 22.5$cm and $a = 0.3$cm in accord with the \citet{Ancey2008} experiments.
-Each simulation was run for $250$ hours of virtual time, a period selected to ensure neat convergence of particle activity and bed elevation statistics.
+In all simulations, I took the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and I set $L = 22.5$cm and $a = 0.3$cm, values from the original \citet{Ancey2008} experiments.
+Each simulation was run for $250$ hours of virtual time, a period selected to ensure convergence of particle activity and bed elevation statistics.
 
@@
 
-We approximately decouple the $n$ and $m$ dynamics in equation (\ref{eq:master}) using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
+The $n$ and $m$ dynamics in equation (\ref{eq:elemaster}) can be approximately decoupled using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
 These inequalites mean many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
-We concentrate on steady state conditions $\partial P/\partial t = 0$ and introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to equation (\ref{eq:master}), with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$.
+Concentrating on steady state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to equation (\ref{eq:elemaster}), with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$ \citep[e.g.][]{Haken1978}.
 This provides the steady state equation
@@
 \end{multline}
-Unforunately, these two equations are no easier to solve than the original master equation, since the coupling between $n$ and $m$ is not reduced in equation (\ref{eq:decomp}).
+Unforunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in equation (\ref{eq:decomp}).
 
@@
 Taking this limit in equation (\ref{eq:decomp}), summing over $m$, and using $\langle m \rangle = 0$ reproduces the \citet{Ancey2008} particle activity model.
-As shown by \citet{Ancey2008}, this has solution
+As shown in section \ref{sec:birthdeath}, this has solution
 \begin{equation} A(n) = \frac{\Gamma(r+n)}{\Gamma(r)n!}p^r(1-p)^n.\label{eq:ancey}\end{equation}
@@
 This mean field equation matches the discrete Ornstein-Uhlenbeck model of bed elevation changes developed by \citet{Martin2014}.
-We summarize that the independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model we present in a mean field approximation when $\kappa$ is insignificant.
+The independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model equation \ref{eq:elemaster} in a mean field approximation when $\kappa$ is insignificant.
 
-In the supplementary information we show the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within equation (\ref{eq:ou}) provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
-As we will demonstrate in section \ref{sec:results}, and as we have already suggested with equation (\ref{eq:active}), this is a poor approximation to the bed elevation variance.
+In the supplementary information I show that the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within equation (\ref{eq:ou}) provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
+As I will demonstrate in section \ref{sec:eleresults}, and as I have already suggested with equation (\ref{eq:active}), this is a poor approximation to the bed elevation variance.
 Nevertheless, this approximation does capture the Gaussian shape of the bed elevation distribution.
 The essential issue with this mean field approach is that the conditional mean particle activity $\langle n | m \rangle$ varies significantly with $m$ in actuality, especially when collective entrainment contributes to the mean entrainment rate $E$.
-We will discuss these points subsequently when developing more refined approximations and presenting numerical results.
+I will discuss these points subsequently when developing more refined approximations and presenting numerical results.
 
@@
 captures the general features of the conditional mean particle activity.
-As we show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
+As I show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
 \begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
@@
 $M_0$ is a normalization constant. 
-As we will demonstrate, this closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:master}), at least for the flow parameters in table \ref{tab:anceyparams}.
+As we will demonstrate, this closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.
 
 \section{Results}
-\label{sec:results}
+\label{sec:eleresults}
 
 From the initial conditions, all simulations show a rapid attainment of steady-state stochastic dynamics of $n$ and $m$ which support a time-independent joint distribution $P(n,m)$.
-We show an elevation time-series in the bottom panel of figure \ref{fig:definition}.
-In order to describe the implications of coupling bedload transport to bed elevation changes, we present the numerical and analytical results for the probability distributions of bedload transport and bed elevations in section \ref{sec:pdf} and the statistical moments of these quantities in section \ref{sec:mom}.
-We isolate the effects of collective entrainment on bed elevation changes in section \ref{sec:colent}, and we present the resting times of sediment undergoing burial in section \ref{sec:rtcdf}.
+The bottom panel of figure \ref{fig:eledefinition} shows an elevation time-series.
+In order to describe the implications of coupling bedload transport to bed elevation changes, the numerical and analytical results for the probability distributions of bedload transport and bed elevations are presented in section \ref{sec:elepdf} and the statistical moments of these quantities in section \ref{sec:elemom}.
+The effects of collective entrainment on bed elevation changes are studied in section \ref{sec:elecolent}, and the resting times of sediment undergoing burial are evaluated in section \ref{sec:elertcdf}.
 
 \subsection{Probability distributions of bedload transport and bed elevations}
-\label{sec:pdf}
+\label{sec:elepdf}
 
-We compute this joint distribution by counting occurrences of the states $(n,m)$ in the simulated time series.
-From this joint distribution we compute marginal distributions $P(n)$ and $P(m)$ by summing over $m$ and $n$ respectively.
+The joint distribution is computed numerically by counting occurrences of the states $(n,m)$ in the simulated time series.
+From this joint distribution the marginal distributions $P(n)$ and $P(m)$ are obtained by summing over $m$ and $n$ respectively.
 A representative subset of these marginal distributions is displayed in figure \ref{fig:pdfs} alongside the approximate results of equations (\ref{eq:ancey}) and (\ref{eq:ou2}).
@@
 \end{figure}
-The mean field equation (\ref{eq:ancey}) for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match equation (\ref{eq:ou2}), having Gaussian profiles consistent with our assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.
+The mean field equation (\ref{eq:ancey}) for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match equation (\ref{eq:ou2}), having Gaussian profiles consistent with the assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.
 
 \subsection{Statistical moments of bed elevation and the particle activity}
-\label{sec:mom}
+\label{sec:elemom}
 
-We calculate the moments of $n$ and $m$ by summing over $P(n,m)$. 
+The moments of $n$ and $m$ are calculated by summing over $P(n,m)$. 
 The $j$th order unconditional moment of the particle activity $n$ derives from
@@
 \begin{equation} \langle n^j|m \rangle = \sum_{n}n^j P(n,m) .\end{equation}
+The moments of $m$ show no dependence on the value of $n$. 
+The mean elevation is always $\langle m \rangle = 0 $ due to the initial assumption of symmetry in the entrainment and deposition rate scaling with $m$. 
 \begin{wrapfigure}{r}{0.5\textwidth}
@@
 \end{wrapfigure}
-We observe no dependence of the moments of $m$ on the value of $n$. 
-The mean elevation is always $\langle m \rangle = 0 $ due to our initial assumption of symmetry in the entrainment and deposition rate scaling with $m$. 
-Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in equation (\ref{eq:ou2}); this result supports our earlier assertion that $l$ is a characteristic length scale of bed elevation fluctuations.
+Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in equation (\ref{eq:ou2}); this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
 The close correspondence between the mean field approximation and the numerical simulations in figure (\ref{fig:pdfs}a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
-We find them to be identical within numerical uncertainty.
+They appear to be identical within numerical uncertainty.
 
@@
 The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
-These results indicate that bedload statistics measurements on short timesclaes could be severely biased by departures from the mean bed elevation.}
+These results indicate that bedload statistics measurements on short timescales could be severely biased by departures from the mean bed elevation.}
 	\label{fig:condmoms}
 \end{figure}
-Figure (\ref{fig:condmoms}a) demonstrates that the \citet{Ancey2008} flow conditions support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
-The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry in this quantity.
+Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparam} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
+The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry.
 Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
 For the parameters of the \citet{Ancey2008} experiments, figure (\ref{fig:condmoms}b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
-We summarize that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
+The summary is that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
 
 \subsection{Collective entrainment and bedload activity fluctuations}
-\label{sec:colent}
-Noting that bed elevations regulate the particle activity moments, we now study the influence of collective entrainment on this effect by modifying the relative proportion of the individual to collective contributions in the mean entrainment rate $E=\lambda + \sigma \langle n \rangle $.
-Using the equilibrium condition $E=D$, we determine the fraction of entrainment due to the collective process is $f = \mu\langle n \rangle/E = \mu/\sigma$.
-Using this fraction, we can hold $E$ constant and modify the prevalence of the collective entrainment process by setting $\lambda = E(1-f)$ and $\mu= \sigma f$.
-As we interpolate $f$ between zero and one, the particle activity component of the master equation \ref{eq:master} interpolates from a purely Poissonian model \citep{Ancey2006} to a negative binomial model \citep{Ancey2008}, isolating the imprint of collective entrainment on particle activity statistics over a dynamic sedimentary bed.
+\label{sec:elecolent}
+Noting that bed elevations regulate the particle activity moments, the next step is to study the influence of collective entrainment on this effect by modifying the relative proportion of the individual to collective contributions in the mean entrainment rate $E=\lambda + \sigma \langle n \rangle $.
+The equilibrium condition $E=D$ implies that the fraction of entrainment due to the collective process is $f = \mu\langle n \rangle/E = \mu/\sigma$.
+This fraction can be used to hold $E$ constant and modify the prevalence of the collective entrainment process by setting $\lambda = E(1-f)$ and $\mu= \sigma f$.
+Interpolating $f$ between zero and one interpolates the particle activity component of the master equation \ref{eq:elemaster} from a purely Poissonian model to a negative binomial model, isolating the effect of collective entrainment on the particle activity statistics over a dynamic sedimentary bed.
 \begin{figure}[!htbp]
@@
 Figure \ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed.
-When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect we introduced in section \ref{sec:mom}.
+When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect identified in section \ref{sec:elemom}.
 
 \subsection{Resting times of sediment undergoing burial}
-\label{sec:rtcdf}
+\label{sec:elertcdf}
 
@@
 \section{Discussion}
-\label{sec:discussion}
+\label{sec:elediscussion}
 
@@
 
-In this paper, we have built on earlier works \citep{Ancey2008,Martin2014} to include the vertical dimension of bed elevation dynamics, study the interplay between bedload transport and bed elevation fluctuations, and investigate resting time distributions of sediment undergoing burial.
-To our knowledge, this model is the first description of bedload transport and bed elevations as a coupled stochastic population model based on individual grains.
+This chapter has modified the birth-death approach to describe sediment transport in a control volume.
+This model was applied to study the interplay between bedload transport and bed elevation fluctuations and to investigate resting time distributions of sediment undergoing burial.
+This model is the first description of bedload transport and bed elevations as a coupled stochastic population model based on individual grains.
 Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
@@
 
-Our inclusion of coupling between the bed elevation and entrainment and deposition rates revealed a novel dependence of particle activity on bed elevation changes,  highlighting a new consequence of the collective entrainment process \citep{Ancey2008,Lee2018}.
+The inclusion of coupling between the bed elevation and entrainment and deposition rates revealed a novel dependence of particle activity on bed elevation changes, highlighting a new consequence of the collective entrainment process \citep{Ancey2008,Lee2018}.
 This coupling develops a significant variation of the particle activity moments with deviations of the bed from its mean elevation.
-We isolated the role of collective entrainment in this bedload activity regulation, and pointed out that particle activity variations with bed elevations dissapear in the absence of collective entrainment.
-Finally, we obtained resting times for sediment undergoing burial within the sedimentary bed by analyzing return times from above in the bed elevation time-series.
-We found heavy-tailed power law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
-The distribution tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.
+These particle activity variations with bed elevation changes disappear in the absence of collective entrainment.
 
-As our model builds on earlier works describing particle activity and bed elevation changes independently, it also reduces to these works in simplified limits when the coupling between the particle activity and bed elevation vanishes.
-With the mean field approach in section \ref{sec:analytical}, we derived the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity as simplified limits of our coupled model.
-While the mean field description of bed elevations over-predicts the bed elevation variance by approximately a factor of two, it does capture the Gaussian shape of the bed elevation distribution, and its conclusions on the tail characteristics of resting time distributions for sediment undergoing burial are identical to ours within the numerical uncertainty: \citet{Martin2014} described a power-law distribution with tail parameter $\alpha \approx 1$ which falls neatly within our estimation $\alpha = 1.18 \pm 0.32$.
-In addition to our original contributions, we have corroborated the models of \citet{Ancey2008} and \citet{Martin2014} from an alternate perspective, showing their results to be mostly robust when accounting for bed elevation changes.
+Finally, the model allowed resolution of resting times for sediment undergoing burial within the sedimentary bed.
+These timescales were obtained by analyzing return times from above in the bed elevation timeseries.
+This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
+The power-law tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.
+
+As the model in this chapter builds on earlier works describing particle activity and bed elevation changes independently, it also reduces to these works in simplified limits when the coupling between the particle activity and the local bed elevation vanishes.
+With the mean field approach in section \ref{sec:eleanalytical}, the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity were derived as simplified limits of the coupled model developed in this chapter.
+While the mean field description of bed elevations over-predicts the bed elevation variance by approximately a factor of two, it does capture the Gaussian shape of the bed elevation distribution, and its conclusions on the tail characteristics of resting time distributions for sediment undergoing burial are identical to the model in this chapter within the numerical uncertainty.
+\citet{Martin2014} described a power-law distribution with tail parameter $\alpha \approx 1$ which falls neatly within the estimation $\alpha = 1.18 \pm 0.32$ presented here.
 
@@
 
-The model we have presented computes statistical characteristics of the bedload particle activity and bed elevation within a control volume by assuming all particles on the bed surface have similar mobility characteristics while sediment transport and bed topography are in equilibrium.
+The model presented in this chapter computes statistical characteristics of the bedload particle activity and bed elevation within a control volume by assuming all particles on the bed surface have similar mobility characteristics while sediment transport and bed topography are in equilibrium.
 In actuality, particles span a range of sizes, and spatial organization occurs both in the forces imparted to particles by the flow \citep{Shih2017,Amir2014} and in the mobility characteristics of particles on the bed surface \citep{Charru2004, Hassan2008, Nelson2014}.
@@
 Models chaining multiple control volumes together have shown spatial correlations in the particle activity as a result of collective entrainment \citep{Heyman2014, Ancey2015}, and similar approaches have also been applied to study correlations in turbulent flows \citep{Gardiner1983}.
-In light of this work, we consider the model we have presented as a preliminary step toward a multiple-cell model of particle activities and bed elevation changes with potential to express spatial correlations between longitudinal profile and particle activity statistics.
+In light of this work, the model presented in this chapter is considered a preliminary step toward a multiple-cell model of particle activities and bed elevation changes with potential to express spatial correlations between longitudinal profile and particle activity statistics.
 
-Like \citet{Martin2014}, we obtained heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
+Like the theory developed by \citet{Martin2014}, the model \ref{eq:elemaster} produces heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
 This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016a}. 
-Our resting time distributions show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of our results.
+The resting time distributions derived in section \ref{sec:elertcdf} show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of the results.
+
 Divergent mean resting time distributions present a paradox, since they imply all particles should eventually be immobile, violating the equilibrium transport assumption.
@@
 
-Finally, we studied probability distribution functions and first and second moments of the particle activity and bed elevation, making novel conclusions about coordination between the statistical characteristics of these quantites which deserve experimental testing.
-In the last decade, particle tracking experiments have emerged \citep{Lajeunesse2010,Roseberry2012,Martin2014, Fathel2015,Heyman2016,Liu2019}, that allow joint resolution of bed elevations and bedload transport.
-A suitably designed experiment could test our prediction that bed elevations regulate particle activity statistics, as essentially represented in figures \ref{fig:condmoms} and \ref{fig:colent}. 
-However, we have left many other statistical characteristics of bedload transport for future studies. 
-For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic.
-Statistical quantities can either be monoscaling or multiscaling across the observation scale \citep{Sornette2006}, and we currently lack physical understanding and general conclusions about the scale dependence of particle activity and bed elevation signals.
-The model we have presented shows statistical monoscaling for both quantites \citep[e.g.][]{Saletti2015}, whereas other experiments indicate statistical multiscaling \citep{Aberle2006,Singh2009,Singh2012}.
-We consider this topic to go beyond the scope of the present work, and we have focused on statistical characteristics at the highest temporal resolutions, with no averaging over the observation scale.
+Finally, this chapter presented the probability distribution functions and first and second moments of the particle activity and bed elevation, indicating novel coordination between the statistical characteristics of these quantites which deserve experimental testing.
+Some relatively recent particle tracking experiments have demonstrated joint resolution of bed elevations and bedload transport \citep{Martin2014,Heyman2016}.
+A suitably designed experiment using this methodology could test the prediction that bed elevations regulate particle activity statistics, as essentially represented in figures \ref{fig:condmoms} and \ref{fig:colent}.
+
+Many other statistical characteristics of bedload transport have been left for future studies. 
+For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic which was previously referenced in section \ref{sec:renewal}.
+Whether bed elevation distributions are also scale-dependent like the sediment flux distribution of chapter \ref{ch:flux} is an interesting question for future research.
 
 \section{Summary and conclusion}
-\label{sec:conclusion}
+\label{sec:eleconclusion}
 
-We developed a stochastic model for particle activity and local bed elevations including feedbacks between elevation changes and the erosion and deposition rates.
+This chapter presented a stochastic model for particle activity and local bed elevations including feedbacks between elevation changes and sediment transport.
 This model includes collective entrainment, whereby moving particles tend to destabilize stationary ones.
-We analyzed this model using a mixture of numerical and analytical methods and provided two key results:
+The model was analyzed using a mixture of numerical and analytical methods, and two key results were presented:
 \begin{enumerate}
@@
 \end{enumerate}
-These results imply measurements of bedload transport statistics could be severely biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
-Next steps are to generalize our model to a multi-cell framework and to study spatial correlations in bed elevation and particle activity statistics.
+These results imply measurements of bedload transport statistics could be biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
+The next step is to generalize the model in this chapter to a multi-cell framework to study channel morphodynamics in a stochastic framework.

### 2021-07-07 app_ch2.tex
very close

@@
 The position probability distribution can be obtained from eq.
-\ref{flippymaster} for the initial conditions $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ using Fourier transforms in space and Laplace transforms in time.
+\ref{eq:flippymaster} for the initial conditions $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ using Fourier transforms in space and Laplace transforms in time.
 Taking these transforms provides

### 2021-07-07 app_ch5.tex
got all references fixed up... except for a few which were overwritten by mendeley like taylor 1920. Still need to check carefully what's actually cited.
rolling on the conclusion
very close

@@
 
-\chapter{Ch. 5 calculations}
+\chapter{Calculations involved in the Collisional Langevin model}
 
 \section{Derivation of Master Equation}
+\label{sec:langmasterderiv}
 To derive the master equation from \ref{eq:langevin}, we temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 d} h$ with $h$ distributed as $f(h)$.
@@
 \section{Derivation of Steady-state solution}
-
+\label{sec:langsteadyderiv}
 Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
@@
 \section{Calculation of the moments}
+\label{sec:langmoments}
 Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
@@
 \section{Weak and strong collision limits}
+\label{sec:langextremes}
 Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
 The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
-Following \cite{Hall1970}, this suggests normalizing the distribution $P(u)$ using 
+Following \cite{Hall1989}, this suggests normalizing the distribution $P(u)$ using 
 \be z = \frac{u - \bar{u}}{\sigma_u} \ee

### 2021-07-07 ch1intro.tex
very close

@@
 	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise eq.
-\ref{eq:lislerando}, while panel (b) shows the tilted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (with mean duration $1/k_E$).
+\ref{eq:lislelangevin}, while panel (b) shows the tilted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (with mean duration $1/k_E$).
 }
@@
 \subsection{Einstein's probabilistic approach}
+\label{sec:einflux}
 
@@
 This can be interpreted as an average over the detailed configurations of individual grains \citep{Coleman2009}.
-This averaging process is expected to obscure the physics whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Schobe2021}.
+This averaging process is expected to obscure the physics whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
 

### 2021-07-07 ch2flipFlop.tex
very close

@@
 
-\subsection{Derivation of the master equation for the sediment position distribution $P(x,t)$}
+\subsection{Derivation of the master equation for the sediment position distribution}
 \label{sec:floppymastereq}
@@
 Such interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here. 
-The dynamical equation \ref{eq:flipfloplangevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+The dynamical equation \ref{eq:flippylangevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
 
-The resulting joint distribution of particle positions and velocities ($P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$) might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
+The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
 

### 2021-07-07 ch3bedEle.tex
very close

@@
 In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
-This model extends the birth-death model approach reviewed in section \ref{sec:elebirthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
+This model extends the birth-death model approach reviewed in section \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
 This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried in the bed subsurface.

### 2021-07-07 ch4downDiff.tex
very close

@@
 
-In this study, we incorporate these two components into Einstein's original approach to describe three diffusion ranges with a physically based model, as called for by \citet{Nikora2002}.
+In this study, I incorporate these two components into Einstein's original approach to describe three diffusion ranges with a physically based model, as called for by \citet{Nikora2002}.
 Einstein was a giant in river geophysics and fostered an entire paradigm of research leveraging and generalizing his stochastic methods \citep{Hubbell1964, Yano1969, Yang1971, Gordon1972, Nakagawa1976,Paintal1971}.
 Einstein's model can be viewed as a pioneering application of the continuous time random walk (CTRW) developed by \citet{Montroll1965} in condensed matter physics to describe the diffusion of charge carriers in solids.
-To incorporate motion intervals and sediment burial, we utilize the multi-state CTRW developed by \citet{Weiss1976, Weiss1994} that extends the CTRW of \citet{Montroll1965}.
-Below, we develop and solve the model in section \ref{sec:model}.
-Then, we discuss the predictions of our model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in sections \ref{sec:discussion} and \ref{sec:conclusion}.
+To incorporate motion intervals and sediment burial, I utilize the multi-state CTRW formalism developed by \citet{Weiss1976, Weiss1994} that extends the CTRW of \citet{Montroll1965}.
+Below, I develop and solve the model in section \ref{sec:model}.
+Then, I discuss the predictions of this model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in sections \ref{sec:discussion} and \ref{sec:conclusion}.
 
@@
 \label{sec:assumptions}
-We construct a three-state random walk where the states are motion, surface rest, and burial, and we label these states as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
-Our target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if we know it started with the initial distribution $p(x,0)=\delta(x)$.
-We characterize times spent moving or resting on the surface by exponential distributions $\psi_2(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
+Here I construct a three-state random walk where the states are motion, surface rest, and burial. These states are labelled as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
+The target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if it is known to have started with the initial distribution $p(x,0)=\delta(x)$.
+Times spent moving or resting on the surface are characterized by exponential distributions $\psi_2(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
 We expect our conclusions will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
-We consider grains in motion to have characteristic velocity $v$ \citep{Lisle1998,Lajeunesse2017}, and we model burial as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried, meaning $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
-We specify the initial conditions with probabilities $\theta_1$ and $\theta_2$ to be in rest and motion at $t=0$, and we require $\theta_1+\theta_2=1$ for normalization.
+Grains in motion are considered to have a characteristic velocity $V$ \citep{Lisle1998,Lajeunesse2017}, and burial is modelled as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried.
+This means $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
+The initial conditions are specified with probabilities $\theta_1$ and $\theta_2$ to be in rest or motion at $t=0$. Normalization requires $\theta_1+\theta_2=1$.
 
 \subsection{Governing equations}
-Using these assumptions, we derive the governing equations for the set of probabilities $\om_{ij}(x,t)$ that a transition occurs from state $i$ to state $j$ at position $x$ and time $t$ using the statistical physics approach to multi-state random walks \citep{Weiss1994,Schmidt2007,Weeks1998}.
+With these assumptions, the governing equations for the set of probabilities $\om_{ij}(x,t)$ that a transition occurs from state $i$ to state $j$ at position $x$ and time $t$ can be derived using the statistical physics approach to multi-state random walks \citep{Weiss1994,Schmidt2007,Weeks1998}.
 Denoting by $g_{ij}(x,t)$ the probability for a particle to displace by $x$ in a time $t$ within the state $i$ before it transitions to the state $j$, the transition probabilities $\om_{ij}(x,t)$ sum over all possible paths to the state $i$ from previous locations and times:
 \be \om_{ij}(x,t) = \theta_i g_{ij}(x,t) + \sum_{k=0}^2 \int_0^x dx' \int_0^t dt' \om_{ki}(x',t')g_{ij}(x-x',t-t').\label{eq:g1}\ee
-Defining another set of probabilities $G_i(x,t)$ that a particle displaces by a distance $x$ in a time $t$ within the state $i$ and possibly remains within the state, we perform a similar sums over paths for the probabilities to be in the state $i$ at $x$, $t$:
+Defining another set of probabilities $G_i(x,t)$ that a particle displaces by a distance $x$ in a time $t$ within the state $i$ and possibly remains within the state, a similar sums over paths for the probabilities to be in the state $i$ at $x$, $t$ produces:
 \be p_i(x,t) = \theta_i G_i(x,t) + \sum_{k=0}^2 \int_0^x dx' \int_0^t dt' \om_{ki}(x',t')G_i(x-x',t-t').\label{eq:g2}\ee
@@
 \be p(x,t) = \sum_{k=0}^2 p_k(x,t) \ee
-This joint density is completely determined from the solutions of equations (\ref{eq:g1}-\ref{eq:g2}). We only need to specify the distributions $g_{ij}$ and $G_i$.
+This joint density is completely determined from the solutions of equations (\ref{eq:g1}-\ref{eq:g2}) given specification of the distributions $g_{ij}$ and $G_i$.
 
@@
 
-We construct these distributions from the assumptions described in section \ref{sec:assumptions}.
-Since particles resting on the bed surface bury in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, we obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$.
-Since motions have velocity $v$ for times distributed as $\psi_2(t)$, we have $g_{21}(x,t) = \delta(x-vt)k_2e^{-k_2 t}$.
+These distributions can be constructed from the assumptions described in section \ref{sec:assumptions}.
+Since particles resting on the bed surface bury in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, these assumptions obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$.
+Since motions have velocity $v$ for times distributed as $\psi_2(t)$, the assumptions provide $g_{21}(x,t) = \delta(x-vt)k_2e^{-k_2 t}$.
 Since burial is quasi-permanent, all other $g_{ij} = 0$.
 The $G_i$ are constructed in the same way except using the cumulative probabilities $\int_t^\infty dt'\psi_i(t) = e^{-k_i t}$, since these characterize motions and rests that are ongoing \citep{Weiss1994}.
-We obtain $G_1(x,t) = \delta(x)e^{-k_1t}$ and $G_2(x,t) = \delta(x-vt)e^{-k_2 t}$.
+Using the cumulative probabilities provides $G_1(x,t) = \delta(x)e^{-k_1t}$ and $G_2(x,t) = \delta(x-vt)e^{-k_2 t}$.
 
-To solve equations (\ref{eq:g1}-\ref{eq:g2}) with these $g_{ij}$ and $G_i$, we take Laplace transforms in space and time $(x,t \rightarrow 
-\eta,s) $ using a method similar to \citet{Weeks1998} to unravel the convolution structure of these equations, eventually obtaining
-\be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}, \label{eq:nicedist}\ee
-where $k' = k_1 + k_2$. Inverting this result using known Laplace transforms \citep{Prudnikov1986,Arfken1985} obtains
+Equations (\ref{eq:g1}-\ref{eq:g2}) with these $g_{ij}$ and $G_i$ are solved using Laplace transforms in space and time $(x,t \rightarrow 
+\eta,s).$ This method, similar to \citet{Weeks1998}, unravels the convolution structure of these equations, eventually producing
+\be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}, \label{eq:diffnicedist}\ee
+where $k' = k_1 + k_2$. Inverting this result using known Laplace transforms \citep{Prudnikov1992a,Arfken1985} obtains
 \begin{align}
@@
 This result generalizes the earlier results of \citet{Lisle1998} and \citet{Einstein1937} to include sediment burial.
-In this equation, we used the shorthand notations $\xi = k_2 x/v$, $\tau = k_1(t-x/v)$, and $\Omega = (\kappa+k_1)/k_1$ \citep{Lisle1998}.
+This equation uses the shorthand notations $\xi = k_2 x/v$, $\tau = k_1(t-x/v)$, and $\Omega = (\kappa+k_1)/k_1$ \citep[c.f.][]{Lisle1998}.
 The $\mathcal{I}_\nu$ are modified Bessel functions of the first kind and the $\mathcal{Q}_\mu$ are generalized Marcum Q-functions defined by $\mathcal{Q}_\mu(x,y) = \int_0^y e^{-z-x}(z/x)^{(\mu-1)/2}\mathcal{I}_{\mu-1}(2\sqrt{xz})dz $ and originally devised for radar detection theory \citep{Marcum1960,Temme1996}.
 The Marcum Q-functions derive from the burial process.
-Since we assumed resting grains could bury with an exponential probability while the resting probability follows a modified Bessel distribution \citep{Einstein1937,Lisle1998}, burial develops the Q-function convolution structure.
+Because resting grains can become buried with an exponential probability, while the probability that particles rest follows a modified Bessel distribution \citep{Einstein1937,Lisle1998}, evaluating the probability that particles rest and become buried generates the Q-function convolution structure.
 
@@
 Colors pertain to different times.
-Units are unspecified, since we aim to demonstrate the general characteristics of $p(x,t)$.
+Units are unspecified, since the aim is to demonstrate the general characteristics of $p(x,t)$.
 Panel (a) shows the case $\kappa=0$ -- no burial.
@@
 Because of burial, the joint distribution tends toward a more uniform distribution than Gaussian.
-		\label{fig:pdfs}}
+		\label{fig:diffpdfs}}
 \end{figure}
 
-Figure \ref{fig:pdfs} depicts the distribution (\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
-When grains do not become buried, as in panel (a) of figure \ref{fig:pdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion and satisfying the central limit theorem.
-When grains become buried, as in panel (b) of figure \ref{fig:pdfs},  the Q-function terms prevent the distribution from approaching a Gaussian at large timescales, exemplifying anomalous diffusion \citep{Weeks1998} and violating the central limit theorem \citep{Metzler2000,Schumer2009}.
+Figure \ref{fig:diffpdfs} depicts the distribution (\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
+When grains do not become buried, as in panel (a) of figure \ref{fig:diffpdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion and satisfying the central limit theorem.
+When grains become buried, as in panel (b) of figure \ref{fig:diffpdfs},  the Q-function terms prevent the distribution from approaching a Gaussian at large timescales, exemplifying anomalous diffusion \citep{Weeks1998} and violating the central limit theorem \citep{Metzler2000,Schumer2009}.
 
@@
 
-To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, we derive the first two moments of position by taking derivatives with respect to $\eta$ of the Laplace space distribution (\ref{eq:nicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}, and we use these to calculate the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$.
+To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, we derive the first two moments of position by taking derivatives with respect to $\eta$ of the Laplace space distribution (\ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}, and these moments produce the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$.
 The first two moments are
@@
 \end{table}
-The positional variance (\ref{eq:var}) is plotted in figure \ref{fig:var} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
-We interpret ``$\gg$'' to mean ``of at least an order of magnitude greater''.
+The positional variance (\ref{eq:var}) is plotted in figure \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
+The notation ``$\gg$'' is interpreted in this context to mean ``of at least an order of magnitude greater''.
 These conditions are most relevant to tracers in gravel-bed rivers, since they represent that grains are initially at rest \citep{Hassan1991,Wu2019}, motions are typically much shorter than rests \citep{Einstein1937,Hubbell1964}, and burial requires a much longer time than typical rests  \citep{Ferguson2002,Hassan1994,Haschenburger2013}.
-Figure \ref{fig:var} demonstrates that under these conditions the variance (\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that we identify as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
-We suggest to call the fourth range geomorphic, since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
+Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
+Following \citet{Hassan2017} I suggest to call the fourth range geomorphic, since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
 
@@
 At local timescales, grains usually either rest or move; at intermediate timescales, they transition between rest and motion; at global timescales, they transition between rest, motion, and burial; and at geomorphic timescales, all grains bury.
-Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as we discuss in section \ref{sec:discussion}.
+Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as discussed in section \ref{sec:discussion}.
 	}
-	\label{fig:var}
+	\label{fig:diffvar}
 \end{figure}
@@
 Limit (2) provides global exponent $1 \leq \gamma \leq 3$ depending on the relative importance of $\kappa$ and $k_1$.
-In the extreme $k_1/\kappa \ll 1$, we find $\gamma=1$ in the global range, while in the opposite extreme $k_1/\kappa \rightarrow \infty$ we find $\gamma=3$.
-We summarize when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, equation (\ref{eq:var}) implies:
+The extreme $k_1/\kappa \ll 1$ produces $\gamma=1$ in the global range, while the opposite extreme $k_1/\kappa \rightarrow \infty$ produces $\gamma=3$.
+To summarize, when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, equation (\ref{eq:var}) implies:
 \begin{enumerate}
@@
 
-We extended \citet{Einstein1937} by including motion and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010}, before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
+This chapter extended \citet{Einstein1937} by including motion, rest, and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010}, before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
 \citet{Nikora2002} highlighted the need for such a physical description, although they suggested to use a two-state random walk between motion and rest states with heavy-tailed resting times, and they did not discuss sediment burial.
 However, other works have demonstrated that a two-state walk with heavy-tailed rests provides two diffusion ranges -- not three \citep{Weeks1996,Fan2016}, and although heavy-tailed resting times have been documented for surface particles \citep{Liu2019,Fraccarollo2019}, they are more often associated with buried particles \citep{Martin2012,Martin2014,Voepel2013,Olinde2015,Pelosi2016, Pierce2020a}, while surface particles retain light-tailed resting times \citep{Einstein1937,Yano1969,Ancey2006,Nakagawa1976}.
-Accordingly, we developed a random walk model of bedload trajectories with light-tailed surface resting times that incorporates sediment burial.
+Accordingly, I developed a random walk model of bedload trajectories with light-tailed surface resting times that incorporates sediment burial.
 
-The local and intermediate range diffusion characteristics resulting from our model correspond closely to the original Nikora et al.
-concepts, while our global range has a different origin than Nikora et al.
+The local and intermediate range diffusion characteristics resulting from this model correspond closely to the original Nikora et al.
+concepts, while the global range has a different origin than Nikora et al.
 envisioned.
 \citet{Nikora2001a} explained that local diffusion results from the non-fractal (smooth) characteristics of bedload trajectories between subsequent interactions with the bed,  while intermediate diffusion results from the fractal (rough) characteristics of bedload trajectories after many interactions with the bed.
-Our model represents these conclusions: non-fractal (and super-diffusive) bedload trajectories exist on timescales short enough that each grain is either resting or moving, while fractal (and normal-diffusive) bedload trajectories exist on timescales when grains are actively switching between motion and rest states.
-We conclude that local and intermediate ranges stem from the interplay between motion and rest timescales, as demonstrated by earlier two-state random walk models \citep{Lisle1998,Lajeunesse2017} and by all Newtonian models that develop sequences of motions and rests \citep{Nikora2001a, Bialik2012}, even those including heavy-tailed rests \citep{Fan2016}.
+This chapter represents these conclusions: non-fractal (and super-diffusive) bedload trajectories exist on timescales short enough that each grain is either resting or moving, while fractal (and normal-diffusive) bedload trajectories exist on timescales when grains are actively switching between motion and rest states.
+I conclude that local and intermediate ranges stem from the interplay between motion and rest timescales, as demonstrated by earlier two-state random walk models \citep{Lisle1998,Lajeunesse2017} and by all Newtonian models that develop sequences of motions and rests \citep{Nikora2001a, Bialik2012}, even those including heavy-tailed rests \citep{Fan2016}.
 
@@
 However, studies have demonstrated that divergent resting times can generate super-diffusion in asymmetric random walks \citep{Weeks1996,Weeks1998}, and both experiments \citep{Bradley2017,Bradley2010} and models \citep{Pelosi2016,Wu2019,Wu2019a} of bedload tracers undergoing burial have demonstrated global range super-diffusion.
-While our results also show global range super-diffusion, they do not necessarily refute the Nikora et al. conclusion of sub-diffusion at long timescales.
-We assumed sediment burial was a permanent condition which developed a non-diffusive geomorphic range.
+While my own results also show global range super-diffusion, they do not necessarily refute the Nikora et al. conclusion of sub-diffusion at long timescales.
+I assumed sediment burial was a permanent condition which developed a non-diffusive geomorphic range.
 In actuality, burial is a temporary condition, because bed scour can exhume buried sediment back into transport \citep{Wu2019a}, probably after heavy-tailed intervals \citep{Voepel2013,Martin2014}.
-We anticipate that a generalization of our model to include heavy-tailed timescales between burial and exhumation would develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in figure \ref{fig:var}.
-If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested by equilibrium transport models \citep{Voepel2013, Martin2014, Pierce2020a} and laboratory experiments \citep{Martin2014,Martin2012}, we expect a super-diffusive geomorphic range \citep{Weeks1998}.
-However, if they decay slower than $T^{-1/2}$, as implicitly suggested by the data of \citet{Olinde2015}, we expect a genuinely sub-diffusive geomorphic range \citep{Weeks1998}, leaving Nikora et al.
+A generalization of the model in this chapter to include heavy-tailed timescales between burial and exhumation might develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in figure \ref{fig:diffvar}.
+If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested in chapter \ref{ch:ch3}, other equilibrium transport models \citep{Voepel2013, Martin2014}, and laboratory experiments \citep{Martin2014,Martin2012}, the geomorphic range is expected to be super-diffusive \citep{Weeks1998}.
+However, if they decay slower than $T^{-1/2}$, as implicitly suggested by the data of \citet{Olinde2015}, the geomorphic range is expected to be genuinely sub-diffusive \citep{Weeks1998}, leaving Nikora et al.
 with the final word on long-time sub-diffusion.
@@
 Meanwhile, in bedload transport research, variable velocities \citep{Lajeunesse2010,Furbish2012,Heyman2016}, correlated motions \citep{Heyman2014,Lee2018,Saletti2020}, and anomalous diffusion \citep{Fathel2016,Bradley2017,Schumer2009} constitute open research issues.
-We believe further developing the linkage between existing bedload models and random walk concepts could rapidly progress our understanding.
+I believe further developing the linkage between existing bedload models and random walk concepts could rapidly progress our understanding.
 
@@
 \label{sec:conclusion}
-We developed a random walk model to describe sediment tracers transporting through a river channel as they gradually become buried, providing a physical description of the local, intermediate, and global diffusion ranges identified by \citet{Nikora2002}.
-Pushing their ideas somewhat further, we proposed a geomorphic range to describe diffusion characteristics at timescales larger than the global range when burial and exhumation both moderate downstream transport.
-At base level, our model demonstrates that (1) durations of sediment motions, (2) durations of sediment rest, and (3) the sediment burial process are sufficient to develop three diffusion ranges that terminate when all tracers become buried.
+In this chapter, I developed a random walk model to describe sediment tracers transporting through a river channel as they gradually become buried, providing a physical description of the local, intermediate, and global diffusion ranges identified by \citet{Nikora2002}.
+Pushing their ideas somewhat further, I followed \citet{Hassan2017} to propose a geomorphic range to describe diffusion characteristics at timescales larger than the global range when burial and exhumation both moderate downstream transport.
+At base level, this work demonstrates that (1) durations of sediment motions, (2) durations of sediment rest, and (3) the sediment burial process are sufficient to develop three diffusion ranges that terminate when all tracers become buried.
+This work provides a complementary perspective on the three range diffusion exhibited in chapter \ref{ch:flux}.
 A next step is to incorporate exhumation to better understand the geomorphic range.
-Ultimately, we emphasize that the multi-state random walk formalism used in this paper implicitly underlies most existing bedload diffusion models and provides a powerful tool for researchers targeting landscape-scale understanding from statistical concepts of the underlying grain-scale dynamics.
+Ultimately, the multi-state random walk formalism used in this chapter has been demonstrated to implicitly underlies most existing bedload diffusion models.
+The random walk formalism is an alternative perspective on the Langevin approaches used elsewhere in this thesis, and it provides a powerful tool for researchers targeting landscape-scale understanding from statistical concepts of the underlying grain-scale dynamics.

### 2021-07-07 ch5langevin.tex
very close

@@
 
-Bulk bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
-Due to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
+Bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
+Owing to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
 These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
-In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations \citep{Ancey2008} and quantifying the dependence of measurements on the observation scale \citep{Turowski2010}.
-Recent indications that sediment transport fluctuations might explain longstanding and unsolved problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
-One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume \citep{Ancey2008,Furbish2012}.
-This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although our understanding of these distributions unfortunately remains limited.
-We have as of yet no consensus on the shape of the bedload velocity distribution, and although models have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, we have as of yet no mechanistic models that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
+In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (secs.
+\ref{sec:birthdeath}-\ref{sec:renewal}).
+Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
+One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (sec.
+\ref{sec:birthdeath}).
+This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although understanding of these distributions remains limited.
+There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in section \ref{sec:langevin} of the introduction have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
 In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.
 
-Particle trajectories are a compromise between turbulent drag and particle-bed collision \citep{Wiberg1985}.
-We therefore anticipate that the Stokes number, which is known to characterize colliding bodies within a viscous fluid \citep{}, is an important dimensionless parameter of the bedload velocity distribution.
-The Stokes number is defined as ... where ....
+High-speed video experiments have measured different streamwise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
+One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
+These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were viscous.
+A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
+In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and super-critical.
+Other experiments display velocity distributions that are intermediate between exponential and Gaussian extremes that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
+The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm.
+Flows were again turbulent and subcritical.
+From this experimental record, the velocity distribution shape apparently does not consistently relate to super or sub-critical flow conditions, whether sediment grains are natural (sand, gravel) or synthetic (glass beads), or whether the flow is laminar or turbulent.
+However there is a loose trend within the particle size.
+Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}.
+This summary suggests that the shape of streamwise bed load velocity distribution could be controlled by the particle size.
 
-High-speed video experiments have measured different streamwise particle velocity distributions without providing much understanding as to why one distribution or another appears in a given set of hydraulic and sedimentary conditions.One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
-These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) having typical Stokes numbers $St \sim 1-10$.
-Their flow conditions are generally sub-critical ($Fr<1$) and turbulent ($Re>5000$) but not always: flows in \citet{Lajeunesse2010} were super-critical ($Fr>1$), and flows in \citet{Charru2004} and \citet{Seizilles2014} were viscous.
-A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
-In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads having higher Stokes numbers ($St \sim 10-500$).
-In all Gaussian cases, flows are turbulent ($Re>5000$) and super-critical ($Fr>1$).
-Two other experiments display velocity distributions that are intermediate between exponential and Gaussian and appear more like a Gamma distribution \citep{Houssais2012, Liu2019}.
-The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions.
-They resolved the velocity distributions for larger grains only.
-The \citet{Liu2019} experiments used uniformly-graded sand having median diameter $1.1$mm.
-Flows were again turbulent and subcritical.
-From this experimental record, we can summarize that the shape of the velocity distribution does not consistentely relate to whether a flow is super or sub-critical ($Fr$), whether sediment grains are natural (sand, gravel) or synthetic (beads), or whether the flow is laminar or turbulent ($Re$).
-However, the typical Stokes numbers of particles do seem to increase monotonically from exponential velocity experiments (where $St \sim 1$), to intermediate (Gamma-like) experiments (where $St \sim 10$), and eventually to Gaussian experiments (where $St \sim 10^2$.) Apparently, the shape of streamwise bed load velocity distributions depends on the particle size.
+Existing models of streamwise bed load velocities can be divided into computational and stochastic physics categories.
+Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modelling particles as spheres interacting through repulsive forces, and the flow using direct simulation of the Navier-Stokes equations or some related approximation (such as large eddy simulation or the St-Venant equations).
+When streamwise particle velocities have been analyzed in such simulations, they show exponential tails \citep{Gonzalez2017,Furbish2013} that agree with one subset of the experimental data.
 
-Existing models of streamwise bed load velocities can be divided into computational and statistical physics categories.
-Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modelling particles as spheres interacting through repulsive forces, and the flow using direct simulation of the Navier-Stokes equations or some related approximation (such as large eddy simulation or the St-Venant equations).
-When streamwise particle velocities have been analyzed in such simulations, they show exponential tails \citep{Gonzalez2017,Furbish2013} that agree with only a subset of the experimental data.
-Statistical physics models have incorporated stochastic driving and resisting terms into the Newtonian dynamics of individual grains to develop a Langevin-like description of bed load particle motions.
-For the downstream velocity $u(t)$, \citet{Fan2014} wrote $\dot{u} = F -\gamma \text{sgn}{u} + \xi(t)$ where $F$ represents the steady component of the fluid forcing, the term involving $\gamma$ is a quasi-static (Coulomb-type) friction term representing momentum dissipation by particle-bed collisions, and $\xi(t)$ is a Gaussian white noise representing variability in these forces.
-This model provides exponential velocity distributions which agree with one subset of experiments.
-\citet{Ancey2014} took a similar approach that includes different forces, solving $\dot{u} = \gamma(\overline{u}-u) + \xi(t)$.
-Here the term involving $\gamma$ is similar to a Stokes drag, except it involves the mean sediment velocity $\overline{u}$, not the fluid velocity as for a ``real" Stokes drag.
-$\xi(t)$ is again a Gaussian white noise representing fluctuations, and the model provides Gaussian velocity distributions that agree with another subset of experiments.
-While these models build insight into grain-scale sediment transport mechanics and provide powerful techniques with which to approach the problem, they have not yet provided a comprehensive explanation for the range of streamwise velocity distributions resolved in experiments.
+Several stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec.
+\ref{sec:langevin}).
+\citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces were lumped into a velocity-dependent force, while Gaussian white noise represents fluctuations.
+Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
+A physical description for the full range of experimentally-observed bedload velocity distributions remains an elusive target.
 
-Here, motivated by the realization that experimental particle velocity distributions vary systematically with the grain size, as summarized above, we hypothesize that the shape of the streamwise velocity distribution is controlled by the momentum dissipation characteristics of particle-bed collisions.
-It is well-known that the elasticity of granular collisions within viscous fluids depends on the particle size.
-In addition, the velocity distributions of granular gases are known to develop increasingly heavier tails than the Gaussian (Boltzmann) form of elastic gases as the inelastiscity of particle-particle collisions is increased.
-Taking inspiration from this established knowledge, we develop below a model for sediment grains in transport as they undergo inelastic particle-bed collisions.
-Our intentions are to test the hypothesis that particle-bed collision characteristics explain the range of experimentally-observed streamwise bed load particle velocity distributions, and to introduce more realistic forces into earlier statistical physics descriptions of individual bed load particle dynamics.
-We develop the model and explain our assumptions in section \ref{sec:model}, then we present the analytical solution and major results in \ref{sec:results}.
-We finally discuss the implications of these results, summarize our findings, and suggest ideas for further research in sections \ref{sec:discussion} and \ref{sec:conclusion}.
+In this chapter I explore the possibility that the shape of the particle velocity distribution is controlled by momentum dissipation during particle-bed collisions.
+In the theory of granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
+Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Chapman1970,Landau1980}.
+Collisions in these models involve episodic kicks to the particle momenta at random times, parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
+In contrast, current stochastic models of bedload transport appear highly simplified, treating collisions instead with static Coulomb or velocity-dependent drag.
+
+Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport, including episodic particle-bed collisions.
+The driving motivation is to test the hypothesis that particle-bed collision characteristics can explain the range of experimentally-observed streamwise bed load particle velocity distributions.
+A secondary one is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
+I develop the model and explain its key assumptions in section \ref{sec:langlangmodel}.
+Then I present the analytical solution of the model and other major results in \ref{sec:langresults}.
+Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in sections \ref{sec:langdiscussion} and \ref{sec:langconclusion}.
 
 \section{Mechanistic description of particle velocities}
-\label{sec:model}
+\label{sec:langmodel}
 
-Figure \ref{fig:1} indicates the configuration we have in mind.
-Nearly spherical and cohesionless particles of diameter $d$ and mass $m$ moving as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow.
-The flow is just strong enough to drive grains into rarefied transport of a kind typical in gravel-bed rivers: particles saltate along the bed in sequences of collisions between events of erosion and deposition; moving particles collide often with stationary particles, but rarely with other moving particles.
+Figure \ref{fig:1} indicates the configuration of the model in this chapter.
+Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow.
+The flow is just strong enough to drive grains into the rarefied transport typical in gravel-bed rivers: particles saltate downstream through a sequence of particle-bed collisions. 
+Moving particles collide often with stationary particles, but rarely with other moving particles.
+
 Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces.
-In contrast to the computational physics approach, we do not aim to characterize the exact timeseries of the forces on an individual particle.
-Instead, we model the ensemble of possible force timeseries that particles could conceivably experience.
-Each possibility implies a different velocity timeseries $u(t)$ in the downstream direction.
-Our objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
-We include the most realistic article-bed collision and fluid forces we can while still allowing for analytical solutions.
+In contrast to the computational physics approach, I do not aim to characterize the exact timeseries of the forces on an individual particle.
+Instead, I model the ensemble of possible force timeseries that particles could conceivably experience.
+Each possibility implies a different possible velocity timeseries $u(t)$ in the downstream direction.
+The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
+I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.
 
-Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Williams2021}.
+\subsection{Episodic collisions}
+Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Schmeeckle2014,Williams2021}.
 The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}.
-Here, we introduce a restitution-like coefficient $\varepsilon$ as indicated in figure \ref{fig:1}.
-This ranges from $\varepsilon=0$ for completely inelastic collisions to $\varepsilon=1$ for completely elastic collisions.
+Here, we introduce a simple elasticity parameter $\varepsilon$ as a first stab at the problem, as indicated in figure \ref{fig:1}.
+This elasticity characterizes the fraction of downstream momentum lost in a collision.
 If the streamwise velocity just prior to a collision is $u$, just after the collision it becomes $\varepsilon u$.
+The elasticity ranges from $\varepsilon=0$, representing a completely inelastic collision, to $\varepsilon=1$, representing a completely elastic collision.
 \begin{figure}
@@
 \end{figure}
-Since this quantity combines effects of particle shape and collision geometry and should vary from one collision to the next, we consider that the fraction of streamwise momentum dissipated per collision $\varepsilon$ lies on a statistical distribution $\rho(\varepsilon)$.
-Similar ideas are available in the granular physics literature \citep{Serero2015}.
-Further assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, we write the collision force in the downstream direction as
+
+Since the elasticity combines effects of particle shape and collision geometry and should vary from one collision to the next, the elasticity $\varepsilon$ is interpreted as a random variable, characterized by a statistical distribution $\rho(\varepsilon)$.
+Some granular gas models have also included random elasticity \citep[e.g.][]{Serero2015}.
+
+Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (sec.
+\ref{sec:einflux}):
 \be F_C(u,t) = - m u \sum_{k=1}^{N_\nu(t)}(1-\varepsilon_k)\delta(t-\tau_k). \label{eq:col} \ee
-Here, $N_\nu(t)$ is the number of collisions in time $t$, the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are elasticity coefficients characterizing the amount by which each collision slows the particle down.
+Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
 This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momentum.
 This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions.
-These conditions are always satisfied for the idealized saltation-type motion depicted in figure \ref{eq:1}.
+These conditions are always satisfied for the idealized saltation trajectories depicted in figure \ref{fig:fig1}.
 
+\subsection{Turbulent fluid forces}
 Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Re_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
 These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
-At realistic $\Re_p$ analytical results are limited, so it is standard practice to turn instead to empirical corrections on the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
+At realistic $\Re_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
 A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
 \rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
-In the present model we set $C_D = \frac{24}{\Re_p}( 1 + 0.194 \Re_p^{0.631})$ \citep{Clift1978,Gonzalez2017} and we do not involve acceleration and history terms for simplicity, although we acknowledge their potential importance for coarse sediment transport \citep{Michaelides1997,Armenio2001}.
+In the present model I set $C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631})$ \citep{Clift1978,Gonzalez2017}, and I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.
 
-Drag forces have been argued to fluctuate rapidly compared to the inertial response times of coarse sediment grains \citep{Fan2014}.
-The magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
-Using these ideas, we make two key simplifications of the drag force above.
-First, we split the drag $F_D$ into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, we represent drag fluctuations as a Gaussian white noise characterized by a particle diffusivity $D$ \citep{Fan2014,Ancey2014}.
-Defining $\bar{V}$ as a representative slip velocity which we specify more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, we express the fluid forces as
-\be F_D(t) = \frac{\pi}{8}
-\rho_f d^2 \bar{C}_D \bar{V}^2 + \sqrt{2 D } \eta(t). \ee
-In this drag force \ref{eq:drag} and the collision force \ref{eq:col}, the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions.
-This set of possiblities defines a statistical ensemble.
+Drag forces have been argued to fluctuate rapidly compared to the inertial response times of submerged grains \citep{Fan2014}, and the magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
+These ideas allow two major simplifications of the above drag force.
+First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is described by a diffusivity $D$ \citep{Fan2014,Ancey2014}.
+Defining $\bar{V}$ as a representative slip velocity to be specified more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, the fluid drag can be written
+\be F_D(t) = \Gamma + \sqrt{2 D } \eta(t), \label{eq:drag}\ee
+where $\Gamma = \frac{\pi}{8}
+\rho_f d^2 \bar{C}_D \bar{V}^2$ is the steady component of the drag.
 
-\subsection{Langevin equation for collisional transport}
+\subsection{Langevin equation for collisional bedload transport}
 
@@
 \be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
-This equation replaces the steady friction terms of earlier models () with an episodic term that provides a more realistic representation of particle-bed collisions during saltation.
-It represents a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
-Collisions introduce ``jumps" in velocity while turbulent generates "diffusion", and the collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
+This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
+Mathematically, equation \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
+Collisions introduce ``jumps" in velocity while turbulent generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
+
 Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
 One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
-Here, the Ito interpretation (lower endpoint integration rule) is the physical choice since the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional.
-Given this integration rule, the remaining issues are to obtain the integro-differential equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
+Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983} since.
+Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
 
@@
 
-The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix \ref{sec:appendixA}, providing
+The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix section \ref{sec:langmasterderiv} with a limiting procedure, providing
 \be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
-In this equation, we introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$. The term
+In this equation, I introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$.
+The term
 \be \mathcal{I}_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
 is a ``collision integral" term representing particle-bed collisions.
+
 Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
-Such equations combining are known as Chapman-Komogorov equations \citep{Gardiner1983}. 
 Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
 This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}.
-Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) goes to infinity.
-Mathematically, the collision integral term represents the probability distribution of the product between $\varepsilon$ and $u$ \citep[c.f.][]{Feller1968}.
+Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
+Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $m u$ \citep[c.f.][]{Feller1968}.
 
-Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so I introduce one further approximation.
-Assuming the distribution of dissipation coefficients $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$ allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
+Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
+Assuming the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
 Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
@@
 where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
+
 In what follows, I drop all but the first two terms in this expansion to obtain the leading order contribution of particle-bed collisions to the velocity distribution.
@@
 I solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$.
-Equation \ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg1/\nu$).
+Equation \ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg \nu^{-1}$).
 
 \section{Results}
-\label{sec:results}
+\label{sec:langresults}
 
 \subsection{Derivation of the bedload velocity distribution}
-\label{sec:solution}
+\label{sec:langsolution}
 Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
@@
 \be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
-which is now a functional differential equation that is nonlocal in velocity due to the last term on the right hand side.
-Such equations seen some attention in the mathematics literature, where they are called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
-In the appendix I solve equation \ref{eq:governer} using Laplace transforms, providing
+which is an advanced functional differential equation.
+This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
+Such equations seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
+
+In the appendix section \ref{sec:langsteadyderiv} I solve equation \ref{eq:governer} using Laplace transforms, providing
 \begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
@@
 \be K_\pm = d(\lambda_+-\lambda_-)\prod_{l=1}^\infty (-d\lambda_\pm^2 \ve^{2l} + \gamma \lambda_\pm \ve^{l} + 1). \ee
-Although this velocity distribution appears quite complicated, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode dissipation constant $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
+Although this velocity distribution appears complicated, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
 
 It is possible to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
-From this method the mean velocity is
-\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve},\ee
-which is scales weakly with the mean fluid drag and sharply with the rate and typical elasticity of collisions.
+These calculations are provided in appendix section \ref{sec:langmoments}, where the mean bedload velocity is derived as
+\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}.\ee
+This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
 The second moment is
@@
 leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
-\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
-This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to fluid properties.
-In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are the leading control on particle velocity fluctuations.
+\be \sigma_u^2 = \frac{2 d + \gamma^2}{1-\ve^2}.\ee
+This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
+In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are a leading control on particle velocity fluctuations.
 
@@
 	\centerline{\includegraphics{./figures/ch5/Fig2pdfs.png}}
-	\caption{Left panels show velocity realizations as gray traces.
+	\caption{Left and right panels are paired.
+Left panels show velocity realizations as gray traces.
 Velocities are calculated from Monte Carlo simulations.
@@
 Flow forces generate fluctuating positive accelerations between collisions.
-Right panels show simulated histograms of particle velocities and exact solutions from equation \ref{eq:steadystate}.}
+Right panels show simulated histograms of particle velocities and exact solutions from equation \ref{eq:steadystate}.
+As elasticity $\ve$ varies, the particle velocity distributions interpolate between exponential (inelastic) and Gaussian (elastic) forms.}
 	\label{fig:fig2}
 \end{figure}
-We can see an apparent transition from exponential-like to Gaussian-like velocity distributions as typical collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
-In between, the full distribution \ref{eq:steadystate} resembles a Gamma distribution, although it is not exactly a Gamma distribution.
+This figure reveals an apparent transition from exponential-like to Gaussian-like velocity distributions as  collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
+In between, the full distribution \ref{eq:steadystate} resembles a Gamma distribution, although it is represented by equation \ref{eq:steadystate}, not a Gamma distribution.
 
 \subsection{Exponential and Gaussian regimes: limits to earlier work}
-\label{sec:modelcomparison}
+\label{sec:langmodelcomparison}
 
-In fact, the apparent transition from exponential to Gaussian in figure \ref{fig:fig2} can be formalized: despite its complex appearance, simple Gaussian and exponential distributions appear as rigorous mathematical limits of equation \ref{eq:steadystate}.
-When particle-bed collisions are completely inelastic, \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic, \ref{eq:steadystate} becomes Gaussian.
-Figure \ref{fig:fig3} demonstrates more closely the approach of the distribution toward these limits.
+In fact, the apparent transition from exponential to Gaussian in figure \ref{fig:fig2} can be demonstrated from equation \ref{eq:steadystate}.
+Despite its complex appearance, simple Gaussian and exponential forms derive from this equation as exact mathematical limits.
+When particle-bed collisions are completely inelastic ($\ne = 0$), \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\nu = 1$), \ref{eq:steadystate} becomes Gaussian.
+Figure \ref{fig:fig3} demonstrates in detail the approach of the distribution toward these limits.
 \begin{figure}
@@
 
-The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is rather easy to see.
+The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see.
 Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for that with $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
@@
 The challenge is that the statistical moments \ref{eq:mean} and \ref{eq:2ndmom} diverge at the same time as the denominator factors of the distribution \ref{eq:steadystate}.
-In appendix \ref{sec:appendixB} I return instead to the original equation \ref{eq:governer} to evaluate this elastic limit, obtaining
+In appendix \ref{sec:langappendixB} I return instead to the original equation \ref{eq:governer} to evaluate this elastic limit, obtaining
 \be P(u) = \frac{1}{\sqrt{2\pi\sigma_u^2}}e^{-\frac{(u-\bar{u})^2}{2\sigma_u^2}}. \label{eq:gaussian}\ee
@@
 \subsection{Comparison with experimental data}
-\label{sec:experimentcomparison}
+\label{sec:langexperimentcomparison}
 
@@
 \section{Discussion}
-\label{sec:discussion}
+\label{sec:langdiscussion}
 
@@
 \section{Conclusion}
+\label{sec:langconclusion}
 We have demonstrated that particle-bed collisions control the shape of the particle velocity distribution.
-\label{sec:conclusion}
+
+\endinput
+
+Particle trajectories are a compromise between turbulent drag and momentum dissipation by particle-bed collisions \citep{Schmeeckle2014}.
+For particles submerged in a viscous fluid, the amount of momentum dissipated by collisions is controlled by the particle Stokes number ($\St$), which characterizes the balance of pre-collisional momentum against the viscous pressure generated in the squeeze film between approaching particles \citep{}.
+The Stokes number is defined by $\St = \rho_s V d/(9\nu \rho)$, where $\rho_s$ and $\rho$ are the solid and fluid densities, $V$ is the approach velocity, $d$ is the particle size, and $\nu$ is the fluid viscosity.
+Momentum dissipation in submerged collisions shows a sharp dependence on $\St$ \citep{Joseph2001, Schmeeckle2001}, and hence also the particle size.
+This suggests particle size may have a nontrivial control over bedload velocity distributions.
+
+Here, motivated by the recognition that experimental particle velocity distributions appear to vary with the grain size, I hypothesize that the shape of the streamwise velocity distribution is controlled by the momentum dissipation characteristics of particle-bed collisions.
+For particles submerged in a viscous fluid, the amount of momentum dissipated by collisions is controlled by the particle Stokes number ($\St$), which characterizes the balance of pre-collisional momentum against the viscous pressure generated in the squeeze film between approaching particles \citep{}.
+The Stokes number is defined by $\St = \rho_s V d/(9\nu \rho)$, where $\rho_s$ and $\rho$ are the solid and fluid densities, $V$ is the approach velocity, $d$ is the particle size, and $\nu$ is the fluid viscosity.
+Momentum dissipation in submerged collisions shows a sharp dependence on $\St$ \citep{Joseph2001, Schmeeckle2001}, and hence also the particle size.
+This suggests particle size may have control over bedload velocity distributions.
+
+The elasticity of granular collisions within a viscous fluid depends on the particle size.
+
+In this drag force, the collision force \ref{eq:col}, the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions.
+This set of possiblities defines a statistical ensemble.

### 2021-07-07 ch6conc.tex
very close

@@
 River science offers no guarantee that these idealized white noises which we happen to understand best are sufficient to describe its phenomena.
-White noises are an idealization which is probably never represented in nature \citep{Gardiner1983,Kubo1972}.
+White noises are an idealization which is probably never represented in nature \citep{Gardiner1983,Kubo1978}.
 In contrast, colored spectra in which fluctuations distribute power unevenly across frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
@@
 
-Fourth, chapter () presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzslav2021}.
+Fourth, chapter () presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
 

### 2021-07-07 app_ch3.tex
this is a nightmare
progress ?

@@
 \chapter{Calculations involved in the bed elevation model}
-
+\label{ch:gill}
 \section{Numerical simulation algorithm}
@@
 The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
-In the following sections, we begin by demonstrating the time intervals $\tau$ between subsequent transitions (i.e., the intervals between transition times) are exponentially distributed within the model we develop in the main text.
-Then we describe the SSA as a consequence of this property.
+The following sections demonstrate that the time intervals $\tau$ between subsequent transitions are exponentially distributed within the model of chapter \ref{ch:ch3}.
+The SSA follows as a consequence of this property.
 
@@
 
-To initialize a simulation, we specify the initial conditions $n_0$ and $m_0$, the model parameters for use in equations (2-7) in the main paper, and the desired simulation duration $t_\text{max}$.
+Simulations are initialized by the initial conditions $n_0$ and $m_0$, the model parameters used in equations \ref{eq:rate1}-\ref{eq:rate4}, and the desired simulation duration $t_\text{max}$.
 The SSA uses these inputs to generate time series of $n$ and $m$ as follows:
@@
 
-By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation (8) of the manuscript over all values of $m$, we obtain the mean field equation for the particle activity:
+By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation (8) of the manuscript over all values of $m$, one can obtain the mean field equation for the particle activity:
 \be 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1)+\gamma(n+1)A(n+1)-(\nu  + \lambda + \mu n + \sigma n + \gamma n)A(n).\ee
@@
 after applying the normalization condition $G(1)=1$.
-From the definition of $G$ we can determine $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in equation (10) of the manuscript.
+From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in equation \ref{eq:ancey} and introduced originally in section \ref{sec:birthdeath}.
 
@@
 
-From the negative binomial distribution we have $\langle n |m \rangle = \langle n \rangle$, so equation (9) provides 
+The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so equation (9) produces 
 \begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m).
@@
 
-We solve this using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$ This works since $z_1$ is small.
+This can be solved using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$ This works since $z_1$ is small.
 Introducing $z$ and writing $\overline{\kappa}=\kappa/z_1$ gives
@@
 \subsection{Closure equation approach for bed elevations}
-We set out to determine an approximate relationship to close $\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$.
+This section produces an approximate relationship to close $\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$.
 \be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
 for the mean particle activity conditional to the elevation $m$ into into manuscript equation (9), noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
-\be 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
-\ee
-At this point, we consider $c$ an undetermined positive constant that may depend on the entrainment, migration, and deposition rates.
+\begin{multline} 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]\\-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
+\end{multline}
+At this point, $c$ is considered an undetermined positive constant that may depend on the entrainment, migration, and deposition rates.
 Taking the Fokker-Planck expansion and requiring the distribution $M(m)$ to vanish at infinity for normalization yields

### 2021-07-07 app_ch5.tex
progress ?

@@
 where
-\be q(z) = d z^2 - i \gamma z + 1 . \label{eq:q} \ee
+\be q(z) = \tilde{D} z^2 - i \tilde{\Gamma} z + 1 . \label{eq:q} \ee
 Recursing $N+1$ times provides
@@
 The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
-\be \lambda_\pm = \frac{\gamma}{2d}\Big[1 \pm \sqrt{ 1 + 4d/\gamma^2} \Big].\ee
+\be \lambda_\pm = \frac{\tilde{\Gamma}}{2\tilde{D}}\Big[1 \pm \sqrt{ 1 + 4\tilde{D}/\tilde{\Gamma}^2} \Big]. \label{eq:lambdas}\ee
 Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
 \be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
-where the coefficients $R_l^\pm$ are the residues of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
+where the coefficients $R_l^\pm$ are the residues (at least up to a constant factor) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
 \be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
-The Fourier transform (\ref{eq:recursion}) has a beautiful feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
+The Fourier transform (\ref{eq:recursion}) has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
 Taking this limit and evaluating the residues provides 
-\begin{multline} \tilde{P}(s) = \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
-	\\+ \frac{1}{d(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
-Finally, inverting the Fourier transforms term by term with contour integration and incorporating (\ref{eq:q}) provides the steady-state solution (\ref{eq:steadystate}).
+\begin{multline} \tilde{P}(s) = \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
+	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
+Finally, inverting the Fourier transforms term by term with contour integration (another use of the integral given in eq.
+\ref{eq:contour}), and incorporating eq.
+(\ref{eq:q}), provides the steady-state solution (\ref{eq:steadystate}).
 
@@
 Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
-\be 0 = Dk (k-1)\langle u^{k-2} \rangle + \Gamma k \langle u^{k-1} \rangle + \nu (\ve^k-1)\langle u^k \rangle. \ee
-$k=1$ provides the mean
-\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}\ee
+\be 0 = \tilde{D} k (k-1)\langle u^{k-2} \rangle + \tilde{\Gamma} k \langle u^{k-1} \rangle + (\ve^k-1)\langle u^k \rangle. \ee
+$k=1$ provides the mean (given in eq. \ref{eq:meanu})
+\be \langle u \rangle = \frac{\tilde{\Gamma}}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}\ee
 while $k=2$ provides the second moment
-\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
+\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
 leading to the velocity variance
-\be \sigma_u = \sqrt{\frac{2 d + \gamma^2}{1-\ve^2}}.\ee
+\be \sigma_u^2 = \frac{2\tilde{D} + \tilde{\Gamma}^2}{1-\ve^2}\ee
+given in equation \ref{eq:varu}.
 \section{Weak and strong collision limits}
 \label{sec:langextremes}
-Now we demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
+Now I demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
 The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
-Following \cite{Hall1989}, this suggests normalizing the distribution $P(u)$ using 
-\be z = \frac{u - \bar{u}}{\sigma_u} \ee
+Following \cite{Hall1989}, this suggests normalizing the distribution $P(u)$ using the scaled variables
+\be z = \frac{u - \bra u \ket}{\sigma_u}, \ee
 and
-\be Q(z) = \sigma_u P(u)\ee
-to seek a differential equation for $Q(z)$ with manageable behavior as $\ve \rightarrow 1$.
-Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" Master equation
-\be (1-\ve^2) \frac{d}{2d + \gamma} Q''(z) - \frac{\gamma\sqrt{1-\ve^2}}{\sqrt{2d+\gamma^2}}Q'(z) - Q(z) + \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\gamma\sqrt{1-\ve^2}}{\ve\sqrt{2d+\gamma^2}} \Big]\Big) = 0 \ee
+\be Q(z) = \sigma_u P(u).\ee
+This produces a differential equation for $Q(z)$ which is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
+Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" master equation
+\begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 \end{multline}
 This equation remains exact and is only a change of variables from (\ref{eq:master}).
-Now we approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
+Now I approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
 \be Q''(z) + z Q'(z) + Q(z) = 0, \ee
-which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$.
+which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$ \citep[e.g.][]{Gardiner1983}.
 This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-07-07 ch5langevin.tex
progress ?

@@
 \begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
-	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(-d \lambda_-^2 \ve^{-2m} + \gamma \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
-The factors $\lambda_\pm$ are defined in the appendix; they are proportional to $\tilde{\Gamma}/\tilde{D}$. 
+	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(- \tilde{D} \lambda_-^2 \ve^{-2m} + \tilde{\Gamma} \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}.
+\end{multline}
+The factors $\lambda_\pm$ are defined in appendix equations \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
 The normalization factors are $K_\pm$ are 
-\be K_\pm = d(\lambda_+-\lambda_-)\prod_{l=1}^\infty (-d\lambda_\pm^2 \ve^{2l} + \gamma \lambda_\pm \ve^{l} + 1). \ee
-Although this velocity distribution appears complicated, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
+\be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
+Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
+The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.
 
@@
 These calculations are provided in appendix section \ref{sec:langmoments}, where the mean bedload velocity is derived as
-\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}.\ee
+\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}. \label{eq:meanu}\ee
 This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
@@
 leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
-\be \sigma_u^2 = \frac{2 d + \gamma^2}{1-\ve^2}.\ee
-This equation demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
+\be \sigma_u^2 = \frac{2 d + \gamma^2}{1-\ve^2}. \label{eq:varu}\ee
+This result demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
 In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are a leading control on particle velocity fluctuations.

### 2021-07-08 app_ch5.tex
got the writing basically drafted. baout to change notation in the rwdifu chapter.

@@
 \label{sec:langsteadyderiv}
-Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:detlangevin} develops the recursion relation
+Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:master} develops the recursion relation
 \be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee

### 2021-07-08 ch1intro.tex
got the writing basically drafted. baout to change notation in the rwdifu chapter.

@@
 
-Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}, 
+Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}. 
 These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Hassan1994,Ferguson2002}.
 
-Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales, from seconds to years \citep{Pretzlav2021}, and each of these processes defies exact description.
+Individual trajectories of particles therefore involve a number of contributing processes which occur over different characteristic timescales, from seconds to years \citep{Pretzlav2021}, and each of these processes has so far evaded any exact description.
 
@@
 The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
-The intermediate range reflects particle motions through sequences of bed encounters, when particles alternately accelerate and decellerate. 
+The intermediate range reflects particle motions through sequences of bed encounters, when particles alternately accelerate and decelerate. 
 The global range refers to particle trajectories through sequential entrainment and deposition events as they cycle between motion and rest.
@@
 Watching painted tracers move through a flume, Einstein came to the conclusion that the movement characteristics of any one particle could not be predicted, so he turned to probabilistic methods to characterize their transport.
+
 Einstein's key insight was to represent particle motions as an alternating sequence of movements and rests having random characteristics.
-As his interest was on the global range of particle motion when many motion-rest alternations have occurred, and the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
+As his interest was on the global range of particle motion when many motion-rest alternations have occurred, and because the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
 With this approximation, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
 Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
+
 Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
-In the process, Einstein formulated an example of what is now called the continuous time random walk, but this was not was not formalized until much later \citep{Montroll1965}.
+In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this was not formalized until much later \citep{Montroll1965}.
 
@@
 A particular realization of this pulsed noise can be written
-\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i), \label{eq:einrando} \ee
-where $N(t)$ is the number of particle entrainments in time $t$, a Poisson random variable distributed as $P(N) = e^{-k_E t} (k_E t)^N/N!$, the $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
-Figure \ref{fig:einsteinfig} panel (a) sketches this noise, and panel (b) sketches the resulting global range trajectory as a sequence of steps and rests.
-Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar in spirit to Newtonian mechanics \citep{Goldstein1997}, but with random driving term (equation \ref{eq:einrando}).
+\be \mu(t) = \sum_{i=1}^{N(t)}s_i \delta(t-t_i). \label{eq:einrando} \ee
+Here $N(t)$ is the number of particle entrainments in time $t$ which is  a Poisson random variable, distributed as $P(N) = e^{-k_E t} (k_E t)^N/N!$.
+The $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
+Figure \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as a sequence of steps and rests.
+
+Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (equation \ref{eq:einrando}).
 \begin{figure}[!htbp]
@@
 Diffusion occurs because step lengths and instants when entrainment events occur vary from one particle to the next.
+
 One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
-\be \langle x \rangle (t) = k_E \ell t, \ee
-so in Einstein's view, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$ given by the product of the entrainment rate and mean step length.
+\be \langle x (t)  \rangle = k_E \ell t. \ee
+This equation indicates that in Einstein's model, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$ given by the product of the entrainment rate and mean step length.
+
 The rate at which particles spread apart due to differences in their trajectories can be represented by the variance of position, $\sigma_x^2  = \langle x^2 \rangle - \langle x \rangle^2$. This gives
@@
 
-The advent of high speed camera experiments of bed load transport produced new insight into the local and intermediate ranges of particle motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
+The advent of high speed camera experiments produced new insight into the local and intermediate ranges of bedload motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
 In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
@@
 \be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
-where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
-This noise is displayed in figure \ref{fig:lislefig} panel (a).
-Some particle trajectories solving equation \ref{eq:lislelangevin} are displayed in figure \ref{fig:lislefig} panel (b).
+%where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
+This noise is displayed in figure \ref{fig:lislefig} panel (a). Some particle trajectories produced by equation \ref{eq:lislelangevin} are displayed in figure \ref{fig:lislefig} panel (b).
 
@@
 The rapid spreading of particles at short timescales occurs because some are completely stationary while others move with velocity $V$.
-The eventual transition to normal diffusion occurs because particles eventually become well-mixed among motion and rest states.
-An analogous diffusion phenomenon is described in \citet{Taylor1920}.
+
+The eventual transition to normal diffusion occurs because particles eventually become well-mixed among motion and rest states. An analogous diffusion phenomenon is described in \citet{Taylor1920}.
 
@@
 
-Some authors have attempted modelled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
-Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included heuristic collision forces to modify particle velocities upon bed contact according to a set of simplified rules \citep{Wiberg1985,Sekine1992,Nino1998}.
+Some authors have modelled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
+Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.
 
@@
 
-Some approaches have described particle velocities in the local and intermediate ranges when velocities fluctuate due to turbulence and particle-bed collisions using Langevin equations, whereby particles are driven by Gaussian white noise \citep{Kubo1978}.
+Another category of models have described bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.
+
 Experimental studies on bedload velocities have provided two major conclusions for the shape of the particle velocity distribution.
-One subset of observations indicates that bedload velocities follow exponential distributions \citep{Lajeunesse2010,Furbish2012,Fathel2015}, and another subset indicates Gaussian distributions \citep{Martin2012,Ancey2014,Heyman2016}.
+One subset of observations shows that bedload velocities follow exponential distributions \citep{Lajeunesse2010,Furbish2012,Fathel2015}, and another subset shows Gaussian distributions \citep{Martin2012,Ancey2014,Heyman2016}.
 
-\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities with a mechanistic model including a noisy driving term to represent fluid turbulence.
-They wrote, for the streamwise particle velocity $u$, a Langevin equation
+\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities ($u$) with the Langevin equation
 \be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
-This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of particle velocity fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
-The fluid drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
+This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
+This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
 The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
@@
 \end{figure}
-Although these descriptions successfully describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying reason why bedload transport has shown multiple velocity distributions remains an open question.
+Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and other distributions have been reported besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
 A comprehensive model of the full range of bedload transport observations is a desirable target.
@@
 Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
-Today, there are two main competing (or complementary) approaches to define the bedload flux. 
-The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $\mathcal{S}$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
-\be q = \int_\mathcal{S} c(\textbf{x},t)\textbf{u}\cdot d\mathcal{\textbf{S}}. \label{eq:surfflux} \ee
+Today, there are two main complementary definitions of the bedload flux.
+
+The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
+\be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
 This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.
+
 The second definition formulates the downstream flux in terms of the number of particles moving within a control volume $V$:
@@
 \subsection{The scaling arguments of Bagnold}
+\label{sec:baggo}
 
@@
 Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
-He assumed that the flow power per unit volume $P_f$ available to move sediment scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
-Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ per unit volume required to sustain particle motion scales as $P_g \propto \Phi/\tau^{1/2}.$
+He assumed that the flow power $P_f$ available to move sediment in a volume scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
+Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$required to sustain particle motion in a volume scales as $P_g \propto \Phi/\tau^{1/2}.$
 Balancing flow energy against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
@@
 
-He partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
+Einstein partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
 Each control volume has downstream length $\ell$ which is also the average particle step length.
@@
 Some authors have argued convincingly that instead, one should calculate $P_n$ as an exceedance probability.
-The central critique is that there is a distinct different between \textit{exactly} and \textit{at least} $n$ entrainment events in $T$.
+The central critique is that there is a distinct difference between \textit{exactly} and \textit{at least} $n$ entrainment events in $T$.
+
 If the entrainment rate of an individual grain is $k_E$ (probability per unit time), then the probability that it entrains \textit{exactly} $n$ times in time $T$, denoted by $p_n$ (distinct from $P_n$) is a Poisson distribution \citep{Cox1965}:
@@
 \be q = E \ell, \ee
-where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit streambed area \citep{Wilcock1997a,Furbish2012}.
+where the quantity $E = \rho k_E$ is differentthe ``areal entrainment rate", representing the number of grains entrained per unit streambed area \citep{Wilcock1997a,Furbish2012}.
 
 Einstein's formulation requires expressions of the single particle entrainment rate $k_E$ and step length $\ell$ in terms of the flow and sediment characteristics.
-Although Einstein and many followers attempted to formulate these connections \citep[e.g.][]{Einstein1950,Grass1970,Paintal1971,Tregnaghi2012,Dey2018}, there are still no comprehensive solutions.
+Although Einstein and many followers have attempted to formulate these connections \citep[e.g.][]{Einstein1950,Grass1970,Paintal1971}, there are still no comprehensive solutions.
+Linking Einstein's entrainment and deposition rates to the grain scale mechanics remains an active area of research \citep[e.g.][]{Tregnaghi2012,Dey2018}.
 
@@
 One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}.
-Their ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's erosion rate $E$ and the complementary deposition rate $D$.
-The model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2018}, in what can be characterized as a mixture of the Einstein and Bagnold strategies.
+This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$.
+These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}, in what can be characterized as a mixture of the Einstein and Bagnold strategies.
 
@@
 Scaling arguments provide relations for $E$, $D$, and $V$ in terms of the fluid shear stress $\tau$, particle size $d$, particle settling velocity $V_s$, and critical shear stress $\tau_c$:
-\be E = a \frac{\tau-\tau_c}{d^3 V_s}, \label{eq:charru1}\ee
-\be D  = b\frac{\gamma V_s}{d}, \ee
-\be V = c + d(\sqrt{\tau}-\sqrt{\tau_c}). \label{eq:charru2}\ee
-The constant coefficients $a$, $b$, $c$, and $d$ are calibrated in experiments.
+\be E = \alpha_1 \frac{\tau-\tau_c}{d^3 V_s}, \label{eq:charru1}\ee
+\be D  = \alpha_2 \frac{\gamma V_s}{d}, \ee
+\be V = \alpha_3 + \alpha_4(\sqrt{\tau}-\sqrt{\tau_c}). \label{eq:charru2}\ee
+The constant coefficients $\alpha_i$ are calibrated in experiments.
 
@@
 Using the scaling relations \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
-\be \gamma \propto \frac{\tau-\tau_c}{d^2 V_s^2}.\ee
+\be \gamma = \frac{\alpha_1}{\alpha_2}\frac{\tau-\tau_c}{d^2 V_s^2}.\ee
 Expressing the mean flux as $\Phi = \gamma V$, the relationship between flux and bed shear stress becomes
-\be \Phi = \frac{A}{d^2 V_s^2}\big(\tau-\tau_c\big)\big[c + d(\sqrt{\tau}-\sqrt{\tau_c})\big]. \ee
+\be \Phi = \frac{\alpha_1}{\alpha_2 d^2 V_s^2}\big(\tau-\tau_c\big)\big[\alpha_3 + \alpha_4(\sqrt{\tau}-\sqrt{\tau_c})\big]. \ee
 This recovers the Bagnold scaling $\Phi \propto \tau^{3/2}$ at large bed shear stresses.
@@
 Einstein's model of the bedload flux is inherently nonlocal in that it aggregates particle motions from all upstream locations \citep{Schumer2009,Tucker2010a,Martin2012}. 
-\citet{Nakagawa1976} and later \citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
-\be q(x,t) = \int_0^\infty dx' F(x-x',t)E(x',t). \ee
-In this equation, motions are instantaneous, and $F(x,t)$ is the probability that a particle entrained at $t=0$ steps a distance $x$ before deposition at time $t$.
-In general, this approach can handle non-uniform conditions by including space and time dependence in $E$ and $F$.
+Originally \citet{Nakagawa1976} and later \citet{Parker2000} formalized this by writing the sediment flux in an explicitly nonlocal form:
+\be q(x,t) = \int_0^\infty dx' F(x',t)E(x-x',t). \ee
+In this equation, motions are considered instantaneous, and $F(x,t)$ is the probability that a particle entrained at $t=0$ steps at least a distance $x$ before deposition at time $t$.
 
@@
 \be q(x,t) = \int_0^\infty dx' \int_0^\infty dt' F(x',t') E(x-x',t-t'), \label{eq:furbo}\ee
-where $F(x,t)$ represents the probability density that particles move \textit{at least} a distance x in time $t$ (right after entrainment).
+where $F(x,t)$ represents the probability density that just-entrained particles move at least a distance x in time $t$ before deposition.
+In general, this approach can handle non-uniform conditions by including space and time dependence in $E$ and $F$.
 
@@
 \be q = \rho_s k_E \int_0^\infty dx' \int_0^\infty dt \theta(Vt-x)k_D\exp(-k_D t) = \rho_s k_E V/k_D. \label{eq:furbflux}\ee
-Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$.
-Equation \ref{eq:furbflux} therefore provides another perspective on Einstein's central result $q= E\ell$, but with the added possibility of describing unsteady conditions \citep{Furbish2012}.
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and qquation \ref{eq:furbflux} provides another perspective on Einstein's central result $q= E\ell$, but with the added ability to describe unsteady conditions \citep{Furbish2012}.
 
@@
 
-\citet{Exner1925} was probably the first to write the mathematical formula linking sediment transport to topographic change.
+\citet{Exner1925} was probably the first to write a mathematical formula linking sediment transport to topographic change.
 He wrote
@@
 \be \partial_x q(x,t) = E - D. \ee 
-Owing to \ref{eq:exner}, this expression of the sediment flux immediately implies the ``entrainment form of the exner equation" \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described with Einstein's rates:
+In combination with equation \ref{eq:exner}, this expression of the sediment flux implies \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described with Einstein's rates:
 \be (1-\phi) \px z(x,t) = D - E.\ee
@@
 This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
-The major limitation of these approaches are that they assume the landscape is a continuous field.
+
+The major limitation of these approaches are that they assume landscapes are continuous.
 This can be interpreted as an average over the detailed configurations of individual grains \citep{Coleman2009}.
-This averaging process is expected to obscure the physics whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
+This averaging process is expected to obscure the relevant landscape evolution processes whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
 
@@
 At the longest timescales, they originate from migrating bedforms \citep{Guala2014} or cycles of aggradation and degradation in pools \citep{Dhont2018}.
-These mechanisms occur in conjunction with other sources of bedload transport fluctuations like grain size sorting \citep{Iseya1987,Cudden2003}, flow discharge variations \citep{Wong2006,Mao2012,Redolfi2018}, or sediment supply perturbations \citep{Lisle1993,Madej2009,Elgueta-Astaburuaga2019}.
+These mechanisms generally occur in conjunction with other sources of bedload transport fluctuations like grain size sorting \citep{Iseya1987,Cudden2003}, flow discharge variations \citep{Wong2006,Mao2012,Redolfi2018}, and sediment supply perturbations \citep{Lisle1993,Madej2009,Elgueta-Astaburuaga2019}.
 
 Despite widespread acceptance that sediment transport rates always fluctuate, we still have relatively little understanding of how to describe these fluctuations.
-\citet{Hamamori1962} was probably the first to calculate a probability distribution for the sediment flux, but this subject fell into relative obscurity for a long period, until it was eventually revisited \citep{Nikora1997,Ancey2006}.
-Now there are several approaches to calculate sediment flux distributions which are described below.
+\citet{Hamamori1962} was probably the first to calculate a probability distribution for the sediment flux, but this subject fell into relative obscurity for a long period, until it was revisited by \citet{Nikora1997}, \citet{Ancey2006}, and others.
 
 Recognition that sediment transport rates fluctuate requires careful interpretation of sediment transport measurements. 
-In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Chartrand2018}, impact sensors \citep{Rickenmann2007}, computer vision \citep{Roseberry2012}, or sediment traps \citep{Papangelakis2016}.
+In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Chartrand2018}, impact sensors \citep{Rickenmann2007}, computer vision techniques \citep{Roseberry2012}, or sediment traps \citep{Papangelakis2016}.
 
-This averaging process in context of fluctuations introduces scale-dependence to sediment transport measurements, whereby averaged transport characteristics depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
+This averaging process in conjunction with transport fluctuations introduces scale-dependence to sediment transport measurements, whereby averaged characteristics depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
 A fundamental question raised by scale-dependence is how large averaging windows must be for sediment flux measurements to converge.
-To date, very few mathematical models have addressed this issue. Several that have are reviewed below.
+To date, very few mathematical models have addressed this issue.
 
@@
 They derived the probability distribution of the flux by counting the number of moving particles in a control volume, considering that this number changes through time as a result of particles migrating into the volume from upstream, entraining, depositing, and migrating out of the volume to downstream.
+
 To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
 
-Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where entrainment is ``birth" and deposition is ``death".
+Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where arrivals of moving particles to the volume is ``birth" and departure is ``death".
 They formulated the probability distribution $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
-\be \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) - [n+1]\alpha P(n+1,t) - \big[\lambda + n (\alpha + \mu)\big],\ee
+\begin{multline} \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) \\ - [n+1]\alpha P(n+1,t) - \big[\lambda + n (\alpha + \mu)\big]P(n,t),\end{multline}
 whose terms describe particle birth (entrainment and migration in) at rate $\lambda$, collective entrainment at rate $\mu$, and particle death (deposition and migration out) at rate $\alpha$.
@@
 The probability distribution of the bedload flux can be computed by the control volume form eq.
-\ref{eq:controlflux} given the probability distribution of particle velocites $P(u)$, if one assumes velocity and activity are independent.
+\ref{eq:controlflux} given the probability distribution of particle velocites $P(u)$ under the assumption that velocity and activity are independent.
 Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep[e.g.][]{Ancey2014}
 \be P(\Phi) = L \sum_{k=0}^\infty P_k(L\Phi) P(n=k). \ee
-One limitation of this approach is that in reality, entrainment and deposition modify the bed elevation, which in turn modifies the rates of entrainment and deposition \citep{Sawai1987,Wong2007}. 
-This highlights a need to evaluate bed elevation change in a birth-death sediment transport model.
+A limitation of this approach is that in reality, entrainment and deposition modify the bed elevation, which in turn modifies the rates of entrainment and deposition \citep{Sawai1987,Wong2007}. 
+This highlights a need to evaluate bed elevation change in birth-death models.
 
@@
 \label{sec:renewal}
-The final bedload transport model to summarize calculates the sediment flux probability distribution from an ``arrival time distribution" representing the statistics of times between subsequent arrivals of particles to a control surface \citep{Turowski2010, Heyman2013,Ancey2020}.
-This distribution changes depending on the observation time $T$ used to calculate the averaged flux, characteristic of scale dependence.
-In statistics, the problem of counting arrival events is called renewal theory \citep{Cox1962}.
+The final bedload transport model to summarize calculates the sediment flux probability distribution from an ``inter-arrival time distribution" representing the statistics of times between subsequent arrivals of particles to a control surface \citep{Turowski2010, Heyman2013,Ancey2020}.
+
+The sediment flux distribution obtained from these approaches shifts depending on the observation time $T$ over which crossing events are observed, so approaches based on inter-arrival times explicitly model scale dependence.
+In statistics, the problem of counting probabilistic arrival events is called renewal theory \citep{Cox1962}.
 
@@
 \be \sigma_q(T)^2 = \frac{\Lambda^2}{T}.\ee
-Even this simple model gives a non-trivial conclusion that the relative uncertainty in a measurement of bedload transport depends on the observation time: $\sigma_q(T)/\langle q(T) \rangle \propto T^{-1/2}.$ The uncertainty diverges as the sampling time becomes short.
+Even this simple model gives a non-trivial conclusion that the relative uncertainty in a measurement of bedload transport depends on the observation time: $\sigma_q(T)/\langle q(T) \rangle \propto T^{-1/2}.$ Our uncertainty in a sediment transport measurement diverges as the observation time becomes short.
 
-The limitation of renewal approaches to calculate the flux is that they do not obviously relate to the dynamics of individual particles.
-In renewal models as they are phrased now, the inter-arrival time distribution is not mechanistic.
+A limitation of renewal approaches to calculate the flux is that they do not obviously relate to the dynamics of individual particles.
+In renewal models as they are phrased now, the inter-arrival time distribution is not based on the underlying mechanics of particle transport.
 
@@
 In the case of mean field models, the devices are heuristic scaling arguments and semi-empirical formulas, while in the case of stochastic models, they are simplified dynamical equations driven by idealized noises, like sequences of pulses, alternating switches, and erratic fluctuations.
-
-This latter methodology is applied throughout the thesis to many of the limitations highlighted during the above review. A detailed outline of the problems to be addressed follows below.
+This thesis applies the latter methodology to many of the limitations highlighted in the review. An outline of the problems to be addressed follows below.
 
@@
 
-The original model by \citet{Einstein1937} describing particle trajectories with instantaneous movements was subsequently improved to include a finite motion time with a constant velocity \citep{Lisle1998,Lajeunesse2017}.
-In reality, particle velocities fluctuate during movements between entrainment and deposition, so there is a need to go one step further and account also for velocity fluctuations of moving particles in Einstein-type models.
+The original model by \citet{Einstein1937} described particle trajectories as a sequence of instantaneous steps.
+This model was subsequently improved include a finite duration of motion at constant velocity \citep{Lisle1998,Lajeunesse2017}.
+
+In reality, particle velocities fluctuate during movements due to turbulence and particle-bed collisions, so there is a need to go one step further and account also for velocity fluctuations of moving particles in Einstein-type models.
 
@@
 
-Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories which includes velocity fluctuations in the motion state.
-The chapter then demonstrates how these trajectories can be used to calculate the scale-dependent probability distribution of the bedload flux, linking together studies from \citet{Einstein1937} to \citet{Ancey2020b} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
+Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories that includes velocity fluctuations in the motion state.
+The chapter also demonstrates how these trajectories can be used to calculate the scale-dependent probability distribution of the bedload flux.
+This joins together and extends studies from \citet{Einstein1937} to \citet{Ancey2020b} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
 
@@
 
-Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Heyman2013,Ancey2014}, which violates conservation of mass.
+Birth-death models calculate the flux probability distribution assuming that entrainment and deposition do not change the local bed elevation \citep{Heyman2013,Ancey2014}, but this assumption violates conservation of mass.
 
@@
 
-Chapter \ref{ch:ch3} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation variability on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movement characteristics of individual grains.
+Chapter \ref{ch:ch3} generalizes the model of \citet{Ancey2008} to include bed elevation changes in order to evaluate the effect of bed elevation change on bedload transport fluctuations and to study how the timescales of sediment burial relate to the movement characteristics of individual grains.
 
@@
 
-Modelling sediment transport as an alterating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface and provides a useful description of bedload transport across intermediate and global timescales, but it does not resolve local or geomorphic timescales.
-
+Modelling sediment transport as an alterating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface.
+Although this provides a useful description of bedload transport across local and intermediate timescales, it requires modification to describe for global and geomorphic timescales.
 At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017}, but burial has not yet been incorporated into Einstein-type models of particle trajectories.
@@
 Finally, the Langevin models of \citet{Fan2014} and \citet{Ancey2014} describe the velocity distributions of sediment moving downstream in the local and intermediate ranges between entrainment and deposition.
-
-These models describe two different end-member sediment velocity distributions, but our understanding of why these different distributions arise is incomplete, and no mathematical models have been developed that describe the full range of distributions observed in experiments.
+These models produce two different end-member sediment velocity distributions, and no mathematical models have been developed that describe the full range of distributions observed in experiments.
 

### 2021-07-08 ch2flipFlop.tex
got the writing basically drafted. baout to change notation in the rwdifu chapter.

@@
 Bed load particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered pockets on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
-Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, particles are once again entrained, and the cycle repeats.
+Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, and particles are once again entrained.
 Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and periods of rest.
-This process is difficult to describe mathematically given the technical stochastic physics required \citep{Furbish2017,Ancey2020}.
+This process is difficult to describe mathematically given the technicality of the required physics.
 
@@
 
-Many researchers have since refined Einstein's approach to provide more realistic descriptions of individual particle motions than Einstein's instantaneous step model.
-One set of efforts has neglected the particle deposition process to calculate the downstream velocity distributions of moving particles after entrainment using mechanistic equations with terms representing turbulent fluid forces and particle-bed collisions \citep{Ancey2014,Fan2014}.
-Another set of works have described particles alternating between motion and rest, considering that motions occur with a constant velocity \citep{Lisle1998,Lajeunesse2017,Pierce2020b}. 
-No models have yet been formulated which describe particles alternating between motion and rest with motions having realistic fluctuating velocities.
+Many researchers have since refined Einstein's approach to provide more realistic descriptions of individual particle motions than his instantaneous step model.
+One set of efforts has neglected the particle deposition process to calculate the downstream velocity distributions of moving particles using mechanistic equations with noise terms representing turbulent fluid forces and particle-bed collisions \citep{Ancey2014,Fan2014}.
+Another set of works have described particles alternating between motion and rest considering that motions occur with a constant velocity \citep{Lisle1998,Lajeunesse2017}. 
+No models have yet been formulated which describes particles alternating between motion and rest with motions having realistic fluctuating velocities, even though this is how bedload particles actually move.
 
@@
 Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006,Furbish2012a}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
-Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale-dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
-To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020,Turowski2010}.
+Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
+To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
 Among those that do, none have yet formulated the sediment flux in terms of the downstream trajectories of individual grains.
@@
 This survey provides context for the two problems addressed in this chapter.
-First, we do not yet have the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution and the dependence of its statistical moments on the observation time.
+First, we lack the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution and the dependence of its statistical moments on the observation time.
 
@@
 \ref{sec:mod}, solve it in sec.
-\ref{sec:res}, and I discuss the implications of my results and future research ideas in secs.
+\ref{sec:res}, and discuss the implications of my results and future research ideas in secs.
 \ref{sec:disc} and \ref{sec:conc}.
@@
 The downstream coordinate is $x$, so that $\dot{x}$ describes a velocity in the downstream direction.
-
 The flow is considered weak enough that interactions among moving grains are very rare, although interactions between moving particles and the bed may be common, characteristic of rarefied transport conditions \citep[e.g.][]{Kumaran2006,Furbish2017}.
-Any variations like particle size, shape, or spatial variations in the overlying fluid flow are neglected for mathematical simplicity.
+Particles are considered to have similar enough shapes and sizes so as to have nearly identical mobility characteristics.
 These conditions allow for all particles to be described as independent from one another but governed by the same underlying dynamical equations.
@@
 Keys in panel (a) demonstrate the average movement time $1/k_D$ and rest time $1/k_E$, while the key in panel (b) shows the average movement velocity $V$.
-Velocity fluctuations lead to tilted stair-step trajectories with unsteady slopes in the $x-t$ plane.
+Velocity fluctuations produce tilted stair-step trajectories with unsteady slopes in the $x-t$ plane.
 This can be compared with figure \ref{fig:lislefig} which does not include fluctuations.}
@@
 First, particles should alternate between motion and rest.
-The transition rate from rest to motion is called entrainment and occurs with probability per unit time (or rate) $k_E$, while the transition from motion to rest is called deposition and occurs with rate $k_D$.
-Second, particles in motion should move with mean velocity $V$ and their instantaneous velocities should jitter around this mean through time due to turbulent drag and particle-bed collisions.
+The entrainment transition rate from rest to motion occurs with probability per unit time (or rate) $k_E$, while the deposition transition from motion to rest occurs with rate $k_D$.
+Second, particles in motion should move with mean velocity $V$ as their instantaneous velocities jitter around this mean value due to turbulent drag and particle-bed collisions.
 
@@
 Some trajectories produced by this equation are sketched in figure \ref{fig:fluxxy0}.
+In this equation, $V$ is the mean particle velocity describing the overall downstream drift of moving particles, while $D$ is a diffusivity [units $L^2/T^3$] describing velocity fluctuations among moving particles.
+The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$. The notation $k=k_E+k_D$ is a shorthand used throughout the thesis.
 
-The mean particle velocity is $V$, and $D$ is a diffusivity [units $L^2/T$] describing velocity fluctuations among moving particles.
-The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$.
-The notation $k=k_E+k_D$ is a shorthand used throughout the thesis.
-
-The Langevin equation \ref{eq:flippylangevin} is a generalization of the model by \citet{Lisle1998} and \citet{Lajeunesse2017}, presented in section \ref{sec:lisle}, to include velocity fluctuations in the motion state.
-It is similar to the ``randomly flashing diffusion" developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}, although it includes a more complex asymmetric noise ($k_E\neq k_D$).
-This equation describes the downstream movement of particles alternating through motion and rest with velocity fluctuations in the motion state.
-This is a more realistic description of bed load trajectories than earlier works in the Einstein paradigm reviewed in section \ref{sec:trajmodels}.
+The Langevin equation \ref{eq:flippylangevin} is a generalization of the model by \citet{Lisle1998} and \citet{Lajeunesse2017}, presented in section \ref{sec:lisle}, to include velocity fluctuations in the motion stateprotocol.
+It is similar to the ``randomly flashing diffusion" developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}, although it is not clear if these models relate mathematically.
+Equation \ref{eq:flippylangevin} describes the downstream movement of particles alternating through motion and rest with velocity fluctuations in the motion state.
+This dynamical equation provides a more realistic description of bed load trajectories than earlier works in the Einstein paradigm (sec.
+\ref{sec:trajmodels}).
 
@@
 The solution of equation \ref{eq:flippylangevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives one possible trajectory of a particle.
-Averaging over the ensemble of all possible trajectories from different realizations of the noises will obtain the probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$.
-This distribution, by construction, will generalize earlier models of bedload alternation between motion and rest which did not include velocity fluctuations among moving particles \citep{Lisle1998,Lajeunesse2017}.
+The probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$ can be formulated as an average over the ensemble of all possible trajectories.
 
-The desired probability distribution of position is formed as $ P(x',t) = \big\bra \delta(x'-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
+This probability distribution of position is formed as $ P(x',t) = \big\bra \delta(x'-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
 \ref{eq:flippylangevin} and the average is over both noises.
 This symbolic equation is not directly useful as taking averages over both noises is a challenging mathematical problem \citep{Hanggi1978}.
-This calculation is completed in the appendix section \ref{sec:appAmaster}, providing the master equation
+This calculation is completed in the appendix in section \ref{sec:appAmaster}, providing the master equation
 \be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0 \label{eq:flippymaster}\ee
 for the probability distribution of position.
+The master equation \ref{eq:flippymaster} is a diffusion-like equation governing the probability distribution of position for individual particles alternating between motion and rest, with the movement velocity being a fluctuating quantity.
 
-The master equation \ref{eq:flippymaster} is a diffusion-like equation governing the probability distribution of position for individual particles as they transport downstream through a sequence of motions and rests, with the movement velocity being a fluctuating quantity.
-
-One can see in particular that taking an the entrainment rate $k_E$ very large, meaning that all particles are generally moving, implies an advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
-Otherwise, with $k_E$ of similar order as $k_D$, there is a finite probability that the particle is at rest, and the advection-diffusion process is often interrupted by deposition, giving rise to the additional terms in eq.
-\ref{eq:flippymaster} reminiscent of equation \ref{eq:lislemaster} from earlier simpler motion-rest models.
-When diffusion is turned off ($D=0$), equation \ref{eq:flippymaster} becomes identical to equation \ref{eq:lislemaster}, as expected.
+One can see in particular that taking the entrainment rate $k_E$ very large, meaning that particles are very often moving, implies an advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
+Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process interrupted, giving rise to the additional terms in eq.
+\ref{eq:flippymaster}, a structure reminiscent of equation \ref{eq:lislemaster} from simpler motion-rest models.
 
@@
 This method is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
-It generalizes the nonlocal formulation (\ref{sec:nonlocal}) to provide the probability distribution of the flux, and it generalizes the renewal approach (sec.
-\ref{sec:renewal}) to include the mechanics of individual particles in downstream transport.
+This generalizes the nonlocal formulation (\ref{sec:nonlocal}) to provide the probability distribution of the flux, and it reframes the renewal approach (sec.
+\ref{sec:renewal}) to in terms of the mechanics of individual particles in transport.
 
@@
 From this initial configuration, particles move downstream through time according to equation \ref{eq:flippylangevin}.
-The flux is calculated as the average rate of particles crossing to the right of $x=0$ after a sampling time $T$.
+The flux is calculated as the average rate of particles crossing to the right of the control surface (at $x=0$) after the sampling time $T$.
 \begin{figure}
@@
 \be q(T) = \frac{1}{T}\sum_{i=1}^N I_i(T). \label{eq:flippyflux} \ee
-In this equation, $I_i(T)$ is an indicator function which is $1$ if the $i$th particle has passed the control surface ($x=0$) at the observation time $T$ and $0$ otherwise.
+In this equation, the $I_i(T)$ are indicator functions which equal $1$ if the $i$th particle has passed the control surface ($x=0$) at the observation time $T$ and $0$ otherwise.
 All particles which have not crossed the control surface (or which have crossed and then crossed back) contribute nothing to the flux.
@@
 \be P(q|T) = \Big \bra \delta\Big( q - \frac{1}{T}\sum_{i=1}^N I_i(T) \Big) \Big\ket. \ee
-To evaluate these averages, taking the Laplace transform over $q$ (i.e. forming the characteristic function of $P(q|T)$) obtains
+These averages are again most easily conducted in Laplace space. 
+Taking the Laplace transform over $q$ (i.e. forming the characteristic function of $P(q|T)$) obtains
 \begin{align} \tilde{P}(s|T) &=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N I_i(T)\Big)\Big\ket \\
@@
 	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra I_i(T) \big\ket \Big] \label{eq:flippycharacteristic}\end{align}
-This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the fact that  $e^{a I} = 1-(1-e^a)I$ because $I$ is either $0$ or $1$.
+This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the observation that  $e^{a I} = 1-(1-e^a)I$ because $I$ is either $0$ or $1$.
 
@@
 \be \bra I_i(t) \ket = \frac{1}{L}\int_L^0 dx' \int_0^\infty dx P(x-x',t) \label{eq:indicator} \ee
-where $P(x,t)$ solves equation \ref{eq:flippymaster}.
+where $P(x,t)$ is the position probability distribution which solves equation \ref{eq:flippymaster}.
 Equation \ref{eq:indicator} describes the probability that the $ith$ particle is found to the right of control surface by time $T$ provided it started at a random location somewhere to the left.
@@
 
-Inserting equation \ref{eq:indicator} into equation \ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, obtaining the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
+Inserting equation \ref{eq:indicator} into equation \ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, producing the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
 \begin{align} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T) \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T) \Big].
 \label{eq:flippychar} \end{align}
-where $\Lambda(T) = \rho \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate constant, similar to the rate constant in the renewal process models of section \ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
+where $\Lambda(T) = \rho \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate \textit{function}, similar to the rate constant in the renewal process models of section \ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
 
-The equation \ref{eq:flippychar} is the characteristic function of a Poisson distribution \citep{Cox1965}.
-Expanding in $e^{-s/T}$ and inverting the Laplace transform provides a key equation of this chapter, the probability distribution of the flux, held contingent on the sampling time $T$:
+Equation \ref{eq:flippychar} is the characteristic function of a Poisson distribution \citep{Cox1965}.
+Expanding in $e^{-s/T}$ and inverting the Laplace transform provides a key equation of this chapter, the probability distribution of the flux held contingent on the sampling time $T$:
 \be P(q|T) = \sum_{l=0}^\infty \frac{\Lambda(T)^l}{l!}e^{-\Lambda(T)}\delta(q-\frac{l}{T}). \label{eq:flippydist}\ee
-This equation implies that the mean flux is $\bra q(T) \ket = \int_0^\infty q P(q|T) dq = \Lambda(T)/T$, and similarly the variance is $\sigma_q^2(T) = \Lambda(T)/T^2$.
-For the case when $\Lambda(T)$ is proportional to the observation time ($\Lambda \propto T$), this equation is identical to the renewal theory approach reviewed in section \ref{sec:renewal}.
-This demonstrates that the renewal approach can be formulated with more generality by considering the dynamics of individual particles as a starting point.
+This equation implies that the mean flux is $\bra q(T) \ket = \int_0^\infty q P(q|T) dq = \Lambda(T)/T$.
+Similarly the variance is $\sigma_q^2(T) = \Lambda(T)/T^2$.
+For the case when $\Lambda(T)$ is proportional to the observation time ($\Lambda \propto T$), these formulas become identical to the renewal theory approach reviewed in section \ref{sec:renewal}.
+This correspondence demonstrates that the renewal approach can be formulated equivalently by considering the dynamics of individual particles as a starting point.
 
@@
 \ref{eq:monster}) as it evolves through time.
-From the initial mixture of motion and rest states, particles advect downstream as they diffuse apart from one another due to differences in their velocities and exchange times between motion and rest.
+From the initial mixture of motion and rest states, particles advect downstream as they diffuse apart from one another due to differences in their velocities and transition times between motion and rest.
 Panel (b) shows the resulting particle diffusion (eq.
 \ref{eq:diffo}).
-At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process (as advection is irrelevant on these timescales).
+At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process.
 For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
 Finally at times longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
-All results are scaled by the mean hop length $l=V/k_D$ and the autocorrelation time $\tau=1/k$ of the motion/rest alternation.
+All results are scaled by the mean hop length $\ell=V/k_D$ and the timescale $\tau=1/k$ of the motion/rest alternation.
 In both plots, the lines are analytical results while the points are the results of Monte Carlo simulations based on evaluating cumulative transition probabilities on a small timestep \citep[e.g.][]{Barik2006}.}
@@
 The master equation \ref{eq:flippymaster} describes the evolution of the probability distribution of position through time.
-Loosely speaking the solution of this equation is expected to be some combination of the \citet{Einstein1937} theory for particle transport (section \ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep{Morse1953a}.
+The solution of this equation should be some combination of the \citet{Einstein1937} theory for particle transport (section \ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep{Morse1953a}.
 
@@
 
-The master equation \ref{eq:flippymaster} is solved by transform calculus in section \ref{sec:fluccymastersol} of the appendices, providing
+The master equation \ref{eq:flippymaster} is solved by transform calculus in section \ref{sec:fluccymastersol} of the appendix, providing
 \begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t) +  \pt \big]\\
@@
 where $\varphi=k_D/k$ is the probability the particle starts at rest.
-This equation generalizes the earlier result \citep[e.g.][]{Lisle1998} for alternation between motions and rests, eq.
-\ref{eq:lisledist}, to include velocity fluctuations within the motion state.
-This is another key result of this chapter.
+This equation generalizes the earlier result  \ref{eq:lisledist} for alternation between motions and rests to include velocity fluctuations within the motion state.
 
-The distribution \ref{eq:monster} is shown evolving through time in figure \ref{fig:flippyfig1} panel (a), where the advection and diffusion characteristics of equation \ref{eq:flippymaster} are both clearly visible.
-The integral in equation \ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the modified Bessel probability density that the particle has been in motion for a period $u$ out of a time $t$ with the advection-diffusion propagator that a particle has travelled a distance $x$ in time $u$.
+The distribution \ref{eq:monster} is shown evolving through time in figure \ref{fig:flippyfig1} panel (a), where the advection and diffusion characteristics of equation \ref{eq:flippymaster} are both evident.
+The integral in equation \ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the Bessel probability that the particle has been in motion for a period $u$ out of a time $t$ with the Gaussian probability that a particle has travelled a distance $x$ in time $u$ within the motion state.
 The prefactors of this integral term can be understood as adapting this distribution to the initial conditions.
@@
 For $l = 1$, this equation generates the mean position $ \bra x \ket(t) = k_E V t/k$, which is unaffected by diffusion (since Gaussian velocity fluctuations are symmetric).
-The case $l=2$ provides the second moment, implying the variance of position
+The case $l=2$ provides the second moment, implying the variance of position is
 \be \sigma_x(t)^2 = \frac{2k_Ek_DV^2}{k^3}\Big( t + \frac{1}{k}e^{-k t} - \frac{1}{k}\Big) + 2\frac{k_E D}{k}t. \label{eq:minor} \ee
-This equation describes a non-trivial multi-scale diffusion phenomenon, whereby the rate at which particles spread apart depends on how long their dynamics have been ongoing.
+This equation describes a non-trivial multi-scale diffusion phenomenon, whereby the rate at which particles spread apart from one another depends on how long their dynamics have been ongoing.
 
-Provided that $2D/V^2 \ll k_D/k^2$, one can determine that equation \ref{eq:minor} produces three different scaling behaviors:
+Provided that $2D/V^2 \ll k_D/k^2$, series expansion of \ref{eq:minor} in $t$ and $1/t$ reveals three different scaling behaviors:
 \be \sigma_x^2 \sim 
@@
 \end{cases}\ee
-Note in the physical condition when $k\approx k_D$, which is satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the condition for existence of three ranges becomes $ \Pe \gg 1,$ where
-$\Pe =V^2/(2k_D D)$ is a Peclet number \citep[e.g.][]{Heyman2014} which measures the relative importance of advection and diffusion in the movements of particles.
+Note in the physical condition when $k\approx k_D$, which is generally satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the above condition for existence of three ranges becomes $ \Pe \gg 1,$ where
+$\Pe =V^2/(2k_D D)$ is a Peclet number \citep[e.g.][]{Heyman2014}. This Peclet number measures the relative importance of advection and diffusion to the downstream movements of particles.
 
 The large Peclet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles very rarely move upstream \citep[e.g.][]{Fathel2015}.
-Incidentally, this limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed.
+This limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed \citep[c.f.][]{Nikora2001,Nikora2002}.
 Figure \ref{fig:flippyfig1} panel (b) sketches this three-range diffusion.
@@
 
-Using the probability distribution of position \ref{eq:monster}, the integrals in equation \ref{eq:flatform} are performed in section \ref{sec:fluxconstant} of the appendix, providing the rate constant
+Using the probability distribution of position \ref{eq:monster}, the integrals in equation \ref{eq:flatform} are performed in section \ref{sec:fluxconstant} of the appendix, providing the rate function
 \begin{multline} 
@@
 
-This is an intricate result for the rate constant in the sediment flux \ref{eq:flippydist}.
+This is an intricate result for the rate function in the sediment flux distribution \ref{eq:flippydist}.
 This mathematical complexity may not be surprising given that equation \ref{eq:flippylangevin} involves two interacting diffusion processes.
@@
 \ref{sec:renewal}) more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
-
-This method to calculate the sediment flux distribution can also be interpreted as a generalization of the nonlocal formulation (sec.
-\ref{sec:nonlocal}) due to \citet{Parker2000,Furbish2012a} to describe sediment flux distributions, rather than just mean values.
+This method can also be viewed as a generalization of the nonlocal formulation (sec.
+\ref{sec:nonlocal}) developed by \citet{Parker2000,Furbish2012a} to describe sediment flux probability distributions, rather than just mean values.
 
@@
 The sediment flux probability distribution in equation \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
-Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
+Poisson distributions have relatively thin tails, meaning their fluctuations are typically small \citep{Ancey2006}.
 In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
-This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider fluctuations.
+This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider transport rate fluctuations.
 
@@
 
-Collective entrainment has been attributed to particle-particle interactions, such as small granular avalanches and collision-induced entrainment, and to fluid-particle interactions, such as large eddies \citep[e.g.][]{Cameron2020} entraining particles en masse as they sweep downstream \citep{Ancey2014,Lee2018}, leading to spatial correlations in sediment transport \citep{Heyman2014}.
+Collective entrainment has been attributed to particle-particle interactions such as small granular avalanches and collision-induced entrainment \citep[e.g.][]{Pahtz2020}, and to fluid-particle interactions, such as large eddies entraining particles en masse as they sweep downstream \citep{Ancey2014,Lee2018,Cameron2020}.
 
-Such interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here. 
+Particle-particle interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here. 
 The dynamical equation \ref{eq:flippylangevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
@@
 
-Models which generalize the approach established in this chapter to include collective motions should be capable of producing wide sediment transport fluctuations, but their development will be a challenging next step for research.
-The present chapter only lays the basic groundwork.
+A generalization of the model developed in this chapter to include collective motions should be capable of producing realistically-wide sediment transport fluctuations.
+Such a development will be a challenging next step for research.
+The present chapter only lays the groundwork.
 
-\subsection{Measurement protocol for the sediment flux}
+\subsection{Measuring the sediment flux}
 
@@
 For example, in the experiments of \citet{Martin2012}, this condition implies convergence times of approximately $T \gg 15 s$.
-With wide sediment velocity fluctuations and long particle movement periods, this convergence condition could become plausibly long.
-In this case, the velocity fluctuations of particles will obscure running mean transport rate data from instruments like light tables and acoustic sensors, which have often relied on short averaging intervals \citep{Zimmermann2008,Tsakiris2014,Mendes2016}.
+With wide sediment velocity fluctuations and long particle movement periods, the required observation timescale could become large.
+In this case, mean sediment flux timeseries obtained with light tables and acoustic sensors \citep{Zimmermann2008,Tsakiris2014,Mendes2016} would require careful interpretation.
 
@@
 
-It has long been recognized that many other processes, like bedform migration \citep{Hamamori1962,Guala2014}, particle sorting processes \citep{}, or entrainment of clusters \citep{Strom2004,Papanicolaou2018} also lend variability to the sediment flux.
-Understanding these processes has been a longstanding problem, in part because different sources of fluctuations mix within measured transport signals \citep[e.g.][]{Hoey1992,Singh2009,Saletti2015,Dhont2018}.
+It has long been recognized that many processes lend variability to the sediment flux, like bedform migration \citep{Hamamori1962,Guala2014}, particle sorting processes \citep{}, or entrainment of clusters \citep{Strom2004,Papanicolaou2018}.
+Understanding these processes has been a challenging problem, in part because different sources of fluctuations mix within measured transport signals \citep[e.g.][]{Hoey1992,Singh2009,Saletti2015,Dhont2018}.
 
-Although the present model has analyzed on individual particles and does not include these processes, we can wonder if better understanding fluctuations due to individual particles will aid understanding of more complex sources.
+Although the present model has analyzed independent particles, so it does not include these processes, we can wonder if better understanding fluctuations due to single-particle sources will aid understanding of more complex collective sources.
 An attempt to subtract the signature of individual particle dynamics from spectra of bedload transport may be a useful step to understand more complex morphological sources of sediment transport variability in channels.
-A comprehensive understanding of the linkage between individual particle dynamics and sediment fluxes, as I have worked toward here, would be the foundation of this approach to the problem.
+A comprehensive understanding of the linkage between individual particle dynamics and sediment fluxes, as I have worked toward here, would provide a foundation for such a subtraction process.
 
-\section{Summary and Conclusion \label{sec:conc}} 
-This chapter introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motions and rests.
-The motion phase included velocity fluctuations, providing the first analytical model of this type.
+\section{Summary \label{sec:conc}} 
+This chapter introduced a two-noise stochastic dynamical equation to describe individual bedload trajectories for particles alternating between motion and rest states.
+The motion state included velocity fluctuations, providing the first analytical model of this type.
 
 The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying dynamics of individual particles.
-The observation timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations in terms of a Peclet number.
+The observation timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations, characterized by a Peclet number.
 
-These results generalize the bedload trajectory models of \citet{Einstein1937}, \citet{Lisle1998}, and \citet{Lajeunesse2017} to include fluctuating movement velocities, and they reframe earlier renewal models of the sediment flux probability distribution \citep{Lajeunesse2010,Ancey2020} to explicitly include the transport dynamics of individual sediment grains.
+These results generalize the bedload trajectory models of \citet{Einstein1937}, \citet{Lisle1998}, and \citet{Lajeunesse2017} to include fluctuating movement velocities.
+This work also builds a particle dynamics framework underneath earlier renewal models of the sediment flux probability distribution \citep{Lajeunesse2010,Ancey2020}.
 
-The work further quantifies the dependence of the sediment flux on the observation scale, and provide guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in simplified conditions such as this model assumes.
+Finally, the work further quantifies the dependence of the sediment flux on the observation scale and provides guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in the idealized conditions in which the model was developed.
 The next step is to include collective motions to produce wider sediment transport fluctuations.

### 2021-07-08 ch4downDiff.tex
got the writing basically drafted. baout to change notation in the rwdifu chapter.

@@
 \label{sec:assumptions}
-Here I construct a three-state random walk where the states are motion, surface rest, and burial. These states are labelled as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
+Particle trajectories are formulated as a three-state random walk where the states are motion, surface rest, and burial. These states are labelled as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
 The target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if it is known to have started with the initial distribution $p(x,0)=\delta(x)$.
 Times spent moving or resting on the surface are characterized by exponential distributions $\psi_2(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
-We expect our conclusions will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
+The end results will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
 Grains in motion are considered to have a characteristic velocity $V$ \citep{Lisle1998,Lajeunesse2017}, and burial is modelled as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried.
@@
 
-To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, we derive the first two moments of position by taking derivatives with respect to $\eta$ of the Laplace space distribution (\ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}, and these moments produce the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$.
+To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution (\ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}.
+These moments produce the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$.
 The first two moments are
@@
 
-\section{Conclusion}
+\section{Summary}
 \label{sec:conclusion}

### 2021-07-08 ch6conc.tex
got the writing basically drafted. baout to change notation in the rwdifu chapter.

@@
 
-The research in this thesis generalizes and extends earlier work to describe sediment transport as a stochastic process.
-In the research, I have characterized the movements of grains along a wider range of timescales than before, framed existing research in terms of Langevin and Master equations, and produced more detailed descriptions of the bedload transport phenomenon, addressing several long-recognized shortcomings of earlier stochastic models.
-In particular I have filled in the connection between sediment fluxes and individual praticle motions, included particle velocity fluctuations in a descriptions of individual particle motions through rest and motion, evaluated the role of bed elevation changes on sediment transport rate fluctuations, and included the process of sediment burial in descriptions of downstream particle movement.
+This thesis has described bedload transport from the statistical physics of individual grains.
+The work has improved upon earlier descriptions.
+It describes the movements of grains along a wider range of timescales than before, highlights Langevin equations, master equations, and random walks as the common threads of stochastic sediment transport modelling, and provides more realistic descriptions of bedload transport.
+In particular, I have
+\begin{enumerate}
+	\item developed the linkage between sediment flux probability distributions and individual particle trajectories (ch. \ref{ch:flux}), 
+	\item described particle trajectories alternating through motion and rest with fluctuating velocities (ch. \ref{ch:flux}),
+	\item evaluated the control of bed elevation changes over sediment transport rates (ch. \ref{ch:ch3}),
+	\item characterized how long particles can remain buried within the sedimentary bed (ch. \ref{ch:ch3}),
+	\item incorporated the process of sediment burial in a description of downstream particle movement (ch. \ref{ch:downDiff}), and
+	\item formulated the velocity distributions of sediment particles including episodic particle-bed collisions (ch. \ref{ch:langevin}).
+\end{enumerate}
+These developments improve understanding of sediment transport in streams.
 
@@
 
-The overarching strategy in all of these developments has been to identify control and response variables, represent variable controls by idealized noises (entrainment and deposition events, particle burial events, turbulent forces, or particle-bed collisions), and then to write dynamical equations relating the respones variables to these stochastic control variables.
-This phrases sedimentary dynamics in terms of Langevin-like equations, stochastic analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for the response variable of interest in the problem and the force $F$ is a stochastic idealization of the control variables.
+The overarching strategy in all of these developments has been to identify control and response variables, represent control variables by idealized noises (entrainment and deposition events, particle burial events, turbulent forces, or particle-bed collisions), and then formulate dynamical equations relating the response variables to these stochastic control variables.
+This strategy phrases sedimentary dynamics in terms of Langevin-like equations, stochastic analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for the response variable of interest in the problem and the force $F$ is a stochastic combination of the control variables.
 
-Once the stochastic dynamical equation was written for a given problem, its solutions were counted () across realizations of the control variables to form a probability distribution of the response variable.
-These distributions were generally governed by integro-differential equations (), and the distributions solving these equations were obtained with transform calculus ().
+Once the stochastic dynamical equations were written for a given problem, its solutions were averaged over realizations of the control variables to produce a master equation, an integro-differential equation governing the probability distributions of the response variables.
 
-In every case, the result is a probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (), particle velocity (), bed elevation (), and sediment transport rate ().
-The same basic strategy should be applicable to a wide range of problems in geomorphology where variability matters \citep{Furbish2021a}.
+The solutions of the master equation in a given problem produce the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (chs.
+\ref{ch:flux} and \ref{ch:downDiff}), the particle velocity (ch.
+\ref{ch:langevin}), the bed elevation (ch.
+\ref{ch:ch3}), and the sediment transport rate (ch.
+\ref{ch:flux}).
+The same stochastic strategy should be applicable to a wide range of problems in geomorphology where phenomena can built up from component parts with apparently noisy characteristics.
 
@@
 The noises used in this thesis include white Gaussian, Poisson, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
-Turbulence was described using Gaussian noise (); while instantaneous steps (), particle arrivals to a control surface (), and particle-bed collisions () were described with Poisson noise.
-Alternation between motion and rest was represented with dichotomous noise ().
-Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise as required ().
+Turbulence was described using Gaussian noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise.
+Alternation between motion and rest was represented with dichotomous noise.
+Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise as required.
 This inclusion of multiple noise sources has not yet to my knowledge been pursued in any earlier stochastic models of sediment transport.
@@
 River science offers no guarantee that these idealized white noises which we happen to understand best are sufficient to describe its phenomena.
-White noises are an idealization which is probably never represented in nature \citep{Gardiner1983,Kubo1978}.
-In contrast, colored spectra in which fluctuations distribute power unevenly across frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
+White noises are an idealization which is probably never realized in nature \citep{Gardiner1983,Kubo1978}.
+In contrast, colored noises in which fluctuations have favoured frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
 In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
-It will be problematic if these phenomena require colored noise for their description, because dynamical equations driven by colored noise can be extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
+It will be problematic if these phenomena require colored noise for their description.
+Dynamical equations driven by colored noise can be extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
 Even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.
@@
 
-The first major contribution of this thesis is in chapter ().
+The first major contribution of this thesis is in chapter \ref{ch:flux}.
 Here, I formulated the probability distribution of the sediment flux from the trajectories of individual particles moving downstream.
-This work unifies the sediment trajectory models originating from Einstein () with the renewal theory approach to calculate the sediment flux ().
-The striking feature of this formulation is that, as a result of the particle dynamics, the mean bedload flux becomes scale-dependent, whereby the expected magnitude of the flux depends on the time-period over which it is observed ().
+This work unifies the sediment trajectory models originating from Einstein (secs.
+\ref{sec:einwalk}-\ref{sec:lisle}) with the renewal approach to calculate the sediment flux (sec.
+\ref{sec:renewal}).
+The striking feature of this formulation is that, as a result of the particle dynamics, the mean bedload flux becomes scale-dependent, whereby the expected magnitude of the flux depends on the time-period over which it is observed.
 
-An expectation that scale-dependence derives from individual particle trajectories was described theoretically by \citet{Ballio2018}, but this had not been described in a mathematical model until now.
-Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (), and an emerging body of research was producing the full probability distribution of the flux, but without referencing individual movement characteristic ().
-This work unifies these two research themes and indicates a path toward a statistical mechanics formulation of bed load sediment transport based on the grain scale physics.
+\citet{Ballio2018} explained that scale-dependence originates from individual particle trajectories, but this had not been described in a mathematical model until now.
+Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (sec.
+\ref{sec:einflux}), and an emerging body of research has producing the full probability distribution of the flux, but without referencing individual movement characteristics (secs.
+\ref{sec:birthdeath}-\ref{sec:renewal}).
+This work unifies these two research themes and presents a statistical mechanics formulation of bed load sediment transport based on individual particle trajectories.
 
@@
 
-Second, in chapter () I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
-Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (), and this was later improved to include the duration of motion ().
-Until this thesis, movement velocities in analytical models were considered constant () which contrasts with our experience (). 
+Second, in chapter \ref{ch:flux} I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
+Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (sec.
+\ref{sec:einwalk}), and this was later improved to include the duration of motion (sec.
+\ref{sec:lisle}).
+Until this thesis, movement velocities in analytical models were considered constant which contrasts with reality. 
 Although some numerical models have described motion/rest cycles with velocity fluctuations \citep{Fan2016,Bialik2012,Schmeeckle2014}, they had not resolved the novel three-range diffusion characteristics these dynamics imply.
 
-This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state (); it introduces a dimensionless Peclet number as an important characteristic of bedload sediment transport (); and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains ().
-These developments constitute a new understanding of bedload movement across its timescales ().
+This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state; it introduces a dimensionless Peclet number as an important characteristic of bedload sediment transport; and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains.
+These developments constitute a new understanding of bedload movement across its timescales.
 
@@
 
-Third, chapter () modified the birth-death description of bedload transport () to include feedbacks between the local bed elevation and the entrainment and deposition rates.
+Third, chapter \ref{ch:ch3} modified the birth-death description of bedload transport (sec.
+\ref{sec:birthdeath}) to include feedbacks between the local bed elevation and the entrainment and deposition rates.
 This allowed for a mathematical investigation of (1) how bed elevation changes affect sediment transport rates and (2) how bed elevation changes control the residence times of particles buried in the bed.
-Sediment transport fluctuations have come under increasing scrutiny with the resurgence of stochastic models in sediment transport (), while the burial times of particles are crucial for using sediment tracers to predict sediment transport ().
+Sediment transport fluctuations have come under increasing scrutiny with the resurgence of stochastic modelling in sediment transport, while the burial times of particles are crucial for using sediment tracers to predict sediment transport.
 Earlier birth-death models had generally considered that entrainment and deposition rates of particles remain constant even though these processes imply bed degradation and aggradation respectively, which are known to modify the entrainment and deposition rates in a negative feedback.
 
-This work demonstrates that this negative feedback buffers bedload transport fluctuations whenever collective entrainment occurs, meaning the magnitude of bedload transport fluctuations depends on the rate of bed elevation change ().
+This work demonstrates that this negative feedback buffers bedload transport fluctuations whenever collective entrainment occurs, meaning the magnitude of bedload transport fluctuations depends on the rate of bed elevation change.
 The residence times of buried particles are random variables that lie on heavy-tailed power-law distributions.
-These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to predict the downstream sediment flux in applications by tracking sediment tracer particles ().
+These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to predict the downstream sediment flux in applications by tracking sediment tracer particles.
 
@@
 
-Fourth, chapter () presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
+Fourth, chapter \ref{ch:downDiff} presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
 
-Until this work, the mechanisms which produce the different spreading rates of particles across the scaling ranges identified by Nikora and coworkers had been uncertain for several decades (), and earlier models works had included what were believed to be the required features without describing three or more scaling ranges ().
-The work further demonstrates that many approaches to describe individual particle motions are just clever reproductions of the continuous time random walk formalism from physics, indicating underlying unity within a diverse body of research and bringing powerful tools from physics to the sediment transport problem.
+Until this work, the mechanisms which produce the different spreading rates of particles across the scaling ranges identified by Nikora and coworkers had been uncertain for several decades, and earlier works had included what were believed to be the required features without describing three or more scaling ranges.
+The work ultimately demonstrates that many approaches to describe individual particle motions are reformulations of the continuous time random walk formalism from physics, indicating underlying unity within a diverse body of research and bringing powerful tools from statistical physics to the sediment transport problem.
 
@@
 
-Finally, chapter () displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
-This work provides the first comprehensive description of all bedload velocity distributions observed in experiments (), while earlier works had described only particular end-member cases ().
-
-This work includes ideas from granular gas theory into sediment transport modelling and provides a concrete example of the control grain-scale processes exert over large scale landscape evolution.
-In particular, since the work implies the shape of the bedload particle velocity distribution may depend on particle size, this has implications for mechanistic understanding of how abrasion by collisions of saltating bedload occurs (), which relates ultimately to the rates at which Earth's bedrock canyons incise.
+Finally, chapter \ref{ch:langevin} displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
+This work provides the first comprehensive description of all bedload velocity distributions observed in experiments (sec.
+\ref{sec:langexperimentcomparison}), while earlier works had described only particular end-member distributions (sec.
+\ref{sec:langevin}).
 
@@
 Simplified experimental geometries based on bimodal sediment beds \citep[e.g.][]{Houssais2012} would be a great context to revisit these issues.
-Extending the motion-rest model of chapter () to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
+Extending the motion-rest model of chapter \ref{ch:ch3} to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
 
@@
 
-The final limitation I will mention is that all of the efforts in this thesis were concentrated on weak bedload transport, where densities of moving particles are low enough that they may interact with the static bed () but never with each other.
-This concentration is justified because weak bedload transport conditions are typical of gravel bed rivers \citep{Ashworth1989,Warburton1992}, but it is nonetheless a limitation given the diversity of processes which move sediment over Earth's surface.
+The final limitation I will mention is that all of the efforts in this thesis were concentrated on weak bedload transport, where densities of moving particles are low enough that they may interact with the static bed but never with each other.
+This assumption is justified because weak bedload transport conditions are typical of gravel bed rivers \citep{Ashworth1989,Warburton1992}, but it is nonetheless a limitation given the diversity of processes which move sediment over Earth's surface.
 
@@
 Weak bedload transport is characterized by fluid forces small in comparison to gravity, and collision forces comparable to gravity.
-Viewed in this way, the work in this thesis represents a minute region of the vast parameter space spanned by Earth's geophysical flows, and geomorphology requires characterization of them all.
-Future studies should continue the effort \citep{Furbish2021a} to describe these processes as different expressions of the same basic building blocks. 
-Statistical mechanics has some versatile blocks.
+Viewed in this way, the work in this thesis targets a minute region of the vast parameter space spanned by Earth's geophysical flows, and geomorphology requires characterization of them all.
+Future studies should continue the effort \citep{Furbish2021a} to describe these processes as different expressions of the same basic statistical mechanics building blocks.
 

### 2021-07-08 clean.sh

@@ -27,5 +27,4 @@ rm -f diss.toc
 rm -f ch6conc.aux
 rm -f ch6conc.log
 rm -f ch5langevin.aux
-rm -r ch5langevin.log
 

### 2021-09-03 revision/formerPapers/jgr/point-by-point.tex
added phd thesis.

\documentclass[11pt]{article}
% General document formatting
\usepackage[margin=0.75in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{titling}
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{booktabs} % for wrapping tabulars in accord with
\bibliographystyle{agu}
\setlength{\droptitle}{-5em}   % This is your set screw

\usepackage{setspace}
\usepackage{etoolbox} % to make a quote command
\AtBeginEnvironment{quote}{\singlespace\vspace{-\topsep}\small}
\AtEndEnvironment{quote}{\vspace{-\topsep}\endsinglespace}

\usepackage{xcolor} % provide colored text blocks

%\usepackage[math]{kurier}
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\begin{document}

\title{Responses to reviewer comments: ``Joint stochastic model of bedload transport and bed elevations: implications for heavy-tailed resting times''}
\author{James K. Pierce \& Marwan A. Hassan}
\maketitle

\section*{Response to associate editor:}

I attach a tentative review paper, which will not be published in this form (I was invited by the JHR editor to write a review paper, but my draft paper is too long for their standards, I have to shorten it and split it into two parts). It could give some ideas about the literature review. Note also that Greg Wilson published two related papers that could interest the authors (Acoustic observations of near-bed sediment concentration and flux statistics above migrating sand dunes, Geophysical Research Letters, 43, 6304-6312, 2016. Anomalous Diffusion of Sand Tracer Particles Under Waves, Journal of Geophysical Research: Earth Surface, 123, 3055-3068, 2018.)

\textcolor{blue}{
Thank you sincerely for your comments and for sharing your work in progress. }

\begin{enumerate}
\item Naively I do not understand how the number of moving and resting particles can be correlated. The fact that there are m = 10 or 10,000 particles in a given control volume should not influence the individual transition probabilities on the number n of moving particles. The authors have made this possible only by using Eq. (1) relating elevation and m. I agree that n should depend on z (probably on its gradient, too, and perhaps on its neighborhood), but I do no see how it can depend on the bed volume or m.

\textcolor{blue}{Only the deviation of bed elevation from the mean value correlates to the number of moving particles in equation (1). This is related to a relative number or volume (area) of particles only, as in $m-m_0$ or $\phi a^2(m-m_0)$. With regard to practically defining these notions, $m_0$ will be contingent on the depth of the control volume measured relative to the mean bed elevation. We hope this is a sufficient clarification.}

\item How the governing equation is solved numerically requires more information, for instance in the form of an electronic supplement (Supporting Information) explaining how the algorithm works, comparing the numerical solutions with analytical ones (when m = 0), etc. Providing the python script and a jupyter notebook is great, but not sufficient. Note also that AGU prefers public data repositories associated with DOIs (see https://copdess.org/enabling-fair-data-project/enabling-fair-data-faqs/).

\textcolor{blue}{Thank you. We have written and included an electronic supplement to address this comment and comment (13) of reviewer \#1.}

\item The authors consider the simplified case with one control volume (or cell) as in my 2008 JFM paper, but the extension to a row of cells shows that there are additional spatial correlations that make the analysis more difficult (see my 2015 JGR paper). Have they tried to consider the extension to a row of cells? Or justify the use of a single cell?

\textcolor{blue}{JKP originally studied a multi-cell model before taking a step back to develop this work, as the interpretation of the multi-cell results proved difficult without a baseline for comparison. Therefore, we present this manuscript as a ``first stab'' at incorporating bed elevation changes in a stochastic model of bedload transport, if we may borrow the quote from your 2010 JGR:ES paper. In the spirit of the simple harmonic oscillator from physics, we present a baseline for comparison and further developments.}

\item The master equation does not admit analytical solutions except for m = 0, but it is possible to obtain approximate analytical solutions (my JGR 2010 and 2015 papers propose some techniques. See the SI of the 2015 paper). The advantage of a Fokker Planck approximation to their master equation is to provide further insights into the behavior of the physical system under consideration.

\textcolor{blue}{response}

\item It will better to split the Discussion into several subsections. This will facilitate the reading, help to structure the contents and remove tangential issues evoked by the reviewers, which distract the reader.

\textcolor{blue}{Thank you. We have split the discussion into sections and also we removed topics we identified as tangential, being motivated by this comment and comment (5) of reviewer \#3.}

\end{enumerate}

\section*{Response to reviewer \#1:}

\begin{enumerate}

\item Qualitatively, I agree with what has been presented in paragraph [5] of Voepel et al. (2013), which discusses some general properties of asymptotic bedload transport. An interesting example here is about the virtual velocity: “If residence times have an infinite-mean power law distribution... the virtual velocity will continue to decay as a power law... it implies that all particles will eventually be immobile – an unattractive idea for a non-aggrading bed.” It seems that an asymptotically heavy-tailed resting time distribution contradicts the steady-state (or equilibrium) transport conditions. Instead, I agree with Voepel et al. (2013) that at some timescale this heavy-tailed power-law decay for resting time distribution should transition into a faster thin-tailed decay, as supported by empirical evidences presented in that paper.

\textcolor{blue}{response}

\item I understand that similar with Martin et al. (2014), this work does not consider a hard reflecting boundary but instead an increasing restoring trend when the bed surface drifts away from the mean, as defined in equations (6) and (7). However, I suspect that this is still a case of vertically unbounded bed (contrast to the confined bed where bedrock exists and no deeper elevation fluctuations are possible, for example). The authors may check the vertical domain of the simulated bed (e.g. as depicted by the tails of the Gaussian distribution in figure 3b), to see if it increases as the simulation spans across different timescales. That is, as simulation duration increases, possibly more extreme elevation values will be sampled. It is not clear how this issue will affect the conclusion.

\textcolor{blue}{The vertical domain of the bed does increase with the simulation duration. However, in accord with the Gaussian distribution of bed elevation the probability of an excursion of distance $h$ from equilibrium is strongly suppressed by a factor proportional to $e^{-h^2}$. In practice, our $1500$hr simulations were sufficient to sample the majority of bed configurations anticipated within such an exponentially suppressed interval.}

\item In addition, I think as far as I am aware, Singh et al 2009 did not mention bed elevation distributions to be symmetric (as mentioned in line 137-138); in fact in Singh et al 2012, they showed bed elevation fluctuations to be asymmetric. Even Wong et al. 2007 argue adeviation from Gaussian behavior at the tails. Similarly, Aberle and Nikora 2006 argue for bed elevation fluctuations pdf to be asymmetric specially with increasing armoring discharge. Given, these recent experimental observations of asymmetric pdfs of bed elevation fluctuations, I wonder if the proposed model needs to be modified in order to obtain the realistic results (asymmetric bed elevation pdfs) and perhaps that would also provide better bedload activity predictions.

\textcolor{blue}{It is true that \citet{Singh2009} did not say this. Thank you reviewer \#1 for identifying this error. We have removed this mis-attribution and restructured the sentence by citing other data showing symmetrical bed elevation pdfs.
With regard to the Singh et al experiments, we performed image analysis on the bed elevation time series plots presented in \citet{Singh2009} and found only a statistically insignificant skew from this analysis, although certainly this methodology is limited. Personal communications with Dr. Singh confirmed the distributions were at only lightly skewed in the 2009 paper. \citet{Singh2012} meanwhile included 15\% sand to the sediment mixture that induced bedform formation and migration. The \citet{Singh2012} distributions are certainly skewed, presumably because of these migrating bedforms. The model we presented must be modified to account for bedforms. This may be a topic for future research. This modification might be possible within a multi-cell approach generalizing \citet{Ancey2015} and mentioned by Associate Editor Ancey in his comments. We mentioned this possibility in our revised discussion. }

\item I also noticed that the pdfs of bed elevations do not change shape (remain Gaussian; Figure 3), however recent work argue against it, i.e. the pdfs change shape with scale (e.g. Aberle and Nikora 2006; Singh et al 2009; 2012). I wonder what authors need to change in the model to account for this multiscale behavior.

\textcolor{blue}{
	Thank you especially reviewer \#1 for this excellent question. It motivated our careful study of the multifractal formalism developed in turbulence research \citep[e.g.,][]{Frisch1985} and subsequently applied to rainfall \citep[e.g.][]{Over1994,Marani2005} and sediment transport timeseries \citep[e.g.][]{Shang2005,Singh2009,Saletti2015}.
	Following these works we applied the moment scaling formalism to the bed elevation and bedload transport timeseries of the manuscript and found statistical monoscaling, meaning the shapes of the pdfs of these quantities do not change with observation scale, contrasting with observations by \citet{Aberle2006} and \citet{Singh2009}. Instead, the scaling of the pdf with observation scale is fully characterized by a single Hurst exponent $H$ as in equation (10) from \citet{Singh2009}. Because of this statistical monoscaling, the bed elevation pdf, for example, can remain Gaussian even though changing the observation scale will modify its height and width. We too are left wondering how a model of bedload transport and bed elevations can capture multiscaling. This is a topic that seems not explored yet in research. Both \citet{Aberle2006} and \citet{Singh2009} studied multiple grain sizes, so the source of multiscaling in bed elevation pdfs may lie in differential mobility. Statistical multiscaling is also associated with intermittency or burstiness of a signal. \citet{Singh2012} observed that migrating bedforms introduce intermittency in the bed elevation timeseries. Therefore, we wonder if bedform development and migration may be a source of statistical multiscaling and scale dependent pdfs. We have introduced this perspective to the discussion in the manuscript, and believe a complete statistical characterization of bedload transport and bed elevation signals to be a still emerging topic of research.}

\item Line 225: How is coupling defined? Is this based on linear correlation between bed elevation and bedload activity?

\textcolor{blue}{Coupling is defined by the parameter $l$ that links bed elevations to bedload mobility through equations (6) and (7). As $l\rightarrow \infty$ bed elevations and bedload transport become totally indepedent, since equations (2-7) of the manuscript become independent of the bed elevation. We have clarified this point near line 225.}

\item The current abstract is very brief and does not provide any information about what is expected in the paper. I think it can be improved by adding some more highlights of results as I believe JGR provides more space for details than currently used.

\textcolor{blue}{Thank you. We have carefully reworked the manuscript's abstract to include the key results and some description of how the model works.}

\item The title. The heavy-tailed resting times are not “derived” but results of numerical simulations. A different word like “implications” is suggested.

\textcolor{blue}{Thank you for this thoughtful suggestion. We directly replaced ``derivation of'' with ``implications for'' and agree this terminology is more representative of the work.}

\item Line 17. What is “bedload fluctuations”? transport rates?

\textcolor{blue}{Fluctuations in $q_s$ or equivalently $n$. We have clarified this point in the abstract.}

\item Line 68: it is?

\textcolor{blue}{Thanks -- Fixed.}

\item Line 70. What about Voepel et al. (2013)? This work also provides evidence of heavy-tailed resting time distributions at some timescales due to burial, as reflected by the empirical results.

\textcolor{blue}{By ``...\citet{Martin2014} have provided the only direct support..'' we mean only Martin et al have clearly and unambiguously resolved burial as the generating mechanism of heavy tailed resting times. Although \citet{Voepel2013} and others (\citep{Martin2014, Pretzlav2016, Olinde2015}) have observed heavy tailed resting times, these experiments did not unambiguously resolve the mechanism generating these heavy tails. In particular, these experiments can not rule out alternative hypotheses that heavy tailed distributions are generated by processes such as stranding on bars \citep[e.g.,][]{Bradley2017} or involvement in stabilizing bedforms or bed structures \citep[e.g.,][]{Church1998,Hassan2008}. We have clarified this point near line 70.}

\item Line 114. ``though'' $\rightarrow$ ``through''?

\textcolor{blue}{Thank you for the careful attention! We corrected this error.}

\item Line 140. Why “anti-symmetrical”?

\textcolor{blue}{Thanks for pointing out this terminology is somewhat confusing. We have rephrased this sentence and avoided ``anti-symmetrical'' altogether.}

\item Line 173. Does this random value mean an equal chance for different types of transitions
(2) – (7)?

\textcolor{blue}{It does not, but this comment with one of Dr. Ancey have helped us realize the details of the simulation algorithm are not fully clear in the paper. Therefore we have added an electronic supplement on the simulation algorithm and rephrased the sentence in question to make at least this issue more clear.}

\item Line 362. “$a<1$”?

\textcolor{blue}{Your suggestion $a<1$ is correct. Thank you and we have fixed this mistake in the revised manuscript.}

\item Line 369. Why “$t^{3.64+-0.45}$”?

\textcolor{blue}{The analysis of \citet{Weeks1996} and \citet{Weeks1998} provide two possibilities for the variance scaling associated with a heavy-tailed resting time and light-tailed step length distribution.}

\item I think it will be useful for the reader to provide more detailed figure captions (e.g. fig 3).

\textcolor{blue}{response}

\end{enumerate}

\section*{Response to reviewer \#2:}

\begin{quote}
The only hesitation that I may have is regarding the lack of adequate comparison of the results with other findings on the topic available in literature but overall I believe that the manuscript should be published in the present form.
\end{quote}

\textcolor{blue}{
Thank you reviewer \#2. We have tried to more carefully compare the results with existing literature in the revised manuscript. In particular, we have added discussion of the (potentially contrasting) the observations of relatively heavy-tailed bed elevation distributions by \citet{Aberle2006} and to the scale dependent and skewed bed elevation distributions identified by \citet{Singh2009} and \citet{Singh2012}. We have also drawn attention to the departures of the bed elevation distributions of \citet{Wong2007} from Gaussian at the tails and highlighted the analysis/discussion of this departure by \citet{Pelosi2014a} and \citet{Pelosi2016}. Finally, we have connected our predictions for the variance scaling of bedload at asymptotically large times to the recent model of \citet{Wu2019} on the dispersion of sediment undergoing burial that predicts $\sigma_x^2 \sim t^\gamma$ with $2\leq \gamma \leq 3$. }

\section*{Response to reviewer \#3:}

\begin{enumerate}
	\item My concern is that while the framework is new, it seems not yielding reasonably novel results. As mentioned in manuscript, the numerical outputs from the theory are mostly known facts (e.g. Martin et al., 2014; Wong et al., 2007) which on one hand confirm the validity of the theory but on the other hand do not generate compelling good stuffs to substantiate its publication at JGR-ES. After all, the evaluation of a new theory is based on how much new results can be brought by it.

	\textcolor{blue}{We believe the novel contribution of this manuscript is that it jointly describes bedload transport and bed elevations as statistical quantities, meaning with this model we can predict how the statistics of each quantity is contingent on the other. }

	\item 
	For example, two highlighted findings in the manuscript are (1) the collapse of T/T0 (1-CDF) at the large time scales and (2) Gaussian like distribution of channel bed elevation. Both are based on the numerical simulations, which is fine, but it is hard to convince the reader how close these results are from the fact. E.g. collapse of T/T0 (1-CDF) may only exist in the numerical outputs and the normalizing T with T0 seems not changing the thickness of tail and thus not changing the nature of anomalous diffusion.

	\textcolor{blue}{response}

	\item Moreover, it is still unclear to me how a heavy tailed PDF emerges from this framework, while the title says "Derivation of heavy-tailed resting times".

	\textcolor{blue}{Thank you for this constructive comment. We have changed the phrasing `` derivation of'' to ``implications for''.}

	\item As to the later one, experiments of Wong et al. (2007) and analyses of Pelosi et al. (2014) also suggested relatively heavy tailed distribution of channel bed elevation. The fact that there are inconsistencies between the new model and experiments without proper explanations weakens the Discussion and Conclusion significantly.

	\textcolor{blue}{Thank you for this comment. \citet{Wong2007} said "Visual inspection suggests a very good agreement [between data and the Gaussian fit] for almost the entire range of potential bed elevation deviation values, with the exception of the extreme tails of the distribution." Looking at their plots, we estimate the departure between theoretical Gaussian distributions and their experimental data at the tails does not exceed $5$\%. Wong et al. did not suggest heavy tailed distributions of bed elevation: rather they highlighted the essential agreement between their results and earlier (Gaussian) conclusions for the bed elevation pdf by \citet{Crickmore1962} and \citet{Pender2001} despite these small departures at the tails. Nor did \citet{Pelosi2014a} present any evidence for heavy tailed bed elevation pdfs. They simply assumed an $\alpha$-stable distribution for the sake of argument. \citet{Pelosi2016} later argued that $\alpha$-stable and Gaussian fits on the \citet{Wong2007} bed elevation data can be ``approximated as identical''. In conclusion, we do not agree that the \citet{Wong2007} or \citet{Pelosi2014} papers support the conclusion of heavy tailed bed elevation pdfs. However, \citet{Aberle2006} have shown heavy tailed bed elevation pdfs from severely degraded beds with multiple grain sizes, and \citet{Singh2012} has demonstrated non-Gaussian bed elevation statistics when bedforms are present. This comment has spurred us to more carefully account for these contrasting observations in the discussion section in the revised manuscript, as we believe these observations suggest important topics for subsequent research. }

	\item Particularly, many tangent materials that appears in the Discussion (although making sense) seems irrelevant to the essence of the work.

	\textcolor{blue}{We identified lines 330 to 358 as tangential and removed them entirely. Thank you for this.}

	\item I recommend a major revision with hope that the authors can dig more compelling results from the model before it is ready for publications. Some suggestions about potential paths forward include (1) showing how this model differs from Martin et al. (2014); (2) a potential explanation of moderately heavy tailed PDF of channel bed elevation from Wong et al (2007).

	\textcolor{blue}{We present analyses of the co-dependence of bedload fluctuations on bed elevation (and vice versa) as more compelling contributions. We have also included a more careful analysis of the difference between our model and \citet{Martin2014}}

\end{enumerate}

\bibliography{biblio}
\end{document}

### 2021-09-03 draft/abs.tex
added phd thesis.

--- 2021-07-04 abs.tex
@@
 
-Because coarse grains control river stability, a central problem in river science is to predict bedload transport, when coarse sediment moves downstream by bouncing, rolling, and sliding.
-This problem is usually approached with continuum mechanics, but this is questionable considering that coarse sediment grains rarely move in densities approximating a continuum.
-An alternative approach based on describing the trajectories of individual grains with statistical physics has become increasingly popular in recent decades, but many fundamental issues remain in this formulation.
-In particular, the connection between individual particle movements and transport rates remains unclear, descriptions of particle trajectories remain highly simplified, feedbacks between topography and sediment transport remain challenging to analyze, processes such as sediment burial known to control the long-term movements of individual grains have not been evaluated, and even fundamental characteristics such as downstream velocities of grains remain incompletely understood.
+Bedload transport is the movement of coarse grains through river channels by bouncing, rolling, and sliding.
+Because coarse grains control river stability, predicting the rate of bedload transport is a fundamental problem in river science.
+This problem is usually approached with continuum mechanics, but this approach is questionable considering that coarse sediment grains rarely move in densities approximating a continuum.
 
-This thesis presents four projects completed in the course of my PhD which address these problems.
-First, I demonstrate how to calculate the sediment flux from the dynamics of individual grains and describe the trajectories of grains alternating through rests and motions having fluctuating velocities, linking the sediment flux to the particle dynamics and describing these with more detail.
-Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting the residence times of particles buried within the bed.
-Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
-Finally I describe particles propelled downstream through episodic collisions with the bed surface with a granular gas-like approach, fitting all experimental data on downstream sediment velocities.
-I conclude by summarizing these developments, discussing their implications for the statistical mechanics sediment transport framework, and suggesting how we can use this progress to better understand landscapes.
+An alternative approach describes bedload transport from the trajectories of individual grains using statistical physics.
+This approach has become increasingly popular in recent decades, but many fundamental issues prevent this approach from being widely adopted.
+In particular, the connection between individual particle trajectories and transport rates remains unclear, and particle trajectory models remain highly simplified.
+Feedbacks between topography and sediment transport remain challenging to analyze, and basic properties of bedload motions like downstream travel velocities remain incompletely understood.
+Buried particles cannot move downstream, but even this simple observation has not been comprehensively described in the statistical physics approach.
+
+This thesis presents four projects completed in my PhD which overcome these issues to provide new understanding of bedload transport from a statistical point of view.
+First, I demonstrate how to calculate the sediment flux from the dynamics of individual grains, and I model the trajectories of grains alternating through motion and rest having fluctuating velocities in the motion state.
+This work links the sediment flux to the grain-scale dynamics and describes particle trajectories in unprecedented detail.
+Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates, and predicting how long particles can stay buried in the riverbed.
+Third, I incorporate this sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
+Finally, I concentrate on bedload dynamics at short timescales, predicting the movement velocities of bedload particles using methods adopted from granular physics.
+I conclude by summarizing these developments, discussing their implications for the statistical description of bedload transport, and suggesting how we can use this modelling progress to better understand landscapes.
 

### 2021-09-03 draft/app_ch3.tex
added phd thesis.

--- 2021-07-07 app_ch3.tex
@@
 \subsection{Times between transitions of any kind}
-Our joint stochastic description of bedload and bed elevation changes is characterized by a set of states $(n,m)$ where $n$ and $m$ are positive integers. 
-Our description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in equations (2-5) in the main text.
+The joint description of bedload transport and bed elevation changes is characterized by a set of states $(n,m)$ where $n$ and $m$ are integers.
+The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in equations (\ref{eq:rate1}-\ref{eq:rate4}).
+
 From the state $(n,m)$, the rate (probability per unit time) for any transition to occur is the sum over all possibilities:
@@
 Using this, the probability that no transition occurs from the state $(n,m)$ in a small time interval $\delta \tau$ is $1-A(n,m)\delta \tau$.
-If we denote by $Q(\tau|n,m)$ the probability density that a transition of any kind occurs from the state $(n,m)$ after a time $\tau$, we can express the probability density that a transition happens after a slightly larger time $\tau + \delta \tau$ as
+If $Q(\tau|n,m)$ denotes the probability density that a transition of any kind occurs from the state $(n,m)$ after a time $\tau$, the probability density that a transition happens after a slightly larger time $\tau + \delta \tau$ can be written
 \be Q(\tau+\delta \tau|n,m) = \big[1-A(n,m)\delta \tau\big]Q(\tau|n,m).\ee
-This equation invokes the Markov property, since it does not involve the past history of states $(n,m)$.
-Taking $\delta\tau \rightarrow 0 $ we find the master equation $\frac{d}{d\tau}Q(\tau|n,m) = -A(n,m)Q(\tau|n,m)$, from which we conclude the time $\tau$ between subsequent transitions is distributed as
+Taking $\delta\tau \rightarrow 0 $ produces a master equation $\frac{d}{d\tau}Q(\tau|n,m) = -A(n,m)Q(\tau|n,m)$, from which it follows that the time $\tau$ between subsequent transitions is distributed as
 \be Q(\tau|n,m) = A(n,m)e^{-A(n,m)\tau}. \label{eq:exp}\ee
-Therefore we have shown the time $\tau$ to the next transition from a state $(n,m)$ is exponentially distributed with mean value $\bar{\tau} = 1/A(n,m).$ In deriving this result, we used the normalization condition $1 = \int_0^\infty Q(\tau|n,m)d\tau.$
-
-This result implies if the stochastic process transitioned to the state $(n,m)$ at a time $t$, the next transition will occur at a time $t+\tau$ with $\tau$ a random variable drawn from the exponential distribution \ref{eq:exp}.
-Therefore, we can determine the times between subsequent transitions by drawing exponentially distributed random numbers.
+This equation demonstrates that the time $\tau$ to the next transition from a state $(n,m)$ is exponentially distributed with mean value $\bar{\tau} = 1/A(n,m).$ This means if the stochastic process transitioned to the state $(n,m)$ at a time $t$, the next transition will occur at a time $t+\tau$ with $\tau$ a random variable drawn from the exponential distribution \ref{eq:exp}.
 
@@
 
-So far, we have determined how to step the time from one transition to the next, but we have not specified how to step the state variables $n$ and $m$ at each transition time. 
-That is, we have not specified the type of transitions that occur at each time step.
-Intuitively, this will depend on the relative magnitudes of the rates from equations (2-5) in the main text: the transition with the highest rate is most likely to occur.
+So far, equation \ref{eq:exp} demonstrates how to step the time from one transition to the next, but it remains unclear how to step the state variables $n$ and $m$ at each transition time.
+There is a need to characterize the type of transition which occurs at a given transition time.
+
+Intuitively, this will depend on the relative magnitudes of the rates from equations (\ref{eq:rate1}-\ref{eq:rate4}).
+The transition with the highest rate is most likely to occur.
 This is formalized by generating the ratios
-\be S = \Bigg\{\frac{R_{MI}(n+1,m|n,m)}{ A(n,m)},\frac{R_E(n+1,m-1|n,m)}{ A(n,m)},\frac{R_D(n-1,m+1|n,m)}{ A(n,m)},\frac{R_{MO}(n-1,m|n,m)}{ A(n,m)}\Bigg\}. \label{eq:rel}\ee
-By construction, $\text{sum}\{S\}=1$.
-By forming cumulative sums of the four ratios, we partition the unit interval $[0,1]$ into four chunks, each associated with a transition -- either migration in, entrainment, deposition, or migration out.
+\begin{multline} S = \Bigg\{\frac{R_{MI}(n+1,m|n,m)}{ A(n,m)},
+	\frac{R_E(n+1,m-1|n,m)}{ A(n,m)},\\
+	\frac{R_D(n-1,m+1|n,m)}{ A(n,m)},
+	\frac{R_{MO}(n-1,m|n,m)}{ A(n,m)}\Bigg\}. \label{eq:rel} \end{multline}
+By construction, $\sum S=1$.
+Forming cumulative sums of the four ratios partitions the unit interval $[0,1]$ into four chunks, each associated with a transition -- either migration in, entrainment, deposition, or migration out.
 The transitions with the highest rates have the largest associated chunks.
-To randomly select the transition that occurs at a transition time, we draw a random number on $[0,1]$ and find which chunk it falls in.
+Drawing a random number on $[0,1]$ can then select the transition which occurs at a given transition time by querying which chunk it falls within.
 
-In summary, to step the process through a single transition, we draw the time interval to the next transition from the distribution (\ref{eq:exp}), then draw a uniform random from $[0,1]$ and use it to select the transition that occurs from the cumulative sum of the ratios in (\ref{eq:rel}).
-The SSA simply iterates this random number generation/selection process to generate exact realizations of the stochastic process: these are series of $n$ and $m$ through time from which we can calculate any statistics of interest.
+In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (\ref{eq:exp}), draws a uniform random from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in (\ref{eq:rel}).
+Iterating this random number generation and selection process simulates exact realizations of the stochastic process.
+These are series of $n$ and $m$ through time from which any statistics of interest can be calculated.
 
@@
 
-By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation (8) of the manuscript over all values of $m$, one can obtain the mean field equation for the particle activity:
-\be 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1)+\gamma(n+1)A(n+1)-(\nu  + \lambda + \mu n + \sigma n + \gamma n)A(n).\ee
+By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation \ref{eq:elemaster} over all values of $m$ obtains a mean field equation for the particle activity:
+\begin{multline} 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1) \\
+	+\gamma(n+1)A(n+1)-(\nu  + \lambda + \mu n + \sigma n + \gamma n)A(n).\end{multline}
 This can be solved by introducing the generating function \citep{Cox1965} $G(x) = \sum_n x^n A(n)$, providing
@@
 
-The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so equation (9) produces 
+The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so equation \ref{eq:elemaster} produces
 \begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m).
 \end{multline}
-Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives equation (11) of the manuscript
+Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives equation \ref{eq:ou}:
 \be 0 = [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1) - 2M(m).\ee
-The \citet{Martin2014} Ornstein-Uhlenbeck model.
 
@@
 \be 0 = -2\overline{\kappa}z_1[zM(z)]' + z_1^2 M''(z). \ee
-Taking into account that this distribution should be normalizable, $\lim_{z\rightarrow \pm \infty}M(z) = 0$, we find
+Taking into account that this distribution should be normalizable, $\lim_{z\rightarrow \pm \infty}M(z) = 0$, the solution is
 \be M = M_0e^{-\kappa (z/z_1)^2}\ee
-as stated in the manuscript.
+as provided in chapter \ref{ch:ch3}.
 
 \subsection{Closure equation approach for bed elevations}
-This section produces an approximate relationship to close $\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$.
+This section produces an approximate relationship to close
+$\langle n | m \rangle$ in terms of $m$ valid to first order in $\kappa$. Writing 
 \be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
-for the mean particle activity conditional to the elevation $m$ into into manuscript equation (9), noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
+for the mean particle activity conditional to the elevation $m$ into \ref{eq:elemaster}, noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
 \begin{multline} 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]\\-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
@@
 \Big[2-\kappa m \frac{(\sigma+\mu)c}{E} \Big]M(m)\Big\}'.\ee
-Finally, integrating, expanding to first order in $\kappa$, and exponentiating to solve for $M$, we find
+Finally, integrating, expanding to first order in $\kappa$, and exponentiating to solve for $M$ produces
 \be M(m) = M_0 \exp\Big\{-\kappa m^2 \Big[1 + \frac{(\sigma-\mu)c}{2E}\Big]\Big\}.\ee
-Since the numerical solutions show $\sigma_m^2 = \frac{1}{4\kappa}$, we determine the closure relation
+Since the numerical solutions show $\sigma_m^2 = \frac{1}{4\kappa}$, this suggests the closure relation
 \be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \ee
-corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation (12) of the manuscript that is plotted with the numerical simulations in manuscript figure (4a).
+corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation \ref{eq:closure} and plotted with the numerical simulations in figure \ref{fig:condmoms}.

### 2021-09-03 draft/ch3bedEle.tex
added phd thesis.

--- 2021-07-07 ch3bedEle.tex
@@
 In equations (\ref{eq:rate2}) and (\ref{eq:rate3}), $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
+$\kappa$ is referred to as a coupling constant since it controls the strength of this feedback. I will later demonstrate the relationship
+\begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
+\end{equation}
+where $l$ is a characteristic length scale of bed elevation change that can be interpreted as the active layer depth \citep{Wong2007,Church2017}.
 $\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
@@
 In conditions when the coupling constant $\kappa$ between the entrainment/deposition rates and the bed elevation is weak, it is also possible to solve the master equation approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
-The Gillespie simulation algorithm is described below in \ref{sec:numerical}, while these analytical approximations are described in \ref{sec:analytical}.
+The Gillespie simulation algorithm is described below in \ref{sec:elenumerical}, while these analytical approximations are described in \ref{sec:eleanalytical}.
 
 \subsection{Numerical study of the joint model}
-\label{sec:numerical}
+\label{sec:elenumerical}
 
@@
 \subsection{Approximate solutions of the joint model}
-\label{sec:analytical}
+\label{sec:eleanalytical}
 
@@
 $M_0$ is a normalization constant. 
-As we will demonstrate, this closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.
+This closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.
 
@@
 \end{figure}
-Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparam} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
+Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
 The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry.
@@
 
-\section{Summary and conclusion}
+\section{Summary}
 \label{sec:eleconclusion}

### 2021-09-03 draft/ch4downDiff.tex
added phd thesis.

--- 2021-07-08 ch4downDiff.tex
@@
 The target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if it is known to have started with the initial distribution $p(x,0)=\delta(x)$.
-Times spent moving or resting on the surface are characterized by exponential distributions $\psi_2(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
+Times spent moving or resting on the surface are characterized by exponential distributions $\psi_M(t)=k_2e^{-k_2 t}$ and $\psi_1(t) = k_1e^{-k_1t}$, since numerous experiments show thin-tailed distributions for these quantities \citep{Fathel2015,Roseberry2012,Einstein1937,Ancey2006,Martin2012}.
 The end results will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.

### 2021-09-03 draft/ch5langevin.tex
added phd thesis.

--- 2021-07-07 ch5langevin.tex
@@
 In the theory of granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
-Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Chapman1970,Landau1980}.
+Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970}.
 Collisions in these models involve episodic kicks to the particle momenta at random times, parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
@@
 A secondary one is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
-I develop the model and explain its key assumptions in section \ref{sec:langlangmodel}.
+I develop the model and explain its key assumptions in section \ref{sec:langmodel}.
 Then I present the analytical solution of the model and other major results in \ref{sec:langresults}.
@@
 
-Figure \ref{fig:1} indicates the configuration of the model in this chapter.
+Figure \ref{fig:fig1} indicates the configuration of the model in this chapter.
 Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow.
@@
 \subsection{Episodic collisions}
-Collision forces dissipate streamwise momentum, partly by converting it to vertical, lateral, or rotational momentum, and partly by deforming particles and generating heat \citep{Schmeeckle2014,Williams2021}.
+Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or rotational momentum, partly by deforming particles and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid \citep{Joseph2001,Schmeeckle2001}.
 The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}.
-Here, we introduce a simple elasticity parameter $\varepsilon$ as a first stab at the problem, as indicated in figure \ref{fig:1}.
+Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in figure \ref{fig:fig1}.
 This elasticity characterizes the fraction of downstream momentum lost in a collision.
@@
 \rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
-In the present model I set $C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631})$ \citep{Clift1978,Gonzalez2017}, and I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.
+In this study, I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.
 
@@
 This result demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
-In contrast, particle velocity fluctuations depend sharply on the parameters representing particle-bed collisions, suggesting collisions are a leading control on particle velocity fluctuations.
+This is consistent with the remarkable similarities sometimes visible between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
+In contrast, particle velocity fluctuations in equation \ref{eq:varu} depend sharply on the parameters representing particle-bed collisions, suggesting collisions are the leading control on particle velocity fluctuations.
 
@@
 The Gaussian limit as $\ve \rightarrow 1$ of \ref{eq:steadystate} is more difficult to evaluate.
-The challenge is that the statistical moments \ref{eq:mean} and \ref{eq:2ndmom} diverge at the same time as the denominator factors of the distribution \ref{eq:steadystate}.
-In appendix \ref{sec:langappendixB} I return instead to the original equation \ref{eq:governer} to evaluate this elastic limit, obtaining
+The challenge is that the statistical moments \ref{eq:mean} and \ref{eq:varu} diverge at the same time as the denominator factors of the distribution \ref{eq:steadystate}.
+In the appendix section \ref{sec:langextremes} I return instead to the original equation \ref{eq:governer} to evaluate the completely elastic limit, obtaining
 \be P(u) = \frac{1}{\sqrt{2\pi\sigma_u^2}}e^{-\frac{(u-\bar{u})^2}{2\sigma_u^2}}. \label{eq:gaussian}\ee
@@
 Comparing eq.
-\ref{eq:steadystate} with experimental data requires values for the steady component of the drag force $\Gamma$, the particle mass $m$, the magnitude of turbulent drag fluctuations $D$, the rate of particle-bed collisions $\nu$, and the most common fraction of momentum $\varepsilon$ dissipated in each collision.
+\ref{eq:steadystate} with experimental data requires values for the steady component of the drag force $\Gamma$, the particle mass $m$, the magnitude of turbulent drag fluctuations $D$, the rate of particle-bed collisions $\nu$, and the mode elasticity of collisions $\varepsilon$.
 
-Below, I compare this formula with the results of six different experiments.
+This section compares eq. \ref{eq:steadystate} with the results of six different experiments.
 In each case, the particle mass is computed from experimental parameters assuming spherical sediment as $m = \pi \rho_s d^3/6$, where $\rho_s$ is the sediment density and $d$ is the particle diameter.
@@
 \be \Gamma =  \frac{\pi}{8}\rho C_D(\Rey_p) d^2 u_\ast^2,\ee
-where $\rho$ is the fluid density, $C_D(\Rey_p)$ is the drag coefficient dependent on particle Reynolds number, and $u_\ast$ is the shear velocity of the flow, considered as a proxy for the average slip velocity between moving particles and the flow.
+where $\rho$ is the fluid density and $C_D(\Rey_p)$ is the drag coefficient, given as \citep{Clift1978,Gonzalez2017}
+\be C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631}). \ee
+Particle Reynolds numbers are estimated using the shear velocity: $\Rey_p= u_\ast d/nu$.
+
 The magnitude $D$ of turbulent fluctuations is eliminated with the mean and variance of the particle velocity.
-The remaining coefficients in the model, the dissipation per collision $\ve$ and the collision rate $\nu$,, are treated as calibration parameters and are tuned to provide best fit between the distribution \ref{eq:steadystate} and the experimental data.
+The remaining coefficients in the model, the dissipation per collision $\ve$ and the collision rate $\nu$, are treated as calibration parameters and are tuned to provide best fit between the distribution \ref{eq:steadystate} and the experimental data.
 
@@
 \end{table}
-In every case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model \ref{eq:langevin} captures most of the underlying physics.
+In every case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model \ref{eq:langevin} captures the essential physics.
 \begin{figure}
@@
 
-Here, I developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
-The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions,
-generalizes earlier approaches available in the literature which did not treat episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
-Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, we have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise.
-Even though the non-Gaussian aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
-We modelled particle-bed collision forces as a sequence of instantaneous impulses where the intervals between successive collisions were characterized as exponential random variables.
-The effect of each collision on the streamwise particle velocities was parameterized by a restitution-like coefficient.
-Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the setting here is somewhat different than grains in air.
-Because particles within a viscous flow in general interact at a distance, we should expect the collision model will become poor when the time between subsequent collisions becomes small.
-Therefore, although the model seems appropriate for saltation, it should be critically examined for ``reptation", when the times between subsequent particle-bed collisions are short.
-Of course, more realistic flow and collision forces could always be incorporated into Langevin equations for bed load transport, although these might require numerical methods for their solution.
-The theory of granular gases provides some indication of the types of more nuanced collision models which are possible \citep{Brilliantov2004}.
+In this final chapter I developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
+The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions.
+This work generalizes earlier approaches available in the literature which did not treat episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
+
+Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise.
+Even though the non-Gaussian and history-dependent aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+
+The model developed in this chapter described particle-bed collision forces as a sequence of instantaneous impulses in an approach that is reminiscent of the kinetic theory of gases \citep{Landau1969}.
+The effect of each collision on the streamwise particle velocities was described by a simple elasticity-like coefficient.
+Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the elasticity coefficient introduced here is not equivalent to the ``coefficient of restitution" typically applied in granular gas theory.
+The coefficient of resitution is defined normal to the contact plane of two particles during a collision, and it characterizes energy loss due to deformation of the colliding particles in the absence of a viscous fluid \citep{Brach1992,Ismail2008}.
+
+In bedload transport, colliding particles are submerged in a viscous fluid, and this introduces additional damping processes besides particle deformation \citep{Joseph2001,Yang2006,Schmeeckle2001}.
+The model developed here did not consider these viscous forces, nor did it explicitly model the geometry of collisions, as $\ve$ was defined as a parameter applying to the downstream velocity only.
+Thus, although I have included episodic particle-bed collisions in a stochastic sediment transport model for the first time, the key parameter ($\ve$) in this formulation has a rather heuristic character and is not clearly related to the coefficient of restitution from granular gas theory.
+Future studies should formulate particle-bed collisions in stochastic sediment transport models considering more details of collision geometry \citep{Sekine1992}, fluid-particle interactions \citep{Marshall2011}, and traditional restitution \citep{Brach1989} using the present study as a starting point.
+
+\subsection{Does the velocity distribution depend on particle size?}
+
+Several researchers have considered why particle velocity distributions differ from one experiment to the next.
+One prevailing view is that it relates to flow hydraulics \citep{Wu2020}.
+Many of the experiments producing Gaussian velocity distributions were conducted in supercritical flows \citep[e.g.][]{Heyman2016,Martin2012,Ancey2014}, whereas many producing exponential distributions were conducted in subcritical flows \citep[e.g.][]{Fathel2015,Charru2004,Seizilles2014}.
+However several experiments run counter to this hypothesis.
+The experiments of \citet{Lajeunesse2010} produced distinctly exponential velocity distributions in flows well within the supercritical regime ($\Fro=1.5$), whereas \citet{Liu2019} produced Gamma-like distributions in the subcritical regime ($\Fro=0.3$).
+
+An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates from granular interactions.
+Particle size enters the Langevin model \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
+The analytical velocity distribution \ref{eq:steadystate} was fit to six different experimental datasets in figure \ref{fig:fig4ch5}, and the best fit parameters required to for these fits were presented in table \ref{tab:calib}.
+These fit parameters suggest collisions are generally less elastic for smaller particle sizes, whereas they are more elastic for larger particle sizes.
+This suggests that the amount of momentum dissipated by collisions may depend on particle size.
+
+Because the elasticity $\ve$ lumps together collision geometry and dissipation effects, it is challenging to attribute the increasing relationship between elasticity and particle size evident in table \ref{tab:calib} directly to particle size.
+Yet this would be consistent with experiments of idealized particle collisions in viscous fluids.
+These demonstrate that momentum dissipation varies sharply with particle size, being more elastic for large particles, and less elastic for small particles \citep{Joseph2001,Yang2006,Schmeeckle2001}, exactly like the trend in the table.
+A definite conclusion that particle size controls the shape of the bedload velocity distribution requires additional experimental and theoretical study.
+This chapter nevertheless produces suggestive ideas.
+
+\section{Summary}
+\label{sec:langconclusion}
+This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
+The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the full range of bedload velocity distributions that have been reported in experimental studies.
+
+This work hints that particle size, and not flow hydraulics, may be primarily responsible for the different bedload velocity distributions obtained in experiments, although future study is required for a definite conclusion.
+The next step is to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions.
+Such an effort would produce additional insight into the controls of transverse, vertical, and rotational movements on the downstream velocities of individual grains.
+
+\endinput
 
@@
 Even in the absence of sorting effects and differential mobility, smaller grains can be expected to carry more control over the largest fluctuations in the overall transport rate, since their velocity distributions have wider tails.
-
-\subsection{Implications for landscape evolution}
-
-Many phenomena of landscape evolution depend on the velocities of individual grains.
-The erosion rates of bedrock rivers are for example parameterized by the impact velocities of individual sediment grains against the bedrock surface \citep{Sklar2004,Li2020,Turowski2021}, on which they have a sensitive dependence.
-It is not much of a stretch to imagine that bedrock erosion could be dominated by the largest impact velocities that originate from the extreme tail of the bedload velocity distribution.
-Here, we have shown that the weight of this tail depends sensitively on the dissipation characteristics of particle-bed collisions, suggesting that another feedback, that between dissipation and ...
-may be at play in the evolution of bedrock canyons.
-
-\section{Conclusion}
-\label{sec:langconclusion}
-We have demonstrated that particle-bed collisions control the shape of the particle velocity distribution.
-
-\endinput
 

### 2021-09-03 draft/formerPapers/com-piece/scraps/com-piece1.tex
added phd thesis.

\documentclass{article}
% General document formatting
\usepackage[margin=0.7in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{unsrtnat}

% \usepackage[math]{kurier}
\usepackage{setspace}       % \onehalfspacing and \singlespacing
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}

\begin{document}

\title{On argumentation and gravel rivers: Einstein and Bagnold in a modern context}
\author{Kevin Pierce}
\maketitle

%% intro

Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
Since the transport of coarse material as bedload is an important control on channel morphology, bedload transport has been the subject of intense research for over a century. 
Despite this effort, our ability to predict bedload transport rates is inadequate.  
In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000. 
Since the bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
However, because bedload transport results from a turbulent flow interacting with granular matter, the full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and among these, two of them have been taken up by generations of researchers, forming two main research paradigms in the physics of bedload transport. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

Each of these approaches is founded on the application of mechanical reasoning to a simplified concept of bedload motion, characterized by empirical parameters, so that both categories of models must be calibrated to applications. 
Also, the simplified concepts of Einstein and Bagnold-type models are different with regard to most aspects of bedload motion.
Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and that neither approach is clearly superior than the other. 
Specifically, both of these mechanical paradigms are expected to generate poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics. 
Accordingly, although each mechanical approach has clear merits and demerits, and it's safe to conclude neither is perfect; the perspective that either is fully superior, for any reason apart from practicality, lacks objective support. 
Because there is no clear path to join the two schools together, proponents of each school toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift that Einstein and Bagnold have left behind. 
We need to ask: is this rift justified by evidence? Are Bagnold and Einstein really so different that they share no common ground, and is one or the other completely wrong? If so, theorists should out the failed approach and concentrate on its successor together. If not, we should work together to synthesize something new out of the strongest elements of each paradigm.

To answer these questions, we should carefully assess the concepts of each paradigm, and try to discern any conditions of bedload transport under which one description provides more or less useful information than the other. 
Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical picture of individual bedload motions, and this picture should be consistent with contemporary experiments. 
Bedload grains typically move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}.
% resting, entrainment by impulse, motion in turbulent flow, collisions with bed, deposition

% explain how bedload transport has different behaviors at low and high stages of bedload transport

% introduce the einstein theory, its revisions, and experiments which support it

% introduce the bagnold theory, its revisions, and experiments which support it

% highlight that bagnold is the high transport stage limit of tau^(3/2), while einstein is really the low transport stage limit where average properties of the flow and grains do not characterize bedload transport

% really, we should we should be joining these theories to describe wolman's intermediate region, which is where channel morphology happens

% however, in intermediate regions, the linkage between morphology and transport becomes more highly coupled; so we really need to integrate knowledge from higher scales with the stochastic theories of einstein and the power dissipation arguments of bagnold -- neither is good enough

% conclude by wrapping problems, stochastic, determinitsic, a synthetic description of intermediate conditions, where feedbacks are accentuated with morphology, with an interesting conclusion about scientific divides and tribalism impeding progress

Sediment grains within gravel-bed rivers typically move intermittently, rolling, sliding, and bouncing along the river bed, in a condition called bedload motion \citep{Einstein1950, Bagnold1973}.
The bedload motion of an individual grain is an alternating series of movements and rests, with these dynamics undoubtly governed by a Newtonian force balance between the solid and fluid phases \citep{Schmeeckle2013, Ji2013}.
However, the forces making up the balance are not exactly known in any context, and the granular \citep{Lamb2008, Frey2011, Booth2014} and fluid \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration. 
Looking at experiments on the mechanics of bedload transport, it's clear that the Einstein and Bagnold schools are simplified from reality.

We should really be attempting a synthesis between them, but to achieve this, we have to assess the experimental findings on the processes underlying bedload transport, as well as the concepts and development of the Einstein and Bagnold schools.  
To make this synthesis, we should be greedy, taking useful ideas from wherever we can find them. 
To this end, we can start from a question: which concepts from each school fit with experimental findings, and how can they be fused together for a predictive theory of bedload transport?

%% bedload experiments

% first describe broad brush ideas. initiation, motion, cessation 
To approach the question, we should develop a clear mechanical understanding of individual bedload motions which is supported by experiments, and we should consider where Einstein and Bagnold's theories fall in relation to it.  
Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact. 
Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed. 
Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.

% now describe gaps in this understanding
This broad mechanical perspective on the processes underlying individual particle motions reveals many gaps in experimental knowledge. 
For one thing, in discussing the

% now describe the einstein theory, and

% initiation
% motion
% cessasion
% rates

%% Einstein and Bagnold

Motion of an individual particle is initiated when a sufficiently strong driving force is applied by the turbulent fluid over a timescale large enough to overcome the resistive forces imparted on that grain by frictional and normal responses from the every bed grain it contacts \citep{Paintal1971, Valyrakis2010, Celik2014}. 
In general, both the duration and magnitude of the applied fluid forces is important \citep{Diplas2008, Valkyrus2010, Celik2014}. 
This is an impulse concept, although in a viscous flow without velocity or pressure fluctuations, this impulse concept reduces to the classical perspective of a critical motion condition based upon velocity or shear stress \citep{Shields1936, Buffington1997}.

This initiation of motion by a turbulent impulse is undoubtedly governed by Newton's laws, although the forces making up the balance are not exactly known in any context, and the granular resistive \citep{Lamb2008} and fluid driving \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration. 
Similarly, the trajectories of particles, once they are in motion, are determined by Newton's laws, contingent on the specifics of turbulent driving forces, including the full time series of turbulent forces, and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The cessastion of motion is poorly understood, but presumably results from a chance coordination of local bed geometry with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.  
Owing to these factors, a Newtonian description of bedload motion based on equations characterizing momentum exchange, is a difficult and probably impossible prospect. 
This difficulty results from the diverse phenomenology of the fluid and granular phases, and especially from the processes that hold them in feedback to induce bedload motion.

Of course, the details of the fluid driving forces onto moving and stationary grains, expressed from the turbulent fluid flow interacting with its randomly arranged and porous gravel boundary, and exchanging momentum with moving bedload grains, are exceedingly complex. 
In feedback with the fluid, the locations of all moving and stationary grains serve as boundary conditions for the Navier-Stokes equations governing the pressure and velocity fields within the fluid, the same fields which drive bedload motion \citep{Nikora2013}.
Even were the flow fields completely known, the way in which the velocity and pressure fields within a fluid exert forces on moving particles is a topic of great uncertainty, especially when the particles under consideration are not spheres \citep{Maxey1983}.
However, in a gravel bed river, these flow fields are defintely unknown, and, of course, the particles are not spheres.  
Even over smooth boundaries, a steady and uniform open channel flow supports spatially extended and transient horseshoe vortices \citep{Adrian2007}: these are generated by rolling packets of fluid which curl up away from the bed: the toe of each horseshoe is relatively elevated to its heel prongs, which drag along the bed and manifest as low-speed streaks in the near-boundary flow \citep{Klein1967, Nino1998}. 
As they move downstream, one conceptual picture is that these coherent structures shed self-similar vortices at smaller and smaller scales, eventually dissipating energy into viscosity as the smallest turbulent scales of the fluid. 
Eventually, from this dissipative cycle, relatively slow near-bed dissipative structures mostly lose coherence and are suddenly swept away by a faster inrush of fluid from above: this cycle of coherent turbulence formation and destruction is referred to as the turbulent burst cycle \citep{}, and experiments have indicated its crucial influence on particle motion \citep{Nino1998, Amir2017}.

In flows over rough boundaries, the idealized horseshoe vortex picture breaks down: the irregular boundary prevents vortices from attaining full coherence, so the velocity and pressure fields within the fluid are informed by the dynamics of these partially coherent vortices, which probably form off of roughness elements, translate, shed smaller vortices, and scatter off of other vortices; meaning the velocity and pressure fields within the fluid are really the empitome of complexity.
Pertaining to stationary bed grains, the formation of vortices off of roughness elements implies sheltering phenomena, whereby all particles on the bed surface do not feel the same fluid forces due to wake effects \citep{Egiazoff1965, AnnaM2017}. 
This shedding can be intermittent or even periodic \citep{}, and it is extended in space, implying regions of exposure propagate non-local effects from the configuration of bed surface grains into the driving forces imparted to those bed grains by the overlying fluid \citep{McEwan2004}. 
This is not to mention the influence of hyporheic flow, or flow below the bed surface, which is poorly understood but has perplexing behaviors \citep{Cooper2017}.  
Apparently, pore regions within the bed surface can support relatively large pressure gradients within the hyporheic fluid, and open porosity on the surface of a gravel bed presumably modifies the pressure and velocity fields in near-bed regions as a result of the hyporheic flow acting as a momentum sink or source, or more probably, both. 
Moving bedload grains, meanwhile, dissipate momentum from the flow and imply changes in its mean and turbulent pressure and velocity fields \citep{Singh2010, Santos2014, Liu2016}. 
Usually, this rich phenomenology of the fluid phase and its interaction with the granular phase is approached by one or two types of gross simplification, which can work well under particular flow and sediment conditions: either we neglect all uncertain aspects of the fluid, characterizing the fluid flow, complete with all its imbricated vortices, coherent or partially coherent, shed in a messy and unpredictable self-similarity, by a time and/or space averaged quantity such as the shear velocity or stress \citep{Yalin1972, Stelczer1983}; or we understand the forcing imparted by the fluid and / or the sheltering effects introduced into the fluid by the granular phase in terms of simplified statistical concepts, such as the concept of a probability distribution (homogeneous in space) of fluid velocity \citep{Einstein1950, Paintal1971} which may be linked to the distributions of lift and drag forces \citep{Hofland2006, Schmeeckle2007}.

As diverse as the details of the fluid phase seem, the details of the granular phase are much less well studied, so considerations are speculative \citep{Frey2011}. 
Borrowing perspectives from granular physics, we can conclude that the bed of static grains is maintained by the coordination of frictional and normal responses at every contact point between two grains within the granular assembly. 
Accordingly, the stability of bed surface grains is contingent on the geometry of contact points with the grains which support them \citep{Coleman1967, Paintal1971}, and this geometry is, in a practical sense, random \citep{Bennett1972}. 
When grains are stationary, the collective frictional, contact, and external forces mutually cancel out on every grain -- action and reaction -- to maintain static equilibrium  \citep{Cundall1979}. 
Frictional forces within the granular assembly depend on granular shape and the material; considerations of granular rheology are, by anyone's judgement, on the list of open problems in modern physics. 
External forces on the granular assembly result from the turbulent fluid shear and impacts from moving bedload. 
Turbulent fluctuations in the fluid forcing or impacts from moving bedload, if they do not induce motion of stationary grains, are necessarily dissipated by adjustments in the force matrix spanning the network of granular contacts within the stationary bed. 
Accordingly, the static equilibrium of the granular bed is, necessarily, under a turbulent flow and subject to collisions from moving bedload particles, maintained by a fluctuating and imbricated matrix of forces across contact points, and, from a mechanical perspective, this object underlies the details of bedload motion.
Whether we can get an appropriate description of bedload transport without incorporating at least some of the details of this granular force balance is an open question. 
So far, most descriptions, whether stochastic or determinsitic, have neglected almost all of its details.

There are indications that, in granular matter, the force matrix between grains self-organizes in such a way as to place most of the force required to maintain static equilibrium across a minority of the contact points in the granular assembly, developing chains of granular forces \citep{Peters2005, Corwin2005, Liu2010}. 
Accordingly, granular forces on individual grains are expected to deviate unpredictably in relation to the orientation of force chains through the granular medium, and these factors are not easy to measure. 
These chains of granular forces have been invoked to explain the otherwise anomalous stability of channel-spanning steps in gravel bed rivers \citep{Zimmerman2010, Saletti2016}. 
Usually, the complexity and potential self-organization of forces within the granular phase, and the coordinated nature of the force balance within the granular phase, is simplified drastically in one of a few ways. 
Either the result of all granular interactions which imply resistance to motion are represented by some averaged parameter(s), such as the critical shear stress \citep{Shields1936, Montgomery1997}, which possibly lies on a distribution due to the random geometry of contact points supporting surface grains or the random effects of sheltering \citep{Wiberg1987, Bridge1992, Ferreira2015}; or they are described in terms of probabilistic concepts, such as the probability of erosion in an instant of time \citep{Einstein1950}, which is either considered an empirical parameter \citep{Ancey2008} or is understood in terms of an exceedance probability of driving over resisting forces, where the resisting forces are usually simplified to include only the particle weight \citep{Einstein1950} or, at most, its normal response off of the supporting contacts \citep{Paintal1971, Dey2018}.
The question of organization in the granular forces impeding the mobility of bed grains is a young inquiry, and its potential influence on bedload motion has not really been quantified \citep{}.

- granular sizes and shape
- sorting, armoring, sheltering effects
- micro meso macro forms , connection with balance of competence and supply 
- collective stability 
- feedbacks with fluid
- basically, none of these aspects key into bagnold or einstein like models at all. 
- grains are spherical and characterized by uniform mobility conditions
- multiple sizes are treated independently or by some representative size

To this point, I've pointed out complexities in the fluid and granular interactions underlying bedload transport, and I've alluded toward the simplifications that stochastic and deterministic approaches have made to glean understanding despite them. 
The optimist might hold that these simplifications are justified: turbulent impulses, coherent turbulence, sheltering phenomena, the energy dissipation of moving bedload, the effects of particle shape on granular friction, the organization of granular forces into chains, the sorting of granular sizes -- sheltering and armoring effects, collective stability in bedforms -- these factors either don't matter, or they matter only at impractically small scales, with no influence on scales of interest to river scientists who might apply bedload transport predictions to understand morphology changes in relation to engineering or ecological problems.
Another perspective is that, although these reach scale or smaller phenomena in the granular-fluid system might be problematic for our understanding of bedload transport, but they pale in comparison to wider issues, such as the input of sediment from landslides, geological controls on channel form, the hydrology of watersheds (with contingence on random weather and shifting climate), or the influence of wood on forest channel dynamics and sediment transport. 
Probably, these thinkers are correct; these factors deeply affect bedload transport rates and channel form; earth science is holistic, and gravel bed rivers probably epitomize this.

Yet we should not conclude that mechanical perspectives on bedload transport are useless for reach scale problems: with this same fallacy, one can argue that chemistry is useless for biology; when, in fact, chemistry and biology are both essential components of medicine. 
In gravel bed rivers, the coupling between the small and the large goes both ways: bedload transport on smaller scales compounds to express morphology on larger scales, but transport is not the only control. 
At best, there is a partial emergence of gravel bed morphology from bedload transport. 
Wood and supply, hydrology and geology -- these play key roles in morphology too, and their influence at large scales goes back down to smaller ones; morphological elements form the boundary of the fluid flow and inform the balance of forces which drive sediment motion, so transport becomes independent of watershed processes only under particular limits. 
- neglecting watershed processes, morphology and sediment transport are still not very connected 
- einstein and bagnold limits are perplexing in relation to morphology -- there are open questions
- for example, one common approach to study the relationship of morphology and transport is by coupling mass conservation principles such as the exner equation with an equation for the flow such as as the shallow water equations. 
- under such a treatment, bedforms do not actually arise: this has motivated alternative stochastic morphodynamics models which do show bedform generation 
- jerolmack and mohrig 2005 and Bohorquez and Ancey 2018
- in light of this, bagnold and einstein type models of bedload transport arise from a considerable simplification: to get them, we must neglect the connectivity of scales within the watershed and the influence of watershed processes on channel dynamics

- so we have some kind of perspective on the full array of processes which may control sediment transport rates in natural channels 
- now we can discuss what sediment transport rates are actually like 
- at relatively low stages of transport, intermittency is highlighted, and sediment moves in a rarefied way. 
- sediment transport rates are highly fluctuating even under the most controlled laboratory conditions
- the critical shear stress is only approximately valuable and below the critical shear stress, transport rates are non-zero 
- at higher stages of transport, bedload transport transitions toward a sheet flow, where particles move in coordination and statistical attributes are minimized relative to mean values
- going forward to describe the approaches of bagnold and einstein, we should keep account of these two extremes of motion characteristics

- the bagnold and einstein approaches were really formulated in order to describe these two extremes
- at lower transport rates, einstein works 
- Einstein observed the transport of sediment grains under typical flow conditions and noted their statistical attributes. 
- he conceptualized sediment motion as a random switching between states of motion and rest
- he characterized the sediment transport rate in terms of the rate of switching between rest and motion, or erosion, and the rate of switching from motion to rest, or deposition. 
- he undrestood the erosion rate in terms of the exceedance probability of the force due to turbulent lift over the force due to weight. 
- thus he did not factor granular forces or drag into particle mobility, but only evaluated water forces in the upward direction. 
- he used these concepts in order to compute the mean bedload rate; 
- notably, he assumed the rates of erosion and deposition were steady, and did not depend on location on the bed, nor did they depend 
- so einstein's model has no influence of morphology: it does not allow for any 
- nevertheless, under suitably uniform conditions, when morphology is suppressed, it describes bedload rates well at low discharges, when its parameters are properly calibrated. 
- More recent Einstein-like models have amended his concepts to include downstream fluid forces and pivoting about granular contacts in order to have sediment motion
- recently, instead of getting mean rates, einstein-models have been extended to predict bedload fluctuations and other statistical moments -- we can now obtain the full probability distribution of the bedload rate from Einstein's concepts
- however, experimentally derived fluctuations using the concepts of Einstein are not wide enough, which has led to the introduction of new potential processes
- granular avalanches, collective effects
- we conclude that einstein-like models provide a clear conceptual framework, but that more work is needed to understand the inputs and processes at play which drive the motion-to-rest switching of bedload grains. 
- they do adequately describe bedload rates at relatively low discharges, when intermittency and irregularity is emphasized in bedload rates
- however, they fail at higher discharges when the divide between motion and rest states becomes blurred: under these conditions, transport becomes a sheet flow where grain-to-grain interactions are emphasized
- notably, there is no limitation of the flow rate on the number of particles which can go into motion
- presumably, this is limited by the power available in the water flow

- at higher transport rates, bagnold like models work 
- bagnold considered sediment transport the result of the balance of power provided from the flow against the power dissipated by friction in the granular phase. 
- he characterized both of these quantities, the driving and the dissipation, by average parameters in time and space including the excess shear stress, and he arrived at a transport formula which exhibits at scaling with average shear stress of tau3/2.
- a few others have generalized bagnold type models (how?) 
- a key component is the so-called bagnold hypothesis, which is that the threshold of motion and the fluid shear stress tend to equilibrate at the bed surface
- parker showed that the bagnold hypothesis was false.
- this scaling had been derived earlier from the careful empirical studies of meyer-peter. 
- nevertheless, in straight laboratory channels, these types of models work well 
- however, in natural channels with issues of supply, morphology, and inhomogeniety in flow and mobility conditions, they do not work, missing the mark by orders of magnitude
- further, at low transport rates, the highly intermittent character of bedload transport calls into question the utility of bagnold-type deterministic models 
- at these discharges, fluctuations are so prevalant that mean values without being supplemented by expected magnitudes of fluctuations lose their meaning
- bagnold is capable of explaining transport rates at lower discharges, however the parameters of bagnold-type models must be fit to unphysical parameters. 
- although bagnold type models have a mechanical basis, they hinge upon invalid concepts and their extension to multiple grain sizes is not obvious

- in practice, mechanical models of bedload transport based upon the concepts of bagnold or einstein do not usually perform better than much simpler models based upon simple power law correlations
- however, since these empirical models do not actually work either, we push on with the mechanistic approaches
- physics can be a guide for model development: when processes get complex we should lean on the principles
- in the case of einstein and bagnold, the question is how to take what we know and can deal with and map it onto physical principles
- einstein used probability arguments in conjunction with the principle of conservation of mass, but his approach lacks deference to the power availble within the fluid flow to drive sediment motion, and it depends on a clean break between states of motion and rest, so while it provides a great statistical description at lower discharges, it breaks down at higher discharges. 
- bagnold-like approaches neglect variation in granular stability or flow competence, and using the bagnold hypothesis, they become able to leverage a power balance argument in conjunction with conservation of mass in order to predict a sediment transport rate. 
- under conditions when these assumptions hold -- steady flow and bagnold assumption -- we expect that bagnold works quite well, and indeed it does
- under such conditions the statistical characteristics of bedload rates are minimized, and transport rates scale as shear stress to the 3/2 power
- we note that either result neglects the coupling between larger scale granular arrangement, organized structures in fluid turbulence, watershed processes, and so on
- maybe some of this is unavoidable 
- however, the most notable deficiency of these models stems from intermediate transport rates: 
- neither model appropriately describes sediment transport away from the limits of high adn low transport
- we conclude that, really, einstein adn bagnold should be viewed as limiting cases of a more general theory
- in fact, they are both short-sighted, with many approximations underpinning them, both acknowledged and unacknowledged
- they are both wrong across most of the regime of sediment transport conditions, 
- and they are both right in at least one extreme regime of sediment transport conditions
- pushing bedload transport understanding forward is in the fusion of these two regimes, and it definitely isn't one or the other.

We conclude what we should really be aiming for is a more complete theory of sediment transport which is rooted in mechanics and confers both statistical and deterministic descriptions under suitable conditions -- purely stochastic, and dominated by fluctuations in fluid driving forces and granular resisting forces at relative low transport rates; and purely deterministic, dominated by average values of fluid driving forces and granular resisting forces at relatively high transport rates. 
The intermediate region, of course, is where the really interesting physics happen -- the river morphology.

Wolman, in his classic theoretical work on geomorphology of river channels \citep{Wolman1968}, introduced the concept of the competent or effective flow. 
To form this concept he highlighted a balance between the frequency and magnitude of hydrological conditions. 
Extremely high river flows may be incredibly geomorphically active, and they may drastically change river morphology over short timescales, but these flows are rare. 
The real geomorphic work is done by intermediate flow conditions which are both relatively common and relatively large enough to overcome the stability of river channels and effect change in their morphology. 
What I'm saying is, following after Wolman, the real geomorphic work is done by intermediate flow conditions over intermediate time scales, and this intermediate range of conditions falls between the two extremes of pure stochastic and deterministic characteristics, the former mostly insensitive of average characterizations of fluid flow and granular resistance, and the latter mostly contingent on them. 
We contend that Wolman's intermediate range is really untouched by any bedload transport theory, stochastic or deterministic.  
In this range, both varying and averaged attributes of the factors driving and resisting bedload motion matter, and incorporating the varying components may require new theoretical approaches, borrowed from granular physics and fluid dynamics, into a fusion of Bagnold-like and Einstein-like bedload transport theories.
Necessarily, these models should be stochastic, because they have to express stochastic transport at lower rates, but at higher rates their output distributions should sharpen, and their predicted fluxes should scale like tau^3/2, in the absence of morphological or watershed-scale influence. 
To develop these approaches, which incorporate everything we know, or select the part of it we consider relevant, we should stop quibbling. 
Transport is deterministic and stochastic; we need to use mechanics because this problem is too complicated to go about without some guiding principles. 
Accordingly, we should integrate and develop the description, and we should do it fast: salmon stocks are at all-time lows, the climate is changing, and forests are being removed at alarming rates. 
Why should we discuss stochastic versus deterministic when each of them works?

\begin{comment}

Where is this coupling in Einstein or Bagnold, or under what limits might they retain predictive power? 
More generally, where are the deficiencies in our understanding of bedload transport, and why do our predictions in natural channels fail?

To arrive at modern bedload transport models, based on mechanical concepts in the schools of Einstein and Bagnold, we're led to neglect almost everything. 
In effect, Bagnold viewed the river as a machine, where water provides the power to move bedload, and friction on moving sediment dissipates it. 
Bagnold represented the power of the fluid phase by the product of the average (time and space) shear velocity with the average shear stress, and he represented the power dissipated by the sediment phase as the energy per unit time to support a given mass of sediment against frictional drag, which he characterized in an average way.

We are saying researchers have crafted two main ways to simplify the vast phenomenology of bedload transport. 
Either we neglect the fluctuating components of the driving and resisting forces implying motion of bedload, and only keep average values, implying deterministic predictions of bedload transport; or we neglect the possibility of a deterministic prediction of bedload transport, and attempt to keep account of the fluctuating components of driving and resisting forces in the most realistic way we can handle, implying a stochastic prediction \citep{Ferreira2015}. 
I tend to believe that these two approaches are both important and valid, and that they describe different regions of the bedload transport parameter space.

The role of these factors on the grain-scale sediment transport is not expected to be reducible: trying to understand all aspects of morphology from the mechanics of sediment transport alone will provide only limited success. 
We know that gravel bed rivers support a wide variety of morphological patterns, which, when classified, form an entire typology -- we call these morphological patterns armoring, clusters, ribs, bars, patches, and so on: the essential point is that morphology can be diverse \citep{Church1998, Hassan2008, Venditti2017}. 
Often, we associate these different morphological patterns with sediment storage or stabilizing roles, and we associate their formation with some regime of hydraulic and sediment supply conditions. 
Bars, for example, form under conditions of relatively high sediment supply which channels do not have the competent flow to evacuate, so they act as storage features which sequester sediment away from the main channel and prevent excessive aggradation. 
Clusters, ribs, stone cells, and so on -- arrangements of grains on the surface of gravel beds, or armoring, the difference in grain sizes between surface and subsurface layers of the gravel bed \citep{Parker1982}, result from conditions of relative sediment starvation, so they act as stabilizing features which prevent excessive channel degradation.

Of course, transport is the realm of rocks, and gravel bed rivers in their undisturbed state often flow through forests, or at least vegetated plains. 
Within forests, wood input from deadfall, landslides, or harvesting activity has crucial influence on sedimentological processees -- transport and storage -- within river channels. 
Relatively large pieces of wood will jam in river channels, which can obstruct the transport of sediment and serve as another type of sediment storage: sediment backs up behind jams of wood in relatively large volumes; as a result, over yearly or decadal timescales, channels can avulse, or change their path, in order to bypass woody jams. Alternatively, the action of bedload and rotting implies the degradation of wood pieces forming jams. 
With enough time, log jams can break, releasing sediment pulses into the downstream reach, and implying transient responses to morphological structures within channels. 
Accordingly, there may be an intense feedback and intermittency in the storage and supply characteristics of natural gravel bed rivers, imparted from vegetation and morphology alone. 
The essential point is that supply, hydrology, and vegetation interplay with sediment transport to express channel morphology, which is usually varies over a wide set of spatial and temporal scales. 
Since the channel morphology affects sediment transport characteristics, and the time and space scales of individual sediment transport events may be different than those of morphological changes, we have a difficult problem: sediment transport and morphology are mixed, and vegetation and sediment supply are outside factors which stir them together. 
Usually, investigators make a way out of this morphology-transport-supply-hydrology link by making some common assumptions: either they contend that the timescale of sediment transport is much smaller than the timescale of morphological change, so that they can consider morphology the result of transport, meaning transport occurs over a static backdrop; or they hold that sediment transport is really nothing more than the shift of morphological elements within the river channel, neglecting all of the grain-scale transport events which compound to express morphological elements or their migration. 
Either approach is oversimple: a landslide can happen at the same timescale as the individual motion of a gravel cobble, so the scales of morphology and transport can overlap; and the migration of larger scale morphological elements within a river channel happen in coordination with and alongside the transport events of individual sediment, so to consider the migration of bars is really to neglect the grain scale physics from which this migration emerges. 
One concludes there is a need for a deeper integration of the feedbacks between morphology and sediment transport in order to really understand and predict change in river channels, and especially of habitat in river channels.

Seeing and accepting the complexity and acknowledging our approximations, or the ways we've taken to escape it, is really the context we need to push river science forward. 
The Archimedian scientist, considering the world of wind, water, fire, and earth, unwilling to acknowledge the existence of something deeper and more complex -- fermions, bosons, molecules, cells -- could not have developed CRISPR, the transistor, or 
Fire is a useful concept but under certain conditions its more useful to see combustion as a consequence of hydrocarbons reacting with oxygen exothermically; and the light it exudes as transitions of electrons between different atomic or molecular states. 
The concept "fire" is not enough to solve all fire-related issues. Throwing water on a gasoline fire will not have the intended effect: under certain limits, different attributes of fire matter.  
Science should not be pursued with blind approximations; these approximations should be calculated and chosen from a more complete incorporation of the known phenomenology. 
Accordingly, river scientists need grain-scale theories of bedload motion that depend on the morphological scales under certain limits; we should understand our average shear stresses contrast with known facts, and are only as meaningful as they are useful; we should break with our ties to engineering or geography and see the science of nature as a more holistic problem at the intersection of a diversity of interacting factors; we can see our approaches to understand rivers as approximations with regimes of validity, and find more comfort breaking with our blind convictions about the way nature should be.

Different approaches to reduce the mechanical complexity of the coupled fluid-granular system, including the connection to emergent morphological structures on larger spatial and temporal scales, that is a gravel bed river, and to obtain models of bedload transport with some degree of tractability, seem to fall mostly into two categories: deterministic and stochastic approaches \citep{Ferreira2015}. 
We are saying researchers have crafted two main ways to simplify the vast phenomenology of bedload transport. 
Either we neglect the fluctuating components of the driving and resisting forces implying motion of bedload, and only keep average values, implying deterministic predictions of bedload transport; or we neglect the possibility of a deterministic prediction of bedload transport, and attempt to keep account of the fluctuating components of driving and resisting forces in the most realistic way we can handle, implying a stochastic prediction \citep{Ferreira2015}. 
I tend to believe that these two approaches are both important and valid, and that they describe different regions of the bedload transport parameter space.

For example, we know that at relatively low transport rates, averaged values of the fluid driving forces and granular resisting forces are not appropriate for describing bedload transport. 
For one, the average shear stress has been held constant while the turbulent properties of fluid flow are varied, modifying transport rates by as much as 400 percent \citep{Sumer2003}. 
For another, the critical shear stress is known to be highly variable, difficult to properly define and measure, and a poor predictor of sediment transport at low transport rates \citep{Paintal1969, Kirchener1990, Montgomery1997, Wilcock2008}. 
Under these relatively low transport rate conditions, stochastic models are a much better descriptor of sediment transport \citep{Yalin1972, Ancey2006, Furbish2012}. 
In these conditions, instantaneous bedload fluxes are as much as 400 percent mean values, and the bedload transport signal is more completely characterized by a probability distribution than it is by a deterministic value \citep{Bohm2004, Singh2009}. 
Meanwhile, stochastic descriptions break down at high rates of bedload transport \citep{Yalin1972}. 
Since stochastic models of bedload motion have been formulated in terms of a random switching between motion and rest states, and at higher transport rates these phases become less cleanly separated, the stochastic approach fails at high rates \citep{Heyman2013}.
Accordingly, the stochastic approach pertains to the relatively low rates of bedload transport, where intermittency is emphasized \citep{Ancey2008}.

Of course, deterministic bedload transport formulas have their own place. 
Opposite to the stochastic regime, at higher rates of sediment transport, fluctuations in the bedload rate tend to become small relative to mean flux values, so that intermittency is minimized and transport rates become more steady \citep{Ancey2008}, in accord with deterministic models of bedload transport such as \citet{Meyer-Peter1948} or \citet{Bagnold1966}.  
Curiously, these mean-field descriptions of bedload motion all predict the bedload flux should scale as the 3/2 power of shear stress at high transport rates \citep{Yalin1972}, and this scaling law has been confirmed by experimental data \citep{Gomez1989, Recking2007, Wong2007} and the most detailed numerical models of sediment transport which have been performed \citep{Elghannay2017}. 
At relatively low stages of transport, meanwhile, the parameters of deterministic models typical must be fit to unphysical values in order to describe mean bedload fluxes \citep{}. 
We conclude what we should really be aiming for is a more complete theory of sediment transport which is rooted in mechanics and confers both statistical and deterministic descriptions under suitable conditions -- purely stochastic, and dominated by fluctuations in fluid driving forces and granular resisting forces at relative low transport rates; and purely deterministic, dominated by average values of fluid driving forces and granular resisting forces at relatively high transport rates. 
The intermediate region, of course, is where the really interesting physics happen -- the river morphology.

Wolman, in his classic theoretical work on geomorphology of river channels \citep{Wolman1968}, introduced the concept of the competent or effective flow. 
Within this concept he introduced a balance between the frequency and magnitude of hydrological conditions. 
Extremely high river flows may be incredibly geomorphically active, and they may drastically change river morphology over short timescales, but these flows are rare. 
The real geomorphic work is done by intermediate flow conditions which are both relatively common and relatively large enough to overcome the stability of river channels and effect change in their morphology. 
What I'm saying is, following after Wolman, the real geomorphic work is done by intermediate flow conditions over intermediate time scales, and this intermediate range of conditions falls between the two extremes of pure stochastic and deterministic characteristics, the former mostly insensitive of average characterizations of fluid flow and granular resistance, and the latter purely contingent on them. 
We contend that Wolman's intermediate range is an intermediate range within which bedload transport cannot be described by any existing theory, rather it be stochastic or deterministic.  
In the intermediate zone, both fluctuating and average quantities matter, and truer geomorphic theories will result from incorporation of the big camp of Shields and the tiny camp of Einstein.

Of course, we already had this conclusion, from a different angle. 
Its well known that deterministic morphodynamics models which are based upon power law transport formulas like \citet{MeyerPeter1948} coupled to simplifications of the Navier-Stokes equations, such as the Saint Venant equations, do not imply the emergence of realistic channel features such as bars \citep{Jerolmack2005}.
So in morphodynamics modeling, deterministic models have a deficiency. 
Meanwhile, stochastic models of morphodynamics, so-far restricted to 1D, do express realistic morphodynamic features which are not expressed by deterministic models, such as antidunes \citep{Bohorquez2018}. 
This suggests stochasticity in bedload motion is necessary that it produce realistic morphological structure within river channels. 
Accordingly, from the physical picture and this suggestion that purely deterministic models are not enough to express morphology, we suggest an armistice. 
Einstein and Shields are not opposed but access different regions of the parameter space of bedload transport. 
The resulting theory of bedload motion will be stochastic, as it should be given the phenomenology of the driving and resisting forces we discussed; but it is deterministic in high transport limit, where distributions sharpen and fluctuations become insignificant relative to averaged characterizations of those driving and resisting forces we can't exactly know. 
We conclude that the two camps should start to speak. Bedload transport is stochastic but also deterministic; and neither camp's approach is fully sufficient to describe the form that river channels take. 
It's time for a treaty, and we should ask Newton to write it.

\end{comment} 
\bibliography{biblio}
\end{document}

### 2021-09-03 draft/formerPapers/ens-rev/review0.tex
added phd thesis.

\section{Review: birth-death theories of sediment transport}

Birth death processes are a special type of stochastic process. 
Stochastic processes track the evolution of a set of random variables through time. 
Birth death processes consider a discrete state space, such as the size of a population. 
Members of the population are born and die: transitions are allowed between adjacent sites in the discrete state space.

Within sediment transport, Einstein was the first to consider a model for the bedload flux in terms of stochastic processes. 
He considered the transport of individual bedload grains to be a random switching between motion and rest states. 
In this way, supplementing this consideration with empirical reasoning, Einstein developed a formula for the mean transport rate \citep{Einstein1950}. 
Einstein determined that the bedload transport rate should follow a Poisson distribution.

Einstein's predecessors have extended his formulations in order to bring them in accord with experimental data and to phrase them in terms of a stronger mathematical foundation drawn stochastic process theory. 
These extensions probably began with \citet{Lisle1998}. 
They linked Einstein's theory of tracer dispersion \citep{Einstein1937}, which is contingent on random transitions between states of motion and rest, into Markov process theory, showing that Einstein's tracer dispersion theory was just a two-state Markov process where the states are motion and rest.  
Then \citet{Ancey2006} went on to rephrase the \citet{Einstein1950} bedload flux derivation in terms of a two-state Markov process, where random transitions between states of motion and rest are included into a Chapman-Kolmogorov equation -- a general mathematical formulation. 
In this way, they derived a probability distribution for a bedload flux which is based upon underlying randomness in the entrainment and deposition of bedload grains. 
Taking the mean of their bedload flux probability distribution reproduces an Einstein-like formula for the mean bedload flux. 
This work is sketched to indicate how Markov process theory can work to describe the bedload flux.

\subsection{Formalizing Einstein: a foundation in Markov process theory}

The starting point for the \citet{Ancey2006} work was a set of experiments: the transport of uniform glass beads was measured in a narrow flume using image processsing techniques. 
Measured fluxes had strong fluctuations: instantaneous fluxes took on values as much as ten times the mean. 
Therefore, they attempted to describe the probability distribution of the bedload flux by revisiting Einstein's theory \citep{Einstein1950} from a Markov process framework.

According to Einstein's way of thinking, an individual sediment grain undergoes a random transition from motion to rest states. We can label these two states as $x=0$ and $x=1$. Then, tracking the evolution of the random variable $x$ through time will fully describe the start-stop motion of the individual grain.

Now we prescribe rates of entrainment and deposition in a small time $dt$. 
If the particle is in motion ($x=1$) at time $t$, the probability it deposits to the rest state ($x=0$) at time $t+dt$ is modelled as a constant $\tau^{-1}$. 
Likewise if the particle is at rest ($x=0$) at time $t$, the probability it entrains to the motion state ($x=1$) at time $t+dt$ is modeled as a constant $\sigma^{-1}$. 
Making the Markov process assumption that the transitions of the particle between $t$ and $t+dt$ are independent of its history, the forward Kolmogorov equations read
\begin{align} 
\pi_0(t+dt) = \tau^{-1} dt \pi_1(t) + (1-\sigma^{-1} dt) \pi_0(t) \\
\pi_1(t+dt) = \sigma^{-1} dt \pi_0(t) + (1-\tau^{-1} dt) \pi_1(t). 
\end{align}

Dividing through by $dt$ and taking the limit as $dt \rightarrow 0$ develops the Master equation for the two state process
\begin{align} \label{eq:anc2006}
\dot{\pi}_0(t) = \tau^{-1} \pi_1(t) - \sigma^{-1} \pi_0(t) \\
\dot{\pi}_1(t) = \sigma^{-1} \pi_0(t) - \tau^{-1} \pi_1(t). 
\end{align} 
This simple model is called a telegraph process \citep{Cox1965}.

To solve for the probabilties $\pi_0(t)$ and $\pi_1(t)$, the two equations can be summed to give $\frac{d}{dt} [ \pi_0(t) +\pi_1(t) ] = 0$, meaning $\pi_0(t) + \pi_1(t) = \text{constant}$. Since motion and rest are mutually exclusive and there are no other possibilities for the particle, $\pi_0(t) + \pi_1(t) = 1$: probabilities are conserved. 
Using this result into \ref{eq:anc2006} gives 
\begin{align} 
\pi_0(t) =  \\
\pi_1(t) = 
\end{align} 
where $\pi_0(0)$ and $\pi_1(0)$ are the initial conditions, which we have not specified.

After a sufficiently long time, the particle forgets its initial conditions and settles into equilibrium transport: 
\begin{align} 
\pi_0 = \\
\pi_1 = 
\end{align}
Therefore, the time dependence and initial conditions are forgotten by the Markov process, and it has a set probability $\pi_0$ to be at rest and $\pi_1$ to be in motion.  
Therefore, the fraction of time the particle spends in motion is $\frac{\tau}{\tau+\sigma} = \xi $, and the fraction of the time it spends in rest is $\frac{\sigma}{\tau+\sigma} = 1-\xi$. 
The probability that the particle is in motion is a Bernoulli distribution.

Now consider that on the bed surface within an observation window there are $N$ particles at any time. 
Each of these particles is undergoing the same random switching between motion and rest states, so that the number of moving particles at any time is 
\be N_m = \sum_{i=1}^N x_i \ee
where $x_i=0,1$ is the Bernoulli random variable of the $i$th particle, with the probability of $x_i=0$ being $\pi_0$ and the probability of $x_1=1$ being $\pi_1$.

Therefore, the probability that there are $N_m =n$ particles in a motion state is a counting problem. If $n$ are in motion then $N-n$ are at rest. The result is 
\be \text{Prob}(n \text{ in motion}) = {{N} \choose {n}} \xi^n (1-\xi)^{N-n}. \ee 
In statistics lingo, the sum of Bernoulli trials is a Binomial distribution. 
This distribution has mean $\xi N$ and variance $\xi(1-\xi)N$. 
Therefore the probability distribution of the number of moving particles within the observation window can be written 
\be \text{Prob}(n) = \text{Bi}[ \xi N, \xi(1-\xi)N].  \ee

If all particles in motion move at velocity $u_p$ and have volume $\nu_p$, then the bedload flux is a random variable $Q = \frac{u_p \nu_p}{L} N_m $. Therefore Ancey et al calculated the bedload flux probability mass function as 
\be \text{Prob}(Q) = \text{Bi} [\frac{u_p \nu_p}{L} \xi N, \xi (1-\xi) \frac{u_p^2 \nu_p^2}{L^2}]. \ee

We conclude that the mean transport rate is 
\be \bar{Q} = \nu_p \frac{N}{L}  \frac{\tau}{\tau+\sigma}   u_p. \ee
This is conceptually similar to Einstein's formula \citep{1950}, and Ancey et al. explored this connection more carefully than we will here \citep{Ancey2006}. 
The essential point is that Einstein took a somewhat different perspective.
The key result of the \citet{Ancey2006} theory is that the mean transport rate is the result of $N$ parallel telegraph processes: each of the $N$ particles on the bed undergoes a switching process between motion and rest states, and the net result is the mean rate $\bar{Q}$ which is governed by a binomial distribution.

In context of the experiments Ancey et al performed, the formulation just described did not capture the magnitude of the bedload fluctuations, although it described mean behavior quite well.  
Their experimental bedload fluctuations were much larger than those expressed by this model. 
Said in a different way, the probability distribution of the experimental bedload transport rate had a heavy tail, while the Binomial distribution of the rate derived from this parallel telegrapher's process theory has a light tail.

Heavy tails are a hallmark of collective motions in physical systems \citep{Sornette2000}. 
Of course, if bedload transport is driven by the advection of large turbulent eddies, a perspective supported by many experiments \citep{Nelson1995, Hofland2006, Celik2014, Amir2014, Shih2017}, then bedload motions will not be independent, but they will be coupled collectively to passing turbulent eddies.
In context of the model just presented, this means the state of each telegraph process at an instant will be coupled to every other. 
Therefore, coherent turbulence could imply heavy tails in the bedload probability distribution.

There are other processes which could imply these collective effects as well. One option is entrainment due to particle collisions. 
Moving particles can impact stationary particles, leading to their entrainment. 
Essentially, this means the entrainment rate will depend on the number of active particles. 
This was the extension made by \citet{Ancey2008} in order to capture the large fluctuations seen in their bedload flux experiments.

\subsection{Capturing fluctuations: The inclusion of collective effects}

In 2008, Ancey et al took a somewhat different perspective in order to describe the heavy tailed bedload probability distribution their experiments showed \citep{Ancey2008}.
Instead of considering the motion of every bedload particle as an independent telegrapher's process, they considered the number of particles within a control volume as governed by a collective birth-death process \citep{Gardiner1983, Gillespie1992, VanKampen1992, Cox1965}.  
Birth of a member of the population is just entrainment. 
Death is deposition. 
The population can also immigrate (move in) to the control volume from upstream, and it can emigrate (move out) of the control volume to downstream. 
Birth, death, immigration, and emigration rates characterize the dynamics of the entire population at once, rather than just the dynamics of independent individuals as the \citet{Ancey2006} work pursued.
In this way, they managed to include a collective entrainment effect.

Consider the probability that there are $N=n$ particles in motion at time $t$. 
This can be denoted $\pi(n,t)$. The random variable $N$ takes on values $0,1,2,/dots$ -- non-negative integers of arbitrary magnitude. Each value has an associated probability $\pi(n,t)$.  The entire population is considered subject to four processes within a small time increment dt. 
First, entrainment of a single bed paricle (birth) can happen with probability $\lambda_0 + \mu n$. 
Here a collective entrainment effect has been included. 
The probability of entrainment in $dt$ grows with the number of particles in motion. 
This is a crude inclusion of the collective effects we just discussed: the advection of turbulent structures leads to correlations in entrainment rates, and the collision of moving particles with stationary particles can induce entrainment as well.

Second, deposition of a single bed particle can happen in $dt$ with probability $\sigma n$. 
The death rate is proportional to the number of individuals in the population. 
Third, immigration of a particle from upstream into the control volume can happen with probability $\lambda_1$. 
Fourth, emigration of a particle from the control volume to downstream can happen with probability $\nu$.

These possible transitions develop a system of equations for the probabilities of finding $n$ moving particles within the control volume: 
\be \pi_n(t+dt) = \alpha dt (n+1)\pi_{n+1}(t) + [\lambda + (n-1)\mu]dt \pi_{n-1}(t) + [1-dt[\lambda + n\alpha + n\mu]]\pi_n(t) \ee
As $dt\rightarrow 0$ these develop a system of differential-difference equations for the probabilities $\pi_n(t)$ of finding $n$ particles in the control volume at time $t$: 
\be \dot{\pi}_n(t) = (n+1) \alpha \pi_{n+1}(t) + (\lambda + (n-1)\mu)\pi_{n-1}(t) - (\lambda +n (\alpha+\mu))\pi_n(t). \label{eq:anc2008} \ee

This birth-death immigration-emigration model can be solved by introducing a probability generating function $G(z,t) = \sum_{n=0}^\infty z^n \pi_n(t)$ \citep{Gardiner1983, Ancey2008}. 
Multiplying \ref{eq:anc2008} by $z^n$ and summing over all $n$ gives, after a careful manipulation of the sums in order to form some function of $G(z,t)$ in every term, 
\be \frac{\partial}{\partial t} G(z,t) = \lambda(z-1)G(z,t) + \{ \sigma + \mu z^2 + \nu - (\mu + \sigma + \nu)z\} \frac{\partial}{\partial z} G(z,t). \ee

This partial differential equation for the probability generating function $G$ can be solved by the method of characteristics \citep{}. 
If there are initially $N_0$ moving particles within the control volume, the solution is 
\be G(z,t) + \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \ee
The factor $K$ is the autocorrelation function $K = \exp(-t(\alpha-\mu)).$

This probability generating function generates probabilities via a Taylor expansion around $z=0$: $G(z,t) = \sum_{n=0}^\infty \frac{z^n}{n!}\big(\frac{\partial}{\partial z}\big)^n G(z,t) |_{z=0}$. Comparing this formula with the definition of $G$, $G(z,t) = \sum_{n=0}^\infty z^n \pi_n(t)$, apparently the coefficient of $z^k$ in the power series expansion of $G(z,t)$ is $\pi_k(t)$.

For times sufficiently large for the system to forget its initial condition and settle into a steady state, the autocorrelations $K(t)\rightarrow 0 $ in the expression of $G$, and the power series expansion of $G(z,t) \rightarrow G(z)$ yields a set of stationary (independent of time) probabilities 
\be \pi_n = \frac{\Gamma(r+n)}{\Gamma(r)n!} p^r (1-p)^n\text{, }n=0,1,\dots.\ee
Here $r=\lambda/\mu$ and $p = 1-\mu/\alpha$. $\Gamma(x) = \int_0^\infty z ^{x-1} e^{-z} dz$ is the well known $\Gamma$-function of mathematical physics.

These stationary probabilities are negative binomial distributions: 
\be \pi_n = \text{NegBin}(n;r,p).\ee
This is the generalization of $N$ parallel telegraph processes which was employed in the \citet{Ancey2006} paper. 
The negative binomial distribution has a heavy tail as a result of collective entrainment $\mu$.

When the collective entrainment process is turned off, $\mu\rightarrow 0$ which recovers the \citet{Ancey2006} result. 
$\mu\rightarrow 0$ implies $r \rightarrow \infty$ and $p \rightarrow 1$. 
Within this limit, the generating function G(z) becomes 
\be G(z) = \exp(-\frac{\lambda}{\alpha}[z-1]).\ee
This is the generating function of a Poisson process, so the probabilities of the number of active particles become Poisson distributed 
\be \pi_n = \frac{r'^n}{n!}e^{-r'}\ee
where the ratio $\lambda/\alpha = r'$ defines $r'$. 
This is the same scaling behavior for large $n$ as the \citet{Ancey2006} model, since the large $n$ limit of a binomial distribution is a Poisson distribution.

The Ancey et al 2008 inclusion of collective entrainment develops a Negative Binomial distribution of the bedload flux. 
Expressing the bedload flux as the number of moving particles within the control volume times their average velocity $u_p$ makes a Negative binomial distribution for bedload flux within the control volume. The mean bedload flux is $ \frac{\lambda}{\alpha-\mu} u_p $ and the variance of the flux is $ \frac{\lambda \alpha}{(\alpha-\mu)^2} u_p^2$: the fluctuations are much wider in the negative binomial distribution, meaning the inclusion of collective entrainment can model the large fluctuations observed in the \citet{Ancey2008} experiments.

One way to view the bedload flux we just presented: it is the number of moving particles times their velocity. 
However, since this model explicitly includes downstream emigration from the control volume, there is another way to define the flux: it is the number of emigration events within a unit of time. 
Counting the number of emigration events expressed from this model within a unit time is not easy. 
\citet{Ma2014b} explored this approach to calculating the bedload flux, and I will review this in a subsequent section.

Two key assumptions of this model are worth examining. 
These set the stage for the subsequent developments in birth-death modelling of the bedload flux which I review next. 
First, this model assumes that stationary bed particles are in infinite supply. 
This is an implicit assumption: Nothing prevents the system from entraining particles ad infinitum through the entrainment processes characterized by rates $\lambda_0$ and $\mu$. 
Second, the model considers sediment transport in an isolated control volume, so the sediment flux is described probabilistically in a locality, but spatial correlations cannot be accounted for: the flux upstream of the control volume is also a fluctuating quantity, implying spatial correlations through immigration events into the control volume considered in this model.

These two observations support two of the more recent developments to the \citet{Ancey2008} birth-death immigration-emigration model: \citet{Turowski2009} generalized this model to finite bed material in order to describe transport within semi-alluvial bedrock environments, where alluvial deposits of limited quantity sit on top of bedrock. When all available particles are entrained and moved away from the control volume, there are none left to entrain, so the entrainment process is turned off. 
Likewise, \citet{Ancey2010, Ancey2014,Ancey2015} extended the birth-death process within a control volume as I reviewed here to an array of spatially distributed control volumes (cells). 
They showed that immigration and emigration processes between cells set up a bedload diffusion, and they connected this theory with earlier sediment diffusion theories by deriving an Exner type equation of sediment diffusion. 
I will now review each of these developments in turn.

Also Heyman 2013 considered that the waiting time distribution need not be exponential. Need more focus on the underlying poisson distributions of these models.

\section{Including sediment availability: Turowski's two species birth death model}

Semi-alluvial channels, where alluvium sits on top of bedrock, will violate one of the key assumptions of the \citet{Ancey2008} framework. 
This model implicitly assumes that the bed surface has an infinite number of particles available for entrainment. 
In reality, this situation will never be satisfied, even in alluvial channels, because there are a wide variety of stabilizing processes acting on bed surfaces \citep{Hassan2008, Venditti2016}.

Turowski was concerned with bedrock scour due to alluvium. 
Therefore he extended the Ancey framework. 
The Ancey framework takes account of one population: the number of moving particles. 
Turowski's extension included a second population into the framework: the number of particles available for entrainment. 
He considered that as a particle entrains into a motion state, the number of particles available for entrainment decreases, and likewise as a particle deposits from the motion state, the number of particles available for entrainment increases. 
Therefore, he considered a probabilisitic population model of two coupled populations: such an inquiry is entirely new in stochastic theory of the bedload flux, although it is commonplace in ecological modeling \citep[e.g.][]{Pielou1977}

Turowski considered a joint probability distribution for the random number $N$ of moving particles and $M$ of particles available for entrainment. 
This can be written $\pi_{n,m}(t)$. 
Within a small time interval $dt$ he considered birth, death, immigration, and emigration transitions governing the probabilities $\pi_{n,m}(t)$, similar to \citet{Ancey2008}. 
Like \citet{Ancey2008}, he also included a collective entrainment contribution, arranging that the model expressed wide Negative Binomial-like fluctuations in the bedload rate.

Turowski's transition probabilities in $dt$ varied by including the number of available particles $m$ into the entrainment and deposition probabilities. He considered a rate of entrainment $(\lambda_0 + \mu n)m$: when the population of particles available for entrainment is zero ($m=0$), the rate of entrainment is zero. The entrainment process, when it occurs, enacts a change of state $(n,m)\rightarrow(n+1,m-1)$: the number of active particles increases by one, while the number of particles available for entrainment decreases by one.

Similarly, Turowski's deposition process was generalized from \citet{Ancey2008}.
Deposition occurs with rate $\sigma n$ in time $dt$, just as in Ancey et al, but when this deposition process occurs it enacts the change $(n,m) \rightarrow (n-1,m+1)$ in the coupled populations. 
The Kolmogorov equation is 
\begin{multline}
P(m,n;t+dt) = P(m,n-1,t)\lambda_0 dt + P(m+1,n-1,t)((m+1)\lambda_1 \\+ (m+1)(n-1)\mu)dt + P(m-1,n+1,t)(n+1)\sigma dt + P(m,n+1,t)(n+1)\nu dt \\+ P(m,n,t)*[1-dt(\lambda_0 + m \lambda_1 + n \nu + n \sigma + mn \mu). \end{multline}
which in the limit $dt \rightarrow 0 $ becomes a Master equation for the joint probabilities $\pi_{n,m}(t)$: 
\begin{multline} 
\frac{d}{dt} P (m,n,t) = \lambda_0 P(m,n-1,t) +\\ ((m+1)\lambda_1 + (m+1)(n-1)\mu)P(m+1,n-1,t) + (n+1)\sigma P(m-1,n+1,t)\\ + (n+1)\nu P(m,n+1,t) - (\lambda_0 + m \lambda_1 + n\nu + n \sigma + mn \mu ) P(m,n,t). 
\end{multline}

Again, as a reminder, the number of active particles is $n$ and the number of particles available for entrainment is $m$: this master equation describes a hirearchy of joint probability distributions for these discrete random varaibles. 
Exact solutions via counting arguments \citep[e.g.][]{Ancey2006} or probability generating functions \citep[e.g.][]{Ancey2008} are evidently no longer tractable. 
\citet{Turowski2009} resorted to a numerical solution.
Numerical algorithms for stochastic birth-death models are relatively easy to implement and are well described in a number of references \citep[e.g.][]{Gillespie1977, Gillespie1992}. 
Turowski pursued one of these algorithms and related his transport rate computations to bedrock erosion and alluvial cover within a bedrock channel.

Now we briefly describe the stochastic simulation algorithm of Gillespie 1977 ... 
Then we reproduce Turowski's results using it ...

We also explore the possibility of an exact solution for the Turowski model.. go harder than he did.

Turowski's master equation reduces to the Ancey et al result as $m \rightarrow \infty$. As $m$ gets very large, $m-1 \approx m \approx m+1$: effectively, large $m$ makes $m$ constant as far as the master equation perceives. 
In this limit the probability $P(m,n,t) \approx P(n,t)$, and since $m$ is effectively constant, one can take $m\lambda_1 \rightarrow \lambda_1'$ and $m\mu \rightarrow \mu'.$ 
With these substitutions the master equation reproduces the \citet{Ancey2008} result. 
For our purposes, given our concern with alluvial channels as outlined in the introduction, the main takeaway of Turowski's work is that the birth-death modelling approach can be applied to study the stochastic dynamics of multiple populations whose variations are correlated. 
This is similar to competing species in ecology \citep{Pielou1977} or to multiple reacting chemical species \citep{Gardiner1983}.

\section{The bedload flux at a point: the counting statistics of emigration events}

Up to this point, the Markovian approaches to understand a fluctuating bedload rate were based upon analyzing the stochastic number of active particles within a control volume. 
The bedload flux definition considered for these studies was somewhat unconventional. 
Typically it is defined as the rate of bedload crossing a vertical plane perpendicular to the flow direction \citep{Ballio2014}.

The studies of \citet{Heyman2013} and \citet{Ma2014b} demonstrated that the birth-death model of \citet{Ancey2008} can also be used to define a flux at a position in space. 
Although the birth-death model of \citet{Ancey2008} counts the number of moving particles at each instant of time, it does not count the number of emigration events which contribute to this number. 
Counting the number of emigration events in an interval of time is a much more difficult problem.
Ma et al (2014) approached this counting problem using a path integral approach: this approach is essentially outlined in \citet{Ohkubo2009}. 
The path integral approach draws heavily from the stochastic simulation algorithm of \citet{Gillespie1977}.

The conventional definition of the bedload flux is an integral of particles through a surface perpendicular to the flow: 
\be Q_s(t) = \int \int_A dA \textbf{u}_s \cdot \textbf{n}, \ee
where $\textbf{n}$ is a unit vector perpendicular to the average flow direction and $\textbf{u}_s$ is the velocity of bed load. 
This instantaneous flux is very difficult to measure, so usually it is averaged over some time interval $\delta t$: 
\be Q_s(t;\delta t) = \frac{1}{\delta t} \int \int_A \int_t^{t+\delta t} \textbf{u}_s \cdot \textbf{n} d\tau dA = \frac{\delta V_s}{\delta t} = v_s \frac{\delta S(t;\delta t)}{\delta t}. \ee
$Q_s(t;\delta t)$ is the volume flux averaged over $\delta t$; $\delta V_s$ is the volume of bed load particles passing through $A$ within $\delta t$; $S(t)$ is the cumulative number of particles passing through the cross section since t=0, and $\delta S(t;dt)$ is the number of particles passing through the surface between $t$ and $t+\delta t$, i.e. $\delta S(t; \delta t) = S(t+\delta t) - S(t)$. 
This equation means the bedload flux at $t$, when viewed in a time-averaged sense, is the number of particles crossing the surface between $t$ and $t+\delta t$. 
Because volume units can always be multiplied in later, the bedload flux is written 
\be q_s(t) = \frac{\delta S(t; \delta t}{\delta t}. \ee

This bedload flux is a random variable.
A fluctuation is defined strictly as $q_s' = q_s - \bra q_s \ket$, and the magnitude of fluctuations is defined without ambiguity by the variance $\bra q_s'^2 \ket$. 
This $\delta S(t;dt)/\delta t$ is the number of emigration events between $t$ and $t+\delta t$, so if this number could be counted, the bedload flux would be known at the downstream boundary of the control volume: the bedload flux would be known at a point in space.

The path integral formulation performs this counting problem. 
For the sake of computation, the time interval t is discretized as $i \Delta t = t$, $i=0,1,2,\dots$.
In the end, the limit $\Delta t \rightarrow 0$ will be taken. 
If the increments $\Delta t$ are sufficiently small, Ma et al argued that emigration events during one of these intervals will be approximately Poissonian so that
\be P[\Delta S_i = n | N_i(t) = N_i] = \exp[-f_3(N_i)\Delta t] \frac{(f_3(N_i) \Delta t)^N}{n!}. \ee
This equation expresses the probability that $n$ particles emigrate between $t$ and $t+\delta t$ is a Poisson distribution with rate parameter $f_3(N_i)\Delta t$, where $f_3(N_i) = \gamma N_i$ is the emigration rate.

Now Ma et al argued that because the entrainment rate $f_1 = \lambda + \mu N \~ O(L)$, the deposition rate $f_2 = \sigma N \~ O(L)$, and the emigration rate $f_3 = \gamma N \~ O(1)$, if the observation window is sufficiently large, entrainment and deposition processes dominate, so that the number of moving particles $N$ is perturbed only weakly by emigration. 
In this limit of sufficiently large observation window size, it then holds that
\be P(\sum \Delta S_i = n | N_1,N_2, \dots) = \exp[-\sum f_3(N_i)\Delta t]\frac{(f_3(N_i)\Delta t)^n}{n!}, \ee
which in the continuum limit becomes a path integral counting emigration events: 
\be P[\delta S = n |N(t): t_0<t<t_0+\delta t] = \exp[-\int f_3(N(\tau))d\tau]\frac{[\int f_3[N(\tau)]d\tau ]^n}{n!}.\ee
Denoting the integral over the process $f_3(N(\tau))$ as $\alpha = \int_{t_0}^{t_0+\delta t} d\tau f_3[N(\tau)]$, the probability for $n$ emigration events in $\delta t$ is written 
\be P(\delta S = n) = \int d\alpha \frac{e^{-\alpha}\alpha^n}{n!}P(\alpha,t). \ee
From this difficult derivation a simple formula emerges: the probability of $n$ emigration events in $\delta t$ is an integral over all possible realizations of the process $N(t)$, where each process is given a weight $e^{-\alpha}\alpha^n/n! P(\alpha,t) d\alpha$. This is called a Cox process. It's a generalized Poisson process where the parameter $\alpha$ is also a random variable. It represents a superposition of a series of Poisson distributions taking every possible path of $N(t)$ into account. 
This is a stochastic description of the bedload transport flux based on the counting statistics of emigration events.

Notably, because of the time integral of $N(t)$ involved in this relationship, it has a memory. It is non-Markovian. 
This character reveals a very interesting fact: the fluctuations of the transport rate depend on the time scale of observation. 
In fact, Ma et al derived the expression for the variance: 
\be \delta t \text{var} [q_s(\delta t)] = \gamma^2 \frac{\lambda (\sigma + \gamma)}{(\sigma + \gamma -\mu)^2} 2 t_c [\delta t - t_c (1-e^{-\delta t/t_c}] + \frac{\gamma\lambda}{\sigma + \gamma -\mu}\delta t \label{eq:3regimes} \ee

where $t_c = 1/(\sigma + \gamma - \mu)$ is the autocorrelation time of $N(t)$, representing memory in the system. 
Apparently, the variance of the bedload flux at a point depends on the observation or averaging time-scale $\delta t$.

In fact, Ma et al discriminated three regimes of behavior from \ref{eq:3regimes}:   
\be
    \delta t \text{var}[q_s(\delta t)] \approx 
\begin{cases}
    \frac{\gamma \lambda}{\sigma + \gamma - \mu}, & 0<\delta t \ll t_l \\
    \delta t \frac{\gamma^2 \lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}   & t_l<\delta t< t_c \\
    2 \gamma^2 \frac{\lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}t_c + \frac{\gamma \lambda}{\sigma + \gamma - \mu}   & t_c \ll \delta t < \infty
\end{cases}
\ee
When this three-regime equation is divided by the mean bedload rate $\text{mean}[q_s(\delta t)]$, the behavior is discriminated by a dimensionless number Ma et al called $Ra$: 
\be Ra(\delta t) \approx \begin{cases} 1 & 0<delta t \ll t_l \\
\frac{delta t}{t_l} & t_l < \delta t < t_c \\ 
\frac{2 t_c}{t_l} + 1 & t_c \ll \delta t < \infty \end{cases} \ee

The essential point is that Ma et al derived three regimes of fluctuations from the \citet{Ancey2008} model. 
Depending on the sampling interval $\delta t$, three different scaling relations are possible for the magnitude of fluctuations in the bedload flux. 
These intervals are discriminated by two timescales $t_c$ and $t_l$. 
There are probably deep consequences of this research which remain to be fully understood. 
At this stage, it's sufficient to note that fluctuations differ with observation timescale, and that this effect can be discriminated into stages with a dimensionless number $Ra$.

The origin of the two timescales still needs to be discussed.

There is another way to get information about the bedload flux at the downstream boundary of the control volume defined by \citet{Ancey2008}, and this was examined by \citet{Heyman2013, Heyman2016}. 
Rather than actually counting the number of emigration events with the complicated path integral approach, Heyman et al. took a somewhat simpler angle: they examined the waiting time between successive emigration events.

\section{The waiting time between emigration events: Bedload flux from waiting time statistics}

These studies are concerned with linking the birth-death formulation, which in the original interpretation deals with transport characteristics within an observation window \citep{Ancey2008}, to the transport characteristics at a point -- the number of particles crossing a plane perpendicular to the flow direction.

\citet{Heyman2013} was concerned with the statistics of the time interval between successive emigration events, while \citet{Ma2014b} focused on actually counting the number of emigration events in a unit time: that is, they computed the bedload flux over the downstream boundary of the control volume \citet{Ancey2008}.

The waiting time between successive transport events is significant in a broad range of transport studies. 
First, there are probabilistic transport rate formulations which use the waiting time distribution as their input \citep{Turowski2010}. 
Second, the waiting time between successive transport events, as we shall show, reflects back on the underlying entrainment and deposition characteristics, so it provides fundamental information into the somewhat mysterious physics of bedload transport.

The original \citet{Einstein1937,Einstein1950} approach reveals that the waiting time between successive transport events should be an exponentially distributed random variable. 
However, Heyman et al (2013) demonstrates that the inclusion of collective entrainment due to \citet{Ancey2008} modifies the waiting time distribution in a non-trivial way. 
The waiting time between successive entrainment events has contributions at two different scales: first, there is a shorter timescale of individual entrainment events, controlled by fluid turbulence and essentially recognized within the \citet{Einstein1937} formalism, and second, there is a longer timescale due to collective entrainment.
These authors showed that within the \citet{Ancey2008} birth-death model, at low transport rates, when entrainment and motion are highly intermittent, the short timescale and the long timescale are well separated. 
While at higher transport rates, these two timescales blend together as individual and collective entrainment contributions become indistinguishable.

I'll now sketch the \citet{Heyman2013} derivation of the statistics of the waiting time between successive emigration events, and show how the two different timescales due to individual and collective entrainment arise. 
For \citet{Heyman2013} the variable of interest is $S_k = \sum_{i=0}^k T-i$: $S_k$ is the time when the $k$th emigration event occurs. $T_1, T_2, \dots $ are the waiting times: the time from $t=0$ to the first emigration event is $T_1$. The time from $T_1$ to the second emigration event is $T_2$, and so on. 
For a Markov process such as the \citet{Ancey2008} model, the waiting times are independant and identically distributed. 
Within the Einstein model we have already seen that the waiting time probability distribution is expoential. 
However, in the \citet{Ancey2008} formalism there is no reason to expect an exponential distribution.

Heyman et al define $F_n(t)$ as the probability that there are $n$ particles in motion and no emigration event occurred in time t: 
\be F_n(t) = Pr(T>t,N(t)=n).\ee
By extension, the probability that there are any number of particles and no emigration event occurred in time $t$ is 
\be F(t) = Pr(T>t) = \sum_{n=0}^\infty F_n(t) .\ee

Now $F(t+\Delta t)$ is equivalent to the probability that any other event but emigration occurs in $\Delta t$. 
Therefore we can write a master equation for $F(t)$: 
\be F_n(t+\Delta t) = F_n(t)[1-(\lambda + n(\sigma+\mu+\gamma))\Delta t] + F_{n+1}(t) \sigma(n+1)\Delta t + F_{n-1}(t)\lambda + \mu(n-1)\Delta t + o(\Delta t).\ee
This equation holds for $n\geq 1$. At $n=0$ deposition and collective entrainment processes are not possible, therefore 
\be F_0(t+\Delta t) = F_0(t) [ 1-\lambda \Delta t] + F_1(t) \sigma \Delta t + o(\Delta t) \ee 
Again dividing by $\Delta t$ and taking $\Delta t \rightarrow 0$ gives the differential difference equations
\begin{align} 
F_0'(t) &= \lambda F_0(t) + \sigma F_1(t) \\
F_n'(t) &= -(\lambda + n(\sigma + \mu + \gamma))F_n(t) + \sigma(n+1)F_{n+1}(t) + [\lambda + \mu(n-1)]F_{n-1}(t) \text{ if } n\geq 1 \end{align} 
for the waiting time distribution.

Summing all of the terms gives the simple equation 
\be \sum_{n=0}^\infty F_n'(t) = -\gamma \sum_{n=0}^\infty nF_n(t), \ee
which, denoting the pdf of waiting times $T$ as $f_T(t)$ gives the simple relationship: 
\be f_T(t) = -F'(t) = \gamma\bra F_n(t)\ket. \ee
Therefore the probability distribution function of waiting times between emigration events, $f_T(t)$, is an average of the probability that there are $n$ particles and no emigration event occurred in $t$ over all $n$.

The general solution of $F_n(t)$ can be obtained from the differential equations with a generating function: 
\be G(z,t) = \sum_{n=0}^\infty F_n(t) z^n. \ee
Multipling the differential equation by $z^n$ and summing over all $n$ gives
\be \frac{\partial G}{\partial t} = (\sigma + \mu z^2 - z(\alpha + \mu) ) \frac{\partial G}{\partial t} + (z-1)\lambda G, \ee
where the shorthand $\alpha = \gamma + \sigma $ has been introduced.

This equation can be solved with the method of characteristics, giving a closed form solution for $G(z,t)$. A useful property of generating functions (and the source of their name) is $\bra F_n \ket = \partial G(z,t) /\partial t |_{z=1}$, therefore the probability distribution of waiting times between emigration events in the \citet{Ancey2008} birth-death model of bedload flux is 
\be f_T(t) = \gamma(z_1-z_2)^{\lambda/\mu} \big( \frac{\alpha-\mu}{A(t)-B(t)} \big)^{\lambda/\mu + 1} e^{-\lambda (1-z_2)t}\big( \frac{\lambda/\mu+1)B(t)}{A(t)-B(t)} [(1-z_2)e^{-(z_1-z_2)t}+z_1-1]+e^{-\mu(z_1-z_2)t}-1 \big)  \ee 
When collective entrainment is turned off ($\mu=0$), the probability distribution becomes $f_T(t) = \frac{\alpha}{\lambda \gamma} e^{-\alpha t/(\lambda \gamma)}$: the exponential waiting time between emigration events is recovered.

This probability distribution for waiting times between emigration events in the \citet{Ancey2008} model has, in general, contributions from two timescales. The relatively common individual entrainment process imparts a 'fast' timescale to the probability distribution of emigration waiting time, while the relatively rare collective entrainment process imparts a 'slower' timescale. 
When collective entrainment is turned off, only the fast timescale is present in the pdf. When the entrainment rate is relatively low, the fast timescale is much shorter than the slow timescale, so that the timescales are well separated and the waiting time distribution is not well approximated by any common distribution in the literature (Gamma, exponential): the behavior of the \citet{Ancey2008} model at low transport rates has defied prediction by any other framework.

The most significant point of the \citet{Heyman2013} paper is that the probability distribution function of waiting times between emigration events can be used to calibrate the \citet{Ancey2008} birth death model to flume experiments. 
This is because short emigration times control $\alpha-\mu$ and long emigration times control $\lambda$. 
The evaluation of waiting times between emigration events supports a calibration of the stochastic birth-death model of \citet{Ancey2008} on flume studies where the transport rate is measured at the outlet of the flume, but the erosion, deposition, or emigration rates are not necessarily measured. 
In contrast to \citet{Ma2014b} the mathematics are not much beyond the original formulation of the \citet{Ancey2008} model, so this work may be more readily accepted by the research community.

\section{The diffusion of bedload from a microstructural basis}

\citet{Ancey2008} developed a generalized version of \citet{Einstein1950} which took account of a new process to describe the large fluctuations seen in experimental bedload flux signals: collective entrainment. 
\citet{Turowski2009} generalized this model to finite sediment availability, while \citet{Heyman2013} and \citet{Ma2014} used the \citet{Ancey2008} model in its original form to understand the distribution of waiting times between emigration events and the bedload flux probability distribution at a point in space, respectively. 
These two works developed new understanding of spatial and temporal aspects of bedload flux: \citet{Heyman2013} demonstrated that the waiting time between successive emigration events expressed contributions from slow and fast timescales related to individual and collective entrainment, respectively. 
\citet{Ma2014} highlighted the role of observation timescale on the magnitude of flutuations, and discriminated three distinct scaling regimes for bedload fluctuations which are contingent on a dimensionless number $Ra(\delta t)$, where $\delta t$ is the timescale of observation.

However, up to this point in early 2014, stochastic birth-death models had focused on a single region of space -- a control volume \citep{Einstein1950, Ancey2006, Ancey2008}, or a single point in space -- the downstream end of the control volume, or the plane across which emigration events happen \citep{Heyman2013, Ma2014b}.
This approach is obviously oversimplified. Variability of sediment transport rates is almost a defining feature of them \citep{Hassan2008, Venditti2016, Nelson2014}. 
There is a need to take account of spatial differences in entrainment and deposition characteristics in calculating bedload fluxes. 
Additionally, there are a wide set of issues centered around bedload diffusion-- or the spreading of bedload particles as they are tracked through time and subjected to random entrainment and deposition events \citep{Hassan2016}. Indeed, this study of bedload diffusion was the original research directive of \citet{Einstein1937}.

Often, bedload diffusion has been understood through deterministic partial differential equation models. These advection diffusion equations have been derived by considering conservation of mass \citep{}, but they were not very well connected to the underlying stochastic dynamics of bedload transport. 
Generating this connection between a microstructural stochastic model like \citet{Ancey2008} and the continuum diffusion equation treatment of bedload diffusion \citep[e.g.][]{Parker2002} was the directive of the next papers on birth-death modeling \citep{Ancey2014,Ancey2015}. 
These papers took a conceptually straightforward extension of the \citet{Ancey2008} model which leads to difficult mathematics.

Rather than considering a single control volume, as in \citet{Ancey2008}, \citet{Ancey2014,Ancey2015} considered an infinite array of adjacent control volumes (cells), indexed by $i=1,2,\dots,M$. 
Bedload particles can entrain and deposit within each cell, and they can also migrate between cells. 
Each of these transitions are characterized by probabilities per unit time in generalization of \citet{Ancey2008} across spatial extent. 
The $i$th cell has a random number of particles $N_i$ in motion within it. 
Therefore the state of the system at an instant of time is fully characterized by the set of these numbers: $\textbf{n} = (n_1(t), n_2(t), \dots n_M(t))$.

One target is the grand probability distribution of the number of active particles within each cell at an arbitrary time: $P(\textbf{n};t)$. 
Another target is the continuum limit: letting the length $\Delta x$ of each cell shrink to zero develops an advection-diffusion equation for bedload diffusion. 
Ancey et al termed this a "Stochastic interpretation of bedload diffusion": and the latter target has much more fundamental scope.
It explores the link between microscale stochastic models of bedload transport and mesoscale deterministic models based upon advection-diffusion equations.

Now I'll sketch the \citet{Ancey2014, Ancey2015} derivation of a birth-death-migration model across an array of cells. 
There are essentially three transitions to take account of: 
\begin{enumerate}
\item Entrainment can occur within each cell ($i$) at probability per unit time $(\lambda_i + \mu_i n_i ) \delta t$
\item Deposition can occur within each cell ($i$) at probability per unit time $ \sigma_i n_i \delta t$
\item Migration can occur from cell $i$ to cell $i+1$ (downstream) at probability per unit time $\nu_i$
\end{enumerate} 
Now some notation is introduced to write the effect of these transitions on $\textbf{n}$ in shorthand. 
If the change of state due to one of these transitions is written as $\Delta \textbf{n}$, then the effect of these transitions can be characterized using vectors $\textbf{r}_i^j$ and $\textbf{r}_i^\pm$. The vectors $ \textbf{r}_i^j$ are of the same dimension as $\textbf{n}$ and all but two of their entries are zero: $r_i = 1$, $r_j = -1$, $r_k=0$ for $k\neq i,j$. The vectors $\textbf{r}_i^\pm$ are the same dimension as $\textbf{n}$ and all but their $i$th entry is zero: $r_i = \pm 1$, $r_j = 0$, $j\neq i$.

Thus the transition probabilities of each process can be written: 
\begin{enumerate}
\item $ p_i^3 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^+; t+ \delta t) = (\lambda_i + \mu_i N_i) \delta t $ -- entrainment
\item $ p_i^2 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^-; t+\delta t) = \sigma_i N_i \delta t $ -- deposition
\item $ p_i^1 = \text{Prob}(\textbf{N} = \textbf{n} \textbf{n} + \textbf{r}_i^{j-1}; t+ \delta t) = \nu_{i-1}N_{i-1} \delta t $ -- migration
\end{enumerate}

These transition probabilities develop a master equation analogous to the one from \citet{Ancey2008}, but generalized to a collection of $M$ cells: 
\begin{multline} \frac{\partial}{\partial t} P(\textbf{n};t) = \sum_{i=1}^M (n_i+1)[P(\textbf{n}+\textbf{r}_{i+1}^j,t) \nu_i + P(\textbf{n}+\textbf{r}_i^+,t)\sigma_i ]\\
+ P(\textbf{n}+\textbf{r}_i^-,t)(\lambda_i + \mu_i(n_i-1))\\
+ P(\textbf{n}+\textbf{r}_i^{j-1},t)\nu_{i-1}n_{i-1}\\
- P(\textbf{n},t)(\nu_{i-1} n_{i-1}+\lambda_i + \mu_i n_{i+1} + \nu_i n_i + \sigma_i n_i ) \end{multline} 
Information about solving this equation is sparse in the literature. 
The closest approach has been in ecology: these birth/death/migration-type stochastic models have been considered in context of ecological population dynamics since at least the sixties \citep{Bailey1968}. 
One exact solution of a very similar stochastic model has been attained using path integral approaches borrowed from quantum field theory \citep{Field2010}.
By any metric, the analytical barrier this multidimensional master equation presents is not trivial, although definitely numerical solutions are possible \citep[e.g.][]{Gillespie1992}.

In order to link this equation to spatial diffusion, \citet{Ancey2014, Ancey2015} resorted to a Poisson space representation \citep{Gardiner1983}. 
In effect, the Poisson transformation converts the discrete random variable $\textbf{N}$ to a continuous random variable $\textbf{a}$ (the Poisson rate). 
The Poisson transform is 
\be P(\textbf{n},t) = \prod_i\int \frac{e^{-a_i}a_i^n}{n!}f(\textbf{a},t)d\textbf{a}, \ee
where $\textbf{a} = (a_i) \geq 0$ for $i=1,2,\dots$ and $f(\textbf{a},t)$ is the multivariable probability density of the continuous vector $\textbf{a}$.

\citet{Ancey2014, Ancey2015} used this transformation to obtain a Langevin equation representation of the dynamics of the random variable $\textbf{a}$.
That is, they obtained a stochastic differential equation representation of $\textbf{a}$: 
\be da_i(t) = (\lambda_i - a_i(\sigma_i-\mu_i) + \nu_{i-1} a_{i-1} - \nu_i a_i)dt + \sqrt{2 \mu_i a_i}dW_i(t),\ee
and the $W_i(t)$ is a white noise term (Wiener Process) on cell $i$. 
The mean steady state solution of this Langevin equation is 
\be \bra a \ket_ss = \frac{\lambda}{\sigma-\mu} \ee 
for all $i$.

Now \citet{Ancey2014,Ancey2015} sought out to find an advection diffusion equation as the limit of cell size $\Delta x \rightarrow 0$, so they went on to impose this limit on the Langevin equation, thereby obtained an advection diffusion equation. 
Unfortunately, the limit is more subtle than expected. 
This did not really work for reasons I need to understand better, and it did not work in ways I need to learn to explain better. 
More study is needed on spatially varying birth-death models, from me, and from others especially.

\section{Spatial correlations in bedload transport: }

\citet{Heyman2014} used the \citet{Ancey2014} model to examine a set of experiments resolving spatial correlations of bedload transport. 
They found a stochastic differential equation for the mean of the poisson rate $\eta$: 
They related this to a point process framework which allowed them to analyze spatial correlations in the bedload rate and relate this to their birth-death model. 
Somehow this whole thing is contingent on their two-point spatial correlation process. 
They state that the whole problem is linear so that it'd be straightforward to extend it to 2d. 
slowing down ...

\section{recap} 
Einstein developed first stochastic theories describing the diffusion and flux of bedload. 
Ancey 2006, building on earlier work from Lisle 1998 and the many investigators who criticized and revised Einstein, formalized Einstein's assumptions over a foundation in Markov process theory, and extended his work by treating the bedload flux within a control volume as the cooperation of many independent two-state Markov processes: each bed particle within the control volume makes random transitions between motion and rest. 
Thus he derived a statistical distribution of the bedload rate which reproduces Einstein's result when the mean is taken. 
However, the magnitude of fluctuations predicted by this model are too small.

Ancey 2008 extended the Einstein theory by linking the transition rates between particles within the observation window with a collective entrainment term: they prescribed that the transition rate from rest to motion depends on the number of active particles. 
This is a simple way to include the effects of coherent turbulence and impact-based entrainment: it is an experimental observation that moving particles tend to come in waves \citep{Drake1988}. 
The Ancey 2008 model was demonstrated capable of describing fluctuations, and all of its parameters have physical meanings, but notably there is no clear suggestion as to how collective and individual entrainment processes can be separated within experiments. 
Clearly, approaches to derive entrainment probabilities are focused on the individual entrainment rate \citep{Dey2018}, and there have been no developments with regard to computing the collective entrainment rate: there is no clear physical model of collective entrainment yet.

Turowski 2009 extended the Ancey work to take account of limited supply, in order to describe semi-alluvial channels where alluvial deposits lie on top of bedrock. 
He continued to work within the Einstein paradigm by describing the transport rate within a control volume, rather than by counting the number of particles leaving the control volume: although the Ancey 2008 work essentially outlines this possibiilty. The transport rate at a point should be considered as the number of emigration events in a unit time.

Heyman et al 2013 and Ma et al 2014 went on to consider the statistics of emigration: they tried to discern what could be learned about the transport rate at a fixed point from the control volume formalism of Ancey et al and Einstein before that. 
Heyman et al was concerned with the statistics of the waiting time between successive emigration events. 
The original Einstein (1937, 1950) assumptions generate an exponential waiting time ditsribution with a timescale related to the entrainment rate.  
However, Heyman et al (2013) showed that when collective entrainment effects were included there are two timescales: one fast timescale related to the individual entrainment rate, and one slower timescale related to collective entrainment effects. 
Thus, the distribution of waiting times between successive emigration events, which is a useful concept for alternative stochastic models of the bedload flux distribution \citep[e.g.][]{Turowski2010}, is a more subtle object if collective entrainment occurs. 
These effects are accentuated at low transport rates, where the fast and slow timescales are more disparate. 
At high rates, the timescales become comparable and the waiting time between successive emigration events blends into an Einstein-like exponential.

All of these approaches considered bedload transport within a control volume or at a point, but indeed, a very large set of contemporary river science studies are concerned with the diffusion or spreading of bed material through a downstream reach of river \citep{Hassan2016}. 
Bedload diffusion is not captured by a model at a single location: often, it has been described using an advection diffusion equation. 
These advection diffusion equations follow from considerations of mass balance within river channels, but they were not derived from any underlying microstructural model until \citep{Ancey2014}. 
\citet{Ancey2014} extended the previous control volume model \citep{Ancey2008} to an array of control volumes, where emigration from the $i$th volume is immigration to the $i+1$th volume.

In an approach widely used in chemical physics \citep{Gardiner1983} and ecology \citep{Bailey1968}, \citet{Ancey2015, Ancey2015} derived an advection diffusion equation from their microstructural model in a limit as the size of each control volume goes to zero while the number becomes infinite. 
Thus, in their interpretation, bedload diffusion emerges in the continuum limit of coupled birth-death immigration-emigration models. 
They rederived the famous Exner equation of bedload diffusion.

Taken together, this research exhibits a coherent progression from the original work of Einstein to a more mathematically and physically based set of more general approaches. 
Within these approaches the bedload flux is understood as a random quantity. 
The mean bedload flux and the magnitude of its fluctuations are modelled in an unambiguous way \citep{Ancey2006, Ancey2008} considering no limitations in sediment availability, and it was shown that collective effects are a necessary inclusion to properly describe fluctuations in bedload transport \citep{Ancey2008}. 
An extension to a more realistic situation of finite supply was developed \citep{Turowski2009}, and the relationships between these control volume based models and the definition of bedload flux as particles crossing a plane perpendicular to the flow were developed \citep{Heyman2013, Ma2014b, Ballio2014}.

These approaches did not address spatial correlations or variations in bedload fluxes, although the problem of bedload diffusing through a reach of channel is of contemporary significance \citep{Hassan2016}. 
This extension was the most recent development in birth-death modelling of bedload \citep{Ancey2014, Ancey2015}. 
By extending the \citet{Ancey2008} model to an array of adjacent control volumes, an advection-diffusion equation describing the concentration of bedload in motion was developed, providing the first connection between a microscale stochastic model of bedload transport with a macroscopic advection-diffusion formulation. 
\citet{Heyman2014} examined the spatial correlations between local bedload fluxes expressed by the \citet{Ancey2014} model. 
This is the state of the art of bedload flux models.

\section{Assumptions, calibraiton problems, and unclear aspects: A criticism of the birth-death approach}

Now that earlier literature is evaluated the scope for generalization can be outlined. 
This scope is a necessarily subjective consideration, and I'd like to highlight that nobody knows what the future of stocahstic modeling will hold. 
First, the various glaring assumptions of the birth-death approaches are highlighted. 
\begin{enumerate}
\item Models only consider one particle size, while natural streams exhibit wide distributions of sizes and bedload transport expresses a wide set of effects related to particle size segregation \citep{Wilcock2003, Parker1982, Chen2008}. 
\item There is no clear means to discriminate collective and individual entrainment processes within experiments. 
\item There is only a one-way feedback in these models: bedload is considered subordinate to the fluid flow, and the effect of bedload transport back onto the fluid flow is considered negligible. In fact, there are measurable influences of bedload transport on the fluid phase \citep{}; it's not yet clear whether a one-way coupled scheme such as this can actually describe natural streams across a realistic range of conditions. 
\item Collective entrainment is somewhat of a catch-all term with the physical mechanisms which may contribute to it unresolved. These may include collective entrianment, the formation and disintegration of particle clusters, interactions between the motion of grains of idfferent size, turbulent structure, local avalanche behavior, and bed form migration. 
\end{enumerate}

\section{Summary: scope for extension}

From the basis provided in this literature I'll now make a few small extended birth-death models.

1. diffusion in 2d (horizontal and vertical)
2. availability of depositional sites on the bed surface
3. Extension to multiple grains sizes: A formal outline

### 2021-09-03 draft/formerPapers/ens-rev/reviewsections/ancey2008_0.tex
added phd thesis.

\subsection{Describing bedload fluctuations: Including correlations between particle motions} 
\label{sec:ancey2008}

\citet{Ancey2006a, Ancey2008} developed a more general birth-death model in order to describe the relatively heavy tailed bedload probability distributions seen in experiments. 
In the last ten years, this model has been restated and more deeply studied by several authors \citep{Turowski2010, Heyman2013, Heyman2014a, Heyman2014, Ma2014}, and has led into exciting generalizations \citep{Turowski2010, Ancey2014, Ancey2015}. 
These are the real focus of this review.

Instead of considering the motion of every bedload particle as an independent process involving transitions between motion and rest states, and summing a collection of these to get the total number of moving particles within the control volume, \citet{Ancey2008} consider the total number of moving particles within the control volume as a state.
In this case, instead of just two states (motion and rest), there is an infinite ladder of states ($n = 0,1,2 \dots$), each representing the number of active particles within the control volume.
The ladder of states is illustrated diagramatically in figure \ref{fig:ladder}.

The advantage of this approach is when all particles are treated simulatenously, the transition rates can be made to depend on the number of active particles. 
The state is a collective property of the moving particles within the control volume. 
This allows interactions between particles to be introduced. 
\citet{Ancey2008} considers the population of moving particles changes due to entrainment (birth) and deposition (death), as before, but they also introduce migration processes for particles to move into (emigration) and out of (immigration) the control volume.
We note that these migration processes do not fundamentally modify the structure of the equations which result.
However, they are physically expected, so their inclusion is rational.

\begin{wrapfigure}{r}{0.5\textwidth}
  \includegraphics[width=.98\linewidth]{./figures/ancey2008.png}
  \caption{Birth, death, immigration, and emigration rates characterize the flow of probability up and down an infinite ladder of states. Each state represents $n$ particles in motion, and $n$ runs across all positive integers including zero. \label{fig:ladder} }
\end{wrapfigure}

The key result of \citet{Ancey2008} is a successful description of wide bedload fluctuations. 
To obtain this, they made a component of the entrainment rate increase with the number of active particles. 
In effect, they introduced a positive feedback on entrainment: when more particles are in motion, entrainment becomes more likely. 
They termed this feedback term "collective entrainment".
The collective entrainment term fundamentally modifies the probability distribution of the bedload flux and lends it a wide tail, so the model is capable of describing realistically large fluctuations. 
Rather than a binomial distribution of the number of moving particles, as in \ref{eq:anceybinomial}, \citet{Ancey2008} derive a negative binomial distribution.
Within their model, collective entrainment is the source of the difference between these two distributions. 
When it is turned off, \citet{Ancey2008} reduces to \citet{Ancey2006}.

Collective entrainment has been attributed to several physical mechanisms. 
Usually, the feedback is considered due to collisions of moving particles with the bed \citep{Ancey2008, Heyman2013, Heyman2014a} and the advection of turbulent structures implying waves of entrainment \citep{Ancey2008, Heyman2014}. 
Another consideration is that collective entrainment is the result of small avalanches, or local rearrangements of the bed surface triggered by the entrainment of an individual particle \citep{Heyman2014, Heyman2014a}. 
Entraining one grain could destabilize all of those contingent on it, so local microstructure \citep{Staron2006} may imply collective entrainment.  
These mechanisms for the collective entrainment term are reasonable, but they only hypotheses at this stage, and research is needed.

The layout of this section is as follows: 
First, I will follow \citet{Ancey2008} to develop a master equation for the number $n$ of active particles within the control volume, obtaining a probability distribution $P(n)$. 
Then we can analyze the mean and variance of $n$ and develop the linkage to the probability distribution of the bedload flux, $P(q_s)$ using the relationship between volume and surface definitions of the flux, equation \ref{eq:fluxy}. 
This flux will be shown to link back to \citet{Einstein1950} as a limiting case. 
The magnitude of fluctuations will be considered, and we will conclude by a discussion of the model's assumptions and successes. 
In subsequent sections, we will examine generalizations of and other ways to study this model.

\subsubsection{Master equation of the collective entrainment model}

Consider the probability that there are $n$ particles in motion at time $t$. 
This can be denoted $P(n,t)$. 
The random variable $n$ takes on values $0,1,2,\dots$ -- non-negative integers of arbitrary magnitude. 
Each of these states has an associated probability $p_n(t)$.  
The population is subject to four transitions within a small time increment $\delta t$: 
\begin{enumerate}
\item 
Entrainment of a single bed particle (birth) can happen in time $\delta t$ with probability $\lambda_0 + \mu n$. 
The second term $\mu n$ is the collective entrainment effect. 
As the number of active particles increases, so does the rate of entrainment. 
This interaction term induces realistically wide bedload fluctuations. 
\item 
Deposition of a single bed particle can happen in $\delta t$ with probability $\sigma n$. 
This is similar to the telegrapher's process considered earlier. The rate of deposition increases with the number of active particles, just as it does implicitly in \citet{Ancey2006}. 
The deposition of each particle is independent of every other.  
\item 
Immigration of a particle from upstream into the control volume can happen in $\delta t$ with probability $\lambda_1$. 
This term does not fundamentally modify the structure of the equations we will derive for the probability distribution of the bedload rate. 
\item 
Emigration of a particle from the control volume to downstream can happen with probability $\nu$. 
This term is significant because it defines a bedload flux across a surface as in $q_s = \int dS \textbf{k} \cdot \textbf{u}_s$. 
Counting emigration events provides an alternate definition of the flux. 
This counting problem was solved by \citet{Ma2014}. 
We will consider this later. 
In this section, we will continue to interpret the flux as in section \ref{sec:fluxdef}, by converting an ensemble average to a volume average.
\end{enumerate} 
These transitions are depicted in figure \ref{fig:anceywindow}. 
The symbols of \ref{sec:ancey2008} are summarized in table \ref{tab:symbols2008}.

\begin{wrapfigure}{l}{0.5\textwidth}
  \includegraphics[width=.98\linewidth]{./figures/controlvol.png}
  \caption{As before, the model is based on analyzing the number of particles within a control volume $V$ which has downstream length $L$. Particles appear into the control volume due to entrainment and emigration. They disappear due to deposition and immigration. The downstream boundary is a stream cross section $S$.  \label{fig:anceywindow} }
\end{wrapfigure}

We can consider the flow of probability up and down the ladder of states ($n=0,1,2,\dots$) which is induced by these transitions, entrainment (including the collective entrainment term), deposition, immigration, and emigration, using the formalism of more general Markov birth-death processes \citep{Cox1965, Pielou1977}. 
By the standard way of developing master equations \citep[e.g.][]{Cox1965}, we find an infinite system of equations for the probabilities of finding $n$ moving particles within the control volume: 
\begin{multline} P(n,t+\delta t) = \alpha \delta t (n+1)P(n+1,t) \\+  [\lambda + (n-1)\mu]\delta t P(n-1,t) \\ + [1-\delta t[\lambda + n\alpha + n\mu]]P(n,t). \end{multline}
Here $\alpha = \nu + \sigma$ summarizes the contributions of emigration and deposition, which both act to decrease the number of particles $n$, and $\lambda = \lambda_0 + \lambda_1$ summarizes the contributions of immigration and (individual) entrainment, which both act to increase the number of particles $n$.

As $\delta t\rightarrow 0$ these equations become a hirearchy of differential-difference equations for the probabilities $P(n,t)$ of finding $n$ particles in the control volume at time $t$: 
\be \frac{d}{dt}P(n,t) = (n+1) \alpha P(n+1,t) + [\lambda + (n-1)\mu]P(n-1,t) - [\lambda +n (\alpha+\mu)]P(n,t). \label{eq:anc2008} \ee

These master equations define a birth-death immigration emigration model \citep[e.g.][]{Cox1965, Gardiner1983}. 
It can be solved for the probabilities $P(n,t)$ by introducing the probability generating function $G(z,t) = \sum_{n=0}^\infty z^n P(n,t)$ \citep{Cox1965, Gardiner1983, Ancey2008}. 
Multiplying \ref{eq:anc2008} by $z^n$ and summing over all $n$ gives, after a careful manipulation of the sums in order to form some function of $G(z,t)$ in every term, 
\be \frac{\partial}{\partial t} G(z,t) = \lambda(z-1)G(z,t) + \{ \sigma + \mu z^2 + \nu - (\mu + \sigma + \nu)z\} \frac{\partial}{\partial z} G(z,t). \ee

In effect, the probability generating function exchanges the discrete variable $n$ for a continous variable $z$. 
This first order partial differential equation for $G$ can be solved by the method of characteristics \citep{Cox1965, Garabedian1964}. 
If there are initially $N_0$ moving particles within the control volume, the solution is 
\be G(z,t) + \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \label{eq:generator}\ee
The factor $K$ is the autocorrelation function $K(t) = \exp(-t(\alpha-\mu)).$

A useful feature of the probability generating function is that it generates the hirearchy of probabilities $P(n,t)$ via the Taylor expansion around $z=0$: $G(z,t) = \sum_{n=0}^\infty \frac{z^n}{n!}\big(\frac{\partial}{\partial z}\big)^n G(z,t) |_{z=0}$ \citep{Cox1965}. Comparing this formula with the definition of $G$, $G(z,t) = \sum_{n=0}^\infty z^n P(n,t)$, the coefficient of $z^n$ in the power series expansion of $G(z,t)$ is  apparently $P(n,t)$. 
The generating function provides the probability of finding $n$ particles in the control volume.

Sufficiently far from the intial time, when $K(t) \approx 0 $ in the expression of $G$, meaning $t$ is much larger than the autocorrelation time $(\alpha-\mu)^{-1}$ so the system has forgotten its initial condition, the power series expansion of $G(z,t) \rightarrow G(z)$ generates a set of stationary probabilities for the number of bedload particles in motion within the control volume: 
\be P(n) = \frac{\Gamma(r+n)}{\Gamma(r)\Gamma(1+n)} p^r (1-p)^n\text{, }n=0,1,\dots. \label{eq:negbin0}\ee
Here $r=\lambda/\mu$ and $p = 1-\mu/\alpha$ characterize the strength of the collective entrainment factor $\mu$ relative to the other transitions which act to increase ($\lambda$) or decrease ($\alpha$) the population.  
$\Gamma(x) = \int_0^\infty z ^{x-1} e^{-z} dz$ is the well-known $\Gamma$ function of mathematical physics \citep{Boas2005, Mathews1971}. It is a generalization of $x!$ to non-integer values of $x$. 
This stationary distribution only exists when $\alpha>\mu$: the rate at which sediment particles appear in the control volume is lower than their disappearance rate. 
The set of symbols introduced in the \citet{Ancey2008} formulation is summarized in table \ref{tab:symbols2008}.

\begin{wraptable}{l}{0.5\linewidth}
\vspace{-10pt}
\caption{The parameters used in section \ref{sec:ancey2008}}\label{tab:symbols2008}
\begin{tabular}{p{3.3cm}p{3.3cm}}\\
\toprule  
Symbol & Meaning \\
\midrule
$\nu$ & emigration  \\  
$\sigma$ & deposition \\  
$\lambda_0$ & immigration  \\ 
$\mu$ & collective entrainment\\
$\lambda_1$ & entrainment\\ 
$\lambda$ &  $\lambda_0 + \lambda_1$ \\ 
$\alpha $& $\sigma+\nu$ \\
$r$ & $\lambda/\mu$ \\
$p$ & $1-\mu/\alpha$ \\
\bottomrule
\end{tabular}
\end{wraptable}

This stationary distribution equation \ref{eq:negbin0} is the negative binomial distribution: 
\be P(n) \sim \text{NegBin}(r,p).\label{eq:negbin}\ee
It is a generalization of the binomial distribution for the number of active particles, equation \ref{eq:anceybinomial}, which was derived in the \citet{Ancey2006} paper.
The negative binomial distribution has a relatively heavy tail, so that large deviations in the number of active particles $n$ are possible. 
These large fluctuations are the result of the collective entrainment rate $\mu$, and they allow the birth-death immigration-emigration model to express bedload fluctuations of realistic magnitude. 
As we will show, they become relatively small as collective entrainment is turned off, $\mu \rightarrow 0$.

\subsubsection{The bedload flux: wide fluctuations and the Einstein limit}

As discussed, the negative binomial distribution supports wide fluctuations, controlled by the collective entrainment parameter $\mu$. 
The mean number of particles in the control volume, taken over the distribution \ref{eq:negbin}, is 
\be \bra n \ket = \sum_{n=0}^\infty n P(n) = \frac{\lambda}{\alpha-\mu}. \label{eq:anceymean}\ee
Likewise the variance is 
\be \bra (\delta n)^2 \ket = \sum_{n=0}^\infty (n-\bra n\ket)^2 P(n) = \frac{ \lambda \alpha}{\alpha-\mu^2},\ee
which means their ratio is 
\be \frac{\bra (\delta n)^2}{\bra n\ket} = \frac{1}{1-\mu/\alpha}.\ee
It is interesting to compare this to equation \ref{eq:2006flucts} bounding the magnitude of fluctuations in the \citet{Ancey2006} model. 
In contrast, within the \citet{Ancey2008} model, the variance must exceed the mean, because as discussed, in steady state $\alpha>\mu$. 
The magnitude of fluctuations can grow arbitrarily with the collective entrainment parameter $\mu$. 
This is a coherent conclusion, since collective entrainment was introduced in order to represent the collective effects of turbulent fluctuations, collision-induced entrainment, and granular avalanches; and with the expressed intent of generating realistically wide bedload fluctuations, exceeding those of the \citet{Ancey2006} model.

Armed with the probability distribution for the number of particles in the control volume, equation \ref{eq:negbin}, and the link between control volume and surface statistics derived in section \ref{sec:fluxdef}, equation \ref{eq:fluxy}, we can derive the probability distribution of the bedload flux in a similar way as equation \ref{eq:anceygauss}.
The number of particles in the control volume has been implicitly assumed to be infinite, so there is no need or means to take a large $N$ limit.

Using the law for transformation of probabilities gives 
\be P(q_s) \approx \frac{\Gamma(r+\frac{Lw}{\nu_p u_s} q_s)}{\Gamma(r)\Gamma(1+\frac{Lw}{\nu_p u_s} q_s)}p^r(1-p)^{\frac{Lw}{\nu_p u_s} q_s}\ee
as the probability distribution of the bedload flux $q_s$.

Now we link back to the \citet{Einstein1950} theory. 
Einstein considered neither immigration ($\lambda_1$), emigration ($\nu$), nor collective entrainment ($\mu$). 
When all of these parameters are set to zero, $\mu = \lambda_1 = \nu = 0$, the parameters $r$ and $p$ of the distribution \ref{eq:negbin0} become $r \rightarrow \infty$ and $p=0$. 
In this limit the distribution \ref{eq:negbin0} tends to a Poisson distribution.

There are two ways to see this. 
The first way is to take the appropriate limits ($t \rightarrow \infty$, $\mu=lambda_1=\nu= 0$)  into the generating function \ref{eq:generator} gives 
\be G(z) = e^{-\lambda_0(z-1)/\sigma},\ee
which inverts to 
\be P(n) = \frac{(\lambda_0/\sigma)^n}{n!}e^{-\lambda_0/\sigma},\label{eq:anceypoisson}\ee
the Poisson distribution.  
The second way is to take limits of the negative binomial distribution equation \ref{eq:negbin0} directly, although this is somewhat tricky.
It involves reparameterizing the distribution by its mean value equation \ref{eq:anceymean} rather than $r$, and then taking limits, using the limit formula $e^{-x} = \lim_{s \rightarrow \infty} (1-x/s)^s$.

In either case, arriving at this Poisson distribution for $n$ in the absence of collective entrainment, we can compute the mean bedload flux using equation \ref{eq:fluxy}: 
\be \bra q_s \ket = \frac{\nu_p u_s}{L w} \sum_{n=0}^\infty n \frac{(\lambda_0/\sigma)^n}{n!}e^{-\lambda_0/\sigma} = \frac{\nu_p u_s}{L w} \frac{\lambda_0}{\sigma}. \ee
To make full contact with the Einstein-like formula of Yalin, equation \ref{eq:yalinflux}, one further step is needed.

In the \citet{Ancey2008} collective entrainment model, there was no limitation on the number of particles within the control volume. 
In the notation of section \ref{sec:ancey2006}, this is equivalent to $N \rightarrow \infty$. 
In \citet{Ancey2006} the fraction of time spent in motion, $\xi$, referred to a single particle. 
We can show in the model under consideration that $\lambda_0/\sigma$ also represents a fraction of time spent in motion, although this is more difficult than it was in the arguments leading to the analogous conclusion for $\xi$.

Similar to equation \ref{eq:}, one can set up a master equation to compute the distribution of time time periods that the number $n$ remains fixed while no transitions except deposition occur.

 that particles spend in motion and to find the mean period spent in motion is $\lambda_0$.

- this model assumes infinite sediment availability. there is no limit to n. This leads to turowski
- the flux is defined in a control volume, emigration provides a pathway to defining the flux at a point. This is Heyman 2013, and Ma 2014. 
- imagine connecting cells, this is ancey2014, which leads into Heyman 2014

\begin{comment}

When the collective entrainment process is turned off, $\mu\rightarrow 0$, the Poisson distribution for the number of moving particles within the control volume is recovered, reproducing the limiting result of \citet{Ancey2006}, which has links to \citet{Einstein1950}. 
We will consider this limit of no collective entrainment. 
$\mu\rightarrow 0$, implies $r \rightarrow \infty$ and $p \rightarrow 1$. 
In this limit, the generating function G(z) becomes 
\be G(z) = \exp(-\frac{\lambda}{\alpha}[z-1]).\ee
Taylor expanding this generating function around $z=0$ generates the hirearchy of probabilities $\pi_n(t)$ in the limit of no collective entrainment:
\be \pi_n = \frac{(\lambda/\alpha)^n}{n!}e^{-\lambda/\alpha}.\ee
Therefore the number of particles in the control volume is Poisson distributed, in accord with the large $L$ limit of the \citet{Ancey2006} model in the previous section \ref{eq:poissonflux}. 
To complete the connection we turn off immigration ($\lambda_0=0$) and emigration ($\nu = 0$) as well, so that the distribution of the number of particles in the control volume reproduces the \citet{Ancey2006} result cleanly: 
\be \pi_n = \frac{(\lambda_1/\sigma)^n}{n!} e^{-\lambda_1/\sigma},\ee
which is the same as equation \ref{eq:poissonflux}. 
We note that immigration ($\lambda_0$ -- to move in) and emigration ($\nu$ -- to move out) do not fundamentally alter the form of the distribution.
Only collective entrainment ($\mu$) does this.

Now we will analyze the probability distribution of bedload flux and the magnitude of its fluctuations. 
Collective entrainment develops a negative binomial distribution for the number of active particles, and in the limit of a large control volume ($L \rightarrow \infty$), the probability distribution for the number of active particles links to the probability distribution of the bedload flux via the relationship \ref{eq:fluxy} of the previous section. 
As a reminder, this relationship was derived using an artificial device. 
To develop an ensemble of particles intersecting a flow cross section $S$ with various positions $\textbf{x}$ and velocities $\textbf{u}_p$, we traded a frequentist interpretation of a conditional probability $P[\textbf{u}_p|\textbf{x}]$ for an ensemble one.
This was only valid in the limit of an infinitely long control volume $L \rightarrow \infty$.

The \citet{Ancey2008} model reveals another way to compute the flux which does not depend on this device. 
Considering the downstream boundary of the control volume as the surface $S$ in the formula \ref{eq:anceyflux}, we can count the number of emigration events across this surface per unit time.  
There are four types of transitions in the \citet{Ancey2008} model, and these occur at random intervals. 
Counting the frequency of only one of these transitions is a non-trivial problem, and it relies on advanced techniques in stochastic mathematics. 
This counting problem was solved by \citet{Ma2014}.
They computed the bedload flux without the need for an exchange of averaging or an artificial control volume limit ($L\rightarrow \infty$). 
We will consider this work later in the review.

For now, consider the approach of the previous section. 
The bedload flux is $q_s \approx \frac{\nu_p u_s}{L} n$, in the limit of a large control volume $L\rightarrow \infty$ and when all particles in motion approach the same velocity $u_s$.
The mean and variance of the negative binomial distribution are 
\be \bra n \ket = \frac{\lambda}{\alpha-\mu} \ee
and 
\be \bra (\delta n)^2 \ket = \frac{\lambda \alpha }{\alpha-\mu}.\ee
Because of collective entrainment, characterized by $\mu$, the variance can exceed the mean in this model: 
\be \frac{\bra (\delta n)^2 \ket}{\bra n \ket } = \frac{1}{1-\mu/\alpha}, \ee
because $1/(1-\mu/\alpha) > 1$ if $\mu > 0 $.

By these relationships, the mean flux becomes 
\be  \bra q_s \ket = \frac{\nu_p u_s}{L} \ee,
and the magnitude of bedload fluctuations is

The Ancey et al 2008 inclusion of collective entrainment develops a Negative Binomial distribution of the bedload flux. 
Expressing the bedload flux as the number of moving particles within the control volume times their average velocity $u_p$ makes a Negative binomial distribution for bedload flux within the control volume. The mean bedload flux is $ \frac{\lambda}{\alpha-\mu} u_p $ and the variance of the flux is $ \frac{\lambda \alpha}{(\alpha-\mu)^2} u_p^2$: the fluctuations are much wider in the negative binomial distribution, meaning the inclusion of collective entrainment can model the large fluctuations observed in the \citet{Ancey2008} experiments.

One way to view the bedload flux we just presented: it is the number of moving particles times their velocity. 
However, since this model explicitly includes downstream emigration from the control volume, there is another way to define the flux: it is the number of emigration events within a unit of time. 
Counting the number of emigration events expressed from this model within a unit time is not easy. 
\citet{Ma2014b} explored this approach to calculating the bedload flux, and I will review this in a subsequent section.

Two key assumptions of this model are worth examining. 
These set the stage for the subsequent developments in birth-death modelling of the bedload flux which I review next. 
First, this model assumes that stationary bed particles are in infinite supply. 
This is an implicit assumption: Nothing prevents the system from entraining particles ad infinitum through the entrainment processes characterized by rates $\lambda_0$ and $\mu$. 
Second, the model considers sediment transport in an isolated control volume, so the sediment flux is described probabilistically in a locality, but spatial correlations cannot be accounted for: the flux upstream of the control volume is also a fluctuating quantity, implying spatial correlations through immigration events into the control volume considered in this model.

These two observations support two of the more recent developments to the \citet{Ancey2008} birth-death immigration-emigration model: \citet{Turowski2009} generalized this model to finite bed material in order to describe transport within semi-alluvial bedrock environments, where alluvial deposits of limited quantity sit on top of bedrock. When all available particles are entrained and moved away from the control volume, there are none left to entrain, so the entrainment process is turned off. 
Likewise, \citet{Ancey2010, Ancey2014,Ancey2015} extended the birth-death process within a control volume as I reviewed here to an array of spatially distributed control volumes (cells). 
They showed that immigration and emigration processes between cells set up a bedload diffusion, and they connected this theory with earlier sediment diffusion theories by deriving an Exner type equation of sediment diffusion. 
I will now review each of these developments in turn. 
\end{comment}

### 2021-09-03 draft/formerPapers/ens-rev/scope.tex
added phd thesis.

--- 2021-09-03 draft/formerPapers/ens-rev/review0.tex
@@
-\section{Review: birth-death theories of sediment transport}
+\section{Scope: identifying future research directions}
+\label{sec:scope}
 
-Birth death processes are a special type of stochastic process. 
-Stochastic processes track the evolution of a set of random variables through time. 
-Birth death processes consider a discrete state space, such as the size of a population. 
-Members of the population are born and die: transitions are allowed between adjacent sites in the discrete state space.
-
-Within sediment transport, Einstein was the first to consider a model for the bedload flux in terms of stochastic processes. 
-He considered the transport of individual bedload grains to be a random switching between motion and rest states. 
-In this way, supplementing this consideration with empirical reasoning, Einstein developed a formula for the mean transport rate \citep{Einstein1950}. 
-Einstein determined that the bedload transport rate should follow a Poisson distribution.
-
-Einstein's predecessors have extended his formulations in order to bring them in accord with experimental data and to phrase them in terms of a stronger mathematical foundation drawn stochastic process theory.
-These extensions probably began with \citet{Lisle1998}. 
-They linked Einstein's theory of tracer dispersion \citep{Einstein1937}, which is contingent on random transitions between states of motion and rest, into Markov process theory, showing that Einstein's tracer dispersion theory was just a two-state Markov process where the states are motion and rest.
-Then \citet{Ancey2006} went on to rephrase the \citet{Einstein1950} bedload flux derivation in terms of a two-state Markov process, where random transitions between states of motion and rest are included into a Chapman-Kolmogorov equation -- a general mathematical formulation.
-In this way, they derived a probability distribution for a bedload flux which is based upon underlying randomness in the entrainment and deposition of bedload grains. 
-Taking the mean of their bedload flux probability distribution reproduces an Einstein-like formula for the mean bedload flux. 
-This work is sketched to indicate how Markov process theory can work to describe the bedload flux.
-
-\subsection{Formalizing Einstein: a foundation in Markov process theory}
-
-The starting point for the \citet{Ancey2006} work was a set of experiments: the transport of uniform glass beads was measured in a narrow flume using image processsing techniques. 
-Measured fluxes had strong fluctuations: instantaneous fluxes took on values as much as ten times the mean. 
-Therefore, they attempted to describe the probability distribution of the bedload flux by revisiting Einstein's theory \citep{Einstein1950} from a Markov process framework.
-
-According to Einstein's way of thinking, an individual sediment grain undergoes a random transition from motion to rest states.
-We can label these two states as $x=0$ and $x=1$.
-Then, tracking the evolution of the random variable $x$ through time will fully describe the start-stop motion of the individual grain.
-
-Now we prescribe rates of entrainment and deposition in a small time $dt$. 
-If the particle is in motion ($x=1$) at time $t$, the probability it deposits to the rest state ($x=0$) at time $t+dt$ is modelled as a constant $\tau^{-1}$. 
-Likewise if the particle is at rest ($x=0$) at time $t$, the probability it entrains to the motion state ($x=1$) at time $t+dt$ is modeled as a constant $\sigma^{-1}$. 
-Making the Markov process assumption that the transitions of the particle between $t$ and $t+dt$ are independent of its history, the forward Kolmogorov equations read
-\begin{align} 
-\pi_0(t+dt) = \tau^{-1} dt \pi_1(t) + (1-\sigma^{-1} dt) \pi_0(t) \\
-\pi_1(t+dt) = \sigma^{-1} dt \pi_0(t) + (1-\tau^{-1} dt) \pi_1(t). 
-\end{align}
-
-Dividing through by $dt$ and taking the limit as $dt \rightarrow 0$ develops the Master equation for the two state process
-\begin{align} \label{eq:anc2006}
-\dot{\pi}_0(t) = \tau^{-1} \pi_1(t) - \sigma^{-1} \pi_0(t) \\
-\dot{\pi}_1(t) = \sigma^{-1} \pi_0(t) - \tau^{-1} \pi_1(t). 
-\end{align} 
-This simple model is called a telegraph process \citep{Cox1965}.
-
-To solve for the probabilties $\pi_0(t)$ and $\pi_1(t)$, the two equations can be summed to give $\frac{d}{dt} [ \pi_0(t) +\pi_1(t) ] = 0$, meaning $\pi_0(t) + \pi_1(t) = \text{constant}$.
-Since motion and rest are mutually exclusive and there are no other possibilities for the particle, $\pi_0(t) + \pi_1(t) = 1$: probabilities are conserved.
-Using this result into \ref{eq:anc2006} gives 
-\begin{align} 
-\pi_0(t) =  \\
-\pi_1(t) = 
-\end{align} 
-where $\pi_0(0)$ and $\pi_1(0)$ are the initial conditions, which we have not specified.
-
-After a sufficiently long time, the particle forgets its initial conditions and settles into equilibrium transport: 
-\begin{align} 
-\pi_0 = \\
-\pi_1 = 
-\end{align}
-Therefore, the time dependence and initial conditions are forgotten by the Markov process, and it has a set probability $\pi_0$ to be at rest and $\pi_1$ to be in motion.  
-Therefore, the fraction of time the particle spends in motion is $\frac{\tau}{\tau+\sigma} = \xi $, and the fraction of the time it spends in rest is $\frac{\sigma}{\tau+\sigma} = 1-\xi$. 
-The probability that the particle is in motion is a Bernoulli distribution.
-
-Now consider that on the bed surface within an observation window there are $N$ particles at any time. 
-Each of these particles is undergoing the same random switching between motion and rest states, so that the number of moving particles at any time is 
-\be N_m = \sum_{i=1}^N x_i \ee
-where $x_i=0,1$ is the Bernoulli random variable of the $i$th particle, with the probability of $x_i=0$ being $\pi_0$ and the probability of $x_1=1$ being $\pi_1$.
-
-Therefore, the probability that there are $N_m =n$ particles in a motion state is a counting problem. If $n$ are in motion then $N-n$ are at rest. The result is 
-\be \text{Prob}(n \text{ in motion}) = {{N} \choose {n}} \xi^n (1-\xi)^{N-n}. \ee 
-In statistics lingo, the sum of Bernoulli trials is a Binomial distribution. 
-This distribution has mean $\xi N$ and variance $\xi(1-\xi)N$. 
-Therefore the probability distribution of the number of moving particles within the observation window can be written 
-\be \text{Prob}(n) = \text{Bi}[ \xi N, \xi(1-\xi)N].  \ee
-
-If all particles in motion move at velocity $u_p$ and have volume $\nu_p$, then the bedload flux is a random variable $Q = \frac{u_p \nu_p}{L} N_m $.
-Therefore Ancey et al calculated the bedload flux probability mass function as
-\be \text{Prob}(Q) = \text{Bi} [\frac{u_p \nu_p}{L} \xi N, \xi (1-\xi) \frac{u_p^2 \nu_p^2}{L^2}]. \ee
-
-We conclude that the mean transport rate is 
-\be \bar{Q} = \nu_p \frac{N}{L}  \frac{\tau}{\tau+\sigma}   u_p. \ee
-This is conceptually similar to Einstein's formula \citep{1950}, and Ancey et al. explored this connection more carefully than we will here \citep{Ancey2006}. 
-The essential point is that Einstein took a somewhat different perspective.
-The key result of the \citet{Ancey2006} theory is that the mean transport rate is the result of $N$ parallel telegraph processes: each of the $N$ particles on the bed undergoes a switching process between motion and rest states, and the net result is the mean rate $\bar{Q}$ which is governed by a binomial distribution.
-
-In context of the experiments Ancey et al performed, the formulation just described did not capture the magnitude of the bedload fluctuations, although it described mean behavior quite well.  
-Their experimental bedload fluctuations were much larger than those expressed by this model. 
-Said in a different way, the probability distribution of the experimental bedload transport rate had a heavy tail, while the Binomial distribution of the rate derived from this parallel telegrapher's process theory has a light tail.
-
-Heavy tails are a hallmark of collective motions in physical systems \citep{Sornette2000}. 
-Of course, if bedload transport is driven by the advection of large turbulent eddies, a perspective supported by many experiments \citep{Nelson1995, Hofland2006, Celik2014, Amir2014, Shih2017}, then bedload motions will not be independent, but they will be coupled collectively to passing turbulent eddies.
-In context of the model just presented, this means the state of each telegraph process at an instant will be coupled to every other. 
-Therefore, coherent turbulence could imply heavy tails in the bedload probability distribution.
-
-There are other processes which could imply these collective effects as well. One option is entrainment due to particle collisions. 
-Moving particles can impact stationary particles, leading to their entrainment. 
-Essentially, this means the entrainment rate will depend on the number of active particles. 
-This was the extension made by \citet{Ancey2008} in order to capture the large fluctuations seen in their bedload flux experiments.
-
-\subsection{Capturing fluctuations: The inclusion of collective effects}
-
-In 2008, Ancey et al took a somewhat different perspective in order to describe the heavy tailed bedload probability distribution their experiments showed \citep{Ancey2008}.
-Instead of considering the motion of every bedload particle as an independent telegrapher's process, they considered the number of particles within a control volume as governed by a collective birth-death process \citep{Gardiner1983, Gillespie1992, VanKampen1992, Cox1965}.
-Birth of a member of the population is just entrainment. 
-Death is deposition. 
-The population can also immigrate (move in) to the control volume from upstream, and it can emigrate (move out) of the control volume to downstream. 
-Birth, death, immigration, and emigration rates characterize the dynamics of the entire population at once, rather than just the dynamics of independent individuals as the \citet{Ancey2006} work pursued.
-In this way, they managed to include a collective entrainment effect.
-
-Consider the probability that there are $N=n$ particles in motion at time $t$. 
-This can be denoted $\pi(n,t)$.
-The random variable $N$ takes on values $0,1,2,/dots$ -- non-negative integers of arbitrary magnitude.
-Each value has an associated probability $\pi(n,t)$.
-The entire population is considered subject to four processes within a small time increment dt.
-First, entrainment of a single bed paricle (birth) can happen with probability $\lambda_0 + \mu n$. 
-Here a collective entrainment effect has been included. 
-The probability of entrainment in $dt$ grows with the number of particles in motion. 
-This is a crude inclusion of the collective effects we just discussed: the advection of turbulent structures leads to correlations in entrainment rates, and the collision of moving particles with stationary particles can induce entrainment as well.
-
-Second, deposition of a single bed particle can happen in $dt$ with probability $\sigma n$. 
-The death rate is proportional to the number of individuals in the population. 
-Third, immigration of a particle from upstream into the control volume can happen with probability $\lambda_1$. 
-Fourth, emigration of a particle from the control volume to downstream can happen with probability $\nu$.
-
-These possible transitions develop a system of equations for the probabilities of finding $n$ moving particles within the control volume: 
-\be \pi_n(t+dt) = \alpha dt (n+1)\pi_{n+1}(t) + [\lambda + (n-1)\mu]dt \pi_{n-1}(t) + [1-dt[\lambda + n\alpha + n\mu]]\pi_n(t) \ee
-As $dt\rightarrow 0$ these develop a system of differential-difference equations for the probabilities $\pi_n(t)$ of finding $n$ particles in the control volume at time $t$: 
-\be \dot{\pi}_n(t) = (n+1) \alpha \pi_{n+1}(t) + (\lambda + (n-1)\mu)\pi_{n-1}(t) - (\lambda +n (\alpha+\mu))\pi_n(t). \label{eq:anc2008} \ee
-
-This birth-death immigration-emigration model can be solved by introducing a probability generating function $G(z,t) = \sum_{n=0}^\infty z^n \pi_n(t)$ \citep{Gardiner1983, Ancey2008}. 
-Multiplying \ref{eq:anc2008} by $z^n$ and summing over all $n$ gives, after a careful manipulation of the sums in order to form some function of $G(z,t)$ in every term, 
-\be \frac{\partial}{\partial t} G(z,t) = \lambda(z-1)G(z,t) + \{ \sigma + \mu z^2 + \nu - (\mu + \sigma + \nu)z\} \frac{\partial}{\partial z} G(z,t). \ee
-
-This partial differential equation for the probability generating function $G$ can be solved by the method of characteristics \citep{}. 
-If there are initially $N_0$ moving particles within the control volume, the solution is 
-\be G(z,t) + \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \ee
-The factor $K$ is the autocorrelation function $K = \exp(-t(\alpha-\mu)).$
-
-This probability generating function generates probabilities via a Taylor expansion around $z=0$: $G(z,t) = \sum_{n=0}^\infty \frac{z^n}{n!}\big(\frac{\partial}{\partial z}\big)^n G(z,t) |_{z=0}$.
-Comparing this formula with the definition of $G$, $G(z,t) = \sum_{n=0}^\infty z^n \pi_n(t)$, apparently the coefficient of $z^k$ in the power series expansion of $G(z,t)$ is $\pi_k(t)$.
-
-For times sufficiently large for the system to forget its initial condition and settle into a steady state, the autocorrelations $K(t)\rightarrow 0 $ in the expression of $G$, and the power series expansion of $G(z,t) \rightarrow G(z)$ yields a set of stationary (independent of time) probabilities
-\be \pi_n = \frac{\Gamma(r+n)}{\Gamma(r)n!} p^r (1-p)^n\text{, }n=0,1,\dots.\ee
-Here $r=\lambda/\mu$ and $p = 1-\mu/\alpha$. $\Gamma(x) = \int_0^\infty z ^{x-1} e^{-z} dz$ is the well known $\Gamma$-function of mathematical physics.
-
-These stationary probabilities are negative binomial distributions: 
-\be \pi_n = \text{NegBin}(n;r,p).\ee
-This is the generalization of $N$ parallel telegraph processes which was employed in the \citet{Ancey2006} paper. 
-The negative binomial distribution has a heavy tail as a result of collective entrainment $\mu$.
-
-When the collective entrainment process is turned off, $\mu\rightarrow 0$ which recovers the \citet{Ancey2006} result. 
-$\mu\rightarrow 0$ implies $r \rightarrow \infty$ and $p \rightarrow 1$. 
-Within this limit, the generating function G(z) becomes 
-\be G(z) = \exp(-\frac{\lambda}{\alpha}[z-1]).\ee
-This is the generating function of a Poisson process, so the probabilities of the number of active particles become Poisson distributed 
-\be \pi_n = \frac{r'^n}{n!}e^{-r'}\ee
-where the ratio $\lambda/\alpha = r'$ defines $r'$. 
-This is the same scaling behavior for large $n$ as the \citet{Ancey2006} model, since the large $n$ limit of a binomial distribution is a Poisson distribution.
-
-The Ancey et al 2008 inclusion of collective entrainment develops a Negative Binomial distribution of the bedload flux. 
-Expressing the bedload flux as the number of moving particles within the control volume times their average velocity $u_p$ makes a Negative binomial distribution for bedload flux within the control volume.
-The mean bedload flux is $ \frac{\lambda}{\alpha-\mu} u_p $ and the variance of the flux is $ \frac{\lambda \alpha}{(\alpha-\mu)^2} u_p^2$: the fluctuations are much wider in the negative binomial distribution, meaning the inclusion of collective entrainment can model the large fluctuations observed in the \citet{Ancey2008} experiments.
-
-One way to view the bedload flux we just presented: it is the number of moving particles times their velocity. 
-However, since this model explicitly includes downstream emigration from the control volume, there is another way to define the flux: it is the number of emigration events within a unit of time. 
-Counting the number of emigration events expressed from this model within a unit time is not easy. 
-\citet{Ma2014b} explored this approach to calculating the bedload flux, and I will review this in a subsequent section.
-
-Two key assumptions of this model are worth examining. 
-These set the stage for the subsequent developments in birth-death modelling of the bedload flux which I review next. 
-First, this model assumes that stationary bed particles are in infinite supply. 
-This is an implicit assumption: Nothing prevents the system from entraining particles ad infinitum through the entrainment processes characterized by rates $\lambda_0$ and $\mu$. 
-Second, the model considers sediment transport in an isolated control volume, so the sediment flux is described probabilistically in a locality, but spatial correlations cannot be accounted for: the flux upstream of the control volume is also a fluctuating quantity, implying spatial correlations through immigration events into the control volume considered in this model.
-
-These two observations support two of the more recent developments to the \citet{Ancey2008} birth-death immigration-emigration model: \citet{Turowski2009} generalized this model to finite bed material in order to describe transport within semi-alluvial bedrock environments, where alluvial deposits of limited quantity sit on top of bedrock.
-When all available particles are entrained and moved away from the control volume, there are none left to entrain, so the entrainment process is turned off.
-Likewise, \citet{Ancey2010, Ancey2014,Ancey2015} extended the birth-death process within a control volume as I reviewed here to an array of spatially distributed control volumes (cells). 
-They showed that immigration and emigration processes between cells set up a bedload diffusion, and they connected this theory with earlier sediment diffusion theories by deriving an Exner type equation of sediment diffusion.
-I will now review each of these developments in turn.
-
-Also Heyman 2013 considered that the waiting time distribution need not be exponential. Need more focus on the underlying poisson distributions of these models.
-
-\section{Including sediment availability: Turowski's two species birth death model}
-
-Semi-alluvial channels, where alluvium sits on top of bedrock, will violate one of the key assumptions of the \citet{Ancey2008} framework. 
-This model implicitly assumes that the bed surface has an infinite number of particles available for entrainment. 
-In reality, this situation will never be satisfied, even in alluvial channels, because there are a wide variety of stabilizing processes acting on bed surfaces \citep{Hassan2008, Venditti2016}.
-
-Turowski was concerned with bedrock scour due to alluvium. 
-Therefore he extended the Ancey framework. 
-The Ancey framework takes account of one population: the number of moving particles. 
-Turowski's extension included a second population into the framework: the number of particles available for entrainment. 
-He considered that as a particle entrains into a motion state, the number of particles available for entrainment decreases, and likewise as a particle deposits from the motion state, the number of particles available for entrainment increases.
-Therefore, he considered a probabilisitic population model of two coupled populations: such an inquiry is entirely new in stochastic theory of the bedload flux, although it is commonplace in ecological modeling \citep[e.g.][]{Pielou1977}
-
-Turowski considered a joint probability distribution for the random number $N$ of moving particles and $M$ of particles available for entrainment. 
-This can be written $\pi_{n,m}(t)$. 
-Within a small time interval $dt$ he considered birth, death, immigration, and emigration transitions governing the probabilities $\pi_{n,m}(t)$, similar to \citet{Ancey2008}. 
-Like \citet{Ancey2008}, he also included a collective entrainment contribution, arranging that the model expressed wide Negative Binomial-like fluctuations in the bedload rate.
-
-Turowski's transition probabilities in $dt$ varied by including the number of available particles $m$ into the entrainment and deposition probabilities.
-He considered a rate of entrainment $(\lambda_0 + \mu n)m$: when the population of particles available for entrainment is zero ($m=0$), the rate of entrainment is zero.
-The entrainment process, when it occurs, enacts a change of state $(n,m)\rightarrow(n+1,m-1)$: the number of active particles increases by one, while the number of particles available for entrainment decreases by one.
-
-Similarly, Turowski's deposition process was generalized from \citet{Ancey2008}.
-Deposition occurs with rate $\sigma n$ in time $dt$, just as in Ancey et al, but when this deposition process occurs it enacts the change $(n,m) \rightarrow (n-1,m+1)$ in the coupled populations. 
-The Kolmogorov equation is 
-\begin{multline}
-P(m,n;t+dt) = P(m,n-1,t)\lambda_0 dt + P(m+1,n-1,t)((m+1)\lambda_1 \\+ (m+1)(n-1)\mu)dt + P(m-1,n+1,t)(n+1)\sigma dt + P(m,n+1,t)(n+1)\nu dt \\+ P(m,n,t)*[1-dt(\lambda_0 + m \lambda_1 + n \nu + n \sigma + mn \mu).
-\end{multline}
-which in the limit $dt \rightarrow 0 $ becomes a Master equation for the joint probabilities $\pi_{n,m}(t)$: 
-\begin{multline} 
-\frac{d}{dt} P (m,n,t) = \lambda_0 P(m,n-1,t) +\\ ((m+1)\lambda_1 + (m+1)(n-1)\mu)P(m+1,n-1,t) + (n+1)\sigma P(m-1,n+1,t)\\ + (n+1)\nu P(m,n+1,t) - (\lambda_0 + m \lambda_1 + n\nu + n \sigma + mn \mu ) P(m,n,t).
-\end{multline}
-
-Again, as a reminder, the number of active particles is $n$ and the number of particles available for entrainment is $m$: this master equation describes a hirearchy of joint probability distributions for these discrete random varaibles.
-Exact solutions via counting arguments \citep[e.g.][]{Ancey2006} or probability generating functions \citep[e.g.][]{Ancey2008} are evidently no longer tractable. 
-\citet{Turowski2009} resorted to a numerical solution.
-Numerical algorithms for stochastic birth-death models are relatively easy to implement and are well described in a number of references \citep[e.g.][]{Gillespie1977, Gillespie1992}. 
-Turowski pursued one of these algorithms and related his transport rate computations to bedrock erosion and alluvial cover within a bedrock channel.
-
-Now we briefly describe the stochastic simulation algorithm of Gillespie 1977 ... 
-Then we reproduce Turowski's results using it ...
-
-We also explore the possibility of an exact solution for the Turowski model.. go harder than he did.
-
-Turowski's master equation reduces to the Ancey et al result as $m \rightarrow \infty$.
-As $m$ gets very large, $m-1 \approx m \approx m+1$: effectively, large $m$ makes $m$ constant as far as the master equation perceives.
-In this limit the probability $P(m,n,t) \approx P(n,t)$, and since $m$ is effectively constant, one can take $m\lambda_1 \rightarrow \lambda_1'$ and $m\mu \rightarrow \mu'.$ 
-With these substitutions the master equation reproduces the \citet{Ancey2008} result. 
-For our purposes, given our concern with alluvial channels as outlined in the introduction, the main takeaway of Turowski's work is that the birth-death modelling approach can be applied to study the stochastic dynamics of multiple populations whose variations are correlated.
-This is similar to competing species in ecology \citep{Pielou1977} or to multiple reacting chemical species \citep{Gardiner1983}.
-
-\section{The bedload flux at a point: the counting statistics of emigration events}
-
-Up to this point, the Markovian approaches to understand a fluctuating bedload rate were based upon analyzing the stochastic number of active particles within a control volume. 
-The bedload flux definition considered for these studies was somewhat unconventional. 
-Typically it is defined as the rate of bedload crossing a vertical plane perpendicular to the flow direction \citep{Ballio2014}.
-
-The studies of \citet{Heyman2013} and \citet{Ma2014b} demonstrated that the birth-death model of \citet{Ancey2008} can also be used to define a flux at a position in space. 
-Although the birth-death model of \citet{Ancey2008} counts the number of moving particles at each instant of time, it does not count the number of emigration events which contribute to this number. 
-Counting the number of emigration events in an interval of time is a much more difficult problem.
-Ma et al (2014) approached this counting problem using a path integral approach: this approach is essentially outlined in \citet{Ohkubo2009}. 
-The path integral approach draws heavily from the stochastic simulation algorithm of \citet{Gillespie1977}.
-
-The conventional definition of the bedload flux is an integral of particles through a surface perpendicular to the flow: 
-\be Q_s(t) = \int \int_A dA \textbf{u}_s \cdot \textbf{n}, \ee
-where $\textbf{n}$ is a unit vector perpendicular to the average flow direction and $\textbf{u}_s$ is the velocity of bed load. 
-This instantaneous flux is very difficult to measure, so usually it is averaged over some time interval $\delta t$: 
-\be Q_s(t;\delta t) = \frac{1}{\delta t} \int \int_A \int_t^{t+\delta t} \textbf{u}_s \cdot \textbf{n} d\tau dA = \frac{\delta V_s}{\delta t} = v_s \frac{\delta S(t;\delta t)}{\delta t}. \ee
-$Q_s(t;\delta t)$ is the volume flux averaged over $\delta t$; $\delta V_s$ is the volume of bed load particles passing through $A$ within $\delta t$; $S(t)$ is the cumulative number of particles passing through the cross section since t=0, and $\delta S(t;dt)$ is the number of particles passing through the surface between $t$ and $t+\delta t$, i.e.
-$\delta S(t; \delta t) = S(t+\delta t) - S(t)$.
-This equation means the bedload flux at $t$, when viewed in a time-averaged sense, is the number of particles crossing the surface between $t$ and $t+\delta t$. 
-Because volume units can always be multiplied in later, the bedload flux is written 
-\be q_s(t) = \frac{\delta S(t; \delta t}{\delta t}. \ee
-
-This bedload flux is a random variable.
-A fluctuation is defined strictly as $q_s' = q_s - \bra q_s \ket$, and the magnitude of fluctuations is defined without ambiguity by the variance $\bra q_s'^2 \ket$. 
-This $\delta S(t;dt)/\delta t$ is the number of emigration events between $t$ and $t+\delta t$, so if this number could be counted, the bedload flux would be known at the downstream boundary of the control volume: the bedload flux would be known at a point in space.
-
-The path integral formulation performs this counting problem. 
-For the sake of computation, the time interval t is discretized as $i \Delta t = t$, $i=0,1,2,\dots$.
-In the end, the limit $\Delta t \rightarrow 0$ will be taken. 
-If the increments $\Delta t$ are sufficiently small, Ma et al argued that emigration events during one of these intervals will be approximately Poissonian so that
-\be P[\Delta S_i = n | N_i(t) = N_i] = \exp[-f_3(N_i)\Delta t] \frac{(f_3(N_i) \Delta t)^N}{n!}. \ee
-This equation expresses the probability that $n$ particles emigrate between $t$ and $t+\delta t$ is a Poisson distribution with rate parameter $f_3(N_i)\Delta t$, where $f_3(N_i) = \gamma N_i$ is the emigration rate.
-
-Now Ma et al argued that because the entrainment rate $f_1 = \lambda + \mu N \~ O(L)$, the deposition rate $f_2 = \sigma N \~ O(L)$, and the emigration rate $f_3 = \gamma N \~ O(1)$, if the observation window is sufficiently large, entrainment and deposition processes dominate, so that the number of moving particles $N$ is perturbed only weakly by emigration.
-In this limit of sufficiently large observation window size, it then holds that
-\be P(\sum \Delta S_i = n | N_1,N_2, \dots) = \exp[-\sum f_3(N_i)\Delta t]\frac{(f_3(N_i)\Delta t)^n}{n!}, \ee
-which in the continuum limit becomes a path integral counting emigration events: 
-\be P[\delta S = n |N(t): t_0<t<t_0+\delta t] = \exp[-\int f_3(N(\tau))d\tau]\frac{[\int f_3[N(\tau)]d\tau ]^n}{n!}.\ee
-Denoting the integral over the process $f_3(N(\tau))$ as $\alpha = \int_{t_0}^{t_0+\delta t} d\tau f_3[N(\tau)]$, the probability for $n$ emigration events in $\delta t$ is written 
-\be P(\delta S = n) = \int d\alpha \frac{e^{-\alpha}\alpha^n}{n!}P(\alpha,t). \ee
-From this difficult derivation a simple formula emerges: the probability of $n$ emigration events in $\delta t$ is an integral over all possible realizations of the process $N(t)$, where each process is given a weight $e^{-\alpha}\alpha^n/n!
-P(\alpha,t) d\alpha$.
-This is called a Cox process.
-It's a generalized Poisson process where the parameter $\alpha$ is also a random variable.
-It represents a superposition of a series of Poisson distributions taking every possible path of $N(t)$ into account.
-This is a stochastic description of the bedload transport flux based on the counting statistics of emigration events.
-
-Notably, because of the time integral of $N(t)$ involved in this relationship, it has a memory. It is non-Markovian. 
-This character reveals a very interesting fact: the fluctuations of the transport rate depend on the time scale of observation. 
-In fact, Ma et al derived the expression for the variance: 
-\be \delta t \text{var} [q_s(\delta t)] = \gamma^2 \frac{\lambda (\sigma + \gamma)}{(\sigma + \gamma -\mu)^2} 2 t_c [\delta t - t_c (1-e^{-\delta t/t_c}] + \frac{\gamma\lambda}{\sigma + \gamma -\mu}\delta t \label{eq:3regimes} \ee
-
-where $t_c = 1/(\sigma + \gamma - \mu)$ is the autocorrelation time of $N(t)$, representing memory in the system. 
-Apparently, the variance of the bedload flux at a point depends on the observation or averaging time-scale $\delta t$.
-
-In fact, Ma et al discriminated three regimes of behavior from \ref{eq:3regimes}:   
-\be
-    \delta t \text{var}[q_s(\delta t)] \approx 
-\begin{cases}
-    \frac{\gamma \lambda}{\sigma + \gamma - \mu}, & 0<\delta t \ll t_l \\
-    \delta t \frac{\gamma^2 \lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}   & t_l<\delta t< t_c \\
-    2 \gamma^2 \frac{\lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}t_c + \frac{\gamma \lambda}{\sigma + \gamma - \mu}   & t_c \ll \delta t < \infty
-\end{cases}
-\ee
-When this three-regime equation is divided by the mean bedload rate $\text{mean}[q_s(\delta t)]$, the behavior is discriminated by a dimensionless number Ma et al called $Ra$: 
-\be Ra(\delta t) \approx \begin{cases} 1 & 0<delta t \ll t_l \\
-\frac{delta t}{t_l} & t_l < \delta t < t_c \\ 
-\frac{2 t_c}{t_l} + 1 & t_c \ll \delta t < \infty \end{cases} \ee
-
-The essential point is that Ma et al derived three regimes of fluctuations from the \citet{Ancey2008} model. 
-Depending on the sampling interval $\delta t$, three different scaling relations are possible for the magnitude of fluctuations in the bedload flux. 
-These intervals are discriminated by two timescales $t_c$ and $t_l$. 
-There are probably deep consequences of this research which remain to be fully understood. 
-At this stage, it's sufficient to note that fluctuations differ with observation timescale, and that this effect can be discriminated into stages with a dimensionless number $Ra$.
-
-The origin of the two timescales still needs to be discussed.
-
-There is another way to get information about the bedload flux at the downstream boundary of the control volume defined by \citet{Ancey2008}, and this was examined by \citet{Heyman2013, Heyman2016}. 
-Rather than actually counting the number of emigration events with the complicated path integral approach, Heyman et al.
-took a somewhat simpler angle: they examined the waiting time between successive emigration events.
-
-\section{The waiting time between emigration events: Bedload flux from waiting time statistics}
-
-These studies are concerned with linking the birth-death formulation, which in the original interpretation deals with transport characteristics within an observation window \citep{Ancey2008}, to the transport characteristics at a point -- the number of particles crossing a plane perpendicular to the flow direction.
-
-\citet{Heyman2013} was concerned with the statistics of the time interval between successive emigration events, while \citet{Ma2014b} focused on actually counting the number of emigration events in a unit time: that is, they computed the bedload flux over the downstream boundary of the control volume \citet{Ancey2008}.
-
-The waiting time between successive transport events is significant in a broad range of transport studies. 
-First, there are probabilistic transport rate formulations which use the waiting time distribution as their input \citep{Turowski2010}. 
-Second, the waiting time between successive transport events, as we shall show, reflects back on the underlying entrainment and deposition characteristics, so it provides fundamental information into the somewhat mysterious physics of bedload transport.
-
-The original \citet{Einstein1937,Einstein1950} approach reveals that the waiting time between successive transport events should be an exponentially distributed random variable. 
-However, Heyman et al (2013) demonstrates that the inclusion of collective entrainment due to \citet{Ancey2008} modifies the waiting time distribution in a non-trivial way. 
-The waiting time between successive entrainment events has contributions at two different scales: first, there is a shorter timescale of individual entrainment events, controlled by fluid turbulence and essentially recognized within the \citet{Einstein1937} formalism, and second, there is a longer timescale due to collective entrainment.
-These authors showed that within the \citet{Ancey2008} birth-death model, at low transport rates, when entrainment and motion are highly intermittent, the short timescale and the long timescale are well separated.
-While at higher transport rates, these two timescales blend together as individual and collective entrainment contributions become indistinguishable.
-
-I'll now sketch the \citet{Heyman2013} derivation of the statistics of the waiting time between successive emigration events, and show how the two different timescales due to individual and collective entrainment arise.
-For \citet{Heyman2013} the variable of interest is $S_k = \sum_{i=0}^k T-i$: $S_k$ is the time when the $k$th emigration event occurs.
-$T_1, T_2, \dots $ are the waiting times: the time from $t=0$ to the first emigration event is $T_1$.
-The time from $T_1$ to the second emigration event is $T_2$, and so on.
-For a Markov process such as the \citet{Ancey2008} model, the waiting times are independant and identically distributed. 
-Within the Einstein model we have already seen that the waiting time probability distribution is expoential. 
-However, in the \citet{Ancey2008} formalism there is no reason to expect an exponential distribution.
-
-Heyman et al define $F_n(t)$ as the probability that there are $n$ particles in motion and no emigration event occurred in time t: 
-\be F_n(t) = Pr(T>t,N(t)=n).\ee
-By extension, the probability that there are any number of particles and no emigration event occurred in time $t$ is 
-\be F(t) = Pr(T>t) = \sum_{n=0}^\infty F_n(t) .\ee
-
-Now $F(t+\Delta t)$ is equivalent to the probability that any other event but emigration occurs in $\Delta t$. 
-Therefore we can write a master equation for $F(t)$: 
-\be F_n(t+\Delta t) = F_n(t)[1-(\lambda + n(\sigma+\mu+\gamma))\Delta t] + F_{n+1}(t) \sigma(n+1)\Delta t + F_{n-1}(t)\lambda + \mu(n-1)\Delta t + o(\Delta t).\ee
-This equation holds for $n\geq 1$. At $n=0$ deposition and collective entrainment processes are not possible, therefore 
-\be F_0(t+\Delta t) = F_0(t) [ 1-\lambda \Delta t] + F_1(t) \sigma \Delta t + o(\Delta t) \ee 
-Again dividing by $\Delta t$ and taking $\Delta t \rightarrow 0$ gives the differential difference equations
-\begin{align} 
-F_0'(t) &= \lambda F_0(t) + \sigma F_1(t) \\
-F_n'(t) &= -(\lambda + n(\sigma + \mu + \gamma))F_n(t) + \sigma(n+1)F_{n+1}(t) + [\lambda + \mu(n-1)]F_{n-1}(t) \text{ if } n\geq 1 \end{align} 
-for the waiting time distribution.
-
-Summing all of the terms gives the simple equation 
-\be \sum_{n=0}^\infty F_n'(t) = -\gamma \sum_{n=0}^\infty nF_n(t), \ee
-which, denoting the pdf of waiting times $T$ as $f_T(t)$ gives the simple relationship: 
-\be f_T(t) = -F'(t) = \gamma\bra F_n(t)\ket. \ee
-Therefore the probability distribution function of waiting times between emigration events, $f_T(t)$, is an average of the probability that there are $n$ particles and no emigration event occurred in $t$ over all $n$.
-
-The general solution of $F_n(t)$ can be obtained from the differential equations with a generating function: 
-\be G(z,t) = \sum_{n=0}^\infty F_n(t) z^n. \ee
-Multipling the differential equation by $z^n$ and summing over all $n$ gives
-\be \frac{\partial G}{\partial t} = (\sigma + \mu z^2 - z(\alpha + \mu) ) \frac{\partial G}{\partial t} + (z-1)\lambda G, \ee
-where the shorthand $\alpha = \gamma + \sigma $ has been introduced.
-
-This equation can be solved with the method of characteristics, giving a closed form solution for $G(z,t)$.
-A useful property of generating functions (and the source of their name) is $\bra F_n \ket = \partial G(z,t) /\partial t |_{z=1}$, therefore the probability distribution of waiting times between emigration events in the \citet{Ancey2008} birth-death model of bedload flux is
-\be f_T(t) = \gamma(z_1-z_2)^{\lambda/\mu} \big( \frac{\alpha-\mu}{A(t)-B(t)} \big)^{\lambda/\mu + 1} e^{-\lambda (1-z_2)t}\big( \frac{\lambda/\mu+1)B(t)}{A(t)-B(t)} [(1-z_2)e^{-(z_1-z_2)t}+z_1-1]+e^{-\mu(z_1-z_2)t}-1 \big)  \ee
-When collective entrainment is turned off ($\mu=0$), the probability distribution becomes $f_T(t) = \frac{\alpha}{\lambda \gamma} e^{-\alpha t/(\lambda \gamma)}$: the exponential waiting time between emigration events is recovered.
-
-This probability distribution for waiting times between emigration events in the \citet{Ancey2008} model has, in general, contributions from two timescales.
-The relatively common individual entrainment process imparts a 'fast' timescale to the probability distribution of emigration waiting time, while the relatively rare collective entrainment process imparts a 'slower' timescale.
-When collective entrainment is turned off, only the fast timescale is present in the pdf.
-When the entrainment rate is relatively low, the fast timescale is much shorter than the slow timescale, so that the timescales are well separated and the waiting time distribution is not well approximated by any common distribution in the literature (Gamma, exponential): the behavior of the \citet{Ancey2008} model at low transport rates has defied prediction by any other framework.
-
-The most significant point of the \citet{Heyman2013} paper is that the probability distribution function of waiting times between emigration events can be used to calibrate the \citet{Ancey2008} birth death model to flume experiments.
-This is because short emigration times control $\alpha-\mu$ and long emigration times control $\lambda$. 
-The evaluation of waiting times between emigration events supports a calibration of the stochastic birth-death model of \citet{Ancey2008} on flume studies where the transport rate is measured at the outlet of the flume, but the erosion, deposition, or emigration rates are not necessarily measured.
-In contrast to \citet{Ma2014b} the mathematics are not much beyond the original formulation of the \citet{Ancey2008} model, so this work may be more readily accepted by the research community.
-
-\section{The diffusion of bedload from a microstructural basis}
-
-\citet{Ancey2008} developed a generalized version of \citet{Einstein1950} which took account of a new process to describe the large fluctuations seen in experimental bedload flux signals: collective entrainment.
-\citet{Turowski2009} generalized this model to finite sediment availability, while \citet{Heyman2013} and \citet{Ma2014} used the \citet{Ancey2008} model in its original form to understand the distribution of waiting times between emigration events and the bedload flux probability distribution at a point in space, respectively.
-These two works developed new understanding of spatial and temporal aspects of bedload flux: \citet{Heyman2013} demonstrated that the waiting time between successive emigration events expressed contributions from slow and fast timescales related to individual and collective entrainment, respectively.
-\citet{Ma2014} highlighted the role of observation timescale on the magnitude of flutuations, and discriminated three distinct scaling regimes for bedload fluctuations which are contingent on a dimensionless number $Ra(\delta t)$, where $\delta t$ is the timescale of observation.
-
-However, up to this point in early 2014, stochastic birth-death models had focused on a single region of space -- a control volume \citep{Einstein1950, Ancey2006, Ancey2008}, or a single point in space -- the downstream end of the control volume, or the plane across which emigration events happen \citep{Heyman2013, Ma2014b}.
-This approach is obviously oversimplified. Variability of sediment transport rates is almost a defining feature of them \citep{Hassan2008, Venditti2016, Nelson2014}. 
-There is a need to take account of spatial differences in entrainment and deposition characteristics in calculating bedload fluxes. 
-Additionally, there are a wide set of issues centered around bedload diffusion-- or the spreading of bedload particles as they are tracked through time and subjected to random entrainment and deposition events \citep{Hassan2016}.
-Indeed, this study of bedload diffusion was the original research directive of \citet{Einstein1937}.
-
-Often, bedload diffusion has been understood through deterministic partial differential equation models.
-These advection diffusion equations have been derived by considering conservation of mass \citep{}, but they were not very well connected to the underlying stochastic dynamics of bedload transport.
-Generating this connection between a microstructural stochastic model like \citet{Ancey2008} and the continuum diffusion equation treatment of bedload diffusion \citep[e.g.][]{Parker2002} was the directive of the next papers on birth-death modeling \citep{Ancey2014,Ancey2015}.
-These papers took a conceptually straightforward extension of the \citet{Ancey2008} model which leads to difficult mathematics.
-
-Rather than considering a single control volume, as in \citet{Ancey2008}, \citet{Ancey2014,Ancey2015} considered an infinite array of adjacent control volumes (cells), indexed by $i=1,2,\dots,M$. 
-Bedload particles can entrain and deposit within each cell, and they can also migrate between cells. 
-Each of these transitions are characterized by probabilities per unit time in generalization of \citet{Ancey2008} across spatial extent. 
-The $i$th cell has a random number of particles $N_i$ in motion within it. 
-Therefore the state of the system at an instant of time is fully characterized by the set of these numbers: $\textbf{n} = (n_1(t), n_2(t), \dots n_M(t))$.
-
-One target is the grand probability distribution of the number of active particles within each cell at an arbitrary time: $P(\textbf{n};t)$. 
-Another target is the continuum limit: letting the length $\Delta x$ of each cell shrink to zero develops an advection-diffusion equation for bedload diffusion. 
-Ancey et al termed this a "Stochastic interpretation of bedload diffusion": and the latter target has much more fundamental scope.
-It explores the link between microscale stochastic models of bedload transport and mesoscale deterministic models based upon advection-diffusion equations.
-
-Now I'll sketch the \citet{Ancey2014, Ancey2015} derivation of a birth-death-migration model across an array of cells. 
-There are essentially three transitions to take account of: 
-\begin{enumerate}
-\item Entrainment can occur within each cell ($i$) at probability per unit time $(\lambda_i + \mu_i n_i ) \delta t$
-\item Deposition can occur within each cell ($i$) at probability per unit time $ \sigma_i n_i \delta t$
-\item Migration can occur from cell $i$ to cell $i+1$ (downstream) at probability per unit time $\nu_i$
-\end{enumerate} 
-Now some notation is introduced to write the effect of these transitions on $\textbf{n}$ in shorthand. 
-If the change of state due to one of these transitions is written as $\Delta \textbf{n}$, then the effect of these transitions can be characterized using vectors $\textbf{r}_i^j$ and $\textbf{r}_i^\pm$.
-The vectors $ \textbf{r}_i^j$ are of the same dimension as $\textbf{n}$ and all but two of their entries are zero: $r_i = 1$, $r_j = -1$, $r_k=0$ for $k\neq i,j$.
-The vectors $\textbf{r}_i^\pm$ are the same dimension as $\textbf{n}$ and all but their $i$th entry is zero: $r_i = \pm 1$, $r_j = 0$, $j\neq i$.
-
-Thus the transition probabilities of each process can be written: 
-\begin{enumerate}
-\item $ p_i^3 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^+; t+ \delta t) = (\lambda_i + \mu_i N_i) \delta t $ -- entrainment
-\item $ p_i^2 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^-; t+\delta t) = \sigma_i N_i \delta t $ -- deposition
-\item $ p_i^1 = \text{Prob}(\textbf{N} = \textbf{n} \textbf{n} + \textbf{r}_i^{j-1}; t+ \delta t) = \nu_{i-1}N_{i-1} \delta t $ -- migration
-\end{enumerate}
-
-These transition probabilities develop a master equation analogous to the one from \citet{Ancey2008}, but generalized to a collection of $M$ cells: 
-\begin{multline} \frac{\partial}{\partial t} P(\textbf{n};t) = \sum_{i=1}^M (n_i+1)[P(\textbf{n}+\textbf{r}_{i+1}^j,t) \nu_i + P(\textbf{n}+\textbf{r}_i^+,t)\sigma_i ]\\
-+ P(\textbf{n}+\textbf{r}_i^-,t)(\lambda_i + \mu_i(n_i-1))\\
-+ P(\textbf{n}+\textbf{r}_i^{j-1},t)\nu_{i-1}n_{i-1}\\
-- P(\textbf{n},t)(\nu_{i-1} n_{i-1}+\lambda_i + \mu_i n_{i+1} + \nu_i n_i + \sigma_i n_i ) \end{multline} 
-Information about solving this equation is sparse in the literature. 
-The closest approach has been in ecology: these birth/death/migration-type stochastic models have been considered in context of ecological population dynamics since at least the sixties \citep{Bailey1968}.
-One exact solution of a very similar stochastic model has been attained using path integral approaches borrowed from quantum field theory \citep{Field2010}.
-By any metric, the analytical barrier this multidimensional master equation presents is not trivial, although definitely numerical solutions are possible \citep[e.g.][]{Gillespie1992}.
-
-In order to link this equation to spatial diffusion, \citet{Ancey2014, Ancey2015} resorted to a Poisson space representation \citep{Gardiner1983}. 
-In effect, the Poisson transformation converts the discrete random variable $\textbf{N}$ to a continuous random variable $\textbf{a}$ (the Poisson rate). 
-The Poisson transform is 
-\be P(\textbf{n},t) = \prod_i\int \frac{e^{-a_i}a_i^n}{n!}f(\textbf{a},t)d\textbf{a}, \ee
-where $\textbf{a} = (a_i) \geq 0$ for $i=1,2,\dots$ and $f(\textbf{a},t)$ is the multivariable probability density of the continuous vector $\textbf{a}$.
-
-\citet{Ancey2014, Ancey2015} used this transformation to obtain a Langevin equation representation of the dynamics of the random variable $\textbf{a}$.
-That is, they obtained a stochastic differential equation representation of $\textbf{a}$: 
-\be da_i(t) = (\lambda_i - a_i(\sigma_i-\mu_i) + \nu_{i-1} a_{i-1} - \nu_i a_i)dt + \sqrt{2 \mu_i a_i}dW_i(t),\ee
-and the $W_i(t)$ is a white noise term (Wiener Process) on cell $i$. 
-The mean steady state solution of this Langevin equation is 
-\be \bra a \ket_ss = \frac{\lambda}{\sigma-\mu} \ee 
-for all $i$.
-
-Now \citet{Ancey2014,Ancey2015} sought out to find an advection diffusion equation as the limit of cell size $\Delta x \rightarrow 0$, so they went on to impose this limit on the Langevin equation, thereby obtained an advection diffusion equation.
-Unfortunately, the limit is more subtle than expected. 
-This did not really work for reasons I need to understand better, and it did not work in ways I need to learn to explain better. 
-More study is needed on spatially varying birth-death models, from me, and from others especially.
-
-\section{Spatial correlations in bedload transport: }
-
-\citet{Heyman2014} used the \citet{Ancey2014} model to examine a set of experiments resolving spatial correlations of bedload transport. 
-They found a stochastic differential equation for the mean of the poisson rate $\eta$: 
-They related this to a point process framework which allowed them to analyze spatial correlations in the bedload rate and relate this to their birth-death model. 
-Somehow this whole thing is contingent on their two-point spatial correlation process. 
-They state that the whole problem is linear so that it'd be straightforward to extend it to 2d. 
-slowing down ...
-
-\section{recap} 
+\subsection{recap} 
 Einstein developed first stochastic theories describing the diffusion and flux of bedload. 
@@
 
-\section{Assumptions, calibraiton problems, and unclear aspects: A criticism of the birth-death approach}
+\subsection{Assumptions, calibraiton problems, and unclear aspects: A criticism of the birth-death approach}
 
-Now that earlier literature is evaluated the scope for generalization can be outlined. 
-This scope is a necessarily subjective consideration, and I'd like to highlight that nobody knows what the future of stocahstic modeling will hold. 
 First, the various glaring assumptions of the birth-death approaches are highlighted. 
@@
 These may include collective entrianment, the formation and disintegration of particle clusters, interactions between the motion of grains of idfferent size, turbulent structure, local avalanche behavior, and bed form migration.
+\item there have been no theories developed to compute the collective entrainment rate from any simplified mechanical model, although there is a very large set of work concerned with calculating the individual entrainment rate from considerations of fluid turbulence and random granular arrangement \citep{Einstein1949, Einstein1950, Grass1970, Paintal1971, Cheng1998, Wu2004, Dey2008, Tregnaghi2012, Dey2018}.
+Calculating the collective entrainment rate from underlying principles appears on the surface very difficult.
+If it is considered to stem from collisions of moving grains with stationary grains, the collective entrainment probability will be related to the probability of collisions with the granular bed.
+If it is considered to stem from granular avalanches, where coherent turbulent structures initiate collections or clusters of grains into motion simultaneously, the collective entrainment probability will depend on the collective dynamics of the granular assembly -- leading immediately into the murky physics of force balance within granular assemblies.
+\item Einstein-like assume a clean division between rest and motion states, which is no doubt an idealization.
+Bedload transport makes a continuous transition from the idealized start-stop motions of Einstein-like models to a less idealized granular flow or creep, where all particles move together in a coorelated fluid-like flow \citep{}, meaning bedload models should only hold at relatively low mobility stages.
+\item At the same time, the divison of motion into only two categories may be flawed.
+A wide collection of studies have highlighted different modes of motion.
+\citet{Einstein1950} considered that bedload was particles moving "in a rolling, sliding, or saltating mode".
+What if rolling, sliding, and saltating modes of motion were treated independently?
+In an \citet{Ancey2006} model of independent particles switching between states, this would require a four-state model.
+There are too many transition probabilities within such a model probably to calibrate the model from experiments.
+Again, in order to consider such a scenario we would need more knowledge of underlying mechanics.
+
+\end{enumerate}
+
+Incorporating multiple grain sizes in birth-death models is possible, in principle. 
+As a first approximation, it is easy to introduce multiple grain sizes without including interactions between grain sizes in the entrainment and deposition probabilities for each size fraction. 
+However, spatial and temporal heterogeniety in the bed surface characteristics are a hallmark of bedload transport of gravel mixtures \citep{Hassan2008}. 
+This leaves models such as \citet{Ancey2008} somewhat without basis when multiple grain sizes are considered. 
+The concept of \citet{Turowski} is essential to include when multiple grain sizes are considered: the bed surface state determines which grains can entrain \citep[e.g.][]{Wilcock2003, Parker1982}, and presumably which grains can deposit, as well.
+The entrainment and deposition rates of each size fraction will need to depend on the bed surface state, and fractional transport will set up spatial hetereogenieties in the bed surface state, so that an \citet{Ancey2014} type model, incorporating the possiblity of spatial hetereogeniety, should be mixed with a \citet{Turowski2009} type model to incorporate the effect of the bed surface state on the entrainment and deposition rates of each size fraction.
+This extension will not be easy, and it will introduce many undetermined parameters: to pin down this transport model of multiple interacting size fractions will take a lot of work.
+
+These stochastic models are admittedly difficult to calibrate. 
+Their calibration requires a large dataset which is prohibitive to measure within natural streams. 
+Therefore, their applicability in real streams remains limited until the inputs of stochastic models can be computed from physical theories based upon practically measurable quantities. 
+This quest for relationships to compute the inputs of stochastic models from measurable quantities has been called "stochastic closure" for the analogous problem in turbulence \citep{Heyman2016}. 
+The difficult issues precluding the application of stochastic models to natural streams should not be downplayed: much more work is needed.
+
+\begin{comment} 
+Birth death processes are a special type of stochastic process. 
+Stochastic processes track the evolution of a set of random variables through time. 
+Birth death processes consider a discrete state space, such as the size of a population. 
+Members of the population are born and die: transitions are allowed between adjacent sites in the discrete state space.
+
+Within sediment transport, Einstein was the first to consider a model for the bedload flux in terms of stochastic processes. 
+He considered the transport of individual bedload grains to be a random switching between motion and rest states. 
+In this way, supplementing this consideration with empirical reasoning, Einstein developed a formula for the mean transport rate \citep{Einstein1950}. 
+Einstein determined that the bedload transport rate should follow a Poisson distribution.
+
+Einstein's predecessors have extended his formulations in order to bring them in accord with experimental data and to phrase them in terms of a stronger mathematical foundation drawn stochastic process theory.
+These extensions probably began with \citet{Lisle1998}. 
+They linked Einstein's theory of tracer dispersion \citep{Einstein1937}, which is contingent on random transitions between states of motion and rest, into Markov process theory, showing that Einstein's tracer dispersion theory was just a two-state Markov process where the states are motion and rest.
+Then \citet{Ancey2006} went on to rephrase the \citet{Einstein1950} bedload flux derivation in terms of a two-state Markov process, where random transitions between states of motion and rest are included into a Chapman-Kolmogorov equation -- a general mathematical formulation.
+In this way, they derived a probability distribution for a bedload flux which is based upon underlying randomness in the entrainment and deposition of bedload grains. 
+Taking the mean of their bedload flux probability distribution reproduces an Einstein-like formula for the mean bedload flux. 
+This work is sketched to indicate how Markov process theory can work to describe the bedload flux.
+
+\begin{enumerate}
+\item all einstein-like models only hold where a clean division between motion and rest states exists.
+this will break down at high transport rates when bedload becomes a sheet flow with continuous particle motion
+\item 
+\item physical and mechanisistic basis of collective entrainment is foggy
+\item models are limited to one grain size 
+\item these models only have one way feedback between grains and fluid
+\item models are difficult to calibrate
+\item models are difficult to actually test -- as they get more flexible they can fit anything.
+they contain ad hoc parameters which must be adjusted.
+so far they have only been fitted on highly idealized uniform or semi-uniform sediment grains
+\item all in all, extension to natural scope is precluded
+\item bedforms induce spatial differences in the model parameters 
+\item "as fluctuations span scales larger than the ones at which bedload can be considered stationary and homogeneous, oor even at scales larger than the experiment size, average equations fail at describing the nonlinear ..
+"
 \end{enumerate}
@@
 3. Extension to multiple grains sizes: A formal outline
+
+\end{comment}

### 2021-09-03 draft/formerPapers/ens-rev/scraps/contemporary.tex
added phd thesis.

\section{Contemporary Markovian sediment transport theories}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sun 1989: a first markov model approach} 
Need to get the reference still

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Lisle et al 1998: Generalizing Einstein 1937 to a finite velocity} 
\citep{Lisle1998}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Papanicolaou 2002: did a thing}
\citep{Papanicolaou2002}
"characterized the sediment entrainment
process using a two-state continuous-time Markov chain"

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Ancey et al 2006: Reformulating Einstein 1950 with a more physically based foundation}
\citep{Ancey2006}

\subsection*{Wu and Yang 2004: Quasi-four state model}
\citep{Wu2004a} is the reference.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Ancey et al 2008: Including collective entrainment into Einstein 1950} 
\citep{Ancey2008}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Ancey et al 2014: From birth-death to diffusion} 
\citep{Coleman2009, Ancey2014}
\begin{enumerate}
\item Exner 1927 -- the exner equation 
\item Coleman and Nikora 2009 -- the exner equation
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Sun et al 2000: model for any size fraction of sediment which is basically Einstein}
\citep{Sun2000}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Tsai et al's three-state markov for mixed size sediment} 
\citep{Tsai2013, Tsai2014, Kuai2016}

adopted a three-state
continuous-time Markov chain model to simulate the particle
movement among three different states for uniform and nonuniform
particle size, respectively.

there is a 2014 and 2013 tsai paper, also investigate the 2016 kuai and tsai paper

"The objective of this paper is to develop a nonhomogeneous
discrete-time 3-state Markov chain model that can efficiently quan-
tify both the bedload and the suspended load discharges under un-
steady flow conditions for mixed size sediment particles"

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Turowski2009 generalized ancey to bedrock situations with finite supply} 
\citep{Turowski2009}

\subsection*{Armanini 2015: Including entrainment probability and hop length distribution in Einstein formulation} 
\citep{Armanini2015}

\begin{comment}

Others to consider

Markovian phase switching for low sediment transport rates 
\begin{itemize}
\item Dey 1999 
\item Dey 2018
\item Einstein et al, 1937
\item Kalinske 1947 
\item Einstein et al 1950 
\item Nakagawa and Tsujimoto [1980] pickup einstein school 
\item van Rijn [1984] pChien, N., and Wan, C. H. (1983). Mechanics of sediment transport.ickup einstein school 
\item Tsujimoto [1991] pickup probability 
\item weeks 1996 assymetrical random walk 
\item Lisle 1998 Sedi xport as birth death process 
\item sun 2000 - mixed size sediment transport as 2 state markov
\item tasi and yang 2013 three-state markov chain model 
\item Ancey et al 2006 - rederivation of einstein from birth-death
\item Ancey et al 2008
\item Ancey et al 2010
\item Ganti 2010  - anomalous diffusion from fractional pdes 
\item Ancey et al 2014 
\item Hu and Guo 2011 -- this is a probability entrainment model which is pretty conventional and not really standout. It contains a lot of good chinese references. Also investigates the concentration of near bed particles. This is not really markovian 
\item Armanini 2014 -- reworking of Einstein 1950 
\item Yalin 1972 / 1977 / 1992-- criticism of Einstein 
\item Armaniini 2018 -- river hydraulics book 
\item Exner 1927 -- the exner equation 
\item Coleman and Nikora 2009 -- the exner equation
\item Ballio 2014 -- def of solid discharge
\item Martin 2014 mean reverting random walk 
\item Fan 2014 random walk from langevin equation
\item Fan 2016 random walk + ad hoc resting times 
\item sun 2015 - an anomalous diffusion model? 
\item Ballio 2018 -- Lagrangian vs Eulerian approach
\item Furbish et al 2016 Jayne Entropy 
\end{itemize}

wu chen did bayesian updating on an entrainment model

Housais argues that sediment motion is a continuous transition driven by granular creep, at which point the partition between phases somewhat breaks down.

\subsection{Stochastic modeling approaches: phenomenological versus mechanistic}

In addition to the division between deterministic (like Bagnold) and stochastic (like Einstein) approaches, another division can be made between the mechanistic and the phenomenological approaches.
Mechanistic approaches rely on underlying physics including fluid and granular properties and their interaction in order to phrase sediment transport processes. 
Phenomenological approaches start from a set of prescribed assumptions with differing amounts of empirical basis. 
Phenomenolgoical approaches include the transport model of Einstein (1950). 
If chaos, the innumerable degrees of freedom, and open system issues guide us toward stochastic approaches, the difficulty of explaining the phenomena of sediment transport apparently encountered when using phenomenological approaches 
The failures of phenomenolgoical approaches include a lack of generalizability, ... . 
(this is all kinda introductory stuff instead)

Investigators have understood the motion of sediment through various simplified conceptual pictures.
Often, a division of sediment motion into phases is made. 
One common scheme is to understand sediment transport as a switching between motion and resting phases. 
This reveals three processes for consideration. These are entrainment, motion, and deposition.
Of course, these are idealized categories. 
At higher transport stages, for example, there must be a breakdown of these three distinct processes. 
They are expected to blur together, or to become less well-defined.
Therefore, most mechanistic-stochastic approaches of interest for this review will be seen to hold only in the limit of low transport stage. 
This is a fundamental limitation of these approaches which we will acknowledge but not attempt to amend.

We will follow the conceptuatlization of sediment transport as emergent from the coorporation of three distinct processes, as has been assumed by many reasearchers since Einstein, certainly. 
Along the course of the review, it has become clear that many approaches have concentrated on only one of these processes, while others have attempted to integrate or convolve these processes together to arrive at a prediction (or a range of predictions) for the sediment transport rate. 
Therefore, studies of entrainment, motion, and deposition are considered separately in this review. 
Subsequently, approaches to integrate or convolve these separate processes together into a sediment transport rate are considered.

\subsection{entrainment}

The erosion process of sediment grains from the rest phase to the motion phase is the most well-studied through a mechanistic lens among entrainment, motion, and deposition.  
A majority of this research has concentrated on the determiniation of the so-called incipient motion criterion. 
Investigators have sought the flow condition corresponding to the initial motion of a sediment grain from the bed. 
Historically, investigators have differed on the flow characteristic determining initial motion. 
Shields (1936) contended entrainment was characterised by an average shear stress, and the influence of this idea has been considerable (Mongomery legend of sheilds, Gomez 1991, Montgomery 1997).  
White (1940) understood entrainment as the result of instantaneous shear stress, and this idea was later taken up by many others (Grass, 1971; Paintal 1971; Hofland 2006).
Others have considered incipient motion as controlled by the instantaeous fluid velocity (White 1940; Kalinske 1947; Einstein, 1950). 
A huge set of theories have been developed to the incipient motion of sediment grains. 
The majority of these studies were developed in a mechanistic-deterministic lens (Wiberg and Smith 1987; Bridge and Bennett 1992).
Other investigators contended that the incipient motion criterion could not be determined precisely. 
Since the 2000s, and indeed much earlier, this fact has been quite well supported by experimental evidence. 
A contemporary perspective on the incipient motion criterion problem is that sediment motion is driven by turbulent fluctuations. 
Research into this movement was initiated by the famous work of Kline (1967) revealing large, coherent, turbulent structures in the flow field of a sheared fluid. 
There were several careful investigations into the role of turbulence in sediment transport inspired by Kline's work, including Nelson 1995 and Sumer 2003. 
A large set of investigations has subsequently quantified the role of these turbulent fluctuations in the initiation of sediment motion. 
Sediment motion is best conceptualized as an extreme value approach, most similar to the thinking of Grass, Paintal, Kalinske, Einstein, and others; but in fact differing fundamentally in one regard. 
Instead of instantaneous fluid velocity or shear stress being the best characteristic of initial sediment motion, a collection of modern studies utilizing real-time fluid velocity or pressure measurements seem to indicate that impulse, a time integrated flow characteristic, along with granular geometry, appears to be the best flow indicator of bed particle incipient motion. 
This body of research into impulse was probably opened by Diplas (2008), although it must have been considered much earlier as a possibility. Diplas et al indicated, by measuring the instantaneous velocity just upstream of particles, that an incipient velocity criterion was not present. 
Sometimes, even if the upstream velocity of the fluid to a sediment grain was above the incipient motion criterion, particles would not entrain. 
They determined that an incipient fluid velocity was necessary but not sufficient for incipient sediment motion, and propoesd that instead, sediment motion was the result of a sufficient fluid impulse. 
Many other investigations have followed up on this impulse hypothesis. 
Valykyris, Celik 2010, etc
The contemporary microscale understanding of sediment entrainment most defensible in light of the considerable research into the interrelationship between fluid dynamics and particle motion, is that impulse, or a driving force integrated through time, not instantaneous fluid velocity or shear stress, is the best coorelate with the incipient motion of sediment grains. 
As this perspective has only emerged in the last ten years, one can argue that all previous entrianment models are somewhat invalidated by it within certain ranges of the parameter spaces over which they purportedly apply. 
In acknowledgement of this, we begin a historical review of mechanistic stochastic entrainment theories. 
When this review is complete, an original ammendment of some of these theories to take account of the perspective that a distribution of impulses drives seidment entrainment is presented.

\begin{enumerate}
\item Einstein 1937
\item Einstein 1950
\item White 1940 
\item Grass 1971
\item Paintal 1971
\item Gessler 1977 
\item Kirchener 1990 
\item Bridge and Bennett 1992
\item Church 1994 
\item Cheng and Chiew 1998
\item Wu and Chou 2003
\item Ma, Xu 2013
\item Papanicolaou 2002
\item Wu and Yang 2004
\item Hofland 2006
\item Valykyris 2011
\item Ling 2012 
argues that impulse based criteria are necessary to work in a grass-like approach 
\item Tregnaghi 2012

\end{enumerate}

\subsection{motion}

\begin{enumerate}
\item Furbish 2015

\end{enumerate}

\subsection{deposition}
\begin{enumerate}
\item Charru 2004 
\end{enumerate}

\subsection{transport}

\begin{enumerate}
\item Einstein 1950 
\item Paintal 1971
\item Ling 1998
\item sun donahue 2000 
\item Tsai and Yang 2001
\item Martin 2014 
\item Ancey 2006
\item Pelosi 2014
\item Sun 2015

\end{enumerate}
\end{comment}

### 2021-09-03 draft/formerPapers/grlOlderVersion/rwdiffu.tex
added phd thesis.

\documentclass[draft,grl]{agujournal2018}
%\documentclass[]{agujournal2018}
\usepackage{apacite}
\usepackage{natbib}

\usepackage{url}
\usepackage{lineno}
\usepackage{trackchanges}
\soulregister\citet7
\soulregister\citep7
\soulregister\ref7
\addeditor{JKP}

%\draftfalse
\journalname{Geophysical Research Letters}

%custom packages
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{comment}
\usepackage{booktabs}
% custom commands
\newcommand\be{\begin{equation}}
\newcommand\ee{\end{equation}} 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}
\newcommand\om{\omega}
\newcommand\tom{\tilde{\omega}}
\newcommand\tg{\tilde{g}}
\newcommand\tp{\tilde{p}}
\newcommand\tG{\tilde{G}}
\newcommand\El{\mathcal{L}}
%\usepackage{layouts}
%\printinunitsof{in}\prntlen{\textwidth} % check scales in the document
\linenumbers

% THIS IS THE LINENO PATCH TO WORK WITH AMSMATH
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
	\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
	\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
	\renewenvironment{#1}%
	{\linenomath\csname old#1\endcsname}%
	{\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
	\patchAmsMathEnvironmentForLineno{#1}%
	\patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
	\patchBothAmsMathEnvironmentsForLineno{equation}%
	\patchBothAmsMathEnvironmentsForLineno{align}%
	\patchBothAmsMathEnvironmentsForLineno{flalign}%
	\patchBothAmsMathEnvironmentsForLineno{alignat}%
	\patchBothAmsMathEnvironmentsForLineno{gather}%
	\patchBothAmsMathEnvironmentsForLineno{multline}%
}
% the patch comes from http://phaseportrait.blogspot.com/2007/08/lineno-and-amsmath-compatibility.html

\begin{document}

\title{Back to Einstein: \change{how to include}{including} burial in fluvial sediment diffusion models\remove{?}}

\authors{James K. Pierce \affil{1}and Marwan A. Hassan\affil{1}}
\affiliation{1}{Department of Geography \\University of British Columbia}
\correspondingauthor{James Kevin Pierce}{kpierce@alumni.ubc.ca}

\begin{keypoints}
\item \change{We develop a random walk model of objects in intermittent transport through an environment with traps}{We develop a random walk model of bedload tracers diffusing through a river as they gradually become buried}
\item \change{Its solution provides three ranges of diffusion, two of which are anomalous}{The model predicts three diffusion ranges as the observation time increases, one normal and two anomalous}
\item \change{We apply the model to sediment transport in rivers to clarify the scale dependence of bedload diffusion}{We relate these ranges to sediment motion, rest, and burial timescales and clarify their generating processes}

\end{keypoints}

\begin{abstract}
Sediment grains transport through gravel bed rivers in cycles of motion and rest.
When grains rest on the surface, material transported from upstream can bury them.
The surface shields buried grains from the flow, so they can be immobile for long time periods.
These immobile periods can dominate sediment diffusion characteristics.
Although investigators have recognized the impact of sediment burial on diffusion, existing models do not usually incorporate it.
In this study, we present the first random walk model incorporating sediment burial and the duration of individual sediment motions and solve it analytically.
The model predicts three sediment diffusion ranges with distinct scaling characteristics in each.
We relate the crossover times dividing these ranges to measurable transport parameters and describe each range from underlying physical processes.
\add{This work has implications for all researchers concerned with coarse material moving in river channels on grain to geomorphic scales.}
\remove{
Our developments provide new geophysical perspective on the anomalous scaling of fluvial sediment transport.}
\note{Who is the audience?}
\end{abstract}

\section{Introduction}

\add{A group of bedload tracers added to a river channel spreads apart through time due to differences in the downstream transport characteristics and conditions encountered by one grain and the next in a process called bedload diffusion.}
Many environmental problems including contaminant transport \citep[e.g.,][]{Macklin2006}, channel stability \citep[e.g.,][]{Hassan2017}, and aquatic habitat restoration \citep[e.g.,][]{Gaeuman2017} rely on our ability to predict this diffusion, so the topic has been researched intensively for over 80 years.
Diffusion is usually quantified by the time dependence of the positional variance of the group of tracers.
With the scaling $\sigma_x^2 \propto t$, the diffusion is said to be normal, since this is found in the classic problems \citep[e.g.,][]{Einstein1905,Taylor1920}.
However, with the scaling $\sigma_x^2 \propto t^\gamma$ with $\gamma \neq 1$, the diffusion is said to be anomalous \citep{Sokolov2012}. 
If $\gamma>1$, it is said to be super-diffusive; while if $\gamma <1$, it is said to be sub-diffusive \citep{Metzler2000}.
Hans Albert Einstein developed the first model of bedload diffusion to describe his comprehensive \add{flume} experiments visually tracking painted grains \remove{in a flume} \citep{Ettema2004,Einstein1937}.
\add{Interpreting individual bedload trajectories as a sequence of random steps and rests, }Einstein originally concluded \change{that bedload transport shows}{that a group of bedload tracers undergoes} normal diffusion.

\change{After Einstein}{More recently}, Nikora and coworkers \add{(hereafter ``Nikora'')} \change{recognized}{realized} coarse sediment \add{tracers} \remove{moving through river channels} can show either anomalous or normal diffusion depending on the \change{timescale of observation}{length of time they have been tracked} \citep{Nikora2001a,Nikora2002}.
\add{From numerical simulations and experimental data, Nikora discerned three scaling ranges $\sigma_x^2 \propto t^\gamma$ as the observation time increased.
They associated the first range with ``local'' timescales being less than the interval between subsequent collisions of moving grains with the bed, the second with ``intermediate'' timescales being less than the interval between successive resting periods of grains, and the third with ``global'' timescales composed of many intermediate timescales.
Nikora proposed super-diffusion in the local range, anomalous or normal diffusion in the intermediate range, and sub-diffusion in the global range. Although Nikora described local range super-diffusion as a consequence of the smoothness of bedload trajectories between subsequent collisions of moving grains with the bed, they did not completely pin down the physical mechanisms generating intermediate and global range diffusion.
In this paper, we refer to diffusion having multiple scaling ranges as being ``scale dependent'', and we set out to better understand the physical processes giving rise to local, intermediate, and global bedload diffusion ranges.}
\remove{
This is a significant issue since it implies diffusion models should be scale dependent, and it renders experimental data contingent on their observation timescales.
Nikora et al. identified three ranges of diffusion from Newtonian simulations and experimental data and termed these ranges local, intermediate, and global in order of increasing timescale.
They determined that $\sigma_x^2$ scales with a different power of time in each range, and they concluded this scaling could be anomalous or normal.}

\change{Recent studies}{Experiments} support \change{the Nikora et al.}{Nikora's conclusion} \change{that bedload diffusion shows multiple scaling ranges}{of multiple bedload diffusion scaling ranges}, but they do not \change{show}{ provide} consensus on the expected number of ranges or their scaling properties.
\add{This lack of consensus likely results from resolution issues.
For example, measurements have been conducted at high temporal resolution for short periods, sampling local and intermediate ranges} \citep[e.g.,][]{Furbish2012}; \add{moderate temporal resolution for moderate periods, sampling only the intermediate range} \citep[e.g.,][]{Einstein1937,Yano1969,Nakagawa1976}; \add{moderate temporal resolution for long periods, sampling the intermediate and global ranges} \citep[e.g.,][]{Martin2012}; \add{or at low temporal resolution for long periods, sampling only the global range} \citep[e.g.,][]{Bradley2010,Phillips2013,Bradley2017}.
\add{As a result, three ranges can be identified by patching together multiple datasets} \citep[e.g.,][]{Zhang2012}\add{, but they are not resolved by any one dataset.}

\add{
Modelling studies display similarly patchy understanding.
Using Newtonian simulations, \citet{Nikora2001a} resolved local and intermediate ranges, while \citet{Bialik2012} resolved all three ranges.
However, direct computations are impractical at field scales and provide limited physical insight, motivating alternative approaches.
One option is to generalize the random walk model of Einstein.
Einstein's model treats bedload trajectories as instantaneous steps interrupted by durations of rest which lie on statistical distributions \citep{Hassan1991}.
This derives only one range of normal diffusion }\citep[e.g.,][]{Einstein1937,Hubbell1964,Nakagawa1976}.
\add{
More recently, researchers generalized Einstein's model in a few different ways to include the interval of motion.  
\citet{Lisle1998} assumed particles moved for random intervals with a deterministic velocity, while \citet{Fan2016} described motions with random intervals and velocities dictated by an underlying Langevin equation.
\citet{Lajeunesse2018} took an alternate perspective developing identical governing equations as \citet{Lisle1998}.
By including the motion interval, these approaches provide two ranges of diffusion -- a local range super-diffusion and an intermediate range normal diffusion, but they do not provide a global range.
To model the global range, \citet{Wu2019} took a different path from Einstein. They retained the instantaneous motion assumption but included the possibility that sediment becomes buried as it rests on the bed.
By including burial, this approach also provides two ranges of diffusion -- an intermediate range normal diffusion and a global range super-diffusion, but it does not provide a local range.
Therefore, although no Einstein-type model of three ranges has been developed, these earlier works suggest the required components are (1) non-zero motion intervals and (2) sediment burial.}

In this study, we develop the first model of bedload diffusion describing local, intermediate, and global \change{ranges of diffusion}{diffusion ranges} by \add{incorporating these two components into Einstein's original model.}\remove{the original diffusion model of \citet{Einstein1937}}
Einstein was a giant in river geophysics and fostered an entire paradigm of research leveraging and generalizing his stochastic methods \citep[e.g.,][]{Hubbell1964, Yano1969, Yang1971, Gordon1972, Nakagawa1976}.
Einstein's model can be viewed as a special case of the continuous time random walk (CTRW) developed by \citet{Montroll1965} in condensed matter physics to describe the diffusion of charge carriers in solids.
\remove{Our generalization of Einstein's model has two components.
First, we include the duration and velocity of motions in place of instantaneous steps, and second, we add in the sediment burial process.}
To \change{develop this generalization}{incorporate motion intervals and sediment burial}, we leverage the multi state CTRW developed by \citet{Weiss1976, Weiss1994} that extends the CTRW of \citet{Montroll1965}.
Below, we develop the model in section \ref{sec:model} and solve it in section \ref{sec:solution}. Finally, we discuss the predictions of our model and its implications for \change{scale dependent bedload transport}{local, intermediate, and global ranges of bedload diffusion} in sections \ref{sec:discussion} and \ref{sec:conclusion}.

\section{Bedload diffusion with burial as a multi state random walk}
\label{sec:model}

Following Einstein, we describe bedload transport as a random walk between mobile and immobile states.
However, in contrast to Einstein, we carefully distinguish two states of immobility: grains can be immobile on the surface, which we refer to as the ``resting state'', or grains can be immobile within the sub-surface, which we refer to as the ``burial state''.
To model bedload transport with both of these immobile states, we construct a three-state random walk where the states are motion, rest, and burial, and we label these states as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
Our development of the governing equations for the three state walk follows \citet{Weiss1994}, and our incorporation of the sediment burial process is similar to \citet{Schmidt2007}.
In our model, times spent moving or resting on the bed surface are random variables characterized by \change{exponential}{probability} distributions, and moving grains have a constant velocity $v$.
Within the model, we consider burial to be a relatively long lasting condition that has some probability to occur when grains resting on the surface are covered by transported sediment.
The probability of burial per unit time (burial rate) is considered constant, so the probability that grains become buried increases with the time they rest.

% describe meaning of sojourns and g_i/G_i/theta_i
Our derivation hinges on the concept of a sojourn in the state $i$ \citep{Weiss1994}.
When a grain enters a state $i$ at some time $t_0$ and position $x_0$, then leaves \change{a}{the} state at some other time $t_1$ and position $x_1$, we say that the grain has completed a sojourn in the state $i$. \add{We denote }the joint probability density for a complete sojourn in the state $i$ of time $t = t_1-t_0$ and displacement $x = x_1-x_0$ \remove{is denoted}{as} $g_i(x,t).$ 
Similarly, we consider incomplete sojourns. If a grain begins a sojourn in the state $i$ at $(t_0,x_0)$ and the sojourn is still ongoing at $(x_1,t_1)$, the joint probability density to find the grain is $G_i(x,t)$.
We refer to $g_i$ and $G_i$ as the complete and incomplete propagators, since they move probability through space-time and are associated respectively with complete and incomplete sojourns.

Our target is the probability distribution $p(x,t)$ to find a grain at $x,t$ if we know it started at $(x,t)=(0,0)$; that is, if it started with the initial distribution $p(x,0)=\delta(x)$.
We denote the initial probabilities to be at rest or in motion as $\theta_1$ and $\theta_2$. Normalization requires $\theta_1+\theta_2=1$.
Our derivation has two main stages.
First, we introduce and solve for a set of joint probabilities \change{associated with}{describing} transitions of a grain from one state to another.
Second, we use these quantities to solve for the probabilities that a grain is in state $i$ having position $x$ at time $t$.
Afterward, we sum these latter distributions over all states $i$ to derive the joint distribution that a grain is in any state at $(x,t)$.

% derive omegas
Now we begin the first stage of the derivation.
Grains at rest may be trapped by burial.
For simplicity, we model burial as long lasting enough to be effectively permanent \citep[e.g.,][]{Wu2019}, and we assume grains resting on the surface can be buried with constant rate $\kappa$.
Equivalently, we could say the mean time required for a resting grain to become buried is $1/\kappa$.
From this assumption, the probability that a grain is not trapped after a time $t$ at rest obeys a survival function: $\Phi_F(t) = e^{-\kappa t}$ ($F$ is for ``free''). Likewise, the probability that it is trapped after resting for a time $t$ is the complement: $\Phi_T(t) = 1-\Phi_F(t)$ ($T$ is for ``trapped'').
We introduce $\omega_{1T}(x,t)$, $\omega_{1F}(x,t)$, and $\omega_2(x,t)$ as the joint probabilities to find a grain at $(x,t)$ having just completed a sojourn.
The subscript ${1T}$ denotes the completion of a rest sojourn due to trapping by burial, while $1F$ denotes the completion of a rest sojourn due to motion.
Similarly, the subscript $2$ denotes the completion of a motion sojourn due to resting.
We link the $\omega$'s in integral equations using an argument similar to \citet{Weiss1994}:
\begin{alignat}{2}
&\om_{1T}(x,t) &&= \theta_1\Phi_T(t)g_1(x,t) + \int_0^x dx' \int_0^t dt' \om_2(x',t')\Phi_T(t-t')g_1(x-x',t-t')\label{eq:x},\\
&\om_{1F}(x,t) &&= \theta_1\Phi_F(t)g_1(x,t) + \int_0^x dx' \int_0^t dt' \om_2(x',t') \Phi_F(t-t') g_1(x-x',t-t'),\label{eq:xy}\\
&\om_2(x,t) &&= \theta_2 g_2(x,t) + \int_0^x dx' \int_0^t dt' \om_{1F}(x',t')g_2(x-x',t-t'). \label{eq:y}
\end{alignat}
The first equation can be understood as follows: $\omega_{1T}(x,t)$ describes the probability \add{density} that a sojourn in the state $1$ ends due to trapping at $(x,t)$. This quantity has two contributions. The first contribution represents the possibility that the grain started at $(x,t)=(0,0)$ in the $i=1$ state (with probability $\theta_1$), propagated a distance $x$ and a time $t$ in the $i=1$ state (with probability density $g_1(x,t)$), was trapped (with probability $\Phi_T(t)$), and is now at $(x,t)$. 
The second contribution describes the possibility that the grain was in a motion sojourn which ended at $(x',t')$ when it came to rest. From here, it propagated from $(x',t')$ to $(x,t)$ at rest (with probability density $g_1(x-x',t-t')$) and was trapped during this sojourn (with probability $\Phi_T(t-t')$).
The reasoning is closely analogous for equations (\ref{eq:xy}) and (\ref{eq:y}).
Once the propagators are specified, we can solve (\ref{eq:x}-\ref{eq:y}) for the $\omega$'s to complete the first stage of the derivation.

% derive p's
The second stage of our derivation involves the joint probabilities of being in state regardless of whether a sojourn has just completed. We denote these by  $p_0(x,t)$ (trapped), $p_1(x,t)$ (rest), and $p_2(x,t)$ (motion), and we involve the $\omega$'s in their definition:
\begin{align}
p_0(x,t) &= \int_0^t dt' \omega_{1T}(x,t-t'), \label{eq:b}\\
p_1(x,t) &= \theta_1 G_1(x,t) + \int_0^x dx' \int_0^t dt' \omega_2(x',t')G_1(x-x',t-t'),\label{eq:a}\\
p_2(x,t) &= \theta_2 G_2(x,t) + \int_0^x dx' \int_0^t dt'  \omega_{1F}(x',t')G_2(x-x',t-t').\label{eq:z}
\end{align}
Equation (\ref{eq:b}) says that grains buried at any $(x,t)$ were trapped at $x$ from any earlier times.
The reasoning in (\ref{eq:a}-\ref{eq:z}) is the same as for (\ref{eq:x}-\ref{eq:y}) except we use the propagators for incomplete sojourns.
Equations (\ref{eq:b}-\ref{eq:z}) can be solved once the propagators are specified and the $\omega$'s are derived from (\ref{eq:x}-\ref{eq:y}).
Finally, we form the total probability density for a grain to be found at $(x,t)$ in any state.
This is simply 
\be p(x,t) = p_0(x,t) + p_1(x,t) + p_2(x,t). \label{eq:dist}\ee
This joint density was our target and is completely determined once (\ref{eq:x}-\ref{eq:z}) are solved.

\section{Specification of propagators and solution of model}
\label{sec:solution}
\add{
We assume sojourn times in the motion and surface rest states are exponentially distributed as this is suggested by experimental data }\citep[e.g.,][]{Yano1969,Einstein1937,Fathel2015, Martin2012}\add{, physical arguments }\citep[e.g.,][]{Furbish2016}\add{, and earlier modelling works} \citep[e.g.,][]{Fan2016,Wu2019}.
\add{We expect our conclusions will not be contingent on these assumptions, since all light tailed distributions are known to provide similar diffusion properties in random walks} \citep[e.g.,][]{Weiss1994,Weeks1998}.
\change{We consider sojourns in the rest state to occur for an exponentially distributed time interval given by the distribution}{We write the distribution of sojourn times in the rest state as}  $\psi_1(t) = k_1 e^{-k_1t}.$
The probability that a sojourn in this state lasts for at least a time $t$ is then $\Psi_1(t) = \int_t^\infty \psi_1(t)dt = e^{-k_1 t}$, and $1/k_1$ is the mean duration of a single rest.
Since grains do not move in the rest sojourn, the probability density that a grain is displaced by a distance $x$ in the rest sojourn in $\delta(x)$.
\change{Hence}{This means} the complete propagator for rest sojourns is $g_1(x,t) = \delta(x)\psi_1(t),$ or 
\be g_1(x,t) = \delta(x)k_1e^{-k_1t}.\label{eq:prop1} \ee
Likewise, the incomplete propagator for a rest sojourn is $G_1(x,t) = \delta(x)\Psi_1(t) = \delta(x)e^{-k_1t} = g_1(x,t)/k_1.$
The motion propagators are reasoned similarly. We consider motions to have a constant velocity $v$ and \change{exponentially distributed durations given by}{sojourn time distribution} $\psi_2(t) = k_2 e^{-k_2t},$
and $1/k_2$ is the mean duration of a single motion.
Since the velocity of motions is constant, the probability to find a grain at position $x$ in a motion sojourn is $\delta(x-vt)$, and the complete propagator for motion sojourns is 
\be g_2(x,t) = \delta(x-vt)k_2e^{-k_2t},\label{eq:prop2}\ee
while the incomplete propagator is $G_2(x,t) = g_2(x,t)/k_2$, as before.

Having defined the propagators  (\ref{eq:prop1}-\ref{eq:prop2}), we set out to solve (\ref{eq:x}-\ref{eq:z}) \change{and}{to} understand the diffusion expressed by the \change{trapping}{three state} model.
\note{Moved a major block to the supplementary information.}
\change{Using transform calculus, the joint distribution $p(x,t)$ is derived in appendix \ref{sec:appendixA}, and the moments $\bra x \ket$, $\bra x^2 \ket$, and ultimately the variance of position $\sigma_x^2(t)$ are derived in appendix \ref{sec:appendixB}.}{To keep the focus on the physics we have pushed the mathematical details into the supplementary information, where we derive the joint distribution $p(x,t)$, the moments $\bra x \ket$ and $\bra x^2 \ket$, and ultimately the variance of position $\sigma_x^2(t)$. }
With the shorthand notations $\xi = k_2 x/v$, $\tau = k_1(t-x/v)$, and $\Omega = (\kappa+k_1)/k_1$ \citep[cf.,][]{Lisle1998}, the joint distribution to find a grain at position $x$ at time $t$ \change{is}{comes out as} 
\begin{multline}
p(x,t) = \theta_1\mathcal{H}(\xi)\mathcal{H}(\tau)\Big[1-\frac{k_1}{\kappa+k_1}\Big(1-e^{-(\kappa+k_1)t}\Big)\Big]\delta(x) \\ + \frac{1}{v}e^{-\Omega \tau - \xi}\mathcal{H}(\xi)\mathcal{H}(\tau)\Big(\theta_1\Big[k_1\mathcal{I}_0\big(2\sqrt{\xi\tau}\big) + k_2\sqrt{\frac{\tau}{\xi}}\mathcal{I}_1\big(2\sqrt{\xi\tau}\big)\Big] \\ + \theta_2\Big[k_1\delta(\tau) + k_2 \mathcal{I}_0\big(2\sqrt{\xi\tau}\big)+k_1 \sqrt{\frac{\xi}{\tau}}\mathcal{I}_1\big(2\sqrt{\xi\tau}\big)\Big]\Big) \\
+ \frac{1}{v}\frac{\kappa k_2}{\kappa + k_1}e^{-\kappa \xi/(\kappa + k_1)}\mathcal{H}(\xi)\mathcal{H}(\tau)\Big[(\theta_1/\Omega)\mathcal{P}_2(\xi/\Omega,\Omega\tau) + \theta_2 \mathcal{P}_1(\xi/\Omega,\Omega\tau)\Big].
\label{eq:pdf}
\end{multline}
$\mathcal{H}$ is the Heaviside step function and we use the convention $\mathcal{H}(0)=1$.
The $\mathcal{I}_\nu$ are modified Bessel functions of the first kind, and the $\mathcal{P}_\mu$ are generalized Marcum Q-functions defined by $\mathcal{P}_\mu(x,y) = \int_0^y e^{-z-x}(z/x)^{(\mu-1)/2}\mathcal{I}_{\mu-1}(2\sqrt{xz})dz $ \citep{Temme1996}. Modified Bessel functions are common in one-dimensional diffusion problems \citep[e.g.,][]{Giddings1955, Daly2010}, and they have long been used to characterize the distribution of bedload tracers moving or resting on the surface \citep[e.g.,][]{Einstein1937,Hubbell1964, Bradley2012}. 
\begin{figure}[t]
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/pdf-plot.png}
	\caption{Joint distributions for a grain to be at position $x$ at time $t$ are displayed for the choice $k_1=0.1$, $k_2=1.0$, $v=2.0$. Grains are considered initially at rest ($\theta_1=1$, $\theta_2=0$). The solid lines are the analytical distribution in equation (\ref{eq:pdf}), while the points are simulation results to show mathematical correctness. Colors pertain to different times. Units are unspecified, since our aim is to demonstrate the general characteristics of $p(x,t)$. Panel (a) shows the case $\kappa=0$ -- the absence of burial.
		In this case, the joint distribution tends toward Gaussian at large times \citep[e.g.,][]{Einstein1937,Lisle1998}. Panel (b) shows the case when grains have rate $\kappa = 0.01$ to become buried while resting.
		Because of burial, the joint distribution tends toward a more uniform distribution than Gaussian. This shows a redistribution of probability to smaller values of $x$ due to the burial process \citep[cf.,][]{Wu2019}. The redistribution is encoded mathematically by the Marcum Q-function terms in equation  (\ref{eq:pdf}). A similar tendency is seen in field studies of tracer dispersion in gravel bed rivers \citep[e.g.,][]{Hassan1994}.}
	\label{fig:pdfs}
\end{figure}

The Marcum Q-functions are \add{new to bedload transport and therefore merit some discussion. They are }convolutions between modified Bessel functions and decaying exponentials\change{. They }{and }were originally devised in relation to radar detection theory \citep{Marcum1960}. 
Conceptually, the Q-functions emerge in our context from the sediment burial process. According to our assumptions, resting grains can become buried in some interval of time with an exponential probability. Meanwhile, the probability that grains are resting \add{on the surface} follows a modified Bessel distribution \citep[e.g.,][]{Einstein1937,Lisle1998}.
As a result, the probability that sediment is resting and becomes buried involves the convolution structure of the Marcum Q-functions.
Consistent with this interpretation, terms involving these convolutions vanish when the burial rate is taken to zero $(\kappa \rightarrow 0)$.
Figure \ref{fig:pdfs} depicts the distribution (\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep[cf.,][]{Barik2006}\add{, illustrating the correctness of our derivations}. A link to the simulation code, which includes descriptive comments, is available in the acknowledgments.
The new feature of our bedload diffusion model exhibited by the joint distribution $p(x,t)$ and missed by most earlier models is the tendency of the tracer distribution to become uniform as time increases.
This pattern is shown in field tracer studies of bedload transport \citep[e.g.,][]{Hassan1994, Ferguson2002}, but has not been addressed in many bedload diffusion models.
When the burial process is turned off, as in panel (a) of figure \ref{fig:pdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion; while the distribution impacted by sediment burial, as in panel (b) of figure \ref{fig:pdfs} becomes uniform-like, exemplifying no diffusion at all and corresponding to the eventual burial of all sediment tracers.

% solution for theta_1=1: moments
To take a closer look at the bedload diffusion, \change{the}{we derive the} first two moments \change{and the positional variance are derived in appendix \ref{sec:appendixB}.
The moments are}{ in the supplementary information. They are}
\begin{align}
\bra x(t) \ket &= A_1 e^{(b-a)t}+B_1e^{-(a+b)t}+C_1, \label{eq:mean}\\
\bra x^2(t) \ket &= A_2(t)e^{(b-a)t}+B_2(t)e^{-(a+b)t}+C_2, \label{eq:second}
\end{align}
\note{restructured}
\add{where the $A_i$, $B_i$, and $C_i$ are polynomials available in table \protect \ref{table:params}}. In these equations, $a = (\kappa + k_1+k_2)/2$ and $b = \sqrt{a^2-\kappa k_2}$ are effective rates having dimensions of inverse time.
\add{From these results the variance comes out as}
\be \sigma_x^2(t) = A(t)e^{(b-a)t} + B(t)e^{-(a+b)t} + C(t). \label{eq:var}\ee
The $A$, $B$, and $C$ are transcendental functions also available in table \ref{table:params}.
This equation represents \change{the scale dependence of bedload diffusion for sediment gradually undergoing burial.}{the diffusion of a group of bedload tracers transporting downstream through intervals of motion and rest as they gradually become buried.}
\begin{table}[!h]
	\centering
	\caption{Polynomials and transcendental functions used in the expressions of the mean (\ref{eq:mean}), second moment (\ref{eq:second}) and variance (\ref{eq:var}) of bedload tracers.}
	\label{table:params}
	\begin{tabular}{c}
		\toprule
		$\begin{aligned}[t]
		&A_1 = \frac{v}{2b}\big[\theta_2+\frac{k_1+\theta_2\kappa}{b-a}\big] \\
		&B_1 = -\frac{v}{2b}\big[\theta_2-\frac{k_1+\theta_2 \kappa}{a+b}\big] \\
		&C_1 =  -\frac{v}{2b}\big[\frac{k_1+\theta_2 \kappa}{b-a}+\frac{k_1+\theta_2 \kappa}{a+b}\big]\\
		&A_2(t) = \frac{v^2}{2b^3}\Big[(bt-1)[k_1+\theta_2(2\kappa + k_1 + b-a)]+\theta_2b \\
		&\hspace{3cm} + \frac{(\kappa+k_1)(\theta_2\kappa+k_1)}{(b-a)^2}[(bt-1)(b-a)-b]\Big]\\
		&B_2(t) = \frac{v^2}{2b^3}\Big[(bt+1)[k_1 + \theta_2(2\kappa+k_1-a-b)]+\theta_2b\\
		&\hspace{3cm} -\frac{(\kappa+k_1)(\theta_2\kappa+k_1)}{(a+b)^2}[(bt+1)(a+b)+b]\Big]\\
		&C_2 = \frac{v^2}{2b^3}(\kappa+k_1)(\theta_2 \kappa + k_1)\Big[\frac{2b-a}{(b-a)^2}+\frac{a+2b}{(a+b)^2}\Big]\\
		&A(t) = A_2(t)-2A_1C_1 - A_1^2\exp[(b-a)t]\\
		&B(t) = B_2(t)-2B_1C_1 - B_1^2\exp[-(a+b)t]\\
		&C(t) = C_2-C_1^2-2A_1B_1\exp[-2at]\\			
		\end{aligned}$\\
		\bottomrule
	\end{tabular}
\end{table}

The positional variance is plotted in figure \ref{fig:var} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
We interpret ``$\gg$'' to mean ``of at least an order of magnitude greater''.
These conditions mean that all grains are initially at rest \citep[cf.,][]{Wu2019}, motion intervals are typically much shorter than rests \citep[cf.,][]{Einstein1937}, and sediment burial requires a much longer time than typical rests.
We concentrate on these conditions in this paper as they are most relevant to bedload diffusion in gravel-bed rivers, where \change{sediment transport is typically rarefied and intermittent}{sediment transport is typically rarefied} and the burial of sediment is relatively slow compared to surface resting times \citep[e.g.,][]{Ferguson2002,Hassan1994}.
Figure \ref{fig:var} demonstrates that under these conditions the variance (\ref{eq:var}) shows three ranges of bedload diffusion each having approximate power law scaling ($\sigma_x^2 \propto t^\gamma$), followed by a fourth range with no diffusion ($\sigma_x^2 = \text{const}$). We associate the first three ranges with the local, intermediate, and global ranges proposed by \citet{Nikora2001a,Nikora2002}, and identify the fourth range as resulting from the eventual burial of all sediment \change{grains}{tracers}.
We suggest to call the fourth range geomorphic, since any further transport in this range can occur only if scour exposes buried grains to the flow \citep[cf.,][]{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
Using this terminology, we summarize when $k_2\gg k_1 \gg \kappa$ and $\theta_1=1$, the variance in equation (\ref{eq:var}) expresses \change{four}{three} scaling ranges: local, intermediate, and global.\remove{, and geomorphic.
We emphasize only the first three of these ranges are diffusive.}
\add{The diffusion terminates when all tracers become buried in the geomorphic range, corresponding to the uniform-like shape of the joint distribution $p(x,t)$ induced by the Marcum Q-functions.}
\begin{figure}[t]	
	\includegraphics[width=\linewidth,keepaspectratio]{./figures/diffusion.png}
	\caption{The variance equation (\ref{eq:var}) is plotted for the parameters $1/k_2 = 1.5$s, $1/k_1 = 30.0$s, and $v=0.1$m/s. These values are comparable to laboratory flume experiments transporting small ($\sim 5$mm) gravels \citep[cf.,][]{Lajeunesse2010,Martin2012}. The timescale of burial is set to $1/\kappa = 7200.0$s (two hours), and the initial condition is rest ($\theta_1=1$). The solid line is equation (\ref{eq:var}) while the points are \change{directly}{numerically} simulated. When $k_2\gg k_1 \gg \kappa$\change{, as is the case}{as} in this plot, there are four distinct scaling ranges of $\sigma_x^2$: local, intermediate, global, and geomorphic. Within each range, a slope key is added to demonstrate the scaling $\sigma_x^2 \propto t^\gamma$. There are three crossovers between these ranges, denoted on the figure by vertical lines. Crossover times $T_L$, $T_I$, and $T_G$ are indicated at the bottom of the plot. They are given in equations (\ref{eq:TL}-\ref{eq:TG}) in terms of the model's input parameters. }
	\label{fig:var}
\end{figure}

% WORK HERE
\section{Discussion}
\label{sec:discussion}

The model we've presented \change{involves}{predicts three diffusion ranges for a group of bedload tracers transporting downstream as they gradually become buried, providing the first analytical description of Nikora's concept.} \change{in terms of}{The model requires} four parameters. These are the characteristic velocity $v$ of moving sediment and three \remove{key}timescales: $1/k_2$, $1/k_1$, and $1/\kappa$.
The timescales represent the mean durations of motion, rest, and rest before burial occurs.
As shown in figure \ref{fig:var}, scaling laws $\sigma_x^2 \propto t^\gamma$ approximately characterize the diffusion in each scaling range.
The exponents $\gamma$ in each range result from competition between different terms in equation (\ref{eq:var}).
Between these ranges, there are crossover regions where the scaling is not a simple power law.
Because these crossover regions are relatively narrow, we can approximately characterize them with crossover times $T_L$, $T_I$, and $T_G$.
These times partition the \remove{diffusion }ranges into $0< t < T_L$ (local), $T_L < t < T_I$ (intermediate), $T_I < t < T_G$ (global), and $T_G < t$ (geomorphic). 
Figure \ref{fig:var} depicts these crossover times as vertical lines.
To understand the scale dependence expressed by the bedload variance in equation (\ref{eq:var}), we need to determine the exponents $\gamma$ of each range and relate the crossover times $T_L$, $T_I$, and $T_G$ between diffusion ranges to the model parameters $v$, $k_1$, $k_2$, and $\kappa$.

% now we figure out what the exponents are 
We determine the diffusion exponents $\gamma$ in the local, intermediate, and global ranges using two limiting cases of equation (\ref{eq:var}): (1) $t\ll 1/\kappa$, and (2) $1/k_2 \rightarrow 0$ while $vk_2 = \text{constant}$.
Limit (1) corresponds to times so short a negligible amount of sediment burial has occurred, while limit (2) corresponds to times so long a motion interval appears as an instantaneous step having mean step length $l=vk_2$.
We evaluate these limits in \change{appendix \ref{sec:appendixC} and}{the supplementary information to} obtain the diffusion exponents.
Local range diffusion has exponent $2 \leq \gamma \leq 3$ depending on the initial conditions $\theta_1$ and $\theta_2$.
When the initial conditions are pure, meaning one of the $\theta_i$ is zero, the local exponent is $\gamma=3$.
When grains start in a mixture of motion and rest states, meaning neither $\theta_i$ is zero, the local exponent is $\gamma=2$.
Intermediate range diffusion always has exponent $\gamma=1$, agreeing with classic flume experiments \citep[e.g.,][]{Einstein1937,Yano1969a,Nakagawa1976}.
Finally, global range diffusion has exponent $1 < \gamma < 3$ depending on the ratio $k_1/\kappa$.
In the extreme case $k_1/\kappa \approx 0 $, we find normal diffusion $\sigma_x^2 \propto t$. 
However, in the opposite extreme $k_1/\kappa \rightarrow \infty$, we find  super-diffusion $\sigma_x^2 \propto t^3$.
For general ratios $\kappa/k_1$, the global exponent lies between these extremes.
For example, when $k_1/\kappa \approx 240$ as in figure \ref{fig:var}, the global range exponent is $\gamma \approx 2.2$. 
We surmise when the model parameters satisfy $k_2\gg k_1 \gg \kappa$, meaning all three diffusion ranges exist, the variance (\ref{eq:var}) implies local range super-diffusion with exponent depending on the $\theta_i$, intermediate range normal diffusion with no dependence on model parameters, and global range super-diffusion with exponent depending on the ratio $k_1/\kappa$.
Finally, the model supports a geomorphic range of no diffusion ($\gamma=0$) associated with the eventual and long term burial of all sediment grains.

Each of the three crossover times relates to a physical process affecting a population of grains.
Exchange between motion and rest states induces the local/intermediate crossover, the onset of sediment burial induces the intermediate/global crossover, and the completion of sediment burial induces the global/geomorphic crossover.
Each process admits two characteristic times, and we formulate the crossover times heuristically as geometric averages of these characteristic times:
\begin{alignat}{2}
&T_L &&= \sqrt{\frac{1}{k_1}\frac{1}{k_2}}, \label{eq:TL}\\
&T_I &&= \sqrt{\frac{1}{k_1}\frac{1}{\kappa}},\label{eq:TI}\\
&T_G &&= \sqrt{\frac{1}{\kappa}\frac{k_1+k_2}{\kappa k_1}} \label{eq:TG}.
\end{alignat}
Figure \ref{fig:var} depicts these relationships as dashed vertical lines.
The times $1/k_1$ and $1/k_2$ characterize exchanges between motion and rest, while $1/k_1$ and $1/\kappa$ characterize the onset of burial originating from grains buried in their first resting sojourns.
Equations (\ref{eq:TL}) and (\ref{eq:TI}) mix these paired times.
The completion of sediment burial and its characteristic timescales originate from the last grains to bury.
In one extreme, all grains rest on the surface, meaning the last grains bury around $1/\kappa$; while in another, a fraction of grains remains in motion for a long time evading burial.
By analogy to the mobile-immobile model of \citet{Ancey2006}, we reason that a fraction $k_1/(k_1+k_2)$ of the free population will be in motion and evading burial at long timescales, and for this fraction we propose an effective trapping rate $\kappa k_1/(k_1+k_2)$.
Equation (\ref{eq:TG}) mixes the reciprocal of this effective rate with $1/\kappa$.
We tested formulas (\ref{eq:TL}-\ref{eq:TG}) for many parameter choices and conclude they adequately represent the crossover locations between diffusion regimes provided the parameters satisfy $k_2\gg k_1 \gg \kappa$ and grains start from rest ($\theta_1=1$).
However, starting grains from motion ($\theta_2=1$) widens the crossover region between local and intermediate ranges, meaning equation (\ref{eq:TL}) loses representative power; while setting relatively very large values of $k_2$ widens the crossover region between global and geomorphic ranges, meaning equation (\ref{eq:TG}) loses representative power.
Ultimately, we emphasize the representation of finite width crossover regions by crossover times is an idealization, so we propose equations (\ref{eq:TL}-\ref{eq:TG}) as heuristics. 
These relations provide useful divisions between the bedload diffusion scaling ranges.

Our \change{four}{three} range bedload diffusion model reduces to earlier works through the two simplified limits previously leveraged to extract the scaling exponents $\gamma$ and \change{explored in appendix \ref{sec:appendixC}.}{described in the supplementary information}.
Limit (1) implies the model developed by \citet{Lisle1998} to describe soil transport within a sheet flow, while limit (2) implies the model developed by \citet{Wu2019} to describe bedload transport with burial.
Either of these cases further simplifies to the classic results of \citet{Einstein1937} and his early followers \citep[e.g.,][]{Hubbell1964, Nakagawa1976, Yano1969} that predict a single range of normal diffusion.
\citet{Lisle1998} generalized the Einstein model to include a finite velocity and duration of motion in place of instantaneous steps, and they derived two ranges of diffusion -- super-diffusive and normal.
\citet{Wu2019} developed an active layer formulation of bedload transport where grains are transferred from the active layer (surface) to the substrate layer (burial) at a constant rate.
They simplified the problem by interpreting motions as instantaneous steps, and they derived two ranges of diffusion -- normal and super-diffusive.
Our model ultimately extends these two works and re-frames them in the formalism of CTRWs \citep[e.g.,][]{Weiss1994} that was implicitly applied by \citet{Einstein1937}.

We offer several implications of our model for bedload diffusion studies in gravel bed streams.
First, our model confines the valid range of scale independent models such as the advection-diffusion equation \citep[e.g.,][]{Bradley2010} and the Einstein model \citep[e.g.,][]{Martin2012} for practical applications such as contaminant transport \citep[e.g][]{Malmon2005,Macklin2006} and aquatic habitat restoration \citep[e.g.,][]{Gaeuman2017}.
We can conclude when the observation timescale satisfies $T_L<t<T_I$, with $T_L$ and $T_I$ given by equations (\ref{eq:TL}) and (\ref{eq:TI}), we expect normal bedload diffusion $\sigma_x^2 = D_d t$.
For example, if grains move for a mean duration $0.5$s, they rest for a mean duration $1$hr, and burial takes, on average $30$hr for resting grains, a normal diffusion model such as \citet{Einstein1937} would be applicable for observation times between $T_L \approx 1.2$min and $T_I \approx 5.5$hr.
Second, the model links bedload transport understanding across scales. 
In practice, we might measure sediment diffusion within a channel on one timescale with intent to apply our knowledge at smaller or larger timescales; usually, experimental limitations constrain the measurement timescale.
For example, we could determine $k_1$ and $k_2$ by measuring the virtual velocity and diffusivity of sediment tracers in the intermediate range \citep[e.g.,][]{Einstein1937,Yano1969a,Nakagawa1976}.
With an estimate of the trapping rate $\kappa$, equation (\ref{eq:var}) provides diffusion characteristics of the global range $T_I<t<T_G$ which is more difficult to study experimentally.

To describe the scale dependence of bedload diffusion, we extended the original Einstein model by building up ideas from condensed matter physics.
Along the way, we incorporated two simplifying assumptions that deserve further research attention. 
First, we treated sediment burial as a quasi permanent condition; and second, we implicitly assumed burial is the only process trapping grains in river channels.
In actuality, burial is a temporary condition linked to scour and fill of the sedimentary bed \citep{Hassan1994,Voepel2013,Martin2014,Wu2019a}, and field experiments show other processes trapping sediment.
For example, grains can become stranded on bars \citep{Ferguson2002,Bradley2017} and embedded in stabilizing structures such as steps or clusters \citep[e.g.][]{Church1998,Hassan2008}.
In principle, the multi state CTRW framework we developed here can describe bedload diffusion with multiple trapping processes by simply adding more states, and it can handle impermanent sojourns in these states by specifying propagators for them.
Unfortunately, we currently lack the experimental knowledge to specify these propagators, since trapping rates and durations are poorly understood in association with particular processes.
For example, recent experiments have attributed resting times to any immobile sediment, and this is known to imply heavy-tailed sediment resting time distributions \citep[e.g.,][]{Olinde2015,Bradley2017}.
However, \change{very few studies have}{only one study has} resolved immobile times of sediment while identifying their underlying mechanism \citep{Martin2014}.
We propose experiments to determine the rates and durations of specific bedload trapping processes are important topics for further study.
A first step might be to measure the burial rate $\kappa$ we applied in this paper; in principle, this could be discerned by timing the period required for tagged surface grains to become buried in flume experiments.
Once better understanding of the rates and durations of sediment trapping processes is in place, generalizations built from the framework we've presented in this paper should better predict bedload diffusion on the global and geomorphic scales.

\section{Conclusion}
\label{sec:conclusion}
We developed a random walk model \remove{with alternating mobile and immobile states and a possibility of trapping from the immobile state, and we used it }to describe \change{sediment transporting through a river channel as it gradually becomes buried}{a group of sediment grains transporting through a river channel as it gradually becomes buried}.
To our knowledge, this is the first analytical description of bedload diffusion \change{across}{capturing} the local, intermediate, and global timescales introduced by \citet{Nikora2001a}.
Pushing the ideas of Nikora somewhat further, we proposed a geomorphic range to describe diffusion characteristics at timescales larger than the global range.
At base level, our model demonstrates the required components for three bedload diffusion ranges: (1) the duration of sediment motions, (2) mobile-immobile switching, and (3) a sediment trapping process.
A next step is to incorporate the bed scour process that re-exposes buried sediment to better understand the global and geomorphic ranges.
Ultimately, we emphasize the multi state random walk formalism used in this paper implicitly underlies most existing bedload diffusion models, and we believe it will readily accommodate efforts to include more physical processes into Einstein's research paradigm.

\appendix

\acknowledgments
J. Pierce acknowledges a helpful exchange with Eduardo Daly during the early stages of this work. He would like to thank Melinda Saunders and Leonardo Golubovic for their careful guidance in mathematics through the years. M. Hassan is supported by an NSERC Discovery grant. The Python simulation code is available at \sloppy
\url{https://github.com/jkpierce/rw-diffu}.

\bibliography{biblio.bib}
\end{document}

### 2021-09-03 revision/ack.tex
added phd thesis.

--- 2021-07-07 ack.tex
@@
 
-Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele, you've formed all of the best parts of my personality and guided me toward this point, and it's impossible to explain how thankful I am for that.
+Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Avis, you've formed all of the best parts of my personality and guided me toward this point, and it's impossible to explain how thankful I am for that.
 Regarding family I will also always be thankful for my great aunt Alice who graciously paid for my undergraduate education.
 
-Thanks to my early teachers who put in the extra care to keep me on track, Jonathan Ramey, Damon Spurlock, Linda Mendez, Linda Afzhalirhad, Phyllis Adkins, Brandon Willard, Marty Ojeda, and Red Hamlin among many others.
+Thanks to my early teachers who put in the extra care to keep me on track, Jonathan Ramey, Linda Ball, Damon Spurlock, Linda Mendez, Linda Afzalirad, Phyllis Adkins, Brandon Willard, Marty Ojeda, and Red Hamlin among many others.
 

### 2021-09-03 revision/app_ch2.tex
added phd thesis.

--- 2021-07-07 app_ch2.tex
@@
 Here I derive the master equation \ref{eq:flippymaster} for the probability distribution of particle position from the Langevin equation \ref{eq:flippylangevin}.
-This is closely based off of the approach in \citet{Balakrishnan1993}.
+This is closely based off on the approach of \citet{Balakrishnan1993}.
 
@@
 
-Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking as time derivative of eq. \ref{eq:logi} gives
+Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of eq. \ref{eq:logi} gives
 \be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket. \label{eq:furut}\ee
@@
 Taking these transforms provides
-\be \bar{\tilde{P}}(g,s) = \frac{s + k +\varphi  Dg^2 - i g V \varphi}{s(s+k)+(Dg^2-igV)(s+k_E)}, \ee
+\be \bar{\tilde{P}}(g,s) = \frac{s + k +\varphi  Dg^2 - i g V \varphi}{s(s+k)+(Dg^2-igV)(s+k_E)},  \label{eq:prev}\ee
 where $\varphi = k_D/k$ is the probability a particle starts at rest.
@@
 The numerator terms involving $g$ are easily expressed as first and second derivatives with respect to $x$. The problem, then is to calculate the inverse Fourier integral of the denominator.
-This can be conducted by finding the zeros of the denominator, expanding in partial fractions, then applying the contour integral \citep[e.g.][]{Arfken1985}
-\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g + i c} dg = \theta(x)e^{-c x}.\ee
+The inverse Fourier transform can be calculated by expanding the denominator of eq. \ref{eq:prev} in partial fractions and applying the contour integral \citep[e.g.][]{Arfken1985}
+\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g + i c} dg = \theta(x)e^{-c x}. \label{eq:contour}\ee
 
@@
 
-Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator, the remaining transform can be evaluated from standard tables \citep{Prudnikov1992a}, eventually giving (after integration over and setting to 1 of the fake parameter),
+Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator, the remaining transform can be evaluated from standard tables \citep{Prudnikov1992a}, eventually giving (after integrating over and setting to $1$ the fake parameter),
 \begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du, \end{multline}
@@
 \be \Lambda(T) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,T).\ee
-This represents the rate at which particles starts at $x_i$, anywhere to the left of $x=0$, and manage to cross $x=0$ by time $T$.
+This represents the rate at which particles start at $x_i$, anywhere to the left of $x=0$, and manage to cross $x=0$ by time $T$.
 

### 2021-09-03 revision/app_ch3.tex
added phd thesis.

--- 2021-09-03 draft/app_ch3.tex
@@
 
-In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (\ref{eq:exp}), draws a uniform random from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in (\ref{eq:rel}).
+In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (\ref{eq:exp}), draws a uniform random number from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in (\ref{eq:rel}).
 Iterating this random number generation and selection process simulates exact realizations of the stochastic process.

### 2021-09-03 revision/app_ch5.tex
added phd thesis.

--- 2021-07-08 app_ch5.tex
@@
 
-Considering the probability $P(u,t+\delta)$ as sum over possible paths from $P(u,t)$ develops 
+Considering the probability $P(u,t+\delta t)$ as a sum over possible paths from $P(u,t)$ develops 
 \begin{align} P(u,t+\delta t) =
@@
 $k=1$ provides the mean (given in eq. \ref{eq:meanu})
-\be \langle u \rangle = \frac{\tilde{\Gamma}}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}\ee
+\be \langle u \rangle =  \frac{\tilde{\Gamma}}{1-\ve} = \frac{\Gamma}{\nu(1-\ve)} \ee
 while $k=2$ provides the second moment
@@
 \label{sec:langextremes}
-Now I demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
+Now I will demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
 The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
-Following \cite{Hall1989}, this suggests normalizing the distribution $P(u)$ using the scaled variables
+Following \cite{Hall1989}, this divergence suggests normalizing the distribution $P(u)$ using the scaled variable
 \be z = \frac{u - \bra u \ket}{\sigma_u}, \ee
-and
+giving a scaled distribution
 \be Q(z) = \sigma_u P(u).\ee
-This produces a differential equation for $Q(z)$ which is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
-Incorporating this transformation into (\ref{eq:master}) provides a ``normalized" master equation
+The resulting differential equation for $Q(z)$ is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
+Incorporating this transformation into (\ref{eq:master}) provides the ``normalized" master equation
 \begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 \end{multline}

### 2021-09-03 revision/ch1intro.tex
added phd thesis.

--- 2021-07-08 ch1intro.tex
@@
 
-Landscapes evolve when water, wind, and ice, driven by gravity, erodes down slopes produced by tectonics.
+Landscapes evolve when water, wind, and ice, under the influence of gravity and sunlight, erode along topographic gradients produced by tectonics.
 Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands.
@@
 Sediment transport in river channels is especially amenable to this quantitative approach since it is mechanistically the result of fluid and granular physics.
-Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of grains \citep{Bagnold1956}.
+Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of sediment grains \citep{Bagnold1956}.
 When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}.
@@
 At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
-Particles moving as bedload are driven downstream by a turbulent fluid flow, but the exact relationships between the fluid flow and the applied forces are not exactly known \citep{Maxey1983,Schmeeckle2007}.
+Particles moving as bedload are driven downstream by a turbulent fluid flow, but precise relationships between the fluid flow and the applied forces remain unknown \citep{Maxey1983,Schmeeckle2007}.
 Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.
@@
 Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}. 
-These buried particles cannot move again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of immobility \citep{Hassan1994,Ferguson2002}.
+These buried particles cannot move along the bed surface again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of relative immobility \citep{Hassan1994,Ferguson2002}.
 
@@
 A common technique to estimate sediment transport is to seed a stream with tracer stones and track their progress downstream \citep{Einstein1937, Takayama1965, Pretzlav2021}.
-In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate tracer velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
+In principle, tracers provide a proxy for the population of grains in a stream, so one can estimate virtual velocities of tracers, then multiply by an estimate of the number of grains available for motion to calculate the overall sediment flux \citep{Wilcock1997a,Ferguson2002}.
 In practice, challenges arise due to the distinct behavior of tracers over the local, intermediate, global, and geomorphic timescales.
-Apparently, tracer velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
-This obscures the relationship between measured tracer velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
+Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
+This obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
 
@@
 \be \langle x (t)  \rangle = k_E \ell t. \ee
-This equation indicates that in Einstein's model, sediment grains move with effective velocity $V_\text{eff} = k_E \ell$ given by the product of the entrainment rate and mean step length.
+This equation indicates that in Einstein's model, sediment grains move with virtual velocity $V_\text{virt} = k_E \ell$ given by the product of the single-particle entrainment rate $k_E$ and mean step length $\ell$.
 
@@
 
-The advent of high speed camera experiments produced new insight into the local and intermediate ranges of bedload motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} never intended for his model to describe.
+The advent of high speed camera experiments produced new insight into the local and intermediate ranges of bedload motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} probably did not intend for his model to describe.
 In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
@@
 Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep{Cundall1979a, Haff1993}.
-The early works utilizing this approach used a two dimensional domain with a highly simplified ``slab" flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{McEwan2001,Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+The early works utilizing this approach used a two dimensional domain with a highly simplified ``slab" flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
 
 The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow.
-The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations, while particles provide the boundary condition for the flow \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
+Particle-particle interactions are represented with the discrete element method, where particles are represented as spheres \citep{Cudnall1979} or polytopes \citep{Wachs2019} while forces are phrased as idealized spring-dashpot models.
+The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
+Particles are represented either as spheres coupled to the flow \citep{}, 
+ while particles provide the boundary condition for the flow \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
 These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
@@
 
-Another category of models have described bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.
+Another category of models describe bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.
 
@@
 This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
-The ``Fokker-Planck equation" governing the probability distribution $P(u,t)$ of the particle velocity can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007} 
+The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007}
 \be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
@@
 Here, the flux is evaluated as a sum over all downstream velocities of particles within the volume (of which there are a fluctuating number), and the division by the downstream length of the control volume is incorporated to count only that proportion of particles near the downstream boundary of the volume.
-These two definitions, $q$ defined with a surface and $\Phi$ defined with a volume, are generally not equivalent \citep{Ancey2020a}.
+These two definitions, $q$ defined with a surface and $\Phi$ defined with a volume, are generally not equivalent \citep{Ballio2014}.
 
@@
 A distinguishing feature of Bagnold's formula is its derivation from mechanistic reasoning, although many of Bagnold's assumptions have since turned out to be incorrect.
-Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and the flow power dissipated by sediment transport shows only a weak correlation the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
+Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and the flow power dissipated by sediment transport shows only a weak correlation with the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
 These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$ and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
@@
 \citet{Einstein1937} formulated his original model of individual particle trajectories in terms of entrainment and deposition rates using the idealization of grains moving downstream through a sequence of instantaneous steps.
-Later, \citet{Einstein1942,Einstein1950} calculated the sediment flux with these same probabilistic ideas, providing an alternative to Bagnold's scaling approach.
+Later, Einstein calculated the sediment flux with these same probabilistic ideas \citep{Einstein1942, Einstein1950}.
+This formulation of the sediment flux would come back to later provide an alternative to Bagnold's scaling approach.
 The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
@@
 \be q = E \ell, \ee
-where the quantity $E = \rho k_E$ is differentthe ``areal entrainment rate", representing the number of grains entrained per unit streambed area \citep{Wilcock1997a,Furbish2012}.
+where the quantity $E = \rho k_E$ is the ``areal entrainment rate", representing the number of grains entrained per unit streambed area \citep{Wilcock1997a,Furbish2012}.
 
@@
 
-Equation \ref{eq:charru} indicates that the mean flux in steady transport conditions is the implicit solution to the equation $E = D.$
+Equation \ref{eq:charru} indicates that the mean flux in steady transport and flow conditions is the implicit solution to the equation $E = D.$
 Using the scaling relations \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
@@
 \be q = \rho_s k_E \int_0^\infty dx' \int_0^\infty dt \theta(Vt-x)k_D\exp(-k_D t) = \rho_s k_E V/k_D. \label{eq:furbflux}\ee
-Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and qquation \ref{eq:furbflux} provides another perspective on Einstein's central result $q= E\ell$, but with the added ability to describe unsteady conditions \citep{Furbish2012}.
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and equation \ref{eq:furbflux} provides another perspective on Einstein's central result $q= E\ell$, but with the added ability to describe unsteady conditions \citep{Furbish2012}.
 
-\subsection{Landscape evolution}
+\subsection{Channel evolution}
 \label{sec:landscape}
@@
 \be \partial_x q(x,t) = E - D. \ee 
-In combination with equation \ref{eq:exner}, this expression of the sediment flux implies \citep{Parker2000,Furbish2012,Fathel2015,Furbish2017} by which topographic evolution can be described with Einstein's rates:
-\be (1-\phi) \px z(x,t) = D - E.\ee
+In combination with equation \ref{eq:exner}, this expression of the sediment flux gradient implies the relation
+\be (1-\phi) \px z(x,t) = D - E\ee
+by which topographic evolution can be described in terms of Einstein's rates.
 In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the eventual result of entrainment from all upstream locations, giving
 \be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
-This ``entrainment form of the Exner equation" \citep{Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
+This ``entrainment form of the Exner equation" \citep{Nakagawa1976, Parker2000, Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
 
-The major limitation of these approaches are that they assume landscapes are continuous.
+The major limitation of these approaches are that they assume land surfaces are continuous.
 This can be interpreted as an average over the detailed configurations of individual grains \citep{Coleman2009}.
-This averaging process is expected to obscure the relevant landscape evolution processes whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
+This averaging process is expected to obscure the relevant morphological evolution processes whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
 
@@
 At moderate timescales they emerge from waves of moving grains \citep{Heyman2014} or the episodic entrainment of clusters of grains \citep{Strom2004,Papanicolaou2018}.
-At the longest timescales, they originate from migrating bedforms \citep{Guala2014} or cycles of aggradation and degradation in pools \citep{Dhont2018}.
+At the longest timescales, they originate from migrating bedforms \citep{Guala2014}, cycles of aggradation and degradation in pools \citep{Dhont2018}.
 These mechanisms generally occur in conjunction with other sources of bedload transport fluctuations like grain size sorting \citep{Iseya1987,Cudden2003}, flow discharge variations \citep{Wong2006,Mao2012,Redolfi2018}, and sediment supply perturbations \citep{Lisle1993,Madej2009,Elgueta-Astaburuaga2019}.
@@
 Recognition that sediment transport rates fluctuate requires careful interpretation of sediment transport measurements. 
-In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Chartrand2018}, impact sensors \citep{Rickenmann2007}, computer vision techniques \citep{Roseberry2012}, or sediment traps \citep{Papangelakis2016}.
+In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Zimmermann2008, Chartrand2018}, impact sensors \citep{Rickenmann2007, Mendes2016}, computer vision techniques \citep{Roseberry2012, Ancey2014}, or sediment traps \citep{Bunte2004, Papangelakis2016}.
 
@@
 A fundamental question raised by scale-dependence is how large averaging windows must be for sediment flux measurements to converge.
-To date, very few mathematical models have addressed this issue.
+To date, very few mathematical models have addressed this issue \citep[e.g.][]{Ancey2020a}.
 
@@
 
-Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where arrivals of moving particles to the volume is ``birth" and departure is ``death".
+Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where arrivals of moving particles to the volume are ``births" and departures are ``death".
 They formulated the probability distribution $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
@@
 
-From eq. \ref{eq:negbin}, the mean number of moving particles is $\langle n \rangle = \lambda/(\alpha-\mu)$, and the variance is
+From eq. \ref{eq:negbin}, the mean number of moving particles is 
+\be \langle n \rangle = \lambda/(\alpha-\mu),\ee
+ and the variance is
 \be \sigma_n^2 = \frac{\lambda \alpha}{(\alpha-\mu)^2}. \ee
-Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\alpha/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, as the distribution \ref{eq:negbin} limits to a Poisson distribution where the mean and variance are equal \citep{Ancey2006}.
+Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\alpha/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, as the distribution (\ref{eq:negbin}) limits to a Poisson distribution where the mean and variance are equal \citep{Ancey2006}.
 
@@
 A limitation of this approach is that in reality, entrainment and deposition modify the bed elevation, which in turn modifies the rates of entrainment and deposition \citep{Sawai1987,Wong2007}. 
-This highlights a need to evaluate bed elevation change in birth-death models.
+This highlights a need to evaluate bed elevation change in statistical sediment transport models.
 
@@
 The original model by \citet{Einstein1937} described particle trajectories as a sequence of instantaneous steps.
-This model was subsequently improved include a finite duration of motion at constant velocity \citep{Lisle1998,Lajeunesse2017}.
+This model was subsequently improved to include a finite duration of motion at constant velocity \citep{Lisle1998,Lajeunesse2017}.
 
@@
 
-Modelling sediment transport as an alterating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface.
-Although this provides a useful description of bedload transport across local and intermediate timescales, it requires modification to describe for global and geomorphic timescales.
+Modelling sediment transport as an alternating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface.
+Although this provides a useful description of bedload transport across local and intermediate timescales, it requires modification to describe longer global and geomorphic timescales.
 At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017}, but burial has not yet been incorporated into Einstein-type models of particle trajectories.
 
-Chapter \ref{ch:downDiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to include sediment motion, rest, and burial.
+Chapter \ref{ch:downDiff} generalizes the model of \citet{Lisle1998} and \citet{Lajeunesse2017} to incorporate a new sediment burial state as suggested by the observations of \citet{Bradley2017}.
+The resulting model evaluates sediment trajectories including sediment motion, rest, and burial.
 This provides the first description of sediment trajectories over local, intermediate, global, and geomorphic timescales.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \endinput
@@
 \subsection{First extensions of the Einstein model}
+\label{
 

### 2021-09-03 revision/ch2flipFlop.tex
added phd thesis.

--- 2021-07-08 ch2flipFlop.tex
@@
 Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, and particles are once again entrained.
-Bed load transport is thus a kind of itinerant motion, characterized by alternation between fluctuating movements and periods of rest.
-This process is difficult to describe mathematically given the technicality of the required physics.
+Bed load transport is thus a kind of intermittent motion, characterized by alternation between fluctuating movements and periods of rest.
 
 To date, descriptions of bed load transport have therefore simplified the problem in various ways to enable progress.
-The foundational work is due to Einstein, who considering bed load motions as instantaneous so he could describe bed load transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
-Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E \ell$, where $k_E$ is the rate at which an individual bed particle entrains into motion, and $\ell$ is the mean length of each downstream step.
+The foundational work is due to Einstein, who considered bed load motions as instantaneous so he could describe bed load transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
+Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E \ell$, where $k_E$ is the rate at which the individual bed particle is entrained into motion, and $\ell$ is the mean length of each downstream step.
 Later, Einstein applied these ideas to calculate the mean downstream flux due to the movements of many such particles \citep{Einstein1950}.
@@
 
-Parallel to these efforts to better describe individual particle trajectories, a complementary set of studies have developed stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Turowski2010,Furbish2012a,Ancey2020}.
+Parallel to these efforts to better describe individual particle trajectories, a complementary set of studies produced stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Turowski2010,Furbish2012a,Ancey2020}.
 Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006,Furbish2012a}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
-Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
+Because these fluctuations occur over disparate time scales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
 To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
-Among those that do, none have yet formulated the sediment flux in terms of the downstream trajectories of individual grains.
+Among those that do include scale dependence, none formulate the sediment flux in terms of the downstream trajectories of individual grains.
 
@@
 
-The Langevin equation \ref{eq:flippylangevin} is a generalization of the model by \citet{Lisle1998} and \citet{Lajeunesse2017}, presented in section \ref{sec:lisle}, to include velocity fluctuations in the motion stateprotocol.
-It is similar to the ``randomly flashing diffusion" developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}, although it is not clear if these models relate mathematically.
+The Langevin equation \ref{eq:flippylangevin} is a generalization of the model by \citet{Lisle1998} and \citet{Lajeunesse2017}, presented in section \ref{sec:lisle}, to include velocity fluctuations in the motion state.
+It is conceptually similar to the ``randomly flashing diffusion" models developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}.
 Equation \ref{eq:flippylangevin} describes the downstream movement of particles alternating through motion and rest with velocity fluctuations in the motion state.
@@
 This generalizes the nonlocal formulation (\ref{sec:nonlocal}) to provide the probability distribution of the flux, and it reframes the renewal approach (sec.
-\ref{sec:renewal}) to in terms of the mechanics of individual particles in transport.
+\ref{sec:renewal}) in terms of the mechanics of individual particles in transport.
 
@@
 	\caption{The left panel indicates the configuration for the flux.
-The particle trajectories within are calculated from equation \ref{eq:langevin}, demonstrating alternation between rest and motion with fluctuating velocity.
+The particle trajectories within demonstrate alternation between rest and motion with fluctuating velocity.
 Particles begin their transport with positions $-L\leq x \leq 0$ at $t=0$, and as depicted in the right panel, the flux is calculated with the number of particles $N_>(T)$ which lie to the right of $x=0$ at the observation time $t=T$, divided by $T$: $q(T) = N_>(T)/T$.
@@
 \ref{eq:diffo}).
-At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process.
-For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
+At time scales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process.
+For larger time scales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
 Finally at times longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
@@
  \end{multline}
-where $\varphi=k_D/k$ is the probability the particle starts at rest.
+where $\varphi=k_D/k$ is the probability the particle starts at rest and $\mathcal{I}_0$ is a modified Bessel function.
 This equation generalizes the earlier result  \ref{eq:lisledist} for alternation between motions and rests to include velocity fluctuations within the motion state.
@@
 The distribution \ref{eq:monster} is shown evolving through time in figure \ref{fig:flippyfig1} panel (a), where the advection and diffusion characteristics of equation \ref{eq:flippymaster} are both evident.
-The integral in equation \ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the Bessel probability that the particle has been in motion for a period $u$ out of a time $t$ with the Gaussian probability that a particle has travelled a distance $x$ in time $u$ within the motion state.
+The integral in equation \ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the Bessel function probability that the particle has been in motion for a period $u$ out of a time $t$ with the Gaussian probability that a particle has travelled a distance $x$ in time $u$ within the motion state.
 The prefactors of this integral term can be understood as adapting this distribution to the initial conditions.
@@
 Poisson distributions have relatively thin tails, meaning their fluctuations are typically small \citep{Ancey2006}.
-In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
+In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux time series \citep{Singh2009, Heyman2013,Benavides2021}.
 This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider transport rate fluctuations.
@@
 With wide sediment velocity fluctuations and long particle movement periods, the required observation timescale could become large.
-In this case, mean sediment flux timeseries obtained with light tables and acoustic sensors \citep{Zimmermann2008,Tsakiris2014,Mendes2016} would require careful interpretation.
+In this case, mean sediment flux time series obtained with light tables and acoustic sensors \citep{Zimmermann2008,Tsakiris2014,Mendes2016} would require careful interpretation.
 
@@
 
-It has long been recognized that many processes lend variability to the sediment flux, like bedform migration \citep{Hamamori1962,Guala2014}, particle sorting processes \citep{}, or entrainment of clusters \citep{Strom2004,Papanicolaou2018}.
+It has long been recognized that many processes lend variability to the sediment flux, including bedform migration \citep{Hamamori1962,Guala2014} and entrainment of clusters \citep{Strom2004,Papanicolaou2018}.
 Understanding these processes has been a challenging problem, in part because different sources of fluctuations mix within measured transport signals \citep[e.g.][]{Hoey1992,Singh2009,Saletti2015,Dhont2018}.
@@
 An attempt to subtract the signature of individual particle dynamics from spectra of bedload transport may be a useful step to understand more complex morphological sources of sediment transport variability in channels.
-A comprehensive understanding of the linkage between individual particle dynamics and sediment fluxes, as I have worked toward here, would provide a foundation for such a subtraction process.
+A comprehensive understanding of the linkage between individual particle dynamics and sediment fluxes, as I have worked toward here, provides the foundation for such a subtraction process.
 
@@
 The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying dynamics of individual particles.
-The observation timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations, characterized by a Peclet number.
+The observation time scales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations, characterized by a Peclet number.
 

### 2021-09-03 revision/ch3bedEle.tex
added phd thesis.

--- 2021-09-03 draft/ch3bedEle.tex
@@
 This model extends the birth-death model approach reviewed in section \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
-This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried in the bed subsurface.
+This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried under the bed surface.
 The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
@@
 Using this method, I simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
-These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
+These values include the range exhibited by the available experimental data on bed elevation time series \citep{Wong2007,Singh2009,Martin2014}.
 
@@
 $M_0$ is a normalization constant. 
-This closure equation approach shows good coorespondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.
+This closure equation approach shows good correspondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.
 
@@
 The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
-These results indicate that bedload statistics measurements on short timescales could be severely biased by departures from the mean bed elevation.}
+These results indicate that bedload statistics measurements on short time scales could be severely biased by departures from the mean bed elevation.}
 	\label{fig:condmoms}
@@
 Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
-The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears to be crude approximation, as it does not capture the asymmetry.
+The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears as a crude approximation since it does not capture the asymmetry.
 Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
@@
 \begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
-A representative subset of these results are displayed in figure \ref{fig:cdfs}.
+Figure \ref{fig:cdfs} displays a representative subset of these results.
 Comparing panels \ref{fig:cdfs}(a) and \ref{fig:cdfs}(c) shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
-However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic timescale $T_0$ is found to collapse away both variations.
+However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic time scale $T_0$ is found to collapse away both variations.
 We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
-This is the mean number of grains leaving the bed per unit time is $E$, and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}). 
-Therefore, bed elevations change with a characteristic speed $v = z_1 E$.
+Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}), bed elevations change with a characteristic speed $v = z_1 E$.
 Since the range of elevation deviations is $l$ (figure \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
@@
 Finally, the model allowed resolution of resting times for sediment undergoing burial within the sedimentary bed.
-These timescales were obtained by analyzing return times from above in the bed elevation timeseries.
-This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
-The power-law tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.
+These time scales were obtained by analyzing return times from above in the bed elevation time series.
+This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long time scales.
+The power-law tails were found to collapse across flow conditions using a time scale formed from the mean erosion rate and the active layer depth.
 
@@
 \end{enumerate}
-These results imply measurements of bedload transport statistics could be biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
+These results imply measurements of bedload transport statistics could be biased at observation time scales smaller than adjustments of the bed elevation time series when collective entrainment occurs.
 The next step is to generalize the model in this chapter to a multi-cell framework to study channel morphodynamics in a stochastic framework.

### 2021-09-03 revision/ch4downDiff.tex
added phd thesis.

--- 2021-09-03 draft/ch4downDiff.tex
@@
 For example, experiments have tracked only moving grains, resolving the local range \citep{Furbish2012,Furbish2017,Fathel2016}; grains resting on the bed surface between movements, resolving the intermediate range \citep{Einstein1937,Yano1969,Nakagawa1976}; grains either moving or resting on the bed surface, likely resolving local and intermediate ranges \citep{Martin2012}; or grains resting on the surface after floods, likely resolving the global range \citep{Phillips2013,Bradley2017}.
-At long timescales, a significant fraction of tracers bury under the bed surface \citep{Hassan1991,Hassan2013,Ferguson2002a,Haschenburger2013,Papangelakis2016}, meaning burial dominates long term diffusion characteristics \citep{Bradley2017,Martin2014,Voepel2013}, possibly at global or even longer ``geomorphic'' timescales \citep{Hassan2017} than Nikora et al.
+At long timescales, a significant fraction of tracers become buried under the bed surface \citep{Hassan1991,Hassan2013,Ferguson2002a,Haschenburger2013,Papangelakis2016}, meaning burial dominates long term diffusion characteristics \citep{Bradley2017,Martin2014,Voepel2013}, possibly at global or even longer ``geomorphic'' timescales \citep{Hassan2017} than Nikora et al.
 originally considered.
@@
 With these assumptions, the governing equations for the set of probabilities $\om_{ij}(x,t)$ that a transition occurs from state $i$ to state $j$ at position $x$ and time $t$ can be derived using the statistical physics approach to multi-state random walks \citep{Weiss1994,Schmidt2007,Weeks1998}.
-Denoting by $g_{ij}(x,t)$ the probability for a particle to displace by $x$ in a time $t$ within the state $i$ before it transitions to the state $j$, the transition probabilities $\om_{ij}(x,t)$ sum over all possible paths to the state $i$ from previous locations and times:
+Denoting by $g_{ij}(x,t)$ the probability for a particle to move a distance $x$ in a time $t$ within the state $i$ before it transitions to the state $j$, the transition probabilities $\om_{ij}(x,t)$ sum over all possible paths to the state $i$ from previous locations and times:
 \be \om_{ij}(x,t) = \theta_i g_{ij}(x,t) + \sum_{k=0}^2 \int_0^x dx' \int_0^t dt' \om_{ki}(x',t')g_{ij}(x-x',t-t').\label{eq:g1}\ee
-Defining another set of probabilities $G_i(x,t)$ that a particle displaces by a distance $x$ in a time $t$ within the state $i$ and possibly remains within the state, a similar sums over paths for the probabilities to be in the state $i$ at $x$, $t$ produces:
+Defining another set of probabilities $G_i(x,t)$ that a particle moves by a distance $x$ in a time $t$ within the state $i$ and possibly remains within the state, a similar sums over paths for the probabilities to be in the state $i$ at $x$, $t$ produces:
 \be p_i(x,t) = \theta_i G_i(x,t) + \sum_{k=0}^2 \int_0^x dx' \int_0^t dt' \om_{ki}(x',t')G_i(x-x',t-t').\label{eq:g2}\ee
@@
 These distributions can be constructed from the assumptions described in section \ref{sec:assumptions}.
-Since particles resting on the bed surface bury in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, these assumptions obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$.
+Since particles resting on the bed surface become buried in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, these assumptions obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$.
 Since motions have velocity $v$ for times distributed as $\psi_2(t)$, the assumptions provide $g_{21}(x,t) = \delta(x-vt)k_2e^{-k_2 t}$.
@@
 Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
-Following \citet{Hassan2017} I suggest to call the fourth range geomorphic, since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
+Following \citet{Hassan2017} I suggest to call the fourth range "geomorphic", since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
 
@@
 Two limiting cases of equation (\ref{eq:var}) provide the scaling exponents $\gamma$ of the diffusion $\sigma_x^2 \propto t^\gamma$ in each range.
-Limit (1) represents times so short a negligible amount of sediment burial has occurred, $t\ll 1/\kappa$, while limit (2) represents times so long motion intervals appear as instantaneous steps of mean length $l=v/k_2$, $1/k_2 \rightarrow 0$ while $v/k_2 = \text{constant}$.
+Limit (1) represents times so short that a negligible amount of sediment burial has occurred, $t\ll 1/\kappa$, while limit (2) represents times so long motion intervals appear as instantaneous steps of mean length $l=v/k_2$, $1/k_2 \rightarrow 0$ while $v/k_2 = \text{constant}$.
 Limit (1) provides local exponent $2 \leq \gamma \leq 3$ depending on the initial conditions $\theta_i$, and intermediate exponent $\gamma=1$.
@@
 
-This chapter extended \citet{Einstein1937} by including motion, rest, and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010}, before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
+This chapter has extended \citet{Einstein1937} by including motion, rest, and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010}, before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
 \citet{Nikora2002} highlighted the need for such a physical description, although they suggested to use a two-state random walk between motion and rest states with heavy-tailed resting times, and they did not discuss sediment burial.
@@
 A next step is to incorporate exhumation to better understand the geomorphic range.
-Ultimately, the multi-state random walk formalism used in this chapter has been demonstrated to implicitly underlies most existing bedload diffusion models.
+Ultimately, the multi-state random walk formalism used in this chapter has been demonstrated to implicitly underlie most existing bedload diffusion models.
 The random walk formalism is an alternative perspective on the Langevin approaches used elsewhere in this thesis, and it provides a powerful tool for researchers targeting landscape-scale understanding from statistical concepts of the underlying grain-scale dynamics.

### 2021-09-03 revision/ch5langevin.tex
added phd thesis.

--- 2021-09-03 draft/ch5langevin.tex
@@
 These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were viscous.
-A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
+A second set of studies show Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
 In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and super-critical.
@@
 \ref{sec:langevin}).
-\citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces were lumped into a velocity-dependent force, while Gaussian white noise represents fluctuations.
+\citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces into a simplified velocity-dependent force, while Gaussian white noise represents fluctuations.
 Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
@@
 Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces.
-In contrast to the computational physics approach, I do not aim to characterize the exact timeseries of the forces on an individual particle.
-Instead, I model the ensemble of possible force timeseries that particles could conceivably experience.
-Each possibility implies a different possible velocity timeseries $u(t)$ in the downstream direction.
+In contrast to the computational physics approach, I do not aim to characterize the exact time series of the forces on an individual particle.
+Instead, I model the ensemble of possible force time series that particles could conceivably experience.
+Each force time series realization implies a different velocity time series $u(t)$ in the downstream direction.
 The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
@@
 \subsection{Turbulent fluid forces}
-Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Re_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
-These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
-At realistic $\Re_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
-A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
-\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
+Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Rey_p = d V/\lambda$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\lambda$.
+These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Rey_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
+At realistic $\Rey_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Rey_p$ formulas \citep{Schmeeckle2007,Clift1978}.
+A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Rey_p$ can be written $F_D = \frac{\pi}{8}
+\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
 In this study, I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.
@@
 These ideas allow two major simplifications of the above drag force.
-First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is described by a diffusivity $D$ \citep{Fan2014,Ancey2014}.
+First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled as a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is characterized by a diffusivity $D$ \citep{Fan2014,Ancey2014}.
 Defining $\bar{V}$ as a representative slip velocity to be specified more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, the fluid drag can be written
@@
 Mathematically, equation \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
-Collisions introduce ``jumps" in velocity while turbulent generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
+Collisions introduce ``jumps" in velocity while turbulence generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
 
@@
 One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
-Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983} since.
+Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
 Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
 
-\subsection{Chapman-Komogorov equation and particle-bed collision integral}
+\subsection{Chapman-Kolmogorov equation and particle-bed collision integral}
 
@@
 The term
-\be \mathcal{I}_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
+\be I_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
 is a ``collision integral" term representing particle-bed collisions.
@@
 This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}.
-Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
-Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $m u$ \citep[c.f.][]{Feller1968}.
+Physically, it corresponds to binary collisions between particles having different masses and random restitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
+Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $(m u)$ \citep[cf.][]{Feller1968}.
 
@@
 \be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
-where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
+where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the modal elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
 
@@
 This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
-Such equations seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
+Such equations have seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.
 
@@
 The factors $\lambda_\pm$ are defined in appendix equations \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
-The normalization factors are $K_\pm$ are 
+The normalization factors $K_\pm$ are 
 \be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
@@
 Despite its complex appearance, simple Gaussian and exponential forms derive from this equation as exact mathematical limits.
-When particle-bed collisions are completely inelastic ($\ne = 0$), \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\nu = 1$), \ref{eq:steadystate} becomes Gaussian.
+When particle-bed collisions are completely inelastic ($\ve = 0$), \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\ve = 1$), \ref{eq:steadystate} becomes Gaussian.
 Figure \ref{fig:fig3} demonstrates in detail the approach of the distribution toward these limits.
@@
 The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see.
-Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for that with $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
+Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for the one at $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
 \be P(u) = \frac{d}{\sqrt{\gamma^2 + 4d}}e^{\frac{\gamma u }{2 d} - \frac{\sqrt{\gamma^2 + 4 d}|u|}{2d}}. \ee
@@
 Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise.
-Even though the non-Gaussian and history-dependent aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+Even though the non-Gaussian and history-dependent aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the time scales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
 
 The model developed in this chapter described particle-bed collision forces as a sequence of instantaneous impulses in an approach that is reminiscent of the kinetic theory of gases \citep{Landau1969}.
-The effect of each collision on the streamwise particle velocities was described by a simple elasticity-like coefficient.
+The effect of each collision on the streamwise particle velocity was described by a simple elasticity-like coefficient.
 Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the elasticity coefficient introduced here is not equivalent to the ``coefficient of restitution" typically applied in granular gas theory.
-The coefficient of resitution is defined normal to the contact plane of two particles during a collision, and it characterizes energy loss due to deformation of the colliding particles in the absence of a viscous fluid \citep{Brach1992,Ismail2008}.
+The coefficient of restitution is defined normal to the contact plane of two particles during a collision, and it characterizes energy loss due to deformation of the colliding particles in the absence of a viscous fluid \citep{Brach1992,Ismail2008}.
 
 In bedload transport, colliding particles are submerged in a viscous fluid, and this introduces additional damping processes besides particle deformation \citep{Joseph2001,Yang2006,Schmeeckle2001}.
-The model developed here did not consider these viscous forces, nor did it explicitly model the geometry of collisions, as $\ve$ was defined as a parameter applying to the downstream velocity only.
+The model developed here does not consider these viscous forces, nor does it explicitly model the geometry of collisions, as $\ve$ was defined as a parameter applying to the downstream velocity only.
 Thus, although I have included episodic particle-bed collisions in a stochastic sediment transport model for the first time, the key parameter ($\ve$) in this formulation has a rather heuristic character and is not clearly related to the coefficient of restitution from granular gas theory.
@@
 Particle size enters the Langevin model \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
-The analytical velocity distribution \ref{eq:steadystate} was fit to six different experimental datasets in figure \ref{fig:fig4ch5}, and the best fit parameters required to for these fits were presented in table \ref{tab:calib}.
+The analytical velocity distribution \ref{eq:steadystate} was fit to six different experimental datasets in figure \ref{fig:fig4ch5}, and the best fit parameters were presented in table \ref{tab:calib}.
 These fit parameters suggest collisions are generally less elastic for smaller particle sizes, whereas they are more elastic for larger particle sizes.
@@
 This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
-The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the full range of bedload velocity distributions that have been reported in experimental studies.
+The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the full range of bedload velocity distributions reported in experimental studies.
 

### 2021-09-03 revision/ch6conc.tex
added phd thesis.

--- 2021-07-08 ch6conc.tex
@@
 \ref{ch:flux}).
-The same stochastic strategy should be applicable to a wide range of problems in geomorphology where phenomena can built up from component parts with apparently noisy characteristics.
+The same stochastic strategy should be applicable to a wide range of problems in geomorphology where phenomena can be built up from component parts with apparently noisy characteristics.
 
@@
 A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modelling are constrained.
-The noises used in this thesis include white Gaussian, Poisson, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
+The noises used in this thesis include white Gaussian, Poisson (shot), and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
 Turbulence was described using Gaussian noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise.
@@
 White noises are an idealization which is probably never realized in nature \citep{Gardiner1983,Kubo1978}.
-In contrast, colored noises in which fluctuations have favoured frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in context of fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
+In contrast, colored noises in which fluctuations have favoured frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
 In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
@@
 Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (sec.
-\ref{sec:einflux}), and an emerging body of research has producing the full probability distribution of the flux, but without referencing individual movement characteristics (secs.
+\ref{sec:einflux}), and an emerging body of research has produced the full probability distribution of the flux without referencing individual movement characteristics (secs.
 \ref{sec:birthdeath}-\ref{sec:renewal}).
@@
 
-Fourth, chapter \ref{ch:downDiff} presenting a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
+Fourth, chapter \ref{ch:downDiff} presents a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (sec.
+\ref{sec:trajmodels}), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
 
@@
 
-\subsection{Landscape dynamics and channel morphology}
+\subsection{Channel dynamics and morphology}
 
 The first limitation concerns the assumption, implicit in every chapter of this thesis, that sediment transport characteristics are steady and uniform in time and space.
-This assumption contrasts with conditions in real gravel-bed rivers, where riffles, bars, and steps coordinate the movements of individual grains \citep{Ashmore1998,McDowell2020}, sediment is supplied in episodic bursts from mass movements \citep{Benda1990, Muller2018}, woody debris coordinates sediment movement \citep{Eaton2012,Reid2019}, and variable flow conditions modify bed texture and sediment availability \citep{Mao2012,Phillips2018}.
+This assumption contrasts with conditions in real gravel-bed rivers, where riffles, bars, and steps coordinate the movements of individual grains \citep{Ashmore1998,McDowell2020}, sediment is supplied in episodic bursts from mass movements \citep{Benda1990, Muller2018}, woody debris directs sediment accumulation \citep{Eaton2012,Reid2019}, and variable flow conditions modify bed texture and sediment availability \citep{Mao2012,Phillips2018}.
 
@@
 The second major limitation of this work concerns grain size.
-In actuality, sediments in rivers spans a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.
+In actuality, sediments in rivers span a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.
 
@@
 Earth's landforms are shaped by numerous transport phenomena from booming debris flows \citep{Iverson1997} to barely perceptible hillslope creep \citep{Deshpande2021}.
-Different phenomena are basically distinguished by the relative importance of fluid, granular, and gravitational forces in sustaining them \citep{Jerolmack2019}.
+Different transport phenomena are basically distinguished by the relative importance of fluid, granular, and gravitational forces in sustaining them \citep{Jerolmack2019}.
 
@@
 Viewed in this way, the work in this thesis targets a minute region of the vast parameter space spanned by Earth's geophysical flows, and geomorphology requires characterization of them all.
-Future studies should continue the effort \citep{Furbish2021a} to describe these processes as different expressions of the same basic statistical mechanics building blocks.
+Future studies should continue the effort \citep[e.g.][]{Furbish2021a} to describe these processes as different expressions of the same basic statistical mechanics building blocks.
 
@@
 
-This thesis described the movements of individual grains along streambeds using probabilistic methods.
-The research has related the overall sediment transport rates responsible for channel evolution to the movements of individual grains.
+This thesis has described the movements of individual grains along streambeds using probabilistic methods.
+The research relates the overall sediment transport rate to the movements of individual grains.
 At base level, the thesis embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as the magnitude of their fluctuations.
 
-The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes without including it, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climate fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving grain size distributions \citep{Parker1982,Andrews1983}.
+The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscales without including variability, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving surface grain size distributions \citep{Parker1982,Andrews1983}.
 

### 2021-09-03 revision/formerPapers/com-piece/com-piece.tex
added phd thesis.

\documentclass{article}
% General document formatting
\usepackage[margin=0.7in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
%\bibliographystyle{unsrtnat}

% \usepackage[math]{kurier}
\usepackage{setspace}       % \onehalfspacing and \singlespacing
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}
\usepackage[utf8]{inputenc}
\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
\doublespacing

\begin{document}

\title{Does gravel flow through rivers, or does it bounce?\\ the complementary ideas of Bagnold and Einstein}
\author{Kevin Pierce}
\maketitle

\begin{abstract}
\noindent Coarse sediment transported as bedload within a turbulent water stream is a mechanical process, which motivated two main approaches to understand it based on physical principles: the stochastic approach of Einstein and the deterministic approach of Bagnold. 
These two approaches have become prevalent research paradigms which support a conceptual divide in the river science community. 
By reviewing the Einstein and Bagnold paradigms and cross-comparing them, I argue this divide is short-sighted. 
Calling on wider knowledge on the dynamics of gravel bed rivers, I speculate about the role of these mechanistic concepts of bedload transport within future geophysical theories of gravel river morphology.  
\end{abstract}

\section{Mechanistic theories of bedload transport}

Gravel-bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological problems \citep{Newson2000, Gaeuman2017}.  
Since the transport of coarse material as bedload is an important control on channel morphology \citep{Church2006, Recking2016}, bedload transport has been carefully researched for over a century. 
Despite this effort, our ability to predict bedload transport rates remains poor.
Predictions of bedload transport can deviate from measured values by factors of 100 or 1000 \citep{Gomez1989, Barry2004, Recking2012}, motivating an intense pursuit of deeper understanding.

Bedload grains move in traction with the bed by rolling, sliding, and bouncing \citep{Einstein1950, Bagnold1973}. 
Since this transport is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases.
This expectation highlights a mechanical description of bedload transport as an attractive prospect.
However, bedload transport results from a turbulent flow interacting with granular matter, so a full mechanical description would be a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to construct mechanical descriptions using supplementary hypotheses. 
From over a century of efforts, two have emerged as especially popular research paradigms. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach, initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

These two paradigms are founded on the application of mechanical reasoning to simplified concepts of bedload motion.
These concepts involve empirical parameters, so these models must be calibrated to applications. 
Viewing bedload transport phenomena in terms of these simplified concepts, proponents of the Einstein and Bagnold paradigms take differing views on most aspects of the process, which often leads to disagreements and miscommunications between proponents of each school. 
This conflict coincides with the observation that neither mechanical approach is, practically speaking, very useful.
Both paradigms generate poorer bedload transport predictions than simple power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}, and because of their semi-empirical attributes, the superiority of either paradigm across the full range of bedload conditions is difficult to judge \citep{Iverson2013}.

Accordingly, although each mechanical approach has merits, the perspective that either is superior is not supported by evidence. 
Meanwhile, there is no clear way to join the two paradigms together, so proponents of each toil away in their respective camps, solving problems within (or with) their chosen paradigm, and failing to communicate science across the conceptual rift between approaches. 
We need to ask: is this rift justified by evidence? 
Are the stochastic and deterministic approaches to understand bedload transport mechanically really so different that they share no common ground, and is one or the other completely wrong? 
To address these questions, we should analyze the concepts, successes, and short-comings of each approach, and use these analyses to form a cross comparison.  
To break anticipation, the short answer is no.

\section{The Einstein paradigm: mechanistic-stochastic}

Einstein introduced stochastic concepts to understand the bedload transport rate by relating it to individual bedload motions \citep{Einstein1942, Einstein1950}.
He conceptualized the individual motion as a series of steps with intervening rests, and he set out to determine the resting time and step length with flume experiments \citep{Einstein1937}. 
During these experiments, he realized that these motion characteristics were not consistent or predictable from one particle to the next, so he was forced to switch to a statistical conception.
Across a population of many particles, Einstein realized the resting time and step length of individual particles generated statistical distributions, which he obtained from his experiments \citep{Einstein1937}.

From this statistical concept of individual bedload motions, Einstein made two key developments. 
First, he calculated the probability distribution of the effective velocity of bedload grains, as they move downstream through cycles of steps and rests. 
From the velocity difference between the fastest and slowest grains, he derived the rate of spreading or diffusion of tracer particles \citep{Einstein1937}.
Second, he derived the random character of the resting period as a mechanical consequence of fluid turbulence, and used this to calculate the mean rate of bedload transport \citep{Einstein1950}. 
The tenacity with which Einstein's stochastic concepts of bedload motion have been modified, critiqued, and improved over the last seventy years is a testament to the originality of his ideas \citep{Ettema2004}.

Generations of researchers have worked within the paradigm established by Einstein, criticizing and improving his concepts without fully invalidating them.  
Characteristics of individual motions have been investigated extensively \citep{Crickmore1962, Hubbell1964, Yano1969, Nakagawa1976, Drake1988, Hassan1991, Habersack2001, Ancey2008, Roseberry2012, Heyman2013, Fathel2015, Heyman2016}, and Einstein's statistical perspective appears essentially correct. 
Since Einstein, other motion statistics been measured and incorporated into stochastic theories of bedload transport, including the velocity, acceleration, and travel time during single particle motions \citep{Drake1988, Radice2006, Ancey2008, Lajeunesse2010, Furbish2012a, Roseberry2012, Fathel2015, Furbish2016,  Heyman2016}. 
Bedload diffusion has been carefully analyzed, and the problem is obviously more nuanced than Einstein considered. 
Diffusion has different characteristics depending on the temporal or spatial scales of consideration \citep{Nikora2002, Zhang2012, Martin2012}, and this probably relates to the effects of particle burial \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014, Bradley2017} and variable movement characteristics from sediment size \citep{Fan2017} on effective velocity differences between moving bedload.

Einstein's mechanical theory of the mean bedload transport rate \citep{Einstein1950} has been more carefully phrased in relation to underlying mechanics, and his concepts have been extended to compute full statistical distributions of the bedload rate \citep{Sun2000, Ancey2006, Ancey2008}. 
Subsequent theories on the resting time of stationary sediment, which is equivalent to the probability of erosion of a grain \citep{Yalin1972}, took more careful account of the mechanics driving particles into motion, including variable supporting configurations of bed surface grains, and the forcing from turbulent lift and drag \citep{Paintal1971, Wu2004, Dey2018}.
Meanwhile, contemporary experiments have increased recognition of the combined role of the magnitude and duration of turbulent fluctuations on the erosion process \citep{Diplas2008, Valyrakis2010, Celik2014}, which Einstein-like models of the resting time (or erosion rate) have only begun to incorporate \citep{Tregnaghi2012,Dey2018}.

Instead of computing mean bedload rates only, contemporary theories have leveraged stochastic mathematics \citep{Cox1965} to compute probability distributions of the bedload flux, deriving mean values and the expected magnitude of bedload fluctuations \citep{Sun2000, Ancey2006, Ancey2008}. 
These descriptions are a more complete description of bedload transport \citep{Ancey2008}, and the higher-order statistics provide additional benchmarks against which models can be tested \citep{Iverson2013}. 
This research links back to the resting time of stationary bedload in a contentious way which requires resolution. 
Einstein's concepts, when translated to theories which predicts the magnitude of bedload fluctuations, generate fluctuations which are too small to match experiments \citep{Ancey2006}. 
Accordingly, some contemporary models have introduced feedbacks in particle motion, which could be physically attributed to the erosion of multiple grains from the bed at once, and these somewhat ad hoc prescriptions imply realistically large fluctuations \citep{Ancey2008, Heyman2013, Ma2014}, meaning their mechanical origin deserves clarification.

From this sketch of research within the Einstein paradigm, we can see that Einstein's initial concepts have generated a great deal of new understanding of the physics underlying bedload transport, but they have also generated many open questions.
The Einstein paradigm is particularly successful at describing bedload fluxes at low transport stages, where instantaneous fluctuations in bedload rates are appreciable relative to mean values, taking values up to 400\% as large \citep{Bohm2004, Ancey2008, Singh2009, Heyman2016, Gonzalez2017}.
Under these conditions, Einstein's stochastic concept, where bedload transport is dominated by the random motions of individual grains, must be closer to the underlying physics.

However, Einstein-type theories have key limitations.
First, at high transport rates, the assumptions of these theories are not realistic. 
In these conditions, the entire bed surface moves as a type of granular flow, with motion several particle diameters below the surface \citep{Jenkins1998, Mouilleron2009, Frey2014, Houssais2015}, so Einstein's partition of transport into distinct periods of motion and rest breaks down \citep{Ancey2008, Heyman2014}.
Second, when channel morphology affects bedload motions, for instance by size sorting of bed material \citep{Nelson2010, Sun2015}, the influence of bedforms \citep{Iseya1987, Cudden2003, Kasprak2014, Hassan2017}, or aggradational/degradational cycles \citep{Dhont2018} there is limited understanding of how this can be represented in the statistics characterizing individual motions. 
Accordingly, we conclude that Einstein-type models are close to the underlying physics of low bedload discharges over a nearly flat bed \citep{Ancey2008, Heyman2013, Ma2014}, but at high bedload discharges, or when morphology affects bedload transport, new developments are required.

\section{The Bagnold paradigm: mechanistic-deterministic}

In contrast, Bagnold understood the bedload rate as a deterministic energy balance between the solid and fluid phases. 
His key concept was that gravity supplies excess energy to the fluid flow, beyond the capacity of turbulence to dissipate, so it is dissipated by the friction of moving bedload against the stationary bed \citep{Bagnold1956, Bagnold1966, Bagnold1973}.
To complete this concept, Bagnold had to define what constituted excess energy within the fluid flow. 
To this end, he made a now-famous hypothesis: when bedload transport is at equilibrium value, so it dissipates all excess power available to move sediment, the average shear stress on the bed will match the threshold of sediment motion.
This is called the Bagnold hypothesis \citep{Seminara2002, Ancey2006}. 
Supplementing his energy balance concept with this hypothesis, Bagnold predicted a bedload rate ($q_b$) which scales, at high stages of transport, as the average fluid shear stress ($\tau$) to the $3/2$ power: $q_b \propto \tau^{3/2}.$
Bagnold's theory requires the specification two parameters: one characterizing the dynamic friction of moving bedload, and another characterizing the threshold of bedload motion \citep{Bagnold1973}.

Bagnold's ideas are enlightening for their pragmatic clarity, and they have been taken up and improved by many researchers. 
His deterministic concept of bedload motion is simplified just enough from the real complexity to make contact with the reliable mechanical principle of conservation of energy, and at high bedload rates, his predicted scaling between bedload rate and average shear stress ($q_b \propto \tau^{3/2}$) is in accord with classic empirical results \citep{MeyerPeter1948}, and, strikingly, is also in accord with the most advanced numerical simulations of bedload transport \citep{Schmeeckle2014}.
However, despite these successes, his simplified assumptions have left behind many open questions for proponents of his ideas.

For example, Bagnold's theory of bedload transport ignores the details of bedload trajectories, which has motivated many efforts to reconcile his average energy budget concepts with the dynamics of individual motions \citep{VanRijn1984, Bridge1984, Wiberg1989, Bridge1992, Nino1998, Bialik2015}. 
Since bedload trajectories are a consequence of turbulent forcing and bed collisions, whose characteristics are random, this work is a juxtaposition of fluctuating and average energy concepts. 
A shortcoming of this research is that the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Schmeeckle2007, Dwivedi2010, Dwivedi2011}, especially when particles are not spherical \citep{Maxey1983a}, with relatively important terms entering the force balance apart from the typical drag and lift \citep{Bialik2015}.
This may partly explain why these Bagnold-type models, based on the Newtonian dynamics of individual bedload motions, remain troubled. 
They usually overpredict bedload rates, and underpredict the energy dissipation of bedload transport \citep{Bridge1984, Bialik2015}.

Some authors have attributed these shortcomings to a lack of inclusion of 
 collisions between moving grains \citep{Lee2002}, and to a lack of understanding of the stresses imparted by moving grains on the stationary bed \citep{Nino1998}.
Both of these criticisms really point toward a deeper problem which links deeply to granular physics.
With its energy budget perspective on bedload motion, which ignores the details of individual motions, Bagnold's original theory essentially phrases bedload motion as an average granular flow with a highly simplified internal friction law, or rheology \citep{Frey2014}. 
Within granular physics, tractable models of rheology are limited to two extremes: there is the simple quasi-static rheology assumed by Bagnold (Coulomb-type), which should be valid in high concentration flows \citep{Berker1992}, and there are models adapted from the kinetic theory of gases \citep{Jenkins1998}, valid in low concentration flows. 
Possibly, bedload is an intermediate concentration flow, which means its description as a type of granular flow demands a theory of rheology which does not exist \citep{Frey2011, Frey2014}. 
There are mixed conclusions from experiments on the validity of Bagnold's friction law \citep{Silbert2001, Frey2014, Houssais2016}, and in context of the shortcomings of trajectory based bedload models, this topic deserves further attention.

From this short review of Bagnold's concepts and the research directions they have generated, it's easy to conclude that Bagnold's work pushed science forward. 
At high bedload discharges, his description appears essentially correct: when bedload transport resembles a collective flow, it scales with average shear stress as $\tau^{3/2}$.
In this regime, his energy dissipation arguments must be close to the relevant physics of bedload transport. 
However, Bagnold's theory yields poor results at low transport stages and over arbitrarily sloping beds, where it fails to characterize bedload transport \citep{Engelund1976, Luque1976, Seminara2002, Martin2000}.
One issue is that measured bedload rates express large fluctuations at low mean discharges, so mean values poorly represent the bedload signal at low rates, and statistical representations are really closer to the underlying physics \citep{Ancey2008}.

The deepest issue with Bagnold at low transport stages is that solid transport rates are not suitably large to dissipate the excess energy available to move bedload grains, meaning Bagold-type models do not fit experimental data without calibrating bulk particle friction coefficients to unphysical values \citep{Engelund1976, Luque1976, Nelson1995, Nino1998}.
For example, \citet{Nino1998} found an effective friction angle of $56.6$ degrees was required to describe their experiments at low transport stage, a value well above the angle of repose of any natural sediment \citep{Miller1966}.
We can conclude that Bagnold's ideas are close to the physics of bedload transport at high transport stages, but his theory fails outside of the conditions in which bedload moves as a collective flow. 
Accordingly, for lower bedload discharges, we need a different perspective.

\section{Stochastic or Deterministic?}

Clearly, both mechanistic paradigms have strengths and weaknesses, but is one or the other relatively closer to the physics of bedload transport?  
Often, the rift between the Einstein and Bagnold paradigms is framed in terms of a divide between stochastic and deterministic approaches, and this is short-sighted. 
Scientists are not justified in ignoring consistently reproduced findings: in natural streams and controlled laboratory measurements, bedload transport rate time-series always appear with random fluctuations. 
In laboratory channels with steady uniform flows and nearly identical or narrowly graded sediment, under conditions where the development of channel morphology is suppressed, instantaneous bedload rates can be up to 400\% mean values \citep{Bohm2004, Ancey2008, Roseberry2012, Heyman2016}.
In real gravel bed rivers and in flume experiments where channel morphology can evolve, bedload fluctuations may be even more predominant \citep{Iseya1987, Kuhnle1988, Hoey1992, Cudden2003, Singh2009, Dhont2018}. 
Experiments consistently show us that bedload transport rates are statistical quantities.

Accordingly, a full characterization of bedload transport should include mean bedload rates, as well as any statistics of the bedload time-series, such as the expected magnitude of instantaneous deviations from mean values. 
In general, bedload transport would be more completely described by a stochastic theory. 
However, this does not mean Einstein is better than Bagnold, or that one or the other is completely wrong.
I have already discussed that Einstein appears closer to the physics at low bedload rates, when the intermittency of individual motions dominates the bedload signal, while Bagnold appears closer at high bedload rates, when bedload motion appears like a collective flow. 
Experimental data reinforce this assignment of the idealized frameworks of Einstein and Bagnold to complementary extremes of transport phenomena.

For example, \citet{Ancey2008} measured the bedload transport rate ($q_s$) of identical glass beads within a narrow flume under a range of steady flow conditions. 
They found the relative strength of bedload fluctuations, characterized by the ratio of the bedload variance to the mean ($\text{var}(q_s)/\overline{q}_s$), scaled with average fluid shear stress. 
This indicates a key trend of bedload transport, at least when bedform development is suppressed: stochastiscity is accentuated at low mean bedload rates, and it is suppressed at high bedload rates. 
They went further to analyze their results in relation to Bagnold's assumption that the energy dissipation of bedload governs its rate. 
From their data, they estimated the fractions of energy dissipated by turbulence and by bedload transport. 
At low bedload discharges, as much as $90$ percent of the energy supplied by gravity was disspated by turbulence (so $10$ percent by bedload), while at high bedload discharges, a much as $75$ percent of  flow energy was dissipated by momentum transfers with moving bedload (so $25$ percent by turbulence).
These trends suggest that bedload transport is actually a compromise between the stochastic concept of Einstein -- that bedload transport is dominated by individual motions, and the deterministic concept of Bagnold at high discharges -- that bedload transport is governed by an excess energy budget.
At low bedload discharges, the data favor Einstein's concepts; while at high bedload discharges, they favor Bagnold's.

It's best for science if the opposition between the stochastic and deterministic viewpoints dissolves, and if river scientists start to see Einstein and Bagnold as what they are: idealized models which pertain to particular limits of the full phenomena of bedload transport. 
Proponents of Einstein build bedload transport rates up from individual particle motions, and contend that the motion characteristics of individual particles are not predictable due to the random character of fluid turbulence and the arrangement of bed grains, and this is correct. 
When bedload discharges are low, so that motions are sparse, the random character of individual motions is emphasized within the bedload rate, and it appears as a statistical quantity.
Under these conditions, a full statistical description based on individual motions is probably more realistic \citep{Ancey2008, Heyman2013, Ma2014, Heyman2016}.
In contrast, proponents of Bagnold understand bedload transport as a dissipative flow, where the details of individual motions are less relevant than their collective dynamics. 
When bedload discharges are high, the random character of individual motions is washed out through inter-granular collisions and mixed through the population of moving grains, so average characteristics dominate the bedload rate, and a deterministic description based upon collective properties of a granular flow is probably more descriptive \citep{Jenkins1998,  Frey2011, Frey2014}.

The physics of bedload transport are apparently complex enough to support more than one theoretical description, each pertinent to a different regime where particular interactions are emphasized, which is not surprising. 
After all, should we be surprised to find the coupled physics of turbulent flows and granular media to be simpler than either alone?
Bagnold and Einstein, or deterministic and stochastic -- these are both meaningful physical descriptions, but of different conditions of bedload transport.
This leads to a few final questions. 
Is one condition or the other of bedload transport more relevant to understand gravel bed rivers physically, and how exactly does bedload transport link into a more complete geophysical understanding of gravel bed rivers?

\section{Toward deeper understanding of gravel bed rivers}

Our understanding of gravel bed rivers, of which bedload transport is only one component, remains far too primitive to supercede descriptive geomorphology.   
\citet{Schumm1977}, in his classic work, outlined morphology as the form and structure of river channels which results from coordination between the supply, storage, and transport of sediment.
In gravel bed rivers, which may flow through forests, or previously glaciated landscapes, sediment supply and storage may be more deeply linked to geology \citep{Rennie2018}, ecology \citep{Wohl2017}, climate \citep{Gregory2006, Slaymaker2009}, and human influence \citep{Hooke2000, Gregory2006, Slaymaker2009}, than they are to bedload transport: so channel morphology could be largely imposed by controls external to river channels.
Yet channel form cannot evolve or develop without the erosion and redistribution of bed sediment, so in gravel bed rivers, and in coordination with these external controls, bedload transport underlies channel morphology \citep{Church2006, Recking2016}.

Gravel bed rivers are facing serious ecological pressures \citep{Frissell1993, Hauer2016}, and this undoubtedly links to changes in their morphology, driven by external controls, while the scope of these pressures pivots on our capacity to understand and counter these morphological changes \citep{Wohl2015}.
Accordingly, we can identify two inquiries which are essential to a deeper understanding of gravel bed rivers.
First, research into the external controls on channel form, like geological and forest processes, needs to continue. 
Storage and supply are just as important as transport in determining channel form. 
Second, we need a quantitative understanding of the erosion and redistribution of bed sediment with channels, and this is where Einstein and Bagnold come in.
However, we must accept the geophysical problem is really more difficult than either scientist envisioned.

Within gravel bed rivers, depending on the particular conditions of sediment supply and storage, a wide diversity of organized structures can be discerned on the bed \citep{Church1998, Hassan2008, Nelson2014, Venditti2017}.  
These morphological elements span a range of spatial scales: clusters, ribs, cells, steps, bars -- from the scale of several grains to several times the width of channels. 
These bedforms modify the stability of bed grains, and redirect hydraulics, in turn coordinating bedload transport \citep{Laronne1976, Lisle1992, Kasprak2014, Recking2016, Hassan2017}.
Simultaneously, bedload transport can form, move, and destory bedforms \citep{Whiting1988, Hoey1992, Cudden2003}, and morphological units such as pools can support quasi-periodic cycles of aggradation and degradation \citep{Jackson1982, Dhont2018}.  
From these observations, we can conclude that real gravel bed rivers express feedbacks between transport and morphology which are far beyond the scope of our simplified models, and we can speculate that these feedbacks are a predominant cause of the noted failure of bedload models when they're applied to natural channels \citep{Gomez1989, Barry2004, Recking2012}.

Within natural rivers, transport of gravels has been divided into three stages to classify its intensity \cite{Jackson1982, Ashworth1989}. 
Stage 1 is finer material overpassing a locally static bed, so any mobile material originates upstream. 
Stage 2 is partial mobility of local bed material, so at any one time, part of the bed is static, but any grain might eventually move. 
Stage 3 is full mobility of bed material, so no part of the bed is static, and all grains move equally, in what could be called a flow. 
We can associate stages 2 and 3 with the complementary concepts of Einstein and Bagnold, if we acknowledge they overlook many details of natural rivers, such as transport-morphology feedbacks.

If we agree with \citet{Wolman1960} that, owing to their relative frequency of occurence, intermediate transport stages predominate over high stages in terms of geomorphic influence, then Einstein's concepts are somewhat more relevant to the geophysics of gravel bed rivers. 
Bedload moves intermittently under most conditions of geomorphic interest. 
However, the Einstein paradigm's inability to understand the interaction of transport with channel morphology, such as the collective migration of bedforms, suggests maybe there's something to learn about this from the energy balance concept of Bagnold. In the classic literature on bedform movement, and its role in bedload fluctuations, these motions are described as a collective migration of very many grains at once \citep{Whiting1988, Hoey1992, Cudden2003}, and is this really so different from Bagnold's concept of bedload motion?

A possible conjecture is that in gravel bed rivers, individual grains and bedforms may both have stochastic motions, and while individual grains move in accord with turbulence, might bedforms move as a type of collective flow? 
Can variable driving and resisting forces govern individual movements of bedload, while an energy balance governs the collective movements of bedforms? 
Or is the mix of phenomena even more intricate? 
One imagines the alternate start stop motions of a bedload sheet, which is not more complex than cycles of aggradation and degradation in pools. 
How could the fluid and grains coordinate to express these phenomena? 
In building a geophysics of gravel bed rivers, we might need to get comfortable with complexity, and to attempt a descriptive synthesis from the concepts of simpler models, like the bedload theories of Einstein and Bagnold.

\bibliographystyle{agu}
\bibliography{biblio}
\end{document}

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece0.tex
added phd thesis.

--- 2021-09-03 draft/formerPapers/com-piece/scraps/com-piece1.tex
@@
 
-\title{On argumentation and gravel rivers: Einstein and Bagnold in a modern context}
+\title{On randomness and gravel rivers: Einstein and Bagnold in modern context}
 \author{Kevin Pierce}
@@
 
-%% intro
+% gravel bed rivers are in trouble 
+% need to understand interaction of transport, storage, supply
+% but transport has been tough to describe due to turbulence and granular interactions
+% fundamentally the problem is to predict the rate given the water flow
+% Currently, there are no general laws, and our ability to predict transport rates in natural streams is inadequate
+% However, there are two prevalent research paradigms:
 
-Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
-Since the transport of coarse material as bedload is an important control on channel morphology, bedload transport has been the subject of intense research for over a century. 
-Despite this effort, our ability to predict bedload transport rates is inadequate.  
-In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000. 
-Since the bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
-However, because bedload transport results from a turbulent flow interacting with granular matter, the full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
-In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and among these, two of them have been taken up by generations of researchers, forming two main research paradigms in the physics of bedload transport.
-These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.
+Gravel bed rivers are troubled natural features, and hosts to a huge set of ecological problems related to changes in their form and function.
+These changes often link to human disturbances of the natural regimes of sediment transport, storage, and supply within these rivers. 
+Accordingly, these processes are cornerstones of river science, and they have been carefully studied for well over a century. 
+Despite this work, with regard to sediment transport in gravel bed rivers, there is very little practical progress. 
+The earliest efforts were based upon correlating empirical data to power laws relating the rate of sediment transport to characteristics of the flow and sediment \citep{MeyerPeter1948}, and this work provided a foundation for the now classic approaches based on mechanistic physics.
+Among these early mechanistic approaches, two have been particularly influential, each initiating its own research school which are dominant today: these are the stochastic approach of \citet{Einstein1937, Einstein1950}, and the deterministic approach of \citet{Bagnold1956, Bagnold1973}.
+A rift exists these two schools, and proponents of one face ire from the other, and they often find debate: these schools take different approaches to understand almost every aspect of sediment transport in gravel bed rivers.
+However, a reality underlies this divide: the most advanced mechanical theories of both schools, from a practical standpoint, remain less useful than modern theories based on fitting power laws to empirical data \citep{}, and apparently no theory among these provides reliable predictions of transport in natural streams \citep{}.
+Accordingly, we should take a step back and look at what's happened.
 
-Each of these approaches is founded on the application of mechanical reasoning to a simplified concept of bedload motion, characterized by empirical parameters, so that both categories of models must be calibrated to applications.
-Also, the simplified concepts of Einstein and Bagnold-type models are different with regard to most aspects of bedload motion.
-Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
-This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and that neither approach is clearly superior than the other. 
-Specifically, both of these mechanical paradigms are expected to generate poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics.
-Accordingly, although each mechanical approach has clear merits and demerits, and it's safe to conclude neither is perfect; the perspective that either is fully superior, for any reason apart from practicality, lacks objective support.
-Because there is no clear path to join the two schools together, proponents of each school toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift that Einstein and Bagnold have left behind.
-We need to ask: is this rift justified by evidence?
-Are Bagnold and Einstein really so different that they share no common ground, and is one or the other completely wrong?
-If so, theorists should out the failed approach and concentrate on its successor together.
-If not, we should work together to synthesize something new out of the strongest elements of each paradigm.
+Among these, probably resulting from the heightened influence of fluid turbulence and granular interactions in gravel bed rivers, sediment transport has been a particularly difficult inquiry. 
+Fundamentally, the problem is this: given the rate of water flow, the form of the channel, and the properties of the sediment grains within a gravel bed river, what is the rate of transport of sediment grains?
+- describe how poor our capacity for prediction is 
+This problem has been studied for over a century, with limited success, and no general laws. 
+We need to understand why.
 
-To answer these questions, we should carefully assess the concepts of each paradigm, and try to discern any conditions of bedload transport under which one description provides more or less useful information than the other.
-Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical picture of individual bedload motions, and this picture should be consistent with contemporary experiments.
-Bedload grains typically move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}.
-% resting, entrainment by impulse, motion in turbulent flow, collisions with bed, deposition
+Sediment grains within gravel-bed rivers typically move intermittently, rolling, sliding, and bouncing along the river bed, a condition called bedload motion \citep{Einstein1950}.
+The bedload motion of an individual grain is an alternating series of start to stop transitions, with these dynamics undoubtly governed by a Newtonian force balance between the solid and fluid phases.
+However, the forces making up the balance are not exactly known in any context, and the granular \citep{Lamb2008} and fluid \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} interactions are known to exhibit considerable variability in magnitude, direction, and duration.
+Owing to this complexity, and to our relative ignorance about the details of the underlying Newtonian basis of bedload transport, two main research schools have emerged, each leveraging a set of approximations and assumptions in order to make use of Newtonian concepts: these are the deterministic school, initiated by Bagnold, and based upon the concept of momentum exchange between the granular and fluid phases; and the stochastic school, initiated by Einstein, and based upon probabilistic transition rates between motion and rest states.
+In the river science community, these two schools are often opposed. 
+In this essay, I argue this division is short-sighted, and probably finds more basis in feelings than facts.
+Looking at the mechanical phenomena of bedload transport which experiments have revealed, it's clear that both stochastic and deterministic models of bedload transport are highly simplified away from the actual mechanics and processes at play.
+They both fail to predict bedload transport to similar degrees, although in different regions of the parameter space; and they each have important successes, of which we should take account. 
+Accordingly, I argue we should really be synthesizing Bagnold and Einstein for a new understanding, and not counterposing them. 
+This synthesis should happen in acknowledgement of the full complexity of the process, and with respect for experimental observations.
 
-% explain how bedload transport has different behaviors at low and high stages of bedload transport
-
-% introduce the einstein theory, its revisions, and experiments which support it
-
-% introduce the bagnold theory, its revisions, and experiments which support it
-
-% highlight that bagnold is the high transport stage limit of tau^(3/2), while einstein is really the low transport stage limit where average properties of the flow and grains do not characterize bedload transport
-
-% really, we should we should be joining these theories to describe wolman's intermediate region, which is where channel morphology happens
-
-% however, in intermediate regions, the linkage between morphology and transport becomes more highly coupled; so we really need to integrate knowledge from higher scales with the stochastic theories of einstein and the power dissipation arguments of bagnold -- neither is good enough
-
-% conclude by wrapping problems, stochastic, determinitsic, a synthetic description of intermediate conditions, where feedbacks are accentuated with morphology, with an interesting conclusion about scientific divides and tribalism impeding progress
-
-Sediment grains within gravel-bed rivers typically move intermittently, rolling, sliding, and bouncing along the river bed, in a condition called bedload motion \citep{Einstein1950, Bagnold1973}.
-The bedload motion of an individual grain is an alternating series of movements and rests, with these dynamics undoubtly governed by a Newtonian force balance between the solid and fluid phases \citep{Schmeeckle2013, Ji2013}.
-However, the forces making up the balance are not exactly known in any context, and the granular \citep{Lamb2008, Frey2011, Booth2014} and fluid \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration.
-Looking at experiments on the mechanics of bedload transport, it's clear that the Einstein and Bagnold schools are simplified from reality.
-
-We should really be attempting a synthesis between them, but to achieve this, we have to assess the experimental findings on the processes underlying bedload transport, as well as the concepts and development of the Einstein and Bagnold schools.
-To make this synthesis, we should be greedy, taking useful ideas from wherever we can find them. 
-To this end, we can start from a question: which concepts from each school fit with experimental findings, and how can they be fused together for a predictive theory of bedload transport?
-
-%% bedload experiments
-
-% first describe broad brush ideas. initiation, motion, cessation 
-To approach the question, we should develop a clear mechanical understanding of individual bedload motions which is supported by experiments, and we should consider where Einstein and Bagnold's theories fall in relation to it.
-Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
-To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
-Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact.
-Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}.
-The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed.
-Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
-Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.
-
-% now describe gaps in this understanding
-This broad mechanical perspective on the processes underlying individual particle motions reveals many gaps in experimental knowledge. 
-For one thing, in discussing the
-
-% now describe the einstein theory, and
-
-% initiation
-% motion
-% cessasion
-% rates
-
-%% Einstein and Bagnold
-
-Motion of an individual particle is initiated when a sufficiently strong driving force is applied by the turbulent fluid over a timescale large enough to overcome the resistive forces imparted on that grain by frictional and normal responses from the every bed grain it contacts \citep{Paintal1971, Valyrakis2010, Celik2014}.
-In general, both the duration and magnitude of the applied fluid forces is important \citep{Diplas2008, Valkyrus2010, Celik2014}. 
-This is an impulse concept, although in a viscous flow without velocity or pressure fluctuations, this impulse concept reduces to the classical perspective of a critical motion condition based upon velocity or shear stress \citep{Shields1936, Buffington1997}.
-
+To begin with, we should discuss the experimentally observed phenomena underpinning bedload transport. 
+Because bedload motion is intermittent, a useful division is to conceptualize it as three stages: initiation, movement, and cessation. 
+The properties of each stage result from a balance of granular and fluid interactions. 
+Motion of an individual particle is initiated when a sufficiently strong driving force is applied by the turbulent fluid over a timescale large enough to overcome the resistive forces imparted on that grain by frictional and normal responses from the every bed grain in contact \citep{Paintal1971, Valyrakis2010, Celik2014}.
+In general, the duration and magnitude of the applied fluid forces is important \citep{Diplas2008}; this is an impulse concept, although in a viscous flow without velocity or pressure fluctuations, this impulse concept reduces to the classical perspective of a critical motion condition based upon velocity or shear stress \citep{Shields1936, Buffington1997}.
 This initiation of motion by a turbulent impulse is undoubtedly governed by Newton's laws, although the forces making up the balance are not exactly known in any context, and the granular resistive \citep{Lamb2008} and fluid driving \citep{Schmeeckle2007, Celik2014, Amir2014, Shih2017} forces are known to exhibit considerable variability in magnitude, direction, and duration.

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece2.tex
added phd thesis.

--- 2021-09-03 draft/formerPapers/com-piece/scraps/com-piece1.tex
@@
 
-\title{On argumentation and gravel rivers: Einstein and Bagnold in a modern context}
+\title{On simplification and gravel rivers: Einstein and Bagnold in a modern context}
 \author{Kevin Pierce}
@@
 
-%% intro
-
+\section{Introduction: Gravel bed rivers and mechanistic theories of bedload motion}
+%% gravel bed rivers, bedload transport, mechanistic theories, Einstein and Bagnold 
 Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
@@
 In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000. 
-Since the bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
+Prediction is difficult because individual bedload grains move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}. 
+Since this bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
 However, because bedload transport results from a turbulent flow interacting with granular matter, the full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
-In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and among these, two of them have been taken up by generations of researchers, forming two main research paradigms in the physics of bedload transport.
-These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.
+In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and from these efforts, two of them have emerged as important research paradigms in the physics of bedload transport.
+These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach, initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.
 
-Each of these approaches is founded on the application of mechanical reasoning to a simplified concept of bedload motion, characterized by empirical parameters, so that both categories of models must be calibrated to applications.
-Also, the simplified concepts of Einstein and Bagnold-type models are different with regard to most aspects of bedload motion.
+% Einstein and Bagnold -- simplified concepts of motion -- semi-empirical -- rift between the two schools -- lack of communication between them 
+These two paradigms are founded on the application of mechanical reasoning to simplified concepts of bedload motion.
+These concepts are characterized by empirical parameters, so that mechanical models must be calibrated to applications. 
+Owing to these simplified concepts, proponents of Einstein and Bagnold-type models have different understanding of most aspects of bedload transport.
 Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
-This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and that neither approach is clearly superior than the other. 
-Specifically, both of these mechanical paradigms are expected to generate poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics.
-Accordingly, although each mechanical approach has clear merits and demerits, and it's safe to conclude neither is perfect; the perspective that either is fully superior, for any reason apart from practicality, lacks objective support.
-Because there is no clear path to join the two schools together, proponents of each school toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift that Einstein and Bagnold have left behind.
-We need to ask: is this rift justified by evidence?
-Are Bagnold and Einstein really so different that they share no common ground, and is one or the other completely wrong?
-If so, theorists should out the failed approach and concentrate on its successor together.
+This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and any superiority of either approach is difficult to judge. 
+For example, either mechanical paradigm generates poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics.
+Accordingly, although each mechanical approach has merits, a perspective that either is uniformly superior for any reason apart from practicality, is not supported by evidence.  
+Because there is no clear path to join the two paradigms together, proponents of each toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift.
+We need to ask: is this scientific divide justified by evidence? 
+Are stochastic and deterministic approaches to understand bedload transport as a mechanical process really so different that they share no common ground, and is one or the other completely wrong? 
+If so, theorists should out the failed approach and focus together on developing its successor. 
 If not, we should work together to synthesize something new out of the strongest elements of each paradigm.
 
-To answer these questions, we should carefully assess the concepts of each paradigm, and try to discern any conditions of bedload transport under which one description provides more or less useful information than the other.
-Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical picture of individual bedload motions, and this picture should be consistent with contemporary experiments.
-Bedload grains typically move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}.
+\section{Bedload fluxes from individual motions: the stochastic paradigm of Einstein}
+% Einstein -- bedload flux is individual motions -- individual motions are random -- start to stop transitions -- entrainment, motion, and deposition
+
+Einstein initiated the stochastic paradigm by relating bedload transport rates to individual particle motions \citep{Einstein1942, Einstein1950}.
+He conceptualized the downstream transport of an individual particle as an alternating series of rests and hops, and he set out to determine the period of rest and distance covered during a hop with flume experiments \citep{Einstein1937}.
+During these experiments, he realized that the resting period and hop distance of individual particles was not consistent or predictable from one particle to the next, so he was forced to switch to a statistical conception.
+Viewing these motion characteristics on a population level involving many particles, Einstein noted the resting period and hop distance of individual particles lie on stable statistical distributions, which he measured \citep{Einstein1937}.
+Initially, he used the random characteristics of individual rests and motions to calculate the probability distribution of the effective velocity at which bedload grains travel as the move downstream through cycles of rest and motion, with relevance to the spreading or diffusion of tracer particles as they move downstream due to their different effective velocity characteristics \citep{Einstein1937}.
+Later, he derived the random character of the resting period as a mechanical consequence of fluid turbulence, and he used this concept to derive the mean rate of bedload transport \citep{Einstein1950}.
+The tenacity with which Einstein's ideas have been followed up, modified, critiqued, and improved over the last seventy years is a testament to his pioneering originality \citep{Ettema2004}, yet the clear pattern of development and refinement along the course of his life provides motivation, showing an obvious balance between hard work and genius \citep{Einstein1937, Einstein1942, Einstein1950, Einstein1964}.
+
+Generations of researchers and proponents of Einstein have criticized, generalized, and reinterpreted his stochastic concept of bedload transport, without fully invalidating his ideas. 
+Statistical characterizations of individual motions have been investigated extensively \citep{Hubbell1964, Yano1969, Nakagawa1976, Hassan1991, Habersack2001, Ancey2008, Roseberry2012, Heyman2016}, and it's clear that his conception of their characteristics as random variables drawn from statistical distributions is essentially correct.
+Since Einstein, new characteristics of particle motion have been considered, and incorporated into stochastic theories of bedload transport, such as the velocity, acceleration, and travel time during single particle motions \citep{Drake1988, Ancey2008, Lajeunesse2010, Roseberry2012, Fathel2015, Heyman2016}.
+Understanding of particle diffusion has advanced a great deal, and the problem is obviously more nuanced than Einstein considered \citep{Einstein1937}.  
+Particle diffusion has different characteristics depending on the temporal or spatial scales of consideration \citep{Nikora2002, Zhang2012, Martin2012}, and this probably relates crucially to the effects of particle burial \citep{Sayre1971, Nakagawa1980, Voepel2013, Martin2014, Bradley2017} and variable movement characteristics due to sediment size \citep{Fan2017} on the differences in effective velocities of moving bedload.
+
+Most notably, Einstein's mechanical theory of the bedload transport rate \citep{Einstein1950} has been reworked a great deal. 
+Subsequent theories on the resting time of stationary sediment, which is equivalent to the rate of erosion of an individual surface grain, took account of the randomness in supporting configuration of bed surface grains, in addition to randomness resulting from fluid turbulence \citep{Paintal1971, Wu2002, Dey2018}, and experiments have increasing recognition of the role of both the magnitude and duration of turbulent fluctuations on particle entrainment \citep{Diplas2008, Valyrakis2010, Celik2014}, which Einstein-like models of the resting time (or erosion rate) have not yet incorporated \citep{Dey2018}.
+Additionally, instead of computing mean bedload rates only, contemporary theories have harnessed the physics of stochastic processes to compute probability distributions of the bedload flux, so that mean values and the expected magnitude of bedload fluctuations are available \citep{Lisle1998, Sun2000, Ancey2006, Ancey2008, Ma2014, Ancey2014}.
+This work reflected back on the resting period of particle motions in a contentious way which still requires resolution. 
+Einstein's concepts, when translated to a theory which predicts the magnitude of bedload fluctuations, produces fluctuations which are too small to describe experiments \citep{Ancey2006}. 
+Accordingly, a set of contemporary models has incorporated collective feedback mechanisms in particle motion, which could be physically attributed to the erosion of multiple grains from the bed at once \citep{Ancey2008, Heyman2013, Ma2014}, as might arise from small granular avalanches \citep{Heyman2013}, the effects of coherent turbulence on particle motion \citep{Nino1998, Amir2014, Santos2014, Shih2017}, or the migration of bedforms \citep{Dhont2018}.
+In any case, research from the stochastic paradigm which Einstein initiated has generated a formidable amount of understanding about the physics underlying bedload transport, and this progress is expected to continue.
+
+However, many elements of Einstein's conception can be criticized. 
+For one thing, there are multiple modes within the statistics characterizing particle rest and motion, and no clear indiciation of the extent to which these must be resolved in bedload transport theory.
+For example, particles can rest on the bed surface, and this resting period can be characterized probabilistically \citep{Einstein1950}, but these resting times are known to be variable across bedforms, when the relative exposure of bed particles to the flow is variable \citep{Crickmore1962, Sayre1967, Yang1971, Nakagawa1980}.
+Also, particles can rest within the bed surface, or in a state of burial \citep{Voepel2013, Martin2014, Olinde2015, Bradley2017}, and the resting statistics associated with the burial process are different than those associated with surface resting.
+Similarly, when bedload particles move, they can fall into different modes of motion. 
+For example, they can roll, slide, or bounce, and within each of these modes they may have different motion statistics \citep{Frey2006, Frey2014}. 
+The degree to which these different modes must be accounted for in stochastic bedload transport theory has not been clarified.  
+For another thing, at high transport stages, Einstein's partition of bedload transport into separate motion and rest periods breaks down.
+Under these conditions, bedload motion takes the form of a sheet flow separating the bed from the water flow, and the entire bed surface moves with prevalent interactions between all moving grains \cite{Jenkins1998, Clark2015}, possibly with a vertical velocity distribution which depends upon distance from the shearing fluid \citep{Bagnold1973, Frey2011, Mouilleron2009, Houssais2015, Ferdowski2016}.
+This stage of bedload transport is expected to require a different description.
+
+\section{Bedload fluxes from collective dissipation: the deterministic paradigm of Bagnold}
+
+%"According to the bagnold hypothesis, the fluid shear stress at the bed equals the threshold value corresponding to incipient motion of particles; the difference %between the total bottom shear stress provides the shear stress in the solid phase.
+(Ancey2006) Interpreting the excess stress as a result of momentum transfers %between the solid and fluid phases makes it possible to compute the number of particles that can be entrained and maintained in motion, and thereby the solid flow %rate"
+
+Bagnold formed a different simplified concept of bedload transport to understand it with mechanical principles, initatiating the deterministic paradigm of mechanistic bedload transport theory. 
+From the physical point of view, Bagnold considered bedload transport in terms of a momentum exchange between solid and liquid phases.
+He balanced the average power supplied to the flow by gravity against the average frictional dissipation by moving bedload, and neglected any details of turbulent fluctuations, irregularity in the configuration of stationary grains, or the trajectories of moving particles.
+Bagnold's formulation hinges on the application of an average dynamic friction coefficient to moving bedload in order to characterize its energy dissipation. 
+His formulation is built upon a key hypothesis: when bedload transport is at its equilibrium value, so it dissipates all available power to move sediment, the tractive force imparted by the moving fluid on the bed surface will match to the threshold for sediment motion.
+This is the so-called Bagnold hypothesis \citep{Engelund1976, Luque1976, Seminara2002, Ancey2006}.
+Bagnold's original formulation predicts a bedload transport rate which scales, at high stages of transport, as the average fluid shear stress to the 3/2 power: $q_b \propto \tau^{3/2},$ and his hypothesis has been taken up by generations of researchers, an enduring legacy which is even more interesting in that his bedload transport research seems to be at most a tertiary component of illustrious life's work, with a larger portion of his time devoted to his career as a decorated military commander in both world wars, and his initially recreational hobby of driving cars around the Sahara desert, which eventually gained him status as a preminent desert explorer \citep{Bagnold1988}.
+
+Bagnold's original concept is enlightening for its pragmatic approach \citep{Ashida1972, Bagnold1973, Engelund1976, Luque1976}: his deterministic picture of bedload motion is simplified just enough from the real complexity to make contact with a foundational and reliable concept of classical physics, the conservation of energy.
+At the same time, his highly simplified assumptions and comparisons between experiements and his results have left many open questions for proponents of his ideas.
+
+% Many investigators have built upon and refined Bagnold's concepts while preserving his main idea that bedload transport rates as a consequence of an average energy budget. 
+% Early on, investigators came to the conclusion that the details of bedload trajectories are not negligible in a Bagnold-like formulation, and that bedload trajectories cannot be understood without accounting for turbulent lift forces.
+% suggesting a contingence of the bagnold formulation on the details of particle motions ( Luque and Van Beek 1976 )
+% For one thing, lifting forces on sediment grains, which are highly variable in accord with fluid turbulence, contribute to the relative amount of time that bedload grains spend in contact with the bed, meaning the dynamic friction coefficient bagnold introduced should scale with fluid lift.
+% accordingly, a large set of bagnold-type treatments focused on including fluid turbulence in-so-far as it affects bedload trajectories, but not in relation to the average power balance which controls, in Bagnold's view, the average number of partilces in motion.
+% These studies effectively contend that bedload is dominated by an average power balance, but the dissipation mechanism within this balance is contingent on fluid turbulence.
+
+For example, Bagnold's total neglect of fluid turbulence and any irregularity between sediment grains left open problems around the degree to which fluid turbulence effects and the differential mobility of grains of varying size can be incorporated into a Bagnold-type framework.
+Likewise, Bagnold's agnostic approach to individual bedload motions within his theory led many investigators to consider how these details can be reconciled with his energy dissipation concept. 
+Despite these simplifications, Bagnold's original theory is aptly descriptive of bedload transport under particular conditions: at high transport stages, bedload transport rates often appear to scale as the average shear stress to the $3/2$ power ($q_b \propto \tau^{3/2}$) in accord with his mechanical prediction, at least over flat beds sheared by a steady uniform flow \citep{MeyerPeter1948,Gomez1989,Schmeeckle2014, Elghannay2017}.
+However, Bagnold's theory is troubled at relatively low transport stages and arbitrarily sloping beds.
+To describe bedload rates at low transport stages, the empirical components of Bagnold's theory must be set to unphysical values \citep{Engelund1976, Luque1976,  Nelson1995, Nino1998}; and his deterministic theory neglects a key attribute of bedload transport rates at low stages, namely that transport is highly intermittent, so that bedload transport rates exhibit strong fluctuations which Bagnold's theory does not constrain \citep{Bohm2004, Ancey2008}.
+Over arbitrarily sloping beds, there are theoretical arguments that Bagnold's hypothesis should flatly fail, even for relatively small bed slopes transverse to the flow direction \citep{Seminara2002}.
+These problems have motivated intense research within the Bagnold paradigm.
+
+Many researchers have concentrated on computing individual particle trajectories from Newtonian physics, and incorporating these details with Bagnold's average energy dissipation concepts, which should fix the number of particles in motion \citep{Abbott1977, Bridge1984, Wiberg1989, Bridge1992, Nino1998, Bialik2015}.
+Because the details of particle trajectories are highly contingent on fluid turbulence and collisions with the irregular bed surface, this incorporation of bedload trajectories can be viewed as a kind of first order inclusion of turbulent forcing and the random arrangement of bed grains \citep{}, which are key elements of Einstein-type models.
+A difficult attribute of this problem is that energy dissipation due to bedload should happen primarily when bedload is in contact with the bed surface, but the relative fraction of time bedload spends in contact with the bed surface is highly contingent on fluid turbulence, and especially fluid lift, and Bagnold neglected both of these aspects.
+Hence, Bagnold's dynamic friction coefficient should not really be a constant across all transport stages, and is instead expected to depend upon the details of bedload trajectories, which are inherently statistical and incompletely understood \citep{Bialik2015}.
+For one thing, the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Maxey1986, Schmeeckle2007, Dwivedi2010, Dwivedi2011}, with relatively important terms in the force balance aside from the typical drag and lift \citep{Bialik2015}.
+These models of individual bedload motions remain somewhat troubled, usually overpredicting the velocity at which bedload grains travel \citep{Bialik2015}, although the research is steadily generating new understanding.
+
+critical shear stresses have been difficult to consistently define and calculate
+and there are indications of difficulty in defining the concept as a consistent consequence of underlying physics
+
+Another deficiency of bagnold's concepts is the failure of the approach over arbitrarily sloping beds
+initially, the observation that bagnold's hypothesis did not lead to physical conclusions even for a relatively negligible transverse tilt in the bed surface
+motivated the development of an alternate formulation which preserved his average energy balance concept while understanding the number of partilces in motion as a balance of erosion and deposition rates
+Parker2003 ammended the bagnold hypothesis by replacing it with an Einstein type hypothesis regarding a balance of erosion and deposition rates 
+they assumed relationships between the entrainment and deposition rates and the excess shear stress which, again, do not agree with 
+Charru 2004 approached a bagnold-type description with a balance of erosion and deposition rates of particles
+which was extended in many ways by the Lajenesse tracer paper and the Charru2006
+However it did not make explicit reference to bagnold concepts, and is somehwat of a parallel development with similar conclusions
+
+There is an overall conclusion that Bagnold's concepts work better when many particles are in motion, and when particles are relatively small. 
+In full mobility bedload (stage 3 transport), there is a continuous bedload layer with several mobile grain layers beneath the surface (Frey2011, Moullerion2009, Jenkins1998, Hsu2004, ).
+Under these conditions, bedload motion takes the form of a granular creep, somewhat closer to the concepts of the earliest bedload transport investigations
+"This condition, which corresponds with the classical notion of bedload transport as a ‘tractive carpet flow’ (DuBoys, 1879; Sohn, 1997), is typical of sand beds and can also occur in gravel beds under strong flows.
+Much of the experimental work on bedload transport has been conducted in this condition"
+Jenkins1998 We focus on sheet flows in which the particles are supported by their collisional interactions rather than by the velocity fluctuations of the turbulent fluid and introduce what we believe to be the simplest theory for the collisional regime that captures its essential features.
+Hsu2004 computes bedload transport in a regime where collision interactions between moving grains are the only upward force, considered dominant to turbulent fluctuations
+Balance equations for the average mass, momentum and energy for the two phases are phrased in terms of concentration–weighted (Favre averaged) velocities
+Mouilleron2009 develops a revised bagnold model based upon a uniform moving particle concentration in order to compute the vertical velocity distribution within a granular sheet flow 
+Once a sheet-layer flow is fully developed, the simple exchanges assumed here may be insufficient to describe reality, as moving particles would be continuously interacting with each others.
+In this limit, continuous models based on energy dissipation inside the bed load layer [Bagnold, 1973] might be more adequate.
+Engelund says "Hence the bed load forms a ))protection shield(( at higher bed load concentrations, which controls the erosion rate."
+
+%PROBLEMS AND OBSERVATIONS:
+%bagnold entirely ignores the details of individual motions.
+There have been many attempts to clarify the Newtonian physics underlying bedload trajectories, and to incorporate these details into a bagnold type framework where an average energy budget fixes the number of particles in motion under bedload equilibrium.
+An interesting issue is that the bulk friction coefficient should depend on the amount of time that bedload grains spend in suspension between successive bed impacts, which may in turn depend on the details of bed trajectories.
+Usually, velocities of bedload are overestimated, which may relate to the ommision of collisons between moving particles Lee2002, or the need to account for shear stress imparted on the bed by saltating particles Nino1998,Bialik2015
+
+%bagnold is inadequate at low discharges VanBeek1976, Engelund1976, Nelson1995, where the solid transport rate is not suitably large to dissipate the excess energy available to move bedload grains Engelund1976, Luque1976, implying the bulk particle friction coefficient must be fit to unphysical values, higher than the bulk friction angle of sediment.
+for example, Nino1998 found a dynamic friction angle of 56.6 degrees was required to describe their experiments at low transport stage, i.e., well above the angle of repose for any natural sediment
+
+% Ancey2008 found that, at low sediment transport stages, up to 90% of the excess energy was dissipated by fluid turbulence, while at relatively high transport stages, as much as 75 percent was dissipated by bedload transport, implying a transition between bagnold and einstein-type regimes
+
+%The bagnold equations show no loss of linear stability in the coupled-fluid granular system so it fails to capture the physics of bedforms Ancey2006
+
+%Bagnold does not adapt to disequilibrium bed load transport or to mixtures of particle sizes Seminara2002,although Bridge1992 attempted to extend it to particle sizes, there is good discussion in Seminara about how this is kind of ridiculous
+
+%bagnold fails away from steady uniform flow conditions and especially over arbitrarily sloping beds Seminara2002.
+The latter problem can be rectified by appealing to Einstein-like concepts of Entrainment and depostion rates, but the former problem is not currently solvable
+
+%The mean-field treatment applied by Bagnold conflicts with a noticeable feature of sediment transport: its intermittent nature at low flow rates. Ancey2006
+
+%very difficult to clarify incipient motion conditions Mongomery1997 or bulk friction angles Kirchener1990 experimentally or theoretically, also there are indications that the optimal value of tau_c to use in bagnold type models is distinct from that which is theoretically obtained to characterize the incipient motion of sediment-- "lso, this formulation appears to be very dependent on the value of t?c, which in general is di?cult to estimate because it does not necessarily correspond to the value associated with the threshold of particle motion in the absence of bedload transport." Nino1998
+
+% Charru2004 attempts to rectify a bagnold framework with empirical data under viscous flows.
+They find the areal density of moving particles varies with time, which they link to granular compaction processes.
+Later on, this framework which borrows bagnoldean concepts with Einsteinian concepts in accord with the Parker2003 extension, is developed to take account of tracer diffusion and to describe the ripple spacing on a sand bed.
+
+% very difficult to extend bagnold concepts to multiple grain sizes: BridgeBennett1992 noted that different grains sizes have different mobilities on the bed surface, and attempted to extend the formulation to multiple grain sizes but implying that the fluid shear stress is reduced to one critical value above the big grains and another, different value over the small grains places a strain on what turbulent flows are capable of doing.
+seminara 2002
+
+%Ancey2008 analyzed energy dissipation by turbulence and by momentum transfers with moving bedload they found at low solid discharges as much as 90 percent of the energy supplied by gravity was disspated by turbulence while at higher discharges, as much as 75 percent can be disspated by momentum transfers with moving bedload so we see a continuous transition between the Einstein regime, in which turbulence is the dominant dissipative mechanism of the fluid flow, and the Bagnold regime, in which friction of moving bedload becomes the dominant mechanism.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Similarity and Difference between Einstein and Bagnold: toward a collaborative synthesis}
+
+% Similarity and Difference -- transport stages -- properties varies across transport stage; 
+"Transport of gravels and larger materials (which always move as bedload except in very exceptional circumstances) has been divided into three stages (Jackson and Beschta, 1982; Ash- worth and Ferguson, 1989; Warburton, 1992): (1) finer mate- rials overpassing a locally static bed: the material resident on the bed does not participate in the transport – the mobile material originates upstream; (2) partial transport of local bed material: at any one time part of the bed remains static, but any grain might eventually move; (3) general motion of the grains on the bed: all grains are equally apt to be moved."
+In light of our discussion of Einstein and Bagnold, we can conclude that each simplified conceptual picture really pertains to a specific stage of bedload motion. 
+In either case, individual bedload trajectories are undoubtedly stochastic: whether they are considered in relation to an Einstein or Bagnold-type formulation. 
+The essential difference between the two formulations is that, within Einstein-type formulations, the bedload flux is a collective result of individual bedload motions, while in Bagnold-type formulations, the individual motions are secondary to an over-arching collective limitation on the number of moving particles.
+Because within the Einstein framework the intermittency of individual motion is dominant,
+
+\section{
+% Synthesis -- Einstein and Bagnold are limit cases, but think like Wolman: both are really missing the real complexity -- the intermediate region
+% so it's not surprising that they both fail -- intermediate region means feedbacks with morphology introduce interactions from higher scales
+% there is a clear separation between the scales of transport and morphology evolution, meaning the feedback is accentuated
+% potential critical phenomena and a mixture of scales driving concordant evolution of bedload transport and morphology
+% implying a potentially unsteady transport rate
+% the next century should be spent exploring these morphology-transport feedbacks in accord with the ideas of classical morphology
+
+\section{Conclusion}
+
+% Mechanistic descriptions are attractive because they provide a possibility to leverage physical principles to improve our understanding of bedload motion
+% Owing to the complexity of fluid turbulence and granular physics, two main research paradigms have emerged, based upon simplified concepts -- Einstein and Bagnold
+% Einstein is based on the random characteristics of discrete particle motions, and describes a bedload rate with stochastic characteristics, where the intermittency of individual particle motions is emphasized, and the rate is better characterized by its statistics than by its mean trend
+% Bagnold is based on the average characteristics of a population of moving particles, and describes a deterministic bedload rate which is fully characterized by the average characteristics of the granular phase and fluid flow.
+% I have the perspective that the full complexity of bedload transport simplifies approximately to the simplified concepts of Einstein in Bagnold in different limit cases
+% Einstein pertains to low transport stages, when not many particles are moving during an observation period, and the role of fluid turbulence and irregularities in bed configuration are emphasized over the average characteristics of the granular and fluid phases, implying a highly fluctuating transport rate dominated by unsteady aspects of the driving and resisting forces
+% Bagnold pertains to high transport stages, when relatively many particles are moving, so there is an effective population averaging over the instantaneous characteristics of individual motions, meaning randomness can cancel out, implying a mostly steady transport rate dominated by the steady aspects of the driving and resisting forces.
+% However, neither of these limit cases is ever fully realized in gravel bed rivers. 
+% There is a need to synthesize these two limit cases from a more sophisticated framework which includes especially the interaction between bed morphology, which is formed by bedload transport and 
+% To make this synthesis, we have to get past our partisan and uncollaborative nature, and realize these mechanical theories for what they are, and quit judging them as right or wrong: they are simplified conceptual pictures of reality, with differing degrees of accord to the real world, depending on the conditions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+To attempt to answer these questions, we should start from a clear physical picture of bedload transport rates, and then consider where Einstein and Bagnold's theories fall in relation to it. 
+Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical understanding of individual bedload motions. 
+Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
+To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
+Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact.
+Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}.
+The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed.
+Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
+Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.
+
 % resting, entrainment by impulse, motion in turbulent flow, collisions with bed, deposition
@@
 % first describe broad brush ideas. initiation, motion, cessation
-To approach the question, we should develop a clear mechanical understanding of individual bedload motions which is supported by experiments, and we should consider where Einstein and Bagnold's theories fall in relation to it.
-Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
-To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
-Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact.
-Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}.
-The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed.
-Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
-Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.
 

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece3.tex
added phd thesis.

--- 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece2.tex
@@
 
-\title{On simplification and gravel rivers: Einstein and Bagnold in a modern context}
+\title{Mechanics of gravel rivers: Einstein and Bagnold in modern context}
 \author{Kevin Pierce}
@@
 
-\section{Introduction: Gravel bed rivers and mechanistic theories of bedload motion}
+\section{Introduction}
 %% gravel bed rivers, bedload transport, mechanistic theories, Einstein and Bagnold 
@@
 
-\section{Bedload fluxes from individual motions: the stochastic paradigm of Einstein}
+\section{The stochastic paradigm of Einstein}
 % Einstein -- bedload flux is individual motions -- individual motions are random -- start to stop transitions -- entrainment, motion, and deposition
@@
 
-\section{Bedload fluxes from collective dissipation: the deterministic paradigm of Bagnold}
-
-%"According to the bagnold hypothesis, the fluid shear stress at the bed equals the threshold value corresponding to incipient motion of particles; the difference %between the total bottom shear stress provides the shear stress in the solid phase.
-(Ancey2006) Interpreting the excess stress as a result of momentum transfers %between the solid and fluid phases makes it possible to compute the number of particles that can be entrained and maintained in motion, and thereby the solid flow %rate"
+\section{The deterministic paradigm of Bagnold}
 
 Bagnold formed a different simplified concept of bedload transport to understand it with mechanical principles, initatiating the deterministic paradigm of mechanistic bedload transport theory. 
-From the physical point of view, Bagnold considered bedload transport in terms of a momentum exchange between solid and liquid phases.
+From a physical point of view, Bagnold considered bedload transport in terms of a momentum exchange between solid and liquid phases.
 He balanced the average power supplied to the flow by gravity against the average frictional dissipation by moving bedload, and neglected any details of turbulent fluctuations, irregularity in the configuration of stationary grains, or the trajectories of moving particles.
 Bagnold's formulation hinges on the application of an average dynamic friction coefficient to moving bedload in order to characterize its energy dissipation. 
-His formulation is built upon a key hypothesis: when bedload transport is at its equilibrium value, so it dissipates all available power to move sediment, the tractive force imparted by the moving fluid on the bed surface will match to the threshold for sediment motion.
+His formulation is built upon a key hypothesis: when bedload transport is at its equilibrium value, so it dissipates all available power to move sediment, the tractive force imparted by the moving fluid on the bed surface will match the threshold of sediment motion.
 This is the so-called Bagnold hypothesis \citep{Engelund1976, Luque1976, Seminara2002, Ancey2006}.
-Bagnold's original formulation predicts a bedload transport rate which scales, at high stages of transport, as the average fluid shear stress to the 3/2 power: $q_b \propto \tau^{3/2},$ and his hypothesis has been taken up by generations of researchers, an enduring legacy which is even more interesting in that his bedload transport research seems to be at most a tertiary component of illustrious life's work, with a larger portion of his time devoted to his career as a decorated military commander in both world wars, and his initially recreational hobby of driving cars around the Sahara desert, which eventually gained him status as a preminent desert explorer \citep{Bagnold1988}.
+Bagnold's original formulation predicts a bedload transport rate which scales, at high stages of transport, as the average fluid shear stress to the $3/2$ power: $q_b \propto \tau^{3/2}.$ 
+His perspective on bedload transport has been taken up by generations of river scientists, and this enduring legacy is even more interesting because his research seems to be at most a tertiary component of his life's work, with a much larger portion of his time devoted to his decorated career as a military commander in both world wars, and his recreational hobby of driving cars around the Sahara desert, which eventually earned him status as a preminent desert explorer \citep{Bagnold1988}.
 
-Bagnold's original concept is enlightening for its pragmatic approach \citep{Ashida1972, Bagnold1973, Engelund1976, Luque1976}: his deterministic picture of bedload motion is simplified just enough from the real complexity to make contact with a foundational and reliable concept of classical physics, the conservation of energy.
-At the same time, his highly simplified assumptions and comparisons between experiements and his results have left many open questions for proponents of his ideas.
+Bagnold's original concept is enlightening for its pragmatic approach \citep{Ashida1972, Bagnold1973, Engelund1976, Luque1976}. 
+His deterministic concept of bedload motion is simplified just enough from the real complexity to make contact with the foundational and reliable mechanical principle of conservation of energy. 
+His theory has been influential for a clear ability to describe bedload transport from mechanical concepts at relatively high transport stages, since it predicts a transport rate which scales as the average shear stress to the $3/2$ power, in accord with a large set of classic experiments and contemporary numerical simulations \citep{MeyerPeter1948,Gomez1989,Schmeeckle2014, Elghannay2017}.
+At the same time, his highly simplified assumptions, and various comparisons between experiments and his theoretical results, have left many open questions for proponents of his ideas. 
+There is at least one criticism of Bagnold's hypothesis that the fluid should be just competent to move sediment on the bed surface under equilibrium bedload transport \citep{Seminara2002}.
+In particular, Bagnold's concept of bedload transport ignores the details of bedload trajectories, motivating many efforts to reconcile his average energy budget concepts with the dynamics of individual bedload motions.
+Finally, his theory is apparently inadequate at low discharges \citep{Engelund1976, Luque1976, Francis1977, Ancey2008}, and this realization has supported a great deal of research into why this is so, and how this problem can be rectified.
 
-% Many investigators have built upon and refined Bagnold's concepts while preserving his main idea that bedload transport rates as a consequence of an average energy budget. 
-% Early on, investigators came to the conclusion that the details of bedload trajectories are not negligible in a Bagnold-like formulation, and that bedload trajectories cannot be understood without accounting for turbulent lift forces.
-% suggesting a contingence of the bagnold formulation on the details of particle motions ( Luque and Van Beek 1976 )
-% For one thing, lifting forces on sediment grains, which are highly variable in accord with fluid turbulence, contribute to the relative amount of time that bedload grains spend in contact with the bed, meaning the dynamic friction coefficient bagnold introduced should scale with fluid lift.
-% accordingly, a large set of bagnold-type treatments focused on including fluid turbulence in-so-far as it affects bedload trajectories, but not in relation to the average power balance which controls, in Bagnold's view, the average number of partilces in motion.
-% These studies effectively contend that bedload is dominated by an average power balance, but the dissipation mechanism within this balance is contingent on fluid turbulence.
+However, the conceptual simplicity of Bagnold's ideas and the essential correctness of his results at high transport rates is striking, so many investigators have worked to refine the details of his formulation while preserving his main concept that bedload transport rates derive from an average budget between gravitational energy and frictional dissipation.
+The time spent in suspension by bedload grains between successive collisons, which is highly variable in accord with fluid turbulence and collisions with the bed surface \citep{Bialik2015}, contribute to the relative amount of time that bedload grains spend in contact with the bed, meaning the dynamic friction coefficient Bagnold introduced should scale with the characteristics of fluid turbulence, or equivalently with the details of bedload trajectories.
+Accordingly, many investigators have tried to understand the turbulent basis of bedload trajectories, and to incorporate these details with Bagnold's average energy budget concepts, which should fix the number of particles in motion \citep{Abbott1977, Bridge1984, Wiberg1989, Bridge1992, Nino1998}.
+This research is a perplexing juxtaposition of average and turbulent energy considerations, and effectively it contends that the number of particles in motion results from an average power balance, while the dissipation mechanism which limits the number of particles in motion is contingent on fluid turbulence and granular collisions.
+A shortcoming of this research is that the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Maxey1986, Schmeeckle2007, Dwivedi2010, Dwivedi2011}, with relatively important terms entering the force balance apart from the typical drag and lift \citep{Bialik2015}.
+Accordingly, these Bagnold-type models incorporating the Newtonian physics of individual bedload motions remain somewhat troubled, usually overpredicting the velocity at which individual grains travel, and therefore also overpredicting bedload transport rates \citep{Bridge1984, Bialik2015}.
+These overpredictions have been attributed to a lack of inclusion of collisions between moving grains \citep{Lee2002}, and a lack of understanding of stresses imparted by moving grains on the stationary bed \citep{Nino1998}, so both of these topics deserve further research attention \citep{Bialik2015}.
 
-For example, Bagnold's total neglect of fluid turbulence and any irregularity between sediment grains left open problems around the degree to which fluid turbulence effects and the differential mobility of grains of varying size can be incorporated into a Bagnold-type framework.
-Likewise, Bagnold's agnostic approach to individual bedload motions within his theory led many investigators to consider how these details can be reconciled with his energy dissipation concept. 
-Despite these simplifications, Bagnold's original theory is aptly descriptive of bedload transport under particular conditions: at high transport stages, bedload transport rates often appear to scale as the average shear stress to the $3/2$ power ($q_b \propto \tau^{3/2}$) in accord with his mechanical prediction, at least over flat beds sheared by a steady uniform flow \citep{MeyerPeter1948,Gomez1989,Schmeeckle2014, Elghannay2017}.
-However, Bagnold's theory is troubled at relatively low transport stages and arbitrarily sloping beds.
-To describe bedload rates at low transport stages, the empirical components of Bagnold's theory must be set to unphysical values \citep{Engelund1976, Luque1976,  Nelson1995, Nino1998}; and his deterministic theory neglects a key attribute of bedload transport rates at low stages, namely that transport is highly intermittent, so that bedload transport rates exhibit strong fluctuations which Bagnold's theory does not constrain \citep{Bohm2004, Ancey2008}.
-Over arbitrarily sloping beds, there are theoretical arguments that Bagnold's hypothesis should flatly fail, even for relatively small bed slopes transverse to the flow direction \citep{Seminara2002}.
-These problems have motivated intense research within the Bagnold paradigm.
+One criticism of Bagnold's hypothesis is that it fails under even a relatively small transverse (cross-stream) tilt in the bed surface \citep{Seminara2002}. 
+In particular, under such conditions it is shown that no bedload transport rate, no matter how large, is sufficient to reduce the fluid shear stress at the bed to the critical value.
+This undermines the Bagnold hypothesis, even under nearly horizontal beds, as it is not robust to realistic deviations from ideal conditions. 
+\citet{Parker2003} ammended the Bagnold hypothesis by replacing it with an Einstein-type concept regarding the balance of erosion and deposition rates, although they assumed somewhat ad hoc relationships to develop this connection.
+Nevertheless, these efforts key into a parallel stream of research which borrows many concepts from Bagnold, where balances between erosion and deposition rates, parameterized by average characteristics of the fluid flow, are conjoined to form bedload transport rates with same scaling patterns as Bagnold's original theories, although they do not defer to the energy balance concept to the same degree as Bagnold \citep{Charru2004, Charru2006, Lajenesse2010, Lajenesse2018}.
 
-Many researchers have concentrated on computing individual particle trajectories from Newtonian physics, and incorporating these details with Bagnold's average energy dissipation concepts, which should fix the number of particles in motion \citep{Abbott1977, Bridge1984, Wiberg1989, Bridge1992, Nino1998, Bialik2015}.
-Because the details of particle trajectories are highly contingent on fluid turbulence and collisions with the irregular bed surface, this incorporation of bedload trajectories can be viewed as a kind of first order inclusion of turbulent forcing and the random arrangement of bed grains \citep{}, which are key elements of Einstein-type models.
-A difficult attribute of this problem is that energy dissipation due to bedload should happen primarily when bedload is in contact with the bed surface, but the relative fraction of time bedload spends in contact with the bed surface is highly contingent on fluid turbulence, and especially fluid lift, and Bagnold neglected both of these aspects.
-Hence, Bagnold's dynamic friction coefficient should not really be a constant across all transport stages, and is instead expected to depend upon the details of bedload trajectories, which are inherently statistical and incompletely understood \citep{Bialik2015}.
-For one thing, the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Maxey1986, Schmeeckle2007, Dwivedi2010, Dwivedi2011}, with relatively important terms in the force balance aside from the typical drag and lift \citep{Bialik2015}.
-These models of individual bedload motions remain somewhat troubled, usually overpredicting the velocity at which bedload grains travel \citep{Bialik2015}, although the research is steadily generating new understanding.
+At low transport stages, Bagnold's theory fails to characterize bedload transport.
+In relation to this, one observation is that bedload transport rates become highly fluctuating at relatively low mean discharges, meaning mean values increasingly represent a simplified characteristic of the bedload flux, and statistical representations are really closer to the underlying physics \citep{Ancey2008}.
+However, the deepest issue with Bagnold under these conditions is that solid transport rates is not suitably large to dissipate the excess energy available to move bedload grains, meaning Bagold-type models do not fit experimental data without calibrating bulk particle friction coefficients to unphysical values \citep{Engelund1976, Luque1976, Nelson1995, Nino1998}.
+For example, \citet{Nino1998} found a dynamic friction angle of $56.6$ degrees was required to describe their experiments at low transport stage, a value well above the angle of repose for any natural sediment.
+This failure of Bagnold at low rates is apparently consequent from his underyling assumption that bedload transport is always a dominant energy dissipation mechanism of the flow. 
+\citet{Ancey2008} analyzed energy dissipation by turbulence and by momentum transfers with moving bedload.
+They found at low solid discharges, as much as $90$ percent of the energy supplied by gravity was disspated by turbulence (so $10$ percent by bedload), while at high solid discharges, a much as $75$ percent was dissipated by momentum transfers with moving bedload (so $25$ percent by turbulence).
+This data suggests that Bagnold's assumptions are only close to the physical reality of bedload transport at higher discharges, and at lower discharges, a different description is required.
 
-critical shear stresses have been difficult to consistently define and calculate
-and there are indications of difficulty in defining the concept as a consistent consequence of underlying physics
+%PROBLEMS AND OBSERVATIONS:
+%bagnold entirely ignores the details of individual motions.
+There have been many attempts to clarify the Newtonian physics underlying bedload trajectories, and to incorporate these details into a bagnold type framework where an average energy budget fixes the number of particles in motion under bedload equilibrium.
+An interesting issue is that the bulk friction coefficient should depend on the amount of time that bedload grains spend in suspension between successive bed impacts, which may in turn depend on the details of bed trajectories.
+Usually, velocities of bedload are overestimated, which may relate to the ommision of collisons between moving particles Lee2002, or the need to account for shear stress imparted on the bed by saltating particles Nino1998,Bialik2015
 
-Another deficiency of bagnold's concepts is the failure of the approach over arbitrarily sloping beds
-initially, the observation that bagnold's hypothesis did not lead to physical conclusions even for a relatively negligible transverse tilt in the bed surface
-motivated the development of an alternate formulation which preserved his average energy balance concept while understanding the number of partilces in motion as a balance of erosion and deposition rates
-Parker2003 ammended the bagnold hypothesis by replacing it with an Einstein type hypothesis regarding a balance of erosion and deposition rates 
-they assumed relationships between the entrainment and deposition rates and the excess shear stress which, again, do not agree with 
-Charru 2004 approached a bagnold-type description with a balance of erosion and deposition rates of particles
-which was extended in many ways by the Lajenesse tracer paper and the Charru2006
-However it did not make explicit reference to bagnold concepts, and is somehwat of a parallel development with similar conclusions
+% Ancey2008 found that, at low sediment transport stages, up to 90% of the excess energy was dissipated by fluid turbulence, while at relatively high transport stages, as much as 75 percent was dissipated by bedload transport, implying a transition between bagnold and einstein-type regimes
+
+%The bagnold equations show no loss of linear stability in the coupled-fluid granular system so it fails to capture the physics of bedforms Ancey2006
+
+%Bagnold does not adapt to disequilibrium bed load transport or to mixtures of particle sizes Seminara2002,although Bridge1992 attempted to extend it to particle sizes, there is good discussion in Seminara about how this is kind of ridiculous
+
+%bagnold fails away from steady uniform flow conditions and especially over arbitrarily sloping beds Seminara2002.
+The latter problem can be rectified by appealing to Einstein-like concepts of Entrainment and depostion rates, but the former problem is not currently solvable
+
+%very difficult to clarify incipient motion conditions Mongomery1997 or bulk friction angles Kirchener1990 experimentally or theoretically, also there are indications that the optimal value of tau_c to use in bagnold type models is distinct from that which is theoretically obtained to characterize the incipient motion of sediment-- "lso, this formulation appears to be very dependent on the value of t?c, which in general is di?cult to estimate because it does not necessarily correspond to the value associated with the threshold of particle motion in the absence of bedload transport." Nino1998
+
+% very difficult to extend bagnold concepts to multiple grain sizes: BridgeBennett1992 noted that different grains sizes have different mobilities on the bed surface, and attempted to extend the formulation to multiple grain sizes but implying that the fluid shear stress is reduced to one critical value above the big grains and another, different value over the small grains places a strain on what turbulent flows are capable of doing.
+seminara 2002
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Similarity and Difference between Einstein and Bagnold: toward a collaborative synthesis}
 
@@
 
-%PROBLEMS AND OBSERVATIONS:
-%bagnold entirely ignores the details of individual motions.
-There have been many attempts to clarify the Newtonian physics underlying bedload trajectories, and to incorporate these details into a bagnold type framework where an average energy budget fixes the number of particles in motion under bedload equilibrium.
-An interesting issue is that the bulk friction coefficient should depend on the amount of time that bedload grains spend in suspension between successive bed impacts, which may in turn depend on the details of bed trajectories.
-Usually, velocities of bedload are overestimated, which may relate to the ommision of collisons between moving particles Lee2002, or the need to account for shear stress imparted on the bed by saltating particles Nino1998,Bialik2015
-
-%bagnold is inadequate at low discharges VanBeek1976, Engelund1976, Nelson1995, where the solid transport rate is not suitably large to dissipate the excess energy available to move bedload grains Engelund1976, Luque1976, implying the bulk particle friction coefficient must be fit to unphysical values, higher than the bulk friction angle of sediment.
-for example, Nino1998 found a dynamic friction angle of 56.6 degrees was required to describe their experiments at low transport stage, i.e., well above the angle of repose for any natural sediment
-
-% Ancey2008 found that, at low sediment transport stages, up to 90% of the excess energy was dissipated by fluid turbulence, while at relatively high transport stages, as much as 75 percent was dissipated by bedload transport, implying a transition between bagnold and einstein-type regimes
-
-%The bagnold equations show no loss of linear stability in the coupled-fluid granular system so it fails to capture the physics of bedforms Ancey2006
-
-%Bagnold does not adapt to disequilibrium bed load transport or to mixtures of particle sizes Seminara2002,although Bridge1992 attempted to extend it to particle sizes, there is good discussion in Seminara about how this is kind of ridiculous
-
-%bagnold fails away from steady uniform flow conditions and especially over arbitrarily sloping beds Seminara2002.
-The latter problem can be rectified by appealing to Einstein-like concepts of Entrainment and depostion rates, but the former problem is not currently solvable
-
-%The mean-field treatment applied by Bagnold conflicts with a noticeable feature of sediment transport: its intermittent nature at low flow rates. Ancey2006
-
-%very difficult to clarify incipient motion conditions Mongomery1997 or bulk friction angles Kirchener1990 experimentally or theoretically, also there are indications that the optimal value of tau_c to use in bagnold type models is distinct from that which is theoretically obtained to characterize the incipient motion of sediment-- "lso, this formulation appears to be very dependent on the value of t?c, which in general is di?cult to estimate because it does not necessarily correspond to the value associated with the threshold of particle motion in the absence of bedload transport." Nino1998
-
-% Charru2004 attempts to rectify a bagnold framework with empirical data under viscous flows.
-They find the areal density of moving particles varies with time, which they link to granular compaction processes.
-Later on, this framework which borrows bagnoldean concepts with Einsteinian concepts in accord with the Parker2003 extension, is developed to take account of tracer diffusion and to describe the ripple spacing on a sand bed.
-
-% very difficult to extend bagnold concepts to multiple grain sizes: BridgeBennett1992 noted that different grains sizes have different mobilities on the bed surface, and attempted to extend the formulation to multiple grain sizes but implying that the fluid shear stress is reduced to one critical value above the big grains and another, different value over the small grains places a strain on what turbulent flows are capable of doing.
-seminara 2002
-
-%Ancey2008 analyzed energy dissipation by turbulence and by momentum transfers with moving bedload they found at low solid discharges as much as 90 percent of the energy supplied by gravity was disspated by turbulence while at higher discharges, as much as 75 percent can be disspated by momentum transfers with moving bedload so we see a continuous transition between the Einstein regime, in which turbulence is the dominant dissipative mechanism of the fluid flow, and the Bagnold regime, in which friction of moving bedload becomes the dominant mechanism.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Similarity and Difference between Einstein and Bagnold: toward a collaborative synthesis}
-
 % Similarity and Difference -- transport stages -- properties varies across transport stage; 
@@
 
-\section{
+\section{Synthesis}
 % Synthesis -- Einstein and Bagnold are limit cases, but think like Wolman: both are really missing the real complexity -- the intermediate region
@@
 In this range, both varying and averaged attributes of the factors driving and resisting bedload motion matter, and incorporating the varying components may require new theoretical approaches, borrowed from granular physics and fluid dynamics, into a fusion of Bagnold-like and Einstein-like bedload transport theories.
-Necessarily, these models should be stochastic, because they have to express stochastic transport at lower rates, but at higher rates their output distributions should sharpen, and their predicted fluxes should scale like tau^3/2, in the absence of morphological or watershed-scale influence.
+Necessarily, these models should be stochastic, because they have to express stochastic transport at lower rates, but at higher rates their output distributions should sharpen, and their predicted fluxes should scale like $\tau^3/2$, in the absence of morphological or watershed-scale influence.
 To develop these approaches, which incorporate everything we know, or select the part of it we consider relevant, we should stop quibbling. 
@@
 
-\begin{comment}
+%\begin{comment} 
+%
+%
+%Where is this coupling in Einstein or Bagnold, or under what limits might they retain predictive power? 
+%More generally, where are the deficiencies in our understanding of bedload transport, and why do our predictions in natural channels fail?%
+%
+%To arrive at modern bedload transport models, based on mechanical concepts in the schools of Einstein and Bagnold, we're led to neglect almost everything. 
+%In effect, Bagnold viewed the river as a machine, where water provides the power to move bedload, and friction on moving sediment dissipates it. 
+%Bagnold represented the power of the fluid phase by the product of the average (time and space) shear velocity with the average shear stress, and he represented the %power dissipated by the sediment phase as the energy per unit time to support a given mass of sediment against frictional drag, which he characterized in an average %way.
 
-Where is this coupling in Einstein or Bagnold, or under what limits might they retain predictive power? 
-More generally, where are the deficiencies in our understanding of bedload transport, and why do our predictions in natural channels fail?
-
-To arrive at modern bedload transport models, based on mechanical concepts in the schools of Einstein and Bagnold, we're led to neglect almost everything. 
-In effect, Bagnold viewed the river as a machine, where water provides the power to move bedload, and friction on moving sediment dissipates it. 
-Bagnold represented the power of the fluid phase by the product of the average (time and space) shear velocity with the average shear stress, and he represented the power dissipated by the sediment phase as the energy per unit time to support a given mass of sediment against frictional drag, which he characterized in an average way.
+%\end{comment}
 
@@
 
-\end{comment} 
+%\end{comment} 
 \bibliography{biblio}

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece4.tex
added phd thesis.

\documentclass{article}
% General document formatting
\usepackage[margin=0.7in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{unsrtnat}

% \usepackage[math]{kurier}
\usepackage{setspace}       % \onehalfspacing and \singlespacing
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}

\begin{document}

\title{Does gravel flow or does it bounce?\\ Bagnold and Einstein in modern context}
\author{Kevin Pierce}
\maketitle

\begin{abstract}
\end{abstract}
\section{Mechanistic theories of bedload transport}

Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
Since the transport of coarse material as bedload is an important control on channel morphology \citep{Church2006, Recking2016}, bedload transport has been the subject of intense research for over a century. 
Despite this effort, our ability to predict bedload transport rates remains poor.  
In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000 \citep{Gomez1989, Barry2004, Recking2012}. 
Prediction is difficult because individual bedload grains move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}. 
Since this bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
However, because bedload transport results from a turbulent flow interacting with granular matter, a full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and from these efforts, two of them have emerged as important research paradigms. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach, initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

These two paradigms are founded on the application of mechanical reasoning to simplified concepts of bedload motion.
These simplified concepts are characterized by empirical parameters, so these mechanistic models must be calibrated to applications. 
Owing to these simplified concepts, proponents of Einstein and Bagnold-type models take different perspectives on most aspects of bedload transport.
Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
This conflict is exascerbated by the observation that neither mechanical approach is, practically speaking, very good; and the general superiority of either approach is difficult to judge. 
For example, either mechanical paradigm generates poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult \citep{Iverson2013}, as neither approach shows uniform superiority across a full range of flow and sediment characteristics. 
Accordingly, although each mechanical approach has merits, the perspective that either is generally superior, for any reason apart from practicality, is not supported by evidence.  
Because there is no clear path to join the two paradigms together, proponents of each toil away in their respective camps, solving problems within (or with) their chosen paradigm's simplified concepts, and there is a failure to communicate science across the stochastic / deterministic rift. 
We need to ask: is this rift justified by evidence? 
Are stochastic and deterministic approaches to understand bedload transport as a mechanical process really so different that they share no common ground, and is one or the other completely wrong? 
The short answer is no, so we should work together to synthesize something new out of the strongest elements of each paradigm.
This paper is about what these elements are, and how we might do this, to understand bedload transport as a mechanical process.

\section{The stochastic paradigm of Einstein}

Einstein initiated the stochastic paradigm to describe bedload transport by relating bedload rates to individual particle motions \citep{Einstein1942, Einstein1950}.
He conceptualized the downstream transport of an individual particle as an alternating series of rests and hops, and he set out to determine the period of rest and distance covered during a hop with flume experiments \citep{Einstein1937}. 
During these experiments, he realized that the rest period and hop distance of individual particles was not consistent or predictable from one particle to the next, so he was forced to switch to a statistical conception. 
Viewing motion characteristics over a population of many particles, Einstein realized the resting period and hop distance of individual particles lie on statistical distributions, and he measured these by tracking many painted tracers \citep{Einstein1937}. 
Initially, he used the random characteristics of individual rests and motions to calculate the probability distribution of the effective velocity at which bedload grains travel as the move downstream through cycles of rest and motion. 
Owing to an effective velocity difference between the fastest and slowest grains, he described a spreading or diffusion of tracer particles as they move downstream \citep{Einstein1937}. 
Later, he derived the random character of the resting period as a mechanical consequence of fluid turbulence, and he used this concept to derive the mean rate of bedload transport \citep{Einstein1950}. 
The tenacity with which Einstein's ideas have been modified, critiqued, and improved over the last seventy years is a testament to his pioneering originality \citep{Ettema2004}, yet his clear pattern of development and refinement along the course of his life shows a motivating balance between hard work and genius \citep{Einstein1937, Einstein1942, Einstein1950, Einstein1964}.

Generations of researchers and proponents of Einstein have built up his stochastic concept of bedload motions, without fully invalidating his underlying ideas. 
Statistical characterizations of individual motions have been investigated extensively \citep{Hubbell1964, Yano1969, Nakagawa1976, Hassan1991, Habersack2001, Ancey2008, Roseberry2012, Heyman2013, Heyman2016}, and his conception of their characteristics as random variables appears essentially correct. 
Individual bedload grains move randomly, but in accord with statistical patterns dictated by turbulence and configuration of bed grains.  
Since Einstein, other characteristics of particle motion have been considered and incorporated into stochastic theories of bedload transport, such as the velocity, acceleration, and travel time during single particle motions \citep{Drake1988, Radice2006, Ancey2008, Lajeunesse2010, Furbish2012a, Roseberry2012, Furbish2015, Fathel2015, Heyman2016}. 
Understanding of particle diffusion has advanced a great deal, and the problem is obviously more nuanced than Einstein considered \citep{Einstein1937}.  
Particle diffusion has different characteristics depending on the temporal or spatial scales of consideration \citep{Nikora2002, Zhang2012, Martin2012}, and this probably relates to the effects of particle burial \citep{Sayre1971, Nakagawa1980, Voepel2013, Martin2014, Bradley2017} and to variable movement characteristics due to sediment size \citep{Fan2017} on the differences in effective velocities of moving bedload.

Einstein's mechanical theory of the bedload transport rate \citep{Einstein1950} has been reworked a great deal. 
Subsequent theories on the resting time of stationary sediment, which is equivalent to the rate of erosion of an individual surface grain \citep{Yalin1972, Papanicolaou2002, Cheng2004}, took account of the variable supporting configuration of bed surface grains, in addition to variable forcing from turbulent fluctuations \citep{Paintal1971, Wu2002, Dey2018}, and experiments have increased recognition of the role of both the magnitude and duration -- turbulent impulse -- of turbulent fluctuations on particle entrainment \citep{Diplas2008, Valyrakis2010, Celik2014}, which Einstein-like models of the resting time (or erosion rate) have not yet incorporated \citep{Dey2018}. 
Instead of computing mean bedload rates only, contemporary theories have leveraged the mathematics of stochastic processes to compute probability distributions of the bedload flux, so that mean values and the expected magnitude of bedload fluctuations are available. \citep{Lisle1998, Sun2000, Ancey2006, Ancey2008, Ma2014, Ancey2014a}. 
These descriptions provide a more physically complete description of bedload transport \citep{Ancey2008}, and higher order statistics of bedload transport provide additional benchmarks against which models can be tested \citep{Iverson2013}. 
This work has reflected back on the resting period between particle movements in a contentious way which still requires resolution. 
Einstein's concepts, when translated to theories which predicts the magnitude of bedload fluctuations, generate fluctuations which are too small to describe experiments \citep{Ancey2006}. 
Accordingly, a set of contemporary models has incorporated collective feedback mechanisms in particle motion, which could be physically attributed to the erosion of multiple grains from the bed at once \citep{Ancey2008, Heyman2013, Ma2014}, as might arise from small granular avalanches \citep{Heyman2013}, the effects of coherent turbulence on particle motion \citep{Nino1998, Amir2014, Santos2014, Shih2017}, or the migration of bedforms \citep{Dhont2018}. 
Research from the stochastic paradigm which Einstein initiated has generated a great deal of new understanding and many open questions about the underlying physics of bedload transport, and this progress is expected to continue.

However, many elements of Einstein's conception are problematic. 
For one thing, there are multiple modes within the statistics characterizing particle rest and motion, and no clear indiciation of the extent to which these must be resolved in bedload transport theory. 
For example, particles can rest on the bed surface, and this resting period can be characterized probabilistically \citep{Einstein1950}, but these resting times are known to be variable across bedforms, as a function of the relative exposure of bed particles to the flow \citep{Crickmore1962, Sayre1967, Yang1971, Nakagawa1980}. 
Also, particles can rest within the bed surface, or in a state of burial \citep{Voepel2013, Martin2014, Olinde2015, Bradley2017}, and the resting statistics associated with the burial process are different than those associated with surface resting. 
Similarly, when bedload particles move, they can do so in different modes. 
For example, they can roll, slide, or bounce, and within each of these modes they may have different motion statistics \citep{Bohm2004, Frey2006, Frey2014}. 
The degree to which these different modes must be accounted for in stochastic bedload transport theory has not been clarified.  
The key limitation of Einstein-type models is at high transport stages. 
Under these conditions, the partition of bedload transport into separate periods of motion and rest breaks down, because the entire bed surface moves as a sheet flow with motion several particle diameters below the apparent surface \citep{Jenkins1998, Mouilleron2009, Houssais2015}. 
We can conclude that, although Einstein-type models appear close to the underlying physics at low bedload discharges \citep{Ancey2008, Heyman2013, Ma2014}, high bedload discharges require a different description altogether.

\section{The deterministic paradigm of Bagnold}

Bagnold formed a different simplified concept of bedload transport to understand it with mechanical principles, initatiating the deterministic paradigm of bedload transport theory. 
From a physical point of view, Bagnold considered bedload transport in terms of a momentum exchange between solid and liquid phases.
He balanced the average power supplied to the flow by gravity against the average frictional dissipation by moving bedload, and neglected any details of turbulent fluctuations, irregularity in the configuration of stationary grains, or the trajectories of moving particles.
Bagnold's formulation hinges on the application of an average dynamic friction coefficient to moving bedload in order to characterize its energy dissipation. 
His formulation is built upon a key hypothesis: when bedload transport is at its equilibrium value, so it dissipates all available power to move sediment, Bagnold held that the tractive force imparted by the moving fluid on the bed surface will match the threshold of sediment motion \citep{Bagnold1973}. 
This is the so-called Bagnold hypothesis \citep{Engelund1976, Luque1976, Seminara2002, Ancey2006}.
Bagnold's original formulation predicts a bedload transport rate which scales, at high stages of transport, as the average fluid shear stress to the $3/2$ power: $q_b \propto \tau^{3/2}.$ 
His perspective on bedload transport has been taken up by generations of river scientists, and this enduring legacy is even more interesting because his scientific research seems to be at most a tertiary component of his life's work, with a much larger portion of his time devoted to his decorated career as a military commander in both world wars, and to his hobby of driving cars around the Sahara desert, which eventually earned him status as a preminent desert explorer \citep{Bagnold1988}.

Bagnold's original concept is enlightening for its pragmatic approach \citep{Ashida1972, Bagnold1973, Engelund1976, Luque1976}. 
His deterministic concept of bedload motion is simplified just enough from the real complexity to make contact with the foundational and reliable mechanical principle of conservation of energy. 
His theory has been influential for a clear ability to describe bedload transport from mechanical concepts at relatively high transport stages, since it predicts a transport rate which scales as the average shear stress to the $3/2$ power, in accord with a large set of experiments and numerical simulations \citep{MeyerPeter1948,Gomez1989,Schmeeckle2014, Elghannay2017}.
At the same time, his highly simplified assumptions, and various comparisons between experiments and his theoretical results, have left many open questions for proponents of his ideas. 
Bagnold's fundamental hypothesis, that the fluid should be just competent to move sediment on the bed surface, has been undermined by theoretical arguments \citep{Seminara2002}, and Bagnold's concept of bedload transport ignores the details of bedload trajectories, which has motivated many efforts to reconcile his average energy budget concept with the dynamics of individual bedload motions. 
Finally, his theory is apparently inadequate at low discharges \citep{Engelund1976, Luque1976, Francis1977, Ancey2008}, and this realization has supported a great deal of research into why this is so, and how this problem can be rectified.

However, the conceptual simplicity of Bagnold's ideas and the essential correctness of his results at high transport rates is striking, so many investigators have worked to refine the details of his formulation while preserving his main concept that bedload transport rates derive from an average budget between gravitational energy and frictional dissipation. 
The time spent in suspension by bedload grains between successive collisons, which is highly variable in accord with fluid turbulence and collisions with the bed surface \citep{Bialik2015}, contribute to the relative amount of time that bedload grains spend in contact with the bed, meaning the dynamic friction coefficient Bagnold introduced should scale with the the details of bedload trajectories, which are themselves contingent on turbulence. 
Accordingly, many investigators have tried to understand the turbulent basis of bedload trajectories, and to incorporate these details with Bagnold's average energy budget concept, which should fix the number of particles in motion and hence the bedload rate \citep{Abbott1977, Bridge1984, Wiberg1989, Bridge1992, Nino1998}.
This research is a perplexing juxtaposition of average and turbulent energy considerations, as it effectively contends that the number of particles in motion results from an average power balance, while the dissipation mechanism which limits the number of particles in motion is contingent on fluid turbulence and granular collisions. 
A shortcoming of this research is that the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Maxey1986, Schmeeckle2007, Dwivedi2010, Dwivedi2011}, with relatively important terms entering the force balance apart from the typical drag and lift \citep{Bialik2015}.
Accordingly, these Bagnold-type models based on the Newtonian physics of individual bedload motions remain troubled, usually overpredicting the velocity at which individual grains travel, and therefore overpredicting bedload rates \citep{Bridge1984, Bialik2015}. 
These shortcomings have been attributed to a lack of inclusion of collisions between moving grains \citep{Lee2002}, and a lack of understanding of stresses imparted by moving grains on the stationary bed \citep{Nino1998}, and both of these topics deserve further research attention \citep{Bialik2015}.

The theoretical argument which undermines Bagnold's hypothesis is based upon the addition of a relatively small transverse (cross-stream) tilt in the bed surface \citep{Seminara2002}. 
Under such conditions, \citet{Seminara2002} showed that no bedload transport rate, no matter how large, is sufficient to reduce the fluid shear stress at the bed to the critical value.
This undermines the Bagnold hypothesis, even for nearly horizontal beds, as the hypothesis is apparently not robust to realistic deviations from ideal conditions. 
\citet{Parker2003} ammended the Bagnold hypothesis by replacing it with an Einstein-type concept regarding the balance of erosion and deposition rates, although they assumed somewhat ad hoc relationships between erosion and deposition rates and the average fluid shear stress to develop this connection. 
Nevertheless, these efforts key into a similar stream of research which borrows many concepts from Bagnold, balancing erosion and deposition rates, parameterized by average characteristics of the fluid flow, to generate bedload transport predictions with the same scaling patterns as Bagnold's original formulas, although these theories do not defer to Bagnold's average energy balance concept \citep{Charru2004, Charru2006, Lajenesse2010, Lajenesse2018}.

At low transport stages, Bagnold's theory fails to characterize bedload transport.
One issue is that bedload transport rates express relatively large fluctuations at relatively low mean discharges, meaning mean values poorly characterize the bedload signal at low rates, and statistical representations are really closer to the underlying physics \citep{Ancey2008}.
However, the deepest issue with Bagnold at low transport stages is that solid transport rates are not suitably large to dissipate the excess energy available to move bedload grains, meaning Bagold-type models do not fit experimental data without calibrating bulk particle friction coefficients to unphysical values \citep{Engelund1976, Luque1976, Nelson1995, Nino1998}.
For example, \citet{Nino1998} found an effective friction angle of $56.6$ degrees was required to describe their experiments at low transport stage, a value well above the angle of repose of any natural sediment \citep{Miller1966}.
This failure of Bagnold-type theories at low rates is a consequence of the underlying assumption that bedload transport is always a dominant energy dissipation mechanism of the flow. 
\citet{Ancey2008} estimated energy dissipation by turbulence and by momentum transfers with moving bedload.
They found at low solid discharges, as much as $90$ percent of the energy supplied by gravity was disspated by turbulence (so $10$ percent by bedload), while at high solid discharges, a much as $75$ percent was dissipated by momentum transfers with moving bedload (so $25$ percent by turbulence). 
The data suggest that Bagnold's energy balance assumptions are close to the physical processes of bedload transport at higher discharges, but at lower discharges, a different physical description is required.

\section{Flowing, bouncing, or both?}

The rift between the Bagnold and Einstein research paradigms is not justified. 
Proponents of Einstein build bedload transport rates up from individual particle motions, and contend that the motion characteristics of individual particles are not predictable due to the random character of fluid turbulence and bed grain configuration. 
When bedload discharges are low, so that motions are relatively rare, the random character of individual motions is emphasized within the bedload rate, and it appears as a statistical quantity, so a full statistical description based on individual motions is closer to the relevant physics \citep{Ancey2008, Heyman2013, Ma2014, Heyman2016}.
At low bedload rates, collective properties of bedload motion, such as the average rate of energy dissipation, are unsteady, so these variables alone do not produce a meaningful description. 
Bagnold's proponents, in contrast, understand bedload transport from the collective properties of bedload motion. 
When bedload discharges are high, so that many individuals combine to form a continuous sheet flow, the random character of individual motions is washed out and mixed through inter-granular collisions and the law of large numbers, so average characteristics dominate the bedload rate, and a deterministic description based upon collective properties of a granular flow is really closer to the relevant physics \citep{Jenkins1998, Hsu2004, Mouilleron2009, Frey2011, Houssais2016, Maurin2018}.
Hence we can view the Einstein and Bagnold paradigms as idealized mechanical concepts of bedload transport which are realistic at two extremes: relatively low and high discharges over a flat bed. 
Bedload particles bounce at low discharges and flow at high discharges: the rift between Bagnold and Einstein is a remnant of our oversimplified picture of bedload transport phenomena, and really they are each pertinent to a simplified extreme of bedload phenomena, while the intermediate region defies physical description.  
We have to remember that the problem of treating bedload transport as a physics problem is not expected to be simple.
Should coupling a turbulent fluid to a granular medium be simpler than either phase alone \citep{Pope2000, Baule2016}?

There are many such problems in physics, where idealized models describe extremes in which particular interactions are more relevant, but intermediate regions, where all interactions matter, remain a profound mystery.
One example is the physics of solids, which are just collections of electrons and ions \citep{Ashcroft1986}.   
On one extreme, we have the ideal metal, where electrons are free to move throughout the solid, unbound to the ions, and they dominate its phenomena. 
They interact with light to make the metal reflective, and their individual characteristics (electrostatic repulsion, Fermi pressure) make the metal exceedingly strong. 
This idealized model is deeply understood and aptly descriptive, and was essentially completed by the 1930s \citep{Ashcroft1986}. 
On another extreme, we have the ideal insulator, where electrons are totally bound to the ions and cannot bound, so that the motion of charge neutral bodies dominate its phenomena. 
They are relatively inert to light, so insulators are white or grey; and their vibrations dominate the flow of heat: these phenomena were essentially described by the 1910s, and in large part by Hans Albert Einstein's father \citep{Ashcroft1986}. 
Yet the intermediate region, where electrons are neither fully free nor bound, has been one of the most tumultuous of modern physics, expressing superconductivity, exotic magnetic states, solids which insulate on the inside which conduct on the surface. 
The physics of the intermediate region are exotic, and their description requires the development of new theoretical approaches. 
This is the condition of Einstein and Bagnold. 
In one extreme we have an intermittent bouncing of grains under fluid shear, and in another extreme we have a collective flow, while the intermediate region is both, or maybe neither; its description requires new theoretical developments, and these are expected to fuse the two extremes.

% introduce dimensionless transport rate and mobility parameter

\section{Toward mechanistic understanding of gravel bed rivers}

But in this analogy, where are the exotic phenomena of the intermediate region? That is, what is exotic about intermediate bedload discharges, and in what way is their mechanical description difficult? 
\citet{Wolman1960}, in their classic theoretical work on geomorphology, introduced a concept of balance between the frequency and magnitude of hydaulic conditions in reworking river channels.  
Although high bedload discharges may be incredibly geomorphically active, and they may drastically change river morphology over short timescales, these discharges are rare. 
The majority of geomorphic change may result from intermediate bedload discharges which are both relatively common and competent enough to overcome the stability of river channels and exert control over their morphology. 
Under many conditions, we can understand part of the emergence, dynamics, and stability of channel morphology, which has elements distributed across a range of spatial scales, as a consequence of intermediate bedload discharges: the same discharges which defy description from Bagnold or Einstein's mechanical paradigms. 
The exotic phenomena of bedload transport, in the intermediate region between Bagnold and Einstein, are the feedbacks between transport and morphology.

From the coordination of bedload transport with sediment supply and storage, a wide diversity of morphological elements are expressed within gravel bed rivers across a range of spatial scales \citep{Brayshaw1984, Church1998, Hassan2008, Nelson2014, Venditti2017}, and these structural elements provide stability to bed grains and modify in-channel hydraulics, in turn coordinating bedload transport \citep{Laronne1976, Lisle1992, Wilcock2003a, Kasprak2014, Recking2016, Hassan2017}. 
Within the intermediate region, between the idealized concepts of Einstein and Bagnold, we have a problem of emergence.
Bedload transport generates and modifies these morphological features, but it also responds to them.  
In gravel bed rivers, these feedbacks mix the intermittency of individual motions with the collective dynamics of morphological features, and express a bedload flux with variations on multiple scales \citep{Hoey1992, Cudden2003, Nelson2010, Saletti2015, Dhont2018}. 
We can speculate that more advanced physical theories of bedload motion, which will be predictive within the intermediate range, will synthesize Einstein and Bagnold's paradigms with a careful understanding of channel morphology to understand this coordination between transport and morphology. 
The success of Bagnold's paradigm at high transport rates shows us that when motion is dense enough to imitate a continuous flow, the transport is governed by its dissipating effects on the fluid flow.
Einstein's paradigm is successful in the opposite extreme; when motions are sparse, they result primarily from variations in driving or resisting forces and have a relatively minimal effect on power dissipation in the fluid flow. 
What we really need is a theory containing both elements: an energy budget with provisions for the formation, degradation, and transport of morphological features; and intermittent motion, stemming from variability in driving and resisting forces on the scale of individual grains. 
This synthesis of Einstein and Bagnold might get us quite far toward addressing problems around gravel bed rivers.

\bibliography{biblio}
\end{document}

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece6.tex
added phd thesis.

\documentclass{article}
% General document formatting
\usepackage[margin=0.7in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[utf8]{inputenc}
\usepackage{subfig}         % side-by-side figures 
% Related to math
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{unsrtnat}

% \usepackage[math]{kurier}
\usepackage{setspace}       % \onehalfspacing and \singlespacing
\newcommand\be{\begin{equation}} % shortcut to start eq envs 
\newcommand\ee{\end{equation}}   % shortcut to end eq envs
\newcommand\ol{\overline}        % shortcut to draw overline 
\newcommand\bra{\langle}
\newcommand\ket{\rangle}

\begin{document}

\title{Does gravel flow in rivers, or does it bounce?\\ similarity and difference between Bagnold and Einstein}
\author{Kevin Pierce}
\maketitle

\begin{abstract}
\end{abstract}
\section{Mechanistic theories of bedload transport}

Gravel-bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological problems \citep{Gaeuman2016}.
Since the transport of coarse material as bedload is an important control on channel morphology \citep{Church2006, Recking2016}, bedload transport has been the subject of intense research for over a century. 
Despite this effort, our ability to predict bedload transport rates remains poor, and predictions of bedload transport can deviate from measured values by factors of 100 or 1000 \citep{Gomez1989, Barry2004, Recking2012}. 
Bedload grains move intermittently by rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}. 
Since this transport is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
However, bedload transport results from a turbulent flow interacting with granular matter, so a full mechanical description would be a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to construct mechanical theories of bedload transport using hypotheses where necessary, and from over a century of effort, two have emerged as important research paradigms. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach, initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

These two paradigms are founded on the application of mechanical reasoning to simplified concepts of bedload motion.
These concepts involve empirical parameters, so these mechanistic models must be calibrated to applications. 
Owing to these simplified concepts, proponents of Einstein and Bagnold-type models take different perspectives on most aspects of bedload transport, which often leads to disagreements and miscommunications between proponents of each school. 
This conflict coincides with the observation that neither mechanical approach is, practically speaking, very good.
Both paradigms generate poorer bedload transport predictions than simple power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}
And the superiority of either paradigm across the full range of bedload conditions is difficult to judge, owing to their empirical parameters \citep{Iverson2013}. 
Accordingly, although each mechanical approach has merits, the perspective that either is generally superior, for any reason apart from practicality, is not supported by evidence. 
Meanwhile, there is no clear way to join the two paradigms together, so proponents of each toil away in their respective camps, solving problems within (or with) their chosen paradigm's simplified concepts, in context of a general failure to communicate science across the conceptual rift. 
We need to ask: is this rift justified by evidence? 
Are stochastic and deterministic approaches to understand bedload transport as a mechanical process really so different that they share no common ground, and is one or the other completely wrong? 
This commentary explores this question.

\section{The Einstein paradigm: mechanistic-stochastic}

% einstein concepts
% einstein successes
% - his concept of indivdual motions as random is essentially correct
% - his research provoked a deeper study and understanding of diffusion and the statistics of particle motion
% - stochastic bedload transport, when coupled to fluid flow, shows a linear instability which implies bedform generation Jerolmack Ancey Bohorquez
% einstein shortcomings
% - we expect bedload to dissipate energy but this is not included in formulations
% - diffusion has been difficult and more complicated than expected 
% - einstein fails at high bedload rates because motion-rest states break down and we have a granular sheet flow
% einstein takeaway 
% - more work is needed, but it... 
% - succeeds at low transport rates, and 
% - stochastic transport generates bedforms

% Einstein concepts
Einstein initiated the stochastic paradigm by relating bedload rates to individual particle motions \citep{Einstein1942, Einstein1950}.
He conceptualized the transport of an individual particle as an alternating series of rests and hops, and he set out to determine the period of rest and distance of one hop with flume experiments \citep{Einstein1937}. 
During these experiments, he realized that the rest period and hop distance of individual particles was not consistent or predictable from one particle to the next, so he was forced to switch to a statistical perspective. 
Viewing motion characteristics over a population of many particles, Einstein realized the resting period and hop distance of individual particles generated statistical distributions, and he measured these distributions in his experiments \citep{Einstein1937}. 
Initially, he used these distributions to calculate the probability distribution of the effective velocity of bedload grains as they move downstream through cycles of rest and motion. 
Because there is an effective velocity difference between the fastest and slowest grains, he derived the rate of spreading or diffusion of tracer particles as they move downstream \citep{Einstein1937}. 
Later, he derived the random character of the resting period as a mechanical consequence of fluid turbulence, and he used this concept to derive the mean rate of bedload transport \citep{Einstein1950}.

% einstein successes
% - his concept of indivdual motions as random is essentially correct
% - his research provoked a deeper study and understanding of diffusion and the statistics of particle motion
% - stochastic bedload transport, when coupled to fluid flow, shows a linear instability which implies bedform generation Jerolmack Ancey Bohorquez
Einstein's stochastic approach to understand bedload transport has been criticised and improved by many researchers, without abandoning his essential ideas. 
Since Einstein, statistical characteristics of individual motions have been investigated extensively and incorporated into stochastic theories of bedload transport, including characteristics he did not originally consider, such as the velocity, acceleration, and travel time of particles \citep{Hubbell1964, Yano1969, Nakagawa1976, Hassan1991, Habersack2001, Ancey2008, Roseberry2012, Fathel2015, Heyman2016}.
Einstein's conception of bedload motion characteristics as random variables appears essentially correct.
Understanding of particle diffusion has advanced a great deal, and the problem is obviously more nuanced than Einstein considered.  
Particle diffusion has different characteristics depending on the temporal or spatial scales of consideration \citep{Nikora2002, Zhang2012, Martin2012}, and this probably relates to the effects of particle burial \citep{Sayre1971, Nakagawa1980, Voepel2013, Martin2014, Bradley2017} and variable movement characteristics from sediment size \citep{Fan2017} on the differences in effective velocities of moving bedload. 
A key success of Einstein's theory is that stochastic bedload transport,

% einstein shortcomings
% - we expect bedload to dissipate energy but this is not included in formulations
% - diffusion has been difficult and more complicated than expected 
% - einstein fails at high bedload rates because motion-rest states break down and we have a granular sheet flow

% einstein takeaway 
% - more work is needed, but it... 
% - succeeds at low transport rates, and 
% - stochastic transport generates bedforms

Generations of researchers have evaluated, criticised, and improved Einstein's stochastic concepts of bedload transport, without abandoning his basic ideas. 
Statistical characteristics of individual motions have been investigated extensively \citep{Hubbell1964, Yano1969, Nakagawa1976, Hassan1991, Habersack2001, Ancey2008, Roseberry2012, Heyman2013, Heyman2016}, and Einstein's conception of bedload motion characteristics as random variables appears essentially correct. 
Individual bedload grains move randomly, but in accord with statistical patterns dictated by turbulence and configuration of bed grains.  
Since Einstein, other characteristics of particle motion have been measured and incorporated into stochastic theories of bedload transport, such as the velocity, acceleration, and travel time during single particle motions \citep{Drake1988, Radice2006, Ancey2008, Lajeunesse2010, Furbish2012a, Roseberry2012, Furbish2015, Fathel2015, Heyman2016}.

Einstein's mechanical theory of the bedload transport rate \citep{Einstein1950} has been reworked a great deal. 
Subsequent theories deriving the resting time of stationary sediment, which is equivalent to the rate of erosion of an individual surface grain \citep{Yalin1972}, took account of variation in the supporting configuration of bed surface grains, in addition to variable forcing from turbulent fluctuations in their Newtonian analyses \citep{Paintal1971, Wu2002, Dey2018}.
Meanwhile, contemporary experiments have increased recognition of the role of both the magnitude and duration, or the impulse, of turbulent fluctuations on particle entrainment \citep{Diplas2008, Valyrakis2010, Celik2014}, which Einstein-like models of the resting time (or erosion rate) have not yet incorporated \citep{Dey2018}. 
Instead of computing mean bedload rates only, contemporary theories have leveraged the mathematics of stochastic processes to compute probability distributions of the bedload flux, deriving mean values and the expected magnitude of bedload fluctuations \citep{Sun2000, Ancey2006, Ancey2008}. 
These descriptions provide a more complete description of bedload transport \citep{Ancey2008}, and higher order statistics of bedload transport provide additional benchmarks to test models against experiments \citep{Iverson2013}. 
This work has linked back on the resting period between particle movements in a contentious way which still requires resolution. 
Einstein's concepts, when translated to theories which predicts the magnitude of bedload fluctuations, generate fluctuations which are too small to describe experiments \citep{Ancey2006}. 
Accordingly, some contemporary models have incorporated collective feedback mechanisms in particle motion, which could be physically attributed to the erosion of multiple grains from the bed at once \citep{Ancey2008, Heyman2013, Ma2014}, as might arise from small granular avalanches \citep{Heyman2013}, the effects of coherent turbulence on particle motion \citep{Nino1998, Amir2014, Santos2014, Shih2017}, or the migration of bedforms \citep{Dhont2018}.

Bedload theories within the Einstein paradigm have provided a great deal of understanding about the physics of bedload transport; and his conception is clearly more representative of bedload fluxes at low transport stages, where instantaneous fluctuations in bedload rates are appreciable relative to mean values, taking values up to 400\% as large \citep{Bohm2004, Ancey2008, Singh2009, Heyman2016}. 
However, Einstein type theories have several key limitations, especially at high transport stages and with spatial or temporal differences in transport characteristics.  
For one thing, Einstein-type theories are energetically flawed, because they make no reference to the energy carried by the flow; intutively, we expect that the water stream cannot hold an unlimited number of particles in motion, because these particles dissipate energy in their transportation, and the flow has a finite amount. 
Under high transport stages, Einstein's partition of bedload transport into separate periods of motion and rest breaks down, because the entire bed surface moves as a sheet flow, with motion several particle diameters below the surface \citep{Jenkins1998, Mouilleron2009, Houssais2015}. 
When transport characteristics differ in space or time, such as differences in the stability of bed sediment caused by morphology, or the spatial sorting of different size grains, researchers of the Einstein paradigm have very limited capacity to predict this difference in bedload transport statistics, meaning much more research is needed. 
Finally, no limiting factor of bedload rates within einstein-type models 
We can conclude although Einstein-type models are close to the underlying physics of low bedload discharges over a nearly flat bed \citep{Ancey2008, Heyman2013, Ma2014}, but at high sediment discharges or for any discharge over an irregular bed, a different description is required.

\section{The Bagnold paradigm: mechanistic-deterministic} 
% bagnold concept
% bagnold successes
% - describes transport rate at high discharges Julien 1994
% - in accord with many classic experiments and contemporary numerical simulations 
% bagnold shortcomings
% - determinisitic formulation contrasts with key feature of bedload rates at low discharges -- fluctuations
% - bagnold fails at low discharges because friction coefficient must be fit to unphysical values 
% - bagnold's hypothesis is flawed, so his formulation gives poor results over arbitrarily sloping beds
% - if bedforms result from a loss of linear stability in coupled fluid-solid system, bagnold does not capture requisite physics Balmforth 2001 
% bagnold takeaway
% - bagnold is realistic at high bedload rates, but 
% - the failure at low rates is a consequence of the underlying assumption that bedload transport is always a dominant energy dissipation mechanism
% - however this contrasts with ancey findings, 
% - we conclude that bagnold's range of validity is limited to high bedload rates over relatively flat beds

% bagnold concept
Bagnold formed a different simplified concept of bedload transport to understand it with mechanical principles, initatiating the deterministic paradigm of bedload theory. 
From a physical point of view, Bagnold considered bedload transport as a momentum exchange between solid and liquid phases.
He balanced the average power supplied to the flow by gravity against the average frictional dissipation by moving bedload, and neglected any details of turbulent fluctuations, irregularities in the configuration of stationary grains, or the trajectories of moving particles.
Bagnold's formulation hinges on the application of an average dynamic friction coefficient to moving bedload, in order to characterize its energy dissipation. 
His formulation is built upon the so-called Bagnold hypothesis \citep{Bagnold1973, Engelund1976, Luque1976, Seminara2002, Ancey2006}: when bedload transport is at its equilibrium value, so it dissipates all available power to move sediment, Bagnold held that the tractive force imparted by the moving fluid on the bed surface will match the threshold of sediment motion.
Bagnold's original formulation predicts a bedload transport rate which scales, at high stages of transport, as the average fluid shear stress to the $3/2$ power: $q_b \propto \tau^{3/2},$ which is in accord, at high bedload discharges, when morphology is poorly developed, with a large set of experiments and numerical simulations \citep{MeyerPeter1948,Gomez1989,Schmeeckle2014, Elghannay2017}.
For this success, his perspective on bedload transport has been taken up by generations of river scientists.
Bagnold's scientific legacy is even more interesting because his research seems to be at most a tertiary component of his life's work, with a much larger portion of his time devoted to his decorated career as a military commander in both world wars, and to his hobby of driving cars around the Sahara desert, which eventually earned him status as a preminent desert explorer \citep{Bagnold1988}.

Bagnold's original concept is enlightening for its pragmatic approach \citep{Ashida1972, Bagnold1973, Engelund1976, Luque1976}. 
His deterministic concept of bedload motion is simplified just enough from the real complexity to make contact with the foundational and reliable mechanical principle of conservation of energy. 
His theory has been influential for a clear ability to describe bedload transport from mechanical concepts at relatively high transport stages, when the influence of morphology is minimized, and for its accord with classic results which also predict a scaling of bedload transport as the average shear stress to the $3/2$ power \citep{MeyerPeter1948}.
At the same time, his highly simplified assumptions have left many open questions for proponents of his ideas, and his key hypothesis has been undermined by theoretical arguments \citep{Seminara2002}.

For example, Bagnold's theory of bedload transport ignores the details of bedload trajectories, which has motivated many efforts to reconcile his average energy budget concepts with the dynamics of individual bedload motions. 
The time spent in suspension by bedload grains between successive collisons, which is highly variable in accord with fluid turbulence and collisions with the bed surface \citep{Bialik2015}, contribute to the relative amount of time that bedload grains spend in contact with the bed, meaning the dynamic friction coefficient Bagnold introduced should scale with the the details of bedload trajectories, which are themselves contingent on turbulence. 
Accordingly, many investigators have tried to understand the turbulent basis of bedload trajectories, and to incorporate these details with Bagnold's average energy budget concept, which should fix the number of particles in motion and hence the bedload rate \citep{Abbott1977, Bridge1984, Wiberg1989, Bridge1992, Nino1998}.
This research is a juxtaposition of average and turbulent energy considerations, as it effectively contends that the number of particles in motion results from an average power balance, while the dissipation mechanism which limits the number of particles in motion is contingent on fluid turbulence and collisions between moving and stationary grains. 
A shortcoming of this research is that the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Schmeeckle2007, Dwivedi2010, Dwivedi2011}, especially when particle shapes deviate from spherical \citep{Maxey1989}, with relatively important terms entering the force balance apart from the typical drag and lift terms \citep{Bialik2015}.
For this reason, and others, these Bagnold-type models based on the Newtonian dynamics of individual bedload motions remain troubled, usually overpredicting bedload rates \citep{Bridge1984, Bialik2015}. 
This problem has been attributed to a lack of inclusion of collisions between moving grains \citep{Lee2002}, and a lack of understanding of stresses imparted by moving grains on the stationary bed \citep{Nino1998}, so both of these topics deserve further research attention \citep{Bialik2015}.

The theoretical argument which undermines Bagnold's hypothesis is based upon the addition of a relatively small transverse (cross-stream) tilt in the bed surface \citep{Seminara2002}. 
Under such conditions, \citet{Seminara2002} showed that no bedload transport rate, no matter how large, is sufficient to reduce the fluid shear stress at the bed to the critical value.
This undermines the Bagnold hypothesis, even for nearly horizontal beds, as the hypothesis is apparently not robust to even minor deviations from ideal conditions. 
\citet{Parker2003} ammended the Bagnold hypothesis by replacing it with an Einstein-type concept regarding the balance of erosion and deposition rates, although they used ad hoc relationships between erosion and deposition rates and the average fluid shear stress to form this connection. 
Nevertheless, these efforts key into a similar stream of research which borrows many concepts from Bagnold, balancing erosion and deposition rates parameterized by average characteristics of the fluid flow, to generate bedload transport predictions with the same scaling patterns as Bagnold's original formulas. Although these theories successfully describe experiments, they do not defer to Bagnold's average energy balance concept \citep{Charru2004, Charru2006, Lajenesse2010, Lajenesse2018}.

The key failure of Bagnold-type theories is at low transport stages, where they fail to characterize bedload transport \citep{Engelund1976, Luque1976, Francis1977, Ancey2008}.
One issue is that measured bedload rates express relative large fluctuations at low mean discharges, so mean values poorly represent the bedload signal at low rates, and statistical representations are really closer to the underlying physics \citep{Ancey2008}.
However, the deepest issue with Bagnold at low transport stages is that solid transport rates are not suitably large to dissipate the excess energy available to move bedload grains, meaning Bagold-type models do not fit experimental data without calibrating bulk particle friction coefficients to unphysical values \citep{Engelund1976, Luque1976, Nelson1995, Nino1998}.
For example, \citet{Nino1998} found an effective friction angle of $56.6$ degrees was required to describe their experiments at low transport stage, a value well above the angle of repose of any natural sediment \citep{Miller1966}.
This failure of Bagnold-type theories at low rates is a consequence of the underlying assumption that bedload transport is always a dominant energy dissipation mechanism of the flow. 
\citet{Ancey2008} estimated energy dissipation by turbulence and by momentum transfers with moving bedload.
They found at low solid discharges, as much as $90$ percent of the energy supplied by gravity was disspated by turbulence (so $10$ percent by bedload), while at high solid discharges, a much as $75$ percent was dissipated by momentum transfers with moving bedload (so $25$ percent by turbulence). 
The data suggest that Bagnold's energy balance assumptions are close to the physical processes of bedload transport at higher discharges over a nearly flat bed, but at lower discharges, or when the bed is irregular, a different physical description is required.

\section{Stochastic or Deterministic?}

% at low bedload discharges in the absence of morphology you have stochastic properties, while at high bedload discharges they become deterministic 
% there is an effective population averaging when many particles are in motion 
% accordingly bedload transport has different characteristics at low and high transport rates, and einstein and bagnold mirror these differences 
% reference ancey's energy dissipation findings
% reference ancey's findings on the scaling of bedload rates

The rift between the Bagnold and Einstein research paradigms is not justified, because these two paradigms pertain to different idealized conditions of bedload transport: relatively high or low solid discharges, respectively, over flat or nearly flat beds.
At low discharges, bedload rates are highly variable, even under the most controlled laboratory conditions available. 
For example, in laboratory experiments at low bedload discharges under uniform steady flows, with identical glass beads or narrowly graded gravel as sediment, over flat or nearly flat beds, instantaneous bedload fluxes can take on values as much as 400\% mean values \citep{Bohm2004, Ancey2008, Heyman2016}. 
However, as mean bedload discharge is increased, the relative strength of bedload fluctuations decreases, so the mean bedload rate becomes more representative of bedload transport \citep{Ancey2008}. 
Proponents of Einstein build bedload transport rates up from individual particle motions, and contend that the motion characteristics of individual particles are not predictable due to the random character of fluid turbulence and bed grain configuration. 
When bedload discharges are low, so that motions are relatively rare, the random character of individual motions is emphasized within the bedload rate, and it appears as a statistical quantity, so a full statistical description based on individual motions is closer to the relevant physics \citep{Ancey2008, Heyman2013, Ma2014, Heyman2016}.
In contrast, proponents of Bagnold understand bedload transport from average characteristics of bedload motion. 
When bedload discharges are high, so that many individuals combine to form a continuous sheet flow, the random character of individual motions is washed out through inter-granular collisions, so average characteristics dominate the bedload rate, and a deterministic description based upon collective properties of a granular flow is really closer to the relevant physics \citep{Jenkins1998, Hsu2004, Mouilleron2009, Frey2011, Houssais2016, Maurin2018}.
Accordingly, an application or comparison of Bagnold and Einstein-type models across a wide range of transport stages is not meaningful, because bedload transport has different characteristics at low and high discharges. 
It's heavily stochastic at low discharges, and quasi-deterministic at high discharges.

% so bagnold and einstein describe two extremes

\section{Toward deeper understanding of gravel bed rivers}

% wolman's frequency and magnitude argument suggests that intermediate bedload conditions can be a dominant control on morphology
% schumm implies this linkage is actually more complex, and involves the interplay of storage and supply as well 
% In light of these perspectives, we can assocaite particular conditions of storage and supply, in conjunction with intermediate bedload conditions, as the conditions which foster morphology creation
% morphology has feedbacks with sediment transport
% meaning intermediate bedload regimes support strong morphology-transport feedbacks

\citet{Wolman1960}, in their classic theoretical work on geomorphology, introduced a concept of balance between the frequency and magnitude of hydaulic conditions in reworking river channels.  
Although high bedload discharges, like those described by the Bagnold paradigm, may be incredibly geomorphically active, and they may drastically change river morphology over short timescales, these discharges are rare. 
The majority of geomorphic change may result from intermediate bedload discharges which are both relatively common and competent enough to overcome the stability of river channels and enact small changes in their morphology over longer intervals of time.  
\citet{Shumm1960}, in another classic work, introduced a complementary perspective of river morphology. 
He viewed it as an interplay between the storage, supply, and transport of sediment. 
Joining these perspectives, we can associate particular conditions of sediment storage and supply, along with intermediate conditions of bedload transport, between the extremes of Bagnold and Einstein, as the genesis of a wide diversity of morphological elements within gravel bed rivers \citep{Brayshaw1984, Church1998, Hassan2008, Nelson2014, Venditti2017}.
These morphological elements, including clusters, ribs, bars, and so on, provide stability to bed grains and modify in-channel hydraulics, in turn coordinating bedload transport \citep{Laronne1976, Lisle1992, Wilcock2003a, Kasprak2014, Recking2016, Hassan2017}. 
We can conclude in the intermediate bedload conditions which neither the Einstein or Bagnold paradigms capture, there is an intricate feedback between the formation, migration, and destruction of collective sedimentary forms.
Individual motions generate morphological structures, which tend to migrate or disintegrate in collective motion.  
The exotic phenomena of bedload transport, in the intermediate region between Bagnold and Einstein, are the feedbacks between transport and morphology.

% the intermittency of individual motion, informed by turbulent fluctuations, develops morphological elements
% under particular storage / supply conditions, 
% these form, migrate, and disintegrate in an interplay between individual and collective transport events
% and this superposition of individual and collective motions imparts distinct signatures to bedload fluctuations across multiple temporal and spatial scales
% accordingly, the phenomenology of the intermediate bedload regime is really a fusion of Einstein and Bagnold
% Bagnold fails to generate bedforms, while Einstein succeeds, meaning a stochastic component of transport is necessary to form morphological elements
% at the same time, these collective structures act at larger scales, so presumably they form appreciable energy sinks, meaning they cannot be described 
% independent of

So in the intermediate range, between the idealized concepts of the Einstein and Bagnold paradigms, we have a problem of emergent interactions. 
Bedload transport generates and modifies morphological features, but it also responds to them.  
In gravel bed rivers, these feedbacks mix individual motions with the collective formation, motion, and destruction of morphological features, such as bars, bedload sheets, or clusters; and as a result, within the intermediate range, bedload fluxes exhibit variations across multiple temporal and spatial scales, which can be associated with these different processes \citep{Hoey1992, Cudden2003, Nelson2010, Saletti2015, Dhont2018}. 
We can speculate that more advanced physical theories of bedload motion will synthesize Einstein and Bagnold's concepts, with a careful understanding of channel morphology, to understand this coordination between transport and morphology.
There are essentially three components to this synthesis. 
First, the success of Bagnold's paradigm at high transport rates shows us that when motion is dense enough to imitate a continuous flow, the transport is governed by its dissipating effects on the fluid flow; so we need to build deeper understanding of the energetics and dynamics of channel morphology. 
Second, Einstein's paradigm is successful in the opposite extreme. 
When motions are sparse, they result from anomalies in driving or resisting forces, and have a realtively minimal effect on power dissipation in the fluid flow. 
However, these irregular motions spawn morphological features, which can migrate or disintegrate in collective motion, more similar to the paradigm of Bagnold. 
Third, need to synthesize energy approaches with more realistic models based on inidivudal motions
since neither einstein or bagnold aren't too developed, should work on these separately
but they should also be fused

so no reason to not work on einstein and bagnold separately 
should probably also work on both

Einstein's paradigm is successful in the opposite extreme; when motions are sparse, they result primarily from variations in driving or resisting forces and have a relatively minimal effect on power dissipation in the fluid flow. 
What we really need is a theory containing both elements: an energy budget with provisions for the formation, degradation, and transport of morphological features; and intermittent motion, stemming from variability in driving and resisting forces on the scale of individual grains. 
This synthesis of Einstein and Bagnold might get us quite far toward addressing problems around gravel bed rivers.

\bibliography{biblio}
\end{document}

### 2021-09-03 revision/formerPapers/ens-rev/abstract.tex
added phd thesis.

\begin{abstract}

A foundational problem in river science is the calculation of the bedload flux, or the rate of movement of bed material. 
This is a significant problem because bed material movement shapes river channels and determines habitat suitability to a large extent.  
Conceptualizing bedload transport as a random switching between motion and rest states leads naturally to an analogy with ecological population modeling. 
Erosion, movement, and deposition are analogous to birth, migration, and death. 
As in ecology, details of these processes are considered difficult to predict, but their rates can be measured. 
These process analogies have opened up new ways to model bedload transport as a stochastic process.
I review approaches which model the bedload flux as a birth-death process. 
These approaches are exciting because they provide a new theoretical framework for understanding the statistical properties of bedload fluxes -- fluctuations, intermittency, spatial and temporal correlations. 
However, existing approaches contain many simplifying assumptions, so the literature review reveals a handful of obvious opportunities for extensions to these birth-death models.

The review is used to define a research trajectory, which I extrapolate from in order to discern a set of possible extensions. 
I develop some of these extensions myself, and present (1) a new birth-death model incorporating bed surface affects in sediment entrainment and deposition for multiple size fractions, (2) a joint probabilistic modeling of bed elevation and the bedload flux, and (3) a two-dimensional birth-death theory of bedload motion, where particles are free to move in longitudinal and transverse directions. 
After these extensions are completed, I conclude with a discussion of the scope and limitations of birth-death approaches, and I frame some possible extensions in terms of interdisciplinary literature in physics, chemistry, and ecology. 
The wider directive of this synthetic review is to foster future research in birth-death modelling of bedload transport, because I believe this theoretical framework provides a foundation for deeper understanding of the relationship between channel morphology and bedload fluctuations, with potential to clarify both of these subjects, each of  contemporary relevance to ecology and engineering.

\end{abstract}

\begin{comment}

Bedload moves intermittently, so it is well suited to describtion by stochastic mathematics.  
In the last several decades, the stochastic approach to modeling bedload transport has been reworked, lending it a firmer mathematical and mechanical foundation.

- the rate of downstream movement of bed material, or the bedload flux, is a foundational and important unsolved problem in river science.
- it has important applications in widespread environmental science
- in the last two decades, stochastic approach has been reworked leading to a firmer mathematical and mechanical foundation 
- Withi
\end{comment}

### 2021-09-03 revision/formerPapers/ens-rev/background.tex
added phd thesis.

\section{Background: the foundational stochastic model of bedload transport} 
\label{sec:background}

\citet{Einstein1950} can be credited with the first attempt to understand the bedload flux as random quantity.  
Einstein understood the movement of individual bedload grains as a random succession of motion and rest intervals.
He termed the transition from rest to motion entrainment, and characterized it by a probability related to fluid turbulence. 
When particles are set in motion, Einstein considered they moved in discrete jumps of the same average distance. 
We define the bedload flux as the volume of grains passing a stream cross section per unit time.
It is a volume of grains per time per unit width of stream, so its units are $\text{area}/\text{time}$.  
Therefore, Einstein's mean flux formula is derived from summing the alternate start-stop motions of all bed particles which cross a surface in a time interval.

The conceptual picture Einstein considered is depicted in figure \ref{fig:yalin}. 
All bed particles are identical and are considered to have the same average geometry when at rest on the bed, meaning their entrainment characteristics are identical. 
For continuity with the rest of the review, we consider sediment particles are spheres of radius $a$, although \citet{Einstein1950} was more general. 
We can follow Einstein to compute the mean number of these particles crossing a flow cross-sectional surface $S$ in an interval of time. 
This will construct the mean bedload flux $\bra q_s \ket$.
In subsequent sections, we consider $q_s$ a random variable, and we will review approaches to derive its full probability distribution.
From these contemporary models, an Einstein-like mean flux formula emerges from taking the mean of the probability distribution for $q_s$.  
Hence we use averaging brackets $\bra \cdot \ket$ for the sake of continuity.

\begin{figure}
  \centering
  \includegraphics[width=.98\linewidth]{./figures/yalindrawing.png}
  \caption{Einstein's conceptual picture (adapted from \citet{Yalin1972}). Particles move in discrete jumps of length $L$ from left to right through an array of adjacent control volumes. The bedload flux is the volume of bedload particles crossing the surface $S$ per unit width and time. \label{fig:yalin}}
\end{figure}

Einstein's derivation goes as follows: we partition the channel into a sequence of identical control volumes $V$. 
Each volume has dimensions $L$, $w$, and $h$ ($V=lwh$). 
$L$ is the downstream length of each control volume, and it is also the average jump distance of particles. 
$w$ and $h$ are the width and height of each control volume.
Let $P_n$ be the probability that any individual grain on the bed is entrained \textit{at least} $n$ times during the time interval $T$. 
That is, let $P_n$ be the probability that an individual grain undergoes \textit{at least} $n$ jumps of length $L$ in an interval $T$, meaning it travels a distance $nL$ or more in $T$. 
Considering that each control volume contains $N_V$ grains at rest within it, it follows that on average
\be N_V P_n. \ee
grains are displaced \textit{at least} a distance $nL$ from a control volume within the time interval $T$.

Now we derive the mean flux. 
Note that the grains passing through a cross-sectional surface $S$ (of area $lw$) in the time $T$ could have come from any upstream location. 
Therefore, the number of grains crossing $S$ in $T$ is 
\be \sum_{n=1}^\infty N_V P_n. \ee
This is the net arrival of grains in the interval $T$ from all upstream control volumes. 
Multiplying by the particle volume $\nu_p$ and dividing by the timescale $T$ and channel width $w$ gives the mean bedload flux (volume per width per time):
\be \bra q_s \ket = \frac{\nu_p N_V}{w T}\sum_{n=1}^\infty P_n = A \frac{a^2}{T} \sum_{n=1}^\infty P_n. \label{eq:einstein1}\ee 
In the second equality, we have introduced $\nu_p = 4\pi a^3/3$, and two assumptions of Einstein. First, is the reasonable assumption that the number $N_V$ of particles on the surface within $V$ is proportional to the ratio of bed surface and particle areas: $N_V \propto Lw/a^2$, and second is the assumption that the travel distance $L$ is proportional to particle size: $L \propto a$. The latter assumption is more difficult to justify \citep[e.g.][]{Yalin1972}. 
$A$ is a constant of proportionality.

Einstein implies the probability that an individual grain travels a distance $nL$ or more in $T$ is equivalent to the probability that an individual grain travels at least a distance $L$ successively $n$ times in $T$. 
Hence $P_n = P_1^n$ by the law of multiplication of probabilities, so that $\sum_{n=1}^\infty P_n = \sum_{n=1}^\infty P_1^n = P_1/(1-P_1)$, using the geometric series (noting $P_1 < 1$ because it is a probability). 
Owing to this substitution, Einstein's mean bedload formula takes the form 
\be  \bra q_s \ket = A \frac{a^2}{T}\frac{P_1}{1-P_1}, \label{eq:einstein2} \ee
where $A$ is a constant of proprotionality.  
The $P_1/(1-P_1)$ structure of Einstein's formula is a distinctive characteristic. 
In order to use the apply the Einstein formula to predict bedload transport in natural streams or experiments, the timescale $T$ and the probability $P_1$ that at least one grain entrains in $T$ must be determined. 
Then the formula can be calibrated to a given setting by adjusting the constant $A$.

In the original work, Einstein considered the timescale $T$ proportional to the "time of the grain" formed by the grain's size $a$ and its settling velocity $w$ in still fluid: $T \propto a/w$. 
We will revisit this assumption shortly. 
Regarding the probability $P_1$, one of Einstein's most influential and enduring ideas within hydraulic engineering is his conception of the probability $P_1$ that at least one particle entrains in $T$. 
He considered that entrainment is driven by the fluctuating lift force imparted by the turbulent fluid, and it is resisted by the submerged weight of grains. 
Therefore, he formulated $P_1$ as an exceedance probability of the lift force over weight, so he wrote $P_1$ as an integral over the probability distribution of fluid shear stress. 
Einstein developed an alternative perspective on the incipient motion of particles that embraced turbulence and did not rely on the critical shear stress concept \citep[e.g.][]{Shields1936}.

This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly random) resisting quantities has been generalized and extended a great deal. 
\citet{Paintal1971} made a significant extension by including random granular geometry of bed particles into the force balance. 
More recently, \citet{Tregnaghi2012} ammended the theory to include the impulse concept: the experimentally verified idea that force magnitude and duration are of equal importance for entrainment \citep[e.g.][]{Diplas2008, Celik2014}. 
Refined theories of the entrainment probability, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2008, Dey2018}, and they are a subject of intensive ongoing research.

As noted by \citet{Yalin1972}, who wrote a careful review and revision of \citet{Einstein1950}, the Einstein theory for $\bra q_s \ket$ is theoretically sound, at least to equation \ref{eq:einstein2}, and provided Einstein's probabilities $P_n$ are interpreted in a careful way. 
I have mimiced Yalin to say $P_n$ is the probability of \textit{at least} $n$ jumps in $T$. 
Had I said $P_n$ was the probability of (only) $n$ jumps in $T$, the relationship $P_n = P_1^n$ would not follow, so the Einstein flux would not develop the notable $P_1/(1-P_1)$ form as in \ref{eq:einstein2}. 
Einstein was not as clear as he could have been on this distinction. 
Notably, the more general birth-death theories which reproduce Einstein as a limiting case \citep[e.g.][]{Ancey2006} use probability concepts relating to (only) one entrainment in a time interval. 
Therefore, to link back to Einstein from these contemporary theories, we need to follow \citet{Yalin1972} to express the mean flux $\bra q_s \ket$ in terms of the probabilities of (only) $n$ entrainments in a time interval, which we can denote by a lower case letter $p_n$. 
These two probabilties, $P_n$ (at least) and $p_n$ (only), have probably been conflated often within the literature.

Yalin appealed to scaling arguments in order to refine Einstein's theory. 
He highlighted that Einstein's prescription of $T \propto a/w$, where $w$ is the settling velocity, was not supported by any experimental evidence or theoretical argument. 
Yalin noted that within Einstein's theory, "$T$ appears in connection to the study of detachment of individual grains due to turbulent fluctuations", from which he concluded that $T$ should be actually be measured in terms of a period $\tau$ of turbulent fluctuations: 
\be T = N_\tau \tau, \label{eq:yalintime}\ee
where $N_\tau$ is the dimensionless number of turbulent fluctuations in $T$ (which is considered unknown but very large).

Yalin introduces a probability $p$ of (only) one entrainment during an individual turbulent fluctuation. 
Since the number $N_\tau$ of turbulent fluctuations in $T$ is very large, he considers this probability is very small, so that the probability of (only) $n$ detachments in $T$ can be computed with the Poisson distribution: 
\be p_n = \frac{(N_\tau p)^n}{n!}e^{-N_\tau p}. \label{eq:yalinpoisson} \ee
The probability $P_n$ of \textit{at least} $n$ detachments in $T$ can be expressed in terms of the probability $p_n$ of (only) $n$ detachments in $T$ as 
\be P_n = \sum_{i=n}^\infty p_i. \label{eq:yalinsum}\ee
Owing to equations \ref{eq:yalintime}, \ref{eq:yalinpoisson}, and \ref{eq:yalinsum}, the mean bedload flux \ref{eq:einstein1} becomes
\be  \bra q_s \ket = A \frac{a^2}{N_\tau \tau} \sum_{n=1}^\infty \sum_{i=n}^\infty p_i = A \frac{a^2}{N_\tau \tau} \sum_{n=1}^\infty n p_n = A \frac{a L}{\tau} p. \label{eq:yalinflux} \ee
The unknown number of turbulent fluctuations $N_\tau$ cancels. 
Hence, Yalin's ammended Einstein formula is proportional to $p$, rather than $P_1/(1-P_1)$ as in \citet{Einstein1950}. 
In Yalin's mind, the probability $p$ is the probability of entrainment of only one grain in a small interval $\tau$ related to the period of turbulent fluctuations.

When Einstein's conceptual picture of bedload transport is revisited from a Markov process framework, the probability distribution $P(q_s)$ of the bedload rate can be derived from his assumptions \citep{Ancey2006}. 
This distribution implies a mean bedload formula $\bra q_s \ket = \sum q_s P(q_s)$ formally similar to Yalin's Einstein-like formula \ref{eq:yalinflux}, and it provides additional information about the magnitude of bedload fluctuations, which are characterized without ambiguity by the variance $\bra \delta q_s^2 \ket$, where $\delta q_s = q_s - \bra q_s \ket$ is the deviation of the bedload flux from the mean.
Jumping ahead a little, the fluctuations $\bra (\delta q_s)^2 \ket$ derived in this way are not large enough to match experimental data \citep{Ancey2006}, and this is because \citet{Einstein1950}, \citet{Yalin1972}, and \citet{Ancey2006} each assumed the transitions between rest and motion were independent between all particles.

\citet{Ancey2008} fixed this problem by describing bedload transport within a more general birth-death process framework \citep{Cox1965}. 
They included a collective interaction into the entrainment rates of bed particles in their model, and with this inclusion they derived bedload fluctuations of realistic magnitude. 
Their collective interaction biases entrainment, making it more likely when particles are already moving, a point we will discuss.  
It leads to clouds of active particles, and a wide tail on the bedload flux probability distribution: both of these features are in accord with experimental observations \citep{Drake1988, Ancey2006, Ancey2008}. 
I'll start the review with \citet{Ancey2006}, which is essentially \citet{Einstein1950} revisited from the birth-death framework, lending it more generality by treating the bedload flux as a probability distribution.   
The mean of this probability distribution reproduces equation \ref{eq:yalinflux}, which is a satisfying historical coherence. 
Followups and refinements of the \citet{Ancey2006} work constitute the bulk of the review in section \ref{sec:review}.

### 2021-09-03 revision/formerPapers/ens-rev/reviewsections/ancey2008.tex
added phd thesis.

--- 2021-09-03 draft/formerPapers/ens-rev/reviewsections/ancey2008_0.tex
@@
 By the standard way of developing master equations \citep[e.g.][]{Cox1965}, we find an infinite system of equations for the probabilities of finding $n$ moving particles within the control volume: 
-\begin{multline} P(n,t+\delta t) = \alpha \delta t (n+1)P(n+1,t) \\+  [\lambda + (n-1)\mu]\delta t P(n-1,t) \\ + [1-\delta t[\lambda + n\alpha + n\mu]]P(n,t). \end{multline}
+\begin{multline} P(n,t+\delta t) = \alpha \delta t (n+1)P(n+1,t) \\+  [\lambda + (n-1)\mu]\delta t P(n-1,t) \\ + [1-\delta t[\lambda + n\alpha + n\mu]]P(n,t). \label{eq:ancey2008master}\end{multline}
 Here $\alpha = \nu + \sigma$ summarizes the contributions of emigration and deposition, which both act to decrease the number of particles $n$, and $\lambda = \lambda_0 + \lambda_1$ summarizes the contributions of immigration and (individual) entrainment, which both act to increase the number of particles $n$.
@@
 If there are initially $N_0$ moving particles within the control volume, the solution is 
-\be G(z,t) + \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \label{eq:generator}\ee
+\be G(z,t) = \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \label{eq:generator}\ee
 The factor $K$ is the autocorrelation function $K(t) = \exp(-t(\alpha-\mu)).$
@@
 The mean number of particles in the control volume, taken over the distribution \ref{eq:negbin}, is 
-\be \bra n \ket = \sum_{n=0}^\infty n P(n) = \frac{\lambda}{\alpha-\mu}. \label{eq:anceymean}\ee
+\be \bra n \ket = \sum_{n=0}^\infty n P(n) = \frac{\lambda \alpha}{\alpha-\mu}. \label{eq:anceymean}\ee
 Likewise the variance is 
-\be \bra (\delta n)^2 \ket = \sum_{n=0}^\infty (n-\bra n\ket)^2 P(n) = \frac{ \lambda \alpha}{\alpha-\mu^2},\ee
+\be \bra \delta n^2 \ket = \sum_{n=0}^\infty (n-\bra n\ket)^2 P(n) = \frac{ \lambda \alpha}{\alpha-\mu^2},\ee
 which means their ratio is 
-\be \frac{\bra (\delta n)^2}{\bra n\ket} = \frac{1}{1-\mu/\alpha}.\ee
+\be \frac{\bra \delta n^2\ket }{\bra n\ket} = \frac{1}{1-\mu/\alpha}.\ee
 It is interesting to compare this to equation \ref{eq:2006flucts} bounding the magnitude of fluctuations in the \citet{Ancey2006} model. 
@@
 Armed with the probability distribution for the number of particles in the control volume, equation \ref{eq:negbin}, and the link between control volume and surface statistics derived in section \ref{sec:fluxdef}, equation \ref{eq:fluxy}, we can derive the probability distribution of the bedload flux in a similar way as equation \ref{eq:anceygauss}.
-The number of particles in the control volume has been implicitly assumed to be infinite, so there is no need or means to take a large $N$ limit.
-
 Using the law for transformation of probabilities gives 
@@
 as the probability distribution of the bedload flux $q_s$.
+The bedload flux, as derived in \citet{Ancey2008}, is a random variable with a negative binomial distribution.
 
 Now we link back to the \citet{Einstein1950} theory. 
-Einstein considered neither immigration ($\lambda_1$), emigration ($\nu$), nor collective entrainment ($\mu$). 
-When all of these parameters are set to zero, $\mu = \lambda_1 = \nu = 0$, the parameters $r$ and $p$ of the distribution \ref{eq:negbin0} become $r \rightarrow \infty$ and $p=0$. 
-In this limit the distribution \ref{eq:negbin0} tends to a Poisson distribution.
+Einstein considered neither immigration ($\lambda_0$), emigration ($\nu$), nor collective entrainment ($\mu$). 
+When all of these parameters are set to zero, $\mu = \lambda_0 = \nu = 0$, the parameters $r$ and $p$ of the distribution \ref{eq:negbin0} become $r \rightarrow \infty$ and $p=0$. 
+With these limits, the distribution \ref{eq:negbin0} tends to a Poisson distribution. 
+There are several ways to show this. 
+The simplest way is to take the appropriate limits ($t \rightarrow \infty$, $\mu=\lambda_1=\nu= 0$)  into the generating function \ref{eq:generator}, giving
+\be G(z) = e^{-\lambda_0(z-1)/\sigma}.\ee
+This inverts to ($P(n) = \frac{1}{n!}(\partial/\partial z)^n G(z)|_{z=1}$)
+\be P(n) = \frac{(\lambda_1/\sigma)^n}{n!}e^{-\lambda_1/\sigma},\label{eq:anceypoisson}\ee
+the Poisson distribution with rate $\lambda_0/\sigma$.
+The \citet{Ancey2008} model links to the limiting result of \citet{Ancey2006}, equation \ref{eq:ancey2006limit2}.
 
-There are two ways to see this. 
-The first way is to take the appropriate limits ($t \rightarrow \infty$, $\mu=lambda_1=\nu= 0$)  into the generating function \ref{eq:generator} gives 
-\be G(z) = e^{-\lambda_0(z-1)/\sigma},\ee
-which inverts to 
-\be P(n) = \frac{(\lambda_0/\sigma)^n}{n!}e^{-\lambda_0/\sigma},\label{eq:anceypoisson}\ee
-the Poisson distribution.  
-The second way is to take limits of the negative binomial distribution equation \ref{eq:negbin0} directly, although this is somewhat tricky.
-It involves reparameterizing the distribution by its mean value equation \ref{eq:anceymean} rather than $r$, and then taking limits, using the limit formula $e^{-x} = \lim_{s \rightarrow \infty} (1-x/s)^s$.
+Using equation \ref{eq:fluxy} with this Poisson distribution, equation \ref{eq:anceypoisson}, we compute the number of active paricles $n$ in the absence of collective entrainment: 
+\be \bra q_s \ket = \frac{\nu_p u_s}{L w} \sum_{n=0}^\infty n \frac{(\lambda_1/\sigma)^n}{n!}e^{-\lambda_1/\sigma} = \frac{\nu_p u_s}{L w} \frac{\lambda_1}{\sigma}. \label{eq:ancey2008mean}\ee
+To connect equation \ref{eq:ancey2008mean} with the Einstein-like formula of Yalin, equation \ref{eq:yalinflux}, requires a delicate interpretation of the parameters $\lambda_0$ and $\sigma$. 
+The rate of entrainment $\lambda_1$ pertains to all particles within the control volume, of which there is a very large number $N \rightarrow \infty$. 
+Meanwhile, the rate of deposition $\sigma$ pertains to an individual particle. 
+These interpretations are consistent with the master equation \ref{eq:ancey2008master}. 
+Hence if we define the probability that (only) one particle entrains in a small interval $\delta t$ as $\lambda'$, then $\lambda = N\lambda'$, and the mean flux is 
+\be \bra q_s \ket = \frac{\nu_p u_s}{L w}\frac{N\lambda'}{\sigma}.\ee
+If we can again interpret the ratio $N/(Lw)$ as proportional to the area of an individual particle, using $\nu_p \propto a^3$, and the travel distance $l\propto a$, we get 
+\be \bra q_s \ket = A \frac{a l}{a/u_s}\frac{\lambda'}{\sigma}, \ee
+again in accord with the Yalin result \ref{eq:yalinflux}.
 
-In either case, arriving at this Poisson distribution for $n$ in the absence of collective entrainment, we can compute the mean bedload flux using equation \ref{eq:fluxy}: 
-\be \bra q_s \ket = \frac{\nu_p u_s}{L w} \sum_{n=0}^\infty n \frac{(\lambda_0/\sigma)^n}{n!}e^{-\lambda_0/\sigma} = \frac{\nu_p u_s}{L w} \frac{\lambda_0}{\sigma}. \ee
-To make full contact with the Einstein-like formula of Yalin, equation \ref{eq:yalinflux}, one further step is needed.
+We have shown that the assumptions of \citet{Einstein1950} can be mixed with stochastic mathematics to develop a more coherent and physically-based theory of the bedload flux, as performed by \citet{Ancey2006}.
+Since the fluctuations predicted by the Gaussian bedload flux of \citet{Ancey2006} are not sufficiently large to describe experiments, we have mimiced \citet{Ancey2008} to include collective motion effects in order to express bedload fluctuations of sufficient magnitude from a birth-death model.
+Turning off the extra processes in the \citet{Ancey2008} model leads to a coherent connection back to a limiting result of \citet{Ancey2006}, and further to back to Yalin's revision of Einstein \citep{Yalin1972}.
 
-In the \citet{Ancey2008} collective entrainment model, there was no limitation on the number of particles within the control volume. 
-In the notation of section \ref{sec:ancey2006}, this is equivalent to $N \rightarrow \infty$. 
-In \citet{Ancey2006} the fraction of time spent in motion, $\xi$, referred to a single particle. 
-We can show in the model under consideration that $\lambda_0/\sigma$ also represents a fraction of time spent in motion, although this is more difficult than it was in the arguments leading to the analogous conclusion for $\xi$.
+However, there is obviously some ambiguity around the number of particles available for motion, which was denoted $N$ in previous sections in this review. 
+Apparently, the \citet{Ancey2008} model considers the $N\rightarrow \infty$ limit, while the probability that any one particle is observed in motion goes to zero, similar to the Poisson limit of the \citet{Ancey2006} model.
+Of course, the number of particles available for motion is not unlimited. 
+Over a purely alluvial bed composed to gravel, it must be limited by the number of particles on the bed surface. 
+Within the idealization of \citet{Einstein1950}, this constraint was expressed by the equation $N a^2 \propto Lw$.
 
-Similar to equation \ref{eq:}, one can set up a master equation to compute the distribution of time time periods that the number $n$ remains fixed while no transitions except deposition occur.
-
- that particles spend in motion and to find the mean period spent in motion is $\lambda_0$.
-
-- this model assumes infinite sediment availability. there is no limit to n. This leads to turowski
-- the flux is defined in a control volume, emigration provides a pathway to defining the flux at a point. This is Heyman 2013, and Ma 2014. 
-- imagine connecting cells, this is ancey2014, which leads into Heyman 2014
-
-\begin{comment}
-
-When the collective entrainment process is turned off, $\mu\rightarrow 0$, the Poisson distribution for the number of moving particles within the control volume is recovered, reproducing the limiting result of \citet{Ancey2006}, which has links to \citet{Einstein1950}.
-We will consider this limit of no collective entrainment. 
-$\mu\rightarrow 0$, implies $r \rightarrow \infty$ and $p \rightarrow 1$. 
-In this limit, the generating function G(z) becomes 
-\be G(z) = \exp(-\frac{\lambda}{\alpha}[z-1]).\ee
-Taylor expanding this generating function around $z=0$ generates the hirearchy of probabilities $\pi_n(t)$ in the limit of no collective entrainment:
-\be \pi_n = \frac{(\lambda/\alpha)^n}{n!}e^{-\lambda/\alpha}.\ee
-Therefore the number of particles in the control volume is Poisson distributed, in accord with the large $L$ limit of the \citet{Ancey2006} model in the previous section \ref{eq:poissonflux}. 
-To complete the connection we turn off immigration ($\lambda_0=0$) and emigration ($\nu = 0$) as well, so that the distribution of the number of particles in the control volume reproduces the \citet{Ancey2006} result cleanly:
-\be \pi_n = \frac{(\lambda_1/\sigma)^n}{n!} e^{-\lambda_1/\sigma},\ee
-which is the same as equation \ref{eq:poissonflux}. 
-We note that immigration ($\lambda_0$ -- to move in) and emigration ($\nu$ -- to move out) do not fundamentally alter the form of the distribution.
-Only collective entrainment ($\mu$) does this.
-
-Now we will analyze the probability distribution of bedload flux and the magnitude of its fluctuations. 
-Collective entrainment develops a negative binomial distribution for the number of active particles, and in the limit of a large control volume ($L \rightarrow \infty$), the probability distribution for the number of active particles links to the probability distribution of the bedload flux via the relationship \ref{eq:fluxy} of the previous section.
-As a reminder, this relationship was derived using an artificial device. 
-To develop an ensemble of particles intersecting a flow cross section $S$ with various positions $\textbf{x}$ and velocities $\textbf{u}_p$, we traded a frequentist interpretation of a conditional probability $P[\textbf{u}_p|\textbf{x}]$ for an ensemble one.
-This was only valid in the limit of an infinitely long control volume $L \rightarrow \infty$.
-
-The \citet{Ancey2008} model reveals another way to compute the flux which does not depend on this device. 
-Considering the downstream boundary of the control volume as the surface $S$ in the formula \ref{eq:anceyflux}, we can count the number of emigration events across this surface per unit time.  
-There are four types of transitions in the \citet{Ancey2008} model, and these occur at random intervals. 
-Counting the frequency of only one of these transitions is a non-trivial problem, and it relies on advanced techniques in stochastic mathematics. 
-This counting problem was solved by \citet{Ma2014}.
-They computed the bedload flux without the need for an exchange of averaging or an artificial control volume limit ($L\rightarrow \infty$). 
-We will consider this work later in the review.
-
-For now, consider the approach of the previous section. 
-The bedload flux is $q_s \approx \frac{\nu_p u_s}{L} n$, in the limit of a large control volume $L\rightarrow \infty$ and when all particles in motion approach the same velocity $u_s$.
-The mean and variance of the negative binomial distribution are 
-\be \bra n \ket = \frac{\lambda}{\alpha-\mu} \ee
-and 
-\be \bra (\delta n)^2 \ket = \frac{\lambda \alpha }{\alpha-\mu}.\ee
-Because of collective entrainment, characterized by $\mu$, the variance can exceed the mean in this model: 
-\be \frac{\bra (\delta n)^2 \ket}{\bra n \ket } = \frac{1}{1-\mu/\alpha}, \ee
-because $1/(1-\mu/\alpha) > 1$ if $\mu > 0 $.
-
-By these relationships, the mean flux becomes 
-\be  \bra q_s \ket = \frac{\nu_p u_s}{L} \ee,
-and the magnitude of bedload fluctuations is
-
-The Ancey et al 2008 inclusion of collective entrainment develops a Negative Binomial distribution of the bedload flux. 
-Expressing the bedload flux as the number of moving particles within the control volume times their average velocity $u_p$ makes a Negative binomial distribution for bedload flux within the control volume.
-The mean bedload flux is $ \frac{\lambda}{\alpha-\mu} u_p $ and the variance of the flux is $ \frac{\lambda \alpha}{(\alpha-\mu)^2} u_p^2$: the fluctuations are much wider in the negative binomial distribution, meaning the inclusion of collective entrainment can model the large fluctuations observed in the \citet{Ancey2008} experiments.
-
-One way to view the bedload flux we just presented: it is the number of moving particles times their velocity. 
-However, since this model explicitly includes downstream emigration from the control volume, there is another way to define the flux: it is the number of emigration events within a unit of time. 
-Counting the number of emigration events expressed from this model within a unit time is not easy. 
-\citet{Ma2014b} explored this approach to calculating the bedload flux, and I will review this in a subsequent section.
-
-Two key assumptions of this model are worth examining. 
-These set the stage for the subsequent developments in birth-death modelling of the bedload flux which I review next. 
-First, this model assumes that stationary bed particles are in infinite supply. 
-This is an implicit assumption: Nothing prevents the system from entraining particles ad infinitum through the entrainment processes characterized by rates $\lambda_0$ and $\mu$. 
-Second, the model considers sediment transport in an isolated control volume, so the sediment flux is described probabilistically in a locality, but spatial correlations cannot be accounted for: the flux upstream of the control volume is also a fluctuating quantity, implying spatial correlations through immigration events into the control volume considered in this model.
-
-These two observations support two of the more recent developments to the \citet{Ancey2008} birth-death immigration-emigration model: \citet{Turowski2009} generalized this model to finite bed material in order to describe transport within semi-alluvial bedrock environments, where alluvial deposits of limited quantity sit on top of bedrock.
-When all available particles are entrained and moved away from the control volume, there are none left to entrain, so the entrainment process is turned off.
-Likewise, \citet{Ancey2010, Ancey2014,Ancey2015} extended the birth-death process within a control volume as I reviewed here to an array of spatially distributed control volumes (cells). 
-They showed that immigration and emigration processes between cells set up a bedload diffusion, and they connected this theory with earlier sediment diffusion theories by deriving an Exner type equation of sediment diffusion.
-I will now review each of these developments in turn. 
-\end{comment}
+In the collective model of \citet{Ancey2008}, limiting the number of particles available for motion is more difficult. 
+The most obvious way would be to consider that the entrainment rate of all particles is proportional to the entrainment rate of a single particle, times the number of particles at rest, such as $\lambda'(N-n)$.
+However, this does not work because particles can also arrive from upstream, meaning $n$ can exceed $N$. 
+In this case, to consider the rate of entrainment decreases with the number of active particles, one would need to disciminate how the particles got into the control volume in the first place. 
+Did they entrain, or did they emigrate?
+This is non-Markovian. 
+A second way to impose that a finite number of bed particles are available for entrainment is to consider the Markovian dynamics of two species at once: the bed surface particles, and the particles in motion.
+This path was taken by \citet{Turowski2009}, and this is the next section of the review.

### 2021-09-03 revision/formerPapers/ens-rev/scraps/mechanics.tex
added phd thesis.

\section{Underlying mechanics of Markov models: toward the mechanistic-stochastic} 
\subsection*{Mechanics versus kinematics} 
words

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Mechanical perspectives on the entrainment probability} 
words

Entrainment probability initially considered by \citet{Einstein1942, Einstein1950}. 
\citet{Grass1970} is very important. 
Many authors followed in the Grass paradigm to define particle entrainment as a convolution integral over a turbulent fluctuation pdf \citep{Paintal1971, Kirchener1990, Ling1995, Laursen1999, Wu2002b, Wu2004a, Wu2003, Papanicolaou2002, Bridge1992, Kirchener1990, Ma2013, Nakagawa1980, Vanrijn1984}
These approaches have been somewhat undermined by the recent developments in the impulse idea \citep{Diplas2008, Valyrakis2011, Valyrakis2013, Celik2014}. 
Investigators have just begun to bring these two research directions into accord with one another \citep{Tregnaghi2012, Dey2018}.

\subsection*{Mechanical perspectives on particle travel characteristics}

hydrodynamics of the particle velocity distribution: 
random walks to describe the distribution: \citep{Ancey2008, Fan2014}
velocity ensembles, \citep{Furbish2012b}, \citep{Furbish2013} 
what controls the particle hop distance distribution? 
Furbish something? where? \citep{Furbish2016}

\citet{Martin2014}: Distribution of waiting times from a mean reverting random walk model

\subsection*{Mechanics of sediment deposition} 
\citep{Charru2004} is a kinematic model. 
There is the recent paper on thresholds of deposition
See what else you can find.

\subsection*{Analyzing assumptions: Are kinematic models enough?}
Houssais argues that bedload transport is a continuous transition from granular creep which would undermine the
partitioning scheme into motion and stationary states.
Possible that multiple motion modes exist meaning more states necessary in markov models.

### 2021-09-03 revision/formerPapers/ens-rev/scraps/outline-intro.tex
added phd thesis.

- bedload flux is a fundamental problem
- bedload flux is the rate of downstream movement of sediment grains 
- bedload transport has strong feedback with stream morphology, so it matters
- matters to lots of stuff: aquatic habitat and energy
- unfortunately, existing approaches are inadequate

- approaches are inadequate because average characterizations of flow and sediment 
do not always correlate with the bedload rate 
- bedload rates can span across orders of magnitude as details of turbulence and bed organization vary, while average characterizations remain constant

- The same local details which correlate with the bedload flux are modified by it. 
- Moving sediment affects turbulent characteristics
- Bedload fluxes modify the stability and arrangement of particles on the bed surface. 
- Thus the bedload flux is in cyclic feedback with the details of turbulence and bed material organization which it correlates with.  
- Through this feedback, the bedload flux inherits randomness.

- As a result, bedload fluxes exhibit wide fluctuations of several times mean values, even as average characterizations of flow and sediment remain steady. 
- These fluctuations occur in the most controlled laboratory experiments, with uniform size glass beads and steady state conditions 
- They occur especially in natural streams, where multiple grain sizes, unsteady flows, and variations in sediment supply only exascerbate the feedbacks imparting randomness to the bedload flux -- the linkages between bed material organization, turbulence, and bedload motion. 
- Apparently, bedload fluxes have statistical characteristics.

- Einstein was the first to recognize the probabilistic character of bedload motion
- He understood bedload motion as a random switching between resting and motion states
- He called the transition from rest to motion entrainment, and he characterized it with a probability which was linked to extreme events in fluid turbulence \citep{Einstein1949, Einstein1950}. 
- He called the transition from motion to rest deposition, and he considered it an implicit function of the bedload flux \citep{Einstein1942, Einstein1950}. 
- By convolving over sequences of entrainment and deposition, employing some semi-empirical arguments along the way, Einstein developed a formula for the average transport rate which was based upon statistics \citep{Einstein1950, Yalin1972, Shen1980, Ancey2006}.

### 2021-09-03 revision/formerPapers/ens-rev/turowski2009.tex
added phd thesis.

--- 2021-09-03 draft/formerPapers/ens-rev/review0.tex
@@
-\section{Review: birth-death theories of sediment transport}
-
-Birth death processes are a special type of stochastic process. 
-Stochastic processes track the evolution of a set of random variables through time. 
-Birth death processes consider a discrete state space, such as the size of a population. 
-Members of the population are born and die: transitions are allowed between adjacent sites in the discrete state space.
-
-Within sediment transport, Einstein was the first to consider a model for the bedload flux in terms of stochastic processes. 
-He considered the transport of individual bedload grains to be a random switching between motion and rest states. 
-In this way, supplementing this consideration with empirical reasoning, Einstein developed a formula for the mean transport rate \citep{Einstein1950}. 
-Einstein determined that the bedload transport rate should follow a Poisson distribution.
-
-Einstein's predecessors have extended his formulations in order to bring them in accord with experimental data and to phrase them in terms of a stronger mathematical foundation drawn stochastic process theory.
-These extensions probably began with \citet{Lisle1998}. 
-They linked Einstein's theory of tracer dispersion \citep{Einstein1937}, which is contingent on random transitions between states of motion and rest, into Markov process theory, showing that Einstein's tracer dispersion theory was just a two-state Markov process where the states are motion and rest.
-Then \citet{Ancey2006} went on to rephrase the \citet{Einstein1950} bedload flux derivation in terms of a two-state Markov process, where random transitions between states of motion and rest are included into a Chapman-Kolmogorov equation -- a general mathematical formulation.
-In this way, they derived a probability distribution for a bedload flux which is based upon underlying randomness in the entrainment and deposition of bedload grains. 
-Taking the mean of their bedload flux probability distribution reproduces an Einstein-like formula for the mean bedload flux. 
-This work is sketched to indicate how Markov process theory can work to describe the bedload flux.
-
-\subsection{Formalizing Einstein: a foundation in Markov process theory}
-
-The starting point for the \citet{Ancey2006} work was a set of experiments: the transport of uniform glass beads was measured in a narrow flume using image processsing techniques. 
-Measured fluxes had strong fluctuations: instantaneous fluxes took on values as much as ten times the mean. 
-Therefore, they attempted to describe the probability distribution of the bedload flux by revisiting Einstein's theory \citep{Einstein1950} from a Markov process framework.
-
-According to Einstein's way of thinking, an individual sediment grain undergoes a random transition from motion to rest states.
-We can label these two states as $x=0$ and $x=1$.
-Then, tracking the evolution of the random variable $x$ through time will fully describe the start-stop motion of the individual grain.
-
-Now we prescribe rates of entrainment and deposition in a small time $dt$. 
-If the particle is in motion ($x=1$) at time $t$, the probability it deposits to the rest state ($x=0$) at time $t+dt$ is modelled as a constant $\tau^{-1}$. 
-Likewise if the particle is at rest ($x=0$) at time $t$, the probability it entrains to the motion state ($x=1$) at time $t+dt$ is modeled as a constant $\sigma^{-1}$. 
-Making the Markov process assumption that the transitions of the particle between $t$ and $t+dt$ are independent of its history, the forward Kolmogorov equations read
-\begin{align} 
-\pi_0(t+dt) = \tau^{-1} dt \pi_1(t) + (1-\sigma^{-1} dt) \pi_0(t) \\
-\pi_1(t+dt) = \sigma^{-1} dt \pi_0(t) + (1-\tau^{-1} dt) \pi_1(t). 
-\end{align}
-
-Dividing through by $dt$ and taking the limit as $dt \rightarrow 0$ develops the Master equation for the two state process
-\begin{align} \label{eq:anc2006}
-\dot{\pi}_0(t) = \tau^{-1} \pi_1(t) - \sigma^{-1} \pi_0(t) \\
-\dot{\pi}_1(t) = \sigma^{-1} \pi_0(t) - \tau^{-1} \pi_1(t). 
-\end{align} 
-This simple model is called a telegraph process \citep{Cox1965}.
-
-To solve for the probabilties $\pi_0(t)$ and $\pi_1(t)$, the two equations can be summed to give $\frac{d}{dt} [ \pi_0(t) +\pi_1(t) ] = 0$, meaning $\pi_0(t) + \pi_1(t) = \text{constant}$.
-Since motion and rest are mutually exclusive and there are no other possibilities for the particle, $\pi_0(t) + \pi_1(t) = 1$: probabilities are conserved.
-Using this result into \ref{eq:anc2006} gives 
-\begin{align} 
-\pi_0(t) =  \\
-\pi_1(t) = 
-\end{align} 
-where $\pi_0(0)$ and $\pi_1(0)$ are the initial conditions, which we have not specified.
-
-After a sufficiently long time, the particle forgets its initial conditions and settles into equilibrium transport: 
-\begin{align} 
-\pi_0 = \\
-\pi_1 = 
-\end{align}
-Therefore, the time dependence and initial conditions are forgotten by the Markov process, and it has a set probability $\pi_0$ to be at rest and $\pi_1$ to be in motion.  
-Therefore, the fraction of time the particle spends in motion is $\frac{\tau}{\tau+\sigma} = \xi $, and the fraction of the time it spends in rest is $\frac{\sigma}{\tau+\sigma} = 1-\xi$. 
-The probability that the particle is in motion is a Bernoulli distribution.
-
-Now consider that on the bed surface within an observation window there are $N$ particles at any time. 
-Each of these particles is undergoing the same random switching between motion and rest states, so that the number of moving particles at any time is 
-\be N_m = \sum_{i=1}^N x_i \ee
-where $x_i=0,1$ is the Bernoulli random variable of the $i$th particle, with the probability of $x_i=0$ being $\pi_0$ and the probability of $x_1=1$ being $\pi_1$.
-
-Therefore, the probability that there are $N_m =n$ particles in a motion state is a counting problem. If $n$ are in motion then $N-n$ are at rest. The result is 
-\be \text{Prob}(n \text{ in motion}) = {{N} \choose {n}} \xi^n (1-\xi)^{N-n}. \ee 
-In statistics lingo, the sum of Bernoulli trials is a Binomial distribution. 
-This distribution has mean $\xi N$ and variance $\xi(1-\xi)N$. 
-Therefore the probability distribution of the number of moving particles within the observation window can be written 
-\be \text{Prob}(n) = \text{Bi}[ \xi N, \xi(1-\xi)N].  \ee
-
-If all particles in motion move at velocity $u_p$ and have volume $\nu_p$, then the bedload flux is a random variable $Q = \frac{u_p \nu_p}{L} N_m $.
-Therefore Ancey et al calculated the bedload flux probability mass function as
-\be \text{Prob}(Q) = \text{Bi} [\frac{u_p \nu_p}{L} \xi N, \xi (1-\xi) \frac{u_p^2 \nu_p^2}{L^2}]. \ee
-
-We conclude that the mean transport rate is 
-\be \bar{Q} = \nu_p \frac{N}{L}  \frac{\tau}{\tau+\sigma}   u_p. \ee
-This is conceptually similar to Einstein's formula \citep{1950}, and Ancey et al. explored this connection more carefully than we will here \citep{Ancey2006}. 
-The essential point is that Einstein took a somewhat different perspective.
-The key result of the \citet{Ancey2006} theory is that the mean transport rate is the result of $N$ parallel telegraph processes: each of the $N$ particles on the bed undergoes a switching process between motion and rest states, and the net result is the mean rate $\bar{Q}$ which is governed by a binomial distribution.
-
-In context of the experiments Ancey et al performed, the formulation just described did not capture the magnitude of the bedload fluctuations, although it described mean behavior quite well.  
-Their experimental bedload fluctuations were much larger than those expressed by this model. 
-Said in a different way, the probability distribution of the experimental bedload transport rate had a heavy tail, while the Binomial distribution of the rate derived from this parallel telegrapher's process theory has a light tail.
-
-Heavy tails are a hallmark of collective motions in physical systems \citep{Sornette2000}. 
-Of course, if bedload transport is driven by the advection of large turbulent eddies, a perspective supported by many experiments \citep{Nelson1995, Hofland2006, Celik2014, Amir2014, Shih2017}, then bedload motions will not be independent, but they will be coupled collectively to passing turbulent eddies.
-In context of the model just presented, this means the state of each telegraph process at an instant will be coupled to every other. 
-Therefore, coherent turbulence could imply heavy tails in the bedload probability distribution.
-
-There are other processes which could imply these collective effects as well. One option is entrainment due to particle collisions. 
-Moving particles can impact stationary particles, leading to their entrainment. 
-Essentially, this means the entrainment rate will depend on the number of active particles. 
-This was the extension made by \citet{Ancey2008} in order to capture the large fluctuations seen in their bedload flux experiments.
-
-\subsection{Capturing fluctuations: The inclusion of collective effects}
-
-In 2008, Ancey et al took a somewhat different perspective in order to describe the heavy tailed bedload probability distribution their experiments showed \citep{Ancey2008}.
-Instead of considering the motion of every bedload particle as an independent telegrapher's process, they considered the number of particles within a control volume as governed by a collective birth-death process \citep{Gardiner1983, Gillespie1992, VanKampen1992, Cox1965}.
-Birth of a member of the population is just entrainment. 
-Death is deposition. 
-The population can also immigrate (move in) to the control volume from upstream, and it can emigrate (move out) of the control volume to downstream. 
-Birth, death, immigration, and emigration rates characterize the dynamics of the entire population at once, rather than just the dynamics of independent individuals as the \citet{Ancey2006} work pursued.
-In this way, they managed to include a collective entrainment effect.
-
-Consider the probability that there are $N=n$ particles in motion at time $t$. 
-This can be denoted $\pi(n,t)$.
-The random variable $N$ takes on values $0,1,2,/dots$ -- non-negative integers of arbitrary magnitude.
-Each value has an associated probability $\pi(n,t)$.
-The entire population is considered subject to four processes within a small time increment dt.
-First, entrainment of a single bed paricle (birth) can happen with probability $\lambda_0 + \mu n$. 
-Here a collective entrainment effect has been included. 
-The probability of entrainment in $dt$ grows with the number of particles in motion. 
-This is a crude inclusion of the collective effects we just discussed: the advection of turbulent structures leads to correlations in entrainment rates, and the collision of moving particles with stationary particles can induce entrainment as well.
-
-Second, deposition of a single bed particle can happen in $dt$ with probability $\sigma n$. 
-The death rate is proportional to the number of individuals in the population. 
-Third, immigration of a particle from upstream into the control volume can happen with probability $\lambda_1$. 
-Fourth, emigration of a particle from the control volume to downstream can happen with probability $\nu$.
-
-These possible transitions develop a system of equations for the probabilities of finding $n$ moving particles within the control volume: 
-\be \pi_n(t+dt) = \alpha dt (n+1)\pi_{n+1}(t) + [\lambda + (n-1)\mu]dt \pi_{n-1}(t) + [1-dt[\lambda + n\alpha + n\mu]]\pi_n(t) \ee
-As $dt\rightarrow 0$ these develop a system of differential-difference equations for the probabilities $\pi_n(t)$ of finding $n$ particles in the control volume at time $t$: 
-\be \dot{\pi}_n(t) = (n+1) \alpha \pi_{n+1}(t) + (\lambda + (n-1)\mu)\pi_{n-1}(t) - (\lambda +n (\alpha+\mu))\pi_n(t). \label{eq:anc2008} \ee
-
-This birth-death immigration-emigration model can be solved by introducing a probability generating function $G(z,t) = \sum_{n=0}^\infty z^n \pi_n(t)$ \citep{Gardiner1983, Ancey2008}. 
-Multiplying \ref{eq:anc2008} by $z^n$ and summing over all $n$ gives, after a careful manipulation of the sums in order to form some function of $G(z,t)$ in every term, 
-\be \frac{\partial}{\partial t} G(z,t) = \lambda(z-1)G(z,t) + \{ \sigma + \mu z^2 + \nu - (\mu + \sigma + \nu)z\} \frac{\partial}{\partial z} G(z,t). \ee
-
-This partial differential equation for the probability generating function $G$ can be solved by the method of characteristics \citep{}. 
-If there are initially $N_0$ moving particles within the control volume, the solution is 
-\be G(z,t) + \Big(\frac{\alpha-\mu}{(K-1)\mu z + \alpha - K \mu}\Big)^{n+\lambda/\mu}\Big( \frac{(K\alpha-\mu)z + \alpha(1-K)}{\alpha-\mu}\Big)^n. \ee
-The factor $K$ is the autocorrelation function $K = \exp(-t(\alpha-\mu)).$
-
-This probability generating function generates probabilities via a Taylor expansion around $z=0$: $G(z,t) = \sum_{n=0}^\infty \frac{z^n}{n!}\big(\frac{\partial}{\partial z}\big)^n G(z,t) |_{z=0}$.
-Comparing this formula with the definition of $G$, $G(z,t) = \sum_{n=0}^\infty z^n \pi_n(t)$, apparently the coefficient of $z^k$ in the power series expansion of $G(z,t)$ is $\pi_k(t)$.
-
-For times sufficiently large for the system to forget its initial condition and settle into a steady state, the autocorrelations $K(t)\rightarrow 0 $ in the expression of $G$, and the power series expansion of $G(z,t) \rightarrow G(z)$ yields a set of stationary (independent of time) probabilities
-\be \pi_n = \frac{\Gamma(r+n)}{\Gamma(r)n!} p^r (1-p)^n\text{, }n=0,1,\dots.\ee
-Here $r=\lambda/\mu$ and $p = 1-\mu/\alpha$. $\Gamma(x) = \int_0^\infty z ^{x-1} e^{-z} dz$ is the well known $\Gamma$-function of mathematical physics.
-
-These stationary probabilities are negative binomial distributions: 
-\be \pi_n = \text{NegBin}(n;r,p).\ee
-This is the generalization of $N$ parallel telegraph processes which was employed in the \citet{Ancey2006} paper. 
-The negative binomial distribution has a heavy tail as a result of collective entrainment $\mu$.
-
-When the collective entrainment process is turned off, $\mu\rightarrow 0$ which recovers the \citet{Ancey2006} result. 
-$\mu\rightarrow 0$ implies $r \rightarrow \infty$ and $p \rightarrow 1$. 
-Within this limit, the generating function G(z) becomes 
-\be G(z) = \exp(-\frac{\lambda}{\alpha}[z-1]).\ee
-This is the generating function of a Poisson process, so the probabilities of the number of active particles become Poisson distributed 
-\be \pi_n = \frac{r'^n}{n!}e^{-r'}\ee
-where the ratio $\lambda/\alpha = r'$ defines $r'$. 
-This is the same scaling behavior for large $n$ as the \citet{Ancey2006} model, since the large $n$ limit of a binomial distribution is a Poisson distribution.
-
-The Ancey et al 2008 inclusion of collective entrainment develops a Negative Binomial distribution of the bedload flux. 
-Expressing the bedload flux as the number of moving particles within the control volume times their average velocity $u_p$ makes a Negative binomial distribution for bedload flux within the control volume.
-The mean bedload flux is $ \frac{\lambda}{\alpha-\mu} u_p $ and the variance of the flux is $ \frac{\lambda \alpha}{(\alpha-\mu)^2} u_p^2$: the fluctuations are much wider in the negative binomial distribution, meaning the inclusion of collective entrainment can model the large fluctuations observed in the \citet{Ancey2008} experiments.
-
-One way to view the bedload flux we just presented: it is the number of moving particles times their velocity. 
-However, since this model explicitly includes downstream emigration from the control volume, there is another way to define the flux: it is the number of emigration events within a unit of time. 
-Counting the number of emigration events expressed from this model within a unit time is not easy. 
-\citet{Ma2014b} explored this approach to calculating the bedload flux, and I will review this in a subsequent section.
-
-Two key assumptions of this model are worth examining. 
-These set the stage for the subsequent developments in birth-death modelling of the bedload flux which I review next. 
-First, this model assumes that stationary bed particles are in infinite supply. 
-This is an implicit assumption: Nothing prevents the system from entraining particles ad infinitum through the entrainment processes characterized by rates $\lambda_0$ and $\mu$. 
-Second, the model considers sediment transport in an isolated control volume, so the sediment flux is described probabilistically in a locality, but spatial correlations cannot be accounted for: the flux upstream of the control volume is also a fluctuating quantity, implying spatial correlations through immigration events into the control volume considered in this model.
-
-These two observations support two of the more recent developments to the \citet{Ancey2008} birth-death immigration-emigration model: \citet{Turowski2009} generalized this model to finite bed material in order to describe transport within semi-alluvial bedrock environments, where alluvial deposits of limited quantity sit on top of bedrock.
-When all available particles are entrained and moved away from the control volume, there are none left to entrain, so the entrainment process is turned off.
-Likewise, \citet{Ancey2010, Ancey2014,Ancey2015} extended the birth-death process within a control volume as I reviewed here to an array of spatially distributed control volumes (cells). 
-They showed that immigration and emigration processes between cells set up a bedload diffusion, and they connected this theory with earlier sediment diffusion theories by deriving an Exner type equation of sediment diffusion.
-I will now review each of these developments in turn.
-
-Also Heyman 2013 considered that the waiting time distribution need not be exponential. Need more focus on the underlying poisson distributions of these models.
-
-\section{Including sediment availability: Turowski's two species birth death model}
+\subsection{What about sediment availability?: Turowski's two species model}
 
@@
 This can be written $\pi_{n,m}(t)$. 
-Within a small time interval $dt$ he considered birth, death, immigration, and emigration transitions governing the probabilities $\pi_{n,m}(t)$, similar to \citet{Ancey2008}. 
-Like \citet{Ancey2008}, he also included a collective entrainment contribution, arranging that the model expressed wide Negative Binomial-like fluctuations in the bedload rate.
+Within a small time interval $dt$ he considered birth, death, immigration, and emigration transitions governing the probabilities $\pi_{n,m}(t)$ similar to \citet{Ancey2008}. 
+Like \citet{Ancey2008}, he also included a collective entrainment contribution, arranging that the model expressed wide Negative-Binomial-like fluctuations in the bedload rate.
 
@@
 This is similar to competing species in ecology \citep{Pielou1977} or to multiple reacting chemical species \citep{Gardiner1983}.
-
-\section{The bedload flux at a point: the counting statistics of emigration events}
-
-Up to this point, the Markovian approaches to understand a fluctuating bedload rate were based upon analyzing the stochastic number of active particles within a control volume. 
-The bedload flux definition considered for these studies was somewhat unconventional. 
-Typically it is defined as the rate of bedload crossing a vertical plane perpendicular to the flow direction \citep{Ballio2014}.
-
-The studies of \citet{Heyman2013} and \citet{Ma2014b} demonstrated that the birth-death model of \citet{Ancey2008} can also be used to define a flux at a position in space. 
-Although the birth-death model of \citet{Ancey2008} counts the number of moving particles at each instant of time, it does not count the number of emigration events which contribute to this number. 
-Counting the number of emigration events in an interval of time is a much more difficult problem.
-Ma et al (2014) approached this counting problem using a path integral approach: this approach is essentially outlined in \citet{Ohkubo2009}. 
-The path integral approach draws heavily from the stochastic simulation algorithm of \citet{Gillespie1977}.
-
-The conventional definition of the bedload flux is an integral of particles through a surface perpendicular to the flow: 
-\be Q_s(t) = \int \int_A dA \textbf{u}_s \cdot \textbf{n}, \ee
-where $\textbf{n}$ is a unit vector perpendicular to the average flow direction and $\textbf{u}_s$ is the velocity of bed load. 
-This instantaneous flux is very difficult to measure, so usually it is averaged over some time interval $\delta t$: 
-\be Q_s(t;\delta t) = \frac{1}{\delta t} \int \int_A \int_t^{t+\delta t} \textbf{u}_s \cdot \textbf{n} d\tau dA = \frac{\delta V_s}{\delta t} = v_s \frac{\delta S(t;\delta t)}{\delta t}. \ee
-$Q_s(t;\delta t)$ is the volume flux averaged over $\delta t$; $\delta V_s$ is the volume of bed load particles passing through $A$ within $\delta t$; $S(t)$ is the cumulative number of particles passing through the cross section since t=0, and $\delta S(t;dt)$ is the number of particles passing through the surface between $t$ and $t+\delta t$, i.e.
-$\delta S(t; \delta t) = S(t+\delta t) - S(t)$.
-This equation means the bedload flux at $t$, when viewed in a time-averaged sense, is the number of particles crossing the surface between $t$ and $t+\delta t$. 
-Because volume units can always be multiplied in later, the bedload flux is written 
-\be q_s(t) = \frac{\delta S(t; \delta t}{\delta t}. \ee
-
-This bedload flux is a random variable.
-A fluctuation is defined strictly as $q_s' = q_s - \bra q_s \ket$, and the magnitude of fluctuations is defined without ambiguity by the variance $\bra q_s'^2 \ket$. 
-This $\delta S(t;dt)/\delta t$ is the number of emigration events between $t$ and $t+\delta t$, so if this number could be counted, the bedload flux would be known at the downstream boundary of the control volume: the bedload flux would be known at a point in space.
-
-The path integral formulation performs this counting problem. 
-For the sake of computation, the time interval t is discretized as $i \Delta t = t$, $i=0,1,2,\dots$.
-In the end, the limit $\Delta t \rightarrow 0$ will be taken. 
-If the increments $\Delta t$ are sufficiently small, Ma et al argued that emigration events during one of these intervals will be approximately Poissonian so that
-\be P[\Delta S_i = n | N_i(t) = N_i] = \exp[-f_3(N_i)\Delta t] \frac{(f_3(N_i) \Delta t)^N}{n!}. \ee
-This equation expresses the probability that $n$ particles emigrate between $t$ and $t+\delta t$ is a Poisson distribution with rate parameter $f_3(N_i)\Delta t$, where $f_3(N_i) = \gamma N_i$ is the emigration rate.
-
-Now Ma et al argued that because the entrainment rate $f_1 = \lambda + \mu N \~ O(L)$, the deposition rate $f_2 = \sigma N \~ O(L)$, and the emigration rate $f_3 = \gamma N \~ O(1)$, if the observation window is sufficiently large, entrainment and deposition processes dominate, so that the number of moving particles $N$ is perturbed only weakly by emigration.
-In this limit of sufficiently large observation window size, it then holds that
-\be P(\sum \Delta S_i = n | N_1,N_2, \dots) = \exp[-\sum f_3(N_i)\Delta t]\frac{(f_3(N_i)\Delta t)^n}{n!}, \ee
-which in the continuum limit becomes a path integral counting emigration events: 
-\be P[\delta S = n |N(t): t_0<t<t_0+\delta t] = \exp[-\int f_3(N(\tau))d\tau]\frac{[\int f_3[N(\tau)]d\tau ]^n}{n!}.\ee
-Denoting the integral over the process $f_3(N(\tau))$ as $\alpha = \int_{t_0}^{t_0+\delta t} d\tau f_3[N(\tau)]$, the probability for $n$ emigration events in $\delta t$ is written 
-\be P(\delta S = n) = \int d\alpha \frac{e^{-\alpha}\alpha^n}{n!}P(\alpha,t). \ee
-From this difficult derivation a simple formula emerges: the probability of $n$ emigration events in $\delta t$ is an integral over all possible realizations of the process $N(t)$, where each process is given a weight $e^{-\alpha}\alpha^n/n!
-P(\alpha,t) d\alpha$.
-This is called a Cox process.
-It's a generalized Poisson process where the parameter $\alpha$ is also a random variable.
-It represents a superposition of a series of Poisson distributions taking every possible path of $N(t)$ into account.
-This is a stochastic description of the bedload transport flux based on the counting statistics of emigration events.
-
-Notably, because of the time integral of $N(t)$ involved in this relationship, it has a memory. It is non-Markovian. 
-This character reveals a very interesting fact: the fluctuations of the transport rate depend on the time scale of observation. 
-In fact, Ma et al derived the expression for the variance: 
-\be \delta t \text{var} [q_s(\delta t)] = \gamma^2 \frac{\lambda (\sigma + \gamma)}{(\sigma + \gamma -\mu)^2} 2 t_c [\delta t - t_c (1-e^{-\delta t/t_c}] + \frac{\gamma\lambda}{\sigma + \gamma -\mu}\delta t \label{eq:3regimes} \ee
-
-where $t_c = 1/(\sigma + \gamma - \mu)$ is the autocorrelation time of $N(t)$, representing memory in the system. 
-Apparently, the variance of the bedload flux at a point depends on the observation or averaging time-scale $\delta t$.
-
-In fact, Ma et al discriminated three regimes of behavior from \ref{eq:3regimes}:   
-\be
-    \delta t \text{var}[q_s(\delta t)] \approx 
-\begin{cases}
-    \frac{\gamma \lambda}{\sigma + \gamma - \mu}, & 0<\delta t \ll t_l \\
-    \delta t \frac{\gamma^2 \lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}   & t_l<\delta t< t_c \\
-    2 \gamma^2 \frac{\lambda(\sigma+\gamma)}{(\sigma+\gamma-\mu)^2}t_c + \frac{\gamma \lambda}{\sigma + \gamma - \mu}   & t_c \ll \delta t < \infty
-\end{cases}
-\ee
-When this three-regime equation is divided by the mean bedload rate $\text{mean}[q_s(\delta t)]$, the behavior is discriminated by a dimensionless number Ma et al called $Ra$: 
-\be Ra(\delta t) \approx \begin{cases} 1 & 0<delta t \ll t_l \\
-\frac{delta t}{t_l} & t_l < \delta t < t_c \\ 
-\frac{2 t_c}{t_l} + 1 & t_c \ll \delta t < \infty \end{cases} \ee
-
-The essential point is that Ma et al derived three regimes of fluctuations from the \citet{Ancey2008} model. 
-Depending on the sampling interval $\delta t$, three different scaling relations are possible for the magnitude of fluctuations in the bedload flux. 
-These intervals are discriminated by two timescales $t_c$ and $t_l$. 
-There are probably deep consequences of this research which remain to be fully understood. 
-At this stage, it's sufficient to note that fluctuations differ with observation timescale, and that this effect can be discriminated into stages with a dimensionless number $Ra$.
-
-The origin of the two timescales still needs to be discussed.
-
-There is another way to get information about the bedload flux at the downstream boundary of the control volume defined by \citet{Ancey2008}, and this was examined by \citet{Heyman2013, Heyman2016}. 
-Rather than actually counting the number of emigration events with the complicated path integral approach, Heyman et al.
-took a somewhat simpler angle: they examined the waiting time between successive emigration events.
-
-\section{The waiting time between emigration events: Bedload flux from waiting time statistics}
-
-These studies are concerned with linking the birth-death formulation, which in the original interpretation deals with transport characteristics within an observation window \citep{Ancey2008}, to the transport characteristics at a point -- the number of particles crossing a plane perpendicular to the flow direction.
-
-\citet{Heyman2013} was concerned with the statistics of the time interval between successive emigration events, while \citet{Ma2014b} focused on actually counting the number of emigration events in a unit time: that is, they computed the bedload flux over the downstream boundary of the control volume \citet{Ancey2008}.
-
-The waiting time between successive transport events is significant in a broad range of transport studies. 
-First, there are probabilistic transport rate formulations which use the waiting time distribution as their input \citep{Turowski2010}. 
-Second, the waiting time between successive transport events, as we shall show, reflects back on the underlying entrainment and deposition characteristics, so it provides fundamental information into the somewhat mysterious physics of bedload transport.
-
-The original \citet{Einstein1937,Einstein1950} approach reveals that the waiting time between successive transport events should be an exponentially distributed random variable. 
-However, Heyman et al (2013) demonstrates that the inclusion of collective entrainment due to \citet{Ancey2008} modifies the waiting time distribution in a non-trivial way. 
-The waiting time between successive entrainment events has contributions at two different scales: first, there is a shorter timescale of individual entrainment events, controlled by fluid turbulence and essentially recognized within the \citet{Einstein1937} formalism, and second, there is a longer timescale due to collective entrainment.
-These authors showed that within the \citet{Ancey2008} birth-death model, at low transport rates, when entrainment and motion are highly intermittent, the short timescale and the long timescale are well separated.
-While at higher transport rates, these two timescales blend together as individual and collective entrainment contributions become indistinguishable.
-
-I'll now sketch the \citet{Heyman2013} derivation of the statistics of the waiting time between successive emigration events, and show how the two different timescales due to individual and collective entrainment arise.
-For \citet{Heyman2013} the variable of interest is $S_k = \sum_{i=0}^k T-i$: $S_k$ is the time when the $k$th emigration event occurs.
-$T_1, T_2, \dots $ are the waiting times: the time from $t=0$ to the first emigration event is $T_1$.
-The time from $T_1$ to the second emigration event is $T_2$, and so on.
-For a Markov process such as the \citet{Ancey2008} model, the waiting times are independant and identically distributed. 
-Within the Einstein model we have already seen that the waiting time probability distribution is expoential. 
-However, in the \citet{Ancey2008} formalism there is no reason to expect an exponential distribution.
-
-Heyman et al define $F_n(t)$ as the probability that there are $n$ particles in motion and no emigration event occurred in time t: 
-\be F_n(t) = Pr(T>t,N(t)=n).\ee
-By extension, the probability that there are any number of particles and no emigration event occurred in time $t$ is 
-\be F(t) = Pr(T>t) = \sum_{n=0}^\infty F_n(t) .\ee
-
-Now $F(t+\Delta t)$ is equivalent to the probability that any other event but emigration occurs in $\Delta t$. 
-Therefore we can write a master equation for $F(t)$: 
-\be F_n(t+\Delta t) = F_n(t)[1-(\lambda + n(\sigma+\mu+\gamma))\Delta t] + F_{n+1}(t) \sigma(n+1)\Delta t + F_{n-1}(t)\lambda + \mu(n-1)\Delta t + o(\Delta t).\ee
-This equation holds for $n\geq 1$. At $n=0$ deposition and collective entrainment processes are not possible, therefore 
-\be F_0(t+\Delta t) = F_0(t) [ 1-\lambda \Delta t] + F_1(t) \sigma \Delta t + o(\Delta t) \ee 
-Again dividing by $\Delta t$ and taking $\Delta t \rightarrow 0$ gives the differential difference equations
-\begin{align} 
-F_0'(t) &= \lambda F_0(t) + \sigma F_1(t) \\
-F_n'(t) &= -(\lambda + n(\sigma + \mu + \gamma))F_n(t) + \sigma(n+1)F_{n+1}(t) + [\lambda + \mu(n-1)]F_{n-1}(t) \text{ if } n\geq 1 \end{align} 
-for the waiting time distribution.
-
-Summing all of the terms gives the simple equation 
-\be \sum_{n=0}^\infty F_n'(t) = -\gamma \sum_{n=0}^\infty nF_n(t), \ee
-which, denoting the pdf of waiting times $T$ as $f_T(t)$ gives the simple relationship: 
-\be f_T(t) = -F'(t) = \gamma\bra F_n(t)\ket. \ee
-Therefore the probability distribution function of waiting times between emigration events, $f_T(t)$, is an average of the probability that there are $n$ particles and no emigration event occurred in $t$ over all $n$.
-
-The general solution of $F_n(t)$ can be obtained from the differential equations with a generating function: 
-\be G(z,t) = \sum_{n=0}^\infty F_n(t) z^n. \ee
-Multipling the differential equation by $z^n$ and summing over all $n$ gives
-\be \frac{\partial G}{\partial t} = (\sigma + \mu z^2 - z(\alpha + \mu) ) \frac{\partial G}{\partial t} + (z-1)\lambda G, \ee
-where the shorthand $\alpha = \gamma + \sigma $ has been introduced.
-
-This equation can be solved with the method of characteristics, giving a closed form solution for $G(z,t)$.
-A useful property of generating functions (and the source of their name) is $\bra F_n \ket = \partial G(z,t) /\partial t |_{z=1}$, therefore the probability distribution of waiting times between emigration events in the \citet{Ancey2008} birth-death model of bedload flux is
-\be f_T(t) = \gamma(z_1-z_2)^{\lambda/\mu} \big( \frac{\alpha-\mu}{A(t)-B(t)} \big)^{\lambda/\mu + 1} e^{-\lambda (1-z_2)t}\big( \frac{\lambda/\mu+1)B(t)}{A(t)-B(t)} [(1-z_2)e^{-(z_1-z_2)t}+z_1-1]+e^{-\mu(z_1-z_2)t}-1 \big)  \ee
-When collective entrainment is turned off ($\mu=0$), the probability distribution becomes $f_T(t) = \frac{\alpha}{\lambda \gamma} e^{-\alpha t/(\lambda \gamma)}$: the exponential waiting time between emigration events is recovered.
-
-This probability distribution for waiting times between emigration events in the \citet{Ancey2008} model has, in general, contributions from two timescales.
-The relatively common individual entrainment process imparts a 'fast' timescale to the probability distribution of emigration waiting time, while the relatively rare collective entrainment process imparts a 'slower' timescale.
-When collective entrainment is turned off, only the fast timescale is present in the pdf.
-When the entrainment rate is relatively low, the fast timescale is much shorter than the slow timescale, so that the timescales are well separated and the waiting time distribution is not well approximated by any common distribution in the literature (Gamma, exponential): the behavior of the \citet{Ancey2008} model at low transport rates has defied prediction by any other framework.
-
-The most significant point of the \citet{Heyman2013} paper is that the probability distribution function of waiting times between emigration events can be used to calibrate the \citet{Ancey2008} birth death model to flume experiments.
-This is because short emigration times control $\alpha-\mu$ and long emigration times control $\lambda$. 
-The evaluation of waiting times between emigration events supports a calibration of the stochastic birth-death model of \citet{Ancey2008} on flume studies where the transport rate is measured at the outlet of the flume, but the erosion, deposition, or emigration rates are not necessarily measured.
-In contrast to \citet{Ma2014b} the mathematics are not much beyond the original formulation of the \citet{Ancey2008} model, so this work may be more readily accepted by the research community.
-
-\section{The diffusion of bedload from a microstructural basis}
-
-\citet{Ancey2008} developed a generalized version of \citet{Einstein1950} which took account of a new process to describe the large fluctuations seen in experimental bedload flux signals: collective entrainment.
-\citet{Turowski2009} generalized this model to finite sediment availability, while \citet{Heyman2013} and \citet{Ma2014} used the \citet{Ancey2008} model in its original form to understand the distribution of waiting times between emigration events and the bedload flux probability distribution at a point in space, respectively.
-These two works developed new understanding of spatial and temporal aspects of bedload flux: \citet{Heyman2013} demonstrated that the waiting time between successive emigration events expressed contributions from slow and fast timescales related to individual and collective entrainment, respectively.
-\citet{Ma2014} highlighted the role of observation timescale on the magnitude of flutuations, and discriminated three distinct scaling regimes for bedload fluctuations which are contingent on a dimensionless number $Ra(\delta t)$, where $\delta t$ is the timescale of observation.
-
-However, up to this point in early 2014, stochastic birth-death models had focused on a single region of space -- a control volume \citep{Einstein1950, Ancey2006, Ancey2008}, or a single point in space -- the downstream end of the control volume, or the plane across which emigration events happen \citep{Heyman2013, Ma2014b}.
-This approach is obviously oversimplified. Variability of sediment transport rates is almost a defining feature of them \citep{Hassan2008, Venditti2016, Nelson2014}. 
-There is a need to take account of spatial differences in entrainment and deposition characteristics in calculating bedload fluxes. 
-Additionally, there are a wide set of issues centered around bedload diffusion-- or the spreading of bedload particles as they are tracked through time and subjected to random entrainment and deposition events \citep{Hassan2016}.
-Indeed, this study of bedload diffusion was the original research directive of \citet{Einstein1937}.
-
-Often, bedload diffusion has been understood through deterministic partial differential equation models.
-These advection diffusion equations have been derived by considering conservation of mass \citep{}, but they were not very well connected to the underlying stochastic dynamics of bedload transport.
-Generating this connection between a microstructural stochastic model like \citet{Ancey2008} and the continuum diffusion equation treatment of bedload diffusion \citep[e.g.][]{Parker2002} was the directive of the next papers on birth-death modeling \citep{Ancey2014,Ancey2015}.
-These papers took a conceptually straightforward extension of the \citet{Ancey2008} model which leads to difficult mathematics.
-
-Rather than considering a single control volume, as in \citet{Ancey2008}, \citet{Ancey2014,Ancey2015} considered an infinite array of adjacent control volumes (cells), indexed by $i=1,2,\dots,M$. 
-Bedload particles can entrain and deposit within each cell, and they can also migrate between cells. 
-Each of these transitions are characterized by probabilities per unit time in generalization of \citet{Ancey2008} across spatial extent. 
-The $i$th cell has a random number of particles $N_i$ in motion within it. 
-Therefore the state of the system at an instant of time is fully characterized by the set of these numbers: $\textbf{n} = (n_1(t), n_2(t), \dots n_M(t))$.
-
-One target is the grand probability distribution of the number of active particles within each cell at an arbitrary time: $P(\textbf{n};t)$. 
-Another target is the continuum limit: letting the length $\Delta x$ of each cell shrink to zero develops an advection-diffusion equation for bedload diffusion. 
-Ancey et al termed this a "Stochastic interpretation of bedload diffusion": and the latter target has much more fundamental scope.
-It explores the link between microscale stochastic models of bedload transport and mesoscale deterministic models based upon advection-diffusion equations.
-
-Now I'll sketch the \citet{Ancey2014, Ancey2015} derivation of a birth-death-migration model across an array of cells. 
-There are essentially three transitions to take account of: 
-\begin{enumerate}
-\item Entrainment can occur within each cell ($i$) at probability per unit time $(\lambda_i + \mu_i n_i ) \delta t$
-\item Deposition can occur within each cell ($i$) at probability per unit time $ \sigma_i n_i \delta t$
-\item Migration can occur from cell $i$ to cell $i+1$ (downstream) at probability per unit time $\nu_i$
-\end{enumerate} 
-Now some notation is introduced to write the effect of these transitions on $\textbf{n}$ in shorthand. 
-If the change of state due to one of these transitions is written as $\Delta \textbf{n}$, then the effect of these transitions can be characterized using vectors $\textbf{r}_i^j$ and $\textbf{r}_i^\pm$.
-The vectors $ \textbf{r}_i^j$ are of the same dimension as $\textbf{n}$ and all but two of their entries are zero: $r_i = 1$, $r_j = -1$, $r_k=0$ for $k\neq i,j$.
-The vectors $\textbf{r}_i^\pm$ are the same dimension as $\textbf{n}$ and all but their $i$th entry is zero: $r_i = \pm 1$, $r_j = 0$, $j\neq i$.
-
-Thus the transition probabilities of each process can be written: 
-\begin{enumerate}
-\item $ p_i^3 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^+; t+ \delta t) = (\lambda_i + \mu_i N_i) \delta t $ -- entrainment
-\item $ p_i^2 = \text{Prob}(\textbf{N} = \textbf{n} + \textbf{r}_i^-; t+\delta t) = \sigma_i N_i \delta t $ -- deposition
-\item $ p_i^1 = \text{Prob}(\textbf{N} = \textbf{n} \textbf{n} + \textbf{r}_i^{j-1}; t+ \delta t) = \nu_{i-1}N_{i-1} \delta t $ -- migration
-\end{enumerate}
-
-These transition probabilities develop a master equation analogous to the one from \citet{Ancey2008}, but generalized to a collection of $M$ cells: 
-\begin{multline} \frac{\partial}{\partial t} P(\textbf{n};t) = \sum_{i=1}^M (n_i+1)[P(\textbf{n}+\textbf{r}_{i+1}^j,t) \nu_i + P(\textbf{n}+\textbf{r}_i^+,t)\sigma_i ]\\
-+ P(\textbf{n}+\textbf{r}_i^-,t)(\lambda_i + \mu_i(n_i-1))\\
-+ P(\textbf{n}+\textbf{r}_i^{j-1},t)\nu_{i-1}n_{i-1}\\
-- P(\textbf{n},t)(\nu_{i-1} n_{i-1}+\lambda_i + \mu_i n_{i+1} + \nu_i n_i + \sigma_i n_i ) \end{multline} 
-Information about solving this equation is sparse in the literature. 
-The closest approach has been in ecology: these birth/death/migration-type stochastic models have been considered in context of ecological population dynamics since at least the sixties \citep{Bailey1968}.
-One exact solution of a very similar stochastic model has been attained using path integral approaches borrowed from quantum field theory \citep{Field2010}.
-By any metric, the analytical barrier this multidimensional master equation presents is not trivial, although definitely numerical solutions are possible \citep[e.g.][]{Gillespie1992}.
-
-In order to link this equation to spatial diffusion, \citet{Ancey2014, Ancey2015} resorted to a Poisson space representation \citep{Gardiner1983}. 
-In effect, the Poisson transformation converts the discrete random variable $\textbf{N}$ to a continuous random variable $\textbf{a}$ (the Poisson rate). 
-The Poisson transform is 
-\be P(\textbf{n},t) = \prod_i\int \frac{e^{-a_i}a_i^n}{n!}f(\textbf{a},t)d\textbf{a}, \ee
-where $\textbf{a} = (a_i) \geq 0$ for $i=1,2,\dots$ and $f(\textbf{a},t)$ is the multivariable probability density of the continuous vector $\textbf{a}$.
-
-\citet{Ancey2014, Ancey2015} used this transformation to obtain a Langevin equation representation of the dynamics of the random variable $\textbf{a}$.
-That is, they obtained a stochastic differential equation representation of $\textbf{a}$: 
-\be da_i(t) = (\lambda_i - a_i(\sigma_i-\mu_i) + \nu_{i-1} a_{i-1} - \nu_i a_i)dt + \sqrt{2 \mu_i a_i}dW_i(t),\ee
-and the $W_i(t)$ is a white noise term (Wiener Process) on cell $i$. 
-The mean steady state solution of this Langevin equation is 
-\be \bra a \ket_ss = \frac{\lambda}{\sigma-\mu} \ee 
-for all $i$.
-
-Now \citet{Ancey2014,Ancey2015} sought out to find an advection diffusion equation as the limit of cell size $\Delta x \rightarrow 0$, so they went on to impose this limit on the Langevin equation, thereby obtained an advection diffusion equation.
-Unfortunately, the limit is more subtle than expected. 
-This did not really work for reasons I need to understand better, and it did not work in ways I need to learn to explain better. 
-More study is needed on spatially varying birth-death models, from me, and from others especially.
-
-\section{Spatial correlations in bedload transport: }
-
-\citet{Heyman2014} used the \citet{Ancey2014} model to examine a set of experiments resolving spatial correlations of bedload transport. 
-They found a stochastic differential equation for the mean of the poisson rate $\eta$: 
-They related this to a point process framework which allowed them to analyze spatial correlations in the bedload rate and relate this to their birth-death model. 
-Somehow this whole thing is contingent on their two-point spatial correlation process. 
-They state that the whole problem is linear so that it'd be straightforward to extend it to 2d. 
-slowing down ...
-
-\section{recap} 
-Einstein developed first stochastic theories describing the diffusion and flux of bedload. 
-Ancey 2006, building on earlier work from Lisle 1998 and the many investigators who criticized and revised Einstein, formalized Einstein's assumptions over a foundation in Markov process theory, and extended his work by treating the bedload flux within a control volume as the cooperation of many independent two-state Markov processes: each bed particle within the control volume makes random transitions between motion and rest.
-Thus he derived a statistical distribution of the bedload rate which reproduces Einstein's result when the mean is taken. 
-However, the magnitude of fluctuations predicted by this model are too small.
-
-Ancey 2008 extended the Einstein theory by linking the transition rates between particles within the observation window with a collective entrainment term: they prescribed that the transition rate from rest to motion depends on the number of active particles.
-This is a simple way to include the effects of coherent turbulence and impact-based entrainment: it is an experimental observation that moving particles tend to come in waves \citep{Drake1988}. 
-The Ancey 2008 model was demonstrated capable of describing fluctuations, and all of its parameters have physical meanings, but notably there is no clear suggestion as to how collective and individual entrainment processes can be separated within experiments.
-Clearly, approaches to derive entrainment probabilities are focused on the individual entrainment rate \citep{Dey2018}, and there have been no developments with regard to computing the collective entrainment rate: there is no clear physical model of collective entrainment yet.
-
-Turowski 2009 extended the Ancey work to take account of limited supply, in order to describe semi-alluvial channels where alluvial deposits lie on top of bedrock. 
-He continued to work within the Einstein paradigm by describing the transport rate within a control volume, rather than by counting the number of particles leaving the control volume: although the Ancey 2008 work essentially outlines this possibiilty.
-The transport rate at a point should be considered as the number of emigration events in a unit time.
-
-Heyman et al 2013 and Ma et al 2014 went on to consider the statistics of emigration: they tried to discern what could be learned about the transport rate at a fixed point from the control volume formalism of Ancey et al and Einstein before that.
-Heyman et al was concerned with the statistics of the waiting time between successive emigration events. 
-The original Einstein (1937, 1950) assumptions generate an exponential waiting time ditsribution with a timescale related to the entrainment rate.  
-However, Heyman et al (2013) showed that when collective entrainment effects were included there are two timescales: one fast timescale related to the individual entrainment rate, and one slower timescale related to collective entrainment effects.
-Thus, the distribution of waiting times between successive emigration events, which is a useful concept for alternative stochastic models of the bedload flux distribution \citep[e.g.][]{Turowski2010}, is a more subtle object if collective entrainment occurs.
-These effects are accentuated at low transport rates, where the fast and slow timescales are more disparate. 
-At high rates, the timescales become comparable and the waiting time between successive emigration events blends into an Einstein-like exponential.
-
-All of these approaches considered bedload transport within a control volume or at a point, but indeed, a very large set of contemporary river science studies are concerned with the diffusion or spreading of bed material through a downstream reach of river \citep{Hassan2016}.
-Bedload diffusion is not captured by a model at a single location: often, it has been described using an advection diffusion equation. 
-These advection diffusion equations follow from considerations of mass balance within river channels, but they were not derived from any underlying microstructural model until \citep{Ancey2014}. 
-\citet{Ancey2014} extended the previous control volume model \citep{Ancey2008} to an array of control volumes, where emigration from the $i$th volume is immigration to the $i+1$th volume.
-
-In an approach widely used in chemical physics \citep{Gardiner1983} and ecology \citep{Bailey1968}, \citet{Ancey2015, Ancey2015} derived an advection diffusion equation from their microstructural model in a limit as the size of each control volume goes to zero while the number becomes infinite.
-Thus, in their interpretation, bedload diffusion emerges in the continuum limit of coupled birth-death immigration-emigration models. 
-They rederived the famous Exner equation of bedload diffusion.
-
-Taken together, this research exhibits a coherent progression from the original work of Einstein to a more mathematically and physically based set of more general approaches. 
-Within these approaches the bedload flux is understood as a random quantity. 
-The mean bedload flux and the magnitude of its fluctuations are modelled in an unambiguous way \citep{Ancey2006, Ancey2008} considering no limitations in sediment availability, and it was shown that collective effects are a necessary inclusion to properly describe fluctuations in bedload transport \citep{Ancey2008}.
-An extension to a more realistic situation of finite supply was developed \citep{Turowski2009}, and the relationships between these control volume based models and the definition of bedload flux as particles crossing a plane perpendicular to the flow were developed \citep{Heyman2013, Ma2014b, Ballio2014}.
-
-These approaches did not address spatial correlations or variations in bedload fluxes, although the problem of bedload diffusing through a reach of channel is of contemporary significance \citep{Hassan2016}.
-This extension was the most recent development in birth-death modelling of bedload \citep{Ancey2014, Ancey2015}. 
-By extending the \citet{Ancey2008} model to an array of adjacent control volumes, an advection-diffusion equation describing the concentration of bedload in motion was developed, providing the first connection between a microscale stochastic model of bedload transport with a macroscopic advection-diffusion formulation.
-\citet{Heyman2014} examined the spatial correlations between local bedload fluxes expressed by the \citet{Ancey2014} model. 
-This is the state of the art of bedload flux models.
-
-\section{Assumptions, calibraiton problems, and unclear aspects: A criticism of the birth-death approach}
-
-Now that earlier literature is evaluated the scope for generalization can be outlined. 
-This scope is a necessarily subjective consideration, and I'd like to highlight that nobody knows what the future of stocahstic modeling will hold. 
-First, the various glaring assumptions of the birth-death approaches are highlighted. 
-\begin{enumerate}
-\item Models only consider one particle size, while natural streams exhibit wide distributions of sizes and bedload transport expresses a wide set of effects related to particle size segregation \citep{Wilcock2003, Parker1982, Chen2008}.
-\item There is no clear means to discriminate collective and individual entrainment processes within experiments. 
-\item There is only a one-way feedback in these models: bedload is considered subordinate to the fluid flow, and the effect of bedload transport back onto the fluid flow is considered negligible.
-In fact, there are measurable influences of bedload transport on the fluid phase \citep{}; it's not yet clear whether a one-way coupled scheme such as this can actually describe natural streams across a realistic range of conditions.
-\item Collective entrainment is somewhat of a catch-all term with the physical mechanisms which may contribute to it unresolved.
-These may include collective entrianment, the formation and disintegration of particle clusters, interactions between the motion of grains of idfferent size, turbulent structure, local avalanche behavior, and bed form migration.
-\end{enumerate}
-
-\section{Summary: scope for extension}
-
-From the basis provided in this literature I'll now make a few small extended birth-death models.
-
-1. diffusion in 2d (horizontal and vertical)
-2. availability of depositional sites on the bed surface
-3. Extension to multiple grains sizes: A formal outline

### 2021-09-03 revision/laySum.tex
added phd thesis.

--- 2021-07-04 laySum.tex
@@
 
-Predicting the movement of gravel through river channels is important to manage environmental issues involving rivers, since moving gravel can destabilize river channels.
-A majority of models assume gravel flows as if it were a heavy fluid, but this description is crude considering how gravel actually moves.
-In this thesis, I instead describe gravel movement as the result of individual grains bouncing and rolling along the streambed.
-Using this concept, I calculate the downstream flow rate of gravel, study how this flow rate depends on the shape of the bed, and predict how quickly individual grains move downstream.
-This research produces a more realistic description of how gravel moves in river channels which helps us understand how rivers change through time.
+Predicting the flow rates of gravel through river channels is important to manage environmental issues involving rivers.
+Most approaches to predict gravel flow consider gravel as if it were a heavy fluid.
+This description oversimplifies the problem, since gravel flow is actually individual grains bouncing and rolling downstream, which is hardly like a fluid.
+
+In this thesis, I consider gravel flow as the result of individual grains.
+Using physics and probability, I calculate how fast individual grains move downstream, then I demonstrate how to use their downstream movement paths to predict overall gravel flow rates.
+This research produces a more realistic description of gravel movement in rivers which helps us understand how rivers change through time.
 

### 2021-09-03 draft/formerPapers/ens-rev/annobiblio/make.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+pandoc -s -o out.pdf anno.md
+echo "anno made" 

### 2021-09-03 draft/formerPapers/ens-rev/scraps/uncomment.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+while read p; do
+    if [["%" -eq  *$p]]; then
+        echo "$p"
+done<$1
+
+

### 2021-09-03 revision/clean.sh

@@ -1,0 +1,30 @@
+#!/bin/bash
+# this cleans out all of the auxilary and log files after compilation
+rm -f abs.aux
+rm -f ack.aux
+rm -f ch1intro.aux
+rm -f ch2flipFlop.aux
+rm -f ch2flipFlop.log
+rm -f ch3bedEle.aux
+rm -f ch3bedEle.log
+rm -f ch4downDiff.aux
+rm -f ch4downDiff.log
+rm -f laySum.aux
+rm -f laySum.log
+rm -f pref.aux
+rm -f pref.log
+rm -f appendix.aux
+rm -f diss.aux
+rm -f diss.bbl
+rm -f diss.blg
+rm -f diss.brf
+rm -f diss.lof
+rm -f diss.log
+rm -f diss.lot
+rm -f diss.out
+rm -f diss.synctex.gz
+rm -f diss.toc
+rm -f ch6conc.aux
+rm -f ch6conc.log
+rm -f ch5langevin.aux
+

### 2021-09-03 revision/formerPapers/grlRrevision2/latexdiff/complier.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+gedit $1 &
+latexmk -pvc -pdf -interaction=nonstopmode $1
+

### 2021-09-23 revision/abs.tex
uh

--- 2021-09-03 draft/abs.tex
@@
 Feedbacks between topography and sediment transport remain challenging to analyze, and basic properties of bedload motions like downstream travel velocities remain incompletely understood.
-Buried particles cannot move downstream, but even this simple observation has not been comprehensively described in the statistical physics approach.
+Buried particles cannot appreciably move downstream, but even this simple observation has not been comprehensively described in the statistical physics approach.
 
@@
 First, I demonstrate how to calculate the sediment flux from the dynamics of individual grains, and I model the trajectories of grains alternating through motion and rest having fluctuating velocities in the motion state.
-This work links the sediment flux to the grain-scale dynamics and describes particle trajectories in unprecedented detail.
-Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates, and predicting how long particles can stay buried in the riverbed.
-Third, I incorporate this sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
-Finally, I concentrate on bedload dynamics at short timescales, predicting the movement velocities of bedload particles using methods adopted from granular physics.
+This links the sediment flux to the grain-scale dynamics and describes particle trajectories with additional detail compared to earlier descriptions.
+Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting how long particles can stay buried in the riverbed.
+Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
+Finally, describe bedload dynamics on short timescales, predicting the movement velocities of bedload particles using methods adopted from granular physics.
 I conclude by summarizing these developments, discussing their implications for the statistical description of bedload transport, and suggesting how we can use this modelling progress to better understand landscapes.

### 2021-09-23 revision/ack.tex
uh

@@
 
-Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Avis, you've formed all of the best parts of my personality and guided me toward this point, and it's impossible to explain how thankful I am for that.
-Regarding family I will also always be thankful for my great aunt Alice who graciously paid for my undergraduate education.
+Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Avis, it's impossible to explain how thankful I am for your guidance.
+Thanks also to my great aunt Alice who graciously paid for my undergraduate education.
 
@@
 Thanks also to my professors at Southern WV Community College who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
-At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble.
-Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science.
-SWVCC was a personal transformation and a great privilege.
+At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble. SWVCC was a personal transformation and a great privilege.
 
@@
 Leo, you are the key person in my development as a researcher.
-Every time I solve a challenging problem with some Bessel functions or contour integrals, I think ``Dr.
-Golubovic would like this one!" I'll be successful if I can one day be as influential an educator as you've been.
 
 To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times.
-I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind).
-I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator for comic relief, and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level.
 May the lounge live on forever!
 
-Now we come to Marwan Hassan.
-Some time ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics.
+To Marwan Hassan, some time ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics.
 I realize now when I said ``statistical physics" your eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well.
 I was lucky to use the right trigger words.
-Marwan, thanks for taking the risk on me, for all of the guidance, for lending me hundreds of books and papers, for giving me freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.
+Marwan, thanks for all of the guidance, for lending me hundreds of books and papers, for the freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.
 
@@
 Thanks to the whole extended research family: Shawn, Matteo, Conor, Nisreen, Yinlue, Alex, Kyle, Leo, Dave, Tobias, Katie, Maria, Elli, Jiamei, Xingyu, Niannian, and Emma, plus the outsider Eaton-ites, Will, Dave, Rose, and Anya.
-In particular thanks to Shawn for all of the shared speculations, and Matteo and Conor -- friendships forged through suffering!
-We'll have to keep in touch and form productive research collaborations once Matteo returns from retirement.
+In particular thanks to Shawn for all of the shared speculations.
 
@@
 Thanks for the thoughtful comments and advice on my PhD.
-I'm not forgetting what I owe you!
-I will always be impressed at your concise way of speaking and unbounded kindness.
-Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modelling in light of real world complexity.
-Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done.
-Maybe they'll bring me closer to your way of thinking.
+Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modeling in light of real world complexity.
 
-I'd be amiss not to mention David Furbish.
-Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation.
+To David Furbish, before I started my PhD when I was still a physics kid exclusively, I recall attending your colloquium at UBC with Marwan's invitation.
 At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes".
 Now I have a thesis on it.
-David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy.
-David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.
+Thanks for the confirmation!
+And to the Furbish people -- Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.
 
 Among UBC Geographers I'd like to thank Nina Hewitt in particular.
-It was always a pleasure and sometimes a sigh of relief to TA your courses, which are so uniquely organized and student-focused.
+It was always a pleasure to work for you on your courses, which are so uniquely organized and student-focused.
 UBC Geography is a much greater department with you in it.
 
-Finally, Mary!
-My PhD has not been great in terms of work-life balance.
-Thanks for sticking it out, putting up with my late bedtimes, early wakeups, the rambling science jargon, and all of the social events I skipped.
-It's time!
-I'm done.
-I'm ready for an extensive tour of the gap and some beautiful Queensland beaches with your family, featuring SPF110 and an incredibly giant umbrella.
-I hope to take up still-lives with Marianne's coaching.
-After that, we're a step closer to settling what colour cat we're gonna get.
-I'm personally now leaning toward white.
+Finally, Mary! My PhD has not been great in terms of work-life balance. Thanks for putting up with my late bedtimes, early wakeups, rambling science jargon, and skipped social events. It's done!
 
 Completing a PhD looks like a personal achievement, but it's really not.
-It's an expression of 30 years of influence and guidance from other people at every stage, my parents, my sisters, my grandmother, and a huge network of teachers and mentors.
-Thank you all for every part of it.
+It's an expression of 30 years of influence and guidance from other people at every stage, my parents, my sisters, my grandmother, and a huge network of teachers and mentors. Thank you all!
 

### 2021-09-23 revision/app_ch2.tex
uh

@@
 The inverse Fourier transform can be calculated by expanding the denominator of eq. \ref{eq:prev} in partial fractions and applying the contour integral \citep[e.g.][]{Arfken1985}
-\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g + i c} dg = \theta(x)e^{-c x}. \label{eq:contour}\ee
+\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g - i c} dg = \sigma(c)\theta(-x\sigma(c))\exp\big(-|c x|\big). \label{eq:contour}\ee
+In this equation $\sigma(c) = \text{sgn}(c)$ is a shorthand for the signum function.
+The above integral can be obtained with the residue theorem by considering different cases for the signs of $x$ and $c$ and applying square contours around the upper or lower-half regions of the complex plane.
 
@@
 
-Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator, the remaining transform can be evaluated from standard tables \citep{Prudnikov1992a}, eventually giving (after integrating over and setting to $1$ the fake parameter),
+Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator (an integral analogue of "Feynman's trick"), the remaining transform can be evaluated from standard tables \citep{Prudnikov1992a}, eventually giving (after integrating over and setting to $1$ the fake parameter),
 \begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du, \end{multline}
-as reported in equation \ref{eq:monster}.
+as reported in eq. \ref{eq:monster}.
 
@@
 \label{sec:fluxconstant}
-The central object required to calculate the sediment flux probability distribution in equation \ref{eq:flippyflux} is 
+The central object required to calculate the sediment flux probability distribution in eq. \ref{eq:flippyflux} is 
 \be \Lambda(T) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,T).\ee
@@
 \be \tilde{\Lambda}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s).\ee
-The Laplace transform of the rate function therefore follows after integrating equation \ref{eq:laplaceapp} twice.
+The Laplace transform of the rate function therefore follows after integrating eq. \ref{eq:laplaceapp} twice.
 
@@
 \end{multline}
-as reported in equation \ref{eq:fluxy}.
+as reported in eq. \ref{eq:fluxy}.
 
@@
 Equation \ref{eq:fluxy} is difficult to work with.
-To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep[e.g.][]{Weiss1994} to invert the Laplace-transformed rate function of equation \ref{eq:laplacefluxrate} at the opposite extreme of $s$.
+To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep{Weiss1994} to invert the Laplace-transformed rate function of eq.
+\ref{eq:laplacefluxrate} at the opposite extreme of $s$.
 
-For example, at short times, expanding equation \ref{eq:laplacefluxrate} as $s\rightarrow \infty$ gives
+For example, at short times, expanding eq. \ref{eq:laplacefluxrate} as $s\rightarrow \infty$ gives
 \be \tilde{\Lambda}(s) = \frac{\rho k_E V}{2 k s^2} + \frac{\rho k_E}{k}\sqrt{\frac{D}{4 s^3}} \ee
@@
 and this inverts to $\Lambda(T) = \rho k_E V T/k$ as the long time solution, providing a mean flux equivalent to the Einstein model (sec. \ref{sec:lisle}).
-These limits are summarized in equation \ref{eq:fluccolimitcases}.
+These limits are summarized in eq. \ref{eq:fluccolimitcases}.

### 2021-09-23 revision/app_ch3.tex
uh

--- 2021-09-03 draft/app_ch3.tex
@@
 It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1991} and \citet{Gillespie2007}.
-The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep[e.g.,][]{Cox1965}.
-The following sections demonstrate that the time intervals $\tau$ between subsequent transitions are exponentially distributed within the model of chapter \ref{ch:ch3}.
+The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep{Cox1965}.
+The following sections demonstrate that the time intervals $\tau$ between subsequent transitions are exponentially distributed within the model of ch.
+\ref{ch:ch3}.
 The SSA follows as a consequence of this property.
@@
 The joint description of bedload transport and bed elevation changes is characterized by a set of states $(n,m)$ where $n$ and $m$ are integers.
-The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in equations (\ref{eq:rate1}-\ref{eq:rate4}).
+The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in eqs. (\ref{eq:rate1}-\ref{eq:rate4}).
 
@@
 
-So far, equation \ref{eq:exp} demonstrates how to step the time from one transition to the next, but it remains unclear how to step the state variables $n$ and $m$ at each transition time.
+So far, eq.
+\ref{eq:exp} demonstrates how to step the time from one transition to the next, but it remains unclear how to step the state variables $n$ and $m$ at each transition time.
 There is a need to characterize the type of transition which occurs at a given transition time.
 
-Intuitively, this will depend on the relative magnitudes of the rates from equations (\ref{eq:rate1}-\ref{eq:rate4}).
+Intuitively, this will depend on the relative magnitudes of the rates from eqs. \ref{eq:rate1}-\ref{eq:rate4}.
 The transition with the highest rate is most likely to occur.
@@
 
-In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (\ref{eq:exp}), draws a uniform random from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in (\ref{eq:rel}).
+In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (eq.
+\ref{eq:exp}), draws a uniform random number from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in eq.
+\ref{eq:rel}.
 Iterating this random number generation and selection process simulates exact realizations of the stochastic process.
@@
 
-Simulations are initialized by the initial conditions $n_0$ and $m_0$, the model parameters used in equations \ref{eq:rate1}-\ref{eq:rate4}, and the desired simulation duration $t_\text{max}$.
+Simulations are initialized by the initial conditions $n_0$ and $m_0$, the model parameters used in eqs.
+\ref{eq:rate1}-\ref{eq:rate4}, and the desired simulation duration $t_\text{max}$.
 The SSA uses these inputs to generate time series of $n$ and $m$ as follows:
@@
 		\State record $(n,m,t)$ \Comment{Build time series of $n$ and $m$}		
-		\State draw $\tau$ from eq. (\ref{eq:exp}) \Comment{Select time to next transition}
+		\State draw $\tau$ from eq. \ref{eq:exp} \Comment{Select time to next transition}
 		\State $t \coloneqq t+\tau$
 		\State draw a random number $r$ in $[0,1]$ \Comment{Select type of transition that occurs}
-		\State compute the ratios $r_1,r_2,r_3,r_4$ in eq. (\ref{eq:rel})
+		\State compute the ratios $r_1,r_2,r_3,r_4$ in eq. \ref{eq:rel}
 		\State form the cumulative sums $r_i = \sum_{1\leq j\leq i}r_j$ 
@@
 
-By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing equation \ref{eq:elemaster} over all values of $m$ obtains a mean field equation for the particle activity:
+By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing eq.
+\ref{eq:elemaster} over all values of $m$ obtains a mean field equation for the particle activity:
 \begin{multline} 0 = \nu A(n-1) + [\lambda + \mu(n-1)]A(n-1) + \sigma(n+1)A(n+1) \\
@@
 after applying the normalization condition $G(1)=1$.
-From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in equation \ref{eq:ancey} and introduced originally in section \ref{sec:birthdeath}.
+From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in eq.
+\ref{eq:ancey} and introduced originally in sec.
+\ref{sec:birthdeath}.
 
@@
 
-The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so equation \ref{eq:elemaster} produces
+The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so eq. \ref{eq:elemaster} produces
 \begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m).
 \end{multline}
-Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives equation \ref{eq:ou}:
+Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives eq. \ref{eq:ou}:
 \be 0 = [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1) - 2M(m).\ee
@@
 \be M = M_0e^{-\kappa (z/z_1)^2}\ee
-as provided in chapter \ref{ch:ch3}.
+as provided in ch. \ref{ch:ch3}.
 
@@
 \be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
-for the mean particle activity conditional to the elevation $m$ into \ref{eq:elemaster}, noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
+for the mean particle activity conditional to the elevation $m$ into eq. \ref{eq:elemaster}, noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
 \begin{multline} 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]\\-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
@@
 \be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \ee
-corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in equation \ref{eq:closure} and plotted with the numerical simulations in figure \ref{fig:condmoms}.
+corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in eq. \ref{eq:closure} and plotted with the numerical simulations in fig. \ref{fig:condmoms}.

### 2021-09-23 revision/app_ch4.tex
uh

--- 2021-07-03 app_ch4.tex
@@
 
-\chapter{Ch. 4 calculations}
+\chapter{Calculations involved in the sediment burial model}
 
@@
 \be \tilde{p}(\eta,s) = \int_0^\infty dx e^{-\eta x}\int_0^\infty dt e^{-st} p(x,t). \label{eq:doubletransform}\ee
-The Laplace-transformed moments of $x$ are linked to derivatives of the double transformed distribution (\ref{eq:doubletransform}) \citep[cf.,][]{Berezhkovskii2002}.
-Equation (\ref{eq:doubletransform}) implies
+The Laplace-transformed moments of $x$ are linked to derivatives of the double transformed distribution (eq. \ref{eq:doubletransform}) \citep[cf.,][]{Berezhkovskii2002}.
+Eq. (\ref{eq:doubletransform}) implies
 \be \bra \tilde{x}(s)^k \ket = (-)^k\partial_\eta^k \tilde{p}(\eta,s)\Big|_{\eta=0}.\label{eq:momenttrick}\ee
@@
 
-Double transforming manuscript equations (1-3) using the definition (\ref{eq:doubletransform}) gives
+Double transforming manuscript equations (1-3) using the definition eq. \ref{eq:doubletransform} gives
 
@@
 \end{align}
-The total probability is $p(x,t) = p_0(x,t) + p_1(x,t) + p_2(x,t)$. Using equations (\ref{eq:A}-\ref{eq:Z}) this becomes, in the double Laplace representation, 
+The total probability is $p(x,t) = p_0(x,t) + p_1(x,t) + p_2(x,t)$. Using eqs. \ref{eq:A}-\ref{eq:Z} this becomes, in the double Laplace representation, 
 \begin{multline}
@@
 \end{multline}
-Plugging the propagators outlined in manuscript equations (8-9) into equation (\ref{eq:lap}) gives 
+Plugging the propagators outlined in manuscript equations (8-9) into eq. \ref{eq:lap} gives 
 \be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}.\label{eq:nicedist}\ee
@@
 \end{multline}
-Using entries 2.2.2.1, 2.2.2.8, and 1.1.1.13 from \citet{Prudnikov1992a} in conjunction with the definition of the Marcum Q-function $ \mathcal{P}_\mu(x,t)$ \citep[e.g.,][]{Temme1996}, and inserting the Heaviside functions to account for the fact that grains can neither travel backwards nor at speeds exceeding $v$, we finally arrive at manuscript equation (10) for the joint distribution $p(x,t)$.
+Using entries 2.2.2.1, 2.2.2.8, and 1.1.1.13 from \citet{Prudnikov1992a} in conjunction with the definition of the Marcum Q-function $ \mathcal{P}_\mu(x,t)$ \citep{Temme1996}, and inserting the Heaviside functions to account for the fact that grains can neither travel backwards nor at speeds exceeding $v$, we finally arrive at manuscript equation (10) for the joint distribution $p(x,t)$.
 
@@
 
-We compute the first two moments of position $x$ and ultimately its variance using equation (\ref{eq:momenttrick}).
-The first two derivatives of the double Laplace transformed distribution (\ref{eq:nicedist}) are
+We compute the first two moments of position $x$ and ultimately its variance using eq.
+\ref{eq:momenttrick}.
+The first two derivatives of the double Laplace transformed distribution eq.
+\ref{eq:nicedist} are
 \be \partial_\eta \tp(\eta,s) = -v \frac{1}{s}\frac{[(s+\kappa + k')s + \kappa k_2][\theta_2(s+\kappa) + k_1]}{[\eta v(s+\kappa +k_1) + (s+ \kappa + k')s+\kappa k_2]^2},\ee
 \be \partial_\eta^2 \tp(\eta,s) = 2v^2 \frac{1}{s} \frac{(s+\kappa+k_1)[(s+\kappa + k')s+\kappa k_2][\theta_2(s+\kappa) + k_1]}{[\eta v(s+\kappa + k_1) + (s+\kappa + k')s+ \kappa k_2]^3}.\ee
-Evaluating these at $\eta=0$ and applying equation (\ref{eq:momenttrick}) provides the Laplace transformed moments
+Evaluating these at $\eta=0$ and applying eq. \ref{eq:momenttrick} provides the Laplace transformed moments
 \be  \frac{\bra\tilde{x}(s)\ket} {v} = \frac{1}{s}\frac{\theta_2(s+\kappa)+k_1}{(s+\kappa+k')s+\kappa k_2} = \frac{1}{s} \frac{\theta_2(s+\kappa)+k_1}{(s+a+b)(s+a-b)}\label{eq:lapmean},\ee
@@
 This equation rearranges to manuscript equation (11).
-The second moment (\ref{eq:lapsecondmom}) is 
+The second moment (eq. \ref{eq:lapsecondmom}) is 
 \begin{multline}
@@
 We determine the diffusion exponents $\gamma$ in the local, intermediate, and global ranges using the two limiting cases described in the discussion of the manuscript.
-Limit (1) is $\kappa \rightarrow 0$. We take this limit in equations (\ref{eq:lapmean}) and (\ref{eq:lapsecondmom}) with initial condition $\theta_1=1$ to obtain
+Limit (1) is $\kappa \rightarrow 0$. We take this limit in eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} with initial condition $\theta_1=1$ to obtain
 \begin{align}
@@
 We further investigate limit (1) for arbitrary initial conditions.
-By applying Tauberian theorems, we assert the $ t \rightarrow 0$ variance is determined by the $s\rightarrow \infty$ limits of (\ref{eq:lapmean}) and (\ref{eq:lapsecondmom}) \citep[e.g.,][]{Weiss1994, Weeks1998}.
+By applying Tauberian theorems, we assert the $ t \rightarrow 0$ variance is determined by the $s\rightarrow \infty$ limits of eqs.
+\ref{eq:lapmean} and \ref{eq:lapsecondmom} \citep[e.g.,][]{Weiss1994, Weeks1998}.
 Expanding these equations in powers of $1/s$ and inverting the resulting transforms gives
@@
 We have taken only leading order terms for any option of $\theta_1$ and $\theta_2$.
-Equation (\ref{eq:init}) shows local range exponent $\gamma=2$ when initial conditions are mixed (both are non-zero) and $\gamma=3$ when initial conditions are pure (one is zero).
+Eq. \ref{eq:init} shows local range exponent $\gamma=2$ when initial conditions are mixed (both are non-zero) and $\gamma=3$ when initial conditions are pure (one is zero).
 
-Limit (2) is $1/k_2 \rightarrow 0$ and $v\rightarrow \infty$ while $v/k_2 = l$. Under this limit, equations (\ref{eq:lapmean}) and (\ref{eq:lapsecondmom}) provide
+Limit (2) is $1/k_2 \rightarrow 0$ and $v\rightarrow \infty$ while $v/k_2 = l$. Under this limit, eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} provide
 \begin{align}
@@
 Expanding for small $t$ provides $\sigma_x^2(t) = 2D_d t$ -- intermediate range normal diffusion, while sending $t\rightarrow \infty$ provides $\sigma_x^2 = (2D_d\kappa + c^2)/\kappa^2$ -- a constant variance in the geomorphic range.
-The global range is characterized by competition between terms in equation (\ref{eq:wuvar}), and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.,][]{Wu2019}.
-Finally, both equations (\ref{eq:li}) and (\ref{eq:wuvar}) reduce to the Einstein result $\sigma_x^2(t) = 2D_d t$ in further simplified limits.
+The global range is characterized by competition between terms in eq. \ref{eq:wuvar} and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.,][]{Wu2019}.
+Finally, both eqs. \ref{eq:li} and \ref{eq:wuvar} reduce to the Einstein result $\sigma_x^2(t) = 2D_d t$ in further simplified limits.

### 2021-09-23 revision/app_ch5.tex
uh

@@
 \label{sec:langmasterderiv}
-To derive the master equation from \ref{eq:langevin}, we temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 d} h$ with $h$ distributed as $f(h)$.
+To derive the master equation from eq.
+\ref{eq:langevin}, we can temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 \tilde{D}} h$ with $h$ distributed as $f(h)$.
 We will later take a GWN limit on this noise.
-With this assumption, integrating \ref{eq:langevin} over a small time interval $\delta t$ considering the Ito interpretation for the collision term provides
+With this assumption, integrating eq.
+\ref{eq:langevin} over a small time interval $\delta t$ and incorporating the Ito interpretation for the collision term provides
 \be     
@@
 \begin{cases}
-	u(t) + \gamma \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
-	u(t) + \sqrt{2 d} h &  \text{with probability } r \delta t \\
+	u(t) + \tilde{\Gamma} \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
+	u(t) + \sqrt{2 \tilde{D}} h &  \text{with probability } r \delta t \\
 	\ve u(t) &  \text{with probability } \nu \delta t
-\end{cases}
+\end{cases}.
 \ee
@@
 \begin{align} P(u,t+\delta t) =
-	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\gamma \delta t) P(w,t)\\ 
-	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 d}h) P(w,t) \\ 
+	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\tilde{\Gamma} \delta t) P(w,t)\\ 
+	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 \tilde{D}}h) P(w,t) \\ 
 	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
@@
 Evaluating all integrals over $\delta$-functions provides 
-\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\gamma \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2d}h) \\+ \nu \delta t \int_0^1 \frac{d\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
-Finally, we take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2d}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanDenBroeck1983}.
+\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\tilde{\Gamma} \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2 \tilde{D}}h) \\+ \nu \delta t \int_0^1 \frac{ \tilde{D}\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
+Finally, take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanDenBroeck1983}.
 This process finally obtains the master equation (\ref{eq:master}).
@@
 \label{sec:langsteadyderiv}
-Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of \ref{eq:master} develops the recursion relation
+Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of eq. \ref{eq:master} develops the recursion relation
 \be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
@@
 \be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
-The polynomials $q(z)$ can always be factored as $q(z) = d(z - i\lambda_-)(z - i\lambda_+)$ where
+The polynomials $q(z)$ can always be factored as $q(z) = \tilde{D}(z - i\lambda_-)(z - i\lambda_+)$ where
 \be \lambda_\pm = \frac{\tilde{\Gamma}}{2\tilde{D}}\Big[1 \pm \sqrt{ 1 + 4\tilde{D}/\tilde{\Gamma}^2} \Big]. \label{eq:lambdas}\ee
@@
 \be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
-The Fourier transform (\ref{eq:recursion}) has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
+The Fourier transform (eq.
+\ref{eq:recursion}) has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
 Taking this limit and evaluating the residues provides 
@@
 	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
-Finally, inverting the Fourier transforms term by term with contour integration (another use of the integral given in eq.
-\ref{eq:contour}), and incorporating eq.
-(\ref{eq:q}), provides the steady-state solution (\ref{eq:steadystate}).
+Finally, we can invert the Fourier transforms term by term with contour integration (another use of the integral given in eq.
+\ref{eq:contour}), and incorporate eq.
+\ref{eq:q} to obtain the steady-state solution (eq.
+\ref{eq:steadystate}).
 
@@
 \label{sec:langmoments}
-Taking (\ref{eq:master}), multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
+Taking eq. \ref{eq:master}, multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
 \be 0 = \tilde{D} k (k-1)\langle u^{k-2} \rangle + \tilde{\Gamma} k \langle u^{k-1} \rangle + (\ve^k-1)\langle u^k \rangle. \ee
@@
 \be \sigma_u^2 = \frac{2\tilde{D} + \tilde{\Gamma}^2}{1-\ve^2}\ee
-given in equation \ref{eq:varu}.
+given in eq. \ref{eq:varu}.
 \section{Weak and strong collision limits}
@@
 Now I will demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
-The limit is challenging since the steady-state distribution \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
+The limit is challenging since the steady-state distribution eq.
+\ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
 Following \cite{Hall1989}, this divergence suggests normalizing the distribution $P(u)$ using the scaled variable
@@
 The resulting differential equation for $Q(z)$ is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
-Incorporating this transformation into (\ref{eq:master}) provides the ``normalized" master equation
+Incorporating this transformation into eq. \ref{eq:master} provides the ``normalized" master equation
 \begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 \end{multline}
-This equation remains exact and is only a change of variables from (\ref{eq:master}).
+This equation remains exact and is only a change of variables from eq. \ref{eq:master}.
 Now I approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
@@
 which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$ \citep[e.g.][]{Gardiner1983}.
-This solution provides \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.
+This solution provides eq. \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-09-23 revision/ch1intro.tex
uh

@@
 
-Landscapes evolve when water, wind, and ice, under the influence of gravity and sunlight, erode along topographic gradients produced by tectonics.
+Landscapes evolve when water, wind, and ice erode along the gradients produced by tectonics.
 Channels initiate along faults and depressions wherever climatic conditions are suitable, incising networks in the landscape and transferring sediments from uplands to lowlands.
@@
 The same turbulence and sediment organization details which correlate with the bedload flux are also modified by it.
-Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulent flow characteristics \citep{Singh2010, Santos2014, Liu2016}.
-The surface arrangement and consolidation of grains \citep{Miller1966,Paintal1971,Dwivedi2012} affects sediment mobility, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018, Masteller2019, Pretzlav2020}.
+Turbulent impulses drive sediment motion \citep{Valyrakis2010, Celik2014, Amir2014, Shih2017}, but moving sediment affects turbulence characteristics \citep{Singh2010, Santos2014, Liu2016}.
+The surface arrangement and consolidation of grains affects sediment mobility \citep{Miller1966,Paintal1971,Dwivedi2012}, but transport rearranges grains and consolidates the bed \citep{Kirchener1990, Charru2004, Allen2018, Masteller2019, Pretzlav2020}.
 Bedload fluxes are in cyclical feedback with their controls \citep{Jerolmack2005}, and this challenges us to step beyond descriptions based on averaged values \citep{Ancey2020b}.
@@
 To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
-To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes and show continuity with my own developments which follow in subsequent chapters.
+To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes, highlight the shortcomings to be addressed, and show continuity with my own developments which follow in subsequent chapters.
 
@@
 At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
-Particles moving as bedload are driven downstream by a turbulent fluid flow, but precise relationships between the fluid flow and the applied forces remain unknown \citep{Maxey1983,Schmeeckle2007}.
+Particles moving as bedload are driven downstream by a turbulent fluid flow, but relationships between the fluid flow and the applied forces are known only approximately \citep{Michaelides1997a,Schmeeckle2007}.
 Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.
@@
 Particles at rest on the bed surface can also become covered by other transported particles \citep{Yang1971}. 
-These buried particles cannot move along the bed surface again until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of relative immobility \citep{Hassan1994,Ferguson2002}.
+These buried particles cannot move appreciably until those burying them have been transported away \citep{Nakagawa1981}, generating long periods of relative immobility \citep{Hassan1994,Ferguson2002}.
 
@@
 \citet{Nikora2001,Nikora2002} provided a conceptual framework which helps to organize this complexity.
-Nikora et al divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
+Nikora et al. divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
 The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
@@
 \subsection{Motivation: tracers and basic understanding}
+
 An early motivation to understand individual particle motions was to predict the efficiency of sediment transport measurements \citep{Ettema2004}, and this motivation still drives a great deal of research into individual particle motions today \citep{Hassan2017,Pretzlav2021}.
@@
 As his interest was on the global range of particle motion when many motion-rest alternations have occurred, and because the duration of particle motions is usually short compared to rests, Einstein made the approximation that individual motions (between entrainment and deposition) are mathematically instantaneous.
-With this approximation, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length, separated by rests of random duration. 
-Implicitly, this picture of sediment trajectories assumes that movement velocities are infinite.
+With this approximation, the downstream movement of sediment becomes an alternate cycle of instantaneous steps of random length and rests of random duration. 
+Implicitly, this concept of sediment trajectories assumes that movement velocities are infinite.
 
 Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
-In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this was not formalized until much later \citep{Montroll1965}.
+In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this approach not formalized until much later \citep{Montroll1965}.
+In Earth science, random walks have been widely applied to hydrologic problems \citep{Berkowitz2006}, but their application to surface processes remains relatively uncommon \citep[e.g.][]{Schumer2009}.
 
@@
 \be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
-where $\dot{x} = dx/dt$, and $\mu(t)$ is a white shot noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
+where $\dot{x} = dx/dt$, and $\mu(t)$ is a Poisson pulse noise \citep{VanDenBroeck1983}, which is essentially a sequence of pulses having random heights with mean height $\ell$ (the mean step length), and random locations in time with mean separation $1/k_E$ (the mean resting time, interpreted as the reciprocal of the entrainment rate $k_E$).
 A particular realization of this pulsed noise can be written
@@
 
-Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (equation \ref{eq:einrando}).
+Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (eq.
+\ref{eq:einrando}).
 \begin{figure}[!htbp]
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
-	\caption{Panel (a) indicates the representation of Einstein's model as an idealized ``white shot noise", as indicated in eq.
+	\caption{Panel (a) indicates the representation of Einstein's model as an idealized Poisson pulse noise, as indicated in eq.
 \ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$).
@@
 %where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
-This noise is displayed in figure \ref{fig:lislefig} panel (a). Some particle trajectories produced by equation \ref{eq:lislelangevin} are displayed in figure \ref{fig:lislefig} panel (b).
+This noise is displayed in fig. \ref{fig:lislefig} panel (a). Some particle trajectories produced by eq. \ref{eq:lislelangevin} are displayed in fig. \ref{fig:lislefig} panel (b).
 
@@
 
-For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of equation \ref{eq:lislemaster} is \citep{Lisle1998}
+For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of eq.
+\ref{eq:lislemaster} is \citep{Lisle1998}
 \begin{multline}  P(x,t) =  e^{-\chi-\tau} \Big[ \frac{k_E}{V}\delta(\tau) + \frac{k_E}{V} \sqrt{\frac{\chi}{\tau}}\mathcal{I}_1(2\sqrt{\chi\tau}) + \frac{k_D}{V}\mathcal{I}_0(2\sqrt{\chi\tau}) \\
@@
 which is a non-trivial result.
-At short times, for $t\ll 1/k$, equation \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}.
+At short times, for $t\ll 1/k$, eq.
+\ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}.
 At long times ($t\gg 1/k$), the diffusion becomes normal again ($\sigma_x^2 = 2 D_\text{eff} t$), with an effective diffusion constant $ D_\text{eff} = k_E k_D V^2/k^3$.
@@
 
-Some authors have modelled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
+Some authors have modeled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
 Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.
@@
 
-The state of the art within this category of sediment transport models is to include two-way coupling between particles and the fluid flow.
+The state of the art within this category of sediment transport models is to simulate both fluid and particle phases simultaneously.
 Particle-particle interactions are represented with the discrete element method, where particles are represented as spheres \citep{Cudnall1979} or polytopes \citep{Wachs2019} while forces are phrased as idealized spring-dashpot models.
-The latter is modelled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
-Particles are represented either as spheres coupled to the flow \citep{}, 
- while particles provide the boundary condition for the flow \citep{Schmeeckle2014,Ji2013,Gonzalez2017,Vowinckel2014,Elghannay2017,Yousefi2020}.
-These computational physics models produce impressive insight into the underlying granular and fluid physics mechanisms producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
+The fluid flow is modeled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
+A majority of these approaches simulate the fluid without including no-slip boundary conditions on the particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although works are beginning to include the fluid boundary conditions on individual particles with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
+These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
 
@@
 This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
-The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from equation \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007}
+The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from eq.
+\ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007}
 \be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
-implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by equation \ref{eq:fanlangevin} is
+implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by eq. \ref{eq:fanlangevin} is
 \be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
@@
 which is the Gaussian velocity distribution from the other subset of the experiments.
-These two models of bedload velocity distributions are summarized in figure \ref{fig:fanAncey}.
+These two models of bedload velocity distributions are summarized in fig. \ref{fig:fanAncey}.
 \begin{figure}[!htbp]
@@
 Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and other distributions have been reported besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
-A comprehensive model of the full range of bedload transport observations is a desirable target.
+A comprehensive model of the full range of bedload transport observations remains a desirable target.
 
 \subsection{The definition of the flux}
+% for some reason the pdf of this figure does not properly render 
 \begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.pdf}
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.png}
 	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$.
-Here, particles are shown moving from time $t-dt$ (faint white particles) to time $t$ (grey particles).
-Each particle's velocity component $u_i$ in the downstream direction is indicated.
-At the instant $t$, the surface definition of the flux \ref{eq:surfflux} gives $q = \frac{1}{L}(u_7+u_8).$, while the volume definition \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_6)$,
+Here, particles are shown moving from time $t-\Delta t$ (faint white particles) to time $t$ (grey particles).
+Each particle's downstream displacement $\Delta x_i$ is indicated.
+These displacements define the downstream particle velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux eq.
+\ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition eq.
+\ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_5)$,
 	showing the non-equivalence of these two definitions.}
@@
 Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
-Today, there are two main complementary definitions of the bedload flux.
+Today, there are two main complementary definitions of the bedload flux being applied in stochastic modeling. Other definitions are discussed comprehensively in \citet{Ballio2014,Ballio2018}.
 
-The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016, Ballio2014}: 
+The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016,Ballio2014}: 
 \be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
-This definition involves the concentration $c$ of particles in space and their velocites at the instant they cross the control surface, which is a somewhat elusive quantity since bedload particles are not a continuous field \citep{Heyman2016}.
+This definition involves the volumetric concentration $c$ of particles and their velocites at the instant they cross the control surface.
+These quantities require careful interpretation because bedload particles do not constitute a continuous field \citep{Furbish2012a}.
+Typically one represents the concentration as a sum of indicator or Dirac delta functions \citep[e.g.][]{}.
 
@@
 Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
-\citet{Einstein1937} formulated his original model of individual particle trajectories in terms of entrainment and deposition rates using the idealization of grains moving downstream through a sequence of instantaneous steps.
+Einstein formulated his original model of individual particle trajectories in terms of entrainment and deposition rates using the idealization of grains moving downstream through a sequence of instantaneous steps \citep{Einstein1937}.
 Later, Einstein calculated the sediment flux with these same probabilistic ideas \citep{Einstein1942, Einstein1950}.
-This formulation of the sediment flux would come back to later provide an alternative to Bagnold's scaling approach.
-The conceptual picture that Einstein considered is depicted in figure \ref{fig:einsteinFluxConcept}.
+This formulation of the sediment flux now provides an alternative to Bagnold's scaling approach.
+The conceptual picture that Einstein considered is depicted in fig. \ref{fig:einsteinFluxConcept}.
  \begin{figure}[!htbp]
@@
 \be p_n = \frac{(k_E T)^n}{n!}e^{-k_E T}.\ee
-This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^n}{n!}e^{-k_E T}, $
+This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^i}{i!}e^{-k_E T}, $
 so Einstein's mean sediment flux across the control surface (eq. \ref{eq:einflux}) becomes
@@
 This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$.
-These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}, in what can be characterized as a mixture of the Einstein and Bagnold strategies.
+These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
+This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its reliance on both probabilistic rates and scaling relations.
 
@@
 Equation \ref{eq:charru} indicates that the mean flux in steady transport and flow conditions is the implicit solution to the equation $E = D.$
-Using the scaling relations \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
+Using the scaling relations eqs. \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
 \be \gamma = \frac{\alpha_1}{\alpha_2}\frac{\tau-\tau_c}{d^2 V_s^2}.\ee
@@
 \be F(x,t) = \int_x^\infty \delta(x-Vt)k_D\exp(-k_D t) dx = \theta(Vt-x)k_D\exp(-k_D t).\ee
-Considering uniform conditions with a density $\rho_s$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the areal entrainment rate can be expressed as $E=\rho_s k_E$, and equation \ref{eq:furbo} provides a mean flux
+Considering uniform conditions with a density $\rho_s$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the areal entrainment rate can be expressed as $E=\rho_s k_E$, and eq.
+\ref{eq:furbo} provides a mean flux
 \be q = \rho_s k_E \int_0^\infty dx' \int_0^\infty dt \theta(Vt-x)k_D\exp(-k_D t) = \rho_s k_E V/k_D. \label{eq:furbflux}\ee
-Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and equation \ref{eq:furbflux} provides another perspective on Einstein's central result $q= E\ell$, but with the added ability to describe unsteady conditions \citep{Furbish2012}.
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and eq.
+\ref{eq:furbflux} provides another perspective on Einstein's central result $q= E\ell$, but with the added ability to describe unsteady conditions \citep{Furbish2012}.
 
@@
 \be \partial_x q(x,t) = E - D. \ee 
-In combination with equation \ref{eq:exner}, this expression of the sediment flux gradient implies the relation
+In combination with eq. \ref{eq:exner}, this expression of the sediment flux gradient implies the relation
 \be (1-\phi) \px z(x,t) = D - E\ee
 by which topographic evolution can be described in terms of Einstein's rates.
-In a nonlocal framework, as in equation \ref{eq:furbo}, sediment deposition can be interpreted as the eventual result of entrainment from all upstream locations, giving
+In a nonlocal framework, as in eq. \ref{eq:furbo}, sediment deposition can be interpreted as the eventual result of entrainment from all upstream locations, giving
 \be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
@@
 \be \sigma_n^2 = \frac{\lambda \alpha}{(\alpha-\mu)^2}. \ee
-Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\alpha/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, as the distribution (\ref{eq:negbin}) limits to a Poisson distribution where the mean and variance are equal \citep{Ancey2006}.
+Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\alpha/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, as the distribution eq.
+\ref{eq:negbin} limits to a Poisson distribution where the mean and variance are equal \citep{Ancey2006}.
 
@@
 \ref{eq:controlflux} given the probability distribution of particle velocites $P(u)$ under the assumption that velocity and activity are independent.
-Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep[e.g.][]{Ancey2014}
+Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep{Ancey2014}
 \be P(\Phi) = L \sum_{k=0}^\infty P_k(L\Phi) P(n=k). \ee
 A limitation of this approach is that in reality, entrainment and deposition modify the bed elevation, which in turn modifies the rates of entrainment and deposition \citep{Sawai1987,Wong2007}. 
-This highlights a need to evaluate bed elevation change in statistical sediment transport models.
+This highlights a need to incorporate bed elevation change in statistical sediment transport models.
 
@@
 The exponential inter-arrival time distribution is $P(t) = \Lambda \exp(-\Lambda t),$ so the rate of particle arrivals is $\Lambda$.
-In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of section \ref{sec:einwalk}:
+In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of sec. \ref{sec:einwalk}:
 \be q(T) = \frac{1}{T}\int_0^T dt' \mu(t'). \label{eq:flucflux}\ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
-as indicated in figure \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$. 
+as indicated in fig. \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$. 
 The flux in eq.
-\ref{eq:flucflux} is a random variable as indicated in figure \ref{fig:ancey} panel (b).
+\ref{eq:flucflux} is a random variable as indicated in fig.
+\ref{fig:ancey} panel (b).
 Its probability distribution $P(q|T)$, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
@@
 
-In addition, although the scale-dependent sediment flux has been calculated from renewal models \citep{Turowski2010,Ancey2020}, these approaches are not based on the dynamics of individual particles in transport, so it remains unclear how the sediment flux distribution and its scale dependence originate from the transport characteristics of individual grains.
+In addition, although the scale-dependent sediment flux has been calculated from renewal models \citep{Turowski2010,Ancey2020a}, these approaches are not based on the dynamics of individual particles in transport, so it remains unclear how the sediment flux distribution and its scale dependence originate from the transport characteristics of individual grains.
 
@@
 
-Modelling sediment transport as an alternating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface.
+Modeling sediment transport as an alternating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface.
 Although this provides a useful description of bedload transport across local and intermediate timescales, it requires modification to describe longer global and geomorphic timescales.
-At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017}, but burial has not yet been incorporated into Einstein-type models of particle trajectories.
+At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017,Bradley2017}, but burial has not yet been incorporated into Einstein-type models of particle trajectories.
 
@@
 Such processes introduce temporal correlations into sediment transport signals which control the sampling interval over which measurements of the mean sediment flux converge \citep{Saletti2015,Singh2009,Singh2012}.
-To date, very few modelling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
+To date, very few modeling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
 

### 2021-09-23 revision/ch2flipFlop.tex
uh

@@
 No models have yet been formulated which describes particles alternating between motion and rest with motions having realistic fluctuating velocities, even though this is how bedload particles actually move.
+A more realistic description of particle velocities during individual movements has potential to improve our understanding of bedform initation and particle-particle interactions in subsequent developments.
 
@@
 To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
-Among those that do include scale dependence, none formulate the sediment flux in terms of the downstream trajectories of individual grains.
+Among those that do include scale dependence, none formulate the sediment flux in terms of the trajectories of individual grains.
 
 This survey provides context for the two problems addressed in this chapter.
-First, we lack the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall downstream sediment flux, including its probability distribution and the dependence of its statistical moments on the observation time.
+First, we lack the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall streamwise sediment flux, including its probability distribution and the dependence of its statistical moments on the observation time.
 
@@
 \be \dot{x}(t) = [V + \sqrt{2D}\xi(t)]\eta(t).  \label{eq:flippylangevin} \ee
-Here $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles.
+Here,  $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles.
 $\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion (with mean duration $1/k_D$), and $\eta=0$, representing rest (with mean duration $1/k_E$).
-Some trajectories produced by this equation are sketched in figure \ref{fig:fluxxy0}.
-In this equation, $V$ is the mean particle velocity describing the overall downstream drift of moving particles, while $D$ is a diffusivity [units $L^2/T^3$] describing velocity fluctuations among moving particles.
 The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$. The notation $k=k_E+k_D$ is a shorthand used throughout the thesis.
+Times spent in motion and rest are respectively distributed as $P(t) = k_D \exp(k_D t)$ and $P(t) = k_E \exp(k_E t)$.
+$V$ is the mean particle velocity describing the overall downstream drift of moving particles, while $D$ is a diffusivity [units $L^2/T^3$] describing velocity fluctuations among moving particles.
 
-The Langevin equation \ref{eq:flippylangevin} is a generalization of the model by \citet{Lisle1998} and \citet{Lajeunesse2017}, presented in section \ref{sec:lisle}, to include velocity fluctuations in the motion state.
-It is conceptually similar to the ``randomly flashing diffusion" models developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}.
 Equation \ref{eq:flippylangevin} describes the downstream movement of particles alternating through motion and rest with velocity fluctuations in the motion state.
-This dynamical equation provides a more realistic description of bed load trajectories than earlier works in the Einstein paradigm (sec.
-\ref{sec:trajmodels}).
+This model can be understood as a generalization of the ``randomly flashing diffusion" models developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}.
+Some trajectories produced by the Langevin equation \ref{eq:flippylangevin} are sketched in figure \ref{fig:fluxxy0}.
+The driving term in panel (a) involves unrealistic fluctuations due to the assumption that velocities evolve as Gaussian white noise, but the integrated particle trajectories in panel (b) are well-behaved and visually similar to those of earlier studies \citep[cf.][]{Fan2016,Bialik2015}.
+Eq.
+\ref{eq:flippylangevin} generalizes the constant velocity model of \citet{Lisle1998} and \citet{Lajeunesse2017} that was presented in section \ref{sec:lisle}, providing the first extension of these works aimed at a more realistic description of bedload particle trajectories through motion and rest.
 
@@
 \ref{eq:flippylangevin} and the average is over both noises.
-This symbolic equation is not directly useful as taking averages over both noises is a challenging mathematical problem \citep{Hanggi1978}.
-This calculation is completed in the appendix in section \ref{sec:appAmaster}, providing the master equation
+This symbolic equation is not directly useful as taking averages over both noises is a challenging mathematical problem \citep[e.g.][]{Hanggi1978}.
+This calculation is completed in the appendix sec.
+\ref{sec:appAmaster}, providing the master equation
 \be (\pt^2 + V \px \pt + k_E V \px + k \pt - D \px^2 \pt - k_E D \px^2) P(x,t) = 0 \label{eq:flippymaster}\ee
-for the probability distribution of position.
-The master equation \ref{eq:flippymaster} is a diffusion-like equation governing the probability distribution of position for individual particles alternating between motion and rest, with the movement velocity being a fluctuating quantity.
+for the position probability distribution.
+The master equation \ref{eq:flippymaster} is a diffusion-like equation governing the probability distribution of position for individual particles alternating between motion and rest, with the movement velocity considered as a fluctuating quantity.
 
-One can see in particular that taking the entrainment rate $k_E$ very large, meaning that particles are very often moving, implies an advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
-Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process interrupted, giving rise to the additional terms in eq.
-\ref{eq:flippymaster}, a structure reminiscent of equation \ref{eq:lislemaster} from simpler motion-rest models.
+One can see in particular that taking the entrainment rate $k_E$ very large, meaning that particles are very often moving, implies a classical advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
+Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process is interrupted, giving rise to the additional terms in eq.
+\ref{eq:flippymaster} and providing an asymmetrical structure reminiscent of equation \ref{eq:lislemaster} from the simpler motion-rest models reviewed before.
 
@@
 This method is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
-This generalizes the nonlocal formulation (\ref{sec:nonlocal}) to provide the probability distribution of the flux, and it reframes the renewal approach (sec.
+This generalizes the nonlocal formulation of sec.
+\ref{sec:nonlocal} to provide the probability distribution of the flux, and it reframes the renewal approach (sec.
 \ref{sec:renewal}) in terms of the mechanics of individual particles in transport.
@@
 The basic idea, as depicted in Figure \ref{fig:flipflopfig1}, is to initially distribute $N$ particles in all states of motion along a domain of length $L$ at some random initial locations $x_i$ to the left of $x=0$.
-Later, the number of particles $N$ and the size $L$ of the domain will be limited to infinity such that their ratio $\rho=N/L$ remains constant. 
+Later, the number of particles $N$ and the size $L$ of the domain will be extended to infinity such that their ratio $\rho=N/L$ remains constant. 
 This limit will provide a configuration similar to the one constructed in the nonlocal formation (sec. \ref{sec:nonlocal}).
@@
 In this equation, the $I_i(T)$ are indicator functions which equal $1$ if the $i$th particle has passed the control surface ($x=0$) at the observation time $T$ and $0$ otherwise.
-All particles which have not crossed the control surface (or which have crossed and then crossed back) contribute nothing to the flux.
+All particles which have not crossed the control surface (or which have crossed and then crossed back) do not contribute to the flux.
 
@@
 The average over initial conditions and possible trajectories for the $i$th particle involved in this characteristic function can be written
-\be \bra I_i(t) \ket = \frac{1}{L}\int_L^0 dx' \int_0^\infty dx P(x-x',t) \label{eq:indicator} \ee
+\be \bra I_i(t) \ket = \frac{1}{L}\int_{-L}^0 dx' \int_0^\infty dx P(x-x',t) \label{eq:indicator} \ee
 where $P(x,t)$ is the position probability distribution which solves equation \ref{eq:flippymaster}.
@@
 From the initial mixture of motion and rest states, particles advect downstream as they diffuse apart from one another due to differences in their velocities and transition times between motion and rest.
+In panel (a), the initial position persists as a Delta-function spike for the black $t=10\tau$ curve.
 Panel (b) shows the resulting particle diffusion (eq.
@@
 \be P(x,0) = \lim_{t\rightarrow 0 } \frac{k_E}{k} \sqrt{\frac{1}{4\pi D t}} \exp\Big[-\frac{(x-Vt)^2}{4Dt}\Big]+ \frac{k_D}{k}\delta(x)\ee
-which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep[c.f.][]{Weiss2002a}.
+which gives $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ \citep[cf.][]{Weiss2002a}.
 
@@
 	 t, & t\gg \frac{1}{k}. \label{eq:diffo}
-\end{cases}\ee
+\end{cases} \ee
 Note in the physical condition when $k\approx k_D$, which is generally satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the above condition for existence of three ranges becomes $ \Pe \gg 1,$ where
@@
 
-The large Peclet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles very rarely move upstream \citep[e.g.][]{Fathel2015}.
-This limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed \citep[c.f.][]{Nikora2001,Nikora2002}.
-Figure \ref{fig:flippyfig1} panel (b) sketches this three-range diffusion.
+The large Peclet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles rarely move upstream any appreciable distance \citep[e.g.][]{Fathel2015}.
+This limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed \citep[cf.][]{Nikora2001,Nikora2002}, at least within the assumptions of the model.
+Figure \ref{fig:flippyfig1} panel (b) sketches the predicted three-range diffusion, which is discussed further in sec. \ref{sec:prollems}.
 
-Below, this anomalous diffusion behavior \citep[c.f.][]{Sokolov2012} is shown to affect the rate constant $\Lambda(t)$ in the sediment flux.
+Below, this anomalous diffusion behavior \citep[cf.][]{Sokolov2012} is shown to affect the rate constant $\Lambda(t)$ in the sediment flux.
 As a result of these multi-scale particle dynamics, the sediment flux becomes scale-dependent in a richer way than the renewal model of section \ref{sec:renewal}.
@@
 
-From the formalism in section \ref{sec:flippyflux}, the central parameter of the sediment flux distribution is the rate function
+From the formalism in sec. \ref{sec:flippyflux}, the central parameter of the sediment flux distribution is the rate function
 \be \Lambda(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t). \label{eq:flatform}\ee
@@
 
-Using the probability distribution of position \ref{eq:monster}, the integrals in equation \ref{eq:flatform} are performed in section \ref{sec:fluxconstant} of the appendix, providing the rate function
+Using the probability distribution of position eq. \ref{eq:monster}, the integrals in eq. \ref{eq:flatform} are performed in sec. \ref{sec:fluxconstant} of the appendix, providing the rate function
 \begin{multline} 
@@
 \end{multline}
-In this equation, the notation $\cev{\partial_t}$ means that the partial time derivative acts from the left of all terms in which it is involved, as in $f(t)\cev{\partial}_t g(t) = \partial_t [f(t)g(t)].$
+In this equation, the notation $\cev{\partial_t}$ means that the partial time derivative acts from the left of all terms in which it is involved, as in $f(t)\cev{\partial_t} g(t) = \partial_t [f(t)g(t)].$
 
-This is an intricate result for the rate function in the sediment flux distribution \ref{eq:flippydist}.
-This mathematical complexity may not be surprising given that equation \ref{eq:flippylangevin} involves two interacting diffusion processes.
-As displayed in figure \ref{fig:fluxconvergence}, as a result of eq.
+This is an intricate result for the rate function in the sediment flux distribution eq.
+\ref{eq:flippydist}.
+This mathematical complexity may not be surprising given that eq.
+\ref{eq:flippylangevin} involves two interacting diffusion processes.
+As displayed in fig.
+\ref{fig:fluxconvergence}, as a result of eq.
 \ref{eq:fluxy} the mean sediment flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $q_0 = \rho k_E V/k = E \ell$ as the observation time becomes much larger than $1/(k_D \Pe)$.
@@
 Stronger velocity fluctuations (smaller $\Pe$) slow the convergence to this limit.
-	In all cases, satisfactory convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by equation \ref{eq:fluccolimitcases}.
+	In all cases, satisfactory convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by eq.
+\ref{eq:fluccolimitcases}.
 Plotted lines are the analytical result eq.
@@
 Further insight into the rate constant of the sediment flux distribution can be gained by investigating extreme cases of the observation time.
-As shown in appendix section \ref{sec:fluxlimits}, equation \ref{eq:fluxy} takes on relatively simple forms at extreme values of $T$: 
+As shown in appendix sec. \ref{sec:fluxlimits}, eq. \ref{eq:fluxy} takes on relatively simple forms at extreme values of $T$: 
 \be \Lambda(T) =
@@
 It is related to the Peclet number and is proportional to the time particles spend in motion ($1/k_D$).
-As a result, when particle velocity fluctuations during motions become large comparable to the mean particle velocity, equation \ref{eq:fluccolimitcases} indicates that the flux becomes slow to converge to the Einstein result.
+As a result, when particle velocity fluctuations during motions become large comparable to the mean particle velocity, eq.
+\ref{eq:fluccolimitcases} indicates that the flux becomes slow to converge to the Einstein result.
 
@@
 
-The sediment flux probability distribution in equation \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
+The sediment flux probability distribution in eq. \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
 Poisson distributions have relatively thin tails, meaning their fluctuations are typically small \citep{Ancey2006}.
@@
 
-Collective entrainment has been attributed to particle-particle interactions such as small granular avalanches and collision-induced entrainment \citep[e.g.][]{Pahtz2020}, and to fluid-particle interactions, such as large eddies entraining particles en masse as they sweep downstream \citep{Ancey2014,Lee2018,Cameron2020}.
+Collective entrainment has been attributed to particle-particle interactions such as small granular avalanches and collision-induced entrainment \citep[e.g.][]{Lee2018, Pahtz2020}, and to fluid-particle interactions, such as coherent structures entraining particles en masse as they sweep downstream \citep{Ancey2014,Cameron2020}.
+Given the prevalance of large coherent structures, gravel beds must often times be conditioned to them.
+Therefore particle-particle interactions seem more plausible a mechanism for collective entrainment.
+In particular, contact forces within granular media are known to organize into ``force chains", where a minority of contacts provide a majority of the granular material's strength \citep[e.g.][]{Radjai1996, Azema2012}.
+Fluid entrainment of a key particle within a near-surface force chain could entail a collective release of the particles it stabilizes, consistent with current understanding of collective entrainment.
 
-Particle-particle interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here. 
-The dynamical equation \ref{eq:flippylangevin} would generalize to $\dot{x}_i(t) = F_i(x_i,t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving terms unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+Such particle-particle interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here, but the basic structure can be sketched.
+The dynamical eq.
+\ref{eq:flippylangevin} would generalize to a stochastic Newtonian form $\ddot{x}_i(t) = F_i(x_i,\dot{x}_i, t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving forces unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+This forcing structure would alternate between moving and resting modes by a dichotomous process \citep[e.g.][]{Bena2006}.
 
-The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, non-ideal gases \citep{Chapman1970,Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
+The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, granular gases \citep{Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
 
-A generalization of the model developed in this chapter to include collective motions should be capable of producing realistically-wide sediment transport fluctuations.
+A suitable generalization of the model developed in this chapter to include particle-particle interactions should be capable of producing realistically-wide sediment transport fluctuations.
 Such a development will be a challenging next step for research.
@@
 
-\subsection{Measuring the sediment flux}
+\subsection{Velocity correlations and bedload diffusion}
+\label{sec:prollems}
 
-Figure \ref{fig:fluxconvergence} indicates that the sediment flux described by equation \ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q = E\ell$ predicted by the Einstein theory.
-Because particles move for durations $1/k_D$ on average, the variance of velocity among moving particles can be written $\sigma_V^2 = 2D/k_D$.
-In terms of this quantity, the convergence condition is  $T\gg \sigma_V^2/(k_D V^2),$ now phrased in terms of easily measurable quantities -- the velocity of particles, the average time spent in motion, and the magnitude of velocity fluctuations.
+Fig.
+\ref{fig:flippyfig1} and eq.
+\ref{eq:diffo} indicate that bedload sediment particles described by the model spread apart with a rate which depends upon observation scale, transitioning through three different ranges consistent with the concept of \citet{Nikora2001a,Nikora2002}.
+Nikora et al. originally proposed that sediment diffusion would be super-diffusive at short (local) timescales, normal at intermediate timescales, and sub-diffusive at long (global) timescales.
+Here, the short range diffusion is normal, not super-diffusive, indicating a disprepancy between the Nikora et al. concept and the present model.
 
-This simple result provides the minimal time required to resolve mean sediment transport rates in terms of measurable properties of particle motions.
-For example, in the experiments of \citet{Martin2012}, this condition implies convergence times of approximately $T \gg 15 s$.
-With wide sediment velocity fluctuations and long particle movement periods, the required observation timescale could become large.
-In this case, mean sediment flux time series obtained with light tables and acoustic sensors \citep{Zimmermann2008,Tsakiris2014,Mendes2016} would require careful interpretation.
+This discrepancy could originate from the representation of velocity fluctuations in eq. \ref{eq:flippylangevin} as Gaussian noise with a vanishing correlation time.
+In actuality, particle velocities are temporally correlated, and this will modify the diffusion characteristics on timescales comparable to the correlation time.
+The simplest modification of eq.
+\ref{eq:flippylangevin} to include velocity correlations would be to replace the Gaussian white noise with an Ornstein-Uhlenbeck noise \citep[e.g.][]{Luczka2005,Hanggi2007}.
+A more challenging (but perhaps more realistic) modification would be to describe the evolution of position by a second-order equation, as in $\ddot{x} = F(x,t)\eta(t)$, where $F$ is an appropriate stochastic driving term switched on and off by the dichotomous noise $\eta(t)$ \citep[e.g.][]{Masoliver1993}.
+Either of these alternatives would introduce correlations to particle velocities and modify the resulting short-range diffusion characteristics.
 
-\subsection{Limitations and next steps}
+The short timescale diffusion may also be affected by the assumption that entrainment and deposition processes are instantaneous within the model.
+\citet{Campagnol2015} showed from experimental data that particle ``unsteadiness" during entrainment and deposition can give rise to diffusion exponents even larger than that for ballistic motion.
+Likely, the representation of entrainment and deposition with an instantaneous alternation obscures the short-timescale physics of these processes \citep[e.g.][]{Valyrakis2010,Celik2014}.
+The simplest modification to include the timescales of entrainment and deposition would introduce a uniform accleration (deceleration) during entrainment (deposition), in effect ``tilting" the sharp motion-rest alternation represented by dichotomous noise in this chapter's model.
+This acceleration could be written $V/\tau_E$, where $\tau_E$ is an ``exchange time", as originally introduced by \citep[cf.][]{Einstein1950}.
+The relatively lower velocity of particles during the unsteady acceleration phase induced by entrainment and deposition can be expected modify the predicted short-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
+
+\subsection{Scale dependent fluxes and channel evolution}
+
+Fig. \ref{fig:fluxconvergence} indicates that the sediment flux described by eq. \ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q = E\ell$ predicted by the Einstein theory.
+Because particles move for durations $1/k_D$ on average, the typical variance of velocity among moving particles can be written $\sigma_V^2 = 2D/k_D$.
+In terms of this quantity, the convergence condition for the flux is  $T\gg \sigma_V^2/(k_D V^2),$ now phrased in terms of easily measurable quantities -- the velocity of particles, the average time spent in motion, and the magnitude of velocity fluctuations.
+
+The Exner equation of sec.
+\ref{landscape} used to describe channel evolution is derived by evaluating mass balance within a control volume \citep[e.g.][]{Coleman2009}.
+When the sediment flux changes values for observation scales $t < T$, as in this chapter, how the channel evolution predicted by the Exner equation is affected remains an open question.
+Future studies should include fluctuating and scale-dependent sediment transport rates into the Exner equation to evaluate their implications.
+
+Presumably, the inclusion of a scale-dependent stochastic transport rate generates a stochastic bed elevation field $h(x)$, generating a probability distribution of field configurations $P(h(x),t)$ which depends on the observation scale.
+The work of \citet{Jerolmack2005} evaluates a stochastic sediment transport rate within the Exner equation, although they did not investigate the probability distribution of bed configurations or its scale dependence.
+Mathematical tools to investigate such a stochastic Exner equation can be found in works on surface growth processes and polymer physics, where the evolution of random fields (like $h(x)$) is common topic of inquiry \citep[e.g.][]{Kawakatsu2001,Barabasi1995, Kardar2007}.
+
+\subsection{Collective motions}
 
@@
 
-Here, I have demonstrated how to describe the sediment flux probability distribution from a dynamical perspective, rather than a kinematic one, using stochastic dynamical equations to describe the downstream trajectories of individual grains (eq.
+Here, I have demonstrated how to describe the sediment flux probability distribution from a dynamical perspective, rather than a kinematic one, using stochastic dynamical equations to describe the trajectories of individual grains (eq.
 \ref{eq:langevin}.
@@
 The first possibility is to introduce correlations between particle movements \citep{Ancey2008,Heyman2013,Ancey2014}, such as might arise when turbulent flow fluctuations entrain many particles at once \citep{Cameron2020} or drive waves of high velocity particles downstream simultaneously \citep{Nino1996}.
-In a theory such as the one I presented here, this first mechanism for wide fluctuations could be described by temporally unsteady velocities, entrainment or deposition rates, or diffusion coefficients in the Langevin model \ref{eq:langevin}.
+In a model like the one I presented here, this first mechanism for wide fluctuations could be described by temporally unsteady velocities, entrainment or deposition rates, or diffusion coefficients in the Langevin model (eq.
+\ref{eq:langevin}).
 The second possibilty is to introduce interactions between moving particles, such as might result from collisions \citep{Lee2018} or wake vortices \citep{Schmeeckle2014} inducing particle entrainment, leading to collective waves of motion \citep{Ancey2014}.
-Including such effects would require a many-body generalization of the Langevin and master equations \ref{eq:langevin} and \ref{eq:master}, where the probability distribution of position is promoted to a joint form $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$, which is puzzling, since the number of particles $N(t)$ in the distribution fluctuates.
-The model of \citet{Ancey2014} hints at this possibility, although its starting point is not a dynamical equation as in \ref{eq:master}.
+Including such effects would require a many-body generalization of the Langevin and master equation \ref{eq:langevin} and \ref{eq:master}, where the probability distribution of position is promoted to a joint form $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$, which is puzzling, since the number of particles $N(t)$ in the distribution fluctuates.
+The model of \citet{Ancey2014} hints at this possibility, although its starting point is not a dynamical equation as in eq .\ref{eq:master}.
 Some hints as to how to generalize our work to include interactions between particles to describe wide-tailed sediment fluxes can be gleaned from literature on reaction-diffusion processes \citep{Hernandez2004,Cardy2006} and interacting gases \citep{Chapman1970,Brilliantov2004}.

### 2021-09-23 revision/ch3bedEle.tex
uh

@@
 Although the displacements of individual grains are a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}.
-In response, researchers have developed a stochastic concepts of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modelled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
+In response, researchers have developed a stochastic concepts of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
 
@@
 In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
-This model extends the birth-death model approach reviewed in section \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
+This model extends the birth-death model approach reviewed in sec. \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
 This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried under the bed surface.
@@
 
-Below, section \ref{sec:elemodel} introduces the stochastic model and section \ref{sec:elesolution} solves it with a mixture of numerical and analytical techniques.
+Below, sec.
+\ref{sec:elemodel} introduces the stochastic model and sec.
+\ref{sec:elesolution} solves it with a mixture of numerical and analytical techniques.
 Model solutions illustrate several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates with bed elevation change.
-Section \ref{sec:eleresults} present these features, while sections \ref{sec:elediscussion} and \ref{sec:eleconclusion} conclude with the implications of these features and the suggestions they raise for future research.
+Sec.
+\ref{sec:eleresults} present these features, while secs.
+\ref{sec:elediscussion} and \ref{sec:eleconclusion} conclude with the implications of these features and the suggestions they raise for future research.
 
-\section{Stochastic model of bedload transport and bed elevations}
+\section{Stochastic model of bedload transport and bed elrefevations}
 \label{sec:elemodel}
 
-The model in this chapter is formulated by considering a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in figure \ref{fig:eledefinition}.
+The model in this chapter is formulated by considering a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in fig.
+\ref{fig:eledefinition}.
 The value $m$ is defined relative to the mean number of grains within the control volume, so that it can be either positive or negative.
@@
 These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
-The four events are depicted as arrows in figure \ref{fig:eledefinition}.
+The four events are depicted as arrows in fig. \ref{fig:eledefinition}.
 As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
 \begin{figure}[!htbp]
-	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/definition-combo.pdf}
+	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/improveddef.png}
 	\caption{Definition sketch of a control volume containing $n$ moving grains and $m$ resting grains.
@@
 The bed is displayed in a degraded state, where $m<0$.
-The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using (\ref{eq:ele}).}
+The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using eq.
+\ref{eq:ele}.}
 	\label{fig:eledefinition}
@@
 \end{align}
-In equations (\ref{eq:rate2}) and (\ref{eq:rate3}), $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
+In eqs. \ref{eq:rate2} and \ref{eq:rate3}, $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
 $\kappa$ is referred to as a coupling constant since it controls the strength of this feedback. I will later demonstrate the relationship
@@
 The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
-The solutions of (\ref{eq:elemaster}) should be expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
+The solutions of eq.
+\ref{eq:elemaster} should be expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
 Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
@@
 
-Unfortunately, equation (\ref{eq:elemaster}) does not admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
+Unfortunately, eq. \ref{eq:elemaster} does not admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
 The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
@@
 In conditions when the coupling constant $\kappa$ between the entrainment/deposition rates and the bed elevation is weak, it is also possible to solve the master equation approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
-The Gillespie simulation algorithm is described below in \ref{sec:elenumerical}, while these analytical approximations are described in \ref{sec:eleanalytical}.
+The Gillespie simulation algorithm is described below in sec. \ref{sec:elenumerical}, while these analytical approximations are described in sec. \ref{sec:eleanalytical}.
 
@@
 As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
-then draw a second random value to choose the type of transition that occurs using relative probabilities formed from equations (\ref{eq:rate1}-\ref{eq:rate4}).
+then draw a second random value to choose the type of transition that occurs using relative probabilities formed from eqs.
+\ref{eq:rate1}-\ref{eq:rate4}.
 The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
@@
 Units are $s^{-1}$ (probability/time).
-Bed elevation changes modulate these rates in accord with (\ref{eq:rate1}-\ref{eq:rate4}).}\label{tab:anceyparams}
+Bed elevation changes modulate these rates in accord with eqs.
+\ref{eq:rate1}-\ref{eq:rate4}.}\label{tab:anceyparams}
 	\begin{tabular}{cccccc} \\ 
@@
 
-The $n$ and $m$ dynamics in equation (\ref{eq:elemaster}) can be approximately decoupled using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
+The $n$ and $m$ dynamics in eq.
+\ref{eq:elemaster} can be approximately decoupled using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
 These inequalites mean many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
-Concentrating on steady state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to equation (\ref{eq:elemaster}), with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$ \citep[e.g.][]{Haken1978}.
+Concentrating on steady state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to eq.
+\ref{eq:elemaster}, with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$ \citep[e.g.][]{Haken1978}.
 This provides the steady state equation
@@
 \end{multline}
-Unforunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in equation (\ref{eq:decomp}).
+Unforunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in eq. \ref{eq:decomp}.
 
 The simplest approximation to these equations holds that $\kappa$ is so small that the dynamics of $n$ are totally independent of $m$: $A(n|m) = A(n)$.
-Taking this limit in equation (\ref{eq:decomp}), summing over $m$, and using $\langle m \rangle = 0$ reproduces the \citet{Ancey2008} particle activity model.
-As shown in section \ref{sec:birthdeath}, this has solution
+Taking this limit in eq.
+\ref{eq:decomp}, summing over $m$, and using $\langle m \rangle = 0$ reproduces the \citet{Ancey2008} particle activity model.
+As shown in sec. \ref{sec:birthdeath}, this has solution
 \begin{equation} A(n) = \frac{\Gamma(r+n)}{\Gamma(r)n!}p^r(1-p)^n.\label{eq:ancey}\end{equation}
 which is a negative binomial distribution for the particle activity with parameters $r=(\nu+\lambda)/\mu$ and $p=1-\mu/(\sigma+\gamma).$
-This result implies $\langle n | m \rangle = \langle n \rangle$, so with the definitions of $E$ and $D$ and the equilibrium condition $E=D$, equation (\ref{eq:approxele}) provides
+This result implies $\langle n | m \rangle = \langle n \rangle$, so with the definitions of $E$ and $D$ and the equilibrium condition $E=D$, eq. \ref{eq:approxele} provides
 \begin{equation}0 \approx [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1)-2M(m). \label{eq:ou} \end{equation}
 This mean field equation matches the discrete Ornstein-Uhlenbeck model of bed elevation changes developed by \citet{Martin2014}.
-The independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model equation \ref{eq:elemaster} in a mean field approximation when $\kappa$ is insignificant.
+The independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model eq.
+\ref{eq:elemaster} in a mean field approximation when $\kappa$ is insignificant.
 
-In the supplementary information I show that the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within equation (\ref{eq:ou}) provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
-As I will demonstrate in section \ref{sec:eleresults}, and as I have already suggested with equation (\ref{eq:active}), this is a poor approximation to the bed elevation variance.
+In the appendix ch.
+\ref{ch:gill} I show that the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within eq.
+\ref{eq:ou} provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
+As I will demonstrate in sec.
+\ref{sec:eleresults}, and as I have already suggested with eq.
+(\ref{eq:active}), this is a poor approximation to the bed elevation variance.
 Nevertheless, this approximation does capture the Gaussian shape of the bed elevation distribution.
@@
 
-A more careful approximate solution to equation (\ref{eq:approxele}) can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into equation (\ref{eq:approxele}) in order to close the equation for $m$ without solving equation (\ref{eq:decomp}).
-From numerical simulations we determine that
+A more careful approximate solution to eq.
+\ref{eq:approxele} can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into eq.
+\ref{eq:approxele} in order to close the equation for $m$ without solving eq.
+\ref{eq:decomp}.
+By comparison with the numerical simulations we determine that
 \begin{equation}
@@
 captures the general features of the conditional mean particle activity.
-As I show in the supplementary information, introducing this closure equation to (\ref{eq:approxele}), making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
+As I show in the supplementary information, introducing this closure relation to eq. \ref{eq:approxele}, making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
 \begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
-representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in equation (\ref{eq:active}).
+representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in eq.
+\ref{eq:active}.
 $M_0$ is a normalization constant. 
-This closure equation approach shows good correspondence with numerical solutions of equation (\ref{eq:elemaster}), at least for the flow parameters in table \ref{tab:anceyparams}.
+This closure equation approach shows good correspondence with numerical solutions of eq. \ref{eq:elemaster}, at least for the flow parameters in tab. \ref{tab:anceyparams}.
 
@@
 From the initial conditions, all simulations show a rapid attainment of steady-state stochastic dynamics of $n$ and $m$ which support a time-independent joint distribution $P(n,m)$.
-The bottom panel of figure \ref{fig:eledefinition} shows an elevation time-series.
-In order to describe the implications of coupling bedload transport to bed elevation changes, the numerical and analytical results for the probability distributions of bedload transport and bed elevations are presented in section \ref{sec:elepdf} and the statistical moments of these quantities in section \ref{sec:elemom}.
-The effects of collective entrainment on bed elevation changes are studied in section \ref{sec:elecolent}, and the resting times of sediment undergoing burial are evaluated in section \ref{sec:elertcdf}.
+The bottom panel of fig.
+\ref{fig:eledefinition} shows an elevation time-series.
+In order to describe the implications of coupling bedload transport to bed elevation changes, the numerical and analytical results for the probability distributions of bedload transport and bed elevations are presented in sec.
+\ref{sec:elepdf} and the statistical moments of these quantities in sec.
+\ref{sec:elemom}.
+The effects of collective entrainment on bed elevation changes are studied in sec.
+\ref{sec:elecolent}, and the resting times of sediment undergoing burial are evaluated in sec.
+\ref{sec:elertcdf}.
 
@@
 From this joint distribution the marginal distributions $P(n)$ and $P(m)$ are obtained by summing over $m$ and $n$ respectively.
-A representative subset of these marginal distributions is displayed in figure \ref{fig:pdfs} alongside the approximate results of equations (\ref{eq:ancey}) and (\ref{eq:ou2}).
+A representative subset of these marginal distributions is displayed in fig. \ref{fig:pdfs} alongside the approximate results of eqs. \ref{eq:ancey} and \ref{eq:ou2}.
 \begin{figure}[!htbp]
@@
 	\caption{Panel (a) presents the probability distribution of particle activity $n$ and panel (b) presents the probability distribution of the relative number of particles $m$ for a representative subset of simulations.
-These distributions represent different flows from table \ref{tab:anceyparams}, distinguished by color, and different values of the active layer depth $l$ (equivalently the coupling constant $\kappa$), distinguished by the marker style.
-The mean field theories (mft) of equations (\ref{eq:ancey}) and (\ref{eq:ou2}) are displayed as solid black lines.}
+These distributions represent different flows from tab.
+\ref{tab:anceyparams}, distinguished by color, and different values of the active layer depth $l$ (equivalently the coupling constant $\kappa$), distinguished by the marker style.
+The mean field theories (mft) of eqs.
+\ref{eq:ancey} and \ref{eq:ou2} are displayed as solid black lines.}
 	\label{fig:pdfs}
 \end{figure}
-The mean field equation (\ref{eq:ancey}) for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match equation (\ref{eq:ou2}), having Gaussian profiles consistent with the assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.
+The mean field eq.
+\ref{eq:ancey} for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match eq.
+\ref{eq:ou2}, having Gaussian profiles consistent with the assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.
 
@@
 	\includegraphics[width=0.5\textwidth,keepaspectratio]{./figures/ch3/variance.pdf}
-	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as posited in equation (\ref{eq:active}): $\sigma_m^2 \approx (l/z_1)^2$. }
+	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as posited in eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$. }
 	\label{fig:var}
 \end{wrapfigure}
-Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in equation (\ref{eq:ou2}); this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
-The close correspondence between the mean field approximation and the numerical simulations in figure (\ref{fig:pdfs}a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
-They appear to be identical within numerical uncertainty.
+Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in eq.
+\ref{eq:ou2}; this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
+The close correspondence between the mean field approximation and the numerical simulations in fig.
+\ref{fig:pdfs}a suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
+They appear identical within numerical uncertainty.
 
 The coupling between bed elevation changes and the erosion and deposition rates develop a strong dependence of the particle activity on $m$.
-Figure (\ref{fig:condmoms}) displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
+Fig.
+\ref{fig:condmoms} displays the mean shift $[\langle n |m \rangle - \langle n \rangle]/\langle n \rangle $ and the variance shift  $[\text{var}(n|m) - \text{var}(n)]/\text{var}(n)$ of the particle activity due to departures of the bed elevation from its mean position.
 \begin{figure}[!htbp]
@@
 This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$.
-The closure equation (\ref{eq:closure}) is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
+The closure equation \ref{eq:closure} is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
 These results indicate that bedload statistics measurements on short time scales could be severely biased by departures from the mean bed elevation.}
@@
 \end{figure}
-Figure (\ref{fig:condmoms}a) demonstrates that the flow conditions in table \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
-The closure model (\ref{eq:closure}) used to derive the approximate bed elevation distribution (\ref{eq:ou2}) is plotted behind the conditional mean profiles in figure (\ref{fig:condmoms}a), where it appears as a crude approximation since it does not capture the asymmetry.
-Nevertheless, figure \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
-For the parameters of the \citet{Ancey2008} experiments, figure (\ref{fig:condmoms}b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
-The summary is that bed elevation changes regulate the particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
+Figure \ref{fig:condmoms}a demonstrates that the flow conditions in tab.
+\ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
+The closure model (eq.
+\ref{eq:closure}) used to derive the approximate bed elevation distribution (eq.
+\ref{eq:ou2}) is plotted behind the conditional mean profiles in fig.
+\ref{fig:condmoms}a, where it appears as a crude approximation since it does not capture the asymmetry.
+Nevertheless, fig. \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
+For the parameters of the \citet{Ancey2008} experiments, fig.
+\ref{fig:condmoms}b displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
+The summary is that bed elevation changes regulate the first and second particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
 
@@
 	\caption{The shift of the mean particle activity in panel (a) and its fluctuations in panel (b) with departures of the bed elevation from its mean.
-All simulations are at flow condition (g) from table \ref{tab:anceyparams} except $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment.
-Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (\ref{eq:ancey}) at large departures from the mean bed elevation.
+All simulations are at flow condition (g) from tab.
+\ref{tab:anceyparams} except $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment.
+Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (eq.
+\ref{eq:ancey}) at large departures from the mean bed elevation.
 Panel (b) shows particle activity fluctuations suppressed by 90\% when $z\approx -3l$ and collective entrainment is the dominant process.
@@
 Figure \ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed.
-When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect identified in section \ref{sec:elemom}.
+When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect identified in sec.
+\ref{sec:elemom}.
 
@@
 Panel (a) shows differing flow conditions at a fixed $l$ value, while panel (c) shows fixed flow conditions at differing $l$.
-When scaled by $T_0$ (\ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d).
+When scaled by $T_0$ (eq.
+\ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d).
 In panels (b) and (d), the black dotted lines indicate a power law decay of the collapsed tails having parameter $\alpha\approx1.18$ .}
@@
 Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
-Using the marginal probability distribution of bed elevations $P(m)$ (figure \ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}:
+Using the marginal probability distribution of bed elevations $P(m)$ (fig.
+\ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}:
 \begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
@@
 We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
-Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (\ref{eq:ele}), bed elevations change with a characteristic speed $v = z_1 E$.
-Since the range of elevation deviations is $l$ (figure \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
+Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (eq.
+\ref{eq:ele}), bed elevations change with a characteristic speed $v = z_1 E$.
+Since the range of elevation deviations is $l$ (fig. \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
 \begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
-When scaling the resting time by this $T_0$, we obtain the collapse shown in figure \ref{fig:cdfs}.
+When scaling the resting time by this $T_0$, we obtain the collapse shown in fig. \ref{fig:cdfs}.
 Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
@@
 As the model in this chapter builds on earlier works describing particle activity and bed elevation changes independently, it also reduces to these works in simplified limits when the coupling between the particle activity and the local bed elevation vanishes.
-With the mean field approach in section \ref{sec:eleanalytical}, the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity were derived as simplified limits of the coupled model developed in this chapter.
+With the mean field approach in sec.
+\ref{sec:eleanalytical}, the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity were derived as simplified limits of the coupled model developed in this chapter.
 While the mean field description of bed elevations over-predicts the bed elevation variance by approximately a factor of two, it does capture the Gaussian shape of the bed elevation distribution, and its conclusions on the tail characteristics of resting time distributions for sediment undergoing burial are identical to the model in this chapter within the numerical uncertainty.
@@
 
-Like the theory developed by \citet{Martin2014}, the model \ref{eq:elemaster} produces heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
+Like the theory developed by \citet{Martin2014}, the model eq.
+\ref{eq:elemaster} produces heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
 This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016a}. 
-The resting time distributions derived in section \ref{sec:elertcdf} show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of the results.
+The resting time distributions derived in sec.
+\ref{sec:elertcdf} show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of the results.
 
@@
 Some relatively recent particle tracking experiments have demonstrated joint resolution of bed elevations and bedload transport \citep{Martin2014,Heyman2016}.
-A suitably designed experiment using this methodology could test the prediction that bed elevations regulate particle activity statistics, as essentially represented in figures \ref{fig:condmoms} and \ref{fig:colent}.
+A suitably designed experiment using this methodology could test the prediction that bed elevations regulate particle activity statistics, as essentially represented in figs.
+\ref{fig:condmoms} and \ref{fig:colent}.
 
 Many other statistical characteristics of bedload transport have been left for future studies. 
-For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic which was previously referenced in section \ref{sec:renewal}.
-Whether bed elevation distributions are also scale-dependent like the sediment flux distribution of chapter \ref{ch:flux} is an interesting question for future research.
+For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic which was previously referenced in sec.
+\ref{sec:renewal}.
+Whether bed elevation distributions are also scale-dependent like the sediment flux distribution of ch. \ref{ch:flux} is an interesting question for future research.
 

### 2021-09-23 revision/ch4downDiff.tex
uh

@@
 To incorporate motion intervals and sediment burial, I utilize the multi-state CTRW formalism developed by \citet{Weiss1976, Weiss1994} that extends the CTRW of \citet{Montroll1965}.
-Below, I develop and solve the model in section \ref{sec:model}.
-Then, I discuss the predictions of this model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in sections \ref{sec:discussion} and \ref{sec:conclusion}.
+Below, I develop and solve the model in sec.
+\ref{sec:model}.
+Then, I discuss the predictions of this model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in secs.
+\ref{sec:discussion} and \ref{sec:conclusion}.
 
@@
 The end results will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
-Grains in motion are considered to have a characteristic velocity $V$ \citep{Lisle1998,Lajeunesse2017}, and burial is modelled as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried.
+Grains in motion are considered to have a characteristic velocity $V$ \citep{Lisle1998,Lajeunesse2017}, and burial is modeled as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried.
 This means $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
@@
 \be p(x,t) = \sum_{k=0}^2 p_k(x,t) \ee
-This joint density is completely determined from the solutions of equations (\ref{eq:g1}-\ref{eq:g2}) given specification of the distributions $g_{ij}$ and $G_i$.
+This joint density is completely determined from the solutions of eqs. \ref{eq:g1}-\ref{eq:g2} given specification of the distributions $g_{ij}$ and $G_i$.
 
@@
 
-These distributions can be constructed from the assumptions described in section \ref{sec:assumptions}.
+These distributions can be constructed from the assumptions described in sec. \ref{sec:assumptions}.
 Since particles resting on the bed surface become buried in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, these assumptions obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$.
@@
 
-Equations (\ref{eq:g1}-\ref{eq:g2}) with these $g_{ij}$ and $G_i$ are solved using Laplace transforms in space and time $(x,t \rightarrow 
+Equations \ref{eq:g1}-\ref{eq:g2} with these $g_{ij}$ and $G_i$ are solved using Laplace transforms in space and time $(x,t \rightarrow 
 \eta,s).$ This method, similar to \citet{Weeks1998}, unravels the convolution structure of these equations, eventually producing
@@
 This result generalizes the earlier results of \citet{Lisle1998} and \citet{Einstein1937} to include sediment burial.
-This equation uses the shorthand notations $\xi = k_2 x/v$, $\tau = k_1(t-x/v)$, and $\Omega = (\kappa+k_1)/k_1$ \citep[c.f.][]{Lisle1998}.
+This equation uses the shorthand notations $\xi = k_2 x/v$, $\tau = k_1(t-x/v)$, and $\Omega = (\kappa+k_1)/k_1$ \citep[cf.][]{Lisle1998}.
 The $\mathcal{I}_\nu$ are modified Bessel functions of the first kind and the $\mathcal{Q}_\mu$ are generalized Marcum Q-functions defined by $\mathcal{Q}_\mu(x,y) = \int_0^y e^{-z-x}(z/x)^{(\mu-1)/2}\mathcal{I}_{\mu-1}(2\sqrt{xz})dz $ and originally devised for radar detection theory \citep{Marcum1960,Temme1996}.
@@
 Grains are considered initially at rest ($\theta_1=1$, $\theta_2=0$).
-The solid lines are the analytical distribution in equation (\ref{eq:pdf}), while the points are numerically simulated, showing the correctness of our derivations.
+The solid lines are the analytical distribution in eq.
+\ref{eq:pdf}), while the points are numerically simulated, showing the correctness of our derivations.
 Colors pertain to different times.
@@
 
-Figure \ref{fig:diffpdfs} depicts the distribution (\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
-When grains do not become buried, as in panel (a) of figure \ref{fig:diffpdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion and satisfying the central limit theorem.
-When grains become buried, as in panel (b) of figure \ref{fig:diffpdfs},  the Q-function terms prevent the distribution from approaching a Gaussian at large timescales, exemplifying anomalous diffusion \citep{Weeks1998} and violating the central limit theorem \citep{Metzler2000,Schumer2009}.
+Figure \ref{fig:diffpdfs} depicts the distribution (eq.
+\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
+When grains do not become buried, as in panel (a) of fig.
+\ref{fig:diffpdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion and satisfying the central limit theorem.
+When grains become buried, as in panel (b) of fig.
+\ref{fig:diffpdfs},  the Q-function terms prevent the distribution from approaching a Gaussian at large timescales, exemplifying anomalous diffusion \citep{Weeks1998} and violating the central limit theorem \citep{Metzler2000,Schumer2009}.
 
@@
 
-To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution (\ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}.
+To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution (eq.
+\ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}.
 These moments produce the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$.
@@
 \be \sigma_x^2(t) = A(t)e^{(b-a)t} + B(t)e^{-(a+b)t} + C(t). \label{eq:var}\ee
-In these equations, $a = (\kappa + k_1+k_2)/2$ and $b = \sqrt{a^2-\kappa k_2}$ are effective rates having dimensions of inverse time, while the $A$, $B$, and $C$ factors are provided in table \ref{table:params}.
+In these equations, $a = (\kappa + k_1+k_2)/2$ and $b = \sqrt{a^2-\kappa k_2}$ are effective rates having dimensions of inverse time, while the $A$, $B$, and $C$ factors are provided in tab.
+\ref{table:params}.
 
@@
 	%\begin{wraptable}{r}{width=5.5cm}
-	\caption{Abbreviations used in the expressions of the mean (\ref{eq:mean}), second moment (\ref{eq:second}) and variance (\ref{eq:var}) of bedload tracers.}
+	\caption{Abbreviations used in the expressions of the mean (eq. \ref{eq:mean}), second moment (eq. \ref{eq:second}) and variance (eq. \ref{eq:var}) of bedload tracers.}
 	\label{table:params}
@@
 \end{table}
-The positional variance (\ref{eq:var}) is plotted in figure \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
+The positional variance (eq. \ref{eq:var}) is plotted in fig. \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
 The notation ``$\gg$'' is interpreted in this context to mean ``of at least an order of magnitude greater''.
 These conditions are most relevant to tracers in gravel-bed rivers, since they represent that grains are initially at rest \citep{Hassan1991,Wu2019}, motions are typically much shorter than rests \citep{Einstein1937,Hubbell1964}, and burial requires a much longer time than typical rests  \citep{Ferguson2002,Hassan1994,Haschenburger2013}.
-Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
+Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (eq.
+\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
 Following \citet{Hassan2017} I suggest to call the fourth range "geomorphic", since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/diffusion.png}
-	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (\ref{eq:var}) with mean motion time $1.5$ s, resting time $30.0$ s, and movement velocity $0.1$ m/s -- values comparable to laboratory experiments transporting small ($5$ mm) gravels \citep{Lajeunesse2010,Martin2012}.
+	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (eq.
+\ref{eq:var}) with mean motion time $1.5$ s, resting time $30.0$ s, and movement velocity $0.1$ m/s -- values comparable to laboratory experiments transporting small ($5$ mm) gravels \citep{Lajeunesse2010,Martin2012}.
 The burial timescale is $7200.0$s (two hours), and grains start from rest ($\theta_1=1$).
-The solid line is equation (\ref{eq:var}), and the points are numerically simulated.
+The solid line is eq.
+\ref{eq:var}, and the points are numerically simulated.
 Panel (b) demonstrates four distinct scaling ranges of $\sigma_x^2$: local, intermediate, global, and geomorphic.
@@
 At local timescales, grains usually either rest or move; at intermediate timescales, they transition between rest and motion; at global timescales, they transition between rest, motion, and burial; and at geomorphic timescales, all grains bury.
-Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as discussed in section \ref{sec:discussion}.
+Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as discussed in sec.
+\ref{sec:discussion}.
 	}
@@
 
-Two limiting cases of equation (\ref{eq:var}) provide the scaling exponents $\gamma$ of the diffusion $\sigma_x^2 \propto t^\gamma$ in each range.
+Two limiting cases of eq.
+\ref{eq:var} provide the scaling exponents $\gamma$ of the diffusion $\sigma_x^2 \propto t^\gamma$ in each range.
 Limit (1) represents times so short that a negligible amount of sediment burial has occurred, $t\ll 1/\kappa$, while limit (2) represents times so long motion intervals appear as instantaneous steps of mean length $l=v/k_2$, $1/k_2 \rightarrow 0$ while $v/k_2 = \text{constant}$.
@@
 The extreme $k_1/\kappa \ll 1$ produces $\gamma=1$ in the global range, while the opposite extreme $k_1/\kappa \rightarrow \infty$ produces $\gamma=3$.
-To summarize, when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, equation (\ref{eq:var}) implies:
+To summarize, when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, eq. \ref{eq:var} implies:
 \begin{enumerate}
@@
 In actuality, burial is a temporary condition, because bed scour can exhume buried sediment back into transport \citep{Wu2019a}, probably after heavy-tailed intervals \citep{Voepel2013,Martin2014}.
-A generalization of the model in this chapter to include heavy-tailed timescales between burial and exhumation might develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in figure \ref{fig:diffvar}.
-If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested in chapter \ref{ch:ch3}, other equilibrium transport models \citep{Voepel2013, Martin2014}, and laboratory experiments \citep{Martin2014,Martin2012}, the geomorphic range is expected to be super-diffusive \citep{Weeks1998}.
+A generalization of the model in this chapter to include heavy-tailed timescales between burial and exhumation might develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in fig.
+\ref{fig:diffvar}.
+If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested in ch.
+\ref{ch:ch3}, other equilibrium transport models \citep{Voepel2013, Martin2014}, and laboratory experiments \citep{Martin2014,Martin2012}, the geomorphic range is expected to be super-diffusive \citep{Weeks1998}.
 However, if they decay slower than $T^{-1/2}$, as implicitly suggested by the data of \citet{Olinde2015}, the geomorphic range is expected to be genuinely sub-diffusive \citep{Weeks1998}, leaving Nikora et al.
@@
 
-The analytical solution of bedload diffusion in equation (\ref{eq:var}) reduces exactly to the analytical solutions of the \citet{Lisle1998} and \citet{Lajeunesse2017} models in the limit without burial ($\kappa \rightarrow 0$), the \citet{Wu2019} model in the limit of instantaneous steps ($k_2 \rightarrow \infty$ and $l = v/k_2$), and the original \citet{Einstein1937} model in the limit of instantaneous steps without burial.
+The analytical solution of bedload diffusion in eq.
+\ref{eq:var} reduces exactly to the analytical solutions of the \citet{Lisle1998} and \citet{Lajeunesse2017} models in the limit without burial ($\kappa \rightarrow 0$), the \citet{Wu2019} model in the limit of instantaneous steps ($k_2 \rightarrow \infty$ and $l = v/k_2$), and the original \citet{Einstein1937} model in the limit of instantaneous steps without burial.
 These reductions demonstrate that the majority of recent bedload diffusion models, whether developed from Exner-type equations \citep{Wu2019,Pelosi2014,Pelosi2016} or advection-diffusion equations \citep{Lisle1998,Lajeunesse2017}, can be viewed equivalently as continuous-time random walks applied to individual bedload trajectories.
@@
 At base level, this work demonstrates that (1) durations of sediment motions, (2) durations of sediment rest, and (3) the sediment burial process are sufficient to develop three diffusion ranges that terminate when all tracers become buried.
-This work provides a complementary perspective on the three range diffusion exhibited in chapter \ref{ch:flux}.
+This work provides a complementary perspective on the three range diffusion exhibited in ch. \ref{ch:flux}.
 A next step is to incorporate exhumation to better understand the geomorphic range.

### 2021-09-23 revision/ch5langevin.tex
uh

@@
 \ref{sec:birthdeath}).
-This formulation requires the instantaneous velocity distributions of sediment particles \citep{Lajeunesse2010}, although understanding of these distributions remains limited.
-There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in section \ref{sec:langevin} of the introduction have described some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe the full range of experimental observations \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
+This formulation requires the instantaneous velocity distributions of sediment particles \citep[e.g.][]{Ancey2020a}, yet unfortunately, understanding of these distributions remains limited.
+There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in sec.
+\ref{sec:langevin} of the introduction describe some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe all experimental observations reported to date \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
 In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.
 
-High-speed video experiments have measured different streamwise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
+High-speed video experiments have measured different stream-wise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
 One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
-These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were viscous.
+These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were turbulent and supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were laminar and (likely) subcritical.
 A second set of studies show Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
-In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and super-critical.
+In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and supercritical.
 Other experiments display velocity distributions that are intermediate between exponential and Gaussian extremes that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
-The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm.
-Flows were again turbulent and subcritical.
-From this experimental record, the velocity distribution shape apparently does not consistently relate to super or sub-critical flow conditions, whether sediment grains are natural (sand, gravel) or synthetic (glass beads), or whether the flow is laminar or turbulent.
+The \citet{Houssais2012} experiments involved a binomial distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm in turbulent and subcritical flow.
+From this experimental record, the velocity distribution shape apparently does not consistently relate to the flow conditions (turbulent/laminar or super/subcritical) or sediment properties (natural sand/gravel or glass beads).
 However there is a loose trend within the particle size.
 Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}.
-This summary suggests that the shape of streamwise bed load velocity distribution could be controlled by the particle size.
+This summary suggests that bed load velocity distribution could be controlled by the particle size.
 
 Existing models of streamwise bed load velocities can be divided into computational and stochastic physics categories.
-Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modelling particles as spheres interacting through repulsive forces, and the flow using direct simulation of the Navier-Stokes equations or some related approximation (such as large eddy simulation or the St-Venant equations).
+Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modeling particles as spheres interacting through repulsive forces and numerically integrating the Navier-Stokes equations (or some approximation of them) to describe the flow forces on grains.
 When streamwise particle velocities have been analyzed in such simulations, they show exponential tails \citep{Gonzalez2017,Furbish2013} that agree with one subset of the experimental data.
 
-Several stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec.
+Stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec.
 \ref{sec:langevin}).
-\citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces into a simplified velocity-dependent force, while Gaussian white noise represents fluctuations.
+\citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces into a simplified velocity-dependent force, again modeling fluctuations with Gaussian white noise.
 Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
@@
 
-In this chapter I explore the possibility that the shape of the particle velocity distribution is controlled by momentum dissipation during particle-bed collisions.
-In the theory of granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
-Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970}.
-Collisions in these models involve episodic kicks to the particle momenta at random times, parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
-In contrast, current stochastic models of bedload transport appear highly simplified, treating collisions instead with static Coulomb or velocity-dependent drag.
+In this chapter I explore the possibility that the particle velocity distribution is controlled by momentum dissipation from particle-bed collisions by incorporating ideas from the physics of granular media.
+A major paradigm in granular dynamics is to describe particles as a continuum or "granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in ch.
+\ref{ch:Introduction} such a continuum assumption is unlikely to be satisfied for weak bedload transport conditions.
+Instead, particles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}.
+In the theory of rarefied granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
+Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970,Briliantov2004}.
+Collisions in these models involve episodic kicks to the particle momenta at random times, usually parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
+Such episodic forcing is distinct from the smooth friction terms included in the current stochastic models describing bedload particle velocities.
 
-Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport, including episodic particle-bed collisions.
+Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport which replaces the smooth friction terms of earlier models by episodic particle-bed collisions.
 The driving motivation is to test the hypothesis that particle-bed collision characteristics can explain the range of experimentally-observed streamwise bed load particle velocity distributions.
-A secondary one is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
-I develop the model and explain its key assumptions in section \ref{sec:langmodel}.
+A secondary motivation is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
+I develop the model and explain its key assumptions in sec.
+\ref{sec:langmodel}.
 Then I present the analytical solution of the model and other major results in \ref{sec:langresults}.
-Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in sections \ref{sec:langdiscussion} and \ref{sec:langconclusion}.
+Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in secs.
+\ref{sec:langdiscussion} and \ref{sec:langconclusion}.
 
@@
 Figure \ref{fig:fig1} indicates the configuration of the model in this chapter.
-Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady turbulent shear flow.
-The flow is just strong enough to drive grains into the rarefied transport typical in gravel-bed rivers: particles saltate downstream through a sequence of particle-bed collisions. 
-Moving particles collide often with stationary particles, but rarely with other moving particles.
+Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady flow.
+The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Jackson1982, Warburton1992,Hassan2001}.
+Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles, but rarely with other moving particles \citep[cf.][]{Williams2020}.
 
-Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces.
+Particles respond to turbulent flow forces and episodic particle-bed collision forces.
 In contrast to the computational physics approach, I do not aim to characterize the exact time series of the forces on an individual particle.
@@
 The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
+Here, the description of bedload particle velocity fluctuations will be more detailed than ch.
+\ref{ch:flux}, but entrainment and deposition processes will be neglected, meaning the model of this chapter is only applicable to moving particles.
 I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.
 
-\subsection{Episodic collisions}
-Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or rotational momentum, partly by deforming particles and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid \citep{Joseph2001,Schmeeckle2001}.
-The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}.
-Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in figure \ref{fig:fig1}.
+\subsection{Episodic collision forces}
+Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid in the interstices between approaching particles \citep{Joseph2001,Schmeeckle2001}.
+The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of resitution may be less relevant than other factors such as collision geometry and bed deformation \citep{}.
+Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in fig.
+\ref{fig:fig1}.
 This elasticity characterizes the fraction of downstream momentum lost in a collision.
@@
 Since the elasticity combines effects of particle shape and collision geometry and should vary from one collision to the next, the elasticity $\varepsilon$ is interpreted as a random variable, characterized by a statistical distribution $\rho(\varepsilon)$.
-Some granular gas models have also included random elasticity \citep[e.g.][]{Serero2015}.
+Some granular gas models have also included random elasticity \citep[e.g.][]{Serero2015}, but this topic has not been deeply explored in the literature.
 
-Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (sec.
+Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (sec.
 \ref{sec:einflux}):
@@
 Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
-This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momentum.
+The time intervals $\Delta \tau_k = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau).$
+This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momenta.
 This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions.
-These conditions are always satisfied for the idealized saltation trajectories depicted in figure \ref{fig:fig1}.
+These conditions should be satisfied for idealized saltation trajectories as depicted in fig.
+\ref{fig:fig1}.
 
 \subsection{Turbulent fluid forces}
-Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Rey_p = d V/\lambda$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\lambda$.
-These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Rey_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
-At realistic $\Rey_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Rey_p$ formulas \citep{Schmeeckle2007,Clift1978}.
-A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Rey_p$ can be written $F_D = \frac{\pi}{8}
-\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
-In this study, I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.
 
-Drag forces have been argued to fluctuate rapidly compared to the inertial response times of submerged grains \citep{Fan2014}, and the magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
-These ideas allow two major simplifications of the above drag force.
-First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled as a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is characterized by a diffusivity $D$ \citep{Fan2014,Ancey2014}.
-Defining $\bar{V}$ as a representative slip velocity to be specified more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, the fluid drag can be written
+Fluid forces on coarse particles are parameterized by the particle Reynolds number $\Rey_p = V d/\lambda$, involving the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\lambda$.
+These forces have been calculated from the Navier-Stokes equations for spherical particles at small (or vanishing) $\Rey_p$ and include drag, virtual mass, buoyancy, fluid pressure gradient, Basset history integral, Saffman lift, and Magnus lift terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
+At realistic $\Rey_p$, analytical results for the fluid forces on a particle remain inaccessible, so it is standard practice to introduce empirical corrections to the small $\Rey_p$ formulas \citep[e.g.][]{Clift1978,Schmeeckle2007}.
+
+Our understanding of the fluid forces remains incomplete.
+Existing formulas were derived in the absence of boundaries, so it is unclear how these formulas should be modified near the bed.
+Even ignoring this complication, the added mass and Basset forces are difficult to compute and therefore impractical to include in analytical models.
+These challenges require researchers to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michalades1997,Armenio2001}.
+As a result, the majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the primary force(s) driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.
+
+The downstream drag force with the $\Rey_p$ correction is usually written 
+\be F_D = \frac{\pi}{8} \rho_f d^2 C_D(\Rey_p) |V|V, \label{eq:draggy}\ee
+where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
+Experiments on stationary particles indicate that drag forces fluctuate rapidly and lie on wide Gaussian-like distributions \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
+The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities should be expected to vary due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
+The nonlinearity in eq. \ref{eq:draggy} will amplify these slip velocity variations, generating large and erratic drag force fluctuations.
+
+Here I assume with earlier computational \citep{Schmeeckle2014,Gonzalez2017,Elghannay2017} and analytical \citep[e.g.][]{Ancey2014,Fan2014} models of bedload transport that fluid drag is the primary force driving sediment particles downstream.
+I further assume with \citet{Fan2014} and \citet{Ancey2014} that drag forces can be represented by a Gaussian white noise.
+This latter assumption is justified insofar as drag forces fluctuate rapidly compared to the inertial response times of particles and the timescales over which particles change height within the flow profile.
+Defining $\bar{C}_D$ as the empirical drag coefficient evaluated at $\bar{V}$ and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep[e.g.][]{Gardiner1983}, the fluid drag force considered in this chapter can be written
 \be F_D(t) = \Gamma + \sqrt{2 D } \eta(t), \label{eq:drag}\ee
@@
 \rho_f d^2 \bar{C}_D \bar{V}^2$ is the steady component of the drag.
+Although this representation of the fluid forces is a simplified approximation, it is similar to the representations used in earlier stochastic models in that it has no dependence on the vertical structure of the flow \citep[e.g.][]{Fan2014,Ancey2014}.
+This level of simplification is required for an analytically tractable model of bedload particle velocities \citep[e.g.][]{Michaelides1997}.
 
@@
 
-With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics becomes
+With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics in the downstream direction becomes
 \be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
 This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
-Mathematically, equation \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
-Collisions introduce ``jumps" in velocity while turbulence generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.
+Mathematically, eq. \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
+Collisions introduce velocity jumps while turbulence generates velocity diffusion. The collision term is multiplicative in the sense that $u$ multiplies the Poisson noise $\xi_{\nu,\ve}(t)$.
 
-Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
-One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
+Models like eq.
+\ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
+One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning eq.
+\ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
 Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
-Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
+Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by eq.
+\ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
 
@@
 
-The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix section \ref{sec:langmasterderiv} with a limiting procedure, providing
+The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
 \be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
-In this equation, I introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$.
+In this equation, I introduced the parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$.
 The term
@@
 Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
-Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
+Nonlocality is introduced by the collision integral (eq. \ref{eq:colint}) which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
 This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}.
-Physically, it corresponds to binary collisions between particles having different masses and random restitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
-Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $(m u)$ \citep[cf.][]{Feller1968}.
+Physically, it corresponds to binary collisions between particles having different masses and random restitution coefficients \citep[cf.][]{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
+Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $(p = m u)$, considering them as uncorrelated random variables \citep[cf.][]{Feller1968}.
 
-Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
-Assuming the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
+Owing to its nonlocality, eq. \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
+To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
+For bedload particle collisions with the bed, we can expect a mode elasticity since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.
+
 Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
@@
 I solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$.
-Equation \ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg \nu^{-1}$).
+Eq.
+\ref{eq:master} indicates that this solution will be a good approximation to the time-dependent problem when particle motions generally survive multiple collisions ($t\gg \nu^{-1}$).
 
@@
 Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
-With the truncation to two terms, equation \ref{eq:master} gives 
+With the truncation to two terms, eq. \ref{eq:master} gives 
 \be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
@@
 
-In the appendix section \ref{sec:langsteadyderiv} I solve equation \ref{eq:governer} using Laplace transforms, providing
+In the appendix sec. \ref{sec:langsteadyderiv} eq. \ref{eq:governer} is solved with Laplace transforms, providing
 \begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
@@
 \end{multline}
-The factors $\lambda_\pm$ are defined in appendix equations \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
+The factors $\lambda_\pm$ are defined in appendix eq. \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
 The normalization factors $K_\pm$ are 
@@
 
-It is possible to derive the moments of this probability distribution by multiplying \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[c.f.][]{Cox1965}.
-These calculations are provided in appendix section \ref{sec:langmoments}, where the mean bedload velocity is derived as
-\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\gamma}{1-\ve}. \label{eq:meanu}\ee
+It is possible to derive the moments of this probability distribution by multiplying eq.
+\ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[cf.][]{Cox1965}.
+These calculations are provided in appendix sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
+\be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}. \label{eq:meanu}\ee
 This result scales linearly with the mean fluid drag and nonlinearly with the rate and typical elasticity of collisions, indicating a strong influence of particle collisions on bedload velocities.
 The second moment is
-\be \langle u^2 \rangle = 2 \frac{d + \gamma \langle u \rangle}{1-\ve^2}, \ee
+\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
 leading to the velocity variance ($\sigma_u^2 = \bra u^2 \ket - \bra u \ket ^2 $)
-\be \sigma_u^2 = \frac{2 d + \gamma^2}{1-\ve^2}. \label{eq:varu}\ee
-This result demonstrates that velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
-This is consistent with the remarkable similarities sometimes visible between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
-In contrast, particle velocity fluctuations in equation \ref{eq:varu} depend sharply on the parameters representing particle-bed collisions, suggesting collisions are the leading control on particle velocity fluctuations.
+\be \sigma_u^2 = \frac{2 \tilde{D}+ \tilde{\Gamma}^2}{1-\ve^2}. \label{eq:varu}\ee
+This result demonstrates that within the model, velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
+This is consistent with the obvious similarities between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
+In contrast, particle velocity fluctuations in eq.
+\ref{eq:varu} depend sharply on the parameters representing particle-bed collisions, suggesting collisions could be the leading control on particle velocity fluctuations.
 
-Figure \ref{fig:fig2} depicts the velocity characteristics of particles for different realizations of the fluid and collisional forces.
+Fig. \ref{fig:fig2} depicts the velocity characteristics of particles for different realizations of the fluid and collisional forces.
 \begin{figure}
@@
 Flow forces generate fluctuating positive accelerations between collisions.
-Right panels show simulated histograms of particle velocities and exact solutions from equation \ref{eq:steadystate}.
+Right panels show simulated histograms of particle velocities and exact solutions from eq.
+\ref{eq:steadystate}.
 As elasticity $\ve$ varies, the particle velocity distributions interpolate between exponential (inelastic) and Gaussian (elastic) forms.}
@@
 This figure reveals an apparent transition from exponential-like to Gaussian-like velocity distributions as collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
-In between, the full distribution \ref{eq:steadystate} resembles a Gamma distribution, although it is represented by equation \ref{eq:steadystate}, not a Gamma distribution.
+In between, the full distribution eq.
+\ref{eq:steadystate} resembles a Gamma distribution, although it is represented by eq.
+\ref{eq:steadystate}, not a Gamma distribution.
 
@@
 
-In fact, the apparent transition from exponential to Gaussian in figure \ref{fig:fig2} can be demonstrated from equation \ref{eq:steadystate}.
+In fact, the apparent transition from exponential to Gaussian in fig.
+\ref{fig:fig2} can be demonstrated analytically from eq.
+\ref{eq:steadystate}.
 Despite its complex appearance, simple Gaussian and exponential forms derive from this equation as exact mathematical limits.
-When particle-bed collisions are completely inelastic ($\ve = 0$), \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\ve = 1$), \ref{eq:steadystate} becomes Gaussian.
-Figure \ref{fig:fig3} demonstrates in detail the approach of the distribution toward these limits.
+When particle-bed collisions are completely inelastic ($\ve = 0$), eq.
+\ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\ve = 1$), eq.
+\ref{eq:steadystate} becomes Gaussian.
+Fig. \ref{fig:fig3} demonstrates in detail the approach of the distribution toward these limits.
 \begin{figure}
@@
 
-The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see.
-Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for the one at $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
-\be P(u) = \frac{d}{\sqrt{\gamma^2 + 4d}}e^{\frac{\gamma u }{2 d} - \frac{\sqrt{\gamma^2 + 4 d}|u|}{2d}}. \ee
+The exponential limit of eq.
+\ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see.
+Taking $\ve \rightarrow 0 $ in eq.
+\ref{eq:steadystate}, all terms in the series except for the one at $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
+\be P(u) = \frac{\tilde{D}}{\sqrt{\tilde{\Gamma}^2 + 4 \tilde{D}}}e^{\frac{\tilde{\Gamma} u }{2 \tilde{D}} - \frac{\sqrt{\tilde{\Gamma}^2 + 4 \tilde{D}}|u|}{2\tilde{D}}}. \ee
 Thus, for bed load transport conditions with typically very inelastic particle-bed collisions, we can expect exponential-like velocities and large deviations from a Gaussian behavior.
 
-The Gaussian limit as $\ve \rightarrow 1$ of \ref{eq:steadystate} is more difficult to evaluate.
-The challenge is that the statistical moments \ref{eq:mean} and \ref{eq:varu} diverge at the same time as the denominator factors of the distribution \ref{eq:steadystate}.
-In the appendix section \ref{sec:langextremes} I return instead to the original equation \ref{eq:governer} to evaluate the completely elastic limit, obtaining
+The Gaussian limit as $\ve \rightarrow 1$ of eq.
+\ref{eq:steadystate} is more difficult to evaluate.
+The challenge is that the statistical moments (eqs.
+\ref{eq:mean} and \ref{eq:varu}) diverge at the same time as the denominator factors of the distribution eq.
+\ref{eq:steadystate}.
+In the appendix sec.
+\ref{sec:langextremes} I return instead to the original equation \ref{eq:governer} to evaluate the completely elastic limit, obtaining
 \be P(u) = \frac{1}{\sqrt{2\pi\sigma_u^2}}e^{-\frac{(u-\bar{u})^2}{2\sigma_u^2}}. \label{eq:gaussian}\ee
@@
 Comparing eq.
-\ref{eq:steadystate} with experimental data requires values for the steady component of the drag force $\Gamma$, the particle mass $m$, the magnitude of turbulent drag fluctuations $D$, the rate of particle-bed collisions $\nu$, and the mode elasticity of collisions $\varepsilon$.
+\ref{eq:steadystate} with experimental data requires values for the steady component of the drag force $\Gamma$, the particle mass $m$, the magnitude of turbulent drag fluctuations $D$, the rate of particle-bed collisions $\nu$, and the modal elasticity of collisions $\varepsilon$.
 
@@
 In each case, the particle mass is computed from experimental parameters assuming spherical sediment as $m = \pi \rho_s d^3/6$, where $\rho_s$ is the sediment density and $d$ is the particle diameter.
-The steady component of the drag force is estimated as
+The characteristic slip velocity entering the steady component of the drag force (eq.
+\ref{eq:drag}) is not obviously related to any one velocity scale of the fluid flow.
+It should really be calculated as an average over different particle heights, particle velocities, and fluid velocities within the flow profile.
+For lack of a better option we simply assume the typical slip velocity scales with the shear velocity of the flow, as it would for a small particle resting on the bed.
+Using this assumption, the steady component of the drag force is estimated as
 \be \Gamma =  \frac{\pi}{8}\rho C_D(\Rey_p) d^2 u_\ast^2,\ee
@@
 Particle Reynolds numbers are estimated using the shear velocity: $\Rey_p= u_\ast d/nu$. 
+The magnitude $D$ of turbulent fluctuations, dissipation per collision $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution (eq.
+\ref{eq:steadystate}) and the experimental data.
 
-The magnitude $D$ of turbulent fluctuations is eliminated with the mean and variance of the particle velocity.
-The remaining coefficients in the model, the dissipation per collision $\ve$ and the collision rate $\nu$, are treated as calibration parameters and are tuned to provide best fit between the distribution \ref{eq:steadystate} and the experimental data.
-
-The parameters for each experiment and the best-fit calibration parameters are collated in table \ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in figure \ref{fig:fig4ch5}.
+The parameters for each experiment and the best-fit calibration parameters are collected in tab.
+\ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in figure \ref{fig:fig4ch5}.
 \begin{table}
@@
 		\end{tabular}
-		\caption{Parameters used to fit the distributions in figure \ref{fig:fig4ch5}.
+		\caption{Parameters used to fit the distributions in fig.
+\ref{fig:fig4ch5}.
 The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St$, and Froude number $\Fro$.
-The final two columns involve values that were tuned to fit the theoretical distribution \ref{eq:steadystate} to the experimental data, the dissipation $\ve$ and the collision rate $\nu$.
+The final two columns involve values that were tuned to fit the theoretical distribution (eq.
+\ref{eq:steadystate}) to the experimental data, the dissipation $\ve$ and the collision rate $\nu$.
 Units are indicated in square brackets.
@@
 \end{table}
-In every case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model \ref{eq:langevin} captures the essential physics.
+In each case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model eq. \ref{eq:langevin} captures the essential physics.
 \begin{figure}
 	\centerline{\includegraphics{./figures/ch5/Fig4expComparison.png}}
-	\caption{This figure compares the available experimental data on velocity distributions and the theoretical velocity distribution \ref{eq:steadystate} with parameters $\nu$, $\varepsilon$, and $D$ treated as calibration parameters.
+	\caption{This figure compares the available experimental data on velocity distributions and the theoretical velocity distribution eq.
+\ref{eq:steadystate} with parameters $\nu$, $\varepsilon$, and $D$ treated as calibration parameters.
 In all cases there is good agreement, indicating that the model is capable of the range of distributions exhibited by the experimental data.} \label{fig:fig4ch5}
@@
 
-Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise.
-Even though the non-Gaussian and history-dependent aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the time scales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise for reasons of analytical tractability \cite[e.g.][]{Michaelides1997}.
+Even though the non-Gaussian, spatio-temporally-varying, and history-dependent aspects of fluid turbulence certainly do impact sediment movement \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is laminar or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the time scales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+Nevertheless, it is possible the vertical flow structure, and in particular the size of particles in comparison to the thickness of the laminar sublayer, may be explanatory of the different velocity distributions resolved in bedload transport experiments, and future studies should investigate this possibility.
 
-The model developed in this chapter described particle-bed collision forces as a sequence of instantaneous impulses in an approach that is reminiscent of the kinetic theory of gases \citep{Landau1969}.
+The model developed in this chapter described particle-bed collision forces as a sequence of instantaneous impulses in an approach that is reminiscent of the kinetic theory of gases \citep{Landau1969,Brilliantov2004}.
 The effect of each collision on the streamwise particle velocity was described by a simple elasticity-like coefficient.
@@
 An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates from granular interactions.
-Particle size enters the Langevin model \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
-The analytical velocity distribution \ref{eq:steadystate} was fit to six different experimental datasets in figure \ref{fig:fig4ch5}, and the best fit parameters were presented in table \ref{tab:calib}.
+Particle size enters the Langevin model eq. \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
+The analytical velocity distribution eq.
+\ref{eq:steadystate} was fit to six different experimental datasets in fig.
+\ref{fig:fig4ch5}, and the best fit parameters were presented in tab.
+\ref{tab:calib}.
 These fit parameters suggest collisions are generally less elastic for smaller particle sizes, whereas they are more elastic for larger particle sizes.
@@
 
-Because the elasticity $\ve$ lumps together collision geometry and dissipation effects, it is challenging to attribute the increasing relationship between elasticity and particle size evident in table \ref{tab:calib} directly to particle size.
+Because the elasticity $\ve$ lumps together collision geometry and dissipation effects, it is challenging to attribute the increasing relationship between elasticity and particle size evident in tab.
+\ref{tab:calib} directly to particle size.
 Yet this would be consistent with experiments of idealized particle collisions in viscous fluids.
@@
 This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
-The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the full range of bedload velocity distributions reported in experimental studies.
+The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the range of bedload velocity distributions which have to date been reported in experimental studies.
 
-This work hints that particle size, and not flow hydraulics, may be primarily responsible for the different bedload velocity distributions obtained in experiments, although future study is required for a definite conclusion.
-The next step is to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions.
-Such an effort would produce additional insight into the controls of transverse, vertical, and rotational movements on the downstream velocities of individual grains.
+This work hints that particle size may be a leading order control over the different bedload velocity distributions obtained in experiments, although future study is required for a definite conclusion.
+Next steps are to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions and to introduce velocity dependence and vertical flow structure in the flow forces acting on moving particles.
+These efforts would produce additional insight into the controls of fluid turbulence and transverse, vertical, and rotational movements on the downstream velocities of individual bedload grains.
 
@@
 
-In this drag force, the collision force \ref{eq:col}, the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions.
+In this drag force, the collision force (eq.
+\ref{eq:col}), the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions.
 This set of possiblities defines a statistical ensemble.

### 2021-09-23 revision/ch6conc.tex
uh

@@
 The work has improved upon earlier descriptions.
-It describes the movements of grains along a wider range of timescales than before, highlights Langevin equations, master equations, and random walks as the common threads of stochastic sediment transport modelling, and provides more realistic descriptions of bedload transport.
+It describes the movements of grains along a wider range of timescales than before, highlights Langevin equations, master equations, and random walks as the common threads of stochastic sediment transport modeling, and provides more realistic descriptions of bedload transport.
 In particular, I have
@@
 
-A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modelling are constrained.
-The noises used in this thesis include white Gaussian, Poisson (shot), and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
+A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modeling are constrained.
+The noises used in this thesis include white Gaussian, Poisson pulse, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
 Turbulence was described using Gaussian noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise.
@@
 
-River science offers no guarantee that these idealized white noises which we happen to understand best are sufficient to describe its phenomena.
-White noises are an idealization which is probably never realized in nature \citep{Gardiner1983,Kubo1978}.
-In contrast, colored noises in which fluctuations have favoured frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
+River science offers no guarantee that these idealized noise sources which we happen to understand best are sufficient to describe its phenomena.
+White noises in particular are an idealization which is probably never realized in nature \citep{Gardiner1983,Kubo1978}.
+In contrast, colored noises in which fluctuations have favored frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
 In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
-It will be problematic if these phenomena require colored noise for their description.
-Dynamical equations driven by colored noise can be extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
-Even if white noise models like those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.
+It will be problematic if these phenomena require colored noise to be incorporated into dynamical equations for an adequate description.
+Dynamical equations driven by colored noise are extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
+Even if white noise models like some of those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.
 
@@
 
-The first major contribution of this thesis is in chapter \ref{ch:flux}.
+The first major contribution of this thesis is in ch.
+\ref{ch:flux}.
 Here, I formulated the probability distribution of the sediment flux from the trajectories of individual particles moving downstream.
@@
 
-Second, in chapter \ref{ch:flux} I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
+Second, in ch. \ref{ch:flux} I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
 Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (sec.
@@
 
-Third, chapter \ref{ch:ch3} modified the birth-death description of bedload transport (sec.
+Third, ch.
+\ref{ch:ch3} modified the birth-death description of bedload transport (sec.
 \ref{sec:birthdeath}) to include feedbacks between the local bed elevation and the entrainment and deposition rates.
 This allowed for a mathematical investigation of (1) how bed elevation changes affect sediment transport rates and (2) how bed elevation changes control the residence times of particles buried in the bed.
-Sediment transport fluctuations have come under increasing scrutiny with the resurgence of stochastic modelling in sediment transport, while the burial times of particles are crucial for using sediment tracers to predict sediment transport.
+Sediment transport fluctuations have come under increasing scrutiny with the resurgence of stochastic modeling in sediment transport, while the burial times of particles are crucial for using sediment tracers to predict sediment transport.
 Earlier birth-death models had generally considered that entrainment and deposition rates of particles remain constant even though these processes imply bed degradation and aggradation respectively, which are known to modify the entrainment and deposition rates in a negative feedback.
@@
 
-Fourth, chapter \ref{ch:downDiff} presents a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (sec.
+Fourth, ch.
+\ref{ch:downDiff} presents a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (sec.
 \ref{sec:trajmodels}), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
@@
 
-Finally, chapter \ref{ch:langevin} displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
+Finally, ch. \ref{ch:langevin} displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
 This work provides the first comprehensive description of all bedload velocity distributions observed in experiments (sec.
@@
 
-There have been a few works on stochastic modelling of sediment transport with multiple grain sizes \citep{Sun2000,Parker2000}, but these approaches are not spatially distributed, so there is as yet no stochastic method to understand sorting processes.
+There have been a few works on stochastic modeling of sediment transport with multiple grain sizes \citep{Sun2000,Parker2000}, but these approaches are not spatially distributed, so there is as yet no stochastic method to understand sorting processes.
 
@@
 Simplified experimental geometries based on bimodal sediment beds \citep[e.g.][]{Houssais2012} would be a great context to revisit these issues.
-Extending the motion-rest model of chapter \ref{ch:ch3} to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
+Extending the motion-rest model of ch.
+\ref{ch:ch3} to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
 

### 2021-09-24 revision/abs.tex
getting close.

@@
 This links the sediment flux to the grain-scale dynamics and describes particle trajectories with additional detail compared to earlier descriptions.
-Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting how long particles can stay buried in the riverbed.
+Second, I include feedbacks between local bed elevations and sediment transport, quantifying the interplay of bed elevation changes and sediment transport rates and predicting how long particles can stay buried in the river bed.
 Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
 Finally, describe bedload dynamics on short timescales, predicting the movement velocities of bedload particles using methods adopted from granular physics.
-I conclude by summarizing these developments, discussing their implications for the statistical description of bedload transport, and suggesting how we can use this modelling progress to better understand landscapes.
-
-\endinput
-
-This research provides solutions to longstanding problems within the statistical approach to sediment transport.
-
-and the mathematical tools applied in this thesis should be of interest to researchers of many other problems. Most
-This research has broad implications across river science where there is increasing acknowledgement of the shortcomings of the continuum hypothesis for predicting dynamics in mountain streams.
-This work extends a wide body of work on individual particle motions which originates from the 1930s, and unifies it with more recent approaches to calculate the bulk bedload flux which have not, until now, been related to the movement characteristics of individual grains.
-
-A central task within Earth science is to understand and predict the evolution of landscapes due to flowing water.
-As rainfall channelizes and flows downhill, it carves out basins and etches in networks of intersecting channels.
-In the highlands of these networks, water arranges boulders and gravels into an intricate array of patterns -- steps, pools, bars, and riffles, which cooperate with vegetation both live and dead to set the stage for much of Earth's biota.
-In the lowlands, channels laden with old mountain sediments, weathered and abraded to fine particles, splay across floodplains and drift between configurations through millennia.
-
-Sediment transport is a main driver of all fluvial dynamics.
-Channel evolution ultimately occurs because individual sediment grains move from one location to another.
-Yet in a majority of modelling studies, landscapes are represented as continua, where the locations of individual grains are averaged away, and sediment transport is represented as a steady stream of mass, rather than the intermittent movements of individual grains
-Useful as this continuum approach may be, many fluvial phenomena are not well-suited for it.
-Sediment transport rates are known to fluctuate widely through space and time, and these fluctuations are understood to initiate bedform development and control channel widths.
-The largest grains in small mountain channels are known to confer the most stability, while violating any assumption of being small compared to the scales of interest.
-To understand fluvial dynamics, we need the capability to model discrete grains, not just continua.
-
-In this thesis, I present four or five years of my theoretical research into the movements of individual sediment grains in river channels.
-This sub-field of Earth science, with a focus on grain-scale process, has existed at least since Hans Albert Einstein in 1937, and within it, modellers have traditionally compromised on severe approximations to balance realism against mathematical difficulty.
-In this enterprise, perfectly flat beds which do not change shape, spherical grains, infinite movement velocities, and turbulence free flows have been typical assumptions to make progress, even though they have little basis in reality.
-The research presented here takes on more mathematical difficulty than before to introduce more realism, adding some bricks to the fortress, and introducing new methods to the field which should work well for other people to make more bricks later.
-I hope you find it useful!
-Happy reading.
+I conclude by summarizing these developments, discussing their implications for the statistical description of bedload transport, and suggesting how we can use this modeling progress to better understand landscapes.

### 2021-09-24 revision/ack.tex
getting close.

@@
 I was lucky to use the right trigger words.
-Marwan, thanks for all of the guidance, for lending me hundreds of books and papers, for the freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.
+Marwan, thanks for all of the guidance, for lending me many dozens of books and papers, for the freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.
 
@@
 Thank you Rui Ferreira and Brett Eaton!
-Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics.
-Your kindness, physics knowledge, and wide-ranging interests beyond physical science are inspiring.
-Thanks for the thoughtful comments and advice on my PhD.
+Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics. Thanks especially for the thoughtful comments on my thesis.
 Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modeling in light of real world complexity.

### 2021-09-24 revision/app_ch2.tex
getting close.

@@
 \chapter{Calculations involved in dynamical sediment flux model}
-\label{sec:appendixA}
+\label{sec:appendixAfluc}
 \section{Derivation of the master equation for particle position}
@@
 
-Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of eq. \ref{eq:logi} gives
+Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of Eq. \ref{eq:logi} gives
 \be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket. \label{eq:furut}\ee
@@
 \be G^{-1}\pt \tilde{P}(g,t) = \pt \bra \eta F\ket = G\bra \eta F \ket + k\big[ \bra \eta \ket \tilde{P}(g,t) - G^{-1} \pt \tilde{P}(g,t) \big]. \ee
-Applying eq. \ref{eq:furut} finally gives
+Applying Eq. \ref{eq:furut} finally gives
 \be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
@@
 \label{sec:fluccymastersol}
-The position probability distribution can be obtained from eq.
+The position probability distribution can be obtained from Eq.
 \ref{eq:flippymaster} for the initial conditions $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ using Fourier transforms in space and Laplace transforms in time.
@@
 The numerator terms involving $g$ are easily expressed as first and second derivatives with respect to $x$. The problem, then is to calculate the inverse Fourier integral of the denominator.
-The inverse Fourier transform can be calculated by expanding the denominator of eq. \ref{eq:prev} in partial fractions and applying the contour integral \citep[e.g.][]{Arfken1985}
+The inverse Fourier transform can be calculated by expanding the denominator of Eq. \ref{eq:prev} in partial fractions and applying the contour integral \citep[e.g.][]{Arfken1985}
 \be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g - i c} dg = \sigma(c)\theta(-x\sigma(c))\exp\big(-|c x|\big). \label{eq:contour}\ee
@@
 \begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du, \end{multline}
-as reported in eq. \ref{eq:monster}.
+as reported in Eq. \ref{eq:monster}.
 
@@
 \label{sec:fluxconstant}
-The central object required to calculate the sediment flux probability distribution in eq. \ref{eq:flippyflux} is 
+The central object required to calculate the sediment flux probability distribution in Eq. \ref{eq:flippyflux} is 
 \be \Lambda(T) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,T).\ee
@@
 \be \tilde{\Lambda}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s).\ee
-The Laplace transform of the rate function therefore follows after integrating eq. \ref{eq:laplaceapp} twice.
+The Laplace transform of the rate function therefore follows after integrating Eq. \ref{eq:laplaceapp} twice.
 
@@
 
-After some work, using eq.
+After some work, using Eq.
 \ref{eq:magic} and tabulated Laplace transform pairs \citep[e.g.][]{Arfken1985,Prudnikov1992a} to perform the inverse Laplace transforms, the net result for the rate constant is
@@
 \end{multline}
-as reported in eq. \ref{eq:fluxy}.
+as reported in Eq. \ref{eq:fluxy}.
 
@@
 Equation \ref{eq:fluxy} is difficult to work with.
-To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep{Weiss1994} to invert the Laplace-transformed rate function of eq.
+To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep{Weiss1994} to invert the Laplace-transformed rate function of Eq.
 \ref{eq:laplacefluxrate} at the opposite extreme of $s$.
 
-For example, at short times, expanding eq. \ref{eq:laplacefluxrate} as $s\rightarrow \infty$ gives
+For example, at short times, expanding Eq. \ref{eq:laplacefluxrate} as $s\rightarrow \infty$ gives
 \be \tilde{\Lambda}(s) = \frac{\rho k_E V}{2 k s^2} + \frac{\rho k_E}{k}\sqrt{\frac{D}{4 s^3}} \ee
@@
 \be \tilde{\Lambda}(s) = \frac{\rho k_E V }{ks^2}, \ee
-and this inverts to $\Lambda(T) = \rho k_E V T/k$ as the long time solution, providing a mean flux equivalent to the Einstein model (sec. \ref{sec:lisle}).
-These limits are summarized in eq. \ref{eq:fluccolimitcases}.
+and this inverts to $\Lambda(T) = \rho k_E V T/k$ as the long time solution, providing a mean flux equivalent to the Einstein model (Sec. \ref{sec:lisle}).
+These limits are summarized in Eq. \ref{eq:fluccolimitcases}.

### 2021-09-24 revision/app_ch3.tex
getting close.

@@
 \section{Numerical simulation algorithm}
-
+\label{sec:arr}
 The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
@@
 The SSA hinges on the defining property of a Markov process. When the transition rates from one state to another are not dependent on the distant past, the process is Markov \citep{Cox1965}.
-The following sections demonstrate that the time intervals $\tau$ between subsequent transitions are exponentially distributed within the model of ch.
+The following sections demonstrate that the time intervals $\tau$ between subsequent transitions are exponentially distributed within the model of Ch.
 \ref{ch:ch3}.
@@
 The joint description of bedload transport and bed elevation changes is characterized by a set of states $(n,m)$ where $n$ and $m$ are integers.
-The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in eqs. (\ref{eq:rate1}-\ref{eq:rate4}).
+The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in Eqs. (\ref{eq:rate1}-\ref{eq:rate4}).
 
@@
 \subsection{Selection of transitions that occur}
-
-So far, eq.
+\label{sec:brr}
+So far, Eq.
 \ref{eq:exp} demonstrates how to step the time from one transition to the next, but it remains unclear how to step the state variables $n$ and $m$ at each transition time.
@@
 
-Intuitively, this will depend on the relative magnitudes of the rates from eqs. \ref{eq:rate1}-\ref{eq:rate4}.
+Intuitively, this will depend on the relative magnitudes of the rates from Eqs. \ref{eq:rate1}-\ref{eq:rate4}.
 The transition with the highest rate is most likely to occur.
@@
 
-In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (eq.
-\ref{eq:exp}), draws a uniform random number from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in eq.
+In summary, to step the process through a single transition, the SSA draws the time interval to the next transition from the distribution (Eq.
+\ref{eq:exp}), draws a uniform random number from $[0,1]$, then uses this uniform random to select the transition that occurs from the cumulative sum of the ratios in Eq.
 \ref{eq:rel}.
@@
 \subsection{Pseudo code for the Gillespie SSA}
-
-Simulations are initialized by the initial conditions $n_0$ and $m_0$, the model parameters used in eqs.
+\label{sec:crr}
+Simulations are initialized by the initial conditions $n_0$ and $m_0$, the model parameters used in Eqs.
 \ref{eq:rate1}-\ref{eq:rate4}, and the desired simulation duration $t_\text{max}$.
-The SSA uses these inputs to generate time series of $n$ and $m$ as follows:
+The SSA uses these inputs to generate timeseries of $n$ and $m$ as follows:
 
@@
 	\While{$t<t_{\text{max}}$;}{	\Comment{Simulation will go until $t$ surpasses $t_\text{max}$}
-		\State record $(n,m,t)$ \Comment{Build time series of $n$ and $m$}		
-		\State draw $\tau$ from eq. \ref{eq:exp} \Comment{Select time to next transition}
+		\State record $(n,m,t)$ \Comment{Build timeseries of $n$ and $m$}		
+		\State draw $\tau$ from Eq. \ref{eq:exp} \Comment{Select time to next transition}
 		\State $t \coloneqq t+\tau$
 		\State draw a random number $r$ in $[0,1]$ \Comment{Select type of transition that occurs}
-		\State compute the ratios $r_1,r_2,r_3,r_4$ in eq. \ref{eq:rel}
+		\State compute the ratios $r_1,r_2,r_3,r_4$ in Eq. \ref{eq:rel}
 		\State form the cumulative sums $r_i = \sum_{1\leq j\leq i}r_j$ 
@@
 
-By assuming the dynamics of the particle activity are totally independent of the bed elevation and summing eq.
+Assuming the dynamics of the particle activity are totally independent of the bed elevation and summing Eq.
 \ref{eq:elemaster} over all values of $m$ obtains a mean field equation for the particle activity:
@@
 after applying the normalization condition $G(1)=1$.
-From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in eq.
-\ref{eq:ancey} and introduced originally in sec.
+From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in Eq.
+\ref{eq:ancey} and introduced originally in Sec.
 \ref{sec:birthdeath}.
@@
 \subsection{Mean field solution of bed elevations}
-
-The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so eq. \ref{eq:elemaster} produces
+\label{sec:drr}
+The negative binomial distribution provides $\langle n |m \rangle = \langle n \rangle$, so Eq. \ref{eq:elemaster} produces
 \begin{multline} 0 = [\lambda + \mu \langle n \rangle][1+\kappa(m+1)]M(m+1) + \sigma \langle n \rangle [1-\kappa(m-1)]M(m-1) \\- \{[\lambda + \mu \langle n \rangle](1+\kappa m) + \sigma \langle n \rangle (1-\kappa m) \}M(m).
 \end{multline}
-Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives eq. \ref{eq:ou}:
+Identifying $E=\lambda + \mu \langle n \rangle$ and $D = \sigma \langle n \rangle$ and incorporating $E=D$ gives Eq. \ref{eq:ou}:
 \be 0 = [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1) - 2M(m).\ee
 
-This can be solved using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$ This works since $z_1$ is small.
+This can be solved using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$.
+This works since $z_1$ is small.
 Introducing $z$ and writing $\overline{\kappa}=\kappa/z_1$ gives
@@
 \be M = M_0e^{-\kappa (z/z_1)^2}\ee
-as provided in ch. \ref{ch:ch3}.
+as provided in Ch. \ref{ch:ch3}.
 
 \subsection{Closure equation approach for bed elevations}
+\label{sec:err}
 This section produces an approximate relationship to close
@@
 \be \langle n | m \rangle \approx \langle n \rangle  - \kappa c m,\ee
-for the mean particle activity conditional to the elevation $m$ into eq. \ref{eq:elemaster}, noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
+for the mean particle activity conditional to the elevation $m$ into Eq. \ref{eq:elemaster}, noting $E=D$, and neglecting terms of $O(\kappa^2)$ provides 
 \begin{multline} 0 \approx \Big[ 1+\kappa(m+1)\Big\{ 1-\frac{\mu c}{E}\Big\}\Big]M(m+1) + \Big[1-\kappa(m-1)\Big\{1+\frac{\sigma c}{E}\Big\}\Big]\\-\Big[2-\kappa m\Big\{\frac{(\sigma+\mu)c}{E}\Big\}\Big]M(m)
@@
 Finally, integrating, expanding to first order in $\kappa$, and exponentiating to solve for $M$ produces
-\be M(m) = M_0 \exp\Big\{-\kappa m^2 \Big[1 + \frac{(\sigma-\mu)c}{2E}\Big]\Big\}.\ee
+\be M(m) = M_0 \exp\Big\{-\kappa m^2 \Big[1 + \frac{(\sigma-\mu)c}{2E}\Big]\Big\}. \label{eq:zap}\ee
 Since the numerical solutions show $\sigma_m^2 = \frac{1}{4\kappa}$, this suggests the closure relation
-\be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \ee
-corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in eq. \ref{eq:closure} and plotted with the numerical simulations in fig. \ref{fig:condmoms}.
+\be \langle n | m \rangle = \langle n \rangle\Big( 1-\frac{2\kappa m}{1-\mu/\sigma}\Big) \label{eq:zim}\ee
+corresponding to $c=2E/(\sigma-\mu)$. This is the closure relationship provided in Eq. \ref{eq:closure} and plotted with the numerical simulations in Fig. \ref{fig:condmoms}.
+The elevation distribution Eq. \ref{eq:ou2} follows by combining Eqs. \ref{eq:zap} and \ref{eq:zim}.

### 2021-09-24 revision/app_ch4.tex
getting close.

@@
 \be \tilde{p}(\eta,s) = \int_0^\infty dx e^{-\eta x}\int_0^\infty dt e^{-st} p(x,t). \label{eq:doubletransform}\ee
-The Laplace-transformed moments of $x$ are linked to derivatives of the double transformed distribution (eq. \ref{eq:doubletransform}) \citep[cf.,][]{Berezhkovskii2002}.
+The Laplace-transformed moments of $x$ are linked to derivatives of the double transformed distribution (Eq. \ref{eq:doubletransform}) \citep[cf.,][]{Berezhkovskii2002}.
 Eq. (\ref{eq:doubletransform}) implies
@@
 
-Double transforming manuscript equations (1-3) using the definition eq. \ref{eq:doubletransform} gives
+Double transforming manuscript equations (1-3) using the definition Eq. \ref{eq:doubletransform} gives
 
@@
 \end{align}
-The total probability is $p(x,t) = p_0(x,t) + p_1(x,t) + p_2(x,t)$. Using eqs. \ref{eq:A}-\ref{eq:Z} this becomes, in the double Laplace representation, 
+The total probability is $p(x,t) = p_0(x,t) + p_1(x,t) + p_2(x,t)$. Using Eqs. \ref{eq:A}-\ref{eq:Z} this becomes, in the double Laplace representation, 
 \begin{multline}
@@
 \end{multline}
-Plugging the propagators outlined in manuscript equations (8-9) into eq. \ref{eq:lap} gives 
+Plugging the propagators outlined in manuscript equations (8-9) into Eq. \ref{eq:lap} gives 
 \be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}.\label{eq:nicedist}\ee
@@
 
-We compute the first two moments of position $x$ and ultimately its variance using eq.
+We compute the first two moments of position $x$ and ultimately its variance using Eq.
 \ref{eq:momenttrick}.
-The first two derivatives of the double Laplace transformed distribution eq.
+The first two derivatives of the double Laplace transformed distribution Eq.
 \ref{eq:nicedist} are
@@
 \be \partial_\eta^2 \tp(\eta,s) = 2v^2 \frac{1}{s} \frac{(s+\kappa+k_1)[(s+\kappa + k')s+\kappa k_2][\theta_2(s+\kappa) + k_1]}{[\eta v(s+\kappa + k_1) + (s+\kappa + k')s+ \kappa k_2]^3}.\ee
-Evaluating these at $\eta=0$ and applying eq. \ref{eq:momenttrick} provides the Laplace transformed moments
+Evaluating these at $\eta=0$ and applying Eq. \ref{eq:momenttrick} provides the Laplace transformed moments
 \be  \frac{\bra\tilde{x}(s)\ket} {v} = \frac{1}{s}\frac{\theta_2(s+\kappa)+k_1}{(s+\kappa+k')s+\kappa k_2} = \frac{1}{s} \frac{\theta_2(s+\kappa)+k_1}{(s+a+b)(s+a-b)}\label{eq:lapmean},\ee
@@
 This equation rearranges to manuscript equation (11).
-The second moment (eq. \ref{eq:lapsecondmom}) is 
+The second moment (Eq. \ref{eq:lapsecondmom}) is 
 \begin{multline}
@@
 We determine the diffusion exponents $\gamma$ in the local, intermediate, and global ranges using the two limiting cases described in the discussion of the manuscript.
-Limit (1) is $\kappa \rightarrow 0$. We take this limit in eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} with initial condition $\theta_1=1$ to obtain
+Limit (1) is $\kappa \rightarrow 0$. We take this limit in Eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} with initial condition $\theta_1=1$ to obtain
 \begin{align}
@@
 We further investigate limit (1) for arbitrary initial conditions.
-By applying Tauberian theorems, we assert the $ t \rightarrow 0$ variance is determined by the $s\rightarrow \infty$ limits of eqs.
+By applying Tauberian theorems, we assert the $ t \rightarrow 0$ variance is determined by the $s\rightarrow \infty$ limits of Eqs.
 \ref{eq:lapmean} and \ref{eq:lapsecondmom} \citep[e.g.,][]{Weiss1994, Weeks1998}.
@@
 
-Limit (2) is $1/k_2 \rightarrow 0$ and $v\rightarrow \infty$ while $v/k_2 = l$. Under this limit, eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} provide
+Limit (2) is $1/k_2 \rightarrow 0$ and $v\rightarrow \infty$ while $v/k_2 = l$. Under this limit, Eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} provide
 \begin{align}
@@
 Expanding for small $t$ provides $\sigma_x^2(t) = 2D_d t$ -- intermediate range normal diffusion, while sending $t\rightarrow \infty$ provides $\sigma_x^2 = (2D_d\kappa + c^2)/\kappa^2$ -- a constant variance in the geomorphic range.
-The global range is characterized by competition between terms in eq. \ref{eq:wuvar} and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.,][]{Wu2019}.
-Finally, both eqs. \ref{eq:li} and \ref{eq:wuvar} reduce to the Einstein result $\sigma_x^2(t) = 2D_d t$ in further simplified limits.
+The global range is characterized by competition between terms in Eq. \ref{eq:wuvar} and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.,][]{Wu2019}.
+Finally, both Eqs. \ref{eq:li} and \ref{eq:wuvar} reduce to the Einstein result $\sigma_x^2(t) = 2D_d t$ in further simplified limits.

### 2021-09-24 revision/app_ch5.tex
getting close.

@@
 \label{sec:langmasterderiv}
-To derive the master equation from eq.
+To derive the master equation from Eq.
 \ref{eq:langevin}, we can temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 \tilde{D}} h$ with $h$ distributed as $f(h)$.
 We will later take a GWN limit on this noise.
-With this assumption, integrating eq.
+With this assumption, integrating Eq.
 \ref{eq:langevin} over a small time interval $\delta t$ and incorporating the Ito interpretation for the collision term provides
@@
 \label{sec:langsteadyderiv}
-Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of eq. \ref{eq:master} develops the recursion relation
+Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of Eq. \ref{eq:master} develops the recursion relation
 \be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
@@
 \be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
-The Fourier transform (eq.
+The Fourier transform (Eq.
 \ref{eq:recursion}) has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
@@
 	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
-Finally, we can invert the Fourier transforms term by term with contour integration (another use of the integral given in eq.
-\ref{eq:contour}), and incorporate eq.
-\ref{eq:q} to obtain the steady-state solution (eq.
+Finally, we can invert the Fourier transforms term by term with contour integration (another use of the integral given in Eq.
+\ref{eq:contour}), and incorporate Eq.
+\ref{eq:q} to obtain the steady-state solution (Eq.
 \ref{eq:steadystate}).
@@
 \label{sec:langmoments}
-Taking eq. \ref{eq:master}, multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
+Taking Eq. \ref{eq:master}, multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
 \be 0 = \tilde{D} k (k-1)\langle u^{k-2} \rangle + \tilde{\Gamma} k \langle u^{k-1} \rangle + (\ve^k-1)\langle u^k \rangle. \ee
-$k=1$ provides the mean (given in eq. \ref{eq:meanu})
+$k=1$ provides the mean (given in Eq. \ref{eq:meanu})
 \be \langle u \rangle =  \frac{\tilde{\Gamma}}{1-\ve} = \frac{\Gamma}{\nu(1-\ve)} \ee
@@
 \be \sigma_u^2 = \frac{2\tilde{D} + \tilde{\Gamma}^2}{1-\ve^2}\ee
-given in eq. \ref{eq:varu}.
+given in Eq. \ref{eq:varu}.
 \section{Weak and strong collision limits}
@@
 Now I will demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
-The limit is challenging since the steady-state distribution eq.
+The limit is challenging since the steady-state distribution Eq.
 \ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
@@
 The resulting differential equation for $Q(z)$ is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
-Incorporating this transformation into eq. \ref{eq:master} provides the ``normalized" master equation
+Incorporating this transformation into Eq. \ref{eq:master} provides the ``normalized" master equation
 \begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 \end{multline}
-This equation remains exact and is only a change of variables from eq. \ref{eq:master}.
+This equation remains exact and is only a change of variables from Eq. \ref{eq:master}.
 Now I approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
@@
 which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$ \citep[e.g.][]{Gardiner1983}.
-This solution provides eq. \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.
+This solution provides Eq. \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.

### 2021-09-24 revision/ch1intro.tex
getting close.

@@
 
-\chapter{Sediment transport and landscape evolution}
+\chapter{Sediment transport and channel evolution}
 \label{ch:Introduction}
@@
 To address these complications, I develop in this thesis a handful of new bedload transport models using methods adopted from statistical physics.
-To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary during this review to indicate common themes, highlight the shortcomings to be addressed, and show continuity with my own developments which follow in subsequent chapters.
+To provide context and motivation, the thesis begins by reviewing earlier approaches to bedload transport which my work is most related to, rephrasing works as necessary to indicate common themes, highlight the shortcomings to be addressed, and show continuity with my own developments which follow in subsequent chapters.
 
@@
 
-A basic problem in sediment transport theory is to predict the downstream movement of individual particles.
+A basic problem in sediment transport modeling is to predict the downstream movement of individual particles.
 At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
-Particles moving as bedload are driven downstream by a turbulent fluid flow, but relationships between the fluid flow and the applied forces are known only approximately \citep{Michaelides1997a,Schmeeckle2007}.
-Downstream movement is resisted by frictional encounters (skidding, pivoting, colliding) between moving particles and the bed, but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.
+Particles moving as bedload are driven downstream by a turbulent fluid flow, but relationships between the fluid flow and the applied forces are known only approximately \citep{Michaelides1997,Schmeeckle2007}.
+Downstream movement is resisted by frictional encounters between moving particles and the bed (skidding, pivoting, colliding), but since the bed is a granular surface, its geometry is difficult to characterize \citep{Gordon1972}, and the role of bed interactions is challenging to describe \citep{Sekine1992, Nino1998}.
 
@@
 Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
-This obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
+This observation time dependence obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
 
@@
 
-Equation \ref{eq:einlangevin} is a kind of dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (eq.
+Equation \ref{eq:einlangevin} is a dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (Eq.
 \ref{eq:einrando}).
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/einsteinConcept.pdf}
-	\caption{Panel (a) indicates the representation of Einstein's model as an idealized Poisson pulse noise, as indicated in eq.
-\ref{eq:einrando}, while panel (b) shows the ``stairstep" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$).
+	\caption{Panel (a) indicates the representation of Einstein's model as an idealized Poisson pulse noise, as indicated in Eq.
+\ref{eq:einrando}, while panel (b) shows the ``stair-step" trajectories of sediment particles moving downstream through cycles of steps (which are instantaneous) and rests (which have mean duration $1/k_E$).
 }
@@
 Here, $\mathcal{I}_\nu$ is a modified Bessel function of order $\nu$. 
-The probability distribution eq. \ref{eq:eindist} fully characterizes the downstream movement of an individual particle alternating randomly through steps and rests.
+The probability distribution Eq. \ref{eq:eindist} fully characterizes the downstream movement of an individual particle alternating randomly through steps and rests.
 
@@
 Advection occurs because particles move downstream.
-Diffusion occurs because step lengths and instants when entrainment events occur vary from one particle to the next.
+Diffusion occurs because step lengths and entrainment times vary from one particle to the next.
 
-One can calculate all moments of the position by multiplying eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
+One can calculate all moments of the position by multiplying Eq. \ref{eq:einmaster} by $x^n$ and integrating over space. This gives the mean position of the particle
 \be \langle x (t)  \rangle = k_E \ell t. \ee
@@
 The advent of high speed camera experiments produced new insight into the local and intermediate ranges of bedload motion \citep{Abbott1977,Francis1973,Drake1988} which \citet{Einstein1937} probably did not intend for his model to describe.
-In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle's height within the flow profile \citep{VanRijn1984,Wiberg1985}.
+In the local range, particles move with a fluctuating velocity due to the variable drag of the turbulent flow \citep{Lajeunesse2010,Fathel2015} and changes in the particle height within the flow profile \citep{VanRijn1984,Wiberg1985}.
 In the intermediate range, particle-bed collisions impart additional variability to sediment velocities \citep{Gordon1972,Martin2013}.
@@
 Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales.
-They approximated particle velocities as constant (neglecting fluctuations), and assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
+They approximated particle velocities as constant (neglecting fluctuations), and they assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
 The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
 \be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
-%where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
-This noise is displayed in fig. \ref{fig:lislefig} panel (a). Some particle trajectories produced by eq. \ref{eq:lislelangevin} are displayed in fig. \ref{fig:lislefig} panel (b).
+where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
+This noise is displayed in Fig. \ref{fig:lislefig} (a). Some particle trajectories produced by Eq. \ref{eq:lislelangevin} are displayed in Fig. \ref{fig:lislefig} (b).
 
-The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt') \rangle$ becomes \citep{Balakrishnan1993}
+The governing equation of the position probability distribution $P(x,t) = \langle \delta(x-\int_0^t V\eta(t')dt') \rangle$ becomes \citep[cf.][]{Balakrishnan1993}
 \be \big(\pt^2 + V \px \pt + k_E V \px + k \pt \big) P(x,t) = 0,\label{eq:lislemaster}\ee
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/lisleConcept.pdf}
-	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise eq.
+	\caption{Panel (a) indicates the generalization of Einstein's model to include the interval of sediment motion between entrainment and deposition, now represented with dichotomous noise Eq.
 \ref{eq:lislelangevin}, while panel (b) shows the tilted stair-step trajectories of sediment particles moving downstream in cycles of motion at velocity $V$ (with mean duration $1/k_D$) and rest (with mean duration $1/k_E$).
@@
 
-For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of eq.
+For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of Eq.
 \ref{eq:lislemaster} is \citep{Lisle1998}
@@
 where $\chi = k_D x/V$ and $\tau = k_E(t-x/V)$ are shorthand notations.
-Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with an effective velocity $V_\text{eff} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state.
+Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with a virtual velocity $V_\text{virt} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state.
 In contrast, the variance develops a two range scaling.
@@
 which is a non-trivial result.
-At short times, for $t\ll 1/k$, eq.
+At short times, for $t\ll 1/k$, Eq.
 \ref{eq:lislevar} shows diffusion $\sigma_x^2 \sim t^2$, which is a faster ``ballistic" rate of spreading than the Einstein model predicts \citep{Sokolov2012}.
@@
 Some authors have modeled local and intermediate ranges of individual particle trajectories by writing approximate Newtonian equations for the dynamics of individual particles and integrating them numerically.
-Early efforts applied time-averaged fluid forces linked to the logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.
+Early efforts applied time-averaged fluid forces linked to a logarithmic mean flow velocity profile \citep{Yalin1963,VanRijn1984}, and later efforts included collision models to modify particle velocities upon bed contact \citep{Wiberg1985,Sekine1992,Nino1998}.
 
-Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep{Cundall1979a, Haff1993}.
-The early works utilizing this approach used a two dimensional domain with a highly simplified ``slab" flow model \citep{Haff1993}, while later works have included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep[e.g.][]{Cundall1979, Wachs2019}.
+The early works utilizing this approach used a two dimensional domain with a simplified ``slab" flow geometry \citep{Haff1993,Gotoh1997}, while later works included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
 
 The state of the art within this category of sediment transport models is to simulate both fluid and particle phases simultaneously.
-Particle-particle interactions are represented with the discrete element method, where particles are represented as spheres \citep{Cudnall1979} or polytopes \citep{Wachs2019} while forces are phrased as idealized spring-dashpot models.
-The fluid flow is modeled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
-A majority of these approaches simulate the fluid without including no-slip boundary conditions on the particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although works are beginning to include the fluid boundary conditions on individual particles with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
+Particle-particle interactions are represented with the discrete element method, while the fluid flow is modeled either by large eddy simulation or direct numerical simulation of the Navier-Stokes equations.
+A majority of these approaches simulate the fluid without including the no-slip boundary conditions on individual particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although some works have resolved these particle-scale fluid boundary conditions with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
 These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
@@
 
-Another category of models describe bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.
+Another category of models describes bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.
 
@@
 
-\citet{Fan2014} set out to describe exponenential-distributed bedload particle velocities ($u$) with the Langevin equation
+\citet{Fan2014} set out to describe exponenentially-distributed bedload particle velocities with the Langevin equation
 \be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
-This equation drives the particle velocity by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
-This drag is resisted by a heuristic particle friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
-The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from eq.
+This equation drives the particle velocity $u$ by a fluid drag $F + \sqrt{2D} \xi(t)$, where $F$ is a constant, $D$ is a diffusivity that characterizes the magnitude of drag fluctuations, and $\xi(t)$ is a Gaussian white noise with unit variance and vanishing mean \citep{Gardiner1983}.
+This drag is resisted by a heuristic Coulomb friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
+This model may have been inspired by models of Brownian motion with Coulomb friction in the physics literature \citep[e.g.][]{DeGennes2005,Touchette2010,Menzel2011}.
+The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from Eq.
 \ref{eq:fanlangevin} as \citep{Risken1989,VanKampen2007}
 \be \frac{\partial}{\partial t} P(u,t) = -\Delta\frac{\partial}{\partial u}\Big[ \text{sgn}(u) P \Big] + D \frac{\partial^2 P}{\partial u^2},\ee
-implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by eq. \ref{eq:fanlangevin} is
+implying that the steady-state velocity distribution ($\pt P(x,t) = 0$) provided by Eq. \ref{eq:fanlangevin} is
 \be P(u) = \frac{\Delta^2-F^2}{2\Delta D}\exp\Big(-\frac{-\Delta |u| + F u}{D}\Big).\ee
@@
 \be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
-where $U$ is the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance.
+where $U$ represents the mean velocity of particles, $D$ characterizes the magnitude of velocity fluctuations, and $\xi(t)$ is again a Gaussian white noise with vanishing mean and unit variance.
 The timescale $t_r$ is a relaxation time over which velocity fluctuations decay.
@@
 which is the Gaussian velocity distribution from the other subset of the experiments.
-These two models of bedload velocity distributions are summarized in fig. \ref{fig:fanAncey}.
+These two models of bedload velocity distributions are summarized in Fig. \ref{fig:fanAncey}.
 \begin{figure}[!htbp]
@@
 \end{figure}
-Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and other distributions have been reported besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
+Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and several experiments have revealed other distributions besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
 A comprehensive model of the full range of bedload transport observations remains a desirable target.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fluxDefinitions.png}
-	\caption{Particle fluxes can be defined with either a control volume $V$ or a control surface $S$.
+	\caption{Particle fluxes can be defined with a control volume $V$ or a control surface $S$.
 Here, particles are shown moving from time $t-\Delta t$ (faint white particles) to time $t$ (grey particles).
 Each particle's downstream displacement $\Delta x_i$ is indicated.
-These displacements define the downstream particle velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux eq.
-\ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition eq.
+These displacements define the downstream velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux Eq.
+\ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition Eq.
 \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_5)$,
@@
 \end{figure}
-Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has even been agreed upon despite over a century of research \citep{Ballio2018}.
-Today, there are two main complementary definitions of the bedload flux being applied in stochastic modeling. Other definitions are discussed comprehensively in \citet{Ballio2014,Ballio2018}.
+Perhaps a surprising summary of bedload transport research is that no one definition of the sediment flux has been agreed upon despite over a century of research \citep{Ballio2018}.
+Today, there are two main complementary definitions of the bedload flux being applied in stochastic modeling (but see \citet{Ballio2014,Ballio2018} for other definitions).
 
-The first definition is reminiscent of continuum mechanics and formulates the flux as a kind of current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016,Ballio2014}: 
+The first definition is reminiscent of continuum mechanics and formulates the flux as a current of sediment across a control surface $S$ \citep{Furbish2012,Heyman2016,Ballio2014}: 
 \be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
@@
 These quantities require careful interpretation because bedload particles do not constitute a continuous field \citep{Furbish2012a}.
-Typically one represents the concentration as a sum of indicator or Dirac delta functions \citep[e.g.][]{}.
+Typically one represents the concentration as a sum of indicator or Dirac delta functions \citep[cf.][Sec.
+8.5.1]{Gardiner1983}.
 
@@
 One of the most influential formulations of the bedload flux is due to \citet{Bagnold1956,Bagnold1966}, who derived a formula for the mean sediment flux using an energy balance approach.
-Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
+Bagnold understood sediment transport as a process which converts flow energy to heat via the effective friction \citep[e.g.][]{Bagnold1954} of grains against the bed as they move downstream through a succession of collisions \citep{Bagnold1973}.
 He assumed that the flow power $P_f$ available to move sediment in a volume scales as $P_f \propto \tau - \tau_c$, where $\tau$ is the average bed shear stress and $\tau_c$ is the threshold shear stress at which particles first begin to move.
-Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$required to sustain particle motion in a volume scales as $P_g \propto \Phi/\tau^{1/2}.$
+Considering that the average volumetric flux of particles is $\Phi$, and particles move with mean velocity proportional to the fluid velocity near the bed, Bagnold hypothesized that the power $P_g$ required to sustain particle motion in a volume scales as $P_g \propto \Phi/\tau^{1/2}.$
 Balancing flow energy against frictional dissipation ($P_f = P_g$) then provides Bagnold's sediment transport formula
 \be \Phi = k(\tau-\tau_c)\tau^{1/2}, \label{eq:bagnold}\ee
-which has shown good correspondence with laboratory data at large transport rates, given careful calibration of the constant factors $k$ and $\tau_c$.
+which has shown good correspondence with laboratory data provided the constant factors $k$ and $\tau_c$ are carefully calibrated.
 
@@
 Bagnold's formulation also faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}.
-The difficulties with Bagnold's approach led to many revisions of his theory, often based on more carefully incorporating the properties of individual particle motions into the sediment transport energy budget \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
+The difficulties with Bagnold's approach have motivated many revisions of his theory.
+These are often based on more carefully incorporating the properties of individual particle motions into the sediment transport energy budget \citep{Engelund1976,Luque1976,Nino1998,Martin2000}.
 
@@
 This formulation of the sediment flux now provides an alternative to Bagnold's scaling approach.
-The conceptual picture that Einstein considered is depicted in fig. \ref{fig:einsteinFluxConcept}.
+The conceptual picture that Einstein considered is depicted in Fig. \ref{fig:einsteinFluxConcept}.
  \begin{figure}[!htbp]
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
-	\caption{Einstein’s conceptual picture (modified from \citet{Yalin1972}). Particles move in discrete jumps of length $\ell$ from
+	\caption{Einstein’s conceptual picture, modified from \citet{Yalin1972}. Particles move in discrete jumps of length $\ell$ from
 left to right through an array of adjacent control volumes. The bedload flux is the rate of particles crossing
@@
 
-Einstein partitioned the channel into a sequence of identical control volumes $V$, and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
+Einstein partitioned the channel into a sequence of identical control volumes ($V$), and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
 Each control volume has downstream length $\ell$ which is also the average particle step length.
@@
 This implies that the probability that it entrains \textit{at least} $n$ times is $P_n = \sum_{i=n}^\infty  \frac{(k_E T)^i}{i!}e^{-k_E T}, $
-so Einstein's mean sediment flux across the control surface (eq. \ref{eq:einflux}) becomes
+so Einstein's mean sediment flux across the control surface (Eq. \ref{eq:einflux}) becomes
 \be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
@@
 Although Einstein and many followers have attempted to formulate these connections \citep[e.g.][]{Einstein1950,Grass1970,Paintal1971}, there are still no comprehensive solutions.
-Linking Einstein's entrainment and deposition rates to the grain scale mechanics remains an active area of research \citep[e.g.][]{Tregnaghi2012,Dey2018}.
+Linking Einstein's entrainment and deposition rates to the grain scale mechanics remains an active area of research that is far from complete \citep[e.g.][]{Tregnaghi2012,Dey2018}.
 
@@
 This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$.
-These model parameters are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
-This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its reliance on both probabilistic rates and scaling relations.
+These rates are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
+This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its incorporate of both probabilistic rates and scaling relations.
 
@@
 Equation \ref{eq:charru} indicates that the mean flux in steady transport and flow conditions is the implicit solution to the equation $E = D.$
-Using the scaling relations eqs. \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
+Using the scaling relations Eqs. \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
 \be \gamma = \frac{\alpha_1}{\alpha_2}\frac{\tau-\tau_c}{d^2 V_s^2}.\ee
@@
 \be q(x,t) = \int_0^\infty dx' \int_0^\infty dt' F(x',t') E(x-x',t-t'), \label{eq:furbo}\ee
-where $F(x,t)$ represents the probability density that just-entrained particles move at least a distance x in time $t$ before deposition.
+where $F(x,t)$ now represents the probability density that just-entrained particles move at least a distance x in time $t$ before deposition.
 In general, this approach can handle non-uniform conditions by including space and time dependence in $E$ and $F$.
@@
 \be F(x,t) = \int_x^\infty \delta(x-Vt)k_D\exp(-k_D t) dx = \theta(Vt-x)k_D\exp(-k_D t).\ee
-Considering uniform conditions with a density $\rho_s$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the areal entrainment rate can be expressed as $E=\rho_s k_E$, and eq.
+Considering uniform conditions with a density $\rho_s$ of particles available for motion on the bed surface, each having entrainment rate $k_E$, the areal entrainment rate can be expressed as $E=\rho_s k_E$, and Eq.
 \ref{eq:furbo} provides a mean flux
 \be q = \rho_s k_E \int_0^\infty dx' \int_0^\infty dt \theta(Vt-x)k_D\exp(-k_D t) = \rho_s k_E V/k_D. \label{eq:furbflux}\ee
-Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and eq.
+Since $1/k_D$ is the average time spent in motion, $V/k_D$ is the average step length $\ell$, and Eq.
 \ref{eq:furbflux} provides another perspective on Einstein's central result $q= E\ell$, but with the added ability to describe unsteady conditions \citep{Furbish2012}.
@@
 
-\citet{Exner1925} was probably the first to write a mathematical formula linking sediment transport to topographic change.
+\citet{Exner1925} was probably the first to write a mathematical formula linking sediment transport to topographic change in a river channel.
 He wrote
 \be (1-\phi)\frac{\partial z}{\partial t}(\textbf{x},t) = -\nabla q (\textbf{x},t). \label{eq:exner}\ee
-This equation links the temporal evolution of the land elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux $q$. The parameter $\phi$ is the bed porosity.
+This equation links the temporal evolution of the bed elevation $z$ at a location $\textbf{x}=(x,y)$ to spatial gradients in the sediment flux $q$. The parameter $\phi$ is the bed porosity.
 
@@
 \be \partial_x q(x,t) = E - D. \ee 
-In combination with eq. \ref{eq:exner}, this expression of the sediment flux gradient implies the relation
+In combination with Eq. \ref{eq:exner}, this expression of the sediment flux gradient implies the relation
 \be (1-\phi) \px z(x,t) = D - E\ee
-by which topographic evolution can be described in terms of Einstein's rates.
-In a nonlocal framework, as in eq. \ref{eq:furbo}, sediment deposition can be interpreted as the eventual result of entrainment from all upstream locations, giving
-\be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x',t')F(x',t') - E(x,t). \ee
-This ``entrainment form of the Exner equation" \citep{Nakagawa1976, Parker2000, Furbish2017} phrases topographic change from a stochastic interpretation of sediment transport.
+by which bed evolution can be described in terms of Einstein's rates.
+In a nonlocal framework, as in Eq. \ref{eq:furbo}, sediment deposition can be interpreted as the eventual result of entrainment from all upstream locations, giving
+\be (1-\phi) \px z(x,t) = \int_0^\infty dx' \int_0^\infty dt' E(x-x',t-t')F(x',t') - E(x,t). \ee
+This ``entrainment form of the Exner equation" \citep{Nakagawa1976, Parker2000, Furbish2017} relates channel morphodynamics to a probabilistic interpretation of sediment transport.
 
-The major limitation of these approaches are that they assume land surfaces are continuous.
-This can be interpreted as an average over the detailed configurations of individual grains \citep{Coleman2009}.
+The major limitation of these approaches are that they assume river beds are continuous.
+This can be justified as an average over the detailed configurations of individual bed grains \citep{Coleman2009}.
 This averaging process is expected to obscure the relevant morphological evolution processes whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
@@
 Recognition that sediment transport rates fluctuate requires careful interpretation of sediment transport measurements. 
-In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Zimmermann2008, Chartrand2018}, impact sensors \citep{Rickenmann2007, Mendes2016}, computer vision techniques \citep{Roseberry2012, Ancey2014}, or sediment traps \citep{Bunte2004, Papangelakis2016}.
+In practice, sediment transport measurements always involve time or space averaging, whether measurements are obtained by light tables \citep{Zimmermann2008, Chartrand2018}, impact sensors \citep{Rickenmann2007, Mendes2016}, computer vision \citep{Roseberry2012, Ancey2014}, or sediment traps \citep{Bunte2004, Papangelakis2016}.
 
-This averaging process in conjunction with transport fluctuations introduces scale-dependence to sediment transport measurements, whereby averaged characteristics depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
+This averaging process in conjunction with transport fluctuations introduces scale-dependence to sediment transport measurements, whereby mean sediment transport rates (and other statistical moments) depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
 A fundamental question raised by scale-dependence is how large averaging windows must be for sediment flux measurements to converge.
@@
 
-To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
+To obtain realistically-wide fluctuations in particle activity, Ancey et al.
+introduced a positive feedback they called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
 
@@
 They formulated the probability distribution $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
-\begin{multline} \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) \\ - [n+1]\alpha P(n+1,t) - \big[\lambda + n (\alpha + \mu)\big]P(n,t),\end{multline}
-whose terms describe particle birth (entrainment and migration in) at rate $\lambda$, collective entrainment at rate $\mu$, and particle death (deposition and migration out) at rate $\alpha$.
+\begin{multline} \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) \\ - [n+1]\alpha P(n+1,t) - \big[\lambda + n (\alpha + \mu)\big]P(n,t).\end{multline}
+The terms on the right hand side of this equation describe respectively particle birth (entrainment and migration in) at rate $\lambda$, collective entrainment at rate $\mu$, and particle death (deposition and migration out) at rate $\alpha$.
 The final term encodes the possibility that $n$ stays constant.
@@
 
-From eq. \ref{eq:negbin}, the mean number of moving particles is 
-\be \langle n \rangle = \lambda/(\alpha-\mu),\ee
+From Eq. \ref{eq:negbin}, the mean number of moving particles is 
+\be \langle n \rangle =\frac{ \lambda}{\alpha-\mu},\ee
  and the variance is
 \be \sigma_n^2 = \frac{\lambda \alpha}{(\alpha-\mu)^2}. \ee
-Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n/\bra n \ket = \sqrt{\alpha/\lambda}$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, as the distribution eq.
-\ref{eq:negbin} limits to a Poisson distribution where the mean and variance are equal \citep{Ancey2006}.
+Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n^2/\bra n \ket = \alpha/(\alpha-\mu)$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to $\sigma_n^2/\bra n\ket = 1$, as the distribution Eq.
+\ref{eq:negbin} limits to a Poisson distribution where the mean and variance are equal \citep[e.g.][]{Ancey2006}.
 
-The probability distribution of the bedload flux can be computed by the control volume form eq.
+The probability distribution of the bedload flux can be computed by the control volume form Eq.
 \ref{eq:controlflux} given the probability distribution of particle velocites $P(u)$ under the assumption that velocity and activity are independent.
-Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep{Ancey2014}
+Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep{Ancey2020a}
 \be P(\Phi) = L \sum_{k=0}^\infty P_k(L\Phi) P(n=k). \ee
@@
 This is observation-scale dependence.
-The nature of this convergence depends on the particle dynamics in an as yet unspecified way.
+The properties of this convergence depend on the particle dynamics in a manner which has not yet been specified.
 }
@@
 The exponential inter-arrival time distribution is $P(t) = \Lambda \exp(-\Lambda t),$ so the rate of particle arrivals is $\Lambda$.
-In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model of sec. \ref{sec:einwalk}:
+In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model reviewed in Sec. \ref{sec:einwalk}:
 \be q(T) = \frac{1}{T}\int_0^T dt' \mu(t'). \label{eq:flucflux}\ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
-as indicated in fig. \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$. 
-The flux in eq.
-\ref{eq:flucflux} is a random variable as indicated in fig.
-\ref{fig:ancey} panel (b).
+as indicated in Fig. \ref{fig:ancey} (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$ -- $P(N)=(\Lambda t)^N/N! \exp(-\Lambda t).$
+The flux in Eq.
+\ref{eq:flucflux} is a random variable as indicated in Fig.
+\ref{fig:ancey} (b).
 Its probability distribution $P(q|T)$, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
-This equation entails an average over all possible realizations of the noise in eq. \ref{eq:poissfluxnoise}, producing \citep{VanKampen2007}
+This equation entails an average over all possible realizations of the noise in Eq. \ref{eq:poissfluxnoise}, producing \citep{VanKampen2007}
 \be P(q|T) = \sum_{l=0}^\infty \frac{\big(\Lambda T\big)^l}{l!}e^{-\Lambda T}\delta(q-\frac{l}{T}),\ee
@@
 All of the works reviewed so far are either mean field models that do not include fluctuations, or stochastic models that calculate probability distributions for the quantities of interest.
-The common thread of all of these models is that they take advantage of various devices to side-step the complex interplay of turbulence and granular physics which ultimately control bedload transport.
+The common thread of all of these models is that they take advantage of various devices to side-step the complex interplay of turbulence and granular physics which ultimately controls bedload transport.
 
-In the case of mean field models, the devices are heuristic scaling arguments and semi-empirical formulas, while in the case of stochastic models, they are simplified dynamical equations driven by idealized noises, like sequences of pulses, alternating switches, and erratic fluctuations.
+In the case of mean field models, the devices are heuristic scaling arguments and semi-empirical formulas, while in the case of stochastic models, they are phenomenological equations driven by idealized noises, like sequences of pulses, alternating switches, and erratic fluctuations.
 This thesis applies the latter methodology to many of the limitations highlighted in the review. An outline of the problems to be addressed follows below.
@@
 The chapter also demonstrates how these trajectories can be used to calculate the scale-dependent probability distribution of the bedload flux.
-This joins together and extends studies from \citet{Einstein1937} to \citet{Ancey2020b} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
+This joins together and extends studies from \citet{Einstein1937} to \citet{Ancey2020a} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
 
@@
 
-Bed elevation changes produce sediment burial, and although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2002,Hassan2017}, our understanding of how long sediment can remain buried is limited.
+Bed elevation changes imply sediment burial, and although sediment burial is known to affect tracer particle motions at geomorphic timescales \citep{Ferguson2002,Hassan2017}, our understanding of how long sediment can remain buried is limited.
 
@@
 
-Modeling sediment transport as an alternating sequence of motion and rest intervals implicitly assumes that particles are either moving or resting on the bed surface.
+Modeling sediment transport as an alternating sequence of motion and rest intervals implies particles are either moving or resting on the bed surface.
 Although this provides a useful description of bedload transport across local and intermediate timescales, it requires modification to describe longer global and geomorphic timescales.
@@
 Finally, the Langevin models of \citet{Fan2014} and \citet{Ancey2014} describe the velocity distributions of sediment moving downstream in the local and intermediate ranges between entrainment and deposition.
-These models produce two different end-member sediment velocity distributions, and no mathematical models have been developed that describe the full range of distributions observed in experiments.
+These models produce two different end-member sediment velocity distributions, but no mathematical models have been developed that describe the full range of distributions observed in experiments.
 
@@
 The improved model is demonstrated to describe all of the different bedload velocity distributions which have been reported in experiments.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\endinput
-
-An alternative statistical formulation of the sediment flux provides an approximate correspondence between these two definitions \citep{Ancey2006,Furbish2012}.
-This hinges on the probability $P[\textbf{u}_p | \textbf{x},t]$ that a particle contacts a control surface $S$ at position $\textbf{x}$ and time $t$ with velocity $\textbf{u}_p$. 
-This conditional probability is considered to result from a very large collection of identical systems selected at random moments in their evolution: that is, the conditional probability is an ensemble quantity \citep{Kittel1958}.
-
-In terms of this ensemble probability, the flux is \citep{Ancey2006}: 
-\be q_s = \int_{\text{all } \textbf{u}_p} \int_S P[\textbf{u}_p | \textbf{x},t] \textbf{u}_p \cdot d \textbf{S} d \textbf{u}_p . \label{eq:anceyflux} \ee
-In steady conditions, when the probability is independent of time -- $\partial P / \partial t = 0$.
-
-An approximate connection to a control volume flux can be derived by swapping the ensemble average for a spatial average by integrating along a control volume.  
-If the control volume is sufficiently long, $L \rightarrow \infty$, it can be seen as a stack of very many independent cross sectional surfaces. 
-These comprise a stack of replicas of $\mathcal{S}$, and at an instant, if spatial correlations in bedload transport are weak, each surface constitutes one configuration of particles intersecting $S$ with some set of positions and velocities contributing to $P$.
-One can then define $P$ by counting occurrences along this stack of replica surfaces with an integral across the control volume.
-
-With this concept of replica surfaces in mind, we can formalize the link with symbols. 
-Consider $n$ particles distributed throughout a volume at some set of positions $\textbf{x}_i$ with some set of velocities $\textbf{u}_i$, where $i=0,1,\dots, n$,
-The cloud of particles can be represented by its (discrete) density in a position-velocity phase space: 
-\be \rho(\textbf{x},\textbf{u}_p) = \sum_{i=1}^n M(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i). \ee  
-Here $M(\textbf{z})$ is a marker function, which is $1$ if $\textbf{z}$ is inside the particle, and $0$ otherwise.
-The marker's volume integral is $\int M(\textbf{x}) dV = \nu_p$, the particle volume.
-
-Using this density at sufficiently large $L$, the conditional probability $P[\textbf{u}_p | \textbf{x}]$ can be written approximately as an integral over the stack of replica surfaces:
-\be P[\textbf{u}_p|\textbf{x}] \approx \frac{1}{L} \int_0^L dx \rho(\textbf{x},\textbf{u}_p). \ee
-The integral runs along the downstream coordinate, and the equality becomes exact as $L \rightarrow \infty$.
-With this swap of ensemble averaging for spatial averaging, the bedload flux becomes 
-\be q_s \approx \int_{\text{all} \textbf{u}_p} \int_S \frac{1}{L} \int_0^L dx \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i) \textbf{u}_p \cdot \textbf{k} dS d \textbf{u}_p = \frac{\nu_p}{L}\sum_{i=1}^n u_i , \ee
-so there is some correspondence between the surface and volume definitions of the flux.
-Yet this is never exact, except in the unphysical case of sediment transport with no spatial correlations (vanishing particle size, discontinuous trajectories).
-Apparently, the control volume and control surface formulations of the sediment flux are not equivalent, so we see in the literature two parallel streams of research.
-
-\subsection{First extensions of the Einstein model}
-\label{
-
-Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest.
-Despite its success, it oversimplifies particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
-
-Einstein's model of particle movement seems to have been overlooked for the first several decades after its development, or possibly overshadowed by Einstein's later work, which uses its essential ideas for more obviously practical purposes.
-Eventually, curiosity into individual particle motions was re-sparked by cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}.
-In relation to these issues, Einstein's model received a great deal of new attention which generated some generalizations of the model.
-
-The first set of modifications to Einstein's approach in this period concerned the form of the distributions used for step lengths and resting times.
-Einsteins flume experiments demonstrated that particle step lengths followed exponential distributions \citep{1937}, and this was confirmed by subsequent studies \citep{Yano1969, Nakagawa1976}.
-Yet these studies were all conducted in steady flows in artificial flumes, not real streams, where sediment transport conditions can be different.
-When sediment transport is observed before and after a flood, rather than during steady flow, the step lengths are better described by a Gamma distribution, this being the sum across the numerous individual steps that occurred during the flood \citep{Hassan1991}.
-When sediment transport occurs in the presence of dunes or other bedforms, similar modifications have been made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965} and the embedding of particles within bedforms (which increases the periods of rest) \citep{Yang1971,Nakagawa1980}.
-
-A full statistical characterization of sediment transport is desirable because transport fluctuations are responsible for the initiation of bedforms \citep{Jerolmack2005,Bohorquez2016} and are beginning to show links to basic geomorphology considerations like the maintenance of stable channel widths \citep{Abramian2019,Abramian2020}.
-This is not to mention that because sediment transport fluctuations are ubiquitous, models which describe only mean bedload fluxes signals are conceptually incomplete.
-
-An additional reason to characterize fluctuations in bed load fluxes is because in reality, sediment flux measurements always involve a sampling interval.
-This introduces questions of the convergence of transport rate measurements \citep{Dhont2018,Ancey2020}.
-Fluctuations originate from processes which span a vast range of characteristic timescales.
-These can occur over a matter of seconds, such as differences in the velocities of grains due to a series of collisions \citep{Benavides2021}, to many hundreds of hours, such as the migrations of dunes and bedload sheets \citep{Hammamori1962,Hoey1992,Guala2014} or the cycling between aggradation and degradation in pools \citep{Dhont2018}.
-Such processes introduce temporal correlations into sediment transport signals which control the sampling interval over which measurements of the mean sediment flux converge \citep{Saletti2015,Singh2009,Singh2012}.
-To date, very few modeling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
-
-These models produce successful descriptions of disparate experimental results, although their components are not clearly interpretable in terms of physical expecatations.
-For example, granular interactions were included in the Fan et al model as a Coulomb friction term.
-While such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo intermittent collisions with the granular bed.
-Although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
-In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{Clift1978}.
-We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
-
-One overarching methodology is the use of idealized noises in combination, where each represents a separate mechanical process affecting the movements of particles.
-This takes on more mathematical complexity than earlier studies, but it also introduces more realism.
-
-Regarding the movements of individual particles, the division into local, intermediate, global, and geomorphic ranges by \citet{Nikora2001,Nikora2002} and \citet{Hassan2017} provides a conceptual scheme by which to organize the current state of research.
-Einstein's original description \citep{Einstein1937} and those of his early followers \citep{} described only the global range associated with timescales over which particles cycle through motion and rest.
-This limitation was overcome in later works which included a finite duration of sediment motion in order to describe both intermediate and global scales, and Newtonian models have also attained descriptions of additional scales, such as local and intermediate \citep{}.
-Meanwhile, Langevin models have derived the velocity characteristics of particles within the local range \citep{}, although they do not manage to account for the full range of experimentally-observed distributions.
-The major shortcoming exposed by this framing in terms of the timescales of particle motion is that no models have yet resolved particles across the full set of timescales, so all of these descriptions are essentially incomplete.
-
-Models of the sediment flux face a similar set of shortcomings.
-Einstein and Bagnold founded complementary perspectives of the sediment flux.
-Both were mean field theories, but one counted particles crossing a surface using probabilistic arguments, while the other considered an average energy balance in a control volume.
-Research has come a long way since these studies, but the essential divide between surface and volume-based flux calculations still exists. These definitions are not equivalent.
-The erosion-deposition model takes steps toward joining these two mean-field theories into a common framework, though it involves many parameters which must be experimentally calibrated.
-A paradigm shift in this research stream came from the recognitition that the sediment flux is a fluctuating quantity best characterized by a probability distribution, and from the acknowledgement that measured values of the sediment flux depend on observation scale, meaning it is a scale-dependent quantity.
-Although simple models of the sediment flux probability distribution have been developed, they stand somewhat apart from the detailed dynamics of individual particles.
-Rather than describing particles moving downstream through cycles of entrainment and deposition as in \citet{Einstein1937}, these methods characterize particle motions by herustic entrainment and deposition rates, or interarrival time distributions, which are somewhat abstract objects, not concretely related.
-Despite the considerable progress these approaches have brought, the lack of connection to the underlying particle mechanics requires attention.
-
-Alongside this central result $q=E\ell$ for the mean flux, Einstein formulated the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle, building insight into the grain-scale mechanics of entrainment.
-His original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
-The entrainment rate is calculated from the exceedance probability of the turbulent lift over the weight, providing an alternative to the critical shear stress concept which explicitly incorporates turbulent fluctuations in the fluid flow.
-This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
-random) resisting quantities \citep{Grass1970} another of Einstein's key contributions.
-\citet{Paintal1971} made a significant extension
-by including the random supporting geometry of bed particles into the force balance.
-More recently, \citet{Tregnaghi2012}
-amended the theory to include both force magnitude and duration, incorporating the recognition that both are relevant for particle entrainment \citep{Diplas2008, Valyrakis2013, Celik2014}.
-Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.

### 2021-09-24 revision/ch2flipFlop.tex
getting close.

@@
 Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006,Furbish2012a}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
-Because these fluctuations occur over disparate time scales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
+Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
 To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
@@
 This distribution exhibits observation-scale dependence as a result of velocity fluctuations among moving particles.
-Below, I develop the new formalism in sec.
-\ref{sec:mod}, solve it in sec.
-\ref{sec:res}, and discuss the implications of my results and future research ideas in secs.
+Below, I develop the new formalism in Sec.
+\ref{sec:mod}, solve it in Sec.
+\ref{sec:res}, and discuss the implications of my results and future research ideas in Secs.
 \ref{sec:disc} and \ref{sec:conc}.
@@
 
-This probability distribution of position is formed as $ P(x',t) = \big\bra \delta(x'-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of eq.
+This probability distribution of position is formed as $ P(x',t) = \big\bra \delta(x'-x(t))\big\ket_{\eta,\xi} $, where $x(t)$ is the formal solution of Eq.
 \ref{eq:flippylangevin} and the average is over both noises.
 This symbolic equation is not directly useful as taking averages over both noises is a challenging mathematical problem \citep[e.g.][]{Hanggi1978}.
-This calculation is completed in the appendix sec.
+This calculation is completed in the appendix Sec.
 \ref{sec:appAmaster}, providing the master equation
@@
 
-One can see in particular that taking the entrainment rate $k_E$ very large, meaning that particles are very often moving, implies a classical advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations \citep{Ancey2014}.
-Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process is interrupted, giving rise to the additional terms in eq.
-\ref{eq:flippymaster} and providing an asymmetrical structure reminiscent of equation \ref{eq:lislemaster} from the simpler motion-rest models reviewed before.
+One can see in particular that taking the entrainment rate $k_E$ very large, meaning that particles are very often moving, implies a classical advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations.
+Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process is interrupted, giving rise to the additional terms in Eq.
+\ref{eq:flippymaster} and providing an asymmetrical structure reminiscent of equation \ref{eq:lislemaster} from the simpler motion-rest models of Sec.
+\ref{sec:lisle}.
 
@@
 
-The probability distribution of the sediment flux can be calculated using the probability distribution of particle position $P(x,t)$ provided as the solution of equation \ref{eq:flippymaster}.
+The probability distribution of the sediment flux can be calculated using the probability distribution of particle position $P(x,t)$ derived as the solution of Eq. \ref{eq:flippymaster}.
 This method is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
-This generalizes the nonlocal formulation of sec.
-\ref{sec:nonlocal} to provide the probability distribution of the flux, and it reframes the renewal approach (sec.
+This generalizes the nonlocal formulation of Sec.
+\ref{sec:nonlocal} to provide the probability distribution of the flux, and it reframes the renewal approach (Sec.
 \ref{sec:renewal}) in terms of the mechanics of individual particles in transport.
@@
 Later, the number of particles $N$ and the size $L$ of the domain will be extended to infinity such that their ratio $\rho=N/L$ remains constant. 
-This limit will provide a configuration similar to the one constructed in the nonlocal formation (sec. \ref{sec:nonlocal}).
+This limit will provide a configuration similar to the one constructed in the nonlocal formation (Sec. \ref{sec:nonlocal}).
 
@@
 	\centerline{\includegraphics{figures/ch2/figure2_slopeKey.pdf}}
-	\caption{Panel (a) indicates the probability distribution of particle position (eq.
+	\caption{Panel (a) indicates the probability distribution of particle position (Eq.
 \ref{eq:monster}) as it evolves through time.
@@
 In panel (a), the initial position persists as a Delta-function spike for the black $t=10\tau$ curve.
-Panel (b) shows the resulting particle diffusion (eq.
+Panel (b) shows the resulting particle diffusion (Eq.
 \ref{eq:diffo}).
-At time scales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process.
-For larger time scales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
+At timescales $t \ll 2D/V^2$, the diffusion is normal since the movement is approximately a standard Brownian diffusion process.
+For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
 Finally at times longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
@@
 Note in the physical condition when $k\approx k_D$, which is generally satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the above condition for existence of three ranges becomes $ \Pe \gg 1,$ where
-$\Pe =V^2/(2k_D D)$ is a Peclet number \citep[e.g.][]{Heyman2014}. This Peclet number measures the relative importance of advection and diffusion to the downstream movements of particles.
+$\Pe =V^2/(2k_D D)$ is a P\'{e}clet number \citep[e.g.][]{Heyman2014}. This P\'{e}clet number measures the relative importance of advection and diffusion to the downstream movements of particles.
 
-The large Peclet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles rarely move upstream any appreciable distance \citep[e.g.][]{Fathel2015}.
+The large P\'{e}clet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles rarely move upstream any appreciable distance \citep[e.g.][]{Fathel2015}.
 This limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed \citep[cf.][]{Nikora2001,Nikora2002}, at least within the assumptions of the model.
-Figure \ref{fig:flippyfig1} panel (b) sketches the predicted three-range diffusion, which is discussed further in sec. \ref{sec:prollems}.
+Figure \ref{fig:flippyfig1} panel (b) sketches the predicted three-range diffusion, which is discussed further in Sec. \ref{sec:prollems}.
 
@@
 
-From the formalism in sec. \ref{sec:flippyflux}, the central parameter of the sediment flux distribution is the rate function
+From the formalism in Sec. \ref{sec:flippyflux}, the central parameter of the sediment flux distribution is the rate function
 \be \Lambda(t) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,t). \label{eq:flatform}\ee
@@
 
-Using the probability distribution of position eq. \ref{eq:monster}, the integrals in eq. \ref{eq:flatform} are performed in sec. \ref{sec:fluxconstant} of the appendix, providing the rate function
+Using the probability distribution of position Eq. \ref{eq:monster}, the integrals in Eq. \ref{eq:flatform} are performed in Sec. \ref{sec:fluxconstant} of the appendix, providing the rate function
 \begin{multline} 
@@
 
-This is an intricate result for the rate function in the sediment flux distribution eq.
+This is an intricate result for the rate function in the sediment flux distribution Eq.
 \ref{eq:flippydist}.
-This mathematical complexity may not be surprising given that eq.
+This mathematical complexity may not be surprising given that Eq.
 \ref{eq:flippylangevin} involves two interacting diffusion processes.
-As displayed in fig.
-\ref{fig:fluxconvergence}, as a result of eq.
+As displayed in Fig.
+\ref{fig:fluxconvergence}, as a result of Eq.
 \ref{eq:fluxy} the mean sediment flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $q_0 = \rho k_E V/k = E \ell$ as the observation time becomes much larger than $1/(k_D \Pe)$.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{figures/ch2/figure3_slopeKey.pdf}
-	\caption{The mean sediment flux is plotted for different values of the Peclet number $\Pe = V^2/(2k_D D)$, characterizing the relative strength of particle velocity fluctuations during motion.
-The flux is normalized by the prediction $q_0 = E \ell$ of the Einstein theory (sec.
+	\caption{The mean sediment flux is plotted for different values of the P\'{e}clet number $\Pe = V^2/(2k_D D)$, characterizing the relative strength of particle velocity fluctuations during motion.
+The flux is normalized by the prediction $q_0 = E \ell$ of the Einstein theory (Sec.
 \ref{sec:einflux}).
@@
 Stronger velocity fluctuations (smaller $\Pe$) slow the convergence to this limit.
-	In all cases, satisfactory convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by eq.
+	In all cases, satisfactory convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by Eq.
 \ref{eq:fluccolimitcases}.
-Plotted lines are the analytical result eq.
+Plotted lines are the analytical result Eq.
 \ref{eq:fluxy}, while the points are the results of Monte Carlo simulations.}
@@
 Further insight into the rate constant of the sediment flux distribution can be gained by investigating extreme cases of the observation time.
-As shown in appendix sec. \ref{sec:fluxlimits}, eq. \ref{eq:fluxy} takes on relatively simple forms at extreme values of $T$: 
+As shown in appendix Sec. \ref{sec:fluxlimits}, Eq. \ref{eq:fluxy} takes on relatively simple forms at extreme values of $T$: 
 \be \Lambda(T) =
@@
 Since the mean flux is $\bra q(T) \ket = \Lambda(T)/T$, the condition on the observation time so that the flux converges to the Einstein value $q_0 = \rho k_E V/k = E \ell$ can be expressed as $T\gg (k_D \Pe)^{-1}$.
-It is related to the Peclet number and is proportional to the time particles spend in motion ($1/k_D$).
-As a result, when particle velocity fluctuations during motions become large comparable to the mean particle velocity, eq.
+It is related to the P\'{e}clet number and is proportional to the time particles spend in motion ($1/k_D$).
+As a result, when particle velocity fluctuations during motions become large comparable to the mean particle velocity, Eq.
 \ref{eq:fluccolimitcases} indicates that the flux becomes slow to converge to the Einstein result.
@@
 
-Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches (sec.
+Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches (Sec.
 \ref{sec:renewal}) more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
-This method can also be viewed as a generalization of the nonlocal formulation (sec.
-\ref{sec:nonlocal}) developed by \citet{Parker2000,Furbish2012a} to describe sediment flux probability distributions, rather than just mean values.
+This method can also be viewed as a generalization of the nonlocal formulation (Sec. \ref{sec:nonlocal}) to describe sediment flux probability distributions, rather than just mean values.
 
@@
 
-The sediment flux probability distribution in eq. \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
+The sediment flux probability distribution in Eq. \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
 Poisson distributions have relatively thin tails, meaning their fluctuations are typically small \citep{Ancey2006}.
-In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux time series \citep{Singh2009, Heyman2013,Benavides2021}.
+In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
 This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider transport rate fluctuations.
@@
 Therefore particle-particle interactions seem more plausible a mechanism for collective entrainment.
-In particular, contact forces within granular media are known to organize into ``force chains", where a minority of contacts provide a majority of the granular material's strength \citep[e.g.][]{Radjai1996, Azema2012}.
-Fluid entrainment of a key particle within a near-surface force chain could entail a collective release of the particles it stabilizes, consistent with current understanding of collective entrainment.
+In particular, contact forces within granular media are known to organize into ``force chains", where a minority of contacts provide a majority of the strength \citep[e.g.][]{Radjai1996, Azema2012}.
+Fluid entrainment of a key particle within a near-surface force chain could entail a collective release of the particles stabilized by this chain, consistent with current understanding of collective entrainment.
 
 Such particle-particle interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here, but the basic structure can be sketched.
-The dynamical eq.
+The dynamical Eq.
 \ref{eq:flippylangevin} would generalize to a stochastic Newtonian form $\ddot{x}_i(t) = F_i(x_i,\dot{x}_i, t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving forces unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
-This forcing structure would alternate between moving and resting modes by a dichotomous process \citep[e.g.][]{Bena2006}.
+As an approximation, this forcing structure might alternate between moving and resting modes by a dichotomous process \citep[e.g.][]{Bena2006}.
 
-The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, granular gases \citep{Brilliantov2004}, or other interacting particle models available in physics literature \citep{Hernandez2004,Escaff2018}.
+The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, granular gases \citep{Brilliantov2004}, or other interacting particle systems available in physics literature \citep{Hernandez2004,Escaff2018}.
 
@@
 Fig.
-\ref{fig:flippyfig1} and eq.
+\ref{fig:flippyfig1} and Eq.
 \ref{eq:diffo} indicate that bedload sediment particles described by the model spread apart with a rate which depends upon observation scale, transitioning through three different ranges consistent with the concept of \citet{Nikora2001a,Nikora2002}.
-Nikora et al. originally proposed that sediment diffusion would be super-diffusive at short (local) timescales, normal at intermediate timescales, and sub-diffusive at long (global) timescales.
-Here, the short range diffusion is normal, not super-diffusive, indicating a disprepancy between the Nikora et al. concept and the present model.
+Nikora et al. originally proposed that sediment diffusion would be super-diffusive at local timescales, normal at intermediate timescales, and sub-diffusive at global timescales.
+Here, the local range diffusion is normal, not super-diffusive, indicating a disprepancy between the Nikora et al. concept and the present model.
 
-This discrepancy could originate from the representation of velocity fluctuations in eq. \ref{eq:flippylangevin} as Gaussian noise with a vanishing correlation time.
+This discrepancy could originate from the representation of velocity fluctuations in Eq. \ref{eq:flippylangevin} as Gaussian noise with a vanishing correlation time.
 In actuality, particle velocities are temporally correlated, and this will modify the diffusion characteristics on timescales comparable to the correlation time.
-The simplest modification of eq.
+The simplest modification of Eq.
 \ref{eq:flippylangevin} to include velocity correlations would be to replace the Gaussian white noise with an Ornstein-Uhlenbeck noise \citep[e.g.][]{Luczka2005,Hanggi2007}.
@@
 \citet{Campagnol2015} showed from experimental data that particle ``unsteadiness" during entrainment and deposition can give rise to diffusion exponents even larger than that for ballistic motion.
-Likely, the representation of entrainment and deposition with an instantaneous alternation obscures the short-timescale physics of these processes \citep[e.g.][]{Valyrakis2010,Celik2014}.
+Likely, the representation of entrainment and deposition with an instantaneous alternation obscures the local-timescale physics of these processes \citep[e.g.][]{Valyrakis2010,Celik2014}.
 The simplest modification to include the timescales of entrainment and deposition would introduce a uniform accleration (deceleration) during entrainment (deposition), in effect ``tilting" the sharp motion-rest alternation represented by dichotomous noise in this chapter's model.
-This acceleration could be written $V/\tau_E$, where $\tau_E$ is an ``exchange time", as originally introduced by \citep[cf.][]{Einstein1950}.
-The relatively lower velocity of particles during the unsteady acceleration phase induced by entrainment and deposition can be expected modify the predicted short-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
+The relatively lower velocity of particles during the unsteady acceleration phase induced by entrainment and deposition should modify the predicted local-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
 
@@
 
-Fig. \ref{fig:fluxconvergence} indicates that the sediment flux described by eq. \ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q = E\ell$ predicted by the Einstein theory.
+Fig. \ref{fig:fluxconvergence} indicates that the sediment flux described by Eq. \ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q = E\ell$ predicted by the Einstein theory.
 Because particles move for durations $1/k_D$ on average, the typical variance of velocity among moving particles can be written $\sigma_V^2 = 2D/k_D$.
@@
 
-The Exner equation of sec.
-\ref{landscape} used to describe channel evolution is derived by evaluating mass balance within a control volume \citep[e.g.][]{Coleman2009}.
+The Exner equation of Sec.
+\ref{sec:landscape} used to describe channel evolution is derived by evaluating mass balance within a control volume \citep[e.g.][]{Coleman2009}.
 When the sediment flux changes values for observation scales $t < T$, as in this chapter, how the channel evolution predicted by the Exner equation is affected remains an open question.
@@
 Presumably, the inclusion of a scale-dependent stochastic transport rate generates a stochastic bed elevation field $h(x)$, generating a probability distribution of field configurations $P(h(x),t)$ which depends on the observation scale.
-The work of \citet{Jerolmack2005} evaluates a stochastic sediment transport rate within the Exner equation, although they did not investigate the probability distribution of bed configurations or its scale dependence.
-Mathematical tools to investigate such a stochastic Exner equation can be found in works on surface growth processes and polymer physics, where the evolution of random fields (like $h(x)$) is common topic of inquiry \citep[e.g.][]{Kawakatsu2001,Barabasi1995, Kardar2007}.
+\citet{Jerolmack2005} includes a stochastic sediment transport rate within the Exner equation, although they only investigated particular realizations of $h(x)$, not the probability distribution of bed configurations or its scale dependence.
+Mathematical tools to investigate such a stochastic Exner equation can be found in works on surface growth processes and polymer physics, where the evolution of random fields is common topic of inquiry \citep[e.g.][]{Kawakatsu2001,Barabasi1995, Kardar2007}.
 
@@
 The probability distribution of the bedload sediment flux was calculated from these particle dynamics, and the resulting flux distribution was demonstrated to adopt scale-dependence from the underlying dynamics of individual particles.
-The observation time scales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations, characterized by a Peclet number.
+The observation timescales over which sediment flux observations converge were expressed in terms of the mean particle movement velocity and the typical magnitude of its fluctuations, characterized by a P\'{e}clet number.
 
@@
 The next step is to include collective motions to produce wider sediment transport fluctuations.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\endinput
-Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
-The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1960}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
-Progress in stochastic approaches to river science \citep[e.g.][]{Furbish2021c,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask again with these classic works to what extent fluctuations shape Earth's surface.
-
-This chapter upscaled from individual particle motions to calculate the sediment flux.
-We can imagine going further, to upscale from the sediment flux to understand channel morphodynamics.
-The sediment flux is stochastic as a result of variability in the individual particle trajectories forming it.
-As a result, the channel morphodynamics should also adopt this stochasticity, although very few studies have evaluated this possibility \citep{Bohorquez2016, Jerolmack2005}.
-
-Earlier approaches to calculate the probability distribution of the sediment flux have counted the fluctuating number of particles in a control volume with population models \citep{Ancey2008,Ancey2014}, or they have evaluated the average rate of particles crossing a control surface using the interarrival time distribution of particles from upstream \citep{Turowski2010,Ancey2020,Heyman2013}, in what is basically an application of renewal theory \citep{Cox1962}.
-These approaches can be described as ``kinematic", in that they do not explicitly account for the trajectories of individual particles as they move downstream.
-
-Here, I have demonstrated how to describe the sediment flux probability distribution from a dynamical perspective, rather than a kinematic one, using stochastic dynamical equations to describe the trajectories of individual grains (eq.
-\ref{eq:langevin}.
-In effect, the interarrival time distribution is replaced by the probability distributions of position for all individual particles available for motion, with their initial positions and velocities considered as random variables.
-Particles moving downstream are counted as they cross the control surface, giving rise to a parameter $\Lambda(T)$, dependent on observation scale, which is analogous to the heuristic rate constant in the renewal theories that were reviewed in section \ref{sec:renewal} \citep{Turowski2010,Ancey2020}
-This new description of the sediment flux from the grain scale particle dynamics, rather than characteristic kinematics derived from them, represents considerable progress toward a mechanistic understanding of sediment transport from statistical physics.
-
-\subsection{Outlook and future research}
-
-The phrasing of sediment fluxes the topographies they moderate as statistical quantities, having mean configurations and fluctuations, provides some context on the severe challenges encountered in predicting landscape change with deterministic equations \citep{}.
-What we observe, looking at Earth's landscapes, may be a single realization, drawn from a spectrum of possibilities, whose precise form evades description. 
-The stochastic perspective allows us to
-
-The sediment flux probability distribution in equation \ref{eq:fluxdist} represents a Poisson distribution with an observation-scale dependent rate.
-Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
-This contrasts with many bedload transport experiments with wide-tailed fluctuations \citep{Ancey2008,Heyman2016,Fathel2015}, especially for weak flow conditions near initiation of particle motion \citep{Benavides2021}, exposing a need to generalize a mechanistic theory of the sediment flux developed here to provide wider fluctuations.
-
-There are essentially two mechanisms available by which wide fluctuations can arise \citep{Goh2008}.
-Both would be ultimately reflected in non-exponential interarrival times of particles \citep[e.g.][]{Turowski2010,Heyman2013}, although their origins would be unambigiously linked to the underlying particle transport mechanics.
-The first possibility is to introduce correlations between particle movements \citep{Ancey2008,Heyman2013,Ancey2014}, such as might arise when turbulent flow fluctuations entrain many particles at once \citep{Cameron2020} or drive waves of high velocity particles downstream simultaneously \citep{Nino1996}.
-In a model like the one I presented here, this first mechanism for wide fluctuations could be described by temporally unsteady velocities, entrainment or deposition rates, or diffusion coefficients in the Langevin model (eq.
-\ref{eq:langevin}).
-The second possibilty is to introduce interactions between moving particles, such as might result from collisions \citep{Lee2018} or wake vortices \citep{Schmeeckle2014} inducing particle entrainment, leading to collective waves of motion \citep{Ancey2014}.
-Including such effects would require a many-body generalization of the Langevin and master equation \ref{eq:langevin} and \ref{eq:master}, where the probability distribution of position is promoted to a joint form $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$, which is puzzling, since the number of particles $N(t)$ in the distribution fluctuates.
-The model of \citet{Ancey2014} hints at this possibility, although its starting point is not a dynamical equation as in eq .\ref{eq:master}.
-Some hints as to how to generalize our work to include interactions between particles to describe wide-tailed sediment fluxes can be gleaned from literature on reaction-diffusion processes \citep{Hernandez2004,Cardy2006} and interacting gases \citep{Chapman1970,Brilliantov2004}.
-The further inclusion of interacting particle models into bedload transport theory would be an exciting development.
-
-\endinput
-
-A simpler approach is to conduct the averages in Fourier space. Integrating eq. 1, using its solution in the probability distribution, then Fourier transforming provides
-\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
-Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen2007} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Shapiro1978}, in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
-\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
-and inverse Fourier transforming provides the desired real-space master equation
-
-At short times $t\ll k^{-1}$, this becomes
-\be \sigma_x(t)^2 \sim \frac{k_E k_D V^2}{k^2} t^2 + \frac{2 k_E D }{k}t, \label{eq:major}\ee
-so provided that $t\ll k^{-1}$, eq. \ref{eq:major} scales as $\sigma_x^2 \sim t^2$ when $t \gg \frac{2Dk}{V^2 k_D}$ and $\sigma_x^2 \sim t$ when $t \ll \frac{2Dk}{V^2 k_D}$.
-Similarly at long times $t\gg k^{-1}$, eq \ref{eq:minor} becomes
-\be \sigma_x^2 \sim \frac{2k_Ek_D V^2}{k^3}t. \ee
-
-Calculated in this way, the sediment flux distribution is observation-scale dependent, describing how sediment fluxes depend on the period over which they are measured.
-I demonstrated that scale-dependence results from velocity fluctuations in the motion state during the movements of individual particles, and I introduced a Peclet number ($\Pe$) to characterize the relative strength of these fluctuations.
-
-We can imagine generalizing the description in equation \ref{eq:flipfloplangevin} to include time-dependent phenomena interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
-
-\subsection{Stochasticity in landscape evolution}
-
-\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{sec:landscape}), writing
-\be (1-\phi) \pt h(x,t) = -\px q(x,t). \label{eq:stocEx}\ee
-In this one-dimensional equation, $h(x,t)$ is a continuous field representing the longitudinal channel profile.
-Traditionally, the Exner equation is evaluated considering the flux as a deterministic quantity \citep{Parker2007,Viparelli2011,An2017}, which provides a solitary trajectory of the bed profile $h(x)$ through time.
-
-Considering the flux in the Exner equation as a stochastic quantity invites us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2016}.
-In contrast to the traditional analysis, a stochastic Exner equation with fluctuating $q(x,t)$ generates an ensemble of possibilities for the channel profile $z(x)$ at each instant.
-This set of possibilities is characterized by a probability distribution $P(\{z(x)\},t)$ for each possible profile.
-
-Mathematically, since the profile is a field, evaluating such a probability distribution of channel profiles requires statistical field theory \citep{Kardar2007}.
-A starting point on this analysis might be $P(\{z(x)\},t) = \bra \delta[z(x,t)-h(x,t)]\ket,$ where $h(x,t)$ involves the stochastic flux $q(x,t)$ from equation \ref{eq:stocEx}.
-In this equation, $\delta[f(x)]$ is a Dirac delta functional, which generalizes the delta function $\delta(x)$ from a single-valued parameter $x$ to a continuous field $f(x)$.
-Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2001,Nelson2004}, but they have not been introduced in geomorphology.
-Such a probabilistic description of bed profiles would be an interesting extension of the methodology applied in this thesis so far to understand landscape evolution in an alternative way.

### 2021-09-24 revision/ch3bedEle.tex
getting close.

@@
 In the first type, individual displacements are considered to result from alternate step-rest sequences, where step lengths and resting times are random variables following statistical distributions \citep{Einstein1937}.
-Differences between the random-walk motions of one grain and the next imply a spreading apart of tracer grains as they transport downstream: bedload tracers undergo diffusion.
+Differences between the random-walk motions of one grain and the next imply a spreading apart (i.e., diffusion) of tracer grains as they transport downstream.
 
@@
 They observed heavy-tailed resting times of buried grains and described their results with a mathematical model similar to an earlier effort by \citet{Voepel2013}.
-Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation time-series \citep{Redner2007}.
-Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, while neither model explicitly describes bedload transport.
+Both of these models treat bed elevation changes as a random walk and interpret resting times as return periods from above in the bed elevation timeseries \citep{Redner2007}.
+Each describes resting time distributions from different experiments, but they rely on different random walk models, and their treatment of bed elevations as a process independent of sediment transport is questionable at first glance, since bedload transport is the source of bed elevation changes \citep{Wong2007}, yet neither model explicitly describes bedload transport.
 Models of sedimentary bed evolution incorporating sediment transport processes might enhance understanding of sediment resting times.
@@
 The second type of stochastic model prescribes rates (probabilities per unit time) to the erosion and deposition events of individual grains within a control volume to calculate the particle activity \citep{Einstein1950}.
-These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and time scales over which they are observed \citep{Singh2009,Singh2012,Saletti2015}.
+These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and timescales over which they are observed \citep{Singh2009,Singh2012,Saletti2015}.
 A recent surge in research activity has generated rapid progress in this subject. 
@@
 In this work, I present the first stochastic model coupling the erosion and deposition of individual bedload grains to local bed elevation changes.
-This model extends the birth-death model approach reviewed in sec. \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
+This model extends the birth-death model approach reviewed in Sec. \ref{sec:birthdeath} to describe the interplay between the bedload flux and the bed elevation within a control volume.
 This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried under the bed surface.
@@
 
-Below, sec.
-\ref{sec:elemodel} introduces the stochastic model and sec.
+Below, Sec.
+\ref{sec:elemodel} introduces the stochastic model and Sec.
 \ref{sec:elesolution} solves it with a mixture of numerical and analytical techniques.
@@
 Sec.
-\ref{sec:eleresults} present these features, while secs.
+\ref{sec:eleresults} present these features, while Secs.
 \ref{sec:elediscussion} and \ref{sec:eleconclusion} conclude with the implications of these features and the suggestions they raise for future research.
 
-\section{Stochastic model of bedload transport and bed elrefevations}
+\section{Stochastic model of bedload transport and bed elevations}
 \label{sec:elemodel}
 
-The model in this chapter is formulated by considering a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in fig.
+The model in this chapter is formulated by considering a volume of downstream length $L$ containing some number $n$ of moving particles in the flow and some number $m$ of stationary particles composing the bed at time $t$, as depicted in Fig.
 \ref{fig:eledefinition}.
@@
 
-All particles are considered approximately spherical with the same diameter $2a$, so their mobility and packing characteristics are consistent.
-Following \citet{Ancey2008}, basically four events can occur at any instant to modify the populations $n$ and $m$. These events are characterized using probabilities per unit time (rates).
+All particles are considered approximately spherical with the same radius $a$, so their mobility and packing characteristics are similar from one particle to the next.
+Following \citet{Ancey2008}, basically four events can occur at any instant to modify the populations $n$ and $m$.
 These events are (1) migration of a moving particle into the volume from upstream ($n \rightarrow n+1$), (2) the entrainment (erosion) of a stationary particle into motion within the volume ($m\rightarrow m-1$ and $n\rightarrow n+1$), (3) the deposition of a moving particle to rest within the volume ($m\rightarrow m+1$ and $n\rightarrow n-1$), and (4) the migration of a moving particle out of the volume to downstream ($n\rightarrow n-1$).
-The four events are depicted as arrows in fig. \ref{fig:eledefinition}.
+The four events are depicted as arrows in Fig. \ref{fig:eledefinition}.
 As the events occur at random intervals, they set up a joint stochastic evolution of the populations $n$ and $m$ characterized by a joint probability distribution $P(n,m,t)$ for the number of particles in motion and rest in the volume at $t$.
@@
 The bed is displayed in a degraded state, where $m<0$.
-The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized time-series of bed elevations computed from $m$ using eq.
+The marginal distributions of $n$ and $m$ are indicated in the upper right, while the bottom panel is a realized timeseries of bed elevations computed from $m$ using Eq.
 \ref{eq:ele}.}
@@
 For simplicity, this chapter utilizes the scaling of \citet{Sawai1987}. This scaling is equivalent to the  \citet{Wong2007} scaling when bed elevation changes are small.
-Because experimental distributions of bed elevations are often symmetrical, \citep{Crickmore1962, Pender2001,Wong2007, Martin2014}, the erosion and deposition feedbacks are considered to have the same strength.
+Because experimental distributions of bed elevations are often symmetrical \citep[e.g.][]{Crickmore1962, Pender2001,Wong2007, Martin2014}, the erosion and deposition feedbacks are considered to have the same strength.
 As bed elevation changes drive up (down) erosion rates, so they drive down (up) deposition rates to the same degree.
@@
 \end{align}
-In eqs. \ref{eq:rate2} and \ref{eq:rate3}, $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
-$\kappa$ is referred to as a coupling constant since it controls the strength of this feedback. I will later demonstrate the relationship
+In Eqs. \ref{eq:rate2} and \ref{eq:rate3}, $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
+Using this coupling constant, I will later demonstrate the relationship
 \begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
@@
 At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
-Away from this elevation, entrainment and deposition are alternatively suppressed or enhanced depending on the sign of $m$, constituting a feedback between bed elevations and erosion / deposition.
+Away from this elevation, entrainment and deposition are alternatively suppressed or enhanced depending on the sign of $m$, constituting a feedback between bed elevations and entrainment and deposition.
 
@@
 The average entrainment and deposition rates $E$ and $D$ over all bed elevations are $E = \lambda +\mu \langle n \rangle$ and $D=\sigma \langle n \rangle$.
-The solutions of eq.
+The solutions of Eq.
 \ref{eq:elemaster} should be expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
 Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
-This Master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
+This master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
 In context of this chapter, the two populations are the moving and stationary grains in the control volume.
@@
 
-Unfortunately, eq. \ref{eq:elemaster} does not admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
+Unfortunately, Eq. \ref{eq:elemaster} does not admit an analytical solution unless $\kappa=0$ (but see \citet{Swift2002} for the generating function method which fails in this case).
 The difficulty originates from the product terms between $n$ and $m$ representing the bed elevation dependence of collective entrainment and deposition rates.
@@
 In conditions when the coupling constant $\kappa$ between the entrainment/deposition rates and the bed elevation is weak, it is also possible to solve the master equation approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
-The Gillespie simulation algorithm is described below in sec. \ref{sec:elenumerical}, while these analytical approximations are described in sec. \ref{sec:eleanalytical}.
+The Gillespie simulation algorithm is described below in Sec. \ref{sec:elenumerical}, while these analytical approximations are described in Sec. \ref{sec:eleanalytical}.
 
@@
 As a result, to step the Markov process through a single transition, one can draw a first random value from the exponential distribution of transition intervals to determine the time of the next transition,
-then draw a second random value to choose the type of transition that occurs using relative probabilities formed from eqs.
+then draw a second random value to choose the type of transition that occurs using relative probabilities formed from Eqs.
 \ref{eq:rate1}-\ref{eq:rate4}.
-The transition is enacted by shifting $t$, $n$ and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
+The transition is enacted by shifting $t$, $n$, and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
 This procedure can be iterated to form an exact realization of the stochastic process \citep{Gillespie2007}.
-The appendix chapter \ref{ch:gill} provides additional background on the stochastic simulation method.
+The appendix Secs. \ref{sec:arr}-\ref{sec:crr} provide additional background on the stochastic simulation method.
 
 Using this method, I simulated 4 transport conditions with 13 different values of $l$ taken across a range from $l=a$ (a single radius) to $l=10a$ (10 radii).
-These values include the range exhibited by the available experimental data on bed elevation time series \citep{Wong2007,Singh2009,Martin2014}.
+These values include the range exhibited by the available experimental data on bed elevation timeseries \citep{Wong2007,Singh2009,Martin2014}.
 
@@
 Units are $s^{-1}$ (probability/time).
-Bed elevation changes modulate these rates in accord with eqs.
+Bed elevation changes modulate these rates in accord with Eqs.
 \ref{eq:rate1}-\ref{eq:rate4}.}\label{tab:anceyparams}
@@
 
-The $n$ and $m$ dynamics in eq.
+The $n$ and $m$ dynamics in Eq.
 \ref{eq:elemaster} can be approximately decoupled using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
-These inequalites mean many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
-Concentrating on steady state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to eq.
+These inequalites imply that many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
+Concentrating on steady state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to Eq.
 \ref{eq:elemaster}, with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$ \citep[e.g.][]{Haken1978}.
@@
 \end{multline}
-Unforunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in eq. \ref{eq:decomp}.
+Unforunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in Eq. \ref{eq:decomp}.
 
 The simplest approximation to these equations holds that $\kappa$ is so small that the dynamics of $n$ are totally independent of $m$: $A(n|m) = A(n)$.
-Taking this limit in eq.
+Taking this limit in Eq.
 \ref{eq:decomp}, summing over $m$, and using $\langle m \rangle = 0$ reproduces the \citet{Ancey2008} particle activity model.
-As shown in sec. \ref{sec:birthdeath}, this has solution
+As shown in Sec. \ref{sec:birthdeath}, this has the solution
 \begin{equation} A(n) = \frac{\Gamma(r+n)}{\Gamma(r)n!}p^r(1-p)^n.\label{eq:ancey}\end{equation}
 which is a negative binomial distribution for the particle activity with parameters $r=(\nu+\lambda)/\mu$ and $p=1-\mu/(\sigma+\gamma).$
-This result implies $\langle n | m \rangle = \langle n \rangle$, so with the definitions of $E$ and $D$ and the equilibrium condition $E=D$, eq. \ref{eq:approxele} provides
+This result implies $\langle n | m \rangle = \langle n \rangle$, so with the definitions of $E$ and $D$ and the equilibrium condition $E=D$, Eq. \ref{eq:approxele} provides
 \begin{equation}0 \approx [1+\kappa(m+1)]M(m+1) + [1-\kappa(m-1)]M(m-1)-2M(m). \label{eq:ou} \end{equation}
 This mean field equation matches the discrete Ornstein-Uhlenbeck model of bed elevation changes developed by \citet{Martin2014}.
-The independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model eq.
+The independent bed elevation and particle activity models of \citet{Martin2014} and \citet{Ancey2008} derive from the model Eq.
 \ref{eq:elemaster} in a mean field approximation when $\kappa$ is insignificant.
 
-In the appendix ch.
-\ref{ch:gill} I show that the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within eq.
+In the appendix Sec.
+\ref{sec:drr} I show that the Fokker-Planck approximation \citep{Gardiner1983} formed by expanding $M(m\pm 1)$ to second order in $m$ within Eq.
 \ref{eq:ou} provides the solution $M(m) \propto \exp(-\kappa m^2)$: this is a normal distribution of bed elevations with variance $\sigma_m^2 \propto \frac{1}{2\kappa}$.
-As I will demonstrate in sec.
-\ref{sec:eleresults}, and as I have already suggested with eq.
+As I will demonstrate in Sec.
+\ref{sec:eleresults}, and as I have already suggested with Eq.
 (\ref{eq:active}), this is a poor approximation to the bed elevation variance.
 Nevertheless, this approximation does capture the Gaussian shape of the bed elevation distribution.
-The essential issue with this mean field approach is that the conditional mean particle activity $\langle n | m \rangle$ varies significantly with $m$ in actuality, especially when collective entrainment contributes to the mean entrainment rate $E$.
-I will discuss these points subsequently when developing more refined approximations and presenting numerical results.
+The essential issue with this mean field approach is that in actuality, the conditional mean particle activity $\langle n | m \rangle$ varies significantly with $m$, especially when collective entrainment contributes to the mean entrainment rate $E$.
 
-A more careful approximate solution to eq.
-\ref{eq:approxele} can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into eq.
-\ref{eq:approxele} in order to close the equation for $m$ without solving eq.
+A more careful approximate solution to Eq.
+\ref{eq:approxele} can be obtained by prescribing a phenomenological equation for $\langle n | m \rangle$ into Eq.
+\ref{eq:approxele} in order to close the equation for $m$ without solving Eq.
 \ref{eq:decomp}.
-By comparison with the numerical simulations we determine that
+As explained in appendix Sec. \ref{sec:err}, comparison with the numerical simulations determines that
 \begin{equation}
@@
 captures the general features of the conditional mean particle activity.
-As I show in the supplementary information, introducing this closure relation to eq. \ref{eq:approxele}, making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
+Introducing this closure relation to Eq. \ref{eq:approxele}, making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
 \begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
-representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result posited in eq.
+representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result of Eq.
 \ref{eq:active}.
 $M_0$ is a normalization constant. 
-This closure equation approach shows good correspondence with numerical solutions of eq. \ref{eq:elemaster}, at least for the flow parameters in tab. \ref{tab:anceyparams}.
+This closure equation approach shows good correspondence with numerical solutions of Eq. \ref{eq:elemaster}, at least for the flow parameters in Tab. \ref{tab:anceyparams}.
 
@@
 From the initial conditions, all simulations show a rapid attainment of steady-state stochastic dynamics of $n$ and $m$ which support a time-independent joint distribution $P(n,m)$.
-The bottom panel of fig.
-\ref{fig:eledefinition} shows an elevation time-series.
-In order to describe the implications of coupling bedload transport to bed elevation changes, the numerical and analytical results for the probability distributions of bedload transport and bed elevations are presented in sec.
-\ref{sec:elepdf} and the statistical moments of these quantities in sec.
+The bottom panel of Fig.
+\ref{fig:eledefinition} shows an elevation timeseries.
+In order to describe the implications of coupling bedload transport to bed elevation changes, the numerical and analytical results for the probability distributions of bedload transport and bed elevations are presented in Sec.
+\ref{sec:elepdf} and the statistical moments of these quantities are presented in Sec.
 \ref{sec:elemom}.
-The effects of collective entrainment on bed elevation changes are studied in sec.
-\ref{sec:elecolent}, and the resting times of sediment undergoing burial are evaluated in sec.
+The effects of collective entrainment on bed elevation changes are studied in Sec.
+\ref{sec:elecolent}, and the resting times of sediment undergoing burial are evaluated in Sec.
 \ref{sec:elertcdf}.
@@
 
-The joint distribution is computed numerically by counting occurrences of the states $(n,m)$ in the simulated time series.
+The joint distribution is computed numerically by counting occurrences of the states $(n,m)$ in the simulated timeseries.
 From this joint distribution the marginal distributions $P(n)$ and $P(m)$ are obtained by summing over $m$ and $n$ respectively.
-A representative subset of these marginal distributions is displayed in fig. \ref{fig:pdfs} alongside the approximate results of eqs. \ref{eq:ancey} and \ref{eq:ou2}.
+A representative subset of these marginal distributions is displayed in Fig. \ref{fig:pdfs} alongside the approximate results of Eqs. \ref{eq:ancey} and \ref{eq:ou2}.
 \begin{figure}[!htbp]
@@
 	\caption{Panel (a) presents the probability distribution of particle activity $n$ and panel (b) presents the probability distribution of the relative number of particles $m$ for a representative subset of simulations.
-These distributions represent different flows from tab.
+These distributions represent different flows from Tab.
 \ref{tab:anceyparams}, distinguished by color, and different values of the active layer depth $l$ (equivalently the coupling constant $\kappa$), distinguished by the marker style.
-The mean field theories (mft) of eqs.
+The mean field theories (mft) of Eqs.
 \ref{eq:ancey} and \ref{eq:ou2} are displayed as solid black lines.}
@@
 \end{figure}
-The mean field eq.
-\ref{eq:ancey} for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match eq.
-\ref{eq:ou2}, having Gaussian profiles consistent with the assumption of a symmetric scaling of erosion and deposition rates with bed elevation changes.
+The mean field equation for the particle activity $n$ (Eq.
+\ref{eq:ancey}) closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match Eq.
+\ref{eq:ou2}, having Gaussian profiles consistent with the assumption of a symmetric scaling between erosion and deposition rates and bed elevation changes.
 
@@
 The moments of $n$ and $m$ are calculated by summing over $P(n,m)$. 
-The $j$th order unconditional moment of the particle activity $n$ derives from
+The $j$th order unconditional moment of the particle activity $n$ is defined as
 \begin{equation} \langle n^j \rangle = \sum_{n}n^jP(n),\end{equation}
@@
 	\includegraphics[width=0.5\textwidth,keepaspectratio]{./figures/ch3/variance.pdf}
-	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as posited in eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$. }
+	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as described by Eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$. }
 	\label{fig:var}
 \end{wrapfigure}
-Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in eq.
+Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in Eq.
 \ref{eq:ou2}; this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
-The close correspondence between the mean field approximation and the numerical simulations in fig.
-\ref{fig:pdfs}a suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
+The close correspondence between the mean field approximation and the numerical simulations in Fig.
+\ref{fig:pdfs}(a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
 They appear identical within numerical uncertainty.
@@
 This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$.
-The closure equation \ref{eq:closure} is plotted in panel (a) Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
-These results indicate that bedload statistics measurements on short time scales could be severely biased by departures from the mean bed elevation.}
+The closure equation \ref{eq:closure} is plotted in panel (a).
+Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
+These results indicate that bedload statistics measurements on short timescales could be severely biased by departures from the mean bed elevation.}
 	\label{fig:condmoms}
 \end{figure}
-Figure \ref{fig:condmoms}a demonstrates that the flow conditions in tab.
+Figure \ref{fig:condmoms}a demonstrates that the flow conditions in Tab.
 \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
-The closure model (eq.
-\ref{eq:closure}) used to derive the approximate bed elevation distribution (eq.
-\ref{eq:ou2}) is plotted behind the conditional mean profiles in fig.
+The closure model (Eq.
+\ref{eq:closure}) used to derive the approximate bed elevation distribution (Eq.
+\ref{eq:ou2}) is plotted behind the conditional mean profiles in Fig.
 \ref{fig:condmoms}a, where it appears as a crude approximation since it does not capture the asymmetry.
-Nevertheless, fig. \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
-For the parameters of the \citet{Ancey2008} experiments, fig.
+Nevertheless, Fig. \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
+For the parameters of the \citet{Ancey2008} experiments, Fig.
 \ref{fig:condmoms}b displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
@@
 	\caption{The shift of the mean particle activity in panel (a) and its fluctuations in panel (b) with departures of the bed elevation from its mean.
-All simulations are at flow condition (g) from tab.
+All simulations are at flow condition (g) from Tab.
 \ref{tab:anceyparams} except $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment.
-Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (eq.
+Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (Eq.
 \ref{eq:ancey}) at large departures from the mean bed elevation.
@@
 Figure \ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed.
-When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect identified in sec.
+When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect identified in Sec.
 \ref{sec:elemom}.
@@
 Panel (a) shows differing flow conditions at a fixed $l$ value, while panel (c) shows fixed flow conditions at differing $l$.
-When scaled by $T_0$ (eq.
+When scaled by $T_0$ (Eq.
 \ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d).
@@
 
-Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the time-series of $m$ \citep[e.g.,][]{Redner2007}.
+Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the timeseries of $m$ \citep[e.g.,][]{Redner2007}.
 Following \citet{Voepel2013} and \citet{Martin2014}, we concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
@@
 Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
-Using the marginal probability distribution of bed elevations $P(m)$ (fig.
+Using the marginal probability distribution of bed elevations $P(m)$ (Fig.
 \ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}:
@@
 We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
-Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (eq.
+Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (Eq.
 \ref{eq:ele}), bed elevations change with a characteristic speed $v = z_1 E$.
-Since the range of elevation deviations is $l$ (fig. \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
+Since the range of elevation deviations is $l$ (Fig. \ref{fig:var}), the time required for the bed to shift through this characteristic distance is $l/v$, or equivalently
 \begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
-When scaling the resting time by this $T_0$, we obtain the collapse shown in fig. \ref{fig:cdfs}.
+When scaling the resting time by this $T_0$, we obtain the collapse shown in Fig. \ref{fig:cdfs}.
 Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
@@
 Einstein developed the first stochastic models of bedload tracer diffusion \citep{Einstein1937} and the bedload flux \citep{Einstein1950}, and his ideas can be viewed as the nexus of an entire paradigm of research that extends into the present day \citep[e.g.,][]{Hubbell1964, Nakagawa1976,Hassan1991,Ancey2008, Wu2019}.
-These models aim to predict bedload transport characteristics from stochastic concepts of individual particle motions.
+These approaches aim to predict bedload transport characteristics from stochastic concepts of individual particle motions.
 With some exceptions \citep{Yang1971,Nakagawa1980,Pelosi2016,Wu2019,Wu2019a}, existing descriptions are spatially one-dimensional, concentrating on the motion of grains in the downstream direction without including the vertical dimension wherein local bed elevation changes imply sediment burial \citep{Voepel2013,Martin2014} and change the mobility of surface grains \citep{Yang1971,Nakagawa1980}.
@@
 Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
-Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments in the literature.
+Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments.
 
@@
 Finally, the model allowed resolution of resting times for sediment undergoing burial within the sedimentary bed.
-These time scales were obtained by analyzing return times from above in the bed elevation time series.
-This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long time scales.
+These timescales were obtained by analyzing return times from above in the bed elevation timeseries.
+This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
 The power-law tails were found to collapse across flow conditions using a time scale formed from the mean erosion rate and the active layer depth.
@@
 As the model in this chapter builds on earlier works describing particle activity and bed elevation changes independently, it also reduces to these works in simplified limits when the coupling between the particle activity and the local bed elevation vanishes.
-With the mean field approach in sec.
+With the mean field approach in Sec.
 \ref{sec:eleanalytical}, the \citet{Martin2014} Ornstein-Uhlenbeck model for bed elevations and the \citet{Ancey2008} birth-death model for the particle activity were derived as simplified limits of the coupled model developed in this chapter.
@@
 
-Like the theory developed by \citet{Martin2014}, the model eq.
+Like the theory developed by \citet{Martin2014}, the model Eq.
 \ref{eq:elemaster} produces heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
 This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016a}. 
-The resting time distributions derived in sec.
+The resting time distributions derived in Sec.
 \ref{sec:elertcdf} show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of the results.
@@
 Some relatively recent particle tracking experiments have demonstrated joint resolution of bed elevations and bedload transport \citep{Martin2014,Heyman2016}.
-A suitably designed experiment using this methodology could test the prediction that bed elevations regulate particle activity statistics, as essentially represented in figs.
+A suitably designed experiment using this methodology could test the prediction that bed elevations regulate particle activity statistics, as essentially represented in Figs.
 \ref{fig:condmoms} and \ref{fig:colent}.
@@
 Many other statistical characteristics of bedload transport have been left for future studies. 
-For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic which was previously referenced in sec.
+For example, the dependence of bedload transport \citep{Saletti2015,Singh2009} and bed elevation statistics \citep{Aberle2006, Singh2009,Singh2012} on the spatial and temporal scales over which they are observed is an emerging research topic which was previously referenced in Sec.
 \ref{sec:renewal}.
-Whether bed elevation distributions are also scale-dependent like the sediment flux distribution of ch. \ref{ch:flux} is an interesting question for future research.
+Whether bed elevation distributions are also scale-dependent like the sediment flux distribution of Ch. \ref{ch:flux} is an interesting question for future research.
 
@@
 \end{enumerate}
-These results imply measurements of bedload transport statistics could be biased at observation time scales smaller than adjustments of the bed elevation time series when collective entrainment occurs.
+These results imply measurements of bedload transport statistics could be biased at observation timescales smaller than adjustments of the bed elevation timeseries when collective entrainment occurs.
 The next step is to generalize the model in this chapter to a multi-cell framework to study channel morphodynamics in a stochastic framework.

### 2021-09-24 revision/ch4downDiff.tex
getting close.

@@
 The majority of these models predict two ranges of diffusion (local and intermediate) without predicting a global range.
-Among these, \citet{Nikora2001a} used synthetic turbulence \citep{Kraichnan1970} with a discrete element method for the granular phase \citep{Cundall1979a}; \citet{Bialik2012} used synthetic turbulence with a random collision model \citep{Sekine1992}; and \citet{Fan2016} used a Langevin equation with probabilistic rests.
+Among these, \citet{Nikora2001a} used synthetic turbulence \citep{Kraichnan1970} with a discrete element method for the granular phase \citep{Cundall1979}; \citet{Bialik2012} used synthetic turbulence with a random collision model \citep{Sekine1992}; and \citet{Fan2016} used a Langevin equation with probabilistic rests.
 To our knowledge, only \citet{Bialik2015} have claimed to capture all three ranges from a Newtonian approach.
@@
 To incorporate motion intervals and sediment burial, I utilize the multi-state CTRW formalism developed by \citet{Weiss1976, Weiss1994} that extends the CTRW of \citet{Montroll1965}.
-Below, I develop and solve the model in sec.
+Below, I develop and solve the model in Sec.
 \ref{sec:model}.
-Then, I discuss the predictions of this model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in secs.
+Then, I discuss the predictions of this model, present its implications for local, intermediate, and global ranges of bedload diffusion, and suggest next steps for bedload diffusion research in Secs.
 \ref{sec:discussion} and \ref{sec:conclusion}.
@@
 \be p(x,t) = \sum_{k=0}^2 p_k(x,t) \ee
-This joint density is completely determined from the solutions of eqs. \ref{eq:g1}-\ref{eq:g2} given specification of the distributions $g_{ij}$ and $G_i$.
+This joint density is completely determined from the solutions of Eqs. \ref{eq:g1}-\ref{eq:g2} given specification of the distributions $g_{ij}$ and $G_i$.
 
@@
 
-These distributions can be constructed from the assumptions described in sec. \ref{sec:assumptions}.
+These distributions can be constructed from the assumptions described in Sec. \ref{sec:assumptions}.
 Since particles resting on the bed surface become buried in a time $t$ with probability $\Phi(t)$, and resting times are distributed as $\psi_1(t)$, these assumptions obtain $g_{12}(x,t) = \delta(x)k_1e^{-k_1t}e^{-\kappa t}$ and $g_{10}(x,t) = \delta(x) k_1 e^{-k_1 t}(1-e^{-\kappa t})$.
@@
 Grains are considered initially at rest ($\theta_1=1$, $\theta_2=0$).
-The solid lines are the analytical distribution in eq.
+The solid lines are the analytical distribution in Eq.
 \ref{eq:pdf}), while the points are numerically simulated, showing the correctness of our derivations.
@@
 
-Figure \ref{fig:diffpdfs} depicts the distribution (eq.
+Figure \ref{fig:diffpdfs} depicts the distribution (Eq.
 \ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
-When grains do not become buried, as in panel (a) of fig.
+When grains do not become buried, as in panel (a) of Fig.
 \ref{fig:diffpdfs}, the distribution becomes Gaussian-like at relatively large observation times, exemplifying normal diffusion and satisfying the central limit theorem.
-When grains become buried, as in panel (b) of fig.
+When grains become buried, as in panel (b) of Fig.
 \ref{fig:diffpdfs},  the Q-function terms prevent the distribution from approaching a Gaussian at large timescales, exemplifying anomalous diffusion \citep{Weeks1998} and violating the central limit theorem \citep{Metzler2000,Schumer2009}.
@@
 
-To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution (eq.
+To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution (Eq.
 \ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}.
@@
 \be \sigma_x^2(t) = A(t)e^{(b-a)t} + B(t)e^{-(a+b)t} + C(t). \label{eq:var}\ee
-In these equations, $a = (\kappa + k_1+k_2)/2$ and $b = \sqrt{a^2-\kappa k_2}$ are effective rates having dimensions of inverse time, while the $A$, $B$, and $C$ factors are provided in tab.
+In these equations, $a = (\kappa + k_1+k_2)/2$ and $b = \sqrt{a^2-\kappa k_2}$ are effective rates having dimensions of inverse time, while the $A$, $B$, and $C$ factors are provided in Tab.
 \ref{table:params}.
@@
 	%\begin{wraptable}{r}{width=5.5cm}
-	\caption{Abbreviations used in the expressions of the mean (eq. \ref{eq:mean}), second moment (eq. \ref{eq:second}) and variance (eq. \ref{eq:var}) of bedload tracers.}
+	\caption{Abbreviations used in the expressions of the mean (Eq. \ref{eq:mean}), second moment (Eq. \ref{eq:second}) and variance (Eq. \ref{eq:var}) of bedload tracers.}
 	\label{table:params}
@@
 \end{table}
-The positional variance (eq. \ref{eq:var}) is plotted in fig. \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
+The positional variance (Eq. \ref{eq:var}) is plotted in Fig. \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
 The notation ``$\gg$'' is interpreted in this context to mean ``of at least an order of magnitude greater''.
 These conditions are most relevant to tracers in gravel-bed rivers, since they represent that grains are initially at rest \citep{Hassan1991,Wu2019}, motions are typically much shorter than rests \citep{Einstein1937,Hubbell1964}, and burial requires a much longer time than typical rests  \citep{Ferguson2002,Hassan1994,Haschenburger2013}.
-Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (eq.
+Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (Eq.
 \ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/diffusion.png}
-	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (eq.
+	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (Eq.
 \ref{eq:var}) with mean motion time $1.5$ s, resting time $30.0$ s, and movement velocity $0.1$ m/s -- values comparable to laboratory experiments transporting small ($5$ mm) gravels \citep{Lajeunesse2010,Martin2012}.
 The burial timescale is $7200.0$s (two hours), and grains start from rest ($\theta_1=1$).
-The solid line is eq.
+The solid line is Eq.
 \ref{eq:var}, and the points are numerically simulated.
@@
 At local timescales, grains usually either rest or move; at intermediate timescales, they transition between rest and motion; at global timescales, they transition between rest, motion, and burial; and at geomorphic timescales, all grains bury.
-Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as discussed in sec.
+Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as discussed in Sec.
 \ref{sec:discussion}.
@@
 
-Two limiting cases of eq.
+Two limiting cases of Eq.
 \ref{eq:var} provide the scaling exponents $\gamma$ of the diffusion $\sigma_x^2 \propto t^\gamma$ in each range.
@@
 The extreme $k_1/\kappa \ll 1$ produces $\gamma=1$ in the global range, while the opposite extreme $k_1/\kappa \rightarrow \infty$ produces $\gamma=3$.
-To summarize, when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, eq. \ref{eq:var} implies:
+To summarize, when $k_2\gg k_1 \gg \kappa$ so all three diffusion ranges exist, Eq. \ref{eq:var} implies:
 \begin{enumerate}
@@
 In actuality, burial is a temporary condition, because bed scour can exhume buried sediment back into transport \citep{Wu2019a}, probably after heavy-tailed intervals \citep{Voepel2013,Martin2014}.
-A generalization of the model in this chapter to include heavy-tailed timescales between burial and exhumation might develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in fig.
+A generalization of the model in this chapter to include heavy-tailed timescales between burial and exhumation might develop four ranges of diffusion, where the long-time decay of the exhumation time distribution would dictate the geomorphic range diffusion characteristics as depicted in Fig.
 \ref{fig:diffvar}.
-If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested in ch.
+If cumulative exhumation times decay faster than $T^{-1/2}$, as suggested in Ch.
 \ref{ch:ch3}, other equilibrium transport models \citep{Voepel2013, Martin2014}, and laboratory experiments \citep{Martin2014,Martin2012}, the geomorphic range is expected to be super-diffusive \citep{Weeks1998}.
@@
 
-The analytical solution of bedload diffusion in eq.
+The analytical solution of bedload diffusion in Eq.
 \ref{eq:var} reduces exactly to the analytical solutions of the \citet{Lisle1998} and \citet{Lajeunesse2017} models in the limit without burial ($\kappa \rightarrow 0$), the \citet{Wu2019} model in the limit of instantaneous steps ($k_2 \rightarrow \infty$ and $l = v/k_2$), and the original \citet{Einstein1937} model in the limit of instantaneous steps without burial.
@@
 At base level, this work demonstrates that (1) durations of sediment motions, (2) durations of sediment rest, and (3) the sediment burial process are sufficient to develop three diffusion ranges that terminate when all tracers become buried.
-This work provides a complementary perspective on the three range diffusion exhibited in ch. \ref{ch:flux}.
+This work provides a complementary perspective on the three range diffusion exhibited in Ch. \ref{ch:flux}.
 A next step is to incorporate exhumation to better understand the geomorphic range.

### 2021-09-24 revision/ch5langevin.tex
getting close.

@@
 These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
-In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (secs.
+In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (Secs.
 \ref{sec:birthdeath}-\ref{sec:renewal}).
 Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
-One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (sec.
+One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (Sec.
 \ref{sec:birthdeath}).
 This formulation requires the instantaneous velocity distributions of sediment particles \citep[e.g.][]{Ancey2020a}, yet unfortunately, understanding of these distributions remains limited.
-There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in sec.
+There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in Sec.
 \ref{sec:langevin} of the introduction describe some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe all experimental observations reported to date \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
@@
 
-Stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec.
+Stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (Sec.
 \ref{sec:langevin}).
@@
 In this chapter I explore the possibility that the particle velocity distribution is controlled by momentum dissipation from particle-bed collisions by incorporating ideas from the physics of granular media.
-A major paradigm in granular dynamics is to describe particles as a continuum or "granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in ch.
+A major paradigm in granular dynamics is to describe particles as a continuum or "granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in Ch.
 \ref{ch:Introduction} such a continuum assumption is unlikely to be satisfied for weak bedload transport conditions.
@@
 In the theory of rarefied granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
-Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970,Briliantov2004}.
+Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970,Brilliantov2004}.
 Collisions in these models involve episodic kicks to the particle momenta at random times, usually parameterized by billiard ball-like models of the underlying rigid body dynamics \citep[e.g.][]{Brach1989}.
@@
 A secondary motivation is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
-I develop the model and explain its key assumptions in sec.
+I develop the model and explain its key assumptions in Sec.
 \ref{sec:langmodel}.
 Then I present the analytical solution of the model and other major results in \ref{sec:langresults}.
-Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in secs.
+Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in Secs.
 \ref{sec:langdiscussion} and \ref{sec:langconclusion}.
@@
 Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady flow.
-The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Jackson1982, Warburton1992,Hassan2001}.
-Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles, but rarely with other moving particles \citep[cf.][]{Williams2020}.
+The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Ashworth1989, Warburton1992}.
+Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles, but rarely with other moving particles \citep[cf.][]{Williams2021}.
 
 Particles respond to turbulent flow forces and episodic particle-bed collision forces.
-In contrast to the computational physics approach, I do not aim to characterize the exact time series of the forces on an individual particle.
-Instead, I model the ensemble of possible force time series that particles could conceivably experience.
-Each force time series realization implies a different velocity time series $u(t)$ in the downstream direction.
+In contrast to the computational physics approach, I do not aim to characterize the exact timeseries of the forces on an individual particle.
+Instead, I model the ensemble of possible force timeseries that particles could conceivably experience.
+Each force timeseries realization implies a different velocity timeseries $u(t)$ in the downstream direction.
 The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
-Here, the description of bedload particle velocity fluctuations will be more detailed than ch.
+Here, the description of bedload particle velocity fluctuations will be more detailed than Ch.
 \ref{ch:flux}, but entrainment and deposition processes will be neglected, meaning the model of this chapter is only applicable to moving particles.
@@
 \subsection{Episodic collision forces}
-Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid in the interstices between approaching particles \citep{Joseph2001,Schmeeckle2001}.
-The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of resitution may be less relevant than other factors such as collision geometry and bed deformation \citep{}.
-Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in fig.
+Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid between approaching particles \citep{Joseph2001,Schmeeckle2001}.
+The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of resitution may be less relevant than other factors such as collision geometry and bed deformation.
+Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in Fig.
 \ref{fig:fig1}.
@@
 	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces.
-During saltation, pre-collisional streamwise velocities $u$ are transformed to postcollisional velocities $\ve u < u$.}
+During saltation, pre-collisional streamwise velocities $u$ are transformed to post-collisional velocities $\ve u < u$.}
 	\label{fig:fig1}
@@
 
-Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (sec.
+Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (Sec.
 \ref{sec:einflux}):
@@
 This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions.
-These conditions should be satisfied for idealized saltation trajectories as depicted in fig.
+These conditions should be satisfied for idealized saltation trajectories as depicted in Fig.
 \ref{fig:fig1}.
@@
 Even ignoring this complication, the added mass and Basset forces are difficult to compute and therefore impractical to include in analytical models.
-These challenges require researchers to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michalades1997,Armenio2001}.
+These challenges require researchers to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michaelides1997,Armenio2001}.
 As a result, the majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the primary force(s) driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.
@@
 The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities should be expected to vary due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
-The nonlinearity in eq. \ref{eq:draggy} will amplify these slip velocity variations, generating large and erratic drag force fluctuations.
+The nonlinearity in Eq. \ref{eq:draggy} will amplify these slip velocity variations, generating large and erratic drag force fluctuations.
 
@@
 This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
-Mathematically, eq. \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
+Mathematically, Eq. \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
 Collisions introduce velocity jumps while turbulence generates velocity diffusion. The collision term is multiplicative in the sense that $u$ multiplies the Poisson noise $\xi_{\nu,\ve}(t)$.
 
-Models like eq.
+Models like Eq.
 \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
-One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning eq.
+One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning Eq.
 \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
 Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
-Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by eq.
+Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by Eq.
 \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
@@
 
-The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
+The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix Sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
 \be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
@@
 Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
-Nonlocality is introduced by the collision integral (eq. \ref{eq:colint}) which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
+Nonlocality is introduced by the collision integral (Eq. \ref{eq:colint}) which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
 This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}.
@@
 
-Owing to its nonlocality, eq. \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
+Owing to its nonlocality, Eq. \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
 To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
@@
 Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
-With the truncation to two terms, eq. \ref{eq:master} gives 
+With the truncation to two terms, Eq. \ref{eq:master} gives 
 \be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
@@
 
-In the appendix sec. \ref{sec:langsteadyderiv} eq. \ref{eq:governer} is solved with Laplace transforms, providing
+In the appendix Sec. \ref{sec:langsteadyderiv} Eq. \ref{eq:governer} is solved with Laplace transforms, providing
 \begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
@@
 \end{multline}
-The factors $\lambda_\pm$ are defined in appendix eq. \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
+The factors $\lambda_\pm$ are defined in appendix Eq. \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
 The normalization factors $K_\pm$ are 
@@
 
-It is possible to derive the moments of this probability distribution by multiplying eq.
+It is possible to derive the moments of this probability distribution by multiplying Eq.
 \ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[cf.][]{Cox1965}.
-These calculations are provided in appendix sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
+These calculations are provided in appendix Sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
 \be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}. \label{eq:meanu}\ee
@@
 This is consistent with the obvious similarities between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
-In contrast, particle velocity fluctuations in eq.
+In contrast, particle velocity fluctuations in Eq.
 \ref{eq:varu} depend sharply on the parameters representing particle-bed collisions, suggesting collisions could be the leading control on particle velocity fluctuations.
@@
 Flow forces generate fluctuating positive accelerations between collisions.
-Right panels show simulated histograms of particle velocities and exact solutions from eq.
+Right panels show simulated histograms of particle velocities and exact solutions from Eq.
 \ref{eq:steadystate}.
@@
 This figure reveals an apparent transition from exponential-like to Gaussian-like velocity distributions as collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
-In between, the full distribution eq.
-\ref{eq:steadystate} resembles a Gamma distribution, although it is represented by eq.
+In between, the full distribution Eq.
+\ref{eq:steadystate} resembles a Gamma distribution, although it is represented by Eq.
 \ref{eq:steadystate}, not a Gamma distribution.
@@
 
-In fact, the apparent transition from exponential to Gaussian in fig.
-\ref{fig:fig2} can be demonstrated analytically from eq.
+In fact, the apparent transition from exponential to Gaussian in Fig.
+\ref{fig:fig2} can be demonstrated analytically from Eq.
 \ref{eq:steadystate}.
 Despite its complex appearance, simple Gaussian and exponential forms derive from this equation as exact mathematical limits.
-When particle-bed collisions are completely inelastic ($\ve = 0$), eq.
-\ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\ve = 1$), eq.
+When particle-bed collisions are completely inelastic ($\ve = 0$), Eq.
+\ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\ve = 1$), Eq.
 \ref{eq:steadystate} becomes Gaussian.
@@
 
-The exponential limit of eq.
+The exponential limit of Eq.
 \ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see.
-Taking $\ve \rightarrow 0 $ in eq.
+Taking $\ve \rightarrow 0 $ in Eq.
 \ref{eq:steadystate}, all terms in the series except for the one at $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
@@
 
-The Gaussian limit as $\ve \rightarrow 1$ of eq.
+The Gaussian limit as $\ve \rightarrow 1$ of Eq.
 \ref{eq:steadystate} is more difficult to evaluate.
-The challenge is that the statistical moments (eqs.
-\ref{eq:mean} and \ref{eq:varu}) diverge at the same time as the denominator factors of the distribution eq.
+The challenge is that the statistical moments (Eqs.
+\ref{eq:mean} and \ref{eq:varu}) diverge at the same time as the denominator factors of the distribution Eq.
 \ref{eq:steadystate}.
-In the appendix sec.
+In the appendix Sec.
 \ref{sec:langextremes} I return instead to the original equation \ref{eq:governer} to evaluate the completely elastic limit, obtaining
@@
 A variety of velocity distributions have been observed in bedload transport experiments, ranging from exponential to Gaussian in shape. 
-This section compares these with the theoretical result eq. \ref{eq:steadystate}.
-Comparing eq.
+This section compares these with the theoretical result Eq. \ref{eq:steadystate}.
+Comparing Eq.
 \ref{eq:steadystate} with experimental data requires values for the steady component of the drag force $\Gamma$, the particle mass $m$, the magnitude of turbulent drag fluctuations $D$, the rate of particle-bed collisions $\nu$, and the modal elasticity of collisions $\varepsilon$.
 
-This section compares eq. \ref{eq:steadystate} with the results of six different experiments.
+This section compares Eq. \ref{eq:steadystate} with the results of six different experiments.
 In each case, the particle mass is computed from experimental parameters assuming spherical sediment as $m = \pi \rho_s d^3/6$, where $\rho_s$ is the sediment density and $d$ is the particle diameter.
-The characteristic slip velocity entering the steady component of the drag force (eq.
+The characteristic slip velocity entering the steady component of the drag force (Eq.
 \ref{eq:drag}) is not obviously related to any one velocity scale of the fluid flow.
@@
 Particle Reynolds numbers are estimated using the shear velocity: $\Rey_p= u_\ast d/nu$. 
-The magnitude $D$ of turbulent fluctuations, dissipation per collision $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution (eq.
+The magnitude $D$ of turbulent fluctuations, dissipation per collision $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution (Eq.
 \ref{eq:steadystate}) and the experimental data.
 
-The parameters for each experiment and the best-fit calibration parameters are collected in tab.
+The parameters for each experiment and the best-fit calibration parameters are collected in Tab.
 \ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in figure \ref{fig:fig4ch5}.
@@
 		\end{tabular}
-		\caption{Parameters used to fit the distributions in fig.
+		\caption{Parameters used to fit the distributions in Fig.
 \ref{fig:fig4ch5}.
 The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St$, and Froude number $\Fro$.
-The final two columns involve values that were tuned to fit the theoretical distribution (eq.
+The final two columns involve values that were tuned to fit the theoretical distribution (Eq.
 \ref{eq:steadystate}) to the experimental data, the dissipation $\ve$ and the collision rate $\nu$.
@@
 \end{table}
-In each case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model eq. \ref{eq:langevin} captures the essential physics.
+In each case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model Eq. \ref{eq:langevin} captures the essential physics.
 \begin{figure}
 	\centerline{\includegraphics{./figures/ch5/Fig4expComparison.png}}
-	\caption{This figure compares the available experimental data on velocity distributions and the theoretical velocity distribution eq.
+	\caption{This figure compares the available experimental data on velocity distributions and the theoretical velocity distribution Eq.
 \ref{eq:steadystate} with parameters $\nu$, $\varepsilon$, and $D$ treated as calibration parameters.
@@
 Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise for reasons of analytical tractability \cite[e.g.][]{Michaelides1997}.
-Even though the non-Gaussian, spatio-temporally-varying, and history-dependent aspects of fluid turbulence certainly do impact sediment movement \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is laminar or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the time scales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+Even though the non-Gaussian, spatio-temporally-varying, and history-dependent aspects of fluid turbulence certainly do impact sediment movement \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is laminar or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
 Nevertheless, it is possible the vertical flow structure, and in particular the size of particles in comparison to the thickness of the laminar sublayer, may be explanatory of the different velocity distributions resolved in bedload transport experiments, and future studies should investigate this possibility.
@@
 An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates from granular interactions.
-Particle size enters the Langevin model eq. \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
-The analytical velocity distribution eq.
-\ref{eq:steadystate} was fit to six different experimental datasets in fig.
-\ref{fig:fig4ch5}, and the best fit parameters were presented in tab.
+Particle size enters the Langevin model Eq. \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
+The analytical velocity distribution Eq.
+\ref{eq:steadystate} was fit to six different experimental datasets in Fig.
+\ref{fig:fig4ch5}, and the best fit parameters were presented in Tab.
 \ref{tab:calib}.
@@
 
-Because the elasticity $\ve$ lumps together collision geometry and dissipation effects, it is challenging to attribute the increasing relationship between elasticity and particle size evident in tab.
+Because the elasticity $\ve$ lumps together collision geometry and dissipation effects, it is challenging to attribute the increasing relationship between elasticity and particle size evident in Tab.
 \ref{tab:calib} directly to particle size.
@@
 These efforts would produce additional insight into the controls of fluid turbulence and transverse, vertical, and rotational movements on the downstream velocities of individual bedload grains.
-
-\endinput
-
-Sediment transport experiments reveal correlations between particle size and the shape of the bed load velocity distribution.
-Experiments with smaller particles tend to give exponential distributions, and those with larger particles give Gaussian distributions.
-In fluid dynamics, the dissipation characteristics of particle-particle collisions in viscous flows are known to depend on the particle size and approach velocity through the Stokes number. 
-In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic grains generate non-Gaussian distributions \citep{Chapman1970,Brilliantov2004}.
-Taken together, these ideas suggest that we might relate the shape of the particle velocity distribution to particle size.
-We can estimate typical Stokes numbers of colliding bed load particles in experiments as $...$, using with the flow shear velocity and mean streamwise sediment velocity to calculate $V$.
-Estimating in this way, transport experiments with exponential velocities have $St \sim 1-10$, those with neither exponential nor Gaussian velocities have $St \sim 10-100$, and those with Gaussian velocities have $St > 100$.
-In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2011,Joseph2001,Yang2006}.
-Although collision geometry and grain shape certainly complicate the narrative, these values of $St$ are consistent with our model conclusion that the shape of the particle velocity depends on the elasticity of collisions.
-
-In real channels, grain sizes often span a wide range.
-A major implication of the dependence of the shape of the particle velocity distribution on grain size is that different grain sizes in a mixture will impart distinct fluctuation signatures to the overall bulk transport rate.
-Even in the absence of sorting effects and differential mobility, smaller grains can be expected to carry more control over the largest fluctuations in the overall transport rate, since their velocity distributions have wider tails.
-
-Particle trajectories are a compromise between turbulent drag and momentum dissipation by particle-bed collisions \citep{Schmeeckle2014}.
-For particles submerged in a viscous fluid, the amount of momentum dissipated by collisions is controlled by the particle Stokes number ($\St$), which characterizes the balance of pre-collisional momentum against the viscous pressure generated in the squeeze film between approaching particles \citep{}.
-The Stokes number is defined by $\St = \rho_s V d/(9\nu \rho)$, where $\rho_s$ and $\rho$ are the solid and fluid densities, $V$ is the approach velocity, $d$ is the particle size, and $\nu$ is the fluid viscosity.
-Momentum dissipation in submerged collisions shows a sharp dependence on $\St$ \citep{Joseph2001, Schmeeckle2001}, and hence also the particle size.
-This suggests particle size may have a nontrivial control over bedload velocity distributions.
-
-Here, motivated by the recognition that experimental particle velocity distributions appear to vary with the grain size, I hypothesize that the shape of the streamwise velocity distribution is controlled by the momentum dissipation characteristics of particle-bed collisions.
-For particles submerged in a viscous fluid, the amount of momentum dissipated by collisions is controlled by the particle Stokes number ($\St$), which characterizes the balance of pre-collisional momentum against the viscous pressure generated in the squeeze film between approaching particles \citep{}.
-The Stokes number is defined by $\St = \rho_s V d/(9\nu \rho)$, where $\rho_s$ and $\rho$ are the solid and fluid densities, $V$ is the approach velocity, $d$ is the particle size, and $\nu$ is the fluid viscosity.
-Momentum dissipation in submerged collisions shows a sharp dependence on $\St$ \citep{Joseph2001, Schmeeckle2001}, and hence also the particle size.
-This suggests particle size may have control over bedload velocity distributions.
-
-The elasticity of granular collisions within a viscous fluid depends on the particle size.
-
-In this drag force, the collision force (eq.
-\ref{eq:col}), the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions.
-This set of possiblities defines a statistical ensemble.

### 2021-09-24 revision/ch6conc.tex
getting close.

@@
 \begin{enumerate}
-	\item developed the linkage between sediment flux probability distributions and individual particle trajectories (ch. \ref{ch:flux}), 
-	\item described particle trajectories alternating through motion and rest with fluctuating velocities (ch. \ref{ch:flux}),
-	\item evaluated the control of bed elevation changes over sediment transport rates (ch. \ref{ch:ch3}),
-	\item characterized how long particles can remain buried within the sedimentary bed (ch. \ref{ch:ch3}),
-	\item incorporated the process of sediment burial in a description of downstream particle movement (ch. \ref{ch:downDiff}), and
-	\item formulated the velocity distributions of sediment particles including episodic particle-bed collisions (ch. \ref{ch:langevin}).
+	\item developed the linkage between sediment flux probability distributions and individual particle trajectories (Ch. \ref{ch:flux}), 
+	\item described particle trajectories alternating through motion and rest with fluctuating velocities (Ch. \ref{ch:flux}),
+	\item evaluated the control of bed elevation changes over sediment transport rates (Ch. \ref{ch:ch3}),
+	\item characterized how long particles can remain buried within the sedimentary bed (Ch. \ref{ch:ch3}),
+	\item incorporated the process of sediment burial in a description of downstream particle movement (Ch. \ref{ch:downDiff}), and
+	\item formulated the velocity distributions of sediment particles including episodic particle-bed collisions (Ch. \ref{ch:langevin}).
 \end{enumerate}
@@
 
-The solutions of the master equation in a given problem produce the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (chs.
-\ref{ch:flux} and \ref{ch:downDiff}), the particle velocity (ch.
-\ref{ch:langevin}), the bed elevation (ch.
-\ref{ch:ch3}), and the sediment transport rate (ch.
+The solutions of the master equation in a given problem produce the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (Chs.
+\ref{ch:flux} and \ref{ch:downDiff}), the particle velocity (Ch.
+\ref{ch:langevin}), the bed elevation (Ch.
+\ref{ch:ch3}), and the sediment transport rate (Ch.
 \ref{ch:flux}).
@@
 A major challenge in this research is that only a handful of noises in statistical physics are comprehensively understood \citep{Horsthemke1984}, so the available options in stochastic modeling are constrained.
-The noises used in this thesis include white Gaussian, Poisson pulse, and dichotomous noises, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
-Turbulence was described using Gaussian noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise.
+The noises used in this thesis include Gaussian white noise, Poisson pulse noise, and dichotomous noise, representing erratic fluctuations, sequences of spikes, and random switches respectively \citep{VanDenBroeck1983}.
+Turbulence was described using Gaussian white noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise.
 Alternation between motion and rest was represented with dichotomous noise.
@@
 In contrast, colored noises in which fluctuations have favored frequencies are well-known to occur in many fluid and granular physics phenomena, most famously in fluid turbulence \citep{Kolmogorov1941,Nikora2000} and granular collapse \citep{Bak1987,Jensen1998}.
-In river science, many phenomena exhibit colored spectra, like the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
-It will be problematic if these phenomena require colored noise to be incorporated into dynamical equations for an adequate description.
-Dynamical equations driven by colored noise are extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
-Even if white noise models like some of those developed in this thesis are not exactly accurate, they remain necessary as a basis for comparison when formulating and solving colored noise models \citep{Fox1986,Moss1989}.
+In river science, many phenomena exhibit colored spectra, including the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
+Unfortunately, dynamical equations driven by colored noise are extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
+In this context, even if the idealized noise models developed in this thesis are not exactly accurate, their development remains necessary as a basis for comparison when formulating and solving colored noise models \citep[e.g.][]{Fox1986,Moss1989}.
 
@@
 
-The first major contribution of this thesis is in ch.
+The first major contribution of this thesis is in Ch.
 \ref{ch:flux}.
 Here, I formulated the probability distribution of the sediment flux from the trajectories of individual particles moving downstream.
-This work unifies the sediment trajectory models originating from Einstein (secs.
-\ref{sec:einwalk}-\ref{sec:lisle}) with the renewal approach to calculate the sediment flux (sec.
+This work unifies the sediment trajectory models originating from Einstein (Secs.
+\ref{sec:einwalk}-\ref{sec:lisle}) with the renewal approach to calculate the sediment flux (Sec.
 \ref{sec:renewal}).
@@
 \citet{Ballio2018} explained that scale-dependence originates from individual particle trajectories, but this had not been described in a mathematical model until now.
-Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (sec.
-\ref{sec:einflux}), and an emerging body of research has produced the full probability distribution of the flux without referencing individual movement characteristics (secs.
+Descriptions of the mean sediment flux from the movement characteristics of individual grains have existed now for a long time (Sec.
+\ref{sec:einflux}), and an emerging body of research has produced the full probability distribution of the flux without referencing individual movement characteristics (Secs.
 \ref{sec:birthdeath}-\ref{sec:renewal}).
-This work unifies these two research themes and presents a statistical mechanics formulation of bed load sediment transport based on individual particle trajectories.
+This work unifies these two research themes and presents a statistical mechanics formulation of the bed load sediment flux probability distribution based on individual particle trajectories.
 
@@
 
-Second, in ch. \ref{ch:flux} I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
-Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (sec.
-\ref{sec:einwalk}), and this was later improved to include the duration of motion (sec.
+Second, in Ch. \ref{ch:flux} I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
+Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (Sec.
+\ref{sec:einwalk}), and this was later improved to include the duration of motion (Sec.
 \ref{sec:lisle}).
-Until this thesis, movement velocities in analytical models were considered constant which contrasts with reality. 
+Until this thesis, movement velocities in analytical models were considered constant, which contrasts with reality. 
 Although some numerical models have described motion/rest cycles with velocity fluctuations \citep{Fan2016,Bialik2012,Schmeeckle2014}, they had not resolved the novel three-range diffusion characteristics these dynamics imply.
 
-This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state; it introduces a dimensionless Peclet number as an important characteristic of bedload sediment transport; and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains.
-These developments constitute a new understanding of bedload movement across its timescales.
+This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state; it introduces a dimensionless P\'{e}clet number as an important characteristic of bedload sediment transport; and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains.
+These developments constitute a new understanding of bedload movement across its timescales and invite generalizations aimed at pushing the approach further to including interactions between particles.
 
@@
 
-Third, ch.
-\ref{ch:ch3} modified the birth-death description of bedload transport (sec.
+Third, Ch.
+\ref{ch:ch3} modified the birth-death description of bedload transport (Sec.
 \ref{sec:birthdeath}) to include feedbacks between the local bed elevation and the entrainment and deposition rates.
@@
 
-Fourth, ch.
-\ref{ch:downDiff} presents a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (sec.
+Fourth, Ch.
+\ref{ch:downDiff} presents a model of sediment trajectories through motion, rest, and burial, describing sediment transport across local, intermediate, global, and geomorphic ranges (Sec.
 \ref{sec:trajmodels}), and producing new understanding of how exactly the distinct spreading characteristics of particles within each of these ranges arise \citep[e.g][]{Pretzlav2021}.
@@
 
-Finally, ch. \ref{ch:langevin} displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
-This work provides the first comprehensive description of all bedload velocity distributions observed in experiments (sec.
-\ref{sec:langexperimentcomparison}), while earlier works had described only particular end-member distributions (sec.
+Finally, Ch. \ref{ch:langevin} displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
+This work provides the first comprehensive description of all bedload velocity distributions observed in experiments to the present time (Sec.
+\ref{sec:langexperimentcomparison}), while earlier works had described only particular end-member distributions (Sec.
 \ref{sec:langevin}).
+This work incorporates ideas from the kinetic theory of granular gases into the description of weak bedload transport and invites generalizations that explicitly include the geometry of particle-particle collisions.
 
@@
 
-This thesis has produced new understanding of how to describe bedload transport using statistical physics, but its approach has many limitations deserving of future research attention.
+This thesis has produced new understanding of how to describe bedload transport using statistical physics, but the developments introduce many limitations deserving of future research attention.
 In some cases these are specific to the models produced in the thesis, but in others, they are shared in common with a majority of models in the stochastic sediment transport research paradigm \citep{Ancey2020,Furbish2021a}.
@@
 
-To date, very little work has concentrated on stochastic sediment transport models in unsteady conditions \citep[e.g.][]{Bohorquez2016}.
-The most obvious scheme to address unsteadiness is to introduce time and space dependence to movement velocities, diffusivities, or entrainment and deposition rates, but how exactly one should express these rates in terms of time and space is a matter for speculation, and would be contingent on a given context given our limited understanding of how these parameters relate to the flow hydraulics \citep[e.g.][]{Heyman2016}.
+To date, very little work has concentrated on developing stochastic sediment transport models for unsteady conditions \citep[e.g.][]{Bohorquez2016}.
+The most obvious scheme to address unsteadiness is to introduce time and space dependence to movement velocities, diffusivities, or entrainment and deposition rates, but how exactly one should express these rates in terms of time and space is a matter for speculation.
+These expressions would be contingent on a given context, given our limited understanding of how these parameters relate to the flow hydraulics and granular physics \citep[e.g.][]{Heyman2016}.
 
-Future studies might investigate in more depth the linkage between flow hydraulics, the geometric arrangement of grains on the bed, and the bed topography with the entrainment and deposition rates of individual particles to develop the foundational knowledge to incorporate temporally or spatially variable flow and sedimentary conditions into stochastic models of sediment transport.
+Future studies might investigate in more depth the linkage between flow hydraulics, the geometric arrangement of grains on the bed, and the bed topography with the entrainment and deposition rates of individual particles to develop our foundational knowledge to later incorporate temporally or spatially variable flow and sedimentary conditions into stochastic models of sediment transport.
 
@@
 Simplified experimental geometries based on bimodal sediment beds \citep[e.g.][]{Houssais2012} would be a great context to revisit these issues.
-Extending the motion-rest model of ch.
+Extending the motion-rest model of Ch.
 \ref{ch:ch3} to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
@@
 The final limitation I will mention is that all of the efforts in this thesis were concentrated on weak bedload transport, where densities of moving particles are low enough that they may interact with the static bed but never with each other.
-This assumption is justified because weak bedload transport conditions are typical of gravel bed rivers \citep{Ashworth1989,Warburton1992}, but it is nonetheless a limitation given the diversity of processes which move sediment over Earth's surface.
+This assumption is justified insofar as weak bedload transport conditions are common within gravel-bed rivers \citep{Ashworth1989,Warburton1992}, but it is nonetheless a limitation given the diversity of processes which move sediment over Earth's surface.
 
@@
 
-The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscales without including variability, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving surface grain size distributions \citep{Parker1982,Andrews1983}.
+The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscales without including variability in models, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving surface grain size distributions \citep{Parker1982,Andrews1983}.
 
-The stochastic descriptions of sediment transport developed in this thesis hint toward a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?
-
-\endinput
-
-Its closest neighbours may be rarefied hillslope transport, where solitary grains tumble down hillslopes \citep{Williams2021}, and intense bedload transport, where particles creep downstream in a dense granular flow, supported by collisions with other moving grains \citep{Frey2014}.
-These phenomena differ only by viscosity or density, not mechanically.
-
-\citet{Ashworth1989} and \citet{Warburton1992} identified three phases of bedload transport in gravel-bed streams that occur as the flow increases.
-Phase 1 is characterized by movement of sand only, while gravel remains stationary; phase 2 involves partial mobility of the smaller gravel sizes; and phase 3 is the full mobility of all grain sizes represented on the bed.
-Phase 2 is most common in gravel-bed rivers, but phase 3 has
+The stochastic descriptions of sediment transport developed in this thesis show a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?

### 2021-09-25 revision/ack.tex
about to track changes.

@@
 
-Finally, Mary! My PhD has not been great in terms of work-life balance. Thanks for putting up with my late bedtimes, early wakeups, rambling science jargon, and skipped social events. It's done!
+Finally, Mary! My PhD has not been great in terms of work-life balance. Thanks for putting up with my late bedtimes, early wake-ups, rambling science jargon, and skipped social events. It's done!
 
@@
 It's an expression of 30 years of influence and guidance from other people at every stage, my parents, my sisters, my grandmother, and a huge network of teachers and mentors. Thank you all!
-
-\endinput
-
-Mom, during my childhood, you worked full time, taught piano lessons, earned Master and Doctoral degrees, played organ every Sunday, consoled us through every difficulty, and engaged us in as many extracurricular activities as we were willing to take on.
-I cannot imagine a better role model of work ethic, open-mindedness, and devotion. I hope I picked a few things up!
-
-Dad, in my childhood I watched you solve an insane diversity of problems without hesitation, from easy ones, like worn out brake pads or stuck chainsaws, to hard ones, like burnt eyes and broken arms.
-On canoeing trips I got lessons in water currents, fish habits, tree species, electrons, engines, wings, hellgrammites, heat transfer, world history, radio waves, and probably everything else I could think of.
-Years of free science lessons!
-You taught me more about science and problem solving than school ever did.
-"Any problem can be fixed with the right tools" is basically the Jim Pierce philosophy.
-It it's incredibly useful in a PhD!
-Thanks for this and so much more.
-
-Kelsey and Kim, my earliest memories are sitting in your bedroom floor struggling through your books, asking for help every couple of paragraphs.
-You made me good at reading and taught me to enjoy it, and this is the basis of everything.
-Kim, you were my first role model, and I think of you all the time.
-Kelsey, I know it goes without saying, but thanks for a lifetime of guidance.
-You're now, as always (all state chorus, math field day, spelling bees, piano skills, grades, .
-.
-.
-), five steps ahead and far more successful.
-On the one hand, I'm deeply proud at your progress through life, with a stable career and now two beautiful kids, but on the other, I'm flatly shocked by it.
-I'll be playing catch up forever!
-
-Gugs, you're easily the most level-headed, good-natured, and reliable person I've ever had contact with, and to think I was lucky enough to be about half raised by you!
-You've been my second mother, my counsellor, at times my daily chauffeur, and my best role model of grace, consistency, and even-handedness.
-I'll never be able to thank you enough!
-One day I'll finally take on your 5am wakeups and figure out how to copycat your completely measured responses to all of life's challenges.
-
-Finally with regard to family, I'd like to thank my great aunt Alice.
-Thanks so much for paying my undergrad education!
-We hardly met before you passed, but we're family, and that was enough.
-It really means a lot.
-
-Next comes the mentors in education.
-
-I'm kinda tired because it's the end, so it's natural to drift on to recollections of my favorite times in life, Watoga and Williams River with my family, encyclopedias in Gugs' living room, watching eagles over the South Branch with dad, discussions with Johnathan Ramey about my education, the glowing fireflies under my great grandpa's tree, ``cross-up, cross-up, squeeze" in Mindy's calc class, or laughter about stolen watermelons with my sisters.
-I can skip years later and flip on the light in the physics lounge to find Megat asleep on the couch, visit Gugs on a trip back home, discuss Tsujimoto with Marwan, or walk through Kitsilano with Mary at the edge of dark.
-The PhD is all of it!
-Thank you all for every part of it.

### 2021-09-25 revision/app_ch2.tex
about to track changes.

@@
 
-\chapter{Calculations involved in dynamical sediment flux model}
+\chapter{Calculations involved in the dynamical bedload flux model}
 \label{sec:appendixAfluc}
@@
 Now I will take time derivatives in order to apply the so-called Furutsu-Novikov formula to evaluate the ensemble average over dichotomous noise \citep{Shapiro1978}.
-Applied to the dichotomous noise in figure \ref{fig:lislefig}, the Furutsu-Novikov formula is, for an arbitrary functional $F$ of the noise,
+Applied to the dichotomous noise in Fig.
+\ref{fig:lislefig}, the Furutsu-Novikov formula is, for an arbitrary functional $F$ of the noise,
 \be \pt \bra \eta(t)F[\eta(t)] \ket = \bra \eta \pt F \ket + k \big[\bra \eta \ket \bra F \ket -\bra \eta F \ket \big]  \label{eq:furuti}.\ee
 
-Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of Eq. \ref{eq:logi} gives
+Making the short-hands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of Eq. \ref{eq:logi} gives
 \be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket. \label{eq:furut}\ee
@@
 The numerator terms encode the initial conditions. The denominator terms are where the real structure of the solution is contained.
-The numerator terms involving $g$ are easily expressed as first and second derivatives with respect to $x$. The problem, then is to calculate the inverse Fourier integral of the denominator.
+The numerator terms involving $g$ can be expressed as first and second derivatives with respect to $x$. The problem, then is to calculate the inverse Fourier integral of the denominator.
 The inverse Fourier transform can be calculated by expanding the denominator of Eq. \ref{eq:prev} in partial fractions and applying the contour integral \citep[e.g.][]{Arfken1985}
 \be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g - i c} dg = \sigma(c)\theta(-x\sigma(c))\exp\big(-|c x|\big). \label{eq:contour}\ee
-In this equation $\sigma(c) = \text{sgn}(c)$ is a shorthand for the signum function.
+In this equation $\sigma(c) = \text{sgn}(c)$ is a shorthand for the sign function.
 The above integral can be obtained with the residue theorem by considering different cases for the signs of $x$ and $c$ and applying square contours around the upper or lower-half regions of the complex plane.
@@
 Rearranging the governing equation gives
-\be \tilde{\hat{P}}(g,s) = \frac{\varphi D g^2 - ig V\varphi  + s + k}{D(s+k_E)}\frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}}\ee
+\be \tilde{\hat{P}}(g,s) = \frac{\varphi D g^2 - ig V\varphi  + s + k}{D(s+k_E)}\frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}}.\ee
 The roots of the denominator are at
@@
 where
-\be R = \sqrt{1 + \frac{4D}{V^2}\frac{s(s+k)}{s+k_E}}\ee
+\be R = \sqrt{1 + \frac{4D}{V^2}\frac{s(s+k)}{s+k_E}}.\ee
 The partial fractions expansion is therefore
-\be \frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}} = \frac{D}{i V R}\Big[ \frac{1}{g-g_+} - \frac{1}{g-g_-}\Big]\ee
-Giving 
-\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi\px + s + k}{VR(s+k_E)}\Bigg[ \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_+} - \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_-}\Bigg]\ee
+\be \frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}} = \frac{D}{i V R}\Big[ \frac{1}{g-g_+} - \frac{1}{g-g_-}\Big],\ee
+giving 
+\begin{multline} \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi\px + s + k}{VR(s+k_E)} \\ \times \Bigg[ \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_+} - \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_-}\Bigg],\end{multline}
 or eventually
-\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\Bigg[ \theta(x) \exp\Big(\frac{Vx}{2D}(1-R)\Big) + \theta(-x)\exp\Big( \frac{Vx}{2D}(1+R)\Big) \Bigg].\ee
+\begin{multline}
+ \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)} \\ \times \Bigg[ \theta(x) \exp\Big(\frac{Vx}{2D}(1-R)\Big) + \theta(-x)\exp\Big( \frac{Vx}{2D}(1+R)\Big) \Bigg].\end{multline}
 This simplifies slightly to
-\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\exp\Big[\frac{Vx}{2D} - \frac{V|x|}{2D}R \Big], \label{eq:laplaceapp}\ee
-where as a reminder, $\varphi = k_D/k$.
+\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\exp\Big[\frac{Vx}{2D} - \frac{V|x|}{2D}R \Big] \label{eq:laplaceapp}.
+\ee
 
 Taking the inverse Laplace transform symbolically provides the desired probability density function
-\be P(x,t) = \mathcal{L}^{-1}\Bigg\{ \frac{1}{V} \big[-\varphi D\px^2 + V\varphi \px + k + s \big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(-\frac{V|x|R}{2D} \Big)}{(s+k_E)R} \Bigg\}(t). \ee
+\begin{multline} P(x,t) = \mathcal{L}^{-1}\Bigg\{ \frac{1}{V} \big[-\varphi D\px^2 + V\varphi \px + k + s \big] \\ \times \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(-\frac{V|x|R}{2D} \Big)}{(s+k_E)R} \Bigg\}(t).
+\end{multline}
 Using the shift property of Laplace transforms \citep[e.g.][]{Arfken1985}, this becomes
-\be P(x,t) = \frac{1}{V}e^{-k_E t} \mathcal{L}^{-1}\Bigg\{ \big[-\varphi D\px^2 + V \varphi \px + k_D + s\big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(\frac{-V|x|R_\ast}{2D}\Big)}{s R_\star} \Bigg\},\ee
+\begin{multline} P(x,t) = \frac{1}{V}e^{-k_E t} \mathcal{L}^{-1}\Bigg\{ \big[-\varphi D\px^2 + V \varphi \px + k_D + s\big] \\ \times \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(\frac{-V|x|R_\ast}{2D}\Big)}{s R_\star} \Bigg\},\end{multline}
 where 
@@
 Using the property that $\mathcal{L}^{-1}\big\{s \tilde{f}\big\} = (\delta(t) + \pt)f(t)$ \citep{Arfken1985} gives
-\be P(x,t) = \frac{1}{V} e^{-k_E t} \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big] \mathcal{L}^{-1}\Bigg\{ \frac{-\exp\Big[\frac{-V|x|R_\ast}{2D}\Big]}{sR_\ast} \Bigg\}(t) \ee
+\begin{multline}
+ P(x,t) = \frac{1}{V} e^{-k_E t} \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \\ \times \exp\Big[\frac{Vx}{2D}\Big] \mathcal{L}^{-1}\Bigg\{ \frac{-\exp\Big[\frac{-V|x|R_\ast}{2D}\Big]}{sR_\ast} \Bigg\}(t).
+\end{multline}
 Finally, using the identity
@@
 \ee
-from \citet{Bateman1953}, pg. 133 (which can be derived by either u-substitution or series expansion in powers of $a$) provides:
-\begin{multline}  P(x,t) = \frac{1}{V} e^{-k_E t}  \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big]\\ \times \int_0^t du \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) \\ \times \mathcal{L}^{-1}\Bigg\{ \frac{\exp\Big[ \frac{-V|x|}{2D}\sqrt{a+b s}\Big]}{\sqrt{a + b s}}\Bigg\}(u)  \label{eq:magic}\end{multline}
-where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$ are shorthands.
+from \citet{Bateman1953}, pg. 133 (which can be derived by either u-substitution or series expansion in powers of $a$) provides
+\begin{multline}  P(x,t) = \frac{1}{V} e^{-k_E t}  \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big]\\ \times \int_0^t du \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) \\ \times \mathcal{L}^{-1}\Bigg\{ \frac{\exp\Big[ \frac{-V|x|}{2D}\sqrt{a+b s}\Big]}{\sqrt{a + b s}}\Bigg\}(u) , \label{eq:magic}\end{multline}
+where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$ are short-hands.
 
@@
 \label{sec:fluxlimits}
-Equation \ref{eq:fluxy} is difficult to work with.
+Eq.
+\ref{eq:fluxy} is difficult to work with.
 To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep{Weiss1994} to invert the Laplace-transformed rate function of Eq.
@@
 This has two scaling limits within it. Provided that $T \ll 4D/V^2/\pi < 2 D/V^2$, the scaling goes as $\Lambda(T) \sim T^{-1/2}$. But if $T\gg 4D/V^2/\pi$, it goes as $\Lambda(t) \sim T$.
-
 For large times, taking $s\rightarrow 0 $ gives

### 2021-09-25 revision/app_ch3.tex
about to track changes.

@@
 \label{sec:arr}
-The Gillespie Stochastic Simulation Algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
+The Gillespie stochastic simulation algorithm (SSA) generates exact realizations of a Markov random process from a sequence of random numbers.
 It was originally developed for chemical physics by \citet{Gillespie1977} and is reviewed in \citet{Gillespie1991} and \citet{Gillespie2007}.
@@
 The joint description of bedload transport and bed elevation changes is characterized by a set of states $(n,m)$ where $n$ and $m$ are integers.
-The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in Eqs. (\ref{eq:rate1}-\ref{eq:rate4}).
+The description involves four possible transitions (migration in, entrainment, deposition, migration out) with rates given in Eqs. \ref{eq:rate1}-\ref{eq:rate4}.
 
@@
 \be Q(\tau|n,m) = A(n,m)e^{-A(n,m)\tau}. \label{eq:exp}\ee
-This equation demonstrates that the time $\tau$ to the next transition from a state $(n,m)$ is exponentially distributed with mean value $\bar{\tau} = 1/A(n,m).$ This means if the stochastic process transitioned to the state $(n,m)$ at a time $t$, the next transition will occur at a time $t+\tau$ with $\tau$ a random variable drawn from the exponential distribution \ref{eq:exp}.
+This equation demonstrates that the time $\tau$ to the next transition from a state $(n,m)$ is exponentially distributed with mean value $\bar{\tau} = 1/A(n,m).$ This means if the stochastic process transitioned to the state $(n,m)$ at a time $t$, the next transition will occur at a time $t+\tau$ with $\tau$ a random variable drawn from the exponential distribution in Eq.
+\ref{eq:exp}.
 
@@
 
-\subsection{Pseudo code for the Gillespie SSA}
+\subsection{Pseudo-code for the Gillespie SSA}
 \label{sec:crr}
@@
 
-\section{Approximate solutions of the Master equation}
+\section{Approximate solutions of the master equation}
 
@@
 after applying the normalization condition $G(1)=1$.
-From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008} and stated in Eq.
-\ref{eq:ancey} and introduced originally in Sec.
+From the definition of $G$ it follows that $A(n) = \frac{1}{n!} \frac{d^nG}{dx^n}|_{x=0}$, giving the negative binomial distribution of particle activity demonstrated by \citet{Ancey2008}, stated in Eq.
+\ref{eq:ancey}, and introduced originally in Sec.
 \ref{sec:birthdeath}.
@@
 
-This can be solved using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete Master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$.
+This can be solved using the Fokker-Planck expansion \citep{Gardiner1983} that effectively converts this discrete master equation for $m$ into a diffusion equation for the quasi-continuous variable $z=z_1 m$.
 This works since $z_1$ is small.
@@
 \be 0 = -2\overline{\kappa}z_1[zM(z)]' + z_1^2 M''(z). \ee
-Taking into account that this distribution should be normalizable, $\lim_{z\rightarrow \pm \infty}M(z) = 0$, the solution is
+Taking into account that this distribution should be normalizable, \newline $\lim_{z\rightarrow \pm \infty}M(z) = 0$, the solution is
 \be M = M_0e^{-\kappa (z/z_1)^2}\ee

### 2021-09-25 revision/app_ch4.tex
about to track changes.

@@
 
-Owing to the convolution structure  of manuscript equations (1-3), their solution is a formidable problem.
+Owing to the convolution structure  of Eqs. \ref{eq:g1} and \ref{eq:g2}, their solution is a formidable problem.
 Luckily, we have the device of Laplace transforms.
-These project integro-differential equations into an alternate space in which convolutions are unraveled \citep[e.g.,][]{Arfken1985}.
+These project integro-differential equations into an alternate space in which convolutions are unraveled \citep[e.g.][]{Arfken1985}.
 The double Laplace transform of a joint probability distribution $p(x,t)$ is defined by 
 \be \tilde{p}(\eta,s) = \int_0^\infty dx e^{-\eta x}\int_0^\infty dt e^{-st} p(x,t). \label{eq:doubletransform}\ee
-The Laplace-transformed moments of $x$ are linked to derivatives of the double transformed distribution (Eq. \ref{eq:doubletransform}) \citep[cf.,][]{Berezhkovskii2002}.
+The Laplace-transformed moments of $x$ are linked to derivatives of the double-transformed distribution Eq. \ref{eq:doubletransform} \citep[cf.][]{Berezhkovskii2002}.
 Eq. (\ref{eq:doubletransform}) implies
 \be \bra \tilde{x}(s)^k \ket = (-)^k\partial_\eta^k \tilde{p}(\eta,s)\Big|_{\eta=0}.\label{eq:momenttrick}\ee
-The operator $\bra \circ \ket$ denotes the ensemble average \citep[e.g.,][]{Kittel1958}.
-This means we can compute the variance of position as $\sigma_x^2(t) = \bra x^2 \ket - \bra x \ket^2 = \El^{-1} \{\bra\tilde{x}^2 \ket;t\} - \El^{-1} \{\bra\tilde{x} \ket;t\}^2$, where $\El^{-1}$ denotes the inverse Laplace transform \citep[e.g.,][]{Arfken1985}.
-This is a powerful tool, since we can use it to derive the positional variance without integrating the distribution in equation (7) of the manuscript.
+This means the variance of position can be derived as as $\sigma_x^2(t) = \bra x^2 \ket - \bra x \ket^2 = \El^{-1} \{\bra\tilde{x}^2 \ket;t\} - \El^{-1} \{\bra\tilde{x} \ket;t\}^2$, where $\El^{-1}$ denotes the inverse Laplace transform \citep[e.g.][]{Arfken1985}.
 
-Double transforming manuscript equations (1-3) using the definition Eq. \ref{eq:doubletransform} gives
-
+Double transforming Eqs. \ref{eq:g1} using the definition Eq. \ref{eq:doubletransform} gives
 \begin{alignat}{2}
-	&\tom_{1T}(\eta,s) &&= \theta_1 \tg_1(\eta,s) + \tom_2(\eta,s)\tg_1(\eta,s)-\tom_{1F}(\eta,s),\\
-	&\tom_{1F}(\eta,s) &&= \theta_1\tg_1(\eta,s+\kappa) + \tom_2(\eta,s)\tg_1(\eta,s+\kappa),\\
-	&\tom_2(\eta,s) &&= \theta_2 \tg_2(\eta,s) + \tom_{1F}(\eta,s)\tg_2(\eta,s).
+	&\tom_{0}(\eta,s) &&= \theta_1 \tg_1(\eta,s) + \tom_2(\eta,s)\tg_1(\eta,s)-\tom_{1}(\eta,s),\\
+	&\tom_{1}(\eta,s) &&= \theta_1\tg_1(\eta,s+\kappa) + \tom_2(\eta,s)\tg_1(\eta,s+\kappa),\\
+	&\tom_2(\eta,s) &&= \theta_2 \tg_2(\eta,s) + \tom_{1}(\eta,s)\tg_2(\eta,s).
 \end{alignat}
@@
 \begin{alignat}{2}
-	&\tom_{1T}(\eta,s) &&= \frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\big\{\tg_1(\eta,s)-\tg_1(\eta,s+\kappa) \big\}, \label{eq:A} \\
-	&\tom_{1F}(\eta,s) &&= \frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\tg_1(\eta,s+\kappa),\\
+	&\tom_{0}(\eta,s) &&= \frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\big\{\tg_1(\eta,s)-\tg_1(\eta,s+\kappa) \big\}, \label{eq:A} \\
+	&\tom_{1}(\eta,s) &&= \frac{\theta_1 + \theta_2 \tg_2(\eta,s)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\tg_1(\eta,s+\kappa),\\
 	&\tom_{2}(\eta,s) &&= \frac{\theta_2 + \theta_1 \tg_1(\eta,s+\kappa)}{1-\tg_1(\eta,s+\kappa)\tg_2(\eta,s)}\tg_2(\eta,s). 
 \end{alignat}
-Double transforming manuscript equations (4-6) gives
+Double transforming Eqs. \ref{eq:g2} gives
 \begin{align}
-	\tp_0(\eta,s) &= \frac{1}{s}\tom_{1T}(\eta,s),\\
+	\tp_0(\eta,s) &= \frac{1}{s}\tom_{0}(\eta,s),\\
 	\tp_1(\eta,s) &= \theta_1 \tG_1(\eta,s) + \tom_2(\eta,s) \tG_1(\eta,s),\\
-	\tp_2(\eta,s) &= \theta_2 \tG_2(\eta,s) + \tom_{1F}(\eta,s)\tG_2(\eta,s).\label{eq:Z}
+	\tp_2(\eta,s) &= \theta_2 \tG_2(\eta,s) + \tom_{1}(\eta,s)\tG_2(\eta,s).\label{eq:Z}
 \end{align}
@@
 \end{multline}
-Plugging the propagators outlined in manuscript equations (8-9) into Eq. \ref{eq:lap} gives 
+Plugging the propagators outlined in Sec. \ref{sec:solution} into Eq. \ref{eq:lap} gives 
 \be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}.\label{eq:nicedist}\ee
-In this equation, $k'=k_1+k_2$, and we have used the normalization requirement of the initial probabilities: $\theta_1 + \theta_2 = 1.$
+In this equation, $k'=k_1+k_2$, and the normalization requirement of the initial probabilities $\theta_1 + \theta_2 = 1$ has been incorporated.
 The double inverse transform of this equation provides the distribution $p(x,t)$.
-We invert the transform over $\eta$ first.
-Using the results 15.103 (transform of exponential), 15.123 (transform of derivative), and 15.141 (transform of Dirac delta function) from \citet{Arfken1985} provides 
+The transform over $\eta$ is inverted first.
+Using the results [15.103] (transform of exponential), [15.123] (transform of derivative), and [15.141] (transform of Dirac delta function) from \citet{Arfken1985} provides 
 \begin{multline} \tp(x,s) = \theta_1 \frac{s+\kappa}{s(s+\kappa + k_1)}\delta(x) + \frac{1}{v} \Big(\frac{(s+\kappa+k')s+\kappa k_2}{s(s+\kappa+k_1)} \\- \frac{\theta_1(s+\kappa)[s(s+\kappa+k_1)+\kappa k_2]}{s(s+\kappa+k_1)^2}\Big)
 	\exp\Big[-\frac{(s+\kappa+k')s+\kappa k_2}{s+\kappa+k_1}\frac{x}{v}\Big].\end{multline}
-Inverting the remaining transform over $s$, applying results 15.152 (substitution), 15.164 (translation), and 15.175 (transform of $te^{kt}$) from \citet{Arfken1985}, and defining the shorthand notations $\tau = k_1(t-x/v)$, $\xi = k_2 x/v$, and $\Omega = (\kappa + k_1)/k_1$, gives the simpler form
+Inverting the remaining transform over $s$, applying results [15.152] (substitution), [15.164] (translation), and [15.175] (transform of $te^{kt}$) from \citet{Arfken1985}, and defining the shorthand notations $\tau = k_1(t-x/v)$, $\xi = k_2 x/v$, and $\Omega = (\kappa + k_1)/k_1$, gives the simpler form
 \begin{multline}
@@
 \end{multline}
-Using entries 2.2.2.1, 2.2.2.8, and 1.1.1.13 from \citet{Prudnikov1992a} in conjunction with the definition of the Marcum Q-function $ \mathcal{P}_\mu(x,t)$ \citep{Temme1996}, and inserting the Heaviside functions to account for the fact that grains can neither travel backwards nor at speeds exceeding $v$, we finally arrive at manuscript equation (10) for the joint distribution $p(x,t)$.
+Using entries [2.2.2.1], [2.2.2.8], and [1.1.1.13] from \citet{Prudnikov1992a} in conjunction with the definition of the generalized Marcum Q-function $ \mathcal{Q}_\mu(x,t)$ \citep{Temme1996}, and inserting the Heaviside functions to account for the fact that grains can neither travel backwards nor at speeds exceeding $v$, then finally produces Eq.
+\ref{eq:pdf} for the joint distribution $p(x,t)$.
 
@@
 
-We compute the first two moments of position $x$ and ultimately its variance using Eq.
+The first two moments of position $x$ and ultimately the variance are computed using Eq.
 \ref{eq:momenttrick}.
-The first two derivatives of the double Laplace transformed distribution Eq.
+The first two derivatives of the double Laplace-transformed distribution Eq.
 \ref{eq:nicedist} are
@@
 \be \partial_\eta^2 \tp(\eta,s) = 2v^2 \frac{1}{s} \frac{(s+\kappa+k_1)[(s+\kappa + k')s+\kappa k_2][\theta_2(s+\kappa) + k_1]}{[\eta v(s+\kappa + k_1) + (s+\kappa + k')s+ \kappa k_2]^3}.\ee
-Evaluating these at $\eta=0$ and applying Eq. \ref{eq:momenttrick} provides the Laplace transformed moments
+Evaluating these at $\eta=0$ and applying Eq. \ref{eq:momenttrick} provides the Laplace-transformed moments
 \be  \frac{\bra\tilde{x}(s)\ket} {v} = \frac{1}{s}\frac{\theta_2(s+\kappa)+k_1}{(s+\kappa+k')s+\kappa k_2} = \frac{1}{s} \frac{\theta_2(s+\kappa)+k_1}{(s+a+b)(s+a-b)}\label{eq:lapmean},\ee
-\be \frac{\bra \tilde{x}^2(s) \ket}{2v^2} = \frac{1}{s} \frac{(s+\kappa+k_1)(\theta_2(s+\kappa)+k_1)}{[(s+\kappa+k')s+\kappa k_2]^2}=  \frac{1}{s}\frac{(s+\kappa+k_1)(\theta_2(s+\kappa)+k_1)}{(s+a+b)^2(s+a-b)^2}.\label{eq:lapsecondmom}\ee
+\begin{multline} \frac{\bra \tilde{x}^2(s) \ket}{2v^2} = \frac{1}{s} \frac{(s+\kappa+k_1)(\theta_2(s+\kappa)+k_1)}{[(s+\kappa+k')s+\kappa k_2]^2} \\=  \frac{1}{s}\frac{(s+\kappa+k_1)(\theta_2(s+\kappa)+k_1)}{(s+a+b)^2(s+a-b)^2}.\label{eq:lapsecondmom}\end{multline}
 The parameters $a= (\kappa+k')/2$ and $b^2 = a^2 -\kappa k_2$ were introduced to factorize the denominators.
-These equations can be inverted using the properties 15.164 (translation), 15.11.1 (integration), and 15.123 (differentiation) from  \citet{Arfken1985} after expansion in partial fractions.
+These equations can be inverted using the properties [15.164] (translation), [15.11.1] (integration), and [15.123] (differentiation) from  \citet{Arfken1985} after expansion in partial fractions.
 For the mean, the calculation is
 \begin{align}
+\begin{split}
 	\frac{2b}{v}\bra x \ket &= \big[\theta_2 + (k_1+\theta_2 \kappa)\int_0^t dt\big]\El^{-1}\Big\{ \frac{1}{s+a-b}-\frac{1}{s+a+b};t\Big\}\\
-	&= \Big[\theta_2 + \frac{k_1+\theta_2\kappa}{b-a}\Big]e^{(b-a)t} - \Big[\theta_2 - \frac{k_1+\theta_2\kappa}{a+b}\Big]e^{-(a+b)t} - \Big[\frac{k_1+\theta_2\kappa}{b-a} + \frac{k_1+\theta_2\kappa}{a+b}\Big].
+	&= \Big[\theta_2 + \frac{k_1+\theta_2\kappa}{b-a}\Big]e^{(b-a)t} - \Big[\theta_2 - \frac{k_1+\theta_2\kappa}{a+b}\Big]e^{-(a+b)t} \\
+ & \hspace{1cm}- \Big[\frac{k_1+\theta_2\kappa}{b-a} + \frac{k_1+\theta_2\kappa}{a+b}\Big].
+\end{split}
 \end{align}
-This equation rearranges to manuscript equation (11).
+This equation rearranges to Eq. \ref{eq:mean}.
 The second moment (Eq. \ref{eq:lapsecondmom}) is 
 \begin{multline}
-	\frac{2b^2}{v^2}\bra x^2 \ket = \Big[\theta_2(\delta(t) + \partial_t) + (\theta_2(2\kappa + k_1)+k_1) + (\kappa+k_1)(\theta_2\kappa+k_1)\int_0^t dt \Big] \\
-	\times \El^{-1}\Big\{ \frac{1}{(s+a-b)^2} + \frac{1}{(s+a+b)^2}-\frac{1}{b(s+a-b)}+\frac{1}{b(s+a+b)};t\Big\}.
+	\frac{2b^2}{v^2}\bra x^2 \ket = \Big[\theta_2(\delta(t) + \partial_t) + (\theta_2(2\kappa + k_1)+k_1) \\ + (\kappa+k_1)(\theta_2\kappa+k_1)\int_0^t dt \Big] \\
+	\times \El^{-1}\Big\{ \frac{1}{(s+a-b)^2} + \frac{1}{(s+a+b)^2} \\-\frac{1}{b(s+a-b)}+\frac{1}{b(s+a+b)};t\Big\}.
 \end{multline}
@@
 \end{multline}
-which evaluates to manuscript equation (12).
-Finally, $\sigma_x^2 = \bra x^2 \ket - \bra x \ket^2$ derives the variance in manuscript equation (13).
+which evaluates to Eq. \ref{eq:second}.
+Finally, $\sigma_x^2 = \bra x^2 \ket - \bra x \ket^2$ derives the variance in Eq. \ref{eq:var}.
 
@@
 
-We determine the diffusion exponents $\gamma$ in the local, intermediate, and global ranges using the two limiting cases described in the discussion of the manuscript.
-Limit (1) is $\kappa \rightarrow 0$. We take this limit in Eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} with initial condition $\theta_1=1$ to obtain
+The diffusion exponents $\gamma$ in the local, intermediate, and global ranges using the two limiting cases described in Sec. \ref{sec:amidonenow}.
+Limit (1) is $\kappa \rightarrow 0$. Taking this limit in Eqs. \ref{eq:lapmean} and \ref{eq:lapsecondmom} with initial condition $\theta_1=1$ obtains
 \begin{align}
@@
 \end{align}
-Inverting these equations provides the variance
+Laplace inverting and combining these equations provides the variance
 \be \sigma_x^2 = 2v^2\frac{k_1}{k'^4}\Big(k_1\big[\frac{1}{2} - k'te^{-k't} - \frac{1}{2} e^{-2k't}\big] + k_2\big[-2+k't + (2+k't)e^{-k't}\big]\Big).\label{eq:li}\ee
@@
 
-We further investigate limit (1) for arbitrary initial conditions.
-By applying Tauberian theorems, we assert the $ t \rightarrow 0$ variance is determined by the $s\rightarrow \infty$ limits of Eqs.
-\ref{eq:lapmean} and \ref{eq:lapsecondmom} \citep[e.g.,][]{Weiss1994, Weeks1998}.
+Limit (1) can be further investigated for arbitrary initial conditions.
+By applying Tauberian theorems, the $ t \rightarrow 0$ variance can be determined from the $s\rightarrow \infty$ limits of Eqs.
+\ref{eq:lapmean} and \ref{eq:lapsecondmom} \citep[e.g.][]{Weiss1994, Weeks1998}.
 Expanding these equations in powers of $1/s$ and inverting the resulting transforms gives
@@
 \be \sigma_x^2(t) \sim v^2\theta_1\theta_2t^2 + \frac{1}{3}v^2(\theta_1k_1+\theta_2k_2)t^3.\label{eq:init}\ee
-We have taken only leading order terms for any option of $\theta_1$ and $\theta_2$.
+This equation includes only leading order terms for any option of $\theta_1$ and $\theta_2$.
 Eq. \ref{eq:init} shows local range exponent $\gamma=2$ when initial conditions are mixed (both are non-zero) and $\gamma=3$ when initial conditions are pure (one is zero).
@@
 Expanding for small $t$ provides $\sigma_x^2(t) = 2D_d t$ -- intermediate range normal diffusion, while sending $t\rightarrow \infty$ provides $\sigma_x^2 = (2D_d\kappa + c^2)/\kappa^2$ -- a constant variance in the geomorphic range.
-The global range is characterized by competition between terms in Eq. \ref{eq:wuvar} and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.,][]{Wu2019}.
+The global range is characterized by competition between terms in Eq. \ref{eq:wuvar} and shows $2 \leq \gamma \leq 3$ depending on the ratio $k_1/\kappa$ \citep[cf.][]{Wu2019}.
 Finally, both Eqs. \ref{eq:li} and \ref{eq:wuvar} reduce to the Einstein result $\sigma_x^2(t) = 2D_d t$ in further simplified limits.

### 2021-09-25 revision/app_ch5.tex
about to track changes.

@@
 
-\chapter{Calculations involved in the Collisional Langevin model}
+\chapter{Calculations involved in the collisional Langevin model}
 
-\section{Derivation of Master Equation}
+\section{Derivation of master equation}
 \label{sec:langmasterderiv}
 To derive the master equation from Eq.
-\ref{eq:langevin}, we can temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 \tilde{D}} h$ with $h$ distributed as $f(h)$.
-We will later take a GWN limit on this noise.
+\ref{eq:langevin}, I temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 \tilde{D}} h$ with $h$ distributed as $f(h)$.
+I will later take a GWN limit from this noise.
 With this assumption, integrating Eq.
-\ref{eq:langevin} over a small time interval $\delta t$ and incorporating the Ito interpretation for the collision term provides
+\ref{eq:langevin} over a small time interval $\delta t$ and incorporating the It\^{o} interpretation for the collision term provides
 \be     
@@
 Considering the probability $P(u,t+\delta t)$ as a sum over possible paths from $P(u,t)$ develops 
-\begin{align} P(u,t+\delta t) =
-	(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\tilde{\Gamma} \delta t) P(w,t)\\ 
-	+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 \tilde{D}}h) P(w,t) \\ 
-	+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
+\begin{align} 
+\begin{split}
+P(u,t+\delta t) =
+	\hspace{4pt}&(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\tilde{\Gamma} \delta t) P(w,t)\\ 
+	&+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 \tilde{D}}h) P(w,t) \\ 
+	&+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
+\end{split}
 \end{align}
 Evaluating all integrals over $\delta$-functions provides 
-\begin{align} P(u,t+\delta t) = (1-r \delta t - \nu \delta t)P(u-\tilde{\Gamma} \delta t, t) \\+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2 \tilde{D}}h) \\+ \nu \delta t \int_0^1 \frac{ \tilde{D}\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).\end{align}
-Finally, take $\delta t \rightarrow 0$ and limit the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ \cite{VanDenBroeck1983}.
-This process finally obtains the master equation (\ref{eq:master}).
-\section{Derivation of Steady-state solution}
+\begin{align}
+\begin{split}
+ P(u,t+\delta t) = \hspace{4pt}&(1-r \delta t - \nu \delta t)P(u-\tilde{\Gamma} \delta t, t) \\&+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2 \tilde{D}}h) \\&+ \nu \delta t \int_0^1 \frac{ \tilde{D}\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).
+\end{split}
+\end{align}
+Finally, taking $\delta t \rightarrow 0$ and limiting the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ obtains the master equation \ref{eq:master} \cite[e.g.][]{VanDenBroeck1983}.
+\section{Derivation of steady-state solution}
 \label{sec:langsteadyderiv}
@@
 Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
-\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] \ee
+\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] ,\ee
 where the coefficients $R_l^\pm$ are the residues (at least up to a constant factor) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
 \be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
-The Fourier transform (Eq.
-\ref{eq:recursion}) has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$ in the limit.
+The Fourier-transformed Eq.
+\ref{eq:recursion} has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$.
 Taking this limit and evaluating the residues provides 
 \begin{multline} \tilde{P}(s) = \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
-	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})} \end{multline}
-Finally, we can invert the Fourier transforms term by term with contour integration (another use of the integral given in Eq.
-\ref{eq:contour}), and incorporate Eq.
-\ref{eq:q} to obtain the steady-state solution (Eq.
-\ref{eq:steadystate}).
+	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})}. \end{multline}
+Finally, the Fourier transforms are inverted term by term with contour integration (another use of the integral given in Eq.
+\ref{eq:contour}), and the definition of $q(z)$ from Eq.
+\ref{eq:q} is incorporated to obtain the steady-state solution quoted in Eq.
+\ref{eq:steadystate}.
 
@@
 $k=1$ provides the mean (given in Eq. \ref{eq:meanu})
-\be \langle u \rangle =  \frac{\tilde{\Gamma}}{1-\ve} = \frac{\Gamma}{\nu(1-\ve)} \ee
+\be \langle u \rangle =  \frac{\tilde{\Gamma}}{1-\ve} = \frac{\Gamma}{\nu(1-\ve)}, \ee
 while $k=2$ provides the second moment
@@
 given in Eq. \ref{eq:varu}.
-\section{Weak and strong collision limits}
+\section{The weak collision limit}
 \label{sec:langextremes}
@@
 Incorporating this transformation into Eq. \ref{eq:master} provides the ``normalized" master equation
-\begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 \end{multline}
+\begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 .\end{multline}
 This equation remains exact and is only a change of variables from Eq. \ref{eq:master}.

### 2021-09-25 revision/ch1intro0.tex
about to track changes.

--- 2021-09-24 revision/ch1intro.tex
@@
 \citet{Nikora2001,Nikora2002} provided a conceptual framework which helps to organize this complexity.
-Nikora et al. divided the downstream tractories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
+Nikora et al. divided the downstream trajectories of individual particles into three timescales, or ``ranges", termed local, intermediate, and global.
 The local range refers to the period of motion between subsequent interactions with the bed, when the particles accelerate downstream within the turbulent flow.
@@
 
-\subsection{Motivation: tracers and basic understanding}
+\subsection{Motivation: Tracers and basic understanding}
 
@@
 
-Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
+Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had traveled a net distance $x$ after a time $t$ has elapsed.
 In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this approach not formalized until much later \citep{Montroll1965}.
@@
 
-Einstein's assumptions imply that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
+Einstein's assumptions imply that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and deposition (step lengths) can be written
 \be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
@@
 The $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
-Figure \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as a sequence of steps and rests.
+Fig. \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as a sequence of steps and rests.
 
@@
 
-For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of Eq.
+For the initial condition that particles have a probability $k_E/k$ to start in motion, reflecting steady-state conditions as far as alternation between motion and rest are concerned \citep[e.g.][]{Ancey2006}, the solution of Eq.
 \ref{eq:lislemaster} is \citep{Lisle1998}
@@
 A majority of these approaches simulate the fluid without including the no-slip boundary conditions on individual particles \citep[e.g.][]{Vowinckel2014,Schmeeckle2014,Gonzalez2017}, although some works have resolved these particle-scale fluid boundary conditions with the immersed boundary method \citep[e.g.][]{Ji2013, Ji2014, Elghannay2017, Yousefi2020}.
-These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bed load transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
+These computational physics models produce impressive insight into the underlying granular and fluid mechanics producing bedload transport \citep{Frey2011}, but analytical models are nevertheless desired for further insight into the problem.
 
@@
 
-\citet{Fan2014} set out to describe exponenentially-distributed bedload particle velocities with the Langevin equation
+\citet{Fan2014} set out to describe exponentially-distributed bedload particle velocities with the Langevin equation
 \be \dot{u}(t) = -\Delta \text{sgn}(u) + F + \sqrt{2D}\xi(t). \label{eq:fanlangevin}\ee
@@
 
-\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
+\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the stream-wise velocity 
 \be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
@@
 \be \frac{\partial}{\partial t} P(u,t) = -\frac{\partial}{\partial u}\Big[\frac{U-u}{t_r}P\Big] + \frac{D}{t_r^2}\frac{\partial^2 P}{\partial u^2},\ee 
-and the steady state solution becomes
+and the steady-state solution becomes
 \be P(u) = \sqrt{\frac{t_r}{2\pi D}} \exp\Big(-\frac{t_r(u-U)^2}{2 D}\Big), \ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
-as indicated in Fig. \ref{fig:ancey} (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$ -- $P(N)=(\Lambda t)^N/N! \exp(-\Lambda t).$
+as indicated in Fig. \ref{fig:ancey} panel (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$ -- $P(N)=(\Lambda t)^N/N! \exp(-\Lambda t).$
 The flux in Eq.
 \ref{eq:flucflux} is a random variable as indicated in Fig.
-\ref{fig:ancey} (b).
+\ref{fig:ancey} panel (b).
 Its probability distribution $P(q|T)$, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.

### 2021-09-25 revision/ch2flipFlop.tex
about to track changes.

@@
 A relativity weak flow shearing a bed of sediment entrains individual particles into a state of motion controlled by turbulent forcing and intermittent collisions with other grains at rest on the bed, generating wide fluctuations in the sediment velocity \citep{Heyman2016,Fathel2015}.
-Bed load particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered pockets on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
+Bedload particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered pockets on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
 Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, and particles are once again entrained.
-Bed load transport is thus a kind of intermittent motion, characterized by alternation between fluctuating movements and periods of rest.
+Bedload transport is thus a kind of intermittent motion, characterized by alternation between fluctuating movements and periods of rest.
 
-To date, descriptions of bed load transport have therefore simplified the problem in various ways to enable progress.
-The foundational work is due to Einstein, who considered bed load motions as instantaneous so he could describe bed load transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
+To date, descriptions of bedload transport have therefore simplified the problem in various ways to enable progress.
+The foundational work is due to Einstein, who considered bedload motions as instantaneous so he could describe bedload transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
 Einstein concluded that particles move downstream with a mean velocity $\bra u \ket = k_E \ell$, where $k_E$ is the rate at which the individual bed particle is entrained into motion, and $\ell$ is the mean length of each downstream step.
@@
 Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
-To date, very few models have calculated the probability distribution of the bed load sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
+To date, very few models have calculated the probability distribution of the bedload sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
 Among those that do include scale dependence, none formulate the sediment flux in terms of the trajectories of individual grains.
@@
 Velocity fluctuations produce tilted stair-step trajectories with unsteady slopes in the $x-t$ plane.
-This can be compared with figure \ref{fig:lislefig} which does not include fluctuations.}
+This can be compared with Fig.
+\ref{fig:lislefig} which does not include fluctuations.}
 	\label{fig:fluxxy0}
@@
 This model can be understood as a generalization of the ``randomly flashing diffusion" models developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}.
-Some trajectories produced by the Langevin equation \ref{eq:flippylangevin} are sketched in figure \ref{fig:fluxxy0}.
-The driving term in panel (a) involves unrealistic fluctuations due to the assumption that velocities evolve as Gaussian white noise, but the integrated particle trajectories in panel (b) are well-behaved and visually similar to those of earlier studies \citep[cf.][]{Fan2016,Bialik2015}.
+Some trajectories produced by the Langevin equation \ref{eq:flippylangevin} are sketched in Fig. \ref{fig:fluxxy0}.
+The driving term in panel (a) displays excessive fluctuations due to the assumption that velocities evolve as Gaussian white noise with no temporal correlations, but the integrated particle trajectories in panel (b) are well-behaved and visually similar to those of earlier studies \citep[cf.][]{Fan2016,Bialik2015}.
+
 Eq.
-\ref{eq:flippylangevin} generalizes the constant velocity model of \citet{Lisle1998} and \citet{Lajeunesse2017} that was presented in section \ref{sec:lisle}, providing the first extension of these works aimed at a more realistic description of bedload particle trajectories through motion and rest.
+\ref{eq:flippylangevin} generalizes the constant velocity model of \citet{Lisle1998} and \citet{Lajeunesse2017} that was presented in section \ref{sec:lisle}, providing the first extension of these works aimed at a refined description of bedload particle trajectories through motion and rest.
 
@@
 The solution of equation \ref{eq:flippylangevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives one possible trajectory of a particle.
-The probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has travelled to position $x$ by time $t$ can be formulated as an average over the ensemble of all possible trajectories.
+The probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has traveled to position $x$ by time $t$ can be formulated as an average over the ensemble of all possible trajectories.
 
@@
 Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process is interrupted, giving rise to the additional terms in Eq.
-\ref{eq:flippymaster} and providing an asymmetrical structure reminiscent of equation \ref{eq:lislemaster} from the simpler motion-rest models of Sec.
+\ref{eq:flippymaster} and providing an asymmetrical structure reminiscent of Eq.
+\ref{eq:lislemaster} from the simpler motion-rest models of Sec.
 \ref{sec:lisle}.
@@
 This generalizes the nonlocal formulation of Sec.
-\ref{sec:nonlocal} to provide the probability distribution of the flux, and it reframes the renewal approach (Sec.
-\ref{sec:renewal}) in terms of the mechanics of individual particles in transport.
+\ref{sec:nonlocal} to provide the probability distribution of the flux, and it re-frames the renewal approach of Sec.
+\ref{sec:renewal} in terms of the mechanics of individual particles in transport.
 
-The basic idea, as depicted in Figure \ref{fig:flipflopfig1}, is to initially distribute $N$ particles in all states of motion along a domain of length $L$ at some random initial locations $x_i$ to the left of $x=0$.
+The basic idea, as depicted in Fig.
+\ref{fig:flipflopfig1}, is to initially distribute $N$ particles in all states of motion along a domain of length $L$ at some random initial locations $x_i$ to the left of $x=0$.
 Later, the number of particles $N$ and the size $L$ of the domain will be extended to infinity such that their ratio $\rho=N/L$ remains constant. 
@@
 
-From this initial configuration, particles move downstream through time according to equation \ref{eq:flippylangevin}.
+From this initial configuration, particles move downstream through time according to Eq. \ref{eq:flippylangevin}.
 The flux is calculated as the average rate of particles crossing to the right of the control surface (at $x=0$) after the sampling time $T$.
@@
 
-The probability distribution of the flux is then the average of equation \ref{eq:flippyflux} across all possible initial configurations of particles and their trajectories
+The probability distribution of the flux is then the average of Eq. \ref{eq:flippyflux} across all possible initial configurations of particles and their trajectories
 \be P(q|T) = \Big \bra \delta\Big( q - \frac{1}{T}\sum_{i=1}^N I_i(T) \Big) \Big\ket. \ee
@@
 \be \bra I_i(t) \ket = \frac{1}{L}\int_{-L}^0 dx' \int_0^\infty dx P(x-x',t) \label{eq:indicator} \ee
-where $P(x,t)$ is the position probability distribution which solves equation \ref{eq:flippymaster}.
-Equation \ref{eq:indicator} describes the probability that the $ith$ particle is found to the right of control surface by time $T$ provided it started at a random location somewhere to the left.
+where $P(x,t)$ is the position probability distribution which solves Eq. \ref{eq:flippymaster}.
+Eq. \ref{eq:indicator} describes the probability that the $ith$ particle is found to the right of control surface by time $T$ provided it started at a random location somewhere to the left.
 These $\bra I_i(t) \ket$ are the components of the flux that depend on the particle dynamics.
 
-Inserting equation \ref{eq:indicator} into equation \ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, producing the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
+Inserting Eq.
+\ref{eq:indicator} into Eq.
+\ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, producing the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
 \begin{align} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T) \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T) \Big].
@@
 
-Equation \ref{eq:flippychar} is the characteristic function of a Poisson distribution \citep{Cox1965}.
+Eq. \ref{eq:flippychar} is the characteristic function of a Poisson distribution \citep{Cox1965}.
 Expanding in $e^{-s/T}$ and inverting the Laplace transform provides a key equation of this chapter, the probability distribution of the flux held contingent on the sampling time $T$:
@@
 All results are scaled by the mean hop length $\ell=V/k_D$ and the timescale $\tau=1/k$ of the motion/rest alternation.
-In both plots, the lines are analytical results while the points are the results of Monte Carlo simulations based on evaluating cumulative transition probabilities on a small timestep \citep[e.g.][]{Barik2006}.}
+In both plots, the lines are analytical results while the points are the results of Monte Carlo simulations based on evaluating cumulative transition probabilities on a small time-step \citep[e.g.][]{Barik2006}.}
 	\label{fig:flippyfig1}
@@
 The master equation \ref{eq:flippymaster} describes the evolution of the probability distribution of position through time.
-The solution of this equation should be some combination of the \citet{Einstein1937} theory for particle transport (section \ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep{Morse1953a}.
+The solution of this equation should be some combination of the \citet{Einstein1937} theory for particle transport (section \ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep[e.g.][]{Morse1953}.
 
@@
 
-The distribution \ref{eq:monster} is shown evolving through time in figure \ref{fig:flippyfig1} panel (a), where the advection and diffusion characteristics of equation \ref{eq:flippymaster} are both evident.
-The integral in equation \ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the Bessel function probability that the particle has been in motion for a period $u$ out of a time $t$ with the Gaussian probability that a particle has travelled a distance $x$ in time $u$ within the motion state.
+The distribution Eq.
+\ref{eq:monster} is shown evolving through time in Fig.
+\ref{fig:flippyfig1} panel (a), where the advection and diffusion characteristics of Eq.
+\ref{eq:flippymaster} are both evident.
+The integral in Eq.
+\ref{eq:monster} encodes the earlier expectation of Einstein model-like behavior mixed with a Gaussian propagator, in that it convolves the Bessel function probability that the particle has been in motion for a period $u$ out of a time $t$ with the Gaussian probability that a particle has traveled a distance $x$ in time $u$ within the motion state.
 The prefactors of this integral term can be understood as adapting this distribution to the initial conditions.
@@
 \subsection{The moments of particle position through motion-rest alternation}
-The moments of position produced by equation \ref{eq:monster} could be derived by integrating it directly, but this is difficult.
-Instead, the moments can be calculated directly from the master equation \ref{eq:flippymaster} \citep{Cox1965}.
 
-Multipling equation \ref{eq:flippymaster} by $x^l$ and integrating over all $x$ provides
+The moments of position produced by Eq.
+\ref{eq:monster} could be derived by integrating it directly, but this is difficult.
+Instead, the moments can be calculated directly from the master equation \ref{eq:flippymaster} \citep[e.g.][]{Cox1965}.
+
+Multiplying Eq. \ref{eq:flippymaster} by $x^l$ and integrating over all $x$ provides
 \begin{multline} \pt^2 \bra x^l \ket -V l \pt \bra x^{l-1} \ket -k_E V l \bra x^{l-1} \ket + k \pt \bra x^{l} \ket\\- D l (l-1) \pt \bra x^{l-2} \ket - k_E D l(l-1) \bra x^{l-2} \ket = 0.\end{multline}
@@
 
-The large P\'{e}clet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocites, so that particles rarely move upstream any appreciable distance \citep[e.g.][]{Fathel2015}.
-This limit also generates the three-range diffusion exemplified by equation \ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed \citep[cf.][]{Nikora2001,Nikora2002}, at least within the assumptions of the model.
-Figure \ref{fig:flippyfig1} panel (b) sketches the predicted three-range diffusion, which is discussed further in Sec. \ref{sec:prollems}.
+The large P\'{e}clet limit ($\Pe\gg 1$) is the physically-relevant condition for bedload transport, where velocity fluctuations are typically small compared to mean downstream movement velocities, so that particles rarely move upstream any appreciable distance \citep[e.g.][]{Fathel2015}.
+This limit also generates the three-range diffusion exemplified by Eq.
+\ref{eq:diffo}, so we should expect bedload particles to spread apart with three different scaling ranges depending on how long they have been observed \citep[cf.][]{Nikora2001,Nikora2002}, at least within the assumptions of the model.
+Fig. \ref{fig:flippyfig1} panel (b) sketches the predicted three-range diffusion, which is discussed further in Sec. \ref{sec:prollems}.
 
@@
 
-Using the probability distribution of position Eq. \ref{eq:monster}, the integrals in Eq. \ref{eq:flatform} are performed in Sec. \ref{sec:fluxconstant} of the appendix, providing the rate function
+Using the probability distribution of position Eq. \ref{eq:monster}, the integrals in Eq. \ref{eq:flatform} are performed in appendix Sec. \ref{sec:fluxconstant}u, providing the rate function
 \begin{multline} 
@@
 Plotted lines are the analytical result Eq.
-\ref{eq:fluxy}, while the points are the results of Monte Carlo simulations.}
+\ref{eq:fluxy}, while the points are the results of Monte Carlo simulations.
+Vertical dotted lines indicate the convergence time $T = (k_D \Pe)^{-1}$ for each condition.
+}
 	\label{fig:fluxconvergence}
@@
 Poisson distributions have relatively thin tails, meaning their fluctuations are typically small \citep{Ancey2006}.
-In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Dhont2018,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Heyman2013,Benavides2021}.
+In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Dhont2018,Benavides2021}.
 This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider transport rate fluctuations.
@@
 Descriptions of sediment transport based on population dynamics in a control volume have produced realistically-wide fluctuations by incorporating a positive feedback between the number of moving particles and the particle entrainment rate called collective entrainment \citep{Ancey2008,Ancey2014}.
-This feedback generates waves of moving particles \citep{Ancey2014, Heyman2015} and produces non-exponential inter-arrival time distributions \citep{Heyman2013} which imply wide-tailed flux distributions when incorporated in the renewal theory \citep[e.g.][]{Turowski2010,Ancey2020}.
+This feedback generates waves of moving particles \citep{Ancey2014, Heyman2014} and produces non-exponential inter-arrival time distributions \citep{Heyman2013} which imply wide-tailed flux distributions when incorporated in the renewal theory \citep[e.g.][]{Turowski2010,Ancey2020}.
 
 Collective entrainment has been attributed to particle-particle interactions such as small granular avalanches and collision-induced entrainment \citep[e.g.][]{Lee2018, Pahtz2020}, and to fluid-particle interactions, such as coherent structures entraining particles en masse as they sweep downstream \citep{Ancey2014,Cameron2020}.
-Given the prevalance of large coherent structures, gravel beds must often times be conditioned to them.
+Given the prevalence of large coherent structures, gravel beds must often times be conditioned to them.
 Therefore particle-particle interactions seem more plausible a mechanism for collective entrainment.
 In particular, contact forces within granular media are known to organize into ``force chains", where a minority of contacts provide a majority of the strength \citep[e.g.][]{Radjai1996, Azema2012}.
-Fluid entrainment of a key particle within a near-surface force chain could entail a collective release of the particles stabilized by this chain, consistent with current understanding of collective entrainment.
+Fluid entrainment of a key particle within a near-surface force chain could entail a collective release of the particles relying on it for stability.
 
@@
 Fig.
-\ref{fig:flippyfig1} and Eq.
-\ref{eq:diffo} indicate that bedload sediment particles described by the model spread apart with a rate which depends upon observation scale, transitioning through three different ranges consistent with the concept of \citet{Nikora2001a,Nikora2002}.
+\ref{fig:flippyfig1} panel (b) indicates that bedload sediment particles described by the model spread apart with a rate which depends upon observation scale, transitioning through three different ranges consistent with the concept of \citet{Nikora2001a,Nikora2002}.
 Nikora et al. originally proposed that sediment diffusion would be super-diffusive at local timescales, normal at intermediate timescales, and sub-diffusive at global timescales.
-Here, the local range diffusion is normal, not super-diffusive, indicating a disprepancy between the Nikora et al. concept and the present model.
+Here, the local range diffusion is normal, not super-diffusive, indicating a discrepancy between the Nikora et al. concept and the present model.
 
@@
 A more challenging (but perhaps more realistic) modification would be to describe the evolution of position by a second-order equation, as in $\ddot{x} = F(x,t)\eta(t)$, where $F$ is an appropriate stochastic driving term switched on and off by the dichotomous noise $\eta(t)$ \citep[e.g.][]{Masoliver1993}.
-Either of these alternatives would introduce correlations to particle velocities and modify the resulting short-range diffusion characteristics.
+Either of these alternatives would introduce correlations to particle velocities and modify the resulting short-range diffusion characteristics, unfortunately at the cost of mathematical difficult.
 
 The short timescale diffusion may also be affected by the assumption that entrainment and deposition processes are instantaneous within the model.
-\citet{Campagnol2015} showed from experimental data that particle ``unsteadiness" during entrainment and deposition can give rise to diffusion exponents even larger than that for ballistic motion.
+\citet{Campagnol2015} showed from experimental data that particle ``unsteadiness" during entrainment and deposition can give rise to diffusion exponents even larger than those for ballistic motion.
 Likely, the representation of entrainment and deposition with an instantaneous alternation obscures the local-timescale physics of these processes \citep[e.g.][]{Valyrakis2010,Celik2014}.
-The simplest modification to include the timescales of entrainment and deposition would introduce a uniform accleration (deceleration) during entrainment (deposition), in effect ``tilting" the sharp motion-rest alternation represented by dichotomous noise in this chapter's model.
-The relatively lower velocity of particles during the unsteady acceleration phase induced by entrainment and deposition should modify the predicted local-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
+The simplest modification to include the timescales of entrainment and deposition would introduce a uniform acceleration (deceleration) during entrainment (deposition), in effect ``tilting" the sharp motion-rest alternation of dichotomous noise.
+The relatively lower velocity of particles during this acceleration phase should modify the predicted local-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
 
@@
 Presumably, the inclusion of a scale-dependent stochastic transport rate generates a stochastic bed elevation field $h(x)$, generating a probability distribution of field configurations $P(h(x),t)$ which depends on the observation scale.
-\citet{Jerolmack2005} includes a stochastic sediment transport rate within the Exner equation, although they only investigated particular realizations of $h(x)$, not the probability distribution of bed configurations or its scale dependence.
+\citet{Jerolmack2005} include a stochastic sediment transport rate within the Exner equation, although they only investigate particular realizations of $h(x)$, not the probability distribution of bed configurations or its scale dependence.
 Mathematical tools to investigate such a stochastic Exner equation can be found in works on surface growth processes and polymer physics, where the evolution of random fields is common topic of inquiry \citep[e.g.][]{Kawakatsu2001,Barabasi1995, Kardar2007}.
@@
 
-Finally, the work further quantifies the dependence of the sediment flux on the observation scale and provides guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in the idealized conditions in which the model was developed.
-The next step is to include collective motions to produce wider sediment transport fluctuations.
+This chapter further quantifies the dependence of the sediment flux on the observation scale and provides guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in the idealized conditions in which the model was developed.
+The next steps are to include collective motions to produce wider sediment transport fluctuations, and to modify the description of velocity fluctuations to include temporal correlations.

### 2021-09-25 revision/ch3bedEle.tex
about to track changes.

@@
 Resting time distributions have been carefully studied in relation to these models because the predicted diffusion characteristics are critically dependent on whether the distribution has a light or heavy tail \citep{Bradley2017,Martin2012,Weeks1998}.
-Resting times have puzzled researchers because early experiments showed exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Nakagawa1976}, while later experiments showed heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
+Resting times have puzzled researchers because early experiments showed exponential distributions \citep{Einstein1937,Hubbell1964, Yano1969,Yano1969a,Nakagawa1976}, while later experiments showed heavy-tailed power-law distributions \citep{Martin2012,Voepel2013, Olinde2015, Pretzlav2016a, Bradley2017, Liu2019}.
 A predominant hypothesis is that power-law distributed resting times originate from buried grains \citep{Martin2014,Voepel2013}. This permits surface grains to retain exponential resting times. 
@@
 
-Because the model is Markovian, it is nevertheless possible to numerically simulate equation (\ref{eq:elemaster}) with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007}. 
+Because the model is Markovian, it is nevertheless possible to numerically simulate Eq. \ref{eq:elemaster} with the Gillespie algorithm \citep{Gillespie1977, Gillespie1991, Gillespie2007}. 
 In conditions when the coupling constant $\kappa$ between the entrainment/deposition rates and the bed elevation is weak, it is also possible to solve the master equation approximately with mean field and Fokker-Planck approaches \citep{Haken1978,Gardiner1983}.
@@
 then draw a second random value to choose the type of transition that occurs using relative probabilities formed from Eqs.
-\ref{eq:rate1}-\ref{eq:rate4}.
+\ref{eq:rate1} - \ref{eq:rate4}.
 The transition is enacted by shifting $t$, $n$, and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
@@
 Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
-In all simulations, I took the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.,][]{Bennett1972}, and I set $L = 22.5$cm and $a = 0.3$cm, values from the original \citet{Ancey2008} experiments.
+In all simulations, I took the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.][]{Bennett1972}, and I set $L = 22.5$cm and $a = 0.3$cm, values from the original \citet{Ancey2008} experiments.
 Each simulation was run for $250$ hours of virtual time, a period selected to ensure convergence of particle activity and bed elevation statistics.
@@
 \ref{eq:elemaster} can be approximately decoupled using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
-These inequalites imply that many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
-Concentrating on steady state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to Eq.
+These inequalities imply that many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
+Concentrating on steady-state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to Eq.
 \ref{eq:elemaster}, with the new distributions normalized as $\sum_m M(m)=1$ and $\sum_n A(n|m)=1$ \citep[e.g.][]{Haken1978}.
-This provides the steady state equation
+This provides the steady-state equation
 \begin{multline}
@@
 \end{multline}
-Unforunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in Eq. \ref{eq:decomp}.
+Unfortunately, these two equations are no easier to solve than the original master equation, since the amount of coupling between $n$ and $m$ is not reduced in Eq. \ref{eq:decomp}.
 
@@
 The mean elevation is always $\langle m \rangle = 0 $ due to the initial assumption of symmetry in the entrainment and deposition rate scaling with $m$.
+
+Fig.
+\ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in Eq.
+\ref{eq:ou2}; this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
 \begin{wrapfigure}{r}{0.5\textwidth}
@@
 \end{wrapfigure}
-Figure \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in Eq.
-\ref{eq:ou2}; this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
-The close correspondence between the mean field approximation and the numerical simulations in Fig.
-\ref{fig:pdfs}(a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
+\indent The close correspondence between the mean field approximation and the numerical simulations in Fig.
+\ref{fig:pdfs} panel (a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.
 They appear identical within numerical uncertainty.
@@
 Panel (a) indicates the mean particle activity shift versus the bed elevation measured in units of $\sigma_z=l$.
-This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiements, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$.
+This shift displays asymmetric dependence on $m$ at the flow conditions of the \citet{Ancey2008} experiments, and departures of the bedload transport mean can be as much as 60\% when the bed is in a severely degraded state with $z\approx -3l$.
 The closure equation \ref{eq:closure} is plotted in panel (a).
@@
 \end{figure}
-Figure \ref{fig:condmoms}a demonstrates that the flow conditions in Tab.
+Figure \ref{fig:condmoms} panel (a) demonstrates that the flow conditions in Tab.
 \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
@@
 \ref{eq:ou2}) is plotted behind the conditional mean profiles in Fig.
-\ref{fig:condmoms}a, where it appears as a crude approximation since it does not capture the asymmetry.
+\ref{fig:condmoms} panel (a), where it appears as a crude approximation since it does not capture the asymmetry.
 Nevertheless, Fig. \ref{fig:var} demonstrates the variance $1/(4\kappa)$ derived from this closure equation is representative of the numerical relationship.
 For the parameters of the \citet{Ancey2008} experiments, Fig.
-\ref{fig:condmoms}b displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
+\ref{fig:condmoms} panel (b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
 The summary is that bed elevation changes regulate the first and second particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
@@
 \end{figure}
-Figure \ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed.
+Fig.
+\ref{fig:colent} depicts the modification of the particle activity mean and variance as the importance of collective entrainment is tuned (through $\lambda$ and $\mu$) with all other parameters fixed.
 When $f=0$, the bed elevation ceases to influence the particle activity mean or variance, while larger fractions increasingly enable the moment regulation effect identified in Sec.
@@
 
-Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the timeseries of $m$ \citep[e.g.,][]{Redner2007}.
-Following \citet{Voepel2013} and \citet{Martin2014}, we concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
+Resting times for sediment undergoing burial are obtained from analyzing the return times from above in the timeseries of $m$ \citep[e.g.][]{Redner2007}.
+Following \citet{Voepel2013} and \citet{Martin2014}, I concentrate on a particular bed elevation $m'$, and find all time intervals separating deposition events at $m=m'$ from erosion events at $m=m'+1$.
 These are the return times from above of the sedimentary bed conditional to the elevation $m'$.
-Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin, we obtain an exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
-Using the marginal probability distribution of bed elevations $P(m)$ (Fig.
-\ref{fig:pdfs}(b)), we derive the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}:
+Binning these conditional return times (using logarithmically-spaced bins to reduce computational load) and counting the occurrences in each bin obtains the exceedance distribution of return times $t_r$ held conditional to the elevation $m'$: $P(T>t_r|m')$.
+The marginal probability distribution of bed elevations $P(m)$ displayed in Fig.
+\ref{fig:pdfs} panel (b) derives the unconditional exceedance distribution of resting times as a sum over all elevations \citep{Yang1971, Nakagawa1980, Voepel2013, Martin2014}:
 \begin{equation} P(T>t_r) = \sum_{m'} P(m') P(T>t_r|m') .\end{equation}
-Figure \ref{fig:cdfs} displays a representative subset of these results.
-Comparing panels \ref{fig:cdfs}(a) and \ref{fig:cdfs}(c) shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
-However, as shown in panels \ref{fig:cdfs}(b) and \ref{fig:cdfs}(d), a characteristic time scale $T_0$ is found to collapse away both variations.
-We obtained this $T_0$ heuristically by considering the characteristic speed of bed elevation change.
+Fig. \ref{fig:cdfs} displays a representative subset of these results.
+Comparing panels (a) and (c) in this figure shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
+However, as shown in panels (b) and u(d), a characteristic timescale $T_0$ is found to collapse away both variations.
+This $T_0$ can be obtained heuristically by considering the characteristic speed of bed elevation change.
 Because the mean number of grains leaving the bed per unit time is $E$ and the removal of a single grain changes the bed elevation by $z_1$ (Eq.
@@
 \begin{equation} T_0 = \frac{l}{z_1 E}.\label{eq:time}\end{equation}
-When scaling the resting time by this $T_0$, we obtain the collapse shown in Fig. \ref{fig:cdfs}.
-Using the log-likelihood estimation technique described by \citet{Newman2005}, we estimate the scaled resting time non-exceedance distributions decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
+Scaling the resting time by this $T_0$ obtains the collapse shown in Fig. \ref{fig:cdfs}.
+Using the log-likelihood estimation technique described by \citet{Newman2005}, the scaled resting time non-exceedance distributions are estimated to decay as a heavy tailed power law with parameter $\alpha = 1.18 \pm 0.32$ for all return times satisfying $T/T_0 > 10^3$.
 These distributions are sufficiently heavy tailed to violate the central limit theorem and drive anomalous super-diffusion of bedload, a result which supports the earlier conclusions of \citet{Voepel2013} and \citet{Martin2014}.
@@
 \subsection{Context of the research}
-Einstein developed the first stochastic models of bedload tracer diffusion \citep{Einstein1937} and the bedload flux \citep{Einstein1950}, and his ideas can be viewed as the nexus of an entire paradigm of research that extends into the present day \citep[e.g.,][]{Hubbell1964, Nakagawa1976,Hassan1991,Ancey2008, Wu2019}.
+Einstein developed the first stochastic models of bedload tracer diffusion \citep{Einstein1937} and the bedload flux \citep{Einstein1950}, and his ideas can be viewed as the nexus of an entire paradigm of research that extends into the present day \citep[e.g.][]{Hubbell1964, Nakagawa1976,Hassan1991,Ancey2008, Wu2019}.
 These approaches aim to predict bedload transport characteristics from stochastic concepts of individual particle motions.
@@
 Numerical solutions and analytical approximations provided negative binomial distributions of bedload activity and normal distributions of bed elevations.
-Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, our results reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments.
+Although experiments under more natural conditions with segregation processes,  migrating bedforms, or sediment supply perturbations have shown particle activity distributions with heavier tails \citep{Dhont2018,Saletti2015} and non-Gaussian bed elevations \citep{Singh2012,Aberle2006}, the results of this chapter reproduce the key features of the most controlled bed elevation \citep{Wong2007,Martin2014} and bedload transport \citep{Heyman2016,Ancey2008} experiments.
 
@@
 This analysis produces heavy-tailed power-law resting times with tail parameters sufficient to drive anomalous diffusion of bedload at long timescales.
-The power-law tails were found to collapse across flow conditions using a time scale formed from the mean erosion rate and the active layer depth.
+The power-law tails were found to collapse across flow conditions using a timescale formed from the mean erosion rate and the active layer depth.
 
@@
 The resting time distributions derived in Sec.
-\ref{sec:elertcdf} show a divergent variance and possibily a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of the results.
+\ref{sec:elertcdf} show a divergent variance and possibly a divergent mean, since this occurs for $\alpha < 1$ \citep{Sornette2006} which is within range of the results.
 
 Divergent mean resting time distributions present a paradox, since they imply all particles should eventually be immobile, violating the equilibrium transport assumption.
-\citet{Voepel2013} demonstrated that a bounded random walk for bed elevations provides a power-law distribution that eventually transitions to a faster thin-tailed decay, allowing for power-law scaling like our result and \citet{Martin2014} without this divergent mean paradox.
+\citet{Voepel2013} demonstrated that a bounded random walk for bed elevations provides a power-law distribution that eventually transitions to a faster thin-tailed decay, allowing for power-law scaling like the result in this chapter and \citet{Martin2014} without this divergent mean paradox.
 One resolution to this issue could come from a spatially distributed model with multiple cells.
@@
 
-Finally, this chapter presented the probability distribution functions and first and second moments of the particle activity and bed elevation, indicating novel coordination between the statistical characteristics of these quantites which deserve experimental testing.
+Finally, this chapter presented the probability distribution functions and first and second moments of the particle activity and bed elevation, indicating novel coordination between the statistical characteristics of these quantities which deserve experimental testing.
 Some relatively recent particle tracking experiments have demonstrated joint resolution of bed elevations and bedload transport \citep{Martin2014,Heyman2016}.

### 2021-09-25 revision/ch4downDiff.tex
about to track changes.

@@
 Diffusion is quantified by the time dependence of the positional variance $\sigma_x^2$ of a group of tracers.
-With the scaling $\sigma_x^2 \propto t$, the diffusion is said to be normal, since this is found in the classic problems \citep{Taylor1920}.
+With the scaling $\sigma_x^2 \propto t$, the diffusion is said to be normal, since this is found in the classic problems \citep{Einstein1905}.
 However, with the scaling $\sigma_x^2 \propto t^\gamma$ with $\gamma \neq 1$, the diffusion is said to be anomalous \citep{Sokolov2012}, with $\gamma>1$ defining super-diffusion and $\gamma<1$ defining sub-diffusion \citep{Metzler2000}.
@@
 originally considered.
-As a result, three diffusion ranges can be identified by patching together multiple datasets \citep{Zhang2012,Nikora2002}, but they are not resolved by any one dataset.
+As a result, three diffusion ranges can be identified by patching together multiple data-sets \citep{Zhang2012,Nikora2002}, but they are not resolved by any one data-set.
 
@@
 Among these, \citet{Nikora2001a} used synthetic turbulence \citep{Kraichnan1970} with a discrete element method for the granular phase \citep{Cundall1979}; \citet{Bialik2012} used synthetic turbulence with a random collision model \citep{Sekine1992}; and \citet{Fan2016} used a Langevin equation with probabilistic rests.
-To our knowledge, only \citet{Bialik2015} have claimed to capture all three ranges from a Newtonian approach.
+To my knowledge, only \citet{Bialik2015} have claimed to capture all three ranges from a Newtonian approach.
 They incorporated a second resting mechanism into their earlier model \citep{Bialik2012}, implicitly suggesting that three diffusion ranges could result from two distinct timescales of sediment rest.
@@
 \label{sec:assumptions}
-Particle trajectories are formulated as a three-state random walk where the states are motion, surface rest, and burial. These states are labelled as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
+Particle trajectories are formulated as a three-state random walk where the states are motion, surface rest, and burial. These states are labeled as $i=2$ (motion), $i=1$ (rest), and $i=0$ (burial).
 The target is the probability distribution $p(x,t)$ to find a grain at position $x$ and time $t$ if it is known to have started with the initial distribution $p(x,0)=\delta(x)$.
@@
 The end results will not be contingent on the specific distributions chosen, since all thin-tailed distributions provide similar diffusion characteristics in random walks \citep{Weiss1994,Weeks1998}.
-Grains in motion are considered to have a characteristic velocity $V$ \citep{Lisle1998,Lajeunesse2017}, and burial is modeled as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried.
+Grains in motion are considered to have a characteristic velocity $v$ \citep{Lisle1998,Lajeunesse2017}, and burial is modeled as long lasting enough to be effectively permanent \citep{Wu2019}, with grains resting on the surface having a probability per unit time $\kappa$ to become buried.
 This means $\Phi(t) = e^{-\kappa t}$ represents the probability that a grain is not buried after resting for a time $t$, while $1-\Phi(t)$ represents the probability that it is buried.
@@
 Finally, the overall probability to be at position $x$ at time $t$ is
-\be p(x,t) = \sum_{k=0}^2 p_k(x,t) \ee
-This joint density is completely determined from the solutions of Eqs. \ref{eq:g1}-\ref{eq:g2} given specification of the distributions $g_{ij}$ and $G_i$.
+\be p(x,t) = \sum_{k=0}^2 p_k(x,t). \ee
+This joint density is completely determined from the solutions of Eqs. \ref{eq:g1} and \ref{eq:g2} given specification of the distributions $g_{ij}$ and $G_i$.
 
@@
 
-Equations \ref{eq:g1}-\ref{eq:g2} with these $g_{ij}$ and $G_i$ are solved using Laplace transforms in space and time $(x,t \rightarrow 
+Eqs. \ref{eq:g1} and \ref{eq:g2} with these $g_{ij}$ and $G_i$ are solved using Laplace transforms in space and time $(x,t \rightarrow 
 \eta,s).$ This method, similar to \citet{Weeks1998}, unravels the convolution structure of these equations, eventually producing
 \be \tilde{p}(\eta,s) = \frac{1}{s}\frac{(s+\kappa + k')s  + \theta_1(s+\kappa )\eta v+ \kappa k_2}{(s+\kappa+k_1)\eta v+(s+\kappa+k')s + \kappa k_2}, \label{eq:diffnicedist}\ee
-where $k' = k_1 + k_2$. Inverting this result using known Laplace transforms \citep{Prudnikov1992a,Arfken1985} obtains
+where $k' = k_1 + k_2$. As explained in appendix Sec. \ref{sec:appendixA}, inverting this result using known Laplace transforms \citep{Prudnikov1992a,Arfken1985} obtains
 \begin{align}
@@
 The solid lines are the analytical distribution in Eq.
-\ref{eq:pdf}), while the points are numerically simulated, showing the correctness of our derivations.
+\ref{eq:pdf}), while the points are numerically simulated, showing the correctness of the derivations.
 Colors pertain to different times.
@@
 
-Figure \ref{fig:diffpdfs} depicts the distribution (Eq.
-\ref{eq:pdf}) alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
+Fig.
+\ref{fig:diffpdfs} depicts the distribution Eq.
+\ref{eq:pdf} alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small time-step \citep{Barik2006}.
 When grains do not become buried, as in panel (a) of Fig.
@@
 \end{table}
-The positional variance (Eq. \ref{eq:var}) is plotted in Fig. \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
+The positional variance Eq. \ref{eq:var} is plotted in Fig. \ref{fig:diffvar} for conditions $\theta_1=1$ and $k_2\gg k_1 \gg \kappa$.
 The notation ``$\gg$'' is interpreted in this context to mean ``of at least an order of magnitude greater''.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch4/diffusion.png}
-	\caption{Panel (a) sketches conceptual trajectories of three grains, while panel (b) depicts the variance (Eq.
-\ref{eq:var}) with mean motion time $1.5$ s, resting time $30.0$ s, and movement velocity $0.1$ m/s -- values comparable to laboratory experiments transporting small ($5$ mm) gravels \citep{Lajeunesse2010,Martin2012}.
-The burial timescale is $7200.0$s (two hours), and grains start from rest ($\theta_1=1$).
+	\caption{Panel (a) sketches conceptual trajectories of grains, while panel (b) depicts Eq.
+\ref{eq:var} with mean motion time $1.5$ s, rest time $30.0$ s, and velocity $0.1$ m/s.
+The burial timescale is $7200$ s (two hours), and grains start from rest ($\theta_1=1$).
 The solid line is Eq.
-\ref{eq:var}, and the points are numerically simulated.
+\ref{eq:var} and points are numerically simulated.
 Panel (b) demonstrates four distinct scaling ranges of $\sigma_x^2$: local, intermediate, global, and geomorphic.
 The first three are diffusive.
-Three crossover times $\tau_L$, $\tau_I$, and $\tau_G$ divide the ranges.
-Within each range, a slope key demonstrates the scaling $\sigma_x^2 \propto t^\gamma$.
+Crossover times $\tau_L$, $\tau_I$, and $\tau_G$ divide the ranges.
+Slope keys demonstrate the scaling $\sigma_x^2 \propto t^\gamma$ in each range.
 Panel (a) demonstrates that different mixtures of motion, rest, and burial states generate the ranges.
-At local timescales, grains usually either rest or move; at intermediate timescales, they transition between rest and motion; at global timescales, they transition between rest, motion, and burial; and at geomorphic timescales, all grains bury.
-Additional slope keys in the local and global ranges of panel (b) illustrate the effect of initial conditions and rest/burial timescales on the diffusion, while the additional slope key within the geomorphic range demonstrates the expected scaling when burial is not permanent, as discussed in Sec.
-\ref{sec:discussion}.
+At local timescales, grains usually either rest or move; at intermediate timescales, they transition between rest and motion; at global timescales, they transition between rest, motion, and burial; and at geomorphic timescales, all grains become buried.
 	}
@@
 
-\subsection{Diffusion exponents and three range scaling}
+\subsection{Diffusion exponents and three-range scaling}
+\label{sec:amidonenow}
 
@@
 
-This chapter has extended \citet{Einstein1937} by including motion, rest, and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010}, before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
+This chapter has extended \citet{Einstein1937} by including motion, rest, and burial processes in a multi-state random walk \citep{Weiss1994,Weeks1998} to demonstrate that a group of bedload tracers moving downstream while gradually becoming buried will generate a super-diffusive local range \citep{Martin2012,Fathel2016,Witz2018}, a normal-diffusive intermediate range \citep{Nakagawa1976,Yano1969}, and a super-diffusive global range \citep{Bradley2017, Bradley2010} before the diffusion eventually terminates in a geomorphic range \citep{Hassan2017}.
 \citet{Nikora2002} highlighted the need for such a physical description, although they suggested to use a two-state random walk between motion and rest states with heavy-tailed resting times, and they did not discuss sediment burial.
@@
 In this chapter, I developed a random walk model to describe sediment tracers transporting through a river channel as they gradually become buried, providing a physical description of the local, intermediate, and global diffusion ranges identified by \citet{Nikora2002}.
-Pushing their ideas somewhat further, I followed \citet{Hassan2017} to propose a geomorphic range to describe diffusion characteristics at timescales larger than the global range when burial and exhumation both moderate downstream transport.
+Pushing their ideas somewhat further, I followed \citet{Hassan2017} to propose a geomorphic range describing particle diffusion characteristics at timescales larger than the global range when burial and exhumation both moderate downstream transport.
 At base level, this work demonstrates that (1) durations of sediment motions, (2) durations of sediment rest, and (3) the sediment burial process are sufficient to develop three diffusion ranges that terminate when all tracers become buried.
-This work provides a complementary perspective on the three range diffusion exhibited in Ch. \ref{ch:flux}.
 A next step is to incorporate exhumation to better understand the geomorphic range.

### 2021-09-25 revision/ch5langevin.tex
about to track changes.

@@
 
-Bed load transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
+Bedload transport rates show wide and frequent fluctuations which originate from coupling between the fluid and granular phases.
 Owing to these fluctuations, measured transport rates often show slow convergence through time \citep{Dhont2018,Turowski2010}, and predicted rates can deviate from measured values by several orders of magnitude \citep{Recking2012,Martin2003}.
 These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
-In recent decades, stochastic formulations of the bed load flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (Secs.
-\ref{sec:birthdeath}-\ref{sec:renewal}).
+In recent decades, stochastic formulations of the bedload flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (Secs.
+\ref{sec:birthdeath} and \ref{sec:renewal}).
 Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
@@
 \ref{sec:birthdeath}).
-This formulation requires the instantaneous velocity distributions of sediment particles \citep[e.g.][]{Ancey2020a}, yet unfortunately, understanding of these distributions remains limited.
+This formulation requires the instantaneous velocity distributions of sediment particles \citep[e.g.][]{Ancey2020a}, yet our understanding of these distributions remains limited.
 There is as of yet no consensus on the shape of the bedload velocity distribution, and although the models described in Sec.
-\ref{sec:langevin} of the introduction describe some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe all experimental observations reported to date \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
+\ref{sec:langevin} describe some extreme end-member distributions \citep[e.g.][]{Fan2014,Ancey2014}, no models have been developed that describe all experimental observations reported to date \citep{Lajeunesse2010,Fathel2015,Heyman2016,Liu2019,Houssais2012}.
 In this chapter, I develop a stochastic model of particle velocities which addresses this shortcoming.
@@
 One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
-These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were turbulent and supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were laminar and (likely) subcritical.
+These experiments involve uniformly-sized small sands or glass beads ($0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were turbulent and supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were laminar and (likely) subcritical.
 A second set of studies show Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
 In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and supercritical.
-Other experiments display velocity distributions that are intermediate between exponential and Gaussian extremes that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
-The \citet{Houssais2012} experiments involved a binomial distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm in turbulent and subcritical flow.
+Other experiments display velocity distributions that are intermediate between exponential and Gaussian that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
+\citet{Houssais2012} used a bimodal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while \citet{Liu2019} used sand having median diameter $1.1$mm in turbulent and subcritical flow.
 From this experimental record, the velocity distribution shape apparently does not consistently relate to the flow conditions (turbulent/laminar or super/subcritical) or sediment properties (natural sand/gravel or glass beads).
@@
 Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}.
-This summary suggests that bed load velocity distribution could be controlled by the particle size.
+This summary suggests that bedload velocity distribution could be controlled by the particle size.
 
-Existing models of streamwise bed load velocities can be divided into computational and stochastic physics categories.
+Existing models of streamwise bedload velocities can be divided into computational and stochastic physics categories.
 Computational models numerically integrate some approximate coupled dynamics for individual grains and the fluid, generally modeling particles as spheres interacting through repulsive forces and numerically integrating the Navier-Stokes equations (or some approximation of them) to describe the flow forces on grains.
@@
 
-Stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (Sec.
+Stochastic models have incorporated fluctuating driving and resisting terms into Newtonian equations of individual grains to develop Langevin-like descriptions of bedload particle motions (Sec.
 \ref{sec:langevin}).
@@
 \ref{ch:Introduction} such a continuum assumption is unlikely to be satisfied for weak bedload transport conditions.
-Instead, particles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}.
+
+Instead,articles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}. 
 In the theory of rarefied granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
@@
 
-Taking inspiration from granular gas theory, I develop in this chapter an improved model for sediment grains in transport which replaces the smooth friction terms of earlier models by episodic particle-bed collisions.
-The driving motivation is to test the hypothesis that particle-bed collision characteristics can explain the range of experimentally-observed streamwise bed load particle velocity distributions.
-A secondary motivation is to introduce more realistic forces into earlier stochastic descriptions of individual bed load particle dynamics. 
+I develop in this chapter an improved model for sediment grains in transport which replaces the smooth friction terms of earlier models by episodic particle-bed collisions.
 I develop the model and explain its key assumptions in Sec.
@@
 Figure \ref{fig:fig1} indicates the configuration of the model in this chapter.
-Nearly spherical cohesionless particles of diameter $d$ and mass $m$ move as bed load down a slope inclined at an angle $\theta$ in a steady flow.
+Nearly spherical particles of diameter $d$ and mass $m$ move as bedload down a slope inclined at an angle $\theta$ in a steady flow.
 The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Ashworth1989, Warburton1992}.
@@
 Each force timeseries realization implies a different velocity timeseries $u(t)$ in the downstream direction.
-The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
+The objective is to calculate the probability distribution $P(u,t)$ of this downstream velocity by averaging over the ensemble of forces.
 Here, the description of bedload particle velocity fluctuations will be more detailed than Ch.
 \ref{ch:flux}, but entrainment and deposition processes will be neglected, meaning the model of this chapter is only applicable to moving particles.
-I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.
+I include the most realistic particle-bed collision and fluid forces possible while still allowing for analytical solutions.
 
@@
 Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid between approaching particles \citep{Joseph2001,Schmeeckle2001}.
-The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of resitution may be less relevant than other factors such as collision geometry and bed deformation.
+The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of restitution may be less relevant than other factors such as collision geometry and bed deformation.
 Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in Fig.
@@
 Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
-The time intervals $\Delta \tau_k = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau).$
+The time intervals $\Delta \tau_k = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau),$ so the number of collisions $n=N_\nu(t)$ in time $t$ is distributed as $P(n) = (\nu t)^n/n!
+\exp(-\nu t).$
 This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momenta.
@@
 Even ignoring this complication, the added mass and Basset forces are difficult to compute and therefore impractical to include in analytical models.
-These challenges require researchers to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michaelides1997,Armenio2001}.
-As a result, the majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the primary force(s) driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.
+These challenges require us to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michaelides1997,Armenio2001}.
+From this same requirement, a majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the only force driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.
 
@@
 Experiments on stationary particles indicate that drag forces fluctuate rapidly and lie on wide Gaussian-like distributions \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
-The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities should be expected to vary due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
-The nonlinearity in Eq. \ref{eq:draggy} will amplify these slip velocity variations, generating large and erratic drag force fluctuations.
+The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities are expected to vary rapidly due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
+The nonlinearity in Eq.
+\ref{eq:draggy} will amplify slip velocity variations, generating large and erratic drag force fluctuations, which only add to the undetermined contributions of the neglected fluid forces.
 
 Here I assume with earlier computational \citep{Schmeeckle2014,Gonzalez2017,Elghannay2017} and analytical \citep[e.g.][]{Ancey2014,Fan2014} models of bedload transport that fluid drag is the primary force driving sediment particles downstream.
-I further assume with \citet{Fan2014} and \citet{Ancey2014} that drag forces can be represented by a Gaussian white noise.
+I further assume with \citet{Fan2014} and \citet{Ancey2014} that the fluctuating component of this force can be represented by a Gaussian white noise.
 This latter assumption is justified insofar as drag forces fluctuate rapidly compared to the inertial response times of particles and the timescales over which particles change height within the flow profile.
-Defining $\bar{C}_D$ as the empirical drag coefficient evaluated at $\bar{V}$ and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep[e.g.][]{Gardiner1983}, the fluid drag force considered in this chapter can be written
+Defining $\bar{C}_D$ as the empirical drag coefficient evaluated at $\bar{V}$ and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep[e.g.][]{Gardiner1983}, the fluid drag force considered in this chapter is
 \be F_D(t) = \Gamma + \sqrt{2 D } \eta(t), \label{eq:drag}\ee
@@
 \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
-Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
+Here, the It\^{o} interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
 Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by Eq.
-\ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.
+\ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u,t)$.
 
@@
 \be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
-In this equation, I introduced the parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$.
-The term
+The parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ scale the steady and fluctuating components of the fluid force against the collision rate $\nu$. The term
 \be I_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
@@
 
-Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
-Nonlocality is introduced by the collision integral (Eq. \ref{eq:colint}) which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
+Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bedload models \citep{Fan2014,Ancey2014}. 
+Nonlocality is introduced by the collision integral of Eq. \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
 This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}.
@@
 To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
-For bedload particle collisions with the bed, we can expect a mode elasticity since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.
+For particle-bed collisions of bedload, we can expect a mode elasticity since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.
 
 Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
-\be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
+\be I_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
 where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the modal elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.
@@
 In what follows, I drop all but the first two terms in this expansion to obtain the leading order contribution of particle-bed collisions to the velocity distribution.
-Higher orders could always be included later by perturbation theory \citep{Morse1953}.
+Higher orders could always be included later by perturbation theory \citep{Morse1953a}.
 I solve the resulting approximate equation in steady-state, when $\partial P(u,t)/\partial t = 0$.
@@
 It is possible to derive the moments of this probability distribution by multiplying Eq.
-\ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[cf.][]{Cox1965}.
-These calculations are provided in appendix Sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
+\ref{eq:governer} by $u$, integrating, and then solving the resulting moment evolution equations \citep[e.g.][]{Cox1965}.
+These calculations are provided in the appendix Sec. \ref{sec:langmoments}, where the mean bedload velocity is derived as
 \be \langle u \rangle = \frac{\Gamma}{\nu(1-\ve)} = \frac{\tilde{\Gamma}}{1-\ve}. \label{eq:meanu}\ee
@@
 This result demonstrates that within the model, velocity fluctuations originate from both the steady and fluctuating components of the flow forces, yet the variance is linear in these factors and is therefore relatively insensitive to the fluid flow.
-This is consistent with the obvious similarities between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
+This is consistent with the similarities apparent between viscous and turbulent flow experiments \citep[e.g.][]{Charru2004,Lajeunesse2010}.
 In contrast, particle velocity fluctuations in Eq.
@@
 \end{figure}
-This figure reveals an apparent transition from exponential-like to Gaussian-like velocity distributions as collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
+Fig.
+\ref{fig:fig2} reveals an apparent transition from exponential-like to Gaussian-like velocity distributions as collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
 In between, the full distribution Eq.
@@
 
-\subsection{Exponential and Gaussian regimes: limits to earlier work}
+\subsection{Exponential and Gaussian regimes: Limits to earlier work}
 \label{sec:langmodelcomparison}
@@
 \be P(u) = \frac{\tilde{D}}{\sqrt{\tilde{\Gamma}^2 + 4 \tilde{D}}}e^{\frac{\tilde{\Gamma} u }{2 \tilde{D}} - \frac{\sqrt{\tilde{\Gamma}^2 + 4 \tilde{D}}|u|}{2\tilde{D}}}. \ee
-Thus, for bed load transport conditions with typically very inelastic particle-bed collisions, we can expect exponential-like velocities and large deviations from a Gaussian behavior.
+Thus, for bedload transport conditions with typically very inelastic particle-bed collisions, we can expect exponential-like velocities and large deviations from a Gaussian behavior.
 
@@
 The challenge is that the statistical moments (Eqs.
-\ref{eq:mean} and \ref{eq:varu}) diverge at the same time as the denominator factors of the distribution Eq.
+\ref{eq:meanu} and \ref{eq:varu}) diverge at the same time as the denominator factors of the distribution Eq.
 \ref{eq:steadystate}.
@@
 A variety of velocity distributions have been observed in bedload transport experiments, ranging from exponential to Gaussian in shape. 
-This section compares these with the theoretical result Eq. \ref{eq:steadystate}.
+This section compares Eq. \ref{eq:steadystate} with the results of six different experiments.
 Comparing Eq.
@@
 
-This section compares Eq. \ref{eq:steadystate} with the results of six different experiments.
 In each case, the particle mass is computed from experimental parameters assuming spherical sediment as $m = \pi \rho_s d^3/6$, where $\rho_s$ is the sediment density and $d$ is the particle diameter.
@@
 It should really be calculated as an average over different particle heights, particle velocities, and fluid velocities within the flow profile.
-For lack of a better option we simply assume the typical slip velocity scales with the shear velocity of the flow, as it would for a small particle resting on the bed.
+For lack of a better option I simply assume the typical slip velocity scales with the shear velocity of the flow, as it would for a small particle resting on the bed.
 Using this assumption, the steady component of the drag force is estimated as
@@
 \be C_D = \frac{24}{\Rey_p}( 1 + 0.194 \Rey_p^{0.631}). \ee
-Particle Reynolds numbers are estimated using the shear velocity: $\Rey_p= u_\ast d/nu$. 
-The magnitude $D$ of turbulent fluctuations, dissipation per collision $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution (Eq.
+Particle Reynolds numbers are estimated using the shear velocity: $\Rey_p= u_\ast d/\lambda$. 
+The magnitude $D$ of turbulent fluctuations, collision elasticity $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution (Eq.
 \ref{eq:steadystate}) and the experimental data.
@@
 The parameters for each experiment and the best-fit calibration parameters are collected in Tab.
-\ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in figure \ref{fig:fig4ch5}.
+\ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in Fig.d \ref{fig:fig4ch5}.
 \begin{table}
@@
 \ref{fig:fig4ch5}.
-The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St$, and Froude number $\Fro$.
+The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St = \rho_s u_\ast d/(9 \lambda)$, and Froude number $\Fro = \bar{u}/\sqrt{g h}$ ($h$ -- depth, $g$ -- gravitational acceleration, $\bar{u}$ -- time and depth averaged flow velocity).
 The final two columns involve values that were tuned to fit the theoretical distribution (Eq.
-\ref{eq:steadystate}) to the experimental data, the dissipation $\ve$ and the collision rate $\nu$.
+\ref{eq:steadystate}) to the experimental data, the elasticity $\ve$ and the collision rate $\nu$.
 Units are indicated in square brackets.
@@
 \end{table}
-In each case, good agreement is obtained between the theoretical and empirical velocity disributions, suggesting that the Langevin model Eq. \ref{eq:langevin} captures the essential physics.
+In each case, good agreement is obtained between the theoretical and empirical velocity distributions, suggesting that the Langevin model Eq. \ref{eq:langevin} captures the essential physics.
 \begin{figure}
@@
 
-In this final chapter I developed a Langevin description of bed load sediment transport which includes episodic collisions between particles and the bed.
+In this final chapter I developed a Langevin description of bedload sediment transport which includes episodic collisions between particles and the bed.
 The model relates the shape of the instantaneous streamwise particle velocity distribution to the elasticity of particle-bed collisions.
-This work generalizes earlier approaches available in the literature which did not treat episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
+This work generalizes earlier approaches available in the literature which did not include episodic collisions \citep{Ancey2014,Fan2014}, and provides a new physical explanation for the different streamwise sediment velocity distributions resolved in experiments.
 
-Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise for reasons of analytical tractability \cite[e.g.][]{Michaelides1997}.
+Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bedload particles as spatially-uniform Gaussian white noise for reasons of analytical tractability \cite[e.g.][]{Michaelides1997}.
 Even though the non-Gaussian, spatio-temporally-varying, and history-dependent aspects of fluid turbulence certainly do impact sediment movement \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is laminar or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
-Nevertheless, it is possible the vertical flow structure, and in particular the size of particles in comparison to the thickness of the laminar sublayer, may be explanatory of the different velocity distributions resolved in bedload transport experiments, and future studies should investigate this possibility.
+Nevertheless, it is possible the vertical flow structure, and in particular the size of particles in comparison to the thickness of the laminar sub-layer, may be explanatory of the different velocity distributions resolved in bedload transport experiments, and future studies should investigate this possibility.
 
@@
 The analytical velocity distribution Eq.
-\ref{eq:steadystate} was fit to six different experimental datasets in Fig.
+\ref{eq:steadystate} was fit to six different experimental data-sets in Fig.
 \ref{fig:fig4ch5}, and the best fit parameters were presented in Tab.
@@
 
-This work hints that particle size may be a leading order control over the different bedload velocity distributions obtained in experiments, although future study is required for a definite conclusion.
-Next steps are to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions and to introduce velocity dependence and vertical flow structure in the flow forces acting on moving particles.
+This work suggests that particle-bed collisions be a leading order control over the different bedload velocity distributions obtained in experiments, although future study is required for a definite conclusion.
+Next steps are to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions and to introduce slip velocity dependence and vertical structure into the flow forces acting on moving particles.
 These efforts would produce additional insight into the controls of fluid turbulence and transverse, vertical, and rotational movements on the downstream velocities of individual bedload grains.

### 2021-09-25 revision/ch6conc.tex
about to track changes.

@@
 \end{enumerate}
-These developments improve understanding of sediment transport in streams.
+These developments bring us closer to a comprehensive mathematical description of sediment transport in streams.
 
@@
 The overarching strategy in all of these developments has been to identify control and response variables, represent control variables by idealized noises (entrainment and deposition events, particle burial events, turbulent forces, or particle-bed collisions), and then formulate dynamical equations relating the response variables to these stochastic control variables.
-This strategy phrases sedimentary dynamics in terms of Langevin-like equations, stochastic analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for the response variable of interest in the problem and the force $F$ is a stochastic combination of the control variables.
+This strategy phrases sedimentary dynamics in terms of Langevin-like equations, stochastic and phenomenological analogues of Newton's $F=ma$, where the acceleration $a$ is swapped for the response variable of interest in the problem and the force $F$ is a stochastic combination of the control variables.
 
@@
 
-The solutions of the master equation in a given problem produce the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (Chs.
+The solutions of the master equation in a given problem provide the probability distribution for the response variable of interest, which in the thesis has included at different points the bedload particle position (Chs.
 \ref{ch:flux} and \ref{ch:downDiff}), the particle velocity (Ch.
@@
 Turbulence was described using Gaussian white noise, while instantaneous steps, particle arrivals to a control surface, and particle-bed collisions were described with Poisson noise.
-Alternation between motion and rest was represented with dichotomous noise.
+Alternation between motion and rest was represented by dichotomous noise.
 Whenever these processes acted in combination, the dynamical equations describing the response variable included multiple sources of idealized noise as required.
@@
 In river science, many phenomena exhibit colored spectra, including the size distributions of bedforms \citep{Nikora1997,Guala2014}, the fluid forces on bed particles \citep{Dwivedi2011, Amir2014}, the roughness characteristics of gravel beds \citep{Aberle2006,Singh2012}, and sediment flux timeseries \citep{Dhont2018,Chartrand2021}.
-Unfortunately, dynamical equations driven by colored noise are extremely challenging to solve \citep{Hanggi1978,Luczka2005,Hanggi2007}.
-In this context, even if the idealized noise models developed in this thesis are not exactly accurate, their development remains necessary as a basis for comparison when formulating and solving colored noise models \citep[e.g.][]{Fox1986,Moss1989}.
+Unfortunately, dynamical equations driven by colored noise are extremely challenging to solve, so progress has been mainly limited to dichotomous and Ornstein-Uhlenbeck noises \citep{Hanggi1978,Luczka2005,Hanggi2007}.
+In this context, even if the idealized noise models developed in this thesis are not exactly accurate, their development remains necessary as a basis for comparison when formulating and solving more challenging colored noise models \citep[e.g.][]{Fox1986,Moss1989}.
 
@@
 
-\subsection{Calculation of the probability distribution of the sediment flux from micromechanics of particle transport}
+\subsection{Calculation of the probability distribution of the sediment flux from micro-mechanics of particle transport}
 
@@
 This work unifies the sediment trajectory models originating from Einstein (Secs.
-\ref{sec:einwalk}-\ref{sec:lisle}) with the renewal approach to calculate the sediment flux (Sec.
+\ref{sec:einwalk} and \ref{sec:lisle}) with the renewal approach to calculate the sediment flux (Sec.
 \ref{sec:renewal}).
-The striking feature of this formulation is that, as a result of the particle dynamics, the mean bedload flux becomes scale-dependent, whereby the expected magnitude of the flux depends on the time-period over which it is observed.
+The striking feature of this formulation is that, as a result of the particle dynamics, the mean bedload flux becomes scale-dependent, whereby the expected magnitude of the flux depends on the timescale over which it is observed.
 
@@
 \ref{sec:einflux}), and an emerging body of research has produced the full probability distribution of the flux without referencing individual movement characteristics (Secs.
-\ref{sec:birthdeath}-\ref{sec:renewal}).
-This work unifies these two research themes and presents a statistical mechanics formulation of the bed load sediment flux probability distribution based on individual particle trajectories.
+\ref{sec:birthdeath} and \ref{sec:renewal}).
+This work unifies these two research themes and provides a statistical mechanics formulation of the bed load sediment flux probability distribution based on individual particle trajectories.
 
@@
 Second, in Ch. \ref{ch:flux} I developed the first analytical description of sediment trajectories through motion and rest including velocity fluctuations within the motion state.
-Einstein originally formulated sediment transport as a sequence of instantaenous steps and rests (Sec.
+Einstein originally formulated sediment transport as a sequence of instantaneous steps and rests (Sec.
 \ref{sec:einwalk}), and this was later improved to include the duration of motion (Sec.
 \ref{sec:lisle}).
-Until this thesis, movement velocities in analytical models were considered constant, which contrasts with reality. 
+Until this thesis, movement velocities in analytical models were considered constant, an unrealistic assumption.
 Although some numerical models have described motion/rest cycles with velocity fluctuations \citep{Fan2016,Bialik2012,Schmeeckle2014}, they had not resolved the novel three-range diffusion characteristics these dynamics imply.
@@
 This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state; it introduces a dimensionless P\'{e}clet number as an important characteristic of bedload sediment transport; and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains.
-These developments constitute a new understanding of bedload movement across its timescales and invite generalizations aimed at pushing the approach further to including interactions between particles.
+These developments constitute a new understanding of bedload movement across its timescales and invite generalizations aimed at pushing the approach further to include interactions between particles and temporal correlations in bedload movement velocities.
 
@@
 This allowed for a mathematical investigation of (1) how bed elevation changes affect sediment transport rates and (2) how bed elevation changes control the residence times of particles buried in the bed.
+
 Sediment transport fluctuations have come under increasing scrutiny with the resurgence of stochastic modeling in sediment transport, while the burial times of particles are crucial for using sediment tracers to predict sediment transport.
@@
 
-This work demonstrates that this negative feedback buffers bedload transport fluctuations whenever collective entrainment occurs, meaning the magnitude of bedload transport fluctuations depends on the rate of bed elevation change.
+This work demonstrates that this negative feedback buffers bedload transport fluctuations whenever collective entrainment occurs, meaning the magnitude of bedload transport fluctuations depends on the relative bed elevation.
 The residence times of buried particles are random variables that lie on heavy-tailed power-law distributions.
-These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to predict the downstream sediment flux in applications by tracking sediment tracer particles.
+These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to determine the downstream sediment flux for applications by tracking sediment tracer particles.
 
@@
 Until this work, the mechanisms which produce the different spreading rates of particles across the scaling ranges identified by Nikora and coworkers had been uncertain for several decades, and earlier works had included what were believed to be the required features without describing three or more scaling ranges.
-The work ultimately demonstrates that many approaches to describe individual particle motions are reformulations of the continuous time random walk formalism from physics, indicating underlying unity within a diverse body of research and bringing powerful tools from statistical physics to the sediment transport problem.
+Ch.
+\ref{ch:downDiff} brings some closure to these studies and demonstrates that many approaches to describe individual particle motions are reformulations of the continuous time random walk formalism from physics, indicating unity within a diverse body of research and bringing powerful tools from statistical physics to the sediment transport problem.
 
@@
 Finally, Ch. \ref{ch:langevin} displayed a new theoretical model of individual grains saltating downstream in a turbulent flow through a sequence of particle-bed collisions.
-This work provides the first comprehensive description of all bedload velocity distributions observed in experiments to the present time (Sec.
+This work provides the first comprehensive description of all bedload velocity distributions sar far observed in experiments (Sec.
 \ref{sec:langexperimentcomparison}), while earlier works had described only particular end-member distributions (Sec.
 \ref{sec:langevin}).
-This work incorporates ideas from the kinetic theory of granular gases into the description of weak bedload transport and invites generalizations that explicitly include the geometry of particle-particle collisions.
+This work incorporates ideas from the kinetic theory of granular gases into the description of weak bedload transport and invites generalizations that explicitly include the geometry of particle-particle collisions and more realistic particle height and slip velocity-dependent fluid forces.
 
@@
 This thesis has produced new understanding of how to describe bedload transport using statistical physics, but the developments introduce many limitations deserving of future research attention.
-In some cases these are specific to the models produced in the thesis, but in others, they are shared in common with a majority of models in the stochastic sediment transport research paradigm \citep{Ancey2020,Furbish2021a}.
+In some cases these are specific to the models produced in the thesis, but in others, they are shared in common with a majority of models in the stochastic sediment transport research paradigm \citep[e.g.][]{Ancey2020,Furbish2021a}.
 
-\subsection{Channel dynamics and morphology}
+\subsection{Channel morphodynamics}
 
@@
 
-The second major limitation of this work concerns grain size.
-In actuality, sediments in rivers span a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.
+The second limitation of this work concerns grain size.
+Sediments in rivers span a range of sizes, and differential mobility based on particle size produces spatial sorting, both vertically as in bed armoring \citep{Parker1982,Wilcock1989,Aberle2006}, and laterally as patch, particle cluster, or riffle development \citep{Nelson2014,Venditti2017}.
 
-There have been a few works on stochastic modeling of sediment transport with multiple grain sizes \citep{Sun2000,Parker2000}, but these approaches are not spatially distributed, so there is as yet no stochastic method to understand sorting processes.
+Only a few works on stochastic modeling of sediment transport involve multiple grain sizes \citep{Sun2000,Parker2000}, but these approaches are not spatially distributed, so there is as yet no stochastic method to understand sorting processes \citep[cf.][]{Ancey2020b}.
 
@@
 Extending the motion-rest model of Ch.
-\ref{ch:ch3} to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice place to start.
+\ref{ch:ch3} to two grain sizes with a matrix formulation, then calibrating its parameters to experiments would be a nice starting point.
 
@@
 
-Earth's landforms are shaped by numerous transport phenomena from booming debris flows \citep{Iverson1997} to barely perceptible hillslope creep \citep{Deshpande2021}.
+Earth's landforms are shaped by numerous transport phenomena from catastrophic debris flows \citep{Iverson1997} to barely perceptible hillslope creep \citep{Deshpande2021}.
 Different transport phenomena are basically distinguished by the relative importance of fluid, granular, and gravitational forces in sustaining them \citep{Jerolmack2019}.
@@
 The research relates the overall sediment transport rate to the movements of individual grains.
-At base level, the thesis embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions which predict mean values as well as the magnitude of their fluctuations.
+At base level, the thesis embraces variability as an intrinsic part of Earth surface dynamics, and it produces descriptions of sediment transport which predict mean values as well as the magnitude of their fluctuations.
 
-The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscales without including variability in models, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving surface grain size distributions \citep{Parker1982,Andrews1983}.
+The founders of process geomorphology always acknowledged the role of variability in landscape evolution \citep{Horton1945,Strahler1952,Langbein1964}, although their main efforts were to develop strategies to describe landscapes without including variability, like averaged shear stresses to avoid fluid turbulence \citep{MeyerPeter1948,Bagnold1954}, competent conditions to replace climatic fluctuations \citep{Wolman1959,Wolman1978}, and representative grain sizes to avoid evolving surface grain size distributions \citep{Parker1982,Andrews1983}.
 
-The stochastic descriptions of sediment transport developed in this thesis show a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?
+The stochastic descriptions of sediment transport developed in this thesis show a methodology we can use to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?

### 2021-09-25 revision/pref.tex
uh
about to track changes.

--- 2021-07-05 pref.tex
@@
 The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
+Suggestions from Michael Church, Rui Ferreira, Roger Beckie, and Peter Ashmore improved the clarity of the writing.
+
 The thesis includes two published works, Chapters \ref{ch:ch3} and \ref{ch:downDiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
-Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to submission of the thesis.
-
-\endinput
-The
-
-This thesis deals with the transport of coarse sediment in flowing water from the perspective it fundamentally results from the movements of individual grains.
-The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.
-
-Geomorphology has long been characterized by observation, description, and the building of conceptual models, not mathematical ones.
-Only relatively recently has the science turned toward quantitative methods, with researchers working to frame observations in terms of underlying processes and to describe them with methods adapted from physics.
-
-Given the complexity of geomorphology problems, a complete reduction of geomorphology to physics would be narrow-minded, but we can nonetheless borrow ways of thinking.
-An extremely successful approach in physics has been to construct idealized models with little intention of direct realism, study them deeply to obtain complete understanding, and then, decades or centuries later, to embellish these models with more sophisticated features to describe real-world phenomena.
-An example is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow shows up in the deepest inquiries of theoretical physics, from quantum matter to cosmological inflation.
-This thesis applies this ``simple harmonic oscillator approach" to some problems in Earth science.
-As a result, there are big assumptions: sediment grains all have the same size, and water flows are constant through time. 
-The objective is to build up simple archetypes which can be embellished later, not to build \textbf{the model} of fluvial geomorphology. 
-That's for later! (And not for me.) CHEERS.
+Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication after acceptance of the thesis.

### 2021-09-28 draft/abs.tex
done

@@
 I conclude by summarizing these developments, discussing their implications for the statistical description of bedload transport, and suggesting how we can use this modelling progress to better understand landscapes.
-
-\endinput
-
-This research provides solutions to longstanding problems within the statistical approach to sediment transport.
-
-and the mathematical tools applied in this thesis should be of interest to researchers of many other problems. Most
-This research has broad implications across river science where there is increasing acknowledgement of the shortcomings of the continuum hypothesis for predicting dynamics in mountain streams.
-This work extends a wide body of work on individual particle motions which originates from the 1930s, and unifies it with more recent approaches to calculate the bulk bedload flux which have not, until now, been related to the movement characteristics of individual grains.
-
-A central task within Earth science is to understand and predict the evolution of landscapes due to flowing water.
-As rainfall channelizes and flows downhill, it carves out basins and etches in networks of intersecting channels.
-In the highlands of these networks, water arranges boulders and gravels into an intricate array of patterns -- steps, pools, bars, and riffles, which cooperate with vegetation both live and dead to set the stage for much of Earth's biota.
-In the lowlands, channels laden with old mountain sediments, weathered and abraded to fine particles, splay across floodplains and drift between configurations through millennia.
-
-Sediment transport is a main driver of all fluvial dynamics.
-Channel evolution ultimately occurs because individual sediment grains move from one location to another.
-Yet in a majority of modelling studies, landscapes are represented as continua, where the locations of individual grains are averaged away, and sediment transport is represented as a steady stream of mass, rather than the intermittent movements of individual grains
-Useful as this continuum approach may be, many fluvial phenomena are not well-suited for it.
-Sediment transport rates are known to fluctuate widely through space and time, and these fluctuations are understood to initiate bedform development and control channel widths.
-The largest grains in small mountain channels are known to confer the most stability, while violating any assumption of being small compared to the scales of interest.
-To understand fluvial dynamics, we need the capability to model discrete grains, not just continua.
-
-In this thesis, I present four or five years of my theoretical research into the movements of individual sediment grains in river channels.
-This sub-field of Earth science, with a focus on grain-scale process, has existed at least since Hans Albert Einstein in 1937, and within it, modellers have traditionally compromised on severe approximations to balance realism against mathematical difficulty.
-In this enterprise, perfectly flat beds which do not change shape, spherical grains, infinite movement velocities, and turbulence free flows have been typical assumptions to make progress, even though they have little basis in reality.
-The research presented here takes on more mathematical difficulty than before to introduce more realism, adding some bricks to the fortress, and introducing new methods to the field which should work well for other people to make more bricks later.
-I hope you find it useful!
-Happy reading.

### 2021-09-28 draft/ack.tex
done

--- 2021-07-07 ack.tex
@@
 Thank you all for every part of it.
-
-\endinput
-
-Mom, during my childhood, you worked full time, taught piano lessons, earned Master and Doctoral degrees, played organ every Sunday, consoled us through every difficulty, and engaged us in as many extracurricular activities as we were willing to take on.
-I cannot imagine a better role model of work ethic, open-mindedness, and devotion. I hope I picked a few things up!
-
-Dad, in my childhood I watched you solve an insane diversity of problems without hesitation, from easy ones, like worn out brake pads or stuck chainsaws, to hard ones, like burnt eyes and broken arms.
-On canoeing trips I got lessons in water currents, fish habits, tree species, electrons, engines, wings, hellgrammites, heat transfer, world history, radio waves, and probably everything else I could think of.
-Years of free science lessons!
-You taught me more about science and problem solving than school ever did.
-"Any problem can be fixed with the right tools" is basically the Jim Pierce philosophy.
-It it's incredibly useful in a PhD!
-Thanks for this and so much more.
-
-Kelsey and Kim, my earliest memories are sitting in your bedroom floor struggling through your books, asking for help every couple of paragraphs.
-You made me good at reading and taught me to enjoy it, and this is the basis of everything.
-Kim, you were my first role model, and I think of you all the time.
-Kelsey, I know it goes without saying, but thanks for a lifetime of guidance.
-You're now, as always (all state chorus, math field day, spelling bees, piano skills, grades, .
-.
-.
-), five steps ahead and far more successful.
-On the one hand, I'm deeply proud at your progress through life, with a stable career and now two beautiful kids, but on the other, I'm flatly shocked by it.
-I'll be playing catch up forever!
-
-Gugs, you're easily the most level-headed, good-natured, and reliable person I've ever had contact with, and to think I was lucky enough to be about half raised by you!
-You've been my second mother, my counsellor, at times my daily chauffeur, and my best role model of grace, consistency, and even-handedness.
-I'll never be able to thank you enough!
-One day I'll finally take on your 5am wakeups and figure out how to copycat your completely measured responses to all of life's challenges.
-
-Finally with regard to family, I'd like to thank my great aunt Alice.
-Thanks so much for paying my undergrad education!
-We hardly met before you passed, but we're family, and that was enough.
-It really means a lot.
-
-Next comes the mentors in education.
-
-I'm kinda tired because it's the end, so it's natural to drift on to recollections of my favorite times in life, Watoga and Williams River with my family, encyclopedias in Gugs' living room, watching eagles over the South Branch with dad, discussions with Johnathan Ramey about my education, the glowing fireflies under my great grandpa's tree, ``cross-up, cross-up, squeeze" in Mindy's calc class, or laughter about stolen watermelons with my sisters.
-I can skip years later and flip on the light in the physics lounge to find Megat asleep on the couch, visit Gugs on a trip back home, discuss Tsujimoto with Marwan, or walk through Kitsilano with Mary at the edge of dark.
-The PhD is all of it!
-Thank you all for every part of it.

### 2021-09-28 draft/ch1intro.tex
done

--- 2021-07-08 ch1intro.tex
@@
 The improved model is demonstrated to describe all of the different bedload velocity distributions which have been reported in experiments.
-
-\endinput
-
-An alternative statistical formulation of the sediment flux provides an approximate correspondence between these two definitions \citep{Ancey2006,Furbish2012}.
-This hinges on the probability $P[\textbf{u}_p | \textbf{x},t]$ that a particle contacts a control surface $S$ at position $\textbf{x}$ and time $t$ with velocity $\textbf{u}_p$. 
-This conditional probability is considered to result from a very large collection of identical systems selected at random moments in their evolution: that is, the conditional probability is an ensemble quantity \citep{Kittel1958}.
-
-In terms of this ensemble probability, the flux is \citep{Ancey2006}: 
-\be q_s = \int_{\text{all } \textbf{u}_p} \int_S P[\textbf{u}_p | \textbf{x},t] \textbf{u}_p \cdot d \textbf{S} d \textbf{u}_p . \label{eq:anceyflux} \ee
-In steady conditions, when the probability is independent of time -- $\partial P / \partial t = 0$.
-
-An approximate connection to a control volume flux can be derived by swapping the ensemble average for a spatial average by integrating along a control volume.  
-If the control volume is sufficiently long, $L \rightarrow \infty$, it can be seen as a stack of very many independent cross sectional surfaces. 
-These comprise a stack of replicas of $\mathcal{S}$, and at an instant, if spatial correlations in bedload transport are weak, each surface constitutes one configuration of particles intersecting $S$ with some set of positions and velocities contributing to $P$.
-One can then define $P$ by counting occurrences along this stack of replica surfaces with an integral across the control volume.
-
-With this concept of replica surfaces in mind, we can formalize the link with symbols. 
-Consider $n$ particles distributed throughout a volume at some set of positions $\textbf{x}_i$ with some set of velocities $\textbf{u}_i$, where $i=0,1,\dots, n$,
-The cloud of particles can be represented by its (discrete) density in a position-velocity phase space: 
-\be \rho(\textbf{x},\textbf{u}_p) = \sum_{i=1}^n M(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i). \ee  
-Here $M(\textbf{z})$ is a marker function, which is $1$ if $\textbf{z}$ is inside the particle, and $0$ otherwise.
-The marker's volume integral is $\int M(\textbf{x}) dV = \nu_p$, the particle volume.
-
-Using this density at sufficiently large $L$, the conditional probability $P[\textbf{u}_p | \textbf{x}]$ can be written approximately as an integral over the stack of replica surfaces:
-\be P[\textbf{u}_p|\textbf{x}] \approx \frac{1}{L} \int_0^L dx \rho(\textbf{x},\textbf{u}_p). \ee
-The integral runs along the downstream coordinate, and the equality becomes exact as $L \rightarrow \infty$.
-With this swap of ensemble averaging for spatial averaging, the bedload flux becomes 
-\be q_s \approx \int_{\text{all} \textbf{u}_p} \int_S \frac{1}{L} \int_0^L dx \sum_{i=1}^n M_a(\textbf{x}-\textbf{x}_i)\delta^3(\textbf{u}_p - \textbf{u}_i) \textbf{u}_p \cdot \textbf{k} dS d \textbf{u}_p = \frac{\nu_p}{L}\sum_{i=1}^n u_i , \ee
-so there is some correspondence between the surface and volume definitions of the flux.
-Yet this is never exact, except in the unphysical case of sediment transport with no spatial correlations (vanishing particle size, discontinuous trajectories).
-Apparently, the control volume and control surface formulations of the sediment flux are not equivalent, so we see in the literature two parallel streams of research.
-
-\subsection{First extensions of the Einstein model}
-
-Einstein's model provides a physically-based model for individual particle trajectories which includes the essential processes at play over global timescales when particles alternate between movement and rest.
-Despite its success, it oversimplifies particle transport, and a number of extensions and improvements have been made in the 85 years since its development.
-
-Einstein's model of particle movement seems to have been overlooked for the first several decades after its development, or possibly overshadowed by Einstein's later work, which uses its essential ideas for more obviously practical purposes.
-Eventually, curiosity into individual particle motions was re-sparked by cold war studies into the fate of radioactive contaminants in river channels \citep{Crickmore1962, Hubbell1964, Sayre1965,Yang1971}, and later by the need to estimate sediment transport rates from tracer measurements \citep{Yano1969, Todorovic1975, Nakagawa1976, Nakagawa1980, Hassan1991}.
-In relation to these issues, Einstein's model received a great deal of new attention which generated some generalizations of the model.
-
-The first set of modifications to Einstein's approach in this period concerned the form of the distributions used for step lengths and resting times.
-Einsteins flume experiments demonstrated that particle step lengths followed exponential distributions \citep{1937}, and this was confirmed by subsequent studies \citep{Yano1969, Nakagawa1976}.
-Yet these studies were all conducted in steady flows in artificial flumes, not real streams, where sediment transport conditions can be different.
-When sediment transport is observed before and after a flood, rather than during steady flow, the step lengths are better described by a Gamma distribution, this being the sum across the numerous individual steps that occurred during the flood \citep{Hassan1991}.
-When sediment transport occurs in the presence of dunes or other bedforms, similar modifications have been made to account for spatial differences in transport characteristics \citep{Crickmore1962, Hubbell1964, Sayre1965} and the embedding of particles within bedforms (which increases the periods of rest) \citep{Yang1971,Nakagawa1980}.
-
-A full statistical characterization of sediment transport is desirable because transport fluctuations are responsible for the initiation of bedforms \citep{Jerolmack2005,Bohorquez2016} and are beginning to show links to basic geomorphology considerations like the maintenance of stable channel widths \citep{Abramian2019,Abramian2020}.
-This is not to mention that because sediment transport fluctuations are ubiquitous, models which describe only mean bedload fluxes signals are conceptually incomplete.
-
-An additional reason to characterize fluctuations in bed load fluxes is because in reality, sediment flux measurements always involve a sampling interval.
-This introduces questions of the convergence of transport rate measurements \citep{Dhont2018,Ancey2020}.
-Fluctuations originate from processes which span a vast range of characteristic timescales.
-These can occur over a matter of seconds, such as differences in the velocities of grains due to a series of collisions \citep{Benavides2021}, to many hundreds of hours, such as the migrations of dunes and bedload sheets \citep{Hammamori1962,Hoey1992,Guala2014} or the cycling between aggradation and degradation in pools \citep{Dhont2018}.
-Such processes introduce temporal correlations into sediment transport signals which control the sampling interval over which measurements of the mean sediment flux converge \citep{Saletti2015,Singh2009,Singh2012}.
-To date, very few modelling works have addressed the observation scale depedence of the bedload flux, whereby measurements of mean fluxes, or statistical moments, shift with observation time \citep{Ancey2020}.
-
-These models produce successful descriptions of disparate experimental results, although their components are not clearly interpretable in terms of physical expecatations.
-For example, granular interactions were included in the Fan et al model as a Coulomb friction term.
-While such approximations are common within Earth science models \citep{Kirkby1971}, they are not entirely satisfying given that bedload particles undergo intermittent collisions with the granular bed.
-Although these motions are sometimes described as ``sliding", bedload particles do not slide in the same sense as a block on a plane, as the granular term in the Fan et al Langevin model indicates.
-In a similar way, the Ancey model does not have a particle-particle interaction term, and it includes what is apparently a Stokes drag term (linear in $u-U$), which is not applicable to bedload transport in water, since bedload particles are typically large compared to the viscous length scale \citep{Clift1978}.
-We have to wonder if a more detailed treatment of particle-bed interactions might produce a more general model of bedload particle velocities for application at the local and intermediate scales.
-
-One overarching methodology is the use of idealized noises in combination, where each represents a separate mechanical process affecting the movements of particles.
-This takes on more mathematical complexity than earlier studies, but it also introduces more realism.
-
-Regarding the movements of individual particles, the division into local, intermediate, global, and geomorphic ranges by \citet{Nikora2001,Nikora2002} and \citet{Hassan2017} provides a conceptual scheme by which to organize the current state of research.
-Einstein's original description \citep{Einstein1937} and those of his early followers \citep{} described only the global range associated with timescales over which particles cycle through motion and rest.
-This limitation was overcome in later works which included a finite duration of sediment motion in order to describe both intermediate and global scales, and Newtonian models have also attained descriptions of additional scales, such as local and intermediate \citep{}.
-Meanwhile, Langevin models have derived the velocity characteristics of particles within the local range \citep{}, although they do not manage to account for the full range of experimentally-observed distributions.
-The major shortcoming exposed by this framing in terms of the timescales of particle motion is that no models have yet resolved particles across the full set of timescales, so all of these descriptions are essentially incomplete.
-
-Models of the sediment flux face a similar set of shortcomings.
-Einstein and Bagnold founded complementary perspectives of the sediment flux.
-Both were mean field theories, but one counted particles crossing a surface using probabilistic arguments, while the other considered an average energy balance in a control volume.
-Research has come a long way since these studies, but the essential divide between surface and volume-based flux calculations still exists. These definitions are not equivalent.
-The erosion-deposition model takes steps toward joining these two mean-field theories into a common framework, though it involves many parameters which must be experimentally calibrated.
-A paradigm shift in this research stream came from the recognitition that the sediment flux is a fluctuating quantity best characterized by a probability distribution, and from the acknowledgement that measured values of the sediment flux depend on observation scale, meaning it is a scale-dependent quantity.
-Although simple models of the sediment flux probability distribution have been developed, they stand somewhat apart from the detailed dynamics of individual particles.
-Rather than describing particles moving downstream through cycles of entrainment and deposition as in \citet{Einstein1937}, these methods characterize particle motions by herustic entrainment and deposition rates, or interarrival time distributions, which are somewhat abstract objects, not concretely related.
-Despite the considerable progress these approaches have brought, the lack of connection to the underlying particle mechanics requires attention.
-
-Alongside this central result $q=E\ell$ for the mean flux, Einstein formulated the entrainment rate $k_E$ of the individual particle in terms of the force balance on the stationary particle, building insight into the grain-scale mechanics of entrainment.
-His original approach considers that entrainment is driven by the fluctuating lift force imparted by the turbulent flow, and it is resisted by the submerged weight the grain.
-The entrainment rate is calculated from the exceedance probability of the turbulent lift over the weight, providing an alternative to the critical shear stress concept which explicitly incorporates turbulent fluctuations in the fluid flow.
-This formulation of entrainment probability in terms of the exceedance of random driving quantities over (possibly
-random) resisting quantities \citep{Grass1970} another of Einstein's key contributions.
-\citet{Paintal1971} made a significant extension
-by including the random supporting geometry of bed particles into the force balance.
-More recently, \citet{Tregnaghi2012}
-amended the theory to include both force magnitude and duration, incorporating the recognition that both are relevant for particle entrainment \citep{Diplas2008, Valyrakis2013, Celik2014}.
-Refined theories of the single-particle entrainment rate, all fundamentally similar to the original ideas of \citet{Einstein1950}, have been carefully reviewed by \citet{Dey2018} and are a topic under active development.

### 2021-09-28 draft/ch2flipFlop.tex
done

--- 2021-07-08 ch2flipFlop.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\endinput
-Introducing such a description with the stochastic Exner equation would lend new mathematical language to the old unsolved problem in geomorphology: how does variability affect landscape evolution?
-The classic works all acknowledge variability \citep{Leopold1952}, although one of their main efforts was to construct strategies to describe landscapes without explicitly accounting for it, producing ideas like competent flows to summarize climatic variability \citep{Wolman1960}, characteristic grain sizes to overcome sorting processes \citep{Parker1982}, or average flow energies to avoid turbulence \citep{Bagnold1954}.
-Progress in stochastic approaches to river science \citep[e.g.][]{Furbish2021c,Ancey2020b} invites us to step beyond averaged descriptions, propagate noises through the governing equations of landscape evolution, and ask again with these classic works to what extent fluctuations shape Earth's surface.
-
-This chapter upscaled from individual particle motions to calculate the sediment flux.
-We can imagine going further, to upscale from the sediment flux to understand channel morphodynamics.
-The sediment flux is stochastic as a result of variability in the individual particle trajectories forming it.
-As a result, the channel morphodynamics should also adopt this stochasticity, although very few studies have evaluated this possibility \citep{Bohorquez2016, Jerolmack2005}.
-
-Earlier approaches to calculate the probability distribution of the sediment flux have counted the fluctuating number of particles in a control volume with population models \citep{Ancey2008,Ancey2014}, or they have evaluated the average rate of particles crossing a control surface using the interarrival time distribution of particles from upstream \citep{Turowski2010,Ancey2020,Heyman2013}, in what is basically an application of renewal theory \citep{Cox1962}.
-These approaches can be described as ``kinematic", in that they do not explicitly account for the trajectories of individual particles as they move downstream.
-
-Here, I have demonstrated how to describe the sediment flux probability distribution from a dynamical perspective, rather than a kinematic one, using stochastic dynamical equations to describe the downstream trajectories of individual grains (eq.
-\ref{eq:langevin}.
-In effect, the interarrival time distribution is replaced by the probability distributions of position for all individual particles available for motion, with their initial positions and velocities considered as random variables.
-Particles moving downstream are counted as they cross the control surface, giving rise to a parameter $\Lambda(T)$, dependent on observation scale, which is analogous to the heuristic rate constant in the renewal theories that were reviewed in section \ref{sec:renewal} \citep{Turowski2010,Ancey2020}
-This new description of the sediment flux from the grain scale particle dynamics, rather than characteristic kinematics derived from them, represents considerable progress toward a mechanistic understanding of sediment transport from statistical physics.
-
-\subsection{Outlook and future research}
-
-The phrasing of sediment fluxes the topographies they moderate as statistical quantities, having mean configurations and fluctuations, provides some context on the severe challenges encountered in predicting landscape change with deterministic equations \citep{}.
-What we observe, looking at Earth's landscapes, may be a single realization, drawn from a spectrum of possibilities, whose precise form evades description. 
-The stochastic perspective allows us to
-
-The sediment flux probability distribution in equation \ref{eq:fluxdist} represents a Poisson distribution with an observation-scale dependent rate.
-Poisson distributions have a relatively thin tail, meaning fluctuations are typically small \citep{Ancey2006}.
-This contrasts with many bedload transport experiments with wide-tailed fluctuations \citep{Ancey2008,Heyman2016,Fathel2015}, especially for weak flow conditions near initiation of particle motion \citep{Benavides2021}, exposing a need to generalize a mechanistic theory of the sediment flux developed here to provide wider fluctuations.
-
-There are essentially two mechanisms available by which wide fluctuations can arise \citep{Goh2008}.
-Both would be ultimately reflected in non-exponential interarrival times of particles \citep[e.g.][]{Turowski2010,Heyman2013}, although their origins would be unambigiously linked to the underlying particle transport mechanics.
-The first possibility is to introduce correlations between particle movements \citep{Ancey2008,Heyman2013,Ancey2014}, such as might arise when turbulent flow fluctuations entrain many particles at once \citep{Cameron2020} or drive waves of high velocity particles downstream simultaneously \citep{Nino1996}.
-In a theory such as the one I presented here, this first mechanism for wide fluctuations could be described by temporally unsteady velocities, entrainment or deposition rates, or diffusion coefficients in the Langevin model \ref{eq:langevin}.
-The second possibilty is to introduce interactions between moving particles, such as might result from collisions \citep{Lee2018} or wake vortices \citep{Schmeeckle2014} inducing particle entrainment, leading to collective waves of motion \citep{Ancey2014}.
-Including such effects would require a many-body generalization of the Langevin and master equations \ref{eq:langevin} and \ref{eq:master}, where the probability distribution of position is promoted to a joint form $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$, which is puzzling, since the number of particles $N(t)$ in the distribution fluctuates.
-The model of \citet{Ancey2014} hints at this possibility, although its starting point is not a dynamical equation as in \ref{eq:master}.
-Some hints as to how to generalize our work to include interactions between particles to describe wide-tailed sediment fluxes can be gleaned from literature on reaction-diffusion processes \citep{Hernandez2004,Cardy2006} and interacting gases \citep{Chapman1970,Brilliantov2004}.
-The further inclusion of interacting particle models into bedload transport theory would be an exciting development.
-
-\endinput
-
-A simpler approach is to conduct the averages in Fourier space. Integrating eq. 1, using its solution in the probability distribution, then Fourier transforming provides
-\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi.\ee
-Taking time derivatives and conducting the averages using known characteristics of averages of exponentials of Gaussian white noise \citep{Gardiner1983,VanKampen2007} and the Furutsu-Norikov procedure for time derivatives of averages involving dichotomous noise \citep{Shapiro1978}, in a method similar to \citep{Balakrishnan1993} provides the Fourier-space master equation
-\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
-and inverse Fourier transforming provides the desired real-space master equation
-
-At short times $t\ll k^{-1}$, this becomes
-\be \sigma_x(t)^2 \sim \frac{k_E k_D V^2}{k^2} t^2 + \frac{2 k_E D }{k}t, \label{eq:major}\ee
-so provided that $t\ll k^{-1}$, eq. \ref{eq:major} scales as $\sigma_x^2 \sim t^2$ when $t \gg \frac{2Dk}{V^2 k_D}$ and $\sigma_x^2 \sim t$ when $t \ll \frac{2Dk}{V^2 k_D}$.
-Similarly at long times $t\gg k^{-1}$, eq \ref{eq:minor} becomes
-\be \sigma_x^2 \sim \frac{2k_Ek_D V^2}{k^3}t. \ee
-
-Calculated in this way, the sediment flux distribution is observation-scale dependent, describing how sediment fluxes depend on the period over which they are measured.
-I demonstrated that scale-dependence results from velocity fluctuations in the motion state during the movements of individual particles, and I introduced a Peclet number ($\Pe$) to characterize the relative strength of these fluctuations.
-
-We can imagine generalizing the description in equation \ref{eq:flipfloplangevin} to include time-dependent phenomena interactions by including time dependence in the movement velocity ($V$), diffusivity ($D$), or entrainment and deposition rates ($k_E$ and $k_D$) as necessary, although the resulting equations would likely require numerical solutions.
-
-\subsection{Stochasticity in landscape evolution}
-
-\citet{Exner1925} originally phrased channel morphodynamics in terms of spatial gradients in the sediment flux (sec. \ref{sec:landscape}), writing
-\be (1-\phi) \pt h(x,t) = -\px q(x,t). \label{eq:stocEx}\ee
-In this one-dimensional equation, $h(x,t)$ is a continuous field representing the longitudinal channel profile.
-Traditionally, the Exner equation is evaluated considering the flux as a deterministic quantity \citep{Parker2007,Viparelli2011,An2017}, which provides a solitary trajectory of the bed profile $h(x)$ through time.
-
-Considering the flux in the Exner equation as a stochastic quantity invites us to reinterpret channel dynamics in a way which few researchers have explored \citep{Jerolmack2005,Bohorquez2016}.
-In contrast to the traditional analysis, a stochastic Exner equation with fluctuating $q(x,t)$ generates an ensemble of possibilities for the channel profile $z(x)$ at each instant.
-This set of possibilities is characterized by a probability distribution $P(\{z(x)\},t)$ for each possible profile.
-
-Mathematically, since the profile is a field, evaluating such a probability distribution of channel profiles requires statistical field theory \citep{Kardar2007}.
-A starting point on this analysis might be $P(\{z(x)\},t) = \bra \delta[z(x,t)-h(x,t)]\ket,$ where $h(x,t)$ involves the stochastic flux $q(x,t)$ from equation \ref{eq:stocEx}.
-In this equation, $\delta[f(x)]$ is a Dirac delta functional, which generalizes the delta function $\delta(x)$ from a single-valued parameter $x$ to a continuous field $f(x)$.
-Such ideas are commonly applied to understand polymers and membranes \citep{Kawakatsu2001,Nelson2004}, but they have not been introduced in geomorphology.
-Such a probabilistic description of bed profiles would be an interesting extension of the methodology applied in this thesis so far to understand landscape evolution in an alternative way.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-09-28 draft/ch5langevin.tex
done

@@
 Such an effort would produce additional insight into the controls of transverse, vertical, and rotational movements on the downstream velocities of individual grains.
-
-\endinput
-
-Sediment transport experiments reveal correlations between particle size and the shape of the bed load velocity distribution.
-Experiments with smaller particles tend to give exponential distributions, and those with larger particles give Gaussian distributions.
-In fluid dynamics, the dissipation characteristics of particle-particle collisions in viscous flows are known to depend on the particle size and approach velocity through the Stokes number. 
-In kinetic theory, it is known that gases of ideal elastic particles generate Gaussian (Boltzmann) velocity distributions, while gases of inelastic grains generate non-Gaussian distributions \citep{Chapman1970,Brilliantov2004}.
-Taken together, these ideas suggest that we might relate the shape of the particle velocity distribution to particle size.
-We can estimate typical Stokes numbers of colliding bed load particles in experiments as $...$, using with the flow shear velocity and mean streamwise sediment velocity to calculate $V$.
-Estimating in this way, transport experiments with exponential velocities have $St \sim 1-10$, those with neither exponential nor Gaussian velocities have $St \sim 10-100$, and those with Gaussian velocities have $St > 100$.
-In experiments relating restitution coefficient to Stokes number for idealized collisions, restitution coefficients vary sharply from $0$ to $1$ as $St$ ranges from $1$ to $500$ \cite{Marshall2011,Joseph2001,Yang2006}.
-Although collision geometry and grain shape certainly complicate the narrative, these values of $St$ are consistent with our model conclusion that the shape of the particle velocity depends on the elasticity of collisions.
-
-In real channels, grain sizes often span a wide range.
-A major implication of the dependence of the shape of the particle velocity distribution on grain size is that different grain sizes in a mixture will impart distinct fluctuation signatures to the overall bulk transport rate.
-Even in the absence of sorting effects and differential mobility, smaller grains can be expected to carry more control over the largest fluctuations in the overall transport rate, since their velocity distributions have wider tails.
-
-Particle trajectories are a compromise between turbulent drag and momentum dissipation by particle-bed collisions \citep{Schmeeckle2014}.
-For particles submerged in a viscous fluid, the amount of momentum dissipated by collisions is controlled by the particle Stokes number ($\St$), which characterizes the balance of pre-collisional momentum against the viscous pressure generated in the squeeze film between approaching particles \citep{}.
-The Stokes number is defined by $\St = \rho_s V d/(9\nu \rho)$, where $\rho_s$ and $\rho$ are the solid and fluid densities, $V$ is the approach velocity, $d$ is the particle size, and $\nu$ is the fluid viscosity.
-Momentum dissipation in submerged collisions shows a sharp dependence on $\St$ \citep{Joseph2001, Schmeeckle2001}, and hence also the particle size.
-This suggests particle size may have a nontrivial control over bedload velocity distributions.
-
-Here, motivated by the recognition that experimental particle velocity distributions appear to vary with the grain size, I hypothesize that the shape of the streamwise velocity distribution is controlled by the momentum dissipation characteristics of particle-bed collisions.
-For particles submerged in a viscous fluid, the amount of momentum dissipated by collisions is controlled by the particle Stokes number ($\St$), which characterizes the balance of pre-collisional momentum against the viscous pressure generated in the squeeze film between approaching particles \citep{}.
-The Stokes number is defined by $\St = \rho_s V d/(9\nu \rho)$, where $\rho_s$ and $\rho$ are the solid and fluid densities, $V$ is the approach velocity, $d$ is the particle size, and $\nu$ is the fluid viscosity.
-Momentum dissipation in submerged collisions shows a sharp dependence on $\St$ \citep{Joseph2001, Schmeeckle2001}, and hence also the particle size.
-This suggests particle size may have control over bedload velocity distributions.
-
-The elasticity of granular collisions within a viscous fluid depends on the particle size.
-
-In this drag force, the collision force \ref{eq:col}, the turbulent fluctuations $\xi(t)$, collision times $\tau_k$, and dissipation coefficients $\varepsilon_k$, can take any values consistent with their distribution and correlation functions.
-This set of possiblities defines a statistical ensemble.

### 2021-09-28 draft/ch6conc.tex
done

--- 2021-07-08 ch6conc.tex
@@
 The stochastic descriptions of sediment transport developed in this thesis hint toward a methodology to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?
-
-\endinput
-
-Its closest neighbours may be rarefied hillslope transport, where solitary grains tumble down hillslopes \citep{Williams2021}, and intense bedload transport, where particles creep downstream in a dense granular flow, supported by collisions with other moving grains \citep{Frey2014}.
-These phenomena differ only by viscosity or density, not mechanically.
-
-\citet{Ashworth1989} and \citet{Warburton1992} identified three phases of bedload transport in gravel-bed streams that occur as the flow increases.
-Phase 1 is characterized by movement of sand only, while gravel remains stationary; phase 2 involves partial mobility of the smaller gravel sizes; and phase 3 is the full mobility of all grain sizes represented on the bed.
-Phase 2 is most common in gravel-bed rivers, but phase 3 has

### 2021-09-28 draft/pref.tex
done

--- 2021-07-05 pref.tex
@@
 Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication subsequent to submission of the thesis.
-
-\endinput
-The
-
-This thesis deals with the transport of coarse sediment in flowing water from the perspective it fundamentally results from the movements of individual grains.
-The research has been motivated by the belief that geomorphology as a science will benefit from mechanistic, process-based, mathematical models of sediment transport processes.
-
-Geomorphology has long been characterized by observation, description, and the building of conceptual models, not mathematical ones.
-Only relatively recently has the science turned toward quantitative methods, with researchers working to frame observations in terms of underlying processes and to describe them with methods adapted from physics.
-
-Given the complexity of geomorphology problems, a complete reduction of geomorphology to physics would be narrow-minded, but we can nonetheless borrow ways of thinking.
-An extremely successful approach in physics has been to construct idealized models with little intention of direct realism, study them deeply to obtain complete understanding, and then, decades or centuries later, to embellish these models with more sophisticated features to describe real-world phenomena.
-An example is the block on the spring, the simple harmonic oscillator, which starts as a toy model studied by every first year physics student, but somehow shows up in the deepest inquiries of theoretical physics, from quantum matter to cosmological inflation.
-This thesis applies this ``simple harmonic oscillator approach" to some problems in Earth science.
-As a result, there are big assumptions: sediment grains all have the same size, and water flows are constant through time. 
-The objective is to build up simple archetypes which can be embellished later, not to build \textbf{the model} of fluvial geomorphology. 
-That's for later! (And not for me.) CHEERS.

### 2021-09-28 revision/abs.tex
done

@@
 
-Bedload transport is the movement of coarse grains through river channels by bouncing, rolling, and sliding.
-Because coarse grains control river stability, predicting the rate of bedload transport is a fundamental problem in river science.
+Bedload transport is the movement of coarse grains through river channels by bouncing, rolling, and sliding.Because coarse grains control river stability, predicting the rate of bedload transport is a fundamental problem in river science.
 This problem is usually approached with continuum mechanics, but this approach is questionable considering that coarse sediment grains rarely move in densities approximating a continuum.
@@
 Third, I incorporate the sediment burial process into a model of downstream sediment transport, predicting how grains move downstream when they can become buried.
-Finally, describe bedload dynamics on short timescales, predicting the movement velocities of bedload particles using methods adopted from granular physics.
+Finally, I describe bedload dynamics on short timescales, predicting the movement velocities of bedload particles using methods adopted from granular physics.
 I conclude by summarizing these developments, discussing their implications for the statistical description of bedload transport, and suggesting how we can use this modeling progress to better understand landscapes.

### 2021-09-28 revision/ack.tex
done

@@
 
-Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Avis, it's impossible to explain how thankful I am for your guidance.
+Most of all, to my mom Calisa Pierce, dad Jim Pierce, sisters Kelsey and Kim, and grandmother Wilma Avis, it's impossible to explain how thankful I am for your guidance.
 Thanks also to my great aunt Alice who graciously paid for my undergraduate education.

### 2021-09-28 revision/app_ch2.tex
done

@@
 
-Making the short-hands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of Eq. \ref{eq:logi} gives
+Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking a time derivative of Eq. \ref{eq:logi} gives
 \be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket. \label{eq:furut}\ee
@@
 \begin{multline}  P(x,t) = \frac{1}{V} e^{-k_E t}  \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big]\\ \times \int_0^t du \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) \\ \times \mathcal{L}^{-1}\Bigg\{ \frac{\exp\Big[ \frac{-V|x|}{2D}\sqrt{a+b s}\Big]}{\sqrt{a + b s}}\Bigg\}(u) , \label{eq:magic}\end{multline}
-where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$ are short-hands.
+where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$ are shorthands.
 
@@
 
+\section{Asymptotic limits of the flux rate function}
 \label{sec:fluxlimits}

### 2021-09-28 revision/app_ch5.tex
done

@@
 I will later take a GWN limit from this noise.
-With this assumption, integrating Eq.
-\ref{eq:langevin} over a small time interval $\delta t$ and incorporating the It\^{o} interpretation for the collision term provides
+Similar derivations are available in some references \citep[e.g.][]{Kanazawa2017,Suweis2011}, and the precise connection between GWN and Poisson noise is discussed in \citet{VanDenBroeck1983}.
+
+From this starting point, integrating Eq. \ref{eq:langevin} over a small time interval $\delta t$ under the It\^{o} interpretation for the collision term provides
 \be     
@@
 \end{align}
-Finally, taking $\delta t \rightarrow 0$ and limiting the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ obtains the master equation \ref{eq:master} \cite[e.g.][]{VanDenBroeck1983}.
+Finally, taking $\delta t \rightarrow 0$ and limiting the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ obtains the master equation \ref{eq:master}.
 \section{Derivation of steady-state solution}
@@
 \be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] ,\ee
-where the coefficients $R_l^\pm$ are the residues (at least up to a constant factor) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
+where the coefficients $R_l^\pm$ are the residues (up to constant factors) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
 \be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee

### 2021-09-28 revision/ch1intro.tex
done

@@
 Sediment moves in different modes depending on the relative importance of the fluid forces against the weight of sediment grains \citep{Bagnold1956}.
-When particles are coarse, as in gravel-bed rivers, the fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}.
+When particles are coarse, as in gravel-bed rivers, fluid forces are relatively weak and particles move as ``bedload", by bouncing, rolling, and sliding along the bed surface \citep{Kalinske1947}.
 In bedload transport conditions, fluid turbulence and the irregular bed surface adopt particular control over the sediment dynamics \citep{Ferreira2015}.
@@
 Unfortunately, existing approaches to compute the bedload flux are inadequate, and predictions often deviate from measured values by orders of magnitude \citep{Gomez1989, Barry2004, Bathurst2007a, Recking2012}.
-Given that the problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020a,Ancey2020}.
+Given that the problem has been researched intensively for over a century now \citep{Gilbert1914}, it is clear that new research strategies are needed \citep{Ancey2020}.
 
@@
 
-A basic problem in sediment transport modeling is to predict the downstream movement of individual particles.
+A basic problem in sediment transport modeling is to predict the downstream movements of individual particles.
 At first glance, this seems an elementary problem in general physics, but numerous challenges appear.
@@
 Moving particles that encounter the bed with sufficiently low velocities can settle into pockets which protect them from the flow \citep{Miller1966} and cause deposition \citep{Charru2004}.
-These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2010,Celik2014}.
-As a result, particles generally alternate through sequences of entrainment and deposition, but the exact times at which these alternations occur is difficult to characterize \citep{Einstein1937}.
+These pockets are not permanent shelter because rearrangement of the surrounding bed by entrainment of neighboring particles or subsurface creep \citep{Houssais2016,Frey2014} can re-expose particles to the flow, and sufficiently strong turbulent fluctuations \citep[e.g.][]{Cameron2020} can overcome shelter even if its geometry is not disturbed \citep{Valyrakis2010,Celik2014}.
+As a result, particles generally alternate through sequences of entrainment and deposition, but the exact times at which these alternations occur are difficult to characterize \citep{Einstein1937}.
 
@@
 Apparently, tracer virtual velocities depend on the observation time, in a phenomenon which has been called ``advective slowdown" \citep{Ferguson2002,Haschenburger2011, Haschenburger2013, Pelosi2016}.
-This observation time dependence obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
+This observation-time dependence obscures the relationship between measured tracer virtual velocities and actual sediment fluxes, exposing a need for further research into how exactly individual particle motions depend on observation scale and what processes generate advective slowdown.
 
@@
 
-Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his PhD was to find the probability density $P(x,t)$ that a particle had travelled a net distance $x$ after a time $t$ has elapsed.
+Einstein's experiments indicated that both step lengths and resting times were well-described by exponential distributions, and the focus of his Ph.D.
+was to find the probability density $P(x,t)$ that a particle had traveled a net distance $x$ after a time $t$ has elapsed.
 In the course of this research, Einstein formulated an example of what is now called the continuous time random walk, but this approach not formalized until much later \citep{Montroll1965}.
-In Earth science, random walks have been widely applied to hydrologic problems \citep{Berkowitz2006}, but their application to surface processes remains relatively uncommon \citep[e.g.][]{Schumer2009}.
+In Earth science, random walks have been widely applied to hydrologic problems \citep{Berkowitz2006}, but their application to Earth surface processes remains relatively uncommon \citep[e.g.][]{Schumer2009}.
 
-Einstein's assumptions imply that if the position of a single sediment particle at a given time is $x(t)$, the assumptions of infinite movement velocity, random resting durations, and random movement distances between entrainment and depositon (step lengths) can be written
+Einstein's assumptions of infinite movement velocity, random resting duration, and random movement distance imply that the particle position $x(t)$ can be expressed by the dynamical equation
 \be \dot{x}(t) = \mu(t), \label{eq:einlangevin}\ee
@@
 Here $N(t)$ is the number of particle entrainments in time $t$ which is a Poisson random variable, distributed as $P(N) = e^{-k_E t} (k_E t)^N/N!$.
-The $t_i$ are distributed according to $P(t) = k_E\exp(-k_E t)$, and the $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
-Figure \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as a sequence of steps and rests.
+The $t_i$ are the times at which particle entrainment events occur.
+These are separated by resting times $\tau = t_{i+1}-t_i$ distributed as $P(\tau) = k_E\exp(-k_E \tau)$.
+The $s_i$ are step lengths distributed according to $P(s) = \ell^{-1}\exp(-s \ell^{-1}).$
+Fig. \ref{fig:einsteinfig} panel (a) sketches the pulsed noise $\mu(t)$, and panel (b) sketches the resulting global range bedload sediment trajectories as an alternating sequence of steps and rests.
 
-Equation \ref{eq:einlangevin} is a dynamical equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term (Eq.
-\ref{eq:einrando}).
+Eq.
+\ref{eq:einlangevin} is a phenomenological equation representing how the position of a sediment grain evolves through time \citep{Kubo1978}, similar to Newtonian mechanics \citep{Goldstein1997} but with random driving term \ref{eq:einrando} and involving the sediment velocity, not acceleration.
 \begin{figure}[!htbp]
@@
 This equation can be solved by standard methods (series solutions or transform calculus) \citep{Arfken1985,Prudnikov1992a} to reproduce the original result of \citet{Einstein1937} for the probability distribution of position of a sediment particle:
-\be P(x,t) = \Big[\delta(x) e^{-k_E t} + e^{-k_E t - x/\ell} \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big)\Big]\theta(x)\theta(t) \label{eq:eindist} \ee
+\be P(x,t) = \Big[\delta(x) e^{-k_E t} + e^{-k_E t - x/\ell} \sqrt{\frac{k_E t}{\ell x}}\mathcal{I}_1\Big(2 \sqrt{\frac{k_E x t}{\ell}}\Big)\Big]\theta(x)\theta(t). \label{eq:eindist} \ee
 Here, $\mathcal{I}_\nu$ is a modified Bessel function of order $\nu$. 
@@
 Studies by \citet{Lisle1998}, and \citet{Lajeunesse2017} generalized the Einstein theory to include intermediate timescales.
-They approximated particle velocities as constant (neglecting fluctuations), and they assumed that the movement times are exponentially distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
-The analogue of Einstein's model equation \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
+They approximated particle velocities as constant (neglecting fluctuations), and they assumed that the movement times are exponentially-distributed random variables just like resting times, but this time characterized by a deposition rate $k_D$, whose reciprocal is the average period of time a particle spends in motion between entrainment and deposition.
+The analogue of Einstein's model Eq. \ref{eq:einlangevin} with a finite movement velocity $V$ can be written
 \be \dot{x} = V\eta(t), \label{eq:lislelangevin}\ee
 where the noise is now a ``dichotomous Markov noise" \citep{VanDenBroeck1990,Bena2006}, which is essentially a random switch or telegraph signal that alternates between ``on" ($\eta(t) = 1$, meaning the particle is moving) and ``off" ($\eta(t) = 0$, meaning the particle is resting) \citep{Cox1965,Horsthemke1984, Masoliver1991, Masoliver1996}.
-This noise is displayed in Fig. \ref{fig:lislefig} (a). Some particle trajectories produced by Eq. \ref{eq:lislelangevin} are displayed in Fig. \ref{fig:lislefig} (b).
+This noise is displayed in Fig. \ref{fig:lislefig} panel (a). Some particle trajectories produced by Eq. \ref{eq:lislelangevin} are displayed in Fig. \ref{fig:lislefig} panel (b).
 
@@
 Incorporating the duration of sediment motion, the mean position of the sediment grain remains linear in time: $\langle x (t) \rangle  = k_E V t/k$, representing movement with a virtual velocity $V_\text{virt} = k_E V/k$, which is the fraction of time spent in motion ($k_E/k)$ \citep{Ancey2006} multiplied by the velocity during the motion state.
-In contrast, the variance develops a two range scaling.
+In contrast to the instantaneous step model, the variance now develops a two range scaling.
 Computing $\sigma_x^2$ provides
@@
 Researchers eventually included realistic granular interactions in many-particle simulations of bedload transport using the discrete element method \citep[e.g.][]{Cundall1979, Wachs2019}.
-The early works utilizing this approach used a two dimensional domain with a simplified ``slab" flow geometry \citep{Haff1993,Gotoh1997}, while later works included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
+The early works utilizing this approach used a two-dimensional domain with a simplified ``slab" flow geometry \citep{Haff1993,Gotoh1997}, while later works included synthetic turbulence to drive particles \citep{Schmeeckle2003,Maurin2015} or clever reduced-complexity representations of the flow \citep{Clark2015,Clark2017}.
 
@@
 
-Another category of models describes bedload particle velocities using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particles dynamics are driven by Gaussian white noise \citep{Kubo1978}.
+One category of analytical models describes bedload particle velocities during the motion state using Langevin equations \citep{Ancey2014,Fan2014}, whereby the particle dynamics are driven by Gaussian white noise \citep{Kubo1978}.
 
@@
 This drag is resisted by a heuristic Coulomb friction term $-\Delta \text{sgn}(u)$, introduced as a proxy for particle-bed collisions.
-This model may have been inspired by models of Brownian motion with Coulomb friction in the physics literature \citep[e.g.][]{DeGennes2005,Touchette2010,Menzel2011}.
+This model is reminiscent of Brownian motion models with Coulomb friction from the physics literature \citep[e.g.][]{DeGennes2005,Touchette2010,Menzel2011}.
 The equation governing the probability distribution $P(u,t)$ of the particle velocity is called a Fokker-Planck equation, and this can be derived from Eq.
@@
 
-\citet{Ancey2014} formulated a Langevin equation to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
+\citet{Ancey2014} formulated a different Langevin model to describe the Gaussian velocity distributions observed in the other subset of experiments. They wrote for the streamwise velocity 
 \be t_r \dot{u}(t) = -(U-u) + \sqrt{2D}\xi(t),\ee
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/fanAncey.pdf}
-	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al model, where Coulomb friction impedes turbulent drag.
+	\caption{Panel (a) demonstrates the trajectory of a particle within the Fan et al.
+model, where Coulomb friction impedes turbulent drag.
 Panel (b) shows the resulting exponential particle velocity distribution, having relatively large fluctuations with mode at $u=0$.
-	Panel (c) shows the analogous trajectory from the Ancey et al model, where turbulent Stokes drag generates the Gaussian velocity distribution seen in (d).
+	Panel (c) shows the analogous trajectory from the Ancey et al.
+model, where fluctuating Stokes drag generates the Gaussian velocity distribution seen in panel (d).
 The latter has relatively narrow and symmetric fluctuations, with a positive mode well above $u=0$.}
@@
 \end{figure}
-Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and several experiments have revealed other distributions besides exponential and Gaussian \citep[e.g.][]{Houssais2012,Liu2019}.
+Although these descriptions describe two end-member sediment velocity distributions, Gaussian and exponential, the underlying mechanical reason why bedload transport has shown multiple velocity distributions remains an open question, and several experiments have revealed other distributions besides exponential and Gaussian which still lack any physical description \citep[e.g.][]{Houssais2012,Liu2019}.
 A comprehensive model of the full range of bedload transport observations remains a desirable target.
@@
 Here, particles are shown moving from time $t-\Delta t$ (faint white particles) to time $t$ (grey particles).
-Each particle's downstream displacement $\Delta x_i$ is indicated.
+The downstream displacement $\Delta x_i$ for each particle is indicated.
 These displacements define the downstream velocities $u_i \approx \Delta x_i/\Delta t.$ At the instant $t$, the surface definition of the flux Eq.
-\ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7).$, while the volume definition Eq.
+\ref{eq:surfflux} gives $q = \frac{1}{L}(u_6+u_7)$, while the volume definition Eq.
 \ref{eq:controlflux} gives $\Phi = \frac{1}{L}(u_1+\dots+u_5)$,
@@
 \be q = \int_S c(\textbf{x},t)\textbf{u}\cdot d\textbf{S}. \label{eq:surfflux} \ee
-This definition involves the volumetric concentration $c$ of particles and their velocites at the instant they cross the control surface.
+This definition involves the volumetric concentration $c$ of particles and their velocities at the instant they cross the control surface.
 These quantities require careful interpretation because bedload particles do not constitute a continuous field \citep{Furbish2012a}.
@@
 Bagnold's assumption that the power available to move sediment scaled with the excess shear stress leads to unphysical results over arbitrarily sloping beds \citep{Seminara2002}, and the flow power dissipated by sediment transport shows only a weak correlation with the sediment flux \citep{Ancey2008}, while Bagnold assumed they were directly proportional.
-These issues have been hinted when calibrating Bagnold's formula to data, where the parameters $k$ and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
-Bagnold's formulation also faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018}.
+These issues are hinted at when calibrating Bagnold's formula to data, where the parameters $k$ and $\tau_c$ take on unphysical values at low transport rates \citep{Nino1996}.
+Bagnold's formulation also faces the notorious challenge of defining the critical shear stress $\tau_c$ for the initiation of sediment transport \citep{Paintal1971,Kirchener1990,Houssais2015,Clark2017,Allen2018,Pahtz2020}.
 The difficulties with Bagnold's approach have motivated many revisions of his theory.
@@
 
-Among these revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
+Among revisions of Bagnold, one category shows a return to the probabilistic ideas of Einstein \citep{Parker2003,Ancey2006}.
 Einstein formulated his original model of individual particle trajectories in terms of entrainment and deposition rates using the idealization of grains moving downstream through a sequence of instantaneous steps \citep{Einstein1937}.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch1/yalinDrawing.pdf}
-	\caption{Einstein’s conceptual picture, modified from \citet{Yalin1972}. Particles move in discrete jumps of length $\ell$ from
-left to right through an array of adjacent control volumes. The bedload flux is the rate of particles crossing
+	\caption{Einstein’s conceptual picture, modified from \citet{Yalin1972}, involves an array of adjacent control volumes. Particles move in discrete jumps of length $\ell$ from
+left to right. The bedload flux is the rate of particles crossing
 the surface $S$ per unit width and time, collected from all upstream control volumes.}
@@
 
-Einstein partitioned the channel into a sequence of identical control volumes ($V$), and calculated the average rate at which particles cross a control surface by aggregating the contributions from each upstream control volume.
+Einstein partitioned the channel into a sequence of identical control volumes ($V$) and calculated the average rate at which particles cross a control surface ($S$) by aggregating the contributions from each upstream control volume.
 Each control volume has downstream length $\ell$ which is also the average particle step length.
@@
 so Einstein's mean sediment flux across the control surface (Eq. \ref{eq:einflux}) becomes
-\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{l=n}^\infty = \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
+\be q = \frac{\rho \ell}{T} \sum_{n=1}^\infty \sum_{i=n}^\infty \frac{(k_E T)^i}{i!}e^{k_E T}= \frac{\rho \ell}{T}\sum_{n=1}^\infty n \frac{(k_E T)^n}{n!}e^{-k_E T} = \rho k_E \ell.\ee
 Noticing that $\rho k_E$ is the entrainment rate of a single grain multiplied by the density of grains available for entrainment on the bed, we can summarize the Einstein theory as 
@@
 Although Einstein and many followers have attempted to formulate these connections \citep[e.g.][]{Einstein1950,Grass1970,Paintal1971}, there are still no comprehensive solutions.
-Linking Einstein's entrainment and deposition rates to the grain scale mechanics remains an active area of research that is far from complete \citep[e.g.][]{Tregnaghi2012,Dey2018}.
+Linking Einstein's entrainment and deposition rates to the grain scale mechanics remains an active area of research with many open problems \citep[e.g.][]{Tregnaghi2012,Dey2018}.
 
@@
 
-One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by \citet{Charru2004,Charru2006}.
+One uniquely successful strategy to relate Einstein's model to flow and sediment characteristics was developed by Charru et al.
+\citep{Charru2004,Charru2006}.
 This ``erosion-deposition model" phrases sediment transport as a mass balance within a control volume using Einstein's entrainment rate $E$ and the complementary deposition rate $D$.
 These rates are related to flow and sediment properties using scaling relations obtained from experiments \citep{Charru2004, Charru2006, Lajeunesse2010,Seizilles2014,Lajeunesse2015}.
-This approach can be characterized as a mixture of the Einstein and Bagnold strategies for its incorporate of both probabilistic rates and scaling relations.
+The Charru et al.
+approach can be characterized as a mixture of the Einstein and Bagnold strategies for its incorporate of both probabilistic rates and scaling relations.
 
@@
 
-Equation \ref{eq:charru} indicates that the mean flux in steady transport and flow conditions is the implicit solution to the equation $E = D.$
-Using the scaling relations Eqs. \ref{eq:charru1} and \ref{eq:charru2} provides the mean particle activity
+Eq. \ref{eq:charru} indicates that the mean flux in steady transport and flow conditions is the implicit solution to the equation $E = D.$
+Using the scaling Eqs. \ref{eq:charru1} and \ref{eq:charru2} produces the mean particle activity
 \be \gamma = \frac{\alpha_1}{\alpha_2}\frac{\tau-\tau_c}{d^2 V_s^2}.\ee
@@
 
-For a simple example of the Furbish et al formalism, consider that particles move with a constant velocity $V$ and have a deposition rate $k_D$. 
-Then the probability density that a particle moves \textit{exactly} a distance $x$ in time $t$ is $f(x,t) = \delta(x-Vt)k_D\exp(-k_D t), $ so the probability that it moves 
+For a simple example of the Furbish et al. formalism, consider that particles move with a constant velocity $V$ and have a deposition rate $k_D$. 
+Then the probability density that a particle moves \textit{exactly} a distance $x$ in time $t$ is $\delta(x-Vt)k_D\exp(-k_D t), $ so the probability that it moves 
 \textit{at least} a distance $x$ in $t$ is
@@
 \citet{Nakagawa1976} and \citet{Tsujimoto1978} developed an alternative statement of the Exner equation using Einstein's entrainment and deposition rates.
-According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition rates:
+According to their formulation, spatial gradients in the sediment flux arise due to local discrepancies in entrainment and deposition:
 \be \partial_x q(x,t) = E - D. \ee 
@@
 The major limitation of these approaches are that they assume river beds are continuous.
-This can be justified as an average over the detailed configurations of individual bed grains \citep{Coleman2009}.
+This can be justified as an average over the detailed spatial configurations of individual bed grains \citep{Coleman2009}.
 This averaging process is expected to obscure the relevant morphological evolution processes whenever individual grains are large compared to the scales of interest \citep[e.g.][]{Shobe2021}.
@@
 Despite widespread acceptance that sediment transport rates always fluctuate, we still have relatively little understanding of how to describe these fluctuations.
-\citet{Hamamori1962} was probably the first to calculate a probability distribution for the sediment flux, but this subject fell into relative obscurity for a long period, until it was revisited by \citet{Nikora1997}, \citet{Ancey2006}, and others.
+\citet{Hamamori1962} was probably the first to calculate a probability distribution for the sediment flux, but this subject fell into relative obscurity for a long period, until it was revisited by \citet{Nikora1997}, \citet{Sun2000}, \citet{Ancey2006}, and others.
 
@@
 
-This averaging process in conjunction with transport fluctuations introduces scale-dependence to sediment transport measurements, whereby mean sediment transport rates (and other statistical moments) depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
+The averaging process in conjunction with transport fluctuations introduces scale-dependence to sediment transport measurements, whereby mean sediment transport rates (and other statistical moments) depend on the averaging timescale \citep{Turowski2010,Campagnol2012,Ancey2020a}.
 A fundamental question raised by scale-dependence is how large averaging windows must be for sediment flux measurements to converge.
@@
 
-\subsection{Birth death models for bedload flucuations}
+\subsection{Birth-death models for bedload fluctuations}
 \label{sec:birthdeath}
@@
 To obtain realistically-wide fluctuations in particle activity, Ancey et al.
-introduced a positive feedback they called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
+introduced a positive feedback they called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the control volume \citep{Ancey2008,Heyman2013}.
 
-Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where arrivals of moving particles to the volume are ``births" and departures are ``death".
-They formulated the probability distribution $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
+Their governing equations are completely analogous to a stochastic population model \citep{Cox1965, Pielou1977} where arrivals of moving particles to the volume are ``births" and departures are ``deaths".
+Ancey et al. formulated the probability distribution $P(n,t)$ of the number of moving grains in the control volume at time $t$ as
 \begin{multline} \pt P(n,t) = \big[\lambda + (n-1)\mu\big] P(n-1,t) \\ - [n+1]\alpha P(n+1,t) - \big[\lambda + n (\alpha + \mu)\big]P(n,t).\end{multline}
-The terms on the right hand side of this equation describe respectively particle birth (entrainment and migration in) at rate $\lambda$, collective entrainment at rate $\mu$, and particle death (deposition and migration out) at rate $\alpha$.
-The final term encodes the possibility that $n$ stays constant.
+The terms on the right hand side of this equation describe respectively particle entrainment and migration in with rate $\lambda$, collective entrainment with rate $\mu$, and deposition and migration out with rate $\alpha$.
+The final term encodes the possibility that $n$ stays constant for a period of time.
 These coupled equations (one for each $n=0,1,\dots$) can be solved by generating functions \citep{Cox1965}, providing the steady-state distribution
@@
 \be \sigma_n^2 = \frac{\lambda \alpha}{(\alpha-\mu)^2}. \ee
-Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n^2/\bra n \ket = \alpha/(\alpha-\mu)$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to $\sigma_n^2/\bra n\ket = 1$, as the distribution Eq.
+Owing to the collective entrainment process, particle activity fluctuations can be arbitrarily wide: $\sigma_n^2/\bra n \ket = \alpha/(\alpha-\mu)$, whereas in the absence of collective entrainment ($\mu=0$), the strength of fluctuations is always pinned to a fixed value, $\sigma_n^2/\bra n\ket = 1$, as the distribution Eq.
 \ref{eq:negbin} limits to a Poisson distribution where the mean and variance are equal \citep[e.g.][]{Ancey2006}.
 
-The probability distribution of the bedload flux can be computed by the control volume form Eq.
-\ref{eq:controlflux} given the probability distribution of particle velocites $P(u)$ under the assumption that velocity and activity are independent.
-Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep{Ancey2020a}
+Using the probability distribution of the number of moving particles Eq.
+\ref{eq:controlflux}, the probability distribution of the bedload flux can be computed with the control volume definition assuming that the particle velocities and the number of moving particles are independent.
+Writing $P_k(u)$ as the probability distribution for the sum of $k$ independent particle velocities ($u=u_1+\dots+u_k$), the sediment flux probability distribution $P(\Phi)$ becomes \citep{Ancey2020a}
 \be P(\Phi) = L \sum_{k=0}^\infty P_k(L\Phi) P(n=k). \ee
@@
 
-\subsection{Renewal theories for scale dependence}
+\subsection{Renewal models for scale dependence}
 \label{sec:renewal}
@@
 
-The sediment flux distribution obtained from these approaches shifts depending on the observation time $T$ over which crossing events are observed, so approaches based on inter-arrival times explicitly model scale dependence.
+The sediment flux distribution obtained from these approaches shifts depending on the observation time $T$ over which particle arrivals are counted, so approaches based on inter-arrival times explicitly model scale dependence.
 In statistics, the problem of counting probabilistic arrival events is called renewal theory \citep{Cox1962}.
@@
 Note that different realizations of the flux converge toward the same value at large observation times, whereas at small observation times, uncertainty in the flux is large.
-This is observation-scale dependence.
+This is scale dependence.
 The properties of this convergence depend on the particle dynamics in a manner which has not yet been specified.
@@
 The exponential inter-arrival time distribution is $P(t) = \Lambda \exp(-\Lambda t),$ so the rate of particle arrivals is $\Lambda$.
-In this case, the flux averaged over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model reviewed in Sec. \ref{sec:einwalk}:
+In this case, the flux observed over a period $T$ can be represented with a Poisson pulse noise, similar to the particle position in the Einstein model reviewed in Sec. \ref{sec:einwalk}:
 \be q(T) = \frac{1}{T}\int_0^T dt' \mu(t'). \label{eq:flucflux}\ee
@@
 \be \mu(t) = \sum_{i=1}^{N(t)}\delta(t-t_i), \label{eq:poissfluxnoise}\ee
-as indicated in Fig. \ref{fig:ancey} (a). In this equation, $N(t)$ is Poisson distributed with rate $\Lambda$ -- $P(N)=(\Lambda t)^N/N! \exp(-\Lambda t).$
+as indicated in Fig. \ref{fig:ancey} panel (a). 
+In this equation, $N(t)$ is distributed as $P(N)=(\Lambda t)^N/N! \exp(-\Lambda t).$
 The flux in Eq.
 \ref{eq:flucflux} is a random variable as indicated in Fig.
-\ref{fig:ancey} (b).
+\ref{fig:ancey} panel (b).
 Its probability distribution $P(q|T)$, which is contingent on the observation time $T$, can be derived by evaluating $P(q|T) = \bra \delta(q-\int_0^T \mu(t') dt'/T) \ket$.
@@
 
-The mean flux derived from this renewal scheme -- $\langle q (T)\rangle = \int_0^\infty dq P(q|T)$ -- is
-\be \langle q (T) \rangle = \Lambda, \ee
+The mean flux $\langle q (T)\rangle = \int_0^\infty dq P(q|T)$ derived from this renewal scheme becomes
+\be \langle q(T) \rangle = \Lambda, \ee
 which is independent of the observation scale, while the magnitude of bedload transport fluctuations scales with $1/T$: 
-\be \sigma_q(T)^2 = \frac{\Lambda^2}{T}.\ee
+\be \sigma_q^2(T) = \frac{\Lambda^2}{T}.\ee
 Even this simple model gives a non-trivial conclusion that the relative uncertainty in a measurement of bedload transport depends on the observation time: $\sigma_q(T)/\langle q(T) \rangle \propto T^{-1/2}.$ Our uncertainty in a sediment transport measurement diverges as the observation time becomes short.
@@
 A limitation of renewal approaches to calculate the flux is that they do not obviously relate to the dynamics of individual particles.
-In renewal models as they are phrased now, the inter-arrival time distribution is not based on the underlying mechanics of particle transport.
+In renewal models as they are phrased now, the inter-arrival time distribution not based on the underlying mechanics of particle transport.
 
@@
 All of the works reviewed so far are either mean field models that do not include fluctuations, or stochastic models that calculate probability distributions for the quantities of interest.
-The common thread of all of these models is that they take advantage of various devices to side-step the complex interplay of turbulence and granular physics which ultimately controls bedload transport.
+The common thread of all of these models is that they take advantage of various devices to side step the complex interplay of turbulence and granular physics which ultimately controls bedload transport.
 
@@
 
-In reality, particle velocities fluctuate during movements due to turbulence and particle-bed collisions, so there is a need to go one step further and account also for velocity fluctuations of moving particles in Einstein-type models.
+In reality, particle velocities fluctuate during movements due to turbulence and particle-bed collisions, so there is a need to go one step further and account also for the velocity fluctuations of moving particles in Einstein-type models.
 
@@
 Chapter \ref{ch:flux} introduces a new model of individual bedload trajectories that includes velocity fluctuations in the motion state.
-The chapter also demonstrates how these trajectories can be used to calculate the scale-dependent probability distribution of the bedload flux.
-This joins together and extends studies from \citet{Einstein1937} to \citet{Ancey2020a} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
+The chapter demonstrates how these trajectories can be used to calculate the scale-dependent probability distribution of the bedload flux.
+This work joins together and extends studies from \citet{Einstein1937} to \citet{Ancey2020a} to develop new understanding of individual particle motions and the scale-dependent sediment flux.
 
@@
 Although this provides a useful description of bedload transport across local and intermediate timescales, it requires modification to describe longer global and geomorphic timescales.
-At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017,Bradley2017}, but burial has not yet been incorporated into Einstein-type models of particle trajectories.
+At longer timescales, particle burial moderates the downstream movements of grains \citep{Hassan2017,Bradley2017}, yet the burial process has not been incorporated into Einstein-type models of particle trajectories.
 
@@
 Finally, the Langevin models of \citet{Fan2014} and \citet{Ancey2014} describe the velocity distributions of sediment moving downstream in the local and intermediate ranges between entrainment and deposition.
-These models produce two different end-member sediment velocity distributions, but no mathematical models have been developed that describe the full range of distributions observed in experiments.
+These models produce two different end-member sediment velocity distributions, but no descriptions have yet been constructed which produce the full range of distributions observed in experiments.
 

### 2021-09-28 revision/ch2flipFlop.tex
done

@@
 
-\chapter{From particle dynamics to the sediment flux}
+\chapter{From particle dynamics to the scale-dependent flux}
 \label{ch:flux}
@@
 Bedload particles move downstream until they are disentrained when they happen to encounter sufficiently sheltered pockets on the bed surface to interrupt their motions \citep{Charru2004,Gordon1972}.
-Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, and particles are once again entrained.
+Eventually, the bed around them rearranges and destroys this shelter, or turbulent fluctuations overcome the shelter \citep{Celik2014,Valyrakis2010}, and resting particles are once again entrained.
 Bedload transport is thus a kind of intermittent motion, characterized by alternation between fluctuating movements and periods of rest.
 
-To date, descriptions of bedload transport have therefore simplified the problem in various ways to enable progress.
+To date, descriptions of bedload transport have simplified the problem in various ways to enable modeling progress.
 The foundational work is due to Einstein, who considered bedload motions as instantaneous so he could describe bedload transport as an alternating sequence of ``steps" and rests having random length and duration \citep{Einstein1937}, in a pioneering application of the continuous time random walk \citep{Montroll1965}.
@@
 Another set of works have described particles alternating between motion and rest considering that motions occur with a constant velocity \citep{Lisle1998,Lajeunesse2017}. 
-No models have yet been formulated which describes particles alternating between motion and rest with motions having realistic fluctuating velocities, even though this is how bedload particles actually move.
-A more realistic description of particle velocities during individual movements has potential to improve our understanding of bedform initation and particle-particle interactions in subsequent developments.
+No models have yet been formulated which describes particles alternating between motion and rest while motions have realistic fluctuating velocities, even though this is how bedload particles actually move.
 
-Parallel to these efforts to better describe individual particle trajectories, a complementary set of studies produced stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Turowski2010,Furbish2012a,Ancey2020}.
+Parallel to these efforts to better describe individual particle trajectories, a complementary set of studies have roduced stochastic formulations of the sediment flux which improve on the mean field description provided by Einstein \citep{Turowski2010,Furbish2012a,Ancey2020}.
 Sediment fluxes generally exhibit wide fluctuations due to variations in the number of moving particles and their velocities \citep{Bohm2005a,Ancey2006,Furbish2012a}, the migration of bedforms and sediment waves \citep{Guala2014,Recking2012}, and a host of other processes \citep{Dhont2018}.
 Because these fluctuations occur over disparate timescales, measurements of mean sediment fluxes depend on the timescale over which they are collected, a phenomenon called scale dependence \citep{Saletti2015,Dhont2018,Singh2009,Turowski2010,Ancey2020}.
-To date, very few models have calculated the probability distribution of the bedload sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and among these, even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
+To date, very few models have calculated the probability distribution of the bedload sediment flux due to individual particle motions \citep{Ancey2008,Ancey2014}, and even fewer have described any observation-scale dependence of the flux \citep{Ancey2020a,Turowski2010}.
 Among those that do include scale dependence, none formulate the sediment flux in terms of the trajectories of individual grains.
@@
 First, we lack the capability to describe individual sediment trajectories through motion and rest including velocity fluctuations in the motion state; and second, we need more understanding of how to connect individual particle trajectories through motion and rest to the overall streamwise sediment flux, including its probability distribution and the dependence of its statistical moments on the observation time.
+These two problems limit our understanding of the sediment flux and undermine our ability to phrase fluvial processes like bedform initiation \citep[e.g.][]{Ashley2021} and bedrock erosion \citep[e.g.][]{Li2020a} in terms of the dynamics of individual grains.
 
 Here, I develop a new statistical physics-based formalism which addresses both of these problems by describing individual particle trajectories with a stochastic Langevin equation.
-This dynamical equation describes alternation between motion and rest at random intervals while particles in motion have velocity that fluctuates around a mean value.
+This dynamical equation describes alternation between motion and rest at random intervals while particles in motion have velocities that fluctuate around a mean value.
 Using the probability distribution of particle position generated by this model, I construct a formalism to derive analytically the probability distribution of the sediment flux.
@@
 \ref{sec:mod}, solve it in Sec.
-\ref{sec:res}, and discuss the implications of my results and future research ideas in Secs.
+\ref{sec:res}, and discuss the implications of these results and future research ideas in Secs.
 \ref{sec:disc} and \ref{sec:conc}.
@@
 	\centerline{\includegraphics{./figures/ch2/fig0mod.pdf}}
-	\caption{Panel (a) sketches a realization of the noise in equation \ref{eq:flippylangevin}, while panel (b) shows the trajectory derived from it (in blue) alongside other possible trajectories.
-Keys in panel (a) demonstrate the average movement time $1/k_D$ and rest time $1/k_E$, while the key in panel (b) shows the average movement velocity $V$.
+	\caption{Panel (a) sketches a realization of the noise in Eq.
+\ref{eq:flippylangevin}, while panel (b) shows the trajectory derived from it (in blue) alongside other possible trajectories.
+Keys in panel (a) indicate the average movement time $1/k_D$ and rest time $1/k_E$, while the key in panel (b) shows the average movement velocity $V$.
 Velocity fluctuations produce tilted stair-step trajectories with unsteady slopes in the $x-t$ plane.
@@
 First, particles should alternate between motion and rest.
-The entrainment transition rate from rest to motion occurs with probability per unit time (or rate) $k_E$, while the deposition transition from motion to rest occurs with rate $k_D$.
-Second, particles in motion should move with mean velocity $V$ as their instantaneous velocities jitter around this mean value due to turbulent drag and particle-bed collisions.
+Entrainment occurs with probability per unit time (or rate) $k_E$, while deposition occurs with rate $k_D$.
+Second, velocities of moving particles should jitter around a mean value $V$ due to turbulent drag and particle-bed collisions.
 
@@
 \be \dot{x}(t) = [V + \sqrt{2D}\xi(t)]\eta(t).  \label{eq:flippylangevin} \ee
-Here,  $\xi(t)$ is a Gaussian white noise having zero mean and unit variance representing velocity fluctuations among moving particles.
-$\eta(t)$ is a dichotomous noise which takes on values $\eta = 1$, representing motion (with mean duration $1/k_D$), and $\eta=0$, representing rest (with mean duration $1/k_E$).
-The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$. The notation $k=k_E+k_D$ is a shorthand used throughout the thesis.
+Here, $V$ is the mean particle velocity while $D$ is a diffusivity [units $L^2/T^3$] representing the magnitude of velocity fluctuations.
+$\xi(t)$ is a Gaussian white noise having zero mean and unit variance which produces the velocity fluctuations.
+$\eta(t)$ is a dichotomous noise generating the motion-rest alternation.
+This noise takes on values $\eta = 1$, representing motion, and $\eta=0$, representing rest.
+The transition rate from $\eta=0$ to $\eta = 1$ is $k_E$, and the transition rate from $\eta=1$ to $\eta= 0$ is $k_D$.
 Times spent in motion and rest are respectively distributed as $P(t) = k_D \exp(k_D t)$ and $P(t) = k_E \exp(k_E t)$.
-$V$ is the mean particle velocity describing the overall downstream drift of moving particles, while $D$ is a diffusivity [units $L^2/T^3$] describing velocity fluctuations among moving particles.
+The notation $k=k_E+k_D$ is a shorthand used throughout this chapter.
 
 Equation \ref{eq:flippylangevin} describes the downstream movement of particles alternating through motion and rest with velocity fluctuations in the motion state.
-This model can be understood as a generalization of the ``randomly flashing diffusion" models developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}.
+This model can be viewed as a generalization of the ``flashing diffusion" models developed by Luczka and coworkers in physics \citep{Luczka1992,Luczka1993,Luczka1995}.
 Some trajectories produced by the Langevin equation \ref{eq:flippylangevin} are sketched in Fig. \ref{fig:fluxxy0}.
-The driving term in panel (a) displays excessive fluctuations due to the assumption that velocities evolve as Gaussian white noise with no temporal correlations, but the integrated particle trajectories in panel (b) are well-behaved and visually similar to those of earlier studies \citep[cf.][]{Fan2016,Bialik2015}.
-
+The driving term in panel (a) displays erratic fluctuations due to the assumption that velocities evolve as Gaussian white noise with no temporal correlations, but the particle trajectories in panel (b) which integrate these fluctuations are well-behaved and visually similar to those of earlier studies \citep[cf.][]{Fan2016,Bialik2015}.
 Eq.
-\ref{eq:flippylangevin} generalizes the constant velocity model of \citet{Lisle1998} and \citet{Lajeunesse2017} that was presented in section \ref{sec:lisle}, providing the first extension of these works aimed at a refined description of bedload particle trajectories through motion and rest.
+\ref{eq:flippylangevin} generalizes the constant velocity model of \citet{Lisle1998} and \citet{Lajeunesse2017} that was presented in Sec.
+\ref{sec:lisle}, providing the first extension of these works aimed at a refined description of bedload particle trajectories through motion and rest.
 
@@
 \label{sec:floppymastereq}
-The solution of equation \ref{eq:flippylangevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives one possible trajectory of a particle.
+The solution of Eq.
+\ref{eq:flippylangevin} for a given realization of the two noises $\eta(t)$ and $\xi(t)$ gives one possible trajectory of a particle.
 The probability distribution $P(x,t)$ that a particle which started at position $x=0$ at time $t=0$ has traveled to position $x$ by time $t$ can be formulated as an average over the ensemble of all possible trajectories.
@@
 One can see in particular that taking the entrainment rate $k_E$ very large, meaning that particles are very often moving, implies a classical advection-diffusion equation $(\pt + V\px -D \px^2)P=0$ for the position, characteristic of a particle moving downstream with Gaussian velocity fluctuations.
-Otherwise, there is a possibility that the particle is at rest and the advection-diffusion process is interrupted, giving rise to the additional terms in Eq.
+Otherwise, there is a possibility that the particle comes to rest and the advection-diffusion process is interrupted, giving rise to the additional terms in Eq.
 \ref{eq:flippymaster} and providing an asymmetrical structure reminiscent of Eq.
-\ref{eq:lislemaster} from the simpler motion-rest models of Sec.
-\ref{sec:lisle}.
+\ref{eq:lislemaster} from the earlier motion-rest model.
 
@@
 The probability distribution of the sediment flux can be calculated using the probability distribution of particle position $P(x,t)$ derived as the solution of Eq. \ref{eq:flippymaster}.
-This method is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
-This generalizes the nonlocal formulation of Sec.
+The method I present here is modified from the approach recently developed by Banerjee and coworkers in physics \citep{Banerjee2020}.
+This approach generalizes the nonlocal formulation of Sec.
 \ref{sec:nonlocal} to provide the probability distribution of the flux, and it re-frames the renewal approach of Sec.
@@
 
-The basic idea, as depicted in Fig.
-\ref{fig:flipflopfig1}, is to initially distribute $N$ particles in all states of motion along a domain of length $L$ at some random initial locations $x_i$ to the left of $x=0$.
+The basic idea, as depicted in Fig. \ref{fig:flipflopfig1}, is to initially distribute $N$ particles in all states of motion along the domain $-L\leq x\leq 0$.
 Later, the number of particles $N$ and the size $L$ of the domain will be extended to infinity such that their ratio $\rho=N/L$ remains constant.
-This limit will provide a configuration similar to the one constructed in the nonlocal formation (Sec. \ref{sec:nonlocal}).
+This limit will provide a configuration similar to the one constructed in the nonlocal formation of Sec. \ref{sec:nonlocal}.
 
 From this initial configuration, particles move downstream through time according to Eq. \ref{eq:flippylangevin}.
-The flux is calculated as the average rate of particles crossing to the right of the control surface (at $x=0$) after the sampling time $T$.
+The flux is calculated as the average rate of particles crossing to the right of the control surface at $x=0$ after the sampling time $T$.
 \begin{figure}
@@
 	\caption{The left panel indicates the configuration for the flux.
-The particle trajectories within demonstrate alternation between rest and motion with fluctuating velocity.
-Particles begin their transport with positions $-L\leq x \leq 0$ at $t=0$, and as depicted in the right panel, the flux is calculated with the number of particles $N_>(T)$ which lie to the right of $x=0$ at the observation time $t=T$, divided by $T$: $q(T) = N_>(T)/T$.
-The probability distribution of $q(T)$ is determined from all possible realizations of the trajectories and initial positions as $N,L$ tend to infinity while the density of particles $\rho=N/L$ to the left of the control surface remains fixed.}
+Here, the observation time increases from the bottom to the top.
+The particle trajectories within demonstrate alternation between rest and motion with fluctuating velocities.
+Particles begin their transport with positions $-L\leq x \leq 0$ at $T=0$, and as depicted in the right panel, the flux is calculated with the number of particles $N_>(T)$ which lie to the right of $x=0$ at the observation time $T$, divided by $T$: $q(T) = N_>(T)/T$.
+The probability distribution of $q(T)$ is determined from all possible realizations of the trajectories and initial positions as $N$ and $L$ tend to infinity while the density of particles $\rho=N/L$ to the left of the control surface is held constant.}
 	\label{fig:flipflopfig1}
@@
 Taking the Laplace transform over $q$ (i.e. forming the characteristic function of $P(q|T)$) obtains
-\begin{align} \tilde{P}(s|T) &=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N I_i(T)\Big)\Big\ket \\
+\begin{align} 
+\begin{split}
+\tilde{P}(s|T) &=  \Big\bra \exp\Big(\frac{s}{T}\sum_{i=1}^N I_i(T)\Big)\Big\ket \\
 	&=  \prod_{i=1}^N \Big\bra \exp\Big(-\frac{s}{T}I_i(T)\Big)\Big\ket \\
-	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra I_i(T) \big\ket \Big] \label{eq:flippycharacteristic}\end{align}
-This progression relies on the independence of averages for each particle (so the average of a product is the product of averages) and the observation that  $e^{a I} = 1-(1-e^a)I$ because $I$ is either $0$ or $1$.
+	&= \prod_{i=1}^N \Big[1-\big(1-e^{-s/T}\big)\big\bra I_i(T) \big\ket \Big].
+\end{split}
+\label{eq:flippycharacteristic}
+\end{align}
+This progression relies on the independence of averages for each particle (so that the average of a product is the product of averages) and the observation that  $e^{a I} = 1-(1-e^a)I$ because $I$ is either $0$ or $1$.
 
 The average over initial conditions and possible trajectories for the $i$th particle involved in this characteristic function can be written
-\be \bra I_i(t) \ket = \frac{1}{L}\int_{-L}^0 dx' \int_0^\infty dx P(x-x',t) \label{eq:indicator} \ee
+\be \bra I_i(t) \ket = \frac{1}{L}\int_{-L}^0 dx' \int_0^\infty dx P(x-x',t), \label{eq:indicator} \ee
 where $P(x,t)$ is the position probability distribution which solves Eq. \ref{eq:flippymaster}.
-Eq. \ref{eq:indicator} describes the probability that the $ith$ particle is found to the right of control surface by time $T$ provided it started at a random location somewhere to the left.
+Eq. \ref{eq:indicator} describes the probability that the $i$th particle is found to the right of control surface by time $T$ provided it started at a random location somewhere to the left.
 These $\bra I_i(t) \ket$ are the components of the flux that depend on the particle dynamics.
@@
 \ref{eq:indicator} into Eq.
-\ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, producing the configuration of the nonlocal formulation in section \ref{sec:nonlocal}, provides
-\begin{align} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T) \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T) \Big].
-\label{eq:flippychar} \end{align}
-where $\Lambda(T) = \rho \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate \textit{function}, similar to the rate constant in the renewal process models of section \ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
+\ref{eq:flippycharacteristic} and taking the limits $L\rightarrow \infty$ and $N \rightarrow \infty$ as the density of particles $\rho = N/L$ remains constant, producing the configuration of the nonlocal formulation in Sec.
+\ref{sec:nonlocal}, provides
+\begin{align}
+\begin{split} \tilde{P}(s|T) &= \lim_{N \rightarrow \infty} \Big(1 - \frac{1}{N}\big(1-e^{-s/T}\big)\Lambda(T) \Big)^N \\ &= \exp \Big[ -\big(1-e^{-s/T}\big)\Lambda(T) \Big]. 
+\end{split}
+\label{eq:flippychar} 
+\end{align}
+where $\Lambda(T) = \rho \int_0^\infty dx \int_0^\infty dx' P(x+x',T)$ is a rate \textit{function}, similar to the rate \textit{constant} in the renewal process models of Sec.
+\ref{sec:renewal}, except that it is now formulated in terms of individual particle trajectories via the probability distribution of particle position $P(x,t)$, which itself originates from the Langevin equation \ref{eq:flippylangevin} governing the particle dynamics.
 
 Eq. \ref{eq:flippychar} is the characteristic function of a Poisson distribution \citep{Cox1965}.
-Expanding in $e^{-s/T}$ and inverting the Laplace transform provides a key equation of this chapter, the probability distribution of the flux held contingent on the sampling time $T$:
+Expanding in $e^{-s/T}$ and inverting the Laplace transforms provides a key equation of this chapter, the probability distribution of the flux held contingent on the sampling time $T$:
 \be P(q|T) = \sum_{l=0}^\infty \frac{\Lambda(T)^l}{l!}e^{-\Lambda(T)}\delta(q-\frac{l}{T}). \label{eq:flippydist}\ee
@@
 Similarly the variance is $\sigma_q^2(T) = \Lambda(T)/T^2$.
-For the case when $\Lambda(T)$ is proportional to the observation time ($\Lambda \propto T$), these formulas become identical to the renewal theory approach reviewed in section \ref{sec:renewal}.
+For the case when $\Lambda(T)$ is proportional to the observation time ($\Lambda \propto T$), these formulas become identical to the renewal theory approach reviewed in Sec.
+\ref{sec:renewal}.
 This correspondence demonstrates that the renewal approach can be formulated equivalently by considering the dynamics of individual particles as a starting point.
@@
 	\centerline{\includegraphics{figures/ch2/figure2_slopeKey.pdf}}
-	\caption{Panel (a) indicates the probability distribution of particle position (Eq.
-\ref{eq:monster}) as it evolves through time.
+	\caption{Panel (a) indicates the probability distribution of particle position \ref{eq:monster} as it evolves through time.
 From the initial mixture of motion and rest states, particles advect downstream as they diffuse apart from one another due to differences in their velocities and transition times between motion and rest.
-In panel (a), the initial position persists as a Delta-function spike for the black $t=10\tau$ curve.
+In panel (a), the initial position persists as a delta function spike for the black $t=10\tau$ curve.
 Panel (b) shows the resulting particle diffusion (Eq.
@@
 For larger timescales, $2D/V^2 \ll t \ll \tau$, particles undergo ballistic diffusion similar to \citet{Lisle1998} as a result of some particles being stationary as others advect.
-Finally at times longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal, formed of particles well-mixed among motion and rest states.
-All results are scaled by the mean hop length $\ell=V/k_D$ and the timescale $\tau=1/k$ of the motion/rest alternation.
-In both plots, the lines are analytical results while the points are the results of Monte Carlo simulations based on evaluating cumulative transition probabilities on a small time-step \citep[e.g.][]{Barik2006}.}
+Finally at times longer than the timescale $\tau = 1/k$ associated with entrainment and deposition, diffusion is again normal since particles are well-mixed among motion and rest states.
+All results are scaled by the mean hop length $\ell=V/k_D$ and the timescale $\tau=1/k$ of the motion-rest alternation.
+In both plots, the lines are analytical results while the points are the results of Monte Carlo simulations based on evaluating cumulative transition probabilities on a small timestep \citep[e.g.][]{Barik2006}.}
 	\label{fig:flippyfig1}
@@
 The master equation \ref{eq:flippymaster} describes the evolution of the probability distribution of position through time.
-The solution of this equation should be some combination of the \citet{Einstein1937} theory for particle transport (section \ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep[e.g.][]{Morse1953}.
+The solution of this equation should be some combination of the \citet{Einstein1937} model for particle transport (Sec.
+\ref{sec:einwalk}) with the Gaussian solution of the advection-diffusion equation \citep[e.g.][]{Morse1953}.
 
@@
 
-The master equation \ref{eq:flippymaster} is solved by transform calculus in section \ref{sec:fluccymastersol} of the appendix, providing
+The master equation \ref{eq:flippymaster} is solved by transform calculus in Sec. \ref{sec:fluccymastersol} of the appendix, providing
 \begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t) +  \pt \big]\\
@@
 where $\varphi=k_D/k$ is the probability the particle starts at rest and $\mathcal{I}_0$ is a modified Bessel function.
-This equation generalizes the earlier result  \ref{eq:lisledist} for alternation between motions and rests to include velocity fluctuations within the motion state.
+This equation generalizes the earlier result Eq.
+\ref{eq:lisledist} for alternation between motions and rests to include velocity fluctuations within the motion state.
 
@@
 \ref{eq:monster} could be derived by integrating it directly, but this is difficult.
-Instead, the moments can be calculated directly from the master equation \ref{eq:flippymaster} \citep[e.g.][]{Cox1965}.
+Instead, the moments can be calculated from the master equation \ref{eq:flippymaster} \citep[e.g.][]{Cox1965}.
 
@@
 
-Provided that $2D/V^2 \ll k_D/k^2$, series expansion of \ref{eq:minor} in $t$ and $1/t$ reveals three different scaling behaviors:
+Provided that $2D/V^2 \ll k_D/k^2$, series expansion of Eq. \ref{eq:minor} in $t$ and $1/t$ reveals three different scaling behaviors:
 \be \sigma_x^2 \sim 
@@
 	 t, & t \ll \frac{2Dk}{V^2 k_D}, \\ 
-	 t^2, &  \frac{2Dk}{V^2 k_D} \ll t \ll \frac{1}{k}, \\
-	 t, & t\gg \frac{1}{k}. \label{eq:diffo}
+	 t^2, &  \frac{2Dk}{V^2 k_D} \ll t \ll \frac{1}{k}. \\
+	 t, & t\gg \frac{1}{k} \label{eq:diffo}
 \end{cases} \ee
-Note in the physical condition when $k\approx k_D$, which is generally satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the above condition for existence of three ranges becomes $ \Pe \gg 1,$ where
-$\Pe =V^2/(2k_D D)$ is a P\'{e}clet number \citep[e.g.][]{Heyman2014}. This P\'{e}clet number measures the relative importance of advection and diffusion to the downstream movements of particles.
+In the physical condition when $k\approx k_D$, which is generally satisfied for bedload transport since motions are typically short compared to rests \citep{Hassan1991,Wu2019}, the above condition for existence of three ranges becomes $ \Pe \gg 1,$ where
+$\Pe =V^2/(2k_D D)$ is a P\'{e}clet number \citep[e.g.][]{Heyman2014}. This dimensionless number measures the relative importance of advection and diffusion to the downstream movements of particles.
 
@@
 Below, this anomalous diffusion behavior \citep[cf.][]{Sokolov2012} is shown to affect the rate constant $\Lambda(t)$ in the sediment flux.
-As a result of these multi-scale particle dynamics, the sediment flux becomes scale-dependent in a richer way than the renewal model of section \ref{sec:renewal}.
+As a result of these multi-scale particle dynamics, the sediment flux becomes scale-dependent in a richer way than the renewal model of Sec.
+\ref{sec:renewal}.
 
@@
 
-Using the probability distribution of position Eq. \ref{eq:monster}, the integrals in Eq. \ref{eq:flatform} are performed in appendix Sec. \ref{sec:fluxconstant}u, providing the rate function
+Using the probability distribution of position Eq. \ref{eq:monster}, the integrals in Eq. \ref{eq:flatform} are performed in appendix Sec. \ref{sec:fluxconstant}, providing the rate function
 \begin{multline} 
@@
 \ref{fig:fluxconvergence}, as a result of Eq.
-\ref{eq:fluxy} the mean sediment flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $q_0 = \rho k_E V/k = E \ell$ as the observation time becomes much larger than $1/(k_D \Pe)$.
+\ref{eq:fluxy} the mean sediment flux takes on a non-trivial scale-dependence, characterized by a decay toward the Einstein prediction $q_0 = \rho k_E V/k \approx E \ell$ as the observation time becomes much larger than $1/(k_D \Pe)$.
 \begin{figure}[!htbp]
 	\includegraphics[width=\linewidth,keepaspectratio]{figures/ch2/figure3_slopeKey.pdf}
-	\caption{The mean sediment flux is plotted for different values of the P\'{e}clet number $\Pe = V^2/(2k_D D)$, characterizing the relative strength of particle velocity fluctuations during motion.
-The flux is normalized by the prediction $q_0 = E \ell$ of the Einstein theory (Sec.
+	\caption{The mean sediment flux is plotted for different values of the P\'{e}clet number $\Pe = V^2/(2k_D D)$, characterizing the relative significance of particle velocity fluctuations during motion.
+The flux is normalized by the prediction $q_0 = E \ell$ of the Einstein model (Sec.
 \ref{sec:einflux}).
-For nonzero velocity fluctuations in the motion state (finite $\Pe$), the Einstein limit is approached as the observation time $T$ grows.
+For nonzero velocity fluctuations in the motion state (moderate $\Pe$), the Einstein limit is approached as the observation time $T$ grows.
 Stronger velocity fluctuations (smaller $\Pe$) slow the convergence to this limit.
-	In all cases, satisfactory convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by Eq.
+	In all cases, convergence of the mean flux is achieved when the observation time satisfies $T\gg 1/(k_D \Pe),$ as expected by Eq.
 \ref{eq:fluccolimitcases}.
@@
 \ref{eq:fluxy}, while the points are the results of Monte Carlo simulations.
-Vertical dotted lines indicate the convergence time $T = (k_D \Pe)^{-1}$ for each condition.
+Vertical dotted lines indicate the convergence time $T = (k_D \Pe)^{-1}$ for each simulation.
 }
@@
 It is related to the P\'{e}clet number and is proportional to the time particles spend in motion ($1/k_D$).
-As a result, when particle velocity fluctuations during motions become large comparable to the mean particle velocity, Eq.
+As a result, when particle velocity fluctuations during motions become appreciable in comparision to the mean particle velocity, Eq.
 \ref{eq:fluccolimitcases} indicates that the flux becomes slow to converge to the Einstein result.
@@
 
-This chapter has generalized the earlier descriptions of individual sediment trajectories \citep[e.g.][]{Lisle1998,Lajeunesse2017} to include velocity fluctuations in the motion state.
+This chapter has generalized earlier descriptions of individual sediment trajectories \citep[e.g.][]{Lisle1998,Lajeunesse2017} to include velocity fluctuations in the motion state.
 
-Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches (Sec.
-\ref{sec:renewal}) more directly in terms of the underlying particle dynamics \citep[e.g.][]{Turowski2010,Ancey2020}.
+Using results from this generalized model as an example, I demonstrated how to calculate the sediment flux probability distribution, phrasing earlier renewal theory approaches \citep[e.g.][]{Turowski2010,Ancey2020} more directly in terms of the underlying particle dynamics.
 This method can also be viewed as a generalization of the nonlocal formulation (Sec. \ref{sec:nonlocal}) to describe sediment flux probability distributions, rather than just mean values.
@@
 
-The sediment flux probability distribution in Eq. \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
-Poisson distributions have relatively thin tails, meaning their fluctuations are typically small \citep{Ancey2006}.
+The sediment flux probability distribution \ref{eq:flippydist} represents a Poisson distribution with an observation scale-dependent rate.
+Poisson distributions have relatively narrow tails, meaning their fluctuations are typically small \citep{Ancey2006}.
 In reality, sediment flux distributions are only Poissonian at high transport rates, whereas in other conditions they have wide tails representing the possibility of extremely large transport fluctuations \citep{Ancey2008,Turowski2010,Saletti2015} which appear as bursts \citep[e.g.][]{Goh2008} in the sediment flux timeseries \citep{Singh2009, Dhont2018,Benavides2021}.
-This highlights a need to generalize the mechanistic theory of the sediment flux I developed here to produce wider transport rate fluctuations.
+This highlights a need to generalize the mechanistic description of the sediment flux I developed here to produce wider transport rate fluctuations.
 
@@
 Given the prevalence of large coherent structures, gravel beds must often times be conditioned to them.
-Therefore particle-particle interactions seem more plausible a mechanism for collective entrainment.
+Therefore particle-particle interactions seem more plausible as a mechanism for collective entrainment.
 In particular, contact forces within granular media are known to organize into ``force chains", where a minority of contacts provide a majority of the strength \citep[e.g.][]{Radjai1996, Azema2012}.
 Fluid entrainment of a key particle within a near-surface force chain could entail a collective release of the particles relying on it for stability.
+Likewise, collisions between moving particles and stationary particles might induce entrainment of otherwise stable surface particles.
 
 Such particle-particle interactions would be challenging to include in a mechanistic model for the sediment flux like I developed here, but the basic structure can be sketched.
-The dynamical Eq.
-\ref{eq:flippylangevin} would generalize to a stochastic Newtonian form $\ddot{x}_i(t) = F_i(x_i,\dot{x}_i, t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving forces unique to each particle, while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
-As an approximation, this forcing structure might alternate between moving and resting modes by a dichotomous process \citep[e.g.][]{Bena2006}.
+Eq.
+\ref{eq:flippylangevin} would generalize to a stochastic Newtonian form $\ddot{x}_i(t) = F_i(x_i,\dot{x}_i, t) + \sum_{i\neq j}G_{ij}(\{x_j\},t)$, where the $F_i$ are the driving forces unique to each particle while the $G_{ij}$ are some (generally stochastic) terms representing interactions between the $i$th and $j$th particles \citep{Goldstein1997}.
+As an approximation, this forcing structure might alternate between moving and resting modes as a dichotomous process \citep[e.g.][]{Bena2006}.
 
-The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction diffusion systems \citep{Pechenik1999, Cardy2008}, granular gases \citep{Brilliantov2004}, or other interacting particle systems available in physics literature \citep{Hernandez2004,Escaff2018}.
+The resulting joint distribution of particle positions and velocities -- $P(x_1,v_1,\dots,x_{N(t)},v_{N(t)},t)$ -- might be formulated by analogy to the theory of reaction-diffusion systems \citep{Pechenik1999, Cardy2008}, granular gases \citep{Brilliantov2004}, or other interacting particle systems available in physics literature \citep{Hernandez2004,Escaff2018}.
 
@@
 Nikora et al. originally proposed that sediment diffusion would be super-diffusive at local timescales, normal at intermediate timescales, and sub-diffusive at global timescales.
-Here, the local range diffusion is normal, not super-diffusive, indicating a discrepancy between the Nikora et al. concept and the present model.
+Here, the local range diffusion is normal, not super-diffusive, indicating a discrepancy between the Nikora et al. concept and the model of this chapter.
 
@@
 \ref{eq:flippylangevin} to include velocity correlations would be to replace the Gaussian white noise with an Ornstein-Uhlenbeck noise \citep[e.g.][]{Luczka2005,Hanggi2007}.
-A more challenging (but perhaps more realistic) modification would be to describe the evolution of position by a second-order equation, as in $\ddot{x} = F(x,t)\eta(t)$, where $F$ is an appropriate stochastic driving term switched on and off by the dichotomous noise $\eta(t)$ \citep[e.g.][]{Masoliver1993}.
+A more challenging (but perhaps more realistic) modification would be to describe the evolution of position by a second-order equation, as in $\ddot{x} = F(x,\dot{x},t)\eta(t)$, where $F$ is an appropriate stochastic driving term switched on and off by the dichotomous noise $\eta(t)$ \citep[e.g.][]{Masoliver1993}.
 Either of these alternatives would introduce correlations to particle velocities and modify the resulting short-range diffusion characteristics, unfortunately at the cost of mathematical difficult.
 
-The short timescale diffusion may also be affected by the assumption that entrainment and deposition processes are instantaneous within the model.
+The short-timescale diffusion may also be affected by the assumption that entrainment and deposition processes are instantaneous within dichotomous noise models.
 \citet{Campagnol2015} showed from experimental data that particle ``unsteadiness" during entrainment and deposition can give rise to diffusion exponents even larger than those for ballistic motion.
 Likely, the representation of entrainment and deposition with an instantaneous alternation obscures the local-timescale physics of these processes \citep[e.g.][]{Valyrakis2010,Celik2014}.
-The simplest modification to include the timescales of entrainment and deposition would introduce a uniform acceleration (deceleration) during entrainment (deposition), in effect ``tilting" the sharp motion-rest alternation of dichotomous noise.
-The relatively lower velocity of particles during this acceleration phase should modify the predicted local-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
+The simplest modification to include the timescales of entrainment and deposition would introduce a uniform acceleration during entrainment and deposition, in effect ``tilting" the sharp motion-rest alternation of the dichotomous noise.
+The relatively lower velocities of particles during the acceleration phase should modify the predicted local-timescale particle diffusion characteristics, possibly producing a faster-than-ballistic spreading as proposed by \citet{Campagnol2015}.
 
-\subsection{Scale dependent fluxes and channel evolution}
+\subsection{Scale-dependent fluxes and channel evolution}
 
-Fig. \ref{fig:fluxconvergence} indicates that the sediment flux described by Eq. \ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q = E\ell$ predicted by the Einstein theory.
-Because particles move for durations $1/k_D$ on average, the typical variance of velocity among moving particles can be written $\sigma_V^2 = 2D/k_D$.
+Fig.
+\ref{fig:fluxconvergence} indicates that the sediment flux described by Eq.
+\ref{eq:fluxy} converges when $T\gg (k_D\ Pe)^{-1}$ to its eventual value $q_0 = E\ell$ predicted by the Einstein model.
+Because particles move for durations $1/k_D$ on average, the typical variance of velocity among moving particles can be estimated as $\sigma_V^2 = 2D/k_D$.
 In terms of this quantity, the convergence condition for the flux is $T\gg \sigma_V^2/(k_D V^2),$ now phrased in terms of easily measurable quantities -- the velocity of particles, the average time spent in motion, and the magnitude of velocity fluctuations.
@@
 
-Presumably, the inclusion of a scale-dependent stochastic transport rate generates a stochastic bed elevation field $h(x)$, generating a probability distribution of field configurations $P(h(x),t)$ which depends on the observation scale.
-\citet{Jerolmack2005} include a stochastic sediment transport rate within the Exner equation, although they only investigate particular realizations of $h(x)$, not the probability distribution of bed configurations or its scale dependence.
-Mathematical tools to investigate such a stochastic Exner equation can be found in works on surface growth processes and polymer physics, where the evolution of random fields is common topic of inquiry \citep[e.g.][]{Kawakatsu2001,Barabasi1995, Kardar2007}.
+Presumably, the inclusion of a scale-dependent stochastic transport rate generates a stochastic bed elevation field $h(x)$, producing a probability distribution of field configurations $P(h(x),t)$ which depends on the observation scale.
+\citet{Jerolmack2005} included a stochastic sediment transport rate within the Exner equation, although they only investigated particular realizations of $h(x)$, not the probability distribution of bed configurations or its scale dependence.
+Mathematical tools to formulate such a stochastic Exner equation might be adopted from works on surface growth and polymer physics, where the evolution of random fields is common topic of inquiry \citep[e.g.][]{Kawakatsu2001,Barabasi1995, Kardar2007}.
 
@@
 
-Although the present model has analyzed independent particles, so it does not include these processes, we can wonder if better understanding fluctuations due to single-particle sources will aid understanding of more complex collective sources.
+Although the present model has analyzed independent particles and does not include these processes, we can wonder if better understanding fluctuations from single-particle sources will improve our understanding of more complex collective sources.
 An attempt to subtract the signature of individual particle dynamics from spectra of bedload transport may be a useful step to understand more complex morphological sources of sediment transport variability in channels.
@@
 
-This chapter further quantifies the dependence of the sediment flux on the observation scale and provides guidance as to the measurement times required to resolve sediment transport without ambiguity, at least in the idealized conditions in which the model was developed.
-The next steps are to include collective motions to produce wider sediment transport fluctuations, and to modify the description of velocity fluctuations to include temporal correlations.
+This chapter further quantifies the dependence of the sediment flux on the observation scale and provides guidance as to the measurement times required to resolve sediment transport without ambiguity, at least under the idealized assumptions for which the model was developed.
+The next steps are to include interactions between particles to produce wider sediment transport fluctuations and to modify the description of velocity fluctuations to include temporal correlations.

### 2021-09-28 revision/ch3bedEle.tex
done

@@
 Although the displacements of individual grains are a mechanical consequence of forces imparted from the flow, bed, and other grains \citep{Wiberg1985, Vowinckel2014,Gonzalez2017}, accurately characterizing these forces within natural channels is practically impossible, especially considering the intense variability these forces display \citep{Schmeeckle2007,Celik2010, Dwivedi2011}.
-In response, researchers have developed a stochastic concepts of bedload transport \citep{Einstein1937}, whereby the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
+In response, researchers have developed a stochastic concept of bedload transport \citep{Einstein1937}, wherein the erosion and deposition of individual grains are modeled as the random results of undetermined forces \citep{Einstein1950,Paintal1971,Ancey2006}.
 
@@
 These approaches aim at a complete statistical characterization of the bedload flux \citep{Furbish2012a,Fathel2015,Furbish2017,Heyman2016}, including probability distributions \citep{Ancey2006,Ancey2008}, spatial and temporal characteristics of its fluctuations \citep{Heyman2014a, Roseberry2012,Dhont2018}, and the dependence of these statistical characteristics on the length and timescales over which they are observed \citep{Singh2009,Singh2012,Saletti2015}.
-A recent surge in research activity has generated rapid progress in this subject. 
+A recent surge in research activity has generated rapid progress in this area. 
 For example, \citet{Ancey2006} demonstrated that a constant erosion rate as originally proposed by \citet{Einstein1950} was insufficient to develop realistically large particle activity fluctuations, so they added a positive feedback between the particle activity and erosion rate they called ``collective entrainment'' \citep{Ancey2008, Heyman2013,Heyman2014, Lee2018}.
@@
 This new model evaluates the effect of collective entrainment on bed elevation change, and it allows for investigations of the residence times of particles buried under the bed surface.
-The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilities per unit time, or rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
+The model has two major assumptions: First, bedload erosion and deposition are characterized by probabilistic rates \citep{Einstein1950, Ancey2008}; and second, these rates are contingent on the local bed elevation, encoding the property that erosion of sediment is emphasized from regions of exposure, while deposition is emphasized in regions of shelter \citep{Sawai1987, Wong2007}.
 
@@
 \ref{sec:elemodel} introduces the stochastic model and Sec.
-\ref{sec:elesolution} solves it with a mixture of numerical and analytical techniques.
+\ref{sec:elesolution} presents solutions of the model from a mixture of numerical and analytical techniques.
 Model solutions illustrate several new features of particle activity and bed elevation statistics that result from feedbacks between the erosion and deposition rates with bed elevation change.
@@
 
-All particles are considered approximately spherical with the same radius $a$, so their mobility and packing characteristics are similar from one particle to the next.
+All particles are considered approximately spherical with the same radius $a$, so their mobility and packing characteristics are similar.
 Following \citet{Ancey2008}, basically four events can occur at any instant to modify the populations $n$ and $m$.
@@
 Migration, entrainment, and deposition are represented by arrows, and the instantaneous bed elevation is depicted by dotted lines.
-The bed is displayed in a degraded state, where $m<0$.
-The marginal distributions of $n$ and $m$ are indicated in the upper right, while the bottom panel is a realized timeseries of bed elevations computed from $m$ using Eq.
+The bed is displayed in a degraded state where $m<0$.
+The marginal distributions of $n$ and $m$ are indicated in the upper right panel, while the bottom panel is a realized timeseries of bed elevations computed from $m$ using Eq.
 \ref{eq:ele}.}
@@
 The populations $n$ and $m$ provide the volumetric bedload flux $\Phi$ and the local bed elevation $z$.
-The mean transport rate is given by $\Phi = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and $\langle n \rangle = 
+The mean transport rate is given by $\Phi = u_s\langle n \rangle/L$, where $u_s$ is the characteristic velocity of moving bedload and \newline $\langle n \rangle = 
 \sum_{n,m}nP(n,m) $ is the mean number of grains in motion \citep{Charru2004, Ancey2008, Furbish2012a}.
@@
 In Eqs. \ref{eq:rate2} and \ref{eq:rate3}, $\kappa$ is a coupling constant between bed elevations and the entrainment and deposition rates.
-Using this coupling constant, I will later demonstrate the relationship
+In terms of $\kappa$ I will later demonstrate the relationship
 \begin{equation}\kappa \approx \big(\frac{z_1}{2l}\big)^2 \label{eq:active}
@@
 where $l$ is a characteristic length scale of bed elevation change that can be interpreted as the active layer depth \citep{Wong2007,Church2017}.
-$\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
+In the rate equations \ref{eq:rate1}-\ref{eq:rate4}, $\nu$ is the rate of migration into the control volume, $\lambda$ is the conventional entrainment rate, $\mu$ is the collective entrainment rate, $\sigma$ is the deposition rate, and $\gamma$ is the rate of migration out of the control volume.
 At $m=0$, these equations reduce to those of the \citet{Ancey2008} model.
@@
 
-The master equation for the probability flow can be written using the Komogorov equation
+The master equation for the probability flow can be written using the Kolmogorov equation
  $\partial P(n,m;t)/\partial t = 
@@
 The solutions of Eq.
-\ref{eq:elemaster} should be expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
+\ref{eq:elemaster} are expected to adjust from whatever initial conditions toward a steady-state distribution $P_s(n,m)$ -- independent of time -- if the constant factors in the transition rates are representative of equilibrium transport conditions.
 Equilibrium requires $E=D$, meaning there is no net change in elevation, and $\nu = \gamma \langle n \rangle$, meaning mass is conserved in the control volume (inflow $=$ outflow).
-This master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well-known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
+This master equation describes a two-species stochastic birth-death model \citep{Cox1965} of a type well known in population ecology \citep{Pielou1977, Swift2002} and chemical physics \citep{Gardiner1983}.
 In context of this chapter, the two populations are the moving and stationary grains in the control volume.
@@
 then draw a second random value to choose the type of transition that occurs using relative probabilities formed from Eqs.
-\ref{eq:rate1} - \ref{eq:rate4}.
+\ref{eq:rate1}-\ref{eq:rate4}.
 The transition is enacted by shifting $t$, $n$, and $m$ by the appropriate values to the type of transition (that is, entrainment is $m\rightarrow m-1$ and $n \rightarrow n+1$, and so on).
@@
 \begin{wraptable}{r}{0.6\textwidth}
-	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ from \citet{Ancey2008}.
+	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ are tabulated from \citet{Ancey2008}.
 Units are $s^{-1}$ (probability/time).
@@
 \end{wraptable}
-For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, I used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in table \ref{tab:anceyparams}.
+For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, I used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in Tab.
+\ref{tab:anceyparams}.
 Flow conditions are labeled (a), (g), (i), and (l), roughly in order of increasing bedload flux (see \citet{Ancey2008} for more details). 
 In all simulations, I took the packing fraction $\phi = 0.6$ -- a typical value for a pile of spheres \citep[e.g.][]{Bennett1972}, and I set $L = 22.5$cm and $a = 0.3$cm, values from the original \citet{Ancey2008} experiments.
-Each simulation was run for $250$ hours of virtual time, a period selected to ensure convergence of particle activity and bed elevation statistics.
+Each simulation was run for $250$hrs of virtual time, a period selected to ensure convergence of particle activity and bed elevation statistics.
 
@@
 \ref{eq:elemaster} can be approximately decoupled using the inequality $l \gg z_1$ (equivalently $\kappa \ll 1$) which holds for large values of the active layer depth $l$.
-These inequalities imply that many entrainment or deposition events are required for an appreciable change in the entrainment or deposition rates.
+These inequalities imply that many individual entrainment or deposition events are required for an appreciable change in the entrainment and deposition rates.
 Concentrating on steady-state conditions $\partial P/\partial t = 0$, one can introduce the exact decomposition $P(n,m) = A(n|m)M(m)$ to Eq.
@@
 As explained in appendix Sec. \ref{sec:err}, comparison with the numerical simulations determines that
+the relationship
 \begin{equation}
@@
 Introducing this closure relation to Eq. \ref{eq:approxele}, making the Fokker-Planck approximation, and neglecting terms of $O(\kappa^2)$ provides 
-\begin{equation} M(m) \approx M_0 e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
+\begin{equation} M(m) \approx \sqrt{\frac{2\kappa}{\pi}} e^{-2\kappa m^2}, \label{eq:ou2}\end{equation}
 representing a Gaussian distribution with variance $\sigma_m^2 = \frac{1}{4\kappa}$ -- smaller than the former mean field theory by a factor of two and in agreement with the result of Eq.
 \ref{eq:active}.
-$M_0$ is a normalization constant. 
 This closure equation approach shows good correspondence with numerical solutions of Eq. \ref{eq:elemaster}, at least for the flow parameters in Tab. \ref{tab:anceyparams}.
@@
 The bottom panel of Fig.
-\ref{fig:eledefinition} shows an elevation timeseries.
+\ref{fig:eledefinition} shows an example elevation timeseries.
 In order to describe the implications of coupling bedload transport to bed elevation changes, the numerical and analytical results for the probability distributions of bedload transport and bed elevations are presented in Sec.
@@
 \end{figure}
-The mean field equation for the particle activity $n$ (Eq.
-\ref{eq:ancey}) closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match Eq.
+The mean field equation \ref{eq:ancey} for the particle activity $n$ closely represents the numerical results, and while there are small differences between numerical and analytical results for the relative number $m$ of resting particles, the numerical solutions approximately match Eq.
 \ref{eq:ou2}, having Gaussian profiles consistent with the assumption of a symmetric scaling between erosion and deposition rates and bed elevation changes.
@@
 	\includegraphics[width=0.5\textwidth,keepaspectratio]{./figures/ch3/variance.pdf}
-	\caption{Data from all simulations demonstrating that the active layer depth $l$ characterizes bed elevation changes as described by Eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$. }
+	\caption{Data from all simulations demonstrates that the active layer depth $l$ characterizes bed elevation changes as described by Eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$. }
 	\label{fig:var}
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/momentsuppression.pdf}
-	\caption{The shifts between particle activity moments conditioned on instantaneous elevations and their over-all mean values.
+	\caption{The particle activity moments conditioned on instantaneous bed elevations shift from their over-all mean values.
 Panel (a) indicates the mean particle activity shift versus the bed elevation measured in units of $\sigma_z=l$.
@@
 The closure equation \ref{eq:closure} is plotted in panel (a).
-Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying shifts of up to 20\% with bed elevations.
+Panel (b) demonstrates a more symmetrical variance shift with some dependence on flow conditions displaying variations of up to 20\% with bed elevation changes.
 These results indicate that bedload statistics measurements on short timescales could be severely biased by departures from the mean bed elevation.}
@@
 \end{figure}
-Figure \ref{fig:condmoms} panel (a) demonstrates that the flow conditions in Tab.
+Fig.
+\ref{fig:condmoms} panel (a) demonstrates that the flow conditions in Tab.
 \ref{tab:anceyparams} support departures of the mean particle activity by as much as 60\% from the overall mean value when the bed is in a degraded state $z\approx -3l$, and the activity can be decreased by 20\% when the bed is in an aggraded state.
-The closure model (Eq.
-\ref{eq:closure}) used to derive the approximate bed elevation distribution (Eq.
-\ref{eq:ou2}) is plotted behind the conditional mean profiles in Fig.
+The closure model \ref{eq:closure} used to derive the approximate bed elevation distribution \ref{eq:ou2} is plotted behind the conditional mean profiles in Fig.
 \ref{fig:condmoms} panel (a), where it appears as a crude approximation since it does not capture the asymmetry.
@@
 \ref{fig:condmoms} panel (b) displays a variance shift with bed elevation changes that is less severe than the mean shift but is nevertheless appreciable, with bed elevations changing the magnitude of bedload activity fluctuations by as much as 20\%.
-The summary is that bed elevation changes regulate the first and second particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
+These results indicate that bed elevation changes regulate the first and second particle activity moments, with a moment suppression effect when the bed is aggraded, and a moment enhancement effect when the bed is degraded.
 
@@
 \label{sec:elecolent}
-Noting that bed elevations regulate the particle activity moments, the next step is to study the influence of collective entrainment on this effect by modifying the relative proportion of the individual to collective contributions in the mean entrainment rate $E=\lambda + \sigma \langle n \rangle $.
+Noting that bed elevations affect the particle activity moments, the next step is to study the influence of collective entrainment on this effect by modifying the relative proportion of the individual to collective contributions in the mean entrainment rate $E=\lambda + \sigma \langle n \rangle $.
 The equilibrium condition $E=D$ implies that the fraction of entrainment due to the collective process is $f = \mu\langle n \rangle/E = \mu/\sigma$.
@@
 	\includegraphics[width=\linewidth,keepaspectratio]{./figures/ch3/colent-suppression.pdf}
-	\caption{The shift of the mean particle activity in panel (a) and its fluctuations in panel (b) with departures of the bed elevation from its mean.
-All simulations are at flow condition (g) from Tab.
-\ref{tab:anceyparams} except $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment.
-Clearly, collective entrainment drives strong departures of the bedload statistics away from the mean field model (Eq.
+	\caption{Panel (a) displays the shift of the mean particle activity with departures of the bed elevation from its mean value, while panel (b) displays the shift of the particle activity fluctuations.
+All simulations are conducted with flow condition (g) from Tab.
+3.1, except that $\lambda$ and $\mu$ are modified to shift the fraction $f=\mu/\sigma$ of the over-all entrainment rate $E$ due to collective entrainment.
+Collective entrainment drives strong departures of the bedload statistics away from the mean field model (Eq.
 \ref{eq:ancey}) at large departures from the mean bed elevation.
 Panel (b) shows particle activity fluctuations suppressed by 90\% when $z\approx -3l$ and collective entrainment is the dominant process.
-When collective entrainment is absent, meaning $\mu/\sigma=0$, this moment regulation effect vanishes: it is a consequence of collective entrainment.}
+When collective entrainment is absent, meaning $\mu/\sigma=0$, this moment regulation effect vanishes, as it is a consequence of collective entrainment.}
 	\label{fig:colent}
@@
 	\caption{Resting time statistics scale differently with transport conditions and the bed elevation variance.
-Panel (a) shows differing flow conditions at a fixed $l$ value, while panel (c) shows fixed flow conditions at differing $l$.
+Panel (a) shows different flow conditions at a fixed $l$ value, while panel (c) shows fixed flow conditions at different $l$.
 When scaled by $T_0$ (Eq.
-\ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d).
-In panels (b) and (d), the black dotted lines indicate a power law decay of the collapsed tails having parameter $\alpha\approx1.18$ .}
+\ref{eq:time}), both types of difference collapse in the tails of the distributions, as shown in panels (b) and (d), where the black dotted lines indicate a power law decay of the collapsed tails having parameter $\alpha\approx1.18$ .}
 	\label{fig:cdfs}
@@
 Comparing panels (a) and (c) in this figure shows two separate variations with input parameters: first, the distributions vary with the flow conditions, and second, they vary with the standard deviation of bed elevations ($l$).
-However, as shown in panels (b) and u(d), a characteristic timescale $T_0$ is found to collapse away both variations.
+However, as shown in panels (b) and (d), a characteristic timescale $T_0$ is found to collapse away both variations.
 This $T_0$ can be obtained heuristically by considering the characteristic speed of bed elevation change.
@@
 
-Like the theory developed by \citet{Martin2014}, the model Eq.
-\ref{eq:elemaster} produces heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
+Like the theory developed by \citet{Martin2014}, the model \ref{eq:elemaster} produces heavy-tailed power-law resting times for sediment undergoing burial by treating bed elevation changes as an unbounded random walk with a mean reverting tendency.
 This result suggests sediment burial can explain the heavy-tailed rests seen in field data \citep{Olinde2015,Bradley2017,Pretzlav2016a}. 

### 2021-09-28 revision/ch4downDiff.tex
done

@@
 originally considered.
-As a result, three diffusion ranges can be identified by patching together multiple data-sets \citep{Zhang2012,Nikora2002}, but they are not resolved by any one data-set.
+As a result, three diffusion ranges can be identified by patching together multiple datasets \citep{Zhang2012,Nikora2002}, but they are not resolved by any one dataset.
 
@@
 \citet{Wu2019} retained Einstein's instantaneous steps but included the possibility that grains can become permanently buried as they rest on the bed, also providing two diffusion ranges -- intermediate and global.
-These earlier works suggest the minimal required components to model three bedload diffusion ranges: (1) exchange between motion and rest intervals and (2) the sediment burial process.
+These earlier works suggest the minimal required components to model three bedload diffusion ranges are (1) exchange between motion and rest intervals and (2) the sediment burial process.
 
@@
 
-\subsection{Joint probability distribution of particle position with burial}
+\subsection{Probability distribution of particle position with burial}
 \label{sec:solution}
@@
 	\caption{Joint distributions for a grain to be at position $x$ at time $t$ are displayed for the choice $k_1=0.1$, $k_2=1.0$, $v=2.0$.
-Grains are considered initially at rest ($\theta_1=1$, $\theta_2=0$).
+Grains are initially at rest ($\theta_1=1$, $\theta_2=0$).
 The solid lines are the analytical distribution in Eq.
-\ref{eq:pdf}), while the points are numerically simulated, showing the correctness of the derivations.
+\ref{eq:pdf}, while the points are numerically simulated, showing the correctness of the mathematical derivations.
 Colors pertain to different times.
@@
 In this case, the joint distribution tends toward Gaussian at large times \citep{Einstein1937,Lisle1998}.
-Panel (b) shows the case when grains have rate $\kappa = 0.01$ to become buried while resting.
+Panel (b) shows the case when grains have rate $\kappa = 0.01$ to become buried while resting on the bed surface.
 Because of burial, the joint distribution tends toward a more uniform distribution than Gaussian.
@@
 \ref{fig:diffpdfs} depicts the distribution Eq.
-\ref{eq:pdf} alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small time-step \citep{Barik2006}.
+\ref{eq:pdf} alongside simulations generated by a direct method based on evaluating the cumulative transition probabilities between states on a small timestep \citep{Barik2006}.
 When grains do not become buried, as in panel (a) of Fig.
@@
 
-To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution (Eq.
-\ref{eq:diffnicedist}) using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}.
+To obtain an analytical formula for tracers diffusing downstream while they gradually become buried, the first two moments of position are derived by taking derivatives with respect to $\eta$ of the Laplace space distribution \ref{eq:diffnicedist} using an approach similar to \citet{Shlesinger1974} and \citet{Weeks1998}.
 These moments produce the positional variance $\sigma_x^2 = \bra x^2\ket - \bra x \ket^2$.
@@
 	%\begin{wraptable}{r}{width=5.5cm}
-	\caption{Abbreviations used in the expressions of the mean (Eq. \ref{eq:mean}), second moment (Eq. \ref{eq:second}) and variance (Eq. \ref{eq:var}) of bedload tracers.}
+	\caption{Abbreviations used in the expressions of the mean (Eq. \ref{eq:mean}), second moment (Eq. \ref{eq:second}) and variance (Eq. \ref{eq:var}) of bedload tracers are defined.}
 	\label{table:params}
@@
 These conditions are most relevant to tracers in gravel-bed rivers, since they represent that grains are initially at rest \citep{Hassan1991,Wu2019}, motions are typically much shorter than rests \citep{Einstein1937,Hubbell1964}, and burial requires a much longer time than typical rests  \citep{Ferguson2002,Hassan1994,Haschenburger2013}.
-Figure \ref{fig:diffvar} demonstrates that under these conditions the variance (Eq.
-\ref{eq:var}) shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
-Following \citet{Hassan2017} I suggest to call the fourth range "geomorphic", since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
+Fig.
+\ref{fig:diffvar} demonstrates that under these conditions the variance \ref{eq:var} shows three diffusion ranges with approximate power law scaling ($\sigma_x^2 \propto t^\gamma$) that are identified as the local, intermediate, and global ranges proposed by Nikora et al., followed by a fourth range of no diffusion ($\sigma_x^2 = \text{const}$) stemming from the burial of all tracers.
+Following \citet{Hassan2017} I suggest to call the fourth range ``geomorphic", since any further transport in this range can occur only if scour re-exposes buried grains to the flow \citep{Nakagawa1980,Voepel2013,Martin2014,Wu2019a}.
 
@@
 	\caption{Panel (a) sketches conceptual trajectories of grains, while panel (b) depicts Eq.
-\ref{eq:var} with mean motion time $1.5$ s, rest time $30.0$ s, and velocity $0.1$ m/s.
-The burial timescale is $7200$ s (two hours), and grains start from rest ($\theta_1=1$).
+\ref{eq:var} with mean motion time $1.5$s, rest time $30.0$s, and velocity $0.1$m/s.
+The burial timescale is $7200$s (two hours), and grains start from rest ($\theta_1=1$).
 The solid line is Eq.
-\ref{eq:var} and points are numerically simulated.
+\ref{eq:var} and the points are numerically simulated.
 Panel (b) demonstrates four distinct scaling ranges of $\sigma_x^2$: local, intermediate, global, and geomorphic.
@@
 Meanwhile, in bedload transport research, variable velocities \citep{Lajeunesse2010,Furbish2012,Heyman2016}, correlated motions \citep{Heyman2014,Lee2018,Saletti2020}, and anomalous diffusion \citep{Fathel2016,Bradley2017,Schumer2009} constitute open research issues.
-I believe further developing the linkage between existing bedload models and random walk concepts could rapidly progress our understanding.
+Further developing the linkage between existing bedload models and random walk concepts could rapidly progress our understanding.
 

### 2021-09-28 revision/ch5langevin.tex
done

@@
 These challenges limit numerous ecological and engineering applications that rely on sediment transport predictions \citep{Gaeuman2017,Malmon2005}.
+
 In recent decades, stochastic formulations of the bedload flux have become increasingly popular for their potential to predict the mean transport rates required by applications while also predicting the magnitude of transport fluctuations and quantifying the dependence of measurements on the observation scale (Secs.
@@
 Recent indications that sediment transport fluctuations might explain longstanding open problems in alluvial channel stability, such as channel width maintenance \citep{Abramian2019,Abramian2020} and bedform initiation \citep{Jerolmack2005,Bohorquez2016} provide additional motivation to develop these stochastic approaches.
+
 One subset of stochastic methods expresses downstream transport rates as a sum over the instantaneous streamwise velocities of all particles in motion within a control volume (Sec.
@@
 
-High-speed video experiments have measured different stream-wise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
+High-speed video experiments have measured different streamwise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
 One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
-These experiments involve uniformly-sized small sands or glass beads ($0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were turbulent and supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were laminar and (likely) subcritical.
+These experiments involve uniformly-sized small sands or glass beads ($0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were turbulent and supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were laminar and likely subcritical.
 A second set of studies show Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}.
@@
 \citet{Houssais2012} used a bimodal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while \citet{Liu2019} used sand having median diameter $1.1$mm in turbulent and subcritical flow.
-From this experimental record, the velocity distribution shape apparently does not consistently relate to the flow conditions (turbulent/laminar or super/subcritical) or sediment properties (natural sand/gravel or glass beads).
-However there is a loose trend within the particle size.
+From this experimental record, the velocity distribution shape apparently does not consistently relate to the flow conditions.
+There is however a loose trend within the particle size.
 Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}.
-This summary suggests that bedload velocity distribution could be controlled by the particle size.
+This summary suggests that streamwise bedload velocity distributions could be controlled by the particle size.
 
@@
 In this chapter I explore the possibility that the particle velocity distribution is controlled by momentum dissipation from particle-bed collisions by incorporating ideas from the physics of granular media.
-A major paradigm in granular dynamics is to describe particles as a continuum or "granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in Ch.
+A major paradigm in granular dynamics is to describe particles as a continuum or ``granular liquid" having an effective rheology \citep[e.g.][]{Jenkins1998,Andreotti2013}, but as discussed in Ch.
 \ref{ch:Introduction} such a continuum assumption is unlikely to be satisfied for weak bedload transport conditions.
 
-Instead,articles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}. 
-In the theory of rarefied granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the ideal Gaussian form expected from perfectly elastic collisions \citep{Brilliantov2004}.
+Instead, particles in these conditions are better characterized as a rarefied granular gas \citep[e.g.][]{Furbish2021}. 
+In granular gases, dissipative collisions are known to cause departures toward an exponential velocity distribution from the Gaussian form expected for perfectly elastic collisions \citep{Brilliantov2004}.
 Gas theory formulates particle-particle collisions as a nonlocal integral within the master equation for the probability distribution of particle velocity called the Boltzmann equation \citep{Landau1969,Chapman1970,Brilliantov2004}.
@@
 I develop in this chapter an improved model for sediment grains in transport which replaces the smooth friction terms of earlier models by episodic particle-bed collisions.
-I develop the model and explain its key assumptions in Sec.
+I present the model and explain its key assumptions in Sec.
 \ref{sec:langmodel}.
-Then I present the analytical solution of the model and other major results in \ref{sec:langresults}.
+Then I explain the analytical solution of the model and other major results in Sec.
+\ref{sec:langresults}.
 Finally I discuss the implications of these results, summarize key findings, and suggest ideas for further research in Secs.
@@
 
-Figure \ref{fig:fig1} indicates the configuration of the model in this chapter.
-Nearly spherical particles of diameter $d$ and mass $m$ move as bedload down a slope inclined at an angle $\theta$ in a steady flow.
-The flow is just strong enough to drive grains into the rarefied transport typical in many gravel-bed rivers \citep[e.g.][]{Ashworth1989, Warburton1992}.
-Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles, but rarely with other moving particles \citep[cf.][]{Williams2021}.
+Fig. \ref{fig:fig1} indicates the configuration of the model in this chapter.
+Nearly spherical particles of diameter $d$ and mass $m$ move as bedload over a granular bed.
+The flow is just strong enough to drive grains into the rarefied transport often typical in gravel-bed rivers \citep[e.g.][]{Ashworth1989, Warburton1992}.
+Particles saltate downstream through a sequence of particle-bed collisions, and moving particles collide often with stationary particles but rarely with other moving particles \citep[cf.][]{Williams2021}.
 
@@
 The objective is to calculate the probability distribution $P(u,t)$ of this downstream velocity by averaging over the ensemble of forces.
-Here, the description of bedload particle velocity fluctuations will be more detailed than Ch.
-\ref{ch:flux}, but entrainment and deposition processes will be neglected, meaning the model of this chapter is only applicable to moving particles.
-I include the most realistic particle-bed collision and fluid forces possible while still allowing for analytical solutions.
+
+The description of bedload particle velocity fluctuations will be more detailed than the simple white noise model Ch.
+\ref{ch:flux}, but alternation between motion and rest will be neglected, meaning the model of this chapter is only applicable to moving particles.
+In analyzing the velocities of moving particles, I include the most realistic particle-bed collision and fluid forces possible while still allowing for analytical solutions.
 
 \subsection{Episodic collision forces}
-Collisions between bedload particles dissipate momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid between approaching particles \citep{Joseph2001,Schmeeckle2001}.
-The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of restitution may be less relevant than other factors such as collision geometry and bed deformation.
-Here, we introduce a simple elasticity parameter $\varepsilon$ as a first attempt at the problem, as indicated in Fig.
+Collisions between bedload particles dissipate translational momentum, partly by converting it to vertical, lateral, or angular momentum, partly by deforming particles and the bed and generating heat \citep{Schmeeckle2014,Williams2021}, and partly by pressurizing the fluid between approaching particles \citep{Joseph2001,Schmeeckle2001}.
+The microscopic details of particle-particle collisions have been thoroughly studied \citep{Brach1989, Lorenz1997,Montaine2011}, although in bedload transport the coefficient of restitution may be less relevant than other factors such as collision geometry and bed deformation \citep{Pahtz2021}.
+Here, I introduce a simple elasticity parameter $\varepsilon$ as a first attempt at modeling dissipation by episodic collisions, as indicated in Fig.
 \ref{fig:fig1}.
@@
 	\centerline{\includegraphics{./figures/ch5/Fig1Concept.png}}
-	\caption{Definition sketch of rarefied sediment transport with turbulent fluid drag and particle-bed collision forces.
-During saltation, pre-collisional streamwise velocities $u$ are transformed to post-collisional velocities $\ve u < u$.}
+	\caption{Rarefied sediment transport involves turbulent fluid drag and episodic particle-bed collisions.
+In the collision model developed in this chapter, pre-collisional streamwise velocities $u$ are transformed instantaneously to post-collisional velocities $\ve$.
+The parameter $\ve$ is an ``elasticity" with $0\leq \ve \leq 1$.}
 	\label{fig:fig1}
@@
 Since the elasticity combines effects of particle shape and collision geometry and should vary from one collision to the next, the elasticity $\varepsilon$ is interpreted as a random variable, characterized by a statistical distribution $\rho(\varepsilon)$.
-Some granular gas models have also included random elasticity \citep[e.g.][]{Serero2015}, but this topic has not been deeply explored in the literature.
+Some granular gas models have also included random elasticity \citep[e.g.][]{Serero2015}, but this topic has not been completely explored in the literature.
 
-Assuming that the number of collisions per unit time is $\nu$ and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (Sec.
+Assuming that the average number of collisions per unit time is $\nu$ (the collision rate) and that the time intervals between subsequent particle-bed collisions are exponentially distributed \citep[e.g.][]{Gordon1972}, the collision force in the downstream direction can be written as a Poisson pulse noise (Sec.
 \ref{sec:einflux}):
@@
 Here, $N_\nu(t)$ is the number of collisions in time $t$ (a Poisson random variable), the $\tau_k$ ($k=1,2,\dots$) are times at which collisions occur, and the $\varepsilon_k$ are the elasticity coefficients, drawn from the distribution $\rho(\varepsilon)$ characterizing the fraction of momentum lost in each particle-bed collision.
-The time intervals $\Delta \tau_k = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau),$ so the number of collisions $n=N_\nu(t)$ in time $t$ is distributed as $P(n) = (\nu t)^n/n!
-\exp(-\nu t).$
+The time intervals $\Delta \tau = \tau_k-\tau_{k-1}$ between collisions are distributed as $P(\Delta \tau) = \nu \exp(-\nu\Delta \tau),$ so the number of collisions $n=N_\nu(t)$ in time $t$ is distributed as $P(n) = (\nu t)^n\exp(-\nu t)/n!
+.$
 This collision force is a sequence of random impulses which are proportional to the pre-collisional streamwise momenta.
 This collision model should be adequate when the contact times between moving and resting particles are small compared to the times between collisions.
-These conditions should be satisfied for idealized saltation trajectories as depicted in Fig.
+These conditions should be satisfied for typical saltation trajectories as in Fig.
 \ref{fig:fig1}.
@@
 Even ignoring this complication, the added mass and Basset forces are difficult to compute and therefore impractical to include in analytical models.
-These challenges require us to adopt the most complex expressions of the fluid forces that are compatible with their modeling goals \citep[e.g.][]{Michaelides1997,Armenio2001}.
-From this same requirement, a majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the only force driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.
+These challenges require us to adopt the most complex expressions of the fluid forces that are compatible with our modeling goals \citep[e.g.][]{Michaelides1997,Armenio2001}.
+From this same requirement, a majority of bedload transport models include fluid drag (and possibly pressure gradient) terms as the only force(s) driving the downstream movement of bedload particles \citep{Ancey2014,Fan2014,Schmeeckle2014,Gonzalez2017,Elghannay2017}.
 
-The downstream drag force with the $\Rey_p$ correction is usually written 
+The downstream drag force with an $\Rey_p$ correction is usually written 
 \be F_D = \frac{\pi}{8} \rho_f d^2 C_D(\Rey_p) |V|V, \label{eq:draggy}\ee
-where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
+where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity calculated as the difference of fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
 Experiments on stationary particles indicate that drag forces fluctuate rapidly and lie on wide Gaussian-like distributions \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
-The drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities are expected to vary rapidly due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
+Drag forces on moving particles have not been experimentally measured, but even in globally steady flow conditions, slip velocities are expected to fluctuate due to particle acceleration, fluid turbulence, passage of other moving particles, and vertical movement of particles within the flow profile.
 The nonlinearity in Eq.
-\ref{eq:draggy} will amplify slip velocity variations, generating large and erratic drag force fluctuations, which only add to the undetermined contributions of the neglected fluid forces.
+\ref{eq:draggy} will amplify slip velocity fluctuations, generating large and erratic drag force variations that only add to the undetermined contributions of the neglected fluid forces.
 
@@
 
-\subsection{Chapman-Kolmogorov equation and particle-bed collision integral}
+\subsection{Master equation and particle-bed collision integral}
 
-The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix Sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
+The equation governing the streamwise velocity distribution $P(u,t)$ is derived in the appendix Sec. \ref{sec:langmasterderiv} with a limiting procedure, providing
 \be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
@@
 
-Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bedload models \citep{Fan2014,Ancey2014}. 
+Eq. \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bedload models \citep{Fan2014,Ancey2014}. 
 Nonlocality is introduced by the collision integral of Eq. \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
@@
 Owing to its nonlocality, Eq. \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
-To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
-For particle-bed collisions of bedload, we can expect a mode elasticity since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.
+To make progress, I assume the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$ to allow for a Kramers-Moyal type expansion of the particle-bed collision integral \citep[e.g.][]{Gardiner1983}.
+Such a mode value in the collision elasticity distribution is present in experiments on rigid body collisions \citep{Glielmo2014}.
+For particle-bed collisions of bedload, a modal elasticity is also expected since experiments indicate a most common collision geometry \citep[e.g.][]{Gordon1972,Martin2013}.
 
@@
 Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values.
-The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.
 
@@
 \ref{eq:steadystate}.
-As elasticity $\ve$ varies, the particle velocity distributions interpolate between exponential (inelastic) and Gaussian (elastic) forms.}
+As the elasticity $\ve$ varies, the particle velocity distributions interpolate between exponential (inelastic) and Gaussian (elastic) forms.}
 	\label{fig:fig2}
@@
 Fig.
-\ref{fig:fig2} reveals an apparent transition from exponential-like to Gaussian-like velocity distributions as collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
+\ref{fig:fig2} reveals an apparent transition from exponential to Gaussian-like velocity distributions as collisions vary from more inelastic ($\ve \rightarrow 0$) to more elastic ($\ve \rightarrow 1$).
 In between, the full distribution Eq.
@@
 	\centerline{\includegraphics{./figures/ch5/Fig3asymptotic.png}}
-	\caption{The particle velocity distribution approaches an exponential distribution in (a) as particle-bed collisions become extremely elastic ($\ve \rightarrow 1$), and it approaches a Gaussian in (b) as they become extremely inelastic ($\ve \rightarrow 0$).
+	\caption{The particle velocity distribution approaches an exponential distribution in panel (a) as particle-bed collisions become extremely elastic ($\ve \rightarrow 1$), and it approaches a Gaussian in panel (b) as they become extremely inelastic ($\ve \rightarrow 0$).
 On the abscissa, the mean sediment velocity is standardized by its mean $\bar{u}$ and standard deviation $\sigma_u$.
@@
 In each case, the particle mass is computed from experimental parameters assuming spherical sediment as $m = \pi \rho_s d^3/6$, where $\rho_s$ is the sediment density and $d$ is the particle diameter.
-The characteristic slip velocity entering the steady component of the drag force (Eq.
-\ref{eq:drag}) is not obviously related to any one velocity scale of the fluid flow.
+The characteristic slip velocity entering the steady component of the drag force in Eq.
+\ref{eq:drag} is not obviously related to any one velocity scale of the fluid flow.
 It should really be calculated as an average over different particle heights, particle velocities, and fluid velocities within the flow profile.
@@
 Particle Reynolds numbers are estimated using the shear velocity: $\Rey_p= u_\ast d/\lambda$. 
-The magnitude $D$ of turbulent fluctuations, collision elasticity $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution (Eq.
-\ref{eq:steadystate}) and the experimental data.
+The magnitude $D$ of turbulent fluctuations, collision elasticity $\ve$, and collision rate $\nu$ are treated as calibration parameters and are tuned to provide best fit between the distribution \ref{eq:steadystate} and the experimental data.
 
 The parameters for each experiment and the best-fit calibration parameters are collected in Tab.
-\ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in Fig.d \ref{fig:fig4ch5}.
+\ref{tab:calib}, while the results of fitting the model to the available experimental data are shown in Fig.
+\ref{fig:fig4ch5}.
 \begin{table}
@@
 \ref{fig:fig4ch5}.
-The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St = \rho_s u_\ast d/(9 \lambda)$, and Froude number $\Fro = \bar{u}/\sqrt{g h}$ ($h$ -- depth, $g$ -- gravitational acceleration, $\bar{u}$ -- time and depth averaged flow velocity).
-The final two columns involve values that were tuned to fit the theoretical distribution (Eq.
-\ref{eq:steadystate}) to the experimental data, the elasticity $\ve$ and the collision rate $\nu$.
+The first five columns involve data from the experiments for particle diameter $d$, shear velocity $u_\ast$, particle Reynolds number $\Rey_p$, Stokes number $\St$, and Froude number $\Fro$.
+The final two columns involve values that were tuned to fit Eq.
+\ref{eq:steadystate}, the elasticity $\ve$ and the collision rate $\nu$.
 Units are indicated in square brackets.
@@
 	\centerline{\includegraphics{./figures/ch5/Fig4expComparison.png}}
-	\caption{This figure compares the available experimental data on velocity distributions and the theoretical velocity distribution Eq.
-\ref{eq:steadystate} with parameters $\nu$, $\varepsilon$, and $D$ treated as calibration parameters.
-In all cases there is good agreement, indicating that the model is capable of the range of distributions exhibited by the experimental data.} \label{fig:fig4ch5}
+	\caption{The available experimental data on velocity distributions is compared to Eq.
+\ref{eq:steadystate} with calibration parameters $\nu$, $\varepsilon$, and $D$.
+In all cases there is good agreement, indicating that the model produces the range of distributions seen in the experimental data.} \label{fig:fig4ch5}
 \end{figure}
@@
 Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bedload particles as spatially-uniform Gaussian white noise for reasons of analytical tractability \cite[e.g.][]{Michaelides1997}.
-Even though the non-Gaussian, spatio-temporally-varying, and history-dependent aspects of fluid turbulence certainly do impact sediment movement \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is laminar or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
-Nevertheless, it is possible the vertical flow structure, and in particular the size of particles in comparison to the thickness of the laminar sub-layer, may be explanatory of the different velocity distributions resolved in bedload transport experiments, and future studies should investigate this possibility.
+Even though the non-Gaussian, space- and time-varying, and history-dependent aspects of fluid turbulence certainly do impact sediment movement \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is laminar or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically large compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.
+Nevertheless, it is possible that the vertical flow structure, and in particular the size of particles in comparison to the thickness of the laminar sub-layer, may be explanatory of the different velocity distributions resolved in bedload transport experiments, and future studies should investigate this possibility.
 
@@
 The model developed here does not consider these viscous forces, nor does it explicitly model the geometry of collisions, as $\ve$ was defined as a parameter applying to the downstream velocity only.
-Thus, although I have included episodic particle-bed collisions in a stochastic sediment transport model for the first time, the key parameter ($\ve$) in this formulation has a rather heuristic character and is not clearly related to the coefficient of restitution from granular gas theory.
-Future studies should formulate particle-bed collisions in stochastic sediment transport models considering more details of collision geometry \citep{Sekine1992}, fluid-particle interactions \citep{Marshall2011}, and traditional restitution \citep{Brach1989} using the present study as a starting point.
+Thus, although I have included episodic particle-bed collisions in a stochastic sediment transport model for the first time, the key parameter $\ve$ in this formulation has a rather heuristic character and is not clearly related to the coefficient of restitution from granular gas theory.
+Future studies should formulate particle-bed collisions in stochastic sediment transport models considering more details of collision geometry \citep{Sekine1992}, fluid-particle interactions \citep{Marshall2011}, and traditional restitution \citep{Brach1989} using the stochastic methods in the present study as a starting point.
 
@@
 Several researchers have considered why particle velocity distributions differ from one experiment to the next.
-One prevailing view is that it relates to flow hydraulics \citep{Wu2020}.
+A prevailing view is that it relates to flow hydraulics \citep{Wu2020}.
 Many of the experiments producing Gaussian velocity distributions were conducted in supercritical flows \citep[e.g.][]{Heyman2016,Martin2012,Ancey2014}, whereas many producing exponential distributions were conducted in subcritical flows \citep[e.g.][]{Fathel2015,Charru2004,Seizilles2014}.
@@
 
-An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates from granular interactions.
+An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates mainly from granular interactions.
 Particle size enters the Langevin model Eq. \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
 The analytical velocity distribution Eq.
-\ref{eq:steadystate} was fit to six different experimental data-sets in Fig.
+\ref{eq:steadystate} was fit to six different experimental datasets in Fig.
 \ref{fig:fig4ch5}, and the best fit parameters were presented in Tab.
@@
 A definite conclusion that particle size controls the shape of the bedload velocity distribution requires additional experimental and theoretical study.
-This chapter nevertheless produces suggestive ideas.
+This work nevertheless produces suggestive ideas.
 
@@
 
-This work suggests that particle-bed collisions be a leading order control over the different bedload velocity distributions obtained in experiments, although future study is required for a definite conclusion.
-Next steps are to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions and to introduce slip velocity dependence and vertical structure into the flow forces acting on moving particles.
+This work suggests that particle-bed collisions may be a leading order control over the velocity distributions of bedload particles, although future study is required for a definite conclusion.
+The next steps are to build up the episodic collision model presented in this chapter to include the geometric details of particle-bed collisions and to introduce slip velocity dependence and vertical structure into the flow forces acting on moving particles.
 These efforts would produce additional insight into the controls of fluid turbulence and transverse, vertical, and rotational movements on the downstream velocities of individual bedload grains.

### 2021-09-28 revision/ch6conc.tex
done

@@
 This thesis has described bedload transport from the statistical physics of individual grains.
-The work has improved upon earlier descriptions.
-It describes the movements of grains along a wider range of timescales than before, highlights Langevin equations, master equations, and random walks as the common threads of stochastic sediment transport modeling, and provides more realistic descriptions of bedload transport.
-In particular, I have
+This work describes the movements of grains along a wider range of timescales than before, highlights Langevin equations, master equations, and random walks as the common threads of stochastic sediment transport modeling, and provides more realistic descriptions of bedload transport.
+In particular, within the thesis I have
 \begin{enumerate}
@@
 \end{enumerate}
-These developments bring us closer to a comprehensive mathematical description of sediment transport in streams.
+These developments bring us closer to a comprehensive mathematical description of sediment transport in streams and bring powerful physics and mathematics tools to longstanding problems in river science.
 
@@
 
-Once the stochastic dynamical equations were written for a given problem, its solutions were averaged over realizations of the control variables to produce a master equation, an integro-differential equation governing the probability distributions of the response variables.
+Once the stochastic dynamical equations were written for a given problem, its solutions were averaged over realizations of the control variables to produce a master equation for the problem, an integro-differential equation governing the probability distributions of the response variables.
 
@@
 
-\subsection{Calculation of the probability distribution of the sediment flux from micro-mechanics of particle transport}
+\subsection{Calculation of the probability distribution of the sediment flux from individual particle trajectories}
 
@@
 This description predicts multiple-range diffusion across local, intermediate, and global scales as a result of particle velocity fluctuations within the motion state; it introduces a dimensionless P\'{e}clet number as an important characteristic of bedload sediment transport; and it relates the timescales at which transitions between local, intermediate, and global timescales occur to the movement characteristics of individual grains.
-These developments constitute a new understanding of bedload movement across its timescales and invite generalizations aimed at pushing the approach further to include interactions between particles and temporal correlations in bedload movement velocities.
+These developments constitute new understanding of bedload movement across its timescales and invite generalizations that include interactions between particles and temporal correlations to bedload movement velocities.
 
@@
 The residence times of buried particles are random variables that lie on heavy-tailed power-law distributions.
-These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to determine the downstream sediment flux for applications by tracking sediment tracer particles.
+These distributions allow for arbitrarily long resting times, which poses challenging implications for researchers attempting to determine downstream sediment fluxes from sediment tracer data.
 
-\subsection{Characterization of how sediment burial affects the downstream transport of sediment particles}
+\subsection{Characterization of sediment burial controls on the downstream transport of sediment particles}
 
@@
 Ch.
-\ref{ch:downDiff} brings some closure to these studies and demonstrates that many approaches to describe individual particle motions are reformulations of the continuous time random walk formalism from physics, indicating unity within a diverse body of research and bringing powerful tools from statistical physics to the sediment transport problem.
+\ref{ch:downDiff} brings some closure to these studies and demonstrates that many approaches to describe individual particle motions are particular adaptations of the continuous time random walk formalism from physics, indicating unity within a diverse body of research and bringing powerful tools from statistical physics to the sediment transport problem.
 
@@
 
-The stochastic descriptions of sediment transport developed in this thesis show a methodology we can use to step beyond averaged descriptions of landscape evolution, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?
+The stochastic descriptions of sediment transport developed in this thesis show a methodology we can use to step beyond averaged descriptions of geomorphic change, propagate noises through the equations governing landscape change, and revisit the old question in geomorphology: How does variability shape Earth's surface?

### 2021-09-28 revision/laySum.tex
done

@@
 Most approaches to predict gravel flow consider gravel as if it were a heavy fluid.
-This description oversimplifies the problem, since gravel flow is actually individual grains bouncing and rolling downstream, which is hardly like a fluid.
+These descriptions oversimplify the problem, since gravel flow is actually composed of individual grains bouncing and rolling downstream, which is hardly like a fluid.
 
@@
 This research produces a more realistic description of gravel movement in rivers which helps us understand how rivers change through time.
-
-\endinput
-The lay or public summary explains the key goals and contributions of
-the research\slash{}scholarly work in terms that can be understood by the
-general public. It must not exceed 150 words in length.

### 2021-09-28 revision/pref.tex
done

@@
 The supervisory committee and especially the research supervisor Marwan Hassan provided guidance in the research.
-Suggestions from Michael Church, Rui Ferreira, Roger Beckie, and Peter Ashmore improved the clarity of the writing.
+Suggestions from Michael Church, Rui Ferreira, Roger Beckie, Peter Ashmore, Bernard Laval, and Brett Eaton improved the clarity of the writing.
 
 The thesis includes two published works, Chapters \ref{ch:ch3} and \ref{ch:downDiff}, available in \textit{Journal of Geophysical Research: Earth Surface} and \textit{Geophysical Research Letters} respectively, with Marwan Hassan as coauthor \citep{Pierce2020a,Pierce2020b}.
-Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication after acceptance of the thesis.
+Chapters \ref{ch:flux} and \ref{ch:langevin} will be submitted for publication following acceptance of the thesis.

### 2021-09-28 renamer.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+#cp ./trackchanges/diss_diff.pdf kevin-pierce-thesis_tc.pdf
+cp ./revision/diss.pdf ubc_2021_november_pierce_james.pdf

### 2021-09-28 revision/formerPapers/ens-rev/annobiblio/make.sh

@@ -1,3 +1,0 @@
-#!/bin/bash
-pandoc -s -o out.pdf anno.md
-echo "anno made" 

### 2021-09-28 revision/formerPapers/ens-rev/scraps/uncomment.sh

@@ -1,7 +1,0 @@
-#!/bin/bash
-while read p; do
-    if [["%" -eq  *$p]]; then
-        echo "$p"
-done<$1
-
-

### 2021-09-28 revision/formerPapers/grlRrevision2/pointbypoint/complier.sh

@@ -1,4 +1,0 @@
-#!/bin/bash
-gedit $1 &
-latexmk -pvc -pdf -interaction=nonstopmode $1
-

### 2021-09-28 trackchanges/diff.sh

@@ -1,0 +1,38 @@
+#!/bin/bash
+old="/home/kp/Desktop/phdThesis/draft"
+new="/home/kp/Desktop/phdThesis/revision"
+files="ack.tex
+abs.tex
+pref.tex
+macros.tex
+app_ch3.tex
+app_ch5.tex
+ch1intro.tex
+ch2flipFlop.tex
+ch3bedEle.tex
+ch4downDiff.tex
+ch5langevin.tex
+ch6conc.tex
+"
+
+#app_ch2.tex
+#app_ch4.tex
+cp "$new/ubcdiss.cls" ./ubcdiss.cls
+cp "$new/biblio.bib" ./biblio.bib
+echo "diffing original and revision files"
+echo "two appendices are excluded as they cause bugs and were modified by an earlier run of diff.sh"
+for f in $files
+do
+    latexdiff --"exclude-textcmd=textcite" --packages=amsmath --math-markup=0 --graphics-markup=0 --config='MATHENV=(?:displaymath|equation*)' $old/$f $new/$f > $f
+    #sed -i -e ' s/\b\be\b/\begin{equation}/g' $f
+    #sed -i -e 's/\b\ee\b/\end{equation}/g' $f
+
+
+done
+echo "finished."
+
+#'--config="PICTUREENV=(?:picture|DIFnomarkup|align)[\w\d*@]*"'
+
+
+f="diss.tex"
+latexdiff --"exclude-textcmd=textcite" --packages=amsmath --math-markup=0 --graphics-markup=0 --config='MATHENV=(?:displaymath|equation*)' $old/$f $new/$f > diss_diff.tex

### 2021-10-14 revision/ch3bedEle.tex
finished.

@@
 
-\begin{wraptable}{r}{0.6\textwidth}
-	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ are tabulated from \citet{Ancey2008}.
-Units are $s^{-1}$ (probability/time).
-Bed elevation changes modulate these rates in accord with Eqs.
-\ref{eq:rate1}-\ref{eq:rate4}.}\label{tab:anceyparams}
-	\begin{tabular}{cccccc} \\ 
+\begin{table}
+	\begin{center}
+	\def~{\hphantom{0}}
+	\begin{tabular}{c|ccccc} 
 		\toprule  
@@
 		(a) & 5.45  & 6.59  & 3.74 & 4.67 & 0.77 \\
-		\midrule
 		(g) & 7.74  & 8.42  & 4.34 & 4.95 & 0.56 \\
-		\midrule
 		(i) & 15.56 & 22.07 & 3.56 & 4.52 & 0.68 \\
-		\midrule
 		(l) & 15.52 & 14.64 & 4.32 & 4.77 & 0.48 \\
-		\bottomrule
 	\end{tabular}
-\end{wraptable}
+	\caption{Migration, entrainment, and deposition rates at $z(m)=0$ are tabulated from \citet{Ancey2008}.
+Units are $s^{-1}$ (probability/time).
+Bed elevation changes modulate these rates in accord with Eqs.
+\ref{eq:rate1}-\ref{eq:rate4}.}
+	\label{tab:anceyparams}
+	\end{center}
+\end{table}
+
 For the migration, entrainment, and deposition parameters representing bedload transport at each flow condition $(\nu, \lambda, \mu, \sigma, \gamma)$, I used the values measured by \citet{Ancey2008} in a series of flume experiments: these are summarized in Tab.
@@
 
+\begin{figure}
+	\centerline{\includegraphics{./figures/ch3/newvariance.pdf}}
+	\caption{Data from all simulations demonstrates that the active layer depth $l$ characterizes bed elevation changes as described by Eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$.}
+	\label{fig:var}
+\end{figure}
+
 Fig.
 \ref{fig:var} demonstrates that the variance of bed elevations is approximately $\sigma_z^2 = z_1^2 \sigma_m^2 = \frac{1}{4\kappa}=l^2$, agreeing with the approximation in Eq.
-\ref{eq:ou2}; this result suggests that $l$ is a characteristic length scale of bed elevation fluctuations.
-\begin{wrapfigure}{r}{0.5\textwidth}
-	\centering
-	\includegraphics[width=0.5\textwidth,keepaspectratio]{./figures/ch3/variance.pdf}
-	\caption{Data from all simulations demonstrates that the active layer depth $l$ characterizes bed elevation changes as described by Eq. \ref{eq:active}: $\sigma_m^2 \approx (l/z_1)^2$. }
-	\label{fig:var}
-\end{wrapfigure}
-\indent The close correspondence between the mean field approximation and the numerical simulations in Fig.
+\ref{eq:ou2}; this result indicates that $l$ is a characteristic length scale of bed elevation fluctuations.
+The close correspondence between the mean field approximation and the numerical simulations in Fig.
 \ref{fig:pdfs} panel (a) suggests the unconditional moments of $n$ correspond closely with the \citet{Ancey2008} result.

### 2021-10-14 revision/figures/ch3/grabvariancefig.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+cp ../../newch3fig/newvariance.pdf .

### 2021-10-19 revision/ch5langevin.tex
completely done for sure.

@@
 
-\section{Summary}
+\section{Conclusion}
 \label{sec:langconclusion}
@@
 These efforts would produce additional insight into the controls of fluid turbulence and transverse, vertical, and rotational movements on the downstream velocities of individual bedload grains.
+
+\section{Derivation of master equation}
+\label{sec:langmasterderiv}
+To derive the master equation from Eq.
+\ref{eq:langevin}, I temporarily consider the Gaussian white noise (GWN) $\xi(t)$ as a Poisson jump process having rate $r$ and jumps $ \sqrt{2 \tilde{D}} h$ with $h$ distributed as $f(h)$.
+I will later take a GWN limit from this noise.
+Similar derivations are available in some references \citep[e.g.][]{Kanazawa2017,Suweis2011}, and the precise connection between GWN and Poisson noise is discussed in \citet{VanDenBroeck1983}.
+
+From this starting point, integrating Eq. \ref{eq:langevin} over a small time interval $\delta t$ under the It\^{o} interpretation for the collision term provides
+\be     
+u(t+\delta t) =
+\begin{cases}
+	u(t) + \tilde{\Gamma} \delta t & \text{with probability } 1- r \delta t - \nu \delta t\\
+	u(t) + \sqrt{2 \tilde{D}} h &  \text{with probability } r \delta t \\
+	\ve u(t) &  \text{with probability } \nu \delta t
+\end{cases}.
+\ee
+
+Considering the probability $P(u,t+\delta t)$ as a sum over possible paths from $P(u,t)$ develops 
+\begin{align} 
+\begin{split}
+P(u,t+\delta t) =
+	\hspace{4pt}&(1- r \delta t - \nu \delta t)\int_{-\infty}^\infty dw  \delta(u-w-\tilde{\Gamma} \delta t) P(w,t)\\ 
+	&+  r \delta t \int_{-\infty}^\infty dw \int_{-\infty}^\infty dh f(h) \delta (u - w - \sqrt{2 \tilde{D}}h) P(w,t) \\ 
+	&+ \nu \delta t \int_{-\infty}^\infty dw \int_0^1 d\ve \rho(\ve)  \delta(u-w\ve) P(w,t).
+\end{split}
+\end{align}
+Evaluating all integrals over $\delta$-functions provides 
+\begin{align}
+\begin{split}
+ P(u,t+\delta t) = \hspace{4pt}&(1-r \delta t - \nu \delta t)P(u-\tilde{\Gamma} \delta t, t) \\&+ r \delta t\int_{-\infty}^\infty dh f(h) P(u+\sqrt{2 \tilde{D}}h) \\&+ \nu \delta t \int_0^1 \frac{ \tilde{D}\ve}{\ve} \rho(\ve) P\big(\frac{u}{\ve}\big).
+\end{split}
+\end{align}
+Finally, taking $\delta t \rightarrow 0$ and limiting the Poisson noise involving $\sqrt{2 \tilde{D}}$ to a Gaussian white noise by taking $r \rightarrow \infty$ as $h \rightarrow 0$ such that $h^2 r = 1$ obtains the master equation \ref{eq:master}.
+\section{Derivation of steady-state solution}
+\label{sec:langsteadyderiv}
+Defining $\tilde{P}(s) = \int_{-\infty}^\infty du e^{i u s} P(u) $ as the Fourier transform (FT) of $P(u)$ and taking the FT of Eq. \ref{eq:master} develops the recursion relation
+\be \tilde{P}(s) = \frac{\tilde{P}(s\ve)}{q(s)}. \ee
+where
+\be q(z) = \tilde{D} z^2 - i \tilde{\Gamma} z + 1 . \label{eq:q} \ee
+Recursing $N+1$ times provides
+\be \tilde{P}(s) = \frac{\tilde{P}(s \ve^{N+1})}{q(s \ve^0)q(s\ve^1)\dots q(s\ve^N)}.\label{eq:recursion}\ee
+The polynomials $q(z)$ can always be factored as $q(z) = \tilde{D}(z - i\lambda_-)(z - i\lambda_+)$ where
+\be \lambda_\pm = \frac{\tilde{\Gamma}}{2\tilde{D}}\Big[1 \pm \sqrt{ 1 + 4\tilde{D}/\tilde{\Gamma}^2} \Big]. \label{eq:lambdas}\ee
+Using these factors to expand $\tilde{P}(s)$ in partial fractions provides
+\be \tilde{P}(s)  = \tilde{P}(s \ve^{N+1}) \sum_{l=0}^N  \Big[ \frac{R_l^-}{s\ve^l - i\lambda_-} + \frac{R_l^+}{s\ve^l - i\lambda_+} \Big] ,\ee
+where the coefficients $R_l^\pm$ are the residues (up to constant factors) of the product $[q(s\ve^0)\dots q(s\ve^N)]^{-1}$:
+\be R_l^\pm = \frac{s \ve^{l} - i\lambda_\pm }{q(s\ve^0)\dots q(s \ve^N)} \Big|_{s= i \lambda_\pm \ve^{-l}}.\ee
+The Fourier-transformed Eq.
+\ref{eq:recursion} has a convenient feature as $N\rightarrow \infty$: since $0<\ve <1$, the prefactor $\tilde{P}( s \ve^{N+1})$ becomes the normalization condition $\tilde{P}(0)=1$ for the probability distribution $P(u)$.
+Taking this limit and evaluating the residues provides 
+\begin{multline} \tilde{P}(s) = \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_- \ve^m)} \sum_{l=0}^\infty \frac{i}{(s\ve^l-i\lambda_-)\prod_{m=1}^l q(i\lambda_- \ve^{-m})} 
+	\\+ \frac{1}{\tilde{D}(\lambda_+ - \lambda_-) \prod_{m=1}^\infty q(i \lambda_+ \ve^m)} \sum_{l=0}^\infty \frac{-i}{(s\ve^l-i \lambda_+)\prod_{m=1}^l q(i\lambda_+ \ve^{-m})}. \end{multline}
+Finally, the Fourier transforms are inverted term by term with contour integration (another use of the integral given in Eq.
+\ref{eq:contour}), and the definition of $q(z)$ from Eq.
+\ref{eq:q} is incorporated to obtain the steady-state solution quoted in Eq.
+\ref{eq:steadystate}.
+
+\section{Calculation of the moments}
+\label{sec:langmoments}
+Taking Eq. \ref{eq:master}, multiplying by $u^k$, integrating over all space, and taking account of normalization of $P(u)$ provides a recursion relation for the moments: 
+\be 0 = \tilde{D} k (k-1)\langle u^{k-2} \rangle + \tilde{\Gamma} k \langle u^{k-1} \rangle + (\ve^k-1)\langle u^k \rangle. \ee
+$k=1$ provides the mean (given in Eq. \ref{eq:meanu})
+\be \langle u \rangle =  \frac{\tilde{\Gamma}}{1-\ve} = \frac{\Gamma}{\nu(1-\ve)}, \ee
+while $k=2$ provides the second moment
+\be \langle u^2 \rangle = 2 \frac{\tilde{D} + \tilde{\Gamma} \langle u \rangle}{1-\ve^2}, \ee
+leading to the velocity variance
+\be \sigma_u^2 = \frac{2\tilde{D} + \tilde{\Gamma}^2}{1-\ve^2}\ee
+given in Eq. \ref{eq:varu}.
+\section{The weak collision limit}
+\label{sec:langextremes}
+Now I will demonstrate that weak collisions imply a Gaussian-like distribution for sediment velocities.
+The limit is challenging since the steady-state distribution Eq.
+\ref{eq:steadystate} and the moments above all diverge as $\ve \rightarrow 1$.
+Following \cite{Hall1989}, this divergence suggests normalizing the distribution $P(u)$ using the scaled variable
+\be z = \frac{u - \bra u \ket}{\sigma_u}, \ee
+giving a scaled distribution
+\be Q(z) = \sigma_u P(u).\ee
+The resulting differential equation for $Q(z)$ is well-behaved in the elastic collision limit $\ve \rightarrow 1$.
+Incorporating this transformation into Eq. \ref{eq:master} provides the ``normalized" master equation
+\begin{multline}(1-\ve^2) \frac{\tilde{D}}{2\tilde{D} + \tilde{\Gamma}} Q''(z) - \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\sqrt{2\tilde{D}+\tilde{\Gamma}^2}}Q'(z) - Q(z) \\+ \frac{1}{\ve} Q\Big(z + \Big[\frac{1-\ve}{\ve}z + \frac{\tilde{\Gamma}\sqrt{1-\ve^2}}{\ve\sqrt{2\tilde{D}+\tilde{\Gamma}^2}} \Big]\Big) = 0 .\end{multline}
+This equation remains exact and is only a change of variables from Eq. \ref{eq:master}.
+Now I approximate the equation for $\ve \rightarrow 1$ by expanding the final term to second order around $z=0$ before setting $\ve = 1$, obtaining
+\be Q''(z) + z Q'(z) + Q(z) = 0, \ee
+which is the classic Ornstein-Uhlenbeck Fokker-Planck equation whose solution is the standard normal distribution for $Q(z)$ \citep[e.g.][]{Gardiner1983}.
+This solution provides Eq. \ref{eq:gaussian} when transformed back to the original variables $P(u)$ and $u$.
